---
title: "Pharmaverse Learning System — Module 0: Setup & Orientation"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  Set up your pharmaverse environment and get oriented with the clinical data
  pipeline: raw data → SDTM → ADaM → TLF.
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120)
```

## Welcome to Pharmaverse

### What is pharmaverse?

**Pharmaverse** is a curated set of open-source R packages developed collaboratively
by the pharmaceutical industry for clinical reporting. Together they cover the
complete clinical data submission pipeline required by the FDA and other regulators.

```
Raw EDC Data
     │
     ▼  [sdtm.oak]
  SDTM Datasets  (.xpt)
     │
     ▼  [admiral]
  ADaM Datasets  (.xpt)
     │
     ▼  [tfrmt / rtables / gt]
  Tables, Listings & Figures  (TLF)
     │
     ▼
  eCTD Submission Package
```

The four modules in this tutorial track that pipeline:

| Module | Package(s) | Topic |
|--------|-----------|-------|
| 0 (this) | — | Ecosystem orientation & setup |
| 1 | `admiral` | Building ADaM datasets |
| 2 | `sdtm.oak`, `xportr` | SDTM mapping & xpt export |
| 3 | `tfrmt`, `rtables`, `gt` | Clinical tables & figures |
| Capstone | all | End-to-end mini-submission |

---

## Installation

### Installing pharmaverse packages

Run this once in your **R console** (not inside a learnr block) before starting the tutorials:

```r
# Install all pharmaverse packages used in this course
install.packages(c(
  "learnr",       # interactive tutorials
  "admiral",      # ADaM construction
  "admiraldev",   # admiral developer utilities
  "xportr",       # SAS transport file export
  "sdtm.oak",     # SDTM mapping
  "tfrmt",        # table formatting
  "rtables",      # clinical tables
  "gt",           # grammar of tables
  "dplyr",        # data manipulation
  "lubridate",    # date handling
  "stringr",      # string manipulation
  "tidyr",        # data reshaping
  "ggplot2"       # figures
))
```

> **Tip:** If `sdtm.oak` isn't on CRAN yet, install from GitHub:
> `remotes::install_github("pharmaverse/sdtm.oak")`

---

## The CDISC Data Standards

### Why do we need ADaM and SDTM?

The FDA requires clinical trial data submissions to follow **CDISC** standards:

- **SDTM** (Study Data Tabulation Model) — raw-ish collected data, organized into
  domain-specific datasets (DM, AE, EX, LB, VS, etc.)
- **ADaM** (Analysis Data Model) — analysis-ready datasets derived from SDTM,
  designed for statistical analysis and direct traceability back to SDTM

```{r quiz-cdisc, echo=FALSE}
quiz(
  caption = "Quick check — CDISC basics",
  question("SDTM datasets are best described as:",
    answer("Analysis-ready datasets with derived variables"),
    answer("Collected trial data organized by clinical domain", correct = TRUE),
    answer("Tables and figures for the clinical study report"),
    answer("R package names"),
    allow_retry = TRUE
  ),
  question("Which statement about ADaM is TRUE?",
    answer("ADaM datasets must be traceable back to SDTM or raw data", correct = TRUE),
    answer("ADaM replaces SDTM — you only need one"),
    answer("ADaM datasets use the same variable names as SDTM"),
    answer("ADaM is only used for safety analyses"),
    allow_retry = TRUE
  )
)
```

---

## Key CDISC Domains

### SDTM domains you'll encounter most

| Domain | Full Name | Contains |
|--------|-----------|----------|
| DM | Demographics | One row per subject, baseline characteristics |
| AE | Adverse Events | Safety data — all reported adverse events |
| EX | Exposure | Drug dosing records |
| LB | Laboratory | Lab test results over time |
| VS | Vital Signs | BP, HR, weight, height, etc. |
| CM | Concomitant Medications | Other meds taken during trial |
| MH | Medical History | Pre-existing conditions |

### ADaM datasets you'll encounter most

| Dataset | Full Name | Contains |
|---------|-----------|----------|
| ADSL | Subject Level | One row per subject — baseline, treatment, flags |
| ADAE | Adverse Events | Safety analysis dataset |
| ADLBC | Laboratory (BDS) | Longitudinal lab with derived flags |
| ADVS | Vital Signs (BDS) | Longitudinal vitals with change-from-baseline |
| ADTTE | Time to Event | Survival/time-to-event analysis dataset |

---

## R Basics Review

### dplyr — the foundation

Admiral and the pharmaverse stack rely heavily on `dplyr`. Let's warm up.

```{r dplyr-warmup, exercise=TRUE, exercise.lines=20}
library(dplyr)

# A tiny fake SDTM DM domain
dm <- tibble::tribble(
  ~USUBJID,       ~AGE, ~SEX, ~RACE,       ~ARM,
  "STUDY-001-001",  45,  "M", "WHITE",     "Drug A",
  "STUDY-001-002",  52,  "F", "BLACK",     "Placebo",
  "STUDY-001-003",  38,  "M", "ASIAN",     "Drug A",
  "STUDY-001-004",  61,  "F", "WHITE",     "Drug A",
  "STUDY-001-005",  29,  "M", "HISPANIC",  "Placebo"
)

# Your turn: filter to Drug A subjects and arrange by AGE descending
dm |>
  filter(ARM == "Drug A") |>
  arrange(desc(AGE))
```

```{r dplyr-warmup-hint}
# Use filter() with == and arrange() with desc()
# dm |> filter(ARM == "Drug A") |> arrange(desc(AGE))
```

### mutate and case_when

One of the most important patterns in admiral is deriving new variables with `mutate()`
and conditional logic with `case_when()`.

```{r mutate-exercise, exercise=TRUE, exercise.lines=25}
library(dplyr)

dm <- tibble::tribble(
  ~USUBJID,       ~AGE, ~SEX, ~ARM,
  "001",  45,  "M", "Drug A",
  "002",  52,  "F", "Placebo",
  "003",  38,  "M", "Drug A",
  "004",  70,  "F", "Drug A",
  "005",  29,  "M", "Placebo"
)

# Task: add a column AGEGR1 with age groups:
#   "<40" for age < 40
#   "40-65" for age 40 to 65
#   ">=65" for age >= 65
dm |>
  mutate(
    AGEGR1 = case_when(
      AGE < 40        ~ "<40",
      AGE >= 40 & AGE < 65 ~ "40-65",
      AGE >= 65       ~ ">=65"
    )
  )
```

```{r mutate-exercise-solution}
library(dplyr)
dm <- tibble::tribble(
  ~USUBJID, ~AGE, ~SEX, ~ARM,
  "001", 45, "M", "Drug A",
  "002", 52, "F", "Placebo",
  "003", 38, "M", "Drug A",
  "004", 70, "F", "Drug A",
  "005", 29, "M", "Placebo"
)
dm |>
  mutate(
    AGEGR1 = case_when(
      AGE < 40             ~ "<40",
      AGE >= 40 & AGE < 65 ~ "40-65",
      AGE >= 65            ~ ">=65"
    )
  )
```

---

## Environment Check

```{r env-check, exercise=TRUE, exercise.lines=15}
# Run this to confirm your pharmaverse packages are installed
pkgs <- c("admiral", "xportr", "tfrmt", "rtables", "gt",
          "dplyr", "lubridate", "stringr")

status <- sapply(pkgs, function(p) {
  if (requireNamespace(p, quietly = TRUE)) "✓ installed" else "✗ MISSING"
})

data.frame(Package = names(status), Status = status, row.names = NULL)
```

---

## You're Ready!

You now understand the pharmaverse ecosystem and have a working environment.

**Next steps:**

- 📦 **Module 1** → `01_admiral_adam.Rmd` — Build ADSL and ADAE with admiral
- 📦 **Module 2** → `02_sdtm_xportr.Rmd` — Map raw data to SDTM and export .xpt files
- 📦 **Module 3** → `03_tlf_generation.Rmd` — Create clinical tables with tfrmt & rtables
- 🏆 **Capstone** → `04_capstone_project.Rmd` — Full end-to-end mini-submission

Open the next module in RStudio with `rmarkdown::run("01_admiral_adam.Rmd")`.
