---
title: "Lesson 04 — R Primer for SAS Programmers, Part 2: The DATA Step in dplyr"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  Comprehensive translation of SAS DATA step patterns into dplyr: filter, select,
  mutate, case_when, summarise, group_by, distinct/count, string operations,
  rowwise, FIRST./LAST. logic, lag/lead, cumulative functions, across(), and
  two clinical capstone projects.
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.blanks = TRUE)

# ── ADSL — 254 subjects (CDISC CDISCPILOT01) ─────────────────────────────────
adsl_raw <- pharmaverseadam::adsl

# HEIGHTBL / WEIGHTBL / BMIBL are present in pharmaverseadam >= 0.1.1.
# For older builds or versions where they are absent, synthesise clinically
# plausible values (Alzheimer's trial demographics, set.seed for reproducibility).
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014)          # CDISC pilot study year
  n  <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F",
               round(rnorm(n, 161, 7)),
               round(rnorm(n, 174, 8)))
  wt <- ifelse(adsl_raw$SEX == "F",
               round(rnorm(n, 68, 12), 1),
               round(rnorm(n, 82, 14), 1))
  adsl_raw <- adsl_raw |>
    mutate(HEIGHTBL = ht,
           WEIGHTBL = wt,
           BMIBL    = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw,
                     BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
}

adsl <- adsl_raw |>
  mutate(
    # In the CDISC pilot all randomised subjects are also ITT
    ITTFL  = SAFFL,
    # Obesity (BMI >= 30) used as a diabetes-risk proxy flag for exercises
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N")
  )

# ── ADAE — treatment-emergent adverse events ─────────────────────────────────
# pharmaverseadam stores treatment as TRTA; rename to TRT01A for consistency
adae_raw <- pharmaverseadam::adae

if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw)) {
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
} else if (!"TRT01A" %in% names(adae_raw)) {
  adae_raw <- left_join(adae_raw,
                        select(adsl, USUBJID, TRT01A),
                        by = "USUBJID")
}

adae <- adae_raw |>
  mutate(
    # Numeric severity: MILD = 1, MODERATE = 2, SEVERE = 3
    AESEVN = case_when(
      AESEV == "MILD"     ~ 1L,
      AESEV == "MODERATE" ~ 2L,
      AESEV == "SEVERE"   ~ 3L,
      TRUE                ~ NA_integer_
    )
  )

# ── ADLB — lab results ────────────────────────────────────────────────────────
adlb <- pharmaverseadam::adlb

# ATOXGR and BTOXGR are stored as character in pharmaverseadam; convert to
# integer so numeric comparisons (>, lag, cummax) behave correctly.
# Values: 0 = no toxicity, 1-4 = NCI CTCAE grades; negative = below-normal range.
if (is.character(adlb$ATOXGR)) {
  adlb <- mutate(adlb,
                 ATOXGR = as.integer(ATOXGR),
                 BTOXGR = as.integer(BTOXGR))
}

# admiral routinely appends derived summary rows to BDS datasets:
#   DTYPE = "MINIMUM", "MAXIMUM", "LOCF", "WOCF", etc.
#   These have sentinel AVISITN values (9997, 9998, ...) and are NOT real visits.
#   Keeping them in longitudinal analyses (lag, lead, worsening flags) produces
#   spurious comparisons — e.g. "POST-BASELINE MAXIMUM" vs. the last real visit.
#   Always exclude them before any visit-sequence work.
if ("DTYPE" %in% names(adlb)) {
  adlb <- filter(adlb, is.na(DTYPE))
}

# Add TRTSDT from ADSL if not already present
if (!"TRTSDT" %in% names(adlb)) {
  adlb <- left_join(adlb,
                    select(adsl, USUBJID, TRTSDT),
                    by = "USUBJID")
}

# ── ADVS — vital signs ───────────────────────────────────────────────────────
advs <- pharmaverseadam::advs
```

## Overview

### Why dplyr matters for pharmaverse

`{dplyr}` is the workhorse of clinical data manipulation in R — a hard dependency of every major pharmaverse package. You'll write more dplyr code in a typical clinical R script than all pharmaverse packages combined.

The motivation: dplyr provides **five small verbs** that cover almost everything you did in the SAS DATA step, plus powerful higher-order tools that SAS requires entire PROCs for.

| dplyr verb | SAS equivalent | What it does |
|---|---|---|
| `filter()` | `where` / subsetting `if` | Keep rows meeting a condition |
| `select()` | `keep` / `drop` | Keep, drop, or reorder columns |
| `mutate()` | Assignment statements | Add or modify columns |
| `summarise()` | `proc means`, `proc summary` | Reduce rows to summary statistics |
| `arrange()` | `proc sort` | Reorder rows |

Combined with `group_by()` (the BY statement) and `|>`, these five verbs — plus a dozen supporting functions — handle every clinical data manipulation pattern you'll encounter.

**Datasets pre-loaded in this lesson — real data from the CDISC CDISCPILOT01 study via `pharmaverseadam`:**

- `adsl` — 254 subjects across 3 treatment arms (Placebo / Xanomeline Low Dose / Xanomeline High Dose)
- `adae` — 1,200+ treatment-emergent AE records spanning 20+ MedDRA system organ classes
- `adlb` — thousands of lab observations: ALT, CREAT, WBC, glucose, and more across screening and on-study visits, with NCI CTCAE toxicity grades
- `advs` — vital signs records: systolic BP, diastolic BP, pulse, weight, and height across multiple visits

These are the same datasets used in admiral, tfrmt, and rtables tutorials throughout this course.

---

```{r quiz-overview, echo=FALSE}
quiz(
  caption = "Before we begin",
  question("Which dplyr verb is equivalent to SAS PROC SORT?",
    answer("mutate()"),
    answer("filter()"),
    answer("arrange()", correct = TRUE),
    answer("summarise()"),
    allow_retry = TRUE
  ),
  question("What SAS concept does group_by() most closely replace?",
    answer("The LENGTH statement"),
    answer("The BY statement in PROC/DATA step", correct = TRUE),
    answer("PROC FORMAT"),
    answer("The ARRAY statement"),
    allow_retry = TRUE
  ),
  question("In R, what does NA propagation mean in practice?",
    answer("NA values are silently removed"),
    answer("NA values are treated as 0"),
    answer("Any arithmetic or comparison involving NA returns NA, unlike SAS which often ignores missing", correct = TRUE),
    answer("NA causes an error"),
    allow_retry = TRUE
  )
)
```

---

## 1 — filter() — the WHERE clause

### Concept

`filter()` keeps rows that match a condition. Multiple conditions separated by commas are treated as AND; use `|` for OR.

```r
adsl |> filter(SAFFL == "Y", AGE >= 65)                          # AND: comma = &
adsl |> filter(SAFFL == "Y" | DTHFL == "Y")                      # OR
adsl |> filter(TRT01A %in% c("Placebo", "Xanomeline Low Dose"))  # IN list
adsl |> filter(!(TRT01A %in% c("Xanomeline High Dose")))         # NOT IN — wrap with !()
adsl |> filter(!is.na(WEIGHTBL))                                  # not missing
adsl |> filter(grepl("HIGH", TRT01A))                            # pattern match
adsl |> filter(WEIGHTBL / (HEIGHTBL / 100)^2 > 30)              # filter on computed expr
```

**`%in%` negation** — to exclude values from a list, wrap the whole `%in%` expression in `!()`:

```r
# SAS:  where trt not in ("Placebo")
# R:    filter(!(TRT01A %in% c("Placebo")))
```

**`grepl()` in filter** — returns TRUE when the pattern is found anywhere in the string:

```r
filter(grepl("Xanomeline", TRT01A))          # matches "Xanomeline Low Dose" AND "High Dose"
filter(grepl("xanomeline", TRT01A, ignore.case = TRUE))  # case-insensitive
```

### Exercise 1a — safety population, elderly or female, on Xanomeline

Filter `adsl` to subjects who satisfy **all three** conditions:

1. Safety population (`SAFFL == "Y"`)
2. Elderly (age ≥ 65) **or** female
3. Randomised to **any** Xanomeline dose — use `grepl()`

Then count results by `TRT01A` and `SEX`.

```{r filter-basic, exercise=TRUE, exercise.lines=9}
adsl |>
  filter(
    SAFFL == "Y",
    ___ | ___,
    grepl(___, TRT01A)
  ) |>
  count(TRT01A, SEX)
```

```{r filter-basic-hint-1}
# Condition 2: AGE >= 65 | SEX == "F"
# Condition 3: grepl("Xanomeline", TRT01A)
```

```{r filter-basic-solution}
adsl |>
  filter(
    SAFFL == "Y",
    AGE >= 65 | SEX == "F",
    grepl("Xanomeline", TRT01A)
  ) |>
  count(TRT01A, SEX)
```

### Exercise 1b — NOT IN list

Filter `adsl` to subjects who are **not** on Xanomeline High Dose. Use `!` + `%in%`.

```{r filter-notin, exercise=TRUE, exercise.lines=5}
adsl |>
  filter(!(TRT01A %in% c(___))) |>
  select(USUBJID, TRT01A)
```

```{r filter-notin-hint-1}
# !(TRT01A %in% c("Xanomeline High Dose"))
```

```{r filter-notin-solution}
adsl |>
  filter(!(TRT01A %in% c("Xanomeline High Dose"))) |>
  select(USUBJID, TRT01A)
```

### Exercise 1c — pattern match with grepl()

Filter `adsl` to subjects whose treatment contains the word `"Xanomeline"` (either dose). Use `grepl()`.

```{r filter-grepl, exercise=TRUE, exercise.lines=5}
adsl |>
  filter(grepl(___, TRT01A)) |>
  count(TRT01A)
```

```{r filter-grepl-hint-1}
# grepl("Xanomeline", TRT01A)
```

```{r filter-grepl-solution}
adsl |>
  filter(grepl("Xanomeline", TRT01A)) |>
  count(TRT01A)
```

### Exercise 1d — complex multi-condition filter

Find subjects who:
- Are in the safety population AND the ITT population
- Are female OR aged 70+
- Have a non-missing WEIGHTBL

```{r filter-complex, exercise=TRUE, exercise.lines=9}
adsl |>
  filter(
    SAFFL == "Y" & ITTFL == "Y",
    ___ | ___,
    ___
  ) |>
  select(USUBJID, AGE, SEX, SAFFL, ITTFL, WEIGHTBL)
```

```{r filter-complex-hint-1}
# SEX == "F" | AGE >= 70
# !is.na(WEIGHTBL)
```

```{r filter-complex-solution}
adsl |>
  filter(
    SAFFL == "Y" & ITTFL == "Y",
    SEX == "F" | AGE >= 70,
    !is.na(WEIGHTBL)
  ) |>
  select(USUBJID, AGE, SEX, SAFFL, ITTFL, WEIGHTBL)
```

### Exercise 1e — filter on computed expression (no pre-step)

Filter `adsl` directly to subjects whose **BMI** (WEIGHTBL / (HEIGHTBL/100)^2) is **above 30** — without creating a BMI column first.

```{r filter-expr, exercise=TRUE, exercise.lines=6}
adsl |>
  filter(___ > 30) |>
  select(USUBJID, HEIGHTBL, WEIGHTBL)
```

```{r filter-expr-hint-1}
# WEIGHTBL / (HEIGHTBL / 100)^2 > 30
```

```{r filter-expr-solution}
adsl |>
  filter(WEIGHTBL / (HEIGHTBL / 100)^2 > 30) |>
  select(USUBJID, HEIGHTBL, WEIGHTBL)
```

### Exercise 1f — serious AEs that are still ongoing (AEENDTC is NA)

Filter `adae` to rows where the AE is **serious** AND still **ongoing** (end date is missing).

```{r filter-ongoing-sae, exercise=TRUE, exercise.lines=6}
adae |>
  filter(___ == "Y", ___) |>
  select(USUBJID, AEDECOD, AESTDTC, AEENDTC, AESER)
```

```{r filter-ongoing-sae-hint-1}
# AESER == "Y", is.na(AEENDTC)
```

```{r filter-ongoing-sae-solution}
adae |>
  filter(AESER == "Y", is.na(AEENDTC)) |>
  select(USUBJID, AEDECOD, AESTDTC, AEENDTC, AESER)
```

---

## 2 — select() — KEEP, DROP, reorder

### Concept

```r
adsl |> select(USUBJID, AGE, SEX)              # keep specific columns
adsl |> select(-STUDYID, -SITEID)              # drop with minus; -c(STUDYID, SITEID) also works
adsl |> select(USUBJID, TRT01A, everything())  # promote columns to front, keep the rest
adsl |> select(ends_with("FL"))                # tidyselect helper
adsl |> select(where(is.numeric))              # predicate helper — all numeric columns
adsl |> relocate(TRT01A, .before = AGE)        # move column position (.before or .after)
```

**Rename while selecting** — put `NEW_NAME = OLD_NAME` inside `select()`. This keeps only the listed columns and renames in one step (more concise than `select()` + `rename()`):

```r
adsl |> select(SUBJECT_ID = USUBJID,   # rename USUBJID → SUBJECT_ID
               TREATMENT  = TRT01A,    # rename TRT01A  → TREATMENT
               AGE, SEX)               # keep as-is
```

| Helper | Selects columns that... |
|---|---|
| `starts_with("TRT")` | start with "TRT" |
| `ends_with("FL")` | end with "FL" |
| `contains("BL")` | contain "BL" anywhere |
| `matches("^A[EV]")` | match a regex |
| `where(is.numeric)` | satisfy a predicate function |

### Exercise 2a — key demographic columns

Select `USUBJID`, `AGE`, `SEX`, `RACE`, and `TRT01A` from `adsl`. No hints needed — straightforward practice.

```{r select-demo-cols, exercise=TRUE, exercise.lines=4}
adsl |>
  select(___)
```

```{r select-demo-cols-solution}
adsl |>
  select(USUBJID, AGE, SEX, RACE, TRT01A)
```

### Exercise 2b — all flag columns + USUBJID

Select `USUBJID` and all flag columns (those ending in `"FL"`) using a tidyselect helper.

```{r select-flags, exercise=TRUE, exercise.lines=4}
adsl |>
  select(USUBJID, ___)
```

```{r select-flags-hint-1}
# ends_with("FL")
```

```{r select-flags-solution}
adsl |>
  select(USUBJID, ends_with("FL"))
```

### Exercise 2c — numeric columns only

Select only columns that are numeric from `adsl`. Use `where(is.numeric)`.

```{r select-numeric, exercise=TRUE, exercise.lines=5}
adsl |>
  select(USUBJID, where(___)) |>
  names()
```

```{r select-numeric-hint-1}
# where(is.numeric)
```

```{r select-numeric-solution}
adsl |>
  select(USUBJID, where(is.numeric)) |>
  names()
```

### Exercise 2d — rename while selecting

In one `select()` call, keep `USUBJID` (rename to `SUBJECT_ID`), `TRT01A` (rename to `TREATMENT`), `AGE`, and `SEX`.

```{r select-rename-select, exercise=TRUE, exercise.lines=6}
adsl |>
  select(
    SUBJECT_ID = ___,
    TREATMENT  = ___,
    AGE,
    SEX
  )
```

```{r select-rename-select-hint-1}
# SUBJECT_ID = USUBJID, TREATMENT = TRT01A
```

```{r select-rename-select-solution}
adsl |>
  select(
    SUBJECT_ID = USUBJID,
    TREATMENT  = TRT01A,
    AGE,
    SEX
  )
```

### Exercise 2e — promote columns with relocate()

In `adsl`, move `TRT01A` to be the **first column** after `USUBJID`, pushing everything else right.

```{r select-relocate, exercise=TRUE, exercise.lines=6}
adsl |>
  relocate(TRT01A, .after = ___) |>
  names() |>
  head(4)
```

```{r select-relocate-hint-1}
# .after = USUBJID
```

```{r select-relocate-solution}
adsl |>
  relocate(TRT01A, .after = USUBJID) |>
  names() |>
  head(4)
```

### Exercise 2f — drop multiple columns, keep the rest

Drop `STUDYID`, `SITEID`, `RFSTDTC`, and `RFENDTC` from `adsl` using the minus notation. (A common real-world pattern: strip study-level admin columns before passing data to a reporting macro.)

```{r select-drop-multi, exercise=TRUE, exercise.lines=5}
adsl |>
  select(-___, -___, -___, -___) |>
  names()
```

```{r select-drop-multi-solution}
adsl |>
  select(-STUDYID, -SITEID, -RFSTDTC, -RFENDTC) |>
  names()
```

---

## 3 — mutate() — assignment statements

### Concept

`mutate()` adds or modifies columns. You can reference columns just created in the same call.

```r
adsl |>
  mutate(
    BMI        = WEIGHTBL / (HEIGHTBL / 100)^2,
    BMI_ROUND  = round(BMI, 1),         # uses BMI from same call
    USUBJID_UC = toupper(USUBJID)
  )
```

**`if_else()` — binary condition (base R):** Use `if_else()` (dplyr) instead of base `ifelse()` because it enforces that both branches return the **same type**, catching type mismatches at run time. This matters in clinical code where a silent type coercion could corrupt a flag.

```r
# Syntax: if_else(condition, value_if_TRUE, value_if_FALSE, missing = value_if_NA)
adsl |> mutate(ELDERLY = if_else(AGE >= 65, "Y", "N"))
adsl |> mutate(ONPBO   = if_else(TRT01A == "Placebo", "Y", NA_character_))
#                                                           ↑ type-safe NA
```

**Type-safe NAs** — when you need to produce `NA` in a typed context, use the appropriate constant:

| Type | Constant |
|---|---|
| character | `NA_character_` |
| numeric (double) | `NA_real_` |
| integer | `NA_integer_` |
| Date | `as.Date(NA)` |

**`as.Date()` — convert strings to Date:** ADaM date columns like `TRTSDT` are stored as `Date` objects. When a date arrives as a character string (ISO 8601), coerce it first:

```r
as.Date("2014-01-15")          # "2014-01-15" → Date
as.Date("15JAN2014", "%d%b%Y") # SAS-style date string
as.Date(ADT)                   # if ADT is already Date, returns unchanged (safe no-op)
```

Date arithmetic works directly: `as.Date("2014-03-01") - as.Date("2014-01-15")` returns `45` days.

**String helpers in `mutate()`** — full coverage is in Section 9, but the three most common ones appear in exercises here:

```r
paste0(SITEID, "-", USUBJID)            # concatenate strings (no separator argument)
substr(USUBJID, 4, 6)                   # extract characters at positions 4–6 (1-based)
nchar(USUBJID)                          # number of characters in string
toupper(RACE)                           # convert to upper case
```

**`case_when()` preview** — Section 4 covers it fully, but Exercise 3d uses it inside `mutate()`. The rule: conditions are checked top-to-bottom; the first matching condition wins; `TRUE ~` is the catch-all else branch:

```r
mutate(ANRIND = case_when(
  AVAL > ANRHI ~ "HIGH",
  AVAL < ANRLO ~ "LOW",
  TRUE         ~ "NORMAL"     # else branch
))
```

### Exercise 3a — BMI, deviation from mean, and unit conversion in one mutate()

In a single `mutate()` call, add **three** columns to `adsl`:

1. `BMI` = weight (kg) / height (m)² — rounded to 1 decimal
2. `BMI_DEV` = each subject's BMI minus the **sample-wide mean BMI** (rounded to 2 decimals) — demonstrates that `mean()` inside `mutate()` operates on the whole column, not per-row
3. `WGTLB` = weight converted to pounds (1 kg = 2.205 lb), rounded to 1 decimal

Then select `USUBJID`, `HEIGHTBL`, `WEIGHTBL`, `BMI`, `BMI_DEV`, `WGTLB`.

```{r mutate-bmi, exercise=TRUE, exercise.lines=12}
adsl |>
  mutate(
    BMI     = round(___ / (___ / 100)^2, 1),
    BMI_DEV = round(BMI - mean(___, na.rm = TRUE), 2),
    WGTLB   = round(WEIGHTBL * ___, 1)
  ) |>
  select(USUBJID, HEIGHTBL, WEIGHTBL, BMI, BMI_DEV, WGTLB)
```

```{r mutate-bmi-hint-1}
# BMI     = WEIGHTBL / (HEIGHTBL / 100)^2
# BMI_DEV = BMI - mean(BMI, na.rm = TRUE)  — mean() of the whole column, scalar subtraction
# WGTLB   = WEIGHTBL * 2.205
```

```{r mutate-bmi-solution}
adsl |>
  mutate(
    BMI     = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1),
    BMI_DEV = round(BMI - mean(BMI, na.rm = TRUE), 2),
    WGTLB   = round(WEIGHTBL * 2.205, 1)
  ) |>
  select(USUBJID, HEIGHTBL, WEIGHTBL, BMI, BMI_DEV, WGTLB)
```

### Exercise 3b — string manipulation

Add a column `SITE_SUBJ` by concatenating `SITEID` and the last four characters of `USUBJID` (the subject number, e.g., `"1015"`), separated by `"-"`. Use `substr()` or `str_sub()` and `paste0()`.

```{r mutate-string, exercise=TRUE, exercise.lines=8}
adsl |>
  mutate(
    SITE_SUBJ = paste0(SITEID, "-", substr(USUBJID, nchar(USUBJID)-___, nchar(USUBJID)))
  ) |>
  select(USUBJID, SITEID, SITE_SUBJ)
```

```{r mutate-string-hint-1}
# In "01-701-1015", the subject number "1015" is the last 4 characters
# substr(x, nchar(x)-3, nchar(x))  OR  str_sub(USUBJID, -4, -1)
```

```{r mutate-string-solution}
adsl |>
  mutate(
    SITE_SUBJ = paste0(SITEID, "-", substr(USUBJID, nchar(USUBJID)-3, nchar(USUBJID)))
  ) |>
  select(USUBJID, SITEID, SITE_SUBJ)
```

### Exercise 3c — percent of reference range in lab data

In `adlb`, add `PCT_ULN` = AVAL as a percentage of the upper limit of normal (`ANRHI`). Round to 1 decimal. Flag rows where PCT_ULN > 100 as `ABOVE_ULN = TRUE`.

```{r mutate-pct-uln, exercise=TRUE, exercise.lines=10}
adlb |>
  mutate(
    PCT_ULN   = round(___ / ANRHI * 100, 1),
    ABOVE_ULN = PCT_ULN > ___
  ) |>
  filter(ABOVE_ULN) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, ANRHI, PCT_ULN, ABOVE_ULN)
```

```{r mutate-pct-uln-hint-1}
# PCT_ULN = round(AVAL / ANRHI * 100, 1)
# ABOVE_ULN = PCT_ULN > 100
```

```{r mutate-pct-uln-solution}
adlb |>
  mutate(
    PCT_ULN   = round(AVAL / ANRHI * 100, 1),
    ABOVE_ULN = PCT_ULN > 100
  ) |>
  filter(ABOVE_ULN) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, ANRHI, PCT_ULN, ABOVE_ULN)
```

### Exercise 3d — reference range indicator without a function

In `advs`, add `ANRIND` by hand:
- `"HIGH"` if AVAL > ANRHI
- `"LOW"` if AVAL < ANRLO
- `"NORMAL"` otherwise

Then count by ANRIND.

```{r mutate-anrind, exercise=TRUE, exercise.lines=14}
advs |>
  mutate(
    ANRIND = case_when(
      AVAL > ___  ~ "HIGH",
      AVAL < ___  ~ "LOW",
      TRUE        ~ "NORMAL"
    )
  ) |>
  count(ANRIND)
```

```{r mutate-anrind-hint-1}
# AVAL > ANRHI → "HIGH"; AVAL < ANRLO → "LOW"
```

```{r mutate-anrind-solution}
advs |>
  mutate(
    ANRIND = case_when(
      AVAL > ANRHI ~ "HIGH",
      AVAL < ANRLO ~ "LOW",
      TRUE         ~ "NORMAL"
    )
  ) |>
  count(ANRIND)
```

### Exercise 3e — derive TRTEMFL (on-treatment flag)

In `adlb`, add `TRTEMFL = "Y"` when `ADT` is **on or after** `TRTSDT` (both are `Date` objects in the real data), otherwise `NA_character_`. Use `if_else()`.

> Note: In the real `pharmaverseadam::adlb`, `ADT` is already a `Date`; `as.Date()` is a safe no-op but shown here as a reminder for data where dates may arrive as strings.

```{r mutate-trtemfl, exercise=TRUE, exercise.lines=10}
adlb |>
  mutate(
    ADT_D    = as.Date(ADT),
    TRTSDT_D = as.Date(TRTSDT),
    TRTEMFL  = if_else(ADT_D >= TRTSDT_D, "Y", NA_character_)
  ) |>
  select(USUBJID, PARAMCD, AVISIT, ADT, TRTSDT, TRTEMFL) |>
  head(8)
```

```{r mutate-trtemfl-hint-1}
# if_else(ADT_D >= TRTSDT_D, "Y", NA_character_)
# if_else() is type-strict: both branches must be the same type (character here)
```

```{r mutate-trtemfl-solution}
adlb |>
  mutate(
    ADT_D    = as.Date(ADT),
    TRTSDT_D = as.Date(TRTSDT),
    TRTEMFL  = if_else(ADT_D >= TRTSDT_D, "Y", NA_character_)
  ) |>
  select(USUBJID, PARAMCD, AVISIT, ADT, TRTSDT, TRTEMFL) |>
  head(8)
```

---

## 4 — case_when() — nested if/else

### Concept

`case_when()` evaluates conditions in order — **first match wins**. The `TRUE ~` catch-all is the else branch.

```r
mutate(
  AGE_CAT = case_when(
    AGE < 18              ~ "Pediatric",
    AGE < 65              ~ "Adult",
    AGE >= 65 & AGE < 75  ~ "Elderly (65-74)",
    AGE >= 75             ~ "Elderly (75+)",
    TRUE                  ~ NA_character_
  )
)
```

**`if_else()` vs `ifelse()`** — Use `if_else()` (dplyr) for binary conditions. It enforces identical types in both branches and is safer in clinical code.

**Typed NAs in `if_else()` and `case_when()`** — when a branch produces a missing value you must supply a type-matching NA constant, not bare `NA` (which is logical):

```r
if_else(TRT01A == "Placebo", "Y", NA_character_)   # ✓ character NA
if_else(AGE > 65, 1L,        NA_integer_)           # ✓ integer NA
if_else(AVAL > 0, AVAL,      NA_real_)              # ✓ numeric NA
if_else(condition, as.Date("2014-01-01"), as.Date(NA)) # ✓ Date NA

# Bare NA fails type check:
if_else(TRT01A == "Placebo", "Y", NA)  # ✗ Error: true/false must be same type
```

### Exercise 4a — five-level age group

Derive `AGEGR2` (a finer grouping than the existing `AGEGR1`) with five levels: `"<45"`, `"45-54"`, `"55-64"`, `"65-74"`, `">=75"`.

> Note: `pharmaverseadam::adsl` already has `AGEGR1` with three levels. Here we derive a five-level version named `AGEGR2` to avoid overwriting it — a common real-world pattern.

```{r casewhen-agegr5, exercise=TRUE, exercise.lines=14}
adsl |>
  mutate(
    AGEGR2 = case_when(
      AGE <  45             ~ "<45",
      AGE <  55             ~ ___,
      AGE <  65             ~ ___,
      AGE <  75             ~ ___,
      TRUE                  ~ ">=75"
    )
  ) |>
  count(AGEGR2)
```

```{r casewhen-agegr5-hint-1}
# < 55 → "45-54", < 65 → "55-64", < 75 → "65-74"
```

```{r casewhen-agegr5-solution}
adsl |>
  mutate(
    AGEGR2 = case_when(
      AGE <  45 ~ "<45",
      AGE <  55 ~ "45-54",
      AGE <  65 ~ "55-64",
      AGE <  75 ~ "65-74",
      TRUE      ~ ">=75"
    )
  ) |>
  count(AGEGR2)
```

### Exercise 4b — BMI category

Add `BMIGR1` to `adsl`. First compute BMI, then categorize:
- `"Underweight"` < 18.5
- `"Normal"` 18.5–24.9
- `"Overweight"` 25–29.9
- `"Obese"` ≥ 30

```{r casewhen-bmi, exercise=TRUE, exercise.lines=14}
adsl |>
  mutate(
    BMI    = WEIGHTBL / (HEIGHTBL / 100)^2,
    BMIGR1 = case_when(
      BMI <  18.5             ~ "Underweight",
      BMI <  25.0             ~ ___,
      BMI <  30.0             ~ ___,
      TRUE                    ~ ___
    )
  ) |>
  count(BMIGR1)
```

```{r casewhen-bmi-hint-1}
# 18.5 to <25: "Normal", 25 to <30: "Overweight", else "Obese"
```

```{r casewhen-bmi-solution}
adsl |>
  mutate(
    BMI    = WEIGHTBL / (HEIGHTBL / 100)^2,
    BMIGR1 = case_when(
      BMI <  18.5 ~ "Underweight",
      BMI <  25.0 ~ "Normal",
      BMI <  30.0 ~ "Overweight",
      TRUE        ~ "Obese"
    )
  ) |>
  count(BMIGR1)
```

### Exercise 4c — two-variable cross-condition

In `adae`, add `RELSEV`: combine relatedness and severity into a single priority category.

- `"HIGH CONCERN"` if `AEREL %in% c("PROBABLE","POSSIBLE")` AND `AESEVN >= 2`
- `"MONITOR"` if `AEREL %in% c("PROBABLE","POSSIBLE")` AND `AESEVN == 1`
- `"BACKGROUND"` otherwise

```{r casewhen-relsev, exercise=TRUE, exercise.lines=14}
adae |>
  mutate(
    RELSEV = case_when(
      AEREL %in% c("PROBABLE","POSSIBLE") & AESEVN >= 2 ~ ___,
      AEREL %in% c("PROBABLE","POSSIBLE") & AESEVN == 1 ~ ___,
      TRUE ~ "BACKGROUND"
    )
  ) |>
  count(RELSEV)
```

```{r casewhen-relsev-hint-1}
# "HIGH CONCERN" and "MONITOR" respectively
```

```{r casewhen-relsev-solution}
adae |>
  mutate(
    RELSEV = case_when(
      AEREL %in% c("PROBABLE","POSSIBLE") & AESEVN >= 2 ~ "HIGH CONCERN",
      AEREL %in% c("PROBABLE","POSSIBLE") & AESEVN == 1 ~ "MONITOR",
      TRUE ~ "BACKGROUND"
    )
  ) |>
  count(RELSEV)
```

### Exercise 4d — if_else() type safety trap

The following code silently returns wrong types in base R. Use `if_else()` to enforce that `NA_real_` is returned (not `NA`) when AVAL is below the lower limit.

```{r casewhen-ifelse, exercise=TRUE, exercise.lines=10}
# Using if_else(): both branches must be numeric
adlb |>
  mutate(
    AVAL_FLAGGED = if_else(
      AVAL < ANRLO,
      NA_real_,      # NA, but typed as numeric
      ___            # otherwise keep AVAL
    )
  ) |>
  filter(PARAMCD == "ALT") |>
  select(USUBJID, AVISIT, AVAL, ANRLO, AVAL_FLAGGED)
```

```{r casewhen-ifelse-hint-1}
# The TRUE branch should return AVAL (numeric)
```

```{r casewhen-ifelse-solution}
adlb |>
  mutate(
    AVAL_FLAGGED = if_else(
      AVAL < ANRLO,
      NA_real_,
      AVAL
    )
  ) |>
  filter(PARAMCD == "ALT") |>
  select(USUBJID, AVISIT, AVAL, ANRLO, AVAL_FLAGGED)
```

```{r quiz-casewhen-adv, echo=FALSE}
quiz(
  question("What does `TRUE ~ NA_character_` at the end of case_when() do?",
    answer("It converts the column to logical type"),
    answer("It is required syntax and does nothing"),
    answer("It matches rows where no earlier condition matched — the default/else branch, here setting them to NA", correct = TRUE),
    answer("It runs only when all conditions are TRUE"),
    allow_retry = TRUE
  ),
  question("Why does `if_else(cond, 1L, NA)` fail in R?",
    answer("NA is not allowed in if_else()"),
    answer("if_else() requires both branches to return the same type — 1L is integer but NA is logical", correct = TRUE),
    answer("if_else() doesn't handle integers"),
    answer("The condition must return numeric"),
    allow_retry = TRUE
  )
)
```

---

## 5 — arrange() — PROC SORT

### Concept

```r
adsl |> arrange(USUBJID)               # ascending
adsl |> arrange(desc(AGE))             # descending
adsl |> arrange(TRT01A, desc(AGE))     # multi-column
adae |> arrange(USUBJID, AESTDTC, desc(AESEVN))
```

Unlike SAS, `arrange()` is only needed when final row order matters (output, lag/lead, cumulative). `group_by()` never requires pre-sorting.

**`desc()` on a logical expression** — applying `desc()` to a condition pushes `TRUE` rows to the top, useful for "deaths first" or "serious AEs first" orderings:

```r
adsl |> arrange(desc(DTHFL == "Y"), TRTSDT)   # died subjects first, then by start date
adae |> arrange(desc(AESER == "Y"), USUBJID)  # serious AEs to top within each subject
```

### Exercise 5a — sort with a logical expression: deaths first

Sort `adsl` so that subjects who died (`DTHFL == "Y"`) appear **at the top**, then within each group sort by `TRTSDT` ascending. Use `desc()` applied to a logical expression to push `TRUE` rows to the top.

Show `USUBJID`, `TRT01A`, `DTHFL`, `TRTSDT` for the first 10 rows.

```{r arrange-clinical, exercise=TRUE, exercise.lines=8}
adsl |>
  arrange(desc(___ == "Y"), ___) |>
  select(USUBJID, TRT01A, DTHFL, TRTSDT) |>
  head(10)
```

```{r arrange-clinical-hint-1}
# desc(DTHFL == "Y") — applying desc() to a logical puts TRUE before FALSE
# Then sort by TRTSDT for tie-breaking within each group
```

```{r arrange-clinical-solution}
adsl |>
  arrange(desc(DTHFL == "Y"), TRTSDT) |>
  select(USUBJID, TRT01A, DTHFL, TRTSDT) |>
  head(10)
```

### Exercise 5b — NAs sort last

By default `arrange()` sorts NAs last (ascending) and first (descending). Filter `adae` to show just the ongoing AEs (`AEENDTC` is NA), sort by `USUBJID`. Confirm the NAs appear.

```{r arrange-na, exercise=TRUE, exercise.lines=6}
adae |>
  arrange(USUBJID, AEENDTC) |>
  filter(is.na(AEENDTC)) |>
  select(USUBJID, AEDECOD, AESTDTC, AEENDTC)
```

---

## 6 — summarise() — PROC MEANS / PROC SUMMARY

### Concept

```r
adsl |>
  filter(SAFFL == "Y") |>
  summarise(
    n           = n(),
    n_distinct  = n_distinct(RACE),
    mean_age    = mean(AGE, na.rm = TRUE),
    median_age  = median(AGE, na.rm = TRUE),
    q25_age     = quantile(AGE, 0.25, na.rm = TRUE),
    q75_age     = quantile(AGE, 0.75, na.rm = TRUE)
  )
```

**Always `na.rm = TRUE`** — R propagates NA where SAS silently skips.

**`sd()` — standard deviation** (base R, no library): `sd(AGE, na.rm = TRUE)`. Combine with `mean()` for descriptive statistics tables:

```r
summarise(
  mean = round(mean(AGE, na.rm = TRUE), 1),
  sd   = round(sd(AGE,   na.rm = TRUE), 1)
)
```

**`mean()` of a logical as a proportion** — in R, `TRUE` = 1 and `FALSE` = 0. Taking the mean of a logical vector gives the proportion of TRUE rows — a clean, idiomatic way to compute percentages:

```r
# Proportion of subjects with DIABFL == "Y"
summarise(pct_diab = round(mean(DIABFL == "Y", na.rm = TRUE) * 100, 1))
# Equivalent SAS: mean(diabfl='Y') * 100
```

### Exercise 6a — full descriptive stats on AGE

For the safety population, compute n, mean, SD, median, Q1, Q3, min, max of `AGE`.

```{r summarise-full, exercise=TRUE, exercise.lines=16}
adsl |>
  filter(SAFFL == "Y") |>
  summarise(
    n        = n(),
    mean     = round(mean(AGE, na.rm = TRUE), 1),
    sd       = round(sd(AGE, na.rm = TRUE), 1),
    median   = median(AGE, na.rm = TRUE),
    q1       = quantile(AGE, ___, na.rm = TRUE),
    q3       = quantile(AGE, ___, na.rm = TRUE),
    min      = min(AGE, na.rm = TRUE),
    max      = max(AGE, na.rm = TRUE)
  )
```

```{r summarise-full-hint-1}
# quantile(AGE, 0.25, ...) for Q1; quantile(AGE, 0.75, ...) for Q3
```

```{r summarise-full-solution}
adsl |>
  filter(SAFFL == "Y") |>
  summarise(
    n      = n(),
    mean   = round(mean(AGE, na.rm = TRUE), 1),
    sd     = round(sd(AGE, na.rm = TRUE), 1),
    median = median(AGE, na.rm = TRUE),
    q1     = quantile(AGE, 0.25, na.rm = TRUE),
    q3     = quantile(AGE, 0.75, na.rm = TRUE),
    min    = min(AGE, na.rm = TRUE),
    max    = max(AGE, na.rm = TRUE)
  )
```

### Exercise 6b — by-treatment summary

Group by `TRT01A` and summarise: n, mean age, and percentage who are diabetic (`DIABFL == "Y"`).

```{r summarise-by-trt, exercise=TRUE, exercise.lines=12}
adsl |>
  filter(SAFFL == "Y") |>
  group_by(___) |>
  summarise(
    n          = n(),
    mean_age   = round(mean(AGE, na.rm = TRUE), 1),
    pct_diab   = round(mean(DIABFL == "Y") * 100, 1),
    .groups    = "drop"
  )
```

```{r summarise-by-trt-hint-1}
# group_by(TRT01A)
# mean(DIABFL == "Y") gives proportion TRUE (since TRUE=1, FALSE=0)
```

```{r summarise-by-trt-solution}
adsl |>
  filter(SAFFL == "Y") |>
  group_by(TRT01A) |>
  summarise(
    n        = n(),
    mean_age = round(mean(AGE, na.rm = TRUE), 1),
    pct_diab = round(mean(DIABFL == "Y") * 100, 1),
    .groups  = "drop"
  )
```

### Exercise 6c — n_distinct count of unique values

For `adae`, summarise: total AE records (`n()`), distinct subjects with at least one AE (`n_distinct(USUBJID)`), and distinct preferred terms (`n_distinct(AEDECOD)`).

```{r summarise-ndistinct, exercise=TRUE, exercise.lines=10}
adae |>
  summarise(
    total_aes    = n(),
    n_subjects   = n_distinct(___),
    n_pt         = n_distinct(___),
    n_soc        = n_distinct(AEBODSYS)
  )
```

```{r summarise-ndistinct-hint-1}
# n_distinct(USUBJID), n_distinct(AEDECOD)
```

```{r summarise-ndistinct-solution}
adae |>
  summarise(
    total_aes  = n(),
    n_subjects = n_distinct(USUBJID),
    n_pt       = n_distinct(AEDECOD),
    n_soc      = n_distinct(AEBODSYS)
  )
```

### Exercise 6d — AE count by body system and severity

Summarise `adae` by `AEBODSYS` and `AESEV`: count records and count distinct subjects. Sort descending by n.

```{r summarise-ae-soc, exercise=TRUE, exercise.lines=10}
adae |>
  group_by(___, ___) |>
  summarise(
    n_ae   = n(),
    n_subj = n_distinct(USUBJID),
    .groups = "drop"
  ) |>
  arrange(desc(n_ae))
```

```{r summarise-ae-soc-hint-1}
# group_by(AEBODSYS, AESEV)
```

```{r summarise-ae-soc-solution}
adae |>
  group_by(AEBODSYS, AESEV) |>
  summarise(
    n_ae   = n(),
    n_subj = n_distinct(USUBJID),
    .groups = "drop"
  ) |>
  arrange(desc(n_ae))
```

---

## 7 — group_by() + mutate() — RETAIN within BY groups

### Concept

`group_by()` + `mutate()` keeps all rows and computes values *within* groups. SAS equivalent: RETAIN + BY.

```r
adlb |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    BASELINE    = first(AVAL[AVISITN == 0]),  # first() grabs one value from a subset
    CHG         = AVAL - BASELINE,
    PCT_SUBJ    = AVAL / sum(AVAL) * 100       # subject's share of total
  ) |>
  ungroup()
```

**`first()` and `last()`** — return the first/last element of a vector. Inside `group_by() + mutate()` this looks up a specific row's value for the entire group:

```r
first(AVAL[AVISITN == 0])   # AVAL at baseline, broadcast to all rows of the group
last(AVAL)                  # AVAL at the last (highest-index) row of the group
```

**`min_rank()` — ranking within a group:** assigns ranks 1, 2, 3 … with ties getting the lowest rank (like SAS `RANK`). Wrap with `desc()` to rank descending:

```r
group_by(PARAMCD, AVISIT) |>
  mutate(RANK = min_rank(AVAL))          # lowest AVAL = rank 1
  mutate(RANK = min_rank(desc(AVAL)))    # highest AVAL = rank 1
```

**`add_count()` — add group size without collapsing:** shorthand for `group_by() |> mutate(n = n()) |> ungroup()`. Keeps all rows and adds a count column:

```r
adae |> add_count(USUBJID, name = "N_AE_SUBJ")   # total AEs per subject, on every row
# identical to:
adae |> group_by(USUBJID) |> mutate(N_AE_SUBJ = n()) |> ungroup()
```

### Exercise 7a — baseline and change from baseline

In `adlb`, add `BL_AVAL` (the AVAL at AVISITN == 0 within each subject × parameter) and `CHG` (AVAL - BL_AVAL) to every row.

```{r groupmutate-chg, exercise=TRUE, exercise.lines=14}
adlb |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    BL_AVAL = first(AVAL[AVISITN == ___]),
    CHG     = AVAL - BL_AVAL
  ) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, BL_AVAL, CHG)
```

```{r groupmutate-chg-hint-1}
# AVISITN == 0 for baseline
# first() returns the first matching value
```

```{r groupmutate-chg-solution}
adlb |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    BL_AVAL = first(AVAL[AVISITN == 0]),
    CHG     = AVAL - BL_AVAL
  ) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, BL_AVAL, CHG)
```

### Exercise 7b — percentage of group total

For each `PARAMCD` and `AVISIT` combination in `adlb`, add `PCT_OF_VISIT_TOTAL`: each subject's AVAL as a percentage of the sum across all subjects at that visit.

```{r groupmutate-pct, exercise=TRUE, exercise.lines=12}
adlb |>
  group_by(PARAMCD, AVISIT) |>
  mutate(
    VISIT_TOTAL    = sum(AVAL, na.rm = TRUE),
    PCT_OF_VISIT_TOTAL = round(___ / VISIT_TOTAL * 100, 1)
  ) |>
  ungroup() |>
  filter(PARAMCD == "ALT") |>
  select(USUBJID, AVISIT, AVAL, VISIT_TOTAL, PCT_OF_VISIT_TOTAL)
```

```{r groupmutate-pct-hint-1}
# PCT_OF_VISIT_TOTAL = round(AVAL / VISIT_TOTAL * 100, 1)
```

```{r groupmutate-pct-solution}
adlb |>
  group_by(PARAMCD, AVISIT) |>
  mutate(
    VISIT_TOTAL        = sum(AVAL, na.rm = TRUE),
    PCT_OF_VISIT_TOTAL = round(AVAL / VISIT_TOTAL * 100, 1)
  ) |>
  ungroup() |>
  filter(PARAMCD == "ALT") |>
  select(USUBJID, AVISIT, AVAL, VISIT_TOTAL, PCT_OF_VISIT_TOTAL)
```

### Exercise 7c — rank within group

For each `PARAMCD` and `AVISIT`, rank subjects by `AVAL` descending (highest = rank 1). Use `min_rank(desc(AVAL))`.

```{r groupmutate-rank, exercise=TRUE, exercise.lines=12}
adlb |>
  group_by(PARAMCD, AVISIT) |>
  mutate(
    RANK_AVAL = min_rank(desc(___))
  ) |>
  ungroup() |>
  filter(PARAMCD == "ALT") |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, RANK_AVAL) |>
  arrange(AVISIT, RANK_AVAL)
```

```{r groupmutate-rank-hint-1}
# min_rank(desc(AVAL))
```

```{r groupmutate-rank-solution}
adlb |>
  group_by(PARAMCD, AVISIT) |>
  mutate(
    RANK_AVAL = min_rank(desc(AVAL))
  ) |>
  ungroup() |>
  filter(PARAMCD == "ALT") |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, RANK_AVAL) |>
  arrange(AVISIT, RANK_AVAL)
```

### Exercise 7d — count AEs per subject and add back (add_count)

`add_count()` is shorthand for `group_by() |> mutate(n = n()) |> ungroup()`. Add `N_AE_SUBJ` (total AEs per subject) to every row of `adae`.

```{r addcount-ex, exercise=TRUE, exercise.lines=8}
adae |>
  add_count(___, name = "N_AE_SUBJ") |>
  select(USUBJID, AEDECOD, N_AE_SUBJ) |>
  arrange(USUBJID)
```

```{r addcount-ex-hint-1}
# add_count(USUBJID, name = "N_AE_SUBJ")
```

```{r addcount-ex-solution}
adae |>
  add_count(USUBJID, name = "N_AE_SUBJ") |>
  select(USUBJID, AEDECOD, N_AE_SUBJ) |>
  arrange(USUBJID)
```

---

## 8 — distinct() and count()

### Concept

These two functions are your fast paths for frequency tables and deduplication:

```r
adsl |> distinct(TRT01A)                      # unique values of one column
adsl |> distinct(TRT01A, SEX)                 # unique combinations
adsl |> distinct(USUBJID, .keep_all = TRUE)   # first row per subject, keep all columns

adae |> count(AEBODSYS)                       # tabulate: returns col + n
adae |> count(TRT01A, AESEV, sort = TRUE)     # cross-tab, sorted descending by n
adae |> count(AEBODSYS, name = "n_soc")       # rename the count column (default is "n")
```

**`count()` arguments:**

| Argument | Effect | Default |
|---|---|---|
| `...` | grouping columns | — |
| `sort = TRUE` | sort descending by count | `FALSE` |
| `name = "col"` | rename the count column | `"n"` |
| `wt = AVAL` | sum `wt` instead of counting rows | `NULL` |

### Exercise 8a — distinct preferred terms per body system

Find distinct (AEBODSYS, AEDECOD) combinations in `adae`, then count how many unique preferred terms each body system has.

```{r distinct-pt, exercise=TRUE, exercise.lines=7}
adae |>
  distinct(AEBODSYS, ___) |>
  count(AEBODSYS, name = "n_pt") |>
  arrange(desc(n_pt))
```

```{r distinct-pt-hint-1}
# distinct(AEBODSYS, AEDECOD) then count(AEBODSYS, ...)
```

```{r distinct-pt-solution}
adae |>
  distinct(AEBODSYS, AEDECOD) |>
  count(AEBODSYS, name = "n_pt") |>
  arrange(desc(n_pt))
```

### Exercise 8b — cross-tabulation with count()

Use `count()` to create a cross-tab of `TRT01A` × `AESEV` in `adae`. Sort by `TRT01A` then descending `n`.

```{r count-crosstab, exercise=TRUE, exercise.lines=7}
adae |>
  count(___, ___, sort = FALSE) |>
  arrange(TRT01A, desc(n))
```

```{r count-crosstab-hint-1}
# count(TRT01A, AESEV)
```

```{r count-crosstab-solution}
adae |>
  count(TRT01A, AESEV, sort = FALSE) |>
  arrange(TRT01A, desc(n))
```

### Exercise 8c — proportion using count + mutate

Use `count()` to tabulate subjects by `TRT01A` in `adsl` (SAFFL == "Y"), then add a `PCT` column showing each arm's percentage of the total.

```{r count-prop, exercise=TRUE, exercise.lines=10}
adsl |>
  filter(SAFFL == "Y") |>
  count(TRT01A) |>
  mutate(PCT = round(n / sum(n) * 100, 1))
```

---

## 9 — String operations in clinical context

### Concept

SAS programmers rely heavily on character functions. The R equivalents (**all base R unless noted — no library needed**):

| SAS | R (base) | R (stringr) |
|---|---|---|
| `upcase()` | `toupper()` | `str_to_upper()` |
| `lowcase()` | `tolower()` | `str_to_lower()` |
| `substr(x,1,3)` | `substr(x,1,3)` | `str_sub(x,1,3)` |
| `index(x,"Y")` | `regexpr("Y",x)` | `str_detect(x,"Y")` |
| `prxmatch` | `grepl()` | `str_detect()` |
| `cats(a,b)` | `paste0(a,b)` | `str_c(a,b)` |
| `trim(x)` | `trimws(x)` | `str_trim(x)` |
| `lengthn(x)` | `nchar(x)` | `str_length(x)` |
| `tranwrd(x,old,new)` | `gsub(old,new,x)` | `str_replace_all(x,old,new)` |
| `put(x, fmt.)` / `catx` | `sprintf(fmt, ...)` | `str_glue(fmt, ...)` |

---

### Key string functions explained

**`substr(x, start, stop)` — extract a substring (base R)**

Extracts characters at positions `start` through `stop` (1-based indexing):

```r
substr("01-701-1015", 1, 2)   # "01"  — study segment
substr("01-701-1015", 4, 6)   # "701" — site segment
substr("01-701-1015", 8, 11)  # "1015"— subject segment

# Vectorised inside mutate():
adsl |> mutate(SITE = substr(USUBJID, 4, 6))

# Extract last N characters: use nchar() to find end position
substr(x, nchar(x) - 3, nchar(x))   # last 4 characters
```

**`nchar(x)` — string length:** returns the number of characters. `nchar("01-701-1015")` = 11.

**`grepl(pattern, x, ignore.case = FALSE)` — pattern detection (base R)**

Returns a logical vector: `TRUE` where the pattern is found. The pattern is a regular expression:

```r
grepl("Xanomeline", TRT01A)                   # matches any Xanomeline arm
grepl("xanomeline", TRT01A, ignore.case=TRUE) # case-insensitive
grepl("^ALT$", PARAMCD)                       # exact match (^ = start, $ = end)
grepl("(NAUSEA|VOMITING)", AEDECOD)           # match either word
grepl("^APPLICATION", AEDECOD)               # starts with "APPLICATION"

# Typical uses:
filter(grepl("Skin", AEBODSYS, ignore.case = TRUE))   # filter on pattern
mutate(SKIN_FL = grepl("Skin", AEBODSYS, ignore.case = TRUE))  # flag column
```

**`gsub(pattern, replacement, x)` — replace all matches (base R)**

Replaces every occurrence of `pattern` with `replacement`. Like SAS `tranwrd()` but accepts full regex:

```r
gsub("BLACK OR AFRICAN AMERICAN", "BLACK", RACE)   # literal string replacement
gsub("\\s+", "_", PARAM)                           # replace whitespace runs with "_"
gsub("^\\s+|\\s+$", "", x)                         # trim whitespace (same as trimws())
gsub("[^A-Za-z0-9]", "", x)                        # keep only alphanumeric characters
```

`sub()` replaces only the **first** match; `gsub()` replaces **all** matches.

**`paste0(..., sep="")` — string concatenation (base R)**

Concatenates strings with no separator (equivalent to SAS `cats()`). Use `paste(..., sep="-")` when you want a separator:

```r
paste0(SITEID, "-", USUBJID)           # "701-01-701-1015"
paste("Site", SITEID, sep = ": ")      # "Site: 701"
paste0(rep("X", 5), collapse = "")     # "XXXXX" — collapse vector to single string
```

**`toupper()` / `tolower()` — case conversion (base R)**

```r
toupper("mild")       # "MILD"
tolower("MODERATE")   # "moderate"
# Vectorised in mutate():
mutate(across(where(is.character), toupper))
```

**`trimws(x, which = "both")` — strip whitespace (base R)**

Removes leading and/or trailing whitespace. `which` can be `"left"`, `"right"`, or `"both"` (default):

```r
trimws("  HEADACHE  ")   # "HEADACHE"
```

---

### Formatted strings: sprintf() and paste0()

**`sprintf()` is a base R function — no library needed.** It works exactly like C's `printf` and SAS's `put()` / `catx()` combination: you write a **format string** with placeholders, then pass the values to substitute in.

```r
sprintf(format_string, value1, value2, ...)
```

**Format placeholders:**

| Placeholder | Meaning | Example | Output |
|---|---|---|---|
| `%s` | character / string | `sprintf("%s", "MALE")` | `"MALE"` |
| `%d` | integer | `sprintf("%d", 42L)` | `"42"` |
| `%f` | floating-point | `sprintf("%f", 3.14)` | `"3.140000"` |
| `%.1f` | float, 1 decimal | `sprintf("%.1f", 3.14159)` | `"3.1"` |
| `%05d` | integer, zero-padded to width 5 | `sprintf("%05d", 7)` | `"00007"` |
| `%-10s` | string, left-aligned in width 10 | `sprintf("%-10s|", "LOW")` | `"LOW       |"` |

**Clinical examples — inside `mutate()`:**

```r
# "[01-701-1015] | Xanomeline High Dose"
adsl |> mutate(LABEL = sprintf("[%s] | %s", USUBJID, TRT01A))

# "AGE: 63 yrs  BMI: 24.8"
adsl |> mutate(SUMMARY = sprintf("AGE: %d yrs  BMI: %.1f", AGE, BMIBL))

# "SITE-007  (n=  3 subjects)"  — fixed-width columns for aligned output
adsl |> group_by(SITEID) |>
  mutate(SITE_LABEL = sprintf("SITE-%-3s  (n=%3d subjects)", SITEID, n())) |>
  ungroup()
```

**`paste0()` vs `sprintf()`** — Use `paste0()` for simple concatenation with no formatting. Use `sprintf()` when you need decimal precision, zero-padding, or fixed-width alignment.

```r
paste0("[", USUBJID, "] | [", TRT01A, "]")   # fine for plain concatenation
sprintf("[%s] | [%s]", USUBJID, TRT01A)       # same result, cleaner template syntax
sprintf("%.1f%%", pct)                         # %% is a literal percent sign
```

---

### Exercise 9a — extract site from USUBJID

In the CDISC pilot, `USUBJID` has the form `"01-701-1015"` (study–site–subject). The site number `"701"` sits at characters 4–6. Use `substr()` to extract it and verify it matches `SITEID`.

```{r strings-substr, exercise=TRUE, exercise.lines=6}
adsl |>
  mutate(SITE_EXTRACTED = substr(USUBJID, ___, ___)) |>
  select(USUBJID, SITEID, SITE_EXTRACTED) |>
  head(4)
```

```{r strings-substr-hint-1}
# In "01-701-1015", the site segment "701" is at positions 4-6
# substr(USUBJID, 4, 6)
```

```{r strings-substr-solution}
adsl |>
  mutate(SITE_EXTRACTED = substr(USUBJID, 4, 6)) |>
  select(USUBJID, SITEID, SITE_EXTRACTED) |>
  head(4)
```

### Exercise 9b — detect pattern and flag

Add `SKIN_AE` = `TRUE` when `AEBODSYS` contains `"Skin"` (case-insensitive). Use `grepl(..., ignore.case = TRUE)`.

```{r strings-detect, exercise=TRUE, exercise.lines=7}
adae |>
  mutate(SKIN_AE = grepl("___", AEBODSYS, ignore.case = TRUE)) |>
  filter(SKIN_AE) |>
  select(USUBJID, AEDECOD, AEBODSYS, SKIN_AE)
```

```{r strings-detect-hint-1}
# grepl("Skin", AEBODSYS, ignore.case = TRUE)
```

```{r strings-detect-solution}
adae |>
  mutate(SKIN_AE = grepl("Skin", AEBODSYS, ignore.case = TRUE)) |>
  filter(SKIN_AE) |>
  select(USUBJID, AEDECOD, AEBODSYS, SKIN_AE)
```

### Exercise 9c — multi-field formatted label with sprintf()

Using `sprintf()`, build a column `SUBJ_SUMMARY` in `adsl` that produces a label like:

```
"[01-701-1015] | Placebo | Age: 63 | BMI: 24.8"
```

The format template is `"[%s] | %s | Age: %d | BMI: %.1f"` — fill in `USUBJID`, `TRT01A`, `AGE`, and `BMIBL`.

```{r strings-label, exercise=TRUE, exercise.lines=8}
adsl |>
  mutate(
    SUBJ_SUMMARY = sprintf("[%s] | %s | Age: %d | BMI: %.1f",
                           ___, ___, ___, ___)
  ) |>
  select(USUBJID, TRT01A, AGE, BMIBL, SUBJ_SUMMARY) |>
  head(4)
```

```{r strings-label-hint-1}
# sprintf("[%s] | %s | Age: %d | BMI: %.1f", USUBJID, TRT01A, AGE, BMIBL)
# %s = string, %d = integer, %.1f = float with 1 decimal
# Arguments fill placeholders left-to-right
```

```{r strings-label-solution}
adsl |>
  mutate(
    SUBJ_SUMMARY = sprintf("[%s] | %s | Age: %d | BMI: %.1f",
                           USUBJID, TRT01A, AGE, BMIBL)
  ) |>
  select(USUBJID, TRT01A, AGE, BMIBL, SUBJ_SUMMARY) |>
  head(4)
```

### Exercise 9d — replace and standardize

In `adsl`, `RACE` uses full CDISC-controlled terminology labels (e.g., `"BLACK OR AFRICAN AMERICAN"`). Use `gsub()` to shorten them for display tables. Also shorten `"AMERICAN INDIAN OR ALASKA NATIVE"` to `"AM IND"`. Use `count()` to verify.

```{r strings-replace, exercise=TRUE, exercise.lines=10}
adsl |>
  mutate(
    RACE_SHORT = RACE,
    RACE_SHORT = gsub("BLACK OR AFRICAN AMERICAN", "BLACK", RACE_SHORT),
    RACE_SHORT = gsub("AMERICAN INDIAN OR ALASKA NATIVE", "AM IND", RACE_SHORT)
  ) |>
  count(RACE_SHORT)
```

---

## 10 — FIRST./LAST. logic: slice_min, slice_max, row_number

### Concept

```r
# SAS: if first.usubjid then output;
adae |> group_by(USUBJID) |>
  slice_min(AESTDTC, n=1, with_ties=FALSE) |> ungroup()

# SAS: if last.usubjid then output;
adae |> group_by(USUBJID) |>
  slice_max(AESTDTC, n=1, with_ties=FALSE) |> ungroup()

# Flag first/last rows (keep all rows)
adae |> arrange(USUBJID, AESTDTC) |> group_by(USUBJID) |>
  mutate(first = row_number() == 1, last = row_number() == n()) |> ungroup()
```

**`slice(i)` — keep rows by position:** After `arrange() + group_by()`, `slice(1)` keeps the first row per group (as ordered by your sort). This is often more convenient than `slice_min/max` when the sort is already applied:

```r
# Worst AE per subject: sort first, then slice(1) to take the top row per group
adae |>
  arrange(USUBJID, desc(AESEVN), AESTDTC) |>
  group_by(USUBJID) |>
  slice(1) |>           # first row after sort = worst AE
  ungroup()

# Also valid: slice(1:3) keeps first 3 rows per group
```

**`row_number()` — sequential integer within group:** Starts at 1 for each new group (equivalent to `_n_` in SAS DATA step after BY):

```r
adae |> arrange(USUBJID, AESTDTC) |>
  group_by(USUBJID) |>
  mutate(AESEQ = row_number()) |>
  ungroup()
```

### Exercise 10a — worst AE per subject

For each subject, keep only the **most severe AE** (highest `AESEVN`). Break ties by earliest `AESTDTC`.

```{r firstlast-worst, exercise=TRUE, exercise.lines=10}
adae |>
  arrange(USUBJID, desc(AESEVN), AESTDTC) |>
  group_by(___) |>
  slice(1) |>
  ungroup() |>
  select(USUBJID, AEDECOD, AESEV, AESEVN, AESTDTC)
```

```{r firstlast-worst-hint-1}
# group_by(USUBJID), slice(1) after sorting
```

```{r firstlast-worst-solution}
adae |>
  arrange(USUBJID, desc(AESEVN), AESTDTC) |>
  group_by(USUBJID) |>
  slice(1) |>
  ungroup() |>
  select(USUBJID, AEDECOD, AESEV, AESEVN, AESTDTC)
```

### Exercise 10b — sequence number per subject (like AESEQ)

Assign a row sequence number `AE_ROWNUM` per subject (1-based, ordered by `AESTDTC`). This is the pattern used to derive AESEQ in ADAE.

```{r firstlast-seq, exercise=TRUE, exercise.lines=10}
adae |>
  arrange(USUBJID, AESTDTC) |>
  group_by(___) |>
  mutate(AE_ROWNUM = row_number()) |>
  ungroup() |>
  select(USUBJID, AE_ROWNUM, AESTDTC, AEDECOD) |>
  head(8)
```

```{r firstlast-seq-hint-1}
# group_by(USUBJID) then row_number()
```

```{r firstlast-seq-solution}
adae |>
  arrange(USUBJID, AESTDTC) |>
  group_by(USUBJID) |>
  mutate(AE_ROWNUM = row_number()) |>
  ungroup() |>
  select(USUBJID, AE_ROWNUM, AESTDTC, AEDECOD) |>
  head(8)
```

### Exercise 10c — last non-missing value per group

In `adlb`, get the **last observed AVAL** per subject × parameter (i.e., highest AVISITN with non-missing AVAL). This is the "last observation carried forward" source record.

```{r firstlast-locf-source, exercise=TRUE, exercise.lines=10}
adlb |>
  filter(!is.na(AVAL)) |>
  group_by(USUBJID, PARAMCD) |>
  slice_max(___, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVISIT, AVISITN, AVAL)
```

```{r firstlast-locf-source-hint-1}
# slice_max(AVISITN, n=1, with_ties=FALSE) to get the latest visit
```

```{r firstlast-locf-source-solution}
adlb |>
  filter(!is.na(AVAL)) |>
  group_by(USUBJID, PARAMCD) |>
  slice_max(AVISITN, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVISIT, AVISITN, AVAL)
```

---

## 11 — lag() and lead() — visit-to-visit changes

### Concept

`lag()` returns the prior value; `lead()` the next value. **Always use inside `group_by()`** so the lag doesn't bleed across subjects.

> **Data type gotcha — always verify column types before comparing:**
> In `pharmaverseadam::adlb`, `ATOXGR` is stored as **character** (`<chr>`), not integer. Running `ATOXGR > ATOXGR_PREV` on character strings does **alphabetic** comparison — `"0" > "-1"` returns `TRUE` because `"-"` sorts before `"0"` lexicographically, which is numerically wrong. The setup chunk in this lesson already converts `ATOXGR` to integer with `as.integer()`, but in your own ADaM pipelines always check `class(df$ATOXGR)` or `str(df)` before numeric operations on grade variables.

> **DTYPE gotcha — exclude admiral-derived rows before any lag/lead analysis:**
> admiral appends derived summary records to BDS datasets alongside the real observation rows. These rows carry a non-missing `DTYPE` value (e.g. `"MAXIMUM"`, `"MINIMUM"`, `"LOCF"`) and use sentinel `AVISITN` values such as **9998** ("POST-BASELINE MAXIMUM") or **9997** ("POST-BASELINE MINIMUM"). They are summary artefacts, not actual study visits.
> If you run `arrange(..., AVISITN) |> lag()` without filtering them out first, the sentinel row sorts last inside each group and the lag comparison fires against the final real visit — incorrectly flagging the derived "maximum" as a worsening event.
> **Always filter them away before longitudinal work:**
> ```r
> adlb |> filter(is.na(DTYPE)) |> arrange(USUBJID, PARAMCD, AVISITN) |> ...
> ```
> The setup chunk in this lesson already applies `filter(is.na(DTYPE))` globally to `adlb`, so exercises here are safe. In your own work, add this filter as a reflex whenever you touch visit-sequence logic on a BDS dataset built with admiral.

**`first(x[condition])` — baseline lookup within a group:** Introduced in Section 7, this pattern is important in lag/lead exercises too. Inside `group_by() + mutate()`, it looks up one row's value and broadcasts it to all rows in the group:

```r
group_by(USUBJID, PARAMCD) |>
  mutate(
    BL_AVAL   = first(AVAL[AVISITN == 0]),   # baseline value on every row
    PCT_CHG_BL = round((AVAL - BL_AVAL) / BL_AVAL * 100, 1)
  )
```

The `[AVISITN == 0]` subset filters to baseline rows first; `first()` returns the single value from that subset. This pattern combines naturally with `lag()` in the same `mutate()` call.

```r
adlb |>
  arrange(USUBJID, PARAMCD, AVISITN) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    AVAL_PREV   = lag(AVAL),
    AVAL_NEXT   = lead(AVAL),
    CHG_PREV    = AVAL - AVAL_PREV,
    PCT_CHG     = (AVAL - AVAL_PREV) / AVAL_PREV * 100
  ) |>
  ungroup()
```

### Exercise 11a — % change from prior visit AND from baseline in one mutate()

In a single `group_by() + mutate()` block, derive **two** percent-change columns for `adlb`:

- `PCT_CHG_PREV` — % change vs the **previous visit** (use `lag()`)
- `PCT_CHG_BL` — % change vs **baseline** (AVISITN == 0; use `first()` to look it up within the group)

Both rounded to 1 decimal. Then filter to ALT post-baseline rows only.

```{r lag-pct-chg, exercise=TRUE, exercise.lines=16}
adlb |>
  arrange(USUBJID, PARAMCD, AVISITN) |>
  group_by(USUBJID, ___) |>
  mutate(
    AVAL_PREV    = lag(AVAL),
    BL_AVAL      = first(AVAL[AVISITN == ___]),
    PCT_CHG_PREV = round((___ - AVAL_PREV) / AVAL_PREV * 100, 1),
    PCT_CHG_BL   = round((___ - BL_AVAL)   / BL_AVAL   * 100, 1)
  ) |>
  ungroup() |>
  filter(PARAMCD == "ALT", AVISITN > 0) |>
  select(USUBJID, AVISIT, AVAL, AVAL_PREV, BL_AVAL, PCT_CHG_PREV, PCT_CHG_BL)
```

```{r lag-pct-chg-hint-1}
# group_by(USUBJID, PARAMCD)
# BL_AVAL = first(AVAL[AVISITN == 0])  — within-group lookup to baseline row
# PCT_CHG_PREV uses lag(AVAL); PCT_CHG_BL uses BL_AVAL (same call, both available)
```

```{r lag-pct-chg-solution}
adlb |>
  arrange(USUBJID, PARAMCD, AVISITN) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    AVAL_PREV    = lag(AVAL),
    BL_AVAL      = first(AVAL[AVISITN == 0]),
    PCT_CHG_PREV = round((AVAL - AVAL_PREV) / AVAL_PREV * 100, 1),
    PCT_CHG_BL   = round((AVAL - BL_AVAL)   / BL_AVAL   * 100, 1)
  ) |>
  ungroup() |>
  filter(PARAMCD == "ALT", AVISITN > 0) |>
  select(USUBJID, AVISIT, AVAL, AVAL_PREV, BL_AVAL, PCT_CHG_PREV, PCT_CHG_BL)
```

### Exercise 11b — detect worsening toxicity grade

Add `GRADE_WORSENED = TRUE` when `ATOXGR` is **strictly higher** than the previous visit's `ATOXGR` (within subject × parameter).

```{r lag-worsened, exercise=TRUE, exercise.lines=14}
adlb |>
  arrange(USUBJID, PARAMCD, AVISITN) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    ATOXGR_PREV    = lag(ATOXGR),
    GRADE_WORSENED = !is.na(ATOXGR_PREV) & ___ > ATOXGR_PREV
  ) |>
  ungroup() |>
  filter(GRADE_WORSENED) |>
  select(USUBJID, PARAMCD, AVISIT, ATOXGR_PREV, ATOXGR, GRADE_WORSENED)
```

```{r lag-worsened-hint-1}
# ATOXGR > ATOXGR_PREV (and guard against NA with !is.na(ATOXGR_PREV))
```

```{r lag-worsened-solution}
adlb |>
  arrange(USUBJID, PARAMCD, AVISITN) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    ATOXGR_PREV    = lag(ATOXGR),
    GRADE_WORSENED = !is.na(ATOXGR_PREV) & ATOXGR > ATOXGR_PREV
  ) |>
  ungroup() |>
  filter(GRADE_WORSENED) |>
  select(USUBJID, PARAMCD, AVISIT, ATOXGR_PREV, ATOXGR, GRADE_WORSENED)
```

### Exercise 11c — consecutive worsening (grade worsened two visits in a row)

Extend the previous: flag `CONSEC_WORSE = TRUE` when ATOXGR increased at **both** the current AND the previous visit compared to their respective prior visits.

```{r lag-consec, exercise=TRUE, exercise.lines=16}
adlb |>
  arrange(USUBJID, PARAMCD, AVISITN) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    ATOXGR_PREV   = lag(ATOXGR),
    WORSE_NOW     = !is.na(ATOXGR_PREV) & ATOXGR > ATOXGR_PREV,
    WORSE_PREV    = lag(WORSE_NOW),                    # was previous row also a worsening?
    CONSEC_WORSE  = WORSE_NOW & !is.na(WORSE_PREV) & ___
  ) |>
  ungroup() |>
  filter(CONSEC_WORSE) |>
  select(USUBJID, PARAMCD, AVISIT, ATOXGR, WORSE_NOW, WORSE_PREV, CONSEC_WORSE)
```

```{r lag-consec-hint-1}
# CONSEC_WORSE = WORSE_NOW & !is.na(WORSE_PREV) & WORSE_PREV == TRUE
```

```{r lag-consec-solution}
adlb |>
  arrange(USUBJID, PARAMCD, AVISITN) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    ATOXGR_PREV  = lag(ATOXGR),
    WORSE_NOW    = !is.na(ATOXGR_PREV) & ATOXGR > ATOXGR_PREV,
    WORSE_PREV   = lag(WORSE_NOW),
    CONSEC_WORSE = WORSE_NOW & !is.na(WORSE_PREV) & WORSE_PREV == TRUE
  ) |>
  ungroup() |>
  filter(CONSEC_WORSE) |>
  select(USUBJID, PARAMCD, AVISIT, ATOXGR, WORSE_NOW, WORSE_PREV, CONSEC_WORSE)
```

---

## 12 — Cumulative functions

### Concept

All cumulative functions below are **base R** — no library needed. All respect `group_by()` so they never bleed across subjects.

```r
cumsum(x)    # running total:    c(1,2,3)  → c(1,3,6)
cummax(x)    # running maximum:  c(1,3,2)  → c(1,3,3)
cummin(x)    # running minimum:  c(3,1,2)  → c(3,1,1)
cumprod(x)   # running product:  c(2,3,4)  → c(2,6,24)
cumany(cond) # latch TRUE:       c(F,F,T,F)→ c(F,F,T,T)  — stays TRUE once fired
cumall(cond) # latch FALSE:      c(T,T,F,T)→ c(T,T,F,F)  — stays FALSE once failed
```

**`cumany(condition)` — "has this ever happened?":** Takes a logical condition and returns TRUE from the first row where it fires, staying TRUE for all subsequent rows in the group. Equivalent to SAS RETAIN with a "once-true" flag:

```r
# EVER_SEVERE = TRUE from the first SEVERE AE onwards, for each subject
adae |>
  arrange(USUBJID, AESTDTC) |>
  group_by(USUBJID) |>
  mutate(EVER_SEVERE = cumany(AESEV == "SEVERE")) |>
  ungroup()
```

**`cumsum(1)` — running row count:** `cumsum()` of a constant `1` produces sequential integers 1, 2, 3 … — a clean way to add a running event count per group:

```r
adae |>
  arrange(USUBJID, AESTDTC) |>
  group_by(USUBJID) |>
  mutate(AE_N_TO_DATE = cumsum(1)) |>    # 1st AE=1, 2nd=2, 3rd=3 ...
  ungroup()
```

### Exercise 12a — worst toxicity grade to date

Add `WORST_GRADE_TD` = running maximum of `ATOXGR` per subject × parameter across visits ordered by `AVISITN`.

```{r cummax-worst, exercise=TRUE, exercise.lines=10}
adlb |>
  arrange(USUBJID, PARAMCD, ___) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(WORST_GRADE_TD = cummax(___)) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVISIT, ATOXGR, WORST_GRADE_TD)
```

```{r cummax-worst-hint-1}
# arrange by AVISITN, then cummax(ATOXGR)
```

```{r cummax-worst-solution}
adlb |>
  arrange(USUBJID, PARAMCD, AVISITN) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(WORST_GRADE_TD = cummax(ATOXGR)) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVISIT, ATOXGR, WORST_GRADE_TD)
```

### Exercise 12b — AE ever occurred flag (cumany)

In `adae`, sorted by USUBJID and AESTDTC, add `EVER_SEVERE` = cumany(AESEVN >= 3) — TRUE from the first severe AE onwards.

```{r cumany-ever, exercise=TRUE, exercise.lines=10}
adae |>
  arrange(USUBJID, AESTDTC) |>
  group_by(USUBJID) |>
  mutate(EVER_SEVERE = cumany(___ >= 3)) |>
  ungroup() |>
  select(USUBJID, AESTDTC, AEDECOD, AESEVN, EVER_SEVERE) |>
  arrange(USUBJID, AESTDTC)
```

```{r cumany-ever-hint-1}
# cumany(AESEVN >= 3)
```

```{r cumany-ever-solution}
adae |>
  arrange(USUBJID, AESTDTC) |>
  group_by(USUBJID) |>
  mutate(EVER_SEVERE = cumany(AESEVN >= 3)) |>
  ungroup() |>
  select(USUBJID, AESTDTC, AEDECOD, AESEVN, EVER_SEVERE) |>
  arrange(USUBJID, AESTDTC)
```

### Exercise 12c — running AE count per subject

Add `AE_COUNT_TO_DATE` = cumulative count of AEs per subject up to and including each row (sorted by date).

```{r cumsum-count, exercise=TRUE, exercise.lines=10}
adae |>
  arrange(USUBJID, AESTDTC) |>
  group_by(USUBJID) |>
  mutate(AE_COUNT_TO_DATE = cumsum(___)) |>  # cumsum of 1 per row = running count
  ungroup() |>
  select(USUBJID, AESTDTC, AEDECOD, AE_COUNT_TO_DATE)
```

```{r cumsum-count-hint-1}
# cumsum(1) — 1 for every row = running row count within subject
# Or: cumsum(rep(1, n()))
```

```{r cumsum-count-solution}
adae |>
  arrange(USUBJID, AESTDTC) |>
  group_by(USUBJID) |>
  mutate(AE_COUNT_TO_DATE = cumsum(1)) |>
  ungroup() |>
  select(USUBJID, AESTDTC, AEDECOD, AE_COUNT_TO_DATE)
```

---

## 13 — across() — apply operations to many columns

### Concept

```r
# Mean of all numeric columns
adsl |> summarise(across(where(is.numeric), ~ mean(.x, na.rm=TRUE)))

# Convert all character columns to upper case (bare function reference, no ~ needed)
adsl |> mutate(across(where(is.character), toupper))

# Chain functions in a lambda — trimws then toupper
adsl |> mutate(across(where(is.character), ~ toupper(trimws(.x))))

# Multiple summary functions, with custom output names
adsl |> summarise(across(c(AGE, HEIGHTBL, WEIGHTBL),
                          list(mean = ~ round(mean(.x, na.rm=TRUE), 1),
                               sd   = ~ round(sd(.x,   na.rm=TRUE), 1)),
                          .names = "{.col}_{.fn}"))   # → AGE_mean, AGE_sd, ...
```

**`.names` — controlling output column names in `across()`:**

The `.names` argument is a glue template using `{.col}` (original column name) and `{.fn}` (function name from the list):

```r
.names = "mean_{.col}"    # → mean_AGE, mean_HEIGHTBL, mean_WEIGHTBL
.names = "{.col}_{.fn}"   # → AGE_mean, AGE_sd  (requires named list of functions)
.names = "{.col}_new"     # → AGE_new, HEIGHTBL_new (literal suffix)
```

**`replace_na()` from `{tidyr}` — substitute NA with a value:**

```r
library(tidyr)

# Single column
adlb |> mutate(ATOXGR = replace_na(ATOXGR, 0))

# All numeric columns at once via across()
adlb |> mutate(across(where(is.numeric), ~ replace_na(.x, 0)))

# Alternatively with coalesce() (base dplyr, no tidyr needed)
adlb |> mutate(ATOXGR = coalesce(ATOXGR, 0L))
```

### Exercise 13a — mean of several baseline vars by treatment

For the safety population, compute mean of `AGE`, `HEIGHTBL`, `WEIGHTBL` by `TRT01A`. Name outputs `mean_AGE`, `mean_HEIGHTBL`, `mean_WEIGHTBL`.

```{r across-means, exercise=TRUE, exercise.lines=12}
adsl |>
  filter(SAFFL == "Y") |>
  group_by(TRT01A) |>
  summarise(
    across(
      c(___, ___, ___),
      ~ round(mean(.x, na.rm = TRUE), 1),
      .names = "mean_{.col}"
    ),
    .groups = "drop"
  )
```

```{r across-means-hint-1}
# c(AGE, HEIGHTBL, WEIGHTBL)
```

```{r across-means-solution}
adsl |>
  filter(SAFFL == "Y") |>
  group_by(TRT01A) |>
  summarise(
    across(
      c(AGE, HEIGHTBL, WEIGHTBL),
      ~ round(mean(.x, na.rm = TRUE), 1),
      .names = "mean_{.col}"
    ),
    .groups = "drop"
  )
```

### Exercise 13b — multiple summary functions in one across()

Compute both mean and SD of `AGE`, `HEIGHTBL`, `WEIGHTBL` in one `across()` call. Name outputs `AGE_mean`, `AGE_sd`, etc.

```{r across-multi-fn, exercise=TRUE, exercise.lines=14}
adsl |>
  filter(SAFFL == "Y") |>
  summarise(
    across(
      c(AGE, HEIGHTBL, WEIGHTBL),
      list(
        mean = ~ round(mean(.x, na.rm = TRUE), 1),
        sd   = ~ round(sd(.x, na.rm = TRUE), 1)
      ),
      .names = "{.col}_{.fn}"
    )
  )
```

### Exercise 13c — chain two string operations inside across()

Apply **two** transformations to every character column in `adsl` in a single `across()` lambda: first `trimws()` to strip any leading/trailing whitespace, then `toupper()` to capitalise. Use the `~ fn(.x)` lambda form so you can chain both calls.

```{r across-toupper, exercise=TRUE, exercise.lines=7}
adsl |>
  mutate(across(where(is.character), ~ toupper(trimws(___)))) |>
  select(USUBJID, TRT01A, RACE, SEX) |>
  head(3)
```

```{r across-toupper-hint-1}
# ~ toupper(trimws(.x))
# .x is the placeholder for the current column — trimws first, then toupper wraps it
```

```{r across-toupper-solution}
adsl |>
  mutate(across(where(is.character), ~ toupper(trimws(.x)))) |>
  select(USUBJID, TRT01A, RACE, SEX) |>
  head(3)
```

### Exercise 13d — coalesce NA with across

Replace all `NA` values in numeric columns with `0` using `across()` and `replace_na()` (from tidyr) or `coalesce()`.

```{r across-replace-na, exercise=TRUE, exercise.lines=8}
library(tidyr)
adlb |>
  mutate(across(where(is.numeric), ~ replace_na(.x, ___))) |>
  filter(PARAMCD == "ALT") |>
  select(USUBJID, AVISIT, AVAL, ATOXGR) |>
  head(5)
```

```{r across-replace-na-hint-1}
# replace_na(.x, 0) — replace NA with 0
```

```{r across-replace-na-solution}
library(tidyr)
adlb |>
  mutate(across(where(is.numeric), ~ replace_na(.x, 0))) |>
  filter(PARAMCD == "ALT") |>
  select(USUBJID, AVISIT, AVAL, ATOXGR) |>
  head(5)
```

---

## 14 — rowwise() — per-row operations

### Concept

Sometimes you need to apply a function *across columns within the same row* — not across rows within a column. `rowwise()` + `c_across()` handles this:

```r
# Maximum of SYSBP, DIABP, PULSE on each row
advs_wide |>
  rowwise() |>
  mutate(MAX_VITAL = max(c_across(c(SYSBP, DIABP, PULSE)), na.rm=TRUE)) |>
  ungroup()
```

For simple arithmetic involving a fixed number of columns, `pmax()` / `pmin()` is faster.

**`any()` and `all()` — aggregate logicals across a vector (base R):**

`any()` returns `TRUE` if at least one element is `TRUE`; `all()` requires every element to be `TRUE`. Inside `rowwise() + c_across()`, they operate across columns of the current row:

```r
# Does this subject have ANY flag set to "Y"?
rowwise() |>
  mutate(ANY_FLAG = any(c_across(ends_with("FL")) == "Y")) |>
  ungroup()

# Are ALL numeric lab values within range?
rowwise() |>
  mutate(ALL_NORMAL = all(c_across(starts_with("AVAL")) < 100, na.rm = TRUE)) |>
  ungroup()
```

**`pmax()` / `pmin()` — parallel (element-wise) max/min:** Operates across columns row-by-row without `rowwise()` overhead. Much faster for a fixed set of columns:

```r
mutate(MAX_BP = pmax(SYSBP, DIABP, na.rm = TRUE))   # larger of two columns per row
mutate(CAPPED = pmin(AVAL, ANRHI, na.rm = TRUE))    # cap AVAL at upper limit of normal
```

### Exercise 14a — row-wise check across multiple flag columns

In `adsl`, add `ANY_FLAG` = TRUE if **any** of `SAFFL`, `ITTFL`, `DTHFL` equals `"Y"` for that row. Use `rowwise()` + `c_across()` and `any()`.

```{r rowwise-any, exercise=TRUE, exercise.lines=10}
adsl |>
  rowwise() |>
  mutate(ANY_FLAG = any(c_across(c(SAFFL, ITTFL, DTHFL)) == "Y")) |>
  ungroup() |>
  select(USUBJID, SAFFL, ITTFL, DTHFL, ANY_FLAG)
```

### Exercise 14b — pmin for element-wise minimum (faster than rowwise)

Without `rowwise()`, use `pmin()` to add `MIN_MEASURE` = the element-wise minimum of `AVAL` and `ANRHI` per row of `adlb`.

```{r rowwise-pmin, exercise=TRUE, exercise.lines=7}
adlb |>
  mutate(MIN_MEASURE = pmin(___, ___, na.rm = TRUE)) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, ANRHI, MIN_MEASURE) |>
  head(6)
```

```{r rowwise-pmin-hint-1}
# pmin(AVAL, ANRHI, na.rm = TRUE)
```

```{r rowwise-pmin-solution}
adlb |>
  mutate(MIN_MEASURE = pmin(AVAL, ANRHI, na.rm = TRUE)) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, ANRHI, MIN_MEASURE) |>
  head(6)
```

---

## 15 — Capstone 1: AE incidence summary table

**The task** — Build a standard AE incidence summary by body system and preferred term, similar to what you'd include in a clinical study report appendix.

The output should have:
- Body system (`AEBODSYS`) and preferred term (`AEDECOD`)
- Number of subjects with ≥1 AE in that PT per treatment arm (`n_subjects`)
- Percentage of subjects in that arm (`pct`)
- Rows sorted by body system then descending subject count on the High Dose arm

This requires: `adsl` (for denominators) + `adae` + multiple `group_by` / `summarise` / `pivot_wider` steps.

```{r capstone-ae-incidence, exercise=TRUE, exercise.lines=55}
# Step 1: denominators per treatment arm (safety population)
denom <- adsl |>
  filter(SAFFL == "Y") |>
  count(TRT01A, name = "N_ARM")

# Step 2: one row per (subject × PT) — deduplicate multiple AEs with same PT per subject
ae_dedup <- adae |>
  distinct(USUBJID, TRT01A, AEBODSYS, AEDECOD)

# Step 3: count subjects per (PT × treatment arm)
ae_counts <- ae_dedup |>
  group_by(TRT01A, AEBODSYS, AEDECOD) |>
  summarise(n_subjects = n_distinct(USUBJID), .groups = "drop")

# Step 4: join denominators and compute percentage
ae_pct <- ae_counts |>
  left_join(denom, by = "TRT01A") |>
  mutate(pct = round(n_subjects / N_ARM * 100, 1))

# Step 5: pivot wide (one column per arm)
ae_wide <- ae_pct |>
  select(AEBODSYS, AEDECOD, TRT01A, pct) |>
  pivot_wider(
    names_from  = TRT01A,
    values_from = pct,
    values_fill = 0
  )

# Step 6: sort by body system, then by High Dose % descending
ae_wide |>
  arrange(AEBODSYS, desc(`Xanomeline High Dose`))
```

---

## 16 — Capstone 2: lab worsening detection pipeline

**The task** — Starting from `adlb`, build a complete lab-monitoring flags dataset that reports, per subject and parameter:

1. Baseline AVAL
2. Post-baseline maximum AVAL
3. Maximum toxicity grade post-baseline
4. Whether there was any grade worsening from baseline
5. Whether there were ≥ 2 consecutive visits with grade worsening
6. The visit at which the maximum grade was first seen

This is a multi-step derivation chain using `group_by`, `mutate`, `lag`, `cummax`, and `summarise`.

```{r capstone-lab-pipeline, exercise=TRUE, exercise.lines=70}
lab_flags <- adlb |>
  # Sort for lag/cumulative operations
  arrange(USUBJID, PARAMCD, AVISITN) |>

  # Step 1: baseline and change from baseline
  group_by(USUBJID, PARAMCD) |>
  mutate(
    BL_AVAL    = first(AVAL[AVISITN == 0]),
    BL_TOXGR   = first(ATOXGR[AVISITN == 0]),
    CHG        = AVAL - BL_AVAL,
    PCHG       = round((AVAL - BL_AVAL) / BL_AVAL * 100, 1)
  ) |>

  # Step 2: worst grade to date and grade worsening vs prior visit
  mutate(
    WORST_GR_TD  = cummax(ATOXGR),
    PREV_TOXGR   = lag(ATOXGR),
    WORSE_THIS   = !is.na(PREV_TOXGR) & ATOXGR > PREV_TOXGR,
    WORSE_PREV   = lag(WORSE_THIS),
    CONSEC_WORSE = WORSE_THIS & !is.na(WORSE_PREV) & WORSE_PREV
  ) |>
  ungroup() |>

  # Step 3: only post-baseline rows for summary
  filter(AVISITN > 0) |>

  # Step 4: summarise per subject × parameter
  group_by(USUBJID, PARAMCD) |>
  summarise(
    BL_AVAL        = first(BL_AVAL),
    BL_TOXGR       = first(BL_TOXGR),
    MAX_POSTBL     = max(AVAL, na.rm = TRUE),
    MAX_TOXGR      = max(ATOXGR, na.rm = TRUE),
    ANY_WORSENING  = any(WORSE_THIS, na.rm = TRUE),
    ANY_CONSEC     = any(CONSEC_WORSE, na.rm = TRUE),
    VISIT_MAX_GR   = AVISIT[which.max(ATOXGR)],
    .groups        = "drop"
  )

lab_flags
```

```{r capstone-lab-extend, exercise=TRUE, exercise.lines=16}
# Extension: from lab_flags, identify subjects with MAX_TOXGR >= 2
# and summarise: how many subjects, and which PARAMCDs are affected most?

lab_flags |>
  filter(MAX_TOXGR >= ___) |>
  group_by(___) |>
  summarise(
    n_subjects = n_distinct(USUBJID),
    .groups    = "drop"
  ) |>
  arrange(desc(n_subjects))
```

```{r capstone-lab-extend-hint-1}
# filter(MAX_TOXGR >= 2), group_by(PARAMCD)
```

```{r capstone-lab-extend-solution}
lab_flags |>
  filter(MAX_TOXGR >= 2) |>
  group_by(PARAMCD) |>
  summarise(
    n_subjects = n_distinct(USUBJID),
    .groups    = "drop"
  ) |>
  arrange(desc(n_subjects))
```

---

## 17 — Final Quiz

```{r final-quiz, echo=FALSE}
quiz(
  caption = "Lesson 04 — comprehensive self-check",

  question(
    "Translate SAS: `where saffl='Y' and age >= 65 and race not in ('UNKNOWN');`",
    answer('filter(SAFFL == "Y", AGE >= 65, !(RACE %in% c("UNKNOWN")))', correct = TRUE),
    answer('filter(SAFFL = "Y" & AGE >= 65 & RACE != "UNKNOWN")'),
    answer('where(SAFFL == "Y" & AGE >= 65)'),
    answer('filter(SAFFL == "Y" & AGE >= 65 & RACE %in% c("UNKNOWN"))'),
    allow_retry = TRUE
  ),

  question(
    "group_by() + summarise() vs group_by() + mutate(): which keeps all rows?",
    answer("summarise() keeps all rows"),
    answer("mutate() keeps all rows and adds a computed column per-group; summarise() collapses to one row per group", correct = TRUE),
    answer("Both collapse rows"),
    answer("Both keep all rows"),
    allow_retry = TRUE
  ),

  question(
    "Why is `lag()` inside `group_by()` safer than SAS's LAG function?",
    answer("R lag() is vectorised"),
    answer("SAS LAG ignores missing"),
    answer("dplyr lag() respects group boundaries — the value never bleeds from the last row of one subject to the first row of the next", correct = TRUE),
    answer("R lag() returns 0 for the first row"),
    allow_retry = TRUE
  ),

  question(
    "What does `add_count(USUBJID, name = 'N')` do?",
    answer("Returns only the count tibble"),
    answer("Removes duplicate USUBJID rows"),
    answer("Adds a column N with the total number of rows per USUBJID — equivalent to group_by(USUBJID) |> mutate(N = n()) |> ungroup()", correct = TRUE),
    answer("Counts unique subjects into a scalar"),
    allow_retry = TRUE
  ),

  question(
    "Which is the preferred way to select all flag columns (SAFFL, ITTFL, DTHFL, DIABFL) at once?",
    answer("select(contains('FLAG'))"),
    answer("select(ends_with('FL'))", correct = TRUE),
    answer("select(starts_with('FL'))"),
    answer("select(where(is.logical))"),
    allow_retry = TRUE
  ),

  question(
    "In `across(where(is.numeric), ~ mean(.x, na.rm=TRUE))`, what does `.x` represent?",
    answer("The row index"),
    answer("The column name as a string"),
    answer("The current column's values (a placeholder in the lambda formula)", correct = TRUE),
    answer("The group key"),
    allow_retry = TRUE
  ),

  question(
    "cumany(ATOXGR >= 3) returns TRUE for a row if:",
    answer("ATOXGR is 3 or higher on that exact row"),
    answer("ATOXGR was 3 or higher at any prior row in the group, OR on this row — it latches TRUE once the condition fires", correct = TRUE),
    answer("All rows in the group have ATOXGR >= 3"),
    answer("The cumulative sum of ATOXGR exceeds 3"),
    allow_retry = TRUE
  ),

  question(
    "What is the difference between `n()` and `n_distinct(USUBJID)` in summarise()?",
    answer("They are identical"),
    answer("n() counts total rows; n_distinct(USUBJID) counts unique subject IDs — useful when subjects can have multiple records", correct = TRUE),
    answer("n_distinct() is only available with group_by()"),
    answer("n() counts distinct values; n_distinct() counts all rows"),
    allow_retry = TRUE
  ),

  question(
    "You have a wide dataset with columns PARAM1, PARAM2, PARAM3. To get the row-wise maximum across these three, the most efficient approach is:",
    answer("rowwise() |> mutate(MAX = max(PARAM1, PARAM2, PARAM3))"),
    answer("pmax(PARAM1, PARAM2, PARAM3, na.rm=TRUE)", correct = TRUE),
    answer("across(c(PARAM1,PARAM2,PARAM3), max)"),
    answer("summarise(MAX = max(c(PARAM1, PARAM2, PARAM3)))"),
    allow_retry = TRUE
  ),

  question(
    "After `arrange(USUBJID, desc(AESEVN)) |> group_by(USUBJID) |> slice(1)`, each group contains:",
    answer("All rows"),
    answer("The row with the lowest AESEVN"),
    answer("The first row as ordered — i.e., the row with the highest AESEVN per subject", correct = TRUE),
    answer("The last row"),
    allow_retry = TRUE
  )
)
```

---

## 18 — SAS → dplyr Cheat Sheet (full)

| SAS pattern | dplyr equivalent |
|---|---|
| `where age > 50` | `filter(AGE > 50)` |
| `where var in ("A","B")` | `filter(var %in% c("A","B"))` |
| `where var not in ("A")` | `filter(!(var %in% c("A")))` |
| `where missing(var)` | `filter(is.na(var))` |
| `where not missing(var)` | `filter(!is.na(var))` |
| `where prxmatch("/HIGH/",trt)` | `filter(grepl("HIGH", TRT))` |
| `keep var1 var2` | `select(var1, var2)` |
| `drop var1` | `select(-var1)` |
| `keep vars starting with 'AE'` | `select(starts_with("AE"))` |
| `keep flag vars` | `select(ends_with("FL"))` |
| `rename old=new` | `rename(new = old)` |
| `proc sort by var` | `arrange(var)` |
| `proc sort by descending var` | `arrange(desc(var))` |
| Assignment in DATA step | `mutate(new = expression)` |
| Nested if/then/else | `case_when(...)` |
| Binary if/else | `if_else(cond, true_val, false_val)` |
| `proc means by group` | `group_by() |> summarise()` |
| RETAIN within BY group | `group_by() |> mutate()` |
| `if first.var` (subset) | `group_by() |> slice_min()` or `slice(1)` |
| `if last.var` (subset) | `group_by() |> slice_max()` or `slice(n())` |
| `first.var` flag column | `mutate(FIRST = row_number() == 1)` |
| `last.var` flag column | `mutate(LAST = row_number() == n())` |
| `aeseq + 1` within BY | `mutate(SEQ = row_number())` |
| LAG function | `lag(var)` — always inside `group_by()` |
| RETAIN cumulative | `cumsum()`, `cummax()`, `cummin()` |
| "has ever happened" | `cumany(condition)` |
| Apply to many vars | `across(c(v1,v2), ~ fn(.x,...))` |
| Apply to all numeric | `across(where(is.numeric), fn)` |
| Freq/cross-tab | `count(var1, var2)` |
| Deduplicate | `distinct(var1, var2)` |
| Row-wise max/min | `pmax()` / `pmin()` |
| Add count within group (keep rows) | `add_count(grp_var, name="N")` |
| `upcase()` | `toupper()` / `str_to_upper()` |
| `substr(x,1,3)` | `substr(x,1,3)` / `str_sub(x,1,3)` |
| `index(x,"Y")>0` | `grepl("Y",x)` / `str_detect(x,"Y")` |
| `tranwrd(x,old,new)` | `gsub(old,new,x)` / `str_replace_all()` |
| `cats(a,b)` | `paste0(a,b)` / `str_c(a,b)` |

---

## Key Takeaways

1. **Five core verbs** — `filter`, `select`, `mutate`, `summarise`, `arrange` — cover 90% of DATA-step work
2. **`case_when()`** is the multi-branch conditional; **`if_else()`** enforces type safety in binary conditions
3. **Always `na.rm = TRUE`** when summarizing — R propagates NA where SAS silently skips
4. **`group_by() + summarise()`** collapses rows; **`group_by() + mutate()`** keeps rows and computes within-group values
5. **`slice_min()` / `slice_max()` / `row_number()`** replace SAS FIRST./LAST. logic cleanly
6. **`lag()` / `lead()`** must be used inside `group_by()` to stay within subject boundaries
7. **`cummax()` / `cumany()`** handle running worst-to-date and "has this ever happened" patterns
8. **`across()`** applies operations to many columns at once — both in `mutate()` and `summarise()`
9. **`count()` / `distinct()` / `add_count()`** are shorthand for common tabulation and deduplication patterns
10. **String helpers** (`grepl`, `substr`, `gsub`, `sprintf`) cover all SAS character function needs
11. **`rowwise()` + `c_across()`** for per-row operations; `pmax()` / `pmin()` for simple row-wise extremes

**Next** → Lesson 05: Joins, SQL patterns, and writing your own functions
