---
title: "Lesson 05 — R Primer for SAS Programmers, Part 3: Joins, SQL, and Functions"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
    ace_theme: tomorrow
runtime: shiny_prerendered
description: >
  Translate PROC SQL and DATA step MERGE into dplyr joins. Diagnose fanout and key
  mismatches. Write reusable R functions using default args, curly-curly, and dot-dot-dot.
  Replace SAS macro loops with purrr::map().
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 05_joins_sql_functions.Rmd
# PURPOSE       : Relational joins (inner, left, anti, cross) and SQL-style window
#                 functions in R: row_number(), rank(), lag/lead,
#                 cumulative aggregates — mapped from SAS PROC SQL and
#                 BY-group processing.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tibble)
library(purrr)
library(tidyr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 60)

# ── Shared datasets ──────────────────────────────────────────────────────────

adsl <- tribble(
  ~USUBJID,       ~AGE, ~SEX, ~RACE,      ~SAFFL, ~EFFFL, ~TRT01A,
  "CDIS-001-001",   63,  "F", "WHITE",    "Y",    "Y",    "Xanomeline High Dose",
  "CDIS-001-002",   52,  "M", "WHITE",    "Y",    "Y",    "Placebo",
  "CDIS-001-003",   70,  "F", "BLACK",    "Y",    "N",    "Xanomeline Low Dose",
  "CDIS-001-004",   45,  "M", "ASIAN",    "Y",    "Y",    "Xanomeline High Dose",
  "CDIS-001-005",   67,  "F", "WHITE",    "Y",    "Y",    "Placebo",
  "CDIS-001-006",   38,  "M", "HISPANIC", "N",    "N",    "Xanomeline Low Dose",
  "CDIS-001-007",   74,  "F", "WHITE",    "Y",    "Y",    "Xanomeline High Dose",
  "CDIS-001-008",   58,  "M", "WHITE",    "Y",    "Y",    "Placebo"
)

adae <- tribble(
  ~USUBJID,       ~AEDECOD,                   ~AESEV,     ~AESEVN, ~AESER, ~AESTDTC,
  "CDIS-001-001", "HEADACHE",                 "MILD",     1,       "N",    "2014-01-05",
  "CDIS-001-001", "NAUSEA",                   "MODERATE", 2,       "N",    "2014-02-10",
  "CDIS-001-003", "DIZZINESS",                "MILD",     1,       "N",    "2014-01-20",
  "CDIS-001-004", "VOMITING",                 "SEVERE",   3,       "Y",    "2014-03-01",
  "CDIS-001-004", "HEADACHE",                 "MILD",     1,       "N",    "2014-01-15",
  "CDIS-001-005", "FATIGUE",                  "MILD",     1,       "N",    "2014-02-22",
  "CDIS-001-007", "APPLICATION SITE PRURITUS","MODERATE", 2,       "N",    "2014-01-30",
  "CDIS-001-009", "NAUSEA",                   "MODERATE", 2,       "N",    "2014-02-08"  # orphan!
)

adlb <- tribble(
  ~USUBJID,       ~PARAMCD, ~AVISIT,    ~AVISITN, ~AVAL, ~ATOXGR,
  "CDIS-001-001", "HGB",    "BASELINE",  0,        13.2,  0,
  "CDIS-001-001", "HGB",    "WEEK 4",    4,        12.1,  1,
  "CDIS-001-001", "ALT",    "BASELINE",  0,        22.0,  0,
  "CDIS-001-001", "ALT",    "WEEK 4",    4,        28.0,  1,
  "CDIS-001-002", "HGB",    "BASELINE",  0,        14.5,  0,
  "CDIS-001-002", "HGB",    "WEEK 4",    4,        13.9,  0,
  "CDIS-001-002", "ALT",    "BASELINE",  0,        35.0,  0,
  "CDIS-001-002", "ALT",    "WEEK 4",    4,        42.0,  1,
  "CDIS-001-003", "HGB",    "BASELINE",  0,        11.8,  1,
  "CDIS-001-003", "HGB",    "WEEK 4",    4,        11.2,  1,
  "CDIS-001-004", "ALT",    "BASELINE",  0,        18.0,  0,
  "CDIS-001-004", "ALT",    "WEEK 4",    4,        20.0,  0
)

# Dataset with a different key column name (for join_by exercise)
ae_renamed <- adae |> rename(SUBJECT_ID = USUBJID)

# Dataset with a whitespace key mismatch (for diagnostics exercise)
adsl_dirty <- adsl |>
  mutate(USUBJID = if_else(USUBJID == "CDIS-001-002",
                           paste0(USUBJID, " "),   # trailing space
                           USUBJID))
```

## Overview

### Joins and functions — the backbone of clinical R

Almost every ADaM dataset is built by joining SDTM domains — AE ← DM, LB ← DM, EX ← DM. Almost every table starts by joining an analysis dataset to ADSL for population flags. And every reusable derivation pattern gets wrapped in a function.

This lesson covers:

- All **six dplyr join verbs** and their SAS equivalents
- Diagnosing the two most common join bugs: **many-to-many fanout** and **key mismatches**
- A brief look at **dbplyr** for database-backed dplyr queries
- Writing **R functions** with default arguments, `{{ }}`, and `...`
- Why **R functions are strictly better than SAS macros**
- **`purrr::map()`** as a replacement for SAS macro loops

> All datasets (`adsl`, `adae`, `adlb`) are pre-loaded in this tutorial.
> Note: `adae` intentionally contains `CDIS-001-009` — a subject **not** in `adsl` —
> to let you practice orphan detection.

---

```{r quiz-overview-05, echo=FALSE}
quiz(
  caption = "Before we begin",
  question("Which dplyr verb is most equivalent to a SAS PROC SQL LEFT JOIN?",
    answer("inner_join()"),
    answer("left_join()", correct = TRUE),
    answer("semi_join()"),
    answer("full_join()"),
    allow_retry = TRUE
  ),
  question("R functions differ from SAS macros most fundamentally because:",
    answer("R functions run faster"),
    answer("R functions are first-class values — they can be passed, returned, and stored like any object", correct = TRUE),
    answer("SAS macros don't support default arguments"),
    answer("R functions don't need to be declared before use"),
    allow_retry = TRUE
  )
)
```

---

## 1  left_join() — keep everything from the left

### Concept

`left_join` is what you reach for ~80% of the time. Every row of the left dataset is preserved; unmatched rows get `NA` for the right-side columns.

```r
adae |>
  left_join(
    adsl |> select(USUBJID, TRT01A, SAFFL),
    by = "USUBJID"
  )
```

SAS equivalent:
```sas
proc sql;
  create table adae_with_trt as
  select a.*, b.trt01a, b.saffl
  from adae a
  left join (select usubjid, trt01a, saffl from adsl) b
  on a.usubjid = b.usubjid;
quit;
```

### Exercise 1a — attach treatment to AE records

Join `TRT01A` and `SAFFL` from `adsl` onto `adae`. Select `USUBJID`, `AEDECOD`, `AESEV`, `TRT01A`, and `SAFFL` in the output.

```{r left-join-basic, exercise=TRUE, exercise.lines=10}
adae |>
  left_join(
    adsl |> select(___, ___, ___),
    by = "___"
  ) |>
  select(USUBJID, AEDECOD, AESEV, TRT01A, SAFFL)
```

```{r left-join-basic-hint-1}
# adsl |> select(USUBJID, TRT01A, SAFFL)
# by = "USUBJID"
```

```{r left-join-basic-solution}
adae |>
  left_join(
    adsl |> select(USUBJID, TRT01A, SAFFL),
    by = "USUBJID"
  ) |>
  select(USUBJID, AEDECOD, AESEV, TRT01A, SAFFL)
```

### Exercise 1b — spot the orphan

Notice that `CDIS-001-009` appears in `adae` but **not** in `adsl`. After a `left_join`, what value will `TRT01A` have for that subject? Run the code and check.

```{r left-join-orphan, exercise=TRUE, exercise.lines=8}
adae |>
  left_join(adsl |> select(USUBJID, TRT01A), by = "USUBJID") |>
  filter(is.na(TRT01A)) |>
  select(USUBJID, AEDECOD, TRT01A)
```

```{r left-join-orphan-hint-1}
# Just run it — the NA row exposes the orphan subject
```

---

## 2  inner_join() — keep only matches

### Concept

`inner_join` drops rows that don't match in **either** dataset. It's useful when you genuinely want to restrict to subjects present in both tables — but `semi_join` (next section) is often cleaner for population filtering.

```r
adae |>
  inner_join(
    adsl |> filter(SAFFL == "Y") |> select(USUBJID, TRT01A),
    by = "USUBJID"
  )
```

### Exercise 2 — restrict to safety population

Use `inner_join` to keep only AE records for subjects where `SAFFL == "Y"`. How many rows are dropped compared to the original `adae`?

```{r inner-join-ex, exercise=TRUE, exercise.lines=12}
# Original row count
nrow(adae)

# After inner_join to safety population
adae_safety <- adae |>
  inner_join(
    adsl |> filter(SAFFL == "___") |> select(USUBJID, TRT01A),
    by = "USUBJID"
  )

nrow(adae_safety)
adae_safety
```

```{r inner-join-ex-hint-1}
# filter(SAFFL == "Y")
```

```{r inner-join-ex-solution}
nrow(adae)

adae_safety <- adae |>
  inner_join(
    adsl |> filter(SAFFL == "Y") |> select(USUBJID, TRT01A),
    by = "USUBJID"
  )

nrow(adae_safety)
adae_safety
```

---

## 3  semi_join() and anti_join() — filtering joins

### Concept

These two are **filtering joins** — they never add columns from the right side.

| Verb | Keeps rows from left where... |
|---|---|
| `semi_join(x, y)` | Key **exists** in y |
| `anti_join(x, y)` | Key does **not** exist in y |

`semi_join` is often cleaner than `inner_join` for population filtering because it doesn't risk adding duplicate columns or fanout from the right side.

`anti_join` is your **QC tool** — use it to find orphans, unmatched records, and data gaps.

### Exercise 3a — semi_join for population filter

Use `semi_join` to keep only AE records for subjects in the safety population. Compare the result to Exercise 2 — same rows, but notice **no `TRT01A` column is added**.

```{r semi-join-ex, exercise=TRUE, exercise.lines=8}
adae |>
  semi_join(
    adsl |> filter(SAFFL == "Y"),
    by = "___"
  )
```

```{r semi-join-ex-hint-1}
# by = "USUBJID"
# Note: no TRT01A column is added — semi_join is a filter only
```

```{r semi-join-ex-solution}
adae |>
  semi_join(
    adsl |> filter(SAFFL == "Y"),
    by = "USUBJID"
  )
```

### Exercise 3b — anti_join for orphan detection

Use `anti_join` to find AE records whose `USUBJID` does **not** appear in `adsl`. This is how you QC for orphan records.

```{r anti-join-ex, exercise=TRUE, exercise.lines=7}
adae |>
  anti_join(adsl, by = "___") |>
  select(USUBJID, AEDECOD, AESEV)
```

```{r anti-join-ex-hint-1}
# by = "USUBJID"
# This should return the CDIS-001-009 row
```

```{r anti-join-ex-solution}
adae |>
  anti_join(adsl, by = "USUBJID") |>
  select(USUBJID, AEDECOD, AESEV)
```

### Exercise 3c — anti_join in reverse: subjects with no AEs

Use `anti_join` in the opposite direction: starting from `adsl`, find subjects in the **safety population** who have **no AE records** at all.

```{r anti-join-reverse, exercise=TRUE, exercise.lines=8}
adsl |>
  filter(SAFFL == "Y") |>
  anti_join(___, by = "USUBJID") |>
  select(USUBJID, TRT01A, AGE)
```

```{r anti-join-reverse-hint-1}
# anti_join(adae, by = "USUBJID")
# Start from adsl (safety pop), find who is NOT in adae
```

```{r anti-join-reverse-solution}
adsl |>
  filter(SAFFL == "Y") |>
  anti_join(adae, by = "USUBJID") |>
  select(USUBJID, TRT01A, AGE)
```

```{r quiz-joins, echo=FALSE}
quiz(
  caption = "Join verbs check",
  question(
    "You want to keep all AE rows for safety-pop subjects but DON'T want to add any ADSL columns. Which verb is cleanest?",
    answer("left_join()"),
    answer("inner_join()"),
    answer("semi_join()", correct = TRUE),
    answer("full_join()"),
    allow_retry = TRUE
  ),
  question(
    "What does anti_join(adae, adsl, by = 'USUBJID') return?",
    answer("All subjects in adsl not in adae"),
    answer("All AE rows for subjects who also appear in adsl"),
    answer("All AE rows for subjects who do NOT appear in adsl", correct = TRUE),
    answer("All rows from both datasets that don't match"),
    allow_retry = TRUE
  )
)
```

---

## 4  full_join() — keep everything from both sides

### Concept

`full_join` keeps every row from both datasets. Rows that don't match on either side get `NA` for the other side's columns. It's most useful for **reconciliation tasks** — finding what's in one dataset but missing from the other.

```r
adsl |>
  full_join(adae, by = "USUBJID")
```

### Exercise 4 — reconciliation

Use `full_join` to combine `adsl` and a small subset of `adae`. Then filter for rows where `AGE` is `NA` (came from adae but not adsl) or `AEDECOD` is `NA` (came from adsl but not adae). What does this reveal?

```{r full-join-ex, exercise=TRUE, exercise.lines=14}
adsl_mini <- adsl |> select(USUBJID, AGE, TRT01A)
adae_mini <- adae |> select(USUBJID, AEDECOD, AESEV)

reconciled <- adsl_mini |>
  full_join(adae_mini, by = "USUBJID")

# Rows in adae with no matching adsl row (AGE will be NA)
reconciled |> filter(is.na(AGE))
```

```{r full-join-ex-hint-1}
# Rows where AGE is NA came from adae only — no matching USUBJID in adsl
# Rows where AEDECOD is NA came from adsl only — subjects with no AEs
```

---

## 5  Joining on differently-named columns

### Concept

When key columns have different names in each dataset, use the named vector `by = c("left_col" = "right_col")` or the modern `join_by()` syntax:

```r
# Named vector style (older but widely used)
left_join(adsl, ae_renamed, by = c("USUBJID" = "SUBJECT_ID"))

# join_by() style (dplyr 1.1+, preferred)
left_join(adsl, ae_renamed, by = join_by(USUBJID == SUBJECT_ID))
```

`join_by()` also unlocks **inequality joins** — for example, joining AE records to a dosing window:
```r
adae |>
  left_join(adcm, by = join_by(USUBJID, ASTDT >= CMSTDT, ASTDT <= CMENDT))
```

### Exercise 5 — join on differently-named key

The dataset `ae_renamed` is a copy of `adae` where `USUBJID` has been renamed to `SUBJECT_ID`. Join it to `adsl` using the correct `by` syntax. Select `USUBJID`, `AEDECOD`, and `TRT01A`.

```{r diff-key-join, exercise=TRUE, exercise.lines=10}
# ae_renamed has SUBJECT_ID instead of USUBJID
# peek at the column names:
names(ae_renamed)

adsl |>
  left_join(ae_renamed, by = join_by(___ == ___)) |>
  select(USUBJID, AEDECOD, TRT01A) |>
  filter(!is.na(AEDECOD))
```

```{r diff-key-join-hint-1}
# join_by(USUBJID == SUBJECT_ID)
# Left side = adsl column name, right side = ae_renamed column name
```

```{r diff-key-join-solution}
names(ae_renamed)

adsl |>
  left_join(ae_renamed, by = join_by(USUBJID == SUBJECT_ID)) |>
  select(USUBJID, AEDECOD, TRT01A) |>
  filter(!is.na(AEDECOD))
```

---

## 6  The most dangerous join bug: many-to-many fanout

### Concept

Fanout occurs when you join two datasets that each have **multiple rows per key**. The result is the Cartesian product per key — rows explode silently.

```
df_a:   3 rows for subject X  (visit-level)
df_b:   5 rows for subject X  (lab-level)
join result:  15 rows for subject X  ← BUG!
```

dplyr 1.1+ warns you with:
```
Warning: Detected an unexpected many-to-many relationship between `x` and `y`.
```

Assert your expectations explicitly using the `relationship` argument:

```r
adae |>
  left_join(adsl, by = "USUBJID", relationship = "many-to-one")
# Errors if any USUBJID appears more than once in adsl
```

### Exercise 6a — witness the fanout

Join `adlb` (multiple rows per subject) to a version of `adae` (also multiple rows per subject) **without a visit key**. Check how many rows you get vs. expected.

```{r fanout-ex, exercise=TRUE, exercise.lines=12}
# Original row counts
nrow(adlb)    # multiple rows per subject
nrow(adae)    # multiple rows per subject

# Join only on USUBJID — watch the row explosion
fanout_result <- adlb |>
  left_join(adae |> select(USUBJID, AEDECOD), by = "USUBJID")

nrow(fanout_result)

# Compare: how many rows did you expect vs. what you got?
fanout_result |> filter(USUBJID == "CDIS-001-001")
```

```{r fanout-ex-hint-1}
# For CDIS-001-001: adlb has 4 rows, adae has 2 rows
# The join produces 4 × 2 = 8 rows for that subject — fanout!
```

### Exercise 6b — assert the relationship

Now join `adae` to `adsl` and explicitly tell dplyr you expect a many-to-one relationship (many AE rows per subject, one ADSL row per subject). If `adsl` had duplicate USUBJIDs, this would raise an error.

```{r relationship-ex, exercise=TRUE, exercise.lines=8}
adae |>
  left_join(
    adsl |> select(USUBJID, TRT01A),
    by = "USUBJID",
    relationship = "___"
  ) |>
  select(USUBJID, AEDECOD, TRT01A)
```

```{r relationship-ex-hint-1}
# relationship = "many-to-one"
# adae has many rows per USUBJID; adsl has exactly one
```

```{r relationship-ex-solution}
adae |>
  left_join(
    adsl |> select(USUBJID, TRT01A),
    by = "USUBJID",
    relationship = "many-to-one"
  ) |>
  select(USUBJID, AEDECOD, TRT01A)
```

---

## 7  Diagnosing key mismatches

### Concept

If a `left_join` produces unexpected `NA` columns, the keys aren't matching. The four most common causes are whitespace, type mismatch, case, and leading-zero differences.

Quick diagnostic toolkit:
```r
# Which keys are in adae but missing from adsl?
adae |> anti_join(adsl, by = "USUBJID") |> distinct(USUBJID)

# Check types
class(adae$USUBJID); class(adsl$USUBJID)

# Whitespace check
adae$USUBJID |> trimws() |> unique()
```

### Exercise 7 — find the whitespace mismatch

`adsl_dirty` is a version of `adsl` where `CDIS-001-002` has a trailing space added to its `USUBJID`. Use `anti_join` to expose the mismatch, then fix it with `trimws()` and confirm the join works.

```{r key-mismatch-ex, exercise=TRUE, exercise.lines=20}
# Step 1: how many adae rows fail to match adsl_dirty?
adae |>
  anti_join(adsl_dirty, by = "USUBJID") |>
  select(USUBJID, AEDECOD)

# Step 2: fix the dirty key and re-check
adsl_clean <- adsl_dirty |>
  mutate(USUBJID = trimws(___))

adae |>
  anti_join(adsl_clean, by = "USUBJID") |>
  select(USUBJID, AEDECOD)
# Now only the genuine orphan (CDIS-001-009) remains
```

```{r key-mismatch-ex-hint-1}
# trimws(USUBJID) strips leading and trailing whitespace
# After cleaning, anti_join should only show CDIS-001-009
```

```{r key-mismatch-ex-solution}
adae |>
  anti_join(adsl_dirty, by = "USUBJID") |>
  select(USUBJID, AEDECOD)

adsl_clean <- adsl_dirty |>
  mutate(USUBJID = trimws(USUBJID))

adae |>
  anti_join(adsl_clean, by = "USUBJID") |>
  select(USUBJID, AEDECOD)
```

---

## 8  dbplyr — dplyr against a database

### Concept (no exercise — database not available here)

`{dbplyr}` lets you write **the same dplyr code** against a database table. It translates your dplyr verbs to SQL behind the scenes and runs the computation on the server. You only pull data into R when you call `collect()`.

```r
library(DBI); library(dbplyr)

con     <- dbConnect(odbc::odbc(), "ClinicalDB")
ae_db   <- tbl(con, "AE_SDTM")          # reference, no data pulled yet

result <- ae_db |>
  filter(AESER == "Y") |>
  group_by(USUBJID) |>
  summarise(n_serious_ae = n()) |>
  collect()                              # now data enters R
```

The key insight: **you can prototype dplyr against a small in-memory tibble**, then point the same code at a 100-million-row database table — the heavy lifting stays on the server.

```{r quiz-dbplyr, echo=FALSE}
quiz(
  caption = "dbplyr check",
  question(
    "In a dbplyr workflow, when does data actually transfer from the database into R?",
    answer("When you write the tbl() reference"),
    answer("When you call filter() or group_by()"),
    answer("When you call collect()", correct = TRUE),
    answer("When you assign the result to a variable"),
    allow_retry = TRUE
  ),
  question(
    "What is the main advantage of dbplyr over writing raw SQL for database queries?",
    answer("dbplyr queries are always faster than SQL"),
    answer("You can write the same dplyr code you use for in-memory tibbles — no SQL needed", correct = TRUE),
    answer("dbplyr doesn't require a database connection"),
    answer("dbplyr automatically creates indexes"),
    allow_retry = TRUE
  )
)
```

---

## 9  Writing your own R functions

### Concept

R's `function` keyword is the analog of SAS's `%macro`. Key differences:

- No `length`/`format` declarations needed — types are dynamic
- The **last evaluated expression** is the return value (`return()` is optional)
- Functions are **first-class objects** — pass them as arguments, store in lists, return from other functions
- Arguments are real R objects, not text substitutions

```r
# SAS:
# %macro add(x, y); %let result = %eval(&x + &y); &result %mend;

# R:
add <- function(x, y) {
  x + y           # last expression is automatically returned
}
add(3, 4)         # 7
```

### Exercise 9a — write your first function

Write a function `flag_elderly` that takes a data frame and a minimum age (default `65`), and adds a column `ELDERLY` that is `"Y"` when `AGE >= min_age`, else `"N"`. Test it on `adsl`.

```{r func-basic, exercise=TRUE, exercise.lines=15}
flag_elderly <- function(data, min_age = ___) {
  data |>
    mutate(
      ELDERLY = if_else(AGE >= ___, "Y", "N")
    )
}

# Test with default (65)
flag_elderly(adsl) |> select(USUBJID, AGE, ELDERLY)
```

```{r func-basic-hint-1}
# function(data, min_age = 65)
# if_else(AGE >= min_age, "Y", "N")
```

```{r func-basic-solution}
flag_elderly <- function(data, min_age = 65) {
  data |>
    mutate(
      ELDERLY = if_else(AGE >= min_age, "Y", "N")
    )
}

flag_elderly(adsl) |> select(USUBJID, AGE, ELDERLY)
```

### Exercise 9b — override the default

Call `flag_elderly()` with `min_age = 70` instead of the default. Which subjects are now flagged?

```{r func-default-override, exercise=TRUE, exercise.lines=8}
flag_elderly <- function(data, min_age = 65) {
  data |>
    mutate(ELDERLY = if_else(AGE >= min_age, "Y", "N"))
}

flag_elderly(adsl, min_age = ___) |>
  select(USUBJID, AGE, ELDERLY)
```

```{r func-default-override-solution}
flag_elderly <- function(data, min_age = 65) {
  data |>
    mutate(ELDERLY = if_else(AGE >= min_age, "Y", "N"))
}

flag_elderly(adsl, min_age = 70) |>
  select(USUBJID, AGE, ELDERLY)
```

---

## 10  Functions with {{ }} — dplyr-aware arguments

### Concept

When you want to pass a **column name** (without quotes) into a dplyr-aware function, use `{{ }}` ("curly-curly"). It tells dplyr to evaluate the argument as a column name rather than a regular R variable.

```r
summarise_by <- function(data, group_col, ...) {
  data |>
    group_by({{ group_col }}) |>
    summarise(n = n(), ...)
}

summarise_by(adsl, TRT01A, mean_age = mean(AGE, na.rm = TRUE))
```

For passing column names as **strings**, use `.data[[col]]`:
```r
summarise_by_str <- function(data, group_col = "TRT01A") {
  data |>
    group_by(.data[[group_col]]) |>
    summarise(n = n())
}
```

### Exercise 10a — write a {{ }} function

Write a function `group_count` that takes a data frame and a grouping column (without quotes), and returns a count per group. Test it with `TRT01A` and then with `SEX`.

```{r curlycurly-ex, exercise=TRUE, exercise.lines=15}
group_count <- function(data, group_col) {
  data |>
    group_by({{ ___ }}) |>
    summarise(n = n(), .groups = "drop")
}

# By treatment
group_count(adsl, TRT01A)

# By sex
group_count(adsl, ___)
```

```{r curlycurly-ex-hint-1}
# group_by({{ group_col }})
# group_count(adsl, SEX)
```

```{r curlycurly-ex-solution}
group_count <- function(data, group_col) {
  data |>
    group_by({{ group_col }}) |>
    summarise(n = n(), .groups = "drop")
}

group_count(adsl, TRT01A)
group_count(adsl, SEX)
```

### Exercise 10b — dot-dot-dot (...) for extra stats

Extend `group_count` to accept `...` and pass those extra expressions to `summarise()`, so callers can add arbitrary summary statistics.

```{r dots-ex, exercise=TRUE, exercise.lines=15}
group_summary <- function(data, group_col, ...) {
  data |>
    group_by({{ group_col }}) |>
    summarise(n = n(), ..., .groups = "drop")
}

# Call with extra stats
group_summary(
  adsl,
  TRT01A,
  mean_age    = mean(AGE, na.rm = TRUE),
  median_age  = median(AGE, na.rm = TRUE)
)
```

```{r dots-ex-hint-1}
# The ... in the function signature captures extra named expressions
# They are forwarded to summarise() via ... in the function body
```

```{r dots-ex-solution}
group_summary <- function(data, group_col, ...) {
  data |>
    group_by({{ group_col }}) |>
    summarise(n = n(), ..., .groups = "drop")
}

group_summary(
  adsl,
  TRT01A,
  mean_age   = mean(AGE, na.rm = TRUE),
  median_age = median(AGE, na.rm = TRUE)
)
```

---

## 11  Anonymous functions — \(x) syntax

### Concept

Anonymous functions are functions defined inline, without a name. R 4.1+ introduced the clean `\(x)` shorthand:

```r
# All equivalent:
function(x) mean(x, na.rm = TRUE)    # long form
\(x) mean(x, na.rm = TRUE)           # short form (R 4.1+)
~ mean(.x, na.rm = TRUE)             # purrr/tidyverse formula style
```

The `\(x)` form is the modern preference — pure base R, reads clearly.

```r
# Use inside across()
adsl |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)))
```

### Exercise 11a — anonymous function with across

Use `across()` with an anonymous function `\(x)` to compute the **median** of every numeric column in `adsl`, grouped by `TRT01A`.

```{r anon-func-ex, exercise=TRUE, exercise.lines=8}
adsl |>
  group_by(TRT01A) |>
  summarise(
    across(where(is.numeric), \(x) ___),
    .groups = "drop"
  )
```

```{r anon-func-ex-hint-1}
# \(x) median(x, na.rm = TRUE)
```

```{r anon-func-ex-solution}
adsl |>
  group_by(TRT01A) |>
  summarise(
    across(where(is.numeric), \(x) median(x, na.rm = TRUE)),
    .groups = "drop"
  )
```

### Exercise 11b — tilde formula style

Repeat the same using the `~` formula syntax (common in older pharmaverse code). Notice `~` uses `.x` instead of `x`.

```{r tilde-func-ex, exercise=TRUE, exercise.lines=8}
adsl |>
  group_by(TRT01A) |>
  summarise(
    across(where(is.numeric), ~ median(___, na.rm = TRUE)),
    .groups = "drop"
  )
```

```{r tilde-func-ex-hint-1}
# ~ median(.x, na.rm = TRUE)
```

```{r tilde-func-ex-solution}
adsl |>
  group_by(TRT01A) |>
  summarise(
    across(where(is.numeric), ~ median(.x, na.rm = TRUE)),
    .groups = "drop"
  )
```

---

## 12  purrr::map() — iteration without macro loops

### Concept

SAS uses macro `%do` loops to repeat an operation. In R, you **map** a function over a vector instead:

```r
# SAS macro loop equivalent
# %do i = 1 %to 4; ... AVISITN = &i. ... %end;

# R with map_dfr (returns a single data frame, rows bound)
library(purrr)

visits <- c(0, 4)
results <- map_dfr(visits, \(v) {
  adlb |>
    filter(AVISITN == v) |>
    summarise(visit = v, mean_aval = mean(AVAL, na.rm = TRUE))
})
```

`map()` variants:

| Function | Returns |
|---|---|
| `map()` | A list |
| `map_dbl()` | A numeric vector |
| `map_chr()` | A character vector |
| `map_lgl()` | A logical vector |
| `map_dfr()` | A data frame (rows bound) |

### Exercise 12a — map over visit numbers

Use `map_dfr()` to compute the mean `AVAL` per parameter (`PARAMCD`) for each visit number in `adlb`. The visit numbers are `c(0, 4)`.

```{r map-ex, exercise=TRUE, exercise.lines=12}
library(purrr)

visits <- c(0, 4)

map_dfr(visits, \(v) {
  adlb |>
    filter(AVISITN == ___) |>
    group_by(PARAMCD) |>
    summarise(
      visit    = v,
      mean_val = mean(AVAL, na.rm = TRUE),
      .groups  = "drop"
    )
})
```

```{r map-ex-hint-1}
# filter(AVISITN == v) — v comes from the map iteration
```

```{r map-ex-solution}
library(purrr)

visits <- c(0, 4)

map_dfr(visits, \(v) {
  adlb |>
    filter(AVISITN == v) |>
    group_by(PARAMCD) |>
    summarise(
      visit    = v,
      mean_val = mean(AVAL, na.rm = TRUE),
      .groups  = "drop"
    )
})
```

### Exercise 12b — map over treatment arms

Use `map_dfr()` to compute the number of subjects and AEs **per treatment arm**. Iterate over the unique values of `TRT01A`.

```{r map-trt, exercise=TRUE, exercise.lines=15}
library(purrr)

trt_arms <- unique(adsl$TRT01A)

map_dfr(trt_arms, \(arm) {
  n_subj <- adsl |>
    filter(TRT01A == ___) |>
    nrow()

  n_ae <- adae |>
    left_join(adsl |> select(USUBJID, TRT01A), by = "USUBJID") |>
    filter(TRT01A == ___) |>
    nrow()

  tibble(treatment = arm, n_subjects = n_subj, n_aes = n_ae)
})
```

```{r map-trt-hint-1}
# filter(TRT01A == arm) — arm is the loop variable
```

```{r map-trt-solution}
library(purrr)

trt_arms <- unique(adsl$TRT01A)

map_dfr(trt_arms, \(arm) {
  n_subj <- adsl |>
    filter(TRT01A == arm) |>
    nrow()

  n_ae <- adae |>
    left_join(adsl |> select(USUBJID, TRT01A), by = "USUBJID") |>
    filter(TRT01A == arm) |>
    nrow()

  tibble(treatment = arm, n_subjects = n_subj, n_aes = n_ae)
})
```

---

## 13  Putting it all together — a derivation function

### The task

Build a function `derive_subject_summary()` that takes `adsl`, `adae`, and `adlb`, and returns one row per subject in the safety population with:

- `USUBJID`, `TRT01A`, `AGE`, `SEX`
- `n_ae` — total number of AEs per subject (0 if none)
- `n_ser_ae` — number of serious AEs (`AESER == "Y"`)
- `hgb_min` — minimum haemoglobin (`PARAMCD == "HGB"`)

### Exercise 13 — build the function

```{r derive-func-ex, exercise=TRUE, exercise.lines=40}
derive_subject_summary <- function(adsl, adae, adlb) {

  # Step 1: count AEs per subject
  ae_counts <- adae |>
    group_by(___) |>
    summarise(
      n_ae     = n(),
      n_ser_ae = sum(AESER == "Y", na.rm = TRUE),
      .groups  = "drop"
    )

  # Step 2: minimum HGB per subject
  hgb_worst <- adlb |>
    filter(PARAMCD == "___") |>
    group_by(USUBJID) |>
    summarise(
      hgb_min = min(AVAL, na.rm = TRUE),
      .groups = "drop"
    )

  # Step 3: join to adsl (safety pop) and fill NA AE counts with 0
  adsl |>
    filter(SAFFL == "Y") |>
    left_join(ae_counts, by = "USUBJID") |>
    left_join(hgb_worst, by = "USUBJID") |>
    mutate(
      n_ae     = replace_na(n_ae,     0),
      n_ser_ae = replace_na(n_ser_ae, 0)
    ) |>
    select(USUBJID, TRT01A, AGE, SEX, n_ae, n_ser_ae, hgb_min)
}

# Run it
derive_subject_summary(adsl, adae, adlb)
```

```{r derive-func-ex-hint-1}
# Step 1: group_by(USUBJID)
# Step 2: filter(PARAMCD == "HGB")
# Steps 3: left_join twice, then replace_na for missing counts
```

```{r derive-func-ex-solution}
derive_subject_summary <- function(adsl, adae, adlb) {

  ae_counts <- adae |>
    group_by(USUBJID) |>
    summarise(
      n_ae     = n(),
      n_ser_ae = sum(AESER == "Y", na.rm = TRUE),
      .groups  = "drop"
    )

  hgb_worst <- adlb |>
    filter(PARAMCD == "HGB") |>
    group_by(USUBJID) |>
    summarise(
      hgb_min = min(AVAL, na.rm = TRUE),
      .groups = "drop"
    )

  adsl |>
    filter(SAFFL == "Y") |>
    left_join(ae_counts, by = "USUBJID") |>
    left_join(hgb_worst, by = "USUBJID") |>
    mutate(
      n_ae     = replace_na(n_ae,     0),
      n_ser_ae = replace_na(n_ser_ae, 0)
    ) |>
    select(USUBJID, TRT01A, AGE, SEX, n_ae, n_ser_ae, hgb_min)
}

derive_subject_summary(adsl, adae, adlb)
```

---

## 14  Self-check Quiz

```{r final-quiz-05, echo=FALSE}
quiz(
  caption = "Lesson 05 — self-check",

  question(
    "Translate: `select a.*, b.trt01a from adae a left join adsl b on a.usubjid = b.usubjid where b.saffl = 'Y'`",
    answer("adae |> inner_join(adsl |> filter(SAFFL=='Y') |> select(USUBJID, TRT01A), by='USUBJID')"),
    answer("adae |> left_join(adsl |> filter(SAFFL=='Y') |> select(USUBJID, TRT01A), by='USUBJID')", correct = TRUE),
    answer("adsl |> filter(SAFFL=='Y') |> left_join(adae, by='USUBJID')"),
    answer("adae |> semi_join(adsl |> filter(SAFFL=='Y'), by='USUBJID')"),
    allow_retry = TRUE
  ),

  question(
    "In dplyr, how do you find subjects in adsl (safety pop) who have NO AE records?",
    answer("adsl |> filter(SAFFL=='Y') |> inner_join(adae, by='USUBJID')"),
    answer("adsl |> filter(SAFFL=='Y') |> anti_join(adae, by='USUBJID')", correct = TRUE),
    answer("adae |> anti_join(adsl, by='USUBJID')"),
    answer("adsl |> filter(SAFFL=='Y') |> full_join(adae, by='USUBJID') |> filter(is.na(AEDECOD))"),
    allow_retry = TRUE
  ),

  question(
    "What does `\\(x) x + 1` mean in R?",
    answer("A regex that matches x followed by + and 1"),
    answer("An anonymous function that takes x and returns x + 1", correct = TRUE),
    answer("A purrr formula shorthand — equivalent to ~ .x + 1 but different syntax"),
    answer("A lambda only usable inside purrr::map()"),
    allow_retry = TRUE
  ),

  question(
    "What is the key purpose of {{ }} (curly-curly) in a dplyr-aware function?",
    answer("It makes the function run faster by pre-evaluating the column"),
    answer("It passes a column name (without quotes) through to dplyr verbs inside the function", correct = TRUE),
    answer("It is required whenever using group_by() inside a function"),
    answer("It converts a string argument to a column reference"),
    allow_retry = TRUE
  ),

  question(
    "What is the difference between purrr::map() and purrr::map_dfr()?",
    answer("map() is faster; map_dfr() is type-safe"),
    answer("map() returns a list; map_dfr() returns a data frame with rows bound together", correct = TRUE),
    answer("map_dfr() only works with data frames as input"),
    answer("There is no difference — map_dfr() is deprecated"),
    allow_retry = TRUE
  ),

  question(
    "The 'many-to-many fanout' join bug occurs when:",
    answer("You forget to include na.rm = TRUE in a summarise"),
    answer("Both the left and right datasets have multiple rows per key, producing a Cartesian product per key", correct = TRUE),
    answer("You join on a column that contains NA values"),
    answer("The key column has different types in each dataset"),
    allow_retry = TRUE
  )
)
```

---

## 15  SAS → dplyr Joins & Functions Cheat Sheet

### Joins

| SAS pattern | dplyr equivalent |
|---|---|
| `PROC SQL LEFT JOIN` | `left_join(x, y, by = "key")` |
| `PROC SQL INNER JOIN` | `inner_join(x, y, by = "key")` |
| `PROC SQL FULL JOIN` | `full_join(x, y, by = "key")` |
| `DATA step MERGE … IF IN(1)` | `left_join()` |
| "Keep X rows where key in Y" | `semi_join(x, y, by = "key")` |
| "Keep X rows where key NOT in Y" | `anti_join(x, y, by = "key")` |
| `ON a.col = b.other_col` | `by = join_by(col == other_col)` |
| PROC SQL pass-through | `dbplyr::tbl()` + `collect()` |
| Post-join row explosion QC | `relationship = "many-to-one"` |

### Functions

| SAS macro pattern | R function equivalent |
|---|---|
| `%macro foo(var)` | `foo <- function(var) { ... }` |
| `&var.` substitution | Just `var` (it's a real variable) |
| `%let x = 5;` | `x <- 5` |
| `%if cond %then; %else;` | `if (cond) ... else ...` |
| `%do i = 1 %to 10;` | `for (i in 1:10)` or `map(1:10, ...)` |
| Return a dataset | `return(my_df)` or last expression |
| Pass column name to inner verb | `{{ col }}` (curly-curly) |
| Pass column name as string | `.data[[col]]` |
| Extra flexible arguments | `...` (dot-dot-dot) |
| Anonymous one-off function | `\(x) expression` (R 4.1+) |

---

## Key Takeaways

1. **Six join verbs** — `left_join` (80%), `semi_join`/`anti_join` (QC), `inner_join`, `full_join`
2. **`anti_join` is your QC friend** — find orphan records and missing matches in seconds
3. **Many-to-many fanout** is the most dangerous silent join bug — assert expectations with `relationship = "..."`
4. **`join_by()`** is the modern syntax for joins on differently-named or inequality columns
5. **R functions are first-class** — pass, return, and compose them freely unlike SAS macros
6. **`{{ }}`** passes unquoted column names into dplyr verbs inside functions
7. **`\(x)`** is the modern anonymous function shorthand — use it inside `across()` and `map()`
8. **`purrr::map_dfr()`** replaces SAS `%do` loops with clean, composable iteration

**Next** → Lesson 06: tidyr reshaping, pivot_longer/pivot_wider, lubridate dates
(`rmarkdown::run("06_tidyr_lubridate.Rmd")`)
