---
title: "Lesson 06 — R Primer for SAS Programmers, Part 4: The Broader Tidyverse"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    ace_theme: tomorrow
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Reshape data with pivot_longer/pivot_wider (PROC TRANSPOSE). Handle CDISC ISO 8601
  dates with lubridate. Manipulate strings with stringr. Control factor levels with
  forcats. Fill and complete sparse visit structures. All illustrated with real
  pharmaverseadam clinical datasets (CDISCPILOT01 Alzheimer's trial).
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 06_tidyverse.Rmd
# PURPOSE       : Core tidyverse tools for clinical data wrangling: tidyr pivoting
#                 (pivot_wider/longer), stringr string manipulation,
#                 lubridate date/time arithmetic, and purrr functional
#                 programming patterns.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(forcats)
library(tibble)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 60)

# ── Load pharmaverseadam datasets ─────────────────────────────────────────────
data(adsl, package = "pharmaverseadam")
data(adae, package = "pharmaverseadam")
data(adlb, package = "pharmaverseadam")
data(advs, package = "pharmaverseadam")
data(adcm, package = "pharmaverseadam")
data(adex, package = "pharmaverseadam")

# Strip admiral-appended derived rows (DTYPE = "MINIMUM", "MAXIMUM", "LOCF" …)
# so exercises work on observed records only
adlb <- adlb |> filter(is.na(DTYPE))
advs <- advs |> filter(is.na(DTYPE))

# ── Wide baseline vital signs snapshot (for pivot_longer exercises) ───────────
# One row per subject with columns BASE_SYSBP, BASE_DIABP, BASE_PULSE
wide_advs <- advs |>
  filter(PARAMCD %in% c("SYSBP", "DIABP", "PULSE"), ABLFL == "Y") |>
  group_by(USUBJID, PARAMCD) |>
  slice(1) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVAL) |>
  pivot_wider(names_from  = PARAMCD,
              values_from = AVAL,
              names_prefix = "BASE_") |>
  filter(!is.na(BASE_SYSBP), !is.na(BASE_DIABP), !is.na(BASE_PULSE))

# ── Wide baseline lab snapshot (for supplementary exercises) ──────────────────
# One row per subject with columns BASE_ALT, BASE_AST, BASE_ALB
wide_labs <- adlb |>
  filter(PARAMCD %in% c("ALT", "AST", "ALB"), ABLFL == "Y") |>
  group_by(USUBJID, PARAMCD) |>
  slice(1) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVAL) |>
  pivot_wider(names_from  = PARAMCD,
              values_from = AVAL,
              names_prefix = "BASE_") |>
  filter(!is.na(BASE_ALT))

# ── Combined TESTCD_CAT column (for separate_wider_delim exercise) ────────────
lab_combined <- adlb |>
  filter(ABLFL == "Y",
         PARAMCD %in% c("ALT", "HGB", "WBC", "AST", "ALB", "PLAT")) |>
  select(USUBJID, LBTESTCD, LBCAT) |>
  mutate(TESTCD_CAT = str_c(LBTESTCD, "_", LBCAT)) |>
  select(USUBJID, TESTCD_CAT) |>
  distinct() |>
  arrange(USUBJID, TESTCD_CAT) |>
  slice_head(n = 8)

# ── Split date columns from adae AESTDTC (for unite exercise) ─────────────────
dates_split <- adae |>
  filter(!is.na(AESTDTC)) |>
  select(USUBJID, AESTDTC) |>
  distinct(USUBJID, .keep_all = TRUE) |>
  mutate(
    YEAR  = str_sub(AESTDTC, 1, 4),
    MONTH = str_sub(AESTDTC, 6, 7),
    DAY   = str_sub(AESTDTC, 9, 10)
  ) |>
  select(USUBJID, YEAR, MONTH, DAY) |>
  slice_head(n = 5)
```

## Overview

### The tidyverse — a shared design philosophy

The tidyverse is a collection of packages built around four principles:

- **Tidy data**: one row per observation, one column per variable, one value per cell
- Functions take a data frame as their first argument and return a data frame
- Function names are **verbs** describing what they do
- The pipe `|>` connects operations into readable pipelines

In clinical work, the packages beyond `{dplyr}` you'll use most are:

| Package | Clinical use | SAS equivalent |
|---|---|---|
| `{tidyr}` | Reshape wide ↔ long (BDS structure); fill sparse visits | `PROC TRANSPOSE`, `RETAIN` |
| `{lubridate}` | Parse & do arithmetic on CDISC ISO 8601 dates | `INPUT(., YYMMDD10.)`, `INTCK`, `INTNX` |
| `{stringr}` | Clean, extract, detect strings | `SUBSTR`, `TRANWRD`, `INDEX` |
| `{forcats}` | Control factor / category order for TLGs | `PROC FORMAT` |

> All datasets are pre-loaded from `{pharmaverseadam}` (CDISCPILOT01 Alzheimer's trial):
> `adsl` (254 subjects), `adae`, `adlb`, `advs`, `adcm`, `adex`, plus helper snapshots
> `wide_advs` and `wide_labs` for reshape exercises.

---

```{r quiz-overview-06, echo=FALSE}
quiz(
  caption = "Before we begin",
  question(
    "CDISC's BDS (Basic Data Structure) format is best described as:",
    answer("One row per subject with parameters in separate columns (wide)"),
    answer("One row per parameter per subject per analysis timepoint (long / tidy)",
           correct = TRUE),
    answer("One row per study with all subjects embedded"),
    answer("Identical to ADSL structure"),
    allow_retry = TRUE
  ),
  question(
    "Which tidyverse package replaces PROC TRANSPOSE?",
    answer("lubridate"),
    answer("stringr"),
    answer("tidyr", correct = TRUE),
    answer("forcats"),
    allow_retry = TRUE
  ),
  question(
    "The pharmaverseadam BDS datasets (adlb, advs) contain rows where DTYPE is
     non-missing (e.g. 'MINIMUM', 'LOCF'). What are these rows?",
    answer("Duplicate source records that should always be kept"),
    answer("Admiral-appended derived / imputed records that must be filtered out
           before most analyses", correct = TRUE),
    answer("Records where the analysis value is missing"),
    answer("Baseline records flagged by ABLFL"),
    allow_retry = TRUE
  )
)
```

---

## 1  Wide vs Long — the tidy data concept

### Concept

**Wide** format — one row per subject, each measurement in its own column:

```
USUBJID      BASE_SYSBP  BASE_DIABP  BASE_PULSE
01-701-1015  130         78          84
01-701-1023  118         72          72
```

**Long (BDS)** format — one row per measurement, with a column identifying which parameter:

```
USUBJID      PARAMCD  AVAL
01-701-1015  SYSBP    130
01-701-1015  DIABP    78
01-701-1015  PULSE    84
01-701-1023  SYSBP    118
...
```

CDISC's BDS datasets (`adlb`, `advs`, `adex`) are long. Correlation analyses, shift tables, and TLG listings sometimes need wide. You'll constantly reshape between them.

Run the code below to compare `wide_advs` and a filtered `advs` side-by-side:

```{r compare-shapes, exercise=TRUE, exercise.lines=8}
# Wide — one row per subject, parameter in columns
wide_advs |> slice_head(n = 4)

# Long — one row per parameter per subject per visit
advs |>
  filter(PARAMCD %in% c("SYSBP", "DIABP", "PULSE"), ABLFL == "Y") |>
  select(USUBJID, PARAMCD, AVAL) |>
  slice_head(n = 8)
```

---

## 2  pivot_longer() — wide to long (PROC TRANSPOSE out=)

### Concept

```r
wide_data |>
  pivot_longer(
    cols      = c(BASE_SYSBP, BASE_DIABP, BASE_PULSE),  # columns to stack
    names_to  = "PARAMCD",                               # column for old names
    values_to = "AVAL"                                   # column for old values
  )
```

Key `cols` options:

- **Explicit list**: `c(BASE_SYSBP, BASE_DIABP, BASE_PULSE)`
- **Exclude key columns**: `-USUBJID`
- **Tidyselect helpers**: `starts_with("BASE_")`, `where(is.numeric)`

### Exercise 2a — basic pivot_longer

Pivot `wide_advs` so each baseline vital sign value is on its own row.
Name the new key column `PARAMCD` and value column `AVAL`.

```{r pivot-longer-basic, exercise=TRUE, exercise.lines=10}
wide_advs |>
  pivot_longer(
    cols      = ___,
    names_to  = "___",
    values_to = "___"
  )
```

```{r pivot-longer-basic-hint-1}
# cols = c(BASE_SYSBP, BASE_DIABP, BASE_PULSE)  OR  cols = -USUBJID
# names_to = "PARAMCD", values_to = "AVAL"
```

```{r pivot-longer-basic-solution}
wide_advs |>
  pivot_longer(
    cols      = -USUBJID,
    names_to  = "PARAMCD",
    values_to = "AVAL"
  )
```

### Exercise 2b — pivot with names_prefix to strip column prefix

Pivot `wide_advs` using `starts_with("BASE_")` as the `cols` selector.
Use `names_prefix` to strip the `"BASE_"` prefix so the resulting `PARAMCD` column
contains `"SYSBP"`, `"DIABP"`, and `"PULSE"` — not `"BASE_SYSBP"`, etc.

```{r pivot-longer-prefix, exercise=TRUE, exercise.lines=10}
wide_advs |>
  pivot_longer(
    cols         = starts_with("___"),
    names_to     = "___",
    values_to    = "___",
    names_prefix = "___"
  )
```

```{r pivot-longer-prefix-hint-1}
# starts_with("BASE_")
# names_prefix = "BASE_"  strips the prefix, leaving "SYSBP", "DIABP", "PULSE"
```

```{r pivot-longer-prefix-solution}
wide_advs |>
  pivot_longer(
    cols         = starts_with("BASE_"),
    names_to     = "___",
    values_to    = "___",
    names_prefix = "BASE_"
  )
```

### Exercise 2c — pivot multiple value columns simultaneously

`wide_labs` has columns `BASE_ALT`, `BASE_AST`, `BASE_ALB`.
A collaborator delivers an **extended** snapshot that also has `CHNG_ALT`, `CHNG_AST`,
`CHNG_ALB` (change from baseline). Pivot both groups at once using `.value` in
`names_to` and a `names_pattern` regex so you get columns `AVAL` (from BASE_*)
and `CHG` (from CHNG_*).

```{r pivot-longer-multivalue, exercise=TRUE, exercise.lines=18}
# Build the extended wide snapshot first
wide_labs_ext <- wide_labs |>
  mutate(
    CHNG_ALT = BASE_ALT * 0.1,   # synthetic change-from-baseline
    CHNG_AST = BASE_AST * 0.08,
    CHNG_ALB = BASE_ALB * -0.05
  )

# Now pivot both BASE_* and CHNG_* into AVAL and CHG columns
wide_labs_ext |>
  pivot_longer(
    cols          = -USUBJID,
    names_to      = c("___", "PARAMCD"),  # .value triggers multi-column output
    names_pattern = "___"                 # regex: capture prefix then testcd
  )
```

```{r pivot-longer-multivalue-hint-1}
# names_to      = c(".value", "PARAMCD")
# names_pattern = "^(BASE|CHNG)_(.+)$"
# .value maps "BASE" → AVAL, "CHNG" → CHG, but we need explicit values_to
# A cleaner approach: names_pattern = "^(BASE|CHNG)_(.+)$"
# and rename BASE->AVAL / CHNG->CHG with a rename_with() afterwards,
# OR: names_to = c("TYPE", "PARAMCD"), then pivot_wider TYPE → AVAL/CHG
```

```{r pivot-longer-multivalue-solution}
wide_labs_ext <- wide_labs |>
  mutate(
    CHNG_ALT = BASE_ALT * 0.1,
    CHNG_AST = BASE_AST * 0.08,
    CHNG_ALB = BASE_ALB * -0.05
  )

wide_labs_ext |>
  pivot_longer(
    cols          = -USUBJID,
    names_to      = c(".value", "PARAMCD"),
    names_pattern = "^(BASE|CHNG)_(.+)$"
  ) |>
  rename(AVAL = BASE, CHG = CHNG)
```

---

## 3  pivot_wider() — long to wide (PROC TRANSPOSE id= var= out=)

### Concept

```r
long_data |>
  pivot_wider(
    names_from  = PARAMCD,   # column whose values become new column names
    values_from = AVAL       # column whose values fill the new columns
  )
```

Common clinical uses: shift tables (ANRIND at baseline vs. on-treatment), correlation
matrices of lab parameters, and wide demography snapshots joined to BDS datasets.

### Exercise 3a — pivot adlb to wide by parameter at each visit

From `adlb`, select SYSBP is in `advs` — let's use lab parameters.
Pivot the **on-treatment** `adlb` records (filter `ONTRTFL == "Y"`, then pick
`ALT`, `AST`, `ALB`) so that each row is one `(USUBJID, AVISIT)` pair with
separate columns per parameter. Select only the columns you need before pivoting.

```{r pivot-wider-basic, exercise=TRUE, exercise.lines=12}
adlb |>
  filter(ONTRTFL == "Y", PARAMCD %in% c("ALT", "AST", "ALB")) |>
  select(USUBJID, AVISIT, PARAMCD, AVAL) |>
  pivot_wider(
    names_from  = ___,
    values_from = ___
  )
```

```{r pivot-wider-basic-hint-1}
# names_from = PARAMCD, values_from = AVAL
# If you see list-columns, a subject had more than one record per AVISIT/PARAMCD.
# Add values_fn = list(AVAL = mean) to collapse them.
```

```{r pivot-wider-basic-solution}
adlb |>
  filter(ONTRTFL == "Y", PARAMCD %in% c("ALT", "AST", "ALB")) |>
  select(USUBJID, AVISIT, PARAMCD, AVAL) |>
  pivot_wider(
    names_from  = PARAMCD,
    values_from = AVAL,
    values_fn   = list(AVAL = mean)   # collapse any duplicates
  )
```

### Exercise 3b — pivot with names_prefix and multiple value columns

Repeat the pivot from 3a, but this time capture **both** `AVAL` (analysis value)
and `CHG` (change from baseline) for each parameter. Add a `"V_"` prefix to avoid
ambiguous column names.

```{r pivot-wider-multicol, exercise=TRUE, exercise.lines=14}
adlb |>
  filter(ONTRTFL == "Y", PARAMCD %in% c("ALT", "AST", "ALB"),
         !is.na(CHG)) |>
  select(USUBJID, AVISIT, PARAMCD, AVAL, CHG) |>
  pivot_wider(
    names_from   = ___,
    values_from  = c(___, ___),
    names_prefix = "___",
    values_fn    = mean
  )
```

```{r pivot-wider-multicol-hint-1}
# names_from = PARAMCD
# values_from = c(AVAL, CHG)
# names_prefix = "V_"
# This creates columns like V_AVAL_ALT, V_AVAL_AST, V_CHG_ALT, etc.
```

```{r pivot-wider-multicol-solution}
adlb |>
  filter(ONTRTFL == "Y", PARAMCD %in% c("ALT", "AST", "ALB"),
         !is.na(CHG)) |>
  select(USUBJID, AVISIT, PARAMCD, AVAL, CHG) |>
  pivot_wider(
    names_from   = PARAMCD,
    values_from  = c(AVAL, CHG),
    names_prefix = "V_",
    values_fn    = mean
  )
```

### Exercise 3c — baseline wide table joined back to adsl

Build a wide baseline lab table (one row per subject, columns `BASE_ALT`, `BASE_AST`,
`BASE_ALB`), then `left_join` it to `adsl` (keeping `USUBJID`, `TRT01A`, `AGE`, `AGEGR1`).

```{r pivot-join-pipeline, exercise=TRUE, exercise.lines=18}
baseline_wide <- adlb |>
  filter(PARAMCD %in% c("ALT", "AST", "ALB"), ABLFL == "Y") |>
  group_by(USUBJID, PARAMCD) |>
  slice(1) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVAL) |>
  pivot_wider(
    names_from   = ___,
    values_from  = ___,
    names_prefix = "BASE_"
  )

adsl |>
  select(USUBJID, TRT01A, AGE, AGEGR1) |>
  left_join(baseline_wide, by = "USUBJID")
```

```{r pivot-join-pipeline-hint-1}
# names_from = PARAMCD, values_from = AVAL
# names_prefix = "BASE_" → columns BASE_ALT, BASE_AST, BASE_ALB
```

```{r pivot-join-pipeline-solution}
baseline_wide <- adlb |>
  filter(PARAMCD %in% c("ALT", "AST", "ALB"), ABLFL == "Y") |>
  group_by(USUBJID, PARAMCD) |>
  slice(1) |>
  ungroup() |>
  select(USUBJID, PARAMCD, AVAL) |>
  pivot_wider(
    names_from   = PARAMCD,
    values_from  = AVAL,
    names_prefix = "BASE_"
  )

adsl |>
  select(USUBJID, TRT01A, AGE, AGEGR1) |>
  left_join(baseline_wide, by = "USUBJID")
```

### Exercise 3d — shift table: ANRIND baseline vs. post-baseline

A shift table shows the reference-range indicator at **baseline** vs. at each
post-baseline visit. Starting from `adlb`, filter to `PARAMCD == "ALT"` and
pivot ANRIND so each row is one `(USUBJID, AVISIT)` and two columns show
`ANRIND_BASE` and `ANRIND_POST`. Hint: you will need to flag baseline rows first.

```{r shift-table, exercise=TRUE, exercise.lines=20}
# Step 1: baseline ANRIND per subject
base_anr <- adlb |>
  filter(PARAMCD == "ALT", ABLFL == "Y") |>
  group_by(USUBJID) |>
  slice(1) |>
  ungroup() |>
  select(USUBJID, ANRIND_BASE = ANRIND)

# Step 2: post-baseline ANRIND per subject per visit
post_anr <- adlb |>
  filter(PARAMCD == "ALT", ONTRTFL == "Y", !is.na(AVISIT)) |>
  select(USUBJID, AVISIT, ANRIND)

# Step 3: join them
post_anr |>
  left_join(___, by = "___") |>
  select(USUBJID, AVISIT, ANRIND_BASE, ANRIND_POST = ANRIND) |>
  count(ANRIND_BASE, ANRIND_POST)
```

```{r shift-table-hint-1}
# left_join(base_anr, by = "USUBJID")
# Then count(ANRIND_BASE, ANRIND_POST) gives the shift counts
```

```{r shift-table-solution}
base_anr <- adlb |>
  filter(PARAMCD == "ALT", ABLFL == "Y") |>
  group_by(USUBJID) |>
  slice(1) |>
  ungroup() |>
  select(USUBJID, ANRIND_BASE = ANRIND)

post_anr <- adlb |>
  filter(PARAMCD == "ALT", ONTRTFL == "Y", !is.na(AVISIT)) |>
  select(USUBJID, AVISIT, ANRIND)

post_anr |>
  left_join(base_anr, by = "USUBJID") |>
  select(USUBJID, AVISIT, ANRIND_BASE, ANRIND_POST = ANRIND) |>
  count(ANRIND_BASE, ANRIND_POST)
```

```{r quiz-pivot, echo=FALSE}
quiz(
  caption = "pivot check",
  question(
    "You call pivot_wider(names_from = PARAMCD, values_from = c(AVAL, CHG)) on
     adlb filtered to three PARAMCDs. How many new columns are created per
     PARAMCD value?",
    answer("One column per PARAMCD"),
    answer("Two columns per PARAMCD — one for AVAL, one for CHG", correct = TRUE),
    answer("Three columns: AVAL, CHG, and a combined column"),
    answer("It depends on the number of distinct AVISITN values"),
    allow_retry = TRUE
  ),
  question(
    "pivot_wider() warns 'Values from AVAL are not uniquely identified'. This means:",
    answer("AVAL contains only missing values"),
    answer("There is more than one AVAL for at least one (USUBJID, AVISIT, PARAMCD)
           combination — the output will contain list-columns", correct = TRUE),
    answer("The names_from column has duplicates"),
    answer("You forgot to supply values_to"),
    allow_retry = TRUE
  ),
  question(
    "What argument in pivot_wider lets you specify how to collapse duplicates —
     for example, taking the mean when there are multiple AVAL values per cell?",
    answer("values_fill"),
    answer("names_repair"),
    answer("values_fn", correct = TRUE),
    answer("id_cols"),
    allow_retry = TRUE
  )
)
```

---

## 4  separate_wider_delim(), unite(), complete(), fill()

### Concept

Four small tools that each replace a different SAS pattern in ADaM / BDS work:

| tidyr function | What it does | SAS analog |
|---|---|---|
| `separate_wider_delim()` | Split **one** column into **many** on a delimiter | `SCAN()` + multiple assignments in DATA step |
| `unite()` | Glue **many** columns into **one** with a separator | `CATX()` / `CAT('-', y, m, d)` |
| `complete()` | Insert **missing rows** for combinations that should exist | `PROC FREQ … /SPARSE` or merging against a `_NULL_` skeleton dataset |
| `fill()` | Propagate the **last non-missing value** into NA rows | `RETAIN var;` + `IF NOT MISSING(var) THEN _var = var;` (LOCF) |

`separate_wider_delim()` and `unite()` are mechanical and easy. The two that
trip SAS users up are `complete()` and `fill()` — covered in detail below.

---

#### 4.1  `complete()` — the BDS "structural zero" problem

In ADaM BDS (`adlb`, `advs`, `adeg`), a row exists only when a measurement was
**actually recorded**. If subject `01-701-1015` skipped Week 8 for HGB, there
is **no Week 8 row at all** for that subject and parameter.

For TLG layouts, this is a bug: the table needs a cell for every
**(subject × visit)** combination, even an empty one. SAS solves this with
`PROC FREQ … /SPARSE` or by merging against a Cartesian skeleton. tidyr solves
it with one call:

```r
adlb_hgb |>
  filter(PARAMCD == "HGB") |>
  complete(USUBJID, AVISIT)
```

That single line says: "find every distinct `USUBJID` and every distinct
`AVISIT` in the data; make sure a row exists for every combination of the two;
for any newly-inserted row, fill non-key columns with `NA`."

**Before `complete()`** — subject is missing Week 8 entirely:

```
USUBJID      PARAMCD  AVISIT      AVAL  TRT01A
01-701-1015  HGB      Baseline    14.2  Placebo
01-701-1015  HGB      Week 2      14.0  Placebo
01-701-1015  HGB      Week 4      14.1  Placebo
                                                     ← no Week 8 row
01-701-1015  HGB      Week 12     13.8  Placebo
```

**After `complete(USUBJID, AVISIT)`** — Week 8 row is materialised with NAs:

```
USUBJID      PARAMCD  AVISIT      AVAL  TRT01A
01-701-1015  HGB      Baseline    14.2  Placebo
01-701-1015  HGB      Week 2      14.0  Placebo
01-701-1015  HGB      Week 4      14.1  Placebo
01-701-1015  HGB      Week 8      NA    NA       ← inserted by complete()
01-701-1015  HGB      Week 12     13.8  Placebo
```

> **Why `TRT01A` is NA on the inserted row** — `complete()` only knows about
> the columns you pass as keys (`USUBJID`, `AVISIT`). Every other column in
> the inserted row defaults to `NA`. That is the entire reason `fill()` exists.

---

#### 4.2  `fill()` — LOCF / propagation of subject-level attributes

After `complete()`, columns like `TRT01A` (constant per subject) are `NA` on
every inserted row. In SAS you would write:

```sas
DATA adlb_fill;
  SET adlb_complete;
  BY USUBJID;
  RETAIN _TRT01A;
  IF NOT MISSING(TRT01A) THEN _TRT01A = TRT01A;
  TRT01A = _TRT01A;
  DROP _TRT01A;
RUN;
```

tidyr collapses that to:

```r
adlb_complete |>
  group_by(USUBJID) |>
  fill(TRT01A, .direction = "downup") |>
  ungroup()
```

`fill()` walks each group and replaces `NA` with the **most recent non-NA**
value above it (or below, depending on `.direction`). The grouping is
critical — without `group_by(USUBJID)` the function leaks values from one
subject into the next, just as a SAS `RETAIN` without a `BY` group would.

**`.direction` options:**

| value | behaviour | clinical use |
|---|---|---|
| `"down"` | NA replaced by the value **above** (default; classic LOCF) | propagate baseline assessment forward |
| `"up"` | NA replaced by the value **below** | back-fill an early missing visit from a later one |
| `"downup"` | Down first, then up for any leading NA gaps | safe default for subject-level constants (`TRT01A`, `SEX`) |
| `"updown"` | Up first, then down | when later visits are more trustworthy than earlier |

**Visual on a single subject's `TRT01A` after `complete()` inserts a Week 8 row:**

```
                       fill(TRT01A,                    fill(TRT01A,
AVISIT      TRT01A      .direction="down")             .direction="downup")
---------   ---------   --------------------           --------------------
Baseline    Placebo     Placebo                        Placebo
Week 2      Placebo     Placebo                        Placebo
Week 4      Placebo     Placebo                        Placebo
Week 8      NA          Placebo  ← carried down        Placebo
Week 12     Placebo     Placebo                        Placebo
```

`complete()` + `fill()` together turn sparse BDS data into a rectangular
analysis-ready grid — exactly what `tern`, `rtables`, and gtsummary expect.

### Exercise 4a — split TESTCD_CAT into test code and category

Split `lab_combined$TESTCD_CAT` (e.g. `"ALT_CHEMISTRY"`) into two columns:
`LBTESTCD` and `LBCAT`.

```{r separate-ex, exercise=TRUE, exercise.lines=8}
lab_combined |>
  separate_wider_delim(
    TESTCD_CAT,
    delim = "___",
    names = c("___", "___")
  )
```

```{r separate-ex-hint-1}
# delim = "_"
# names = c("LBTESTCD", "LBCAT")
```

```{r separate-ex-solution}
lab_combined |>
  separate_wider_delim(
    TESTCD_CAT,
    delim = "_",
    names = c("LBTESTCD", "LBCAT")
  )
```

### Exercise 4b — unite year/month/day back into ISO 8601

`dates_split` contains `YEAR`, `MONTH`, `DAY` as separate character columns
(sourced from `adae$AESTDTC`). Combine them into a single `AESTDTC` column
in `YYYY-MM-DD` format.

```{r unite-ex, exercise=TRUE, exercise.lines=8}
dates_split |>
  unite("___", c(___, ___, ___), sep = "___")
```

```{r unite-ex-hint-1}
# unite("AESTDTC", c(YEAR, MONTH, DAY), sep = "-")
```

```{r unite-ex-solution}
dates_split |>
  unite("AESTDTC", c(YEAR, MONTH, DAY), sep = "-")
```

### Exercise 4c — complete() to expose missing visit records

A TLG requires every subject in the safety population to appear at every scheduled
visit, even if no lab result was recorded. Starting from `adlb`, filter to
`PARAMCD == "HGB"` and the safety population (`SAFFL == "Y"`), then use `complete()`
to ensure all `(USUBJID, AVISIT)` combinations exist. Keep only scheduled visits
(`!is.na(AVISIT)`).

```{r complete-ex, exercise=TRUE, exercise.lines=14}
adlb |>
  filter(PARAMCD == "HGB", SAFFL == "Y", !is.na(AVISIT)) |>
  select(USUBJID, TRT01A, AVISIT, AVISITN, AVAL) |>
  complete(___, ___) |>
  arrange(USUBJID, AVISITN) |>
  slice_head(n = 15)
```

```{r complete-ex-hint-1}
# complete(USUBJID, AVISIT)
# After complete(), rows for missing visits have NA for AVAL, TRT01A, AVISITN.
# Notice that AVISITN and TRT01A become NA for the gap rows.
```

```{r complete-ex-solution}
adlb |>
  filter(PARAMCD == "HGB", SAFFL == "Y", !is.na(AVISIT)) |>
  select(USUBJID, TRT01A, AVISIT, AVISITN, AVAL) |>
  complete(USUBJID, AVISIT) |>
  arrange(USUBJID, AVISITN) |>
  slice_head(n = 15)
```

### Exercise 4d — fill() to propagate TRT01A into gap rows

Extend 4c: after `complete()`, use `fill()` to propagate `TRT01A` and `AVISITN`
downward within each subject so the gap rows are labelled correctly.

```{r fill-ex, exercise=TRUE, exercise.lines=18}
visit_order <- adlb |>
  filter(!is.na(AVISIT)) |>
  distinct(AVISIT, AVISITN) |>
  arrange(AVISITN)

adlb |>
  filter(PARAMCD == "HGB", SAFFL == "Y", !is.na(AVISIT)) |>
  select(USUBJID, TRT01A, AVISIT, AVISITN, AVAL) |>
  complete(USUBJID, AVISIT) |>
  group_by(USUBJID) |>
  fill(___, ___, .direction = "___") |>
  ungroup() |>
  left_join(visit_order |> select(AVISIT, AVISITN), by = "AVISIT") |>
  arrange(USUBJID, AVISITN.y) |>
  slice_head(n = 15)
```

```{r fill-ex-hint-1}
# fill(TRT01A, AVISITN, .direction = "downup")
# "downup" fills forward first, then fills any remaining NAs at the top
# backward — so both leading and trailing gaps are covered.
```

```{r fill-ex-solution}
visit_order <- adlb |>
  filter(!is.na(AVISIT)) |>
  distinct(AVISIT, AVISITN) |>
  arrange(AVISITN)

adlb |>
  filter(PARAMCD == "HGB", SAFFL == "Y", !is.na(AVISIT)) |>
  select(USUBJID, TRT01A, AVISIT, AVISITN, AVAL) |>
  complete(USUBJID, AVISIT) |>
  group_by(USUBJID) |>
  fill(TRT01A, .direction = "downup") |>
  ungroup() |>
  left_join(visit_order |> select(AVISIT, AVISITN_NUM = AVISITN),
            by = "AVISIT") |>
  arrange(USUBJID, AVISITN_NUM) |>
  slice_head(n = 15)
```

```{r quiz-tidyr, echo=FALSE}
quiz(
  caption = "tidyr check",
  question(
    "After calling complete(USUBJID, AVISIT) on an adlb subset, rows added for
     missing visits have NA in columns that were not listed in complete().
     How would you fill TRT01A (which is constant per subject) into those gaps?",
    answer("mutate(TRT01A = coalesce(TRT01A, first(TRT01A)))"),
    answer("group_by(USUBJID) |> fill(TRT01A, .direction = 'downup') |> ungroup()",
           correct = TRUE),
    answer("replace_na(list(TRT01A = 'Unknown'))"),
    answer("pivot_wider(names_from = TRT01A, values_from = AVAL)"),
    allow_retry = TRUE
  ),
  question(
    "separate_wider_delim() fails with 'Expected 2 pieces. Additional pieces
     discarded.' What is the most likely cause for CDISC lab data?",
    answer("The delimiter appears more than once in some values — for example,
           a LBCAT containing an underscore", correct = TRUE),
    answer("The column has missing values"),
    answer("The names argument has too many elements"),
    answer("delim must be a single character only"),
    allow_retry = TRUE
  )
)
```

---

## 5  lubridate — CDISC ISO 8601 dates

### Concept

CDISC stores raw dates as ISO 8601 character strings (`YYYY-MM-DD`). You must convert
them to R `Date` objects for arithmetic. In `{pharmaverseadam}`, source-domain date
character columns (e.g. `RFSTDTC`, `AESTDTC`) co-exist with pre-parsed `Date` columns
(e.g. `TRTSDT`, `ASTDT`, `ADT`) — you need both skills.

```r
library(lubridate)

ymd("2014-01-05")          # Date from ISO 8601 string → 2014-01-05
year(d); month(d); yday(d) # extract parts
d + days(7)                # 2014-01-12
d + months(1)              # 2014-02-05  (use %m+% months(1) to stay in-bounds)
as.numeric(d2 - d1)        # integer days between two Dates
```

### Exercise 5a — parse ISO 8601 strings from RFSTDTC

`adsl$RFSTDTC` is the date of first study treatment as a character string
(`"YYYY-MM-DD"`). Parse it into a proper R `Date` column called `RFSTDT`.
Show `USUBJID`, `RFSTDTC`, and `RFSTDT`, and confirm the class of the new column.

```{r lubridate-parse, exercise=TRUE, exercise.lines=10}
adsl |>
  mutate(RFSTDT = ___(RFSTDTC)) |>
  select(USUBJID, RFSTDTC, RFSTDT) |>
  slice_head(n = 5) |>
  mutate(col_class = class(RFSTDT))
```

```{r lubridate-parse-hint-1}
# ymd(RFSTDTC) converts an ISO 8601 character column to a Date object
```

```{r lubridate-parse-solution}
adsl |>
  mutate(RFSTDT = ymd(RFSTDTC)) |>
  select(USUBJID, RFSTDTC, RFSTDT) |>
  slice_head(n = 5) |>
  mutate(col_class = class(RFSTDT))
```

### Exercise 5b — manually compute study day and verify against ASTDY

In `adae`, `AESTDTC` is the ISO 8601 AE start date character string.
The CDISC ADaM study-day formula is:

```
ASTDY = AESTDT − TRTSDT + 1   (if AESTDT ≥ TRTSDT)
ASTDY = AESTDT − TRTSDT       (if AESTDT < TRTSDT)
```

Parse `AESTDTC`, join `TRTSDT` from `adsl`, compute `ASTDY_CALC` using the
formula above, then compare it to the pre-computed `ASTDY` column in `adae`.

```{r lubridate-arith, exercise=TRUE, exercise.lines=18}
adae |>
  mutate(AESTDT = ymd(___)) |>
  left_join(adsl |> select(USUBJID, TRTSDT), by = "USUBJID") |>
  mutate(
    DIFF       = as.numeric(AESTDT - TRTSDT),
    ASTDY_CALC = if_else(DIFF >= 0, DIFF + 1L, DIFF)
  ) |>
  select(USUBJID, AESTDTC, TRTSDT, ASTDY, ASTDY_CALC) |>
  filter(!is.na(ASTDY)) |>
  slice_head(n = 8)
```

```{r lubridate-arith-hint-1}
# ymd(AESTDTC) — parse the character column
# as.numeric(AESTDT - TRTSDT) gives the numeric day difference
# CDISC: if diff >= 0 add 1 (no day 0); if diff < 0 keep as is (pre-dose)
```

```{r lubridate-arith-solution}
adae |>
  mutate(AESTDT = ymd(AESTDTC)) |>
  left_join(adsl |> select(USUBJID, TRTSDT), by = "USUBJID") |>
  mutate(
    DIFF       = as.numeric(AESTDT - TRTSDT),
    ASTDY_CALC = if_else(DIFF >= 0, DIFF + 1L, DIFF)
  ) |>
  select(USUBJID, AESTDTC, TRTSDT, ASTDY, ASTDY_CALC) |>
  filter(!is.na(ASTDY)) |>
  slice_head(n = 8)
```

### Exercise 5c — extract date parts from the ADT column in adlb

`adlb$ADT` is already an R `Date` object. For the `ALT` parameter, extract the
**year**, **calendar month as an abbreviated label** (`label = TRUE`), and
**day of year** from `ADT`. Show `USUBJID`, `AVISIT`, `ADT`, and the three new columns.

```{r lubridate-parts, exercise=TRUE, exercise.lines=12}
adlb |>
  filter(PARAMCD == "ALT", !is.na(ADT), !is.na(AVISIT)) |>
  mutate(
    YEAR   = ___(ADT),
    MON    = ___(ADT, label = TRUE),
    DOY    = ___(ADT)
  ) |>
  select(USUBJID, AVISIT, ADT, YEAR, MON, DOY) |>
  distinct() |>
  slice_head(n = 8)
```

```{r lubridate-parts-hint-1}
# year(ADT), month(ADT, label = TRUE), yday(ADT)
```

```{r lubridate-parts-solution}
adlb |>
  filter(PARAMCD == "ALT", !is.na(ADT), !is.na(AVISIT)) |>
  mutate(
    YEAR = year(ADT),
    MON  = month(ADT, label = TRUE),
    DOY  = yday(ADT)
  ) |>
  select(USUBJID, AVISIT, ADT, YEAR, MON, DOY) |>
  distinct() |>
  slice_head(n = 8)
```

### Exercise 5d — verify treatment duration vs TRTDURD

`adsl` contains pre-computed `TRTDURD` (treatment duration in days, integer).
Using `TRTSDT` and `TRTEDT` (both already `Date` objects), manually compute
`TRTDURD_CALC = TRTEDT − TRTSDT + 1` and compare to `TRTDURD`. Flag any
mismatches.

```{r lubridate-trtdur, exercise=TRUE, exercise.lines=12}
adsl |>
  filter(!is.na(TRTEDT)) |>
  mutate(
    TRTDURD_CALC = as.numeric(___ - ___) + 1L,
    MISMATCH     = TRTDURD != TRTDURD_CALC
  ) |>
  select(USUBJID, TRT01A, TRTSDT, TRTEDT, TRTDURD, TRTDURD_CALC, MISMATCH) |>
  slice_head(n = 8)
```

```{r lubridate-trtdur-hint-1}
# as.numeric(TRTEDT - TRTSDT) + 1L
# CDISC: duration = end_date - start_date + 1 (inclusive of both endpoints)
```

```{r lubridate-trtdur-solution}
adsl |>
  filter(!is.na(TRTEDT)) |>
  mutate(
    TRTDURD_CALC = as.numeric(TRTEDT - TRTSDT) + 1L,
    MISMATCH     = TRTDURD != TRTDURD_CALC
  ) |>
  select(USUBJID, TRT01A, TRTSDT, TRTEDT, TRTDURD, TRTDURD_CALC, MISMATCH) |>
  slice_head(n = 8)
```

### Exercise 5e — month arithmetic: follow-up window from TRTEDT

Calculate each subject's 30-day and 90-day post-treatment follow-up end dates from
`TRTEDT`. Use `%m+%` (safe month arithmetic that clamps to month-end) instead of
`+ months()` to avoid an impossible date like February 30.

```{r lubridate-followup, exercise=TRUE, exercise.lines=10}
adsl |>
  filter(!is.na(TRTEDT)) |>
  mutate(
    FU30D  = TRTEDT + ___,     # +30 days
    FU3MO  = TRTEDT %m+% ___   # +3 calendar months (safe)
  ) |>
  select(USUBJID, TRTEDT, FU30D, FU3MO) |>
  slice_head(n = 6)
```

```{r lubridate-followup-hint-1}
# TRTEDT + days(30)
# TRTEDT %m+% months(3)  — %m+% clamps to the last day of the month if needed
```

```{r lubridate-followup-solution}
adsl |>
  filter(!is.na(TRTEDT)) |>
  mutate(
    FU30D = TRTEDT + days(30),
    FU3MO = TRTEDT %m+% months(3)
  ) |>
  select(USUBJID, TRTEDT, FU30D, FU3MO) |>
  slice_head(n = 6)
```

```{r quiz-lubridate, echo=FALSE}
quiz(
  caption = "lubridate check",
  question(
    "Which lubridate function parses the CDISC standard ISO 8601 date string
     '2014-03-15'?",
    answer("mdy('2014-03-15')"),
    answer("dmy('2014-03-15')"),
    answer("ymd('2014-03-15')", correct = TRUE),
    answer("parse_date('2014-03-15')"),
    allow_retry = TRUE
  ),
  question(
    "You compute as.numeric(TRTEDT - TRTSDT) for a subject and get 167.
     But TRTDURD in the CDISC ADaM dataset is 168. Why the difference?",
    answer("lubridate uses 0-based indexing for dates"),
    answer("CDISC ADaM defines duration as end_date - start_date + 1
           (both endpoints inclusive), so the difference is always 1 more
           than simple subtraction", correct = TRUE),
    answer("There is a bug in the pharmaverseadam dataset"),
    answer("TRTEDT is stored in UTC and TRTSDT in local time"),
    allow_retry = TRUE
  ),
  question(
    "What is the risk of using + months(1) instead of %m+% months(1) in
     lubridate?",
    answer("It rounds to the nearest month boundary"),
    answer("It adds 30 days instead of 1 calendar month"),
    answer("It can produce an invalid date — e.g. January 31 + 1 month becomes
           NA or March 3 depending on platform", correct = TRUE),
    answer("It only works on POSIXct, not Date objects"),
    allow_retry = TRUE
  )
)
```

---

## 6  stringr — consistent string manipulation

### Concept

`{stringr}` gives every base R string function a consistent `str_*` prefix and
puts the string as the first argument — making pipes natural.

| Task | base R | stringr |
|---|---|---|
| Detect pattern | `grepl()` | `str_detect()` |
| Extract match | `regmatches()` | `str_extract()` |
| Replace | `gsub()` | `str_replace_all()` |
| Substring | `substr()` | `str_sub()` |
| Trim whitespace | `trimws()` | `str_trim()` |
| Collapse interior spaces | — | `str_squish()` |
| Pad with zeros | `formatC()` | `str_pad()` |
| Upper/lower/title | `toupper()` | `str_to_upper()` / `str_to_title()` |
| Concatenate | `paste0()` | `str_c()` |
| Count matches | `lengths(gregexpr())` | `str_count()` |

### Exercise 6a — detect a body-system pattern in adae

Filter `adae` (treatment-emergent AEs only: `TRTEMFL == "Y"`) to rows where
`AEBODSYS` contains the word `"skin"` (case-insensitive). Count how many such
AEs there are per treatment arm (`TRT01A`).

```{r stringr-detect, exercise=TRUE, exercise.lines=10}
library(stringr)

adae |>
  filter(TRTEMFL == "Y",
         str_detect(AEBODSYS, regex("___", ignore_case = TRUE))) |>
  count(TRT01A, AEBODSYS, sort = TRUE)
```

```{r stringr-detect-hint-1}
# regex("skin", ignore_case = TRUE)
# The Xanomeline arms should show many skin/subcutaneous AEs from patch application
```

```{r stringr-detect-solution}
library(stringr)

adae |>
  filter(TRTEMFL == "Y",
         str_detect(AEBODSYS, regex("skin", ignore_case = TRUE))) |>
  count(TRT01A, AEBODSYS, sort = TRUE)
```

### Exercise 6b — str_sub: extract site and subject number from USUBJID

`adsl$USUBJID` follows the pattern `"01-SSS-NNNN"` (study-site-subject).
Use `str_sub()` to extract:

- `SITEID_CALC`: characters 4–6 (the 3-digit site code)
- `SUBJID_CALC`: characters 8–11 (the 4-digit subject number)

Verify `SITEID_CALC` matches the pre-existing `SITEID` column.

```{r stringr-sub, exercise=TRUE, exercise.lines=10}
adsl |>
  mutate(
    SITEID_CALC  = str_sub(USUBJID, ___, ___),
    SUBJID_CALC  = str_sub(USUBJID, ___, ___)
  ) |>
  select(USUBJID, SITEID, SITEID_CALC, SUBJID, SUBJID_CALC) |>
  slice_head(n = 6)
```

```{r stringr-sub-hint-1}
# USUBJID = "01-701-1015"
# Characters:  1234567890123...
# Site code starts at position 4, ends at 6: str_sub(USUBJID, 4, 6)
# Subject number starts at 8, ends at 11:  str_sub(USUBJID, 8, 11)
```

```{r stringr-sub-solution}
adsl |>
  mutate(
    SITEID_CALC = str_sub(USUBJID, 4, 6),
    SUBJID_CALC = str_sub(USUBJID, 8, 11)
  ) |>
  select(USUBJID, SITEID, SITEID_CALC, SUBJID, SUBJID_CALC) |>
  slice_head(n = 6)
```

### Exercise 6c — str_extract with regex: pull the numeric subject ID

Use `str_extract()` with a regex to capture the **last numeric segment** of
`USUBJID` (i.e. the 4-digit subject number after the final dash). This approach
works even if the format changes — unlike fixed positions.

```{r stringr-extract, exercise=TRUE, exercise.lines=8}
adsl |>
  mutate(SUBJ_NUM = str_extract(USUBJID, "___")) |>
  select(USUBJID, SUBJ_NUM) |>
  slice_head(n = 6)
```

```{r stringr-extract-hint-1}
# Regex to match digits at end: "\\d+$"
# Or more specifically: "(?<=-)[0-9]+$"  (digits after the last dash)
```

```{r stringr-extract-solution}
adsl |>
  mutate(SUBJ_NUM = str_extract(USUBJID, "(?<=-)[0-9]+$")) |>
  select(USUBJID, SUBJ_NUM) |>
  slice_head(n = 6)
```

### Exercise 6d — str_pad: zero-pad the subject number for custom IDs

Suppose a reporting system requires subject IDs in the format `"SUBJ-NNNN"` where
NNNN is zero-padded to 4 digits. Using `SUBJID` from `adsl` (already a character
column like `"1015"`), construct this new ID.

```{r stringr-pad, exercise=TRUE, exercise.lines=8}
adsl |>
  mutate(
    SUBJ_PADDED = str_c("SUBJ-", str_pad(SUBJID, width = ___, pad = "___"))
  ) |>
  select(USUBJID, SUBJID, SUBJ_PADDED) |>
  slice_head(n = 6)
```

```{r stringr-pad-hint-1}
# str_pad(SUBJID, width = 4, pad = "0")
# str_c("SUBJ-", ...) concatenates the prefix
```

```{r stringr-pad-solution}
adsl |>
  mutate(
    SUBJ_PADDED = str_c("SUBJ-", str_pad(SUBJID, width = 4, pad = "0"))
  ) |>
  select(USUBJID, SUBJID, SUBJ_PADDED) |>
  slice_head(n = 6)
```

### Exercise 6e — title case and squish for display-ready AE terms

`adae$AEDECOD` contains MedDRA Preferred Terms in **ALL CAPS**.
For a TLG, convert to title case with `str_to_title()`. Then apply
`str_squish()` to collapse any irregular whitespace. Count the top 10
treatment-emergent preferred terms per treatment arm.

```{r stringr-pipeline, exercise=TRUE, exercise.lines=12}
adae |>
  filter(TRTEMFL == "Y") |>
  mutate(
    AEDECOD_DISPLAY = AEDECOD |> ___() |> ___()
  ) |>
  count(TRT01A, AEDECOD_DISPLAY, sort = TRUE) |>
  group_by(TRT01A) |>
  slice_max(n, n = 5) |>
  ungroup()
```

```{r stringr-pipeline-hint-1}
# str_to_title() then str_squish()
# AEDECOD |> str_to_title() |> str_squish()
```

```{r stringr-pipeline-solution}
adae |>
  filter(TRTEMFL == "Y") |>
  mutate(
    AEDECOD_DISPLAY = AEDECOD |> str_to_title() |> str_squish()
  ) |>
  count(TRT01A, AEDECOD_DISPLAY, sort = TRUE) |>
  group_by(TRT01A) |>
  slice_max(n, n = 5) |>
  ungroup()
```

### Exercise 6f — str_replace_all and str_count across multiple columns

Using `across()`, apply `str_to_upper()` to every character column in
`adsl` that starts with `"TRT"`. Then use `str_count()` to count how many
subjects have a treatment arm name containing the word `"DOSE"`.

```{r stringr-across, exercise=TRUE, exercise.lines=12}
# Part 1: uppercase all TRT* character columns
adsl |>
  mutate(across(starts_with("TRT") & where(is.character),
                ___ )) |>
  select(USUBJID, TRT01A, TRT01P) |>
  slice_head(n = 4)

# Part 2: count subjects whose TRT01A contains "DOSE"
adsl |>
  summarise(n_dose_arms = sum(str_count(TRT01A, "___") > 0))
```

```{r stringr-across-hint-1}
# across(starts_with("TRT") & where(is.character), str_to_upper)
# str_count(TRT01A, "DOSE") > 0  — TRUE when "DOSE" appears at least once
```

```{r stringr-across-solution}
adsl |>
  mutate(across(starts_with("TRT") & where(is.character),
                str_to_upper)) |>
  select(USUBJID, TRT01A, TRT01P) |>
  slice_head(n = 4)

adsl |>
  summarise(n_dose_arms = sum(str_count(TRT01A, "DOSE") > 0))
```

```{r quiz-stringr, echo=FALSE}
quiz(
  caption = "stringr check",
  question(
    "What is the difference between str_trim() and str_squish()?",
    answer("str_trim() is faster; str_squish() is more accurate"),
    answer("str_trim() removes only leading/trailing whitespace;
           str_squish() also collapses multiple interior spaces to one",
           correct = TRUE),
    answer("str_squish() removes all whitespace including inside words"),
    answer("They are identical — str_squish() is just an alias"),
    allow_retry = TRUE
  ),
  question(
    "str_extract(USUBJID, '(?<=-)[0-9]+$') returns '1015' for '01-701-1015'.
     What does the (?<=-) part of the regex do?",
    answer("It matches only digits"),
    answer("It anchors the match to the end of the string"),
    answer("It is a lookbehind assertion: matches digits that are immediately
           preceded by a dash, without including the dash in the result",
           correct = TRUE),
    answer("It matches the entire string after the first dash"),
    allow_retry = TRUE
  ),
  question(
    "Translate SAS: tranwrd(AEDECOD, 'NOS', 'not otherwise specified') to R:",
    answer("str_detect(AEDECOD, 'NOS')"),
    answer("str_extract(AEDECOD, 'NOS')"),
    answer("str_replace_all(AEDECOD, 'NOS', 'not otherwise specified')",
           correct = TRUE),
    answer("str_sub(AEDECOD, 1, 3)"),
    allow_retry = TRUE
  )
)
```

---

## 7  forcats — controlling factor levels for TLGs

### Concept

A **factor** is R's categorical type. Internally it is an integer vector with a
**levels** attribute — a lookup table from integer to label. That is exactly
what a SAS user-defined format does: PROC FORMAT defines the labels, and
`ORDER=FORMATTED` controls display order in PROC FREQ / PROC REPORT.

```r
fct(c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"),
    levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"))
#> [1] Placebo              Xanomeline Low Dose  Xanomeline High Dose
#> Levels: Placebo Xanomeline Low Dose Xanomeline High Dose
```

#### Why it matters in TLGs

`adsl$TRT01A` is a character column. If you build a frequency table from it,
**R sorts alphabetically by default** — that is the wrong order clinically:

```
                            character sort                 factor with explicit levels
                            (default R / SAS               (clinical order — what the
                            ORDER=INTERNAL)                regulator expects)
USUBJID  TRT01A             TRT01A          n              TRT01A               n
─────────────────────       ───────────────────            ──────────────────────
A  Placebo                  Placebo          86            Placebo               86
B  Xanomeline High Dose     Xanomeline High  84            Xanomeline Low Dose   84
C  Xanomeline Low Dose      Xanomeline Low   84            Xanomeline High Dose  84
                            ▲ alphabetical                 ▲ regulator-friendly
```

Without converting `TRT01A` to a factor with explicit `levels`, every TLG
column will land in the wrong order — `High Dose` will appear **before**
`Low Dose`, just as `PROC FREQ` with `ORDER=INTERNAL` sorts them.

The other reason: **regression reference category**. The first factor level is
the baseline group in any `lm()` / `glm()` / `coxph()` model. If you fit a
logistic regression of AE risk against treatment with the default character
encoding, R will pick `"Placebo"` alphabetically (often what you want —
lucky), but for `AESEV` (`"MILD"`, `"MODERATE"`, `"SEVERE"`), R picks
`"MILD"` alphabetically, which **is** the clinical reference — also lucky.
**Do not rely on luck.** Set the factor levels explicitly.

#### The 6 forcats functions you will actually use

| function | task | clinical example |
|---|---|---|
| `fct(x, levels = ...)` | Build a factor with **exact** level order | order `TRT01A` Placebo → Low → High before a TLG |
| `fct_recode()` | Rename levels (display labels) | `"Xanomeline High Dose"` → `"High Dose"` for narrow columns |
| `fct_collapse()` | Merge several levels into one | combine `"Xanomeline Low Dose"` + `"Xanomeline High Dose"` into `"Active"` |
| `fct_lump_n(n = 5)` | Keep the top N levels, lump the rest into `"Other"` | top-5 AE body systems table |
| `fct_infreq()` | Reorder levels by frequency, most common first | AE preferred-term listing sorted by occurrence |
| `fct_relevel()` | Move specific levels to specific positions | force `"Placebo"` to be the reference in a regression |

#### Where factors live in the pharmaverse workflow

```
              ADaM .xpt              factor conversion              TLG layer
              (character)            (right before display)         (tern, gtsummary,
                                                                     rtables, gt)
adsl$TRT01A   "Placebo"        ──▶   fct(TRT01A, levels = c(    ──▶ column header
              "Xanomeline...."         "Placebo",                    in correct order
                                       "Xanomeline Low Dose",
                                       "Xanomeline High Dose"))
```

**The rule:** Keep ADaM columns as **character** (XPT has no factor type).
Convert to a factor with **explicit `levels`** immediately before building a
TLG, model, or plot — never earlier, never in the ADaM derivation step.

### Exercise 7a — ordered treatment levels with fct()

Convert `TRT01A` in `adsl` to a factor with the correct clinical order:
Placebo → Low Dose → High Dose. Verify with `levels()` and `table()`.

```{r factor-create, exercise=TRUE, exercise.lines=14}
adsl_fct <- adsl |>
  mutate(
    TRT01A = fct(TRT01A,
                 levels = c("___",
                            "___",
                            "___"))
  )

levels(adsl_fct$TRT01A)
table(adsl_fct$TRT01A)
```

```{r factor-create-hint-1}
# levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose")
```

```{r factor-create-solution}
adsl_fct <- adsl |>
  mutate(
    TRT01A = fct(TRT01A,
                 levels = c("Placebo",
                            "Xanomeline Low Dose",
                            "Xanomeline High Dose"))
  )

levels(adsl_fct$TRT01A)
table(adsl_fct$TRT01A)
```

### Exercise 7b — ordered severity with fct() and cross-tabulation

Severity (`AESEV`) in `adae` is a character with three values: `"MILD"`,
`"MODERATE"`, `"SEVERE"`. Create a factor with the correct clinical order and
produce a cross-tabulation of severity vs. treatment arm for treatment-emergent AEs.

```{r factor-aesev, exercise=TRUE, exercise.lines=16}
adae |>
  filter(TRTEMFL == "Y") |>
  mutate(
    AESEV_F  = fct(AESEV,  levels = c("___", "___", "___")),
    TRT01A_F = fct(TRT01A, levels = c("Placebo",
                                      "Xanomeline Low Dose",
                                      "Xanomeline High Dose"))
  ) |>
  count(TRT01A_F, AESEV_F) |>
  pivot_wider(names_from = AESEV_F, values_from = n, values_fill = 0)
```

```{r factor-aesev-hint-1}
# levels = c("MILD", "MODERATE", "SEVERE")
# The pivot_wider at the end creates a severity-by-treatment cross-tab
```

```{r factor-aesev-solution}
adae |>
  filter(TRTEMFL == "Y") |>
  mutate(
    AESEV_F  = fct(AESEV,  levels = c("MILD", "MODERATE", "SEVERE")),
    TRT01A_F = fct(TRT01A, levels = c("Placebo",
                                      "Xanomeline Low Dose",
                                      "Xanomeline High Dose"))
  ) |>
  count(TRT01A_F, AESEV_F) |>
  pivot_wider(names_from = AESEV_F, values_from = n, values_fill = 0)
```

### Exercise 7c — fct_recode: shorten labels for TLG column headers

Rename the treatment levels so they fit in a narrow table column:

- `"Xanomeline Low Dose"` → `"Low Dose"`
- `"Xanomeline High Dose"` → `"High Dose"`
- `"Placebo"` stays as-is

```{r factor-recode, exercise=TRUE, exercise.lines=14}
adsl |>
  mutate(
    TRT01A_SHORT = factor(TRT01A) |>
      fct_recode(
        "___" = "Xanomeline Low Dose",
        "___" = "Xanomeline High Dose"
      )
  ) |>
  count(TRT01A, TRT01A_SHORT)
```

```{r factor-recode-hint-1}
# fct_recode(TRT01A, "Low Dose" = "Xanomeline Low Dose",
#                    "High Dose" = "Xanomeline High Dose")
```

```{r factor-recode-solution}
adsl |>
  mutate(
    TRT01A_SHORT = factor(TRT01A) |>
      fct_recode(
        "Low Dose"  = "Xanomeline Low Dose",
        "High Dose" = "Xanomeline High Dose"
      )
  ) |>
  count(TRT01A, TRT01A_SHORT)
```

### Exercise 7d — fct_lump_n: collapse rare AE body systems

An AE summary table should show the **top 5 body systems** by frequency and lump
all others into `"Other"`. Use `fct_lump_n()` on `AEBODSYS` and then order by
decreasing frequency with `fct_infreq()`.

```{r factor-lump, exercise=TRUE, exercise.lines=12}
adae |>
  filter(TRTEMFL == "Y") |>
  mutate(
    SOC = AEBODSYS |>
            fct_lump_n(n = ___, other_level = "Other") |>
            fct_infreq() |>
            fct_rev()
  ) |>
  count(SOC, sort = FALSE) |>
  print(n = Inf)
```

```{r factor-lump-hint-1}
# fct_lump_n(AEBODSYS, n = 5, other_level = "Other") keeps top 5 by frequency
# fct_infreq() orders levels by frequency (most frequent first)
# fct_rev() reverses so the most frequent appears at the top when plotted
```

```{r factor-lump-solution}
adae |>
  filter(TRTEMFL == "Y") |>
  mutate(
    SOC = AEBODSYS |>
            fct_lump_n(n = 5, other_level = "Other") |>
            fct_infreq() |>
            fct_rev()
  ) |>
  count(SOC, sort = FALSE) |>
  print(n = Inf)
```

### Exercise 7e — fct_collapse: merge active arms into a single category

For an exposure-adjusted analysis, you want to compare any-active treatment versus
placebo. Collapse `"Xanomeline Low Dose"` and `"Xanomeline High Dose"` into a
single level `"Active"`.

```{r factor-collapse, exercise=TRUE, exercise.lines=12}
adsl |>
  mutate(
    TRTGRP = factor(TRT01A) |>
      fct_collapse(
        "Active" = c("___", "___")
      )
  ) |>
  count(TRTGRP) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

```{r factor-collapse-hint-1}
# fct_collapse("Active" = c("Xanomeline Low Dose", "Xanomeline High Dose"))
# "Placebo" is not listed so it keeps its original name
```

```{r factor-collapse-solution}
adsl |>
  mutate(
    TRTGRP = factor(TRT01A) |>
      fct_collapse(
        "Active" = c("Xanomeline Low Dose", "Xanomeline High Dose")
      )
  ) |>
  count(TRTGRP) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

```{r quiz-factors, echo=FALSE}
quiz(
  caption = "forcats check",
  question(
    "Why do you typically NOT store factors in exported ADaM datasets (.xpt)?",
    answer("Factors are too slow for large datasets"),
    answer("XPT/SAS transport files don't have a factor type — everything is
           character or numeric", correct = TRUE),
    answer("admiral functions don't accept factor columns"),
    answer("Factors lose their levels when you use filter()"),
    allow_retry = TRUE
  ),
  question(
    "You apply fct_infreq(AEBODSYS) before building a frequency table. What does
     this do?",
    answer("Sorts rows of the data frame by frequency"),
    answer("Reorders the factor levels so the most frequent level comes first",
           correct = TRUE),
    answer("Removes levels with frequency below a threshold"),
    answer("Converts the factor back to a character column"),
    allow_retry = TRUE
  ),
  question(
    "fct_lump_n(x, n = 5) keeps the top 5 levels. What happens to the rest?",
    answer("They are dropped from the data frame"),
    answer("They are converted to NA"),
    answer("They are merged into an 'Other' level (name configurable via
           other_level)", correct = TRUE),
    answer("They remain as individual levels at the end of the factor"),
    allow_retry = TRUE
  )
)
```

---

## 8  End-to-end CDISC pipeline — safety summary with multiple datasets

### The task

Build a treatment-emergent AE safety summary that:

1. Filter `adae` to treatment-emergent AEs (`TRTEMFL == "Y"`)
2. Join treatment arm and duration from `adsl`
3. Compute per-subject AE counts and flag **serious early-onset** AEs
   (`AESER == "Y"` and `ASTDY <= 30`)
4. Summarise by treatment arm: count of AEs, count of serious AEs, count of
   subjects with ≥1 serious early-onset AE
5. Clean `AEDECOD` to title case for display
6. Order treatment arm as a factor in clinical order

```{r cdisc-pipeline, exercise=TRUE, exercise.lines=40}
library(lubridate)
library(stringr)
library(forcats)

# Step 1 & 2 — filter TE AEs and join adsl
ae_base <- adae |>
  filter(TRTEMFL == "___") |>
  left_join(
    adsl |> select(USUBJID, TRT01A, TRTDURD),
    by = "USUBJID"
  ) |>
  mutate(
    # Step 3a — clean AE term
    AEDECOD_DISPLAY = AEDECOD |> ___() |> ___(),
    # Step 3b — flag serious early-onset
    SER_EARLY_FL = if_else(AESER == "Y" & !is.na(ASTDY) & ASTDY <= 30,
                           "Y", "N")
  )

# Step 4 — subject-level: did they have ≥1 serious early-onset AE?
subj_ser_early <- ae_base |>
  group_by(USUBJID, TRT01A) |>
  summarise(ANY_SER_EARLY = if_else(any(SER_EARLY_FL == "Y"), "Y", "N"),
            .groups = "drop")

# Step 5 — treatment-arm summary
ae_base |>
  group_by(TRT01A) |>
  summarise(
    N_AE      = n(),
    N_SER_AE  = sum(AESER == "Y", na.rm = TRUE),
    .groups   = "drop"
  ) |>
  left_join(
    subj_ser_early |>
      group_by(TRT01A) |>
      summarise(N_SUBJ_SER_EARLY = sum(ANY_SER_EARLY == "Y"), .groups = "drop"),
    by = "TRT01A"
  ) |>
  # Step 6 — order treatment arm
  mutate(
    TRT01A = fct(TRT01A,
                 levels = c("Placebo",
                            "Xanomeline Low Dose",
                            "Xanomeline High Dose"))
  ) |>
  arrange(TRT01A)
```

```{r cdisc-pipeline-hint-1}
# TRTEMFL == "Y"
# str_to_title() |> str_squish() for AEDECOD_DISPLAY
# fct(TRT01A, levels = c("Placebo","Xanomeline Low Dose","Xanomeline High Dose"))
```

```{r cdisc-pipeline-solution}
library(lubridate)
library(stringr)
library(forcats)

ae_base <- adae |>
  filter(TRTEMFL == "Y") |>
  left_join(
    adsl |> select(USUBJID, TRT01A, TRTDURD),
    by = "USUBJID"
  ) |>
  mutate(
    AEDECOD_DISPLAY = AEDECOD |> str_to_title() |> str_squish(),
    SER_EARLY_FL    = if_else(AESER == "Y" & !is.na(ASTDY) & ASTDY <= 30,
                              "Y", "N")
  )

subj_ser_early <- ae_base |>
  group_by(USUBJID, TRT01A) |>
  summarise(ANY_SER_EARLY = if_else(any(SER_EARLY_FL == "Y"), "Y", "N"),
            .groups = "drop")

ae_base |>
  group_by(TRT01A) |>
  summarise(
    N_AE     = n(),
    N_SER_AE = sum(AESER == "Y", na.rm = TRUE),
    .groups  = "drop"
  ) |>
  left_join(
    subj_ser_early |>
      group_by(TRT01A) |>
      summarise(N_SUBJ_SER_EARLY = sum(ANY_SER_EARLY == "Y"), .groups = "drop"),
    by = "TRT01A"
  ) |>
  mutate(
    TRT01A = fct(TRT01A,
                 levels = c("Placebo",
                            "Xanomeline Low Dose",
                            "Xanomeline High Dose"))
  ) |>
  arrange(TRT01A)
```

---

## 9  Self-check Quiz

```{r final-quiz-06, echo=FALSE}
quiz(
  caption = "Lesson 06 — comprehensive self-check",

  # ── tidyr ────────────────────────────────────────────────────────────────────
  question(
    "You have adlb in long format and call pivot_wider(names_from = PARAMCD,
     values_from = c(AVAL, BASE)). For three PARAMCDs, how many new columns are
     produced?",
    answer("3 — one per PARAMCD"),
    answer("6 — two per PARAMCD (AVAL_ALT, BASE_ALT, AVAL_HGB, …)",
           correct = TRUE),
    answer("2 — AVAL and BASE"),
    answer("Depends on the number of visits"),
    allow_retry = TRUE
  ),

  question(
    "What does values_fill = list(AVAL = 0) do in pivot_wider()?",
    answer("Replaces missing values in the original AVAL column with 0"),
    answer("Fills the newly created wide-format cells that have no matching
           long-format row with 0", correct = TRUE),
    answer("Rounds all AVAL values to 0 decimal places"),
    answer("Sets the default AVAL for all rows before pivoting"),
    allow_retry = TRUE
  ),

  question(
    "complete(USUBJID, AVISIT) adds rows for all subject-visit combinations that
     were absent. What does fill() do AFTER complete()?",
    answer("Removes the rows that complete() just added"),
    answer("Fills NA values in other columns by propagating the last
           (or next) observed value within a group", correct = TRUE),
    answer("Fills missing AVAL values by linear interpolation"),
    answer("It is equivalent to replace_na()"),
    allow_retry = TRUE
  ),

  question(
    "In pivot_longer(), you set names_to = c('.value', 'VISIT') and
     names_pattern = '^(AVAL|CHG)_(\\\\d+)$'. What does .value do?",
    answer("It creates a literal column called '.value'"),
    answer("It signals that the matched piece should determine WHICH output
           column the data goes into — producing separate AVAL and CHG columns",
           correct = TRUE),
    answer("It drops the column name prefix entirely"),
    answer("It stores the column names in a metadata attribute"),
    allow_retry = TRUE
  ),

  # ── lubridate ────────────────────────────────────────────────────────────────
  question(
    "Translate SAS: intck('day', trtsdt, aestdt) to R:",
    answer("days(aestdt, trtsdt)"),
    answer("difftime(aestdt, trtsdt, units = 'days')"),
    answer("as.numeric(ymd(aestdt) - ymd(trtsdt))", correct = TRUE),
    answer("interval(trtsdt, aestdt) / days(1)"),
    allow_retry = TRUE
  ),

  question(
    "You need to compute the 30-day follow-up date after treatment end for every
     subject, and some treatment end dates fall on January 31. What is the safest
     R expression?",
    answer("TRTEDT + 30"),
    answer("TRTEDT + days(30)", correct = TRUE),
    answer("TRTEDT %m+% months(1)"),
    answer("ymd(TRTEDT) + month(1)"),
    allow_retry = TRUE
  ),

  question(
    "adsl$TRTSDT is already an R Date. Which of the following correctly computes
     treatment duration in completed days (inclusive of both endpoints)?",
    answer("as.numeric(TRTEDT - TRTSDT)"),
    answer("difftime(TRTEDT, TRTSDT)"),
    answer("as.numeric(TRTEDT - TRTSDT) + 1", correct = TRUE),
    answer("interval(TRTSDT, TRTEDT) / ddays(1)"),
    allow_retry = TRUE
  ),

  # ── stringr ──────────────────────────────────────────────────────────────────
  question(
    "What is the difference between str_trim() and str_squish()?",
    answer("str_trim() is faster; str_squish() is more accurate"),
    answer("str_trim() removes leading/trailing whitespace only; str_squish()
           also collapses multiple interior spaces to one", correct = TRUE),
    answer("str_squish() removes all whitespace including inside words"),
    answer("They are identical — str_squish() is just an alias"),
    allow_retry = TRUE
  ),

  question(
    "str_detect(AEBODSYS, regex('skin', ignore_case = TRUE)) returns TRUE for
     which of these values?",
    answer("'Nervous system disorders'", correct = TRUE),
    answer("'Gastrointestinal disorders'"),
    answer("'Musculoskeletal and connective tissue disorders'"),
    answer("'Blood and lymphatic system disorders'"),
    allow_retry = TRUE
  ),

  question(
    "You want to apply str_to_upper() to every character column in adsl whose
     name starts with 'TRT'. What is the correct dplyr call?",
    answer("mutate(TRT01A = str_to_upper(TRT01A), TRT01P = str_to_upper(TRT01P))"),
    answer("mutate(across(starts_with('TRT') & where(is.character),
           str_to_upper))", correct = TRUE),
    answer("rename_with(str_to_upper, starts_with('TRT'))"),
    answer("transform(adsl, TRT = str_to_upper)"),
    allow_retry = TRUE
  ),

  # ── forcats ──────────────────────────────────────────────────────────────────
  question(
    "When should you convert a character column to a factor in pharmaverse
     workflows?",
    answer("As early as possible, right after reading the data"),
    answer("Never — all pharmaverse packages require character columns"),
    answer("Immediately before building a TLG or fitting a model, not in stored
           ADaM datasets", correct = TRUE),
    answer("Only when the column has more than 5 unique values"),
    allow_retry = TRUE
  ),

  question(
    "fct_collapse(TRT01A, 'Active' = c('Xanomeline Low Dose',
     'Xanomeline High Dose')) produces a factor with how many levels?",
    answer("2 — Active and Placebo", correct = TRUE),
    answer("3 — unchanged"),
    answer("1 — everything is collapsed"),
    answer("It errors because 'Active' was not an existing level"),
    allow_retry = TRUE
  ),

  question(
    "You call fct_lump_n(AEBODSYS, n = 5) on a column with 22 distinct SOC
     values. What is the maximum number of distinct levels in the result?",
    answer("5"),
    answer("6 — the top 5 plus 'Other'", correct = TRUE),
    answer("22 — lump_n only reorders, it does not reduce"),
    answer("Depends on whether ties exist at the 5th position"),
    allow_retry = TRUE
  ),

  # ── cross-cutting / CDISC ────────────────────────────────────────────────────
  question(
    "In pharmaverseadam ADLB and ADVS, some rows have non-missing DTYPE values
     (e.g. 'MINIMUM', 'LOCF'). What should you do before most analyses?",
    answer("Keep them — they represent the primary analysis records"),
    answer("Filter them out with filter(is.na(DTYPE)) to retain only observed
           records", correct = TRUE),
    answer("Replace their AVAL with NA"),
    answer("Pivot them to a separate wide dataset"),
    allow_retry = TRUE
  ),

  question(
    "Which of the following tidyverse verbs is the R equivalent of the SAS
     RETAIN statement when filling in a subject's treatment arm across newly
     added visit rows after complete()?",
    answer("mutate(TRT01A = lag(TRT01A))"),
    answer("group_by(USUBJID) |> fill(TRT01A, .direction = 'downup') |>
           ungroup()", correct = TRUE),
    answer("replace_na(list(TRT01A = 'Placebo'))"),
    answer("coalesce(TRT01A, first(TRT01A))"),
    allow_retry = TRUE
  )
)
```

---

## 10  SAS → Tidyverse Cheat Sheet

### Reshaping

| SAS | R (tidyr) |
|---|---|
| `PROC TRANSPOSE out= prefix=` | `pivot_longer(cols=, names_to=, values_to=)` |
| `PROC TRANSPOSE id= var= out=` | `pivot_wider(names_from=, values_from=)` |
| `scan(var, 1, '_')` | `separate_wider_delim(col, delim='_', names=c(...))` |
| Build ISO date from parts | `unite('AESTDTC', c(YEAR,MONTH,DAY), sep='-')` |
| Ensure all category combinations | `complete(USUBJID, AVISIT)` |
| `RETAIN` (carry forward) | `group_by(id) |> fill(col) |> ungroup()` |

### Dates

| SAS | R (lubridate) |
|---|---|
| `input(var, yymmdd10.)` | `ymd(var)` |
| `put(date, date9.)` | `format(date, "%d%b%Y")` |
| `intck('day', d1, d2)` | `as.numeric(d2 - d1)` |
| `intck('day', d1, d2) + 1` (inclusive) | `as.numeric(d2 - d1) + 1` |
| `intnx('month', d, 3)` | `d %m+% months(3)` |
| `intnx('day',  d, 30)` | `d + days(30)` |
| `year(date)` | `year(date)` |
| `month(date)` | `month(date)` |
| `day(date)` | `day(date)` |

### Strings

| SAS | R (stringr) |
|---|---|
| `index(s, 'x') > 0` | `str_detect(s, 'x')` |
| `substr(s, 1, 3)` | `str_sub(s, 1, 3)` |
| `tranwrd(s, 'old', 'new')` | `str_replace_all(s, 'old', 'new')` |
| `upcase(s)` / `lowcase(s)` | `str_to_upper(s)` / `str_to_lower(s)` |
| `propcase(s)` | `str_to_title(s)` |
| `compbl(s)` | `str_squish(s)` |
| `strip(s)` | `str_trim(s)` |
| `put(n, z4.)` (zero-pad) | `str_pad(n, width=4, pad='0')` |
| `catx('-', a, b)` | `str_c(a, b, sep='-')` |
| `prxmatch('/pat/i', s) > 0` | `str_detect(s, regex('pat', ignore_case=TRUE))` |
| Apply to multiple columns | `mutate(across(where(is.character), str_squish))` |

### Factors

| SAS | R (forcats) |
|---|---|
| `PROC FORMAT` value statement | `fct(x, levels=c(...))` |
| Change format order | `fct_relevel(x, 'first_level')` |
| Rename a level | `fct_recode(x, 'new'='old')` |
| Merge levels | `fct_collapse(x, 'merged'=c('a','b'))` |
| Keep top N, lump rest | `fct_lump_n(x, n=5)` |
| Order by frequency | `fct_infreq(x)` |
| Reverse level order | `fct_rev(x)` |

---

## Key Takeaways

1. **Tidy data** — one row per observation, one column per variable. CDISC BDS is tidy (long format).
2. **`pivot_longer()`** = PROC TRANSPOSE wide→long; **`pivot_wider()`** = PROC TRANSPOSE long→wide. Always pre-filter DTYPE records before pivoting pharmaverseadam BDS datasets.
3. **`separate_wider_delim()`** splits combined codes; **`unite()`** assembles ISO 8601 dates from parts; **`complete()`** + **`fill()`** handle sparse visit structures — the R equivalent of SAS RETAIN and format-driven templates.
4. **`lubridate::ymd()`** parses CDISC ISO 8601 strings; Date subtraction gives days directly. CDISC duration = difference + 1 (inclusive endpoints). Use `%m+%` for safe month arithmetic.
5. **`stringr`** gives consistent `str_*` functions. Use `across()` to apply them to many columns at once.
6. **Factors** control TLG column/row order — convert character → factor at the TLG step, not in stored ADaMs. Use `fct_lump_n()` + `fct_infreq()` for frequency-ordered summaries.

**R foundations complete!** Module 2 begins with `{pharmaverseraw}` and `{sdtm.oak}`.

(`rmarkdown::run("07_pharmaverseraw_sdtm_intro.Rmd")`)
