---
title: "Lesson 07 — {pharmaverseraw}: Test EDC Data for Pharmaverse Examples"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    ace_theme: tomorrow
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Understand where pharmaverseraw sits in the data pipeline (raw → SDTM → ADaM).
  Explore realistic raw EDC data, compare it to real pharmaversesdtm SDTM output,
  manually preview every derivation sdtm.oak will later automate, perform cross-domain
  joins, compute clinical variables, and understand what annotated CRFs provide.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 07_pharmaverseraw.Rmd
# PURPOSE       : Exploration of pharmaversesdtm and pharmaverseadam reference datasets:
#                 SDTM domain structure, ADaM dataset conventions, and
#                 navigating CDISCPILOT01 raw study data.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(lubridate)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 60)

# ── 1. Raw datasets ───────────────────────────────────────────────────────────
# Curated teaching datasets. NOTE: the real {pharmaverseraw} package (v0.1.1)
# ships EDC-native column names (STUDY, IT.AETERM, IT.AESEV, no SITEID, …) that
# do NOT match this lesson's schema or exercises. We therefore always build the
# curated tribbles below so every exercise (count(SITEID), USUBJID derivation,
# referential-integrity joins) resolves regardless of installed packages.

{

  # ae_raw: mirrors pharmaverseraw AE structure (PATNUM not USUBJID, no AESEQ,
  # no AEDECOD / AEBODSYS, dates as plain ISO 8601 strings without 'DTC' suffix)
  ae_raw <- tribble(
    ~STUDYID,     ~PATNUM, ~SITEID, ~AETERM,                       ~AESTDT,       ~AEENDT,       ~AESEV,      ~AESER, ~AEOUT,                          ~AEREL,
    "TEST_STUDY",    1015,   "701",  "HEADACHE",                   "2020-07-15",  "2020-07-17",  "MILD",      "N",    "RECOVERED/RESOLVED",            "POSSIBLE",
    "TEST_STUDY",    1015,   "701",  "NAUSEA",                     "2020-08-02",  "2020-08-05",  "MODERATE",  "N",    "RECOVERED/RESOLVED",            "PROBABLE",
    "TEST_STUDY",    1023,   "701",  "DIZZINESS",                  "2020-07-20",  NA,            "MILD",      "N",    "NOT RECOVERED/NOT RESOLVED",    "POSSIBLE",
    "TEST_STUDY",    1023,   "701",  "APPLICATION SITE ERYTHEMA",  "2020-09-10",  "2020-09-12",  "MODERATE",  "N",    "RECOVERED/RESOLVED",            "PROBABLE",
    "TEST_STUDY",    1034,   "702",  "FALL",                       "2020-07-28",  "2020-07-28",  "SEVERE",    "Y",    "RECOVERED/RESOLVED",            "NOT RELATED",
    "TEST_STUDY",    1034,   "702",  "HEADACHE",                   "2020-08-14",  "2020-08-16",  "MILD",      "N",    "RECOVERED/RESOLVED",            "REMOTE",
    "TEST_STUDY",    1056,   "703",  "FATIGUE",                    "2020-10-01",  "2020-10-10",  "MILD",      "N",    "RECOVERED/RESOLVED",            "POSSIBLE",
    "TEST_STUDY",    1056,   "703",  "VOMITING",                   "2020-10-03",  "2020-10-04",  "SEVERE",    "Y",    "RECOVERED/RESOLVED",            "PROBABLE",
    "TEST_STUDY",    1078,   "703",  "PRURITUS",                   "2020-11-02",  "2020-11-15",  "MILD",      "N",    "RECOVERED/RESOLVED",            "PROBABLE",
    "TEST_STUDY",    1099,   "704",  "INSOMNIA",                   "2020-08-20",  "2020-09-01",  "MODERATE",  "N",    "RECOVERED/RESOLVED",            "POSSIBLE"
  )

  # dm_raw: mirrors pharmaverseraw DM structure (BRTHDAT not BRTHDTC, PATNUM not USUBJID)
  dm_raw <- tribble(
    ~STUDYID,     ~PATNUM, ~SITEID, ~BRTHDAT,      ~SEX, ~RACE,                          ~ETHNIC,                 ~ARM,                    ~RFSTDAT,
    "TEST_STUDY",    1015,   "701",  "1955-04-12",  "F",  "WHITE",                        "NOT HISPANIC OR LATINO", "Xanomeline High Dose",  "2020-07-01",
    "TEST_STUDY",    1023,   "701",  "1948-09-30",  "M",  "WHITE",                        "NOT HISPANIC OR LATINO", "Placebo",               "2020-07-01",
    "TEST_STUDY",    1034,   "702",  "1961-07-22",  "F",  "BLACK OR AFRICAN AMERICAN",    "NOT HISPANIC OR LATINO", "Xanomeline Low Dose",   "2020-07-01",
    "TEST_STUDY",    1056,   "703",  "1940-11-05",  "M",  "WHITE",                        "NOT HISPANIC OR LATINO", "Xanomeline High Dose",  "2020-07-01",
    "TEST_STUDY",    1078,   "703",  "1953-02-18",  "F",  "ASIAN",                        "NOT HISPANIC OR LATINO", "Placebo",               "2020-07-01"
    # Note: PATNUM 1099 is intentionally absent from dm_raw → referential integrity exercise
  )

  # ds_raw: disposition events
  ds_raw <- tribble(
    ~STUDYID,     ~PATNUM, ~SITEID, ~DSTERM,                          ~DSSTDT,       ~DSCAT,
    "TEST_STUDY",    1015,   "701",  "COMPLETED",                     "2021-01-15",  "DISPOSITION EVENT",
    "TEST_STUDY",    1023,   "701",  "ADVERSE EVENT",                 "2020-10-05",  "DISPOSITION EVENT",
    "TEST_STUDY",    1034,   "702",  "COMPLETED",                     "2021-01-15",  "DISPOSITION EVENT",
    "TEST_STUDY",    1056,   "703",  "SUBJECT/GUARDIAN DECISION",     "2020-11-20",  "DISPOSITION EVENT",
    "TEST_STUDY",    1078,   "703",  "COMPLETED",                     "2021-01-15",  "DISPOSITION EVENT",
    "TEST_STUDY",    1015,   "701",  "PROTOCOL VISIT COMPLETED",      "2020-07-01",  "PROTOCOL MILESTONE",
    "TEST_STUDY",    1023,   "701",  "PROTOCOL VISIT COMPLETED",      "2020-07-01",  "PROTOCOL MILESTONE",
    "TEST_STUDY",    1034,   "702",  "PROTOCOL VISIT COMPLETED",      "2020-07-01",  "PROTOCOL MILESTONE"
  )
}

# ── 2. SDTM reference — real pharmaversesdtm output ──────────────────────────
# pharmaversesdtm is installed (v1.3.1). These are the CDISC pilot SDTM datasets
# that represent the TARGET structure raw data must be mapped to.
library(pharmaversesdtm)

ae_sdtm_real <- pharmaversesdtm::ae   # 1,191 rows, 35 columns — full AE SDTM
dm_sdtm_real <- pharmaversesdtm::dm   # 306 rows,  25 columns — full DM SDTM
ds_sdtm_real <- pharmaversesdtm::ds   # disposition
suppae_real  <- pharmaversesdtm::suppae  # supplemental AE qualifiers

# ── 3. Small illustrative ae_sdtm mirroring what sdtm.oak produces from ae_raw ──
# (Same study as ae_raw — kept for the "before vs after" comparison exercises)
ae_sdtm <- tribble(
  ~STUDYID,      ~DOMAIN, ~USUBJID,               ~AESEQ, ~AETERM,                       ~AEDECOD,                      ~AEBODSYS,                              ~AESTDTC,      ~AEENDTC,      ~AESEV,      ~AESER, ~AEOUT,                       ~AEREL,
  "TEST_STUDY",  "AE",    "TEST_STUDY-701-1015",    1,     "HEADACHE",                    "Headache",                    "Nervous system disorders",             "2020-07-15",  "2020-07-17",  "MILD",      "N",    "RECOVERED/RESOLVED",         "POSSIBLE",
  "TEST_STUDY",  "AE",    "TEST_STUDY-701-1015",    2,     "NAUSEA",                      "Nausea",                      "Gastrointestinal disorders",           "2020-08-02",  "2020-08-05",  "MODERATE",  "N",    "RECOVERED/RESOLVED",         "PROBABLE",
  "TEST_STUDY",  "AE",    "TEST_STUDY-701-1023",    1,     "DIZZINESS",                   "Dizziness",                   "Nervous system disorders",             "2020-07-20",  NA,            "MILD",      "N",    "NOT RECOVERED/NOT RESOLVED", "POSSIBLE",
  "TEST_STUDY",  "AE",    "TEST_STUDY-701-1023",    2,     "APPLICATION SITE ERYTHEMA",   "Application site erythema",   "General disorders and admin site cond","2020-09-10",  "2020-09-12",  "MODERATE",  "N",    "RECOVERED/RESOLVED",         "PROBABLE",
  "TEST_STUDY",  "AE",    "TEST_STUDY-702-1034",    1,     "FALL",                        "Fall",                        "Injury, poisoning and proc complications","2020-07-28","2020-07-28",  "SEVERE",    "Y",    "RECOVERED/RESOLVED",         "NOT RELATED",
  "TEST_STUDY",  "AE",    "TEST_STUDY-702-1034",    2,     "HEADACHE",                    "Headache",                    "Nervous system disorders",             "2020-08-14",  "2020-08-16",  "MILD",      "N",    "RECOVERED/RESOLVED",         "REMOTE",
  "TEST_STUDY",  "AE",    "TEST_STUDY-703-1056",    1,     "FATIGUE",                     "Fatigue",                     "General disorders and admin site cond","2020-10-01",  "2020-10-10",  "MILD",      "N",    "RECOVERED/RESOLVED",         "POSSIBLE",
  "TEST_STUDY",  "AE",    "TEST_STUDY-703-1056",    2,     "VOMITING",                    "Vomiting",                    "Gastrointestinal disorders",           "2020-10-03",  "2020-10-04",  "SEVERE",    "Y",    "RECOVERED/RESOLVED",         "PROBABLE",
  "TEST_STUDY",  "AE",    "TEST_STUDY-703-1078",    1,     "PRURITUS",                    "Pruritus",                    "Skin and subcutaneous tissue disorders","2020-11-02",  "2020-11-15",  "MILD",      "N",    "RECOVERED/RESOLVED",         "PROBABLE"
)
```

## Overview

### The pharmaverse data chain

`{pharmaverseraw}` is the **starting point** of the pharmaverse test data chain — simulated EDC output that feeds directly into `{sdtm.oak}`:

```
{pharmaverseraw}      ← you are here (Lesson 07)
   │  raw EDC data
   │  via {sdtm.oak}  ← Lessons 08–10
   ▼
{pharmaversesdtm}     ← SDTM datasets  (used live in this lesson)
   │
   │  via {admiral}   ← Lessons 11+
   ▼
{pharmaverseadam}     ← ADaM datasets
   │
   │  via {tfrmt/rtables}
   ▼
   TLFs (Tables, Listings, Figures)
```

This lesson uses **two complementary data sources**:

- **`ae_raw`, `dm_raw`, `ds_raw`** — raw EDC-style datasets (real `{pharmaverseraw}` if installed, otherwise realistic fallback tribbles mirroring its structure)
- **`ae_sdtm_real`, `dm_sdtm_real`** — actual `{pharmaversesdtm}` SDTM output from the CDISC pilot study, loaded live so you can compare real SDTM structure against raw

> **Run this to see which source is active:**
> ```r
> if (requireNamespace("pharmaverseraw", quietly=TRUE)) "pharmaverseraw" else "fallback tribbles"
> ```

---

```{r quiz-pipeline, echo=FALSE}
quiz(
  caption = "Pipeline orientation",
  question(
    "In the pharmaverse test data chain, what does {pharmaverseraw} feed into?",
    answer("{admiral} directly"),
    answer("{sdtm.oak} to produce SDTM datasets", correct = TRUE),
    answer("{pharmaverseadam} directly"),
    answer("{tfrmt} for table generation"),
    allow_retry = TRUE
  ),
  question(
    "pharmaverseraw is best described as:",
    answer("A package that builds SDTM datasets automatically"),
    answer("A data package — it bundles raw datasets and annotated CRFs with minimal computation",
           correct = TRUE),
    answer("A validation framework for SDTM conformance checking"),
    answer("A replacement for {pharmaversesdtm}"),
    allow_retry = TRUE
  ),
  question(
    "In this lesson, pharmaversesdtm is used alongside pharmaverseraw. Why?",
    answer("They contain the same datasets — one is just a backup of the other"),
    answer("pharmaversesdtm provides the SDTM target structure, so you can compare
           what raw data looks like versus what a finished SDTM dataset looks like",
           correct = TRUE),
    answer("pharmaversesdtm must be loaded before pharmaverseraw can function"),
    answer("pharmaversesdtm supplies the MedDRA coding tables"),
    allow_retry = TRUE
  )
)
```

---

## 1  Design philosophy — agnostic by construction

### EDC-agnostic

Real raw data comes from Veeva Vault, Medidata Rave, Castor, REDCap, and dozens of others — each with its own column naming and quirks. `pharmaverseraw` deliberately uses naming conventions that **don't favour any specific EDC**. What you learn here applies to any raw data source.

### Standards-agnostic

Some columns follow **CDASH** (Clinical Data Acquisition Standards Harmonization — the CDISC standard for CRF design), while others don't. This mirrors real-world variability: your sponsor might have CDASH-conformant CRFs for some pages and bespoke naming for others. The data is *intentionally* a bit inconsistent — it is a realistic playground, not a clean reference.

```{r quiz-philosophy, echo=FALSE}
quiz(
  caption = "Design philosophy",
  question(
    "Why is pharmaverseraw deliberately EDC-agnostic?",
    answer("To make it compatible only with Medidata Rave"),
    answer("So the SDTM mapping patterns you learn apply generically to any EDC,
           not just one system", correct = TRUE),
    answer("Because no CDISC standard governs raw data"),
    answer("To avoid licensing issues with EDC vendors"),
    allow_retry = TRUE
  ),
  question(
    "CDASH stands for:",
    answer("Clinical Data Analysis and Submission Harmonization"),
    answer("Clinical Data Acquisition Standards Harmonization — the CDISC standard
           for CRF design and raw variable naming", correct = TRUE),
    answer("Controlled Dose Assignment for Standard Hosts"),
    answer("Clinical Domain Attribute Schema Hierarchy"),
    allow_retry = TRUE
  ),
  question(
    "A sponsor's raw AE data uses the column name 'ADVERSE_REACTION_START' instead
     of the CDASH-recommended 'AESTDAT'. Which pharmaverseraw design principle
     does this illustrate?",
    answer("EDC-agnostic — real raw data uses whatever naming the EDC enforces,
           so mapping code must handle any name", correct = TRUE),
    answer("Standards-agnostic means all column names are randomised"),
    answer("CDASH forbids 'ADVERSE_REACTION_START' as a column name"),
    answer("The data is wrong and must be corrected before sdtm.oak can run"),
    allow_retry = TRUE
  )
)
```

---

## 2  Exploring the raw datasets

### Available domains

| Raw dataset | Maps to SDTM | Description |
|---|---|---|
| `dm_raw` | DM | Demographics — one row per subject |
| `ae_raw` | AE | Adverse Events — one row per AE per subject |
| `ds_raw` | DS | Disposition events |
| `ec_raw`, `ex_raw` | EC, EX | Exposure (planned / actual) |

The naming convention is consistent: `<domain>_raw`. Check installed versions with
`data(package = "pharmaverseraw")`.

### Exercise 2a — structural contrast: raw vs real SDTM

Run the code below to compare the column count and names of `ae_raw` (raw EDC) against
the real `pharmaversesdtm::ae` dataset. Note which columns exist only in SDTM.

```{r explore-structural, exercise=TRUE, exercise.lines=16}
cat("=== ae_raw: EDC columns ===\n")
cat(sprintf("Rows: %d   Columns: %d\n", nrow(ae_raw), ncol(ae_raw)))
names(ae_raw)

cat("\n=== pharmaversesdtm::ae: SDTM columns ===\n")
cat(sprintf("Rows: %d   Columns: %d\n", nrow(ae_sdtm_real), ncol(ae_sdtm_real)))

cat("\nColumns in SDTM AE that do NOT exist in raw (must be derived/coded):\n")
setdiff(names(ae_sdtm_real), names(ae_raw))

cat("\nColumns in raw that do NOT appear in SDTM AE (EDC-internal, dropped):\n")
setdiff(names(ae_raw), names(ae_sdtm_real))
```

```{r explore-structural-hint-1}
# Just run it and observe carefully.
# Key SDTM columns absent from raw:
#   DOMAIN (hardcode), USUBJID (derive), AESEQ (derive),
#   AEDECOD / AEBODSYS / AESOC (MedDRA external coding)
# Key raw columns absent from SDTM:
#   PATNUM (replaced by USUBJID), STUDYID/SITEID embedded in USUBJID,
#   AESTDT / AEENDT renamed to AESTDTC / AEENDTC
```

### Exercise 2b — inspect record counts and key variables

How many AE records exist per site in `ae_raw`? Then, for comparison, show the top
5 most common AE preferred terms (`AEDECOD`) in the real SDTM dataset.

```{r count-ae, exercise=TRUE, exercise.lines=12}
# Raw: AEs per site
cat("Raw AEs per site:\n")
ae_raw |> count(SITEID, sort = TRUE) |> print()

# Real SDTM: top 5 preferred terms
cat("\nReal pharmaversesdtm::ae — top 5 AE preferred terms:\n")
ae_sdtm_real |>
  count(___, sort = TRUE) |>
  slice_head(n = 5)
```

```{r count-ae-hint-1}
# count(AEDECOD, sort = TRUE) on ae_sdtm_real
```

```{r count-ae-solution}
ae_raw |> count(SITEID, sort = TRUE)

ae_sdtm_real |>
  count(AEDECOD, sort = TRUE) |>
  slice_head(n = 5)
```

### Exercise 2c — severity distribution: raw vs real SDTM

Compare the severity profile between `ae_raw` (raw `AESEV`) and the real
`pharmaversesdtm::ae` (also `AESEV`). Do the severity distributions look similar?
What does this tell you about the representativeness of the test data?

```{r severity-compare, exercise=TRUE, exercise.lines=14}
cat("=== ae_raw severity distribution ===\n")
ae_raw |>
  count(AESEV) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  print()

cat("\n=== pharmaversesdtm::ae severity distribution ===\n")
ae_sdtm_real |>
  count(___) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  print()
```

```{r severity-compare-hint-1}
# count(AESEV) on ae_sdtm_real
```

```{r severity-compare-solution}
ae_raw |>
  count(AESEV) |>
  mutate(pct = round(100 * n / sum(n), 1))

ae_sdtm_real |>
  count(AESEV) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

---

## 3  The four mapping patterns

Every raw → SDTM variable transformation falls into one of four categories:

| Pattern | Description | AE example |
|---|---|---|
| **Hardcode** | Same value for every row | `DOMAIN = "AE"` |
| **Direct map** | Copy/rename from source column | `AETERM` → `AETERM`; `AESTDT` → `AESTDTC` |
| **Derive** | Compute from one or more source columns | `USUBJID` from `STUDYID + SITEID + PATNUM`; `AESEQ` from `row_number()` |
| **External** | From an outside process not present in raw | `AEDECOD` / `AEBODSYS` from MedDRA coding |

```{r quiz-patterns, echo=FALSE}
quiz(
  caption = "Mapping patterns",
  question(
    "AEDECOD (the MedDRA preferred term for an AE) is an example of which pattern?",
    answer("Hardcode — it's always 'Adverse Event'"),
    answer("Direct map — it's the same as AETERM"),
    answer("External — it comes from medical coding (MedDRA), not from the raw CRF data",
           correct = TRUE),
    answer("Derive — it is computed from AETERM using a formula"),
    allow_retry = TRUE
  ),
  question(
    "DOMAIN = 'AE' for every row in the AE dataset is which pattern?",
    answer("Direct map"),
    answer("Hardcode — the same value is assigned to every row regardless of input",
           correct = TRUE),
    answer("Derive"),
    answer("External"),
    allow_retry = TRUE
  ),
  question(
    "AESEQ — the within-subject sequence number — is which pattern, and why?",
    answer("Hardcode — it's always 1"),
    answer("Direct map — it exists in the raw data as a row counter"),
    answer("Derive — it must be computed (e.g. via row_number()) because it doesn't
           exist anywhere in the raw EDC extract", correct = TRUE),
    answer("External — it comes from the data management system"),
    allow_retry = TRUE
  ),
  question(
    "A sponsor's raw DM dataset has a column 'TRT_ASSIGNED'. In SDTM DM this
     becomes ARM. Which mapping pattern is this?",
    answer("Hardcode"),
    answer("Direct map — copy the value and rename the column", correct = TRUE),
    answer("Derive"),
    answer("External"),
    allow_retry = TRUE
  )
)
```

---

## 4  Manually building SDTM from raw data

In Lesson 08 you'll use `{sdtm.oak}` functions to automate these steps. Here you do them by hand — building the mental model that makes `sdtm.oak` intuitive.

### Exercise 4a — derive USUBJID

In raw data the subject is identified by `PATNUM` (an EDC integer). SDTM requires
`USUBJID` in the format `STUDYID-SITEID-PATNUM`. Derive it using `str_c()` and verify
it matches the `USUBJID` pattern seen in `ae_sdtm_real`.

```{r derive-usubjid, exercise=TRUE, exercise.lines=12}
# Derive from ae_raw
ae_with_id <- ae_raw |>
  mutate(
    USUBJID = str_c(___, ___, ___, sep = "-")
  ) |>
  select(USUBJID, PATNUM, SITEID, AETERM)

ae_with_id |> slice_head(n = 4)

# Compare format to real SDTM
cat("\nReal SDTM USUBJID examples:\n")
ae_sdtm_real |> distinct(USUBJID) |> slice_head(n = 4)
```

```{r derive-usubjid-hint-1}
# str_c(STUDYID, SITEID, PATNUM, sep = "-")
# Notice: real SDTM uses "01-701-1015" pattern (study prefix-site-subject)
# Our test study uses "TEST_STUDY-701-1015" — same logic, different study name
```

```{r derive-usubjid-solution}
ae_with_id <- ae_raw |>
  mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
  select(USUBJID, PATNUM, SITEID, AETERM)

ae_with_id |> slice_head(n = 4)

ae_sdtm_real |> distinct(USUBJID) |> slice_head(n = 4)
```

### Exercise 4b — derive AESEQ with tie-breaking

`AESEQ` is the within-subject AE sequence number. It must be unique per `USUBJID` and
is sorted by AE start date. Add a secondary sort on `AETERM` to break ties when two
AEs start on the same day.

```{r derive-aeseq, exercise=TRUE, exercise.lines=14}
ae_raw |>
  mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
  arrange(USUBJID, AESTDT, ___) |>       # secondary sort for tie-breaking
  group_by(USUBJID) |>
  mutate(AESEQ = row_number()) |>
  ungroup() |>
  select(USUBJID, AESEQ, AETERM, AESTDT) |>
  arrange(USUBJID, AESEQ)
```

```{r derive-aeseq-hint-1}
# arrange(USUBJID, AESTDT, AETERM) — alphabetical AETERM breaks same-date ties
# This is a production best practice: always specify a deterministic secondary sort
```

```{r derive-aeseq-solution}
ae_raw |>
  mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
  arrange(USUBJID, AESTDT, AETERM) |>
  group_by(USUBJID) |>
  mutate(AESEQ = row_number()) |>
  ungroup() |>
  select(USUBJID, AESEQ, AETERM, AESTDT) |>
  arrange(USUBJID, AESEQ)
```

### Exercise 4c — hardcode DOMAIN and rename date columns in one pipeline

Combine two mapping patterns: hardcode `DOMAIN = "AE"` and rename the raw date
columns to SDTM convention (`AESTDT` → `AESTDTC`, `AEENDT` → `AEENDTC`). The `C`
suffix signals that SDTM stores dates as collected **character** strings in ISO 8601.

```{r hardcode-rename, exercise=TRUE, exercise.lines=10}
ae_raw |>
  mutate(DOMAIN = "___") |>
  rename(
    AESTDTC = ___,
    AEENDTC = ___
  ) |>
  select(DOMAIN, AETERM, AESTDTC, AEENDTC, AESEV) |>
  slice_head(n = 4)
```

```{r hardcode-rename-hint-1}
# DOMAIN = "AE"
# rename(AESTDTC = AESTDT, AEENDTC = AEENDT)
```

```{r hardcode-rename-solution}
ae_raw |>
  mutate(DOMAIN = "AE") |>
  rename(AESTDTC = AESTDT, AEENDTC = AEENDT) |>
  select(DOMAIN, AETERM, AESTDTC, AEENDTC, AESEV) |>
  slice_head(n = 4)
```

### Exercise 4d — cross-domain join: enrich AEs with demographics

In the raw data, `ae_raw` and `dm_raw` are separate CRF pages joined by `PATNUM`
and `SITEID`. Join them to bring subject demographics (birth date, sex, treatment arm)
into the AE dataset — the foundation for any treatment-stratified AE analysis.

```{r cross-domain-join, exercise=TRUE, exercise.lines=16}
ae_raw |>
  mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
  left_join(
    dm_raw |>
      mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
      select(USUBJID, BRTHDAT, SEX, ARM),
    by = "___"
  ) |>
  select(USUBJID, AETERM, AESTDT, AESEV, SEX, ARM, BRTHDAT) |>
  arrange(USUBJID, AESTDT)
```

```{r cross-domain-join-hint-1}
# by = "USUBJID"
# The join fails gracefully for subjects in ae_raw with no dm_raw record
# (PATNUM 1099) — those rows get NA for demographics.
```

```{r cross-domain-join-solution}
ae_raw |>
  mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
  left_join(
    dm_raw |>
      mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
      select(USUBJID, BRTHDAT, SEX, ARM),
    by = "USUBJID"
  ) |>
  select(USUBJID, AETERM, AESTDT, AESEV, SEX, ARM, BRTHDAT) |>
  arrange(USUBJID, AESTDT)
```

### Exercise 4e — compute AGE at AE onset

Extend the cross-domain join: compute `AGE_AT_AE` — the subject's age in completed
years at the time of each AE start, derived from `BRTHDAT` and `AESTDT`.
This is a realistic clinical programming task that combines demographic and event data.

```{r age-at-ae, exercise=TRUE, exercise.lines=18}
ae_raw |>
  mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
  left_join(
    dm_raw |>
      mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
      select(USUBJID, BRTHDAT, ARM),
    by = "USUBJID"
  ) |>
  filter(!is.na(BRTHDAT)) |>
  mutate(
    AESTDT_D  = ymd(AESTDT),
    BRTHDAT_D = ymd(BRTHDAT),
    AGE_AT_AE = floor(as.numeric(___ - ___) / 365.25)
  ) |>
  select(USUBJID, ARM, AETERM, AESTDT, AGE_AT_AE, AESEV) |>
  arrange(USUBJID, AESTDT)
```

```{r age-at-ae-hint-1}
# floor(as.numeric(AESTDT_D - BRTHDAT_D) / 365.25)
# floor() ensures completed years (same as SAS floor(intck('day',birth,event)/365.25))
```

```{r age-at-ae-solution}
ae_raw |>
  mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
  left_join(
    dm_raw |>
      mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
      select(USUBJID, BRTHDAT, ARM),
    by = "USUBJID"
  ) |>
  filter(!is.na(BRTHDAT)) |>
  mutate(
    AESTDT_D  = ymd(AESTDT),
    BRTHDAT_D = ymd(BRTHDAT),
    AGE_AT_AE = floor(as.numeric(AESTDT_D - BRTHDAT_D) / 365.25)
  ) |>
  select(USUBJID, ARM, AETERM, AESTDT, AGE_AT_AE, AESEV) |>
  arrange(USUBJID, AESTDT)
```

### Exercise 4f — full SDTM AE shell pipeline

Combine all mapping steps into one pipeline: USUBJID, DOMAIN, AESEQ, date rename,
and AEREL (already present as a direct map). This is exactly what `sdtm.oak` will
automate in Lesson 08 — one function call per step.

```{r ae-sdtm-manual, exercise=TRUE, exercise.lines=22}
ae_raw |>
  # Derive USUBJID
  mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-")) |>
  # Hardcode DOMAIN
  mutate(DOMAIN = "AE") |>
  # Derive AESEQ with deterministic sort
  arrange(USUBJID, AESTDT, AETERM) |>
  group_by(USUBJID) |>
  mutate(AESEQ = row_number()) |>
  ungroup() |>
  # Rename date columns to SDTM DTC convention
  rename(AESTDTC = AESTDT, AEENDTC = AEENDT) |>
  # Select SDTM columns (AEDECOD / AEBODSYS excluded — come from MedDRA coding)
  select(STUDYID, DOMAIN, USUBJID, AESEQ,
         AETERM, AESTDTC, AEENDTC, AESEV, AESER, AEREL, AEOUT) |>
  arrange(USUBJID, AESEQ)
```

```{r ae-sdtm-manual-hint-1}
# Just run it — all pieces are assembled.
# Notice what is still MISSING vs ae_sdtm_real:
#   AEDECOD, AEBODSYS, AESOC, AELLTCD, AEPTCD, AEHLT etc. (MedDRA coding)
# sdtm.oak provides hardcode_ct(), assign_no_ct(), oak_id_vars()
# for each step above. Lesson 08 introduces each one.
```

---

## 5  Date quality and referential integrity

### Exercise 5a — validate date format and detect partial dates

Raw dates are not always well-formed ISO 8601. Use `str_detect()` to classify
`AESTDT` values as full dates (`YYYY-MM-DD`), partial dates (`YYYY-MM`), or invalid.
Then parse only the full dates and flag the rest.

```{r date-check, exercise=TRUE, exercise.lines=16}
ae_raw |>
  mutate(
    DATE_TYPE = case_when(
      str_detect(AESTDT, "^\\d{4}-\\d{2}-\\d{2}$") ~ "FULL",
      str_detect(AESTDT, "^\\d{4}-\\d{2}$")        ~ "PARTIAL_YM",
      str_detect(AESTDT, "^\\d{4}$")                ~ "PARTIAL_Y",
      TRUE                                           ~ "INVALID"
    ),
    AESTDTC_D = if_else(DATE_TYPE == "FULL", ymd(AESTDT), NA)
  ) |>
  select(PATNUM, AETERM, AESTDT, DATE_TYPE, AESTDTC_D)
```

```{r date-check-hint-1}
# ^\\d{4}-\\d{2}-\\d{2}$  matches exactly YYYY-MM-DD
# ^\\d{4}-\\d{2}$         matches YYYY-MM (partial — day missing)
# CDISC ISO 8601 allows partial dates: "2020-07" is a valid AESTDTC value
```

### Exercise 5b — compute AE duration in days

For AEs where both start and end date are present, compute `AEDUR` — the number of
days the AE lasted. Use the CDISC inclusive convention: end − start + 1.
Then summarise the mean and max duration by severity.

```{r ae-duration, exercise=TRUE, exercise.lines=14}
ae_raw |>
  filter(!is.na(AEENDT)) |>
  mutate(
    AESTDT_D = ymd(AESTDT),
    AEENDT_D = ymd(AEENDT),
    AEDUR    = as.numeric(___ - ___) + 1L
  ) |>
  group_by(AESEV) |>
  summarise(
    N        = n(),
    MEAN_DUR = round(mean(AEDUR), 1),
    MAX_DUR  = max(AEDUR),
    .groups  = "drop"
  )
```

```{r ae-duration-hint-1}
# as.numeric(AEENDT_D - AESTDT_D) + 1L  — inclusive of both endpoints
```

```{r ae-duration-solution}
ae_raw |>
  filter(!is.na(AEENDT)) |>
  mutate(
    AESTDT_D = ymd(AESTDT),
    AEENDT_D = ymd(AEENDT),
    AEDUR    = as.numeric(AEENDT_D - AESTDT_D) + 1L
  ) |>
  group_by(AESEV) |>
  summarise(
    N        = n(),
    MEAN_DUR = round(mean(AEDUR), 1),
    MAX_DUR  = max(AEDUR),
    .groups  = "drop"
  )
```

### Exercise 5c — referential integrity check

A subject present in `ae_raw` but absent from `dm_raw` is a data quality issue —
you cannot create a complete SDTM AE record without the demographics to derive
USUBJID and populate the ISM columns. Identify any orphaned AE subjects.

```{r referential-integrity, exercise=TRUE, exercise.lines=14}
ae_subjects <- ae_raw |> distinct(STUDYID, SITEID, PATNUM)
dm_subjects <- dm_raw |> distinct(STUDYID, SITEID, PATNUM)

# Subjects with AEs but NO demographics
orphaned <- anti_join(ae_subjects, dm_subjects,
                      by = c("___", "___", "___"))

cat("Subjects in ae_raw with no matching dm_raw record:\n")
print(orphaned)

cat("\nAEs that cannot be fully mapped to SDTM without a DM record:\n")
ae_raw |>
  semi_join(orphaned, by = c("STUDYID", "SITEID", "PATNUM")) |>
  select(STUDYID, SITEID, PATNUM, AETERM, AESEV)
```

```{r referential-integrity-hint-1}
# anti_join(ae_subjects, dm_subjects, by = c("STUDYID", "SITEID", "PATNUM"))
# PATNUM 1099 is in ae_raw but deliberately absent from dm_raw —
# this represents a real DM data issue that must be resolved before SDTM mapping.
```

```{r referential-integrity-solution}
ae_subjects <- ae_raw |> distinct(STUDYID, SITEID, PATNUM)
dm_subjects <- dm_raw |> distinct(STUDYID, SITEID, PATNUM)

orphaned <- anti_join(ae_subjects, dm_subjects,
                      by = c("STUDYID", "SITEID", "PATNUM"))

orphaned

ae_raw |>
  semi_join(orphaned, by = c("STUDYID", "SITEID", "PATNUM")) |>
  select(STUDYID, SITEID, PATNUM, AETERM, AESEV)
```

---

## 6  dm_raw — deriving analysis-ready variables

Rather than simply renaming columns, clinical programmers derive new variables
from demographic raw data. Two common derivations: age groups and the ISO 8601
birth date column.

### Exercise 6a — derive AGEGR1 from birth date

`dm_raw$BRTHDAT` gives the birth date. Using a study reference date of `"2020-07-01"`
(first dosing date), compute `AGE` in completed years, then classify into `AGEGR1`
using the Alzheimer's trial grouping (`<65` / `65-80` / `>80`).

```{r dm-agegr1, exercise=TRUE, exercise.lines=16}
study_start <- ymd("2020-07-01")

dm_raw |>
  mutate(
    USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-"),
    AGE     = floor(as.numeric(study_start - ymd(BRTHDAT)) / 365.25),
    AGEGR1  = case_when(
      AGE <  65           ~ "___",
      AGE <= 80           ~ "___",
      TRUE                ~ "___"
    )
  ) |>
  select(USUBJID, BRTHDAT, AGE, AGEGR1, SEX, ARM)
```

```{r dm-agegr1-hint-1}
# case_when:
#   AGE < 65  ~ "<65"
#   AGE <= 80 ~ "65-80"
#   TRUE      ~ ">80"
# Compare: pharmaverseadam::adsl$AGEGR1 uses exactly this grouping
```

```{r dm-agegr1-solution}
study_start <- ymd("2020-07-01")

dm_raw |>
  mutate(
    USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-"),
    AGE     = floor(as.numeric(study_start - ymd(BRTHDAT)) / 365.25),
    AGEGR1  = case_when(
      AGE <  65 ~ "<65",
      AGE <= 80 ~ "65-80",
      TRUE      ~ ">80"
    )
  ) |>
  select(USUBJID, BRTHDAT, AGE, AGEGR1, SEX, ARM)
```

### Exercise 6b — rename to SDTM conventions and add hardcoded variables

Complete a partial DM SDTM shell: derive USUBJID, rename `BRTHDAT` → `BRTHDTC`,
rename `RFSTDAT` → `RFSTDTC`, hardcode `DOMAIN = "DM"`, and add `AGEU = "YEARS"`.
Then compare your output column names to `dm_sdtm_real`.

```{r dm-sdtm-shell, exercise=TRUE, exercise.lines=18}
dm_shell <- dm_raw |>
  mutate(
    USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-"),
    DOMAIN  = "___",
    AGEU    = "___"
  ) |>
  rename(
    BRTHDTC = BRTHDAT,
    RFSTDTC = RFSTDAT
  ) |>
  select(STUDYID, DOMAIN, USUBJID, RFSTDTC, BRTHDTC, AGEU, SEX, RACE, ETHNIC, ARM)
  # Note: AGE is derived from BRTHDAT (see 6a); it is not a raw dm_raw column,
  # so it is intentionally omitted here.

# Columns in real SDTM DM not yet in our shell:
setdiff(names(dm_sdtm_real), names(dm_shell))
```

```{r dm-sdtm-shell-hint-1}
# DOMAIN = "DM", AGEU = "YEARS"
# The select() needs: AGE is derived in 6a but not present in dm_raw directly
# Simplification: select without AGE for now — it's derived separately
# setdiff() will show what SDTM DM has that we haven't mapped yet
```

```{r dm-sdtm-shell-solution}
dm_shell <- dm_raw |>
  mutate(
    USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-"),
    DOMAIN  = "DM",
    AGEU    = "YEARS"
  ) |>
  rename(
    BRTHDTC = BRTHDAT,
    RFSTDTC = RFSTDAT
  ) |>
  select(STUDYID, DOMAIN, USUBJID, RFSTDTC, BRTHDTC, AGEU, SEX, RACE, ETHNIC, ARM)

setdiff(names(dm_sdtm_real), names(dm_shell))
```

---

## 7  ds_raw — mapping to CDISC controlled terminology

CDISC requires that `DSDECOD` (the standardised disposition reason) uses **controlled
terminology** — a fixed vocabulary defined by the CDISC CT package. Raw `DSTERM`
values are free-text and must be mapped to these standard values.

### Exercise 7a — map DSTERM to DSDECOD using case_when

Map `ds_raw$DSTERM` to the CDISC controlled term for `DSDECOD`. Focus on
`DSCAT == "DISPOSITION EVENT"` records (the study-completion dispositions).

| Raw DSTERM | CDISC DSDECOD |
|---|---|
| `COMPLETED` | `COMPLETED` |
| `ADVERSE EVENT` | `ADVERSE EVENT` |
| `SUBJECT/GUARDIAN DECISION` | `WITHDRAWAL BY SUBJECT` |
| `LACK OF EFFICACY` | `LACK OF EFFICACY` |
| `PROTOCOL VIOLATION` | `PROTOCOL DEVIATION` |
| Any other | `OTHER` |

```{r ds-dsdecod, exercise=TRUE, exercise.lines=18}
ds_raw |>
  mutate(
    USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-"),
    DSDECOD = case_when(
      DSTERM == "COMPLETED"                 ~ "COMPLETED",
      DSTERM == "ADVERSE EVENT"             ~ "ADVERSE EVENT",
      DSTERM == "SUBJECT/GUARDIAN DECISION" ~ "___",
      DSTERM == "LACK OF EFFICACY"          ~ "___",
      DSTERM == "PROTOCOL VIOLATION"        ~ "___",
      TRUE                                  ~ "OTHER"
    )
  ) |>
  select(USUBJID, DSTERM, DSDECOD, DSCAT, DSSTDT) |>
  arrange(USUBJID, DSCAT)
```

```{r ds-dsdecod-hint-1}
# "SUBJECT/GUARDIAN DECISION" → "WITHDRAWAL BY SUBJECT"
# "LACK OF EFFICACY"          → "LACK OF EFFICACY"
# "PROTOCOL VIOLATION"        → "PROTOCOL DEVIATION"
# Compare to pharmaversesdtm::ds$DSDECOD to see all real CT values used
```

```{r ds-dsdecod-solution}
ds_raw |>
  mutate(
    USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-"),
    DSDECOD = case_when(
      DSTERM == "COMPLETED"                 ~ "COMPLETED",
      DSTERM == "ADVERSE EVENT"             ~ "ADVERSE EVENT",
      DSTERM == "SUBJECT/GUARDIAN DECISION" ~ "WITHDRAWAL BY SUBJECT",
      DSTERM == "LACK OF EFFICACY"          ~ "LACK OF EFFICACY",
      DSTERM == "PROTOCOL VIOLATION"        ~ "PROTOCOL DEVIATION",
      TRUE                                  ~ "OTHER"
    )
  ) |>
  select(USUBJID, DSTERM, DSDECOD, DSCAT, DSSTDT) |>
  arrange(USUBJID, DSCAT)
```

### Exercise 7b — cross-reference with real pharmaversesdtm::ds

Compare the DSDECOD values you derived above against those in the real
`pharmaversesdtm::ds` dataset to verify your CT mapping is correct.

```{r ds-compare, exercise=TRUE, exercise.lines=12}
cat("DSDECOD values in your mapped ds_raw (DISPOSITION EVENT only):\n")
ds_raw |>
  filter(DSCAT == "DISPOSITION EVENT") |>
  mutate(
    DSDECOD = case_when(
      DSTERM == "COMPLETED"                 ~ "COMPLETED",
      DSTERM == "ADVERSE EVENT"             ~ "ADVERSE EVENT",
      DSTERM == "SUBJECT/GUARDIAN DECISION" ~ "WITHDRAWAL BY SUBJECT",
      TRUE ~ "OTHER"
    )
  ) |>
  count(DSDECOD)

cat("\nDSDECOD values in real pharmaversesdtm::ds:\n")
ds_sdtm_real |>
  count(DSDECOD, sort = TRUE) |>
  slice_head(n = 8)
```

---

## 8  Annotated CRFs (aCRFs)

### Concept — what an aCRF tells you

An **annotated CRF** is a PDF of each CRF page with annotations showing which SDTM
domain and variable each collected field maps to. They are:

- A **regulatory requirement** — FDA expects them in eCTD Module 5
- A **programming guide** — before writing any `sdtm.oak` mapping, check the aCRF to understand the intended raw→SDTM intent

`pharmaverseraw` bundles its aCRFs under `inst/acrf/`. Locate them with:

```r
acrf_dir <- system.file("acrf", package = "pharmaverseraw")
list.files(acrf_dir)
```

**Typical aCRF annotation example:**

```
CRF Page: Adverse Events
─────────────────────────────────────────────────
Field: "Adverse Event Term"  → AE.AETERM
Field: "Start Date"          → AE.AESTDTC
Field: "End Date"            → AE.AEENDTC
Field: "Severity"            → AE.AESEV
Field: "Serious?"            → AE.AESER
[No annotation]              ← AEDECOD (from MedDRA coding — not on CRF)
[No annotation]              ← AEBODSYS (from MedDRA coding — not on CRF)
[No annotation]              ← AESEQ (derived — not collected)
```

```{r quiz-acrf, echo=FALSE}
quiz(
  caption = "aCRF check",
  question(
    "Annotated CRFs (aCRFs) are required by the FDA as part of which eCTD module?",
    answer("Module 2 — Clinical Summaries"),
    answer("Module 3 — Quality"),
    answer("Module 5 — Clinical Study Reports", correct = TRUE),
    answer("Module 1 — Administrative Information"),
    allow_retry = TRUE
  ),
  question(
    "Why are AEDECOD and AEBODSYS NOT annotated on the AE CRF page?",
    answer("They were accidentally omitted from the aCRF"),
    answer("They are derived computationally from AETERM using a formula"),
    answer("They come from MedDRA medical coding — an external process performed
           after data collection, not something captured on a CRF",
           correct = TRUE),
    answer("They are always the same value as AETERM"),
    allow_retry = TRUE
  ),
  question(
    "You receive a new study's aCRF and notice the field 'Sponsor Subject Number'
     is annotated as DM.SUBJID. What does this tell you about the mapping?",
    answer("SUBJID is a hardcode — always the same value"),
    answer("SUBJID is a direct map from the raw sponsor subject number field",
           correct = TRUE),
    answer("SUBJID must be derived using a complex formula"),
    answer("SUBJID requires MedDRA coding"),
    allow_retry = TRUE
  )
)
```

---

## 9  Exploring real pharmaversesdtm output

The datasets you just built manually are structurally similar to what `sdtm.oak`
produces. Now explore the real output to deepen your understanding.

### Exercise 9a — AE severity by treatment arm in real SDTM

Join `ae_sdtm_real` with `dm_sdtm_real` to compute a cross-tabulation of
AE severity (`AESEV`) by treatment arm (`ARM`). This is the same join you performed
with raw data in Exercise 4d, now using proper SDTM variables.

```{r sdtm-ae-by-arm, exercise=TRUE, exercise.lines=16}
ae_sdtm_real |>
  left_join(
    dm_sdtm_real |> select(USUBJID, ARM),
    by = "USUBJID"
  ) |>
  filter(!is.na(AESEV)) |>
  count(ARM, AESEV) |>
  pivot_wider(names_from = AESEV, values_from = n, values_fill = 0) |>
  arrange(ARM)
```

```{r sdtm-ae-by-arm-hint-1}
# left_join on USUBJID — both ae and dm have this column in SDTM
# This works because SDTM guarantees USUBJID consistency across domains
```

### Exercise 9b — explore the SUPP domain (supplemental qualifiers)

SDTM has a maximum column width constraint. Variables that don't fit in the main domain
or represent non-standard qualifiers live in a **supplemental** (`SUPP--`) dataset.
Explore `suppae_real` to understand its structure.

```{r suppae-explore, exercise=TRUE, exercise.lines=14}
cat("=== suppae structure ===\n")
glimpse(suppae_real)

cat("\n=== Unique qualifier names (QNAM) and their labels (QLABEL) ===\n")
suppae_real |>
  distinct(QNAM, QLABEL) |>
  arrange(QNAM)

cat("\n=== Example rows for one qualifier ===\n")
suppae_real |>
  filter(QNAM == first(suppae_real$QNAM)) |>
  slice_head(n = 4)
```

```{r suppae-explore-hint-1}
# SUPP domains link back to the parent domain via:
#   RDOMAIN = parent domain (e.g. "AE")
#   USUBJID = subject identifier
#   IDVAR   = the key variable in the parent domain (e.g. "AESEQ")
#   IDVARVAL = the value of that key (e.g. "1", "2")
#   QNAM    = the new variable name
#   QVAL    = the value
# This 'tall' structure avoids adding many sparse columns to the main AE domain.
```

### Exercise 9c — verify your manual pipeline against real SDTM structure

Compare the column names your manual pipeline produces to those in `ae_sdtm_real`.
Which SDTM columns are still absent and why?

```{r verify-vs-real, exercise=TRUE, exercise.lines=20}
manual_ae <- ae_raw |>
  mutate(USUBJID = str_c(STUDYID, SITEID, PATNUM, sep = "-"),
         DOMAIN  = "AE") |>
  arrange(USUBJID, AESTDT, AETERM) |>
  group_by(USUBJID) |>
  mutate(AESEQ = row_number()) |>
  ungroup() |>
  rename(AESTDTC = AESTDT, AEENDTC = AEENDT) |>
  select(STUDYID, DOMAIN, USUBJID, AESEQ,
         AETERM, AESTDTC, AEENDTC, AESEV, AESER, AEREL, AEOUT)

cat("Columns produced by manual pipeline:\n")
names(manual_ae)

cat("\nColumns in pharmaversesdtm::ae NOT yet in manual pipeline:\n")
setdiff(names(ae_sdtm_real), names(manual_ae))

cat("\nThese fall into which mapping pattern?\n")
cat("  AEDECOD, AEBODSYS, AESOC, AESOCCD, etc. → EXTERNAL (MedDRA coding)\n")
cat("  AESPID → DIRECT MAP (sponsor-provided ID, absent from our ae_raw)\n")
cat("  AESTDY, AEENDY → DERIVE (study day = date - RFSTDTC + 1)\n")
```

---

## 10  When to use (and not use) this package

```{r quiz-use-cases, echo=FALSE}
quiz(
  caption = "Use cases",
  question(
    "Which of the following is an appropriate use of {pharmaverseraw}?",
    answer("Validating your company's production SDTM pipeline for a regulatory submission"),
    answer("Practicing sdtm.oak derivations and comparing results to pharmaversesdtm",
           correct = TRUE),
    answer("Benchmarking SDTM mapping performance at scale"),
    answer("Running a real clinical study's SDTM mapping"),
    allow_retry = TRUE
  ),
  question(
    "Why would you NOT use pharmaverseraw for real study work?",
    answer("It is too large to load into memory"),
    answer("It uses the wrong R version"),
    answer("It is simulated test data — real studies require actual EDC extracts
           representative of their specific data", correct = TRUE),
    answer("It doesn't support the DM domain"),
    allow_retry = TRUE
  )
)
```

---

## 11  Installing the real packages

```r
install.packages("pharmaverseraw")    # raw EDC test data
install.packages("pharmaversesdtm")   # SDTM target datasets

# Development version
remotes::install_github("pharmaverse/pharmaverseraw", ref = "main")

# Verify and list datasets
library(pharmaverseraw)
data(package = "pharmaverseraw")  # lists ae_raw, dm_raw, ds_raw, ex_raw, ec_raw ...
glimpse(ae_raw)

# Find bundled aCRFs
list.files(system.file("acrf", package = "pharmaverseraw"))
```

### Exercise 11 — environment check

```{r pkg-check, exercise=TRUE, exercise.lines=14}
pkgs <- c("pharmaverseraw", "pharmaversesdtm", "pharmaverseadam",
          "admiral", "sdtm.oak", "xportr")

status <- sapply(pkgs, function(p) {
  if (requireNamespace(p, quietly = TRUE)) {
    paste0("✓ v", packageVersion(p))
  } else {
    "✗ not installed"
  }
})

data.frame(Package = names(status), Status = unname(status), row.names = NULL)
```

---

## 12  Self-check Quiz

```{r final-quiz-07, echo=FALSE}
quiz(
  caption = "Lesson 07 — comprehensive self-check",

  question(
    "In pharmaverseraw, the raw dataset for adverse events is named:",
    answer("adae_raw"),
    answer("ae_raw", correct = TRUE),
    answer("AE_RAW"),
    answer("adverse_events_raw"),
    allow_retry = TRUE
  ),

  question(
    "Why does PATNUM in raw data become USUBJID in SDTM?",
    answer("PATNUM and USUBJID are identical — just renamed"),
    answer("EDC systems use site-local patient numbers; SDTM requires a globally
           unique subject ID built from STUDYID + SITEID + PATNUM", correct = TRUE),
    answer("USUBJID is shorter and easier to type"),
    answer("The FDA requires all identifiers to start with 'U'"),
    allow_retry = TRUE
  ),

  question(
    "A subject appears in ae_raw but has no matching record in dm_raw. What is
     the most important consequence for SDTM mapping?",
    answer("The AE records will be silently dropped by sdtm.oak"),
    answer("AESEQ cannot be computed for that subject"),
    answer("USUBJID can still be derived correctly — there is no consequence"),
    answer("You cannot validate the USUBJID derivation or populate required DM-based
           ISM columns without a DM record — this is a data quality issue that must
           be resolved before mapping", correct = TRUE),
    allow_retry = TRUE
  ),

  question(
    "AESTDTC in SDTM has the suffix 'C'. What does this mean?",
    answer("The value was collected as a numeric code that maps to a character label"),
    answer("The column stores the date as a character string in ISO 8601 format",
           correct = TRUE),
    answer("The date was collected via a checkbox"),
    answer("'C' is an arbitrary suffix with no meaning"),
    allow_retry = TRUE
  ),

  question(
    "CDISC ISO 8601 allows partial dates such as '2020-07' (year-month only, no day).
     When would you expect a partial AESTDTC in real data?",
    answer("Never — SDTM always requires a full YYYY-MM-DD date"),
    answer("When the subject recalls only the month and year of their AE onset,
           and the day was genuinely not captured on the CRF", correct = TRUE),
    answer("When the AE was resolved before the end date was recorded"),
    answer("Only for serious adverse events"),
    allow_retry = TRUE
  ),

  question(
    "Which four mapping patterns cover all raw→SDTM variable transformations?",
    answer("Copy, Rename, Calculate, Lookup"),
    answer("Hardcode, Direct map, Derive, External", correct = TRUE),
    answer("Assign, Conditionally assign, Sequence, Code"),
    answer("CDASH, Non-CDASH, MedDRA, WHO-Drug"),
    allow_retry = TRUE
  ),

  question(
    "You use anti_join(ae_subjects, dm_subjects) to check referential integrity.
     What does a non-empty result tell you?",
    answer("All subjects are correctly represented in both datasets"),
    answer("Some subjects have more AEs than demographics records"),
    answer("There are subjects in ae_raw with no matching dm_raw record — a data
           quality issue requiring resolution before SDTM mapping", correct = TRUE),
    answer("The join keys are mismatched"),
    allow_retry = TRUE
  ),

  question(
    "When computing AESEQ using group_by(USUBJID) |> mutate(AESEQ = row_number()),
     why is it important to arrange() first?",
    answer("arrange() is required before group_by() — it is a syntax rule"),
    answer("row_number() assigns sequence based on current row order, so sorting by
           AESTDT (and a tie-breaker) ensures AESEQ is chronologically reproducible
           and deterministic", correct = TRUE),
    answer("Without arrange(), row_number() returns missing values"),
    answer("It doesn't matter — AESEQ is just an arbitrary counter"),
    allow_retry = TRUE
  ),

  question(
    "What is the purpose of the SUPP (supplemental) domain in SDTM?",
    answer("To store duplicate records for quality checking"),
    answer("To hold variables that cannot be accommodated in the main domain due to
           SDTM column constraints or because they are non-standard qualifiers",
           correct = TRUE),
    answer("To archive raw data alongside SDTM"),
    answer("To provide a long-format version of the main domain"),
    allow_retry = TRUE
  ),

  question(
    "A sponsor maps DSTERM = 'SUBJECT/GUARDIAN DECISION' to DSDECOD = 'WITHDRAWAL
     BY SUBJECT'. This mapping step is an example of which pattern?",
    answer("Hardcode — the value is always 'WITHDRAWAL BY SUBJECT'"),
    answer("Direct map — the values are identical"),
    answer("Derive — the CDISC CT value is computed from the raw free-text term
           using conditional logic (case_when)", correct = TRUE),
    answer("External — the mapping comes from a WHO-Drug lookup table"),
    allow_retry = TRUE
  ),

  question(
    "In pharmaversesdtm::ae, AEDECOD uses MedDRA preferred terms in title case
     (e.g. 'Headache'), while raw AETERM uses all-caps ('HEADACHE'). Why?",
    answer("It is a display preference only — both formats are equally valid in SDTM"),
    answer("SDTM requires all text fields in lower case"),
    answer("AEDECOD is populated from the MedDRA dictionary during medical coding,
           which uses the MedDRA standard casing (title case preferred terms)",
           correct = TRUE),
    answer("sdtm.oak automatically converts to title case during mapping"),
    allow_retry = TRUE
  ),

  question(
    "AGE_AT_AE was computed as floor(as.numeric(AESTDT_D - BRTHDAT_D) / 365.25).
     Why use 365.25 rather than 365?",
    answer("365 gives a result in months, not years"),
    answer("365.25 accounts for leap years, giving a more accurate estimate of
           completed years over multi-year spans", correct = TRUE),
    answer("The CDISC standard mandates 365.25 for all age calculations"),
    answer("floor() only works correctly when dividing by 365.25"),
    allow_retry = TRUE
  )
)
```

---

## Key Takeaways

1. **`{pharmaverseraw}`** is the first link in the chain: raw EDC → SDTM → ADaM. It is a data package, not a computation package. Use `ae_raw`, `dm_raw`, `ds_raw` naming.
2. **`{pharmaversesdtm}`** is the matching SDTM output — load it with `library(pharmaversesdtm)` to use real SDTM reference datasets for structural comparison.
3. **EDC-agnostic and standards-agnostic** by design — what you learn applies to any EDC, not just one system.
4. **The four mapping patterns** — hardcode, direct map, derive, external — cover every raw→SDTM transformation. `sdtm.oak` provides a function for each.
5. **Referential integrity matters**: use `anti_join()` to detect subjects present in one domain but missing from another before mapping begins.
6. **`case_when()`** is the primary tool for mapping raw free-text to CDISC controlled terminology (DSDECOD, AEREL, etc.) — far superior to nested `if_else()`.
7. **SUPP domains** hold variables that overflow SDTM column limits; they link back via `RDOMAIN`, `USUBJID`, `IDVAR`, and `IDVARVAL`.
8. **Annotated CRFs** ship with `{pharmaverseraw}` and show the intended raw→SDTM field mapping. They're also a regulatory deliverable (eCTD Module 5).

**Next** → Lesson 08: `{sdtm.oak}` Part 1 — the OAK algorithm framework

(`rmarkdown::run("08_sdtm_oak_part1.Rmd")`)
