---
title: "Lesson 07 — {pharmaverseraw}: Test EDC Data for Pharmaverse Examples"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
    ace_theme: tomorrow
runtime: shiny_prerendered
description: >
  Understand where pharmaverseraw sits in the data pipeline (raw → SDTM → ADaM).
  Explore realistic raw EDC data, compare it to SDTM, manually preview the key
  derivations sdtm.oak will later automate, and understand what annotated CRFs provide.
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tibble)
library(lubridate)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 60)

# ── Mock raw datasets that mirror pharmaverseraw structure ───────────────────
# These reproduce the column patterns documented in pharmaverseraw v0.1.0.
# If pharmaverseraw is installed you would use:  data("ae_raw", package="pharmaverseraw")

ae_raw <- tribble(
  ~STUDYID,     ~PATNUM, ~SITEID, ~AETERM,                       ~AESTDT,       ~AEENDT,       ~AESEV,      ~AESER, ~AEOUT,
  "TEST_STUDY",    1015,   "701",  "HEADACHE",                   "2020-07-15",  "2020-07-17",  "MILD",      "N",    "RECOVERED/RESOLVED",
  "TEST_STUDY",    1015,   "701",  "NAUSEA",                     "2020-08-02",  "2020-08-05",  "MODERATE",  "N",    "RECOVERED/RESOLVED",
  "TEST_STUDY",    1023,   "701",  "DIZZINESS",                  "2020-07-20",  NA,            "MILD",      "N",    "NOT RECOVERED/NOT RESOLVED",
  "TEST_STUDY",    1023,   "701",  "APPLICATION SITE ERYTHEMA",  "2020-09-10",  "2020-09-12",  "MODERATE",  "N",    "RECOVERED/RESOLVED",
  "TEST_STUDY",    1034,   "702",  "FALL",                       "2020-07-28",  "2020-07-28",  "SEVERE",    "Y",    "RECOVERED/RESOLVED",
  "TEST_STUDY",    1034,   "702",  "HEADACHE",                   "2020-08-14",  "2020-08-16",  "MILD",      "N",    "RECOVERED/RESOLVED",
  "TEST_STUDY",    1056,   "703",  "FATIGUE",                    "2020-10-01",  "2020-10-10",  "MILD",      "N",    "RECOVERED/RESOLVED",
  "TEST_STUDY",    1056,   "703",  "VOMITING",                   "2020-10-03",  "2020-10-04",  "SEVERE",    "Y",    "RECOVERED/RESOLVED"
)

dm_raw <- tribble(
  ~STUDYID,     ~PATNUM, ~SITEID, ~BRTHDAT,      ~SEX, ~RACE,                    ~ARM,
  "TEST_STUDY",    1015,   "701",  "1955-04-12",  "F",  "WHITE",                  "Xanomeline High Dose",
  "TEST_STUDY",    1023,   "701",  "1948-09-30",  "M",  "WHITE",                  "Placebo",
  "TEST_STUDY",    1034,   "702",  "1961-07-22",  "F",  "BLACK OR AFRICAN AMERICAN", "Xanomeline Low Dose",
  "TEST_STUDY",    1056,   "703",  "1940-11-05",  "M",  "WHITE",                  "Xanomeline High Dose",
  "TEST_STUDY",    1078,   "703",  "1953-02-18",  "F",  "ASIAN",                  "Placebo"
)

ds_raw <- tribble(
  ~STUDYID,     ~PATNUM, ~SITEID, ~DSTERM,                         ~DSSTDT,
  "TEST_STUDY",    1015,   "701",  "COMPLETED",                    "2021-01-15",
  "TEST_STUDY",    1023,   "701",  "ADVERSE EVENT",                "2020-10-05",
  "TEST_STUDY",    1034,   "702",  "COMPLETED",                    "2021-01-15",
  "TEST_STUDY",    1056,   "703",  "SUBJECT/GUARDIAN DECISION",    "2020-11-20",
  "TEST_STUDY",    1078,   "703",  "COMPLETED",                    "2021-01-15"
)

# The SDTM-shaped OUTPUT (what pharmaversesdtm contains)
ae_sdtm <- tribble(
  ~STUDYID,       ~DOMAIN, ~USUBJID,             ~AESEQ, ~AETERM,                      ~AEDECOD,               ~AEBODSYS,                  ~AESTDTC,      ~AEENDTC,      ~AESEV,      ~AESER,
  "TEST_STUDY",   "AE",    "TEST_STUDY-701-1015",  1,     "HEADACHE",                  "Headache",             "Nervous system disorders", "2020-07-15",  "2020-07-17",  "MILD",      "N",
  "TEST_STUDY",   "AE",    "TEST_STUDY-701-1015",  2,     "NAUSEA",                    "Nausea",               "Gastrointestinal disorders","2020-08-02",  "2020-08-05",  "MODERATE",  "N",
  "TEST_STUDY",   "AE",    "TEST_STUDY-701-1023",  1,     "DIZZINESS",                 "Dizziness",            "Nervous system disorders", "2020-07-20",  NA,            "MILD",      "N",
  "TEST_STUDY",   "AE",    "TEST_STUDY-701-1023",  2,     "APPLICATION SITE ERYTHEMA", "Application site erythema","General disorders",     "2020-09-10",  "2020-09-12",  "MODERATE",  "N",
  "TEST_STUDY",   "AE",    "TEST_STUDY-702-1034",  1,     "FALL",                      "Fall",                 "Injury, poisoning",        "2020-07-28",  "2020-07-28",  "SEVERE",    "Y",
  "TEST_STUDY",   "AE",    "TEST_STUDY-702-1034",  2,     "HEADACHE",                  "Headache",             "Nervous system disorders", "2020-08-14",  "2020-08-16",  "MILD",      "N",
  "TEST_STUDY",   "AE",    "TEST_STUDY-703-1056",  1,     "FATIGUE",                   "Fatigue",              "General disorders",        "2020-10-01",  "2020-10-10",  "MILD",      "N",
  "TEST_STUDY",   "AE",    "TEST_STUDY-703-1056",  2,     "VOMITING",                  "Vomiting",             "Gastrointestinal disorders","2020-10-03",  "2020-10-04",  "SEVERE",    "Y"
)
```

## Overview

### The pharmaverse data chain

`{pharmaverseraw}` is the **starting point** of the pharmaverse test data chain — simulated EDC output that feeds directly into `{sdtm.oak}`:

```
{pharmaverseraw}      ← you are here (Lesson 07)
   │  raw EDC data
   │  via {sdtm.oak}  ← Lessons 08–10
   ▼
{pharmaversesdtm}     ← SDTM datasets
   │
   │  via {admiral}   ← Lessons 11+
   ▼
{pharmaverseadam}     ← ADaM datasets
   │
   │  via {tfrmt/rtables}
   ▼
   TLFs (Tables, Listings, Figures)
```

This package was released in mid-2025 to fill a long-standing gap: before it, the raw → SDTM step had no canonical test data. `pharmaverseraw` itself does almost **no computation** — it is a pure **data package** bundling raw datasets and annotated CRFs.

> **In this tutorial** the datasets `ae_raw`, `dm_raw`, and `ds_raw` mirror the real
> `pharmaverseraw` package structure. `ae_sdtm` mirrors `pharmaversesdtm::ae`.
> If the packages are installed on your system, try `data("ae_raw", package = "pharmaverseraw")`.

---

```{r quiz-pipeline, echo=FALSE}
quiz(
  caption = "Pipeline orientation",
  question(
    "In the pharmaverse test data chain, what does {pharmaverseraw} feed into?",
    answer("{admiral} directly"),
    answer("{sdtm.oak} to produce SDTM datasets", correct = TRUE),
    answer("{pharmaverseadam} directly"),
    answer("{tfrmt} for table generation"),
    allow_retry = TRUE
  ),
  question(
    "pharmaverseraw is best described as:",
    answer("A package that builds SDTM datasets automatically"),
    answer("A data package — it bundles raw datasets and annotated CRFs with minimal computation", correct = TRUE),
    answer("A validation framework for SDTM conformance checking"),
    answer("A replacement for {pharmaversesdtm}"),
    allow_retry = TRUE
  )
)
```

---

## 1  Design philosophy — agnostic by construction

### EDC-agnostic

Real raw data comes from Veeva Vault, Medidata Rave, Castor, REDCap, and dozens of others — each with its own column naming and quirks. `pharmaverseraw` deliberately uses naming conventions that **don't favour any specific EDC**. What you learn here applies to any raw data source.

### Standards-agnostic

Some columns follow **CDASH** (Clinical Data Acquisition Standards Harmonization — the CDISC standard for CRF design), while others don't. This mirrors real-world variability: your sponsor might have CDASH-conformant CRFs for some pages and bespoke naming for others.

The test data is *intentionally* a bit inconsistent — it's a realistic playground, not a clean reference.

```{r quiz-philosophy, echo=FALSE}
quiz(
  caption = "Design philosophy",
  question(
    "Why is pharmaverseraw deliberately EDC-agnostic?",
    answer("To make it compatible only with Medidata Rave"),
    answer("So the SDTM mapping patterns you learn apply generically to any EDC, not just one system", correct = TRUE),
    answer("Because no CDISC standard governs raw data"),
    answer("To avoid licensing issues with EDC vendors"),
    allow_retry = TRUE
  ),
  question(
    "CDASH stands for:",
    answer("Clinical Data Analysis and Submission Harmonization"),
    answer("Clinical Data Acquisition Standards Harmonization — the CDISC standard for CRF design and raw variable naming", correct = TRUE),
    answer("Controlled Dose Assignment for Standard Hosts"),
    answer("Clinical Domain Attribute Schema Hierarchy"),
    allow_retry = TRUE
  )
)
```

---

## 2  Exploring the raw datasets

### Available domains

The initial v0.1.0 release covers:

| Raw dataset | Maps to SDTM | Description |
|---|---|---|
| `dm_raw` | DM | Demographics — one row per subject |
| `ae_raw` | AE | Adverse Events |
| `ec_raw`, `ex_raw` | EC, EX | Exposure (planned / actual) |
| `ds_raw` | DS | Disposition events |

The naming convention is consistent: `<domain>_raw`. Check installed versions with `data(package = "pharmaverseraw")`.

### Exercise 2a — inspect ae_raw

Use `glimpse()` to inspect the structure of `ae_raw`. Note the column names carefully — how do they differ from what you'd see in SDTM?

```{r explore-ae, exercise=TRUE, exercise.lines=6}
glimpse(ae_raw)
```

```{r explore-ae-hint-1}
# Notice: PATNUM not USUBJID, AESTDT not AESTDTC, no AESEQ, no AEDECOD
```

### Exercise 2b — inspect dm_raw

Inspect `dm_raw`. What is the subject identifier called? What date format is used for birthdate?

```{r explore-dm, exercise=TRUE, exercise.lines=5}
glimpse(dm_raw)
```

### Exercise 2c — count records

How many AE records exist per site? Use `count()`.

```{r count-ae-by-site, exercise=TRUE, exercise.lines=5}
ae_raw |>
  count(___, sort = TRUE)
```

```{r count-ae-by-site-hint-1}
# count(SITEID)
```

```{r count-ae-by-site-solution}
ae_raw |>
  count(SITEID, sort = TRUE)
```

### Exercise 2d — list column names side by side

Compare the column names of `ae_raw` and `ae_sdtm` side by side. Which columns are present in SDTM but absent from raw, and vice versa?

```{r colnames-compare, exercise=TRUE, exercise.lines=12}
# Columns in ae_raw
raw_cols  <- names(ae_raw)

# Columns in ae_sdtm
sdtm_cols <- names(ae_sdtm)

# In SDTM but NOT in raw
cat("In SDTM only:\n")
setdiff(sdtm_cols, raw_cols)
```

```{r colnames-compare-hint-1}
# setdiff(sdtm_cols, raw_cols) — in SDTM but missing from raw
# setdiff(raw_cols, sdtm_cols) — in raw but missing from SDTM
```

---

## 3  What raw data has — and what it doesn't

### The gap between raw and SDTM

Run this comparison to see raw and SDTM rows for the same subject:

```{r raw-sdtm-compare, exercise=TRUE, exercise.lines=14}
cat("=== RAW: subject 1015 ===\n")
ae_raw |>
  filter(PATNUM == 1015) |>
  select(STUDYID, PATNUM, AETERM, AESTDT, AESEV, AESER) |>
  print()

cat("\n=== SDTM: subject TEST_STUDY-701-1015 ===\n")
ae_sdtm |>
  filter(USUBJID == "TEST_STUDY-701-1015") |>
  select(USUBJID, AESEQ, AETERM, AEDECOD, AEBODSYS, AESTDTC, AESEV, AESER) |>
  print()
```

```{r raw-sdtm-compare-hint-1}
# Just run it and observe — focus on what SDTM has that raw doesn't:
# USUBJID (derived), DOMAIN (hardcoded), AESEQ (derived),
# AEDECOD & AEBODSYS (from MedDRA coding), AESTDTC (renamed/formatted)
```

### The four mapping patterns

Every raw → SDTM variable falls into one of four categories:

| Pattern | Description | Example |
|---|---|---|
| **Hardcode** | Same value for every row — no input needed | `DOMAIN = "AE"` |
| **Direct map** | Copy value directly, possibly rename | `AETERM` from `ae_raw$AETERM` |
| **Derive** | Compute from one or more source columns | `USUBJID` from `STUDYID + SITEID + PATNUM` |
| **External** | From an outside process — not in raw data | `AEDECOD` / `AEBODSYS` from MedDRA coding |

```{r quiz-patterns, echo=FALSE}
quiz(
  caption = "Mapping patterns",
  question(
    "AEDECOD (the MedDRA preferred term for an AE) is an example of which mapping pattern?",
    answer("Hardcode — it's always 'Adverse Event'"),
    answer("Direct map — it's the same as AETERM"),
    answer("External — it comes from medical coding (MedDRA), not from the raw CRF data", correct = TRUE),
    answer("Derive — it's computed from AETERM using a formula"),
    allow_retry = TRUE
  ),
  question(
    "DOMAIN = 'AE' for every row in the AE dataset is an example of which pattern?",
    answer("Direct map"),
    answer("Hardcode — the same value is assigned to every row regardless of input", correct = TRUE),
    answer("Derive"),
    answer("External"),
    allow_retry = TRUE
  ),
  question(
    "Which variable must be DERIVED (not directly mapped) because it doesn't exist anywhere in the raw data?",
    answer("AETERM — copied from ae_raw$AETERM"),
    answer("AESEV — copied from ae_raw$AESEV"),
    answer("AESEQ — the within-subject sequence number that must be computed", correct = TRUE),
    answer("AESER — copied from ae_raw$AESER"),
    allow_retry = TRUE
  )
)
```

---

## 4  Manually previewing SDTM derivations

In Lesson 08 you'll use `{sdtm.oak}` functions to automate these. First, do them by hand — it builds the mental model for what `sdtm.oak` is actually doing.

### Exercise 4a — derive USUBJID

In raw data the subject is identified by `PATNUM` (an EDC integer). SDTM requires `USUBJID` in the format `STUDYID-SITEID-PATNUM`. Derive it using `paste()` or `str_c()`.

```{r derive-usubjid, exercise=TRUE, exercise.lines=8}
ae_raw |>
  mutate(
    USUBJID = paste(___, ___, ___, sep = "-")
  ) |>
  select(USUBJID, PATNUM, AETERM)
```

```{r derive-usubjid-hint-1}
# paste(STUDYID, SITEID, PATNUM, sep = "-")
# This mirrors what sdtm.oak's oak_id_vars concept handles
```

```{r derive-usubjid-solution}
ae_raw |>
  mutate(
    USUBJID = paste(STUDYID, SITEID, PATNUM, sep = "-")
  ) |>
  select(USUBJID, PATNUM, AETERM)
```

### Exercise 4b — derive AESEQ

`AESEQ` is the within-subject sequence number — a row counter per USUBJID. After deriving USUBJID, use `group_by()` + `mutate(row_number())` to add it.

```{r derive-aeseq, exercise=TRUE, exercise.lines=12}
ae_raw |>
  mutate(USUBJID = paste(STUDYID, SITEID, PATNUM, sep = "-")) |>
  arrange(USUBJID, AESTDT) |>
  group_by(___) |>
  mutate(AESEQ = ___) |>
  ungroup() |>
  select(USUBJID, AESEQ, AETERM, AESTDT)
```

```{r derive-aeseq-hint-1}
# group_by(USUBJID)
# mutate(AESEQ = row_number())
```

```{r derive-aeseq-solution}
ae_raw |>
  mutate(USUBJID = paste(STUDYID, SITEID, PATNUM, sep = "-")) |>
  arrange(USUBJID, AESTDT) |>
  group_by(USUBJID) |>
  mutate(AESEQ = row_number()) |>
  ungroup() |>
  select(USUBJID, AESEQ, AETERM, AESTDT)
```

### Exercise 4c — hardcode DOMAIN

Add a `DOMAIN` column with the hardcoded value `"AE"` for every row.

```{r hardcode-domain, exercise=TRUE, exercise.lines=6}
ae_raw |>
  mutate(DOMAIN = "___") |>
  select(DOMAIN, AETERM, AESEV)
```

```{r hardcode-domain-solution}
ae_raw |>
  mutate(DOMAIN = "AE") |>
  select(DOMAIN, AETERM, AESEV)
```

### Exercise 4d — rename date columns to SDTM convention

In raw data the AE start date is `AESTDT`. In SDTM it must be named `AESTDTC` (the `C` suffix means "collected character", ISO 8601). Use `rename()` to align.

```{r rename-dtc, exercise=TRUE, exercise.lines=7}
ae_raw |>
  rename(
    AESTDTC = ___,
    AEENDTC = ___
  ) |>
  select(AETERM, AESTDTC, AEENDTC)
```

```{r rename-dtc-hint-1}
# rename(AESTDTC = AESTDT, AEENDTC = AEENDT)
```

```{r rename-dtc-solution}
ae_raw |>
  rename(
    AESTDTC = AESTDT,
    AEENDTC = AEENDT
  ) |>
  select(AETERM, AESTDTC, AEENDTC)
```

### Exercise 4e — put it all together: a manual AE SDTM shell

Combine all four patterns into one pipeline that produces a partial SDTM AE dataset. This is exactly what `sdtm.oak` will automate in Lesson 08.

```{r ae-sdtm-manual, exercise=TRUE, exercise.lines=20}
ae_raw |>
  # 1. Derive USUBJID
  mutate(USUBJID = paste(STUDYID, SITEID, PATNUM, sep = "-")) |>

  # 2. Hardcode DOMAIN
  mutate(DOMAIN = "AE") |>

  # 3. Derive AESEQ (sequence within subject, sorted by start date)
  arrange(USUBJID, AESTDT) |>
  group_by(USUBJID) |>
  mutate(AESEQ = row_number()) |>
  ungroup() |>

  # 4. Rename date columns to SDTM convention
  rename(AESTDTC = AESTDT, AEENDTC = AEENDT) |>

  # 5. Select SDTM columns (AEDECOD excluded — comes from MedDRA coding)
  select(STUDYID, DOMAIN, USUBJID, AESEQ,
         AETERM, AESTDTC, AEENDTC, AESEV, AESER) |>

  arrange(USUBJID, AESEQ)
```

```{r ae-sdtm-manual-hint-1}
# Just run it — all the pieces are already there.
# Notice: AEDECOD and AEBODSYS are absent because they require MedDRA coding.
# sdtm.oak provides functions for every step above — hardcode_ct(), assign_no_ct(),
# oak_id_vars(), etc. Lesson 08 will introduce each one.
```

---

## 5  Parsing raw dates

Raw date strings may not always be in ISO 8601 format. Always validate before converting.

### Exercise 5a — check date format consistency

Inspect the `AESTDT` column in `ae_raw`. Are all values ISO 8601 (`YYYY-MM-DD`) format? Use `str_detect()` to flag any that don't match.

```{r date-check, exercise=TRUE, exercise.lines=10}
ae_raw |>
  mutate(
    IS_ISO = str_detect(AESTDT, "^\\d{4}-\\d{2}-\\d{2}$"),
    AESTDT_PARSED = if_else(IS_ISO, lubridate::ymd(AESTDT), NA)
  ) |>
  select(PATNUM, AESTDT, IS_ISO, AESTDT_PARSED)
```

```{r date-check-hint-1}
# ^\\d{4}-\\d{2}-\\d{2}$ matches YYYY-MM-DD exactly
# IS_ISO tells you if the string is already well-formed
```

### Exercise 5b — compute AE duration in days

For AEs where both start and end date are present, compute `AEDUR` — the number of days the AE lasted (end − start + 1 is the clinical convention for inclusive days).

```{r ae-duration, exercise=TRUE, exercise.lines=10}
ae_raw |>
  filter(!is.na(AEENDT)) |>
  mutate(
    AESTDT_D = ymd(AESTDT),
    AEENDT_D = ymd(AEENDT),
    AEDUR    = as.numeric(___ - ___) + 1   # inclusive days
  ) |>
  select(PATNUM, AETERM, AESTDT, AEENDT, AEDUR)
```

```{r ae-duration-hint-1}
# as.numeric(AEENDT_D - AESTDT_D) + 1
```

```{r ae-duration-solution}
ae_raw |>
  filter(!is.na(AEENDT)) |>
  mutate(
    AESTDT_D = ymd(AESTDT),
    AEENDT_D = ymd(AEENDT),
    AEDUR    = as.numeric(AEENDT_D - AESTDT_D) + 1
  ) |>
  select(PATNUM, AETERM, AESTDT, AEENDT, AEDUR)
```

---

## 6  dm_raw — demographics mapping preview

### Exercise 6a — derive USUBJID from dm_raw

`dm_raw` has `STUDYID`, `SITEID`, and `PATNUM`. Derive `USUBJID` the same way you did for AE.

```{r dm-usubjid, exercise=TRUE, exercise.lines=7}
dm_raw |>
  mutate(USUBJID = paste(___, ___, ___, sep = "___")) |>
  select(USUBJID, SEX, RACE, BRTHDAT)
```

```{r dm-usubjid-hint-1}
# paste(STUDYID, SITEID, PATNUM, sep = "-")
```

```{r dm-usubjid-solution}
dm_raw |>
  mutate(USUBJID = paste(STUDYID, SITEID, PATNUM, sep = "-")) |>
  select(USUBJID, SEX, RACE, BRTHDAT)
```

### Exercise 6b — identify raw→SDTM naming gaps in DM

In SDTM DM, the birthdate column is called `BRTHDTC`. In `dm_raw` it is called `BRTHDAT`. Rename and add a hardcoded `DOMAIN` column.

```{r dm-rename, exercise=TRUE, exercise.lines=8}
dm_raw |>
  mutate(DOMAIN = "___") |>
  rename(BRTHDTC = ___) |>
  select(DOMAIN, BRTHDTC, SEX, RACE)
```

```{r dm-rename-hint-1}
# DOMAIN = "DM"
# rename(BRTHDTC = BRTHDAT)
```

```{r dm-rename-solution}
dm_raw |>
  mutate(DOMAIN = "DM") |>
  rename(BRTHDTC = BRTHDAT) |>
  select(DOMAIN, BRTHDTC, SEX, RACE)
```

---

## 7  ds_raw — disposition mapping preview

Disposition (`ds_raw`) records why and when subjects left the study.

### Exercise 7 — flag completers vs discontinuers

In `ds_raw`, subjects who left early have a `DSTERM` other than `"COMPLETED"`. Add a flag `DISC_FL` that is `"Y"` if the subject discontinued, `"N"` if they completed.

```{r ds-flag, exercise=TRUE, exercise.lines=10}
ds_raw |>
  mutate(
    USUBJID = paste(STUDYID, SITEID, PATNUM, sep = "-"),
    DISC_FL = if_else(DSTERM == "___", "N", "Y")
  ) |>
  select(USUBJID, DSTERM, DISC_FL, DSSTDT)
```

```{r ds-flag-hint-1}
# if_else(DSTERM == "COMPLETED", "N", "Y")
```

```{r ds-flag-solution}
ds_raw |>
  mutate(
    USUBJID = paste(STUDYID, SITEID, PATNUM, sep = "-"),
    DISC_FL = if_else(DSTERM == "COMPLETED", "N", "Y")
  ) |>
  select(USUBJID, DSTERM, DISC_FL, DSSTDT)
```

---

## 8  Annotated CRFs (aCRFs)

### Concept — what an aCRF tells you

An **annotated CRF** is a PDF of each CRF page with annotations showing which SDTM domain and variable each collected field maps to. They are:

- A **regulatory requirement** — FDA expects them in eCTD Module 5
- A **learning tool** — before writing any `sdtm.oak` mapping, check the aCRF to understand the intended raw→SDTM intent

`pharmaverseraw` bundles its aCRFs under `inst/acrf/`. Locate them with:

```r
acrf_dir <- system.file("acrf", package = "pharmaverseraw")
list.files(acrf_dir)
```

**Typical aCRF annotation example:**

```
CRF Page: Demographics
─────────────────────────────────────────────────
Field: "Date of Birth"       → DM.BRTHDTC
Field: "Sex"                 → DM.SEX
Field: "Race"                → DM.RACE
Field: "Ethnic Group"        → DM.ETHNIC

CRF Page: Adverse Events
─────────────────────────────────────────────────
Field: "Adverse Event Term"  → AE.AETERM
Field: "Start Date"          → AE.AESTDTC
Field: "End Date"            → AE.AEENDTC
Field: "Severity"            → AE.AESEV
Field: "Serious?"            → AE.AESER
```

Notice: `AEDECOD`, `AEBODSYS`, and `AESOC` are **not** annotated here — they don't come from the CRF; they come from MedDRA coding done post-collection.

```{r quiz-acrf, echo=FALSE}
quiz(
  caption = "aCRF check",
  question(
    "Annotated CRFs (aCRFs) are required by the FDA as part of which eCTD module?",
    answer("Module 2 — Clinical Summaries"),
    answer("Module 3 — Quality"),
    answer("Module 5 — Clinical Study Reports", correct = TRUE),
    answer("Module 1 — Administrative Information"),
    allow_retry = TRUE
  ),
  question(
    "Why are AEDECOD and AEBODSYS NOT annotated on the AE CRF page?",
    answer("They were accidentally omitted"),
    answer("They are derived computationally from AETERM using a formula"),
    answer("They come from MedDRA medical coding — an external process after data collection", correct = TRUE),
    answer("They are always the same value as AETERM"),
    allow_retry = TRUE
  )
)
```

---

## 9  pharmaversesdtm — the expected output

### Comparing raw to its SDTM target

Use `ae_sdtm` (which mirrors `pharmaversesdtm::ae`) to build your mental model of what SDTM mapping produces.

### Exercise 9a — columns in SDTM that need to be derived

Which columns exist in `ae_sdtm` but are absent from `ae_raw`? These are the ones `sdtm.oak` must create.

```{r sdtm-gaps, exercise=TRUE, exercise.lines=8}
in_sdtm_not_raw <- setdiff(names(ae_sdtm), names(ae_raw))
in_raw_not_sdtm <- setdiff(names(ae_raw), names(ae_sdtm))

cat("Columns in SDTM but NOT in raw (must be derived/hardcoded/coded):\n")
print(in_sdtm_not_raw)

cat("\nColumns in raw but NOT in SDTM (EDC-internal, dropped):\n")
print(in_raw_not_sdtm)
```

### Exercise 9b — verify your manual output matches SDTM structure

Run your manual AE pipeline from Exercise 4e and compare its column set to `ae_sdtm`. Are the shared columns present?

```{r verify-manual-ae, exercise=TRUE, exercise.lines=22}
manual_ae <- ae_raw |>
  mutate(USUBJID = paste(STUDYID, SITEID, PATNUM, sep = "-"),
         DOMAIN  = "AE") |>
  arrange(USUBJID, AESTDT) |>
  group_by(USUBJID) |>
  mutate(AESEQ = row_number()) |>
  ungroup() |>
  rename(AESTDTC = AESTDT, AEENDTC = AEENDT) |>
  select(STUDYID, DOMAIN, USUBJID, AESEQ,
         AETERM, AESTDTC, AEENDTC, AESEV, AESER)

# Columns your manual pipeline has
cat("Manual pipeline columns:\n"); names(manual_ae)

# Columns in ae_sdtm that you're still missing
cat("\nStill missing from manual (need sdtm.oak or MedDRA):\n")
setdiff(names(ae_sdtm), names(manual_ae))
```

```{r verify-manual-ae-hint-1}
# You'll be missing: AEDECOD, AEBODSYS (need MedDRA coding)
# sdtm.oak handles all other mapping steps with cleaner, auditable functions
```

---

## 10  When to use (and not use) this package

```{r quiz-use-cases, echo=FALSE}
quiz(
  caption = "Use cases",
  question(
    "Which of the following is an appropriate use of {pharmaverseraw}?",
    answer("Validating your company's production SDTM pipeline for a regulatory submission"),
    answer("Practicing sdtm.oak derivations and comparing results to pharmaversesdtm", correct = TRUE),
    answer("Benchmarking SDTM performance at scale"),
    answer("Running a real clinical study's SDTM mapping"),
    allow_retry = TRUE
  ),
  question(
    "Why would you NOT use pharmaverseraw for real study work?",
    answer("It is too large to load into memory"),
    answer("It uses the wrong R version"),
    answer("It is simulated test data — real studies require actual EDC extracts representative of their specific data", correct = TRUE),
    answer("It doesn't support the DM domain"),
    allow_retry = TRUE
  )
)
```

---

## 11  Installing the real package

To use the real `pharmaverseraw` (and follow along with official vignettes):

```r
# Stable CRAN release
install.packages("pharmaverseraw")
install.packages("pharmaversesdtm")

# Development version (latest domains not yet on CRAN)
remotes::install_github("pharmaverse/pharmaverseraw", ref = "main")

# List all available datasets
data(package = "pharmaverseraw")

# Load and inspect
library(pharmaverseraw)
data("ae_raw")
glimpse(ae_raw)

# Find the bundled aCRFs
list.files(system.file("acrf", package = "pharmaverseraw"))
```

### Exercise 11 — environment check

Run this code to see which pharmaverse data packages are available in your current environment:

```{r pkg-check, exercise=TRUE, exercise.lines=12}
pkgs <- c("pharmaverseraw", "pharmaversesdtm", "pharmaverseadam",
          "admiral", "sdtm.oak", "xportr")

status <- sapply(pkgs, function(p) {
  if (requireNamespace(p, quietly = TRUE)) {
    paste0("✓ ", packageVersion(p))
  } else {
    "✗ not installed"
  }
})

data.frame(Package = names(status), Status = unname(status), row.names = NULL)
```

---

## 12  Self-check Quiz

```{r final-quiz-07, echo=FALSE}
quiz(
  caption = "Lesson 07 — self-check",

  question(
    "In pharmaverseraw, the raw dataset for adverse events is named:",
    answer("adae_raw"),
    answer("ae_raw", correct = TRUE),
    answer("AE_RAW"),
    answer("adverse_events_raw"),
    allow_retry = TRUE
  ),

  question(
    "Why does PATNUM in raw data become USUBJID in SDTM?",
    answer("PATNUM and USUBJID are identical — just renamed"),
    answer("EDC systems use site-local patient numbers; SDTM requires a globally unique subject ID built from STUDYID + SITEID + PATNUM", correct = TRUE),
    answer("USUBJID is shorter and easier to type"),
    answer("The FDA requires all identifiers to start with 'U'"),
    allow_retry = TRUE
  ),

  question(
    "Which of these four mapping patterns requires data from OUTSIDE the raw CRF?",
    answer("Hardcode (DOMAIN = 'AE')"),
    answer("Direct map (AETERM → AETERM)"),
    answer("Derive (AESEQ from row_number())"),
    answer("External — MedDRA coding for AEDECOD / AEBODSYS", correct = TRUE),
    allow_retry = TRUE
  ),

  question(
    "AESTDTC in SDTM has the suffix 'C' (for 'collected character'). What does this tell you?",
    answer("The value was collected as a numeric code that maps to a character label"),
    answer("The column stores the date as a character string in ISO 8601 format", correct = TRUE),
    answer("The date was collected via a checkbox (C = checked)"),
    answer("The 'C' suffix is arbitrary — it has no meaning"),
    allow_retry = TRUE
  ),

  question(
    "What is the single biggest thing pharmaverseraw was released to provide?",
    answer("Faster SDTM building performance"),
    answer("A canonical raw test dataset so the complete raw → SDTM → ADaM pipeline can be demonstrated end to end", correct = TRUE),
    answer("A replacement for pharmaversesdtm"),
    answer("A validated SDTM conformance checker"),
    allow_retry = TRUE
  )
)
```

---

## Key Takeaways

1. **`{pharmaverseraw}`** is the first link in the chain: raw EDC → SDTM → ADaM. It's a data package, not a computation package.
2. **EDC-agnostic and standards-agnostic** by design — what you learn applies to any EDC, not just one system.
3. Datasets are named **`<domain>_raw`**: `ae_raw`, `dm_raw`, `ds_raw`, etc.
4. Raw data is missing **USUBJID** (derive it), **sequence numbers** (derive them), **ISO 8601 date suffixes** (rename), and **coded terms** (MedDRA provides those externally).
5. The **four mapping patterns** — hardcode, direct map, derive, external — cover every raw→SDTM transformation. `sdtm.oak` provides a function for each.
6. **Annotated CRFs** ship with the package and show the intended raw→SDTM field mapping. They're also a regulatory deliverable (eCTD Module 5).
7. **`{pharmaversesdtm}`** is the matching output — use it as a reference to check your mapping work.

**Next** → Lesson 08: `{sdtm.oak}` Part 1 — the OAK algorithm framework
(`rmarkdown::run("08_sdtm_oak_part1.Rmd")`)
