---
title: "Lesson 08 — {sdtm.oak} Part 1: Concepts and Architecture"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
    ace_theme: tomorrow
runtime: shiny_prerendered
description: >
  Understand why sdtm.oak exists, learn all six core algorithm functions, master the
  oak_id_vars threading concept, read and apply a CT spec, and use condition_add for
  conditional mapping — all with runnable exercises built on mock implementations that
  mirror the real package behaviour.
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tibble)
library(lubridate)
library(stringr)
library(rlang)
library(pharmaversesdtm)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 60)

# ════════════════════════════════════════════════════════════════════════════
# MOCK sdtm.oak IMPLEMENTATIONS
# These reproduce the core behaviour of the real sdtm.oak functions so every
# exercise runs without the package installed.  The real functions add
# additional validation, logging, and metadata-driven features not shown here.
# Once you install sdtm.oak, replace these with library(sdtm.oak).
# ════════════════════════════════════════════════════════════════════════════

oak_id_vars <- function() c("oak_id", "raw_source", "patient_number")

generate_oak_id_vars <- function(raw_dat, pat_var, raw_src) {
  raw_dat |>
    mutate(
      oak_id         = row_number(),
      raw_source     = raw_src,
      patient_number = .data[[pat_var]]
    )
}

assign_no_ct <- function(raw_dat, raw_var, tgt_var,
                         id_vars = oak_id_vars(), tgt_dat = NULL) {
  result <- raw_dat |>
    select(any_of(c(id_vars, raw_var))) |>
    rename(!!sym(tgt_var) := !!sym(raw_var))
  if (!is.null(tgt_dat)) {
    left_join(tgt_dat, result, by = intersect(id_vars, names(tgt_dat)))
  } else {
    result
  }
}

hardcode_no_ct <- function(raw_dat, raw_var, tgt_var, tgt_val,
                           id_vars = oak_id_vars()) {
  raw_dat |>
    filter(!is.na(.data[[raw_var]])) |>
    select(any_of(c(id_vars, raw_var))) |>
    mutate(!!sym(tgt_var) := tgt_val) |>
    select(-any_of(raw_var))
}

assign_ct <- function(raw_dat, raw_var, tgt_var, ct_spec, ct_clst,
                      id_vars = oak_id_vars()) {
  lookup <- ct_spec |>
    filter(codelist_code == ct_clst) |>
    select(collected_value, term_value) |>
    distinct()
  unmapped <- raw_dat |>
    pull(!!sym(raw_var)) |>
    unique() |>
    setdiff(lookup$collected_value) |>
    na.omit()
  if (length(unmapped) > 0)
    message("assign_ct: unmapped value(s) in '", raw_var, "': ",
            paste(unmapped, collapse = ", "))
  raw_dat |>
    select(any_of(c(id_vars, raw_var))) |>
    left_join(lookup, by = setNames("collected_value", raw_var)) |>
    select(-any_of(raw_var)) |>
    rename(!!sym(tgt_var) := term_value)
}

hardcode_ct <- function(raw_dat, raw_var, tgt_var, tgt_val,
                        ct_spec, ct_clst, id_vars = oak_id_vars()) {
  valid_vals <- ct_spec |>
    filter(codelist_code == ct_clst) |>
    pull(term_value) |>
    unique()
  if (!tgt_val %in% valid_vals)
    warning("hardcode_ct: '", tgt_val, "' not in codelist ", ct_clst)
  hardcode_no_ct(raw_dat, raw_var, tgt_var, tgt_val, id_vars)
}

condition_add <- function(raw_dat, condition) {
  cond <- enquo(condition)
  raw_dat |> mutate(.oak_condition = !!cond)
}

# Apply a conditioned dataset — strips the flag and filters to active rows
apply_condition <- function(conditioned_dat) {
  conditioned_dat |>
    filter(.oak_condition) |>
    select(-.oak_condition)
}

# ── Shared raw datasets ──────────────────────────────────────────────────────

vs_raw <- tribble(
  ~STUDYID,     ~PATNUM, ~SITEID, ~VISIT,       ~VISDAT,       ~VISTM,  ~SYS_BP, ~DIA_BP, ~PULSE, ~SUBPOS,
  "TEST_STUDY",    1015,  "701",  "SCREENING",  "15-JUL-2020", "09:00",    128,      82,     72,  "Sitting",
  "TEST_STUDY",    1015,  "701",  "WEEK 4",     "12-AUG-2020", "10:15",    132,      84,     74,  "sitting",
  "TEST_STUDY",    1015,  "701",  "WEEK 8",     "09-SEP-2020", "09:45",    126,      80,     70,  "SITTING",
  "TEST_STUDY",    1023,  "701",  "SCREENING",  "16-JUL-2020", "08:30",    145,      90,     80,  "Sitting",
  "TEST_STUDY",    1023,  "701",  "WEEK 4",     "13-AUG-2020", "11:00",    138,      88,     78,  "sit",
  "TEST_STUDY",    1034,  "702",  "SCREENING",  "17-JUL-2020", "14:00",    120,      76,     65,  "SITTING",
  "TEST_STUDY",    1034,  "702",  "WEEK 4",     "14-AUG-2020", "09:30",    NA,       NA,     68,  NA
)

ae_raw <- tribble(
  ~STUDYID,     ~PATNUM, ~SITEID, ~AETERM,                      ~AESTDT,       ~AESEV,      ~AESER,
  "TEST_STUDY",    1015,  "701",  "HEADACHE",                   "2020-07-20",  "MILD",      "N",
  "TEST_STUDY",    1015,  "701",  "NAUSEA",                     "2020-08-05",  "MODERATE",  "N",
  "TEST_STUDY",    1023,  "701",  "DIZZINESS",                  "2020-07-22",  "MILD",      "N",
  "TEST_STUDY",    1034,  "702",  "FALL",                       "2020-07-28",  "SEVERE",    "Y",
  "TEST_STUDY",    1034,  "702",  "HEADACHE",                   "2020-08-15",  "mild",      "N"  # dirty CT
)

cm_raw <- tribble(
  ~STUDYID,     ~PATNUM, ~SITEID, ~IT.CMTRT,     ~IT.CMTRTOTH,    ~CMDOSE, ~CMDOSU,
  "TEST_STUDY",    1015,  "701",  "IBUPROFEN",    NA,              400,     "mg",
  "TEST_STUDY",    1015,  "701",  "OTHER",        "HERBAL TEA",    NA,      NA,
  "TEST_STUDY",    1023,  "701",  "PARACETAMOL",  NA,              500,     "mg",
  "TEST_STUDY",    1034,  "702",  NA,             "VITAMIN D",     1000,    "IU"
)

# ── Real pharmaversesdtm datasets (for comparison exercises) ─────────────────
vs_sdtm_real  <- pharmaversesdtm::vs
ae_sdtm_real  <- pharmaversesdtm::ae
dm_sdtm_real  <- pharmaversesdtm::dm

# Controlled terminology spec
study_ct <- tribble(
  ~codelist_code, ~codelist_name,  ~term_code, ~term_value,            ~collected_value,
  "C66741",       "VS Test Code",  "SYSBP",    "SYSBP",                "SYSBP",
  "C66741",       "VS Test Code",  "DIABP",    "DIABP",                "DIABP",
  "C66741",       "VS Test Code",  "PULSE",    "PULSE",                "PULSE",
  "C67153",       "VS Test Name",  "SYSBP",    "Systolic Blood Pressure", "Systolic Blood Pressure",
  "C67153",       "VS Test Name",  "DIABP",    "Diastolic Blood Pressure","Diastolic Blood Pressure",
  "C67153",       "VS Test Name",  "PULSE",    "Pulse Rate",           "Pulse Rate",
  "C66770",       "VS Unit",       "mmHg",     "mmHg",                 "mmHg",
  "C66770",       "VS Unit",       "beats/min","beats/min",            "beats/min",
  "C71148",       "Position",      "SITTING",  "SITTING",              "Sitting",
  "C71148",       "Position",      "SITTING",  "SITTING",              "sitting",
  "C71148",       "Position",      "SITTING",  "SITTING",              "sit",
  "C71148",       "Position",      "SITTING",  "SITTING",              "SITTING",
  "C71148",       "Position",      "STANDING", "STANDING",             "Standing",
  "C71148",       "Position",      "SUPINE",   "SUPINE",               "Supine",
  "C66742",       "AE Severity",   "MILD",     "MILD",                 "MILD",
  "C66742",       "AE Severity",   "MILD",     "MILD",                 "mild",
  "C66742",       "AE Severity",   "MODERATE", "MODERATE",             "MODERATE",
  "C66742",       "AE Severity",   "SEVERE",   "SEVERE",               "SEVERE"
)
```

## Overview

### Why a special package for SDTM mapping?

You could build SDTM datasets with nothing but `dplyr` — as you did manually in Lesson 07. So why does pharmaverse maintain `{sdtm.oak}`?

The answer is **standardisation through algorithms**. Roche's analysis of their internal SDTM work found that ~80% of 13,000+ SDTM variable mappings across 6 therapeutic areas could be expressed as instances of just **22 reusable algorithm patterns**. Each one is now a function. The benefits:

| Benefit | What it means in practice |
|---|---|
| **Consistency** | Every "hardcode with CT check" uses the same function — no ad-hoc dplyr |
| **Validation** | Validate `assign_ct()` once; every use of it inherits that validation |
| **Metadata-readiness** | Algorithm + arguments can live in a spec spreadsheet → path to automation |

`sdtm.oak` is co-developed by Roche, Pfizer, GSK, Vertex, and Merck, sponsored by **CDISC COSA** (Open Source Alliance). It is in the 0.x lifecycle — stable enough for production use, with new domains added each release.

> **About this tutorial's code.** The exercises use *mock implementations* of the sdtm.oak
> functions that reproduce their core behaviour. This means every exercise runs without the
> package installed. Real function calls look identical — just add `library(sdtm.oak)` and
> remove the mock block. Where the real function adds extra validation or features, a
> callout box notes the difference.

---

```{r quiz-why-oak, echo=FALSE}
quiz(
  caption = "Why sdtm.oak?",
  question(
    "The core insight behind sdtm.oak is that SDTM mapping patterns are:",
    answer("Unique to every domain and cannot be generalised"),
    answer("Highly repetitive — ~80% of mappings reduce to a small set of algorithm patterns", correct = TRUE),
    answer("Best expressed as SAS macros"),
    answer("Fully automatable without any code"),
    allow_retry = TRUE
  ),
  question(
    "Which organisation sponsors the collaborative development of sdtm.oak?",
    answer("FDA"),
    answer("PHUSE"),
    answer("CDISC COSA (Open Source Alliance)", correct = TRUE),
    answer("ICH"),
    allow_retry = TRUE
  )
)
```

---

## 1  The Algorithm Catalog

### Six core algorithms — one function per pattern

| Algorithm | Function | When to use |
|---|---|---|
| **assign_no_ct** | `assign_no_ct()` | Copy a raw variable → SDTM variable, no CT validation |
| **assign_ct** | `assign_ct()` | Copy a raw variable → SDTM variable, recode against CT |
| **hardcode_no_ct** | `hardcode_no_ct()` | Set SDTM variable to a fixed string, no CT check |
| **hardcode_ct** | `hardcode_ct()` | Set SDTM variable to a fixed string, validated against CT |
| **assign_datetime** | `assign_datetime()` | Parse date/time column(s) → ISO 8601 `--DTC` variable |
| **condition_add** | `condition_add()` | Apply subsequent mappings only when a condition is TRUE |

Plus essential utility functions:

| Function | Purpose |
|---|---|
| `generate_oak_id_vars()` | Attach row-tracing identifiers to a raw dataset |
| `oak_id_vars()` | Return the names of the three identifier columns |
| `derive_seq()` | Derive `--SEQ` (within-subject sequence number) |
| `derive_study_day()` | Calculate `--DY` from ISO 8601 dates |
| `read_ct_spec()` | Load a CT spec from a CSV file |

```{r quiz-catalog, echo=FALSE}
quiz(
  caption = "Algorithm matching",
  question(
    "DOMAIN = 'AE' (same value for every row, no CT needed). Which function?",
    answer("assign_no_ct()"),
    answer("assign_ct()"),
    answer("hardcode_no_ct()", correct = TRUE),
    answer("assign_datetime()"),
    allow_retry = TRUE
  ),
  question(
    "AESEV mapped from raw IT.AESEV — must be MILD, MODERATE, or SEVERE. Which function?",
    answer("assign_no_ct() — just copy the value"),
    answer("assign_ct() — copy and validate/recode against the severity codelist", correct = TRUE),
    answer("hardcode_ct() — it's a fixed value"),
    answer("condition_add() — only map when AESEV is not blank"),
    allow_retry = TRUE
  ),
  question(
    "VSTESTCD = 'SYSBP' for all rows sourced from the SYS_BP raw column, validated against the VS Test Code codelist. Which function?",
    answer("assign_ct()"),
    answer("hardcode_no_ct()"),
    answer("hardcode_ct() — it's a fixed value that must pass CT validation", correct = TRUE),
    answer("assign_datetime()"),
    allow_retry = TRUE
  ),
  question(
    "VSDTC must be built from separate raw columns VISDAT ('15-JUL-2020') and VISTM ('09:00'). Which function?",
    answer("assign_no_ct() with two raw_var arguments"),
    answer("assign_datetime() — it combines and reformats to ISO 8601", correct = TRUE),
    answer("hardcode_no_ct() with a paste() in tgt_val"),
    answer("condition_add()"),
    allow_retry = TRUE
  )
)
```

---

## 2  generate_oak_id_vars() — attach row-tracing identifiers

### Concept

Before any algorithm can run, every raw row needs three identity columns that track it throughout the mapping chain:

| Column | Content | Purpose |
|---|---|---|
| `oak_id` | 1, 2, 3, … | Unique row identifier within the raw source |
| `raw_source` | e.g. `"vs_raw"` | Labels which raw dataset this row came from |
| `patient_number` | copy of `PATNUM` | Subject identifier carried forward |

Together these are `oak_id_vars()`. Every algorithm function accepts `id_vars = oak_id_vars()` so it can merge derived values back onto the correct rows.

```r
# Real sdtm.oak:
library(sdtm.oak)
vs_raw <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")
oak_id_vars()   # c("oak_id", "raw_source", "patient_number")
```

### Exercise 2a — attach OAK identifiers to vs_raw

Run `generate_oak_id_vars()` on `vs_raw`. Inspect the result and confirm the three new columns.

```{r gen-oak-id, exercise=TRUE, exercise.lines=8}
vs_id <- generate_oak_id_vars(
  raw_dat = vs_raw,
  pat_var = "___",    # the raw patient ID column
  raw_src = "___"     # label for this source
)

vs_id |> select(oak_id, raw_source, patient_number, VISIT, SYS_BP)
```

```{r gen-oak-id-hint-1}
# pat_var = "PATNUM"  (the column holding the patient number)
# raw_src = "vs_raw"  (any descriptive label you choose)
```

```{r gen-oak-id-solution}
vs_id <- generate_oak_id_vars(
  raw_dat = vs_raw,
  pat_var = "PATNUM",
  raw_src = "vs_raw"
)
vs_id |> select(oak_id, raw_source, patient_number, VISIT, SYS_BP)
```

### Exercise 2b — inspect oak_id_vars()

Call `oak_id_vars()` and confirm the three column names it returns. Then check they all exist in `vs_id`.

```{r oak-id-vars, exercise=TRUE, exercise.lines=8}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

# The three identifier column names
id_cols <- oak_id_vars()
id_cols

# Verify they are all present in vs_id
all(id_cols %in% names(vs_id))
```

```{r quiz-oak-id, echo=FALSE}
quiz(
  caption = "oak_id_vars check",
  question(
    "Why does sdtm.oak keep ALL rows (including those with NA raw values) rather than filtering them out early?",
    answer("To save memory — filtering is slower than keeping all rows"),
    answer("So every raw row remains traceable through the full algorithm chain and can be audited", correct = TRUE),
    answer("Because NA rows may still receive hardcoded DOMAIN values"),
    answer("The algorithms automatically skip NA rows anyway"),
    allow_retry = TRUE
  )
)
```

### Exercise 2c — compare raw vs real SDTM structure

`vs_sdtm_real` is the actual pharmaversesdtm::vs SDTM dataset — a properly completed VS domain. Use `names()` and `setdiff()` to answer two questions:

1. Which SDTM VS variables are **not** directly present in our raw `vs_raw`? (These need OAK derivation.)
2. What are the `VSTESTCD` values in the real dataset — do they match the codes we've been using?

```{r compare-raw-sdtm, exercise=TRUE, exercise.lines=14}
# Columns in the real SDTM VS domain
names(vs_sdtm_real)

# Columns in our raw source — what's missing from the raw side?
setdiff(names(vs_sdtm_real), names(vs_raw))

# What topic codes are actually in the real domain?
vs_sdtm_real |> distinct(VSTESTCD, VSTEST) |> arrange(VSTESTCD)
```

```{r compare-raw-sdtm-hint-1}
# setdiff(A, B) gives elements in A that are NOT in B.
# Variables like DOMAIN, STUDYID, USUBJID, VSTESTCD, VSTEST, VSORRES, VSORRESU,
# VSPOS, VSDTC, VSSEQ etc. appear in the real SDTM but not in vs_raw — they all
# need to be derived by OAK algorithm calls.
```

```{r quiz-compare-raw-sdtm, echo=FALSE}
quiz(
  caption = "Raw vs SDTM gap",
  question(
    "VSSEQ is present in the real pharmaversesdtm::vs but not in vs_raw. Which sdtm.oak utility function is responsible for deriving it?",
    answer("assign_no_ct()"),
    answer("hardcode_no_ct()"),
    answer("derive_seq()", correct = TRUE),
    answer("generate_oak_id_vars()"),
    allow_retry = TRUE
  )
)
```

---

## 3  assign_no_ct() — direct copy, no CT

### Concept

The simplest algorithm: copy a raw variable into an SDTM variable with no controlled terminology validation.

```r
# Real sdtm.oak syntax:
cm_trt <- assign_no_ct(
  raw_dat = cm_raw,
  raw_var = "IT.CMTRT",    # raw column
  tgt_var = "CMTRT",       # SDTM target column
  id_vars = oak_id_vars()
)
```

**Priority chaining** — a key feature: if two `assign_no_ct()` calls write to the same `tgt_var`, the second call fills only the rows that are still `NA` from the first. This gives you a clean "try this source first, fall back to that source" pattern:

```r
# Fill CMTRT from IT.CMTRT; for rows where that's NA, fill from IT.CMTRTOTH
cm_trt <- cm_raw |>
  assign_no_ct(raw_var = "IT.CMTRT",    tgt_var = "CMTRT") |>
  assign_no_ct(raw_var = "IT.CMTRTOTH", tgt_var = "CMTRT")
```

### Exercise 3a — basic assign_no_ct

Map `AETERM` from `ae_raw` (first attach OAK IDs). Show the result with id columns and `AETERM`.

```{r assign-no-ct-basic, exercise=TRUE, exercise.lines=10}
ae_id <- generate_oak_id_vars(ae_raw, pat_var = "PATNUM", raw_src = "ae_raw")

ae_term <- assign_no_ct(
  raw_dat = ae_id,
  raw_var = "___",
  tgt_var = "___",
  id_vars = oak_id_vars()
)

ae_term
```

```{r assign-no-ct-basic-hint-1}
# raw_var = "AETERM"  (the raw column)
# tgt_var = "AETERM"  (SDTM target — same name here, just a direct copy)
```

```{r assign-no-ct-basic-solution}
ae_id <- generate_oak_id_vars(ae_raw, pat_var = "PATNUM", raw_src = "ae_raw")

ae_term <- assign_no_ct(
  raw_dat = ae_id,
  raw_var = "AETERM",
  tgt_var = "AETERM",
  id_vars = oak_id_vars()
)

ae_term
```

### Exercise 3b — priority chaining

`cm_raw` has two treatment columns: `IT.CMTRT` (the primary treatment name) and `IT.CMTRTOTH` ("Other, specify" free text). Use priority chaining to fill `CMTRT` from the primary column first, then fall back to the "Other" column for rows where the primary was NA or `"OTHER"`.

First attach OAK IDs, then chain two `assign_no_ct()` calls.

```{r assign-no-ct-chain, exercise=TRUE, exercise.lines=14}
cm_id <- generate_oak_id_vars(cm_raw, pat_var = "PATNUM", raw_src = "cm_raw")

# Replace "OTHER" with NA so the second source can fill it
cm_id <- cm_id |>
  mutate(IT.CMTRT = if_else(IT.CMTRT == "OTHER", NA_character_, IT.CMTRT))

cm_trt <- cm_id |>
  assign_no_ct(raw_var = "IT.CMTRT",    tgt_var = "CMTRT", id_vars = oak_id_vars()) |>
  # Now merge the second source (IT.CMTRTOTH fills remaining NAs)
  left_join(
    cm_id |> assign_no_ct(raw_var = "IT.CMTRTOTH", tgt_var = "CMTRT_OTH",
                           id_vars = oak_id_vars()),
    by = oak_id_vars()
  ) |>
  mutate(CMTRT = coalesce(CMTRT, CMTRT_OTH)) |>
  select(-CMTRT_OTH)

cm_trt |> select(patient_number, CMTRT)
```

```{r assign-no-ct-chain-hint-1}
# coalesce(CMTRT, CMTRT_OTH) fills NA positions in CMTRT with the "Other" value
# In real sdtm.oak, the second assign_no_ct() call handles this automatically
# because it only fills NAs in the existing tgt_dat
```

### Exercise 3c — condition_add as a guard for OAK mappings

In the previous exercise you used `if_else()` to replace `"OTHER"` with NA before chaining. In OAK, `condition_add()` is the idiomatic way to guard a mapping to a subset of rows — keeping all rows traceable.

Rewrite the primary-source step so that `CMTRT` is mapped from `IT.CMTRT` **only** for rows where `IT.CMTRT` is not NA **and** not `"OTHER"`. Use `condition_add()` followed by `apply_condition()`.

```{r assign-no-ct-condition, exercise=TRUE, exercise.lines=16}
cm_id <- generate_oak_id_vars(cm_raw, pat_var = "PATNUM", raw_src = "cm_raw")

# Map CMTRT only when IT.CMTRT is not NA and not "OTHER"
cmtrt_primary <- cm_id |>
  condition_add(!is.na(IT.CMTRT) & IT.CMTRT != "___") |>
  apply_condition() |>
  assign_no_ct(raw_var = "IT.CMTRT", tgt_var = "CMTRT", id_vars = oak_id_vars())

# Map from the "Other, specify" column for the remaining rows
cmtrt_other <- cm_id |>
  condition_add(is.na(IT.CMTRT) | IT.CMTRT == "OTHER") |>
  apply_condition() |>
  assign_no_ct(raw_var = "IT.CMTRTOTH", tgt_var = "CMTRT", id_vars = oak_id_vars())

bind_rows(cmtrt_primary, cmtrt_other) |>
  select(patient_number, CMTRT) |>
  arrange(patient_number)
```

```{r assign-no-ct-condition-hint-1}
# condition_add(!is.na(IT.CMTRT) & IT.CMTRT != "OTHER")
# The two conditioned subsets are mutually exclusive — bind_rows() merges them.
# In real sdtm.oak, this pattern avoids any pre-processing mutation of the raw data.
```

```{r assign-no-ct-condition-solution}
cm_id <- generate_oak_id_vars(cm_raw, pat_var = "PATNUM", raw_src = "cm_raw")

cmtrt_primary <- cm_id |>
  condition_add(!is.na(IT.CMTRT) & IT.CMTRT != "OTHER") |>
  apply_condition() |>
  assign_no_ct(raw_var = "IT.CMTRT", tgt_var = "CMTRT", id_vars = oak_id_vars())

cmtrt_other <- cm_id |>
  condition_add(is.na(IT.CMTRT) | IT.CMTRT == "OTHER") |>
  apply_condition() |>
  assign_no_ct(raw_var = "IT.CMTRTOTH", tgt_var = "CMTRT", id_vars = oak_id_vars())

bind_rows(cmtrt_primary, cmtrt_other) |>
  select(patient_number, CMTRT) |>
  arrange(patient_number)
```

---

## 4  assign_ct() — copy with controlled terminology recoding

### Concept

When the SDTM variable is subject to a CT codelist, `assign_ct()` both maps the value **and** recodes it from whatever the raw data says to the standardised CDISC term.

```r
# Real sdtm.oak syntax:
vs_pos <- assign_ct(
  raw_dat = vs_raw,
  raw_var = "SUBPOS",         # raw: "Sitting", "sitting", "sit"
  tgt_var = "VSPOS",          # SDTM: must be "SITTING"
  ct_spec = study_ct,
  ct_clst = "C71148",         # Position codelist
  id_vars = oak_id_vars()
)
```

The CT spec maps `"Sitting"` → `"SITTING"`, `"sitting"` → `"SITTING"`, `"sit"` → `"SITTING"`. If a raw value is not in the spec, the function **alerts you** — it never silently passes non-conformant values through.

### Exercise 4a — inspect the CT spec

Examine `study_ct`. Find all collected values that map to `"SITTING"` under codelist `C71148`. This is how OAK handles raw data inconsistency.

```{r inspect-ct-spec, exercise=TRUE, exercise.lines=8}
study_ct |>
  filter(codelist_code == "___") |>
  select(codelist_code, collected_value, term_value)
```

```{r inspect-ct-spec-hint-1}
# codelist_code == "C71148" (Position codelist)
```

```{r inspect-ct-spec-solution}
study_ct |>
  filter(codelist_code == "C71148") |>
  select(codelist_code, collected_value, term_value)
```

### Exercise 4b — apply assign_ct to VSPOS

Map `SUBPOS` from `vs_raw` to SDTM `VSPOS`, recoding all variants of "sitting" to the standard `"SITTING"`.

```{r assign-ct-vspos, exercise=TRUE, exercise.lines=10}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vspos <- assign_ct(
  raw_dat = vs_id,
  raw_var = "___",        # raw column with dirty values
  tgt_var = "___",        # SDTM target variable
  ct_spec = study_ct,
  ct_clst = "___",        # Position codelist
  id_vars = oak_id_vars()
)

vspos
```

```{r assign-ct-vspos-hint-1}
# raw_var = "SUBPOS"
# tgt_var = "VSPOS"
# ct_clst = "C71148"
```

```{r assign-ct-vspos-solution}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vspos <- assign_ct(
  raw_dat = vs_id,
  raw_var = "SUBPOS",
  tgt_var = "VSPOS",
  ct_spec = study_ct,
  ct_clst = "C71148",
  id_vars = oak_id_vars()
)

vspos
```

### Exercise 4c — catch an unmapped value

`ae_raw$AESEV` has one row with the value `"mild"` (lowercase). Map it using `assign_ct()` with the AE Severity codelist (`C66742`). Notice whether the mock alerts you about the unmapped value or silently passes it.

```{r assign-ct-catch, exercise=TRUE, exercise.lines=10}
ae_id <- generate_oak_id_vars(ae_raw, pat_var = "PATNUM", raw_src = "ae_raw")

aesev <- assign_ct(
  raw_dat = ae_id,
  raw_var = "AESEV",
  tgt_var = "AESEV",
  ct_spec = study_ct,
  ct_clst = "C66742",
  id_vars = oak_id_vars()
)

aesev   # Which row has NA in AESEV?
```

```{r assign-ct-catch-hint-1}
# "mild" (lowercase) maps to "MILD" in our CT spec — it SHOULD be handled.
# Check study_ct for C66742 to see if "mild" is listed as a collected_value.
# If it IS missing from the spec, that row gets NA and you see the alert.
```

```{r quiz-assign-ct, echo=FALSE}
quiz(
  caption = "assign_ct check",
  question(
    "What happens in assign_ct() when a raw value is not found in the CT spec?",
    answer("The function silently passes the original raw value through"),
    answer("The function errors and stops the pipeline"),
    answer("The target variable gets NA for that row and the function alerts the user", correct = TRUE),
    answer("The function substitutes the closest matching term automatically"),
    allow_retry = TRUE
  )
)
```

---

## 5  hardcode_no_ct() and hardcode_ct()

### Concept

When the SDTM value is **fixed** (not pulled from raw data), use a hardcode function:

```r
# No CT check needed (DOMAIN is always "AE" — that's obvious)
ae_domain <- hardcode_no_ct(
  raw_dat = ae_raw,
  raw_var = "AETERM",    # identifies which rows are "active"
  tgt_var = "DOMAIN",
  tgt_val = "AE",
  id_vars = oak_id_vars()
)

# CT check needed (VSTESTCD must be in the VS Test Code codelist)
vs_testcd <- hardcode_ct(
  raw_dat = vs_raw,
  raw_var = "SYS_BP",
  tgt_var = "VSTESTCD",
  tgt_val = "SYSBP",
  ct_spec = study_ct,
  ct_clst = "C66741",
  id_vars = oak_id_vars()
)
```

> **Why use `raw_var` at all in a hardcode call?**
> It identifies the *row population* — the rows that exist for this variable.
> In a Findings domain, each topic variable (SYSBP, DIABP, PULSE) comes from a
> different raw column. `raw_var = "SYS_BP"` means "create one SDTM row for
> each non-NA row in SYS_BP."

### Exercise 5a — hardcode DOMAIN

Attach OAK IDs to `ae_raw`, then hardcode `DOMAIN = "AE"` for every row where `AETERM` is not missing.

```{r hardcode-no-ct, exercise=TRUE, exercise.lines=10}
ae_id <- generate_oak_id_vars(ae_raw, pat_var = "PATNUM", raw_src = "ae_raw")

ae_domain <- hardcode_no_ct(
  raw_dat = ae_id,
  raw_var = "___",     # the topic variable (identifies active rows)
  tgt_var = "___",
  tgt_val = "___",
  id_vars = oak_id_vars()
)

ae_domain
```

```{r hardcode-no-ct-hint-1}
# raw_var = "AETERM"  (rows with non-NA AETERM are active)
# tgt_var = "DOMAIN", tgt_val = "AE"
```

```{r hardcode-no-ct-solution}
ae_id <- generate_oak_id_vars(ae_raw, pat_var = "PATNUM", raw_src = "ae_raw")

ae_domain <- hardcode_no_ct(
  raw_dat = ae_id,
  raw_var = "AETERM",
  tgt_var = "DOMAIN",
  tgt_val = "AE",
  id_vars = oak_id_vars()
)

ae_domain
```

### Exercise 5b — hardcode_ct with validation

Hardcode `VSTESTCD = "SYSBP"` for rows sourced from `SYS_BP`, validated against codelist `C66741`. Then try setting `tgt_val = "INVALIDCODE"` — does the mock warn you?

```{r hardcode-ct-ex, exercise=TRUE, exercise.lines=18}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

# Valid hardcode
vs_sysbp_code <- hardcode_ct(
  raw_dat = vs_id,
  raw_var = "SYS_BP",
  tgt_var = "VSTESTCD",
  tgt_val = "___",
  ct_spec = study_ct,
  ct_clst = "C66741",
  id_vars = oak_id_vars()
)

vs_sysbp_code

# Now try an invalid value — watch for the warning message
hardcode_ct(vs_id, "SYS_BP", "VSTESTCD", "INVALIDCODE", study_ct, "C66741")
```

```{r hardcode-ct-ex-hint-1}
# tgt_val = "SYSBP"  (a valid entry in the C66741 codelist)
# "INVALIDCODE" is not in the spec → the function warns you
```

```{r hardcode-ct-ex-solution}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vs_sysbp_code <- hardcode_ct(
  raw_dat = vs_id,
  raw_var  = "SYS_BP",
  tgt_var  = "VSTESTCD",
  tgt_val  = "SYSBP",
  ct_spec  = study_ct,
  ct_clst  = "C66741",
  id_vars  = oak_id_vars()
)

vs_sysbp_code
hardcode_ct(vs_id, "SYS_BP", "VSTESTCD", "INVALIDCODE", study_ct, "C66741")
```

### Exercise 5c — hardcode a domain-wide STUDYID

`STUDYID` is the same fixed string for every row in a domain — a classic `hardcode_no_ct()` job. The `raw_var` here identifies which rows are "active": rows with a non-NA `AETERM`.

Hardcode `STUDYID = "TEST_STUDY"` for every AE record. Then verify that all 5 rows in `ae_raw` received the value.

```{r hardcode-studyid, exercise=TRUE, exercise.lines=14}
ae_id <- generate_oak_id_vars(ae_raw, pat_var = "PATNUM", raw_src = "ae_raw")

ae_studyid <- hardcode_no_ct(
  raw_dat = ae_id,
  raw_var = "___",          # activating variable
  tgt_var = "___",
  tgt_val = "___",
  id_vars = oak_id_vars()
)

# How many rows and what is the unique STUDYID value?
ae_studyid |>
  count(STUDYID)
```

```{r hardcode-studyid-hint-1}
# raw_var = "AETERM"  (rows with non-NA AETERM get the hardcoded value)
# tgt_var = "STUDYID", tgt_val = "TEST_STUDY"
```

```{r hardcode-studyid-solution}
ae_id <- generate_oak_id_vars(ae_raw, pat_var = "PATNUM", raw_src = "ae_raw")

ae_studyid <- hardcode_no_ct(
  raw_dat = ae_id,
  raw_var = "AETERM",
  tgt_var = "STUDYID",
  tgt_val = "TEST_STUDY",
  id_vars = oak_id_vars()
)

ae_studyid |> count(STUDYID)
```

```{r quiz-hardcode-studyid, echo=FALSE}
quiz(
  caption = "hardcode_no_ct vs hardcode_ct",
  question(
    "STUDYID uses hardcode_no_ct() rather than hardcode_ct(). The best reason is:",
    answer("STUDYID is short and doesn't need validation"),
    answer("STUDYID is not subject to a controlled terminology codelist — any sponsor-assigned value is valid", correct = TRUE),
    answer("hardcode_ct() cannot handle string values"),
    answer("STUDYID is always the same length so CT checks are redundant"),
    allow_retry = TRUE
  )
)
```

---

## 6  assign_datetime() — ISO 8601 from raw date/time columns

### Concept

SDTM `--DTC` variables must be ISO 8601 strings (`YYYY-MM-DD` or `YYYY-MM-DDTHH:MM:SS`). Raw data may provide dates in a variety of formats, sometimes with a separate time column.

```r
# Real sdtm.oak syntax:
vs_dtc <- assign_datetime(
  raw_dat = vs_raw,
  raw_var = c("VISDAT", "VISTM"),           # date col, time col
  tgt_var = "VSDTC",
  raw_fmt = c(list(c("d-m-y", "dd-mmm-yyyy")), "H:M"),
  id_vars = oak_id_vars()
)
```

The function:
- Accepts a **list of input formats** per component (handles `"15-JUL-2020"` and `"15-07-2020"` in the same column)
- Combines date + time into `"2020-07-15T09:00"`
- Returns partial ISO 8601 when time is missing: `"2020-07-15"`
- Exposes parse failures via `problems()` — **never silently drops bad rows**

### Exercise 6 — manual ISO 8601 construction (mirroring assign_datetime logic)

The mock in this tutorial doesn't implement `assign_datetime()` (date parsing is complex). Instead, practice the underlying logic manually using `lubridate`. Parse `VISDAT` from `"dd-mmm-yyyy"` format and combine with `VISTM` into an ISO 8601 datetime string.

```{r datetime-manual, exercise=TRUE, exercise.lines=14}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vs_id |>
  mutate(
    # Parse the "15-JUL-2020" format (day-month abbreviation-year)
    VISDAT_D  = dmy(VISDAT),
    # Combine with time to produce ISO 8601 datetime string
    VSDTC     = if_else(
      !is.na(VISTM),
      paste0(as.character(VISDAT_D), "T", VISTM, ":00"),
      as.character(VISDAT_D)
    )
  ) |>
  select(patient_number, VISIT, VISDAT, VISTM, VSDTC)
```

```{r datetime-manual-hint-1}
# dmy() parses day-month-year format ("15-JUL-2020")
# paste0(date, "T", time, ":00") builds "2020-07-15T09:00:00"
# This is exactly what assign_datetime() automates with format detection
```

```{r quiz-datetime, echo=FALSE}
quiz(
  caption = "assign_datetime check",
  question(
    "When assign_datetime() cannot parse a date value, it:",
    answer("Silently substitutes today's date"),
    answer("Errors and stops the entire pipeline"),
    answer("Returns NA for that row and records the failure in problems() for review", correct = TRUE),
    answer("Attempts the next format in the format list indefinitely"),
    allow_retry = TRUE
  ),
  question(
    "A raw VS record has a date ('15-JUL-2020') but no time. assign_datetime() will produce:",
    answer("An error because ISO 8601 requires a time component"),
    answer("'2020-07-15T00:00:00' (midnight default)"),
    answer("'2020-07-15' — a valid partial ISO 8601 date string", correct = TRUE),
    answer("NA because the datetime is incomplete"),
    allow_retry = TRUE
  )
)
```

---

## 7  condition_add() — conditional mapping

### Concept

`condition_add()` creates a **conditioned data frame** — a tibble tagged with a logical column `.oak_condition` that marks which rows are "active" for the next algorithm call. The critical design choice: rows are **tagged, not filtered** — all rows remain present for traceability.

```r
# Real sdtm.oak syntax:
vs_orres_conditional <- vs_raw |>
  condition_add(SYS_BP != "" & !is.na(SYS_BP)) |>
  assign_no_ct(raw_var = "SYS_BP", tgt_var = "VSORRES", id_vars = oak_id_vars())
```

The condition is consumed by the next algorithm — only the active rows receive the mapped value. Inactive rows get `NA`.

This is OAK's answer to SAS's `IF condition THEN` in a DATA step.

### Exercise 7a — inspect a conditioned data frame

Apply `condition_add()` to flag only rows where `SYS_BP` is not NA. Print the result and observe the `.oak_condition` column.

```{r condition-inspect, exercise=TRUE, exercise.lines=8}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vs_cond <- condition_add(vs_id, !is.na(SYS_BP))

vs_cond |> select(oak_id, patient_number, VISIT, SYS_BP, .oak_condition)
```

```{r condition-inspect-hint-1}
# .oak_condition is TRUE when SYS_BP is not NA, FALSE when it is
# Row 7 (PATNUM 1034, WEEK 4) has SYS_BP = NA → .oak_condition = FALSE
```

### Exercise 7b — conditional assign_no_ct

Map `SYS_BP` to `VSORRES` **only** for rows where `SYS_BP` is not NA. Use `condition_add()` then `apply_condition()` (a helper in this tutorial that strips the flag and keeps active rows).

```{r condition-map, exercise=TRUE, exercise.lines=12}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vs_orres <- vs_id |>
  condition_add(!is.na(___)) |>
  apply_condition() |>           # keep only active rows (real sdtm.oak handles this internally)
  assign_no_ct(
    raw_var = "SYS_BP",
    tgt_var = "VSORRES",
    id_vars = oak_id_vars()
  )

vs_orres   # should have 6 rows — row with NA SYS_BP is excluded
```

```{r condition-map-hint-1}
# condition_add(!is.na(SYS_BP))
# apply_condition() is a tutorial helper — in real sdtm.oak, the subsequent
# algorithm call handles the conditioned data frame automatically
```

```{r condition-map-solution}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vs_orres <- vs_id |>
  condition_add(!is.na(SYS_BP)) |>
  apply_condition() |>
  assign_no_ct(
    raw_var = "SYS_BP",
    tgt_var = "VSORRES",
    id_vars = oak_id_vars()
  )

vs_orres
```

```{r quiz-condition-add, echo=FALSE}
quiz(
  caption = "condition_add check",
  question(
    "Why does condition_add() tag rows rather than filter them out?",
    answer("Filtering is slower in R"),
    answer("So all raw rows remain in the dataset and every row can be traced and audited back to its source", correct = TRUE),
    answer("Because the condition may change later in the pipeline"),
    answer("Filtering breaks the oak_id sequence"),
    allow_retry = TRUE
  )
)
```

### Exercise 7c — compound condition: non-missing AND non-screening visits

In clinical trials, baseline (SCREENING) vital signs are sometimes handled separately from on-treatment visits. Map `VSORRES` from `SYS_BP` only for rows where **both** conditions hold:

- `SYS_BP` is not NA, **and**
- `VISIT` is not `"SCREENING"`

How many rows pass this compound condition?

```{r compound-condition, exercise=TRUE, exercise.lines=14}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vs_ontreatment <- vs_id |>
  condition_add(!is.na(SYS_BP) & VISIT != "___") |>
  apply_condition() |>
  assign_no_ct(raw_var = "SYS_BP", tgt_var = "VSORRES", id_vars = oak_id_vars())

# Expect: WEEK 4 and WEEK 8 rows only (not SCREENING, not the NA row)
vs_ontreatment |> select(patient_number, VSORRES)
nrow(vs_ontreatment)
```

```{r compound-condition-hint-1}
# VISIT != "SCREENING" removes all three screening rows.
# The NA SYS_BP row (patient 1034, WEEK 4) also fails the !is.na(SYS_BP) check.
# You should end up with 3 rows: patient 1015 WEEK4+WEEK8, patient 1023 WEEK4.
```

```{r compound-condition-solution}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vs_ontreatment <- vs_id |>
  condition_add(!is.na(SYS_BP) & VISIT != "SCREENING") |>
  apply_condition() |>
  assign_no_ct(raw_var = "SYS_BP", tgt_var = "VSORRES", id_vars = oak_id_vars())

vs_ontreatment |> select(patient_number, VSORRES)
nrow(vs_ontreatment)
```

---

## 8  The CT spec in detail

### Spec format

A CT spec is a plain data frame (usually read from a CSV) with these key columns:

| Column | Description |
|---|---|
| `codelist_code` | NCI code for the codelist (e.g. `"C66742"`) |
| `codelist_name` | Human-readable name (e.g. `"AE Severity"`) |
| `term_code` | CDISC submission value code |
| `term_value` | The standard CDISC term string to write into SDTM |
| `collected_value` | Raw value(s) that map to this term (many rows per term) |

Multiple `collected_value` rows for the same `term_value` is how OAK handles dirty raw data — one standard output, many acceptable raw inputs.

```r
# Real sdtm.oak:
study_ct <- read_ct_spec("path/to/study_ct.csv")
# Or use the built-in example:
study_ct <- read_ct_spec_example("ct-01-cm")
```

### Exercise 8a — explore the full CT spec

Inspect `study_ct`. How many distinct codelists are present? For codelist `C66742`, how many raw collected values map to `"MILD"`?

```{r ct-spec-explore, exercise=TRUE, exercise.lines=12}
# Number of distinct codelists
study_ct |> distinct(codelist_code, codelist_name)

# How many raw values map to MILD in C66742?
study_ct |>
  filter(codelist_code == "C66742", term_value == "MILD") |>
  select(collected_value, term_value)
```

### Exercise 8b — add a new entry to the CT spec

The spec for `C71148` (Position) is missing the raw value `"standing"` (lowercase). Add it as a mapping to `"STANDING"`. Use `bind_rows()`.

```{r ct-spec-extend, exercise=TRUE, exercise.lines=12}
new_entry <- tibble(
  codelist_code = "C71148",
  codelist_name = "Position",
  term_code     = "STANDING",
  term_value    = "___",
  collected_value = "___"
)

study_ct_extended <- bind_rows(study_ct, new_entry)

# Verify the new entry
study_ct_extended |>
  filter(codelist_code == "C71148") |>
  select(collected_value, term_value)
```

```{r ct-spec-extend-hint-1}
# term_value    = "STANDING"
# collected_value = "standing"
```

```{r ct-spec-extend-solution}
new_entry <- tibble(
  codelist_code   = "C71148",
  codelist_name   = "Position",
  term_code       = "STANDING",
  term_value      = "STANDING",
  collected_value = "standing"
)

study_ct_extended <- bind_rows(study_ct, new_entry)

study_ct_extended |>
  filter(codelist_code == "C71148") |>
  select(collected_value, term_value)
```

---

## 9  The OAK Pipeline Pattern — putting it all together

### Shape of a complete topic variable

A typical OAK pipeline for one **topic variable** (e.g., SYSBP) chains:

1. `hardcode_ct()` → `VSTESTCD = "SYSBP"` (what was measured)
2. `hardcode_ct()` → `VSTEST = "Systolic Blood Pressure"` (long label)
3. `assign_no_ct()` → `VSORRES = SYS_BP` (the collected value)
4. `hardcode_no_ct()` → `VSORRESU = "mmHg"` (units)
5. `assign_ct()` → `VSPOS` from `SUBPOS` (position, with CT recode)

Then you repeat for DIABP and PULSE, `bind_rows()` all three, and add domain-wide columns (STUDYID, USUBJID, VSSEQ, VSDTC).

### Exercise 9a — build the SYSBP topic variable

Chain all five steps above for SYSBP. Use the mock functions and `study_ct`. The result should be a tibble with one row per measurement of systolic BP.

```{r oak-sysbp-pipeline, exercise=TRUE, exercise.lines=35}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

# Focus on rows where SYS_BP was recorded
vs_sysbp <- vs_id |>
  condition_add(!is.na(SYS_BP)) |>
  apply_condition()

# Step 1: VSTESTCD = "SYSBP"
step1 <- hardcode_ct(vs_sysbp, "SYS_BP", "VSTESTCD", "SYSBP",
                     study_ct, "C66741", oak_id_vars())

# Step 2: VSTEST = "Systolic Blood Pressure"
step2 <- hardcode_ct(vs_sysbp, "SYS_BP", "VSTEST",
                     "Systolic Blood Pressure",
                     study_ct, "C67153", oak_id_vars())

# Step 3: VSORRES = SYS_BP (the actual reading)
step3 <- assign_no_ct(vs_sysbp, "SYS_BP", "VSORRES", oak_id_vars())

# Step 4: VSORRESU = "mmHg"
step4 <- hardcode_no_ct(vs_sysbp, "SYS_BP", "VSORRESU", "mmHg", oak_id_vars())

# Step 5: VSPOS from SUBPOS (recode with CT)
step5 <- assign_ct(vs_sysbp, "SUBPOS", "VSPOS", study_ct, "C71148", oak_id_vars())

# Join all steps on oak_id_vars()
vs_sysbp_full <- step1 |>
  left_join(step2, by = oak_id_vars()) |>
  left_join(step3, by = oak_id_vars()) |>
  left_join(step4, by = oak_id_vars()) |>
  left_join(step5, by = oak_id_vars())

vs_sysbp_full |> select(patient_number, VSTESTCD, VSTEST, VSORRES, VSORRESU, VSPOS)
```

```{r oak-sysbp-pipeline-hint-1}
# Each step() returns a tibble with oak_id_vars() + the derived SDTM column.
# left_join by oak_id_vars() aligns them all correctly.
# In real sdtm.oak, a single chained pipeline with tgt_dat= argument handles this.
```

### Exercise 9b — build DIABP and PULSE, then bind_rows all three

Now repeat the pattern for `DIA_BP` → `DIABP` and `PULSE` → `PULSE`. Use `hardcode_ct()` to set VSTESTCD and VSTEST from the CT spec, then `bind_rows()` the three topic tibbles into one VS snapshot. How many rows does the combined dataset have?

```{r oak-full-vs-pipeline, exercise=TRUE, exercise.lines=50}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

# ── SYSBP (completed for you) ───────────────────────────────────────────────
vs_sysbp <- vs_id |> condition_add(!is.na(SYS_BP)) |> apply_condition()

sysbp_domain <- hardcode_ct(vs_sysbp, "SYS_BP", "VSTESTCD", "SYSBP", study_ct, "C66741", oak_id_vars()) |>
  left_join(hardcode_ct(vs_sysbp, "SYS_BP", "VSTEST", "Systolic Blood Pressure", study_ct, "C67153", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_no_ct(vs_sysbp, "SYS_BP", "VSORRES", oak_id_vars()), by = oak_id_vars()) |>
  left_join(hardcode_no_ct(vs_sysbp, "SYS_BP", "VSORRESU", "mmHg", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_ct(vs_sysbp, "SUBPOS", "VSPOS", study_ct, "C71148", oak_id_vars()), by = oak_id_vars())

# ── DIABP — your turn ───────────────────────────────────────────────────────
vs_diabp <- vs_id |> condition_add(!is.na(___)) |> apply_condition()

diabp_domain <- hardcode_ct(vs_diabp, "DIA_BP", "VSTESTCD", "___", study_ct, "C66741", oak_id_vars()) |>
  left_join(hardcode_ct(vs_diabp, "DIA_BP", "VSTEST", "___", study_ct, "C67153", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_no_ct(vs_diabp, "___", "VSORRES", oak_id_vars()), by = oak_id_vars()) |>
  left_join(hardcode_no_ct(vs_diabp, "DIA_BP", "VSORRESU", "mmHg", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_ct(vs_diabp, "SUBPOS", "VSPOS", study_ct, "C71148", oak_id_vars()), by = oak_id_vars())

# ── PULSE ────────────────────────────────────────────────────────────────────
vs_pulse <- vs_id |> condition_add(!is.na(PULSE)) |> apply_condition()

pulse_domain <- hardcode_ct(vs_pulse, "PULSE", "VSTESTCD", "PULSE", study_ct, "C66741", oak_id_vars()) |>
  left_join(hardcode_ct(vs_pulse, "PULSE", "VSTEST", "Pulse Rate", study_ct, "C67153", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_no_ct(vs_pulse, "PULSE", "VSORRES", oak_id_vars()), by = oak_id_vars()) |>
  left_join(hardcode_no_ct(vs_pulse, "PULSE", "VSORRESU", "beats/min", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_ct(vs_pulse, "SUBPOS", "VSPOS", study_ct, "C71148", oak_id_vars()), by = oak_id_vars())

# ── Combine all three topic variables ────────────────────────────────────────
vs_domain <- bind_rows(sysbp_domain, diabp_domain, pulse_domain) |>
  arrange(patient_number, VSTESTCD)

vs_domain |> select(patient_number, VSTESTCD, VSTEST, VSORRES, VSORRESU, VSPOS)
nrow(vs_domain)
```

```{r oak-full-vs-pipeline-hint-1}
# DIABP: condition_add(!is.na(DIA_BP))
# VSTESTCD = "DIABP", VSTEST = "Diastolic Blood Pressure"
# raw_var for VSORRES = "DIA_BP"
# VSORRESU = "mmHg" (same units as SYSBP)
# Expected total rows: 7 SYSBP + 7 DIABP + 7 PULSE = 21 rows
# (all 7 raw rows have PULSE recorded; only 6 have SYS_BP / DIA_BP)
```

```{r oak-full-vs-pipeline-solution}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")

vs_sysbp <- vs_id |> condition_add(!is.na(SYS_BP)) |> apply_condition()
sysbp_domain <- hardcode_ct(vs_sysbp, "SYS_BP", "VSTESTCD", "SYSBP", study_ct, "C66741", oak_id_vars()) |>
  left_join(hardcode_ct(vs_sysbp, "SYS_BP", "VSTEST", "Systolic Blood Pressure", study_ct, "C67153", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_no_ct(vs_sysbp, "SYS_BP", "VSORRES", oak_id_vars()), by = oak_id_vars()) |>
  left_join(hardcode_no_ct(vs_sysbp, "SYS_BP", "VSORRESU", "mmHg", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_ct(vs_sysbp, "SUBPOS", "VSPOS", study_ct, "C71148", oak_id_vars()), by = oak_id_vars())

vs_diabp <- vs_id |> condition_add(!is.na(DIA_BP)) |> apply_condition()
diabp_domain <- hardcode_ct(vs_diabp, "DIA_BP", "VSTESTCD", "DIABP", study_ct, "C66741", oak_id_vars()) |>
  left_join(hardcode_ct(vs_diabp, "DIA_BP", "VSTEST", "Diastolic Blood Pressure", study_ct, "C67153", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_no_ct(vs_diabp, "DIA_BP", "VSORRES", oak_id_vars()), by = oak_id_vars()) |>
  left_join(hardcode_no_ct(vs_diabp, "DIA_BP", "VSORRESU", "mmHg", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_ct(vs_diabp, "SUBPOS", "VSPOS", study_ct, "C71148", oak_id_vars()), by = oak_id_vars())

vs_pulse <- vs_id |> condition_add(!is.na(PULSE)) |> apply_condition()
pulse_domain <- hardcode_ct(vs_pulse, "PULSE", "VSTESTCD", "PULSE", study_ct, "C66741", oak_id_vars()) |>
  left_join(hardcode_ct(vs_pulse, "PULSE", "VSTEST", "Pulse Rate", study_ct, "C67153", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_no_ct(vs_pulse, "PULSE", "VSORRES", oak_id_vars()), by = oak_id_vars()) |>
  left_join(hardcode_no_ct(vs_pulse, "PULSE", "VSORRESU", "beats/min", oak_id_vars()), by = oak_id_vars()) |>
  left_join(assign_ct(vs_pulse, "SUBPOS", "VSPOS", study_ct, "C71148", oak_id_vars()), by = oak_id_vars())

vs_domain <- bind_rows(sysbp_domain, diabp_domain, pulse_domain) |>
  arrange(patient_number, VSTESTCD)

vs_domain |> select(patient_number, VSTESTCD, VSTEST, VSORRES, VSORRESU, VSPOS)
nrow(vs_domain)
```

### Exercise 9c — compare your VS domain against the real pharmaversesdtm::vs

You've built a VS domain from a small mock raw dataset. The real pharmaversesdtm::vs was produced from the full CDISCPILOT01 study. Use `setdiff()` to check which SDTM VS variables you derived versus which the real domain has — a practical gap analysis.

```{r compare-vs-domains, exercise=TRUE, exercise.lines=12}
# Variables you derived in your vs_domain:
# (re-run the pipeline just to get the column list)
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")
vs_sysbp_ex <- vs_id |> condition_add(!is.na(SYS_BP)) |> apply_condition()
my_vs_cols <- c("VSTESTCD", "VSTEST", "VSORRES", "VSORRESU", "VSPOS")

# SDTM VS variables in the real pharmaversesdtm::vs
real_vs_cols <- names(vs_sdtm_real)

# What's in the real domain that you haven't derived yet?
setdiff(real_vs_cols, my_vs_cols)

# What unique test codes exist in the real domain?
vs_sdtm_real |> distinct(VSTESTCD) |> arrange(VSTESTCD)
```

```{r compare-vs-domains-hint-1}
# You'll see variables like STUDYID, DOMAIN, USUBJID, VSSEQ, VSDTC, VISITNUM,
# VISIT, EPOCH, VSBLFL, VSDTC etc. — all needing additional OAK calls.
# VSTESTCD values in the real data include HEIGHT, WEIGHT, TEMP, and others
# beyond the three you mapped — each would be a separate topic variable chain.
```

---

## 10  Self-check Quiz

```{r final-quiz-08, echo=FALSE}
quiz(
  caption = "Lesson 08 — self-check",

  question(
    "What are the three columns that make up oak_id_vars()?",
    answer("oak_id, PATNUM, STUDYID"),
    answer("oak_id, raw_source, patient_number", correct = TRUE),
    answer("row_id, dataset_name, subject_id"),
    answer("oak_id, raw_source, USUBJID"),
    allow_retry = TRUE
  ),

  question(
    "You need to map AESER from raw IT.AESER — values must be Y or N per CDISC CT. Which function?",
    answer("assign_no_ct() — just copy the value"),
    answer("hardcode_ct() — it's always Y or N"),
    answer("assign_ct() — copy the value and validate/recode against the Y/N codelist", correct = TRUE),
    answer("condition_add() — only map when IT.AESER is not blank"),
    allow_retry = TRUE
  ),

  question(
    "What does priority chaining with two assign_no_ct() calls to the same tgt_var achieve?",
    answer("The second call overwrites all values set by the first"),
    answer("It causes an error because a target variable can only be mapped once"),
    answer("The second call fills only the rows that are still NA after the first — a fallback pattern", correct = TRUE),
    answer("Both calls run independently and you choose which result to keep"),
    allow_retry = TRUE
  ),

  question(
    "In a Findings domain (e.g., VS), the raw_var argument in hardcode_ct() identifies:",
    answer("The codelist to validate against"),
    answer("The raw column whose non-NA rows define which SDTM rows to create for this topic variable", correct = TRUE),
    answer("The variable that will be recoded using the CT spec"),
    answer("The date column for the observation"),
    allow_retry = TRUE
  ),

  question(
    "Which statement about condition_add() is TRUE?",
    answer("It permanently removes rows that don't meet the condition"),
    answer("It creates a conditioned data frame — all rows remain but only active rows receive the next mapping", correct = TRUE),
    answer("It is equivalent to dplyr::filter()"),
    answer("It must be paired with condition_remove() to work"),
    allow_retry = TRUE
  ),

  question(
    "A CT spec row has collected_value = 'sitting' and term_value = 'SITTING'. This means:",
    answer("The raw data must always contain exactly 'SITTING'"),
    answer("When assign_ct() encounters 'sitting' in the raw data, it writes 'SITTING' to the SDTM variable", correct = TRUE),
    answer("Both 'sitting' and 'SITTING' are valid SDTM outputs"),
    answer("The spec entry is invalid — collected and term values must match exactly"),
    allow_retry = TRUE
  ),

  question(
    "The CT spec column 'collected_value' can have MULTIPLE rows pointing to the same 'term_value'. This design allows:",
    answer("Multiple SDTM variables to share the same codelist"),
    answer("Dirty raw data (case variants, abbreviations, synonyms) to all map to a single standard CDISC term", correct = TRUE),
    answer("The same raw value to produce different SDTM outputs based on domain"),
    answer("Codelists to be shared across studies without modification"),
    allow_retry = TRUE
  ),

  question(
    "You call generate_oak_id_vars() before any algorithm. If you forget this step and call assign_no_ct() directly on vs_raw, the most likely outcome is:",
    answer("assign_no_ct() adds oak_id_vars automatically so nothing changes"),
    answer("An error because the required oak_id, raw_source, and patient_number columns are missing from the data", correct = TRUE),
    answer("The function silently uses row numbers as a fallback identifier"),
    answer("The output omits the id columns but the mapping still succeeds"),
    allow_retry = TRUE
  ),

  question(
    "assign_datetime() is told a date column contains two possible formats: 'dd-mmm-yyyy' (e.g. '15-JUL-2020') and 'yyyy-mm-dd' (e.g. '2020-07-15'). When a value matches neither format, the function:",
    answer("Picks the first format and parses it anyway"),
    answer("Stops the pipeline with a fatal error"),
    answer("Returns NA for that row and records the failure in problems() so it can be audited later", correct = TRUE),
    answer("Converts the value to a character string as-is"),
    allow_retry = TRUE
  ),

  question(
    "In the full OAK pipeline for one topic variable (e.g., SYSBP in a VS domain), results from each algorithm call are joined back together using which column(s)?",
    answer("USUBJID alone"),
    answer("PATNUM and VISIT"),
    answer("The three oak_id_vars() columns: oak_id, raw_source, patient_number", correct = TRUE),
    answer("The target variable name"),
    allow_retry = TRUE
  ),

  question(
    "A programmer needs to map CMTRT for 'Other, specify' rows. They use: condition_add(IT.CMTRT == 'OTHER') followed by assign_no_ct(raw_var = 'IT.CMTRTOTH', tgt_var = 'CMTRT'). What is the result for rows where IT.CMTRT is NOT 'OTHER'?",
    answer("Those rows get CMTRT = IT.CMTRT (the primary value)"),
    answer("Those rows get CMTRT = NA — the condition excluded them from this mapping", correct = TRUE),
    answer("Those rows cause an error"),
    answer("Those rows are permanently deleted from the dataset"),
    allow_retry = TRUE
  ),

  question(
    "sdtm.oak was co-developed by Roche, Pfizer, GSK, Vertex, and Merck and is sponsored by CDISC COSA. Why does cross-company collaboration matter for a mapping package specifically?",
    answer("It lowers licensing costs for each company"),
    answer("Shared algorithm implementations mean a validation done once by any company benefits all — no duplicate re-validation effort", correct = TRUE),
    answer("It ensures the package meets FDA regulatory submission requirements automatically"),
    answer("Each company contributes different programming languages so the package works everywhere"),
    allow_retry = TRUE
  )
)
```

---

## Key Takeaways

1. **`sdtm.oak` turns SDTM mapping into 6 reusable algorithm functions** — eliminating ad-hoc dplyr that looks different every time.
2. **`generate_oak_id_vars()`** is always the first step — it attaches `oak_id`, `raw_source`, and `patient_number` so every row is traceable through the chain.
3. **Four mapping patterns** → four function pairs: copy (assign_no_ct / assign_ct), hardcode (hardcode_no_ct / hardcode_ct).
4. **`assign_ct()` and `hardcode_ct()`** validate against a CT spec — unmapped values produce NAs with alerts, never silent data quality issues.
5. **`assign_datetime()`** handles raw date/time → ISO 8601, with multi-format support and a `problems()` audit trail.
6. **`condition_add()`** tags rows for conditional mapping without losing any rows — essential for traceability.
7. **One topic variable = one chain of algorithm calls**, joined by `oak_id_vars()`. Repeat per topic, then `bind_rows()`.

**Next** → Lesson 09: `{sdtm.oak}` Part 2 — building the VS and AE domains end-to-end
(`rmarkdown::run("09_sdtm_oak_part2.Rmd")`)
