---
title: "Lesson 09 — {sdtm.oak} Part 2: Building SDTM Domains End-to-End"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
    ace_theme: tomorrow
runtime: shiny_prerendered
description: >
  Build a complete VS (Findings) domain and AE (Events) domain from raw data using
  the sdtm.oak algorithm pipeline. Apply derive_seq, derive_study_day, and derive_blfl.
  Handle multiple raw sources and compare output to pharmaversesdtm.
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tibble)
library(lubridate)
library(stringr)
library(rlang)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 90)

# ════════════════════════════════════════════════════════════════════════════
# MOCK sdtm.oak FUNCTIONS  (same core as Lesson 08, extended here)
# Replace with library(sdtm.oak) when the package is installed.
# ════════════════════════════════════════════════════════════════════════════

oak_id_vars <- function() c("oak_id", "raw_source", "patient_number")

generate_oak_id_vars <- function(raw_dat, pat_var, raw_src) {
  raw_dat |>
    mutate(
      oak_id         = row_number(),
      raw_source     = raw_src,
      patient_number = .data[[pat_var]]
    )
}

assign_no_ct <- function(raw_dat, raw_var, tgt_var, id_vars = oak_id_vars()) {
  raw_dat |>
    select(any_of(c(id_vars, raw_var))) |>
    rename(!!sym(tgt_var) := !!sym(raw_var))
}

hardcode_no_ct <- function(raw_dat, raw_var, tgt_var, tgt_val, id_vars = oak_id_vars()) {
  raw_dat |>
    filter(!is.na(.data[[raw_var]])) |>
    select(any_of(id_vars)) |>
    mutate(!!sym(tgt_var) := tgt_val)
}

assign_ct <- function(raw_dat, raw_var, tgt_var, ct_spec, ct_clst, id_vars = oak_id_vars()) {
  lookup <- ct_spec |> filter(codelist_code == ct_clst) |>
    select(collected_value, term_value) |> distinct()
  unmapped <- setdiff(na.omit(unique(pull(raw_dat, !!sym(raw_var)))), lookup$collected_value)
  if (length(unmapped)) message("assign_ct: unmapped in '", raw_var, "': ", paste(unmapped, collapse=", "))
  raw_dat |>
    select(any_of(c(id_vars, raw_var))) |>
    left_join(lookup, by = setNames("collected_value", raw_var)) |>
    select(-any_of(raw_var)) |>
    rename(!!sym(tgt_var) := term_value)
}

hardcode_ct <- function(raw_dat, raw_var, tgt_var, tgt_val, ct_spec, ct_clst, id_vars = oak_id_vars()) {
  valid <- ct_spec |> filter(codelist_code == ct_clst) |> pull(term_value) |> unique()
  if (!tgt_val %in% valid) warning("hardcode_ct: '", tgt_val, "' not in codelist ", ct_clst)
  hardcode_no_ct(raw_dat, raw_var, tgt_var, tgt_val, id_vars)
}

condition_add <- function(raw_dat, condition) {
  cond <- enquo(condition)
  raw_dat |> mutate(.oak_condition = !!cond)
}

apply_condition <- function(conditioned_dat) {
  conditioned_dat |> filter(.oak_condition) |> select(-.oak_condition)
}

# ── NEW in Lesson 09 ──────────────────────────────────────────────────────

# derive_seq: assign a 1-based row sequence number per USUBJID
derive_seq <- function(data, tgt_var, rec_vars = "USUBJID") {
  data |>
    arrange(across(all_of(rec_vars))) |>
    group_by(USUBJID) |>
    mutate(!!sym(tgt_var) := row_number()) |>
    ungroup()
}

# derive_study_day: compute --DY with no-Day-0 convention
# CDISC rule: days >= 0 from ref → day+1; days < 0 → unchanged (negative)
derive_study_day <- function(data, dm_domain, tgdt, refdt, study_day_var) {
  ref_dates <- dm_domain |> select(USUBJID, all_of(refdt))
  data |>
    left_join(ref_dates, by = "USUBJID") |>
    mutate(
      .tgt_d  = ymd(.data[[tgdt]]),
      .ref_d  = ymd(.data[[refdt]]),
      !!sym(study_day_var) := case_when(
        is.na(.tgt_d) | is.na(.ref_d) ~ NA_real_,
        as.numeric(.tgt_d - .ref_d) >= 0 ~ as.numeric(.tgt_d - .ref_d) + 1,
        TRUE ~ as.numeric(.tgt_d - .ref_d)
      )
    ) |>
    select(-all_of(refdt), -.tgt_d, -.ref_d)
}

# derive_blfl: flag the last pre-dose observation per USUBJID × VSTESTCD
derive_blfl <- function(data, dm_domain, tgt_var, ref_var,
                        baseline_visits = NULL, date_var = "VSDTC", test_var = "VSTESTCD") {
  ref_dates <- dm_domain |> select(USUBJID, all_of(ref_var))
  data |>
    left_join(ref_dates, by = "USUBJID") |>
    mutate(
      .ref_d  = ymd(.data[[ref_var]]),
      .obs_d  = ymd(.data[[date_var]]),
      .is_bl  = !is.na(.obs_d) & .obs_d <= .ref_d &
                (is.null(baseline_visits) | .data[["VISIT"]] %in% baseline_visits)
    ) |>
    group_by(USUBJID, across(any_of(test_var))) |>
    mutate(
      !!sym(tgt_var) := if_else(
        .is_bl & .obs_d == suppressWarnings(max(.obs_d[.is_bl], na.rm = TRUE)),
        "Y", NA_character_
      )
    ) |>
    ungroup() |>
    select(-all_of(ref_var), -.ref_d, -.obs_d, -.is_bl)
}

# ── Shared datasets ──────────────────────────────────────────────────────────

vs_raw <- tribble(
  ~STUDYID,     ~PATNUM, ~SITEID, ~VISIT,       ~VISDAT,       ~VISTM,  ~SYS_BP, ~DIA_BP, ~PULSE,
  "TEST_STUDY",    1015,  "701",  "SCREENING",  "10-JUL-2020", "09:00",    128,      82,     72,
  "TEST_STUDY",    1015,  "701",  "WEEK 4",     "07-AUG-2020", "10:15",    132,      84,     74,
  "TEST_STUDY",    1015,  "701",  "WEEK 8",     "04-SEP-2020", "09:45",    126,      80,     70,
  "TEST_STUDY",    1023,  "701",  "SCREENING",  "11-JUL-2020", "08:30",    145,      90,     80,
  "TEST_STUDY",    1023,  "701",  "WEEK 4",     "08-AUG-2020", "11:00",    138,      88,     78,
  "TEST_STUDY",    1034,  "702",  "SCREENING",  "12-JUL-2020", "14:00",    120,      76,     65,
  "TEST_STUDY",    1034,  "702",  "WEEK 4",     "09-AUG-2020", "09:30",     NA,      NA,     68
)

ae_raw <- tribble(
  ~STUDYID,     ~PATNUM, ~SITEID, ~AETERM,                      ~AESTDT,       ~AEENDT,       ~AESEV,      ~AESER, ~AEOUT,
  "TEST_STUDY",    1015,  "701",  "HEADACHE",                   "2020-07-20",  "2020-07-22",  "MILD",      "N",    "RECOVERED/RESOLVED",
  "TEST_STUDY",    1015,  "701",  "NAUSEA",                     "2020-08-05",  "2020-08-08",  "MODERATE",  "N",    "RECOVERED/RESOLVED",
  "TEST_STUDY",    1023,  "701",  "DIZZINESS",                  "2020-07-22",  NA,            "MILD",      "N",    "NOT RECOVERED/NOT RESOLVED",
  "TEST_STUDY",    1034,  "702",  "FALL",                       "2020-07-28",  "2020-07-28",  "SEVERE",    "Y",    "RECOVERED/RESOLVED",
  "TEST_STUDY",    1034,  "702",  "HEADACHE",                   "2020-08-15",  "2020-08-17",  "MILD",      "N",    "RECOVERED/RESOLVED"
)

# SAE supplement raw (separate raw source for multi-source exercise)
sae_raw <- tribble(
  ~STUDYID,     ~PATNUM, ~SITEID, ~AETERM,   ~AESTDT,       ~AEENDT,       ~AESEV,   ~AESER,
  "TEST_STUDY",    1034,  "702",  "FALL",    "2020-07-28",  "2020-07-28",  "SEVERE", "Y"
)

# DM reference dataset (provides first-dose date for study-day calculation)
dm_ref <- tribble(
  ~USUBJID,               ~RFXSTDTC,     ~RFSTDTC,
  "TEST_STUDY-701-1015",  "2020-07-15",  "2020-07-15",
  "TEST_STUDY-701-1023",  "2020-07-16",  "2020-07-16",
  "TEST_STUDY-702-1034",  "2020-07-17",  "2020-07-17"
)

# SDTM AE "canonical" output (mirrors pharmaversesdtm::ae structure)
ae_sdtm_ref <- tribble(
  ~STUDYID,       ~DOMAIN, ~USUBJID,               ~AESEQ, ~AETERM,     ~AESTDTC,      ~AEENDTC,      ~AESEV,      ~AESER, ~AEOUT,
  "TEST_STUDY",   "AE",    "TEST_STUDY-701-1015",    1,     "HEADACHE",  "2020-07-20",  "2020-07-22",  "MILD",      "N",    "RECOVERED/RESOLVED",
  "TEST_STUDY",   "AE",    "TEST_STUDY-701-1015",    2,     "NAUSEA",    "2020-08-05",  "2020-08-08",  "MODERATE",  "N",    "RECOVERED/RESOLVED",
  "TEST_STUDY",   "AE",    "TEST_STUDY-701-1023",    1,     "DIZZINESS", "2020-07-22",  NA,            "MILD",      "N",    "NOT RECOVERED/NOT RESOLVED",
  "TEST_STUDY",   "AE",    "TEST_STUDY-702-1034",    1,     "FALL",      "2020-07-28",  "2020-07-28",  "SEVERE",    "Y",    "RECOVERED/RESOLVED",
  "TEST_STUDY",   "AE",    "TEST_STUDY-702-1034",    2,     "HEADACHE",  "2020-08-15",  "2020-08-17",  "MILD",      "N",    "RECOVERED/RESOLVED"
)

# CT spec
study_ct <- tribble(
  ~codelist_code, ~codelist_name,  ~term_code, ~term_value,                   ~collected_value,
  "C66741",       "VS Test Code",  "SYSBP",    "SYSBP",                       "SYSBP",
  "C66741",       "VS Test Code",  "DIABP",    "DIABP",                       "DIABP",
  "C66741",       "VS Test Code",  "PULSE",    "PULSE",                       "PULSE",
  "C67153",       "VS Test Name",  "SYSBP",    "Systolic Blood Pressure",     "Systolic Blood Pressure",
  "C67153",       "VS Test Name",  "DIABP",    "Diastolic Blood Pressure",    "Diastolic Blood Pressure",
  "C67153",       "VS Test Name",  "PULSE",    "Pulse Rate",                  "Pulse Rate",
  "C66770",       "VS Unit",       "mmHg",     "mmHg",                        "mmHg",
  "C66770",       "VS Unit",       "bpm",      "beats/min",                   "beats/min",
  "C66742",       "AE Severity",   "MILD",     "MILD",                        "MILD",
  "C66742",       "AE Severity",   "MODERATE", "MODERATE",                    "MODERATE",
  "C66742",       "AE Severity",   "SEVERE",   "SEVERE",                      "SEVERE",
  "NY",           "No Yes",        "N",        "N",                           "N",
  "NY",           "No Yes",        "Y",        "Y",                           "Y",
  "C66768",       "AE Outcome",    "RECO",     "RECOVERED/RESOLVED",          "RECOVERED/RESOLVED",
  "C66768",       "AE Outcome",    "NREC",     "NOT RECOVERED/NOT RESOLVED",  "NOT RECOVERED/NOT RESOLVED"
)
```

## Overview

### From algorithms to full domains

In Lesson 08 you learned the individual algorithm functions. This lesson puts them together to build **complete SDTM domains** from raw data. You'll work through two domain types:

| Class | Domain | Raw → SDTM pattern |
|---|---|---|
| **Findings** | VS (Vital Signs) | Wide raw row → **many** SDTM rows (one per parameter) |
| **Events** | AE (Adverse Events) | One raw row → **one** SDTM row |

The standard build order for any domain:

```
1. generate_oak_id_vars()       attach row-tracing identifiers
2. Topic block(s)               map TESTCD/TERM/TRT + qualifiers
3. bind_rows()                  combine topic blocks (Findings only)
4. mutate()                     STUDYID, DOMAIN, USUBJID
5. derive_seq()                 --SEQ sequence number
6. derive_study_day()           --DY from reference date
7. derive_blfl()                --BLFL baseline flag (Findings)
8. select() / relocate()        SDTM column order
```

---

```{r quiz-classes, echo=FALSE}
quiz(
  caption = "SDTM observation classes",
  question(
    "Vital Signs (VS) is a Findings domain. Which structural fact distinguishes Findings from Events?",
    answer("Findings have more variables than Events"),
    answer("Findings use a topic variable (--TESTCD) meaning one raw wide row becomes multiple long SDTM rows", correct = TRUE),
    answer("Findings only contain numeric data; Events contain text"),
    answer("Findings domains start with 'F'; Events domains start with 'E'"),
    allow_retry = TRUE
  ),
  question(
    "Which set correctly lists the four SDTM observation classes?",
    answer("Findings, Events, Interventions, Findings About", correct = TRUE),
    answer("Findings, Events, Demographics, Special Purpose"),
    answer("Findings, AE, EX, CM"),
    answer("BDS, ADSL, ADAE, ADLB"),
    allow_retry = TRUE
  ),
  question(
    "For an Interventions domain like EX (Exposure), the topic variable is:",
    answer("EXTESTCD"),
    answer("EXTERM"),
    answer("EXTRT — the treatment name", correct = TRUE),
    answer("EXOBJ"),
    allow_retry = TRUE
  )
)
```

---

## Part 1 — Building VS (Findings) {.tabset}

## 1  Prepare raw data

### Step 1 of 8: attach OAK identifiers

Always the first step before any algorithm call.

```{r vs-setup, exercise=TRUE, exercise.lines=8}
vs_id <- generate_oak_id_vars(
  raw_dat = vs_raw,
  pat_var = "PATNUM",
  raw_src = "vs_raw"
)

vs_id |> select(oak_id, raw_source, patient_number, VISIT, SYS_BP, DIA_BP, PULSE)
```

---

## 2  Build the SYSBP topic block

### Steps 2–3 for one parameter

For each parameter we need four SDTM columns:

| Column | Source | Function |
|---|---|---|
| `VSTESTCD` | fixed `"SYSBP"` | `hardcode_ct()` |
| `VSTEST` | fixed `"Systolic Blood Pressure"` | `hardcode_ct()` |
| `VSORRES` | raw `SYS_BP` | `assign_no_ct()` |
| `VSORRESU` | fixed `"mmHg"` | `hardcode_no_ct()` |

Each function call returns a small tibble with `oak_id_vars()` + the one new column. Join them on `oak_id_vars()`.

```{r sysbp-block, exercise=TRUE, exercise.lines=30}
vs_id <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")

# Only rows where SYS_BP was recorded
vs_sysbp_rows <- vs_id |> filter(!is.na(SYS_BP))

# Four columns — four calls
sysbp_testcd  <- hardcode_ct(vs_sysbp_rows, "SYS_BP", "VSTESTCD",
                              "SYSBP", study_ct, "C66741")
sysbp_test    <- hardcode_ct(vs_sysbp_rows, "SYS_BP", "VSTEST",
                              "Systolic Blood Pressure", study_ct, "C67153")
sysbp_orres   <- assign_no_ct(vs_sysbp_rows, "SYS_BP", "VSORRES")
sysbp_orresu  <- hardcode_no_ct(vs_sysbp_rows, "SYS_BP", "VSORRESU", "mmHg")

# Join all on oak_id_vars()
vs_sysbp <- sysbp_testcd |>
  left_join(sysbp_test,   by = oak_id_vars()) |>
  left_join(sysbp_orres,  by = oak_id_vars()) |>
  left_join(sysbp_orresu, by = oak_id_vars())

vs_sysbp
```

```{r sysbp-block-hint-1}
# Note: vs_sysbp_rows is pre-filtered to non-NA SYS_BP rows
# hardcode_no_ct and hardcode_ct: raw_var identifies the row population,
#   the tgt_val is what gets written into the SDTM column
# All four narrow tibbles share oak_id_vars() → join them cleanly
```

---

## 3  Build the DIABP and PULSE topic blocks

### Exercise — write the DIABP block yourself

Following exactly the same pattern as SYSBP, build the DIABP topic block:

- `VSTESTCD` = `"DIABP"` (codelist `C66741`)
- `VSTEST` = `"Diastolic Blood Pressure"` (codelist `C67153`)
- `VSORRES` from raw `DIA_BP`
- `VSORRESU` = `"mmHg"` (no CT needed)

```{r diabp-block, exercise=TRUE, exercise.lines=25}
vs_id <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")
vs_diabp_rows <- vs_id |> filter(!is.na(DIA_BP))

diabp_testcd  <- hardcode_ct(vs_diabp_rows, "DIA_BP", "VSTESTCD",
                              "___", study_ct, "C66741")
diabp_test    <- hardcode_ct(vs_diabp_rows, "DIA_BP", "VSTEST",
                              "___", study_ct, "C67153")
diabp_orres   <- assign_no_ct(vs_diabp_rows, "___", "VSORRES")
diabp_orresu  <- hardcode_no_ct(vs_diabp_rows, "DIA_BP", "VSORRESU", "___")

vs_diabp <- diabp_testcd |>
  left_join(diabp_test,   by = oak_id_vars()) |>
  left_join(diabp_orres,  by = oak_id_vars()) |>
  left_join(diabp_orresu, by = oak_id_vars())

vs_diabp
```

```{r diabp-block-hint-1}
# "DIABP" for VSTESTCD tgt_val
# "Diastolic Blood Pressure" for VSTEST tgt_val
# raw_var = "DIA_BP" for assign_no_ct (the raw column with the reading)
# "mmHg" for VSORRESU
```

```{r diabp-block-solution}
vs_id <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")
vs_diabp_rows <- vs_id |> filter(!is.na(DIA_BP))

diabp_testcd  <- hardcode_ct(vs_diabp_rows, "DIA_BP", "VSTESTCD", "DIABP", study_ct, "C66741")
diabp_test    <- hardcode_ct(vs_diabp_rows, "DIA_BP", "VSTEST", "Diastolic Blood Pressure", study_ct, "C67153")
diabp_orres   <- assign_no_ct(vs_diabp_rows, "DIA_BP", "VSORRES")
diabp_orresu  <- hardcode_no_ct(vs_diabp_rows, "DIA_BP", "VSORRESU", "mmHg")

vs_diabp <- diabp_testcd |>
  left_join(diabp_test,   by = oak_id_vars()) |>
  left_join(diabp_orres,  by = oak_id_vars()) |>
  left_join(diabp_orresu, by = oak_id_vars())

vs_diabp
```

### Exercise — write the PULSE block with less scaffolding

Pulse rate: `VSTESTCD = "PULSE"`, `VSTEST = "Pulse Rate"`, `VSORRES` from `PULSE`, `VSORRESU = "beats/min"` (codelist `C66770` for units).

```{r pulse-block, exercise=TRUE, exercise.lines=20}
vs_id   <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")
vs_pulse_rows <- vs_id |> filter(!is.na(PULSE))

# Build pulse_testcd, pulse_test, pulse_orres, pulse_orresu
# then join them into vs_pulse

vs_pulse <- ___

vs_pulse
```

```{r pulse-block-hint-1}
# pulse_testcd  <- hardcode_ct(vs_pulse_rows, "PULSE", "VSTESTCD", "PULSE", study_ct, "C66741")
# pulse_test    <- hardcode_ct(vs_pulse_rows, "PULSE", "VSTEST", "Pulse Rate", study_ct, "C67153")
# pulse_orres   <- assign_no_ct(vs_pulse_rows, "PULSE", "VSORRES")
# pulse_orresu  <- hardcode_no_ct(vs_pulse_rows, "PULSE", "VSORRESU", "beats/min")
```

```{r pulse-block-solution}
vs_id         <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")
vs_pulse_rows <- vs_id |> filter(!is.na(PULSE))

pulse_testcd  <- hardcode_ct(vs_pulse_rows, "PULSE", "VSTESTCD", "PULSE",     study_ct, "C66741")
pulse_test    <- hardcode_ct(vs_pulse_rows, "PULSE", "VSTEST",   "Pulse Rate",study_ct, "C67153")
pulse_orres   <- assign_no_ct(vs_pulse_rows, "PULSE", "VSORRES")
pulse_orresu  <- hardcode_no_ct(vs_pulse_rows, "PULSE", "VSORRESU", "beats/min")

vs_pulse <- pulse_testcd |>
  left_join(pulse_test,   by = oak_id_vars()) |>
  left_join(pulse_orres,  by = oak_id_vars()) |>
  left_join(pulse_orresu, by = oak_id_vars())

vs_pulse
```

---

## 4  bind_rows — the implicit transpose

### Step 3 of 8: combine topic blocks

`bind_rows()` stacks the three topic blocks into one long-format dataset — one row per parameter per visit. This is the **Findings transpose**: wide raw → long SDTM.

```{r vs-bind-rows, exercise=TRUE, exercise.lines=35}
# Rebuild all three blocks (condensed)
vs_id <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")

build_block <- function(rows, raw_col, testcd, test, unit, unit_clst) {
  tc <- hardcode_ct(rows, raw_col, "VSTESTCD", testcd, study_ct, "C66741")
  tt <- hardcode_ct(rows, raw_col, "VSTEST",   test,   study_ct, "C67153")
  or <- assign_no_ct(rows, raw_col, "VSORRES")
  ou <- hardcode_no_ct(rows, raw_col, "VSORRESU", unit)
  tc |> left_join(tt, by = oak_id_vars()) |>
       left_join(or, by = oak_id_vars()) |>
       left_join(ou, by = oak_id_vars())
}

vs_sysbp <- build_block(vs_id |> filter(!is.na(SYS_BP)), "SYS_BP",
                         "SYSBP", "Systolic Blood Pressure", "mmHg", "C66770")
vs_diabp  <- build_block(vs_id |> filter(!is.na(DIA_BP)), "DIA_BP",
                         "DIABP", "Diastolic Blood Pressure","mmHg", "C66770")
vs_pulse  <- build_block(vs_id |> filter(!is.na(PULSE)),  "PULSE",
                         "PULSE", "Pulse Rate", "beats/min", "C66770")

# Bind them
vs <- bind_rows(vs_sysbp, vs_diabp, vs_pulse)

# How many rows? (should be sum of non-NA measurements across all three params)
nrow(vs)
vs |> count(VSTESTCD)
```

```{r vs-bind-rows-hint-1}
# bind_rows() stacks data frames with the same columns
# Rows for SYS_BP NA visits are already excluded (we filtered before building)
# count(VSTESTCD) confirms the row distribution across parameters
```

---

## 5  Domain-wide keys

### Step 4 of 8: STUDYID, DOMAIN, USUBJID

Use plain `dplyr::mutate()` for values that apply to every row regardless of topic.

```{r vs-domain-keys, exercise=TRUE, exercise.lines=20}
# Rebuild vs (condensed)
vs_id <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")
build_block <- function(rows, rc, tc, tt, u) {
  hardcode_ct(rows, rc, "VSTESTCD", tc, study_ct, "C66741") |>
    left_join(hardcode_ct(rows, rc, "VSTEST",   tt, study_ct, "C67153"), by = oak_id_vars()) |>
    left_join(assign_no_ct(rows, rc, "VSORRES"),                          by = oak_id_vars()) |>
    left_join(hardcode_no_ct(rows, rc, "VSORRESU", u),                    by = oak_id_vars())
}
vs <- bind_rows(
  build_block(vs_id |> filter(!is.na(SYS_BP)), "SYS_BP", "SYSBP", "Systolic Blood Pressure", "mmHg"),
  build_block(vs_id |> filter(!is.na(DIA_BP)), "DIA_BP", "DIABP", "Diastolic Blood Pressure","mmHg"),
  build_block(vs_id |> filter(!is.na(PULSE)),  "PULSE",  "PULSE", "Pulse Rate", "beats/min")
)

# Add domain-wide keys
vs <- vs |>
  mutate(
    STUDYID = "TEST_STUDY",
    DOMAIN  = "___",
    USUBJID = paste("TEST_STUDY", patient_number, sep = "___")
  )

vs |> select(STUDYID, DOMAIN, USUBJID, VSTESTCD, VSORRES) |> head(8)
```

```{r vs-domain-keys-hint-1}
# DOMAIN = "VS"
# sep = "-"  →  USUBJID = paste("TEST_STUDY", patient_number, sep = "-")
```

```{r vs-domain-keys-solution}
vs_id <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")
build_block <- function(rows, rc, tc, tt, u) {
  hardcode_ct(rows, rc, "VSTESTCD", tc, study_ct, "C66741") |>
    left_join(hardcode_ct(rows, rc, "VSTEST",   tt, study_ct, "C67153"), by = oak_id_vars()) |>
    left_join(assign_no_ct(rows, rc, "VSORRES"),                          by = oak_id_vars()) |>
    left_join(hardcode_no_ct(rows, rc, "VSORRESU", u),                    by = oak_id_vars())
}
vs <- bind_rows(
  build_block(vs_id |> filter(!is.na(SYS_BP)), "SYS_BP", "SYSBP", "Systolic Blood Pressure", "mmHg"),
  build_block(vs_id |> filter(!is.na(DIA_BP)), "DIA_BP", "DIABP", "Diastolic Blood Pressure","mmHg"),
  build_block(vs_id |> filter(!is.na(PULSE)),  "PULSE",  "PULSE", "Pulse Rate", "beats/min")
)
vs <- vs |>
  mutate(
    STUDYID = "TEST_STUDY",
    DOMAIN  = "VS",
    USUBJID = paste("TEST_STUDY", patient_number, sep = "-")
  )
vs |> select(STUDYID, DOMAIN, USUBJID, VSTESTCD, VSORRES) |> head(8)
```

---

## 6  derive_seq() — sequence number

### Step 5 of 8: assign VSSEQ

`derive_seq()` assigns a 1-based row number within each subject. The `rec_vars` argument determines the uniqueness key — typically `c("USUBJID", "VSTESTCD", "VISIT")`.

```{r vs-seq, exercise=TRUE, exercise.lines=25}
vs_id <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")
build_block <- function(rows, rc, tc, tt, u) {
  hardcode_ct(rows, rc, "VSTESTCD", tc, study_ct, "C66741") |>
    left_join(hardcode_ct(rows, rc, "VSTEST",   tt, study_ct, "C67153"), by = oak_id_vars()) |>
    left_join(assign_no_ct(rows, rc, "VSORRES"),                          by = oak_id_vars()) |>
    left_join(hardcode_no_ct(rows, rc, "VSORRESU", u),                    by = oak_id_vars()) |>
    left_join(vs_id |> select(all_of(oak_id_vars()), VISIT), by = oak_id_vars())
}
vs <- bind_rows(
  build_block(vs_id |> filter(!is.na(SYS_BP)), "SYS_BP", "SYSBP", "Systolic Blood Pressure", "mmHg"),
  build_block(vs_id |> filter(!is.na(DIA_BP)), "DIA_BP", "DIABP", "Diastolic Blood Pressure","mmHg"),
  build_block(vs_id |> filter(!is.na(PULSE)),  "PULSE",  "PULSE", "Pulse Rate", "beats/min")
) |>
  mutate(STUDYID = "TEST_STUDY", DOMAIN = "VS",
         USUBJID = paste("TEST_STUDY", patient_number, sep = "-"))

# Add VSSEQ
vs <- derive_seq(vs, tgt_var = "___", rec_vars = c("USUBJID", "VSTESTCD", "VISIT"))

vs |> select(USUBJID, VSTESTCD, VISIT, VSSEQ, VSORRES)
```

```{r vs-seq-hint-1}
# tgt_var = "VSSEQ"
# derive_seq() numbers rows 1, 2, 3 ... within each USUBJID
```

```{r vs-seq-solution}
vs_id <- generate_oak_id_vars(vs_raw, "PATNUM", "vs_raw")
build_block <- function(rows, rc, tc, tt, u) {
  hardcode_ct(rows, rc, "VSTESTCD", tc, study_ct, "C66741") |>
    left_join(hardcode_ct(rows, rc, "VSTEST", tt, study_ct, "C67153"), by = oak_id_vars()) |>
    left_join(assign_no_ct(rows, rc, "VSORRES"),  by = oak_id_vars()) |>
    left_join(hardcode_no_ct(rows, rc, "VSORRESU", u), by = oak_id_vars()) |>
    left_join(vs_id |> select(all_of(oak_id_vars()), VISIT), by = oak_id_vars())
}
vs <- bind_rows(
  build_block(vs_id |> filter(!is.na(SYS_BP)), "SYS_BP","SYSBP","Systolic Blood Pressure","mmHg"),
  build_block(vs_id |> filter(!is.na(DIA_BP)), "DIA_BP","DIABP","Diastolic Blood Pressure","mmHg"),
  build_block(vs_id |> filter(!is.na(PULSE)),  "PULSE", "PULSE","Pulse Rate","beats/min")
) |>
  mutate(STUDYID="TEST_STUDY", DOMAIN="VS",
         USUBJID=paste("TEST_STUDY", patient_number, sep="-"))

vs <- derive_seq(vs, tgt_var = "VSSEQ", rec_vars = c("USUBJID", "VSTESTCD", "VISIT"))
vs |> select(USUBJID, VSTESTCD, VISIT, VSSEQ, VSORRES)
```

---

## 7  derive_study_day() — VSDY

### Step 6 of 8: days from first dose

`derive_study_day()` computes days from a reference date (first dose, `RFXSTDTC` in DM) to the observation date, applying the **no-Day-0 convention**: the reference day itself is Day 1; days before it are negative.

```{r vs-study-day, exercise=TRUE, exercise.lines=22}
# Add VSDTC from raw VISDAT (simplified parse for tutorial)
vs_with_dtc <- vs |>   # using vs built in Exercise 6 — re-run below if needed
  left_join(vs_raw |>
              mutate(USUBJID = paste("TEST_STUDY", SITEID, PATNUM, sep="-"),
                     VSDTC = as.character(dmy(VISDAT))) |>
              select(USUBJID, VISIT, VSDTC),
            by = c("USUBJID", "VISIT"))

# Derive VSDY using dm_ref for reference dates
vs_with_dy <- derive_study_day(
  data        = vs_with_dtc,
  dm_domain   = dm_ref,
  tgdt        = "VSDTC",
  refdt       = "___",          # first-dose date column in dm_ref
  study_day_var = "VSDY"
)

vs_with_dy |> select(USUBJID, VSTESTCD, VISIT, VSDTC, VSDY)
```

```{r vs-study-day-hint-1}
# refdt = "RFXSTDTC"  (the first-dose date column in dm_ref)
# SCREENING rows should be negative VSDY (before first dose on 2020-07-15)
# WEEK 4 rows should be positive VSDY
```

```{r vs-study-day-solution}
vs_with_dtc <- vs |>
  left_join(vs_raw |>
              mutate(USUBJID = paste("TEST_STUDY", SITEID, PATNUM, sep="-"),
                     VSDTC = as.character(dmy(VISDAT))) |>
              select(USUBJID, VISIT, VSDTC),
            by = c("USUBJID", "VISIT"))

vs_with_dy <- derive_study_day(
  data          = vs_with_dtc,
  dm_domain     = dm_ref,
  tgdt          = "VSDTC",
  refdt         = "RFXSTDTC",
  study_day_var = "VSDY"
)
vs_with_dy |> select(USUBJID, VSTESTCD, VISIT, VSDTC, VSDY)
```

```{r quiz-study-day, echo=FALSE}
quiz(
  caption = "Study day convention",
  question(
    "Under the CDISC no-Day-0 convention, if a subject's first dose is 2020-07-15 and an observation is on 2020-07-15 (same day), VSDY is:",
    answer("0"),
    answer("1 — the reference day itself is Day 1", correct = TRUE),
    answer("-1"),
    answer("NA — Day 0 is undefined"),
    allow_retry = TRUE
  ),
  question(
    "An observation on 2020-07-14 (one day before first dose on 2020-07-15) produces VSDY =",
    answer("0"),
    answer("-1 — one day before Day 1", correct = TRUE),
    answer("1"),
    answer("-2"),
    allow_retry = TRUE
  )
)
```

---

## 8  derive_blfl() — baseline flag

### Step 7 of 8: VSBLFL

`derive_blfl()` flags the **last pre-dose observation** per subject per test as `"Y"`. It checks:
- The observation date ≤ first dose date (`RFXSTDTC`)
- The visit is in a list of qualifying baseline visits
- Among qualifying rows, the chronologically last one gets `"Y"`

```{r vs-blfl, exercise=TRUE, exercise.lines=18}
vs_blfl <- derive_blfl(
  data             = vs_with_dy,
  dm_domain        = dm_ref,
  tgt_var          = "VSBLFL",
  ref_var          = "RFXSTDTC",
  baseline_visits  = c("SCREENING"),
  date_var         = "VSDTC",
  test_var         = "VSTESTCD"
)

vs_blfl |>
  select(USUBJID, VSTESTCD, VISIT, VSDTC, VSDY, VSBLFL) |>
  arrange(USUBJID, VSTESTCD, VISIT)
```

```{r vs-blfl-hint-1}
# Only SCREENING visits (before first dose) should get VSBLFL = "Y"
# WEEK 4 and WEEK 8 visits (after first dose) should be NA
```

---

## 9  VS Capstone — full pipeline in one block

### Exercise — build the entire VS domain end-to-end

This exercise brings together all 8 steps. Fill in the blanks to complete the pipeline.

```{r vs-capstone, exercise=TRUE, exercise.lines=55}
# ── Step 1: OAK IDs ──────────────────────────────────────────────────────────
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "___", raw_src = "vs_raw")

# ── Step 2–3: Topic blocks + bind_rows ───────────────────────────────────────
build_block <- function(rows, rc, tc, tt, u) {
  hardcode_ct(rows, rc, "VSTESTCD", tc, study_ct, "C66741") |>
    left_join(hardcode_ct(rows, rc, "VSTEST",   tt, study_ct, "C67153"), by = oak_id_vars()) |>
    left_join(assign_no_ct(rows, rc, "VSORRES"),                          by = oak_id_vars()) |>
    left_join(hardcode_no_ct(rows, rc, "VSORRESU", u),                    by = oak_id_vars()) |>
    left_join(vs_id |> select(all_of(oak_id_vars()), VISIT, VISDAT),      by = oak_id_vars())
}

vs <- bind_rows(
  build_block(vs_id |> filter(!is.na(SYS_BP)), "SYS_BP","SYSBP","Systolic Blood Pressure","mmHg"),
  build_block(vs_id |> filter(!is.na(DIA_BP)), "DIA_BP","DIABP","Diastolic Blood Pressure","mmHg"),
  build_block(vs_id |> filter(!is.na(PULSE)),  "PULSE", "PULSE","Pulse Rate","beats/min")
)

# ── Step 4: Domain-wide keys ─────────────────────────────────────────────────
vs <- vs |>
  mutate(
    STUDYID = "TEST_STUDY",
    DOMAIN  = "VS",
    USUBJID = paste("TEST_STUDY", patient_number, sep = "-"),
    VSDTC   = as.character(dmy(VISDAT))
  ) |>
  select(-VISDAT)

# ── Step 5: --SEQ ────────────────────────────────────────────────────────────
vs <- derive_seq(vs, tgt_var = "VSSEQ",
                 rec_vars = c("USUBJID", "___", "VISIT"))

# ── Step 6: --DY ─────────────────────────────────────────────────────────────
vs <- derive_study_day(vs, dm_ref, tgdt = "VSDTC",
                       refdt = "RFXSTDTC", study_day_var = "VSDY")

# ── Step 7: --BLFL ───────────────────────────────────────────────────────────
vs <- derive_blfl(vs, dm_ref, tgt_var = "VSBLFL",
                  ref_var = "RFXSTDTC", baseline_visits = c("SCREENING"))

# ── Step 8: Column order ─────────────────────────────────────────────────────
vs_final <- vs |>
  select(STUDYID, DOMAIN, USUBJID, VSSEQ,
         VSTESTCD, VSTEST, VSORRES, VSORRESU,
         VISIT, VSDTC, VSDY, VSBLFL)

vs_final
```

```{r vs-capstone-hint-1}
# Step 1: pat_var = "PATNUM"
# Step 5: rec_vars = c("USUBJID", "VSTESTCD", "VISIT")
```

```{r vs-capstone-solution}
vs_id <- generate_oak_id_vars(vs_raw, pat_var = "PATNUM", raw_src = "vs_raw")
build_block <- function(rows, rc, tc, tt, u) {
  hardcode_ct(rows, rc, "VSTESTCD", tc, study_ct, "C66741") |>
    left_join(hardcode_ct(rows, rc, "VSTEST",   tt, study_ct, "C67153"), by = oak_id_vars()) |>
    left_join(assign_no_ct(rows, rc, "VSORRES"),                          by = oak_id_vars()) |>
    left_join(hardcode_no_ct(rows, rc, "VSORRESU", u),                    by = oak_id_vars()) |>
    left_join(vs_id |> select(all_of(oak_id_vars()), VISIT, VISDAT),      by = oak_id_vars())
}
vs <- bind_rows(
  build_block(vs_id |> filter(!is.na(SYS_BP)), "SYS_BP","SYSBP","Systolic Blood Pressure","mmHg"),
  build_block(vs_id |> filter(!is.na(DIA_BP)), "DIA_BP","DIABP","Diastolic Blood Pressure","mmHg"),
  build_block(vs_id |> filter(!is.na(PULSE)),  "PULSE", "PULSE","Pulse Rate","beats/min")
) |>
  mutate(STUDYID="TEST_STUDY", DOMAIN="VS",
         USUBJID=paste("TEST_STUDY", patient_number, sep="-"),
         VSDTC=as.character(dmy(VISDAT))) |>
  select(-VISDAT)

vs <- derive_seq(vs, "VSSEQ", c("USUBJID","VSTESTCD","VISIT"))
vs <- derive_study_day(vs, dm_ref, "VSDTC", "RFXSTDTC", "VSDY")
vs <- derive_blfl(vs, dm_ref, "VSBLFL", "RFXSTDTC", c("SCREENING"))

vs |> select(STUDYID, DOMAIN, USUBJID, VSSEQ, VSTESTCD, VSTEST, VSORRES, VSORRESU, VISIT, VSDTC, VSDY, VSBLFL)
```

---

## Part 2 — Building AE (Events)

## 10  Events are simpler — one raw row → one SDTM row

### No topic blocks needed

Because there's no transposition, AE is a straight pipeline:

```
attach OAK IDs
→ assign/hardcode each SDTM column
→ mutate domain keys
→ derive_seq → derive_study_day
```

### Exercise 10a — map AE columns

Attach OAK IDs to `ae_raw`, then map these columns:

| SDTM column | Source | Function |
|---|---|---|
| `AETERM` | raw `AETERM` | `assign_no_ct()` |
| `AESEV` | raw `AESEV`, CT `C66742` | `assign_ct()` |
| `AESER` | raw `AESER`, CT `NY` | `assign_ct()` |
| `AESTDTC` | raw `AESTDT` (rename) | `assign_no_ct()` |
| `AEENDTC` | raw `AEENDT` (rename) | `assign_no_ct()` |

```{r ae-map-cols, exercise=TRUE, exercise.lines=30}
ae_id <- generate_oak_id_vars(ae_raw, pat_var = "PATNUM", raw_src = "ae_raw")

ae_term    <- assign_no_ct(ae_id, "AETERM",  "AETERM")
ae_sev     <- assign_ct(ae_id,   "AESEV",   "AESEV",   study_ct, "C66742")
ae_ser     <- assign_ct(ae_id,   "AESER",   "AESER",   study_ct, "NY")
ae_stdt    <- assign_no_ct(ae_id, "AESTDT", "AESTDTC")
ae_endt    <- assign_no_ct(ae_id, "AEENDT", "AEENDTC")
ae_out     <- assign_ct(ae_id,   "___",     "AEOUT",   study_ct, "C66768")

# Join all
ae <- ae_term |>
  left_join(ae_sev,  by = oak_id_vars()) |>
  left_join(ae_ser,  by = oak_id_vars()) |>
  left_join(ae_stdt, by = oak_id_vars()) |>
  left_join(ae_endt, by = oak_id_vars()) |>
  left_join(ae_out,  by = oak_id_vars())

ae
```

```{r ae-map-cols-hint-1}
# ae_out: raw_var = "AEOUT", tgt_var = "AEOUT", ct_clst = "C66768"
```

```{r ae-map-cols-solution}
ae_id <- generate_oak_id_vars(ae_raw, pat_var = "PATNUM", raw_src = "ae_raw")

ae_term <- assign_no_ct(ae_id, "AETERM", "AETERM")
ae_sev  <- assign_ct(ae_id,   "AESEV",  "AESEV",  study_ct, "C66742")
ae_ser  <- assign_ct(ae_id,   "AESER",  "AESER",  study_ct, "NY")
ae_stdt <- assign_no_ct(ae_id, "AESTDT","AESTDTC")
ae_endt <- assign_no_ct(ae_id, "AEENDT","AEENDTC")
ae_out  <- assign_ct(ae_id,   "AEOUT", "AEOUT",   study_ct, "C66768")

ae <- ae_term |>
  left_join(ae_sev,  by = oak_id_vars()) |>
  left_join(ae_ser,  by = oak_id_vars()) |>
  left_join(ae_stdt, by = oak_id_vars()) |>
  left_join(ae_endt, by = oak_id_vars()) |>
  left_join(ae_out,  by = oak_id_vars())
ae
```

### Exercise 10b — domain keys, AESEQ, and AESTDY

Add `STUDYID`, `DOMAIN`, `USUBJID`, `AESEQ`, and `AESTDY` to the AE domain.

```{r ae-keys-seq, exercise=TRUE, exercise.lines=20}
ae_full <- ae |>
  mutate(
    STUDYID = "TEST_STUDY",
    DOMAIN  = "AE",
    USUBJID = paste("TEST_STUDY", patient_number, sep = "-")
  ) |>
  derive_seq(tgt_var = "AESEQ",
             rec_vars = c("USUBJID", "AETERM", "AESTDTC")) |>
  derive_study_day(
    dm_domain     = dm_ref,
    tgdt          = "AESTDTC",
    refdt         = "___",
    study_day_var = "AESTDY"
  )

ae_full |>
  select(STUDYID, DOMAIN, USUBJID, AESEQ, AETERM, AESTDTC, AESTDY, AESEV, AESER)
```

```{r ae-keys-seq-hint-1}
# refdt = "RFSTDTC"  (reference start date in dm_ref)
```

```{r ae-keys-seq-solution}
ae_full <- ae |>
  mutate(
    STUDYID = "TEST_STUDY",
    DOMAIN  = "AE",
    USUBJID = paste("TEST_STUDY", patient_number, sep = "-")
  ) |>
  derive_seq(tgt_var = "AESEQ",
             rec_vars = c("USUBJID", "AETERM", "AESTDTC")) |>
  derive_study_day(
    dm_domain     = dm_ref,
    tgdt          = "AESTDTC",
    refdt         = "RFSTDTC",
    study_day_var = "AESTDY"
  )
ae_full |>
  select(STUDYID, DOMAIN, USUBJID, AESEQ, AETERM, AESTDTC, AESTDY, AESEV, AESER)
```

---

## 11  Multiple raw sources → one domain

### Pattern: process each source independently, then bind_rows

Some studies collect AEs on a main CRF and serious AEs on a separate SAE CRF. Each source gets its own `generate_oak_id_vars()` with a different `raw_src` label.

```{r multi-source, exercise=TRUE, exercise.lines=28}
# Source 1: main AE
ae_main_id <- generate_oak_id_vars(ae_raw,  "PATNUM", raw_src = "ae_main")
# Source 2: SAE supplement
sae_id     <- generate_oak_id_vars(sae_raw, "PATNUM", raw_src = "ae_sae")

# Build each source (AETERM only for brevity)
ae_from_main <- assign_no_ct(ae_main_id, "AETERM", "AETERM") |>
  mutate(USUBJID = paste("TEST_STUDY", ae_main_id$SITEID, ae_main_id$PATNUM, sep="-"))

ae_from_sae  <- assign_no_ct(sae_id,    "AETERM", "AETERM") |>
  mutate(USUBJID = paste("TEST_STUDY", sae_id$SITEID, sae_id$PATNUM, sep="-"))

# Combine — raw_source tells us which CRF each row came from
ae_combined <- bind_rows(ae_from_main, ae_from_sae)

ae_combined |>
  select(raw_source, patient_number, USUBJID, AETERM) |>
  arrange(USUBJID)
```

```{r multi-source-hint-1}
# raw_source distinguishes "ae_main" from "ae_sae" rows
# After bind_rows(), derive_seq() would renumber AESEQ across both sources together
# The SAE row for PATNUM 1034 appears twice: once from ae_main, once from ae_sae
# In production you'd deduplicate before deriving AESEQ
```

---

## 12  Comparing to pharmaversesdtm

### Validate your output against the canonical SDTM

```{r vs-canonical-compare, exercise=TRUE, exercise.lines=20}
# ae_full is our derived AE; ae_sdtm_ref mirrors pharmaversesdtm::ae
my_cols      <- names(ae_full)
canonical_cols <- names(ae_sdtm_ref)

cat("In canonical but NOT in our output (still to derive):\n")
print(setdiff(canonical_cols, my_cols))

cat("\nIn our output but NOT in canonical (OAK-internal columns):\n")
print(setdiff(my_cols, canonical_cols))

# Row count match?
cat("\nOur rows:", nrow(ae_full), "  Canonical rows:", nrow(ae_sdtm_ref), "\n")

# AESEQ and AETERM values match?
inner_join(
  ae_full      |> select(USUBJID, AESEQ, AETERM),
  ae_sdtm_ref  |> select(USUBJID, AESEQ, AETERM),
  by = c("USUBJID", "AESEQ"),
  suffix = c("_ours", "_canonical")
) |>
  mutate(MATCH = AETERM_ours == AETERM_canonical)
```

```{r vs-canonical-compare-hint-1}
# Columns in canonical not in ours: likely AEDECOD, AEBODSYS, AESOC (need MedDRA)
# OAK-internal cols in ours: oak_id, raw_source, patient_number
# Row counts and AETERM/AESEQ should align perfectly
```

---

## 13  Common Gotchas

```{r quiz-gotchas, echo=FALSE}
quiz(
  caption = "Common pitfalls",
  question(
    "You forget id_vars = oak_id_vars() on the second call in a topic block. What likely happens?",
    answer("The function errors with 'argument id_vars missing'"),
    answer("The merge uses the default id_vars and may produce a many-to-many join or misaligned rows", correct = TRUE),
    answer("The second column is simply added as a new column with no join"),
    answer("Nothing — id_vars is optional and has no effect"),
    allow_retry = TRUE
  ),
  question(
    "In a Findings domain hardcode call like hardcode_ct(vs_raw, 'SYS_BP', 'VSTESTCD', 'SYSBP', ...), using raw_var = 'patient_number' instead of 'SYS_BP' would cause:",
    answer("A CT validation error because patient_number is not in the codelist"),
    answer("An SDTM row to be created for EVERY raw row, even visits where SYS_BP was not recorded", correct = TRUE),
    answer("No change — raw_var is ignored for hardcode calls"),
    answer("The VSTESTCD column to be filled with patient numbers"),
    allow_retry = TRUE
  ),
  question(
    "After assign_datetime() some VSDTC rows are NA. The correct next step is:",
    answer("Filter out the NA rows and continue"),
    answer("Re-run with na.rm = TRUE"),
    answer("Call problems() to see which rows failed parsing and fix the raw_fmt specification", correct = TRUE),
    answer("Hardcode NA dates to '9999-12-31'"),
    allow_retry = TRUE
  ),
  question(
    "Two AE raw datasets (ae_main and sae_raw) are processed separately. Both use generate_oak_id_vars with different raw_src values. After bind_rows, why might the same AE event appear twice?",
    answer("bind_rows always deduplicates by AETERM"),
    answer("Because the same serious AE is recorded on both the main CRF and the SAE CRF — deduplication must be applied before derive_seq", correct = TRUE),
    answer("derive_seq will automatically remove duplicates"),
    answer("Different raw_src values prevent duplicates from appearing"),
    allow_retry = TRUE
  )
)
```

---

## 14  Self-check Quiz

```{r final-quiz-09, echo=FALSE}
quiz(
  caption = "Lesson 09 — self-check",

  question(
    "Why does a Findings domain like VS need a 'topic block + bind_rows' pattern?",
    answer("CDISC requires it for all non-DM domains"),
    answer("Raw VS data is wide (one row per visit with multiple BP columns); SDTM VS is long (one row per measurement), so the wide columns must be individually processed and stacked", correct = TRUE),
    answer("bind_rows is needed to attach multiple OAK identifier sets"),
    answer("Each topic block maps a different subject"),
    allow_retry = TRUE
  ),
  question(
    "In derive_seq(), rec_vars = c('USUBJID', 'VSTESTCD', 'VISIT') means the sequence number resets to 1 when:",
    answer("Any column in rec_vars changes — i.e. for each unique USUBJID (regardless of test or visit)"),
    answer("USUBJID changes — a new subject always starts at 1"),
    answer("Within each USUBJID, rows are numbered 1, 2, 3 … sorted by the rec_vars combination", correct = TRUE),
    answer("VSTESTCD changes"),
    allow_retry = TRUE
  ),
  question(
    "An AE starts on 2020-07-14 and the reference start date (RFSTDTC) is 2020-07-16. AESTDY =",
    answer("-1"),
    answer("-2 — two days before Day 1", correct = TRUE),
    answer("1"),
    answer("0"),
    allow_retry = TRUE
  ),
  question(
    "AEDECOD and AEBODSYS are absent from your OAK AE pipeline. Why?",
    answer("They are optional SDTM variables not required for submissions"),
    answer("They come from external MedDRA medical coding — not from the raw CRF data or OAK mapping", correct = TRUE),
    answer("They are derived by derive_seq() in a later step"),
    answer("sdtm.oak does not support Events domains yet"),
    allow_retry = TRUE
  ),
  question(
    "The raw_source column set by generate_oak_id_vars() is most useful when:",
    answer("Validating CT spec conformance"),
    answer("Two raw datasets feed one SDTM domain — raw_source identifies which CRF each row came from for provenance and QC", correct = TRUE),
    answer("Computing study day"),
    answer("Deriving USUBJID from PATNUM"),
    allow_retry = TRUE
  )
)
```

---

## Key Takeaways

1. **Four SDTM classes**: Findings (wide→long, topic blocks), Events (one-to-one), Interventions (one-to-one), Findings About (rare).
2. **Findings pipeline = topic blocks + bind_rows**: each raw measurement column becomes a separate block of rows, then all blocks are stacked.
3. **Events pipeline** is simpler: assign/hardcode each SDTM column, then domain keys, `derive_seq()`, `derive_study_day()`.
4. **`derive_seq()`** assigns a 1-based row number within USUBJID — the choice of `rec_vars` determines uniqueness.
5. **`derive_study_day()`** implements the CDISC no-Day-0 convention: reference day = Day 1, days before = negative.
6. **`derive_blfl()`** flags the last pre-dose observation per subject per test — encodes the full CDISC baseline rule.
7. **Multiple raw sources** → process independently, `bind_rows()`, deduplicate before `derive_seq()`. The `raw_source` column tracks provenance.
8. **Validate against `pharmaversesdtm`** using `setdiff(names(...))` and `inner_join()` comparisons.

**Next** → Lesson 10: `{sdtm.oak}` Part 3 — SUPP-- domains and RELREC
(`rmarkdown::run("10_sdtm_oak_part3.Rmd")`)
