---
title: "Lesson 10 — sdtm.oak Part 3: SUPP-- Domains and RELREC"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build SUPP-- (Supplemental Qualifier) datasets, understand RELREC structure,
  and handle non-standard SDTM constructs using dplyr patterns.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 10_sdtm_oak_part3.Rmd
# PURPOSE       : sdtm.oak Part 3 — RELREC domain for related records, define-XML
#                 generation with xportr, and submission-ready XPT
#                 export with variable-level metadata.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(pharmaversesdtm)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── Mock sdtm.oak functions (same as Lessons 08/09) ──────────────────────────

oak_id_vars <- function() c("oak_id", "raw_source", "patient_number")

generate_oak_id_vars <- function(raw_dat, pat_var, raw_src) {
  raw_dat |>
    mutate(
      oak_id        = row_number(),
      raw_source    = raw_src,
      patient_number = .data[[pat_var]]
    )
}

assign_no_ct <- function(raw_dat, raw_var, tgt_var, id_vars = oak_id_vars()) {
  raw_dat |>
    select(any_of(c(id_vars, raw_var))) |>
    rename(!!sym(tgt_var) := !!sym(raw_var))
}

hardcode_no_ct <- function(raw_dat, raw_var, tgt_var, tgt_val, id_vars = oak_id_vars()) {
  raw_dat |>
    filter(!is.na(.data[[raw_var]])) |>
    select(any_of(id_vars)) |>
    mutate(!!sym(tgt_var) := tgt_val)
}

assign_ct <- function(raw_dat, raw_var, tgt_var, ct_spec, ct_clst, id_vars = oak_id_vars()) {
  lookup <- ct_spec |>
    filter(codelist_code == ct_clst) |>
    select(collected_value, term_value) |>
    distinct()
  raw_dat |>
    select(any_of(c(id_vars, raw_var))) |>
    left_join(lookup, by = setNames("collected_value", raw_var)) |>
    select(-any_of(raw_var)) |>
    rename(!!sym(tgt_var) := term_value)
}

hardcode_ct <- function(raw_dat, raw_var, tgt_var, tgt_val, ct_spec, ct_clst, id_vars = oak_id_vars()) {
  valid <- ct_spec |> filter(codelist_code == ct_clst) |> pull(term_value) |> unique()
  if (!tgt_val %in% valid)
    warning("hardcode_ct: '", tgt_val, "' not in codelist ", ct_clst)
  hardcode_no_ct(raw_dat, raw_var, tgt_var, tgt_val, id_vars)
}

condition_add <- function(raw_dat, condition) {
  cond <- enquo(condition)
  raw_dat |> mutate(.oak_condition = !!cond)
}

apply_condition <- function(conditioned_dat) {
  conditioned_dat |> filter(.oak_condition) |> select(-.oak_condition)
}

derive_seq <- function(data, tgt_var, rec_vars = "USUBJID") {
  data |>
    arrange(across(all_of(rec_vars))) |>
    group_by(USUBJID) |>
    mutate(!!sym(tgt_var) := row_number()) |>
    ungroup()
}

# ── Mock generate_sdtm_supp ───────────────────────────────────────────────────

generate_sdtm_supp <- function(parent_dataset, raw_dat, id_var, qnam_definitions) {
  rows <- lapply(qnam_definitions, function(qdef) {
    parent_dataset |>
      select(STUDYID, USUBJID, all_of(id_var), patient_number) |>
      inner_join(
        raw_dat |>
          select(patient_number, any_of(qdef$raw_var)),
        by = "patient_number"
      ) |>
      filter(!is.na(.data[[qdef$raw_var]]) & .data[[qdef$raw_var]] != "") |>
      transmute(
        STUDYID,
        RDOMAIN  = substr(id_var, 1, 2),
        USUBJID,
        IDVAR    = id_var,
        IDVARVAL = as.character(.data[[id_var]]),
        QNAM     = qdef$qnam,
        QLABEL   = qdef$qlabel,
        QVAL     = as.character(.data[[qdef$raw_var]]),
        QORIG    = qdef$qorig,
        QEVAL    = qdef$qeval
      )
  })
  bind_rows(rows)
}

# ── Sample datasets ───────────────────────────────────────────────────────────

# Raw AE data with extra sponsor-specific columns
ae_raw <- tibble(
  PATNUM       = c(1015L, 1015L, 1023L, 1042L, 1042L),
  STUDYID      = "TEST_STUDY",
  SITEID       = c("701", "701", "702", "703", "703"),
  AETERM       = c("HEADACHE", "NAUSEA", "FATIGUE", "DIZZINESS", "RASH"),
  AESTDT       = c("2020-07-15", "2020-08-01", "2020-07-20", "2020-09-10", "2020-09-12"),
  AEENDT       = c("2020-07-17", "2020-08-03", "2020-07-25", "2020-09-12", "2020-09-20"),
  AESEV        = c("MILD", "MODERATE", "MILD", "SEVERE", "MILD"),
  AESER        = c("N", "N", "N", "Y", "N"),
  AEOUT        = c("RECOVERED/RESOLVED", "RECOVERED/RESOLVED",
                   "RECOVERED/RESOLVED", "RECOVERING/RESOLVING", "RECOVERED/RESOLVED"),
  SPONSOR_AE_ID = c("AE001", "AE002", "AE003", "AE004", "AE005"),
  INV_AE_CAT   = c("NEUROLOGICAL", "GI", "GENERAL", "NEUROLOGICAL", "DERMATOLOGICAL"),
  AE_SAFETY_MTG = c("Y", "N", "N", "Y", "N")
)

# Raw EX data with sponsor link column
ex_raw <- tibble(
  PATNUM     = c(1015L, 1015L, 1023L, 1042L),
  STUDYID    = "TEST_STUDY",
  SITEID     = c("701", "701", "702", "703"),
  EXTRT      = "STUDY DRUG",
  EXDOSE     = c(100, 100, 50, 100),
  EXDOSU     = "mg",
  EXSTDT     = c("2020-07-01", "2020-08-01", "2020-07-05", "2020-09-01"),
  EXENDT     = c("2020-07-31", "2020-08-31", "2020-07-31", "2020-09-30"),
  EXSPID     = c("EX001", "EX002", "EX003", "EX004")
)

# AE SDTM (pre-built, as if from Lesson 09)
ae <- tibble(
  STUDYID  = "TEST_STUDY",
  DOMAIN   = "AE",
  USUBJID  = c("TEST_STUDY-701-1015", "TEST_STUDY-701-1015",
               "TEST_STUDY-702-1023", "TEST_STUDY-703-1042", "TEST_STUDY-703-1042"),
  AESEQ    = c(1L, 2L, 1L, 1L, 2L),
  AETERM   = c("HEADACHE", "NAUSEA", "FATIGUE", "DIZZINESS", "RASH"),
  AESTDTC  = c("2020-07-15", "2020-08-01", "2020-07-20", "2020-09-10", "2020-09-12"),
  AEENDTC  = c("2020-07-17", "2020-08-03", "2020-07-25", "2020-09-12", "2020-09-20"),
  AESEV    = c("MILD", "MODERATE", "MILD", "SEVERE", "MILD"),
  AESER    = c("N", "N", "N", "Y", "N"),
  AEOUT    = c("RECOVERED/RESOLVED", "RECOVERED/RESOLVED",
               "RECOVERED/RESOLVED", "RECOVERING/RESOLVING", "RECOVERED/RESOLVED"),
  patient_number = c(1015L, 1015L, 1023L, 1042L, 1042L)
)

# EX SDTM (pre-built)
ex <- tibble(
  STUDYID  = "TEST_STUDY",
  DOMAIN   = "EX",
  USUBJID  = c("TEST_STUDY-701-1015", "TEST_STUDY-701-1015",
               "TEST_STUDY-702-1023", "TEST_STUDY-703-1042"),
  EXSEQ    = c(1L, 2L, 1L, 1L),
  EXTRT    = "STUDY DRUG",
  EXDOSE   = c(100, 100, 50, 100),
  EXDOSU   = "mg",
  EXSTDTC  = c("2020-07-01", "2020-08-01", "2020-07-05", "2020-09-01"),
  EXENDTC  = c("2020-07-31", "2020-08-31", "2020-07-31", "2020-09-30"),
  EXSPID   = c("EX001", "EX002", "EX003", "EX004"),
  patient_number = c(1015L, 1015L, 1023L, 1042L)
)

# AE-to-EX link table (for RELREC)
ae_ex_link <- tibble(
  PATNUM        = c(1015L, 1042L),
  AE_SPONSOR_ID = c("AE001", "AE004"),
  EX_SPONSOR_ID = c("EX001", "EX004")
)

# ── Real pharmaversesdtm SUPP-- data (for comparison exercises) ───────────────
suppae_real <- pharmaversesdtm::suppae

# Study CT spec
study_ct <- tribble(
  ~codelist_code, ~collected_value, ~term_value,
  "NY",           "Y",              "Y",
  "NY",           "N",              "N",
  "AESEV",        "MILD",           "MILD",
  "AESEV",        "MODERATE",       "MODERATE",
  "AESEV",        "SEVERE",         "SEVERE"
)
```

## Welcome

This is Lesson 10 of the pharmaverse tutorial series — **sdtm.oak Part 3: SUPP-- Domains and RELREC**.

You've built standard SDTM domains. Now we handle the edge cases: what happens when your CRF collects variables that don't belong in any standard SDTM domain? And how do you formally link records *across* domains?

This lesson covers:

- The **SUPP--** (Supplemental Qualifier) dataset — shape, purpose, and when to use it
- Building SUPP-- both automatically with `generate_sdtm_supp()` and manually with `dplyr`
- **RELREC** — the Related Records dataset, built manually with `dplyr`
- **CO** domain (Comments) and trial design domains — what they are and how to build them
- Validation rules that are unique to these structures

All `sdtm.oak` functions in these exercises are mocked — they run without any package installed and produce identical output to the real functions.

---

## 1 — The "extra variables" problem

### Why SUPP-- exists

SDTM standardizes variable names and meanings: every AE has AETERM, AESEV, AESER, etc. But real studies collect more than the standard allows. A sponsor might track:

- Internal severity grades (0–5 ordinal) not covered by the AESEV codelist
- Investigator narratives
- Sponsor-specific classifications or flags
- Custom dates

These variables matter for analysis but have no home in the standard domain. SDTM's answer is **SUPP--** (Supplemental Qualifier datasets): `SUPPAE`, `SUPPDM`, `SUPPLB`, etc.

```{r supp-intro-demo}
# Peek at what "extra" columns our ae_raw has beyond standard SDTM
setdiff(names(ae_raw), c("PATNUM", "STUDYID", "SITEID",
                          "AETERM", "AESTDT", "AEENDT",
                          "AESEV", "AESER", "AEOUT"))
```

Those three extra columns (`SPONSOR_AE_ID`, `INV_AE_CAT`, `AE_SAFETY_MTG`) are SUPP-- candidates — they were collected on the CRF but don't belong in the AE domain.

### Exercise 1b — programmatically identify SUPP-- candidates

In practice, raw datasets can have dozens of columns. Rather than inspecting them manually, use `setdiff()` to compare raw column names against the known standard SDTM AE variable list. Anything left over is a SUPP-- candidate.

```{r supp-autodetect, exercise=TRUE, exercise.lines=18}
# Standard SDTM AE variables (from SDTMIG v3.4)
sdtm_ae_vars <- c(
  "STUDYID", "DOMAIN", "USUBJID", "AESEQ", "AESPID",
  "AETERM", "AELLT", "AELLTCD", "AEDECOD", "AEPTCD",
  "AEHLT", "AEHLTCD", "AEHLGT", "AEHLGTCD", "AEBODSYS",
  "AEBDSYCD", "AESOC", "AESOCCD", "AESEV", "AESER",
  "AEACN", "AEREL", "AEOUT", "AESCAN", "AESCONG",
  "AESDISAB", "AESDTH", "AESHOSP", "AESLIFE", "AESOD",
  "AESMIE", "AECONTRT", "AESTDTC", "AEENDTC", "AESTDY",
  "AEENDY", "AEDUR", "AETOXGR", "EPOCH"
)

# Raw columns that are NOT in the SDTM AE standard → SUPP-- candidates
raw_ae_cols <- names(ae_raw)

supp_candidates <- setdiff(raw_ae_cols, c(sdtm_ae_vars, "PATNUM", "SITEID",
                                           "AESTDT", "AEENDT"))  # raw-format names
cat("SUPP-- candidates found in ae_raw:\n")
print(supp_candidates)

# How many?
cat("\nTotal SUPP-- candidates:", length(supp_candidates), "\n")
```

```{r supp-autodetect-hint-1}
# setdiff(raw_ae_cols, known_vars) returns raw columns with no SDTM home.
# PATNUM, SITEID are raw identifiers (become part of USUBJID — not SUPP--).
# AESTDT / AEENDT are raw date formats (become AESTDTC / AEENDTC — not SUPP--).
# The three remaining columns are genuine SUPP-- candidates.
```

---

## 2 — The SUPP-- shape

### Fixed long format

A SUPP-- dataset has a fixed column structure defined by SDTMIG:

| Column | Meaning |
|---|---|
| `STUDYID` | Study identifier |
| `RDOMAIN` | Parent domain (e.g., `"AE"`) |
| `USUBJID` | Subject |
| `IDVAR` | Variable in parent that identifies the row (e.g., `"AESEQ"`) |
| `IDVARVAL` | Value of that variable for this row (e.g., `"1"`) |
| `QNAM` | Qualifier name (≤8 chars, uppercase) |
| `QLABEL` | Human-readable label for QNAM |
| `QVAL` | The value (always character) |
| `QORIG` | Origin: `"CRF"`, `"DERIVED"`, `"ASSIGNED"` |
| `QEVAL` | Evaluator: `"INVESTIGATOR"`, `"SPONSOR"`, etc. |

**Key rule**: instead of adding non-standard columns to `ae`, you create rows in `suppae` — one row per (subject × parent record × non-standard variable).

```{r supp-shape-demo}
# A hand-built example of what a SUPPAE row looks like
tibble(
  STUDYID  = "TEST_STUDY",
  RDOMAIN  = "AE",
  USUBJID  = "TEST_STUDY-701-1015",
  IDVAR    = "AESEQ",
  IDVARVAL = "1",
  QNAM     = "AESPID",
  QLABEL   = "Sponsor-specific AE ID",
  QVAL     = "AE001",
  QORIG    = "CRF",
  QEVAL    = "INVESTIGATOR"
)
```

### Subject-level vs record-level

- **Record-level**: IDVAR + IDVARVAL point to a specific row in the parent. Example: a comment about AE #2.
- **Subject-level**: IDVAR and IDVARVAL are `""` (empty string). The qualifier applies to the subject as a whole. Common in SUPPDM — e.g., a stratification group that doesn't fit any DM variable.

---

## 3 — What goes to SUPP-- vs the parent domain?

```{r quiz-supp-vs-parent, echo=FALSE}
quiz(
  question(
    "Your CRF has a field 'AEBODSYS' (Body System). Where does it go?",
    answer("SUPPAE — it's non-standard"),
    answer("AE domain — AEBODSYS is in the SDTMIG", correct = TRUE),
    answer("It depends on the sponsor"),
    random_answer_order = TRUE
  ),
  question(
    "Your CRF has a field 'INV_AE_GRADE' (Investigator grade 0-5). Where does it go?",
    answer("AE domain — it's about adverse events"),
    answer("SUPPAE — it's not in the SDTMIG AE variable list", correct = TRUE),
    answer("You just drop it"),
    random_answer_order = TRUE
  ),
  question(
    "A SUPPAE row with blank IDVAR and IDVARVAL means:",
    answer("It's a record-level qualifier for a specific AE"),
    answer("It's a subject-level qualifier, not tied to a specific AE row", correct = TRUE),
    answer("The IDVAR was not collected"),
    random_answer_order = TRUE
  )
)
```

---

## 4 — Building SUPP-- manually with dplyr

The manual approach is the most transparent. Walk through it once to understand exactly what the structure requires.

**Strategy**: join the parent domain (which has AESEQ) to the raw data (which has the non-standard values), then `pivot_longer()` to go from wide to long, then add the fixed SUPP-- columns.

```{r supp-manual-demo}
# Step 1: Attach non-standard raw columns to the parent using patient_number
ae_with_extras <- ae |>
  select(STUDYID, USUBJID, AESEQ, patient_number) |>
  inner_join(
    ae_raw |>
      generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw") |>
      select(patient_number, SPONSOR_AE_ID, INV_AE_CAT),
    by = "patient_number"
  )

ae_with_extras
```

```{r supp-manual-pivot}
# Step 2: pivot_longer to get one row per QNAM
ae_long <- ae_with_extras |>
  pivot_longer(
    cols      = c(SPONSOR_AE_ID, INV_AE_CAT),
    names_to  = "QNAM",
    values_to = "QVAL"
  ) |>
  filter(!is.na(QVAL) & QVAL != "")

ae_long
```

```{r supp-manual-final}
# Step 3: add fixed SUPP-- columns
suppae_manual <- ae_long |>
  mutate(
    RDOMAIN  = "AE",
    IDVAR    = "AESEQ",
    IDVARVAL = as.character(AESEQ),
    QLABEL   = case_when(
      QNAM == "SPONSOR_AE_ID" ~ "Sponsor-specific AE ID",
      QNAM == "INV_AE_CAT"    ~ "Investigator AE Category"
    ),
    QORIG = "CRF",
    QEVAL = "INVESTIGATOR"
  ) |>
  select(STUDYID, RDOMAIN, USUBJID, IDVAR, IDVARVAL,
         QNAM, QLABEL, QVAL, QORIG, QEVAL)

suppae_manual
```

### Exercise: Add a third qualifier to SUPPAE

The `ae_raw` dataset also has `AE_SAFETY_MTG` — whether the AE was discussed at a safety review meeting. Add it to the SUPPAE dataset. Start from `ae_with_extras` above but include `AE_SAFETY_MTG` in the `pivot_longer()`.

```{r supp-three-cols, exercise=TRUE, exercise.lines=30}
# Modify this to also include AE_SAFETY_MTG
ae_with_extras3 <- ae |>
  select(STUDYID, USUBJID, AESEQ, patient_number) |>
  inner_join(
    ae_raw |>
      generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw") |>
      select(patient_number, SPONSOR_AE_ID, INV_AE_CAT, AE_SAFETY_MTG),
    by = "patient_number"
  )

# Now pivot all three columns and add SUPP-- structure
suppae_3 <- ae_with_extras3 |>
  pivot_longer(
    cols      = _____________,   # include all three columns
    names_to  = "QNAM",
    values_to = "QVAL"
  ) |>
  filter(!is.na(QVAL) & QVAL != "") |>
  mutate(
    RDOMAIN  = "AE",
    IDVAR    = "AESEQ",
    IDVARVAL = as.character(AESEQ),
    QLABEL   = case_when(
      QNAM == "SPONSOR_AE_ID"  ~ "Sponsor-specific AE ID",
      QNAM == "INV_AE_CAT"     ~ "Investigator AE Category",
      QNAM == "AE_SAFETY_MTG"  ~ _____________   # fill in the label
    ),
    QORIG = "CRF",
    QEVAL = "INVESTIGATOR"
  ) |>
  select(STUDYID, RDOMAIN, USUBJID, IDVAR, IDVARVAL,
         QNAM, QLABEL, QVAL, QORIG, QEVAL)

suppae_3
```

```{r supp-three-cols-hint-1}
# In pivot_longer, list all three column names:
# cols = c(SPONSOR_AE_ID, INV_AE_CAT, AE_SAFETY_MTG)
# For QLABEL:
# QNAM == "AE_SAFETY_MTG" ~ "Safety Review Meeting Flag"
```

```{r supp-three-cols-solution}
ae_with_extras3 <- ae |>
  select(STUDYID, USUBJID, AESEQ, patient_number) |>
  inner_join(
    ae_raw |>
      generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw") |>
      select(patient_number, SPONSOR_AE_ID, INV_AE_CAT, AE_SAFETY_MTG),
    by = "patient_number"
  )

suppae_3 <- ae_with_extras3 |>
  pivot_longer(
    cols      = c(SPONSOR_AE_ID, INV_AE_CAT, AE_SAFETY_MTG),
    names_to  = "QNAM",
    values_to = "QVAL"
  ) |>
  filter(!is.na(QVAL) & QVAL != "") |>
  mutate(
    RDOMAIN  = "AE",
    IDVAR    = "AESEQ",
    IDVARVAL = as.character(AESEQ),
    QLABEL   = case_when(
      QNAM == "SPONSOR_AE_ID"  ~ "Sponsor-specific AE ID",
      QNAM == "INV_AE_CAT"     ~ "Investigator AE Category",
      QNAM == "AE_SAFETY_MTG"  ~ "Safety Review Meeting Flag"
    ),
    QORIG = "CRF",
    QEVAL = "INVESTIGATOR"
  ) |>
  select(STUDYID, RDOMAIN, USUBJID, IDVAR, IDVARVAL,
         QNAM, QLABEL, QVAL, QORIG, QEVAL)

suppae_3
```

---

## 5 — Subject-level SUPPDM

Subject-level SUPP-- rows have blank IDVAR and IDVARVAL — the qualifier applies to the whole subject, not a specific record.

A DM domain rarely has a sequential key the way AE does. For subject-level items, set both `IDVAR = ""` and `IDVARVAL = ""`.

### Exercise: Build a subject-level SUPPDM

Suppose `dm_raw` has a column `STRAT_GRP` (stratification group) that doesn't map to any standard DM variable. Build a SUPPDM with one row per subject.

```{r suppdm-subject, exercise=TRUE, exercise.lines=28}
# Mock DM datasets
dm_raw <- tibble(
  PATNUM    = c(1015L, 1023L, 1042L),
  STUDYID   = "TEST_STUDY",
  SITEID    = c("701", "702", "703"),
  BIRTHDT   = c("1975-03-12", "1962-08-22", "1988-11-05"),
  SEX       = c("M", "F", "M"),
  RACE      = c("WHITE", "ASIAN", "WHITE"),
  STRAT_GRP = c("A", "B", "A")   # <- non-standard, goes to SUPPDM
)

dm <- tibble(
  STUDYID = "TEST_STUDY",
  DOMAIN  = "DM",
  USUBJID = c("TEST_STUDY-701-1015", "TEST_STUDY-702-1023", "TEST_STUDY-703-1042"),
  patient_number = c(1015L, 1023L, 1042L)
)

# Build SUPPDM: subject-level (IDVAR = "", IDVARVAL = "")
suppdm <- dm |>
  select(STUDYID, USUBJID, patient_number) |>
  inner_join(
    dm_raw |> select(PATNUM, STRAT_GRP) |> rename(patient_number = PATNUM),
    by = "patient_number"
  ) |>
  filter(!is.na(STRAT_GRP)) |>
  transmute(
    STUDYID,
    RDOMAIN  = ___,
    USUBJID,
    IDVAR    = ___,        # subject-level: blank
    IDVARVAL = ___,        # subject-level: blank
    QNAM     = "STRAGRP",
    QLABEL   = "Stratification Group",
    QVAL     = STRAT_GRP,
    QORIG    = "CRF",
    QEVAL    = "INVESTIGATOR"
  )

suppdm
```

```{r suppdm-subject-hint-1}
# RDOMAIN = "DM"
# IDVAR   = ""   (empty string — subject level)
# IDVARVAL = ""  (empty string — subject level)
```

```{r suppdm-subject-solution}
dm_raw <- tibble(
  PATNUM    = c(1015L, 1023L, 1042L),
  STUDYID   = "TEST_STUDY",
  SITEID    = c("701", "702", "703"),
  BIRTHDT   = c("1975-03-12", "1962-08-22", "1988-11-05"),
  SEX       = c("M", "F", "M"),
  RACE      = c("WHITE", "ASIAN", "WHITE"),
  STRAT_GRP = c("A", "B", "A")
)

dm <- tibble(
  STUDYID = "TEST_STUDY",
  DOMAIN  = "DM",
  USUBJID = c("TEST_STUDY-701-1015", "TEST_STUDY-702-1023", "TEST_STUDY-703-1042"),
  patient_number = c(1015L, 1023L, 1042L)
)

suppdm <- dm |>
  select(STUDYID, USUBJID, patient_number) |>
  inner_join(
    dm_raw |> select(PATNUM, STRAT_GRP) |> rename(patient_number = PATNUM),
    by = "patient_number"
  ) |>
  filter(!is.na(STRAT_GRP)) |>
  transmute(
    STUDYID,
    RDOMAIN  = "DM",
    USUBJID,
    IDVAR    = "",
    IDVARVAL = "",
    QNAM     = "STRAGRP",
    QLABEL   = "Stratification Group",
    QVAL     = STRAT_GRP,
    QORIG    = "CRF",
    QEVAL    = "INVESTIGATOR"
  )

suppdm
```

---

## 6 — `generate_sdtm_supp()` (sdtm.oak automation)

`sdtm.oak` v0.2.0+ provides `generate_sdtm_supp()` to automate the dplyr pattern above. You pass the parent dataset, the raw data, the row-identifier variable, and a list of qualifier definitions.

```{r supp-oak-demo}
# Using generate_sdtm_supp() — mocked here with identical behavior
suppae_oak <- generate_sdtm_supp(
  parent_dataset  = ae,
  raw_dat         = ae_raw |> generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw"),
  id_var          = "AESEQ",
  qnam_definitions = list(
    list(
      qnam    = "AESPID",
      qlabel  = "Sponsor-specific AE ID",
      raw_var = "SPONSOR_AE_ID",
      qorig   = "CRF",
      qeval   = "INVESTIGATOR"
    ),
    list(
      qnam    = "AESCAT",
      qlabel  = "Investigator AE Category",
      raw_var = "INV_AE_CAT",
      qorig   = "CRF",
      qeval   = "INVESTIGATOR"
    )
  )
)

suppae_oak
```

### Exercise: Add the safety meeting qualifier via `generate_sdtm_supp()`

Call `generate_sdtm_supp()` with three qualifiers: `AESPID`, `AESCAT`, and a new one for `AE_SAFETY_MTG` (QNAM = `"AESMTG"`, label = `"Safety Review Meeting Flag"`).

```{r supp-oak-three, exercise=TRUE, exercise.lines=30}
ae_raw_with_ids <- ae_raw |>
  generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw")

suppae_three <- generate_sdtm_supp(
  parent_dataset  = ae,
  raw_dat         = ae_raw_with_ids,
  id_var          = "AESEQ",
  qnam_definitions = list(
    list(
      qnam    = "AESPID",
      qlabel  = "Sponsor-specific AE ID",
      raw_var = "SPONSOR_AE_ID",
      qorig   = "CRF",
      qeval   = "INVESTIGATOR"
    ),
    list(
      qnam    = "AESCAT",
      qlabel  = "Investigator AE Category",
      raw_var = "INV_AE_CAT",
      qorig   = "CRF",
      qeval   = "INVESTIGATOR"
    ),
    # Add the third qualifier here
    list(
      qnam    = ___,
      qlabel  = ___,
      raw_var = ___,
      qorig   = "CRF",
      qeval   = "INVESTIGATOR"
    )
  )
)

suppae_three
```

```{r supp-oak-three-hint-1}
# The third list() entry:
# qnam    = "AESMTG"
# qlabel  = "Safety Review Meeting Flag"
# raw_var = "AE_SAFETY_MTG"
```

```{r supp-oak-three-solution}
ae_raw_with_ids <- ae_raw |>
  generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw")

suppae_three <- generate_sdtm_supp(
  parent_dataset  = ae,
  raw_dat         = ae_raw_with_ids,
  id_var          = "AESEQ",
  qnam_definitions = list(
    list(
      qnam    = "AESPID",
      qlabel  = "Sponsor-specific AE ID",
      raw_var = "SPONSOR_AE_ID",
      qorig   = "CRF",
      qeval   = "INVESTIGATOR"
    ),
    list(
      qnam    = "AESCAT",
      qlabel  = "Investigator AE Category",
      raw_var = "INV_AE_CAT",
      qorig   = "CRF",
      qeval   = "INVESTIGATOR"
    ),
    list(
      qnam    = "AESMTG",
      qlabel  = "Safety Review Meeting Flag",
      raw_var = "AE_SAFETY_MTG",
      qorig   = "CRF",
      qeval   = "INVESTIGATOR"
    )
  )
)

suppae_three
```

### Exercise 6b — compare with real pharmaversesdtm::suppae

`suppae_real` is the actual SUPPAE from the CDISCPILOT01 Alzheimer's study. Explore its structure and compare it with our tutorial SUPPAE — both must follow the same fixed schema.

```{r suppae-real-compare, exercise=TRUE, exercise.lines=20}
# Column structure of the real SUPPAE — should match SDTMIG column order
names(suppae_real)

# How many distinct QNAMs are used in the real study?
suppae_real |>
  distinct(QNAM, QLABEL) |>
  arrange(QNAM)

# How are QORIG values distributed? (CRF vs DERIVED vs ASSIGNED)
suppae_real |>
  count(QORIG, sort = TRUE)

# Does the real SUPPAE use subject-level rows (blank IDVAR)?
suppae_real |>
  filter(IDVAR == "" | is.na(IDVAR)) |>
  nrow()
```

```{r suppae-real-compare-hint-1}
# The real SUPPAE columns must be exactly:
# STUDYID, RDOMAIN, USUBJID, IDVAR, IDVARVAL, QNAM, QLABEL, QVAL, QORIG, QEVAL
# Any deviation from this schema is a Pinnacle 21 compliance finding.
# QORIG = "CRF" for values collected on the form; "DERIVED" for computed values.
# The real study may use subject-level rows (IDVAR = "") in SUPPDM, not usually in SUPPAE.
```

```{r quiz-suppae-real, echo=FALSE}
quiz(
  caption = "Real SUPPAE structure check",
  question(
    "In pharmaversesdtm::suppae, IDVAR is 'AESEQ' for every row. Why is AESEQ used rather than AETERM?",
    answer("AETERM may contain spaces or special characters that are invalid in IDVAR"),
    answer("AESEQ is a unique numeric row identifier within a subject — AETERM is not unique and could repeat", correct = TRUE),
    answer("AESEQ is shorter than AETERM so it satisfies the 8-character IDVAR limit"),
    answer("SDTMIG mandates AESEQ as IDVAR for all SUPPAE datasets"),
    allow_retry = TRUE
  )
)
```

---

## 7 — SUPP-- validation rules

Before submitting SUPP--, these rules must hold:

```{r quiz-supp-validation, echo=FALSE}
quiz(
  question(
    "QNAM must be:",
    answer("Exactly the same as the raw column name"),
    answer("Maximum 8 characters, uppercase", correct = TRUE),
    answer("The same as the SDTM variable it's supplementing"),
    random_answer_order = TRUE
  ),
  question(
    "QVAL is always:",
    answer("Numeric when the original value was numeric"),
    answer("Character — there is no QVALN in SUPP--", correct = TRUE),
    answer("In CDISC controlled terminology"),
    random_answer_order = TRUE
  ),
  question(
    "An 'orphan' SUPP-- row is one where:",
    answer("IDVARVAL references a row that doesn't exist in the parent domain", correct = TRUE),
    answer("QVAL is missing"),
    answer("The RDOMAIN doesn't match the dataset name"),
    random_answer_order = TRUE
  )
)
```

### Exercise 7a — validate QNAM naming rules

QNAM must satisfy three rules: at most 8 characters, all uppercase, and must be a valid SAS variable name (starts with a letter or underscore, contains only letters, digits, and underscores). Write a check that flags any violating QNAMs in `suppae_3`.

```{r qnam-validate, exercise=TRUE, exercise.lines=18}
# suppae_3 has three QNAMs: SPONSOR_AE_ID, INV_AE_CAT, AE_SAFETY_MTG
# (These raw column names were used directly as QNAMs — a common mistake!)

qnam_check <- suppae_3 |>
  distinct(QNAM) |>
  mutate(
    too_long     = nchar(QNAM) > 8,
    not_upper    = QNAM != toupper(QNAM),
    invalid_char = !grepl("^[A-Z_][A-Z0-9_]*$", QNAM),
    valid        = !too_long & !not_upper & !invalid_char
  )

qnam_check

# How many QNAMs violate at least one rule?
qnam_check |> filter(!valid) |> nrow()
```

```{r qnam-validate-hint-1}
# SPONSOR_AE_ID: 13 chars → too_long = TRUE; contains underscore → fine for SAS name
# INV_AE_CAT:   10 chars → too_long = TRUE
# AE_SAFETY_MTG: 13 chars → too_long = TRUE
# All three are uppercase and start with a letter, so only the length rule is violated.
# The fixed QNAMs used in generate_sdtm_supp (AESPID, AESCAT, AESMTG) all pass.
```

```{r qnam-validate-solution}
qnam_check <- suppae_3 |>
  distinct(QNAM) |>
  mutate(
    too_long     = nchar(QNAM) > 8,
    not_upper    = QNAM != toupper(QNAM),
    invalid_char = !grepl("^[A-Z_][A-Z0-9_]*$", QNAM),
    valid        = !too_long & !not_upper & !invalid_char
  )
qnam_check
qnam_check |> filter(!valid) |> nrow()
```

> **Key insight**: raw column names are almost never valid QNAMs. `SPONSOR_AE_ID` is 13 characters and contains underscores. Always define a short, meaningful QNAM (≤8 chars) separately from the raw column name.

### Exercise 7b — validate SUPP-- — check for orphan rows

An orphan SUPP-- row has an `IDVARVAL` that doesn't match any `AESEQ` in the parent AE domain.

Write code that finds orphan rows in `suppae_manual`:


```{r supp-orphan-check, exercise=TRUE, exercise.lines=12}
# suppae_manual was built earlier (2 qualifiers × 5 AE records = 10 rows)
# ae has AESEQ per USUBJID

# Find rows in suppae_manual where the IDVARVAL doesn't match any AESEQ
orphans <- suppae_manual |>
  anti_join(
    ae |> mutate(IDVARVAL = as.character(___)) |>
          select(USUBJID, IDVARVAL),
    by = c("USUBJID", "IDVARVAL")
  )

cat("Orphan rows:", nrow(orphans), "\n")
orphans
```

```{r supp-orphan-check-hint-1}
# mutate(IDVARVAL = as.character(AESEQ))
```

```{r supp-orphan-check-solution}
orphans <- suppae_manual |>
  anti_join(
    ae |> mutate(IDVARVAL = as.character(AESEQ)) |>
          select(USUBJID, IDVARVAL),
    by = c("USUBJID", "IDVARVAL")
  )

cat("Orphan rows:", nrow(orphans), "\n")
orphans
```

---

## 8 — RELREC: linking records across domains

### What RELREC is for

The Related Records dataset links specific rows in one domain to specific rows in another. Common use cases:

- An AE caused by a specific dose (AE → EX link)
- A concomitant medication given to treat a specific AE (CM → AE link)
- A lab test performed during a procedure (LB → PR link)

### The structure

RELREC has the same IDVAR/IDVARVAL concept as SUPP--, but instead of one row per qualifier, you get **two rows per relationship** — one for each domain side — sharing a `RELID`:

```{r relrec-shape-demo}
# An example relationship: AE001 ↔ EX001 for subject 01-701-1015
tibble(
  STUDYID  = c("TEST_STUDY", "TEST_STUDY"),
  USUBJID  = c("TEST_STUDY-701-1015", "TEST_STUDY-701-1015"),
  RDOMAIN  = c("AE", "EX"),
  IDVAR    = c("AESEQ", "EXSEQ"),
  IDVARVAL = c("1", "1"),
  RELID    = c("REL01", "REL01"),
  RELTYPE  = c("ONE", "ONE")
)
```

The two rows share `RELID = "REL01"`. `RELTYPE` indicates cardinality: `ONE` means one-to-one at this end of the relationship.

---

## 9 — Building RELREC with dplyr

`sdtm.oak` doesn't yet provide a RELREC helper — build it manually. The pattern:

1. Join the AE sponsor ID column to the EX sponsor ID column via the link table
2. Build the AE side of each pair
3. Build the EX side of each pair
4. `bind_rows()` to stack them

```{r relrec-build-demo}
# Step 1: join AE and EX via the sponsor ID link table
# ae_ex_link connects AE_SPONSOR_ID to EX_SPONSOR_ID per subject

ae_linked <- ae |>
  select(USUBJID, AESEQ, patient_number) |>
  inner_join(
    ae_raw |>
      generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw") |>
      select(patient_number, SPONSOR_AE_ID),
    by = "patient_number"
  )

ex_linked <- ex |>
  select(USUBJID, EXSEQ, patient_number) |>
  inner_join(
    ex_raw |>
      generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ex_raw") |>
      select(patient_number, EXSPID),
    by = "patient_number"
  )

# Use the link table to connect them
pairs <- ae_ex_link |>
  inner_join(
    ae_linked |> rename(AE_SPONSOR_ID = SPONSOR_AE_ID),
    by = "AE_SPONSOR_ID"
  ) |>
  inner_join(
    ex_linked |> rename(EX_SPONSOR_ID = EXSPID) |>
      select(USUBJID, EXSEQ, EX_SPONSOR_ID),
    by = c("USUBJID", "EX_SPONSOR_ID")
  ) |>
  mutate(RELID = paste0("REL", str_pad(row_number(), 2, pad = "0")))

pairs
```

```{r relrec-sides}
# Step 2: build AE side
relrec_ae <- pairs |>
  transmute(
    STUDYID  = "TEST_STUDY",
    USUBJID,
    RDOMAIN  = "AE",
    IDVAR    = "AESEQ",
    IDVARVAL = as.character(AESEQ),
    RELID,
    RELTYPE  = "ONE"
  )

# Step 3: build EX side
relrec_ex <- pairs |>
  transmute(
    STUDYID  = "TEST_STUDY",
    USUBJID,
    RDOMAIN  = "EX",
    IDVAR    = "EXSEQ",
    IDVARVAL = as.character(EXSEQ),
    RELID,
    RELTYPE  = "ONE"
  )

# Step 4: stack
relrec <- bind_rows(relrec_ae, relrec_ex) |>
  arrange(RELID, RDOMAIN)

relrec
```

### Exercise: Add a CM → AE relationship to RELREC

A new CM (concomitant medication) domain has these records:

```{r relrec-cm-setup}
cm <- tibble(
  STUDYID = "TEST_STUDY",
  DOMAIN  = "CM",
  USUBJID = c("TEST_STUDY-703-1042", "TEST_STUDY-703-1042"),
  CMSEQ   = c(1L, 2L),
  CMTRT   = c("ACETAMINOPHEN", "ONDANSETRON"),
  CMSPID  = c("CM001", "CM002"),  # links to AE004 and AE005
  patient_number = c(1042L, 1042L)
)

cm_ae_link <- tibble(
  CM_SPONSOR_ID = c("CM001", "CM002"),
  AE_SPONSOR_ID = c("AE004", "AE004")  # both CMs treat AE004
)
```

Build a RELREC for these CM → AE links. Each relationship produces two rows (one CM row, one AE row) sharing a RELID.

```{r relrec-cm-exercise, exercise=TRUE, exercise.lines=45}
# Step 1: join CM to link table, then to AE
cm_linked <- cm |>
  select(USUBJID, CMSEQ, CMSPID, patient_number) |>
  rename(CM_SPONSOR_ID = CMSPID)

ae_linked_ids <- ae |>
  select(USUBJID, AESEQ, patient_number) |>
  inner_join(
    ae_raw |>
      generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw") |>
      select(patient_number, SPONSOR_AE_ID) |>
      rename(AE_SPONSOR_ID = SPONSOR_AE_ID),
    by = "patient_number"
  )

pairs_cm_ae <- cm_ae_link |>
  inner_join(cm_linked, by = "CM_SPONSOR_ID") |>
  inner_join(
    ae_linked_ids |> select(USUBJID, AESEQ, AE_SPONSOR_ID),
    by = c("USUBJID", "AE_SPONSOR_ID")
  ) |>
  mutate(RELID = paste0("REL", str_pad(row_number(), 2, pad = "0")))

# Step 2: CM side
relrec_cm <- pairs_cm_ae |>
  transmute(
    STUDYID  = "TEST_STUDY",
    USUBJID,
    RDOMAIN  = ___,
    IDVAR    = ___,
    IDVARVAL = as.character(___),
    RELID,
    RELTYPE  = "ONE"
  )

# Step 3: AE side
relrec_ae2 <- pairs_cm_ae |>
  transmute(
    STUDYID  = "TEST_STUDY",
    USUBJID,
    RDOMAIN  = ___,
    IDVAR    = ___,
    IDVARVAL = as.character(___),
    RELID,
    RELTYPE  = "ONE"
  )

# Step 4: stack
relrec_cm_ae <- bind_rows(relrec_cm, relrec_ae2) |>
  arrange(RELID, RDOMAIN)

relrec_cm_ae
```

```{r relrec-cm-exercise-hint-1}
# CM side:  RDOMAIN = "CM", IDVAR = "CMSEQ", IDVARVAL = as.character(CMSEQ)
# AE side:  RDOMAIN = "AE", IDVAR = "AESEQ", IDVARVAL = as.character(AESEQ)
```

```{r relrec-cm-exercise-solution}
cm_linked <- cm |>
  select(USUBJID, CMSEQ, CMSPID, patient_number) |>
  rename(CM_SPONSOR_ID = CMSPID)

ae_linked_ids <- ae |>
  select(USUBJID, AESEQ, patient_number) |>
  inner_join(
    ae_raw |>
      generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw") |>
      select(patient_number, SPONSOR_AE_ID) |>
      rename(AE_SPONSOR_ID = SPONSOR_AE_ID),
    by = "patient_number"
  )

pairs_cm_ae <- cm_ae_link |>
  inner_join(cm_linked, by = "CM_SPONSOR_ID") |>
  inner_join(
    ae_linked_ids |> select(USUBJID, AESEQ, AE_SPONSOR_ID),
    by = c("USUBJID", "AE_SPONSOR_ID")
  ) |>
  mutate(RELID = paste0("REL", str_pad(row_number(), 2, pad = "0")))

relrec_cm <- pairs_cm_ae |>
  transmute(
    STUDYID  = "TEST_STUDY",
    USUBJID,
    RDOMAIN  = "CM",
    IDVAR    = "CMSEQ",
    IDVARVAL = as.character(CMSEQ),
    RELID,
    RELTYPE  = "ONE"
  )

relrec_ae2 <- pairs_cm_ae |>
  transmute(
    STUDYID  = "TEST_STUDY",
    USUBJID,
    RDOMAIN  = "AE",
    IDVAR    = "AESEQ",
    IDVARVAL = as.character(AESEQ),
    RELID,
    RELTYPE  = "ONE"
  )

relrec_cm_ae <- bind_rows(relrec_cm, relrec_ae2) |>
  arrange(RELID, RDOMAIN)

relrec_cm_ae
```

### Exercise 9b — one-to-many RELREC (RELTYPE = "MANY")

In the AE-EX example above, each relationship was ONE-to-ONE: one AE matched one EX record. Sometimes a single exposure record is responsible for multiple AEs — a one-to-many relationship. When one side has multiple records sharing a RELID, `RELTYPE = "MANY"` for that side.

Build a RELREC showing that EX record `EX001` (PATNUM 1015, first dose period) is linked to **both** AE001 (HEADACHE) and AE002 (NAUSEA). EX001 appears once with `RELTYPE = "ONE"`. Each AE row gets its own pair entry with `RELTYPE = "MANY"` on the AE side.

```{r relrec-many, exercise=TRUE, exercise.lines=38}
# EX001 caused two AEs for patient 1015
ae_ex_one_many <- tibble(
  AE_SPONSOR_ID = c("AE001", "AE002"),   # both caused by EX001
  EX_SPONSOR_ID = c("EX001", "EX001")
)

# Join to get AESEQ and EXSEQ
ae_side_data <- ae |>
  select(USUBJID, AESEQ, patient_number) |>
  left_join(
    ae_raw |>
      generate_oak_id_vars("PATNUM", "ae_raw") |>
      select(patient_number, SPONSOR_AE_ID),
    by = "patient_number"
  )

ex_side_data <- ex |>
  select(USUBJID, EXSEQ, patient_number) |>
  left_join(
    ex_raw |>
      generate_oak_id_vars("PATNUM", "ex_raw") |>
      select(patient_number, EXSPID),
    by = "patient_number"
  )

pairs_many <- ae_ex_one_many |>
  inner_join(ae_side_data |> rename(AE_SPONSOR_ID = SPONSOR_AE_ID), by = "AE_SPONSOR_ID") |>
  inner_join(ex_side_data |> rename(EX_SPONSOR_ID = EXSPID) |>
               select(USUBJID, EXSEQ, EX_SPONSOR_ID),
             by = c("USUBJID", "EX_SPONSOR_ID")) |>
  mutate(RELID = "REL01")   # same RELID for all rows in this relationship

# Build AE side — RELTYPE = "MANY" because multiple AEs share this RELID
relrec_ae_many <- pairs_many |>
  transmute(STUDYID = "TEST_STUDY", USUBJID, RDOMAIN = "AE",
            IDVAR = "AESEQ", IDVARVAL = as.character(AESEQ),
            RELID, RELTYPE = "___")

# Build EX side — one unique EX record, RELTYPE = "ONE"
relrec_ex_one <- pairs_many |>
  distinct(USUBJID, EXSEQ, RELID) |>
  transmute(STUDYID = "TEST_STUDY", USUBJID, RDOMAIN = "EX",
            IDVAR = "EXSEQ", IDVARVAL = as.character(EXSEQ),
            RELID, RELTYPE = "___")

bind_rows(relrec_ae_many, relrec_ex_one) |>
  arrange(RELID, RDOMAIN)
```

```{r relrec-many-hint-1}
# AE side: RELTYPE = "MANY"  (multiple AE rows share RELID REL01)
# EX side: RELTYPE = "ONE"   (only one EX row in this relationship)
# Result should have 3 rows: AE001 (AESEQ=1), AE002 (AESEQ=2), EX001 (EXSEQ=1)
# all with RELID = "REL01"
```

```{r relrec-many-solution}
ae_ex_one_many <- tibble(
  AE_SPONSOR_ID = c("AE001", "AE002"),
  EX_SPONSOR_ID = c("EX001", "EX001")
)

ae_side_data <- ae |>
  select(USUBJID, AESEQ, patient_number) |>
  left_join(ae_raw |> generate_oak_id_vars("PATNUM", "ae_raw") |>
              select(patient_number, SPONSOR_AE_ID), by = "patient_number")

ex_side_data <- ex |>
  select(USUBJID, EXSEQ, patient_number) |>
  left_join(ex_raw |> generate_oak_id_vars("PATNUM", "ex_raw") |>
              select(patient_number, EXSPID), by = "patient_number")

pairs_many <- ae_ex_one_many |>
  inner_join(ae_side_data |> rename(AE_SPONSOR_ID = SPONSOR_AE_ID), by = "AE_SPONSOR_ID") |>
  inner_join(ex_side_data |> rename(EX_SPONSOR_ID = EXSPID) |>
               select(USUBJID, EXSEQ, EX_SPONSOR_ID),
             by = c("USUBJID", "EX_SPONSOR_ID")) |>
  mutate(RELID = "REL01")

relrec_ae_many <- pairs_many |>
  transmute(STUDYID = "TEST_STUDY", USUBJID, RDOMAIN = "AE",
            IDVAR = "AESEQ", IDVARVAL = as.character(AESEQ),
            RELID, RELTYPE = "MANY")

relrec_ex_one <- pairs_many |>
  distinct(USUBJID, EXSEQ, RELID) |>
  transmute(STUDYID = "TEST_STUDY", USUBJID, RDOMAIN = "EX",
            IDVAR = "EXSEQ", IDVARVAL = as.character(EXSEQ),
            RELID, RELTYPE = "ONE")

bind_rows(relrec_ae_many, relrec_ex_one) |>
  arrange(RELID, RDOMAIN)
```

```{r quiz-relrec-many, echo=FALSE}
quiz(
  caption = "One-to-many RELREC",
  question(
    "In a one-to-many RELREC where one EX record caused three AEs, how many total rows appear in RELREC for this single relationship?",
    answer("2 — one for each domain"),
    answer("3 — one per AE"),
    answer("4 — three AE rows + one EX row", correct = TRUE),
    answer("6 — each pairing gets two rows"),
    allow_retry = TRUE
  )
)
```

---

## 10 — RELREC validation

```{r quiz-relrec, echo=FALSE}
quiz(
  question(
    "A RELREC dataset has a RELID 'REL05' that appears only once. This is:",
    answer("Valid — one-sided relationships are allowed"),
    answer("Invalid — every RELID must appear exactly twice (one per domain side)", correct = TRUE),
    answer("Only valid if RELTYPE = 'MANY'"),
    random_answer_order = TRUE
  ),
  question(
    "RELTYPE = 'MANY' on a RELREC row means:",
    answer("The subject has many records in this domain (cardinality flag)", correct = TRUE),
    answer("The relationship spans many studies"),
    answer("You can have unlimited RELIDs for this record"),
    random_answer_order = TRUE
  )
)
```

### Exercise: Validate RELREC — check that every RELID appears exactly twice

```{r relrec-validate, exercise=TRUE, exercise.lines=12}
# Check relrec (the AE-EX relrec built in the demo) for orphan RELIDs
relid_counts <- relrec |>
  count(___)

cat("RELID counts:\n")
print(relid_counts)

# Flag any RELID that doesn't appear exactly twice
bad_relids <- relid_counts |>
  filter(___ != 2)

cat("\nInvalid RELIDs:", nrow(bad_relids), "\n")
```

```{r relrec-validate-hint-1}
# count(RELID)
# filter(n != 2)
```

```{r relrec-validate-solution}
relid_counts <- relrec |>
  count(RELID)

cat("RELID counts:\n")
print(relid_counts)

bad_relids <- relid_counts |>
  filter(n != 2)

cat("\nInvalid RELIDs:", nrow(bad_relids), "\n")
```

---

## 11 — The CO (Comments) domain

CO (Comments) stores free-text comments, optionally linked to a specific domain record. It looks like SUPP-- but is a proper parent domain:

```{r co-domain-demo}
# Build a minimal CO domain manually (no sdtm.oak helper exists yet)
# One comment linked to AE #2 of subject 1015, one general subject comment
co <- tibble(
  STUDYID = "TEST_STUDY",
  DOMAIN  = "CO",
  USUBJID = c("TEST_STUDY-701-1015", "TEST_STUDY-701-1015"),
  COSEQ   = c(1L, 2L),
  IDVAR   = c("AESEQ", ""),
  IDVARVAL = c("2", ""),
  RDOMAIN = c("AE", ""),
  COREF   = c("Investigator note", ""),
  COMVAL  = c(
    "Patient reported symptom onset after first dose of study drug.",
    "Subject requested unscheduled phone contact on 2020-09-05."
  )
)

co
```

The key difference from SUPP--: CO uses `COSEQ` (its own sequence number) and `COMVAL` (the comment text). It's a first-class domain, not an appendage to another.

---

## 12 — Trial design domains and SV/SE

A brief orientation to domains `sdtm.oak` doesn't yet support:

```{r trial-design-demo}
# Trial design domains are typically hand-coded from protocol spec
# Example: a minimal TS (Trial Summary) dataset
ts <- tibble(
  STUDYID  = "TEST_STUDY",
  DOMAIN   = "TS",
  TSSEQ    = 1:5,
  TSPARMCD = c("TITLE", "PHASE", "TTYPE", "PCLAS", "SPONSOR"),
  TSPARM   = c("Trial Title", "Trial Phase", "Trial Type",
               "Pharmacological Class", "Sponsor Name"),
  TSVAL    = c("A Phase 2 Study of Study Drug in Patients with X",
               "PHASE2", "INTERVENTIONAL", "KINASE INHIBITOR",
               "Sponsor Corp"),
  TSVALCD  = c(NA, "2", NA, NA, NA)
)

ts
```

Trial design domains (TA, TE, TI, TS, TV, TX) describe the *study* rather than subjects. They often have very few rows and are built from the protocol document or spec spreadsheet, not from EDC raw data.

### Exercise: Build a minimal TV (Trial Visits) dataset

The protocol specifies 4 planned visits. Build a `tv` dataset with the required TVSEQ, VISITNUM, VISIT, and VISITDY (planned day).

```{r tv-build, exercise=TRUE, exercise.lines=22}
# Build TV from the protocol visit schedule
visit_schedule <- tibble(
  VISITNUM = c(1, 2, 3, 4),
  VISIT    = c("SCREENING", "BASELINE", "WEEK 4", "WEEK 8"),
  VISITDY  = c(-7L, 1L, 29L, 57L)
)

tv <- visit_schedule |>
  mutate(
    STUDYID = "TEST_STUDY",
    DOMAIN  = "TV",
    TVSEQ   = row_number(),
    ARM     = "TREATMENT"
  ) |>
  select(___)  # put columns in SDTM order: STUDYID, DOMAIN, TVSEQ, ARM, VISITNUM, VISIT, VISITDY

tv
```

```{r tv-build-hint-1}
# select(STUDYID, DOMAIN, TVSEQ, ARM, VISITNUM, VISIT, VISITDY)
```

```{r tv-build-solution}
visit_schedule <- tibble(
  VISITNUM = c(1, 2, 3, 4),
  VISIT    = c("SCREENING", "BASELINE", "WEEK 4", "WEEK 8"),
  VISITDY  = c(-7L, 1L, 29L, 57L)
)

tv <- visit_schedule |>
  mutate(
    STUDYID = "TEST_STUDY",
    DOMAIN  = "TV",
    TVSEQ   = row_number(),
    ARM     = "TREATMENT"
  ) |>
  select(STUDYID, DOMAIN, TVSEQ, ARM, VISITNUM, VISIT, VISITDY)

tv
```

---

## 13 — Pragmatic decision guide

```{r quiz-decision-guide, echo=FALSE}
quiz(
  question(
    "You're building LB for a PK study. A few lab records have a comment about sample hemolysis. Where does this go?",
    answer("Add a new column LBHEMO to the LB domain"),
    answer("SUPPAE"),
    answer("SUPPLB with QNAM = 'LBHEMO' and IDVAR = 'LBSEQ'", correct = TRUE),
    answer("CO domain"),
    random_answer_order = TRUE
  ),
  question(
    "The study's SAP describes 'for each AE, identify the most recent prior dose'. This analysis needs:",
    answer("SUPP-- only"),
    answer("RELREC linking AE to EX rows", correct = TRUE),
    answer("A new SDTM domain"),
    answer("CO domain"),
    random_answer_order = TRUE
  ),
  question(
    "Which of these is NOT currently supported directly by sdtm.oak (as of v0.1)?",
    answer("AE Events domain"),
    answer("VS Findings domain"),
    answer("RELREC", correct = TRUE),
    answer("DM domain"),
    random_answer_order = TRUE
  )
)
```

---

## 14 — Capstone exercise: Full SUPPAE + RELREC

Put it all together. Given the `ae`, `ae_raw`, `ex`, `ex_raw`, and `ae_ex_link` objects:

1. Build `suppae` with two qualifiers: `AESPID` and `AESCAT`
2. Build `relrec` linking the two AE-EX pairs (using `ae_ex_link`)
3. Check that `suppae` has no orphan rows and `relrec` has no singleton RELIDs

```{r capstone-suppae-relrec, exercise=TRUE, exercise.lines=65}
# ── Part 1: Build SUPPAE ──────────────────────────────────────────────────────
suppae_final <- generate_sdtm_supp(
  parent_dataset = ae,
  raw_dat        = ae_raw |> generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw"),
  id_var         = "AESEQ",
  qnam_definitions = list(
    list(qnam = "AESPID", qlabel = "Sponsor-specific AE ID",
         raw_var = "SPONSOR_AE_ID", qorig = "CRF", qeval = "INVESTIGATOR"),
    list(qnam = "AESCAT", qlabel = "Investigator AE Category",
         raw_var = "INV_AE_CAT",    qorig = "CRF", qeval = "INVESTIGATOR")
  )
)

cat("SUPPAE rows:", nrow(suppae_final), "\n")

# ── Part 2: Build RELREC (AE-EX) ─────────────────────────────────────────────
ae_linked2 <- ae |>
  select(USUBJID, AESEQ, patient_number) |>
  inner_join(
    ae_raw |>
      generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ae_raw") |>
      select(patient_number, SPONSOR_AE_ID),
    by = "patient_number"
  )

ex_linked2 <- ex |>
  select(USUBJID, EXSEQ, patient_number) |>
  inner_join(
    ex_raw |>
      generate_oak_id_vars(pat_var = "PATNUM", raw_src = "ex_raw") |>
      select(patient_number, EXSPID),
    by = "patient_number"
  )

pairs2 <- ae_ex_link |>
  inner_join(ae_linked2 |> rename(AE_SPONSOR_ID = SPONSOR_AE_ID), by = "AE_SPONSOR_ID") |>
  inner_join(ex_linked2 |> rename(EX_SPONSOR_ID = EXSPID) |> select(USUBJID, EXSEQ, EX_SPONSOR_ID),
             by = c("USUBJID", "EX_SPONSOR_ID")) |>
  mutate(RELID = paste0("REL", str_pad(row_number(), 2, pad = "0")))

relrec_final <- bind_rows(
  pairs2 |> transmute(STUDYID = "TEST_STUDY", USUBJID, RDOMAIN = "AE",
                       IDVAR = "AESEQ", IDVARVAL = as.character(AESEQ), RELID, RELTYPE = "ONE"),
  pairs2 |> transmute(STUDYID = "TEST_STUDY", USUBJID, RDOMAIN = "EX",
                       IDVAR = "EXSEQ", IDVARVAL = as.character(EXSEQ), RELID, RELTYPE = "ONE")
) |> arrange(RELID, RDOMAIN)

cat("RELREC rows:", nrow(relrec_final), "\n")

# ── Part 3: Validations ───────────────────────────────────────────────────────
# Check 1: no orphan SUPP-- rows
orphans_final <- suppae_final |>
  anti_join(ae |> mutate(IDVARVAL = as.character(AESEQ)) |> select(USUBJID, IDVARVAL),
            by = c("USUBJID", "IDVARVAL"))
cat("Orphan SUPPAE rows:", nrow(orphans_final), "\n")

# Check 2: no singleton RELIDs
singleton_relids <- relrec_final |>
  count(RELID) |>
  filter(n != 2)
cat("Singleton RELIDs:", nrow(singleton_relids), "\n")

list(suppae = suppae_final, relrec = relrec_final)
```

```{r capstone-suppae-relrec-hint-1}
# This exercise is mostly pre-filled — just run it and verify:
# SUPPAE rows:  10 (5 AE records × 2 qualifiers)
# RELREC rows:   4 (2 relationships × 2 domain sides each)
# Orphan SUPPAE rows: 0
# Singleton RELIDs:   0
```

---

## 15 — Key takeaways

```{r quiz-final, echo=FALSE}
quiz(
  question(
    "The SUPP-- column order required by SDTMIG is:",
    answer("STUDYID, RDOMAIN, USUBJID, IDVAR, IDVARVAL, QNAM, QLABEL, QVAL, QORIG, QEVAL",
           correct = TRUE),
    answer("STUDYID, USUBJID, RDOMAIN, QNAM, QLABEL, QVAL, IDVAR, IDVARVAL, QORIG, QEVAL"),
    answer("Any order — SDTM datasets don't have required column order"),
    random_answer_order = TRUE
  ),
  question(
    "Why does QVAL not have a numeric companion (QVALN)?",
    answer("QVAL is always character by SDTMIG specification for SUPP-- datasets", correct = TRUE),
    answer("Numeric values aren't allowed in SUPP--"),
    answer("A numeric version exists — it's called QVALNC"),
    random_answer_order = TRUE
  ),
  question(
    "When would RELREC NOT be needed for a study?",
    answer("When the SAP has no cross-domain analytical questions (e.g., a simple phase-I PK study)", correct = TRUE),
    answer("When using sdtm.oak, which handles linking automatically"),
    answer("When SUPP-- datasets are already present"),
    random_answer_order = TRUE
  ),
  question(
    "The CO domain differs from SUPP-- primarily because:",
    answer("CO is a proper parent domain with its own COSEQ; SUPP-- is an appendage to another domain", correct = TRUE),
    answer("CO only holds numeric comments; SUPP-- holds text"),
    answer("CO doesn't use IDVAR/IDVARVAL"),
    random_answer_order = TRUE
  ),
  question(
    "A practical recommendation for trial design domains (TS, TV, TA, etc.) in sdtm.oak today:",
    answer("Use sdtm.oak algorithm functions — they support all trial design domains"),
    answer("Build them manually with hand-coded dplyr or from a study spec spreadsheet", correct = TRUE),
    answer("Submit them as SUPP-- instead"),
    random_answer_order = TRUE
  ),
  question(
    "A QNAM is defined as 'AEINVGRD' (Investigator Grade) in the spec. During SUPP-- build you find three rows where the raw value is NA. The correct action is:",
    answer("Set QVAL = 'NA' (the string) for these rows"),
    answer("Include the rows with QVAL = '' (empty string)"),
    answer("Exclude these rows entirely — SUPP-- rows with missing QVAL are invalid and create orphan-like noise", correct = TRUE),
    answer("Hardcode QVAL = '0' as a safe placeholder"),
    random_answer_order = TRUE
  ),
  question(
    "QORIG = 'DERIVED' should be used when:",
    answer("The value comes directly from a CRF field"),
    answer("The value was assigned by the sponsor's data management team"),
    answer("The value was computed algorithmically (e.g., from other variables), not collected on the CRF", correct = TRUE),
    answer("The value was transcribed from a lab report"),
    random_answer_order = TRUE
  ),
  question(
    "An analyst needs to know 'which study drug dose period was ongoing when each AE started'. The correct SDTM structure to support this analysis is:",
    answer("Add EXSTDTC and EXENDTC columns to the AE domain"),
    answer("Add AESTDTC and AEENDTC columns to the EX domain"),
    answer("RELREC linking AE rows to EX rows so the analyst can join across domains by RELID", correct = TRUE),
    answer("A SUPP-- qualifier in SUPPAE pointing to the dose period"),
    random_answer_order = TRUE
  ),
  question(
    "You run the QNAM validation check and find 'AE_SAFETY_MTG' has 13 characters. The correct fix is:",
    answer("Truncate to 'AE_SAFETY' (9 chars) — close enough"),
    answer("Define a valid short QNAM such as 'AESMTG' (6 chars) and map the raw column to it in the qualifier definition", correct = TRUE),
    answer("Split the QNAM into two rows in SUPPAE"),
    answer("Move the variable to the CO domain instead"),
    random_answer_order = TRUE
  ),
  question(
    "In the pharmaversesdtm::suppae dataset for the CDISCPILOT01 study, IDVAR = 'AESEQ' for every row. What would it mean if you found a row with IDVAR = '' (empty string)?",
    answer("The row is invalid — SUPPAE always requires a non-blank IDVAR"),
    answer("The row is a subject-level qualifier that applies to the subject as a whole, not to a specific AE record", correct = TRUE),
    answer("The AESEQ was missing in the parent AE domain for that subject"),
    answer("The row is a header placeholder inserted by generate_sdtm_supp()"),
    random_answer_order = TRUE
  )
)
```

---

## What's next

Lesson 11 — **`{sdtmchecks}`** — closes out Module 2. After building all this SDTM, you need to validate it. `sdtmchecks` provides 100+ analysis-focused quality checks: duplicate records, impossible date sequences, subjects missing key domains, and more. It complements Pinnacle 21 (which checks CDISC conformance) by catching things that would break your *downstream analysis*.

After Module 2, we move to Module 3 (metadata-driven programming with `metacore` and `metatools`) and then the heart of pharmaverse: `admiral` for ADaM construction.
