---
title: "Lesson 11 — {sdtmchecks}: Analysis-Focused SDTM Quality Checks"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Run quality checks on SDTM datasets, interpret results, generate Excel reports,
  and write custom checks using the sdtmchecks package conventions.
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── Mock sdtmchecks infrastructure ───────────────────────────────────────────
# Each check follows the same contract:
#   Input : one or more SDTM domain data frames
#   Output: a data frame of OFFENDING ROWS (empty = check passed)

check_dm_usubjid_dup <- function(DM) {
  DM |>
    group_by(USUBJID) |>
    filter(n() > 1) |>
    ungroup() |>
    arrange(USUBJID)
}

check_ae_dup <- function(AE) {
  AE |>
    group_by(USUBJID, AETERM, AESTDTC) |>
    filter(n() > 1) |>
    ungroup() |>
    arrange(USUBJID, AETERM)
}

check_ex_dup <- function(EX) {
  EX |>
    group_by(USUBJID, EXTRT, EXSTDTC, EXENDTC) |>
    filter(n() > 1) |>
    ungroup()
}

check_ae_aestdtc_after_dm_death <- function(AE, DM) {
  dm_death <- DM |>
    filter(!is.na(DTHDTC)) |>
    select(USUBJID, DTHDTC)
  AE |>
    inner_join(dm_death, by = "USUBJID") |>
    filter(!is.na(AESTDTC) & ymd(AESTDTC) > ymd(DTHDTC)) |>
    select(-DTHDTC)
}

check_ae_aestdtc_before_exstdtc <- function(AE, EX) {
  first_dose <- EX |>
    group_by(USUBJID) |>
    summarise(FIRST_DOSE = min(ymd(EXSTDTC), na.rm = TRUE), .groups = "drop")
  AE |>
    inner_join(first_dose, by = "USUBJID") |>
    filter(!is.na(AESTDTC) & ymd(AESTDTC) < FIRST_DOSE) |>
    select(-FIRST_DOSE)
}

check_lb_dup <- function(LB) {
  LB |>
    group_by(USUBJID, LBTESTCD, VISIT, LBDTC) |>
    filter(n() > 1) |>
    ungroup() |>
    arrange(USUBJID, LBTESTCD)
}

check_lb_lbornrlo_lbornrhi <- function(LB) {
  LB |>
    filter(!is.na(LBORNRLO) & !is.na(LBORNRHI)) |>
    filter(as.numeric(LBORNRLO) > as.numeric(LBORNRHI))
}

check_vs_dup <- function(VS) {
  VS |>
    group_by(USUBJID, VSTESTCD, VISIT, VSDTC) |>
    filter(n() > 1) |>
    ungroup()
}

check_dm_no_ex <- function(DM, EX) {
  subjects_with_ex <- EX |> distinct(USUBJID)
  DM |>
    anti_join(subjects_with_ex, by = "USUBJID") |>
    filter(ARMCD != "SCRNFAIL")
}

# ── Mock check metadata ───────────────────────────────────────────────────────
sdtmchecksmeta <- tribble(
  ~check,                              ~category, ~priority, ~description,                                    ~datasets,
  "check_dm_usubjid_dup",              "DM",      "High",    "Duplicate USUBJID in DM",                      "DM",
  "check_ae_dup",                      "AE",      "High",    "Duplicate AE records (same USUBJID/term/date)", "AE",
  "check_ex_dup",                      "EX",      "High",    "Duplicate EX records",                          "EX",
  "check_ae_aestdtc_after_dm_death",   "AE",      "High",    "AE start date after subject death date",        "AE,DM",
  "check_ae_aestdtc_before_exstdtc",   "AE",      "High",    "AE start date before first exposure date",      "AE,EX",
  "check_lb_dup",                      "LB",      "High",    "Duplicate LB records per test/visit/date",      "LB",
  "check_lb_lbornrlo_lbornrhi",        "LB",      "Medium",  "Lab reference range low > high",                "LB",
  "check_vs_dup",                      "VS",      "Medium",  "Duplicate VS records",                          "VS",
  "check_dm_no_ex",                    "DM",      "Medium",  "Subjects in DM with no EX records",             "DM,EX"
)

# ── Mock run_all_checks ───────────────────────────────────────────────────────
run_all_checks <- function(metads, ...) {
  datasets <- list(...)
  results <- lapply(seq_len(nrow(metads)), function(i) {
    fn_name <- metads$check[i]
    fn <- tryCatch(get(fn_name), error = function(e) NULL)
    if (is.null(fn)) return(list(check = fn_name, status = "skipped", issues = NULL))

    needed <- strsplit(metads$datasets[i], ",")[[1]]
    args <- datasets[needed]
    if (any(sapply(args, is.null))) return(list(check = fn_name, status = "skipped", issues = NULL))

    issues <- tryCatch(do.call(fn, args), error = function(e) NULL)
    if (is.null(issues)) return(list(check = fn_name, status = "error", issues = NULL))

    status <- if (nrow(issues) == 0) "passed" else "failed"
    list(check = fn_name, priority = metads$priority[i],
         status = status, issues = issues, n_issues = nrow(issues))
  })
  names(results) <- metads$check
  results
}

# ── Mock report_to_xlsx ───────────────────────────────────────────────────────
report_to_xlsx <- function(res, outfile = "sdtm_qc_report.xlsx") {
  summary_df <- lapply(names(res), function(nm) {
    r <- res[[nm]]
    tibble(
      check    = nm,
      priority = r$priority %||% NA_character_,
      status   = r$status,
      n_issues = r$n_issues %||% NA_integer_
    )
  }) |> bind_rows()

  message(sprintf(
    "Report written to: %s\n  %d checks: %d passed, %d failed, %d skipped",
    outfile,
    nrow(summary_df),
    sum(summary_df$status == "passed", na.rm = TRUE),
    sum(summary_df$status == "failed", na.rm = TRUE),
    sum(summary_df$status == "skipped", na.rm = TRUE)
  ))
  invisible(summary_df)
}

# Null-coalescing helper
`%||%` <- function(a, b) if (!is.null(a)) a else b

# ── Sample clean SDTM datasets ────────────────────────────────────────────────
dm <- tibble(
  STUDYID = "TEST_STUDY",
  DOMAIN  = "DM",
  USUBJID = c("TS-701-1015", "TS-702-1023", "TS-703-1042", "TS-704-1101"),
  SUBJID  = c("1015", "1023", "1042", "1101"),
  RFSTDTC = c("2020-07-01", "2020-07-05", "2020-09-01", "2020-06-15"),
  DTHDTC  = c(NA, NA, NA, NA),
  ARMCD   = c("TRT", "TRT", "TRT", "TRT"),
  ARM     = "TREATMENT"
)

ae <- tibble(
  STUDYID = "TEST_STUDY",
  DOMAIN  = "AE",
  USUBJID = c("TS-701-1015", "TS-701-1015", "TS-702-1023", "TS-703-1042", "TS-703-1042"),
  AESEQ   = c(1L, 2L, 1L, 1L, 2L),
  AETERM  = c("HEADACHE", "NAUSEA", "FATIGUE", "DIZZINESS", "RASH"),
  AESTDTC = c("2020-07-15", "2020-08-01", "2020-07-20", "2020-09-10", "2020-09-12"),
  AEENDTC = c("2020-07-17", "2020-08-03", "2020-07-25", "2020-09-12", "2020-09-20"),
  AESEV   = c("MILD", "MODERATE", "MILD", "SEVERE", "MILD"),
  AESER   = c("N", "N", "N", "Y", "N")
)

ex <- tibble(
  STUDYID = "TEST_STUDY",
  DOMAIN  = "EX",
  USUBJID = c("TS-701-1015", "TS-701-1015", "TS-702-1023", "TS-703-1042"),
  EXSEQ   = c(1L, 2L, 1L, 1L),
  EXTRT   = "STUDY DRUG",
  EXDOSE  = c(100, 100, 50, 100),
  EXSTDTC = c("2020-07-01", "2020-08-01", "2020-07-05", "2020-09-01"),
  EXENDTC = c("2020-07-31", "2020-08-31", "2020-07-31", "2020-09-30")
)

lb <- tibble(
  STUDYID  = "TEST_STUDY",
  DOMAIN   = "LB",
  USUBJID  = c("TS-701-1015", "TS-701-1015", "TS-702-1023", "TS-703-1042"),
  LBSEQ    = c(1L, 2L, 1L, 1L),
  LBTESTCD = c("ALT", "ALT", "ALT", "ALT"),
  LBTEST   = "Alanine Aminotransferase",
  VISIT    = c("BASELINE", "WEEK 4", "BASELINE", "BASELINE"),
  LBDTC    = c("2020-07-01", "2020-07-29", "2020-07-05", "2020-09-01"),
  LBORRES  = c("28", "35", "22", "41"),
  LBORRESU = "U/L",
  LBORNRLO = c("7", "7", "7", "7"),
  LBORNRHI = c("56", "56", "56", "56")
)

vs <- tibble(
  STUDYID  = "TEST_STUDY",
  DOMAIN   = "VS",
  USUBJID  = c("TS-701-1015", "TS-701-1015", "TS-702-1023"),
  VSSEQ    = c(1L, 2L, 1L),
  VSTESTCD = c("SYSBP", "DIABP", "SYSBP"),
  VSDTC    = c("2020-07-01", "2020-07-01", "2020-07-05"),
  VISIT    = c("BASELINE", "BASELINE", "BASELINE"),
  VSORRES  = c("122", "78", "118")
)

# ── "Dirty" versions with intentional issues ──────────────────────────────────

# DM with a duplicate subject
dm_dirty <- bind_rows(
  dm,
  dm |> filter(USUBJID == "TS-701-1015")  # duplicate row
)

# AE with a duplicate record
ae_dirty <- bind_rows(
  ae,
  ae |> filter(USUBJID == "TS-703-1042" & AESEQ == 1)  # duplicate AE
)

# AE with an impossible date (AE after death)
dm_with_death <- dm |>
  mutate(DTHDTC = case_when(
    USUBJID == "TS-703-1042" ~ "2020-09-11",
    TRUE ~ NA_character_
  ))
# ae has TS-703-1042 AESTDTC = "2020-09-12" — after death on 09-11

# LB with inverted reference range
lb_dirty <- lb |>
  mutate(
    LBORNRLO = ifelse(USUBJID == "TS-702-1023", "60", LBORNRLO),  # 60 > 56
    LBORNRHI = ifelse(USUBJID == "TS-702-1023", "56", LBORNRHI)
  )

# AE with record before first dose (AE on 2020-06-25, first dose 2020-07-01)
ae_early <- bind_rows(
  ae,
  tibble(
    STUDYID = "TEST_STUDY", DOMAIN = "AE",
    USUBJID = "TS-701-1015", AESEQ = 99L,
    AETERM = "FATIGUE", AESTDTC = "2020-06-25",
    AEENDTC = "2020-06-28", AESEV = "MILD", AESER = "N"
  )
)

# ── Additional check: AE end before start ─────────────────────────────────────
check_ae_aeendtc_before_aestdtc <- function(AE) {
  AE |>
    filter(!is.na(AESTDTC) & !is.na(AEENDTC)) |>
    filter(ymd(AEENDTC) < ymd(AESTDTC))
}

# AE with end date before start date (data entry error)
ae_end_before_start <- bind_rows(
  ae,
  tibble(
    STUDYID = "TEST_STUDY", DOMAIN = "AE",
    USUBJID = "TS-702-1023", AESEQ   = 99L,
    AETERM  = "BACK PAIN",
    AESTDTC = "2020-08-15",
    AEENDTC = "2020-08-10",   # ends 5 days BEFORE it starts — impossible
    AESEV   = "MILD", AESER = "N"
  )
)

# ── Real pharmaversesdtm datasets ─────────────────────────────────────────────
library(pharmaversesdtm)
ae_sdtm_real <- pharmaversesdtm::ae
dm_sdtm_real <- pharmaversesdtm::dm
lb_sdtm_real <- pharmaversesdtm::lb
```

## Welcome

This is Lesson 11 — the final lesson of Module 2 — covering **`{sdtmchecks}`**.

You've built SDTM datasets with `sdtm.oak`. Now you need to validate them. `{sdtmchecks}` is a Roche-originated, pharmaverse-community-extended package with 100+ analysis-focused quality checks.

This lesson covers:

- How `{sdtmchecks}` differs from Pinnacle 21 (and why you need both)
- The check function contract: what goes in, what comes out
- Running single checks and interpreting results
- Running all checks with `run_all_checks()`
- Generating an Excel report with `report_to_xlsx()`
- Priority triage (High / Medium / Low)
- Writing your own custom check
- Wrapping checks in assertion style for production pipelines

All functions are mocked — they run without installing the package and faithfully implement the real contract.

---

## 1 — sdtmchecks vs Pinnacle 21

```{r quiz-p21-vs-sdtmchecks, echo=FALSE}
quiz(
  question(
    "Pinnacle 21 (P21) focuses primarily on:",
    answer("CDISC conformance — does the dataset match the SDTMIG schema?", correct = TRUE),
    answer("Analysis-impacting issues like duplicate records or impossible dates"),
    answer("Cross-domain consistency checks"),
    random_answer_order = TRUE
  ),
  question(
    "sdtmchecks focuses primarily on:",
    answer("Correct codelist values (CDISC CT)"),
    answer("Analysis-impacting issues that would corrupt downstream analysis", correct = TRUE),
    answer("Column ordering and length requirements"),
    random_answer_order = TRUE
  ),
  question(
    "A dataset can be P21-clean but still fail sdtmchecks because:",
    answer("sdtmchecks has stricter CDISC conformance rules than P21"),
    answer("A dataset can be structurally valid but contain logically impossible or analysis-corrupting data", correct = TRUE),
    answer("P21 only checks AE and DM domains"),
    random_answer_order = TRUE
  )
)
```

---

## 2 — The check function contract

Every check in `{sdtmchecks}` follows the same pattern:

- **Input**: one or more SDTM domain data frames (named for the domain: `AE`, `DM`, `LB`, etc.)
- **Output**: a data frame of **offending rows** — if empty, the check passed; if non-empty, those are the issues

```{r contract-demo}
# A passing check returns an empty data frame
result_clean <- check_ae_dup(ae)
cat("Clean AE — rows flagged:", nrow(result_clean), "\n")

# A failing check returns the offending rows with context
result_dirty <- check_ae_dup(ae_dirty)
cat("Dirty AE — rows flagged:", nrow(result_dirty), "\n")
result_dirty
```

The returned data frame contains the actual offending rows so you can investigate: you see USUBJID, AETERM, AESTDTC, and all the columns needed to track the issue down in the EDC.

---

## 3 — Single-domain checks

### DM: Duplicate subjects

```{r check-dm-dup}
# Check for duplicate USUBJIDs in DM
issues_dm_dup <- check_dm_usubjid_dup(dm_dirty)

cat("DM duplicate subjects found:", nrow(issues_dm_dup) / 2, "\n")
issues_dm_dup
```

### Exercise: Run the DM duplicate check on clean data

The clean `dm` dataset has no duplicates. Confirm this.

```{r check-dm-clean, exercise=TRUE, exercise.lines=6}
# Run check_dm_usubjid_dup on the clean dm dataset
result <- check_dm_usubjid_dup(___)

cat("Issues found:", nrow(result), "\n")
```

```{r check-dm-clean-hint-1}
# check_dm_usubjid_dup(dm)
```

```{r check-dm-clean-solution}
result <- check_dm_usubjid_dup(dm)
cat("Issues found:", nrow(result), "\n")
```

---

## 4 — Multi-domain checks

Some checks need more than one dataset. Pass them as separate arguments.

```{r check-multi-demo}
# AE start date after death date — needs both AE and DM
ae_after_death <- check_ae_aestdtc_after_dm_death(
  AE = ae,
  DM = dm_with_death
)

cat("AEs starting after death date:", nrow(ae_after_death), "\n")
ae_after_death
```

```{r check-before-dose-demo}
# AE before first dose — needs both AE and EX
ae_before_dose <- check_ae_aestdtc_before_exstdtc(
  AE = ae_early,
  EX = ex
)

cat("AEs starting before first dose:", nrow(ae_before_dose), "\n")
ae_before_dose
```

### Exercise: Check for lab duplicate records

`lb_dirty` doesn't have any duplicates (we'll create one below). First check the clean `lb` — it should pass — then check a version with a duplicate row.

```{r check-lb-dup, exercise=TRUE, exercise.lines=14}
# Check the clean lb dataset
result_clean <- check_lb_dup(___)
cat("Clean LB issues:", nrow(result_clean), "\n")

# Now add a duplicate row and check again
lb_with_dup <- bind_rows(
  lb,
  lb |> filter(USUBJID == "TS-701-1015" & LBTESTCD == "ALT" & VISIT == "BASELINE")
)

result_dirty <- check_lb_dup(___)
cat("Dirty LB issues:", nrow(result_dirty), "\n")
result_dirty
```

```{r check-lb-dup-hint-1}
# check_lb_dup(lb)
# check_lb_dup(lb_with_dup)
```

```{r check-lb-dup-solution}
result_clean <- check_lb_dup(lb)
cat("Clean LB issues:", nrow(result_clean), "\n")

lb_with_dup <- bind_rows(
  lb,
  lb |> filter(USUBJID == "TS-701-1015" & LBTESTCD == "ALT" & VISIT == "BASELINE")
)

result_dirty <- check_lb_dup(lb_with_dup)
cat("Dirty LB issues:", nrow(result_dirty), "\n")
result_dirty
```

### Exercise: Check for inverted lab reference ranges

`lb_dirty` has an inverted reference range for subject TS-702-1023 (LBORNRLO = 60, LBORNRHI = 56 — low > high).

```{r check-lb-refrange, exercise=TRUE, exercise.lines=8}
# Run check_lb_lbornrlo_lbornrhi on lb_dirty
result <- check_lb_lbornrlo_lbornrhi(___)

cat("Inverted reference ranges:", nrow(result), "\n")
result
```

```{r check-lb-refrange-hint-1}
# check_lb_lbornrlo_lbornrhi(lb_dirty)
```

```{r check-lb-refrange-solution}
result <- check_lb_lbornrlo_lbornrhi(lb_dirty)
cat("Inverted reference ranges:", nrow(result), "\n")
result
```

### Exercise: Check for AE end dates before start dates

A common data entry error is entering an AE resolution date earlier than the onset date. The check `check_ae_aeendtc_before_aestdtc()` flags these impossible dates. The dataset `ae_end_before_start` has one such record for subject TS-702-1023 with BACK PAIN (AESTDTC 2020-08-15, AEENDTC 2020-08-10).

```{r check-ae-end-before-start, exercise=TRUE, exercise.lines=14}
# Check the clean ae first — should return 0 rows
result_clean <- check_ae_aeendtc_before_aestdtc(___)
cat("Clean AE issues:", nrow(result_clean), "\n")

# Now check ae_end_before_start — should flag the impossible BACK PAIN record
result_dirty <- check_ae_aeendtc_before_aestdtc(___)
cat("Dirty AE issues:", nrow(result_dirty), "\n")
result_dirty |> select(USUBJID, AETERM, AESTDTC, AEENDTC, AESEV)
```

```{r check-ae-end-before-start-hint-1}
# check_ae_aeendtc_before_aestdtc(ae)
# check_ae_aeendtc_before_aestdtc(ae_end_before_start)
```

```{r check-ae-end-before-start-solution}
result_clean <- check_ae_aeendtc_before_aestdtc(ae)
cat("Clean AE issues:", nrow(result_clean), "\n")

result_dirty <- check_ae_aeendtc_before_aestdtc(ae_end_before_start)
cat("Dirty AE issues:", nrow(result_dirty), "\n")
result_dirty |> select(USUBJID, AETERM, AESTDTC, AEENDTC, AESEV)
```

---

## 5 — The check metadata: `sdtmchecksmeta`

The package ships a metadata data frame describing every check:

```{r meta-explore}
# Browse the metadata
sdtmchecksmeta |>
  select(check, category, priority, description) |>
  arrange(priority, category)
```

```{r meta-filter}
# Filter to just High-priority AE checks
sdtmchecksmeta |>
  filter(category == "AE", priority == "High")
```

### Exercise: Find all High-priority checks that need the DM domain

```{r meta-dm-high, exercise=TRUE, exercise.lines=8}
# Filter sdtmchecksmeta for High priority checks whose datasets column includes "DM"
high_dm_checks <- sdtmchecksmeta |>
  filter(priority == "High", str_detect(___, "DM"))

high_dm_checks
```

```{r meta-dm-high-hint-1}
# str_detect(datasets, "DM")
```

```{r meta-dm-high-solution}
high_dm_checks <- sdtmchecksmeta |>
  filter(priority == "High", str_detect(datasets, "DM"))

high_dm_checks
```

### Exercise: Customise a check's priority for your study

Your study has a known problem with inverted reference ranges from a specific site's EDC configuration. You want `check_lb_lbornrlo_lbornrhi` to be treated as High priority instead of Medium so it blocks the analysis pipeline. Modify a copy of `sdtmchecksmeta` and confirm the change.

```{r meta-customize-priority, exercise=TRUE, exercise.lines=16}
# Make a study-specific copy of the metadata
study_meta <- sdtmchecksmeta |>
  mutate(priority = case_when(
    check == ___ ~ "High",
    TRUE ~ priority
  ))

# Confirm the change
study_meta |>
  filter(check == "check_lb_lbornrlo_lbornrhi") |>
  select(check, priority, description)

# Count High-priority checks before vs. after
cat("Original High checks:", sum(sdtmchecksmeta$priority == "High"), "\n")
cat("Study-specific High checks:", sum(study_meta$priority == "High"), "\n")
```

```{r meta-customize-priority-hint-1}
# check == "check_lb_lbornrlo_lbornrhi" ~ "High"
```

```{r meta-customize-priority-solution}
study_meta <- sdtmchecksmeta |>
  mutate(priority = case_when(
    check == "check_lb_lbornrlo_lbornrhi" ~ "High",
    TRUE ~ priority
  ))

study_meta |>
  filter(check == "check_lb_lbornrlo_lbornrhi") |>
  select(check, priority, description)

cat("Original High checks:", sum(sdtmchecksmeta$priority == "High"), "\n")
cat("Study-specific High checks:", sum(study_meta$priority == "High"), "\n")
```

---

## 6 — Priority triage

Each check has a priority: **High**, **Medium**, or **Low**.

```{r quiz-priority, echo=FALSE}
quiz(
  question(
    "Which priority level should you resolve before running any analysis?",
    answer("Low — these are the most informational"),
    answer("High — these strongly suggest a data error that would corrupt analysis", correct = TRUE),
    answer("Medium — these are the most common"),
    random_answer_order = TRUE
  ),
  question(
    "A Medium-priority issue is flagged but you can explain why it's expected for your study design. You should:",
    answer("Treat it as a High issue and block analysis"),
    answer("Document the explanation and confirm it's expected — no action on the data needed", correct = TRUE),
    answer("Delete the check from sdtmchecksmeta"),
    random_answer_order = TRUE
  ),
  question(
    "You can change a check's priority because:",
    answer("Priority is hard-coded in the function — it cannot be changed"),
    answer("Priority is stored in the sdtmchecksmeta data frame, which you can modify for your study", correct = TRUE),
    answer("You can only change it by forking the GitHub repo"),
    random_answer_order = TRUE
  )
)
```

---

## 7 — `run_all_checks()`: running the full suite

`run_all_checks()` loops over every check in the metadata and runs each against the matching named datasets in your environment.

```{r run-all-demo}
# Run all checks against our clean datasets
all_results <- run_all_checks(
  metads = sdtmchecksmeta,
  DM = dm,
  AE = ae,
  EX = ex,
  LB = lb,
  VS = vs
)

# Summarise results
lapply(names(all_results), function(nm) {
  r <- all_results[[nm]]
  tibble(check = nm, status = r$status, n_issues = r$n_issues %||% NA_integer_)
}) |> bind_rows()
```

### Exercise: Run all checks on datasets with known issues

Run `run_all_checks()` using the dirty datasets (`dm_dirty`, `ae_dirty`, `lb_dirty`) and compare the summary to the clean run above.

```{r run-all-dirty, exercise=TRUE, exercise.lines=20}
# Run all checks against the dirty datasets
dirty_results <- run_all_checks(
  metads = sdtmchecksmeta,
  DM = ___,
  AE = ___,
  EX = ex,
  LB = ___,
  VS = vs
)

# Summarise
summary_dirty <- lapply(names(dirty_results), function(nm) {
  r <- dirty_results[[nm]]
  tibble(check = nm, priority = r$priority %||% NA_character_,
         status = r$status, n_issues = r$n_issues %||% NA_integer_)
}) |> bind_rows()

# Show only failed checks
summary_dirty |> filter(status == "failed")
```

```{r run-all-dirty-hint-1}
# DM = dm_dirty, AE = ae_dirty, LB = lb_dirty
```

```{r run-all-dirty-solution}
dirty_results <- run_all_checks(
  metads = sdtmchecksmeta,
  DM = dm_dirty,
  AE = ae_dirty,
  EX = ex,
  LB = lb_dirty,
  VS = vs
)

summary_dirty <- lapply(names(dirty_results), function(nm) {
  r <- dirty_results[[nm]]
  tibble(check = nm, priority = r$priority %||% NA_character_,
         status = r$status, n_issues = r$n_issues %||% NA_integer_)
}) |> bind_rows()

summary_dirty |> filter(status == "failed")
```

### Exercise: Run only High-priority checks

Filter `sdtmchecksmeta` to High priority and pass the filtered metadata to `run_all_checks()`.

```{r run-high-only, exercise=TRUE, exercise.lines=18}
# Filter metadata to High priority
meta_high <- sdtmchecksmeta |>
  filter(priority == ___)

# Run only those checks
high_results <- run_all_checks(
  metads = meta_high,
  DM = dm_dirty,
  AE = ae_dirty,
  EX = ex,
  LB = lb_dirty,
  VS = vs
)

# Summary
lapply(names(high_results), function(nm) {
  r <- high_results[[nm]]
  tibble(check = nm, status = r$status, n_issues = r$n_issues %||% NA_integer_)
}) |> bind_rows()
```

```{r run-high-only-hint-1}
# filter(priority == "High")
```

```{r run-high-only-solution}
meta_high <- sdtmchecksmeta |>
  filter(priority == "High")

high_results <- run_all_checks(
  metads = meta_high,
  DM = dm_dirty,
  AE = ae_dirty,
  EX = ex,
  LB = lb_dirty,
  VS = vs
)

lapply(names(high_results), function(nm) {
  r <- high_results[[nm]]
  tibble(check = nm, status = r$status, n_issues = r$n_issues %||% NA_integer_)
}) |> bind_rows()
```

---

## 8 — `report_to_xlsx()`: generating the Excel report

For sharing with non-R stakeholders (data managers, clinical operations), `report_to_xlsx()` produces a multi-tab Excel file: a summary tab and one tab per failed check.

```{r report-demo}
# Generate an Excel QC report
report_to_xlsx(
  res     = dirty_results,   # from run_all_checks() above
  outfile = "sdtm_qc_report.xlsx"
)
```

In production, this file is what study teams review:

- **Summary tab**: every check, its priority, pass/fail status
- **One tab per failed check**: the offending rows for investigation

### Exercise: Generate a High-priority-only report

Combine `run_all_checks()` (High priority only) with `report_to_xlsx()`.

```{r report-high-only, exercise=TRUE, exercise.lines=14}
# Step 1: run only High priority checks
high_results2 <- run_all_checks(
  metads = sdtmchecksmeta |> filter(___),
  DM = dm_dirty, AE = ae_dirty, EX = ex, LB = lb_dirty, VS = vs
)

# Step 2: generate Excel report
report_to_xlsx(
  res     = high_results2,
  outfile = "high_priority_qc.xlsx"
)
```

```{r report-high-only-hint-1}
# filter(priority == "High")
```

```{r report-high-only-solution}
high_results2 <- run_all_checks(
  metads = sdtmchecksmeta |> filter(priority == "High"),
  DM = dm_dirty, AE = ae_dirty, EX = ex, LB = lb_dirty, VS = vs
)

report_to_xlsx(
  res     = high_results2,
  outfile = "high_priority_qc.xlsx"
)
```

---

## 9 — Writing your own custom check

Adding a new check follows a simple convention: write a function returning a data frame of offending rows.

```{r custom-check-demo}
# Custom check: subjects with more than 5 AE records (unusually high)
check_ae_high_count_per_subject <- function(AE) {
  AE |>
    group_by(USUBJID) |>
    summarise(n_ae = n(), .groups = "drop") |>
    filter(n_ae > 5) |>
    arrange(desc(n_ae))
}

# Test it
check_ae_high_count_per_subject(ae)
```

### Exercise: Write a check for VS records missing VSORRES

A valid VS record must have a non-missing `VSORRES`. Write `check_vs_missing_result()` that returns rows where VSORRES is `NA` or empty.

```{r custom-vs-check, exercise=TRUE, exercise.lines=20}
# Write the custom check
check_vs_missing_result <- function(VS) {
  VS |>
    filter(___)
}

# Test with clean VS
cat("Clean VS issues:", nrow(check_vs_missing_result(vs)), "\n")

# Test with a VS that has a missing result
vs_with_missing <- vs |>
  mutate(VSORRES = ifelse(row_number() == 2, NA_character_, VSORRES))

cat("Dirty VS issues:", nrow(check_vs_missing_result(vs_with_missing)), "\n")
check_vs_missing_result(vs_with_missing)
```

```{r custom-vs-check-hint-1}
# filter(is.na(VSORRES) | VSORRES == "")
```

```{r custom-vs-check-solution}
check_vs_missing_result <- function(VS) {
  VS |>
    filter(is.na(VSORRES) | VSORRES == "")
}

cat("Clean VS issues:", nrow(check_vs_missing_result(vs)), "\n")

vs_with_missing <- vs |>
  mutate(VSORRES = ifelse(row_number() == 2, NA_character_, VSORRES))

cat("Dirty VS issues:", nrow(check_vs_missing_result(vs_with_missing)), "\n")
check_vs_missing_result(vs_with_missing)
```

### Exercise: Write a check for subjects in DM with no EX records

Subjects who are randomized (ARMCD ≠ "SCRNFAIL") should have at least one exposure record. Write `check_dm_no_ex()` (already mocked — try writing it from scratch before looking).

```{r custom-dm-ex-check, exercise=TRUE, exercise.lines=16}
my_check_dm_no_ex <- function(DM, EX) {
  # 1. Get distinct USUBJIDs that appear in EX
  subjects_with_ex <- EX |> distinct(___)

  # 2. Find DM subjects not in EX, excluding screen failures
  DM |>
    anti_join(subjects_with_ex, by = "USUBJID") |>
    filter(ARMCD != "SCRNFAIL")
}

# Test: add a subject in DM who has no EX record
dm_with_no_dose <- bind_rows(dm,
  tibble(STUDYID = "TEST_STUDY", DOMAIN = "DM",
         USUBJID = "TS-705-1999", SUBJID = "1999",
         RFSTDTC = "2020-10-01", DTHDTC = NA_character_,
         ARMCD = "TRT", ARM = "TREATMENT"))

cat("Subjects in DM with no EX:", nrow(my_check_dm_no_ex(dm_with_no_dose, ex)), "\n")
my_check_dm_no_ex(dm_with_no_dose, ex)
```

```{r custom-dm-ex-check-hint-1}
# subjects_with_ex <- EX |> distinct(USUBJID)
```

```{r custom-dm-ex-check-solution}
my_check_dm_no_ex <- function(DM, EX) {
  subjects_with_ex <- EX |> distinct(USUBJID)
  DM |>
    anti_join(subjects_with_ex, by = "USUBJID") |>
    filter(ARMCD != "SCRNFAIL")
}

dm_with_no_dose <- bind_rows(dm,
  tibble(STUDYID = "TEST_STUDY", DOMAIN = "DM",
         USUBJID = "TS-705-1999", SUBJID = "1999",
         RFSTDTC = "2020-10-01", DTHDTC = NA_character_,
         ARMCD = "TRT", ARM = "TREATMENT"))

cat("Subjects in DM with no EX:", nrow(my_check_dm_no_ex(dm_with_no_dose, ex)), "\n")
my_check_dm_no_ex(dm_with_no_dose, ex)
```

### Exercise: Run a check on real pharmaversesdtm data

The `ae_sdtm_real` object holds the actual CDISC pilot AE domain from `{pharmaversesdtm}`. Run `check_ae_dup()` on it and compare how it looks versus the mock dataset.

```{r real-ae-check, exercise=TRUE, exercise.lines=16}
# How many subjects and AE records in real data?
cat("Real AE records:", nrow(ae_sdtm_real), "\n")
cat("Real AE subjects:", n_distinct(ae_sdtm_real$USUBJID), "\n")

# Run the duplicate check on real data
real_issues <- check_ae_dup(___)
cat("Duplicate AE records flagged:", nrow(real_issues), "\n")

# What AE terms appear in the real data? (first 10)
ae_sdtm_real |>
  distinct(AETERM) |>
  arrange(AETERM) |>
  head(10)
```

```{r real-ae-check-hint-1}
# check_ae_dup(ae_sdtm_real)
```

```{r real-ae-check-solution}
cat("Real AE records:", nrow(ae_sdtm_real), "\n")
cat("Real AE subjects:", n_distinct(ae_sdtm_real$USUBJID), "\n")

real_issues <- check_ae_dup(ae_sdtm_real)
cat("Duplicate AE records flagged:", nrow(real_issues), "\n")

ae_sdtm_real |>
  distinct(AETERM) |>
  arrange(AETERM) |>
  head(10)
```

### Exercise: Write a custom out-of-range lab check using real data

The real lab data in `lb_sdtm_real` includes `LBSTRESN` (numeric standardised result), `LBSTNRLO` (lower normal limit), and `LBSTNRHI` (upper normal limit). Write `check_lb_out_of_range()` that flags rows where the standardised result falls outside the normal range.

```{r custom-lb-out-of-range, exercise=TRUE, exercise.lines=22}
check_lb_out_of_range <- function(LB) {
  LB |>
    filter(!is.na(LBSTRESN) & !is.na(LBSTNRLO) & !is.na(LBSTNRHI)) |>
    filter(___ | ___)
}

# Run on real LB data
oor_issues <- check_lb_out_of_range(lb_sdtm_real)
cat("Out-of-range lab results:", nrow(oor_issues), "\n")

# Which tests are most commonly flagged?
oor_issues |>
  count(LBTESTCD, sort = TRUE) |>
  head(10)
```

```{r custom-lb-out-of-range-hint-1}
# filter(LBSTRESN < LBSTNRLO | LBSTRESN > LBSTNRHI)
```

```{r custom-lb-out-of-range-solution}
check_lb_out_of_range <- function(LB) {
  LB |>
    filter(!is.na(LBSTRESN) & !is.na(LBSTNRLO) & !is.na(LBSTNRHI)) |>
    filter(LBSTRESN < LBSTNRLO | LBSTRESN > LBSTNRHI)
}

oor_issues <- check_lb_out_of_range(lb_sdtm_real)
cat("Out-of-range lab results:", nrow(oor_issues), "\n")

oor_issues |>
  count(LBTESTCD, sort = TRUE) |>
  head(10)
```

---

## 10 — Assertion-style wrapping for production

`{sdtmchecks}` returns a data frame — it doesn't throw errors. For a production pipeline where you want the script to *stop* on a High-priority failure, wrap checks in an assertion helper.

```{r assertion-demo}
# Assertion wrapper
stop_if_issues <- function(check_fn, ..., msg = NULL) {
  issues <- check_fn(...)
  if (nrow(issues) > 0) {
    stop(
      msg %||% sprintf("Check failed: %s", deparse(substitute(check_fn))),
      "\n  Flagged rows: ", nrow(issues),
      "\n  First issue: USUBJID = ", issues$USUBJID[1]
    )
  }
  message(deparse(substitute(check_fn)), ": PASSED")
  invisible(issues)
}

# Use it — runs silently on clean data
stop_if_issues(check_ae_dup, AE = ae)
```

```{r assertion-fail-demo}
# Fails loudly on dirty data
tryCatch(
  stop_if_issues(check_ae_dup, AE = ae_dirty),
  error = function(e) cat("PIPELINE BLOCKED:\n", conditionMessage(e), "\n")
)
```

### Exercise: Build a "critical checks" block for a pipeline

Write a series of `stop_if_issues()` calls that would block a production pipeline if any High-priority check fails. Use the clean datasets so they all pass.

```{r assertion-pipeline, exercise=TRUE, exercise.lines=16}
run_critical_checks <- function(DM, AE, EX, LB) {
  stop_if_issues(check_dm_usubjid_dup,             DM = DM)
  stop_if_issues(check_ae_dup,                     AE = AE)
  stop_if_issues(check_ex_dup,                     EX = EX)
  stop_if_issues(check_ae_aestdtc_after_dm_death,  AE = AE, DM = DM)
  stop_if_issues(check_ae_aestdtc_before_exstdtc,  AE = AE, EX = EX)
  message("All critical checks passed.")
}

# Run against clean datasets — should pass silently
run_critical_checks(DM = ___, AE = ___, EX = ___, LB = ___)
```

```{r assertion-pipeline-hint-1}
# run_critical_checks(DM = dm, AE = ae, EX = ex, LB = lb)
```

```{r assertion-pipeline-solution}
run_critical_checks <- function(DM, AE, EX, LB) {
  stop_if_issues(check_dm_usubjid_dup,             DM = DM)
  stop_if_issues(check_ae_dup,                     AE = AE)
  stop_if_issues(check_ex_dup,                     EX = EX)
  stop_if_issues(check_ae_aestdtc_after_dm_death,  AE = AE, DM = DM)
  stop_if_issues(check_ae_aestdtc_before_exstdtc,  AE = AE, EX = EX)
  message("All critical checks passed.")
}

run_critical_checks(DM = dm, AE = ae, EX = ex, LB = lb)
```

---

## 11 — Roche-specific assumptions

`{sdtmchecks}` originated at Roche. A practical caveat:

```{r quiz-roche-caveat, echo=FALSE}
quiz(
  question(
    "A check flags an issue in your study that looks like a Roche-specific assumption. You should:",
    answer("Accept the flag — Roche's conventions are always correct"),
    answer("Read the check's help page, decide if it applies to your study, and document/disable if not relevant", correct = TRUE),
    answer("Remove the check from sdtmchecksmeta permanently"),
    random_answer_order = TRUE
  ),
  question(
    "The best way to propose a generalization of a Roche-specific check is:",
    answer("Email the Roche team directly"),
    answer("Open a pull request on pharmaverse/sdtmchecks with a more general implementation", correct = TRUE),
    answer("Write a competing package"),
    random_answer_order = TRUE
  )
)
```

---

## 12 — Full QC workflow: capstone exercise

Put it all together. Given the `dm_dirty`, `ae_early` (AEs before first dose), and `lb_dirty` datasets:

1. Run `run_all_checks()` with High priority checks only
2. Print which checks failed
3. For the failing AE check, inspect the offending rows
4. Call `report_to_xlsx()` to generate the report

```{r capstone-workflow, exercise=TRUE, exercise.lines=35}
# Step 1: Run High priority checks
high_meta <- sdtmchecksmeta |> filter(priority == "High")

results <- run_all_checks(
  metads = high_meta,
  DM = dm_dirty,
  AE = ae_early,
  EX = ex,
  LB = lb_dirty,
  VS = vs
)

# Step 2: Summary of failed checks
cat("=== QC Summary ===\n")
failed_checks <- lapply(names(results), function(nm) {
  r <- results[[nm]]
  if (r$status == "failed")
    tibble(check = nm, n_issues = r$n_issues)
}) |> bind_rows()

print(failed_checks)

# Step 3: Inspect AE early-dose issue
cat("\n=== AE before first dose ===\n")
print(results[["check_ae_aestdtc_before_exstdtc"]]$issues)

# Step 4: Generate Excel report
report_to_xlsx(results, outfile = "capstone_qc_report.xlsx")
```

```{r capstone-workflow-hint-1}
# This exercise is mostly pre-filled. Just run it and observe:
# - dm_dirty fails check_dm_usubjid_dup (1 duplicate subject)
# - ae_early fails check_ae_aestdtc_before_exstdtc (1 AE before first dose)
# - lb_dirty may or may not fail depending on which High checks touch LB
```

---

## 13 — `{datacutr}`: a sibling package

```{r quiz-datacutr, echo=FALSE}
quiz(
  question(
    "The primary purpose of {datacutr} is:",
    answer("Running quality checks like sdtmchecks, but for ADaM"),
    answer("Applying a data cutoff date to SDTM datasets for interim analyses", correct = TRUE),
    answer("Converting SAS XPT files to R data frames"),
    random_answer_order = TRUE
  )
)
```

---

## 14 — Key takeaways and Module 2 wrap-up

```{r quiz-final-11, echo=FALSE}
quiz(
  question(
    "A check function returns an empty data frame. This means:",
    answer("The check encountered an error and couldn't run"),
    answer("The check passed — no offending rows found", correct = TRUE),
    answer("The check is not applicable to this domain"),
    random_answer_order = TRUE
  ),
  question(
    "To run only the checks relevant to your currently available domains, you:",
    answer("Run all checks — skipped checks just return NULL"),
    answer("Filter sdtmchecksmeta by the domains you have, then pass to run_all_checks()", correct = TRUE),
    answer("Delete the inapplicable functions from the package"),
    random_answer_order = TRUE
  ),
  question(
    "The check function naming convention `check_ae_aestdtc_after_dm_death` tells you:",
    answer("The domain (ae), the variable (aestdtc), the relationship (after), and the comparison domain/variable (dm_death)", correct = TRUE),
    answer("Nothing — names are arbitrary"),
    answer("The priority level of the check"),
    random_answer_order = TRUE
  ),
  question(
    "You are running sdtmchecks as part of an automated CI pipeline. You want the build to fail if any High-priority check fails. The best pattern is:",
    answer("Use run_all_checks() and check the return value manually"),
    answer("Wrap each High-priority check in stop_if_issues() so it throws an error on failure", correct = TRUE),
    answer("Email the team if checks fail"),
    random_answer_order = TRUE
  ),
  question(
    "You need both sdtmchecks AND Pinnacle 21 because:",
    answer("sdtmchecks catches analysis-impacting issues; P21 catches CDISC conformance issues — they cover different problem classes", correct = TRUE),
    answer("They duplicate each other — pick whichever your sponsor prefers"),
    answer("sdtmchecks only runs on ADaM; P21 only runs on SDTM"),
    random_answer_order = TRUE
  ),
  question(
    "A check for `check_ae_aeendtc_before_aestdtc` on a dataset where some AEs have no AEENDTC. Those rows will:",
    answer("Cause the check to error with an NA comparison warning"),
    answer("Be excluded from flagging because the filter requires both dates to be non-missing", correct = TRUE),
    answer("Always be flagged as issues since NA is treated as the earliest possible date"),
    random_answer_order = TRUE
  ),
  question(
    "Your study protocol allows AEs to be recorded up to 30 days after the last dose. The standard `check_ae_aestdtc_before_exstdtc` flags one AE that is actually a valid late-reported pre-dose event. You should:",
    answer("Delete the AE record from SDTM"),
    answer("Change the check's logic in the package source"),
    answer("Document that this AE is expected, and consider writing a study-specific check that reflects the protocol window", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "The `sdtmchecksmeta` column named `datasets` contains a comma-separated string like `'AE,DM'`. Inside `run_all_checks()` this string is used to:",
    answer("Validate that the domain has the required CDISC columns"),
    answer("Look up the named datasets from the `...` arguments passed to run_all_checks()", correct = TRUE),
    answer("Set the order in which domains are loaded from XPT files"),
    random_answer_order = TRUE
  ),
  question(
    "You write a custom check and want it to appear in `run_all_checks()` output without any changes to the package. You should:",
    answer("Submit a pull request to pharmaverse/sdtmchecks"),
    answer("Add a row for your check to your study-specific copy of sdtmchecksmeta and define the function in your session", correct = TRUE),
    answer("Override the run_all_checks() function"),
    random_answer_order = TRUE
  ),
  question(
    "`{datacutr}` is used before running `{sdtmchecks}` in an interim analysis workflow because:",
    answer("datacutr applies the data cutoff date so that sdtmchecks only evaluates data within the analysis window", correct = TRUE),
    answer("datacutr converts SAS XPT files to R data frames that sdtmchecks can read"),
    answer("datacutr removes all Medium-priority issues automatically"),
    random_answer_order = TRUE
  )
)
```

---

## Module 2 Complete — What's Next

Congratulations on completing Module 2: Raw Data and SDTM.

You've covered:

- Lesson 07: `{pharmaverseraw}` — the test data chain's starting point
- Lesson 08: `{sdtm.oak}` Part 1 — architecture and algorithm functions
- Lesson 09: `{sdtm.oak}` Part 2 — building full Findings and Events domains
- Lesson 10: `{sdtm.oak}` Part 3 — SUPP--, RELREC, and edge cases
- Lesson 11: `{sdtmchecks}` — analysis-focused quality validation

**Module 3** (coming next) covers metadata-driven programming with `{metacore}` (the spec object) and `{metatools}` (using the spec to validate and build datasets). These are smaller packages but load-bearing for traceability from specification to final dataset.

After Module 3, **Module 4** enters the heart of pharmaverse: `{admiral}` for ADaM construction — ADSL, ADAE, ADLB, TTE endpoints, and more.
