---
title: "Lesson 12 — {metacore}: The Centralized Spec Object"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build and query a metacore spec object — the R equivalent of an Excel
  specification or Define-XML. Learn the six tables, select_dataset(),
  and metadata-driven programming principles.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 12_metacore.Rmd
# PURPOSE       : metacore package for CDISC metadata management: creating and validating
#                 metacore specification objects, variable-level and
#                 dataset-level metadata, and integration with admiral
#                 and xportr.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── Mock metacore infrastructure ─────────────────────────────────────────────
# A metacore object is an R6 container with 6 tibbles.
# We mock it as an S3 list with a print method and a validator.

metacore <- function(ds_spec, ds_vars, var_spec,
                     value_spec = tibble(variable = character(), dataset = character(),
                                         where = character(), code_id = character()),
                     code_list  = tibble(code_id = character(), name = character(),
                                         code = character(), decode = character()),
                     derivations = tibble(derivation_id = character(),
                                          derivation = character())) {
  # Validate cross-table consistency
  missing_datasets <- setdiff(unique(ds_vars$dataset), ds_spec$dataset)
  if (length(missing_datasets) > 0)
    stop("ds_vars references datasets not in ds_spec: ",
         paste(missing_datasets, collapse = ", "))

  missing_vars <- setdiff(unique(ds_vars$variable), var_spec$variable)
  if (length(missing_vars) > 0)
    stop("ds_vars references variables not in var_spec: ",
         paste(missing_vars, collapse = ", "))

  if (nrow(value_spec) > 0) {
    bad_codes <- setdiff(value_spec$code_id[!is.na(value_spec$code_id)],
                         unique(code_list$code_id))
    if (length(bad_codes) > 0)
      stop("value_spec references code_ids not in code_list: ",
           paste(bad_codes, collapse = ", "))
  }

  obj <- list(ds_spec = ds_spec, ds_vars = ds_vars, var_spec = var_spec,
              value_spec = value_spec, code_list = code_list,
              derivations = derivations)
  class(obj) <- c("Metacore", "list")
  obj
}

print.Metacore <- function(x, ...) {
  cat("── Metacore Object ─────────────────────────────────────────\n")
  cat(sprintf("  ds_spec    : %d dataset(s)  — %s\n",
              nrow(x$ds_spec), paste(x$ds_spec$dataset, collapse = ", ")))
  cat(sprintf("  ds_vars    : %d variable-dataset mappings\n", nrow(x$ds_vars)))
  cat(sprintf("  var_spec   : %d variables\n", nrow(x$var_spec)))
  cat(sprintf("  value_spec : %d value-level entries\n", nrow(x$value_spec)))
  cat(sprintf("  code_list  : %d codelist entries\n", nrow(x$code_list)))
  cat(sprintf("  derivations: %d derivations\n", nrow(x$derivations)))
  invisible(x)
}

# DatasetMeta: metacore subset to one dataset
select_dataset <- function(metacore_obj, dataset) {
  if (!dataset %in% metacore_obj$ds_spec$dataset)
    stop("Dataset '", dataset, "' not found in metacore object")

  vars_in_ds   <- metacore_obj$ds_vars |> filter(dataset == !!dataset) |> pull(variable)
  ds_spec2     <- metacore_obj$ds_spec  |> filter(dataset == !!dataset)
  ds_vars2     <- metacore_obj$ds_vars  |> filter(dataset == !!dataset)
  var_spec2    <- metacore_obj$var_spec |> filter(variable %in% vars_in_ds)
  value_spec2  <- metacore_obj$value_spec  |> filter(dataset == !!dataset | is.na(dataset))
  code_ids     <- unique(value_spec2$code_id[!is.na(value_spec2$code_id)])
  code_list2   <- metacore_obj$code_list |> filter(code_id %in% code_ids)
  derivs2      <- metacore_obj$derivations

  obj <- list(dataset = dataset,
              ds_spec = ds_spec2, ds_vars = ds_vars2, var_spec = var_spec2,
              value_spec = value_spec2, code_list = code_list2,
              derivations = derivs2)
  class(obj) <- c("DatasetMeta", "Metacore", "list")
  obj
}

print.DatasetMeta <- function(x, ...) {
  cat(sprintf("── DatasetMeta: %s ──────────────────────────────────────\n", x$dataset))
  cat(sprintf("  Variables  : %d\n", nrow(x$ds_vars)))
  cat(sprintf("  Codelists  : %d code_ids\n", length(unique(x$code_list$code_id))))
  cat(sprintf("  Derivations: %d\n", nrow(x$derivations)))
  invisible(x)
}

# ── Pre-built example metacore objects ───────────────────────────────────────

# -- Codelists --
adsl_code_list <- tribble(
  ~code_id,  ~name,    ~code, ~decode,
  "SEX_CL",  "SEX",    "M",   "Male",
  "SEX_CL",  "SEX",    "F",   "Female",
  "SEX_CL",  "SEX",    "U",   "Unknown",
  "RACE_CL", "RACE",   "1",   "AMERICAN INDIAN OR ALASKA NATIVE",
  "RACE_CL", "RACE",   "2",   "ASIAN",
  "RACE_CL", "RACE",   "3",   "BLACK OR AFRICAN AMERICAN",
  "RACE_CL", "RACE",   "4",   "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER",
  "RACE_CL", "RACE",   "5",   "WHITE",
  "RACE_CL", "RACE",   "6",   "MULTIPLE",
  "RACE_CL", "RACE",   "8",   "UNKNOWN",
  "RACE_CL", "RACE",   "9",   "NOT REPORTED",
  "ETHNIC_CL","ETHNIC","1",   "HISPANIC OR LATINO",
  "ETHNIC_CL","ETHNIC","2",   "NOT HISPANIC OR LATINO",
  "ETHNIC_CL","ETHNIC","3",   "UNKNOWN",
  "TRT_CL",  "TRT01A", "1",   "XANOMELINE HIGH DOSE",
  "TRT_CL",  "TRT01A", "2",   "XANOMELINE LOW DOSE",
  "TRT_CL",  "TRT01A", "3",   "PLACEBO"
)

# -- var_spec --
adsl_var_spec <- tribble(
  ~variable,  ~label,                          ~type,        ~length, ~format,
  "STUDYID",  "Study Identifier",               "character",  8,       NA,
  "USUBJID",  "Unique Subject Identifier",      "character",  32,      NA,
  "SUBJID",   "Subject Identifier for Study",   "character",  8,       NA,
  "SITEID",   "Study Site Identifier",          "character",  3,       NA,
  "AGE",      "Age (years)",                    "numeric",    8,       NA,
  "AGEU",     "Age Units",                      "character",  5,       NA,
  "AGEGR1",   "Age Group 1",                    "character",  5,       NA,
  "AGEGR1N",  "Age Group 1 (N)",                "numeric",    8,       NA,
  "SEX",      "Sex",                            "character",  1,       NA,
  "SEXN",     "Sex (N)",                        "numeric",    8,       NA,
  "RACE",     "Race",                           "character",  60,      NA,
  "RACEN",    "Race (N)",                       "numeric",    8,       NA,
  "ETHNIC",   "Ethnicity",                      "character",  25,      NA,
  "TRT01P",   "Planned Treatment for Period 01","character",  40,      NA,
  "TRT01A",   "Actual Treatment for Period 01", "character",  40,      NA,
  "TRT01AN",  "Actual Treatment for Period 01 (N)","numeric", 8,       NA,
  "TRTSDT",   "Date of First Exposure",         "numeric",    8,       "DATE9.",
  "TRTEDT",   "Date of Last Exposure",          "numeric",    8,       "DATE9.",
  "TRTDURD",  "Total Treatment Duration (Days)","numeric",    8,       NA,
  "RANDDT",   "Date of Randomization",          "numeric",    8,       "DATE9.",
  "EOSDT",    "Date of End of Study",           "numeric",    8,       "DATE9.",
  "SAFFL",    "Safety Population Flag",         "character",  1,       NA,
  "ITTFL",    "Intent-To-Treat Population Flag","character",  1,       NA
)

# -- ds_spec --
adsl_ds_spec <- tibble(
  dataset   = c("ADSL", "ADAE"),
  structure  = c("One record per subject", "One record per subject per adverse event"),
  label      = c("Subject-Level Analysis Dataset", "Adverse Events Analysis Dataset")
)

# -- ds_vars --
adsl_ds_vars <- tibble(
  dataset   = "ADSL",
  variable  = adsl_var_spec$variable,
  order     = seq_along(adsl_var_spec$variable),
  mandatory = c(TRUE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, FALSE,
                TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, TRUE, FALSE,
                FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE),
  core      = c("Required", "Required", "Required", "Permissible",
                "Required", "Permissible", "Permissible", "Permissible",
                "Required", "Permissible", "Required", "Permissible",
                "Permissible", "Required", "Required", "Permissible",
                "Permissible", "Permissible", "Permissible", "Permissible",
                "Permissible", "Permissible", "Permissible")
)

# Add a few ADAE rows too (to demonstrate multi-dataset metacore)
adae_var_spec <- tribble(
  ~variable, ~label,                        ~type,        ~length, ~format,
  "STUDYID", "Study Identifier",             "character",  8,       NA,
  "USUBJID", "Unique Subject Identifier",    "character",  32,      NA,
  "AETERM",  "Reported Term for AE",         "character",  200,     NA,
  "AESEV",   "Severity/Intensity",           "character",  5,       NA,
  "AESEQ",   "Sequence Number",              "numeric",    8,       NA
)

adae_ds_vars <- tibble(
  dataset   = "ADAE",
  variable  = adae_var_spec$variable,
  order     = 1:5,
  mandatory = c(TRUE, TRUE, TRUE, FALSE, TRUE),
  core      = c("Required", "Required", "Required", "Permissible", "Required")
)

combined_var_spec <- bind_rows(adsl_var_spec,
  adae_var_spec |> filter(!variable %in% adsl_var_spec$variable))

combined_ds_vars <- bind_rows(adsl_ds_vars, adae_ds_vars)

# -- value_spec --
adsl_value_spec <- tribble(
  ~variable, ~dataset, ~where, ~code_id,
  "SEX",     "ADSL",   NA,     "SEX_CL",
  "RACE",    "ADSL",   NA,     "RACE_CL",
  "ETHNIC",  "ADSL",   NA,     "ETHNIC_CL",
  "TRT01A",  "ADSL",   NA,     "TRT_CL"
)

# -- derivations --
adsl_derivations <- tribble(
  ~derivation_id, ~derivation,
  "DR_TRTSDT",  "Treatment start date: first dose date from EX (EXDOSE > 0 or placebo)",
  "DR_TRTEDT",  "Treatment end date: last dose end date from EX",
  "DR_TRTDURD", "TRTEDT - TRTSDT + 1",
  "DR_SAFFL",   "Y if subject has at least one EX record with EXDOSE > 0 or placebo",
  "DR_RACEN",   "Numeric from RACE codelist RACE_CL",
  "DR_AGEGR1",  "Age group: <18 / 18-64 / >=65"
)

# Build the full multi-dataset metacore
md_full <- metacore(
  ds_spec     = adsl_ds_spec,
  ds_vars     = combined_ds_vars,
  var_spec    = combined_var_spec,
  value_spec  = adsl_value_spec,
  code_list   = adsl_code_list,
  derivations = adsl_derivations
)

# Single-dataset DatasetMeta for ADSL
md_adsl <- select_dataset(md_full, "ADSL")

# Small mock subject-level dataset (as-if from DM)
dm_sample <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101"),
  SUBJID  = c("1015", "1023", "1042", "1101"),
  SITEID  = c("701", "701", "702", "702"),
  AGE     = c(63L, 64L, 71L, 56L),
  AGEU    = "YEARS",
  SEX     = c("M", "F", "M", "F"),
  RACE    = c("WHITE", "WHITE", "ASIAN", "WHITE"),
  ETHNIC  = c("NOT HISPANIC OR LATINO", "NOT HISPANIC OR LATINO",
              "NOT HISPANIC OR LATINO", "HISPANIC OR LATINO"),
  ARM     = c("Xanomeline High Dose", "Xanomeline Low Dose",
              "Placebo", "Xanomeline High Dose"),
  ACTARM  = c("XANOMELINE HIGH DOSE", "XANOMELINE LOW DOSE",
              "PLACEBO", "XANOMELINE HIGH DOSE")
)

# ── Real pharmaverseadam data for gap analysis ────────────────────────────────
library(pharmaverseadam)
adsl_real <- pharmaverseadam::adsl
```

## Welcome

This is Lesson 12 — the start of Module 3 — covering **`{metacore}`**: the centralized spec object.

In SAS workflows, the study specification lives in Excel and programmers transcribe it manually into code. When the spec changes, code drifts. `{metacore}` solves this by turning the spec into a **programmable R object** that your code reads from directly.

This lesson covers:

- Why metadata-driven programming matters
- The six tables that form a metacore object
- Building a metacore from scratch (and from Excel/Define-XML)
- Using `select_dataset()` to get a `DatasetMeta` for one dataset
- Querying the spec for labels, codelists, and derivation logic

All metacore functions are mocked and fully functional — no package installation required.

---

## 1 — The problem with "spec in Excel, code transcribed"

```{r quiz-why-metacore, echo=FALSE}
quiz(
  question(
    "Why is the 'spec lives in Excel, programmers transcribe it' workflow brittle?",
    answer("Excel is too slow to open"),
    answer("When the spec changes, all code that hard-codes spec details must be found and updated manually", correct = TRUE),
    answer("Regulators don't accept Excel-based specs"),
    random_answer_order = TRUE
  ),
  question(
    "In a metadata-driven pipeline, when the spec changes you:",
    answer("Update every R script that has a label or codelist"),
    answer("Update the spec source (Excel/Define-XML), rebuild the metacore object, and re-run — downstream code picks up changes automatically", correct = TRUE),
    answer("Re-do dual programming from scratch"),
    random_answer_order = TRUE
  )
)
```

---

## 2 — The six tables

A metacore object holds six tibbles. Let's explore them:

```{r explore-tables}
# Print the object — summary of all six tables
print(md_full)
```

```{r explore-ds-spec}
# Dataset-level info: names, labels, structure
md_full$ds_spec
```

```{r explore-ds-vars}
# Which variables belong to each dataset, in what order, with what flags
md_full$ds_vars |>
  filter(dataset == "ADSL") |>
  arrange(order) |>
  head(10)
```

```{r explore-var-spec}
# Variable-level info: label, type, length — shared across datasets
md_full$var_spec |>
  filter(variable %in% c("AGE", "SEX", "RACE", "USUBJID")) |>
  select(variable, label, type, length)
```

```{r explore-code-list}
# Codelist definitions
md_full$code_list |>
  filter(code_id == "SEX_CL")
```

```{r explore-value-spec}
# Which codelist applies to which variable in which dataset
md_full$value_spec
```

```{r explore-derivations}
# Derivation logic for derived variables
md_full$derivations
```

---

## 3 — Querying the spec

Because each table is a plain tibble, you use dplyr to query it:

### Exercise: Find all mandatory variables in ADSL

Use `md_full$ds_vars` to find variables where `dataset == "ADSL"` and `mandatory == TRUE`.

```{r query-mandatory, exercise=TRUE, exercise.lines=10}
# Find all mandatory ADSL variables
mandatory_adsl <- md_full$ds_vars |>
  filter(dataset == ___, mandatory == ___) |>
  arrange(order) |>
  pull(variable)

cat("Mandatory ADSL variables:\n")
cat(paste(mandatory_adsl, collapse = ", "), "\n")
```

```{r query-mandatory-hint-1}
# filter(dataset == "ADSL", mandatory == TRUE)
```

```{r query-mandatory-solution}
mandatory_adsl <- md_full$ds_vars |>
  filter(dataset == "ADSL", mandatory == TRUE) |>
  arrange(order) |>
  pull(variable)

cat("Mandatory ADSL variables:\n")
cat(paste(mandatory_adsl, collapse = ", "), "\n")
```

### Exercise: Get the label for TRTDURD

```{r query-label, exercise=TRUE, exercise.lines=6}
# Pull the label for TRTDURD from var_spec
label_trtdurd <- md_full$var_spec |>
  filter(variable == ___) |>
  pull(___)

cat("TRTDURD label:", label_trtdurd, "\n")
```

```{r query-label-hint-1}
# filter(variable == "TRTDURD") |> pull(label)
```

```{r query-label-solution}
label_trtdurd <- md_full$var_spec |>
  filter(variable == "TRTDURD") |>
  pull(label)

cat("TRTDURD label:", label_trtdurd, "\n")
```

### Exercise: Get the full RACE codelist

Find all entries in `code_list` that belong to the `"RACE_CL"` codelist.

```{r query-codelist, exercise=TRUE, exercise.lines=8}
race_codelist <- md_full$code_list |>
  filter(code_id == ___) |>
  select(code, decode) |>
  arrange(as.integer(code))

race_codelist
```

```{r query-codelist-hint-1}
# filter(code_id == "RACE_CL")
```

```{r query-codelist-solution}
race_codelist <- md_full$code_list |>
  filter(code_id == "RACE_CL") |>
  select(code, decode) |>
  arrange(as.integer(code))

race_codelist
```

### Exercise: Look up which codelist applies to TRT01A in ADSL

```{r query-which-codelist, exercise=TRUE, exercise.lines=10}
# Step 1: find the code_id for TRT01A in value_spec
trt_code_id <- md_full$value_spec |>
  filter(variable == ___, dataset == "ADSL") |>
  pull(___)

cat("code_id for TRT01A:", trt_code_id, "\n")

# Step 2: retrieve the codelist entries
md_full$code_list |>
  filter(code_id == trt_code_id)
```

```{r query-which-codelist-hint-1}
# filter(variable == "TRT01A", dataset == "ADSL") |> pull(code_id)
```

```{r query-which-codelist-solution}
trt_code_id <- md_full$value_spec |>
  filter(variable == "TRT01A", dataset == "ADSL") |>
  pull(code_id)

cat("code_id for TRT01A:", trt_code_id, "\n")

md_full$code_list |>
  filter(code_id == trt_code_id)
```

### Exercise: Gap analysis — spec variables vs. real data

`adsl_real` is the actual `pharmaverseadam::adsl` dataset. Use `setdiff()` to identify:

1. Variables defined in the spec but absent from the real dataset (planned but not yet derived, or simplified away from the full spec)
2. Variables present in the real dataset but absent from our simplified spec (columns we didn't document)

This is a standard QC step before every ADaM submission.

```{r query-gap-analysis, exercise=TRUE, exercise.lines=18}
# Extract ADSL variable list from the spec
spec_vars <- md_full$ds_vars |>
  filter(dataset == "ADSL") |>
  pull(variable)

# 1. In spec but NOT in the real dataset
in_spec_not_data <- setdiff(___, names(adsl_real))
cat("In spec, not in real data:\n")
cat(paste(in_spec_not_data, collapse = ", "), "\n\n")

# 2. In real data but NOT in our simplified spec (show first 15)
in_data_not_spec <- setdiff(names(adsl_real), ___) |> head(15)
cat("In real data, not in spec (first 15):\n")
cat(paste(in_data_not_spec, collapse = ", "), "\n")
```

```{r query-gap-analysis-hint-1}
# setdiff(spec_vars, names(adsl_real))   # in spec, missing from data
# setdiff(names(adsl_real), spec_vars)   # in data, not in spec
```

```{r query-gap-analysis-solution}
spec_vars <- md_full$ds_vars |>
  filter(dataset == "ADSL") |>
  pull(variable)

in_spec_not_data <- setdiff(spec_vars, names(adsl_real))
cat("In spec, not in real data:\n")
cat(paste(in_spec_not_data, collapse = ", "), "\n\n")

in_data_not_spec <- setdiff(names(adsl_real), spec_vars) |> head(15)
cat("In real data, not in spec (first 15):\n")
cat(paste(in_data_not_spec, collapse = ", "), "\n")
```

> **Why this matters**: Variables in the spec but absent from the data are un-derived — a submission risk. Variables in the data but absent from the spec are undocumented. Both should be resolved before data lock.

---

## 4 — `select_dataset()`: focusing on one dataset

A full-study metacore object holds metadata for every dataset. When building a single ADaM (e.g., ADSL), you need only that dataset's slice. `select_dataset()` produces a `DatasetMeta` — a metacore subclass restricted to one dataset.

```{r select-dataset-demo}
# Full multi-dataset metacore
print(md_full)

# Subset to ADSL only
md_adsl_local <- select_dataset(md_full, "ADSL")
print(md_adsl_local)
```

The `DatasetMeta` has the same 6 tables, but filtered — only `ADSL` variables, only the codelists ADSL needs.

```{r select-dataset-compare}
# Variables in full vs. ADSL-only
cat("ds_vars in full metacore:  ", nrow(md_full$ds_vars), "\n")
cat("ds_vars in ADSL DatasetMeta:", nrow(md_adsl$ds_vars), "\n")
```

### Exercise: Select the ADAE dataset

```{r select-adae, exercise=TRUE, exercise.lines=8}
# Create a DatasetMeta for ADAE
md_adae <- select_dataset(md_full, ___)

print(md_adae)

# What are the ADAE variables?
md_adae$ds_vars |> pull(variable)
```

```{r select-adae-hint-1}
# select_dataset(md_full, "ADAE")
```

```{r select-adae-solution}
md_adae <- select_dataset(md_full, "ADAE")
print(md_adae)
md_adae$ds_vars |> pull(variable)
```

### Exercise: Confirm that selecting a non-existent dataset errors

```{r select-invalid, exercise=TRUE, exercise.lines=6}
# What happens when you select a dataset that doesn't exist?
tryCatch(
  select_dataset(md_full, "ADVS"),
  error = function(e) cat("Error caught:", conditionMessage(e), "\n")
)
```

---

## 5 — The `mandatory` flag and `core`

```{r quiz-mandatory, echo=FALSE}
quiz(
  question(
    "In metacore, `mandatory = TRUE` means:",
    answer("The variable is nice to have but optional"),
    answer("The variable cannot have blank values in the delivered dataset", correct = TRUE),
    answer("The variable is derived, not collected"),
    random_answer_order = TRUE
  ),
  question(
    "The `core` field can be 'Required', 'Expected', or 'Permissible'. 'Required' means:",
    answer("The variable is in the SDTMIG and must be present in every submission", correct = TRUE),
    answer("The variable is recommended but optional"),
    answer("The variable is sponsor-specific"),
    random_answer_order = TRUE
  ),
  question(
    "In metacore 0.3.0, `keep` was renamed to `mandatory` because:",
    answer("'keep' was too long to type"),
    answer("'mandatory' better aligns with CDISC terminology for required variables", correct = TRUE),
    answer("The keep flag was removed entirely"),
    random_answer_order = TRUE
  )
)
```

---

## 6 — Building a metacore object from scratch

The constructor `metacore()` validates all six tables on construction. Let's build a minimal one:

```{r build-manual-demo}
# Step 1: the six tables
my_ds_spec <- tibble(
  dataset   = "ADSL",
  structure  = "One record per subject",
  label      = "Subject-Level Analysis Dataset"
)

my_ds_vars <- tibble(
  dataset   = "ADSL",
  variable  = c("STUDYID", "USUBJID", "AGE", "SEX"),
  order     = 1:4,
  mandatory = c(TRUE, TRUE, TRUE, TRUE),
  core      = c("Required", "Required", "Required", "Required")
)

my_var_spec <- tibble(
  variable = c("STUDYID", "USUBJID", "AGE", "SEX"),
  label    = c("Study Identifier", "Unique Subject Identifier", "Age (years)", "Sex"),
  type     = c("character", "character", "numeric", "character"),
  length   = c(8L, 32L, 8L, 1L),
  format   = c(NA, NA, NA, NA)
)

my_value_spec <- tibble(
  variable = "SEX",
  dataset  = "ADSL",
  where    = NA_character_,
  code_id  = "SEX_CL"
)

my_code_list <- tribble(
  ~code_id, ~name, ~code, ~decode,
  "SEX_CL", "SEX", "M",   "Male",
  "SEX_CL", "SEX", "F",   "Female",
  "SEX_CL", "SEX", "U",   "Unknown"
)

my_derivations <- tibble(
  derivation_id = "DR_AGE",
  derivation    = "From DM.AGE, no derivation needed"
)

# Step 2: construct — validation happens here
md_manual <- metacore(
  ds_spec     = my_ds_spec,
  ds_vars     = my_ds_vars,
  var_spec    = my_var_spec,
  value_spec  = my_value_spec,
  code_list   = my_code_list,
  derivations = my_derivations
)

print(md_manual)
```

### Exercise: Trigger a validation error

Add a variable `"WEIGHT"` to `ds_vars` that doesn't exist in `var_spec`, and observe the error:

```{r build-invalid, exercise=TRUE, exercise.lines=25}
bad_ds_vars <- tibble(
  dataset   = "ADSL",
  variable  = c("STUDYID", "USUBJID", "AGE", "SEX", "WEIGHT"),  # WEIGHT not in var_spec
  order     = 1:5,
  mandatory = c(TRUE, TRUE, TRUE, TRUE, FALSE),
  core      = c("Required", "Required", "Required", "Required", "Permissible")
)

tryCatch(
  metacore(
    ds_spec     = my_ds_spec,
    ds_vars     = bad_ds_vars,
    var_spec    = my_var_spec,
    value_spec  = my_value_spec,
    code_list   = my_code_list,
    derivations = my_derivations
  ),
  error = function(e) cat("Validation caught:\n", conditionMessage(e), "\n")
)
```

### Exercise: Fix the validation error — add WEIGHT to `var_spec`

The previous exercise failed because `WEIGHT` was in `ds_vars` but not in `var_spec`. Fix this by adding a `WEIGHT` row to `var_spec` first, then rebuild successfully.

```{r build-extend, exercise=TRUE, exercise.lines=28}
# Step 1: add WEIGHT to var_spec
my_var_spec_v2 <- bind_rows(
  my_var_spec,
  tibble(
    variable = ___,
    label    = "Weight (kg)",
    type     = "numeric",
    length   = 8L,
    format   = NA_character_
  )
)

# Step 2: add WEIGHT to ds_vars
my_ds_vars_v2 <- bind_rows(
  my_ds_vars,
  tibble(dataset = "ADSL", variable = "WEIGHT",
         order = 5L, mandatory = FALSE, core = "Permissible")
)

# Step 3: build — should succeed now
md_v2 <- metacore(
  ds_spec     = my_ds_spec,
  ds_vars     = ___,
  var_spec    = ___,
  value_spec  = my_value_spec,
  code_list   = my_code_list,
  derivations = my_derivations
)

cat("WEIGHT in var_spec:", "WEIGHT" %in% md_v2$var_spec$variable, "\n")
cat("WEIGHT in ds_vars:", "WEIGHT" %in% md_v2$ds_vars$variable, "\n")
print(md_v2)
```

```{r build-extend-hint-1}
# variable = "WEIGHT"
# ds_vars = my_ds_vars_v2, var_spec = my_var_spec_v2
```

```{r build-extend-solution}
my_var_spec_v2 <- bind_rows(
  my_var_spec,
  tibble(
    variable = "WEIGHT",
    label    = "Weight (kg)",
    type     = "numeric",
    length   = 8L,
    format   = NA_character_
  )
)

my_ds_vars_v2 <- bind_rows(
  my_ds_vars,
  tibble(dataset = "ADSL", variable = "WEIGHT",
         order = 5L, mandatory = FALSE, core = "Permissible")
)

md_v2 <- metacore(
  ds_spec     = my_ds_spec,
  ds_vars     = my_ds_vars_v2,
  var_spec    = my_var_spec_v2,
  value_spec  = my_value_spec,
  code_list   = my_code_list,
  derivations = my_derivations
)

cat("WEIGHT in var_spec:", "WEIGHT" %in% md_v2$var_spec$variable, "\n")
cat("WEIGHT in ds_vars:", "WEIGHT" %in% md_v2$ds_vars$variable, "\n")
print(md_v2)
```

---

## 7 — The `DatasetMeta` requirement

```{r quiz-dataset-meta, echo=FALSE}
quiz(
  question(
    "Why do metatools functions require a DatasetMeta (not a full metacore)?",
    answer("DatasetMeta is faster to compute with"),
    answer("It prevents ambiguity — when working on ADSL, you don't want code accidentally using ADAE metadata", correct = TRUE),
    answer("Full metacore objects can't be passed to functions"),
    random_answer_order = TRUE
  ),
  question(
    "In a typical session, you create one DatasetMeta per dataset because:",
    answer("You need a separate metacore object for each dataset build, ensuring metadata is scoped correctly", correct = TRUE),
    answer("select_dataset() modifies the original object"),
    answer("Full metacore objects can only hold one dataset"),
    random_answer_order = TRUE
  )
)
```

---

## 8 — Reading a derivation from the spec

In production, your code should *implement* the spec's derivation text exactly. The `derivations` table stores the written logic:

```{r read-derivation-demo}
# What does the spec say about how SAFFL should be derived?
md_full$derivations |>
  filter(derivation_id == "DR_SAFFL") |>
  pull(derivation)
```

### Exercise: Retrieve the derivation for TRTSDT and TRTDURD

```{r read-derivation-ex, exercise=TRUE, exercise.lines=10}
# Get derivation text for TRTSDT
cat("TRTSDT derivation:\n")
md_full$derivations |>
  filter(derivation_id == ___) |>
  pull(derivation) |>
  cat()

cat("\n\nTRTDURD derivation:\n")
md_full$derivations |>
  filter(derivation_id == ___) |>
  pull(derivation) |>
  cat()
```

```{r read-derivation-ex-hint-1}
# derivation_id == "DR_TRTSDT"
# derivation_id == "DR_TRTDURD"
```

```{r read-derivation-ex-solution}
cat("TRTSDT derivation:\n")
md_full$derivations |>
  filter(derivation_id == "DR_TRTSDT") |>
  pull(derivation) |>
  cat()

cat("\n\nTRTDURD derivation:\n")
md_full$derivations |>
  filter(derivation_id == "DR_TRTDURD") |>
  pull(derivation) |>
  cat()
```

### Exercise: Add a new derivation for `AGEGR1N`

The spec has been updated: `AGEGR1N` (numeric age group companion) needs a derivation entry. Use `bind_rows()` to add it to the derivations table, then confirm it's retrievable.

```{r add-derivation-ex, exercise=TRUE, exercise.lines=20}
# Add a new derivation row
updated_derivations <- bind_rows(
  md_full$derivations,
  tibble(
    derivation_id = ___,
    derivation    = ___
  )
)

# Confirm count increased
cat("Original derivation count:", nrow(md_full$derivations), "\n")
cat("Updated derivation count: ", nrow(updated_derivations), "\n\n")

# Retrieve the new entry
cat("AGEGR1N derivation:\n")
updated_derivations |>
  filter(derivation_id == "DR_AGEGR1N") |>
  pull(derivation) |>
  cat()
```

```{r add-derivation-ex-hint-1}
# derivation_id = "DR_AGEGR1N"
# derivation    = "Numeric companion to AGEGR1: 1 = <65, 2 = >=65"
```

```{r add-derivation-ex-solution}
updated_derivations <- bind_rows(
  md_full$derivations,
  tibble(
    derivation_id = "DR_AGEGR1N",
    derivation    = "Numeric companion to AGEGR1: 1 = <65, 2 = >=65"
  )
)

cat("Original derivation count:", nrow(md_full$derivations), "\n")
cat("Updated derivation count: ", nrow(updated_derivations), "\n\n")

cat("AGEGR1N derivation:\n")
updated_derivations |>
  filter(derivation_id == "DR_AGEGR1N") |>
  pull(derivation) |>
  cat()
```

---

## 9 — The pipeline perspective

```{r quiz-pipeline, echo=FALSE}
quiz(
  question(
    "Which package typically consumes the metacore object to perform actual dataset derivations?",
    answer("{xportr}"),
    answer("{metatools}", correct = TRUE),
    answer("{sdtm.oak}"),
    random_answer_order = TRUE
  ),
  question(
    "Where in the pipeline should the metacore object be constructed?",
    answer("Inside each programmer's derivation script"),
    answer("In one shared constructor script, saved as .rds, used by all downstream scripts", correct = TRUE),
    answer("It's created on-the-fly by admiral"),
    random_answer_order = TRUE
  ),
  question(
    "The metacore object is read-only after construction because:",
    answer("R6 objects can't be modified"),
    answer("Treating it as immutable ensures the spec is the single source of truth — no accidental in-place edits", correct = TRUE),
    answer("metatools modifies it and returns a new copy"),
    random_answer_order = TRUE
  )
)
```

---

## 10 — Capstone: Build and interrogate a spec

Build a metacore for a minimal ADVS dataset (Vital Signs) and answer questions about it.

```{r capstone-metacore, exercise=TRUE, exercise.lines=65}
# ── Build a metacore for ADVS ──────────────────────────────────────────────

advs_ds_spec <- tibble(
  dataset   = "ADVS",
  structure  = "One record per subject per parameter per visit",
  label      = "Vital Signs Analysis Dataset"
)

advs_ds_vars <- tibble(
  dataset   = "ADVS",
  variable  = c("STUDYID", "USUBJID", "PARAMCD", "PARAM",
                "AVAL", "ADT", "VISIT", "VISITNUM",
                "ABLFL", "ANL01FL"),
  order     = 1:10,
  mandatory = c(TRUE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE),
  core      = c("Required", "Required", "Required", "Permissible",
                "Required", "Permissible", "Permissible", "Permissible",
                "Permissible", "Permissible")
)

advs_var_spec <- tribble(
  ~variable,  ~label,                              ~type,        ~length, ~format,
  "STUDYID",  "Study Identifier",                  "character",  8,       NA,
  "USUBJID",  "Unique Subject Identifier",          "character",  32,      NA,
  "PARAMCD",  "Parameter Code",                    "character",  8,       NA,
  "PARAM",    "Parameter",                         "character",  40,      NA,
  "AVAL",     "Analysis Value",                    "numeric",    8,       NA,
  "ADT",      "Analysis Date",                     "numeric",    8,       "DATE9.",
  "VISIT",    "Visit Name",                        "character",  200,     NA,
  "VISITNUM", "Visit Number",                      "numeric",    8,       NA,
  "ABLFL",    "Baseline Record Flag",              "character",  1,       NA,
  "ANL01FL",  "Analysis Flag 01",                  "character",  1,       NA
)

advs_code_list <- tribble(
  ~code_id,    ~name,     ~code,    ~decode,
  "PARAMCD_CL","PARAMCD", "SYSBP",  "Systolic Blood Pressure (mmHg)",
  "PARAMCD_CL","PARAMCD", "DIABP",  "Diastolic Blood Pressure (mmHg)",
  "PARAMCD_CL","PARAMCD", "PULSE",  "Pulse Rate (beats/min)",
  "PARAMCD_CL","PARAMCD", "TEMP",   "Temperature (C)",
  "PARAMCD_CL","PARAMCD", "WEIGHT", "Weight (kg)"
)

advs_value_spec <- tibble(
  variable = "PARAMCD", dataset = "ADVS",
  where = NA_character_, code_id = "PARAMCD_CL"
)

advs_derivations <- tribble(
  ~derivation_id, ~derivation,
  "DR_ADT",  "Analysis date: from VSDTC in SDTM VS domain",
  "DR_AVAL", "Analysis value: numeric from VSORRES/VSSTRESC"
)

# Build the metacore
md_advs <- metacore(
  ds_spec     = advs_ds_spec,
  ds_vars     = advs_ds_vars,
  var_spec    = advs_var_spec,
  value_spec  = advs_value_spec,
  code_list   = advs_code_list,
  derivations = advs_derivations
)

# ── Questions to answer ────────────────────────────────────────────────────

# Q1: How many mandatory ADVS variables are there?
n_mandatory <- md_advs$ds_vars |>
  filter(mandatory == TRUE) |>
  nrow()
cat("Mandatory variables:", n_mandatory, "\n")

# Q2: What parameters are defined in the ADVS codelist?
params <- md_advs$code_list |>
  filter(code_id == "PARAMCD_CL") |>
  pull(code)
cat("Parameters:", paste(params, collapse = ", "), "\n")

# Q3: Select the dataset (produces a DatasetMeta)
md_advs_single <- select_dataset(md_advs, "ADVS")
print(md_advs_single)
```

```{r capstone-metacore-hint-1}
# This exercise is pre-filled — run it as-is to see the full output.
# Expected: 3 mandatory variables (STUDYID, USUBJID, PARAMCD, AVAL = 4 TRUE)
# Parameters: SYSBP, DIABP, PULSE, TEMP, WEIGHT
```

---

## 11 — Key takeaways

```{r quiz-final-12, echo=FALSE}
quiz(
  question(
    "The six tables in a metacore object are:",
    answer("ds_spec, ds_vars, var_spec, value_spec, code_list, derivations", correct = TRUE),
    answer("datasets, variables, codelists, types, labels, origins"),
    answer("define_xml, study_spec, ct_spec, vlm, derivations, outputs"),
    random_answer_order = TRUE
  ),
  question(
    "When you call metacore() and a code_id in value_spec doesn't exist in code_list, the constructor:",
    answer("Silently ignores the missing code_id"),
    answer("Throws an error immediately, catching the spec inconsistency at load time", correct = TRUE),
    answer("Returns a warning but continues"),
    random_answer_order = TRUE
  ),
  question(
    "A DatasetMeta is a metacore subclass that:",
    answer("Has fewer tables than a full metacore"),
    answer("Is restricted to a single dataset — required by metatools functions", correct = TRUE),
    answer("Can only be used with admiral"),
    random_answer_order = TRUE
  ),
  question(
    "The `var_spec` table is shared across all datasets in a metacore object because:",
    answer("Each dataset must define its own copy of every variable"),
    answer("Variable definitions (label, type, length) are properties of the variable itself, not of the dataset it appears in — sharing avoids inconsistencies", correct = TRUE),
    answer("metatools requires all variables to appear in the first dataset alphabetically"),
    random_answer_order = TRUE
  ),
  question(
    "`setdiff(spec_vars, names(my_dataset))` returns variables that are:",
    answer("In the dataset but undocumented in the spec"),
    answer("In the spec but not yet present in the derived dataset — a potential submission gap", correct = TRUE),
    answer("Variables that exist in both the spec and the dataset"),
    random_answer_order = TRUE
  ),
  question(
    "The `code_list` table uses a `code_id` foreign key (e.g., 'SEX_CL') rather than embedding codelist values in `var_spec` directly because:",
    answer("R can't store lists inside a tibble column"),
    answer("A single codelist (e.g., RACE) can be reused by multiple variables or datasets without duplication", correct = TRUE),
    answer("The code_id column must match the SAS format name exactly"),
    random_answer_order = TRUE
  ),
  question(
    "In a metadata-driven pipeline, the best time to construct the metacore object is:",
    answer("Inside each individual derivation script, right before it's needed"),
    answer("Once, in a shared constructor script, saved as .rds and loaded by all downstream scripts", correct = TRUE),
    answer("After all ADaM datasets have been derived"),
    random_answer_order = TRUE
  ),
  question(
    "The `order` column in `ds_vars` is important because:",
    answer("It controls the column order in the final delivered dataset, which must match the spec", correct = TRUE),
    answer("It determines which variables are checked first by Pinnacle 21"),
    answer("It sets the sort order of subjects within the dataset"),
    random_answer_order = TRUE
  ),
  question(
    "A flag variable like SAFFL (values 'Y' / 'N') does NOT need a codelist entry in `code_list` or `value_spec` because:",
    answer("Flag variables are always excluded from CDISC codelists"),
    answer("metatools auto-detects flag variables"),
    answer("The permissible values are defined by a controlled terminology term (NY codelist) referenced in the submission package, not a study-specific codelist", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "When you call `select_dataset(md_full, 'ADVS')` on a metacore that has no ADVS dataset:",
    answer("It returns an empty DatasetMeta with zero variables"),
    answer("It throws an error telling you ADVS is not in the metacore object", correct = TRUE),
    answer("It returns the full metacore unchanged"),
    random_answer_order = TRUE
  )
)
```

---

## What's next

Lesson 13 — **`{metatools}`** — uses the metacore objects we've built here to do real data work: construct base datasets from SDTM sources, derive numeric companion variables from codelists, and run conformance checks that verify your final dataset matches the spec.

The metadata-driven approach pays off fully once you see how `create_var_from_codelist()` uses the RACE codelist from your metacore object to derive RACEN — one line of code, spec-driven, no maintenance when the codelist changes.
