---
title: "Lesson 13 — {metatools}: Using the Spec to Build and Validate Datasets"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Use metacore objects to build datasets spec-first, derive codelist-driven
  companion variables, apply labels, and run conformance checks.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 13_metatools.Rmd
# PURPOSE       : metatools package for metadata-driven dataset derivation: check_ct(),
#                 create_iso8601(), drop_unspec_vars(), metatools-guided
#                 ADaM variable creation, and spec-conformance
#                 validation.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── Re-create metacore infrastructure (same as Lesson 12) ────────────────────
metacore <- function(ds_spec, ds_vars, var_spec,
                     value_spec  = tibble(variable = character(), dataset = character(),
                                          where = character(), code_id = character()),
                     code_list   = tibble(code_id = character(), name = character(),
                                          code = character(), decode = character()),
                     derivations = tibble(derivation_id = character(),
                                          derivation = character())) {
  missing_datasets <- setdiff(unique(ds_vars$dataset), ds_spec$dataset)
  if (length(missing_datasets) > 0)
    stop("ds_vars references datasets not in ds_spec: ", paste(missing_datasets, collapse = ", "))
  missing_vars <- setdiff(unique(ds_vars$variable), var_spec$variable)
  if (length(missing_vars) > 0)
    stop("ds_vars references variables not in var_spec: ", paste(missing_vars, collapse = ", "))
  obj <- list(ds_spec = ds_spec, ds_vars = ds_vars, var_spec = var_spec,
              value_spec = value_spec, code_list = code_list, derivations = derivations)
  class(obj) <- c("Metacore", "list"); obj
}

print.Metacore <- function(x, ...) {
  cat("── Metacore Object ─────────────────────────────────────────\n")
  cat(sprintf("  Datasets: %s\n", paste(x$ds_spec$dataset, collapse = ", ")))
  cat(sprintf("  Variables: %d  |  Codelists: %d entries  |  Derivations: %d\n",
              nrow(x$var_spec), nrow(x$code_list), nrow(x$derivations)))
  invisible(x)
}

select_dataset <- function(metacore_obj, dataset) {
  if (!dataset %in% metacore_obj$ds_spec$dataset)
    stop("Dataset '", dataset, "' not found")
  vars_in_ds  <- metacore_obj$ds_vars |> filter(dataset == !!dataset) |> pull(variable)
  value_spec2 <- metacore_obj$value_spec |> filter(dataset == !!dataset | is.na(dataset))
  code_ids    <- unique(value_spec2$code_id[!is.na(value_spec2$code_id)])
  obj <- list(
    dataset = dataset,
    ds_spec     = metacore_obj$ds_spec  |> filter(dataset == !!dataset),
    ds_vars     = metacore_obj$ds_vars  |> filter(dataset == !!dataset),
    var_spec    = metacore_obj$var_spec |> filter(variable %in% vars_in_ds),
    value_spec  = value_spec2,
    code_list   = metacore_obj$code_list |> filter(code_id %in% code_ids),
    derivations = metacore_obj$derivations
  )
  class(obj) <- c("DatasetMeta", "Metacore", "list"); obj
}

# ── Mock metatools functions ──────────────────────────────────────────────────

# combine_supp: fold SUPP-- (long) back into parent domain (wide)
combine_supp <- function(parent_dataset, supp_dataset) {
  if (nrow(supp_dataset) == 0) return(parent_dataset)
  supp_wide <- supp_dataset |>
    select(USUBJID, QNAM, QVAL) |>
    pivot_wider(names_from = QNAM, values_from = QVAL)
  left_join(parent_dataset, supp_wide, by = "USUBJID")
}

# build_from_derived: construct base dataset from source datasets per spec
build_from_derived <- function(metacore_obj, ds_list, predecessor_only = FALSE, keep = FALSE) {
  spec_vars  <- metacore_obj$ds_vars |> arrange(order) |> pull(variable)
  var_labels <- metacore_obj$var_spec |> select(variable, label, type)

  # Collect all columns available across source datasets
  all_cols <- lapply(names(ds_list), function(nm) {
    ds <- ds_list[[nm]]
    tibble(variable = names(ds), src = nm)
  }) |> bind_rows() |> distinct(variable, .keep_all = TRUE)

  vars_available <- intersect(spec_vars, all_cols$variable)
  vars_missing   <- setdiff(spec_vars, all_cols$variable)

  # Start from the first source dataset
  base_ds <- ds_list[[1]]

  # Add columns from other source datasets that aren't yet present
  for (nm in names(ds_list)[-1]) {
    src_ds <- ds_list[[nm]]
    new_cols <- setdiff(names(src_ds), names(base_ds))
    key_cols <- intersect(c("USUBJID", "STUDYID"), names(base_ds))
    if (length(new_cols) > 0 && length(key_cols) > 0) {
      base_ds <- left_join(base_ds,
                           src_ds |> select(any_of(c(key_cols, new_cols))),
                           by = key_cols)
    }
  }

  # Add NA columns for spec variables that don't exist in any source
  for (v in vars_missing) {
    base_ds[[v]] <- NA_character_
  }

  # Re-order per spec
  base_ds <- base_ds |> select(any_of(spec_vars), everything())
  base_ds
}

# create_var_from_codelist: derive numeric companion or decoded label from a codelist
create_var_from_codelist <- function(dataset, metacore_obj, input_var, out_var,
                                     decode_to_code = FALSE) {
  input_nm <- deparse(substitute(input_var))
  out_nm   <- deparse(substitute(out_var))

  # Find code_id for input_var
  code_id_val <- metacore_obj$value_spec |>
    filter(variable == input_nm) |>
    pull(code_id) |>
    first()

  if (is.na(code_id_val) || length(code_id_val) == 0)
    stop("No codelist found for variable: ", input_nm)

  cl <- metacore_obj$code_list |>
    filter(code_id == code_id_val) |>
    select(code, decode) |>
    distinct()

  if (!decode_to_code) {
    # code → numeric (when input is the decode, map to numeric code)
    lookup <- cl |>
      mutate(numeric_code = suppressWarnings(as.numeric(code))) |>
      select(decode, numeric_code)
    dataset <- dataset |>
      select(-any_of(out_nm)) |>          # drop NA placeholder if build_from_derived() added it
      left_join(lookup, by = setNames("decode", input_nm)) |>
      rename(!!out_nm := numeric_code)
  } else {
    # code → decode (expand to long label)
    lookup <- cl |> select(code, decode)
    dataset <- dataset |>
      select(-any_of(out_nm)) |>          # drop NA placeholder if build_from_derived() added it
      left_join(lookup, by = setNames("code", input_nm)) |>
      rename(!!out_nm := decode)
  }
  dataset
}

# apply_variable_labels: set label attribute on each column per spec
apply_variable_labels <- function(dataset, metacore_obj) {
  for (v in names(dataset)) {
    lbl <- metacore_obj$var_spec |> filter(variable == v) |> pull(label)
    if (length(lbl) == 1 && !is.na(lbl)) attr(dataset[[v]], "label") <- lbl
  }
  dataset
}

# check_variables: verify dataset has exactly the variables in the spec
check_variables <- function(dataset, metacore_obj) {
  spec_vars <- metacore_obj$ds_vars |> pull(variable)
  data_vars <- names(dataset)
  missing_vars <- setdiff(spec_vars, data_vars)
  extra_vars   <- setdiff(data_vars, spec_vars)
  if (length(missing_vars) > 0)
    warning("Variables in spec but missing from dataset: ", paste(missing_vars, collapse = ", "))
  if (length(extra_vars) > 0)
    warning("Variables in dataset but not in spec: ", paste(extra_vars, collapse = ", "))
  if (length(missing_vars) == 0 && length(extra_vars) == 0)
    message("check_variables: PASSED — dataset variables match spec exactly")
  invisible(list(missing = missing_vars, extra = extra_vars))
}

# check_ct_data: verify coded variables are in their codelist
check_ct_data <- function(dataset, metacore_obj) {
  issues <- list()
  for (i in seq_len(nrow(metacore_obj$value_spec))) {
    vs_row  <- metacore_obj$value_spec[i, ]
    vname   <- vs_row$variable
    code_id <- vs_row$code_id
    if (!vname %in% names(dataset)) next
    valid_vals <- metacore_obj$code_list |>
      filter(code_id == !!code_id) |>
      pull(decode)
    invalid <- dataset[[vname]][!is.na(dataset[[vname]]) & !dataset[[vname]] %in% valid_vals]
    if (length(invalid) > 0)
      issues[[vname]] <- unique(invalid)
  }
  if (length(issues) == 0)
    message("check_ct_data: PASSED — all coded variables within codelists")
  else
    warning("check_ct_data: FAILED — invalid values found: ",
            paste(sapply(names(issues), function(n) paste0(n, "=", issues[[n]])), collapse = "; "))
  invisible(issues)
}

# check_unique_keys: verify the dataset's key columns identify rows uniquely
check_unique_keys <- function(dataset, metacore_obj, key_vars = c("STUDYID", "USUBJID")) {
  key_cols <- intersect(key_vars, names(dataset))
  dups <- dataset |>
    group_by(across(all_of(key_cols))) |>
    filter(n() > 1) |>
    ungroup()
  if (nrow(dups) == 0)
    message("check_unique_keys: PASSED — ", paste(key_cols, collapse = "+"),
            " uniquely identifies rows")
  else
    warning("check_unique_keys: FAILED — ", nrow(dups), " duplicate rows on ",
            paste(key_cols, collapse = "+"))
  invisible(dups)
}

# ── Data ──────────────────────────────────────────────────────────────────────

# Spec: ADSL metacore object
adsl_code_list <- tribble(
  ~code_id,   ~name,    ~code, ~decode,
  "SEX_CL",   "SEX",    "M",   "Male",
  "SEX_CL",   "SEX",    "F",   "Female",
  "SEX_CL",   "SEX",    "U",   "Unknown",
  "RACE_CL",  "RACE",   "1",   "AMERICAN INDIAN OR ALASKA NATIVE",
  "RACE_CL",  "RACE",   "2",   "ASIAN",
  "RACE_CL",  "RACE",   "3",   "BLACK OR AFRICAN AMERICAN",
  "RACE_CL",  "RACE",   "4",   "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER",
  "RACE_CL",  "RACE",   "5",   "WHITE",
  "RACE_CL",  "RACE",   "6",   "MULTIPLE",
  "RACE_CL",  "RACE",   "8",   "UNKNOWN",
  "RACE_CL",  "RACE",   "9",   "NOT REPORTED"
)

adsl_var_spec <- tribble(
  ~variable, ~label,                           ~type,        ~length, ~format,
  "STUDYID", "Study Identifier",               "character",  8,       NA,
  "USUBJID", "Unique Subject Identifier",      "character",  32,      NA,
  "AGE",     "Age (years)",                    "numeric",    8,       NA,
  "SEX",     "Sex",                            "character",  1,       NA,
  "SEXN",    "Sex (N)",                        "numeric",    8,       NA,
  "RACE",    "Race",                           "character",  60,      NA,
  "RACEN",   "Race (N)",                       "numeric",    8,       NA,
  "TRT01A",  "Actual Treatment for Period 01", "character",  40,      NA,
  "TRTSDT",  "Date of First Exposure",         "numeric",    8,       "DATE9.",
  "TRTEDT",  "Date of Last Exposure",          "numeric",    8,       "DATE9.",
  "TRTDURD", "Total Treatment Duration (Days)","numeric",    8,       NA,
  "SAFFL",   "Safety Population Flag",         "character",  1,       NA,
  "ITTFL",   "Intent-To-Treat Population Flag","character",  1,       NA
)

adsl_ds_spec <- tibble(
  dataset   = "ADSL",
  structure  = "One record per subject",
  label      = "Subject-Level Analysis Dataset"
)

adsl_ds_vars <- tibble(
  dataset   = "ADSL",
  variable  = adsl_var_spec$variable,
  order     = seq_along(adsl_var_spec$variable),
  mandatory = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, FALSE, TRUE,
                FALSE, FALSE, FALSE, FALSE, FALSE),
  core      = c("Required", "Required", "Required", "Required", "Permissible",
                "Required", "Permissible", "Required", "Permissible",
                "Permissible", "Permissible", "Permissible", "Permissible")
)

adsl_value_spec <- tribble(
  ~variable, ~dataset, ~where,          ~code_id,
  "SEX",     "ADSL",   NA_character_,   "SEX_CL",
  "RACE",    "ADSL",   NA_character_,   "RACE_CL"
)

adsl_derivations <- tribble(
  ~derivation_id, ~derivation,
  "DR_TRTSDT",  "First non-zero dose date from EX",
  "DR_TRTDURD", "TRTEDT - TRTSDT + 1",
  "DR_SAFFL",   "Y if any EX record with EXDOSE > 0 or placebo",
  "DR_RACEN",   "Numeric code from RACE_CL codelist"
)

md <- metacore(adsl_ds_spec, adsl_ds_vars, adsl_var_spec,
               adsl_value_spec, adsl_code_list, adsl_derivations)
md_adsl <- select_dataset(md, "ADSL")

# Source DM data
dm <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101"),
  AGE     = c(63L, 64L, 71L, 56L),
  SEX     = c("M", "F", "M", "F"),
  RACE    = c("WHITE", "WHITE", "ASIAN", "WHITE"),
  TRT01A  = c("XANOMELINE HIGH DOSE", "XANOMELINE LOW DOSE", "PLACEBO", "XANOMELINE HIGH DOSE")
)

# SUPPDM (subject-level supplemental qualifiers — single QNAM)
suppdm <- tibble(
  STUDYID  = "CDISCPILOT01",
  RDOMAIN  = "DM",
  USUBJID  = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101"),
  IDVAR    = "",
  IDVARVAL = "",
  QNAM     = "STRATGRP",
  QLABEL   = "Stratification Group",
  QVAL     = c("A", "B", "A", "B"),
  QORIG    = "CRF",
  QEVAL    = "INVESTIGATOR"
)

# SUPPDM with two supplemental variables (STRATGRP + ETHNIC2)
suppdm_multi <- tibble(
  STUDYID  = "CDISCPILOT01",
  RDOMAIN  = "DM",
  USUBJID  = rep(c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101"), 2),
  IDVAR    = "",
  IDVARVAL = "",
  QNAM     = c(rep("STRATGRP", 4), rep("ETHNIC2", 4)),
  QLABEL   = c(rep("Stratification Group", 4), rep("Ethnicity Detail", 4)),
  QVAL     = c("A", "B", "A", "B",
               "NOT HISPANIC OR LATINO", "NOT HISPANIC OR LATINO",
               "NOT HISPANIC OR LATINO", "HISPANIC OR LATINO"),
  QORIG    = "CRF",
  QEVAL    = "INVESTIGATOR"
)

# ── Real pharmaverseadam data ─────────────────────────────────────────────────
library(pharmaverseadam)
adsl_real <- pharmaverseadam::adsl
```

## Welcome

This is Lesson 13 — **`{metatools}`**: Using the Spec to Build and Validate Datasets.

Where `{metacore}` stores the spec, `{metatools}` *acts on it*. Every metatools function takes a `DatasetMeta` and uses the metadata to do something useful with your data.

This lesson covers the five things metatools does:

1. `combine_supp()` — fold SUPP-- back into the parent domain
2. `build_from_derived()` — spec-driven base dataset construction
3. `create_var_from_codelist()` — derive companion variables from codelists
4. `apply_variable_labels()` — apply spec labels to dataset columns
5. `check_variables()`, `check_ct_data()`, `check_unique_keys()` — spec conformance checks

All functions are mocked with identical behavior to the real package.

---

## 1 — Five things metatools does

```{r quiz-five-things, echo=FALSE}
quiz(
  question(
    "Which metatools function would you call FIRST when building an ADSL that includes SUPPDM variables?",
    answer("build_from_derived()"),
    answer("combine_supp() — to fold SUPPDM into DM so the combined columns are available for the base build", correct = TRUE),
    answer("check_variables()"),
    random_answer_order = TRUE
  ),
  question(
    "create_var_from_codelist() is the metadata-driven alternative to:",
    answer("A hard-coded case_when() that maps character values to numeric codes", correct = TRUE),
    answer("derive_vars_dt() from admiral"),
    answer("pivot_wider() from tidyr"),
    random_answer_order = TRUE
  )
)
```

---

## 2 — `combine_supp()`: fold SUPP-- into the parent domain

SUPPDM is long-format (one row per subject per QNAM). For ADSL, you want the SUPPDM variables as proper columns. `combine_supp()` pivots SUPPDM wide and joins it to DM.

```{r combine-supp-demo}
# Before: DM and SUPPDM are separate
cat("DM columns:", paste(names(dm), collapse = ", "), "\n\n")
cat("SUPPDM sample:\n")
print(suppdm |> select(USUBJID, QNAM, QVAL))
```

```{r combine-supp-result}
# After: SUPPDM's STRATGRP column is now in dm_combined
dm_combined <- combine_supp(dm, suppdm)

cat("dm_combined columns:", paste(names(dm_combined), collapse = ", "), "\n\n")
dm_combined
```

### Exercise: Check that combine_supp adds the correct column and values

```{r combine-supp-check, exercise=TRUE, exercise.lines=10}
dm_combined2 <- combine_supp(dm, suppdm)

# Verify the STRATGRP column is present and has the right values
cat("STRATGRP present:", "STRATGRP" %in% names(dm_combined2), "\n")
cat("STRATGRP values:", paste(dm_combined2$STRATGRP, collapse = ", "), "\n")

# What happens with a subject missing from SUPPDM? (none missing here, but check for NA)
cat("NAs in STRATGRP:", sum(is.na(dm_combined2$STRATGRP)), "\n")
```

### Exercise: Combine a SUPPDM with two supplemental variables

`suppdm_multi` contains two QNAMs: `STRATGRP` and `ETHNIC2`. `combine_supp()` pivots all QNAMs wide in one call — no loop needed.

```{r combine-supp-multi, exercise=TRUE, exercise.lines=14}
# Combine both SUPPDM variables into DM in one call
dm_multi <- combine_supp(dm, ___)

# Confirm both new columns exist
cat("Columns added to DM:\n")
cat(paste(setdiff(names(dm_multi), names(dm)), collapse = ", "), "\n\n")

# Show the result
dm_multi |> select(USUBJID, STRATGRP, ETHNIC2)
```

```{r combine-supp-multi-hint-1}
# combine_supp(dm, suppdm_multi)
```

```{r combine-supp-multi-solution}
dm_multi <- combine_supp(dm, suppdm_multi)

cat("Columns added to DM:\n")
cat(paste(setdiff(names(dm_multi), names(dm)), collapse = ", "), "\n\n")

dm_multi |> select(USUBJID, STRATGRP, ETHNIC2)
```

> **Real-world note**: A DM domain can have many SUPP-- variables collected across dozens of CRF pages. `combine_supp()` handles all QNAMs in one call regardless of how many there are, making it far more maintainable than individual `pivot_wider()` + `left_join()` chains.

---

## 3 — `build_from_derived()`: spec-driven base dataset

`build_from_derived()` constructs a starting dataset with columns in the spec's order, populated where source datasets provide values, and `NA` where derivations are still needed.

```{r build-from-derived-demo}
# Build base ADSL from DM (with SUPPDM already combined)
adsl_base <- build_from_derived(
  metacore_obj = md_adsl,
  ds_list      = list(dm = dm_combined),
  predecessor_only = FALSE,
  keep = FALSE
)

glimpse(adsl_base)
```

```{r build-from-derived-structure}
# The key feature: columns are in SPEC ORDER, not source order
names(adsl_base)

# Derived columns exist as NA — waiting to be filled
cat("\nTRTSDT (not yet derived):", sum(is.na(adsl_base$TRTSDT)), "NAs out of", nrow(adsl_base), "\n")
```

### Exercise: Confirm column order matches the spec

The spec defines `order` for each variable. Check that the first 6 columns of `adsl_base` match the spec order.

```{r build-order-check, exercise=TRUE, exercise.lines=12}
# What order does the spec define?
spec_order <- md_adsl$ds_vars |>
  arrange(order) |>
  pull(variable)

cat("Spec order (first 6):", paste(head(spec_order, 6), collapse = " → "), "\n")
cat("Data order (first 6):", paste(head(names(adsl_base), 6), collapse = " → "), "\n")
cat("Match:", identical(head(spec_order, 6), head(names(adsl_base), 6)), "\n")
```

### Exercise: Compare `adsl_base` output to the real pharmaverseadam ADSL

`adsl_real` is the fully derived `pharmaverseadam::adsl`. The `adsl_base` we just built is the spec-ordered skeleton before admiral derivations. Use `setdiff()` to see which variables the real ADSL has that our base is still missing, and which spec variables the real dataset doesn't carry.

```{r build-real-gap, exercise=TRUE, exercise.lines=16}
# Variables in real adsl not in our spec-skeleton (fully derived but not in our mini spec)
in_real_not_base <- setdiff(names(adsl_real), names(adsl_base))
cat("In real ADSL, not in our base (", length(in_real_not_base), "vars ):\n", sep = "")
cat(paste(head(in_real_not_base, 12), collapse = ", "), "...\n\n")

# Variables spec planned that real ADSL already has
shared <- intersect(names(adsl_base), names(adsl_real))
cat("Variables present in both base and real ADSL:", length(shared), "\n")
cat(paste(shared, collapse = ", "), "\n\n")

# Variables in our spec still NA in adsl_base (not yet derived)
na_cols <- names(adsl_base)[sapply(adsl_base, function(x) all(is.na(x)))]
cat("Spec columns still all-NA (need admiral derivation):", paste(na_cols, collapse = ", "), "\n")
```

```{r build-real-gap-hint-1}
# setdiff(names(adsl_real), names(adsl_base))  — in real, not in base
# intersect(names(adsl_base), names(adsl_real)) — in both
# sapply(adsl_base, function(x) all(is.na(x)))  — all-NA columns
```

```{r build-real-gap-solution}
in_real_not_base <- setdiff(names(adsl_real), names(adsl_base))
cat("In real ADSL, not in our base (", length(in_real_not_base), "vars):\n", sep = "")
cat(paste(head(in_real_not_base, 12), collapse = ", "), "...\n\n")

shared <- intersect(names(adsl_base), names(adsl_real))
cat("Variables present in both base and real ADSL:", length(shared), "\n")
cat(paste(shared, collapse = ", "), "\n\n")

na_cols <- names(adsl_base)[sapply(adsl_base, function(x) all(is.na(x)))]
cat("Spec columns still all-NA (need admiral derivation):", paste(na_cols, collapse = ", "), "\n")
```

---

## 4 — `create_var_from_codelist()`: spec-driven companion variables

CDISC ADaM convention: categorical character variables have numeric companions (SEX→SEXN, RACE→RACEN). The values come from the codelist. `create_var_from_codelist()` reads the codelist from the spec and derives the companion automatically.

```{r create-racen-demo}
# Derive RACEN from RACE using the RACE_CL codelist
adsl_with_racen <- adsl_base |>
  create_var_from_codelist(md_adsl, input_var = RACE, out_var = RACEN)

adsl_with_racen |>
  select(USUBJID, RACE, RACEN)
```

```{r create-sexn-demo}
# Derive SEXN from SEX using the SEX_CL codelist
adsl_with_sexn <- adsl_with_racen |>
  create_var_from_codelist(md_adsl, input_var = SEX, out_var = SEXN)

adsl_with_sexn |>
  select(USUBJID, SEX, SEXN, RACE, RACEN)
```

### Exercise: Chain both companion derivations

Starting from `adsl_base`, derive SEXN and RACEN in a single pipeline.

```{r derive-companions, exercise=TRUE, exercise.lines=14}
adsl_with_companions <- adsl_base |>
  create_var_from_codelist(md_adsl, input_var = ___, out_var = ___) |>
  create_var_from_codelist(md_adsl, input_var = ___, out_var = ___)

adsl_with_companions |>
  select(USUBJID, SEX, SEXN, RACE, RACEN)
```

```{r derive-companions-hint-1}
# create_var_from_codelist(md_adsl, input_var = SEX, out_var = SEXN)
# create_var_from_codelist(md_adsl, input_var = RACE, out_var = RACEN)
```

```{r derive-companions-solution}
adsl_with_companions <- adsl_base |>
  create_var_from_codelist(md_adsl, input_var = SEX,  out_var = SEXN) |>
  create_var_from_codelist(md_adsl, input_var = RACE, out_var = RACEN)

adsl_with_companions |>
  select(USUBJID, SEX, SEXN, RACE, RACEN)
```

### Exercise: What is the manual (non-metatools) equivalent of create_var_from_codelist for RACEN?

Reproduce the RACEN derivation using `case_when()` directly, without using the spec.

```{r manual-racen, exercise=TRUE, exercise.lines=20}
# Manual approach (what you'd write without metacore/metatools)
adsl_manual_racen <- adsl_base |>
  mutate(RACEN = case_when(
    RACE == "AMERICAN INDIAN OR ALASKA NATIVE"          ~ 1,
    RACE == "ASIAN"                                     ~ 2,
    RACE == "BLACK OR AFRICAN AMERICAN"                 ~ 3,
    RACE == "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER" ~ 4,
    RACE == "WHITE"                                     ~ 5,
    RACE == "MULTIPLE"                                  ~ 6,
    RACE == "UNKNOWN"                                   ~ 8,
    RACE == "NOT REPORTED"                              ~ 9,
    TRUE                                                ~ NA_real_
  ))

# Compare to the spec-driven result
adsl_spec_racen <- adsl_base |>
  create_var_from_codelist(md_adsl, input_var = RACE, out_var = RACEN)

cat("Manual RACEN:    ", adsl_manual_racen$RACEN, "\n")
cat("Spec-driven RACEN:", adsl_spec_racen$RACEN, "\n")
cat("Identical:", identical(adsl_manual_racen$RACEN, adsl_spec_racen$RACEN), "\n")
```

### Exercise: Apply `create_var_from_codelist` to real pharmaverseadam data

`adsl_real` has 254 subjects and a real `RACE` column. Derive `RACEN` from it using the spec codelist, then summarise the race distribution.

```{r create-var-real, exercise=TRUE, exercise.lines=20}
# Take just the columns we need from the real adsl
adsl_real_mini <- adsl_real |>
  select(USUBJID, RACE)

# Derive RACEN using the codelist in md_adsl
adsl_real_with_racen <- adsl_real_mini |>
  create_var_from_codelist(md_adsl, input_var = ___, out_var = ___)

# Summarise RACE distribution with derived numeric code
adsl_real_with_racen |>
  count(RACE, RACEN, sort = TRUE)
```

```{r create-var-real-hint-1}
# create_var_from_codelist(md_adsl, input_var = RACE, out_var = RACEN)
```

```{r create-var-real-solution}
adsl_real_mini <- adsl_real |>
  select(USUBJID, RACE)

adsl_real_with_racen <- adsl_real_mini |>
  create_var_from_codelist(md_adsl, input_var = RACE, out_var = RACEN)

adsl_real_with_racen |>
  count(RACE, RACEN, sort = TRUE)
```

> **What to observe**: Any RACE values in the real data that don't match a codelist decode will produce `NA` for RACEN. A `NA` count here would indicate either a data entry variant or a gap in the codelist — exactly the kind of issue `check_ct_data()` is designed to surface.

---

## 5 — `apply_variable_labels()`

Variable labels are metadata attributes set on each column. They're not shown by default but are used by export tools (`{xportr}`) and reporting packages (`gtsummary`, `tern`).

```{r apply-labels-demo}
# Before applying labels
cat("Label on AGE (before):", attr(adsl_base$AGE, "label"), "\n")

# Apply labels from the spec
adsl_labeled <- adsl_base |>
  apply_variable_labels(md_adsl)

# After applying labels
cat("Label on AGE (after): ", attr(adsl_labeled$AGE, "label"), "\n")
cat("Label on SEX (after): ", attr(adsl_labeled$SEX, "label"), "\n")
cat("Label on TRT01A (after):", attr(adsl_labeled$TRT01A, "label"), "\n")
```

### Exercise: Verify labels on multiple columns

```{r verify-labels, exercise=TRUE, exercise.lines=14}
adsl_labeled2 <- adsl_base |> apply_variable_labels(md_adsl)

# Extract labels for the first 5 columns
col_labels <- sapply(names(adsl_labeled2)[1:5], function(v) {
  lbl <- attr(adsl_labeled2[[v]], "label")
  if (is.null(lbl)) "(none)" else lbl
})

tibble(
  variable = names(col_labels),
  label    = col_labels
)
```

---

## 6 — Conformance checks

Three checks run at the end of a derivation to verify the dataset matches its spec.

### `check_variables()`: Are the right variables present?

```{r check-variables-demo}
# A dataset that matches the spec exactly — should pass
adsl_complete <- adsl_base  # all spec columns present (with NAs for derived ones)
result <- check_variables(adsl_complete, md_adsl)
```

```{r check-variables-missing}
# A dataset missing a required variable — should warn
adsl_missing_col <- adsl_base |> select(-TRTDURD)
result_bad <- check_variables(adsl_missing_col, md_adsl)
cat("Missing:", paste(result_bad$missing, collapse = ", "), "\n")
```

### Exercise: `check_variables()` on a dataset with an extra column

```{r check-variables-extra, exercise=TRUE, exercise.lines=10}
# Add an extra column not in the spec
adsl_extra <- adsl_base |> mutate(EXTRA_COL = "unexpected")

result_extra <- check_variables(adsl_extra, md_adsl)
cat("Missing variables:", paste(result_extra$missing, collapse = ", "), "\n")
cat("Extra variables:", paste(result_extra$extra, collapse = ", "), "\n")
```

---

### `check_ct_data()`: Are coded variable values in their codelists?

```{r check-ct-demo}
# Clean dataset — all SEX values are M/F which map to Male/Female via SEX_CL
# (The codelist decode column stores "Male"/"Female"; SEX stores "M"/"F")
# Since our data uses "M"/"F" and the lookup is code→decode, let's show the check

# Introduce a bad value
adsl_bad_sex <- adsl_base |>
  mutate(SEX = ifelse(USUBJID == "01-702-1042", "MALE", SEX))  # "MALE" not in codelist decode

result_ct <- check_ct_data(adsl_bad_sex, md_adsl)
```

```{r check-ct-pass}
# Clean: standard values
check_ct_data(adsl_base, md_adsl)
```

### Exercise: `check_ct_data()` with an invalid RACE value

```{r check-ct-race, exercise=TRUE, exercise.lines=10}
# Introduce an invalid RACE value
adsl_bad_race <- adsl_base |>
  mutate(RACE = ifelse(row_number() == 3, "OTHER", RACE))

check_ct_data(adsl_bad_race, md_adsl)
```

---

### `check_unique_keys()`: Is the key unique?

```{r check-keys-demo}
# Clean — each USUBJID appears once (ADSL is one row per subject)
check_unique_keys(adsl_base, md_adsl)
```

```{r check-keys-dup}
# Introduce a duplicate
adsl_dup <- bind_rows(adsl_base, adsl_base[1, ])
result_dup <- check_unique_keys(adsl_dup, md_adsl)
cat("Duplicate rows detected:", nrow(result_dup), "\n")
```

### Exercise: Run all three checks in sequence as a final QC block

```{r qc-block, exercise=TRUE, exercise.lines=12}
# Run all three checks on a clean dataset
# Expected: all three pass
run_final_qc <- function(dataset, metacore_obj) {
  message("=== Final QC Block ===")
  check_variables(dataset, metacore_obj)
  check_ct_data(dataset, metacore_obj)
  check_unique_keys(dataset, metacore_obj)
  message("=== QC Complete ===")
}

run_final_qc(adsl_base, md_adsl)
```

---

## 7 — The full metadata-driven ADSL pattern

Putting it all together — the canonical metatools workflow before admiral derivations:

```{r full-pattern-demo}
# Step 1: Combine SUPP-- into parent domain
dm_with_supp <- combine_supp(dm, suppdm)

# Step 2: Build the spec-ordered base dataset
adsl_step2 <- build_from_derived(md_adsl, ds_list = list(dm = dm_with_supp))

# Step 3: Derive codelist-driven companion variables
adsl_step3 <- adsl_step2 |>
  create_var_from_codelist(md_adsl, input_var = SEX,  out_var = SEXN) |>
  create_var_from_codelist(md_adsl, input_var = RACE, out_var = RACEN)

# Step 4: Apply variable labels
adsl_step4 <- adsl_step3 |>
  apply_variable_labels(md_adsl)

# Step 5: Run conformance checks
check_variables(adsl_step4, md_adsl)
check_unique_keys(adsl_step4, md_adsl)

# Show the result
glimpse(adsl_step4)
```

### Exercise: Add a stratification group derivation to the pipeline

After `combine_supp()`, the `dm_with_supp` dataset has a `STRATGRP` column from SUPPDM. Continue the pipeline to include this step, then check that `STRATGRP` is present in the final dataset.

```{r full-pattern-strat, exercise=TRUE, exercise.lines=20}
# Step 1: Combine
dm_with_supp2 <- combine_supp(dm, suppdm)

# Step 2: Build base
adsl_ex <- build_from_derived(md_adsl, ds_list = list(dm = dm_with_supp2))

# Step 3: Derive companions
adsl_ex <- adsl_ex |>
  create_var_from_codelist(md_adsl, input_var = SEX,  out_var = SEXN) |>
  create_var_from_codelist(md_adsl, input_var = RACE, out_var = RACEN)

# Step 4: Does dm_with_supp2 have STRATGRP?
cat("STRATGRP in dm_with_supp2:", "STRATGRP" %in% names(dm_with_supp2), "\n")

# Inspect the pipeline output
adsl_ex |> select(USUBJID, SEX, SEXN, RACE, RACEN) |> print()
```

---

## 8 — When to use (and when to skip) metacore/metatools

```{r quiz-when-to-use, echo=FALSE}
quiz(
  question(
    "For a phase-I exploratory analysis that won't be submitted to FDA, metacore/metatools is:",
    answer("Always required"),
    answer("Optional — the metadata discipline may be more overhead than value for exploratory work", correct = TRUE),
    answer("Forbidden for exploratory work"),
    random_answer_order = TRUE
  ),
  question(
    "For a submission-quality ADSL going to FDA, metatools is valuable because:",
    answer("FDA mandates the use of metatools specifically"),
    answer("It enables spec-driven derivation, conformance checking, and Define-XML serialization — all required for submissions", correct = TRUE),
    answer("It makes the dataset smaller"),
    random_answer_order = TRUE
  ),
  question(
    "metatools functions require a DatasetMeta (not a full metacore) because:",
    answer("Full metacore objects are too slow"),
    answer("It prevents ambiguity — scoping to one dataset ensures metadata doesn't accidentally bleed across ADaMs", correct = TRUE),
    answer("Full metacore objects can't have value_spec entries"),
    random_answer_order = TRUE
  )
)
```

---

## 9 — Key takeaways

```{r quiz-final-13, echo=FALSE}
quiz(
  question(
    "The correct order of the standard metatools ADSL workflow is:",
    answer("build_from_derived → combine_supp → create_var_from_codelist → apply_variable_labels → check_*"),
    answer("combine_supp → build_from_derived → create_var_from_codelist → apply_variable_labels → check_*", correct = TRUE),
    answer("check_* → build_from_derived → combine_supp → create_var_from_codelist → apply_variable_labels"),
    random_answer_order = TRUE
  ),
  question(
    "create_var_from_codelist() is more maintainable than a case_when() because:",
    answer("It runs faster"),
    answer("When the codelist changes in the spec, the derivation automatically reflects the change without touching the R script", correct = TRUE),
    answer("case_when() doesn't work with character variables"),
    random_answer_order = TRUE
  ),
  question(
    "Which function checks that all values of a coded variable (e.g., SEX) are in its codelist?",
    answer("check_variables()"),
    answer("check_ct_data()", correct = TRUE),
    answer("check_unique_keys()"),
    random_answer_order = TRUE
  ),
  question(
    "`combine_supp()` with a SUPPDM that has three different QNAMs will produce:",
    answer("Only the first QNAM as a new column — you need one call per QNAM"),
    answer("Three new columns joined to the parent domain — one per QNAM", correct = TRUE),
    answer("A warning that only a single QNAM is supported"),
    random_answer_order = TRUE
  ),
  question(
    "A subject present in DM but missing from SUPPDM will have what value in the SUPPDM-derived column after `combine_supp()`?",
    answer("The subject's row is dropped from the result"),
    answer("An error is thrown"),
    answer("NA — because combine_supp uses a left join, keeping all parent domain rows", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "`build_from_derived()` produces `NA` for a spec variable when:",
    answer("The variable is mandatory"),
    answer("The variable is not present in any of the source datasets passed in `ds_list`", correct = TRUE),
    answer("The variable type is 'numeric'"),
    random_answer_order = TRUE
  ),
  question(
    "`apply_variable_labels()` stores labels as:",
    answer("A separate tibble column called 'label'"),
    answer("Attribute metadata on each column vector — readable by export tools like xportr", correct = TRUE),
    answer("A PROC CONTENTS-style report"),
    random_answer_order = TRUE
  ),
  question(
    "`check_variables()` returns a warning when a variable is in the dataset but NOT in the spec. This usually means:",
    answer("The variable is a required variable that was accidentally deleted"),
    answer("A temporary intermediate variable was not dropped, or a variable was added that isn't documented in the spec", correct = TRUE),
    answer("The dataset has too many rows"),
    random_answer_order = TRUE
  ),
  question(
    "You run `create_var_from_codelist()` and some output values are `NA`. The most likely cause is:",
    answer("The function has a bug — it always produces some NAs"),
    answer("The input variable has values that don't match any decode in the codelist — a data quality issue to investigate", correct = TRUE),
    answer("NA output means the codelist derivation succeeded"),
    random_answer_order = TRUE
  ),
  question(
    "metatools functions take a `DatasetMeta` rather than a full `metacore` object because:",
    answer("DatasetMeta objects load faster from disk"),
    answer("Scoping to one dataset avoids accidental use of another dataset's metadata (e.g., ADAE codelists when building ADSL)", correct = TRUE),
    answer("Full metacore objects don't have a value_spec table"),
    random_answer_order = TRUE
  )
)
```

---

## What's next

Module 3 is complete. The metacore/metatools patterns you've learned here thread through the rest of pharmaverse:

- **Lesson 14 (admiral Part 1)**: admiral's date helpers and `derive_vars_merged()` — the foundation of every ADaM
- **Lesson 15 (admiral Part 2)**: a full ADSL end-to-end, combining metatools (base dataset + codelist companions) with admiral (treatment dates, population flags, death cause)

The metacore object you built in these two lessons is the same one `{xportr}` (Module 9) uses to apply spec labels and lengths when exporting the final XPT file for submission.
