---
title: "Lesson 14 — {admiral} Part 1: Foundations and Philosophy"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.1
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Admiral's design manifesto, function naming convention, exprs(), date/time
  helpers, and derive_vars_merged() — the foundation of every ADaM build.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 14_admiral_part1.Rmd
# PURPOSE       : admiral Part 1 — package architecture, derivation function taxonomy,
#                 template-driven workflow, derive_vars_merged(), and
#                 foundational ADaM dataset construction patterns.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── Mock admiral functions ────────────────────────────────────────────────────

# convert_blanks_to_na: replace "" with NA in all character columns
convert_blanks_to_na <- function(dataset) {
  dataset |>
    mutate(across(where(is.character), ~ if_else(.x == "", NA_character_, .x)))
}

# ── Date helpers ──────────────────────────────────────────────────────────────

# Impute a partial DTC string
.impute_dtc <- function(dtc, date_imputation = "first", time_imputation = "first") {
  sapply(dtc, function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    parts <- str_split(x, "T")[[1]]
    date_part <- parts[1]
    time_part <- if (length(parts) > 1) parts[2] else NA_character_

    # Pad partial dates
    segs <- str_split(date_part, "-")[[1]]
    if (length(segs) == 1) {
      segs <- c(segs, if (date_imputation == "last") "12" else "01")
    }
    if (length(segs) == 2) {
      last_day <- if (date_imputation == "last") {
        as.character(lubridate::days_in_month(as.Date(paste0(segs[1], "-", segs[2], "-01"))))
      } else "01"
      segs <- c(segs, last_day)
    }
    full_date <- paste(segs, collapse = "-")

    if (!is.na(time_part)) {
      paste0(full_date, "T", time_part)
    } else {
      full_date
    }
  }, USE.NAMES = FALSE)
}

.imputation_flag <- function(dtc, imputed_dtc, mode = "date") {
  sapply(seq_along(dtc), function(i) {
    if (is.na(dtc[i])) return(NA_character_)
    orig  <- str_split(dtc[i], "-")[[1]]
    if (length(orig) == 1) return(if (mode == "date") "Y" else "M")
    if (length(orig) == 2) return("D")
    return(NA_character_)
  }, USE.NAMES = FALSE)
}

# derive_vars_dt: convert DTC character → numeric Date + imputation flag
derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation = "none",
                            flag_imputation = "auto") {
  .e <- substitute(dtc); dtc_nm <- if (is.character(.e)) .e else deparse(.e)
  dt_nm   <- paste0(new_vars_prefix, "DT")
  dtf_nm  <- paste0(new_vars_prefix, "DTF")
  raw_dtc <- dataset[[dtc_nm]]
  imputed <- .impute_dtc(raw_dtc, date_imputation = date_imputation)
  dataset[[dt_nm]]  <- suppressWarnings(as.numeric(as.Date(imputed)))
  dataset[[dtf_nm]] <- .imputation_flag(raw_dtc, imputed, "date")
  dataset
}

# derive_vars_dtm: convert DTC → numeric POSIXct datetime + flags
derive_vars_dtm <- function(dataset, dtc, new_vars_prefix,
                             date_imputation = "none",
                             time_imputation = "none",
                             flag_imputation = "auto") {
  .e <- substitute(dtc); dtc_nm <- if (is.character(.e)) .e else deparse(.e)
  dtm_nm  <- paste0(new_vars_prefix, "DTM")
  dtf_nm  <- paste0(new_vars_prefix, "DTF")
  tmf_nm  <- paste0(new_vars_prefix, "TMF")
  raw_dtc <- dataset[[dtc_nm]]

  ti_str  <- switch(time_imputation, "first" = "00:00:00", "last" = "23:59:59", "00:00:00")
  imputed <- sapply(raw_dtc, function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    if (!str_detect(x, "T")) paste0(x, "T", ti_str) else x
  }, USE.NAMES = FALSE)
  imputed <- .impute_dtc(imputed, date_imputation = date_imputation)
  dataset[[dtm_nm]] <- suppressWarnings(as.numeric(as.POSIXct(str_replace(imputed, "T", " "), tz = "UTC")))
  dataset[[dtf_nm]] <- .imputation_flag(raw_dtc, imputed, "date")
  dataset[[tmf_nm]] <- if_else(!str_detect(raw_dtc, "T") & !is.na(raw_dtc),
                                if (time_imputation == "first") "H" else "H", NA_character_)
  dataset
}

# derive_vars_dtm_to_dt: datetime → date (strip time)
derive_vars_dtm_to_dt <- function(dataset, source_vars) {
  sv_names <- sapply(source_vars, deparse)
  for (nm in sv_names) {
    out_nm <- str_replace(nm, "DTM$", "DT")
    dataset[[out_nm]] <- trunc(dataset[[nm]] / 86400)  # seconds → days
  }
  dataset
}

# derive_vars_dy: compute study day variables (no Day 0)
derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref_nm  <- deparse(substitute(reference_date))
  sv_names <- sapply(source_vars, deparse)
  for (nm in sv_names) {
    out_nm <- str_replace(nm, "DT$", "DY")
    ref_days <- dataset[[ref_nm]]
    src_days <- dataset[[nm]]
    raw_diff <- src_days - ref_days
    dataset[[out_nm]] <- case_when(
      is.na(raw_diff)  ~ NA_real_,
      raw_diff >= 0    ~ raw_diff + 1,
      TRUE             ~ raw_diff
    )
  }
  dataset
}

# exprs: capture unevaluated arguments as a list of expressions (mirrors rlang::exprs)
exprs <- function(...) as.list(substitute(list(...)))[-1]

# derive_vars_merged: left-join with filter, order, first/last
derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL,
                                order = NULL, filter_add = NULL, mode = NULL,
                                missing_values = NULL) {
  by_names <- sapply(by_vars, deparse)

  # Apply filter_add
  filter_expr <- substitute(filter_add)
  if (!is.null(filter_expr) && deparse(filter_expr) != "NULL") {
    dataset_add <- dataset_add |> filter(!!filter_expr)
  }

  # Order
  if (!is.null(order)) {
    ord_names <- sapply(order, deparse)
    ord_descs <- grepl("^desc\\(", ord_names)
    ord_clean <- str_replace_all(ord_names, "^desc\\(|\\)$", "")
    for (i in rev(seq_along(ord_clean))) {
      if (ord_descs[i]) {
        dataset_add <- dataset_add |> arrange(desc(.data[[ord_clean[i]]]))
      } else {
        dataset_add <- dataset_add |> arrange(.data[[ord_clean[i]]])
      }
    }
  }

  # First/last selection
  if (!is.null(mode)) {
    fn <- if (mode == "first") dplyr::slice_head else dplyr::slice_tail
    dataset_add <- dataset_add |>
      group_by(across(all_of(by_names))) |>
      fn(n = 1) |>
      ungroup()
  }

  # Rename/select new vars
  if (!is.null(new_vars)) {
    nv_names <- sapply(new_vars, deparse)
    nv_map   <- setNames(names(nv_names), nv_names)  # new_name = old_name
    keep_cols <- c(by_names, names(nv_map))
    dataset_add <- dataset_add |>
      select(any_of(keep_cols)) |>
      rename(any_of(nv_map))
    new_col_names <- nv_names
  } else {
    new_col_names <- setdiff(names(dataset_add), by_names)
  }

  dataset |>
    left_join(
      dataset_add |> select(any_of(c(by_names, new_col_names))),
      by = by_names
    )
}

# derive_var_extreme_flag: mark first/last record per group
derive_var_extreme_flag <- function(dataset, by_vars, order, new_var, mode = "first") {
  by_names  <- sapply(by_vars, deparse)
  ord_names <- sapply(order, deparse)
  out_nm    <- deparse(substitute(new_var))
  ord_descs <- grepl("^desc\\(", ord_names)
  ord_clean <- str_replace_all(ord_names, "^desc\\(|\\)$", "")

  sorted <- dataset
  for (i in rev(seq_along(ord_clean))) {
    if (ord_descs[i]) {
      sorted <- sorted |> arrange(desc(.data[[ord_clean[i]]]))
    } else {
      sorted <- sorted |> arrange(.data[[ord_clean[i]]])
    }
  }

  flagged <- sorted |>
    group_by(across(all_of(by_names))) |>
    mutate(.row_rank = row_number()) |>
    ungroup() |>
    mutate(!!out_nm := case_when(
      (mode == "first" & .row_rank == 1) |
      (mode == "last"  & .row_rank == max(.row_rank)) ~ "Y",
      TRUE ~ NA_character_
    )) |>
    select(-.row_rank)

  # Merge flag back in original row order
  dataset |>
    left_join(flagged |> select(any_of(c(by_names, ord_clean, out_nm))),
              by = c(by_names, ord_clean))
}

# derive_var_merged_exist_flag: Y/NA based on existence in another dataset
derive_var_merged_exist_flag <- function(dataset, dataset_add, by_vars,
                                          new_var, condition) {
  by_names <- sapply(by_vars, deparse)
  out_nm   <- deparse(substitute(new_var))
  cond_expr <- substitute(condition)
  flagged  <- dataset_add |>
    filter(!!cond_expr) |>
    distinct(across(all_of(by_names))) |>
    mutate(!!out_nm := "Y")
  dataset |> left_join(flagged, by = by_names)
}

# derive_vars_duration: compute duration between two date variables
derive_vars_duration <- function(dataset, new_var, start_date, end_date,
                                  out_unit = "days", add_one = TRUE) {
  out_nm  <- deparse(substitute(new_var))
  start_nm <- deparse(substitute(start_date))
  end_nm   <- deparse(substitute(end_date))
  raw_diff <- dataset[[end_nm]] - dataset[[start_nm]]
  if (out_unit == "days") {
    dataset[[out_nm]] <- case_when(
      is.na(raw_diff) ~ NA_real_,
      add_one         ~ raw_diff + 1,
      TRUE            ~ raw_diff
    )
  } else if (out_unit == "months") {
    dataset[[out_nm]] <- round(as.numeric(raw_diff) / 30.4375, 1)
  }
  dataset
}

# derive_var_trtdurd: convenience wrapper for treatment duration
derive_var_trtdurd <- function(dataset, new_var = TRTDURD,
                                start_date = TRTSDT, end_date = TRTEDT) {
  out_nm   <- deparse(substitute(new_var))
  start_nm <- deparse(substitute(start_date))
  end_nm   <- deparse(substitute(end_date))
  dataset[[out_nm]] <- dataset[[end_nm]] - dataset[[start_nm]] + 1
  dataset
}

# call_derivation: apply one function multiple times with different variable params
call_derivation <- function(dataset, derivation, variable_params, ...) {
  shared_args <- list(...)
  for (vp in variable_params) {
    args_i <- c(list(dataset = dataset), vp, shared_args)
    dataset <- do.call(derivation, args_i)
  }
  dataset
}

# params: capture bare column names as strings so do.call() can pass them safely
params <- function(...) {
  dots <- as.list(substitute(list(...)))[-1]
  lapply(dots, function(x) if (is.symbol(x)) deparse(x) else x)
}

# ── Sample data ───────────────────────────────────────────────────────────────

# SDTM-like datasets with deliberate blanks (to test convert_blanks_to_na)
dm_raw <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101", "01-703-1211"),
  AGE     = c(63L, 64L, 71L, 56L, 45L),
  SEX     = c("M", "F", "M", "F", "M"),
  RACE    = c("WHITE", "WHITE", "ASIAN", "WHITE", "BLACK OR AFRICAN AMERICAN"),
  ARM     = c("Xanomeline High Dose", "Xanomeline Low Dose",
              "Placebo", "Xanomeline High Dose", "Placebo"),
  ACTARM  = c("XANOMELINE HIGH DOSE", "XANOMELINE LOW DOSE",
              "PLACEBO", "XANOMELINE HIGH DOSE", "PLACEBO"),
  RFXSTDTC = c("2014-01-02", "2012-08-05", "2013-07-09", "2013-04-14", "2014-03-22")
)

ex_raw <- tibble(
  STUDYID  = "CDISCPILOT01",
  USUBJID  = c("01-701-1015", "01-701-1015",
               "01-701-1023", "01-702-1042", "01-702-1101",
               "01-703-1211", "01-703-1211"),
  EXTRT    = c("XANOMELINE", "XANOMELINE",
               "XANOMELINE", "PLACEBO", "XANOMELINE",
               "PLACEBO", "PLACEBO"),
  EXDOSE   = c(81, 81, 54, 0, 81, 0, 0),
  EXSEQ    = c(1L, 2L, 1L, 1L, 1L, 1L, 2L),
  EXSTDTC  = c("2014-01-02", "2014-02-02",
               "2012-08-05", "2013-07-09", "2013-04-14",
               "2014-03-22", "2014-04-22"),
  EXENDTC  = c("2014-01-31", "2014-02-28",
               "2012-11-01", "2013-09-15", "2013-06-30",
               "2014-04-21", "2014-05-21")
)

# LB for study day derivation example
lb_raw <- tibble(
  STUDYID  = "CDISCPILOT01",
  USUBJID  = c("01-701-1015", "01-701-1015", "01-701-1023"),
  LBTESTCD = c("ALT", "ALT", "ALT"),
  LBDTC    = c("2014-01-02", "2014-02-10", "2012-08-05"),
  LBORRES  = c("28", "35", "22")
)

# Datasets with blanks (to demo convert_blanks_to_na)
dm_with_blanks <- dm_raw |>
  mutate(ARM = if_else(USUBJID == "01-703-1211", "", ARM))

# Pre-processed clean versions
dm  <- convert_blanks_to_na(dm_raw)
ex  <- convert_blanks_to_na(ex_raw)
lb  <- convert_blanks_to_na(lb_raw)

# ── Real pharmaversesdtm data (with fallback to mock if package missing) ─────
if (requireNamespace("pharmaversesdtm", quietly = TRUE)) {
  dm_real <- pharmaversesdtm::dm
  ex_real <- pharmaversesdtm::ex
  lb_real <- pharmaversesdtm::lb
} else {
  # Fallback: reuse the mock SDTM datasets defined above so all chunks still run
  dm_real <- dm_raw
  ex_real <- ex_raw
  lb_real <- lb_raw
}
```

## Welcome

This is Lesson 14 — **`{admiral}` Part 1: Foundations and Philosophy**.

Admiral is the pharmaverse flagship for building ADaM datasets. Before writing a single derivation, you need to internalize two things: the manifesto (admiral is a *toolbox*, not a black box) and the function naming convention (which tells you 80% of what a function does before you read the docs).

This lesson covers:

- The admiral manifesto — why it matters
- The function prefix convention: `derive_var_*`, `derive_vars_*`, `derive_param_*`, `compute_*`
- `exprs()` — the expression argument pattern
- `convert_blanks_to_na()` — why it's always your first call
- Date/time helpers: `derive_vars_dt()`, `derive_vars_dtm()`, `derive_vars_dy()`
- `derive_vars_merged()` — the most important function
- `derive_var_extreme_flag()` and `derive_var_merged_exist_flag()`

All admiral functions are mocked and run without package installation.

---

## 1 — The manifesto

Admiral is a *toolbox*. There is no `make_adsl()` function that magically produces an ADSL from raw SDTM. Instead, you assemble ADaMs by **chaining many small, auditable function calls** — each doing one identifiable thing. This is the single most important mindset shift for a SAS programmer.

### 1.1 Toolbox vs. canned PROC vs. macro library

A SAS programmer entering pharmaverse usually arrives with one of three mental models. None of them is quite right, and the differences matter.

| Mental model | Where SAS folks know it from | Why admiral is different |
|---|---|---|
| **Canned PROC** (`PROC LIFETEST`, `PROC FREQ`) | One call, dozens of options, opaque output dataset | admiral has *no* canned PROC — every variable you add is its own function call, fully visible in the code |
| **Macro library** (e.g. `%adslmacro`, `%trtemfl`) | A black box: pass `%let drug=XAN;`, get back an ADSL | admiral is more granular than a typical macro library — even a single derivation like TRTSDT is one function call, not bundled with TRTEDT |
| **Function-level utility library** (`%macro daysbtwn`) | Tiny helpers you `%include` from a shared folder | This is closest. admiral is a library of small, single-purpose functions that you compose explicitly in your program. |

The pharmaverse manifesto puts it this way: admiral functions have **clear purpose, no multi-purpose functions**. There is `derive_vars_dt()` and there is `derive_vars_dtm()`. There is no `derive_vars_dt_or_dtm()` with a `type=` switch. The maintainers refuse to add such a function on principle, because it would make code less predictable.

### 1.2 "Monolithic PROC" vs. "many small derivations"

```
   /* SAS DATA STEP — one giant block,           |   # admiral — many small calls,
      hard to audit, harder to test piece by      |     each one independently verifiable
      piece */                                    |
   data adsl;                                     |   adsl <- dm |>
     set dm;                                      |     derive_vars_dt(
     RFXSTDT = input(RFXSTDTC, yymmdd10.);        |       dtc = RFXSTDTC,
     if RFXSTDT ne . then TRTSDT = RFXSTDT;       |       new_vars_prefix = "RFXST"
     /* ... 200 more lines of derivations ... */  |     ) |>
     SAFFL = 'N';                                 |     derive_vars_merged(
     if EXDOSE > 0 then SAFFL = 'Y';              |       dataset_add = ex,
     /* ... and so on ... */                      |       new_vars    = exprs(TRTSDT = EXSTDT),
   run;                                           |       by_vars     = exprs(STUDYID, USUBJID),
                                                  |       mode        = "first"
                                                  |     ) |>
                                                  |     derive_var_merged_exist_flag(
                                                  |       dataset_add = ex,
                                                  |       new_var     = SAFFL,
                                                  |       condition   = EXDOSE > 0
                                                  |     )
```

> **SAS analog mental model:** Admiral is to ADaM derivation what a well-organized macro folder is to a SAS production area — except the macros are versioned, regression-tested, on CRAN, and named so consistently that you can guess a function's behavior from its name.

### 1.3 The 3-year deprecation cycle (vs. the SAS tradition)

Pharma data flows have a long memory. A Phase III pivotal trial may have been programmed in 2022 and is still being re-run for a regulatory request in 2026. SAS solves this by **never deprecating anything** — `PROC LIFETEST` from SAS 6 still runs in SAS 9.4. Admiral takes a different approach:

- **Year 1:** the function is marked deprecated, emits a warning at runtime, but still works.
- **Year 2:** the function emits an error message pointing to its replacement, but still runs if you set an option.
- **Year 3:** the function is removed entirely; you must rewrite the call.

The point is to give study teams time to migrate without surprise breakage mid-cycle. If you `library(admiral)` in 2026 and your code from 2023 throws a deprecation warning, you have until 2026 to fix it.

> **SAS analog mental model:** Imagine if SAS announced in 2025 that `PROC MEANS NWAY` would print a warning, in 2026 it would print an error you had to suppress, and in 2027 it would be removed entirely. That would be the admiral lifecycle.

```{r quiz-manifesto, echo=FALSE}
quiz(
  question(
    "The admiral maintainers describe the package as 'a toolbox, not an automation tool'. This means:",
    answer("You call make_adsl() and admiral builds the dataset for you"),
    answer("You assemble ADaMs by chaining many small, auditable function calls — no black-box 'make ADSL' functions", correct = TRUE),
    answer("admiral only handles one ADaM at a time"),
    random_answer_order = TRUE
  ),
  question(
    "Why does the manifesto emphasize 'clear purpose, no multi-purpose functions'?",
    answer("Regulators require single-purpose functions"),
    answer("It makes function names predictable and code auditable — each line does exactly one identifiable thing", correct = TRUE),
    answer("Multi-purpose functions are slower in R"),
    random_answer_order = TRUE
  ),
  question(
    "Admiral's 3-year deprecation cycle (warning → error → removal) exists because:",
    answer("The development team needs time to update examples"),
    answer("Studies run for years; a mid-study package update should never break existing code without ample warning time", correct = TRUE),
    answer("CRAN requires 3-year cycles"),
    random_answer_order = TRUE
  ),
  question(
    "Which best describes admiral's relationship to a typical SAS macro library like `%adslmacro`?",
    answer("admiral is a single big macro you parameterize and call once"),
    answer("admiral is more granular — a typical macro might bundle TRTSDT, TRTEDT, and TRTDURD into one call; admiral splits them into three named functions", correct = TRUE),
    answer("admiral and SAS macro libraries have no analog"),
    random_answer_order = TRUE
  ),
  question(
    "Why doesn't admiral ship a `derive_dates(type = 'dt'|'dtm')` switch instead of two separate functions?",
    answer("R doesn't support optional arguments"),
    answer('It violates the "clear purpose, no multi-purpose functions" rule — predictable, single-purpose names make code more auditable and easier to grep', correct = TRUE),
    answer("The pharmaverse maintainers haven't gotten around to it yet"),
    random_answer_order = TRUE
  )
)
```

---

## 2 — Function naming convention

Admiral function names are highly structured. Once you internalize the prefix taxonomy, you can guess what a function does from its name alone — and you can grep the package for the exact function you need.

### 2.1 The six prefixes

| Prefix | Singular/Plural | What it does | Closest SAS analog | Example function names |
|---|---|---|---|---|
| `derive_var_*` | singular | Adds **one** variable to a dataset | A `DATA` step assignment: `TRTDURD = TRTEDT - TRTSDT + 1;` | `derive_var_trtdurd()`, `derive_var_extreme_flag()`, `derive_var_merged_exist_flag()` |
| `derive_vars_*` | plural | Adds **multiple related** variables in one call (e.g. DT + DTF flag) | A macro that emits several `OUTPUT` columns | `derive_vars_dt()`, `derive_vars_dtm()`, `derive_vars_merged()`, `derive_vars_duration()` |
| `derive_param_*` | parameter | Adds a **new parameter row** (PARAMCD) to a BDS dataset | `PROC TRANSPOSE` then `SET old new;` to append a derived parameter | `derive_param_bmi()`, `derive_param_bsa()`, `derive_param_extreme_record()` |
| `compute_*` | scalar | Returns a **scalar value** (does not modify a dataset) | A SAS function used inside a DATA step: `AGE = YRDIF(BRTHDT, RFSTDT, 'AGE');` | `compute_age_years()`, `compute_duration()`, `compute_bmi()` |
| `get_*` | retrieval | Returns a **value or object** from inputs (often a tibble of derived records) | `PROC SQL SELECT ... INTO :macrovar` — fetch and stash | `get_summary_records()`, `get_one_to_many_dataset()` |
| `filter_*` | reducer | Returns a **subset** of an input dataset based on a clinical rule | `DATA out; SET in; IF condition; RUN;` or `PROC SQL WHERE` | `filter_extreme()`, `filter_relative()`, `filter_date_sources()` |

### 2.2 ASCII tree of the prefix taxonomy

```
admiral function name
│
├── derive_*           ──→ "modifies a dataset by adding columns"
│   ├── derive_var_*       (singular: ONE new column)
│   │     e.g. derive_var_trtdurd, derive_var_extreme_flag
│   ├── derive_vars_*      (plural: MULTIPLE columns, usually related — DT + DTF, etc.)
│   │     e.g. derive_vars_dt, derive_vars_dtm, derive_vars_merged
│   └── derive_param_*     (adds a new PARAMETER ROW to a BDS dataset)
│         e.g. derive_param_bmi, derive_param_extreme_record
│
├── compute_*          ──→ "pure: scalar in → scalar out, no dataset"
│         e.g. compute_age_years, compute_bmi, compute_duration
│
├── get_*              ──→ "fetch / retrieve some derived object"
│         e.g. get_summary_records, get_one_to_many_dataset
│
└── filter_*           ──→ "shrink a dataset based on a clinical rule"
          e.g. filter_extreme, filter_relative, filter_date_sources
```

### 2.3 When to reach for which prefix

| You need to... | Pick this prefix |
|---|---|
| Add TRTSDT to ADSL using a date from EX | `derive_vars_merged()` — it joins another dataset and adds variable(s) |
| Add a `BASE` (baseline value) column derived from a flag | `derive_var_base()` — single variable |
| Add a new BMI row alongside HEIGHT and WEIGHT in ADVS | `derive_param_bmi()` — new parameter row, not a new column |
| Compute someone's age in years given two dates, no dataset | `compute_age_years()` — scalar function |
| Pull the latest non-missing date from a long lab dataset | `get_summary_records()` or `derive_vars_merged()` with `mode = "last"` |
| Subset AEs to only "treatment-emergent" rows | `filter_*` family (or a plain `dplyr::filter()` for simple cases) |

### 2.4 SAS analog comparison: function naming

```
   /* SAS — no convention; macro names    |   # admiral — convention enforced
      may be anything */                  |
   %adsl_dates;        /* what does this  |   adsl |>
                          do? */          |     derive_vars_dt(...)      # CLEARLY adds vars (plural)
   %trtem_flag;        /* one var or      |     derive_var_trtemfl(...)  # CLEARLY adds ONE var
                          many? */        |     derive_param_bmi(...)   # CLEARLY adds a PARAM row
   %compute_age;       /* dataset or      |     compute_age_years(...)  # CLEARLY a scalar function
                          scalar? */      |
```

> **SAS analog mental model:** Imagine you converted your shop's macro library to follow a strict naming rule where the first word of every macro name tells you exactly what kind of thing it returns. That convention is what admiral enforces at the package level — you read the name and you know whether to expect a column, a row, a scalar, or a subset.

### 2.5 Anatomy of a name

Take `derive_var_merged_exist_flag()`:

```
derive_  var_  merged_  exist_  flag
   │      │       │         │      │
   │      │       │         │      └── what's added: a flag (Y/NA)
   │      │       │         └── the rule for setting it: existence in another dataset
   │      │       └── how it's computed: by merging
   │      └── singular: ONE new variable is added
   └── dataset-mutating action
```

You can read the function and predict its signature before opening the help page:

- It takes a `dataset` (the one we are adding to).
- It takes a `dataset_add` (the one we look up in).
- It takes `by_vars` (the join keys).
- It takes a `condition` (the existence rule).
- It produces a single new column whose name you provide via `new_var`.

```{r quiz-naming, echo=FALSE}
quiz(
  question(
    "derive_vars_dt() (plural 'vars') vs derive_var_trtdurd() (singular 'var') — the difference is:",
    answer("'vars' functions add multiple related variables at once; 'var' functions add a single variable", correct = TRUE),
    answer("'vars' is newer than 'var'"),
    answer("'var' functions are faster"),
    random_answer_order = TRUE
  ),
  question(
    "derive_param_bmi() belongs to which prefix family?",
    answer("derive_var_* — adding one variable"),
    answer("derive_param_* — adding a new parameter row to a BDS dataset", correct = TRUE),
    answer("compute_* — pure scalar computation"),
    random_answer_order = TRUE
  ),
  question(
    "compute_age_years() returns:",
    answer("A dataset with a new AGE column"),
    answer("A scalar numeric value without modifying a dataset", correct = TRUE),
    answer("A plot of age distribution"),
    random_answer_order = TRUE
  ),
  question(
    "You are asked to add a BMI row to a long-format ADVS dataset (so PARAMCD=BMI appears alongside HEIGHT and WEIGHT). Which prefix?",
    answer("derive_var_*"),
    answer("derive_param_* — you are adding a new parameter row, not a column", correct = TRUE),
    answer("compute_*"),
    random_answer_order = TRUE
  ),
  question(
    "A function named `filter_extreme()` most likely:",
    answer("Adds a flag for extreme records"),
    answer("Returns a subset of the dataset keeping only extreme records", correct = TRUE),
    answer("Removes outliers using a statistical rule"),
    random_answer_order = TRUE
  )
)
```

### 2.6 Exercise: Guess the prefix

For each task below, reason about which prefix family the right admiral function belongs to. Use the dataset names and SAS-side intent to choose.

```{r naming-guess, exercise=TRUE, exercise.lines=18}
# For each row, replace ___ with one of:
#   "derive_var", "derive_vars", "derive_param", "compute", "get", "filter"
prefix_guess <- tribble(
  ~task,                                                 ~prefix,
  "Add TRTSDT (one new variable) to ADSL from EX",        ___,
  "Add both ASTDT and ASTDTF (two related vars) to ADAE", ___,
  "Add a derived BMI row (PARAMCD='BMI') to ADVS",        ___,
  "Compute someone's age in years from two date vectors", ___,
  "Reduce AE rows down to treatment-emergent only",       ___
)
prefix_guess
```

```{r naming-guess-hint-1}
# Singular vs plural variable count drives "derive_var" vs "derive_vars".
# A new row in a BDS dataset is "derive_param".
# A scalar in / scalar out function is "compute".
```

```{r naming-guess-hint-2}
# Reducing rows = "filter_". Adding a single column = "derive_var_".
```

```{r naming-guess-solution}
prefix_guess <- tribble(
  ~task,                                                 ~prefix,
  "Add TRTSDT (one new variable) to ADSL from EX",        "derive_var",
  "Add both ASTDT and ASTDTF (two related vars) to ADAE", "derive_vars",
  "Add a derived BMI row (PARAMCD='BMI') to ADVS",        "derive_param",
  "Compute someone's age in years from two date vectors", "compute",
  "Reduce AE rows down to treatment-emergent only",       "filter"
)
prefix_guess
```

> **Read the output:** look at the `prefix` column. Notice that "Add TRTSDT" alone is `derive_var_*` because TRTSDT is *one* variable — even though it requires a join to derive it (joins don't change the prefix family; the *count of variables produced* does).

---

## 3 — `convert_blanks_to_na()`: always your first call

SAS XPT files store missing character values as `""` (empty string), not `NA`. R's `is.na()` returns `FALSE` for empty strings — so downstream derivations silently produce wrong results if you skip this step.

### 3.1 SAS analog comparison

| admiral function | What it does | Closest SAS analog | SAS code sketch |
|---|---|---|---|
| `convert_blanks_to_na()` | Replace `""` with `NA` in every character column | A DATA step pass that converts `if VAR=" " then call missing(VAR);` | `data dm; set dm; if ARM=" " then ARM=" "; ... run;` (SAS treats blank-only strings as missing automatically; R does not) |

The reason this is needed is a quirk of the SAS XPT file format and how `haven::read_xpt()` reads it into R.

### 3.2 Why XPT stores `""` and not `.`

In SAS, *numeric* missing is the single character `.`. *Character* missing is a string of one or more blanks (`" "`, `"  "`, or longer). SAS handles both transparently — `if ARM=" " then` matches whether ARM is one blank or thirty.

When SAS writes an XPT (a Version 5 transport file), character missings are stored as their literal blank padding. When `haven::read_xpt()` reads that file into R, it left-trims those blanks. A character variable that was missing in SAS becomes `""` (a zero-length string) in R.

R's `is.na("")` returns `FALSE`. R's `is.na(NA)` returns `TRUE`. These are different — and admiral derivations that branch on `is.na()` (which is most of them) will silently produce the wrong answer if you forget this conversion.

```
   /* SAS — missing character handled       |   # R after read_xpt() — blank is NOT NA
      transparently */                      |
   if ARM = " " then put "missing";         |   "" |> is.na()      # FALSE  (oops!)
   if missing(ARM) then put "missing";      |   NA_character_ |> is.na()  # TRUE
   /* Both detect "missing" correctly */    |
                                            |   # Fix: convert blanks BEFORE any derivation
                                            |   dm <- convert_blanks_to_na(dm)
                                            |   dm$ARM |> is.na()   # TRUE  (now correct)
```

### 3.3 Anatomy of the call

```r
convert_blanks_to_na(
  dataset    # the input data frame; every character column is scanned
)
```

- Only **character columns** are touched. Numeric, logical, date, and factor columns pass through unchanged.
- The function replaces only **fully empty strings** (`""`). It does **not** trim whitespace, so a value of `" "` (one literal space) is *not* converted. If you suspect padded blanks, do `mutate(across(where(is.character), str_trim))` first.
- It is **idempotent** — calling it twice is the same as calling it once.

> **SAS analog mental model:** `convert_blanks_to_na()` is the R equivalent of running a `DATA` step pass with `array c[*] _character_; do i = 1 to dim(c); if c[i]=" " then call missing(c[i]); end;` — except admiral does it for you in one line, and it is the *only* line you should ever write between `read_xpt()` and your first derivation.

### 3.4 Contrast with `read_xpt()` and `PROC IMPORT`

| Reader | Blank handling | Action required |
|---|---|---|
| `haven::read_xpt()` | Leaves blank as `""` | **MUST** call `convert_blanks_to_na()` |
| `haven::read_sas()` (SAS7BDAT) | Leaves blank as `""` | Same — must convert |
| `pharmaverseadam::adsl` (built tibble) | Already uses `NA` | Optional but harmless |
| SAS `PROC IMPORT` (SAS-side) | Reads CSV blanks as missing automatically | N/A — different ecosystem |
| `read_xpt()` in older `SASxport` package | Sometimes converted, sometimes not | Inconsistent — always convert defensively |

> **Defensive habit:** call `convert_blanks_to_na()` on *every* SDTM dataset you import, even if you think it is already clean. It is cheap, idempotent, and saves you from one of the most insidious bug classes in pharmaverse.

```{r blanks-demo}
# Demonstrate the problem
cat("Empty string is.na test:", is.na(""), "\n")
cat("Proper NA is.na test:    ", is.na(NA), "\n")

# dm_with_blanks has ARM = "" for subject 01-703-1211
dm_with_blanks |>
  filter(USUBJID == "01-703-1211") |>
  select(USUBJID, ARM)
```

```{r blanks-fix}
# After convert_blanks_to_na()
dm_fixed <- convert_blanks_to_na(dm_with_blanks)

dm_fixed |>
  filter(USUBJID == "01-703-1211") |>
  select(USUBJID, ARM)

cat("ARM is NA after conversion:", is.na(dm_fixed$ARM[dm_fixed$USUBJID == "01-703-1211"]), "\n")
```

> **Read the output:** look at the cat() line that prints `is.na test`. Before conversion, the ARM column for subject `01-703-1211` is the literal `""` (zero-length string) — and `is.na("")` returns FALSE. After `convert_blanks_to_na()` the same cell is genuine `NA_character_` and `is.na()` returns TRUE. Every admiral derivation that branches on `is.na()` (most do) only works correctly after this transformation.

### 3.5 Exercise: Convert blanks and verify with is.na

```{r blanks-exercise, exercise=TRUE, exercise.lines=14}
# Create a dataset with blanks in multiple columns
test_ds <- tibble(
  USUBJID = c("S001", "S002", "S003"),
  ARM     = c("TRT", "", "TRT"),
  RACE    = c("WHITE", "ASIAN", ""),
  AGE     = c(45L, NA_integer_, 62L)
)

# Convert blanks
test_clean <- convert_blanks_to_na(test_ds)

# Check: how many NAs in ARM before and after?
cat("NAs in ARM (before):", sum(is.na(test_ds$ARM)), "\n")
cat("NAs in ARM (after): ", sum(is.na(test_clean$ARM)), "\n")
cat("NAs in RACE (after):", sum(is.na(test_clean$RACE)), "\n")
test_clean
```

### 3.6 Exercise: Blank audit on real pharmaversesdtm DM

The real `dm_real` arrives as if read from a SAS XPT. Check whether it carries any blank strings, then apply `convert_blanks_to_na()` defensively. This pattern belongs at the top of every admiral script regardless of whether blanks are expected.

```{r blanks-real-dm, exercise=TRUE, exercise.lines=18}
# Count blank strings per character column in dm_real
blank_counts <- dm_real |>
  select(where(is.character)) |>
  summarise(across(everything(), ~ sum(.x == "", na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_blanks") |>
  filter(n_blanks > 0)

cat("Columns with blank strings in dm_real:\n")
if (nrow(blank_counts) == 0) cat("  (none found — dataset is already clean)\n") else print(blank_counts)

# Apply defensively anyway
dm_real_clean <- convert_blanks_to_na(dm_real)

# How many NAs exist per character column?
na_summary <- sapply(select(dm_real_clean, where(is.character)),
                     function(x) sum(is.na(x)))
cat("\nNA counts in character columns after conversion:\n")
print(na_summary[na_summary > 0])
```

```{r blanks-real-dm-hint-1}
# summarise(across(everything(), ~ sum(.x == "", na.rm = TRUE)))
# The real value: even if zero blanks exist today, a study database refresh
# could introduce them — calling convert_blanks_to_na() costs nothing.
```

```{r blanks-real-dm-solution}
blank_counts <- dm_real |>
  select(where(is.character)) |>
  summarise(across(everything(), ~ sum(.x == "", na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_blanks") |>
  filter(n_blanks > 0)

if (nrow(blank_counts) == 0) cat("(none found)\n") else print(blank_counts)

dm_real_clean <- convert_blanks_to_na(dm_real)

na_summary <- sapply(select(dm_real_clean, where(is.character)),
                     function(x) sum(is.na(x)))
print(na_summary[na_summary > 0])
```

> **Read the output:** look at the `na_summary` printout — every entry is a count of *genuine* NAs that admiral derivations will correctly treat as missing. Without `convert_blanks_to_na()`, any cell that used to be `""` would have been silently treated as a present value of length zero.

### 3.7 Exercise: Diagnose the wrong-result bug

The chunk below shows what happens when you skip `convert_blanks_to_na()` and try to count "subjects with no recorded arm." Fix the bug.

```{r blanks-diagnose, exercise=TRUE, exercise.lines=18}
test_ds <- tibble(
  USUBJID = c("S001", "S002", "S003", "S004"),
  ARM     = c("Placebo", "", "Active", "")   # S002 and S004 have blank ARM
)

# Wrong: counts ZERO because is.na("") is FALSE
n_missing_wrong <- test_ds |>
  filter(is.na(ARM)) |>
  nrow()
cat("Subjects with missing ARM (WRONG):", n_missing_wrong, "\n")

# Fix: convert blanks to NA first
test_ds_fixed <- ___(test_ds)
n_missing_right <- test_ds_fixed |>
  filter(is.na(ARM)) |>
  nrow()
cat("Subjects with missing ARM (RIGHT):", n_missing_right, "\n")
```

```{r blanks-diagnose-hint-1}
# Wrap test_ds in convert_blanks_to_na() before the filter.
```

```{r blanks-diagnose-hint-2}
# test_ds_fixed <- convert_blanks_to_na(test_ds)
```

```{r blanks-diagnose-solution}
test_ds <- tibble(
  USUBJID = c("S001", "S002", "S003", "S004"),
  ARM     = c("Placebo", "", "Active", "")
)

n_missing_wrong <- test_ds |>
  filter(is.na(ARM)) |>
  nrow()
cat("Subjects with missing ARM (WRONG):", n_missing_wrong, "\n")

test_ds_fixed <- convert_blanks_to_na(test_ds)
n_missing_right <- test_ds_fixed |>
  filter(is.na(ARM)) |>
  nrow()
cat("Subjects with missing ARM (RIGHT):", n_missing_right, "\n")
```

> **Read the output:** the WRONG count is 0 — `filter(is.na(ARM))` matches nothing because `""` is not `NA` in R. After conversion, the RIGHT count is 2. This is the exact bug class that motivates the existence of `convert_blanks_to_na()`.

### 3.8 Knowledge check

```{r quiz-blanks, echo=FALSE}
quiz(
  question(
    "Why does `haven::read_xpt()` produce empty strings instead of NAs for missing character values?",
    answer("haven has a bug"),
    answer("The XPT file format stores character missing as padded blanks; haven left-trims them to '' instead of converting them to NA", correct = TRUE),
    answer("R does not support character NAs"),
    random_answer_order = TRUE
  ),
  question(
    "A SAS programmer writes `if missing(ARM) then ...`. The closest R equivalent (after `convert_blanks_to_na()`) is:",
    answer("if (ARM == 'NA') ..."),
    answer("if (is.na(ARM)) ...", correct = TRUE),
    answer("if (ARM == '') ..."),
    random_answer_order = TRUE
  ),
  question(
    "`convert_blanks_to_na()` does NOT trim whitespace. A value of `' '` (one literal space) is:",
    answer("Converted to NA"),
    answer("Left unchanged — only zero-length '' is converted", correct = TRUE),
    answer("Dropped from the dataset"),
    random_answer_order = TRUE
  ),
  question(
    "Calling `convert_blanks_to_na()` twice in a row will:",
    answer("Throw an error"),
    answer("Have no extra effect — the function is idempotent", correct = TRUE),
    answer("Convert NAs back to blanks"),
    random_answer_order = TRUE
  )
)
```

---

## 4 — `derive_vars_dt()`: ISO 8601 DTC → numeric Date

The workhorse for converting character SDTM dates to numeric dates suitable for arithmetic.

### 4.1 What it produces

For every call, `derive_vars_dt()` adds **two** variables (hence the plural `vars`): the numeric date and an imputation flag describing what (if anything) was filled in.

| Output variable | Type | Content |
|---|---|---|
| `<prefix>DT` | numeric (days since 1970-01-01) | The fully resolved analysis date |
| `<prefix>DTF` | character | `"D"` (day imputed), `"M"` (month+day imputed), `"Y"` (year imputed), `NA` (nothing imputed) |

### 4.2 SAS analog comparison

| admiral function | What it does | Closest SAS analog | SAS code sketch |
|---|---|---|---|
| `derive_vars_dt()` | DTC string → numeric date + imputation flag | `INPUT()` with informat, plus a length-check macro | `RFXSTDT = INPUT(RFXSTDTC, E8601DA10.);` plus `if length(RFXSTDTC)=7 then RFXSTDTF="D";` |
| `derive_vars_dt(date_imputation="first")` | Impute missing day to "01" | Macro that pads partial date with `01` | `if length(RFXSTDTC)=7 then RFXSTDT = INPUT(cats(RFXSTDTC,"-01"), E8601DA10.);` |
| `derive_vars_dt(date_imputation="last")` | Impute missing day to month-end | Macro with `INTNX('month',d,0,'E')` | `RFXSTDT = INTNX('MONTH', INPUT(cats(RFXSTDTC,"-01"), E8601DA10.), 0, 'E');` |

### 4.3 ASCII side-by-side

```
   /* SAS — informat input + manual flag */     |   # admiral — one call, both vars
   data adsl;                                   |   adsl |>
     set adsl;                                  |     derive_vars_dt(
     RFXSTDT = input(RFXSTDTC, E8601DA10.);     |       dtc             = RFXSTDTC,
     if length(RFXSTDTC)=7 then do;             |       new_vars_prefix = "RFXST",
       RFXSTDT = input(cats(RFXSTDTC,"-01"),    |       date_imputation = "first"
                       E8601DA10.);             |     )
       RFXSTDTF = "D";                          |
     end;                                       |
     else if length(RFXSTDTC)=4 then do;        |
       RFXSTDT = input(cats(RFXSTDTC,"-01-01"), |
                       E8601DA10.);             |
       RFXSTDTF = "M";                          |
     end;                                       |
   run;                                         |
```

### 4.4 Anatomy of the call

```r
derive_vars_dt(
  dataset         = dm,         # input dataset
  dtc             = RFXSTDTC,   # character ISO 8601 column to parse
  new_vars_prefix = "RFXST",    # name stem; output is RFXSTDT + RFXSTDTF
  date_imputation = "first",    # "first", "last", or "none"
  flag_imputation = "auto"      # whether to emit the *DTF flag automatically
)
```

- **`dtc`** is the *character* date column in ISO 8601 form (`"2014-01-02"` or partial `"2014-01"` or `"2014"`). It is passed unquoted, tidy-evaluation style.
- **`new_vars_prefix`** is a stem. If you pass `"RFXST"`, you get `RFXSTDT` (numeric) and `RFXSTDTF` (flag). If you pass `"A"`, you get `ADT` and `ADTF` (the canonical CDISC analysis-date naming).
- **`date_imputation`** controls how partial dates are filled in. `"first"` pads to January 1st / first of month. `"last"` pads to December 31st / month-end. `"none"` returns NA for any partial date.
- **`flag_imputation`** defaults to `"auto"` — admiral emits the DTF flag any time it had to impute. Set to `"none"` to suppress the flag column entirely (rare; usually you want the audit trail).

### 4.5 ISO 8601 partial-date imputation rules

| DTC input | Length | `imputation="first"` → `<prefix>DT` | DTF flag | `imputation="last"` → `<prefix>DT` | DTF flag |
|---|---|---|---|---|---|
| `"2014-01-02"` | 10 | 2014-01-02 | `NA` (no imputation) | 2014-01-02 | `NA` |
| `"2014-01"` | 7 | 2014-01-01 | `"D"` | 2014-01-31 | `"D"` |
| `"2014"` | 4 | 2014-01-01 | `"M"` | 2014-12-31 | `"M"` |
| `"--"` or `""` | 0 | `NA` | `NA` | `NA` | `NA` |
| `"2014-02"` (last) | 7 | 2014-02-01 | `"D"` | 2014-02-28 *(or 29 in leap year)* | `"D"` |

Note that month-end imputation is **calendar-aware** — February 2014 resolves to the 28th, February 2016 to the 29th. SAS's `INTNX('MONTH', d, 0, 'E')` does the same thing.

### 4.6 SAS analog mental model

> **SAS analog mental model:** `derive_vars_dt()` is your PROC equivalent of `INPUT(DTC, E8601DA10.)` + a length-check macro for partial dates + an automatic flag variable, all in one call. In SAS you would typically write a `%macro impute_date(dtc=, prefix=)` that does the length test and conditional `INTNX`. admiral ships that macro as a versioned, tested function.

### 4.7 Common pitfalls vs SAS `INPUT()`

| Pitfall | SAS behavior | admiral behavior |
|---|---|---|
| Partial date `"2014-01"` | `INPUT("2014-01", E8601DA10.)` → `.` (missing) | With `date_imputation="first"` → `2014-01-01` |
| Empty/missing DTC | `INPUT("", E8601DA10.)` → `.` | Returns `NA` regardless of `date_imputation` |
| Trailing `T` (datetime form) `"2014-01-02T08:30"` | `INPUT()` with `DATE` informat → `.` | admiral strips the time portion gracefully |
| Day=15 imputation (common SAS shop convention) | Custom macro: `date('15Jan2014')` | admiral does **not** support `"mid"` imputation; if you need day-15 you must override post hoc |

> **Note:** Some SAS shops use day-15 imputation for partial dates (the "middle of the month" convention) on the argument that it minimizes mean absolute imputation error. Admiral's built-in choices are only `"first"` and `"last"`. If your SAP mandates day-15, derive with `"first"` then overwrite: `mutate(RFXSTDT = if_else(RFXSTDTF == "D", RFXSTDT + 14, RFXSTDT))`.

```{r derive-dt-demo}
# Pre-process DM to get a numeric date from RFXSTDTC
dm_dates <- dm |>
  derive_vars_dt(
    new_vars_prefix  = "RFXST",
    dtc              = RFXSTDTC,
    date_imputation  = "first"
  )

dm_dates |>
  select(USUBJID, RFXSTDTC, RFXSTDT, RFXSTDTF) |>
  mutate(RFXSTDT_human = as.Date(RFXSTDT, origin = "1970-01-01"))
```

```{r derive-dt-partial}
# With a partial date — imputation kicks in
partial_dm <- tibble(
  USUBJID  = c("S001", "S002"),
  STARTDTC = c("2014-07", "2014-07-15")   # S001 has no day
)

partial_dm |>
  derive_vars_dt(
    new_vars_prefix = "START",
    dtc = STARTDTC,
    date_imputation = "first"
  ) |>
  mutate(STARTDT_human = as.Date(STARTDT, origin = "1970-01-01"))
```

> **Read the output:** S001 had `STARTDTC = "2014-07"` (year+month, no day). With `date_imputation = "first"`, the resolved `STARTDT` is `2014-07-01` and `STARTDTF` is `"D"` (flag = "day was imputed"). S002 had a full DTC, so its `STARTDTF` is `NA` (no imputation needed). The flag column is your audit trail — never throw it away.

### 4.8 Exercise: Derive a date with "last" imputation

Impute the same partial date dataset using `date_imputation = "last"`. Subject S001's date should now resolve to the last day of July 2014.

```{r derive-dt-last, exercise=TRUE, exercise.lines=12}
partial_dm2 <- tibble(
  USUBJID  = c("S001", "S002"),
  STARTDTC = c("2014-07", "2014-07-15")
)

partial_dm2 |>
  derive_vars_dt(
    new_vars_prefix = "START",
    dtc             = STARTDTC,
    date_imputation = ___
  ) |>
  mutate(STARTDT_human = as.Date(STARTDT, origin = "1970-01-01"))
```

```{r derive-dt-last-hint-1}
# date_imputation = "last"
# Expected: S001 → 2014-07-31
```

```{r derive-dt-last-solution}
partial_dm2 <- tibble(
  USUBJID  = c("S001", "S002"),
  STARTDTC = c("2014-07", "2014-07-15")
)

partial_dm2 |>
  derive_vars_dt(
    new_vars_prefix = "START",
    dtc             = STARTDTC,
    date_imputation = "last"
  ) |>
  mutate(STARTDT_human = as.Date(STARTDT, origin = "1970-01-01"))
```

### 4.9 Exercise: Derive numeric dates from real pharmaversesdtm DM

`dm_real` contains `RFXSTDTC` — the reference date for first treatment exposure. Derive `RFXSTDT` from it, then check how many subjects had partial dates that triggered the imputation flag.

```{r derive-dt-real, exercise=TRUE, exercise.lines=18}
# Preview the raw RFXSTDTC values
cat("RFXSTDTC sample (first 8):\n")
cat(paste(head(dm_real$RFXSTDTC, 8), collapse = ", "), "\n\n")

# How many look partial (< 10 chars = missing day or month)?
partial_n <- sum(nchar(dm_real$RFXSTDTC) < 10 & !is.na(dm_real$RFXSTDTC))
cat("Partial RFXSTDTC values:", partial_n, "\n\n")

# Derive RFXSTDT (numeric date) with first-day imputation
dm_real_dates <- dm_real |>
  derive_vars_dt(
    dtc             = ___,
    new_vars_prefix = "RFXST",
    date_imputation = "first"
  )

# How many imputation flags were set?
cat("RFXSTDTF imputation flag counts:\n")
table(dm_real_dates$RFXSTDTF, useNA = "ifany")
```

```{r derive-dt-real-hint-1}
# dtc = RFXSTDTC
# A DTF flag of NA means the date was complete — no imputation needed.
# A flag of "D" means the day was missing and imputed.
```

```{r derive-dt-real-solution}
cat("RFXSTDTC sample (first 8):\n")
cat(paste(head(dm_real$RFXSTDTC, 8), collapse = ", "), "\n\n")

partial_n <- sum(nchar(dm_real$RFXSTDTC) < 10 & !is.na(dm_real$RFXSTDTC))
cat("Partial RFXSTDTC values:", partial_n, "\n\n")

dm_real_dates <- dm_real |>
  derive_vars_dt(
    dtc             = RFXSTDTC,
    new_vars_prefix = "RFXST",
    date_imputation = "first"
  )

cat("RFXSTDTF imputation flag counts:\n")
table(dm_real_dates$RFXSTDTF, useNA = "ifany")
```

> **Read the output:** the `table()` of `RFXSTDTF` tells you the imputation profile of the dataset — how many rows had complete DTCs (NA flag), how many had day-only imputation ("D"), etc. In a real ADaM build this is the table you paste into your derivation listings as evidence that you handled partial dates correctly.

### 4.10 Exercise: Compare "first" vs "last" on the same partial DTC

Apply both imputation modes to the same partial DTC vector and inspect the difference. This is the test you would run when your SAP changes between drafts.

```{r derive-dt-firstlast, exercise=TRUE, exercise.lines=20}
test_partial <- tibble(
  USUBJID = c("S001", "S002", "S003", "S004"),
  ASTDTC  = c("2014-02", "2014",   "2014-11-15", "2016-02")
  #            ^ no day  ^ no mo   ^ complete    ^ feb in leap year
)

# Apply "first" imputation
res_first <- test_partial |>
  derive_vars_dt(
    dtc             = ASTDTC,
    new_vars_prefix = "AST",
    date_imputation = ___
  ) |>
  mutate(ASTDT_human = as.Date(ASTDT, origin = "1970-01-01")) |>
  select(USUBJID, ASTDTC, ASTDT_first = ASTDT_human, ASTDTF_first = ASTDTF)

# Apply "last" imputation
res_last <- test_partial |>
  derive_vars_dt(
    dtc             = ASTDTC,
    new_vars_prefix = "AST",
    date_imputation = ___
  ) |>
  mutate(ASTDT_human = as.Date(ASTDT, origin = "1970-01-01")) |>
  select(USUBJID, ASTDT_last = ASTDT_human, ASTDTF_last = ASTDTF)

left_join(res_first, res_last, by = "USUBJID")
```

```{r derive-dt-firstlast-hint-1}
# Set date_imputation = "first" in the first call,
# and date_imputation = "last" in the second.
```

```{r derive-dt-firstlast-hint-2}
# "first" pads to the earliest valid day (01-01 or month-01).
# "last" pads to the latest valid day (12-31 or month-end, leap-aware).
```

```{r derive-dt-firstlast-solution}
test_partial <- tibble(
  USUBJID = c("S001", "S002", "S003", "S004"),
  ASTDTC  = c("2014-02", "2014", "2014-11-15", "2016-02")
)

res_first <- test_partial |>
  derive_vars_dt(
    dtc             = ASTDTC,
    new_vars_prefix = "AST",
    date_imputation = "first"
  ) |>
  mutate(ASTDT_human = as.Date(ASTDT, origin = "1970-01-01")) |>
  select(USUBJID, ASTDTC, ASTDT_first = ASTDT_human, ASTDTF_first = ASTDTF)

res_last <- test_partial |>
  derive_vars_dt(
    dtc             = ASTDTC,
    new_vars_prefix = "AST",
    date_imputation = "last"
  ) |>
  mutate(ASTDT_human = as.Date(ASTDT, origin = "1970-01-01")) |>
  select(USUBJID, ASTDT_last = ASTDT_human, ASTDTF_last = ASTDTF)

left_join(res_first, res_last, by = "USUBJID")
```

> **Read the output:** S003 has identical first and last dates (the DTC was complete and the flag is NA). S001 (`2014-02`) resolves to `2014-02-01` vs `2014-02-28`. S004 (`2016-02`, a leap year) resolves to `2016-02-01` vs `2016-02-29` — proof that admiral's month-end imputation is calendar-aware. S002 (`2014`) shows month+day imputation, with flag `"M"`.

### 4.11 Knowledge check

```{r quiz-dt, echo=FALSE}
quiz(
  question(
    "A subject's `RFSTDTC` is `'2014'`. With `date_imputation = 'last'`, what does `RFSTDT` become?",
    answer("2014-01-01"),
    answer("2014-12-31", correct = TRUE),
    answer("NA — admiral cannot impute year-only"),
    random_answer_order = TRUE
  ),
  question(
    "A SAS programmer writes `RFSTDT = INPUT(RFSTDTC, E8601DA10.);` and the DTC is `'2014-02'`. SAS returns `.` (missing). admiral's `derive_vars_dt(date_imputation='first')` returns:",
    answer("Also missing"),
    answer("2014-02-01 with DTF = 'D'", correct = TRUE),
    answer("An error"),
    random_answer_order = TRUE
  ),
  question(
    "The `DTF` flag value `'M'` means:",
    answer("The month-only portion was wrong"),
    answer("Both the month and the day had to be imputed (DTC was year-only)", correct = TRUE),
    answer("The date is missing"),
    random_answer_order = TRUE
  ),
  question(
    "You need to impute partial dates to day-15 (the 'middle of the month' SAS convention). Does admiral support this directly?",
    answer("Yes, via `date_imputation = 'mid'`"),
    answer("No — only 'first' and 'last' are built in; you must override post hoc with mutate()", correct = TRUE),
    answer("Yes, but only for SAS-compatible XPT inputs"),
    random_answer_order = TRUE
  ),
  question(
    "After `derive_vars_dt(new_vars_prefix = 'AST', ...)`, which two columns will be present?",
    answer("ASTDT and ASTDY"),
    answer("ASTDT and ASTDTF", correct = TRUE),
    answer("ASTDT and ASTDTC"),
    random_answer_order = TRUE
  )
)
```

---

## 5 — `derive_vars_dtm()`: DTC → numeric Datetime

For treatment-window calculations you often need datetimes (not just dates), because two records can share a date but have different times of day. `derive_vars_dtm()` is the sister function of `derive_vars_dt()` — same idea, but for full ISO 8601 datetime strings (`"2014-01-02T08:30:00"`).

### 5.1 What it produces

| Output variable | Type | Content |
|---|---|---|
| `<prefix>DTM` | numeric (seconds since 1970-01-01 UTC) | Fully resolved analysis datetime |
| `<prefix>DTF` | character | Date imputation flag — `"D"`, `"M"`, `"Y"`, or `NA` |
| `<prefix>TMF` | character | Time imputation flag — `"H"` (hour), `"M"` (minute), `"S"` (second), or `NA` |

### 5.2 SAS analog comparison

| admiral function | What it does | Closest SAS analog | SAS code sketch |
|---|---|---|---|
| `derive_vars_dtm()` | DTC string → numeric datetime + date/time flags | `INPUT()` with `E8601DT19.` informat, plus length checks | `EXSTDTM = INPUT(EXSTDTC, E8601DT19.);` |
| `time_imputation = "first"` | Pad missing time to `00:00:00` | Macro: append `'T00:00:00'` if length=10 | `if length(EXSTDTC)=10 then EXSTDTM = DHMS(INPUT(EXSTDTC,YYMMDD10.),0,0,0);` |
| `time_imputation = "last"` | Pad missing time to `23:59:59` | Macro: append `'T23:59:59'` if length=10 | `if length(EXSTDTC)=10 then EXSTDTM = DHMS(INPUT(EXSTDTC,YYMMDD10.),23,59,59);` |
| `derive_vars_dtm_to_dt()` | Strip time → date | `DATEPART(dtm)` SAS function | `EXSTDT = DATEPART(EXSTDTM);` |

### 5.3 ASCII side-by-side

```
   /* SAS — INPUT then pad time */          |   # admiral — one call
   data ex;                                 |   ex |>
     set ex;                                |     derive_vars_dtm(
     if length(EXSTDTC)>=19 then            |       dtc             = EXSTDTC,
       EXSTDTM = INPUT(EXSTDTC,             |       new_vars_prefix = "EXST",
                       E8601DT19.);         |       date_imputation = "first",
     else if length(EXSTDTC)=10 then        |       time_imputation = "first"
       EXSTDTM = DHMS(INPUT(EXSTDTC,        |     )
                            YYMMDD10.),     |
                      0,0,0);               |
     if length(EXSTDTC)<19 then             |
       EXSTTMF = "H";                       |
   run;                                     |
```

### 5.4 Anatomy of the call

```r
derive_vars_dtm(
  dataset         = ex,         # input dataset
  dtc             = EXSTDTC,    # ISO 8601 datetime column
  new_vars_prefix = "EXST",     # stem; produces EXSTDTM, EXSTDTF, EXSTTMF
  date_imputation = "first",    # how to handle partial date portion
  time_imputation = "first",    # how to handle missing time portion
  flag_imputation = "auto"      # emit DTF/TMF audit flags
)
```

- **`time_imputation`** is the new argument vs `derive_vars_dt()`. `"first"` pads to `00:00:00`. `"last"` pads to `23:59:59`. This matters most for end-of-day datetimes (e.g. AE end time, treatment end time).
- The output is **numeric seconds since 1970-01-01 UTC** — the standard R `POSIXct` representation. Display with `as.POSIXct(x, origin = "1970-01-01", tz = "UTC")`.
- Three flag columns instead of two: the extra `TMF` flag distinguishes "time was imputed" from "date was imputed".

### 5.5 Time zone discussion

Admiral assumes **UTC** internally. There is no `tz` argument on `derive_vars_dtm()`. This is by design — clinical data in CDISC SDTM is recorded with the assumption that DTC strings are local clock time of the site, and the analysis collapses everything to a study-level convention without DST conversions.

| System | Default time zone |
|---|---|
| SAS | Local server time, modified by `OPTIONS TIMEZONE='...'` |
| admiral | UTC always |
| `lubridate::ymd_hms()` | UTC unless `tz=` is passed |

> **Practical implication:** when you derive a treatment-emergent flag with `TRTSDTM <= AESTDTM`, both sides are UTC numerics. Comparison is exact-second precision. There is no daylight saving time gotcha because nothing is local.

### 5.6 SAS analog mental model

> **SAS analog mental model:** `derive_vars_dtm()` is `INPUT(DTC, E8601DT19.)` plus a length-conditional `DHMS()` to pad missing time, plus two audit flags, all bundled in one call. The output is a *numeric seconds-since-epoch* analog of SAS's `DATETIME20.` format.

### 5.7 Datetime → date conversion (`derive_vars_dtm_to_dt()`)

CDISC ADaM convention is that every datetime variable has a companion date variable — `ASTDTM` always pairs with `ASTDT`. You derive the date by stripping the time portion.

| admiral | SAS analog |
|---|---|
| `derive_vars_dtm_to_dt(source_vars = exprs(EXSTDTM, EXENDTM))` | `EXSTDT = DATEPART(EXSTDTM); EXENDT = DATEPART(EXENDTM);` |

### 5.8 The "no Day 0" gotcha is not here

A subtle point worth stating clearly: `derive_vars_dtm()` does **not** apply the no-Day-0 rule. It is a pure DTC-to-datetime converter. The +1-day, no-Day-0 logic only kicks in at `derive_vars_dy()` (the study day function in the next section). Mixing these up is one of the most common SAS-to-admiral translation errors.

```{r derive-dtm-demo}
# Pre-process EX with start and end datetimes
ex_ext <- ex |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first",
    time_imputation = "first"   # impute missing time to 00:00:00
  ) |>
  derive_vars_dtm(
    dtc             = EXENDTC,
    new_vars_prefix = "EXEN",
    date_imputation = "last",
    time_imputation = "last"    # impute missing time to 23:59:59
  )

ex_ext |>
  select(USUBJID, EXSTDTC, EXSTDTM, EXSTTMF, EXENDTC, EXENDTM)
```

> **Read the output:** `EXSTDTM` and `EXENDTM` are numeric seconds-since-1970. `EXSTTMF` is the time-imputation flag — it tells you "the start time was missing in the raw DTC, so we filled 00:00:00 in." The DTC strings in the input were date-only (`"2014-01-02"` not `"2014-01-02T08:30"`), so every row has a time flag set.

```{r dtm-to-dt-demo}
# Convert datetimes back to dates (strip time component)
ex_ext2 <- ex_ext |>
  derive_vars_dtm_to_dt(source_vars = exprs(EXSTDTM, EXENDTM))

ex_ext2 |>
  select(USUBJID, EXSTDTM, EXSTDT, EXENDTM, EXENDT) |>
  head(4)
```

> **Read the output:** notice that `EXSTDT` is derived directly from `EXSTDTM` by stripping the time portion (dividing by 86400 seconds per day and truncating to a whole day). This is the admiral equivalent of SAS's `DATEPART()` function.

### 5.9 Exercise: Derive a datetime with last-time imputation

When an AE end date is recorded as a date-only DTC (`"2014-03-15"`), the analysis should treat it as "the AE could have ended any time up to end-of-day on that date." Use `time_imputation = "last"`.

```{r derive-dtm-last, exercise=TRUE, exercise.lines=18}
mock_ae <- tibble(
  USUBJID  = c("S001", "S002", "S003"),
  AEENDTC  = c("2014-03-15",          # date only
               "2014-03-15T18:45:00", # full datetime
               "2014-03")             # partial (year-month only)
)

ae_end <- mock_ae |>
  derive_vars_dtm(
    dtc             = AEENDTC,
    new_vars_prefix = "AEEN",
    date_imputation = ___,    # for "2014-03", pad day to month-end
    time_imputation = ___     # for date-only inputs, pad time to 23:59:59
  )

ae_end |>
  mutate(AEENDTM_human = as.POSIXct(AEENDTM, origin = "1970-01-01", tz = "UTC")) |>
  select(USUBJID, AEENDTC, AEENDTM_human, AEENDTF, AEENTMF)
```

```{r derive-dtm-last-hint-1}
# AE end semantics: "last possible moment" interpretation.
# date_imputation = "last" pads partial date to month-end.
# time_imputation = "last" pads missing time to 23:59:59.
```

```{r derive-dtm-last-hint-2}
# Both arguments should be "last".
```

```{r derive-dtm-last-solution}
mock_ae <- tibble(
  USUBJID  = c("S001", "S002", "S003"),
  AEENDTC  = c("2014-03-15",
               "2014-03-15T18:45:00",
               "2014-03")
)

ae_end <- mock_ae |>
  derive_vars_dtm(
    dtc             = AEENDTC,
    new_vars_prefix = "AEEN",
    date_imputation = "last",
    time_imputation = "last"
  )

ae_end |>
  mutate(AEENDTM_human = as.POSIXct(AEENDTM, origin = "1970-01-01", tz = "UTC")) |>
  select(USUBJID, AEENDTC, AEENDTM_human, AEENDTF, AEENTMF)
```

> **Read the output:** S001 had a date-only DTC — the resolved datetime is `2014-03-15 23:59:59` (time padded to end of day) with `AEENTMF = "H"`. S002 had a complete DTC and both flags are NA. S003 had only year-month — the date was padded to `2014-03-31 23:59:59` with `AEENDTF = "D"` and `AEENTMF = "H"`.

### 5.10 Knowledge check

```{r quiz-dtm, echo=FALSE}
quiz(
  question(
    "What does the `TMF` flag track that `DTF` does not?",
    answer("Time zone changes"),
    answer("Imputation of the time portion (hour/minute/second), as distinct from the date portion", correct = TRUE),
    answer("Whether the datetime is in UTC"),
    random_answer_order = TRUE
  ),
  question(
    "A SAS programmer is used to setting OPTIONS TIMEZONE='America/New_York'. In admiral:",
    answer("There is an equivalent `set_admiral_timezone()` function"),
    answer("All datetimes are UTC by design; admiral has no timezone option", correct = TRUE),
    answer("Pass tz='America/New_York' to derive_vars_dtm()"),
    random_answer_order = TRUE
  ),
  question(
    "After `derive_vars_dtm()` with `new_vars_prefix = 'EXST'`, how many new columns are added?",
    answer("One (EXSTDTM)"),
    answer("Three (EXSTDTM, EXSTDTF, EXSTTMF)", correct = TRUE),
    answer("Two (EXSTDTM and EXSTDT)"),
    random_answer_order = TRUE
  ),
  question(
    "`derive_vars_dtm_to_dt()` is the admiral analog of which SAS function?",
    answer("INTNX('day', dtm, 0)"),
    answer("DATEPART(dtm)", correct = TRUE),
    answer("INPUT(dtm, DATE9.)"),
    random_answer_order = TRUE
  )
)
```

---

## 6 — `derive_vars_dy()`: study day (no Day 0)

CDISC ADaM convention: the day of the reference event is **Day 1**. The day before is **Day -1**. **There is no Day 0**. This rule trips up almost every SAS programmer's first ADaM build because every SAS date function (`INTCK`, simple subtraction) gives you a continuous integer with zero in the middle.

### 6.1 The "no Day 0" rule, in one sentence

```
   raw_diff = event_date - reference_date
   if raw_diff >= 0 then ADY = raw_diff + 1
   else                    ADY = raw_diff
```

The `+1` shift applies only on the **positive** side of the reference. The negative side is left alone. The net effect is that Day 0 is skipped over.

### 6.2 SAS analog comparison

| admiral function | What it does | Closest SAS analog | SAS code sketch |
|---|---|---|---|
| `derive_vars_dy()` | Compute study day from reference date, with no Day 0 | A DATA step with `INTCK('day', ref, dt)` plus a sign-aware +1 | `if dt >= ref then ADY = dt - ref + 1; else ADY = dt - ref;` |

### 6.3 Side-by-side

```
   /* SAS — typical "no Day 0" pattern */     |   # admiral
   data lb;                                   |   lb |>
     set lb;                                  |     derive_vars_dy(
     if ADT ne . and TRTSDT ne . then do;     |       reference_date = TRTSDT,
       if ADT >= TRTSDT then                  |       source_vars    = exprs(ADT)
         ADY = ADT - TRTSDT + 1;              |     )
       else                                   |
         ADY = ADT - TRTSDT;                  |
     end;                                     |
   run;                                       |
```

### 6.4 Anatomy of the call

```r
derive_vars_dy(
  dataset        = lb,                  # input dataset (rows where you want ADY)
  reference_date = TRTSDT,              # the Day 1 anchor (typically TRTSDT or RFSTDT)
  source_vars    = exprs(ADT, AENDT)    # one or more dates to convert to study days
)
```

- **`reference_date`** is the anchor day — Day 1. Almost always `TRTSDT` for treatment-relative analyses, or `RFSTDT` for randomization-relative.
- **`source_vars`** accepts one or more date columns. For each `X_DT` in the list, admiral creates `X_DY`. So `exprs(ADT, AENDT)` produces both `ADY` and `AENDY` in one call.
- Both inputs must be **numeric dates** (output of `derive_vars_dt()`). Passing a character DTC will error.

### 6.5 Worked table

| Reference date | Event date | raw_diff | ADY (after no-Day-0) |
|---|---|---|---|
| 2014-01-10 | 2014-01-15 | 5 | **6** (+1 shift) |
| 2014-01-10 | 2014-01-10 | 0 | **1** (+1 shift — the reference day itself is Day 1) |
| 2014-01-10 | 2014-01-09 | -1 | **-1** (no shift on negative side) |
| 2014-01-10 | 2014-01-08 | -2 | **-2** |
| 2014-01-10 | NA | NA | **NA** |

> **SAS analog mental model:** `derive_vars_dy()` is the function form of the SAS one-liner `ADY = ADT - TRTSDT + (ADT >= TRTSDT);` — same logic, but it works on a vector of dates in one call and handles NAs without warnings.

### 6.6 Why "no Day 0"?

The convention dates back to clinical pharmacology — when reporting an AE that occurred "two days after dosing started," there's no useful concept of "Day 0 = the day of dosing AND the day before dosing." Either you count the dose day as Day 1 (and shift everything after) or you make it Day 0 (and have to explain why screening Day -1 is right next to dosing Day 0). The CDISC convention chose the former. Once you internalize it, all your tables and listings line up.

### 6.7 Common pitfall: forgetting to merge the reference date first

`derive_vars_dy()` looks up `reference_date` *from the dataset you pass in*. If your LB dataset doesn't already have a TRTSDT column, you must merge it on first (typically with `derive_vars_merged()` or a plain `left_join()` from ADSL).

```{r study-day-demo}
# First: derive numeric analysis dates for LB
lb_dates <- lb |>
  derive_vars_dt(dtc = LBDTC, new_vars_prefix = "A")

# Then merge TRTSDT onto LB (from dm) and derive ADY
lb_with_ref <- lb_dates |>
  left_join(
    dm |>
      derive_vars_dt(dtc = RFXSTDTC, new_vars_prefix = "TRTS") |>
      select(USUBJID, TRTSDT),
    by = "USUBJID"
  )

lb_with_day <- lb_with_ref |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ADT)
  )

lb_with_day |>
  mutate(
    ADT_human    = as.Date(ADT,    origin = "1970-01-01"),
    TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")
  ) |>
  select(USUBJID, ADT_human, TRTSDT_human, ADY)
```

> **Read the output:** the `ADY` column is the integer study day. Subject `01-701-1015` has its first LB observation on the same day as TRTSDT (Day 1, not Day 0). Subject `01-701-1023` has the same property. Walk through the math: `ADT - TRTSDT = 0`, then `+1` because the diff is `>= 0`, so `ADY = 1`. This is the rule visible in the data.

### 6.8 Exercise: Explain the "no Day 0" convention

```{r quiz-study-day, echo=FALSE}
quiz(
  question(
    "A subject's treatment starts on 2014-01-02. An AE occurs on 2014-01-02. What is the study day of the AE?",
    answer("Day 0"),
    answer("Day 1 — the reference date itself is always Day 1", correct = TRUE),
    answer("Day -1"),
    random_answer_order = TRUE
  ),
  question(
    "An AE occurs one day BEFORE the reference date (2014-01-01, with ref = 2014-01-02). The study day is:",
    answer("Day 0"),
    answer("Day -1 — there is no Day 0; the day before Day 1 is Day -1", correct = TRUE),
    answer("Day -2"),
    random_answer_order = TRUE
  )
)
```

### 6.9 Exercise: Derive study day for an EX dataset

```{r study-day-ex, exercise=TRUE, exercise.lines=18}
# Add numeric date to EX
ex_with_dates <- ex |>
  derive_vars_dt(dtc = EXSTDTC, new_vars_prefix = "A")

# Merge TRTSDT from DM (first exposure date per subject)
trtsdt_lookup <- dm |>
  derive_vars_dt(dtc = RFXSTDTC, new_vars_prefix = "TRTS") |>
  select(USUBJID, TRTSDT)

ex_study_day <- ex_with_dates |>
  left_join(trtsdt_lookup, by = "USUBJID") |>
  derive_vars_dy(
    reference_date = ___,
    source_vars    = exprs(___)
  )

ex_study_day |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01"),
         TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")) |>
  select(USUBJID, EXSTDTC, ADT_human, TRTSDT_human, ADY)
```

```{r study-day-ex-hint-1}
# reference_date = TRTSDT
# source_vars = exprs(ADT)
```

```{r study-day-ex-solution}
ex_with_dates <- ex |>
  derive_vars_dt(dtc = EXSTDTC, new_vars_prefix = "A")

trtsdt_lookup <- dm |>
  derive_vars_dt(dtc = RFXSTDTC, new_vars_prefix = "TRTS") |>
  select(USUBJID, TRTSDT)

ex_study_day <- ex_with_dates |>
  left_join(trtsdt_lookup, by = "USUBJID") |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ADT)
  )

ex_study_day |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01"),
         TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")) |>
  select(USUBJID, EXSTDTC, ADT_human, TRTSDT_human, ADY)
```

> **Read the output:** every `ADY` value is `>= 1` (since these are dosing records on or after first treatment). If you ran the same derivation on an AE that happened before treatment, you would see negative ADY values — and Day -1 would jump straight to Day 1, skipping Day 0.

### 6.10 Exercise: Two source vars in one call

`derive_vars_dy()` can compute several study-day columns in one shot — useful for AEs where you need both `ASTDY` (start) and `AENDY` (end).

```{r study-day-two, exercise=TRUE, exercise.lines=20}
mock_ae <- tibble(
  USUBJID = c("S001", "S001", "S002"),
  AESTDTC = c("2014-01-15", "2014-02-01", "2012-08-10"),
  AEENDTC = c("2014-01-20", "2014-02-05", "2012-08-12"),
  TRTSDT  = as.numeric(as.Date(c("2014-01-10", "2014-01-10", "2012-08-05")))
)

ae_with_dy <- mock_ae |>
  derive_vars_dt(dtc = AESTDTC, new_vars_prefix = "AST") |>
  derive_vars_dt(dtc = AEENDTC, new_vars_prefix = "AEN") |>
  derive_vars_dy(
    reference_date = ___,
    source_vars    = exprs(___, ___)
  )

ae_with_dy |>
  mutate(
    ASTDT_h = as.Date(ASTDT, origin = "1970-01-01"),
    AENDT_h = as.Date(AENDT, origin = "1970-01-01"),
    TRTSDT_h = as.Date(TRTSDT, origin = "1970-01-01")
  ) |>
  select(USUBJID, TRTSDT_h, ASTDT_h, ASTDY, AENDT_h, AENDY)
```

```{r study-day-two-hint-1}
# reference_date = TRTSDT
# Source variables are the analysis date columns we just derived: ASTDT and AENDT.
```

```{r study-day-two-hint-2}
# source_vars = exprs(ASTDT, AENDT)
# Output columns: ASTDY and AENDY (the trailing DT becomes DY).
```

```{r study-day-two-solution}
mock_ae <- tibble(
  USUBJID = c("S001", "S001", "S002"),
  AESTDTC = c("2014-01-15", "2014-02-01", "2012-08-10"),
  AEENDTC = c("2014-01-20", "2014-02-05", "2012-08-12"),
  TRTSDT  = as.numeric(as.Date(c("2014-01-10", "2014-01-10", "2012-08-05")))
)

ae_with_dy <- mock_ae |>
  derive_vars_dt(dtc = AESTDTC, new_vars_prefix = "AST") |>
  derive_vars_dt(dtc = AEENDTC, new_vars_prefix = "AEN") |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ASTDT, AENDT)
  )

ae_with_dy |>
  mutate(
    ASTDT_h = as.Date(ASTDT, origin = "1970-01-01"),
    AENDT_h = as.Date(AENDT, origin = "1970-01-01"),
    TRTSDT_h = as.Date(TRTSDT, origin = "1970-01-01")
  ) |>
  select(USUBJID, TRTSDT_h, ASTDT_h, ASTDY, AENDT_h, AENDY)
```

> **Read the output:** for S001's first AE, `ASTDY = 6` (event on 2014-01-15, ref on 2014-01-10, so raw_diff=5, +1 = 6) and `AENDY = 11`. The "+1 on positives" pattern is visible. Verify by hand that no row has ADY=0.

### 6.11 Knowledge check

```{r quiz-dy, echo=FALSE}
quiz(
  question(
    "A reference date is 2014-01-10. An AE start date is 2014-01-10 (same day). What is ASTDY?",
    answer("0"),
    answer("1 — the reference day is Day 1", correct = TRUE),
    answer("NA"),
    random_answer_order = TRUE
  ),
  question(
    "A reference date is 2014-01-10. A screening LB was drawn on 2014-01-09. What is ADY?",
    answer("-1 — no +1 shift on the negative side", correct = TRUE),
    answer("0"),
    answer("-2 (because of the no-Day-0 rule)"),
    random_answer_order = TRUE
  ),
  question(
    "What is the SAS one-liner equivalent of `derive_vars_dy()`?",
    answer("ADY = ADT - TRTSDT"),
    answer("ADY = ADT - TRTSDT + (ADT >= TRTSDT)", correct = TRUE),
    answer("ADY = INTCK('day', TRTSDT, ADT)"),
    random_answer_order = TRUE
  ),
  question(
    "You call `derive_vars_dy(reference_date = TRTSDT)` but get an error that TRTSDT is not found. The likely cause:",
    answer("TRTSDT must be character"),
    answer("TRTSDT is not yet on the input dataset — merge it on first (e.g., with derive_vars_merged() from ADSL)", correct = TRUE),
    answer("The pipe operator was used incorrectly"),
    random_answer_order = TRUE
  )
)
```

---

## 7 — `derive_vars_merged()`: the most important function

`derive_vars_merged()` is the workhorse of every admiral pipeline. You will call it 10-30 times per ADaM build. Understand it deeply and most of admiral becomes a remix of patterns you already know.

### 7.1 What it does in one sentence

Left-join one variable (or several) from a *source* dataset onto a *base* dataset, optionally filtering the source first, optionally sorting and picking the first or last row per by-group, optionally renaming on arrival, and optionally filling defaults for unmatched rows.

That single sentence packs five SQL primitives — SUBSET, JOIN, GROUP, ORDER, LIMIT — into one call.

### 7.2 SAS analog comparison

| `derive_vars_merged()` argument | What it controls          | Closest SAS counterpart                                |
|---------------------------------|---------------------------|--------------------------------------------------------|
| `dataset`                       | Left-hand (base) table    | The dataset on `DATA=` of the outer DATA step          |
| `dataset_add`                   | Right-hand (source) table | The second dataset in a `MERGE a b; BY USUBJID;`       |
| `by_vars`                       | Join keys                 | `BY USUBJID;` after `MERGE`                            |
| `filter_add`                    | Pre-merge subset          | `WHERE` clause applied to the right side               |
| `order`                         | Tie-breaker sort          | `PROC SORT BY` ahead of `FIRST.` / `LAST.` processing  |
| `mode = "first" \| "last"`      | Pick one row per group    | `IF FIRST.USUBJID;` / `IF LAST.USUBJID;`               |
| `new_vars`                      | Rename-on-arrival         | `RENAME=(EXSTDTM=TRTSDTM)` on the source `MERGE`       |
| `missing_values`                | Default for unmatched     | `COALESCE(TRTSDTM, '01JAN9999'd)` after the merge      |
| `check_type`                    | Many-to-one duplication   | `PROC SORT NODUPKEY` + a downstream existence check    |

Where a single SAS DATA step would chain together a `PROC SORT`, a `PROC SQL` subquery, a `MERGE`, and a `WHERE` plus `RETAIN`, `derive_vars_merged()` declares the whole intent in one call. The mental model "build the right-hand side, then left-join" maps very cleanly to PROC SQL.

### 7.3 ASCII side-by-side: PROC SQL vs `derive_vars_merged()`

```
   proc sql;                              |   adsl_base |>
     create table adsl_trt as             |     derive_vars_merged(
     select a.*,                          |       dataset_add = ex_dtm,
            min(b.EXSTDTM) as TRTSDTM     |       filter_add  =
       from adsl_base as a                |         EXDOSE > 0 |
       left join ex_dtm as b              |         (EXDOSE == 0 &
         on a.STUDYID = b.STUDYID         |          str_detect(EXTRT, "PLACEBO")),
        and a.USUBJID = b.USUBJID         |       new_vars    =
        and (b.EXDOSE > 0                 |         exprs(TRTSDTM = EXSTDTM),
             or (b.EXDOSE = 0             |       order       =
                 and b.EXTRT              |         exprs(EXSTDTM, EXSEQ),
                 contains "PLACEBO"))     |       mode        = "first",
      group by a.STUDYID, a.USUBJID;      |       by_vars     =
   quit;                                  |         exprs(STUDYID, USUBJID))
```

Both produce identical results. The PROC SQL form crams JOIN, WHERE, and GROUP-BY-MIN into one query; the admiral form factors them into named arguments that read like English. The admiral form is also auditable: each argument can be inspected, validated, and explained independently.

### 7.4 ASCII side-by-side: DATA-step MERGE vs `derive_vars_merged()`

```
   proc sort data=ex_dtm out=ex_sorted;   |   (admiral does this internally
     where EXDOSE > 0                     |    once filter_add and order
        or (EXDOSE = 0 and                |    are provided)
            index(EXTRT, "PLACEBO"));     |
     by STUDYID USUBJID EXSTDTM EXSEQ;    |
   run;                                   |
                                          |
   data adsl_trt;                         |   adsl_trt <- adsl_base |>
     merge adsl_base (in=a)               |     derive_vars_merged(
           ex_sorted (in=b                |       dataset_add = ex_dtm,
                      keep=STUDYID        |       filter_add  = ...,
                           USUBJID        |       new_vars    =
                           EXSTDTM        |         exprs(TRTSDTM = EXSTDTM),
                      rename=             |       order       =
                       (EXSTDTM=TRTSDTM));|         exprs(EXSTDTM, EXSEQ),
     by STUDYID USUBJID;                  |       mode        = "first",
     if a;                                |       by_vars     =
     if first.USUBJID;                    |         exprs(STUDYID, USUBJID))
   run;                                   |
```

The SAS workflow needs a separate PROC SORT, a careful `IF a;` to keep all left rows, and an `IF FIRST.USUBJID;` to pick one row per subject. `derive_vars_merged()` packs all four steps into one call.

### 7.5 Anatomy of the call

```r
adsl_base |>
  derive_vars_merged(
    dataset_add = ex_dtm,                    # right-hand source (after upstream prep)
    by_vars     = exprs(STUDYID, USUBJID),   # join keys — base × source
    filter_add  = EXDOSE > 0 |               # WHERE clause on source ONLY
                  (EXDOSE == 0 &
                   str_detect(EXTRT, "PLACEBO")),
    order       = exprs(EXSTDTM, EXSEQ),     # tie-breaker sort within group
    mode        = "first",                   # pick FIRST after the order — i.e. earliest
    new_vars    = exprs(TRTSDTM = EXSTDTM)   # rename EXSTDTM to TRTSDTM during join
  )
```

- **`dataset_add`** must be a tibble or data frame, already pre-processed (dates parsed, blanks handled, etc.). Admiral does not chain upstream functions inside the call — pre-process explicitly so the audit trail is visible.
- **`by_vars`** is always wrapped in `exprs()` so the bare column names are captured as expressions, not evaluated. Multi-column keys are common (`exprs(STUDYID, USUBJID, PARAMCD)`).
- **`filter_add`** runs *before* the join. It scopes which rows in the source are eligible. The expression sees columns of `dataset_add` only.
- **`order`** is a list of bare column names (also wrapped in `exprs()`). The sort runs ascending unless wrapped in `desc()`.
- **`mode`** is one of `"first"`, `"last"`, or omitted. If omitted, the source is expected to be unique by `by_vars` already — admiral checks and errors out if it is not (controlled by `check_type`).
- **`new_vars`** is a named-list-of-expressions: `exprs(NEWNAME = OLDNAME, ALSONEW = SRCCOL)`. Without it, *every* non-key column of the source is brought across — usually too many.

> **SAS analog mental model:** `derive_vars_merged()` is what a SAS programmer would write if they could declare a `MERGE` with built-in `WHERE`, `BY`, `IF FIRST.` / `IF LAST.`, and `RENAME=` flags, all in one statement. The composite-argument design replaces three or four small DATA steps with one named, auditable call.

### 7.6 The `by_vars` argument: matching keys

The cardinality rule is enforced. If the source has more than one row per `by_vars` combination *after* `filter_add` and *before* `mode` selection, admiral errors out unless `mode` is set. This protects against the SAS pitfall where a one-to-many MERGE silently duplicates the left side.

- **One-to-one join** (e.g., source already unique by `STUDYID × USUBJID`): omit `mode`, admiral validates uniqueness.
- **Many-to-one join** (e.g., multiple EX records per subject, you want one): set `mode = "first"` or `"last"` with a sensible `order`.
- **Many-to-many join**: not directly supported. Pre-process the source to many-to-one first.

### 7.7 The `dataset_add` argument: the right-side source

`dataset_add` is the source table to pull variables *from*. The base dataset (the one piped into `derive_vars_merged()`) is the table being augmented. Mnemonic: *add* — what you are *adding to* the base.

A common pattern is to prepare `dataset_add` in a `WORK` step right before the call:

```r
ex_dtm <- ex |>
  derive_vars_dtm(dtc = EXSTDTC, new_vars_prefix = "EXST")

adsl_base |> derive_vars_merged(dataset_add = ex_dtm, ...)
```

This separation mirrors the SAS habit of building a `WORK.ex_sorted` intermediate before the `MERGE`.

### 7.8 The `filter_add` argument: pre-merge subset

`filter_add` is evaluated **inside `dataset_add`** before the join. Two implications:

1. You can reference columns that exist in `dataset_add` but not in the base — e.g., `EXDOSE > 0` works even though `EXDOSE` is not in `adsl_base`.
2. You cannot reference columns of the base — for that, you need `derive_vars_joined()` (covered in Lesson 17).

> **Common pitfall:** writing `filter_add = TRTSDT >= 1` and expecting it to apply to the base — it errors because `TRTSDT` does not exist in `dataset_add`.

### 7.9 The `order` argument: tie-breaker sort

`order` controls the row ordering inside each `by_vars` group *before* `mode` picks one. Mode `"first"` picks the earliest row by this ordering; `"last"` picks the latest.

The order argument almost always includes the analysis date as the primary sort key plus the sequence variable (`EXSEQ`, `AESEQ`) as a deterministic tie-breaker:

```r
order = exprs(EXSTDTM, EXSEQ)
```

If two records share the same EXSTDTM, the lower EXSEQ wins (with `mode = "first"`). This deterministic tie-break protects against subtle bugs when the source is re-sorted or refreshed.

### 7.10 The `mode` argument: first vs last vs NULL

| `mode` value      | Effect                                              | Use case                          |
|-------------------|-----------------------------------------------------|-----------------------------------|
| `"first"`         | After `order`, take row 1 per `by_vars` group       | TRTSDT, ABLFL, first dose date    |
| `"last"`          | After `order`, take last row per group              | TRTEDT, last visit date           |
| omitted / `NULL`  | Require source uniqueness per `by_vars`             | One-to-one lookups                |

### 7.11 The `new_vars` argument: rename-on-arrival

`new_vars = exprs(TRTSDTM = EXSTDTM)` reads as "create `TRTSDTM` in the base by taking `EXSTDTM` from the source." Without this argument, *every* non-key source column comes across with its original name. That is rarely what you want — be explicit.

Multiple renames are common:

```r
new_vars = exprs(
  TRTSDTM = EXSTDTM,
  TRTEDTM = EXENDTM,
  TRTSEQ  = EXSEQ
)
```

### 7.12 The `missing_values` argument: filling unmatched rows

If a base row has no matching source row (after `filter_add`), the new variable is `NA` by default. `missing_values = exprs(TRTSDTM = as.POSIXct("1900-01-01"))` fills with a sentinel.

> **SAS analog:** the `COALESCE(x, sentinel)` you'd write after the MERGE.

### 7.13 Why not just `left_join() + filter + slice`?

You could write the equivalent in raw dplyr:

```r
adsl_trt <- adsl_base |>
  left_join(
    ex_dtm |>
      filter(EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) |>
      arrange(EXSTDTM, EXSEQ) |>
      group_by(STUDYID, USUBJID) |>
      slice_head(n = 1) |>
      ungroup() |>
      transmute(STUDYID, USUBJID, TRTSDTM = EXSTDTM),
    by = c("STUDYID", "USUBJID")
  )
```

That works. But it is 8 lines of conditional, sequential dplyr that has to be re-read carefully every time. `derive_vars_merged()` standardizes the intent: every reviewer recognizes the pattern instantly, and the named arguments make the audit trail readable. Consistency across 30 calls in an ADSL program is the win.

### 7.14 Common pitfalls

1. **Forgetting `mode`** when the source is many-to-one. Admiral errors out — that is the design — but the message can confuse first-time users.
2. **Wrong `order` direction.** "First exposure" needs ascending EXSTDTM; "last exposure end" needs ascending EXENDTM with `mode = "last"`. Wrapping in `desc()` flips it.
3. **`filter_add` referencing base columns.** Move the filter to `restrict_derivation()` outside the call (covered in Lesson 19), or pre-filter the base before the pipe.
4. **`new_vars` omitted, source has 30 columns.** All 30 land in the base, polluting the output. Always be explicit.
5. **Source not pre-processed.** If `EXSTDTM` was never derived, you get `NA`s silently. Pre-process the source with `derive_vars_dtm()` first.

### 7.15 Worked demo: TRTSDTM step-by-step

```{r derive-merged-demo}
# Merge TRTSDT from DM onto a base dataset
adsl_base <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM)

ex_dtm <- ex |>
  derive_vars_dtm(dtc = EXSTDTC, new_vars_prefix = "EXST",
                  date_imputation = "first", time_imputation = "first") |>
  derive_vars_dtm(dtc = EXENDTC, new_vars_prefix = "EXEN",
                  date_imputation = "last",  time_imputation = "last")

# Derive TRTSDTM: first qualifying exposure datetime
adsl_trt <- adsl_base |>
  derive_vars_merged(
    dataset_add = ex_dtm,
    filter_add  = EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")),
    new_vars    = exprs(TRTSDTM = EXSTDTM),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_trt |>
  select(USUBJID, ACTARM, TRTSDTM) |>
  mutate(TRTSDTM_human = as.POSIXct(TRTSDTM, origin = "1970-01-01", tz = "UTC"))
```

> **Read the output:** five subjects, five TRTSDTM values. Subject `01-701-1015` had two EX records (EXSEQ 1 and 2); `mode = "first"` ordered by EXSTDTM picked the earliest (2014-01-02). The downstream layout step will format this datetime as the table-shell "First Exposure" cell.

### 7.16 Exercise: Derive TRTEDTM (last qualifying exposure end datetime)

Use `mode = "last"` and `EXENDTM` instead of `EXSTDTM`.

```{r derive-merged-end, exercise=TRUE, exercise.lines=18}
adsl_trte <- adsl_trt |>
  derive_vars_merged(
    dataset_add = ex_dtm,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXENDTM),
    new_vars    = exprs(TRTEDTM = ___),
    order       = exprs(___, EXSEQ),
    mode        = ___,
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_trte |>
  select(USUBJID, TRTSDTM, TRTEDTM) |>
  mutate(
    TRTS_human = as.POSIXct(TRTSDTM, origin = "1970-01-01", tz = "UTC"),
    TRTE_human = as.POSIXct(TRTEDTM, origin = "1970-01-01", tz = "UTC")
  )
```

```{r derive-merged-end-hint-1}
# new_vars = exprs(TRTEDTM = EXENDTM)
# order    = exprs(EXENDTM, EXSEQ)
# mode     = "last"
```

```{r derive-merged-end-solution}
adsl_trte <- adsl_trt |>
  derive_vars_merged(
    dataset_add = ex_dtm,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXENDTM),
    new_vars    = exprs(TRTEDTM = EXENDTM),
    order       = exprs(EXENDTM, EXSEQ),
    mode        = "last",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_trte |>
  select(USUBJID, TRTSDTM, TRTEDTM) |>
  mutate(
    TRTS_human = as.POSIXct(TRTSDTM, origin = "1970-01-01", tz = "UTC"),
    TRTE_human = as.POSIXct(TRTEDTM, origin = "1970-01-01", tz = "UTC")
  )
```

### Exercise: derive_vars_merged on real pharmaversesdtm data

Build a minimal ADSL skeleton from `dm_real` and derive `TRTSDT` (first treatment start date) from `ex_real`. This mirrors the first `derive_vars_merged()` call in every real ADSL program.

```{r derive-merged-real, exercise=TRUE, exercise.lines=22}
# Step 1: convert blanks and add numeric EXSTDT to real EX
ex_real_dt <- convert_blanks_to_na(ex_real) |>
  derive_vars_dt(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first"
  )

# Step 2: ADSL skeleton — keep first 10 treated subjects from real DM
adsl_real_skel <- convert_blanks_to_na(dm_real) |>
  filter(!is.na(RFXSTDTC)) |>
  slice_head(n = 10) |>
  select(STUDYID, USUBJID, ARM, ACTARM)

# Step 3: merge first exposure date per subject
adsl_real_trts <- adsl_real_skel |>
  derive_vars_merged(
    dataset_add = ex_real_dt,
    filter_add  = !is.na(EXSTDT),
    new_vars    = exprs(TRTSDT = ___),
    order       = exprs(___),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_real_trts |>
  mutate(TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")) |>
  select(USUBJID, ARM, TRTSDT_human)
```

```{r derive-merged-real-hint-1}
# new_vars = exprs(TRTSDT = EXSTDT)
# order    = exprs(EXSTDT)
```

```{r derive-merged-real-solution}
ex_real_dt <- convert_blanks_to_na(ex_real) |>
  derive_vars_dt(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first"
  )

adsl_real_skel <- convert_blanks_to_na(dm_real) |>
  filter(!is.na(RFXSTDTC)) |>
  slice_head(n = 10) |>
  select(STUDYID, USUBJID, ARM, ACTARM)

adsl_real_trts <- adsl_real_skel |>
  derive_vars_merged(
    dataset_add = ex_real_dt,
    filter_add  = !is.na(EXSTDT),
    new_vars    = exprs(TRTSDT = EXSTDT),
    order       = exprs(EXSTDT),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_real_trts |>
  mutate(TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")) |>
  select(USUBJID, ARM, TRTSDT_human)
```

### 7.18 New exercise: Merge a calculated max dose per subject

Build a tiny "max-dose lookup" from `ex` (the cumulative-maximum EXDOSE per subject), then merge it onto `adsl_base` as `MAXDOSE`. Mode must be `"last"` and the order must put the highest dose at the bottom of the group.

```{r derive-merged-maxdose, exercise=TRUE, exercise.lines=18}
# Source: highest single-record EXDOSE per subject
ex_max_lookup <- ex |>
  group_by(STUDYID, USUBJID) |>
  summarise(MAXDOSE_SRC = max(EXDOSE, na.rm = TRUE), .groups = "drop")

adsl_max <- adsl_base |>
  derive_vars_merged(
    dataset_add = ex_max_lookup,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(___ = ___)
    # No mode/order needed — the lookup is already unique per subject.
  )

adsl_max |> select(USUBJID, ACTARM, MAXDOSE)
```

```{r derive-merged-maxdose-hint-1}
# new_vars = exprs(MAXDOSE = MAXDOSE_SRC)
```

```{r derive-merged-maxdose-hint-2}
# Because ex_max_lookup is already one row per subject, omit `mode`.
# Admiral will validate uniqueness — if duplicates existed, it would error.
```

```{r derive-merged-maxdose-solution}
ex_max_lookup <- ex |>
  group_by(STUDYID, USUBJID) |>
  summarise(MAXDOSE_SRC = max(EXDOSE, na.rm = TRUE), .groups = "drop")

adsl_max <- adsl_base |>
  derive_vars_merged(
    dataset_add = ex_max_lookup,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(MAXDOSE = MAXDOSE_SRC)
  )

adsl_max |> select(USUBJID, ACTARM, MAXDOSE)
```

> **Read the output:** every subject has a single `MAXDOSE` because the lookup was pre-aggregated. This pattern — "summarise upstream, then merge" — is the most common simplification for SAS programmers used to a `PROC SUMMARY` followed by a `MERGE`.

### 7.19 Knowledge check

```{r quiz-sec7, echo=FALSE}
quiz(
  question(
    "Which SAS construct is the closest single-statement analog of `derive_vars_merged()` with `mode = \"first\"`?",
    answer("`PROC MEANS DATA= NOPRINT;` with `OUTPUT OUT=`"),
    answer("`PROC SORT` then `DATA step` with `MERGE BY USUBJID;` and `IF FIRST.USUBJID;`", correct = TRUE),
    answer("`PROC FREQ` with `BY USUBJID;`"),
    random_answer_order = TRUE
  ),
  question(
    "`filter_add` is applied:",
    answer("After the join, against the merged dataset"),
    answer("Before the join, against `dataset_add` only", correct = TRUE),
    answer("On both sides of the join simultaneously"),
    random_answer_order = TRUE
  ),
  question(
    "You set `mode = \"first\"` and `order = exprs(EXSTDTM, EXSEQ)`. Two records share the same EXSTDTM. Which one is kept?",
    answer("Both — admiral returns one row per match"),
    answer("Whichever appears first in the source — ordering is non-deterministic"),
    answer("The record with the lower EXSEQ — EXSEQ is the deterministic tie-breaker", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "Omitting `new_vars` causes:",
    answer("Admiral to error out demanding the argument"),
    answer("Every non-key column of the source to be brought onto the base — usually too many", correct = TRUE),
    answer("Only the first numeric column to be brought across"),
    random_answer_order = TRUE
  ),
  question(
    "`derive_vars_merged()` errors with \"check_type = 'error': dataset_add contains duplicates by by_vars\". The fix is:",
    answer("Add `check_type = \"none\"` to suppress the error"),
    answer("Set `mode = \"first\"` (or `\"last\"`) with a sensible `order` — this is many-to-one, declare it explicitly", correct = TRUE),
    answer("Switch to `derive_var_extreme_flag()`"),
    random_answer_order = TRUE
  )
)
```

---

## 8 — `derive_var_extreme_flag()`: first/last record flags

### 8.1 What it does

Inside a `by_vars` group, sort rows by `order`, then mark exactly one row with the value `"Y"` in a new flag column. All other rows get `NA`. This is the admiral idiom for **`ABLFL`** (baseline flag), **first/last visit flags**, **first/last AE per term**, and any other one-of-many indicator that lives on the source dataset rather than as a merged column.

### 8.2 SAS analog comparison

| Concept                | SAS                                                      | admiral                                                |
|------------------------|----------------------------------------------------------|--------------------------------------------------------|
| Sort first             | `PROC SORT BY USUBJID LBTESTCD ADT;`                     | `order = exprs(ADT)` inside the call                   |
| Detect first per group | `IF FIRST.LBTESTCD;`                                     | `mode = "first"`                                       |
| Detect last per group  | `IF LAST.LBTESTCD;`                                      | `mode = "last"`                                        |
| Set the flag           | `IF FIRST.LBTESTCD THEN ABLFL = "Y";`                    | `new_var = ABLFL` (admiral writes `"Y"` automatically) |
| Keep all rows          | The DATA step naturally keeps all rows; only the flag changes | derive_var_extreme_flag mutates a new column in place |

> **SAS analog mental model:** `derive_var_extreme_flag()` is `PROC SORT BY ... ; DATA ...; SET sorted; BY ...; IF FIRST.x THEN FL = "Y";` — but the sort, the BY group, and the flag assignment are declared in one named-argument call instead of three SAS steps.

### 8.3 ASCII side-by-side

```
   proc sort data=lb out=lb_sorted;       |   lb_flagged <- lb |>
     by STUDYID USUBJID LBTESTCD ADT;     |     derive_vars_dt(
   run;                                   |       dtc = LBDTC,
                                          |       new_vars_prefix = "A") |>
   data lb_flagged;                       |     derive_var_extreme_flag(
     set lb_sorted;                       |       by_vars = exprs(STUDYID,
     by STUDYID USUBJID LBTESTCD;         |                       USUBJID,
     if first.LBTESTCD                    |                       LBTESTCD),
       then ABLFL = "Y";                  |       order   = exprs(ADT),
   run;                                   |       new_var = ABLFL,
                                          |       mode    = "first")
```

The SAS form needs an explicit PROC SORT then a DATA-step BY processing. Admiral folds both into a single call.

### 8.4 Anatomy of the call

```r
derive_var_extreme_flag(
  dataset  = lb,                                  # base, piped in
  by_vars  = exprs(STUDYID, USUBJID, LBTESTCD),   # group within which "first/last" applies
  order    = exprs(ADT),                          # sort within group ascending by date
  new_var  = ABLFL,                               # column name for the "Y"/NA flag
  mode     = "first"                              # "first" → earliest; "last" → most recent
)
```

- **`by_vars`** must uniquely partition the rows you want to flag. For baseline, partition by `(STUDYID, USUBJID, LBTESTCD)` so each (subject × test) gets exactly one ABLFL.
- **`order`** drives the "first/last" semantic. Wrap a variable in `desc()` to flip direction without changing `mode`.
- **`new_var`** is a bare name. The flag column is created on the output regardless of whether it existed before; if it did exist, it is overwritten.
- **`mode`** values: `"first"` (lowest in sorted order) and `"last"` (highest).
- **Filter via `restrict_derivation()`** if you need to flag only a subset (e.g., pre-treatment records only). Pre-filtering before the pipe is also fine.

### 8.5 Step-by-step decomposition

1. **Sort:** apply `order` within `by_vars` groups. Default is ascending; wrap in `desc()` to reverse.
2. **Group:** within each `by_vars` combination, isolate the sorted rows.
3. **Rank:** assign `row_number()` 1..N inside the group.
4. **Flag:** set `new_var = "Y"` where rank == 1 (`mode = "first"`) or rank == max (`mode = "last"`); set `NA` elsewhere.
5. **Return:** the base dataset with the new flag column appended.

> **CDISC convention:** the flag is `"Y"` or `NA`, **never** `"N"`. ADaM IG explicitly forbids `"N"` in flag variables. This matches the SAS `IF FIRST.x THEN FL="Y";` pattern (no ELSE clause).

### 8.6 Worked demo

```{r extreme-flag-demo}
# Add an analysis date to LB, then flag the baseline (first) record per subject per test
lb_flagged <- lb |>
  derive_vars_dt(dtc = LBDTC, new_vars_prefix = "A") |>
  derive_var_extreme_flag(
    by_vars  = exprs(STUDYID, USUBJID, LBTESTCD),
    order    = exprs(ADT),
    new_var  = ABLFL,
    mode     = "first"
  )

lb_flagged |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01")) |>
  select(USUBJID, LBTESTCD, ADT_human, ABLFL)
```

> **Read the output:** subject `01-701-1015` has two ALT records (2014-01-02 and 2014-02-10). The earlier one carries `ABLFL = "Y"`; the later carries `NA`. The same logic, applied per `(USUBJID, LBTESTCD)`, would scale to thousands of subjects and dozens of tests without changing one character of code.

### 8.7 Common pitfalls

1. **Wrong `by_vars` granularity.** Forgetting `LBTESTCD` would flag the first record per *subject*, not per *test* — every test except the first would lose its baseline.
2. **Wrong `order` direction.** "Most recent" needs `mode = "last"` *or* `order = exprs(desc(ADT))` with `mode = "first"`. Pick one and stick with it.
3. **Forgetting to derive ADT first.** Sorting on `LBDTC` (character) would order alphabetically — `"2014-02-10"` sorts after `"2014-01-02"` but `"2014-1-2"` (unzero-padded) would sort wrong. Always sort on the numeric `ADT`, not the character `DTC`.
4. **Pre-treatment vs post-treatment confusion.** ABLFL should usually be defined as the *last pre-treatment* record, not the *first overall* record. Pre-filter `lb |> filter(ADT < TRTSDT)` before flagging, or use `restrict_derivation()`.

### 8.8 Exercise: Flag the LAST (most recent) lab record per subject per test

Change `mode` to `"last"` to flag the most recent observation instead.

```{r extreme-flag-last, exercise=TRUE, exercise.lines=14}
lb_last_flagged <- lb |>
  derive_vars_dt(dtc = LBDTC, new_vars_prefix = "A") |>
  derive_var_extreme_flag(
    by_vars = exprs(STUDYID, USUBJID, LBTESTCD),
    order   = exprs(ADT),
    new_var = LAST_FL,
    mode    = ___
  )

lb_last_flagged |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01")) |>
  select(USUBJID, LBTESTCD, ADT_human, LAST_FL)
```

```{r extreme-flag-last-hint-1}
# mode = "last"
```

```{r extreme-flag-last-solution}
lb_last_flagged <- lb |>
  derive_vars_dt(dtc = LBDTC, new_vars_prefix = "A") |>
  derive_var_extreme_flag(
    by_vars = exprs(STUDYID, USUBJID, LBTESTCD),
    order   = exprs(ADT),
    new_var = LAST_FL,
    mode    = "last"
  )

lb_last_flagged |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01")) |>
  select(USUBJID, LBTESTCD, ADT_human, LAST_FL)
```

### 8.9 New exercise: Flag the most-severe AE per subject

Build a tiny AE table and flag the *highest-severity* AE per subject. Use `mode = "last"` with `order = exprs(AESEVN)` — the highest severity will sort to the end.

```{r extreme-flag-ae, exercise=TRUE, exercise.lines=20}
ae_mock <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("S001","S001","S001","S002","S002","S003"),
  AESEQ   = 1:6,
  AEDECOD = c("HEADACHE","NAUSEA","FATIGUE","HEADACHE","FATIGUE","FATIGUE"),
  AESEVN  = c(1, 2, 3, 1, 2, 1)   # 1=mild, 2=moderate, 3=severe
)

ae_worst <- ae_mock |>
  derive_var_extreme_flag(
    by_vars = exprs(STUDYID, ___),
    order   = exprs(___),
    new_var = AWORSTFL,
    mode    = ___
  )

ae_worst
```

```{r extreme-flag-ae-hint-1}
# by_vars: one row per (STUDYID, USUBJID); we want one worst per subject.
# order: AESEVN — higher = more severe.
# mode: "last" picks the highest in ascending sort.
```

```{r extreme-flag-ae-hint-2}
# by_vars = exprs(STUDYID, USUBJID)
# order   = exprs(AESEVN)
# mode    = "last"
```

```{r extreme-flag-ae-solution}
ae_mock <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("S001","S001","S001","S002","S002","S003"),
  AESEQ   = 1:6,
  AEDECOD = c("HEADACHE","NAUSEA","FATIGUE","HEADACHE","FATIGUE","FATIGUE"),
  AESEVN  = c(1, 2, 3, 1, 2, 1)
)

ae_worst <- ae_mock |>
  derive_var_extreme_flag(
    by_vars = exprs(STUDYID, USUBJID),
    order   = exprs(AESEVN),
    new_var = AWORSTFL,
    mode    = "last"
  )

ae_worst
```

> **Read the output:** S001 has three AEs ranging from mild to severe; only the severe row carries `AWORSTFL = "Y"`. S002 has two AEs; the moderate one wins. S003 has only one AE; it wins by default. This is the canonical pattern for "worst-event" derivations in OCCDS (Lesson 17) and BDS (Lesson 16).

### 8.10 Knowledge check

```{r quiz-sec8, echo=FALSE}
quiz(
  question(
    "`derive_var_extreme_flag()` writes `\"N\"` to non-extreme records. True or false?",
    answer("True"),
    answer("False — non-extreme records receive `NA`. CDISC ADaM IG forbids `\"N\"` in flag variables.", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "Which SAS construct mirrors `derive_var_extreme_flag(... , mode = \"first\")`?",
    answer("`PROC FREQ;` with `BY USUBJID;`"),
    answer("`PROC SORT BY USUBJID ADT;` followed by `DATA step; SET; BY USUBJID; IF FIRST.USUBJID THEN FL = \"Y\";`", correct = TRUE),
    answer("`PROC SQL;` `SELECT DISTINCT USUBJID FROM lb`"),
    random_answer_order = TRUE
  ),
  question(
    "To flag the LATEST visit per subject per test, the safest declaration is:",
    answer("`mode = \"first\"`, `order = exprs(desc(ADT))` — equivalent to `mode = \"last\"`, `order = exprs(ADT)`", correct = TRUE),
    answer("Two separate calls, one for `first` and one for `last`"),
    answer("`mode = \"last\"`, `order = exprs(desc(ADT))` — double-reversal gives latest"),
    random_answer_order = TRUE
  ),
  question(
    "ABLFL should be defined as:",
    answer("The first record overall per subject"),
    answer("The last pre-treatment record per subject per parameter — requiring a pre-filter or `restrict_derivation()`", correct = TRUE),
    answer("The record with the lowest AVAL"),
    random_answer_order = TRUE
  )
)
```

---

## 9 — `derive_var_merged_exist_flag()`: population flags

### 9.1 What it does

For each row in the base dataset, look up the matching `by_vars` group in another dataset. If at least one row in that group satisfies the `condition`, set the new flag to `"Y"`. If no matching group exists, set `NA`. This is the admiral idiom for **SAFFL, ITTFL, RANDFL, FASFL, PPROTFL** — the entire family of population flags.

### 9.2 The "Y or NA, never N" CDISC rule

CDISC ADaM IG Section 4.1.2 specifies that flag variables (`*FL`) take the values `"Y"` or `NA`. **`"N"` is forbidden.** This is *not* how a SAS programmer typically writes flag logic:

```
   SAS habit (wrong for CDISC):           |   admiral (right):
                                          |
   data adsl;                             |   adsl |>
     set adsl;                            |     derive_var_merged_exist_flag(
     if condition                         |       dataset_add = ex,
       then SAFFL = "Y";                  |       by_vars     = exprs(USUBJID),
       else SAFFL = "N";  /* WRONG */     |       new_var     = SAFFL,
   run;                                   |       condition   = EXDOSE > 0)
```

The reason is downstream tabulation: a `"N"` row would appear as an explicit zero in a frequency table, while `NA` correctly drops out of the denominator. A subject who never met the safety condition is simply absent from the safety population, not actively "no."

> **SAS analog:** the closest SAS analog is `IF condition THEN FL = "Y";` with **no ELSE**, deliberately leaving non-matchers as missing.

### 9.3 SAS analog comparison

| `derive_var_merged_exist_flag()` argument | Closest SAS equivalent                                          |
|-------------------------------------------|-----------------------------------------------------------------|
| `dataset_add`                             | The source dataset in a `PROC SQL EXISTS` subquery               |
| `by_vars`                                 | The correlated key columns                                       |
| `condition`                               | The `WHERE` clause inside the subquery                           |
| `new_var`                                 | The output column on the parent dataset                          |
| Default value when no match               | `NA` — corresponds to leaving the variable unset (no ELSE)       |

### 9.4 ASCII side-by-side

```
   proc sql;                              |   adsl_for_saffl <- adsl_base |>
     create table adsl_for_saffl as       |     derive_var_merged_exist_flag(
     select a.*,                          |       dataset_add = ex,
            case when exists              |       by_vars     =
              (select 1 from ex as b      |         exprs(STUDYID, USUBJID),
                where a.STUDYID =         |       new_var     = SAFFL,
                      b.STUDYID           |       condition   =
                  and a.USUBJID =         |         EXDOSE > 0 |
                      b.USUBJID           |         (EXDOSE == 0 &
                  and (b.EXDOSE > 0       |          str_detect(EXTRT, "PLACEBO")))
                    or (b.EXDOSE = 0      |
                        and b.EXTRT       |
                        contains          |
                        "PLACEBO")))      |
            then "Y" end as SAFFL         |
       from adsl_base as a;               |
   quit;                                  |
```

The PROC SQL `CASE WHEN EXISTS` subquery is the closest single-statement SAS equivalent. The DATA-step alternative is uglier (sort + merge + retain + reset). Either way, `derive_var_merged_exist_flag()` collapses many lines into a named-argument call.

### 9.5 Anatomy of the call

```r
derive_var_merged_exist_flag(
  dataset     = adsl_base,                   # base (left), piped in
  dataset_add = ex,                          # source (right) — does at least one row exist matching condition?
  by_vars     = exprs(STUDYID, USUBJID),     # correlate by these keys
  new_var     = SAFFL,                       # output flag column name
  condition   = EXDOSE > 0 |                 # the EXISTS condition; sees columns of dataset_add only
                (EXDOSE == 0 &
                 str_detect(EXTRT, "PLACEBO"))
)
```

- **`condition`** is an unevaluated expression. Quote it with `exprs()` if you build it dynamically; otherwise write it inline.
- **`by_vars`** must reference columns present in both base and source — typically `STUDYID + USUBJID` for population flags.
- The function does **not** sort or pick rows. It is a pure EXISTS lookup. Any single row in the source meeting `condition` triggers `"Y"`.
- The output is the base dataset with one new column. No row count change.

> **SAS analog mental model:** `derive_var_merged_exist_flag()` is a `PROC SQL CASE WHEN EXISTS (correlated subquery) THEN "Y" END`. There is no ELSE branch — that is intentional and correct for CDISC.

### 9.6 Common pitfalls

1. **Writing `condition = "EXDOSE > 0"` as a string.** Pass an unquoted expression, not a string.
2. **Adding an `else_val = "N"` argument.** It does not exist — admiral enforces the CDISC rule by not offering an "N" path.
3. **Forgetting to filter to safety-eligible subjects upstream.** Sometimes the desired population is "treated AND alive at first dose," requiring a multi-condition `condition` or a pre-filtered `dataset_add`.
4. **Source has duplicates that don't match `by_vars`.** Duplicate matches still produce one `"Y"` — duplicates do *not* trigger an error here (unlike `derive_vars_merged()`), because EXISTS only cares if *any* row matches.

### 9.7 Worked demo

```{r exist-flag-demo}
adsl_for_saffl <- adsl_base |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))
  )

adsl_for_saffl |> select(USUBJID, ACTARM, SAFFL)
```

> **Read the output:** every subject has `SAFFL = "Y"` here because every subject in `ex` has at least one record meeting the condition. Subjects who appear in `adsl_base` but **not** in `ex` would carry `SAFFL = NA` — they were never dosed, so they cannot be in the safety population.

### 9.8 Knowledge check on the CDISC flag convention

```{r quiz-flag-convention, echo=FALSE}
quiz(
  question(
    "CDISC ADaM IG specifies that flag variables like SAFFL take values:",
    answer('"Y" if met, "N" if not'),
    answer('"Y" if met, NA if not — never use "N"', correct = TRUE),
    answer("1 if met, 0 if not"),
    random_answer_order = TRUE
  )
)
```

### 9.9 Exercise: Derive ITTFL based on whether RANDDT exists

A subject is in the ITT population if they have a randomization date (RANDDT is not NA). For now, we'll simulate RANDDT from DM:

```{r ittfl-exercise, exercise=TRUE, exercise.lines=16}
# Create a mock randomization event dataset
rand_events <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101"),
  # 01-703-1211 is a screen failure — no randomization
  RANDDT  = as.Date(c("2014-01-02", "2012-08-05", "2013-07-09", "2013-04-14"))
)

# Derive ITTFL: "Y" for subjects with a randomization event
adsl_ittfl <- adsl_base |>
  derive_var_merged_exist_flag(
    dataset_add = rand_events,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = ITTFL,
    condition   = !is.na(___)
  )

adsl_ittfl |> select(USUBJID, ACTARM, ITTFL)
```

```{r ittfl-exercise-hint-1}
# condition = !is.na(RANDDT)
```

```{r ittfl-exercise-solution}
rand_events <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101"),
  RANDDT  = as.Date(c("2014-01-02", "2012-08-05", "2013-07-09", "2013-04-14"))
)

adsl_ittfl <- adsl_base |>
  derive_var_merged_exist_flag(
    dataset_add = rand_events,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = ITTFL,
    condition   = !is.na(RANDDT)
  )

adsl_ittfl |> select(USUBJID, ACTARM, ITTFL)
```

### 9.10 New exercise: Derive PPROTFL (per-protocol)

Per-protocol = randomized AND treated AND no major protocol deviations. Build a small `devs` table of major deviations, then derive `PPROTFL`. A subject is per-protocol if they are NOT in the deviation list.

This is a *complementary* exist-flag: existence of a row → `"N"` semantically, which CDISC writes as flipping the logic so existence → `NA`, absence → `"Y"`. The trick is to compute PPROTFL in two steps: first flag deviations, then derive PPROTFL based on the absence.

```{r pprotfl-exercise, exercise=TRUE, exercise.lines=22}
# Subjects with major deviations
major_devs <- tibble(
  STUDYID    = "CDISCPILOT01",
  USUBJID    = c("01-701-1015"),      # one subject deviates
  DVSEV      = "MAJOR"
)

# Step 1: flag deviation existence
adsl_dev <- adsl_base |>
  derive_var_merged_exist_flag(
    dataset_add = major_devs,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = DEVFL,
    condition   = DVSEV == ___
  )

# Step 2: PPROTFL = "Y" where DEVFL is NA
adsl_pp <- adsl_dev |>
  mutate(PPROTFL = if_else(is.na(___), "Y", NA_character_))

adsl_pp |> select(USUBJID, ACTARM, DEVFL, PPROTFL)
```

```{r pprotfl-exercise-hint-1}
# condition = DVSEV == "MAJOR"
# is.na(DEVFL) — the subject is in PP if they did NOT trigger DEVFL
```

```{r pprotfl-exercise-hint-2}
# Final: if_else(is.na(DEVFL), "Y", NA_character_)
```

```{r pprotfl-exercise-solution}
major_devs <- tibble(
  STUDYID    = "CDISCPILOT01",
  USUBJID    = c("01-701-1015"),
  DVSEV      = "MAJOR"
)

adsl_dev <- adsl_base |>
  derive_var_merged_exist_flag(
    dataset_add = major_devs,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = DEVFL,
    condition   = DVSEV == "MAJOR"
  )

adsl_pp <- adsl_dev |>
  mutate(PPROTFL = if_else(is.na(DEVFL), "Y", NA_character_))

adsl_pp |> select(USUBJID, ACTARM, DEVFL, PPROTFL)
```

> **Read the output:** four subjects get `PPROTFL = "Y"` (no deviation); subject `01-701-1015` gets `PPROTFL = NA` (deviation flagged). Note the strict CDISC rule: PPROTFL is never `"N"`. A SAS programmer would write `ELSE PPROTFL = "N";` — admiral writes `NA_character_` instead, preserving the missing-as-not-met semantics for downstream tabulation.

### 9.11 Knowledge check

```{r quiz-sec9, echo=FALSE}
quiz(
  question(
    "`derive_var_merged_exist_flag()` is closest to which SAS construct?",
    answer("`PROC SQL ... CASE WHEN EXISTS (correlated subquery) THEN \"Y\" END`", correct = TRUE),
    answer("`PROC SUMMARY NWAY OUT=`"),
    answer("`PROC FREQ TABLES ... / NOFREQ;`"),
    random_answer_order = TRUE
  ),
  question(
    "ITTFL is `NA` (not `\"N\"`) for non-ITT subjects because:",
    answer("Admiral has a bug"),
    answer("CDISC ADaM IG Section 4.1.2 forbids `\"N\"` in flag variables; `NA` correctly drops from tabulation denominators", correct = TRUE),
    answer("`\"N\"` and `NA` are equivalent in R"),
    random_answer_order = TRUE
  ),
  question(
    "The source dataset has 3 duplicate rows that all match the condition. The flag receives:",
    answer("`\"YYY\"` — one `Y` per match"),
    answer("`\"Y\"` — the function reports existence, not count", correct = TRUE),
    answer("An error — duplicates are forbidden"),
    random_answer_order = TRUE
  ),
  question(
    "To derive PPROTFL (per-protocol), the most reliable pattern is:",
    answer("Flag deviation existence with `derive_var_merged_exist_flag`, then set `PPROTFL = \"Y\"` where the deviation flag is `NA`", correct = TRUE),
    answer("Call `derive_var_merged_exist_flag` directly with `condition = !exists(deviation)`"),
    answer("Use `derive_vars_merged()` with `mode = \"first\"`"),
    random_answer_order = TRUE
  )
)
```

---

## 10 — `call_derivation()`: DRY derivation of multiple variables

### 10.1 The DRY problem

A SAS programmer building EXSTDTM and EXENDTM would typically write *two* near-identical `derive_vars_dtm()` calls that differ only in the source variable and the imputation direction. Multiply that pattern over ADT/ADY derivations for 8 lab parameters and you end up with 16 nearly-identical calls. `call_derivation()` is admiral's "DRY" wrapper that runs one derivation function multiple times with shared and per-iteration arguments.

### 10.2 SAS analog comparison

| `call_derivation()` argument | What it controls                          | SAS analog                                                |
|------------------------------|-------------------------------------------|-----------------------------------------------------------|
| `derivation`                 | The function to invoke each iteration     | The macro `%name` being expanded                          |
| `variable_params`            | List of per-iteration argument sets       | A `%DO i = 1 %TO N` loop body with positional macro args  |
| `...` (shared args)          | Arguments held constant across iterations | Macro variables defined once outside the loop             |

> **SAS analog mental model:** `call_derivation()` is a `%DO` loop over a list of parameter sets, calling the same derivation macro each iteration with one set, while shared parameters stay constant.

### 10.3 ASCII side-by-side

```
   %macro dtm_one(dtc, prefix, time_imp); |   ex_both_dtm <- ex |>
     data ex;                             |     call_derivation(
       set ex;                            |       derivation = derive_vars_dtm,
       &prefix.DTM = input(&dtc,          |       variable_params = list(
                          E8601DT19.);    |         params(dtc = EXSTDTC,
       /* (imputation logic) */           |                new_vars_prefix = "EXST"),
     run;                                 |         params(dtc = EXENDTC,
   %mend;                                 |                new_vars_prefix = "EXEN",
                                          |                time_imputation = "last")
   %dtm_one(EXSTDTC, EXST, first);        |       ),
   %dtm_one(EXENDTC, EXEN, last);         |       date_imputation = "first"
                                          |     )
```

### 10.4 The `exprs()` vs `params()` distinction

This catches everyone the first time.

- `exprs(EXSTDTC, EXSEQ)` captures bare *column-name expressions* for use *inside* an admiral derivation, where tidy-evaluation kicks in.
- `params(dtc = EXSTDTC, new_vars_prefix = "EXST")` captures *named arguments* for use as one row of `variable_params`. Each `params(...)` block represents one iteration's argument set.

Mnemonically: `exprs()` makes a list of column references; `params()` makes a list of named arguments.

### 10.5 Anatomy of the call

```r
ex |>
  call_derivation(
    derivation      = derive_vars_dtm,           # what function to invoke
    variable_params = list(                      # what changes per iteration
      params(dtc = EXSTDTC, new_vars_prefix = "EXST"),
      params(dtc = EXENDTC, new_vars_prefix = "EXEN",
             time_imputation = "last")
    ),
    date_imputation = "first"                    # what stays constant (shared args)
  )
```

- **`derivation`** is the unquoted function name. Admiral resolves it at runtime — any `derive_*` function with consistent argument conventions works here.
- **`variable_params`** is a list of `params(...)` blocks. Each block becomes one invocation of `derivation`.
- **Shared args** (`date_imputation` above) are passed via `...`. They are merged into every iteration's call.
- A per-iteration arg shadows a shared arg: in the example, `time_imputation = "last"` for EXEN overrides any shared `time_imputation`.

### 10.6 When to use call_derivation vs explicit chained calls

| Situation                                                   | Preferred                           |
|-------------------------------------------------------------|-------------------------------------|
| Two derivations differ only in source variable              | `call_derivation()`                 |
| Three or more parameters share the same derivation pattern  | `call_derivation()`                 |
| Each derivation has substantially different logic           | Explicit chained calls (clarity wins) |
| You need different *functions* per iteration                | Explicit chained calls — call_derivation runs one function |

### 10.7 Worked demo

```{r call-derivation-demo}
# Instead of two separate derive_vars_dtm() calls:
ex_both_dtm <- ex |>
  call_derivation(
    derivation = derive_vars_dtm,
    variable_params = list(
      params(dtc = EXSTDTC, new_vars_prefix = "EXST"),
      params(dtc = EXENDTC, new_vars_prefix = "EXEN", time_imputation = "last")
    ),
    date_imputation = "first"   # shared arg
  )

ex_both_dtm |>
  select(USUBJID, EXSTDTC, EXSTDTM, EXENDTC, EXENDTM) |>
  head(3)
```

> **Read the output:** both `EXSTDTM` and `EXENDTM` appear in the same output despite a single `call_derivation()` call. EXSTDTM used the shared `date_imputation = "first"` with the default `time_imputation = "first"`; EXENDTM used the shared `date_imputation = "first"` overridden in its own block to `time_imputation = "last"`. Per-iteration args win over shared args.

### 10.8 New exercise: derive_vars_dt across three DTCs in one call

You have three DTC variables on a mock DM table — `BRTHDTC`, `RFSTDTC`, `RFXSTDTC`. Derive the numeric versions (`BRTHDT`, `RFSTDT`, `RFXSTDT`) in one `call_derivation()` block.

```{r call-derivation-3dtc, exercise=TRUE, exercise.lines=22}
dm_multi <- tibble(
  USUBJID  = c("S001", "S002", "S003"),
  BRTHDTC  = c("1950-04-12", "1948-10-30", "1952-07-22"),
  RFSTDTC  = c("2014-01-02", "2014-01-15", "2014-02-01"),
  RFXSTDTC = c("2014-01-04", "2014-01-18", "2014-02-04")
)

dm_3dt <- dm_multi |>
  call_derivation(
    derivation      = ___,
    variable_params = list(
      params(dtc = BRTHDTC,  new_vars_prefix = "BRTH"),
      params(dtc = RFSTDTC,  new_vars_prefix = "RFST"),
      params(dtc = RFXSTDTC, new_vars_prefix = "RFXST")
    ),
    date_imputation = "first"
  )

dm_3dt |>
  mutate(
    BRTHDT_h = as.Date(BRTHDT,  origin = "1970-01-01"),
    RFSTDT_h = as.Date(RFSTDT,  origin = "1970-01-01"),
    RFXSTDT_h = as.Date(RFXSTDT, origin = "1970-01-01")
  ) |>
  select(USUBJID, BRTHDT_h, RFSTDT_h, RFXSTDT_h)
```

```{r call-derivation-3dtc-hint-1}
# derivation = derive_vars_dt
```

```{r call-derivation-3dtc-hint-2}
# derivation = derive_vars_dt
# The same date_imputation applies to all three iterations because it
# is passed as a shared argument.
```

```{r call-derivation-3dtc-solution}
dm_multi <- tibble(
  USUBJID  = c("S001", "S002", "S003"),
  BRTHDTC  = c("1950-04-12", "1948-10-30", "1952-07-22"),
  RFSTDTC  = c("2014-01-02", "2014-01-15", "2014-02-01"),
  RFXSTDTC = c("2014-01-04", "2014-01-18", "2014-02-04")
)

dm_3dt <- dm_multi |>
  call_derivation(
    derivation      = derive_vars_dt,
    variable_params = list(
      params(dtc = BRTHDTC,  new_vars_prefix = "BRTH"),
      params(dtc = RFSTDTC,  new_vars_prefix = "RFST"),
      params(dtc = RFXSTDTC, new_vars_prefix = "RFXST")
    ),
    date_imputation = "first"
  )

dm_3dt |>
  mutate(
    BRTHDT_h = as.Date(BRTHDT,  origin = "1970-01-01"),
    RFSTDT_h = as.Date(RFSTDT,  origin = "1970-01-01"),
    RFXSTDT_h = as.Date(RFXSTDT, origin = "1970-01-01")
  ) |>
  select(USUBJID, BRTHDT_h, RFSTDT_h, RFXSTDT_h)
```

> **Read the output:** three new numeric date columns appear, all from a single derivation invocation pattern. Without `call_derivation()`, you would write `derive_vars_dt()` three times — 12-15 lines of near-duplicate code. The DRY win is small for three derivations but becomes large at 10+ parameters in ADLB.

### 10.9 Knowledge check

```{r quiz-sec10, echo=FALSE}
quiz(
  question(
    "`call_derivation()` is most useful when:",
    answer("Each derivation uses a totally different function"),
    answer("The same derivation function is applied to multiple parameter sets that share most arguments", correct = TRUE),
    answer("You want to bypass argument validation"),
    random_answer_order = TRUE
  ),
  question(
    "`params(dtc = EXSTDTC, new_vars_prefix = \"EXST\")` and `exprs(EXSTDTC, EXSEQ)` differ how?",
    answer("They are synonyms"),
    answer("`params()` captures named arguments for one iteration; `exprs()` captures a list of bare column-name expressions", correct = TRUE),
    answer("`params()` works only inside `call_derivation`; `exprs()` works only inside `derive_vars_*`"),
    random_answer_order = TRUE
  ),
  question(
    "A shared `date_imputation = \"first\"` argument is overridden in one `params()` block. Which value applies for that iteration?",
    answer("The shared value (`first`)"),
    answer("The per-iteration value — it shadows the shared value", correct = TRUE),
    answer("Admiral errors out — conflicting arguments are forbidden"),
    random_answer_order = TRUE
  ),
  question(
    "The closest SAS analog to `call_derivation()` is:",
    answer("`PROC SQL` with `UNION ALL`"),
    answer("A `%macro` invoked inside a `%DO i = 1 %TO N` loop with positional arguments", correct = TRUE),
    answer("`PROC TRANSPOSE`"),
    random_answer_order = TRUE
  )
)
```

---

## 11 — Key takeaways

```{r quiz-final-14, echo=FALSE}
quiz(
  question(
    "You need to get 'the most recent non-missing lab date from LB for each subject'. Which function?",
    answer("derive_vars_merged() with mode = 'last'", correct = TRUE),
    answer("derive_param_*()",
           message = "derive_param_* adds new parameter rows, not date columns"),
    answer("derive_vars_dy()"),
    random_answer_order = TRUE
  ),
  question(
    "A subject's RFXSTDTC is '2014-01'. With date_imputation = 'first', TRTSDT resolves to:",
    answer("2014-01-01", correct = TRUE),
    answer("2014-01-31"),
    answer("NA — partial dates aren't supported"),
    random_answer_order = TRUE
  ),
  question(
    "The exprs() function is used in admiral because:",
    answer("It's faster than character strings"),
    answer("It passes variable names as unquoted tidy-evaluation expressions, consistent with dplyr style", correct = TRUE),
    answer("It encrypts variable names for security"),
    random_answer_order = TRUE
  ),
  question(
    "derive_vars_merged() vs derive_vars_joined() — when should you prefer the former?",
    answer("When the filter condition uses only variables from the source dataset", correct = TRUE),
    answer("When the filter involves comparing dates from both datasets"),
    answer("Always — derive_vars_joined() is deprecated"),
    random_answer_order = TRUE
  ),
  question(
    "The `DTF` imputation flag produced by `derive_vars_dt()` is `NA` when:",
    answer("Imputation failed"),
    answer("The original DTC was complete — no day or month was missing, so no imputation was needed", correct = TRUE),
    answer("date_imputation = 'none' was specified"),
    random_answer_order = TRUE
  ),
  question(
    "`convert_blanks_to_na()` is always called first because SAS XPT files represent missing character values as:",
    answer("NA in R"),
    answer("NULL"),
    answer('Empty strings (\"\") — which R\'s is.na() returns FALSE for, silently corrupting downstream derivations', correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "`derive_vars_duration()` with `add_one = TRUE` is correct for treatment duration because:",
    answer("A subject treated for exactly one day (start == end) should have duration 1, not 0 — the inclusive counting convention requires adding 1", correct = TRUE),
    answer("CDISC forbids duration of zero"),
    answer("add_one adjusts for the no-Day-0 convention"),
    random_answer_order = TRUE
  ),
  question(
    "A subject has a study day of -3. This means the event occurred:",
    answer("3 days after the reference date"),
    answer("3 days before the reference date — Day -1 is immediately before Day 1, and Day -3 is two days further back", correct = TRUE),
    answer("On the reference date (Day 1 and Day -3 are equivalent)"),
    random_answer_order = TRUE
  ),
  question(
    "`derive_var_extreme_flag()` with `mode = 'first'` is used to derive ABLFL because:",
    answer("'first' means alphabetically first USUBJID"),
    answer("The baseline record is defined as the earliest (first in date order) pre-treatment observation", correct = TRUE),
    answer("ABLFL must always be the record with the highest ADT"),
    random_answer_order = TRUE
  ),
  question(
    "`call_derivation()` is most useful when:",
    answer("You need to debug a single derive_vars_dtm() call"),
    answer("You need to apply the same derivation function to multiple sets of variables (e.g., EXSTDTC and EXENDTC) with shared arguments, avoiding copy-paste", correct = TRUE),
    answer("You are calling admiral functions from a loop in base R"),
    random_answer_order = TRUE
  ),
  question(
    "Which admiral prefix tells you the function adds *one* variable as a *new row* (not a new column) to a BDS dataset?",
    answer("`derive_var_*`"),
    answer("`derive_vars_*`"),
    answer("`derive_param_*`", correct = TRUE),
    answer("`compute_*`"),
    random_answer_order = TRUE
  ),
  question(
    "A SAS programmer writes `PROC SORT BY USUBJID ADT;` then `DATA; SET; BY USUBJID; IF FIRST.USUBJID;`. The closest admiral equivalent is:",
    answer("`derive_vars_merged()` with `mode = \"first\"`"),
    answer("`derive_var_extreme_flag()` with `mode = \"first\"` — flags one row per group", correct = TRUE),
    answer("`derive_var_merged_exist_flag()`"),
    random_answer_order = TRUE
  ),
  question(
    "A subject's EXSTDTC partial date is `\"2014\"` (year only). With `date_imputation = \"first\"`, the resulting EXSTDT is:",
    answer("NA — admiral refuses year-only dates"),
    answer("2014-01-01 — admiral imputes month and day to January 1", correct = TRUE),
    answer("2014-12-31"),
    random_answer_order = TRUE
  ),
  question(
    "`derive_vars_dtm()` with `time_imputation = \"last\"` is correct when:",
    answer("You want to bias the datetime toward the start of the day"),
    answer("You want to bias the datetime toward the end of the day — typically for a treatment-window END boundary", correct = TRUE),
    answer("The source DTC is character"),
    random_answer_order = TRUE
  ),
  question(
    "Which is NOT a valid value of the `DTF` imputation flag?",
    answer("`\"D\"`"),
    answer("`\"M\"`"),
    answer("`\"Y\"`"),
    answer("`\"H\"` — that is for time imputation (TMF), not date (DTF)", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "In a multi-period crossover study, the closest SAS analog to admiral's APERIOD/APERSDT/APEREDT triple is:",
    answer("A `PROC FORMAT` lookup"),
    answer("A DATA-step with `IF date1 <= DT <= date2 THEN APERIOD = 1;` plus `ELSE IF date3 <= DT <= date4 THEN APERIOD = 2;`", correct = TRUE),
    answer("A `PROC LIFETEST` covariate"),
    random_answer_order = TRUE
  ),
  question(
    "`derive_var_merged_exist_flag(condition = AESER == \"Y\")` sets `new_var = NA` for subjects with no matching row. Is this CDISC-compliant?",
    answer("No — CDISC requires `\"N\"`"),
    answer("Yes — CDISC ADaM IG requires `\"Y\"` or `NA`; `\"N\"` is forbidden in flag variables", correct = TRUE),
    answer("Only for SAFFL — other flag variables can use `\"N\"`"),
    random_answer_order = TRUE
  ),
  question(
    "`derive_vars_merged()` with no `mode` argument and a source containing 3 rows per USUBJID will:",
    answer("Silently keep the first row (default behavior)"),
    answer("Error out — `check_type = \"error\"` is the default, refusing to silently duplicate the left side", correct = TRUE),
    answer("Cross-join 3× the base"),
    random_answer_order = TRUE
  ),
  question(
    "The SAS habit `IF condition THEN FL = \"Y\"; ELSE FL = \"N\";` violates CDISC because:",
    answer("ELSE branches are slow in SAS"),
    answer("`\"N\"` rows appear as explicit zeros in downstream PROC FREQ tabulations; CDISC requires absence (`NA`) so non-population subjects drop from the denominator", correct = TRUE),
    answer("`\"N\"` is reserved for non-numeric flags"),
    random_answer_order = TRUE
  )
)
```

---

## 12 — Lesson 14 Summary {.congratulations}

You now know how to:

- **Internalize the manifesto** — admiral is a toolbox of small, audit-friendly derivations, not a black-box ADaM factory. Each derive_* function does one identifiable thing, mapping cleanly to a SAS DATA step, PROC SQL clause, or macro.
- **Decode function names** — `derive_var_*` (one new column), `derive_vars_*` (multiple related columns), `derive_param_*` (a new BDS row), `compute_*` (a scalar value), `get_*` (a lookup), `filter_*` (a row subset). The prefix tells you 80% of what a function does.
- **Sanitize SAS XPT imports with `convert_blanks_to_na()`** — XPT character missing is `""`, not `.`. Run this defensively at the top of every admiral script to prevent silently wrong downstream derivations.
- **Convert ISO-8601 DTC strings to numeric dates with `derive_vars_dt()`** — handle partial-date imputation (`"first"` or `"last"`), and read the DTF flag (`"D"`/`"M"`/`"Y"`/`NA`) to know what was imputed. Mental analog: SAS `INPUT(DTC, YYMMDD10.)` plus a partial-date macro plus a flag variable, in one call.
- **Convert DTC to datetime with `derive_vars_dtm()`** — same as `dt` but adds time-of-day imputation and produces a POSIXct-like numeric. Use `time_imputation = "first"` for window starts, `"last"` for window ends.
- **Compute study day with `derive_vars_dy()`** — the no-Day-0 convention. Reference date is Day 1; the day before is Day -1; there is no Day 0. SAS analog: `INTCK('day', ref, dt) + 1` for non-negative differences.
- **Master `derive_vars_merged()`** — the workhorse left-join with built-in `filter_add`, `order`, `mode`, `new_vars`, and `missing_values`. One call replaces a PROC SORT + DATA-step MERGE + IF FIRST. processing.
- **Flag first/last records with `derive_var_extreme_flag()`** — the admiral counterpart to SAS `IF FIRST.x;` / `IF LAST.x;`. CDISC convention: `"Y"` or `NA`, never `"N"`.
- **Build population flags with `derive_var_merged_exist_flag()`** — the EXISTS subquery in admiral form. The CDISC `"Y"` or `NA` rule is enforced by the function's API (no `else_val` parameter).
- **DRY-up repeated derivations with `call_derivation()`** — equivalent to a SAS `%DO` loop wrapping a macro call. Shared arguments stay constant; per-iteration `params()` blocks override.
- **Map every admiral call to a SAS analog** — every derivation in Lesson 14 has a clear PROC SQL or DATA-step equivalent. The mental shift is from "imperative DATA-step chain" to "declarative named-argument call."

You also know:

- All flag variables follow the CDISC `"Y" or NA` rule (never `"N"`).
- All admiral expressions are wrapped in `exprs()` (column references) or `params()` (named argument sets) — tidy-eval primitives that mirror dplyr's NSE style.
- The 3-year deprecation cycle is admiral's promise: a function that exists today will at minimum warn (year 1), error (year 2), then disappear (year 3) — giving multi-year studies time to migrate.

---

## What's next

Lesson 15 — **`{admiral}` Part 2** — builds a full ADSL end-to-end using everything in this lesson. You'll see treatment dates, treatment duration, disposition dates, cause of death (multi-source), last known alive date, population flags, and age groups — all from the pharmaverse SDTM datasets. Every admiral function you've practiced here gets used in a real derivation context.
