---
title: "Lesson 14 — {admiral} Part 1: Foundations and Philosophy"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Admiral's design manifesto, function naming convention, exprs(), date/time
  helpers, and derive_vars_merged() — the foundation of every ADaM build.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 14_admiral_part1.Rmd
# PURPOSE       : admiral Part 1 — package architecture, derivation function taxonomy,
#                 template-driven workflow, derive_vars_merged(), and
#                 foundational ADaM dataset construction patterns.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── Mock admiral functions ────────────────────────────────────────────────────

# convert_blanks_to_na: replace "" with NA in all character columns
convert_blanks_to_na <- function(dataset) {
  dataset |>
    mutate(across(where(is.character), ~ if_else(.x == "", NA_character_, .x)))
}

# ── Date helpers ──────────────────────────────────────────────────────────────

# Impute a partial DTC string
.impute_dtc <- function(dtc, date_imputation = "first", time_imputation = "first") {
  sapply(dtc, function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    parts <- str_split(x, "T")[[1]]
    date_part <- parts[1]
    time_part <- if (length(parts) > 1) parts[2] else NA_character_

    # Pad partial dates
    segs <- str_split(date_part, "-")[[1]]
    if (length(segs) == 1) {
      segs <- c(segs, if (date_imputation == "last") "12" else "01")
    }
    if (length(segs) == 2) {
      last_day <- if (date_imputation == "last") {
        as.character(lubridate::days_in_month(as.Date(paste0(segs[1], "-", segs[2], "-01"))))
      } else "01"
      segs <- c(segs, last_day)
    }
    full_date <- paste(segs, collapse = "-")

    if (!is.na(time_part)) {
      paste0(full_date, "T", time_part)
    } else {
      full_date
    }
  }, USE.NAMES = FALSE)
}

.imputation_flag <- function(dtc, imputed_dtc, mode = "date") {
  sapply(seq_along(dtc), function(i) {
    if (is.na(dtc[i])) return(NA_character_)
    orig  <- str_split(dtc[i], "-")[[1]]
    if (length(orig) == 1) return(if (mode == "date") "Y" else "M")
    if (length(orig) == 2) return("D")
    return(NA_character_)
  }, USE.NAMES = FALSE)
}

# derive_vars_dt: convert DTC character → numeric Date + imputation flag
derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation = "none",
                            flag_imputation = "auto") {
  .e <- substitute(dtc); dtc_nm <- if (is.character(.e)) .e else deparse(.e)
  dt_nm   <- paste0(new_vars_prefix, "DT")
  dtf_nm  <- paste0(new_vars_prefix, "DTF")
  raw_dtc <- dataset[[dtc_nm]]
  imputed <- .impute_dtc(raw_dtc, date_imputation = date_imputation)
  dataset[[dt_nm]]  <- suppressWarnings(as.numeric(as.Date(imputed)))
  dataset[[dtf_nm]] <- .imputation_flag(raw_dtc, imputed, "date")
  dataset
}

# derive_vars_dtm: convert DTC → numeric POSIXct datetime + flags
derive_vars_dtm <- function(dataset, dtc, new_vars_prefix,
                             date_imputation = "none",
                             time_imputation = "none",
                             flag_imputation = "auto") {
  .e <- substitute(dtc); dtc_nm <- if (is.character(.e)) .e else deparse(.e)
  dtm_nm  <- paste0(new_vars_prefix, "DTM")
  dtf_nm  <- paste0(new_vars_prefix, "DTF")
  tmf_nm  <- paste0(new_vars_prefix, "TMF")
  raw_dtc <- dataset[[dtc_nm]]

  ti_str  <- switch(time_imputation, "first" = "00:00:00", "last" = "23:59:59", "00:00:00")
  imputed <- sapply(raw_dtc, function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    if (!str_detect(x, "T")) paste0(x, "T", ti_str) else x
  }, USE.NAMES = FALSE)
  imputed <- .impute_dtc(imputed, date_imputation = date_imputation)
  dataset[[dtm_nm]] <- suppressWarnings(as.numeric(as.POSIXct(str_replace(imputed, "T", " "), tz = "UTC")))
  dataset[[dtf_nm]] <- .imputation_flag(raw_dtc, imputed, "date")
  dataset[[tmf_nm]] <- if_else(!str_detect(raw_dtc, "T") & !is.na(raw_dtc),
                                if (time_imputation == "first") "H" else "H", NA_character_)
  dataset
}

# derive_vars_dtm_to_dt: datetime → date (strip time)
derive_vars_dtm_to_dt <- function(dataset, source_vars) {
  sv_names <- sapply(source_vars, deparse)
  for (nm in sv_names) {
    out_nm <- str_replace(nm, "DTM$", "DT")
    dataset[[out_nm]] <- trunc(dataset[[nm]] / 86400)  # seconds → days
  }
  dataset
}

# derive_vars_dy: compute study day variables (no Day 0)
derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref_nm  <- deparse(substitute(reference_date))
  sv_names <- sapply(source_vars, deparse)
  for (nm in sv_names) {
    out_nm <- str_replace(nm, "DT$", "DY")
    ref_days <- dataset[[ref_nm]]
    src_days <- dataset[[nm]]
    raw_diff <- src_days - ref_days
    dataset[[out_nm]] <- case_when(
      is.na(raw_diff)  ~ NA_real_,
      raw_diff >= 0    ~ raw_diff + 1,
      TRUE             ~ raw_diff
    )
  }
  dataset
}

# exprs: capture unevaluated arguments as a list of expressions (mirrors rlang::exprs)
exprs <- function(...) as.list(substitute(list(...)))[-1]

# derive_vars_merged: left-join with filter, order, first/last
derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL,
                                order = NULL, filter_add = NULL, mode = NULL,
                                missing_values = NULL) {
  by_names <- sapply(by_vars, deparse)

  # Apply filter_add
  filter_expr <- substitute(filter_add)
  if (!is.null(filter_expr) && deparse(filter_expr) != "NULL") {
    dataset_add <- dataset_add |> filter(!!filter_expr)
  }

  # Order
  if (!is.null(order)) {
    ord_names <- sapply(order, deparse)
    ord_descs <- grepl("^desc\\(", ord_names)
    ord_clean <- str_replace_all(ord_names, "^desc\\(|\\)$", "")
    for (i in rev(seq_along(ord_clean))) {
      if (ord_descs[i]) {
        dataset_add <- dataset_add |> arrange(desc(.data[[ord_clean[i]]]))
      } else {
        dataset_add <- dataset_add |> arrange(.data[[ord_clean[i]]])
      }
    }
  }

  # First/last selection
  if (!is.null(mode)) {
    fn <- if (mode == "first") dplyr::slice_head else dplyr::slice_tail
    dataset_add <- dataset_add |>
      group_by(across(all_of(by_names))) |>
      fn(n = 1) |>
      ungroup()
  }

  # Rename/select new vars
  if (!is.null(new_vars)) {
    nv_names <- sapply(new_vars, deparse)
    nv_map   <- setNames(names(nv_names), nv_names)  # new_name = old_name
    keep_cols <- c(by_names, names(nv_map))
    dataset_add <- dataset_add |>
      select(any_of(keep_cols)) |>
      rename(any_of(nv_map))
    new_col_names <- nv_names
  } else {
    new_col_names <- setdiff(names(dataset_add), by_names)
  }

  dataset |>
    left_join(
      dataset_add |> select(any_of(c(by_names, new_col_names))),
      by = by_names
    )
}

# derive_var_extreme_flag: mark first/last record per group
derive_var_extreme_flag <- function(dataset, by_vars, order, new_var, mode = "first") {
  by_names  <- sapply(by_vars, deparse)
  ord_names <- sapply(order, deparse)
  out_nm    <- deparse(substitute(new_var))
  ord_descs <- grepl("^desc\\(", ord_names)
  ord_clean <- str_replace_all(ord_names, "^desc\\(|\\)$", "")

  sorted <- dataset
  for (i in rev(seq_along(ord_clean))) {
    if (ord_descs[i]) {
      sorted <- sorted |> arrange(desc(.data[[ord_clean[i]]]))
    } else {
      sorted <- sorted |> arrange(.data[[ord_clean[i]]])
    }
  }

  flagged <- sorted |>
    group_by(across(all_of(by_names))) |>
    mutate(.row_rank = row_number()) |>
    ungroup() |>
    mutate(!!out_nm := case_when(
      (mode == "first" & .row_rank == 1) |
      (mode == "last"  & .row_rank == max(.row_rank)) ~ "Y",
      TRUE ~ NA_character_
    )) |>
    select(-.row_rank)

  # Merge flag back in original row order
  dataset |>
    left_join(flagged |> select(any_of(c(by_names, ord_clean, out_nm))),
              by = c(by_names, ord_clean))
}

# derive_var_merged_exist_flag: Y/NA based on existence in another dataset
derive_var_merged_exist_flag <- function(dataset, dataset_add, by_vars,
                                          new_var, condition) {
  by_names <- sapply(by_vars, deparse)
  out_nm   <- deparse(substitute(new_var))
  cond_expr <- substitute(condition)
  flagged  <- dataset_add |>
    filter(!!cond_expr) |>
    distinct(across(all_of(by_names))) |>
    mutate(!!out_nm := "Y")
  dataset |> left_join(flagged, by = by_names)
}

# derive_vars_duration: compute duration between two date variables
derive_vars_duration <- function(dataset, new_var, start_date, end_date,
                                  out_unit = "days", add_one = TRUE) {
  out_nm  <- deparse(substitute(new_var))
  start_nm <- deparse(substitute(start_date))
  end_nm   <- deparse(substitute(end_date))
  raw_diff <- dataset[[end_nm]] - dataset[[start_nm]]
  if (out_unit == "days") {
    dataset[[out_nm]] <- case_when(
      is.na(raw_diff) ~ NA_real_,
      add_one         ~ raw_diff + 1,
      TRUE            ~ raw_diff
    )
  } else if (out_unit == "months") {
    dataset[[out_nm]] <- round(as.numeric(raw_diff) / 30.4375, 1)
  }
  dataset
}

# derive_var_trtdurd: convenience wrapper for treatment duration
derive_var_trtdurd <- function(dataset, new_var = TRTDURD,
                                start_date = TRTSDT, end_date = TRTEDT) {
  out_nm   <- deparse(substitute(new_var))
  start_nm <- deparse(substitute(start_date))
  end_nm   <- deparse(substitute(end_date))
  dataset[[out_nm]] <- dataset[[end_nm]] - dataset[[start_nm]] + 1
  dataset
}

# call_derivation: apply one function multiple times with different variable params
call_derivation <- function(dataset, derivation, variable_params, ...) {
  shared_args <- list(...)
  for (vp in variable_params) {
    args_i <- c(list(dataset = dataset), vp, shared_args)
    dataset <- do.call(derivation, args_i)
  }
  dataset
}

# params: capture bare column names as strings so do.call() can pass them safely
params <- function(...) {
  dots <- as.list(substitute(list(...)))[-1]
  lapply(dots, function(x) if (is.symbol(x)) deparse(x) else x)
}

# ── Sample data ───────────────────────────────────────────────────────────────

# SDTM-like datasets with deliberate blanks (to test convert_blanks_to_na)
dm_raw <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101", "01-703-1211"),
  AGE     = c(63L, 64L, 71L, 56L, 45L),
  SEX     = c("M", "F", "M", "F", "M"),
  RACE    = c("WHITE", "WHITE", "ASIAN", "WHITE", "BLACK OR AFRICAN AMERICAN"),
  ARM     = c("Xanomeline High Dose", "Xanomeline Low Dose",
              "Placebo", "Xanomeline High Dose", "Placebo"),
  ACTARM  = c("XANOMELINE HIGH DOSE", "XANOMELINE LOW DOSE",
              "PLACEBO", "XANOMELINE HIGH DOSE", "PLACEBO"),
  RFXSTDTC = c("2014-01-02", "2012-08-05", "2013-07-09", "2013-04-14", "2014-03-22")
)

ex_raw <- tibble(
  STUDYID  = "CDISCPILOT01",
  USUBJID  = c("01-701-1015", "01-701-1015",
               "01-701-1023", "01-702-1042", "01-702-1101",
               "01-703-1211", "01-703-1211"),
  EXTRT    = c("XANOMELINE", "XANOMELINE",
               "XANOMELINE", "PLACEBO", "XANOMELINE",
               "PLACEBO", "PLACEBO"),
  EXDOSE   = c(81, 81, 54, 0, 81, 0, 0),
  EXSEQ    = c(1L, 2L, 1L, 1L, 1L, 1L, 2L),
  EXSTDTC  = c("2014-01-02", "2014-02-02",
               "2012-08-05", "2013-07-09", "2013-04-14",
               "2014-03-22", "2014-04-22"),
  EXENDTC  = c("2014-01-31", "2014-02-28",
               "2012-11-01", "2013-09-15", "2013-06-30",
               "2014-04-21", "2014-05-21")
)

# LB for study day derivation example
lb_raw <- tibble(
  STUDYID  = "CDISCPILOT01",
  USUBJID  = c("01-701-1015", "01-701-1015", "01-701-1023"),
  LBTESTCD = c("ALT", "ALT", "ALT"),
  LBDTC    = c("2014-01-02", "2014-02-10", "2012-08-05"),
  LBORRES  = c("28", "35", "22")
)

# Datasets with blanks (to demo convert_blanks_to_na)
dm_with_blanks <- dm_raw |>
  mutate(ARM = if_else(USUBJID == "01-703-1211", "", ARM))

# Pre-processed clean versions
dm  <- convert_blanks_to_na(dm_raw)
ex  <- convert_blanks_to_na(ex_raw)
lb  <- convert_blanks_to_na(lb_raw)

# ── Real pharmaversesdtm data ─────────────────────────────────────────────────
library(pharmaversesdtm)
dm_real <- pharmaversesdtm::dm
ex_real <- pharmaversesdtm::ex
lb_real <- pharmaversesdtm::lb
```

## Welcome

This is Lesson 14 — **`{admiral}` Part 1: Foundations and Philosophy**.

Admiral is the pharmaverse flagship for building ADaM datasets. Before writing a single derivation, you need to internalize two things: the manifesto (admiral is a *toolbox*, not a black box) and the function naming convention (which tells you 80% of what a function does before you read the docs).

This lesson covers:

- The admiral manifesto — why it matters
- The function prefix convention: `derive_var_*`, `derive_vars_*`, `derive_param_*`, `compute_*`
- `exprs()` — the expression argument pattern
- `convert_blanks_to_na()` — why it's always your first call
- Date/time helpers: `derive_vars_dt()`, `derive_vars_dtm()`, `derive_vars_dy()`
- `derive_vars_merged()` — the most important function
- `derive_var_extreme_flag()` and `derive_var_merged_exist_flag()`

All admiral functions are mocked and run without package installation.

---

## 1 — The manifesto

```{r quiz-manifesto, echo=FALSE}
quiz(
  question(
    "The admiral maintainers describe the package as 'a toolbox, not an automation tool'. This means:",
    answer("You call make_adsl() and admiral builds the dataset for you"),
    answer("You assemble ADaMs by chaining many small, auditable function calls — no black-box 'make ADSL' functions", correct = TRUE),
    answer("admiral only handles one ADaM at a time"),
    random_answer_order = TRUE
  ),
  question(
    "Why does the manifesto emphasize 'clear purpose, no multi-purpose functions'?",
    answer("Regulators require single-purpose functions"),
    answer("It makes function names predictable and code auditable — each line does exactly one identifiable thing", correct = TRUE),
    answer("Multi-purpose functions are slower in R"),
    random_answer_order = TRUE
  ),
  question(
    "Admiral's 3-year deprecation cycle (warning → error → removal) exists because:",
    answer("The development team needs time to update examples"),
    answer("Studies run for years; a mid-study package update should never break existing code without ample warning time", correct = TRUE),
    answer("CRAN requires 3-year cycles"),
    random_answer_order = TRUE
  )
)
```

---

## 2 — Function naming convention

```{r quiz-naming, echo=FALSE}
quiz(
  question(
    "derive_vars_dt() (plural 'vars') vs derive_var_trtdurd() (singular 'var') — the difference is:",
    answer("'vars' functions add multiple related variables at once; 'var' functions add a single variable", correct = TRUE),
    answer("'vars' is newer than 'var'"),
    answer("'var' functions are faster"),
    random_answer_order = TRUE
  ),
  question(
    "derive_param_bmi() belongs to which prefix family?",
    answer("derive_var_* — adding one variable"),
    answer("derive_param_* — adding a new parameter row to a BDS dataset", correct = TRUE),
    answer("compute_* — pure scalar computation"),
    random_answer_order = TRUE
  ),
  question(
    "compute_age_years() returns:",
    answer("A dataset with a new AGE column"),
    answer("A scalar numeric value without modifying a dataset", correct = TRUE),
    answer("A plot of age distribution"),
    random_answer_order = TRUE
  )
)
```

---

## 3 — `convert_blanks_to_na()`: always your first call

SAS XPT files store missing character values as `""` (empty string), not `NA`. R's `is.na()` returns `FALSE` for empty strings — so downstream derivations silently produce wrong results if you skip this step.

```{r blanks-demo}
# Demonstrate the problem
cat("Empty string is.na test:", is.na(""), "\n")
cat("Proper NA is.na test:    ", is.na(NA), "\n")

# dm_with_blanks has ARM = "" for subject 01-703-1211
dm_with_blanks |>
  filter(USUBJID == "01-703-1211") |>
  select(USUBJID, ARM)
```

```{r blanks-fix}
# After convert_blanks_to_na()
dm_fixed <- convert_blanks_to_na(dm_with_blanks)

dm_fixed |>
  filter(USUBJID == "01-703-1211") |>
  select(USUBJID, ARM)

cat("ARM is NA after conversion:", is.na(dm_fixed$ARM[dm_fixed$USUBJID == "01-703-1211"]), "\n")
```

### Exercise: Convert blanks and verify with is.na

```{r blanks-exercise, exercise=TRUE, exercise.lines=14}
# Create a dataset with blanks in multiple columns
test_ds <- tibble(
  USUBJID = c("S001", "S002", "S003"),
  ARM     = c("TRT", "", "TRT"),
  RACE    = c("WHITE", "ASIAN", ""),
  AGE     = c(45L, NA_integer_, 62L)
)

# Convert blanks
test_clean <- convert_blanks_to_na(test_ds)

# Check: how many NAs in ARM before and after?
cat("NAs in ARM (before):", sum(is.na(test_ds$ARM)), "\n")
cat("NAs in ARM (after): ", sum(is.na(test_clean$ARM)), "\n")
cat("NAs in RACE (after):", sum(is.na(test_clean$RACE)), "\n")
test_clean
```

### Exercise: Blank audit on real pharmaversesdtm DM

The real `dm_real` arrives as if read from a SAS XPT. Check whether it carries any blank strings, then apply `convert_blanks_to_na()` defensively. This pattern belongs at the top of every admiral script regardless of whether blanks are expected.

```{r blanks-real-dm, exercise=TRUE, exercise.lines=18}
# Count blank strings per character column in dm_real
blank_counts <- dm_real |>
  select(where(is.character)) |>
  summarise(across(everything(), ~ sum(.x == "", na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_blanks") |>
  filter(n_blanks > 0)

cat("Columns with blank strings in dm_real:\n")
if (nrow(blank_counts) == 0) cat("  (none found — dataset is already clean)\n") else print(blank_counts)

# Apply defensively anyway
dm_real_clean <- convert_blanks_to_na(dm_real)

# How many NAs exist per character column?
na_summary <- sapply(select(dm_real_clean, where(is.character)),
                     function(x) sum(is.na(x)))
cat("\nNA counts in character columns after conversion:\n")
print(na_summary[na_summary > 0])
```

```{r blanks-real-dm-hint-1}
# summarise(across(everything(), ~ sum(.x == "", na.rm = TRUE)))
# The real value: even if zero blanks exist today, a study database refresh
# could introduce them — calling convert_blanks_to_na() costs nothing.
```

```{r blanks-real-dm-solution}
blank_counts <- dm_real |>
  select(where(is.character)) |>
  summarise(across(everything(), ~ sum(.x == "", na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_blanks") |>
  filter(n_blanks > 0)

if (nrow(blank_counts) == 0) cat("(none found)\n") else print(blank_counts)

dm_real_clean <- convert_blanks_to_na(dm_real)

na_summary <- sapply(select(dm_real_clean, where(is.character)),
                     function(x) sum(is.na(x)))
print(na_summary[na_summary > 0])
```

---

## 4 — `derive_vars_dt()`: ISO 8601 DTC → numeric Date

The workhorse for converting character SDTM dates to numeric dates suitable for arithmetic.

```{r derive-dt-demo}
# Pre-process DM to get a numeric date from RFXSTDTC
dm_dates <- dm |>
  derive_vars_dt(
    new_vars_prefix  = "RFXST",
    dtc              = RFXSTDTC,
    date_imputation  = "first"
  )

dm_dates |>
  select(USUBJID, RFXSTDTC, RFXSTDT, RFXSTDTF) |>
  mutate(RFXSTDT_human = as.Date(RFXSTDT, origin = "1970-01-01"))
```

```{r derive-dt-partial}
# With a partial date — imputation kicks in
partial_dm <- tibble(
  USUBJID  = c("S001", "S002"),
  STARTDTC = c("2014-07", "2014-07-15")   # S001 has no day
)

partial_dm |>
  derive_vars_dt(
    new_vars_prefix = "START",
    dtc = STARTDTC,
    date_imputation = "first"
  ) |>
  mutate(STARTDT_human = as.Date(STARTDT, origin = "1970-01-01"))
```

### Exercise: Derive a date with "last" imputation

Impute the same partial date dataset using `date_imputation = "last"`. Subject S001's date should now resolve to the last day of July 2014.

```{r derive-dt-last, exercise=TRUE, exercise.lines=12}
partial_dm2 <- tibble(
  USUBJID  = c("S001", "S002"),
  STARTDTC = c("2014-07", "2014-07-15")
)

partial_dm2 |>
  derive_vars_dt(
    new_vars_prefix = "START",
    dtc             = STARTDTC,
    date_imputation = ___
  ) |>
  mutate(STARTDT_human = as.Date(STARTDT, origin = "1970-01-01"))
```

```{r derive-dt-last-hint-1}
# date_imputation = "last"
# Expected: S001 → 2014-07-31
```

```{r derive-dt-last-solution}
partial_dm2 <- tibble(
  USUBJID  = c("S001", "S002"),
  STARTDTC = c("2014-07", "2014-07-15")
)

partial_dm2 |>
  derive_vars_dt(
    new_vars_prefix = "START",
    dtc             = STARTDTC,
    date_imputation = "last"
  ) |>
  mutate(STARTDT_human = as.Date(STARTDT, origin = "1970-01-01"))
```

### Exercise: Derive numeric dates from real pharmaversesdtm DM

`dm_real` contains `RFXSTDTC` — the reference date for first treatment exposure. Derive `RFXSTDT` from it, then check how many subjects had partial dates that triggered the imputation flag.

```{r derive-dt-real, exercise=TRUE, exercise.lines=18}
# Preview the raw RFXSTDTC values
cat("RFXSTDTC sample (first 8):\n")
cat(paste(head(dm_real$RFXSTDTC, 8), collapse = ", "), "\n\n")

# How many look partial (< 10 chars = missing day or month)?
partial_n <- sum(nchar(dm_real$RFXSTDTC) < 10 & !is.na(dm_real$RFXSTDTC))
cat("Partial RFXSTDTC values:", partial_n, "\n\n")

# Derive RFXSTDT (numeric date) with first-day imputation
dm_real_dates <- dm_real |>
  derive_vars_dt(
    dtc             = ___,
    new_vars_prefix = "RFXST",
    date_imputation = "first"
  )

# How many imputation flags were set?
cat("RFXSTDTF imputation flag counts:\n")
table(dm_real_dates$RFXSTDTF, useNA = "ifany")
```

```{r derive-dt-real-hint-1}
# dtc = RFXSTDTC
# A DTF flag of NA means the date was complete — no imputation needed.
# A flag of "D" means the day was missing and imputed.
```

```{r derive-dt-real-solution}
cat("RFXSTDTC sample (first 8):\n")
cat(paste(head(dm_real$RFXSTDTC, 8), collapse = ", "), "\n\n")

partial_n <- sum(nchar(dm_real$RFXSTDTC) < 10 & !is.na(dm_real$RFXSTDTC))
cat("Partial RFXSTDTC values:", partial_n, "\n\n")

dm_real_dates <- dm_real |>
  derive_vars_dt(
    dtc             = RFXSTDTC,
    new_vars_prefix = "RFXST",
    date_imputation = "first"
  )

cat("RFXSTDTF imputation flag counts:\n")
table(dm_real_dates$RFXSTDTF, useNA = "ifany")
```

---

## 5 — `derive_vars_dtm()`: DTC → numeric Datetime

For treatment-window calculations you often need datetimes (not just dates), because two records can share a date but have different times of day.

```{r derive-dtm-demo}
# Pre-process EX with start and end datetimes
ex_ext <- ex |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first",
    time_imputation = "first"   # impute missing time to 00:00:00
  ) |>
  derive_vars_dtm(
    dtc             = EXENDTC,
    new_vars_prefix = "EXEN",
    date_imputation = "last",
    time_imputation = "last"    # impute missing time to 23:59:59
  )

ex_ext |>
  select(USUBJID, EXSTDTC, EXSTDTM, EXSTTMF, EXENDTC, EXENDTM)
```

```{r dtm-to-dt-demo}
# Convert datetimes back to dates (strip time component)
ex_ext2 <- ex_ext |>
  derive_vars_dtm_to_dt(source_vars = exprs(EXSTDTM, EXENDTM))

ex_ext2 |>
  select(USUBJID, EXSTDTM, EXSTDT, EXENDTM, EXENDT) |>
  head(4)
```

---

## 6 — `derive_vars_dy()`: study day (no Day 0)

CDISC ADaM convention: the day of the reference event is Day 1. The day before is Day -1. There is no Day 0.

```{r study-day-demo}
# First: derive numeric analysis dates for LB
lb_dates <- lb |>
  derive_vars_dt(dtc = LBDTC, new_vars_prefix = "A")

# Then merge TRTSDT onto LB (from dm) and derive ADY
lb_with_ref <- lb_dates |>
  left_join(
    dm |>
      derive_vars_dt(dtc = RFXSTDTC, new_vars_prefix = "TRTS") |>
      select(USUBJID, TRTSDT),
    by = "USUBJID"
  )

lb_with_day <- lb_with_ref |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ADT)
  )

lb_with_day |>
  mutate(
    ADT_human    = as.Date(ADT,    origin = "1970-01-01"),
    TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")
  ) |>
  select(USUBJID, ADT_human, TRTSDT_human, ADY)
```

### Exercise: Explain the "no Day 0" convention

```{r quiz-study-day, echo=FALSE}
quiz(
  question(
    "A subject's treatment starts on 2014-01-02. An AE occurs on 2014-01-02. What is the study day of the AE?",
    answer("Day 0"),
    answer("Day 1 — the reference date itself is always Day 1", correct = TRUE),
    answer("Day -1"),
    random_answer_order = TRUE
  ),
  question(
    "An AE occurs one day BEFORE the reference date (2014-01-01, with ref = 2014-01-02). The study day is:",
    answer("Day 0"),
    answer("Day -1 — there is no Day 0; the day before Day 1 is Day -1", correct = TRUE),
    answer("Day -2"),
    random_answer_order = TRUE
  )
)
```

### Exercise: Derive study day for an EX dataset

```{r study-day-ex, exercise=TRUE, exercise.lines=18}
# Add numeric date to EX
ex_with_dates <- ex |>
  derive_vars_dt(dtc = EXSTDTC, new_vars_prefix = "A")

# Merge TRTSDT from DM (first exposure date per subject)
trtsdt_lookup <- dm |>
  derive_vars_dt(dtc = RFXSTDTC, new_vars_prefix = "TRTS") |>
  select(USUBJID, TRTSDT)

ex_study_day <- ex_with_dates |>
  left_join(trtsdt_lookup, by = "USUBJID") |>
  derive_vars_dy(
    reference_date = ___,
    source_vars    = exprs(___)
  )

ex_study_day |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01"),
         TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")) |>
  select(USUBJID, EXSTDTC, ADT_human, TRTSDT_human, ADY)
```

```{r study-day-ex-hint-1}
# reference_date = TRTSDT
# source_vars = exprs(ADT)
```

```{r study-day-ex-solution}
ex_with_dates <- ex |>
  derive_vars_dt(dtc = EXSTDTC, new_vars_prefix = "A")

trtsdt_lookup <- dm |>
  derive_vars_dt(dtc = RFXSTDTC, new_vars_prefix = "TRTS") |>
  select(USUBJID, TRTSDT)

ex_study_day <- ex_with_dates |>
  left_join(trtsdt_lookup, by = "USUBJID") |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ADT)
  )

ex_study_day |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01"),
         TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")) |>
  select(USUBJID, EXSTDTC, ADT_human, TRTSDT_human, ADY)
```

---

## 7 — `derive_vars_merged()`: the most important function

`derive_vars_merged()` is the left-join equivalent with filter, ordering, and first/last selection built in. You'll use it for almost every variable that comes from a non-DM source.

```{r derive-merged-demo}
# Merge TRTSDT from DM onto a base dataset
adsl_base <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM)

ex_dtm <- ex |>
  derive_vars_dtm(dtc = EXSTDTC, new_vars_prefix = "EXST",
                  date_imputation = "first", time_imputation = "first") |>
  derive_vars_dtm(dtc = EXENDTC, new_vars_prefix = "EXEN",
                  date_imputation = "last",  time_imputation = "last")

# Derive TRTSDTM: first qualifying exposure datetime
adsl_trt <- adsl_base |>
  derive_vars_merged(
    dataset_add = ex_dtm,
    filter_add  = EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")),
    new_vars    = exprs(TRTSDTM = EXSTDTM),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_trt |>
  select(USUBJID, ACTARM, TRTSDTM) |>
  mutate(TRTSDTM_human = as.POSIXct(TRTSDTM, origin = "1970-01-01", tz = "UTC"))
```

### Exercise: Derive TRTEDTM (last qualifying exposure end datetime)

Use `mode = "last"` and `EXENDTM` instead of `EXSTDTM`.

```{r derive-merged-end, exercise=TRUE, exercise.lines=18}
adsl_trte <- adsl_trt |>
  derive_vars_merged(
    dataset_add = ex_dtm,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXENDTM),
    new_vars    = exprs(TRTEDTM = ___),
    order       = exprs(___, EXSEQ),
    mode        = ___,
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_trte |>
  select(USUBJID, TRTSDTM, TRTEDTM) |>
  mutate(
    TRTS_human = as.POSIXct(TRTSDTM, origin = "1970-01-01", tz = "UTC"),
    TRTE_human = as.POSIXct(TRTEDTM, origin = "1970-01-01", tz = "UTC")
  )
```

```{r derive-merged-end-hint-1}
# new_vars = exprs(TRTEDTM = EXENDTM)
# order    = exprs(EXENDTM, EXSEQ)
# mode     = "last"
```

```{r derive-merged-end-solution}
adsl_trte <- adsl_trt |>
  derive_vars_merged(
    dataset_add = ex_dtm,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXENDTM),
    new_vars    = exprs(TRTEDTM = EXENDTM),
    order       = exprs(EXENDTM, EXSEQ),
    mode        = "last",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_trte |>
  select(USUBJID, TRTSDTM, TRTEDTM) |>
  mutate(
    TRTS_human = as.POSIXct(TRTSDTM, origin = "1970-01-01", tz = "UTC"),
    TRTE_human = as.POSIXct(TRTEDTM, origin = "1970-01-01", tz = "UTC")
  )
```

### Exercise: derive_vars_merged on real pharmaversesdtm data

Build a minimal ADSL skeleton from `dm_real` and derive `TRTSDT` (first treatment start date) from `ex_real`. This mirrors the first `derive_vars_merged()` call in every real ADSL program.

```{r derive-merged-real, exercise=TRUE, exercise.lines=22}
# Step 1: convert blanks and add numeric EXSTDT to real EX
ex_real_dt <- convert_blanks_to_na(ex_real) |>
  derive_vars_dt(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first"
  )

# Step 2: ADSL skeleton — keep first 10 treated subjects from real DM
adsl_real_skel <- convert_blanks_to_na(dm_real) |>
  filter(!is.na(RFXSTDTC)) |>
  slice_head(n = 10) |>
  select(STUDYID, USUBJID, ARM, ACTARM)

# Step 3: merge first exposure date per subject
adsl_real_trts <- adsl_real_skel |>
  derive_vars_merged(
    dataset_add = ex_real_dt,
    filter_add  = !is.na(EXSTDT),
    new_vars    = exprs(TRTSDT = ___),
    order       = exprs(___),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_real_trts |>
  mutate(TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")) |>
  select(USUBJID, ARM, TRTSDT_human)
```

```{r derive-merged-real-hint-1}
# new_vars = exprs(TRTSDT = EXSTDT)
# order    = exprs(EXSTDT)
```

```{r derive-merged-real-solution}
ex_real_dt <- convert_blanks_to_na(ex_real) |>
  derive_vars_dt(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first"
  )

adsl_real_skel <- convert_blanks_to_na(dm_real) |>
  filter(!is.na(RFXSTDTC)) |>
  slice_head(n = 10) |>
  select(STUDYID, USUBJID, ARM, ACTARM)

adsl_real_trts <- adsl_real_skel |>
  derive_vars_merged(
    dataset_add = ex_real_dt,
    filter_add  = !is.na(EXSTDT),
    new_vars    = exprs(TRTSDT = EXSTDT),
    order       = exprs(EXSTDT),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_real_trts |>
  mutate(TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01")) |>
  select(USUBJID, ARM, TRTSDT_human)
```

---

## 8 — `derive_var_extreme_flag()`: first/last record flags

```{r extreme-flag-demo}
# Add an analysis date to LB, then flag the baseline (first) record per subject per test
lb_flagged <- lb |>
  derive_vars_dt(dtc = LBDTC, new_vars_prefix = "A") |>
  derive_var_extreme_flag(
    by_vars  = exprs(STUDYID, USUBJID, LBTESTCD),
    order    = exprs(ADT),
    new_var  = ABLFL,
    mode     = "first"
  )

lb_flagged |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01")) |>
  select(USUBJID, LBTESTCD, ADT_human, ABLFL)
```

### Exercise: Flag the LAST (most recent) lab record per subject per test

Change `mode` to `"last"` to flag the most recent observation instead.

```{r extreme-flag-last, exercise=TRUE, exercise.lines=14}
lb_last_flagged <- lb |>
  derive_vars_dt(dtc = LBDTC, new_vars_prefix = "A") |>
  derive_var_extreme_flag(
    by_vars = exprs(STUDYID, USUBJID, LBTESTCD),
    order   = exprs(ADT),
    new_var = LAST_FL,
    mode    = ___
  )

lb_last_flagged |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01")) |>
  select(USUBJID, LBTESTCD, ADT_human, LAST_FL)
```

```{r extreme-flag-last-hint-1}
# mode = "last"
```

```{r extreme-flag-last-solution}
lb_last_flagged <- lb |>
  derive_vars_dt(dtc = LBDTC, new_vars_prefix = "A") |>
  derive_var_extreme_flag(
    by_vars = exprs(STUDYID, USUBJID, LBTESTCD),
    order   = exprs(ADT),
    new_var = LAST_FL,
    mode    = "last"
  )

lb_last_flagged |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01")) |>
  select(USUBJID, LBTESTCD, ADT_human, LAST_FL)
```

---

## 9 — `derive_var_merged_exist_flag()`: population flags

Population flags (SAFFL, ITTFL, RANDFL) take the value `"Y"` if a condition is met in another dataset, or `NA` otherwise. Never `"N"` — CDISC ADaM IG specifies this.

```{r exist-flag-demo}
adsl_for_saffl <- adsl_base |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))
  )

adsl_for_saffl |> select(USUBJID, ACTARM, SAFFL)
```

```{r quiz-flag-convention, echo=FALSE}
quiz(
  question(
    "CDISC ADaM IG specifies that flag variables like SAFFL take values:",
    answer('"Y" if met, "N" if not'),
    answer('"Y" if met, NA if not — never use "N"', correct = TRUE),
    answer("1 if met, 0 if not"),
    random_answer_order = TRUE
  )
)
```

### Exercise: Derive ITTFL based on whether RANDDT exists

A subject is in the ITT population if they have a randomization date (RANDDT is not NA). For now, we'll simulate RANDDT from DM:

```{r ittfl-exercise, exercise=TRUE, exercise.lines=16}
# Create a mock randomization event dataset
rand_events <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101"),
  # 01-703-1211 is a screen failure — no randomization
  RANDDT  = as.Date(c("2014-01-02", "2012-08-05", "2013-07-09", "2013-04-14"))
)

# Derive ITTFL: "Y" for subjects with a randomization event
adsl_ittfl <- adsl_base |>
  derive_var_merged_exist_flag(
    dataset_add = rand_events,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = ITTFL,
    condition   = !is.na(___)
  )

adsl_ittfl |> select(USUBJID, ACTARM, ITTFL)
```

```{r ittfl-exercise-hint-1}
# condition = !is.na(RANDDT)
```

```{r ittfl-exercise-solution}
rand_events <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101"),
  RANDDT  = as.Date(c("2014-01-02", "2012-08-05", "2013-07-09", "2013-04-14"))
)

adsl_ittfl <- adsl_base |>
  derive_var_merged_exist_flag(
    dataset_add = rand_events,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = ITTFL,
    condition   = !is.na(RANDDT)
  )

adsl_ittfl |> select(USUBJID, ACTARM, ITTFL)
```

---

## 10 — `call_derivation()`: DRY derivation of multiple variables

When you need to apply the same function multiple times with slightly different arguments, `call_derivation()` keeps the code concise.

```{r call-derivation-demo}
# Instead of two separate derive_vars_dtm() calls:
ex_both_dtm <- ex |>
  call_derivation(
    derivation = derive_vars_dtm,
    variable_params = list(
      params(dtc = EXSTDTC, new_vars_prefix = "EXST"),
      params(dtc = EXENDTC, new_vars_prefix = "EXEN", time_imputation = "last")
    ),
    date_imputation = "first"   # shared arg
  )

ex_both_dtm |>
  select(USUBJID, EXSTDTC, EXSTDTM, EXENDTC, EXENDTM) |>
  head(3)
```

---

## 11 — Key takeaways

```{r quiz-final-14, echo=FALSE}
quiz(
  question(
    "You need to get 'the most recent non-missing lab date from LB for each subject'. Which function?",
    answer("derive_vars_merged() with mode = 'last'", correct = TRUE),
    answer("derive_param_*()",
           message = "derive_param_* adds new parameter rows, not date columns"),
    answer("derive_vars_dy()"),
    random_answer_order = TRUE
  ),
  question(
    "A subject's RFXSTDTC is '2014-01'. With date_imputation = 'first', TRTSDT resolves to:",
    answer("2014-01-01", correct = TRUE),
    answer("2014-01-31"),
    answer("NA — partial dates aren't supported"),
    random_answer_order = TRUE
  ),
  question(
    "The exprs() function is used in admiral because:",
    answer("It's faster than character strings"),
    answer("It passes variable names as unquoted tidy-evaluation expressions, consistent with dplyr style", correct = TRUE),
    answer("It encrypts variable names for security"),
    random_answer_order = TRUE
  ),
  question(
    "derive_vars_merged() vs derive_vars_joined() — when should you prefer the former?",
    answer("When the filter condition uses only variables from the source dataset", correct = TRUE),
    answer("When the filter involves comparing dates from both datasets"),
    answer("Always — derive_vars_joined() is deprecated"),
    random_answer_order = TRUE
  ),
  question(
    "The `DTF` imputation flag produced by `derive_vars_dt()` is `NA` when:",
    answer("Imputation failed"),
    answer("The original DTC was complete — no day or month was missing, so no imputation was needed", correct = TRUE),
    answer("date_imputation = 'none' was specified"),
    random_answer_order = TRUE
  ),
  question(
    "`convert_blanks_to_na()` is always called first because SAS XPT files represent missing character values as:",
    answer("NA in R"),
    answer("NULL"),
    answer('Empty strings (\"\") — which R\'s is.na() returns FALSE for, silently corrupting downstream derivations', correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "`derive_vars_duration()` with `add_one = TRUE` is correct for treatment duration because:",
    answer("A subject treated for exactly one day (start == end) should have duration 1, not 0 — the inclusive counting convention requires adding 1", correct = TRUE),
    answer("CDISC forbids duration of zero"),
    answer("add_one adjusts for the no-Day-0 convention"),
    random_answer_order = TRUE
  ),
  question(
    "A subject has a study day of -3. This means the event occurred:",
    answer("3 days after the reference date"),
    answer("3 days before the reference date — Day -1 is immediately before Day 1, and Day -3 is two days further back", correct = TRUE),
    answer("On the reference date (Day 1 and Day -3 are equivalent)"),
    random_answer_order = TRUE
  ),
  question(
    "`derive_var_extreme_flag()` with `mode = 'first'` is used to derive ABLFL because:",
    answer("'first' means alphabetically first USUBJID"),
    answer("The baseline record is defined as the earliest (first in date order) pre-treatment observation", correct = TRUE),
    answer("ABLFL must always be the record with the highest ADT"),
    random_answer_order = TRUE
  ),
  question(
    "`call_derivation()` is most useful when:",
    answer("You need to debug a single derive_vars_dtm() call"),
    answer("You need to apply the same derivation function to multiple sets of variables (e.g., EXSTDTC and EXENDTC) with shared arguments, avoiding copy-paste", correct = TRUE),
    answer("You are calling admiral functions from a loop in base R"),
    random_answer_order = TRUE
  )
)
```

---

## What's next

Lesson 15 — **`{admiral}` Part 2** — builds a full ADSL end-to-end using everything in this lesson. You'll see treatment dates, treatment duration, disposition dates, cause of death (multi-source), last known alive date, population flags, and age groups — all from the pharmaverse SDTM datasets. Every admiral function you've practiced here gets used in a real derivation context.
