---
title: "Lesson 15 — {admiral} Part 2: Building ADSL End-to-End"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Build a production-grade ADSL step by step: treatment dates, durations,
  disposition dates, death cause, population flags, and age groups.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 15_admiral_part2_adsl.Rmd
# PURPOSE       : admiral Part 2 — ADSL derivation: treatment variables, disposition
#                 endpoints, baseline flags (SAFFL, ITTFL), study day
#                 calculations, and population-level subject-level
#                 analysis dataset construction.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# exprs: capture unevaluated arguments as a list of expressions (mirrors rlang::exprs)
exprs <- function(...) as.list(substitute(list(...)))[-1]

# ── Mock admiral functions (same as Lesson 14 + new ones) ────────────────────

convert_blanks_to_na <- function(dataset) {
  dataset |>
    mutate(across(where(is.character), ~ if_else(.x == "", NA_character_, .x)))
}

.impute_dtc <- function(dtc, date_imputation = "first") {
  sapply(dtc, function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    segs <- str_split(str_split(x, "T")[[1]][1], "-")[[1]]
    if (length(segs) == 1) segs <- c(segs, if (date_imputation == "last") "12" else "01")
    if (length(segs) == 2) {
      d <- if (date_imputation == "last")
        as.character(days_in_month(as.Date(paste0(segs[1], "-", segs[2], "-01")))) else "01"
      segs <- c(segs, d)
    }
    paste(segs, collapse = "-")
  }, USE.NAMES = FALSE)
}

derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation = "none", flag_imputation = "auto") {
  dtc_nm <- deparse(substitute(dtc))
  dt_nm  <- paste0(new_vars_prefix, "DT")
  dtf_nm <- paste0(new_vars_prefix, "DTF")
  raw    <- dataset[[dtc_nm]]
  imp    <- .impute_dtc(raw, date_imputation)
  dataset[[dt_nm]]  <- suppressWarnings(as.numeric(as.Date(imp)))
  dataset[[dtf_nm]] <- sapply(raw, function(x) {
    if (is.na(x)) return(NA_character_)
    n <- length(str_split(x, "-")[[1]])
    if (n == 1) "Y" else if (n == 2) "D" else NA_character_
  }, USE.NAMES = FALSE)
  dataset
}

derive_vars_dtm <- function(dataset, dtc, new_vars_prefix,
                             date_imputation = "none", time_imputation = "none",
                             flag_imputation = "auto") {
  dtc_nm <- deparse(substitute(dtc))
  dtm_nm <- paste0(new_vars_prefix, "DTM")
  dtf_nm <- paste0(new_vars_prefix, "DTF")
  tmf_nm <- paste0(new_vars_prefix, "TMF")
  raw    <- dataset[[dtc_nm]]
  ti_str <- switch(time_imputation, "first" = "00:00:00", "last" = "23:59:59", "00:00:00")
  imp    <- sapply(raw, function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    x2 <- if (!str_detect(x, "T")) paste0(x, "T", ti_str) else x
    dp <- .impute_dtc(str_split(x2, "T")[[1]][1], date_imputation)
    tp <- str_split(x2, "T")[[1]][2]
    paste0(dp, "T", tp)
  }, USE.NAMES = FALSE)
  dataset[[dtm_nm]] <- suppressWarnings(
    as.numeric(as.POSIXct(str_replace(imp, "T", " "), tz = "UTC")))
  dataset[[dtf_nm]] <- sapply(raw, function(x) {
    if (is.na(x)) NA_character_ else if (length(str_split(x, "-")[[1]]) < 3) "D" else NA_character_
  }, USE.NAMES = FALSE)
  dataset[[tmf_nm]] <- if_else(!str_detect(raw, "T") & !is.na(raw),
                                if (time_imputation == "first") "H" else "H", NA_character_)
  dataset
}

derive_vars_dtm_to_dt <- function(dataset, source_vars) {
  for (nm in sapply(source_vars, deparse)) {
    dataset[[str_replace(nm, "DTM$", "DT")]] <- trunc(dataset[[nm]] / 86400)
  }
  dataset
}

derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref_nm <- deparse(substitute(reference_date))
  for (nm in sapply(source_vars, deparse)) {
    d <- dataset[[nm]] - dataset[[ref_nm]]
    dataset[[str_replace(nm, "DT$", "DY")]] <- case_when(
      is.na(d) ~ NA_real_, d >= 0 ~ d + 1, TRUE ~ d)
  }
  dataset
}

derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL,
                                order = NULL, filter_add = NULL, mode = NULL,
                                missing_values = NULL) {
  by_nm <- sapply(by_vars, deparse)
  fe    <- substitute(filter_add)
  # collapse deparse to single string — avoids length > 1 error with compound expressions
  if (!is.null(fe) && paste(deparse(fe), collapse = " ") != "NULL")
    dataset_add <- dataset_add |> filter(!!fe)
  if (!is.null(order)) {
    ord_nm   <- sapply(order, deparse)
    ord_desc <- grepl("^desc\\(", ord_nm)
    ord_cl   <- str_replace_all(ord_nm, "^desc\\(|\\)$", "")
    for (i in rev(seq_along(ord_cl)))
      dataset_add <- if (ord_desc[i]) dataset_add |> arrange(desc(.data[[ord_cl[i]]]))
                     else dataset_add |> arrange(.data[[ord_cl[i]]])
  }
  if (!is.null(mode)) {
    fn <- if (mode == "first") dplyr::slice_head else dplyr::slice_tail
    dataset_add <- dataset_add |> group_by(across(all_of(by_nm))) |> fn(n = 1) |> ungroup()
  }
  if (!is.null(new_vars)) {
    nv_nm <- sapply(new_vars, deparse)           # c(new_name = "old_name")
    dataset_add <- dataset_add |>
      select(any_of(c(by_nm, unname(nv_nm)))) |> # select by OLD column names
      rename(any_of(nv_nm))                       # rename(any_of(c(new="old"))) ✓
    new_col_names <- names(nv_nm)                 # the NEW column names for the join
  } else {
    new_col_names <- setdiff(names(dataset_add), by_nm)
  }
  dataset |> left_join(dataset_add |> select(any_of(c(by_nm, new_col_names))), by = by_nm)
}

derive_var_trtdurd <- function(dataset, new_var = TRTDURD,
                                start_date = TRTSDT, end_date = TRTEDT) {
  out <- deparse(substitute(new_var))
  s   <- deparse(substitute(start_date))
  e   <- deparse(substitute(end_date))
  dataset[[out]] <- dataset[[e]] - dataset[[s]] + 1
  dataset
}

derive_vars_duration <- function(dataset, new_var, start_date, end_date,
                                  out_unit = "days", add_one = TRUE) {
  out <- deparse(substitute(new_var))
  s   <- deparse(substitute(start_date))
  e   <- deparse(substitute(end_date))
  d   <- dataset[[e]] - dataset[[s]]
  dataset[[out]] <- case_when(
    is.na(d) ~ NA_real_,
    out_unit == "months" ~ round(as.numeric(d) / 30.4375, 1),
    add_one  ~ d + 1,
    TRUE     ~ d
  )
  dataset
}

derive_var_merged_exist_flag <- function(dataset, dataset_add, by_vars, new_var, condition) {
  by_nm <- sapply(by_vars, deparse)
  out   <- deparse(substitute(new_var))
  ce    <- substitute(condition)
  flagged <- dataset_add |> filter(!!ce) |> distinct(across(all_of(by_nm))) |>
    mutate(!!out := "Y")
  dataset |> left_join(flagged, by = by_nm)
}

# event_source: describe one possible source of an event for derive_vars_extreme_event
event_source <- function(dataset_name, filter = NULL, date, set_values_to) {
  list(
    dataset_name   = dataset_name,
    filter         = substitute(filter),
    date           = substitute(date),
    set_values_to  = set_values_to
  )
}

# derive_vars_extreme_event: find first/last event across multiple sources
derive_vars_extreme_event <- function(dataset, by_vars, events, order, mode,
                                       source_datasets, new_vars) {
  by_nm <- sapply(by_vars, deparse)

  event_rows <- lapply(events, function(ev) {
    src <- source_datasets[[ev$dataset_name]]
    # Apply filter (collapse deparse to avoid length > 1 in compound expressions)
    if (!is.null(ev$filter) && paste(deparse(ev$filter), collapse = " ") != "NULL")
      src <- src |> filter(!!ev$filter)
    if (nrow(src) == 0) return(NULL)
    # Evaluate date expression
    src <- src |> mutate(.EVENT_DATE = !!ev$date)
    # Apply set_values_to: evaluate each expression (handles function calls like convert_dtc_to_dt())
    for (i in seq_along(ev$set_values_to)) {
      out_nm   <- names(ev$set_values_to)[i]
      src_expr <- ev$set_values_to[[i]]
      src <- src |> mutate(!!out_nm := !!src_expr)
    }
    src |> select(any_of(c(by_nm, ".EVENT_DATE", names(ev$set_values_to))))
  }) |> bind_rows()

  if (nrow(event_rows) == 0) {
    # new_vars are unnamed exprs e.g. exprs(DTHCAUS, DTHDT) — use values, not names
    for (nm in sapply(new_vars, deparse)) dataset[[nm]] <- NA_character_
    return(dataset)
  }

  # Sort by date
  ord_nm  <- sapply(order, deparse)
  if (mode == "first")
    event_rows <- event_rows |> arrange(across(all_of(intersect(ord_nm, names(event_rows)))))
  else
    event_rows <- event_rows |> arrange(desc(across(all_of(intersect(ord_nm, names(event_rows))))))

  # Take first/last per subject
  fn <- if (mode == "first") dplyr::slice_head else dplyr::slice_tail
  best <- event_rows |>
    group_by(across(all_of(by_nm))) |>
    fn(n = 1) |>
    ungroup() |>
    select(-any_of(".EVENT_DATE"))

  dataset |> left_join(best, by = by_nm)
}

convert_dtc_to_dt <- function(dtc) {
  suppressWarnings(as.numeric(as.Date(.impute_dtc(dtc, "first"))))
}

# ── Source SDTM datasets ──────────────────────────────────────────────────────

dm <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "DM",
  USUBJID  = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101", "01-703-1211"),
  SUBJID   = c("1015", "1023", "1042", "1101", "1211"),
  SITEID   = c("701", "701", "702", "702", "703"),
  AGE      = c(63L, 64L, 71L, 56L, 45L),
  AGEU     = "YEARS",
  SEX      = c("M", "F", "M", "F", "M"),
  RACE     = c("WHITE", "WHITE", "ASIAN", "WHITE", "BLACK OR AFRICAN AMERICAN"),
  ETHNIC   = c("NOT HISPANIC OR LATINO", "NOT HISPANIC OR LATINO",
               "NOT HISPANIC OR LATINO", "HISPANIC OR LATINO", "NOT HISPANIC OR LATINO"),
  ARM      = c("Xanomeline High Dose", "Xanomeline Low Dose",
               "Placebo", "Xanomeline High Dose", "Placebo"),
  ACTARM   = c("XANOMELINE HIGH DOSE", "XANOMELINE LOW DOSE",
               "PLACEBO", "XANOMELINE HIGH DOSE", "PLACEBO"),
  RFXSTDTC = c("2014-01-02", "2012-08-05", "2013-07-09", "2013-04-14", "2014-03-22"),
  RFXENDTC = c("2014-06-15", "2012-12-22", "2013-12-15", "2013-10-30", "2014-09-01"),
  DTHDTC   = c(NA, NA, NA, NA, NA)
) |> convert_blanks_to_na()

ex <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "EX",
  USUBJID  = c("01-701-1015", "01-701-1015",
               "01-701-1023", "01-702-1042",
               "01-702-1101", "01-703-1211"),
  EXTRT    = c("XANOMELINE", "XANOMELINE",
               "XANOMELINE", "PLACEBO",
               "XANOMELINE", "PLACEBO"),
  EXDOSE   = c(81, 81, 54, 0, 81, 0),
  EXSEQ    = c(1L, 2L, 1L, 1L, 1L, 1L),
  EXSTDTC  = c("2014-01-02", "2014-02-02",
               "2012-08-05", "2013-07-09",
               "2013-04-14", "2014-03-22"),
  EXENDTC  = c("2014-01-31", "2014-06-15",
               "2012-12-22", "2013-12-15",
               "2013-10-30", "2014-09-01")
) |> convert_blanks_to_na()

ds <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "DS",
  USUBJID  = c("01-701-1015", "01-701-1015",
               "01-701-1023", "01-702-1042",
               "01-702-1101", "01-703-1211"),
  DSSEQ    = c(1L, 2L, 1L, 1L, 1L, 1L),
  DSCAT    = c("PROTOCOL MILESTONE", "DISPOSITION EVENT",
               "DISPOSITION EVENT", "DISPOSITION EVENT",
               "PROTOCOL MILESTONE", "DISPOSITION EVENT"),
  DSDECOD  = c("RANDOMIZED", "COMPLETED",
               "COMPLETED", "ADVERSE EVENT",
               "RANDOMIZED", "COMPLETED"),
  DSTERM   = c("RANDOMIZED", "COMPLETED STUDY",
               "COMPLETED STUDY", "DISCONTINUED DUE TO AE",
               "RANDOMIZED", "COMPLETED STUDY"),
  DSSTDTC  = c("2014-01-02", "2014-06-15",
               "2012-12-22", "2013-09-15",
               "2013-04-14", "2014-09-01")
) |> convert_blanks_to_na()

ae <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "AE",
  USUBJID  = c("01-701-1015", "01-702-1042", "01-702-1042"),
  AESEQ    = c(1L, 1L, 2L),
  AETERM   = c("HEADACHE", "DIZZINESS", "RASH"),
  AEDECOD  = c("Headache", "Dizziness", "Rash"),
  AESTDTC  = c("2014-03-15", "2013-08-10", "2013-09-01"),
  AEOUT    = c("RECOVERED/RESOLVED", "RECOVERED/RESOLVED", "RECOVERING/RESOLVING"),
  AEDTHDTC = c(NA, NA, NA)
) |> convert_blanks_to_na()

lb <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "LB",
  USUBJID  = c("01-701-1015", "01-701-1015", "01-701-1023", "01-702-1042"),
  LBTESTCD = c("ALT", "ALT", "ALT", "ALT"),
  LBDTC    = c("2014-01-02", "2014-04-01", "2012-08-05", "2013-07-09"),
  LBORRES  = c("28", "35", "22", "41")
) |> convert_blanks_to_na()

# ── Real pharmaverse datasets (for validation exercises) ──────────────────────
library(pharmaversesdtm)
library(pharmaverseadam)
dm_real   <- pharmaversesdtm::dm |> convert_blanks_to_na()
ex_real   <- pharmaversesdtm::ex |> convert_blanks_to_na()
adsl_real <- pharmaverseadam::adsl
```

## Welcome

This is Lesson 15 — **`{admiral}` Part 2: Building ADSL End-to-End**.

An ADSL has one row per subject and holds everything true about that subject: demographics, treatment dates, durations, population flags, and outcomes. It's built first — every other ADaM references it.

This lesson walks through a complete ADSL build using admiral, step by step:

1. Base dataset from DM
2. Treatment start/end datetimes → dates → duration
3. Disposition dates (randomization, end of study)
4. Cause of death and last known alive date (multi-source)
5. Population flags (SAFFL, ITTFL, RANDFL)
6. Age groups (AGEGR1, AGEGR1N)

---

## 1 — What's in an ADSL?

```{r quiz-adsl-overview, echo=FALSE}
quiz(
  question(
    "ADSL has one row per:",
    answer("Subject per visit"),
    answer("Subject", correct = TRUE),
    answer("Subject per parameter"),
    random_answer_order = TRUE
  ),
  question(
    "Why is ADSL typically the first ADaM you build?",
    answer("It's the smallest and easiest"),
    answer("Every other ADaM joins to ADSL for subject-level metadata like treatment dates and population flags", correct = TRUE),
    answer("Regulators require it first"),
    random_answer_order = TRUE
  ),
  question(
    "TRTSDTM vs TRTSDT — the difference is:",
    answer("TRTSDTM is a datetime (with time component); TRTSDT is a date-only version", correct = TRUE),
    answer("They are identical — both store dates"),
    answer("TRTSDT comes from DS; TRTSDTM comes from EX"),
    random_answer_order = TRUE
  )
)
```

---

## 2 — Setup and base dataset

```{r adsl-base-setup}
# Step 0: All SDTM already loaded and blanks converted (see setup)
# Step 1: Create base ADSL from DM
adsl <- dm |>
  mutate(
    TRT01P = ARM,
    TRT01A = ACTARM
  )

glimpse(adsl)
```

---

## 3 — Pre-process EX and DS

Before merging, derive datetime variables on the source datasets.

```{r preprocess-ex}
# EX: add start and end datetimes
ex_ext <- ex |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first",
    time_imputation = "first"
  ) |>
  derive_vars_dtm(
    dtc             = EXENDTC,
    new_vars_prefix = "EXEN",
    date_imputation = "last",
    time_imputation = "last"
  )

ex_ext |>
  select(USUBJID, EXSTDTC, EXSTDTM, EXENDTC, EXENDTM) |>
  mutate(
    EXST_human = as.POSIXct(EXSTDTM, origin = "1970-01-01", tz = "UTC"),
    EXEN_human = as.POSIXct(EXENDTM, origin = "1970-01-01", tz = "UTC")
  )
```

```{r preprocess-ds}
# DS: add date variable from DSSTDTC
ds_ext <- ds |>
  derive_vars_dt(
    dtc             = DSSTDTC,
    new_vars_prefix = "DSST",
    date_imputation = "first"
  )

ds_ext |>
  select(USUBJID, DSCAT, DSDECOD, DSSTDTC, DSSTDT) |>
  mutate(DSSTDT_human = as.Date(DSSTDT, origin = "1970-01-01"))
```

---

## 4 — Treatment start and end datetimes

The standard EX filter: `(EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))` — counts active doses AND zero-dose placebo records.

```{r derive-trt-dates}
# Derive TRTSDTM (first qualifying dose) and TRTEDTM (last qualifying dose end)
adsl <- adsl |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM, TRTSTMF = EXSTTMF),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXENDTM),
    new_vars    = exprs(TRTEDTM = EXENDTM, TRTETMF = EXENTMF),
    order       = exprs(EXENDTM, EXSEQ),
    mode        = "last",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl |>
  select(USUBJID, ACTARM, TRTSDTM, TRTEDTM) |>
  mutate(
    TRTS_human = as.POSIXct(TRTSDTM, origin = "1970-01-01", tz = "UTC"),
    TRTE_human = as.POSIXct(TRTEDTM, origin = "1970-01-01", tz = "UTC")
  )
```

### Exercise: Convert datetimes to dates and compute treatment duration

```{r trt-dates-to-dt, exercise=TRUE, exercise.lines=20}
# Step 1: convert TRTSDTM/TRTEDTM → TRTSDT/TRTEDT
adsl_dates <- adsl |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM))

# Step 2: derive TRTDURD = TRTEDT - TRTSDT + 1
adsl_dur <- adsl_dates |>
  derive_var_trtdurd()

adsl_dur |>
  select(USUBJID, ACTARM, TRTSDT, TRTEDT, TRTDURD) |>
  mutate(
    TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_human = as.Date(TRTEDT, origin = "1970-01-01")
  )
```

### Exercise: Validate TRTSDT against real pharmaverseadam data

Use `dm_real` and `ex_real` (the full CDISCPILOT01 SDTM datasets from pharmaversesdtm) to re-derive TRTSDT for the first 10 subjects, then compare to the published values in `adsl_real`.

> **Why this matters:** In real submissions, your ADSL is compared to an independently derived QC dataset. Discrepancies usually mean a wrong filter, wrong mode, or a missing `convert_blanks_to_na()`.

```{r trt-dates-real-validate, exercise=TRUE, exercise.lines=52}
# Step 1: subset to 10 subjects that appear in adsl_real
subs10 <- adsl_real$USUBJID[1:10]

dm10 <- dm_real |> filter(USUBJID %in% subs10)

ex10_ext <- ex_real |>
  filter(USUBJID %in% subs10) |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first",
    time_imputation = "first"
  )

# Step 2: derive TRTSDTM → TRTSDT using the standard EX filter
derived10 <- dm10 |>
  derive_vars_merged(
    dataset_add = ex10_ext,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM)) |>
  mutate(TRTSDT_derived = as.Date(TRTSDT, origin = "1970-01-01"))

# Step 3: pull reference from pharmaverseadam and compare
ref <- adsl_real |>
  filter(USUBJID %in% subs10) |>
  select(USUBJID, TRTSDT_ref = TRTSDT)

comparison <- left_join(
  derived10 |> select(USUBJID, ARM, TRTSDT_derived),
  ref, by = "USUBJID"
) |>
  mutate(
    d_derived = as.integer(TRTSDT_derived),
    d_ref     = as.integer(as.Date(TRTSDT_ref, origin = "1970-01-01")),
    MATCH     = (d_derived == d_ref) | (is.na(d_derived) & is.na(d_ref))
  ) |>
  select(USUBJID, ARM, TRTSDT_derived, TRTSDT_ref, MATCH)

cat("Matching subjects:", sum(comparison$MATCH, na.rm = TRUE), "/ 10\n\n")
print(comparison)
```

```{r trt-dates-real-validate-hint-1}
# The key EX filter: (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
# mode = "first"  picks the earliest qualifying dose start (= first treatment date)
# Any mismatch here suggests a filter, mode, or blank-handling discrepancy
```

```{r trt-dates-real-validate-solution}
subs10 <- adsl_real$USUBJID[1:10]
dm10 <- dm_real |> filter(USUBJID %in% subs10)

ex10_ext <- ex_real |>
  filter(USUBJID %in% subs10) |>
  derive_vars_dtm(
    dtc = EXSTDTC, new_vars_prefix = "EXST",
    date_imputation = "first", time_imputation = "first"
  )

derived10 <- dm10 |>
  derive_vars_merged(
    dataset_add = ex10_ext,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM)) |>
  mutate(TRTSDT_derived = as.Date(TRTSDT, origin = "1970-01-01"))

ref <- adsl_real |>
  filter(USUBJID %in% subs10) |>
  select(USUBJID, TRTSDT_ref = TRTSDT)

left_join(derived10 |> select(USUBJID, ARM, TRTSDT_derived), ref, by = "USUBJID") |>
  mutate(
    d_derived = as.integer(TRTSDT_derived),
    d_ref     = as.integer(as.Date(TRTSDT_ref, origin = "1970-01-01")),
    MATCH     = (d_derived == d_ref) | (is.na(d_derived) & is.na(d_ref))
  ) |>
  select(USUBJID, ARM, TRTSDT_derived, TRTSDT_ref, MATCH) |>
  print()
```

---

## 5 — Disposition dates

Both RANDDT and EOSDT come from DS with different DSDECOD filters:

```{r derive-disp-dates}
adsl <- adsl |>
  # Randomization date
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "PROTOCOL MILESTONE" & DSDECOD == "RANDOMIZED",
    new_vars    = exprs(RANDDT = DSSTDT),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  # End of study date
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(EOSDT = DSSTDT),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl |>
  select(USUBJID, RANDDT, EOSDT) |>
  mutate(
    RANDDT_human = as.Date(RANDDT, origin = "1970-01-01"),
    EOSDT_human  = as.Date(EOSDT,  origin = "1970-01-01")
  )
```

### Exercise: Derive the discontinuation reason (DCSREAS)

Merge the DS term for "DISPOSITION EVENT" records onto ADSL as `DCSREAS`.

```{r derive-dcsreas, exercise=TRUE, exercise.lines=15}
adsl_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == ___ & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = ___),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_dcsreas |>
  select(USUBJID, ACTARM, DCSREAS)
```

```{r derive-dcsreas-hint-1}
# filter_add = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE"
# new_vars = exprs(DCSREAS = DSTERM)
```

```{r derive-dcsreas-solution}
adsl_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_dcsreas |>
  select(USUBJID, ACTARM, DCSREAS)
```

---

## 6 — Cause of death: multi-source derivation

DTHCAUS can come from DS (death recorded as a disposition event) OR from AE (fatal adverse event). `derive_vars_extreme_event()` handles multi-source lookups elegantly.

```{r death-source-demo}
# Source 1: DS records for death
src_ds_death <- event_source(
  dataset_name = "ds",
  filter       = DSDECOD == "DEATH",
  date         = DSSTDT,
  set_values_to = exprs(
    DTHCAUS = DSTERM,
    DTHDT   = DSSTDT
  )
)

# Source 2: AE records with fatal outcome
src_ae_death <- event_source(
  dataset_name = "ae",
  filter       = AEOUT == "FATAL",
  date         = convert_dtc_to_dt(AEDTHDTC),
  set_values_to = exprs(
    DTHCAUS = AEDECOD,
    DTHDT   = convert_dtc_to_dt(AEDTHDTC)
  )
)

# Derive DTHCAUS and DTHDT from whichever source has the event
adsl <- adsl |>
  derive_vars_extreme_event(
    by_vars         = exprs(STUDYID, USUBJID),
    events          = list(src_ds_death, src_ae_death),
    order           = exprs(DTHDT),
    mode            = "first",
    source_datasets = list(ds = ds_ext, ae = ae),
    new_vars        = exprs(DTHCAUS, DTHDT)
  )

adsl |>
  select(USUBJID, DTHCAUS, DTHDT) |>
  mutate(DTHDT_human = as.Date(DTHDT, origin = "1970-01-01"))
```

### Exercise: Verify no subjects have a death event in our sample data

Our sample `ds` has no `DSDECOD == "DEATH"` and `ae` has no `AEOUT == "FATAL"`. Confirm DTHCAUS is NA for all subjects.

```{r death-check, exercise=TRUE, exercise.lines=8}
# How many subjects have a non-NA DTHCAUS?
n_deaths <- adsl |>
  filter(!is.na(___)) |>
  nrow()

cat("Subjects with DTHCAUS:", n_deaths, "\n")
cat("Expected: 0 — no deaths in sample data\n")
```

```{r death-check-hint-1}
# filter(!is.na(DTHCAUS))
```

```{r death-check-solution}
n_deaths <- adsl |>
  filter(!is.na(DTHCAUS)) |>
  nrow()

cat("Subjects with DTHCAUS:", n_deaths, "\n")
cat("Expected: 0 — no deaths in sample data\n")
```

---

## 7 — Last known alive date (LSTALVDT)

LSTALVDT is the latest date we have *any* evidence the subject was alive. It comes from multiple sources — we take the maximum (latest) across all of them.

```{r lstalvdt-demo}
# Source 1: latest AE start date (subject was alive when the AE started)
src_ae_alive <- event_source(
  dataset_name = "ae",
  filter       = !is.na(AESTDTC),
  date         = convert_dtc_to_dt(AESTDTC),
  set_values_to = exprs(LSTALVDT = convert_dtc_to_dt(AESTDTC))
)

# Source 2: latest lab collection date
src_lb_alive <- event_source(
  dataset_name = "lb",
  filter       = !is.na(LBDTC),
  date         = convert_dtc_to_dt(LBDTC),
  set_values_to = exprs(LSTALVDT = convert_dtc_to_dt(LBDTC))
)

adsl <- adsl |>
  derive_vars_extreme_event(
    by_vars         = exprs(STUDYID, USUBJID),
    events          = list(src_ae_alive, src_lb_alive),
    order           = exprs(LSTALVDT),
    mode            = "last",    # latest date across all sources
    source_datasets = list(ae = ae, lb = lb),
    new_vars        = exprs(LSTALVDT)
  )

adsl |>
  select(USUBJID, LSTALVDT) |>
  mutate(LSTALVDT_human = as.Date(LSTALVDT, origin = "1970-01-01"))
```

```{r quiz-lstalvdt, echo=FALSE}
quiz(
  question(
    "For LSTALVDT you use mode = 'last' because:",
    answer("You want the most recent date at which the subject was confirmed alive", correct = TRUE),
    answer("You want the earliest date"),
    answer("Both first and last produce the same result"),
    random_answer_order = TRUE
  )
)
```

---

## 8 — Treatment duration

```{r trt-duration}
# TRTSDT and TRTEDT are now available — derive TRTDURD
adsl <- adsl |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM)) |>
  derive_var_trtdurd()

adsl |>
  select(USUBJID, TRTSDT, TRTEDT, TRTDURD) |>
  mutate(
    TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_human = as.Date(TRTEDT, origin = "1970-01-01")
  )
```

### Exercise: Derive TRTDURM (treatment duration in months)

Use `derive_vars_duration()` with `out_unit = "months"`.

```{r trt-duration-months, exercise=TRUE, exercise.lines=12}
adsl_with_months <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURM,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = ___,
    add_one    = FALSE
  )

adsl_with_months |>
  select(USUBJID, TRTDURD, TRTDURM)
```

```{r trt-duration-months-hint-1}
# out_unit = "months"
```

```{r trt-duration-months-solution}
adsl_with_months <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURM,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = "months",
    add_one    = FALSE
  )

adsl_with_months |>
  select(USUBJID, TRTDURD, TRTDURM)
```

---

## 9 — Population flags

### SAFFL: Safety population

A subject is in the safety population if they received at least one dose (positive dose OR placebo).

```{r saffl-demo}
adsl <- adsl |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
  )

adsl |> select(USUBJID, ACTARM, SAFFL)
```

### ITTFL and RANDFL: Intent-to-treat and randomized populations

These are derived from RANDDT — a subject is in ITT if they were randomized.

```{r ittfl-demo}
adsl <- adsl |>
  mutate(
    ITTFL  = if_else(!is.na(RANDDT), "Y", NA_character_),
    RANDFL = if_else(!is.na(RANDDT), "Y", NA_character_)
  )

adsl |> select(USUBJID, SAFFL, ITTFL, RANDFL)
```

```{r quiz-flag-values, echo=FALSE}
quiz(
  question(
    "Why do population flags use 'Y' or NA, not 'Y' or 'N'?",
    answer("CDISC ADaM IG specifies 'Y'/NA — a blank means 'not in population'; 'N' is redundant and not standard", correct = TRUE),
    answer("'N' is a reserved value in R"),
    answer("It saves storage space"),
    random_answer_order = TRUE
  )
)
```

### Exercise: Derive COMPLFL (completion flag)

A subject is a completer if their DCSREAS is `"COMPLETED STUDY"`. Derive `COMPLFL` using a direct `mutate()`.

```{r complfl-exercise, exercise=TRUE, exercise.lines=10}
# First, get DCSREAS onto adsl (derive it from ds)
adsl_with_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  mutate(COMPLFL = if_else(DCSREAS == ___, "Y", NA_character_))

adsl_with_dcsreas |>
  select(USUBJID, ACTARM, DCSREAS, COMPLFL)
```

```{r complfl-exercise-hint-1}
# if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_)
```

```{r complfl-exercise-solution}
adsl_with_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_))

adsl_with_dcsreas |>
  select(USUBJID, ACTARM, DCSREAS, COMPLFL)
```

---

## 10 — Age groups

AGEGR1 (character) and AGEGR1N (numeric companion) are the standard age categorization:

```{r age-groups-demo}
adsl <- adsl |>
  mutate(
    AGEGR1  = case_when(
      AGE < 18             ~ "<18",
      between(AGE, 18, 64) ~ "18-64",
      AGE >= 65            ~ ">=65"
    ),
    AGEGR1N = case_when(
      AGE < 18             ~ 1,
      between(AGE, 18, 64) ~ 2,
      AGE >= 65            ~ 3
    )
  )

adsl |>
  select(USUBJID, AGE, AGEGR1, AGEGR1N)
```

### Exercise: Add a finer age categorization (AGEGR2)

Create `AGEGR2` with categories: `<40`, `40-54`, `55-64`, `>=65`.

```{r agegr2-exercise, exercise=TRUE, exercise.lines=14}
adsl_agegr2 <- adsl |>
  mutate(
    AGEGR2 = case_when(
      AGE < 40             ~ "<40",
      between(AGE, 40, 54) ~ "40-54",
      between(AGE, 55, 64) ~ "55-64",
      AGE >= 65            ~ ">=65"
    )
  )

adsl_agegr2 |>
  select(USUBJID, AGE, AGEGR1, AGEGR2)
```

---

## 11 — Putting it all together: the complete ADSL build

Here is the full ADSL derivation script, condensed. Run it to see the complete dataset.

```{r adsl-full-build, exercise=TRUE, exercise.lines=80}
# ── Pre-process source datasets ───────────────────────────────────────────────
ex_ext_full <- ex |>
  derive_vars_dtm(dtc = EXSTDTC, new_vars_prefix = "EXST",
                  date_imputation = "first", time_imputation = "first") |>
  derive_vars_dtm(dtc = EXENDTC, new_vars_prefix = "EXEN",
                  date_imputation = "last",  time_imputation = "last")

ds_ext_full <- ds |>
  derive_vars_dt(dtc = DSSTDTC, new_vars_prefix = "DSST", date_imputation = "first")

# ── Build ADSL ────────────────────────────────────────────────────────────────
adsl_full <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM) |>

  # 1. Treatment datetimes
  derive_vars_merged(
    dataset_add = ex_ext_full,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM), order = exprs(EXSTDTM, EXSEQ),
    mode = "first", by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_merged(
    dataset_add = ex_ext_full,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXENDTM),
    new_vars    = exprs(TRTEDTM = EXENDTM), order = exprs(EXENDTM, EXSEQ),
    mode = "last", by_vars = exprs(STUDYID, USUBJID)) |>

  # 2. Convert to dates + duration
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM)) |>
  derive_var_trtdurd() |>

  # 3. Disposition dates
  derive_vars_merged(
    dataset_add = ds_ext_full,
    filter_add  = DSCAT == "PROTOCOL MILESTONE" & DSDECOD == "RANDOMIZED",
    new_vars    = exprs(RANDDT = DSSTDT), by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_merged(
    dataset_add = ds_ext_full,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(EOSDT = DSSTDT, DCSREAS = DSTERM),
    by_vars = exprs(STUDYID, USUBJID)) |>

  # 4. Population flags
  derive_var_merged_exist_flag(
    dataset_add = ex, by_vars = exprs(STUDYID, USUBJID), new_var = SAFFL,
    condition   = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))) |>
  mutate(ITTFL  = if_else(!is.na(RANDDT), "Y", NA_character_),
         RANDFL = if_else(!is.na(RANDDT), "Y", NA_character_),
         COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_)) |>

  # 5. Age groups
  mutate(
    AGEGR1  = case_when(AGE < 18 ~ "<18", between(AGE, 18, 64) ~ "18-64", AGE >= 65 ~ ">=65"),
    AGEGR1N = case_when(AGE < 18 ~ 1,     between(AGE, 18, 64) ~ 2,       AGE >= 65 ~ 3)
  )

# ── Summary ───────────────────────────────────────────────────────────────────
cat("ADSL rows (one per subject):", nrow(adsl_full), "\n")
cat("ADSL columns:", ncol(adsl_full), "\n\n")

adsl_full |>
  select(USUBJID, ACTARM, TRTSDT, TRTEDT, TRTDURD,
         SAFFL, ITTFL, RANDFL, COMPLFL, AGEGR1, AGEGR1N) |>
  mutate(
    TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_human = as.Date(TRTEDT, origin = "1970-01-01")
  ) |>
  select(-TRTSDT, -TRTEDT) |>
  print()
```

---

## 11.5 — P21 Derivation Diagnostics

Clinical programmers submitting to FDA routinely receive Pinnacle 21 (P21) validation reports that flag derivation errors in ADSL. This section presents three realistic bugs — each accompanied by a **P21-style derivation method specification** exactly as it would appear in a define.xml submission file. Your task: read the spec, study the broken code, identify what is wrong, and fix it.

> **Format for each exercise:** (1) the derivation specification text, (2) the broken programmer code, (3) the P21 error message received. Fill in the blank to correct the code.

---

### Diagnostic 1 — SAFFL: Safety Analysis Set Flag

**Derivation method (define.xml / Pinnacle 21 field):**

> *SAFFL = 'Y' if the subject has at least one exposure record in EX where EXDOSE > 0, OR where EXDOSE = 0 AND EXTRT contains the string 'PLACEBO'. Subjects with no qualifying EX record should have SAFFL missing (not 'N'). ADaM IG §4.5. Source variables: EX.EXDOSE, EX.EXTRT.*

The programmer submitted:

```r
adsl_draft <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM) |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = EXDOSE > 0          # ← bug here
  )
```

**P21 report:** *"SAFFL is missing for 2 Placebo subjects. All subjects with an EX record should have SAFFL populated per derivation spec."*

```{r p21-bug-saffl, exercise=TRUE, exercise.lines=18}
# Fix the SAFFL derivation so Placebo subjects (EXDOSE = 0, EXTRT contains "PLACEBO")
# are correctly included in the Safety Analysis Set.

adsl_fixed_saffl <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM) |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = ___
  )

adsl_fixed_saffl |> select(USUBJID, ACTARM, SAFFL)
# Expected: all 5 subjects have SAFFL = "Y"
```

```{r p21-bug-saffl-hint-1}
# The condition must cover BOTH active-dose records AND zero-dose placebo records:
# condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
```

```{r p21-bug-saffl-solution}
adsl_fixed_saffl <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM) |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
  )

adsl_fixed_saffl |> select(USUBJID, ACTARM, SAFFL)
# All 5 subjects should now have SAFFL = "Y"
```

---

### Diagnostic 2 — TRTDURD: Treatment Duration in Days

**Derivation method (define.xml / Pinnacle 21 field):**

> *TRTDURD = number of days of study treatment exposure. Calculated as TRTEDT − TRTSDT + 1 (inclusive of both the first and last treatment dates). Unit: days. Source variables: derived TRTSDT, TRTEDT.*

The programmer submitted:

```r
adsl_dur_draft <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURD,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = "days",
    add_one    = FALSE         # ← bug here
  )
```

**P21 report:** *"TRTDURD is 1 day less than expected for all subjects. Example: USUBJID 01-701-1015 — TRTSDT = 2014-01-02, TRTEDT = 2014-06-15. Expected TRTDURD = 165; observed = 164."*

```{r p21-bug-trtdurd, exercise=TRUE, exercise.lines=22}
# adsl already has TRTSDT and TRTEDT from the demo steps above.
# Fix the duration derivation to match the spec (inclusive: TRTEDT - TRTSDT + 1).

adsl_fixed_dur <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURD_fixed,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = "days",
    add_one    = ___           # fix this argument
  )

# Compare broken vs fixed vs expected
adsl_fixed_dur |>
  select(USUBJID, TRTSDT, TRTEDT, TRTDURD, TRTDURD_fixed) |>
  mutate(
    TRTSDT_d = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_d = as.Date(TRTEDT, origin = "1970-01-01"),
    expected = as.integer(TRTEDT_d - TRTSDT_d) + 1L
  )
```

```{r p21-bug-trtdurd-hint-1}
# add_one = TRUE  ← the spec says "inclusive of both endpoints" = TRTEDT - TRTSDT + 1
# add_one = FALSE gives TRTEDT - TRTSDT (excludes the start day) — 1 day short
```

```{r p21-bug-trtdurd-solution}
adsl_fixed_dur <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURD_fixed,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = "days",
    add_one    = TRUE
  )

adsl_fixed_dur |>
  select(USUBJID, TRTSDT, TRTEDT, TRTDURD, TRTDURD_fixed) |>
  mutate(
    TRTSDT_d = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_d = as.Date(TRTEDT, origin = "1970-01-01"),
    expected = as.integer(TRTEDT_d - TRTSDT_d) + 1L
  )
```

---

### Diagnostic 3 — COMPLFL: Study Completion Flag

**Derivation method (define.xml / Pinnacle 21 field):**

> *COMPLFL = 'Y' if the subject completed the study per protocol (DCSREAS = 'COMPLETED STUDY'). Subjects who did not complete should have COMPLFL missing. Per ADaM IG §4.5, flag variables take values 'Y' or missing — the value 'N' is not permitted. Source variables: derived DCSREAS.*

The programmer submitted:

```r
adsl_comp_draft <- adsl_with_dcsreas |>
  mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", "N"))   # ← bug here
```

**P21 report:** *"AD0018 — COMPLFL contains the value 'N'. Flag variables (suffix -FL) must only contain 'Y' or missing per ADaM IG. Permissible values violated."*

```{r p21-bug-complfl, exercise=TRUE, exercise.lines=22}
# Build adsl_with_dcsreas (self-contained — ds_ext is available from the demo chunks above)
adsl_with_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  )

# Fix: replace "N" with the correct ADaM IG value for subjects not in the flag group
adsl_fixed_complfl <- adsl_with_dcsreas |>
  mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", ___))

adsl_fixed_complfl |>
  select(USUBJID, ACTARM, DCSREAS, COMPLFL)
# Expected: completers → "Y"; discontinuers → NA (not "N")
```

```{r p21-bug-complfl-hint-1}
# ADaM IG: flag variables are "Y" or NA — the value "N" is never valid.
# Use NA_character_ as the false branch:
# mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_))
```

```{r p21-bug-complfl-solution}
adsl_with_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_fixed_complfl <- adsl_with_dcsreas |>
  mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_))

adsl_fixed_complfl |>
  select(USUBJID, ACTARM, DCSREAS, COMPLFL)
```

---

## 12 — Validation strategy

```{r quiz-validation, echo=FALSE}
quiz(
  question(
    "The standard validation approach for submission-quality ADSL is:",
    answer("One programmer writes it and a reviewer checks the code"),
    answer("Two programmers independently build the ADSL; outputs are compared with {diffdf}; discrepancies are resolved", correct = TRUE),
    answer("Run check_variables() from metatools — that's sufficient"),
    random_answer_order = TRUE
  ),
  question(
    "The add_one = FALSE convention in derive_vars_duration() applies to:",
    answer("TRTDURD — treatment duration in days"),
    answer("Elapsed-time variables like LDDTHELD, where you count calendar gap without an endpoint adjustment", correct = TRUE),
    answer("All duration variables"),
    random_answer_order = TRUE
  ),
  question(
    "The filter (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, 'PLACEBO'))) ensures:",
    answer("Only active doses count"),
    answer("Active doses AND zero-dose placebo records both count as exposure", correct = TRUE),
    answer("Missing doses are excluded"),
    random_answer_order = TRUE
  )
)
```

---

## 13 — Key takeaways

```{r quiz-final-15, echo=FALSE}
quiz(
  question(
    "A programmer derives SAFFL using `condition = EXDOSE > 0`. P21 reports SAFFL is missing for all Placebo subjects. What is the correct fix?",
    answer("Change to `condition = EXDOSE >= 0`"),
    answer("Add `mutate(SAFFL = if_else(ARM == 'Placebo', 'Y', SAFFL))` after the flag derivation"),
    answer("Use `condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, 'PLACEBO')))`", correct = TRUE),
    answer("Filter EX to only rows where EXDOSE > 0 before calling derive_var_merged_exist_flag()"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The derivation spec says TRTDURD = TRTEDT − TRTSDT + 1 (inclusive). The programmer uses `add_one = FALSE`. A subject treated January 1–31 gets TRTDURD = 30. The spec expects 31. What does `add_one = FALSE` calculate?",
    answer("The elapsed gap between two dates, not counting the start day — TRTEDT − TRTSDT with no +1 adjustment", correct = TRUE),
    answer("Duration rounded down to avoid double-counting"),
    answer("Duration in calendar months instead of days"),
    answer("Exactly the same result as add_one = TRUE"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "P21 rule AD0018 flags COMPLFL = 'N'. Why does 'N' violate ADaM IG and what replaces it?",
    answer("'N' is not a valid R logical — use FALSE instead"),
    answer("ADaM IG §4.5 specifies flag variables take values 'Y' or missing (NA_character_) — 'N' is prohibited", correct = TRUE),
    answer("'N' causes a metadata mismatch with the define.xml codelist for FL variables"),
    answer("'N' is valid in SDTM but not in ADaM — use 0 instead"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "LSTALVDT is the *last* known alive date. A programmer accidentally writes `mode = 'first'` in derive_vars_extreme_event(). What is the consequence?",
    answer("The function throws an error — 'first' is not a valid mode for alive-date sources"),
    answer("LSTALVDT is set to the *earliest* date the subject was observed alive, underestimating their known survival time", correct = TRUE),
    answer("LSTALVDT is replaced by the death date for subjects who died"),
    answer("No consequence — for alive-date derivations, 'first' and 'last' produce identical results"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The AGEGR1N derivation uses `between(AGE, 18, 65) ~ 2` for the '18-64' group. A 65-year-old ends up in AGEGR1N = 2 instead of AGEGR1N = 3 ('>=65'). What is the bug?",
    answer("`between(AGE, 18, 65)` includes 65 (it is inclusive on both ends) — the fix is `between(AGE, 18, 64)`", correct = TRUE),
    answer("case_when() evaluates all conditions and picks the last match, overwriting group 3 with group 2"),
    answer("Age 65 belongs in group 2 per CDISC ADaM IG guidance"),
    answer("The bug is only in the character AGEGR1 variable, not AGEGR1N"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "A programmer forgets convert_blanks_to_na() on DM before building ADSL. Some subjects have ARM = '' (empty string). What goes wrong downstream?",
    answer("admiral immediately errors when it encounters empty strings in character columns"),
    answer("Empty strings pass `!is.na(ARM)` checks, so those subjects appear to have a valid ARM — but str_detect() and equality comparisons on ARM then return NA or FALSE unexpectedly, producing wrong flags", correct = TRUE),
    answer("Nothing — admiral automatically coerces empty strings to NA internally"),
    answer("Empty strings are identical to NA in all R comparisons"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The spec says EOSDT is the date of the subject's disposition event, excluding screen failures. A programmer writes `filter_add = DSDECOD == 'COMPLETED'`. Which subjects will have a missing EOSDT?",
    answer("Only subjects who completed the study — they are included, not excluded"),
    answer("All subjects who discontinued (adverse event, withdrew consent, lost to follow-up) because their DSDECOD is not 'COMPLETED'", correct = TRUE),
    answer("Subjects with partial disposition dates only"),
    answer("All subjects — the filter should be on DSCAT, not DSDECOD"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "EXSTDTC = '2014-07' (year-month partial date, no day). The programmer calls derive_vars_dtm(..., date_imputation = 'first', time_imputation = 'first'). What datetime will EXSTDTM represent?",
    answer("2014-07-15T12:00:00 — admiral imputes the middle of the period"),
    answer("NA — admiral cannot impute partial dates with a missing day component"),
    answer("2014-07-01T00:00:00 — 'first' imputes the first day of the month and midnight time", correct = TRUE),
    answer("2014-07-31T23:59:59 — admiral always imputes to the end of the period by default"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "Why is derive_vars_extreme_event() preferred over two sequential derive_vars_merged() calls when deriving DTHCAUS from both DS and AE?",
    answer("derive_vars_extreme_event() executes in a single database pass and is more memory efficient"),
    answer("Sequential derive_vars_merged() calls overwrite the first source's value for subjects found in both sources — derive_vars_extreme_event() selects the true first/last event across all sources simultaneously", correct = TRUE),
    answer("derive_vars_merged() cannot join on date columns — only character keys"),
    answer("There is no practical difference; both approaches produce identical results for all subjects"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The standard admiral pipeline for treatment start date is: (1) derive_vars_dtm() on EX, (2) derive_vars_merged() onto ADSL, (3) derive_vars_dtm_to_dt(). Why does step 3 come last — why not derive the date directly in step 1?",
    answer("Admiral stores datetimes internally as numerics; deriving the datetime first preserves precision and time-flag metadata (TRTSTMF); dtm_to_dt() then creates the date-only variable cleanly from the stored datetime", correct = TRUE),
    answer("derive_vars_merged() can only operate on datetime (POSIXct) columns, not Date columns"),
    answer("Step 3 is only required for submission datasets — it is optional for internal QC"),
    answer("TRTSDT must be sourced from DM.RFXSTDTC, not EX — step 3 reads it from the correct domain"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

---

## What's next

Lesson 16 — **`{admiral}` Part 3** — covers BDS (Basic Data Structure) ADaMs: ADLB, ADVS, ADEG. These are long-format datasets with one row per subject per parameter per visit. The derivation patterns reuse `derive_vars_merged()`, `derive_var_extreme_flag()`, and the date helpers — but now applied to analysis-value reshaping and baseline flagging across many parameters.

After Lesson 16: OCCDS (ADAE, ADCM) in Lesson 17, time-to-event (ADTTE) in Lesson 18, and advanced admiral patterns in Lesson 19.
