---
title: "Lesson 15 — {admiral} Part 2: Building ADSL End-to-End"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.1
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build a production-grade ADSL step by step: treatment dates, durations,
  disposition dates, death cause, population flags, and age groups.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 15_admiral_part2_adsl.Rmd
# PURPOSE       : admiral Part 2 — ADSL derivation: treatment variables, disposition
#                 endpoints, baseline flags (SAFFL, ITTFL), study day
#                 calculations, and population-level subject-level
#                 analysis dataset construction.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# exprs: capture unevaluated arguments as a list of expressions (mirrors rlang::exprs)
exprs <- function(...) as.list(substitute(list(...)))[-1]

# ── Mock admiral functions (same as Lesson 14 + new ones) ────────────────────

convert_blanks_to_na <- function(dataset) {
  dataset |>
    mutate(across(where(is.character), ~ if_else(.x == "", NA_character_, .x)))
}

.impute_dtc <- function(dtc, date_imputation = "first") {
  sapply(dtc, function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    segs <- str_split(str_split(x, "T")[[1]][1], "-")[[1]]
    if (length(segs) == 1) segs <- c(segs, if (date_imputation == "last") "12" else "01")
    if (length(segs) == 2) {
      d <- if (date_imputation == "last")
        as.character(days_in_month(as.Date(paste0(segs[1], "-", segs[2], "-01")))) else "01"
      segs <- c(segs, d)
    }
    paste(segs, collapse = "-")
  }, USE.NAMES = FALSE)
}

derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation = "none", flag_imputation = "auto") {
  dtc_nm <- deparse(substitute(dtc))
  dt_nm  <- paste0(new_vars_prefix, "DT")
  dtf_nm <- paste0(new_vars_prefix, "DTF")
  raw    <- dataset[[dtc_nm]]
  imp    <- .impute_dtc(raw, date_imputation)
  dataset[[dt_nm]]  <- suppressWarnings(as.numeric(as.Date(imp)))
  dataset[[dtf_nm]] <- sapply(raw, function(x) {
    if (is.na(x)) return(NA_character_)
    n <- length(str_split(x, "-")[[1]])
    if (n == 1) "Y" else if (n == 2) "D" else NA_character_
  }, USE.NAMES = FALSE)
  dataset
}

derive_vars_dtm <- function(dataset, dtc, new_vars_prefix,
                             date_imputation = "none", time_imputation = "none",
                             flag_imputation = "auto") {
  dtc_nm <- deparse(substitute(dtc))
  dtm_nm <- paste0(new_vars_prefix, "DTM")
  dtf_nm <- paste0(new_vars_prefix, "DTF")
  tmf_nm <- paste0(new_vars_prefix, "TMF")
  raw    <- dataset[[dtc_nm]]
  ti_str <- switch(time_imputation, "first" = "00:00:00", "last" = "23:59:59", "00:00:00")
  imp    <- sapply(raw, function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    x2 <- if (!str_detect(x, "T")) paste0(x, "T", ti_str) else x
    dp <- .impute_dtc(str_split(x2, "T")[[1]][1], date_imputation)
    tp <- str_split(x2, "T")[[1]][2]
    paste0(dp, "T", tp)
  }, USE.NAMES = FALSE)
  dataset[[dtm_nm]] <- suppressWarnings(
    as.numeric(as.POSIXct(str_replace(imp, "T", " "), tz = "UTC")))
  dataset[[dtf_nm]] <- sapply(raw, function(x) {
    if (is.na(x)) NA_character_ else if (length(str_split(x, "-")[[1]]) < 3) "D" else NA_character_
  }, USE.NAMES = FALSE)
  dataset[[tmf_nm]] <- if_else(!str_detect(raw, "T") & !is.na(raw),
                                if (time_imputation == "first") "H" else "H", NA_character_)
  dataset
}

derive_vars_dtm_to_dt <- function(dataset, source_vars) {
  for (nm in sapply(source_vars, deparse)) {
    dataset[[str_replace(nm, "DTM$", "DT")]] <- trunc(dataset[[nm]] / 86400)
  }
  dataset
}

derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref_nm <- deparse(substitute(reference_date))
  for (nm in sapply(source_vars, deparse)) {
    d <- dataset[[nm]] - dataset[[ref_nm]]
    dataset[[str_replace(nm, "DT$", "DY")]] <- case_when(
      is.na(d) ~ NA_real_, d >= 0 ~ d + 1, TRUE ~ d)
  }
  dataset
}

derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL,
                                order = NULL, filter_add = NULL, mode = NULL,
                                missing_values = NULL) {
  by_nm <- sapply(by_vars, deparse)
  fe    <- substitute(filter_add)
  # collapse deparse to single string — avoids length > 1 error with compound expressions
  if (!is.null(fe) && paste(deparse(fe), collapse = " ") != "NULL")
    dataset_add <- dataset_add |> filter(!!fe)
  if (!is.null(order)) {
    ord_nm   <- sapply(order, deparse)
    ord_desc <- grepl("^desc\\(", ord_nm)
    ord_cl   <- str_replace_all(ord_nm, "^desc\\(|\\)$", "")
    for (i in rev(seq_along(ord_cl)))
      dataset_add <- if (ord_desc[i]) dataset_add |> arrange(desc(.data[[ord_cl[i]]]))
                     else dataset_add |> arrange(.data[[ord_cl[i]]])
  }
  if (!is.null(mode)) {
    fn <- if (mode == "first") dplyr::slice_head else dplyr::slice_tail
    dataset_add <- dataset_add |> group_by(across(all_of(by_nm))) |> fn(n = 1) |> ungroup()
  }
  if (!is.null(new_vars)) {
    nv_nm <- sapply(new_vars, deparse)           # c(new_name = "old_name")
    dataset_add <- dataset_add |>
      select(any_of(c(by_nm, unname(nv_nm)))) |> # select by OLD column names
      rename(any_of(nv_nm))                       # rename(any_of(c(new="old"))) ✓
    new_col_names <- names(nv_nm)                 # the NEW column names for the join
  } else {
    new_col_names <- setdiff(names(dataset_add), by_nm)
  }
  dataset |> left_join(dataset_add |> select(any_of(c(by_nm, new_col_names))), by = by_nm)
}

derive_var_trtdurd <- function(dataset, new_var = TRTDURD,
                                start_date = TRTSDT, end_date = TRTEDT) {
  out <- deparse(substitute(new_var))
  s   <- deparse(substitute(start_date))
  e   <- deparse(substitute(end_date))
  dataset[[out]] <- dataset[[e]] - dataset[[s]] + 1
  dataset
}

derive_vars_duration <- function(dataset, new_var, start_date, end_date,
                                  out_unit = "days", add_one = TRUE) {
  out <- deparse(substitute(new_var))
  s   <- deparse(substitute(start_date))
  e   <- deparse(substitute(end_date))
  d   <- dataset[[e]] - dataset[[s]]
  dataset[[out]] <- case_when(
    is.na(d) ~ NA_real_,
    out_unit == "months" ~ round(as.numeric(d) / 30.4375, 1),
    add_one  ~ d + 1,
    TRUE     ~ d
  )
  dataset
}

derive_var_merged_exist_flag <- function(dataset, dataset_add, by_vars, new_var, condition) {
  by_nm <- sapply(by_vars, deparse)
  out   <- deparse(substitute(new_var))
  ce    <- substitute(condition)
  flagged <- dataset_add |> filter(!!ce) |> distinct(across(all_of(by_nm))) |>
    mutate(!!out := "Y")
  dataset |> left_join(flagged, by = by_nm)
}

# event_source: describe one possible source of an event for derive_vars_extreme_event
event_source <- function(dataset_name, filter = NULL, date, set_values_to) {
  list(
    dataset_name   = dataset_name,
    filter         = substitute(filter),
    date           = substitute(date),
    set_values_to  = set_values_to
  )
}

# derive_vars_extreme_event: find first/last event across multiple sources
derive_vars_extreme_event <- function(dataset, by_vars, events, order, mode,
                                       source_datasets, new_vars) {
  by_nm <- sapply(by_vars, deparse)

  event_rows <- lapply(events, function(ev) {
    src <- source_datasets[[ev$dataset_name]]
    # Apply filter (collapse deparse to avoid length > 1 in compound expressions)
    if (!is.null(ev$filter) && paste(deparse(ev$filter), collapse = " ") != "NULL")
      src <- src |> filter(!!ev$filter)
    if (nrow(src) == 0) return(NULL)
    # Evaluate date expression
    src <- src |> mutate(.EVENT_DATE = !!ev$date)
    # Apply set_values_to: evaluate each expression (handles function calls like convert_dtc_to_dt())
    for (i in seq_along(ev$set_values_to)) {
      out_nm   <- names(ev$set_values_to)[i]
      src_expr <- ev$set_values_to[[i]]
      src <- src |> mutate(!!out_nm := !!src_expr)
    }
    src |> select(any_of(c(by_nm, ".EVENT_DATE", names(ev$set_values_to))))
  }) |> bind_rows()

  if (nrow(event_rows) == 0) {
    # new_vars are unnamed exprs e.g. exprs(DTHCAUS, DTHDT) — use values, not names
    for (nm in sapply(new_vars, deparse)) dataset[[nm]] <- NA_character_
    return(dataset)
  }

  # Sort by date
  ord_nm  <- sapply(order, deparse)
  if (mode == "first")
    event_rows <- event_rows |> arrange(across(all_of(intersect(ord_nm, names(event_rows)))))
  else
    event_rows <- event_rows |> arrange(desc(across(all_of(intersect(ord_nm, names(event_rows))))))

  # Take first/last per subject
  fn <- if (mode == "first") dplyr::slice_head else dplyr::slice_tail
  best <- event_rows |>
    group_by(across(all_of(by_nm))) |>
    fn(n = 1) |>
    ungroup() |>
    select(-any_of(".EVENT_DATE"))

  dataset |> left_join(best, by = by_nm)
}

convert_dtc_to_dt <- function(dtc) {
  suppressWarnings(as.numeric(as.Date(.impute_dtc(dtc, "first"))))
}

# ── Source SDTM datasets ──────────────────────────────────────────────────────

dm <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "DM",
  USUBJID  = c("01-701-1015", "01-701-1023", "01-702-1042", "01-702-1101", "01-703-1211"),
  SUBJID   = c("1015", "1023", "1042", "1101", "1211"),
  SITEID   = c("701", "701", "702", "702", "703"),
  AGE      = c(63L, 64L, 71L, 56L, 45L),
  AGEU     = "YEARS",
  SEX      = c("M", "F", "M", "F", "M"),
  RACE     = c("WHITE", "WHITE", "ASIAN", "WHITE", "BLACK OR AFRICAN AMERICAN"),
  ETHNIC   = c("NOT HISPANIC OR LATINO", "NOT HISPANIC OR LATINO",
               "NOT HISPANIC OR LATINO", "HISPANIC OR LATINO", "NOT HISPANIC OR LATINO"),
  ARM      = c("Xanomeline High Dose", "Xanomeline Low Dose",
               "Placebo", "Xanomeline High Dose", "Placebo"),
  ACTARM   = c("XANOMELINE HIGH DOSE", "XANOMELINE LOW DOSE",
               "PLACEBO", "XANOMELINE HIGH DOSE", "PLACEBO"),
  RFXSTDTC = c("2014-01-02", "2012-08-05", "2013-07-09", "2013-04-14", "2014-03-22"),
  RFXENDTC = c("2014-06-15", "2012-12-22", "2013-12-15", "2013-10-30", "2014-09-01"),
  DTHDTC   = c(NA, NA, NA, NA, NA)
) |> convert_blanks_to_na()

ex <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "EX",
  USUBJID  = c("01-701-1015", "01-701-1015",
               "01-701-1023", "01-702-1042",
               "01-702-1101", "01-703-1211"),
  EXTRT    = c("XANOMELINE", "XANOMELINE",
               "XANOMELINE", "PLACEBO",
               "XANOMELINE", "PLACEBO"),
  EXDOSE   = c(81, 81, 54, 0, 81, 0),
  EXSEQ    = c(1L, 2L, 1L, 1L, 1L, 1L),
  EXSTDTC  = c("2014-01-02", "2014-02-02",
               "2012-08-05", "2013-07-09",
               "2013-04-14", "2014-03-22"),
  EXENDTC  = c("2014-01-31", "2014-06-15",
               "2012-12-22", "2013-12-15",
               "2013-10-30", "2014-09-01")
) |> convert_blanks_to_na()

ds <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "DS",
  USUBJID  = c("01-701-1015", "01-701-1015",
               "01-701-1023", "01-702-1042",
               "01-702-1101", "01-703-1211"),
  DSSEQ    = c(1L, 2L, 1L, 1L, 1L, 1L),
  DSCAT    = c("PROTOCOL MILESTONE", "DISPOSITION EVENT",
               "DISPOSITION EVENT", "DISPOSITION EVENT",
               "PROTOCOL MILESTONE", "DISPOSITION EVENT"),
  DSDECOD  = c("RANDOMIZED", "COMPLETED",
               "COMPLETED", "ADVERSE EVENT",
               "RANDOMIZED", "COMPLETED"),
  DSTERM   = c("RANDOMIZED", "COMPLETED STUDY",
               "COMPLETED STUDY", "DISCONTINUED DUE TO AE",
               "RANDOMIZED", "COMPLETED STUDY"),
  DSSTDTC  = c("2014-01-02", "2014-06-15",
               "2012-12-22", "2013-09-15",
               "2013-04-14", "2014-09-01")
) |> convert_blanks_to_na()

ae <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "AE",
  USUBJID  = c("01-701-1015", "01-702-1042", "01-702-1042"),
  AESEQ    = c(1L, 1L, 2L),
  AETERM   = c("HEADACHE", "DIZZINESS", "RASH"),
  AEDECOD  = c("Headache", "Dizziness", "Rash"),
  AESTDTC  = c("2014-03-15", "2013-08-10", "2013-09-01"),
  AEOUT    = c("RECOVERED/RESOLVED", "RECOVERED/RESOLVED", "RECOVERING/RESOLVING"),
  AEDTHDTC = c(NA, NA, NA)
) |> convert_blanks_to_na()

lb <- tibble(
  STUDYID  = "CDISCPILOT01",
  DOMAIN   = "LB",
  USUBJID  = c("01-701-1015", "01-701-1015", "01-701-1023", "01-702-1042"),
  LBTESTCD = c("ALT", "ALT", "ALT", "ALT"),
  LBDTC    = c("2014-01-02", "2014-04-01", "2012-08-05", "2013-07-09"),
  LBORRES  = c("28", "35", "22", "41")
) |> convert_blanks_to_na()

# ── Real pharmaverse datasets (with fallback if packages missing) ────────────
if (requireNamespace("pharmaversesdtm", quietly = TRUE)) {
  dm_real <- pharmaversesdtm::dm |> convert_blanks_to_na()
  ex_real <- pharmaversesdtm::ex |> convert_blanks_to_na()
} else {
  dm_real <- dm
  ex_real <- ex
}
if (requireNamespace("pharmaverseadam", quietly = TRUE)) {
  adsl_real <- pharmaverseadam::adsl
} else {
  adsl_real <- dm |> mutate(TRT01A = ACTARM, TRT01P = ARM)
}
```

## Welcome

This is Lesson 15 — **`{admiral}` Part 2: Building ADSL End-to-End**.

An ADSL has one row per subject and holds everything true about that subject: demographics, treatment dates, durations, population flags, and outcomes. It's built first — every other ADaM references it.

This lesson walks through a complete ADSL build using admiral, step by step:

1. Base dataset from DM
2. Treatment start/end datetimes → dates → duration
3. Disposition dates (randomization, end of study)
4. Cause of death and last known alive date (multi-source)
5. Population flags (SAFFL, ITTFL, RANDFL)
6. Age groups (AGEGR1, AGEGR1N)

---

## 1 — What's in an ADSL?

### 1.1 The "one row per subject" contract

ADSL is the **Subject Level Analysis Dataset**. By CDISC ADaM convention, it holds *exactly one row per subject* — no exceptions. Every other ADaM dataset (ADAE, ADLB, ADVS, ADTTE, etc.) joins to ADSL by `USUBJID` to pick up treatment dates, population flags, demographics, and disposition information.

> **SAS analog — "one row per subject":** In SAS you would build this with `PROC SUMMARY NWAY` or a `BY USUBJID; IF FIRST.USUBJID;` DATA step. The pharmaverse equivalent is implicit — every admiral derivation routine treats `USUBJID` (plus `STUDYID`) as the unique key and refuses to create duplicates.

### 1.2 The canonical ADSL variable inventory

A production ADSL typically carries 30 to 60 variables. They split into six functional groups:

| Group | Example variables | What they tell you |
|---|---|---|
| **Identifiers** | `STUDYID`, `USUBJID`, `SUBJID`, `SITEID` | Who, where, what study |
| **Demographics** | `AGE`, `AGEU`, `SEX`, `RACE`, `ETHNIC`, `COUNTRY` | Patient characteristics, mostly copied from DM |
| **Treatment** | `ARM`, `ACTARM`, `TRT01P`, `TRT01A`, `TRTSDT`, `TRTEDT`, `TRTSDTM`, `TRTEDTM`, `TRTDURD` | Planned vs actual treatment + exposure window |
| **Disposition** | `RANDDT`, `EOSDT`, `EOSSTT`, `DCSREAS`, `DCSDT` | End-of-study status, why they left |
| **Outcome** | `DTHFL`, `DTHDT`, `DTHCAUS`, `LSTALVDT` | Death and last-known-alive metadata |
| **Population flags** | `SAFFL`, `ITTFL`, `RANDFL`, `COMPLFL`, `PPROTFL` | Which analysis populations the subject belongs to |
| **Categorical groupings** | `AGEGR1`, `AGEGR1N`, `RACEGR1`, `REGION1` | Pre-computed grouping variables for tables |

### 1.3 SAS-analog cheat sheet for ADSL derivations

| Derived variable | What it represents | Closest SAS analog | Gotcha vs SAS |
|---|---|---|---|
| `TRTSDT` | First active-treatment date | `PROC SQL; SELECT MIN(EXSTDT) AS TRTSDT FROM ex WHERE EXDOSE>0 GROUP BY USUBJID;` | Need to include placebo zero-dose rows; admiral does it via `filter_add` |
| `TRTEDT` | Last active-treatment date | `PROC SQL; SELECT MAX(EXENDT)...` | Must impute missing day/time before MAX |
| `TRTDURD` | Treatment duration (days, inclusive) | `INTCK('day', TRTSDT, TRTEDT) + 1` | The `+1` is mandatory; `add_one = TRUE` in admiral |
| `RANDDT` | Randomization date | DATA step: `IF DSDECOD='RANDOMIZED' THEN RANDDT=DSSTDT;` then `RETAIN` | admiral does the filter + merge in one call |
| `EOSDT` | End-of-study date | DATA step against DS filtered to `DSCAT='DISPOSITION EVENT'` | Don't forget to exclude `SCREEN FAILURE` |
| `DTHCAUS` | Cause of death | `COALESCE(DTHCAUS_DS, DTHCAUS_AE, DTHCAUS_DM)` | admiral uses `derive_vars_extreme_event` for priority-ranked sources |
| `LSTALVDT` | Last known alive date | `MAX(of date1-date20)` over many SDTM domains | admiral takes the *latest* event date across event sources |
| `SAFFL` | Safety population flag | `IF first.USUBJID AND EXDOSE>0 THEN SAFFL='Y';` | CDISC says use `'Y'` or **missing**, never `'N'` |
| `AGEGR1` | Age band | `PROC FORMAT; VALUE agegr LOW-<18='<18' 18-64='18-64' 65-HIGH='>=65';` | admiral uses `case_when` — no separate FORMAT catalog |

### 1.4 Why admiral feels different from SAS

In SAS, an ADSL is typically built across **10–20 DATA steps and PROC SQL joins**, with `WORK.` intermediates piling up. Programmers spend most of their time naming intermediate datasets and remembering which one is current.

admiral collapses this to roughly **15 function calls in a single pipe**. Each call is *declarative* — you say *what* you want (the latest qualifying EX record per subject), not *how* to compute it (sort, retain, first-dot-USUBJID).

> **SAS analog mental model:** Think of each `derive_vars_*()` call as a self-contained, well-named macro `%DERIVE_TRTSDT()`. The macro already knows the pattern (filter → sort → first record per subject → merge). You only specify what's variable: which dataset, which filter, which output name.

```
  PROC SQL;                                |   adsl <- adsl |>
    CREATE TABLE trt AS                    |     derive_vars_merged(
      SELECT USUBJID, MIN(EXSTDT) AS TRTSDT|       dataset_add = ex,
      FROM ex                              |       filter_add  = EXDOSE > 0,
      WHERE EXDOSE > 0                     |       new_vars    = exprs(TRTSDT = EXSTDT),
      GROUP BY USUBJID;                    |       order       = exprs(EXSTDT),
  QUIT;                                    |       mode        = "first",
  DATA adsl;                               |       by_vars     = exprs(USUBJID))
    MERGE dm(IN=a) trt(IN=b);              |
    BY USUBJID;                            |
    IF a;                                  |
  RUN;                                     |
```

```{r quiz-adsl-overview, echo=FALSE}
quiz(
  question(
    "ADSL has one row per:",
    answer("Subject per visit"),
    answer("Subject", correct = TRUE),
    answer("Subject per parameter"),
    random_answer_order = TRUE
  ),
  question(
    "Why is ADSL typically the first ADaM you build?",
    answer("It's the smallest and easiest"),
    answer("Every other ADaM joins to ADSL for subject-level metadata like treatment dates and population flags", correct = TRUE),
    answer("Regulators require it first"),
    random_answer_order = TRUE
  ),
  question(
    "TRTSDTM vs TRTSDT — the difference is:",
    answer("TRTSDTM is a datetime (with time component); TRTSDT is a date-only version", correct = TRUE),
    answer("They are identical — both store dates"),
    answer("TRTSDT comes from DS; TRTSDTM comes from EX"),
    random_answer_order = TRUE
  ),
  question(
    "Which SAS construct most closely mirrors the way admiral builds ADSL in a single pipe?",
    answer("A long DATA step with many KEEP/DROP statements"),
    answer("A series of named %macro calls chained by reusing the same library dataset", correct = TRUE),
    answer("PROC IML"),
    answer("PROC FCMP"),
    random_answer_order = TRUE
  ),
  question(
    "ADSL has exactly one row per subject. Which SAS PROC most directly produces a one-row-per-subject summary from a many-row source?",
    answer("PROC SUMMARY NWAY", correct = TRUE),
    answer("PROC TRANSPOSE"),
    answer("PROC SORT NODUPKEY"),
    answer("PROC FORMAT"),
    random_answer_order = TRUE
  )
)
```

---

## 2 — Setup and base dataset

### 2.1 Why we start from DM

Every ADSL begins life as a copy of DM (Demographics), one row per subject. DM already carries the identifiers (`STUDYID`, `USUBJID`), demographics (`AGE`, `SEX`, `RACE`), and planned treatment (`ARM`, `ACTARM`). The ADSL build is essentially the process of *augmenting* DM with derived variables from other SDTM domains (EX, DS, AE, LB, etc.).

| Setup step | What it does | SAS analog |
|---|---|---|
| `convert_blanks_to_na(dm)` | Replace `""` (empty character) with `NA` | None — SAS treats `""` and missing the same |
| `mutate(TRT01P = ARM)` | Initialize planned treatment for period 1 | `TRT01P = ARM;` inside DATA step |
| `mutate(TRT01A = ACTARM)` | Initialize actual treatment for period 1 | `TRT01A = ACTARM;` |

### 2.2 The `convert_blanks_to_na()` rationale

XPT (SAS Transport) files store missing character values as empty strings `""`. R distinguishes `""` from `NA` — and most admiral derivations test `!is.na(x)`. If you skip `convert_blanks_to_na()`, an empty-string `ARM` will pass `!is.na(ARM)` checks but fail `ARM == "Placebo"` equality. Result: silent miss-flagging of subjects.

> **SAS analog:** In SAS, the character missing value is just an empty string and there is no distinction. You never had to think about this. In R it is mandatory — and pharmaverse handles it with one helper.

```
   SAS WORK.DM after PROC IMPORT      |   R after PROC IMPORT-equivalent (haven::read_xpt)
                                      |
   USUBJID    ARM                     |   USUBJID    ARM
   01-701-1   "Placebo"               |   01-701-1   "Placebo"
   01-701-2   "" (treated as missing) |   01-701-2   ""   ← NOT NA in R!
                                      |
   IF ARM = "" THEN ...               |   if_else(is.na(ARM), ...)   would MISS this row
                                      |   ↑ until we run convert_blanks_to_na()
```

### 2.3 Anatomy of the base-dataset call

```r
adsl <- dm |>                      # start from DM (already blank-converted in setup)
  mutate(
    TRT01P = ARM,                  # planned treatment for period 1 = planned arm
    TRT01A = ACTARM                # actual treatment for period 1 = actual arm
  )
```

> **SAS analog mental model:** This is the equivalent of `DATA adsl; SET dm; TRT01P = ARM; TRT01A = ACTARM; RUN;`. The native pipe `|>` plays the role of `SET` followed by assignment statements.

```{r adsl-base-setup}
# Step 0: All SDTM already loaded and blanks converted (see setup)
# Step 1: Create base ADSL from DM
adsl <- dm |>
  mutate(
    TRT01P = ARM,
    TRT01A = ACTARM
  )

glimpse(adsl)
```

> **Read the output:** `glimpse()` prints one column per row — the analog of `PROC CONTENTS` in SAS. You should see 5 rows (one per subject) and the original DM columns plus the two new ones (`TRT01P`, `TRT01A`). If you see 6+ rows here something is wrong — DM must be unique on USUBJID.

### 2.4 Exercise: Inspect the base ADSL

Count rows and columns and confirm there are no duplicate `USUBJID`s before any derivation runs.

```{r adsl-base-inspect, exercise=TRUE, exercise.lines=10}
adsl_base <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM)

# How many rows, columns, and unique subjects?
cat("rows:    ", nrow(adsl_base), "\n")
cat("cols:    ", ncol(adsl_base), "\n")
cat("unique:  ", n_distinct(adsl_base$___), "\n")
```

```{r adsl-base-inspect-hint-1}
# Unique subject identifier is USUBJID
```

```{r adsl-base-inspect-hint-2}
# n_distinct(adsl_base$USUBJID) should equal nrow(adsl_base)
```

```{r adsl-base-inspect-solution}
adsl_base <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM)

cat("rows:    ", nrow(adsl_base), "\n")
cat("cols:    ", ncol(adsl_base), "\n")
cat("unique:  ", n_distinct(adsl_base$USUBJID), "\n")
```

> **Read the output:** rows == unique subjects (5 == 5) — this confirms the "one row per subject" contract.

---

## 3 — Pre-process EX and DS

### 3.1 Why pre-process at all?

In SAS, you might be tempted to derive datetime variables inline inside the same DATA step that builds ADSL. In admiral the convention is the opposite: **pre-process the source datasets** so they each carry the derived numerical date/datetime columns *before* you ever merge them onto ADSL.

| Derived variable | What it represents | SAS analog | Gotcha vs SAS |
|---|---|---|---|
| `EXSTDTM` (numeric) | EX start datetime imputed from `EXSTDTC` | `EXSTDTM = INPUT(EXSTDTC, IS8601DT.);` | admiral imputes partial dates; SAS `INPUT` returns missing |
| `EXENDTM` | EX end datetime | `EXENDTM = INPUT(EXENDTC, IS8601DT.);` | use `date_imputation="last"` for end dates |
| `DSSTDT` (numeric date) | DS event date | `DSSTDT = INPUT(SUBSTR(DSSTDTC,1,10), YYMMDD10.);` | use `derive_vars_dt()` for date-only output |
| `EXSTTMF`, `EXSTDTF` | Imputation flags ("D" for day imputed, "Y" for year only) | None — SAS expects a separate flag DATA step | admiral emits them automatically |

### 3.2 Separation of concerns

Pre-processing achieves three goals that are hard to get from a single mega-DATA-step:

1. **Auditability.** A define.xml entry for `EXSTDTM` points at one function call on EX. Reviewers can re-run it in isolation.
2. **Testability.** You can unit-test `ex_ext` against a known answer set without rebuilding ADSL.
3. **Reuse.** The same `ex_ext` and `ds_ext` feed *both* TRTSDT/TRTEDT derivations *and* SAFFL/EOSDT. Pre-computing once avoids re-imputing dates twice.

> **SAS analog:** Think of `ex_ext` and `ds_ext` as named `WORK.` intermediate datasets — `WORK.EX_EXT`, `WORK.DS_EXT`. The pipe just makes the chaining explicit and lets the names die when the pipeline ends.

### 3.3 The `slice_derivation` / `restrict_derivation` companions

For derivations that need *different behavior for different subsets* (e.g. impute year for AE start dates but not for AE end dates), admiral provides `slice_derivation()` and `restrict_derivation()`. These wrap any `derive_vars_*` call and dispatch to different imputation rules based on a filter. They're outside the scope of this lesson but are the canonical answer to the SAS programmer's question: *"How do I do an IF-THEN-ELSE around a whole imputation?"*

### 3.4 Anatomy of `derive_vars_dtm()` on EX

```r
ex_ext <- ex |>
  derive_vars_dtm(
    dtc             = EXSTDTC,   # input ISO 8601 character column (SAS char(19))
    new_vars_prefix = "EXST",    # output prefix → creates EXSTDTM, EXSTDTF, EXSTTMF
    date_imputation = "first",   # missing day  → 01; missing month → 01
    time_imputation = "first"    # missing time → 00:00:00
  )
```

> **SAS analog mental model:** This single call replaces a DATA step that would (a) `INPUT()` the character to numeric, (b) handle partial dates with a custom format or `SUBSTR`+`IF MISSING()` block, and (c) emit a `DTF`/`TMF` flag column manually. admiral wraps all three in one function with auditable arguments.

### 3.5 EX pre-processing in action

Before merging, derive datetime variables on the source datasets.

```{r preprocess-ex}
# EX: add start and end datetimes
ex_ext <- ex |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first",
    time_imputation = "first"
  ) |>
  derive_vars_dtm(
    dtc             = EXENDTC,
    new_vars_prefix = "EXEN",
    date_imputation = "last",
    time_imputation = "last"
  )

ex_ext |>
  select(USUBJID, EXSTDTC, EXSTDTM, EXENDTC, EXENDTM) |>
  mutate(
    EXST_human = as.POSIXct(EXSTDTM, origin = "1970-01-01", tz = "UTC"),
    EXEN_human = as.POSIXct(EXENDTM, origin = "1970-01-01", tz = "UTC")
  )
```

```{r preprocess-ds}
# DS: add date variable from DSSTDTC
ds_ext <- ds |>
  derive_vars_dt(
    dtc             = DSSTDTC,
    new_vars_prefix = "DSST",
    date_imputation = "first"
  )

ds_ext |>
  select(USUBJID, DSCAT, DSDECOD, DSSTDTC, DSSTDT) |>
  mutate(DSSTDT_human = as.Date(DSSTDT, origin = "1970-01-01"))
```

> **Read the output:** `EXSTDTM` and `EXENDTM` are numeric seconds-since-1970 (the internal admiral representation). The `_human` columns show the date you'd actually see in a CRF. For the second EX record of subject 1015 (`EXSTDTC = "2014-02-02"`), confirm `EXST_human` reads `2014-02-02 00:00:00 UTC` — that confirms time imputation to midnight worked.

### 3.6 Exercise: Pre-process EX with end-of-day time imputation

Re-run `derive_vars_dtm()` on `ex` but impute time as `23:59:59` instead of `00:00:00` for both start and end dates. This mirrors a "last possible second of the day" convention some sponsors use for windowing.

```{r ex-preprocess-lastlast, exercise=TRUE, exercise.lines=18}
ex_late <- ex |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first",
    time_imputation = ___        # impute to end of day
  ) |>
  derive_vars_dtm(
    dtc             = EXENDTC,
    new_vars_prefix = "EXEN",
    date_imputation = "last",
    time_imputation = ___
  )

ex_late |>
  select(USUBJID, EXSTDTC, EXSTDTM, EXENDTC, EXENDTM) |>
  mutate(EXST_human = as.POSIXct(EXSTDTM, origin = "1970-01-01", tz = "UTC"))
```

```{r ex-preprocess-lastlast-hint-1}
# time_imputation = "last" → 23:59:59
```

```{r ex-preprocess-lastlast-hint-2}
# Both time_imputation arguments should be "last"
```

```{r ex-preprocess-lastlast-solution}
ex_late <- ex |>
  derive_vars_dtm(
    dtc = EXSTDTC, new_vars_prefix = "EXST",
    date_imputation = "first", time_imputation = "last"
  ) |>
  derive_vars_dtm(
    dtc = EXENDTC, new_vars_prefix = "EXEN",
    date_imputation = "last",  time_imputation = "last"
  )

ex_late |>
  select(USUBJID, EXSTDTC, EXSTDTM, EXENDTC, EXENDTM) |>
  mutate(EXST_human = as.POSIXct(EXSTDTM, origin = "1970-01-01", tz = "UTC"))
```

> **Read the output:** `EXST_human` now reads `2014-01-02 23:59:59` for subject 1015 — the time-imputation flag worked.

### 3.7 Exercise: Pre-process DS for date-only output

Use `derive_vars_dt()` (note: `dt`, not `dtm`) on `ds$DSSTDTC` with `date_imputation = "first"` and prefix `"DSST"`. Confirm the new variable is `DSSTDT` (date) plus `DSSTDTF` (imputation flag).

```{r ds-preprocess-exercise, exercise=TRUE, exercise.lines=12}
ds_minimal <- ds |>
  derive_vars_dt(
    dtc             = ___,
    new_vars_prefix = ___,
    date_imputation = "first"
  )

ds_minimal |>
  select(USUBJID, DSCAT, DSSTDTC, DSSTDT, DSSTDTF)
```

```{r ds-preprocess-exercise-hint-1}
# dtc = DSSTDTC, new_vars_prefix = "DSST"
```

```{r ds-preprocess-exercise-solution}
ds_minimal <- ds |>
  derive_vars_dt(
    dtc = DSSTDTC, new_vars_prefix = "DSST", date_imputation = "first"
  )

ds_minimal |>
  select(USUBJID, DSCAT, DSSTDTC, DSSTDT, DSSTDTF)
```

> **Read the output:** `DSSTDT` is a numeric date (days since 1970). `DSSTDTF` is `NA` here because every `DSSTDTC` is a full date — there was nothing to impute.

---

## 4 — Treatment start and end datetimes

### 4.1 The treatment-window concept (TRTSDT vs RFXSTDT)

Two date variables look superficially similar but mean very different things:

| Variable | Source | Meaning | Used by |
|---|---|---|---|
| `RFXSTDTC` | DM (per SDTM IG) | Date/time of *first study treatment* — populated by the data manager | SDTM submission, traceability |
| `TRTSDT` | derived from EX in ADSL | First date the subject *actually received* a qualifying dose | Every downstream ADaM via study day calcs |

In a perfect data world they agree. In the real world they sometimes diverge — for example, when EX records the first administration of a placebo run-in but RFXSTDTC was already populated from the screening visit. **You always derive `TRTSDT` from EX yourself**; you never trust RFXSTDTC alone.

| Derived variable | What it represents | Closest SAS analog | Gotcha vs SAS |
|---|---|---|---|
| `TRTSDTM` | Numeric datetime of first qualifying dose | `PROC SQL; SELECT MIN(EXSTDTM)... GROUP BY USUBJID;` | Need partial-date imputation BEFORE the MIN |
| `TRTSDT` | Date-only version of TRTSDTM | `TRTSDT = DATEPART(TRTSDTM);` | admiral uses `derive_vars_dtm_to_dt()` for consistency |
| `TRTEDTM` | Last qualifying dose end datetime | `PROC SQL; SELECT MAX(EXENDTM)... GROUP BY USUBJID;` | Use `time_imputation = "last"` to push to 23:59:59 |
| `TRTSTMF` | Time imputation flag for TRTSDTM | None — SAS programmers track it manually | admiral auto-emits — copy along with TRTSDTM |

### 4.2 The placebo + dose>0 filter (why both)

The canonical filter in admiral templates is:

```r
filter_add = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
```

Read this in English: *"Include the record if the subject received a non-zero dose, OR if the dose is zero but they were receiving placebo."*

Why both? Because **placebo IS treatment** in CDISC terms — a placebo subject IS in the safety population and DOES have a TRTSDT. If you filter `EXDOSE > 0` alone, your placebo arm vanishes from every downstream analysis. This is one of the top-five P21 findings.

> **SAS analog:** The same logic in a DATA step:
> ```
> IF EXDOSE > 0 OR (EXDOSE = 0 AND INDEX(EXTRT, 'PLACEBO') > 0) THEN OUTPUT;
> ```
> The OR clause is mandatory; never simplify it away.

### 4.3 First exposure: `derive_vars_merged()` with `mode = "first"`

`derive_vars_merged()` is the workhorse of admiral. It performs four steps that in SAS would each be a separate program:

1. Apply `filter_add` to the source dataset (filter).
2. Sort by `order` (sort).
3. Pick the `mode = "first"` (or last) record per `by_vars` group (first.USUBJID).
4. Merge the named `new_vars` onto the input dataset (merge).

```
   PROC SORT DATA=ex_ext OUT=ex_first;       |   derive_vars_merged(
     BY USUBJID EXSTDTM EXSEQ;                |     dataset_add = ex_ext,
   RUN;                                       |     filter_add  = (EXDOSE > 0 |
   DATA ex_first;                             |                    (EXDOSE == 0 &
     SET ex_first;                            |                    str_detect(EXTRT, "PLACEBO"))),
     BY USUBJID;                              |     order       = exprs(EXSTDTM, EXSEQ),
     WHERE (EXDOSE > 0) OR                    |     mode        = "first",
           (EXDOSE = 0 AND                    |     new_vars    = exprs(TRTSDTM = EXSTDTM,
            INDEX(EXTRT,'PLACEBO')>0);        |                          TRTSTMF = EXSTTMF),
     IF FIRST.USUBJID;                        |     by_vars     = exprs(STUDYID, USUBJID))
     RENAME EXSTDTM=TRTSDTM EXSTTMF=TRTSTMF;  |
   RUN;                                       |
   DATA adsl;                                 |
     MERGE adsl(IN=a) ex_first(KEEP=USUBJID   |
       TRTSDTM TRTSTMF);                      |
     BY USUBJID; IF a;                        |
   RUN;                                       |
```

### 4.4 Last exposure: `derive_vars_merged()` with `mode = "last"`

For TRTEDTM the pattern is identical except (a) the order column is `EXENDTM`, (b) `mode = "last"`, and (c) the imputation rule on the source was `time_imputation = "last"` (push missing times to 23:59:59 so a missing-time row doesn't *appear* earlier than a 14:00 row).

### 4.5 Imputing missing exposure end dates

Real EX datasets routinely have partial dates: `EXENDTC = "2014-06"` (year-month, no day) or `EXENDTC = ""` (missing entirely). The `date_imputation` / `time_imputation` arguments on `derive_vars_dtm()` resolve these to *some* datetime so MIN/MAX has something to compare. Convention:

| Variable | date_imputation | time_imputation | Why |
|---|---|---|---|
| TRTSDTM (start) | `"first"` | `"first"` | A missing day is conservatively early — first of month, midnight |
| TRTEDTM (end) | `"last"` | `"last"` | A missing day is conservatively late — last of month, 23:59:59 |

This convention is the **"furthest exposure window" rule** — when you don't know exactly when treatment started or stopped, you imagine the widest possible window. It gives you a conservative TRTDURD (longer rather than shorter) and avoids accidentally putting an AE *before* TRTSDT just because the day was missing.

### 4.6 Anatomy of the TRTEDTM call

```r
derive_vars_merged(
  dataset_add = ex_ext,                                 # pre-processed EX with EXENDTM derived
  filter_add  = (EXDOSE > 0 |                           # active dose...
                 (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &  # ...or zero-dose placebo
                !is.na(EXENDTM),                        # AND end datetime known
  new_vars    = exprs(TRTEDTM = EXENDTM,                # rename in-flight: EXENDTM → TRTEDTM
                      TRTETMF = EXENTMF),               # carry over the time-imputation flag
  order       = exprs(EXENDTM, EXSEQ),                  # sort: end dt, then seq as tiebreaker
  mode        = "last",                                 # pick the LATEST end per subject
  by_vars     = exprs(STUDYID, USUBJID)                 # group key — never just USUBJID
)
```

> **SAS analog mental model:** Imagine a macro `%LAST_EX(dataset=ex_ext, var=EXENDTM, output=TRTEDTM)` — admiral's argument names map 1:1 to the macro's parameters. The `order=` argument is the macro's `BY` statement; `mode="last"` is `LAST.USUBJID`; `new_vars=` is `RENAME=()`.

### 4.7 SAS equivalent: PROC SQL with MAX/MIN

The closest SQL formulation in SAS:

```
PROC SQL;
  CREATE TABLE trt AS
    SELECT  USUBJID,
            MIN(EXSTDTM) AS TRTSDTM FORMAT=IS8601DT.,
            MAX(EXENDTM) AS TRTEDTM FORMAT=IS8601DT.
    FROM    ex_ext
    WHERE  (EXDOSE > 0) OR (EXDOSE = 0 AND INDEX(EXTRT,'PLACEBO') > 0)
    GROUP BY USUBJID;
QUIT;

DATA adsl;
  MERGE adsl(IN=a) trt;
  BY USUBJID;
  IF a;
RUN;
```

That's two PROC SQL aggregations + one MERGE step — three programs. admiral does it in two `derive_vars_merged()` calls, each self-documenting.

### 4.8 Common pitfalls

| Pitfall | Symptom | Fix |
|---|---|---|
| Forgot `(EXDOSE == 0 & PLACEBO)` clause | Placebo subjects have `TRTSDT = NA`, missing from safety summaries | Use the canonical filter verbatim |
| Used `mode = "first"` for end date | `TRTEDT` is the earliest END (often = same day as start) | `mode = "last"` for end variables |
| Forgot `time_imputation = "last"` for end | Missing-time end rows look earlier than full-time end rows | Use `"last"` time imputation for end variables |
| Joined on `USUBJID` only, missing `STUDYID` | Works in single-study data; fails silently in pooled analyses | Always `by_vars = exprs(STUDYID, USUBJID)` |
| Did not `convert_blanks_to_na()` on EX | `is.na(EXTRT)` evaluates FALSE for empty strings; filter passes wrong rows | Apply at SDTM ingestion |

### 4.9 Live derivation

The standard EX filter: `(EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))` — counts active doses AND zero-dose placebo records.

```{r derive-trt-dates}
# Derive TRTSDTM (first qualifying dose) and TRTEDTM (last qualifying dose end)
adsl <- adsl |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM, TRTSTMF = EXSTTMF),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXENDTM),
    new_vars    = exprs(TRTEDTM = EXENDTM, TRTETMF = EXENTMF),
    order       = exprs(EXENDTM, EXSEQ),
    mode        = "last",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl |>
  select(USUBJID, ACTARM, TRTSDTM, TRTEDTM) |>
  mutate(
    TRTS_human = as.POSIXct(TRTSDTM, origin = "1970-01-01", tz = "UTC"),
    TRTE_human = as.POSIXct(TRTEDTM, origin = "1970-01-01", tz = "UTC")
  )
```

> **Read the output:** Subject `01-701-1015` got dosed twice (rows 1 and 2 in EX). `TRTS_human` should be `2014-01-02` (first dose start) and `TRTE_human` should be `2014-06-15` (last dose end). Subject `01-702-1042` is on Placebo (`EXDOSE = 0`) — *they should still get TRTSDTM populated* (the placebo OR clause). If you see NA here, your filter is wrong.

### 4.10 Exercise: Convert datetimes to dates and compute treatment duration

```{r trt-dates-to-dt, exercise=TRUE, exercise.lines=20}
# Step 1: convert TRTSDTM/TRTEDTM → TRTSDT/TRTEDT
adsl_dates <- adsl |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM))

# Step 2: derive TRTDURD = TRTEDT - TRTSDT + 1
adsl_dur <- adsl_dates |>
  derive_var_trtdurd()

adsl_dur |>
  select(USUBJID, ACTARM, TRTSDT, TRTEDT, TRTDURD) |>
  mutate(
    TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_human = as.Date(TRTEDT, origin = "1970-01-01")
  )
```

### 4.11 Exercise: Validate TRTSDT against real pharmaverseadam data

Use `dm_real` and `ex_real` (the full CDISCPILOT01 SDTM datasets from pharmaversesdtm) to re-derive TRTSDT for the first 10 subjects, then compare to the published values in `adsl_real`.

> **Why this matters:** In real submissions, your ADSL is compared to an independently derived QC dataset. Discrepancies usually mean a wrong filter, wrong mode, or a missing `convert_blanks_to_na()`.

```{r trt-dates-real-validate, exercise=TRUE, exercise.lines=52}
# Step 1: subset to 10 subjects that appear in adsl_real
subs10 <- adsl_real$USUBJID[1:10]

dm10 <- dm_real |> filter(USUBJID %in% subs10)

ex10_ext <- ex_real |>
  filter(USUBJID %in% subs10) |>
  derive_vars_dtm(
    dtc             = EXSTDTC,
    new_vars_prefix = "EXST",
    date_imputation = "first",
    time_imputation = "first"
  )

# Step 2: derive TRTSDTM → TRTSDT using the standard EX filter
derived10 <- dm10 |>
  derive_vars_merged(
    dataset_add = ex10_ext,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM)) |>
  mutate(TRTSDT_derived = as.Date(TRTSDT, origin = "1970-01-01"))

# Step 3: pull reference from pharmaverseadam and compare
ref <- adsl_real |>
  filter(USUBJID %in% subs10) |>
  select(USUBJID, TRTSDT_ref = TRTSDT)

comparison <- left_join(
  derived10 |> select(USUBJID, ARM, TRTSDT_derived),
  ref, by = "USUBJID"
) |>
  mutate(
    d_derived = as.integer(TRTSDT_derived),
    d_ref     = as.integer(as.Date(TRTSDT_ref, origin = "1970-01-01")),
    MATCH     = (d_derived == d_ref) | (is.na(d_derived) & is.na(d_ref))
  ) |>
  select(USUBJID, ARM, TRTSDT_derived, TRTSDT_ref, MATCH)

cat("Matching subjects:", sum(comparison$MATCH, na.rm = TRUE), "/ 10\n\n")
print(comparison)
```

```{r trt-dates-real-validate-hint-1}
# The key EX filter: (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
# mode = "first"  picks the earliest qualifying dose start (= first treatment date)
# Any mismatch here suggests a filter, mode, or blank-handling discrepancy
```

```{r trt-dates-real-validate-solution}
subs10 <- adsl_real$USUBJID[1:10]
dm10 <- dm_real |> filter(USUBJID %in% subs10)

ex10_ext <- ex_real |>
  filter(USUBJID %in% subs10) |>
  derive_vars_dtm(
    dtc = EXSTDTC, new_vars_prefix = "EXST",
    date_imputation = "first", time_imputation = "first"
  )

derived10 <- dm10 |>
  derive_vars_merged(
    dataset_add = ex10_ext,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM)) |>
  mutate(TRTSDT_derived = as.Date(TRTSDT, origin = "1970-01-01"))

ref <- adsl_real |>
  filter(USUBJID %in% subs10) |>
  select(USUBJID, TRTSDT_ref = TRTSDT)

left_join(derived10 |> select(USUBJID, ARM, TRTSDT_derived), ref, by = "USUBJID") |>
  mutate(
    d_derived = as.integer(TRTSDT_derived),
    d_ref     = as.integer(as.Date(TRTSDT_ref, origin = "1970-01-01")),
    MATCH     = (d_derived == d_ref) | (is.na(d_derived) & is.na(d_ref))
  ) |>
  select(USUBJID, ARM, TRTSDT_derived, TRTSDT_ref, MATCH) |>
  print()
```

> **Read the output:** "Matching subjects: 10 / 10" means your derivation is byte-identical to pharmaverseadam's TRTSDT. Anything less and you have a filter or imputation discrepancy to investigate.

### 4.12 Exercise: Derive TRTSDTM only from active doses (deliberately wrong)

Re-derive `TRTSDTM` using only `EXDOSE > 0` — *without* the placebo OR clause. Then compare to the correct version and identify which subjects end up missing.

```{r trt-wrong-filter, exercise=TRUE, exercise.lines=22}
adsl_wrong <- dm |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add  = ___ & !is.na(EXSTDTM),    # active dose ONLY — no placebo clause
    new_vars    = exprs(TRTSDTM_wrong = EXSTDTM),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  )

# adsl already has TRTSDTM derived correctly above
left_join(
  adsl |> select(USUBJID, ACTARM, TRTSDTM_correct = TRTSDTM),
  adsl_wrong |> select(USUBJID, TRTSDTM_wrong),
  by = "USUBJID"
) |>
  mutate(missing_in_wrong = is.na(TRTSDTM_wrong) & !is.na(TRTSDTM_correct))
```

```{r trt-wrong-filter-hint-1}
# filter_add = EXDOSE > 0  ← the deliberately broken filter
```

```{r trt-wrong-filter-hint-2}
# Look for rows where missing_in_wrong = TRUE — those are the Placebo subjects
# whose TRTSDTM disappears because EXDOSE = 0.
```

```{r trt-wrong-filter-solution}
adsl_wrong <- dm |>
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add  = EXDOSE > 0 & !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM_wrong = EXSTDTM),
    order       = exprs(EXSTDTM, EXSEQ),
    mode        = "first",
    by_vars     = exprs(STUDYID, USUBJID)
  )

left_join(
  adsl |> select(USUBJID, ACTARM, TRTSDTM_correct = TRTSDTM),
  adsl_wrong |> select(USUBJID, TRTSDTM_wrong),
  by = "USUBJID"
) |>
  mutate(missing_in_wrong = is.na(TRTSDTM_wrong) & !is.na(TRTSDTM_correct))
```

> **Read the output:** Two subjects (`01-702-1042` and `01-703-1211`, both Placebo) have `missing_in_wrong = TRUE`. Their TRTSDTM vanished because the wrong filter excluded `EXDOSE = 0`. In a real submission this would show up in the safety table as "Placebo n = 0".

### 4.13 Exercise: Derive TRTEDT directly with end-of-day imputation

Skip the datetime step entirely: use `derive_vars_dt()` directly on `EXENDTC` to produce `EXENDT` (a date), then merge the LAST per subject as `TRTEDT`.

```{r trtedt-direct, exercise=TRUE, exercise.lines=20}
ex_dt_only <- ex |>
  derive_vars_dt(
    dtc             = EXENDTC,
    new_vars_prefix = "EXEN",
    date_imputation = ___      # impute partial dates to end of month
  )

adsl_direct <- dm |>
  derive_vars_merged(
    dataset_add = ex_dt_only,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXENDT),
    new_vars    = exprs(TRTEDT_direct = EXENDT),
    order       = exprs(EXENDT, EXSEQ),
    mode        = ___,          # last end date per subject
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_direct |>
  select(USUBJID, ACTARM, TRTEDT_direct) |>
  mutate(TRTEDT_human = as.Date(TRTEDT_direct, origin = "1970-01-01"))
```

```{r trtedt-direct-hint-1}
# date_imputation = "last"  for end dates
```

```{r trtedt-direct-hint-2}
# mode = "last" picks the maximum date per subject
```

```{r trtedt-direct-solution}
ex_dt_only <- ex |>
  derive_vars_dt(
    dtc = EXENDTC, new_vars_prefix = "EXEN", date_imputation = "last"
  )

adsl_direct <- dm |>
  derive_vars_merged(
    dataset_add = ex_dt_only,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) &
                  !is.na(EXENDT),
    new_vars    = exprs(TRTEDT_direct = EXENDT),
    order       = exprs(EXENDT, EXSEQ),
    mode        = "last",
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_direct |>
  select(USUBJID, ACTARM, TRTEDT_direct) |>
  mutate(TRTEDT_human = as.Date(TRTEDT_direct, origin = "1970-01-01"))
```

> **Read the output:** Each subject's `TRTEDT_human` is the latest `EXENDTC` from their EX records. This is the "shortcut" pattern — useful when you don't need TRTEDTM (datetime) downstream, only TRTEDT.

### 4.14 Knowledge check — treatment dates

```{r quiz-trt-dates, echo=FALSE}
quiz(
  question(
    "The canonical EX filter for TRTSDTM derivation includes a placebo OR clause. Why?",
    answer("Because PLACEBO is the SAS function used to flag placebo records"),
    answer("Because placebo subjects record EXDOSE = 0 but are still part of the safety population and need TRTSDTM populated", correct = TRUE),
    answer("Because admiral requires both clauses for syntactic reasons"),
    random_answer_order = TRUE
  ),
  question(
    "What is the SAS PROC SQL equivalent of derive_vars_merged(mode = 'first', by_vars = exprs(USUBJID), order = exprs(EXSTDTM))?",
    answer("PROC SQL SELECT FIRST(USUBJID) FROM ex"),
    answer("PROC SQL SELECT MIN(EXSTDTM) AS TRTSDTM FROM ex GROUP BY USUBJID", correct = TRUE),
    answer("PROC SQL SELECT * FROM ex WHERE EXSTDTM IS NOT NULL"),
    random_answer_order = TRUE
  ),
  question(
    "For TRTEDTM you set time_imputation = 'last'. What would go wrong with time_imputation = 'first'?",
    answer("admiral would throw an error — 'first' is invalid for end variables"),
    answer("A row with a missing time would impute to 00:00:00 and could be sorted *earlier* than rows with valid afternoon times, producing the wrong MAX", correct = TRUE),
    answer("The TRTEDTM would still be correct; the flag would just be wrong"),
    random_answer_order = TRUE
  ),
  question(
    "A pooled-study ADSL must group by STUDYID and USUBJID, not USUBJID alone. Why?",
    answer("STUDYID is mandatory for the metadata"),
    answer("USUBJID could in theory repeat across studies in a pooled dataset — STUDYID makes the key truly unique", correct = TRUE),
    answer("STUDYID drives the JOIN performance"),
    random_answer_order = TRUE
  ),
  question(
    "RFXSTDTC (from DM) and TRTSDT (derived in ADSL) usually agree. When they disagree, which one should ADSL trust?",
    answer("RFXSTDTC — it was set by the data manager"),
    answer("TRTSDT derived from EX — it reflects the actual administered exposure, which is what every downstream analysis needs", correct = TRUE),
    answer("Whichever is earlier"),
    random_answer_order = TRUE
  )
)
```

---

## 5 — Disposition dates

### 5.1 What "disposition" means in SDTM/ADaM

In SDTM, the DS (Disposition) domain records every milestone in a subject's journey: screening, randomization, treatment phase end, end-of-study. Each row has a `DSCAT` (the bucket — `"PROTOCOL MILESTONE"` or `"DISPOSITION EVENT"`) and a `DSDECOD` (the codified verb — `"RANDOMIZED"`, `"COMPLETED"`, `"ADVERSE EVENT"`, `"DEATH"`, etc.).

ADSL extracts three derived facts from DS:

| Derived variable | What it represents | Closest SAS analog | Gotcha vs SAS |
|---|---|---|---|
| `RANDDT` | Date of randomization | DATA step `IF DSDECOD='RANDOMIZED' THEN RANDDT=DSSTDT;` + RETAIN | Subjects who screen-failed never randomized → RANDDT = NA |
| `EOSDT` | End of study date | DATA step `IF DSCAT='DISPOSITION EVENT' THEN EOSDT=DSSTDT;` | Must exclude `"SCREEN FAILURE"` rows |
| `EOSSTT` | End of study status | DATA step `IF DSDECOD='COMPLETED' THEN EOSSTT='COMPLETED'; ELSE EOSSTT='DISCONTINUED';` | Tri-state: completed / discontinued / ongoing |
| `DCSREAS` | Discontinuation reason | DATA step `IF DSCAT='DISPOSITION EVENT' THEN DCSREAS=DSTERM;` | Free-text DSTERM; sponsors sometimes use DSDECOD |
| `DCSDT` | Discontinuation date | Same date as EOSDT for discontinuers | NA for completers |

### 5.2 The tri-state EOSSTT

End-of-study status has three valid values, not two:

| `EOSSTT` | When | Example DSDECOD |
|---|---|---|
| `"COMPLETED"` | Subject finished the protocol | `"COMPLETED"` |
| `"DISCONTINUED"` | Subject left early for any reason | `"ADVERSE EVENT"`, `"WITHDRAWAL BY SUBJECT"`, `"LOST TO FOLLOW-UP"` |
| `"ONGOING"` | Study is still running and subject hasn't reached EOS yet | (no DISPOSITION EVENT row exists yet) |

> **SAS analog:** A typical SAS implementation:
> ```
> IF DSCAT = 'DISPOSITION EVENT' AND DSDECOD = 'COMPLETED' THEN EOSSTT = 'COMPLETED';
> ELSE IF DSCAT = 'DISPOSITION EVENT' THEN EOSSTT = 'DISCONTINUED';
> ELSE EOSSTT = 'ONGOING';
> ```
> The R equivalent uses `case_when()` — same three branches, no fall-through bug because case_when stops at first TRUE.

### 5.3 Anatomy of the RANDDT call

```r
derive_vars_merged(
  dataset_add = ds_ext,                                   # DS with DSSTDT already derived
  filter_add  = DSCAT == "PROTOCOL MILESTONE" &           # randomization is a milestone...
                DSDECOD == "RANDOMIZED",                  # ...specifically this one
  new_vars    = exprs(RANDDT = DSSTDT),                   # promote the date as RANDDT
  by_vars     = exprs(STUDYID, USUBJID)                   # one record per subject expected
)
```

Note there's no `mode =` argument — because the filter is so tight (one record per subject) there's nothing to pick "first" or "last" from. If subjects could have multiple `"RANDOMIZED"` rows you would add `mode = "first"` to be defensive.

### 5.4 SAS PROC SQL equivalent

```
PROC SQL;
  CREATE TABLE rand AS
    SELECT  USUBJID, DSSTDT AS RANDDT
    FROM    ds
    WHERE   DSCAT = 'PROTOCOL MILESTONE' AND DSDECOD = 'RANDOMIZED';
QUIT;
DATA adsl; MERGE adsl(IN=a) rand; BY USUBJID; IF a; RUN;
```

> **SAS analog mental model:** Each `derive_vars_merged()` call replaces "PROC SQL + DATA-step MERGE" — about 10 lines of SAS condensed to 5 named arguments of a function call.

### 5.5 Live derivation

Both RANDDT and EOSDT come from DS with different DSDECOD filters:

```{r derive-disp-dates}
adsl <- adsl |>
  # Randomization date
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "PROTOCOL MILESTONE" & DSDECOD == "RANDOMIZED",
    new_vars    = exprs(RANDDT = DSSTDT),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  # End of study date
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(EOSDT = DSSTDT),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl |>
  select(USUBJID, RANDDT, EOSDT) |>
  mutate(
    RANDDT_human = as.Date(RANDDT, origin = "1970-01-01"),
    EOSDT_human  = as.Date(EOSDT,  origin = "1970-01-01")
  )
```

> **Read the output:** All five subjects should have both `RANDDT` and `EOSDT` populated — every subject in our mock data was randomized AND had a DISPOSITION EVENT. Subject `01-702-1042` (the AE discontinuation) has `EOSDT_human = 2013-09-15` — earlier than their `RFXENDTC` because they discontinued mid-treatment.

### 5.6 Exercise: Derive the discontinuation reason (DCSREAS)

Merge the DS term for "DISPOSITION EVENT" records onto ADSL as `DCSREAS`.

```{r derive-dcsreas, exercise=TRUE, exercise.lines=15}
adsl_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == ___ & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = ___),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_dcsreas |>
  select(USUBJID, ACTARM, DCSREAS)
```

```{r derive-dcsreas-hint-1}
# filter_add = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE"
# new_vars = exprs(DCSREAS = DSTERM)
```

```{r derive-dcsreas-solution}
adsl_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_dcsreas |>
  select(USUBJID, ACTARM, DCSREAS)
```

> **Read the output:** Subject `01-702-1042` should show `DCSREAS = "DISCONTINUED DUE TO AE"` — sourced from DS.DSTERM. The other four subjects, who completed, all show `DCSREAS = "COMPLETED STUDY"`.

### 5.7 Exercise: Derive the tri-state EOSSTT

Build `EOSSTT` using `case_when()`. Use `DSDECOD == "COMPLETED"` for the COMPLETED branch and treat any other DISPOSITION EVENT as DISCONTINUED.

```{r eosstt-exercise, exercise=TRUE, exercise.lines=16}
adsl_eosstt <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSDECOD = DSDECOD),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  mutate(
    EOSSTT = case_when(
      DCSDECOD == ___           ~ "COMPLETED",
      !is.na(DCSDECOD)          ~ "DISCONTINUED",
      TRUE                      ~ ___          # no DISPOSITION row → still on study
    )
  )

adsl_eosstt |> select(USUBJID, ACTARM, DCSDECOD, EOSSTT)
```

```{r eosstt-exercise-hint-1}
# DCSDECOD == "COMPLETED"  ~ "COMPLETED"
# TRUE branch returns "ONGOING"
```

```{r eosstt-exercise-hint-2}
# In our sample data every subject has a DISPOSITION row, so no one is ONGOING.
# In a real study mid-conduct, subjects without a DS row stay ONGOING.
```

```{r eosstt-exercise-solution}
adsl_eosstt <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSDECOD = DSDECOD),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  mutate(
    EOSSTT = case_when(
      DCSDECOD == "COMPLETED" ~ "COMPLETED",
      !is.na(DCSDECOD)        ~ "DISCONTINUED",
      TRUE                    ~ "ONGOING"
    )
  )

adsl_eosstt |> select(USUBJID, ACTARM, DCSDECOD, EOSSTT)
```

> **Read the output:** Four subjects → `EOSSTT = "COMPLETED"`, one (1042, AE) → `"DISCONTINUED"`. No one is `"ONGOING"` because our mock data is end-of-study; in a snapshot during conduct you'd see ONGOINGs.

### 5.8 Knowledge check — disposition

```{r quiz-disposition, echo=FALSE}
quiz(
  question(
    "EOSSTT is a tri-state variable. What are the three permitted values?",
    answer("'Y', 'N', NA"),
    answer("'COMPLETED', 'DISCONTINUED', 'ONGOING'", correct = TRUE),
    answer("'PASS', 'FAIL', 'PENDING'"),
    random_answer_order = TRUE
  ),
  question(
    "A subject screen-failed (DSDECOD = 'SCREEN FAILURE'). What does the standard filter_add do to them?",
    answer("Includes them with EOSDT set to screening date"),
    answer("Excludes them — they never reached randomization or treatment, so they're not in any analysis population", correct = TRUE),
    answer("Includes them with a special flag SCRNFLAG = 'Y'"),
    random_answer_order = TRUE
  ),
  question(
    "The SAS equivalent of `derive_vars_merged(filter_add = DSCAT == 'PROTOCOL MILESTONE' & DSDECOD == 'RANDOMIZED', new_vars = exprs(RANDDT = DSSTDT))` is:",
    answer("DATA adsl; SET ds; IF DSCAT='PROTOCOL MILESTONE' AND DSDECOD='RANDOMIZED'; RANDDT=DSSTDT; RUN; — then MERGE back by USUBJID", correct = TRUE),
    answer("PROC RANK GROUPS=2 OUT=randomization"),
    answer("PROC TRANSPOSE OUT=rand"),
    random_answer_order = TRUE
  ),
  question(
    "Why is DCSREAS sourced from DSTERM (the free-text reason) rather than DSDECOD (the coded reason)?",
    answer("DSTERM is required; DSDECOD is optional"),
    answer("DSDECOD has only a handful of codes; DSTERM carries the protocol-specific reason wording sponsors want in the final ADaM", correct = TRUE),
    answer("DSDECOD is dropped from XPT exports"),
    random_answer_order = TRUE
  )
)
```

---

## 6 — Cause of death: multi-source derivation

### 6.1 Why death has multiple sources

In a clinical trial, a subject's death is captured in at least three places, often four:

| Source | Field carrying cause | Field carrying date |
|---|---|---|
| DM | `DTHDTC` (when known at randomization) | `DTHDTC` |
| DS | `DSDECOD == "DEATH"`, `DSTERM` (cause text) | `DSSTDTC` |
| AE | `AEOUT == "FATAL"`, `AEDECOD` (preferred term) | `AEDTHDTC` |
| Sometimes CE/EX/lab | adverse-event-like records | various |

The DTHCAUS variable in ADSL must take the *first* (earliest) of these with a defined priority order. That's the multi-source derivation pattern.

### 6.2 SAS analog: COALESCE() and DATA-step priority

In SAS you'd typically build per-source intermediate tables and then COALESCE:

```
PROC SQL;
  SELECT USUBJID,
         COALESCE(DTHCAUS_DS, DTHCAUS_AE, DTHCAUS_DM) AS DTHCAUS,
         COALESCE(DTHDT_DS,   DTHDT_AE,   DTHDT_DM)   AS DTHDT  FORMAT=DATE9.
    FROM dthsources;
QUIT;
```

The drawback: `COALESCE` picks the first *non-missing* source, not the earliest *date*. If DM has a wrong cause and DS has the right one, COALESCE picks DM. admiral's `derive_vars_extreme_event()` fixes this by sorting all source events together and picking the true first/last by date.

| Derived variable | What it represents | Closest SAS analog | Gotcha vs SAS |
|---|---|---|---|
| `DTHDT` | Death date | `COALESCE(DTHDT_DS, DTHDT_AE, ...)` | admiral picks by event order, not source order |
| `DTHCAUS` | Cause of death | `COALESCE(DTHCAUS_DS, ...)` | Free text from the source row |
| `DTHFL` | Death flag | `IF NOT MISSING(DTHDT) THEN DTHFL='Y';` | Derived from DTHDT existence |

### 6.3 The `event_source()` builder

Each event source is described by a small "list" object that says: "look in this dataset, with this filter, taking this date column, and copying these output values." It's a lightweight declarative DSL — exactly the kind of thing SAS macros end up reinventing.

```r
src_ds_death <- event_source(
  dataset_name = "ds",                           # which input dataset key to look up
  filter       = DSDECOD == "DEATH",             # apply this filter to that dataset
  date         = DSSTDT,                          # this is the event date column
  set_values_to = exprs(                          # assign these output columns:
    DTHCAUS = DSTERM,                            #   DTHCAUS comes from DSTERM
    DTHDT   = DSSTDT                              #   DTHDT comes from DSSTDT
  )
)
```

### 6.4 Anatomy of `derive_vars_extreme_event()`

```r
adsl <- adsl |>
  derive_vars_extreme_event(
    by_vars         = exprs(STUDYID, USUBJID),    # one record per subject
    events          = list(src_ds_death,           # candidate sources, ordered by priority
                           src_ae_death),
    order           = exprs(DTHDT),                # sort events by this column
    mode            = "first",                     # take the EARLIEST death record
    source_datasets = list(ds = ds_ext, ae = ae),  # lookup table: name → actual dataset
    new_vars        = exprs(DTHCAUS, DTHDT)        # which output columns survive
  )
```

> **SAS analog mental model:** Think of `events = list(...)` as a SAS macro array `&srcds1`, `&srcds2`. The macro loops over them, filters each, stacks the result, sorts by `order`, and picks `FIRST.USUBJID`. admiral's wrapper just gives the pattern a name and a tidy signature.

### 6.5 Live derivation

DTHCAUS can come from DS (death recorded as a disposition event) OR from AE (fatal adverse event). `derive_vars_extreme_event()` handles multi-source lookups elegantly.

```{r death-source-demo}
# Source 1: DS records for death
src_ds_death <- event_source(
  dataset_name = "ds",
  filter       = DSDECOD == "DEATH",
  date         = DSSTDT,
  set_values_to = exprs(
    DTHCAUS = DSTERM,
    DTHDT   = DSSTDT
  )
)

# Source 2: AE records with fatal outcome
src_ae_death <- event_source(
  dataset_name = "ae",
  filter       = AEOUT == "FATAL",
  date         = convert_dtc_to_dt(AEDTHDTC),
  set_values_to = exprs(
    DTHCAUS = AEDECOD,
    DTHDT   = convert_dtc_to_dt(AEDTHDTC)
  )
)

# Derive DTHCAUS and DTHDT from whichever source has the event
adsl <- adsl |>
  derive_vars_extreme_event(
    by_vars         = exprs(STUDYID, USUBJID),
    events          = list(src_ds_death, src_ae_death),
    order           = exprs(DTHDT),
    mode            = "first",
    source_datasets = list(ds = ds_ext, ae = ae),
    new_vars        = exprs(DTHCAUS, DTHDT)
  )

adsl |>
  select(USUBJID, DTHCAUS, DTHDT) |>
  mutate(DTHDT_human = as.Date(DTHDT, origin = "1970-01-01"))
```

> **Read the output:** All five subjects show `DTHCAUS = NA` and `DTHDT = NA` because our mock DS has no `DSDECOD == "DEATH"` and AE has no `AEOUT == "FATAL"`. In a study with deaths you would see the cause text and date populated for the affected subjects only.

### 6.6 Exercise: Verify no subjects have a death event in our sample data

Our sample `ds` has no `DSDECOD == "DEATH"` and `ae` has no `AEOUT == "FATAL"`. Confirm DTHCAUS is NA for all subjects.

```{r death-check, exercise=TRUE, exercise.lines=8}
# How many subjects have a non-NA DTHCAUS?
n_deaths <- adsl |>
  filter(!is.na(___)) |>
  nrow()

cat("Subjects with DTHCAUS:", n_deaths, "\n")
cat("Expected: 0 — no deaths in sample data\n")
```

```{r death-check-hint-1}
# filter(!is.na(DTHCAUS))
```

```{r death-check-solution}
n_deaths <- adsl |>
  filter(!is.na(DTHCAUS)) |>
  nrow()

cat("Subjects with DTHCAUS:", n_deaths, "\n")
cat("Expected: 0 — no deaths in sample data\n")
```

> **Read the output:** `Subjects with DTHCAUS: 0` confirms no death events. In a real study, this is a useful sanity check before tabulating mortality.

### 6.7 Exercise: Derive DTHFL from DTHDT existence

`DTHFL` is the death flag — `"Y"` if any death record exists, `NA` otherwise. Derive it from DTHDT.

```{r dthfl-exercise, exercise=TRUE, exercise.lines=10}
adsl_dthfl <- adsl |>
  mutate(
    DTHFL = if_else(!is.na(___), "Y", NA_character_)
  )

adsl_dthfl |> select(USUBJID, DTHDT, DTHCAUS, DTHFL)
```

```{r dthfl-exercise-hint-1}
# Test !is.na(DTHDT) — the death date column
```

```{r dthfl-exercise-hint-2}
# DTHFL = if_else(!is.na(DTHDT), "Y", NA_character_)
# Per ADaM IG, flag is "Y" or missing — never "N"
```

```{r dthfl-exercise-solution}
adsl_dthfl <- adsl |>
  mutate(
    DTHFL = if_else(!is.na(DTHDT), "Y", NA_character_)
  )

adsl_dthfl |> select(USUBJID, DTHDT, DTHCAUS, DTHFL)
```

> **Read the output:** All five rows show `DTHFL = NA` — consistent with our death-free mock data. In a real study, any row with a populated DTHDT would flip to `"Y"`. The classic bug is using `"N"` for the false branch — ADaM IG forbids that for `*FL` variables.

### 6.8 Knowledge check — cause of death

```{r quiz-dthcaus, echo=FALSE}
quiz(
  question(
    "DTHCAUS can come from DS, AE, or DM. Why is derive_vars_extreme_event() better than a SAS COALESCE() chain?",
    answer("COALESCE picks the first non-missing source by *priority order*; derive_vars_extreme_event() picks the first source by *event date*, which is the correct semantic", correct = TRUE),
    answer("COALESCE only works on numeric columns in SAS"),
    answer("derive_vars_extreme_event() is faster than COALESCE"),
    random_answer_order = TRUE
  ),
  question(
    "DTHFL = 'Y' when DTHDT is non-missing. What value does DTHFL take for a non-died subject per ADaM IG?",
    answer("'N'"),
    answer("missing (NA)", correct = TRUE),
    answer("'A' (alive)"),
    random_answer_order = TRUE
  ),
  question(
    "The SAS construct closest to admiral's `event_source()` + `derive_vars_extreme_event(mode='first')` is:",
    answer("PROC FORMAT"),
    answer("Multiple PROC SQL queries with UNION ALL into a stack, then `BY USUBJID date; IF FIRST.USUBJID;`", correct = TRUE),
    answer("PROC MEANS"),
    random_answer_order = TRUE
  ),
  question(
    "A subject has both an AE with AEOUT='FATAL' on 2014-05-10 AND a DS row with DSDECOD='DEATH' on 2014-05-12. With mode='first' and order=exprs(DTHDT), which date wins?",
    answer("2014-05-10 (the AE date — earliest event)", correct = TRUE),
    answer("2014-05-12 (the DS date — DS has priority over AE)"),
    answer("The function errors because of conflicting dates"),
    random_answer_order = TRUE
  )
)
```

---

## 7 — Last known alive date (LSTALVDT)

### 7.1 Clinical meaning

`LSTALVDT` = latest date with *any* evidence subject alive. Used downstream as censoring date in Overall Survival (Lesson 18). Wrong LSTALVDT → wrong OS censor → wrong KM curve.

Evidence sources (any of these prove subject alive at that date):

- AE record start date (AE happened → subject alive)
- LB sample collection date (sample collected → subject alive)
- VS measurement date
- EX dosing date
- DS disposition date if non-death

Output = max across all sources.

### 7.2 SAS-analog comparison

| admiral concept            | What it does                                | SAS counterpart                                                              |
|----------------------------|---------------------------------------------|------------------------------------------------------------------------------|
| `event_source(dataset, filter, date)` | Declare one candidate-date source  | `PROC SQL` `SELECT MAX(date) FROM src WHERE filter GROUP BY USUBJID;`         |
| `derive_vars_extreme_event(events=)`  | Stack many sources, pick max/min   | `DATA all; SET ae(in=a) lb(in=b) vs(in=c); ...; PROC SQL SELECT MAX(date)` |
| `mode = "last"`            | Pick latest date across sources             | `MAX(of d1-dN)` after a multi-source MERGE / SET                             |
| `set_values_to = exprs(LSTALVDT = ...)` | Project the chosen date into ADSL  | Final `MERGE adsl maxdates BY USUBJID;`                                      |

> **SAS mental model:** SAS pattern = `PROC SQL` UNION ALL of (USUBJID, date) from each source, then `SELECT USUBJID, MAX(date) GROUP BY USUBJID`, then MERGE back to ADSL. admiral `derive_vars_extreme_event` collapses all 3 steps.

### 7.3 ASCII side-by-side

```
   proc sql;                              |   src_ae <- event_source(
     create table alive_candidates as     |     dataset_name = "ae",
       select USUBJID, AESTDT as DT       |     filter = !is.na(AESTDTC),
         from ae                          |     date = convert_dtc_to_dt(AESTDTC),
        where AESTDTC ne ""               |     set_values_to =
       union all                          |       exprs(LSTALVDT =
       select USUBJID, LBDT as DT         |         convert_dtc_to_dt(AESTDTC)))
         from lb                          |
        where LBDTC ne "";                |   src_lb <- event_source(
                                          |     dataset_name = "lb",
     create table lstalv as               |     filter = !is.na(LBDTC),
       select USUBJID, max(DT) LSTALVDT   |     date = convert_dtc_to_dt(LBDTC),
         from alive_candidates            |     set_values_to =
        group by USUBJID;                 |       exprs(LSTALVDT =
                                          |         convert_dtc_to_dt(LBDTC)))
     create table adsl as                 |
       select a.*, b.LSTALVDT             |   adsl |>
         from adsl a                      |     derive_vars_extreme_event(
         left join lstalv b               |       by_vars = exprs(STUDYID,USUBJID),
           on a.USUBJID = b.USUBJID;      |       events  = list(src_ae, src_lb),
   quit;                                  |       order   = exprs(LSTALVDT),
                                          |       mode    = "last",
                                          |       source_datasets = list(ae=ae,lb=lb),
                                          |       new_vars = exprs(LSTALVDT))
```

### 7.4 Anatomy of `event_source()`

```r
event_source(
  dataset_name  = "ae",                          # name (key into source_datasets list)
  filter        = !is.na(AESTDTC),               # row predicate in src — drop rows w/o date
  date          = convert_dtc_to_dt(AESTDTC),    # candidate date expr
  set_values_to = exprs(                         # what to project onto ADSL row
    LSTALVDT = convert_dtc_to_dt(AESTDTC)
  )
)
```

- `dataset_name` → string. Must match key in `source_datasets = list(ae=ae,...)` passed to `derive_vars_extreme_event`.
- `filter` → unevaluated expr. Drops rows lacking a usable date.
- `date` → expr returning numeric Date. Used by `order` to rank candidates.
- `set_values_to` → named-list-of-exprs. Output column(s) to write onto ADSL.

### 7.5 Anatomy of `derive_vars_extreme_event()`

```r
adsl |>
  derive_vars_extreme_event(
    by_vars         = exprs(STUDYID, USUBJID),     # ADSL key
    events          = list(src_ae, src_lb),         # candidate sources
    order           = exprs(LSTALVDT),              # sort within group
    mode            = "last",                       # "last" = max date
    source_datasets = list(ae = ae, lb = lb),       # actual data frames
    new_vars        = exprs(LSTALVDT)               # which projected vars to keep
  )
```

`mode="last"` + `order=exprs(LSTALVDT)` = "max LSTALVDT wins per subject across all sources."

### 7.6 Common pitfalls

1. Forget death record exclusion → death date may sneak in as "alive evidence." Filter DS dispositions excluding death.
2. Use character `AESTDTC` (ISO string) directly in `order` → alphabetic sort → wrong winner for partial dates. Always wrap in `convert_dtc_to_dt()`.
3. Forget `source_datasets=list(...)` arg → admiral throws "dataset not found." Names in list must match `dataset_name` strings.
4. Use `mode="first"` accidentally → earliest evidence wins instead of latest → LSTALVDT becomes screening date. Always `"last"`.

### 7.7 Live derivation

```{r lstalvdt-demo}
# Source 1: latest AE start date (subject was alive when the AE started)
src_ae_alive <- event_source(
  dataset_name = "ae",
  filter       = !is.na(AESTDTC),
  date         = convert_dtc_to_dt(AESTDTC),
  set_values_to = exprs(LSTALVDT = convert_dtc_to_dt(AESTDTC))
)

# Source 2: latest lab collection date
src_lb_alive <- event_source(
  dataset_name = "lb",
  filter       = !is.na(LBDTC),
  date         = convert_dtc_to_dt(LBDTC),
  set_values_to = exprs(LSTALVDT = convert_dtc_to_dt(LBDTC))
)

adsl <- adsl |>
  derive_vars_extreme_event(
    by_vars         = exprs(STUDYID, USUBJID),
    events          = list(src_ae_alive, src_lb_alive),
    order           = exprs(LSTALVDT),
    mode            = "last",    # latest date across all sources
    source_datasets = list(ae = ae, lb = lb),
    new_vars        = exprs(LSTALVDT)
  )

adsl |>
  select(USUBJID, LSTALVDT) |>
  mutate(LSTALVDT_human = as.Date(LSTALVDT, origin = "1970-01-01"))
```

> **Read output:** one LSTALVDT per subject = max(AE start date, LB collection date). Downstream OS censor date pulls directly from this column.

### 7.8 Exercise: Add VS as a third alive-evidence source

Add `src_vs_alive` (mock — assume `vs` dataset with `VSDTC`). Stack three sources.

```{r lstalvdt-three-src, exercise=TRUE, exercise.lines=24}
# Mock vs dataset
vs <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015","01-701-1023"),
  VSTESTCD = "SYSBP",
  VSDTC    = c("2014-03-15","2012-09-01")
)

src_vs_alive <- event_source(
  dataset_name = "vs",
  filter       = !is.na(___),
  date         = convert_dtc_to_dt(___),
  set_values_to = exprs(LSTALVDT = convert_dtc_to_dt(VSDTC))
)

adsl_3src <- adsl |>
  derive_vars_extreme_event(
    by_vars = exprs(STUDYID, USUBJID),
    events  = list(src_ae_alive, src_lb_alive, ___),
    order   = exprs(LSTALVDT),
    mode    = "last",
    source_datasets = list(ae = ae, lb = lb, vs = vs),
    new_vars = exprs(LSTALVDT)
  )

adsl_3src |> select(USUBJID, LSTALVDT) |>
  mutate(LSTALVDT_human = as.Date(LSTALVDT, origin = "1970-01-01"))
```

```{r lstalvdt-three-src-hint-1}
# filter = !is.na(VSDTC)
# date   = convert_dtc_to_dt(VSDTC)
```

```{r lstalvdt-three-src-hint-2}
# events = list(src_ae_alive, src_lb_alive, src_vs_alive)
# source_datasets must include vs=vs
```

```{r lstalvdt-three-src-solution}
vs <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015","01-701-1023"),
  VSTESTCD = "SYSBP",
  VSDTC    = c("2014-03-15","2012-09-01")
)

src_vs_alive <- event_source(
  dataset_name = "vs",
  filter       = !is.na(VSDTC),
  date         = convert_dtc_to_dt(VSDTC),
  set_values_to = exprs(LSTALVDT = convert_dtc_to_dt(VSDTC))
)

adsl_3src <- adsl |>
  derive_vars_extreme_event(
    by_vars = exprs(STUDYID, USUBJID),
    events  = list(src_ae_alive, src_lb_alive, src_vs_alive),
    order   = exprs(LSTALVDT),
    mode    = "last",
    source_datasets = list(ae = ae, lb = lb, vs = vs),
    new_vars = exprs(LSTALVDT)
  )

adsl_3src |> select(USUBJID, LSTALVDT) |>
  mutate(LSTALVDT_human = as.Date(LSTALVDT, origin = "1970-01-01"))
```

> **Read output:** subject `01-701-1015` LSTALVDT now reflects VS date (`2014-03-15`) if it exceeds AE/LB max. Pattern scales: add EX, DS-non-death, FA dates the same way.

### 7.9 Knowledge check — LSTALVDT

```{r quiz-lstalvdt, echo=FALSE}
quiz(
  question(
    "`mode = \"last\"` chosen because:",
    answer("Want most recent date subject confirmed alive", correct = TRUE),
    answer("Want earliest date"),
    answer("Both first and last produce same result"),
    random_answer_order = TRUE
  ),
  question(
    "Closest SAS analog to `derive_vars_extreme_event()` for LSTALVDT?",
    answer("`PROC MEANS NMISS`"),
    answer("`PROC SQL` UNION ALL of source dates + `SELECT USUBJID, MAX(date) GROUP BY USUBJID` + MERGE back", correct = TRUE),
    answer("`PROC TRANSPOSE`"),
    random_answer_order = TRUE
  ),
  question(
    "Why must DS death records be excluded from alive-evidence sources?",
    answer("DS has no DTC variable"),
    answer("Death date proves subject NOT alive — including it makes LSTALVDT = date of death, biasing OS censoring", correct = TRUE),
    answer("admiral errors on DS"),
    random_answer_order = TRUE
  ),
  question(
    "`source_datasets = list(ae = ae, lb = lb)` purpose?",
    answer("Tells admiral which real data frames correspond to the `dataset_name` strings in `event_source()`", correct = TRUE),
    answer("Sorts the source data"),
    answer("Validates schema"),
    random_answer_order = TRUE
  )
)
```

---

## 8 — Treatment duration

### 8.1 The +1 inclusive-counting rule

TRTDURD = TRTEDT − TRTSDT + 1. The `+1` exists because both endpoints are *inclusive*: a subject treated only on 2014-01-05 (start = end) has duration **1 day**, not 0.

SAS counterpart: `INTCK('day', TRTSDT, TRTEDT) + 1` — same `+1`.

| Subject scenario       | TRTSDT     | TRTEDT     | Diff | +1 → TRTDURD |
|------------------------|------------|------------|------|--------------|
| One-day exposure       | 2014-01-05 | 2014-01-05 | 0    | 1            |
| Two-day exposure       | 2014-01-05 | 2014-01-06 | 1    | 2            |
| Standard 28-day cycle  | 2014-01-02 | 2014-01-29 | 27   | 28           |

> **Pitfall:** raw subtraction without `+1` undercounts by 1 every time. Causes downstream PK-exposure metrics to be wrong by ~3-4% in 28-day cycles.

### 8.2 SAS-analog comparison

| admiral                    | What it does                              | SAS equivalent                                   |
|----------------------------|-------------------------------------------|--------------------------------------------------|
| `derive_var_trtdurd()`     | TRTEDT − TRTSDT + 1 in days               | `TRTDURD = INTCK('day', TRTSDT, TRTEDT) + 1;`    |
| `derive_vars_duration(out_unit="months")` | Convert to months         | `TRTDURM = INTCK('month', TRTSDT, TRTEDT, 'CONTINUOUS');` |
| `derive_vars_duration(out_unit="years")`  | Convert to years          | `TRTDURY = YRDIF(TRTSDT, TRTEDT, 'ACTUAL');`     |
| `add_one = TRUE/FALSE`     | Toggle inclusive vs exclusive             | Explicit `+1` (or not) in SAS expression         |

### 8.3 ASCII side-by-side

```
   data adsl;                             |   adsl |>
     set adsl;                            |     derive_var_trtdurd(
     TRTDURD = TRTEDT - TRTSDT + 1;       |       new_var    = TRTDURD,
   run;                                   |       start_date = TRTSDT,
                                          |       end_date   = TRTEDT)
```

### 8.4 Anatomy of `derive_vars_duration()`

```r
derive_vars_duration(
  dataset    = adsl,        # piped in
  new_var    = TRTDURM,     # output var name
  start_date = TRTSDT,      # numeric date or datetime
  end_date   = TRTEDT,      # numeric date or datetime
  out_unit   = "months",    # "days" | "weeks" | "months" | "years"
  add_one    = FALSE        # +1 inclusive (default TRUE for days, FALSE for months/years)
)
```

- `add_one = TRUE` only makes sense for "days" — days are atomic. Months/years are continuous quantities; `+1` would shift by a whole unit incorrectly.
- `out_unit` controls divisor: months = `/30.4375`, years = `/365.25`.

### 8.5 Common pitfalls

1. `add_one = TRUE` with `out_unit = "months"` → off by 1 whole month. Disable for non-day units.
2. TRTSDT/TRTEDT not yet derived → silently NA result. Run duration AFTER treatment-date derivation.
3. Datetime instead of date inputs → fractional days. Use `derive_vars_dtm_to_dt()` first.

### 8.6 Live derivation

```{r trt-duration}
# TRTSDT and TRTEDT are now available — derive TRTDURD
adsl <- adsl |>
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM)) |>
  derive_var_trtdurd()

adsl |>
  select(USUBJID, TRTSDT, TRTEDT, TRTDURD) |>
  mutate(
    TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_human = as.Date(TRTEDT, origin = "1970-01-01")
  )
```

> **Read output:** TRTDURD = (TRTEDT − TRTSDT) + 1. Sanity check: subject treated 2014-01-02 → 2014-01-31 should show 30 days.

### 8.7 Exercise: Derive TRTDURM (treatment duration in months)

Use `derive_vars_duration()` with `out_unit = "months"`.

```{r trt-duration-months, exercise=TRUE, exercise.lines=12}
adsl_with_months <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURM,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = ___,
    add_one    = FALSE
  )

adsl_with_months |>
  select(USUBJID, TRTDURD, TRTDURM)
```

```{r trt-duration-months-hint-1}
# out_unit = "months"
```

```{r trt-duration-months-solution}
adsl_with_months <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURM,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = "months",
    add_one    = FALSE
  )

adsl_with_months |>
  select(USUBJID, TRTDURD, TRTDURM)
```

### 8.8 Exercise: Derive TRTDURY (treatment duration in years)

```{r trt-duration-years, exercise=TRUE, exercise.lines=12}
adsl_with_years <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURY,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = ___,
    add_one    = FALSE
  )

adsl_with_years |> select(USUBJID, TRTDURD, TRTDURY)
```

```{r trt-duration-years-hint-1}
# out_unit = "years"
```

```{r trt-duration-years-hint-2}
# add_one stays FALSE — years are continuous, +1 makes no sense
```

```{r trt-duration-years-solution}
adsl_with_years <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURY,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = "years",
    add_one    = FALSE
  )

adsl_with_years |> select(USUBJID, TRTDURD, TRTDURY)
```

> **Read output:** TRTDURY = (TRTEDT − TRTSDT) / 365.25. Used in PK exposure-time covariates.

### 8.9 Knowledge check — treatment duration

```{r quiz-trtdur, echo=FALSE}
quiz(
  question(
    "Why does `derive_var_trtdurd()` add 1?",
    answer("SAS legacy"),
    answer("Both endpoints inclusive — subject treated only on day X has duration 1, not 0", correct = TRUE),
    answer("To round up"),
    random_answer_order = TRUE
  ),
  question(
    "Closest SAS expression for TRTDURD?",
    answer("`INTCK('day', TRTSDT, TRTEDT) + 1`", correct = TRUE),
    answer("`TRTEDT - TRTSDT`"),
    answer("`DATDIF(TRTSDT, TRTEDT, 'ACT/ACT')`"),
    random_answer_order = TRUE
  ),
  question(
    "Why set `add_one = FALSE` for `out_unit = \"months\"`?",
    answer("Months continuous → `+1` shifts result by whole month → wrong", correct = TRUE),
    answer("admiral errors otherwise"),
    answer("Performance"),
    random_answer_order = TRUE
  )
)
```

---

## 9 — Population flags

### 9.1 What are populations?

Analysis populations = pre-defined subject subsets. Each table footer says "Population: SAFFL='Y'". Inclusion logic comes from protocol.

| Flag      | Meaning                              | Inclusion logic                                  |
|-----------|--------------------------------------|--------------------------------------------------|
| `SAFFL`   | Safety population                    | Received ≥ 1 dose (active or placebo)            |
| `ITTFL`   | Intent-to-treat                      | Randomized (RANDDT not NA)                        |
| `RANDFL`  | Randomized                           | Same as ITT in most studies                       |
| `FASFL`   | Full Analysis Set                    | ITT + ≥ 1 baseline efficacy assessment            |
| `PPROTFL` | Per-protocol                         | FAS minus major protocol deviations               |
| `COMPLFL` | Study completion                     | DCSREAS = "COMPLETED STUDY"                       |

### 9.2 The CDISC `"Y" or NA, never "N"` rule

ADaM IG Section 4.1.2: flag vars (`*FL`) take `"Y"` or `NA`. **`"N"` forbidden.**

```
   SAS habit (WRONG for CDISC):           |   admiral / CDISC (RIGHT):
   if condition then SAFFL = "Y";         |   condition = ...        →  "Y"
                else SAFFL = "N";  /*X*/  |   no match               →  NA
```

Reason: PROC FREQ on `SAFFL` should show only `"Y"` row. `"N"` rows leak into denominators → wrong percentages.

### 9.3 SAS-analog comparison

| admiral concept                  | SAS counterpart                                                       |
|----------------------------------|-----------------------------------------------------------------------|
| `derive_var_merged_exist_flag()` | `PROC SQL CASE WHEN EXISTS (subquery) THEN "Y" END`                   |
| `condition = EXDOSE > 0 \| ...`  | `WHERE` clause inside EXISTS subquery                                 |
| Output `"Y"` or `NA`             | No `ELSE` branch in `IF condition THEN FL="Y";`                       |
| `if_else(!is.na(RANDDT), "Y", NA_character_)` | `IF NOT MISSING(RANDDT) THEN ITTFL = "Y";`             |

### 9.4 ASCII side-by-side (SAFFL)

```
   proc sql;                              |   adsl |>
     create table adsl as                 |     derive_var_merged_exist_flag(
     select a.*,                          |       dataset_add = ex,
            case when exists              |       by_vars =
              (select 1 from ex b         |         exprs(STUDYID, USUBJID),
                where a.USUBJID = b.USUBJID|      new_var = SAFFL,
                  and (b.EXDOSE > 0       |       condition =
                       or (b.EXDOSE = 0   |         EXDOSE > 0 |
                       and b.EXTRT        |         (EXDOSE == 0 &
                       contains "PLACEBO")))|        str_detect(EXTRT,
            then "Y" end as SAFFL         |                    "PLACEBO")))
       from adsl a;                       |
   quit;                                  |
```

### 9.5 Anatomy of `derive_var_merged_exist_flag()`

```r
derive_var_merged_exist_flag(
  dataset     = adsl,                         # piped in
  dataset_add = ex,                           # source — search for ≥1 matching row per subject
  by_vars     = exprs(STUDYID, USUBJID),      # join key
  new_var     = SAFFL,                        # output flag var
  condition   = EXDOSE > 0 |                  # row predicate; sees columns of dataset_add only
                (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))
)
```

- `condition` is unquoted expression. Evaluated against rows of `dataset_add` only.
- Function has **no** `else_val` arg — enforces CDISC rule by API design.
- Output = base + one new column. Row count unchanged.

> **SAS mental model:** correlated `CASE WHEN EXISTS (...)` subquery in PROC SQL. No `ELSE NULL` needed — admiral defaults to `NA` automatically.

### 9.6 SAFFL: Safety population

A subject is in the safety population if they received at least one dose (positive dose OR placebo).

```{r saffl-demo}
adsl <- adsl |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
  )

adsl |> select(USUBJID, ACTARM, SAFFL)
```

> **Read output:** every subject with ≥ 1 qualifying EX row gets `SAFFL = "Y"`. Subjects without EX rows would get `NA` (screen failures, non-dosed). Downstream PROC FREQ would show only the `"Y"` row — exactly what tables need.

### 9.7 ITTFL and RANDFL: Intent-to-treat and randomized populations

These are derived from RANDDT — a subject is in ITT if they were randomized.

```{r ittfl-demo}
adsl <- adsl |>
  mutate(
    ITTFL  = if_else(!is.na(RANDDT), "Y", NA_character_),
    RANDFL = if_else(!is.na(RANDDT), "Y", NA_character_)
  )

adsl |> select(USUBJID, SAFFL, ITTFL, RANDFL)
```

> **Note:** `mutate(ITTFL = if_else(!is.na(RANDDT), "Y", NA_character_))` is the manual analog of `derive_var_merged_exist_flag` when the qualifying variable is already on ADSL. Either works.

### 9.8 New exercise — Derive PPROTFL (per-protocol)

Per-protocol = randomized AND treated AND no major protocol deviations. Two-step pattern: flag deviation existence, then PPROTFL = `"Y"` where deviation flag is NA.

```{r pprotfl-ex15, exercise=TRUE, exercise.lines=22}
# Mock major deviations
major_devs <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015"),
  DVSEV   = "MAJOR"
)

adsl_dev <- adsl |>
  derive_var_merged_exist_flag(
    dataset_add = major_devs,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = DEVFL,
    condition   = DVSEV == ___
  )

adsl_pp <- adsl_dev |>
  mutate(PPROTFL = if_else(
    SAFFL == "Y" & is.na(___),
    "Y",
    NA_character_
  ))

adsl_pp |> select(USUBJID, SAFFL, DEVFL, PPROTFL)
```

```{r pprotfl-ex15-hint-1}
# condition = DVSEV == "MAJOR"
# is.na(DEVFL) — PP requires no major deviation
```

```{r pprotfl-ex15-hint-2}
# PPROTFL = if_else(SAFFL == "Y" & is.na(DEVFL), "Y", NA_character_)
```

```{r pprotfl-ex15-solution}
major_devs <- tibble(
  STUDYID = "CDISCPILOT01",
  USUBJID = c("01-701-1015"),
  DVSEV   = "MAJOR"
)

adsl_dev <- adsl |>
  derive_var_merged_exist_flag(
    dataset_add = major_devs,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = DEVFL,
    condition   = DVSEV == "MAJOR"
  )

adsl_pp <- adsl_dev |>
  mutate(PPROTFL = if_else(
    SAFFL == "Y" & is.na(DEVFL),
    "Y",
    NA_character_
  ))

adsl_pp |> select(USUBJID, SAFFL, DEVFL, PPROTFL)
```

> **Read output:** subject `01-701-1015` gets `DEVFL="Y"` → `PPROTFL=NA`. Other safety-population subjects with no deviation get `PPROTFL="Y"`.

### 9.9 Knowledge check — population flags

```{r quiz-flag-values, echo=FALSE}
quiz(
  question(
    "Why population flags use `\"Y\"` or `NA`, not `\"Y\"` or `\"N\"`?",
    answer("CDISC ADaM IG Section 4.1.2: blank = 'not in population'; `\"N\"` is forbidden — would leak into PROC FREQ denominators", correct = TRUE),
    answer("`\"N\"` reserved in R"),
    answer("Saves storage space"),
    random_answer_order = TRUE
  ),
  question(
    "`derive_var_merged_exist_flag()` ≈ which SAS construct?",
    answer("`PROC SQL CASE WHEN EXISTS (correlated subquery) THEN \"Y\" END`", correct = TRUE),
    answer("`PROC FREQ`"),
    answer("`PROC TABULATE`"),
    random_answer_order = TRUE
  ),
  question(
    "ITT inclusion logic typically?",
    answer("Received ≥ 1 dose"),
    answer("Randomized (RANDDT not NA)", correct = TRUE),
    answer("Completed study"),
    random_answer_order = TRUE
  ),
  question(
    "PPROTFL derivation pattern?",
    answer("Direct `derive_var_merged_exist_flag` with `condition = no_deviation`"),
    answer("Flag deviation existence → `PPROTFL = \"Y\"` where deviation flag is NA AND SAFFL = \"Y\"", correct = TRUE),
    answer("`PROC SORT NODUPKEY`"),
    random_answer_order = TRUE
  )
)
```

### 9.10 Exercise: Derive COMPLFL (completion flag)

A subject is a completer if their DCSREAS is `"COMPLETED STUDY"`. Derive `COMPLFL` using a direct `mutate()`.

```{r complfl-exercise, exercise=TRUE, exercise.lines=10}
# First, get DCSREAS onto adsl (derive it from ds)
adsl_with_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  mutate(COMPLFL = if_else(DCSREAS == ___, "Y", NA_character_))

adsl_with_dcsreas |>
  select(USUBJID, ACTARM, DCSREAS, COMPLFL)
```

```{r complfl-exercise-hint-1}
# if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_)
```

```{r complfl-exercise-solution}
adsl_with_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_))

adsl_with_dcsreas |>
  select(USUBJID, ACTARM, DCSREAS, COMPLFL)
```

---

## 10 — Age groups

### 10.1 Why age groups?

Demographics tables present age both continuous (mean ± SD) and categorical (AGEGR1 strata). AGEGR1 enables stratified subgroup analysis: "AE incidence in age ≥ 65 stratum" etc.

### 10.2 SAS-analog comparison

| admiral concept                | SAS counterpart                                                       |
|--------------------------------|-----------------------------------------------------------------------|
| `case_when(AGE < 18 ~ "<18", ...)` | `PROC FORMAT VALUE agegrp LOW-17="<18" 18-64="18-64" 65-HIGH=">=65";` |
| AGEGR1 (character)             | `FORMAT AGE agegrp.;` then use in `PROC FREQ`                          |
| AGEGR1N (numeric)              | Companion ordering var: `IF AGE<18 THEN AGEGR1N=1;` etc.              |
| `mutate(...)` direct           | `DATA` step with `IF/THEN/ELSE` chain                                  |

> **SAS mental model:** PROC FORMAT for the labels + a DATA step `IF/ELSE` chain for the numeric companion. admiral fuses both into a single `case_when()` pair.

### 10.3 ASCII side-by-side

```
   proc format;                           |   adsl |>
     value agegrp                         |     mutate(
       low-17  = "<18"                    |       AGEGR1 = case_when(
       18-64   = "18-64"                  |         AGE < 18             ~ "<18",
       65-high = ">=65";                  |         between(AGE, 18, 64) ~ "18-64",
   run;                                   |         AGE >= 65            ~ ">=65"),
                                          |       AGEGR1N = case_when(
   data adsl;                             |         AGE < 18             ~ 1,
     set adsl;                            |         between(AGE, 18, 64) ~ 2,
     length AGEGR1 $8;                    |         AGE >= 65            ~ 3))
     if AGE < 18 then do;                 |
       AGEGR1 = "<18";  AGEGR1N = 1; end; |
     else if 18 <= AGE <= 64 then do;     |
       AGEGR1 = "18-64"; AGEGR1N = 2; end;|
     else if AGE >= 65 then do;           |
       AGEGR1 = ">=65"; AGEGR1N = 3; end; |
   run;                                   |
```

### 10.4 Anatomy of `case_when()` for age groups

```r
mutate(
  AGEGR1  = case_when(
    AGE < 18             ~ "<18",       # ordered: <18 evaluated first
    between(AGE, 18, 64) ~ "18-64",     # next condition
    AGE >= 65            ~ ">=65"       # catch-all upper
  ),
  AGEGR1N = case_when(
    AGE < 18             ~ 1L,          # 1L = integer companion
    between(AGE, 18, 64) ~ 2L,
    AGE >= 65            ~ 3L
  )
)
```

- `case_when` evaluates conditions top-to-bottom, first match wins.
- Always pair AGEGR1 (character display) with AGEGR1N (numeric order). PROC FREQ orders categories by AGEGR1N alphabetically-on-numeric so `>=65` sorts after `18-64`.
- Use `L` suffix on integer literals (`1L`) → AGEGR1N becomes integer, not double.

### 10.5 Common pitfalls

1. Missing AGEGR1N → PROC FREQ orders alphabetically: `"18-64" < "<18" < ">=65"` → wrong row order in table.
2. Overlapping bins (`AGE <= 18` then `18 <= AGE <= 64`) → 18-year-olds match both, first wins by case_when semantics but logic intent unclear. Use strict inequalities + `between()`.
3. Forget `NA` handling → subjects with AGE=NA fall through all conditions → AGEGR1=NA. Add explicit `TRUE ~ NA_character_` row only if needed; default `case_when` already returns NA on no match.

### 10.6 Live derivation

```{r age-groups-demo}
adsl <- adsl |>
  mutate(
    AGEGR1  = case_when(
      AGE < 18             ~ "<18",
      between(AGE, 18, 64) ~ "18-64",
      AGE >= 65            ~ ">=65"
    ),
    AGEGR1N = case_when(
      AGE < 18             ~ 1,
      between(AGE, 18, 64) ~ 2,
      AGE >= 65            ~ 3
    )
  )

adsl |>
  select(USUBJID, AGE, AGEGR1, AGEGR1N)
```

> **Read output:** AGEGR1 and AGEGR1N align row-by-row. PROC FREQ TABLES AGEGR1 / ORDER=FORMATTED ↔ pivot on AGEGR1N for correct visual ordering.

### 10.7 Exercise: Add a finer age categorization (AGEGR2)

Create `AGEGR2` with categories: `<40`, `40-54`, `55-64`, `>=65`.

```{r agegr2-exercise, exercise=TRUE, exercise.lines=14}
adsl_agegr2 <- adsl |>
  mutate(
    AGEGR2 = case_when(
      AGE < 40             ~ "<40",
      between(AGE, 40, 54) ~ "40-54",
      between(AGE, 55, 64) ~ "55-64",
      AGE >= 65            ~ ">=65"
    )
  )

adsl_agegr2 |>
  select(USUBJID, AGE, AGEGR1, AGEGR2)
```

> **Read output:** AGEGR2 partitions adults more finely — useful for elderly subgroup analyses (`55-64` vs `>=65`).

### 10.8 Knowledge check — age groups

```{r quiz-agegr, echo=FALSE}
quiz(
  question(
    "Closest SAS analog of `case_when()` for AGEGR1?",
    answer("`PROC FREQ`"),
    answer("`PROC FORMAT VALUE agegrp ...` + `FORMAT AGE agegrp.;`", correct = TRUE),
    answer("`PROC TRANSPOSE`"),
    random_answer_order = TRUE
  ),
  question(
    "Why pair AGEGR1 (char) with AGEGR1N (num)?",
    answer("PROC FREQ orders alphabetically on AGEGR1 → `\"18-64\" < \"<18\" < \">=65\"` wrong order. AGEGR1N forces correct numeric ordering.", correct = TRUE),
    answer("ADaM IG mandates redundancy"),
    answer("Speed"),
    random_answer_order = TRUE
  ),
  question(
    "`between(AGE, 18, 65)` includes endpoint 65?",
    answer("No — exclusive on upper"),
    answer("Yes — `between()` is inclusive on both ends. Use `between(AGE, 18, 64)` if 65 belongs in next group.", correct = TRUE),
    answer("Only if specified"),
    random_answer_order = TRUE
  )
)
```

---

## 11 — Putting it all together: the complete ADSL build

Here is the full ADSL derivation script, condensed. Run it to see the complete dataset.

```{r adsl-full-build, exercise=TRUE, exercise.lines=80}
# ── Pre-process source datasets ───────────────────────────────────────────────
ex_ext_full <- ex |>
  derive_vars_dtm(dtc = EXSTDTC, new_vars_prefix = "EXST",
                  date_imputation = "first", time_imputation = "first") |>
  derive_vars_dtm(dtc = EXENDTC, new_vars_prefix = "EXEN",
                  date_imputation = "last",  time_imputation = "last")

ds_ext_full <- ds |>
  derive_vars_dt(dtc = DSSTDTC, new_vars_prefix = "DSST", date_imputation = "first")

# ── Build ADSL ────────────────────────────────────────────────────────────────
adsl_full <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM) |>

  # 1. Treatment datetimes
  derive_vars_merged(
    dataset_add = ex_ext_full,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXSTDTM),
    new_vars    = exprs(TRTSDTM = EXSTDTM), order = exprs(EXSTDTM, EXSEQ),
    mode = "first", by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_merged(
    dataset_add = ex_ext_full,
    filter_add  = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXENDTM),
    new_vars    = exprs(TRTEDTM = EXENDTM), order = exprs(EXENDTM, EXSEQ),
    mode = "last", by_vars = exprs(STUDYID, USUBJID)) |>

  # 2. Convert to dates + duration
  derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM)) |>
  derive_var_trtdurd() |>

  # 3. Disposition dates
  derive_vars_merged(
    dataset_add = ds_ext_full,
    filter_add  = DSCAT == "PROTOCOL MILESTONE" & DSDECOD == "RANDOMIZED",
    new_vars    = exprs(RANDDT = DSSTDT), by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_merged(
    dataset_add = ds_ext_full,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(EOSDT = DSSTDT, DCSREAS = DSTERM),
    by_vars = exprs(STUDYID, USUBJID)) |>

  # 4. Population flags
  derive_var_merged_exist_flag(
    dataset_add = ex, by_vars = exprs(STUDYID, USUBJID), new_var = SAFFL,
    condition   = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))) |>
  mutate(ITTFL  = if_else(!is.na(RANDDT), "Y", NA_character_),
         RANDFL = if_else(!is.na(RANDDT), "Y", NA_character_),
         COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_)) |>

  # 5. Age groups
  mutate(
    AGEGR1  = case_when(AGE < 18 ~ "<18", between(AGE, 18, 64) ~ "18-64", AGE >= 65 ~ ">=65"),
    AGEGR1N = case_when(AGE < 18 ~ 1,     between(AGE, 18, 64) ~ 2,       AGE >= 65 ~ 3)
  )

# ── Summary ───────────────────────────────────────────────────────────────────
cat("ADSL rows (one per subject):", nrow(adsl_full), "\n")
cat("ADSL columns:", ncol(adsl_full), "\n\n")

adsl_full |>
  select(USUBJID, ACTARM, TRTSDT, TRTEDT, TRTDURD,
         SAFFL, ITTFL, RANDFL, COMPLFL, AGEGR1, AGEGR1N) |>
  mutate(
    TRTSDT_human = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_human = as.Date(TRTEDT, origin = "1970-01-01")
  ) |>
  select(-TRTSDT, -TRTEDT) |>
  print()
```

---

## 11.5 — P21 Derivation Diagnostics

Clinical programmers submitting to FDA routinely receive Pinnacle 21 (P21) validation reports that flag derivation errors in ADSL. This section presents three realistic bugs — each accompanied by a **P21-style derivation method specification** exactly as it would appear in a define.xml submission file. Your task: read the spec, study the broken code, identify what is wrong, and fix it.

> **Format for each exercise:** (1) the derivation specification text, (2) the broken programmer code, (3) the P21 error message received. Fill in the blank to correct the code.

---

### Diagnostic 1 — SAFFL: Safety Analysis Set Flag

**Derivation method (define.xml / Pinnacle 21 field):**

> *SAFFL = 'Y' if the subject has at least one exposure record in EX where EXDOSE > 0, OR where EXDOSE = 0 AND EXTRT contains the string 'PLACEBO'. Subjects with no qualifying EX record should have SAFFL missing (not 'N'). ADaM IG §4.5. Source variables: EX.EXDOSE, EX.EXTRT.*

The programmer submitted:

```r
adsl_draft <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM) |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = EXDOSE > 0          # ← bug here
  )
```

**P21 report:** *"SAFFL is missing for 2 Placebo subjects. All subjects with an EX record should have SAFFL populated per derivation spec."*

```{r p21-bug-saffl, exercise=TRUE, exercise.lines=18}
# Fix the SAFFL derivation so Placebo subjects (EXDOSE = 0, EXTRT contains "PLACEBO")
# are correctly included in the Safety Analysis Set.

adsl_fixed_saffl <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM) |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = ___
  )

adsl_fixed_saffl |> select(USUBJID, ACTARM, SAFFL)
# Expected: all 5 subjects have SAFFL = "Y"
```

```{r p21-bug-saffl-hint-1}
# The condition must cover BOTH active-dose records AND zero-dose placebo records:
# condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
```

```{r p21-bug-saffl-solution}
adsl_fixed_saffl <- dm |>
  mutate(TRT01P = ARM, TRT01A = ACTARM) |>
  derive_var_merged_exist_flag(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_var     = SAFFL,
    condition   = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
  )

adsl_fixed_saffl |> select(USUBJID, ACTARM, SAFFL)
# All 5 subjects should now have SAFFL = "Y"
```

---

### Diagnostic 2 — TRTDURD: Treatment Duration in Days

**Derivation method (define.xml / Pinnacle 21 field):**

> *TRTDURD = number of days of study treatment exposure. Calculated as TRTEDT − TRTSDT + 1 (inclusive of both the first and last treatment dates). Unit: days. Source variables: derived TRTSDT, TRTEDT.*

The programmer submitted:

```r
adsl_dur_draft <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURD,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = "days",
    add_one    = FALSE         # ← bug here
  )
```

**P21 report:** *"TRTDURD is 1 day less than expected for all subjects. Example: USUBJID 01-701-1015 — TRTSDT = 2014-01-02, TRTEDT = 2014-06-15. Expected TRTDURD = 165; observed = 164."*

```{r p21-bug-trtdurd, exercise=TRUE, exercise.lines=22}
# adsl already has TRTSDT and TRTEDT from the demo steps above.
# Fix the duration derivation to match the spec (inclusive: TRTEDT - TRTSDT + 1).

adsl_fixed_dur <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURD_fixed,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = "days",
    add_one    = ___           # fix this argument
  )

# Compare broken vs fixed vs expected
adsl_fixed_dur |>
  select(USUBJID, TRTSDT, TRTEDT, TRTDURD, TRTDURD_fixed) |>
  mutate(
    TRTSDT_d = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_d = as.Date(TRTEDT, origin = "1970-01-01"),
    expected = as.integer(TRTEDT_d - TRTSDT_d) + 1L
  )
```

```{r p21-bug-trtdurd-hint-1}
# add_one = TRUE  ← the spec says "inclusive of both endpoints" = TRTEDT - TRTSDT + 1
# add_one = FALSE gives TRTEDT - TRTSDT (excludes the start day) — 1 day short
```

```{r p21-bug-trtdurd-solution}
adsl_fixed_dur <- adsl |>
  derive_vars_duration(
    new_var    = TRTDURD_fixed,
    start_date = TRTSDT,
    end_date   = TRTEDT,
    out_unit   = "days",
    add_one    = TRUE
  )

adsl_fixed_dur |>
  select(USUBJID, TRTSDT, TRTEDT, TRTDURD, TRTDURD_fixed) |>
  mutate(
    TRTSDT_d = as.Date(TRTSDT, origin = "1970-01-01"),
    TRTEDT_d = as.Date(TRTEDT, origin = "1970-01-01"),
    expected = as.integer(TRTEDT_d - TRTSDT_d) + 1L
  )
```

---

### Diagnostic 3 — COMPLFL: Study Completion Flag

**Derivation method (define.xml / Pinnacle 21 field):**

> *COMPLFL = 'Y' if the subject completed the study per protocol (DCSREAS = 'COMPLETED STUDY'). Subjects who did not complete should have COMPLFL missing. Per ADaM IG §4.5, flag variables take values 'Y' or missing — the value 'N' is not permitted. Source variables: derived DCSREAS.*

The programmer submitted:

```r
adsl_comp_draft <- adsl_with_dcsreas |>
  mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", "N"))   # ← bug here
```

**P21 report:** *"AD0018 — COMPLFL contains the value 'N'. Flag variables (suffix -FL) must only contain 'Y' or missing per ADaM IG. Permissible values violated."*

```{r p21-bug-complfl, exercise=TRUE, exercise.lines=22}
# Build adsl_with_dcsreas (self-contained — ds_ext is available from the demo chunks above)
adsl_with_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  )

# Fix: replace "N" with the correct ADaM IG value for subjects not in the flag group
adsl_fixed_complfl <- adsl_with_dcsreas |>
  mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", ___))

adsl_fixed_complfl |>
  select(USUBJID, ACTARM, DCSREAS, COMPLFL)
# Expected: completers → "Y"; discontinuers → NA (not "N")
```

```{r p21-bug-complfl-hint-1}
# ADaM IG: flag variables are "Y" or NA — the value "N" is never valid.
# Use NA_character_ as the false branch:
# mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_))
```

```{r p21-bug-complfl-solution}
adsl_with_dcsreas <- adsl |>
  derive_vars_merged(
    dataset_add = ds_ext,
    filter_add  = DSCAT == "DISPOSITION EVENT" & DSDECOD != "SCREEN FAILURE",
    new_vars    = exprs(DCSREAS = DSTERM),
    by_vars     = exprs(STUDYID, USUBJID)
  )

adsl_fixed_complfl <- adsl_with_dcsreas |>
  mutate(COMPLFL = if_else(DCSREAS == "COMPLETED STUDY", "Y", NA_character_))

adsl_fixed_complfl |>
  select(USUBJID, ACTARM, DCSREAS, COMPLFL)
```

---

## 12 — Validation strategy

```{r quiz-validation, echo=FALSE}
quiz(
  question(
    "The standard validation approach for submission-quality ADSL is:",
    answer("One programmer writes it and a reviewer checks the code"),
    answer("Two programmers independently build the ADSL; outputs are compared with {diffdf}; discrepancies are resolved", correct = TRUE),
    answer("Run check_variables() from metatools — that's sufficient"),
    random_answer_order = TRUE
  ),
  question(
    "The add_one = FALSE convention in derive_vars_duration() applies to:",
    answer("TRTDURD — treatment duration in days"),
    answer("Elapsed-time variables like LDDTHELD, where you count calendar gap without an endpoint adjustment", correct = TRUE),
    answer("All duration variables"),
    random_answer_order = TRUE
  ),
  question(
    "The filter (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, 'PLACEBO'))) ensures:",
    answer("Only active doses count"),
    answer("Active doses AND zero-dose placebo records both count as exposure", correct = TRUE),
    answer("Missing doses are excluded"),
    random_answer_order = TRUE
  )
)
```

---

## 13 — Key takeaways

```{r quiz-final-15, echo=FALSE}
quiz(
  question(
    "A programmer derives SAFFL using `condition = EXDOSE > 0`. P21 reports SAFFL is missing for all Placebo subjects. What is the correct fix?",
    answer("Change to `condition = EXDOSE >= 0`"),
    answer("Add `mutate(SAFFL = if_else(ARM == 'Placebo', 'Y', SAFFL))` after the flag derivation"),
    answer("Use `condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, 'PLACEBO')))`", correct = TRUE),
    answer("Filter EX to only rows where EXDOSE > 0 before calling derive_var_merged_exist_flag()"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The derivation spec says TRTDURD = TRTEDT − TRTSDT + 1 (inclusive). The programmer uses `add_one = FALSE`. A subject treated January 1–31 gets TRTDURD = 30. The spec expects 31. What does `add_one = FALSE` calculate?",
    answer("The elapsed gap between two dates, not counting the start day — TRTEDT − TRTSDT with no +1 adjustment", correct = TRUE),
    answer("Duration rounded down to avoid double-counting"),
    answer("Duration in calendar months instead of days"),
    answer("Exactly the same result as add_one = TRUE"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "P21 rule AD0018 flags COMPLFL = 'N'. Why does 'N' violate ADaM IG and what replaces it?",
    answer("'N' is not a valid R logical — use FALSE instead"),
    answer("ADaM IG §4.5 specifies flag variables take values 'Y' or missing (NA_character_) — 'N' is prohibited", correct = TRUE),
    answer("'N' causes a metadata mismatch with the define.xml codelist for FL variables"),
    answer("'N' is valid in SDTM but not in ADaM — use 0 instead"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "LSTALVDT is the *last* known alive date. A programmer accidentally writes `mode = 'first'` in derive_vars_extreme_event(). What is the consequence?",
    answer("The function throws an error — 'first' is not a valid mode for alive-date sources"),
    answer("LSTALVDT is set to the *earliest* date the subject was observed alive, underestimating their known survival time", correct = TRUE),
    answer("LSTALVDT is replaced by the death date for subjects who died"),
    answer("No consequence — for alive-date derivations, 'first' and 'last' produce identical results"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The AGEGR1N derivation uses `between(AGE, 18, 65) ~ 2` for the '18-64' group. A 65-year-old ends up in AGEGR1N = 2 instead of AGEGR1N = 3 ('>=65'). What is the bug?",
    answer("`between(AGE, 18, 65)` includes 65 (it is inclusive on both ends) — the fix is `between(AGE, 18, 64)`", correct = TRUE),
    answer("case_when() evaluates all conditions and picks the last match, overwriting group 3 with group 2"),
    answer("Age 65 belongs in group 2 per CDISC ADaM IG guidance"),
    answer("The bug is only in the character AGEGR1 variable, not AGEGR1N"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "A programmer forgets convert_blanks_to_na() on DM before building ADSL. Some subjects have ARM = '' (empty string). What goes wrong downstream?",
    answer("admiral immediately errors when it encounters empty strings in character columns"),
    answer("Empty strings pass `!is.na(ARM)` checks, so those subjects appear to have a valid ARM — but str_detect() and equality comparisons on ARM then return NA or FALSE unexpectedly, producing wrong flags", correct = TRUE),
    answer("Nothing — admiral automatically coerces empty strings to NA internally"),
    answer("Empty strings are identical to NA in all R comparisons"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The spec says EOSDT is the date of the subject's disposition event, excluding screen failures. A programmer writes `filter_add = DSDECOD == 'COMPLETED'`. Which subjects will have a missing EOSDT?",
    answer("Only subjects who completed the study — they are included, not excluded"),
    answer("All subjects who discontinued (adverse event, withdrew consent, lost to follow-up) because their DSDECOD is not 'COMPLETED'", correct = TRUE),
    answer("Subjects with partial disposition dates only"),
    answer("All subjects — the filter should be on DSCAT, not DSDECOD"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "EXSTDTC = '2014-07' (year-month partial date, no day). The programmer calls derive_vars_dtm(..., date_imputation = 'first', time_imputation = 'first'). What datetime will EXSTDTM represent?",
    answer("2014-07-15T12:00:00 — admiral imputes the middle of the period"),
    answer("NA — admiral cannot impute partial dates with a missing day component"),
    answer("2014-07-01T00:00:00 — 'first' imputes the first day of the month and midnight time", correct = TRUE),
    answer("2014-07-31T23:59:59 — admiral always imputes to the end of the period by default"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "Why is derive_vars_extreme_event() preferred over two sequential derive_vars_merged() calls when deriving DTHCAUS from both DS and AE?",
    answer("derive_vars_extreme_event() executes in a single database pass and is more memory efficient"),
    answer("Sequential derive_vars_merged() calls overwrite the first source's value for subjects found in both sources — derive_vars_extreme_event() selects the true first/last event across all sources simultaneously", correct = TRUE),
    answer("derive_vars_merged() cannot join on date columns — only character keys"),
    answer("There is no practical difference; both approaches produce identical results for all subjects"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The standard admiral pipeline for treatment start date is: (1) derive_vars_dtm() on EX, (2) derive_vars_merged() onto ADSL, (3) derive_vars_dtm_to_dt(). Why does step 3 come last — why not derive the date directly in step 1?",
    answer("Admiral stores datetimes internally as numerics; deriving the datetime first preserves precision and time-flag metadata (TRTSTMF); dtm_to_dt() then creates the date-only variable cleanly from the stored datetime", correct = TRUE),
    answer("derive_vars_merged() can only operate on datetime (POSIXct) columns, not Date columns"),
    answer("Step 3 is only required for submission datasets — it is optional for internal QC"),
    answer("TRTSDT must be sourced from DM.RFXSTDTC, not EX — step 3 reads it from the correct domain"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```

---

## 14 — Lesson 15 Summary {.congratulations}

You now know how to build a production ADSL end-to-end:

- **Frame ADSL** — one row per subject, 30-40 canonical variables (identifiers, treatment, disposition, demographics, population flags, derived dates). SAS analog: a `PROC SUMMARY NWAY OUT=` collapsed dataset built from multiple SDTM domains.
- **Pre-process EX and DS** — separation of concerns: pre-derive numeric dates and clean blanks BEFORE merging onto ADSL. WORK-dataset analog to the SAS habit of building intermediate `WORK` datasets between steps.
- **Derive treatment dates (TRTSDT/TRTSDTM, TRTEDT/TRTEDTM)** — `derive_vars_merged()` with `mode = "first"/"last"`, the dose-greater-than-zero-OR-placebo filter, and `derive_vars_dtm_to_dt()` to project date-only versions. SAS analog: `PROC SQL SELECT MIN(EXSTDT) FROM ex WHERE EXDOSE > 0 GROUP BY USUBJID;` plus MERGE back.
- **Derive disposition dates (RANDDT, EOSDT, DCSREAS, EOSSTT)** — pull from DS via `derive_vars_merged()` with `filter_add = DSCAT == "DISPOSITION EVENT"`. Tri-state EOSSTT via `case_when()`.
- **Derive cause of death (DTHCAUS, DTHDT, DTHFL)** — multi-source via `derive_vars_extreme_event()` with `event_source()` objects from DS and AE. SAS analog: `COALESCE(DTHCAUS_DS, DTHCAUS_AE)` after multi-source DATA-step priority logic.
- **Derive LSTALVDT** — max of (AE start, LB collection, VS, EX, DS-non-death) via `derive_vars_extreme_event()` with `mode = "last"`. Drives downstream OS censoring.
- **Treatment duration (TRTDURD, TRTDURM, TRTDURY)** — `derive_var_trtdurd()` and `derive_vars_duration()`. Inclusive-day rule: `+1` for days, `add_one = FALSE` for months/years.
- **Population flags (SAFFL, ITTFL, RANDFL, COMPLFL, PPROTFL)** — `derive_var_merged_exist_flag()` with `condition = ...`. Strict CDISC rule: `"Y"` or `NA`, **never** `"N"`.
- **Age groups (AGEGR1/AGEGR1N, AGEGR2)** — paired character + numeric via `case_when()`. PROC FORMAT analog. The numeric companion fixes alphabetical-ordering pitfalls in PROC FREQ.
- **Validate against pharmaverseadam reference** — diffdf-style comparison: re-derive then check row-by-row match to published ADSL.
- **Read P21 diagnostics** — three classic submission bugs: SAFFL missing placebo subjects, TRTDURD off-by-one, COMPLFL using `"N"`.

You also know:

- All flag vars follow CDISC `"Y" or NA` rule. No ELSE branch.
- `derive_vars_merged()` is the workhorse — used 10+ times in one ADSL build.
- `derive_vars_extreme_event()` is the multi-source variant — used for DTHCAUS, LSTALVDT.
- Always pair character display vars (AGEGR1) with numeric companions (AGEGR1N) for correct table ordering.
- The build order matters: dates → merged vars → flags → groups → duration. Reversing creates `NA` cascades.

---

## What's next

Lesson 16 — **`{admiral}` Part 3** — covers BDS (Basic Data Structure) ADaMs: ADLB, ADVS, ADEG. These are long-format datasets with one row per subject per parameter per visit. The derivation patterns reuse `derive_vars_merged()`, `derive_var_extreme_flag()`, and the date helpers — but now applied to analysis-value reshaping and baseline flagging across many parameters.

After Lesson 16: OCCDS (ADAE, ADCM) in Lesson 17, time-to-event (ADTTE) in Lesson 18, and advanced admiral patterns in Lesson 19.
