---
title: "Lesson 16 — {admiral} Part 3: BDS Datasets (ADVS, ADLB, ADEG)"
author: "Hamid Tavakoli, MD, MSc"
tutorial:
  version: 1.1
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build BDS-shaped ADaMs — ADVS, ADLB, ADEG — covering merge, PARAMCD,
  derived parameters, baseline, change-from-baseline, and shift derivations.
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(rlang)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ─── NSE helpers (mocks of rlang::exprs and admiral params) ───────────────────
exprs <- function(...) as.list(substitute(list(...)))[-1]

params <- function(...) {
  parent_env <- parent.frame()
  dots <- as.list(substitute(list(...)))[-1]
  lapply(dots, function(x) {
    if (is.symbol(x)) deparse(x) else eval(x, envir = parent_env)
  })
}

# ─── XPT-style blank cleaner ──────────────────────────────────────────────────
convert_blanks_to_na <- function(dataset) {
  dataset |>
    mutate(across(where(is.character), ~ if_else(.x == "", NA_character_, .x)))
}

# ─── Date helpers ─────────────────────────────────────────────────────────────
derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation = "none",
                            flag_imputation = "auto") {
  .e <- substitute(dtc); dtc_nm <- if (is.character(.e)) .e else deparse(.e)
  dt_nm  <- paste0(new_vars_prefix, "DT")
  dtf_nm <- paste0(new_vars_prefix, "DTF")
  raw    <- dataset[[dtc_nm]]
  imputed <- sapply(raw, function(x) {
    if (is.na(x) || x == "") return(NA_character_)
    segs <- str_split(x, "-")[[1]]
    if (length(segs) == 1) segs <- c(segs, if (date_imputation == "last") "12" else "01")
    if (length(segs) == 2) segs <- c(segs, if (date_imputation == "last") "28" else "01")
    paste(segs, collapse = "-")
  }, USE.NAMES = FALSE)
  dataset[[dt_nm]]  <- suppressWarnings(as.numeric(as.Date(imputed)))
  dataset[[dtf_nm]] <- sapply(raw, function(x) {
    if (is.na(x)) return(NA_character_)
    segs <- str_split(x, "-")[[1]]
    if (length(segs) == 1) "M" else if (length(segs) == 2) "D" else NA_character_
  }, USE.NAMES = FALSE)
  dataset
}

derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref_nm  <- deparse(substitute(reference_date))
  src_nms <- sapply(source_vars, deparse)
  for (sv in src_nms) {
    dy_nm <- sub("DT$", "DY", sv)
    if (sv %in% names(dataset) && ref_nm %in% names(dataset)) {
      v <- dataset[[sv]]; r <- dataset[[ref_nm]]
      dataset[[dy_nm]] <- ifelse(is.na(v) | is.na(r), NA_real_,
        ifelse(v >= r, as.numeric(v - r) + 1, as.numeric(v - r)))
    }
  }
  dataset
}

# ─── Generic merge helpers ────────────────────────────────────────────────────
derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL,
                                order = NULL, filter_add = NULL, mode = NULL, ...) {
  by_nm <- sapply(by_vars, deparse)
  fe <- substitute(filter_add)
  if (!is.null(fe) && deparse(fe) != "NULL")
    dataset_add <- dataset_add |> filter(!!fe)
  if (!is.null(order)) {
    ord_nms <- sapply(order, function(x) {
      d <- deparse(x); gsub("desc\\(|\\)", "", d)
    })
    desc_flags <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
    for (i in rev(seq_along(ord_nms))) {
      if (ord_nms[i] %in% names(dataset_add)) {
        dataset_add <- if (desc_flags[i])
          dataset_add |> arrange(desc(.data[[ord_nms[i]]]))
        else dataset_add |> arrange(.data[[ord_nms[i]]])
      }
    }
  }
  if (!is.null(mode)) {
    fn <- if (mode == "first") dplyr::slice_head else dplyr::slice_tail
    dataset_add <- dataset_add |>
      group_by(across(all_of(by_nm))) |> fn(n = 1) |> ungroup()
  }
  if (!is.null(new_vars)) {
    src_names <- sapply(new_vars, deparse)
    out_names <- names(new_vars)
    if (is.null(out_names)) out_names <- rep("", length(src_names))
    out_names <- ifelse(out_names == "" | is.na(out_names),
                        src_names, out_names)
    # rename map: c(NEW = OLD)
    rn_map <- setNames(src_names, out_names)
    keep <- intersect(c(by_nm, src_names), names(dataset_add))
    dataset_add <- dataset_add |> select(all_of(keep))
    # apply renames where new differs from old
    for (i in seq_along(src_names)) {
      if (out_names[i] != src_names[i] && src_names[i] %in% names(dataset_add)) {
        names(dataset_add)[names(dataset_add) == src_names[i]] <- out_names[i]
      }
    }
    new_cols <- out_names
  } else new_cols <- setdiff(names(dataset_add), by_nm)
  left_join(dataset,
            dataset_add |> select(any_of(c(by_nm, new_cols))),
            by = by_nm)
}

derive_vars_merged_lookup <- function(dataset, dataset_add, by_vars,
                                       new_vars = NULL, ...) {
  by_nm <- sapply(by_vars, deparse)
  if (!is.null(new_vars)) {
    nms <- sapply(new_vars, deparse)
    keep <- intersect(c(by_nm, nms), names(dataset_add))
    dataset_add <- dataset_add |> select(all_of(keep))
  }
  left_join(dataset, dataset_add, by = by_nm)
}

# ─── Flag helpers ─────────────────────────────────────────────────────────────
derive_var_extreme_flag <- function(dataset, by_vars, order, new_var,
                                     mode = "first") {
  .e <- substitute(new_var); new_nm <- if (is.character(.e)) .e else deparse(.e)
  by_nm <- sapply(by_vars, deparse)
  ord_nms <- sapply(order, function(x) {
    d <- deparse(x); gsub("desc\\(|\\)", "", d)
  })
  desc_flags <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
  ord_df <- dataset
  for (i in rev(seq_along(ord_nms))) {
    if (ord_nms[i] %in% names(ord_df)) {
      ord_df <- if (desc_flags[i])
        arrange(ord_df, desc(.data[[ord_nms[i]]]))
      else arrange(ord_df, .data[[ord_nms[i]]])
    }
  }
  by_present <- intersect(by_nm, names(ord_df))
  if (mode == "first") {
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(1) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)], new_nm)))
  } else {
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(n()) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)], new_nm)))
  }
  if (new_nm %in% names(dataset)) dataset[[new_nm]] <- NULL
  left_join(dataset, flagged, by = intersect(names(flagged), names(dataset)))
}

restrict_derivation <- function(dataset, derivation, args = list(),
                                 filter = NULL) {
  fe <- substitute(filter)
  in_rows <- if (!is.null(fe)) {
    tryCatch(which(eval(fe, dataset, parent.frame())),
             error = function(e) seq_len(nrow(dataset)))
  } else seq_len(nrow(dataset))
  out_rows <- setdiff(seq_len(nrow(dataset)), in_rows)
  sub_data <- dataset[in_rows, , drop = FALSE]
  all_args <- c(list(dataset = sub_data), args)
  derived  <- tryCatch(do.call(derivation, all_args),
                       error = function(e) sub_data)
  new_cols <- setdiff(names(derived), names(dataset))
  for (nc in new_cols) dataset[[nc]] <- NA
  dataset[in_rows, names(derived)] <- derived[seq_len(length(in_rows)),
                                              names(derived)]
  dataset
}

derivation_slice <- function(filter = NULL, args = list()) {
  list(filter = substitute(filter), args = args)
}

slice_derivation <- function(dataset, derivation, args = list(), ...) {
  slices <- list(...)
  result <- dataset
  covered <- integer(0)
  for (sl in slices) {
    fe <- sl$filter
    idx <- if (!is.null(fe)) {
      tryCatch(which(eval(fe, result, parent.frame())),
               error = function(e) integer(0))
    } else seq_len(nrow(result))
    idx <- setdiff(idx, covered)
    if (length(idx) == 0) next
    merged_args <- modifyList(args, sl$args)
    sub_data <- result[idx, , drop = FALSE]
    all_args <- c(list(dataset = sub_data), merged_args)
    derived  <- tryCatch(do.call(derivation, all_args),
                         error = function(e) sub_data)
    new_cols <- setdiff(names(derived), names(result))
    for (nc in new_cols) result[[nc]] <- NA
    result[idx, names(derived)] <- derived[seq_len(length(idx)),
                                            names(derived)]
    covered <- c(covered, idx)
  }
  result
}

# ─── ANRIND, ONTRTFL ──────────────────────────────────────────────────────────
derive_var_anrind <- function(dataset) {
  dataset |> mutate(ANRIND = case_when(
    is.na(AVAL) ~ NA_character_,
    AVAL < ANRLO ~ "LOW",
    AVAL > ANRHI ~ "HIGH",
    TRUE ~ "NORMAL"
  ))
}

derive_var_ontrtfl <- function(dataset, start_date, ref_start_date,
                                ref_end_date, ref_end_window = 0, ...) {
  sd_nm <- deparse(substitute(start_date))
  rs_nm <- deparse(substitute(ref_start_date))
  re_nm <- deparse(substitute(ref_end_date))
  dataset |> mutate(ONTRTFL = case_when(
    is.na(.data[[sd_nm]]) | is.na(.data[[rs_nm]]) ~ NA_character_,
    .data[[sd_nm]] >= .data[[rs_nm]] &
      .data[[sd_nm]] <= (.data[[re_nm]] + ref_end_window) ~ "Y",
    TRUE ~ NA_character_
  ))
}

# ─── Baseline & change helpers ────────────────────────────────────────────────
derive_var_base <- function(dataset, by_vars, source_var, new_var) {
  src_nm <- deparse(substitute(source_var))
  new_nm <- deparse(substitute(new_var))
  by_nm  <- sapply(by_vars, deparse)
  by_present <- intersect(by_nm, names(dataset))
  baseline <- dataset |>
    filter(!is.na(ABLFL) & ABLFL == "Y") |>
    select(all_of(c(by_present, src_nm))) |>
    rename(!!new_nm := !!src_nm)
  if (new_nm %in% names(dataset)) dataset[[new_nm]] <- NULL
  left_join(dataset, baseline, by = by_present)
}

derive_var_chg  <- function(dataset) dataset |>
  mutate(CHG = AVAL - BASE)

derive_var_pchg <- function(dataset) dataset |>
  mutate(PCHG = if_else(BASE != 0 & !is.na(BASE),
                        (AVAL - BASE) / BASE * 100, NA_real_))

derive_var_shift <- function(dataset, new_var, from_var, to_var) {
  fn <- deparse(substitute(from_var))
  tn <- deparse(substitute(to_var))
  nn <- deparse(substitute(new_var))
  dataset |> mutate(!!nn := if_else(
    is.na(.data[[fn]]) | is.na(.data[[tn]]),
    NA_character_,
    paste0(.data[[fn]], " to ", .data[[tn]])
  ))
}

# ─── Derived parameter helpers ────────────────────────────────────────────────
derive_param_bmi <- function(dataset, by_vars, weight_code = "WEIGHT",
                              height_code = "HEIGHT",
                              constant_by_vars = NULL,
                              set_values_to = list(), ...) {
  by_nm <- sapply(by_vars, deparse)
  ws <- dataset |> filter(PARAMCD == weight_code) |>
    select(any_of(c(by_nm, "AVAL"))) |> rename(WT = AVAL)
  hs <- if (!is.null(constant_by_vars)) {
    cb <- sapply(constant_by_vars, deparse)
    dataset |> filter(PARAMCD == height_code) |>
      group_by(across(all_of(cb))) |>
      summarise(HT = first(AVAL[!is.na(AVAL)]), .groups = "drop")
  } else {
    dataset |> filter(PARAMCD == height_code) |>
      select(any_of(c(by_nm, "AVAL"))) |> rename(HT = AVAL)
  }
  join_keys <- if (!is.null(constant_by_vars)) sapply(constant_by_vars, deparse)
              else intersect(by_nm, names(hs))
  bmi_rows <- ws |> left_join(hs, by = join_keys) |>
    mutate(AVAL = round(WT / (HT / 100)^2, 1)) |>
    select(-WT, -HT)
  set_vals <- lapply(set_values_to, function(x) {
    if (is.character(x)) x else if (is.numeric(x)) x else eval(x)
  })
  for (nm in names(set_vals)) bmi_rows[[nm]] <- set_vals[[nm]]
  bind_rows(dataset, bmi_rows)
}

derive_param_map <- function(dataset, by_vars, set_values_to = list(), ...) {
  by_nm <- sapply(by_vars, deparse)
  sbp <- dataset |> filter(PARAMCD == "SYSBP") |>
    select(any_of(c(by_nm, "AVAL"))) |> rename(SBP = AVAL)
  dbp <- dataset |> filter(PARAMCD == "DIABP") |>
    select(any_of(c(by_nm, "AVAL"))) |> rename(DBP = AVAL)
  map_rows <- sbp |> left_join(dbp, by = intersect(by_nm, names(dbp))) |>
    mutate(AVAL = round((SBP + 2 * DBP) / 3, 1)) |>
    select(-SBP, -DBP)
  set_vals <- lapply(set_values_to, function(x) {
    if (is.character(x)) x else if (is.numeric(x)) x else eval(x)
  })
  for (nm in names(set_vals)) map_rows[[nm]] <- set_vals[[nm]]
  bind_rows(dataset, map_rows)
}

# ─── Sample data: VS, ADSL, PARAMCD lookup ───────────────────────────────────
vs <- tribble(
  ~STUDYID,       ~USUBJID, ~VSTESTCD, ~VSTEST,         ~VSSTRESN, ~VSSTRESC,
  ~VSDTC,         ~VISIT,    ~VSORNRLO, ~VSORNRHI,
  "CDISCPILOT01", "01-001", "SYSBP",  "Sys BP",           135, "135",
  "2024-01-10",   "SCREENING", 90, 140,
  "CDISCPILOT01", "01-001", "DIABP",  "Dia BP",            85,  "85",
  "2024-01-10",   "SCREENING", 60,  90,
  "CDISCPILOT01", "01-001", "SYSBP",  "Sys BP",           150, "150",
  "2024-02-08",   "WEEK 4",    90, 140,
  "CDISCPILOT01", "01-001", "DIABP",  "Dia BP",            95,  "95",
  "2024-02-08",   "WEEK 4",    60,  90,
  "CDISCPILOT01", "01-001", "WEIGHT", "Weight",            68,  "68",
  "2024-01-10",   "SCREENING", NA, NA,
  "CDISCPILOT01", "01-001", "HEIGHT", "Height",           165, "165",
  "2024-01-10",   "SCREENING", NA, NA,
  "CDISCPILOT01", "01-002", "SYSBP",  "Sys BP",           120, "120",
  "2024-01-12",   "SCREENING", 90, 140,
  "CDISCPILOT01", "01-002", "DIABP",  "Dia BP",            75,  "75",
  "2024-01-12",   "SCREENING", 60,  90,
  "CDISCPILOT01", "01-002", "SYSBP",  "Sys BP",           118, "118",
  "2024-02-12",   "WEEK 4",    90, 140,
  "CDISCPILOT01", "01-002", "DIABP",  "Dia BP",            72,  "72",
  "2024-02-12",   "WEEK 4",    60,  90,
  "CDISCPILOT01", "01-002", "WEIGHT", "Weight",            82,  "82",
  "2024-01-12",   "SCREENING", NA, NA,
  "CDISCPILOT01", "01-002", "HEIGHT", "Height",           178, "178",
  "2024-01-12",   "SCREENING", NA, NA
) |> mutate(VSDTC = as.character(VSDTC))

adsl_bds <- tribble(
  ~STUDYID,       ~USUBJID, ~TRTSDT, ~TRTEDT, ~TRT01A,                ~TRT01P,                ~SEX, ~AGE, ~SAFFL,
  "CDISCPILOT01", "01-001", 19735,   19783,   "Xanomeline High Dose", "Xanomeline High Dose", "F",  52,   "Y",
  "CDISCPILOT01", "01-002", 19737,   19783,   "Placebo",              "Placebo",              "M",  61,   "Y"
)

param_lookup <- tribble(
  ~VSTESTCD, ~PARAMCD, ~PARAM,                              ~PARAMN, ~PARCAT1,
  "SYSBP",   "SYSBP",  "Systolic Blood Pressure (mmHg)",    1,       "Hemodynamic",
  "DIABP",   "DIABP",  "Diastolic Blood Pressure (mmHg)",   2,       "Hemodynamic",
  "PULSE",   "PULSE",  "Pulse Rate (beats/min)",            3,       "Hemodynamic",
  "WEIGHT",  "WEIGHT", "Weight (kg)",                       10,      "Anthropometric",
  "HEIGHT",  "HEIGHT", "Height (cm)",                       11,      "Anthropometric"
)

# Real pharmaverseadam fallback
if (requireNamespace("pharmaverseadam", quietly = TRUE)) {
  advs_real <- pharmaverseadam::advs
  adlb_real <- pharmaverseadam::adlb |> filter(is.na(DTYPE))
  adsl_real <- pharmaverseadam::adsl
} else {
  advs_real <- vs |>
    left_join(select(adsl_bds, USUBJID, TRT01A, TRTSDT), by = "USUBJID")
  adlb_real <- vs |>
    left_join(select(adsl_bds, USUBJID, TRT01A, TRTSDT), by = "USUBJID")
  adsl_real <- adsl_bds
}
```

## Welcome

Lesson 16 — **`{admiral}` Part 3: BDS Datasets**.

You will build a complete **ADVS** (vital signs) step by step. The same patterns apply to **ADLB** (labs) and **ADEG** (ECGs). Together these are the **BDS — Basic Data Structure** ADaMs: long-format, one row per (subject × parameter × visit × timepoint).

What you learn here:

- Why BDS is **long** (and what SAS programmers must unlearn from `PROC TRANSPOSE`).
- The canonical **13-step ADVS pipeline**.
- Baseline (`ABLFL`, `BASE`, `BNRIND`) and change-from-baseline (`CHG`, `PCHG`, `SHIFT1`).
- Conditional derivation: `restrict_derivation()` vs `slice_derivation()`.
- ADLB toxicity grades (NCI CTCAE) and ADEG triplicate ECG timepoints.
- Three Pinnacle 21 (P21) diagnostic bugs.

---

## 1 — The BDS shape

### 1.1 The biggest mental shift from SAS

In SAS, vital signs land in `VS` as one row per measurement and you transpose to wide for analysis (`PROC TRANSPOSE`). In ADaM **BDS**, you stay long forever — analysis tools read long; layout flips to wide.

| Direction               | SAS habit                                         | admiral / BDS                           |
|-------------------------|---------------------------------------------------|-----------------------------------------|
| Source → analysis       | `PROC TRANSPOSE` to wide                          | Stay long; one row per (subj × param × visit × timepoint) |
| Filter one parameter    | `KEEP _NAME_ = 'SYSBP'` after transpose           | `filter(PARAMCD == "SYSBP")`             |
| Analysis model input    | Re-transpose to long again                        | Already long                             |
| Display table output    | Re-transpose to wide                              | Layout engine (Lesson 32+) flips it      |

> **SAS mental model:** BDS = "what `PROC SQL` would emit if you `UNION ALLed` `PROC MEANS NWAY OUT=` for every parameter." Long stays. Width happens only at the display layer.

### 1.2 The BDS primary key

Every BDS dataset has the same key:

| Variable    | Meaning                                       | Example          |
|-------------|-----------------------------------------------|------------------|
| `USUBJID`   | Unique subject identifier                     | `01-701-1015`    |
| `PARAMCD`   | Parameter code                                | `SYSBP`          |
| `AVISIT`    | Analysis visit                                | `WEEK 4`         |
| `ATPT`      | Analysis timepoint within visit (optional)    | `2 HR POST DOSE` |
| `ADT`       | Analysis date                                 | `2024-02-05`     |
| `DTYPE`     | Derivation type (`NA` for source, `LOCF` etc.)| `NA`             |

### 1.3 PARAM, PARAMCD, PARAMN — the trio

| Variable   | Type      | Role                                    | Example                            |
|------------|-----------|-----------------------------------------|------------------------------------|
| `PARAMCD`  | character | Short code — used for filtering         | `SYSBP`                            |
| `PARAM`    | character | Full label with units — used in titles  | `Systolic Blood Pressure (mmHg)`   |
| `PARAMN`   | numeric   | Sort order — used for output ordering   | `1`                                |

Three columns hold the same conceptual information at three abstraction levels. The PROC FORMAT analog: a control dataset has `START`, `LABEL`, and `SEXCL` columns. PARAM/PARAMCD/PARAMN play the same role for BDS rows.

### 1.4 AVAL versus AVALC

| Variable | Purpose                                  | Used for                               |
|----------|------------------------------------------|----------------------------------------|
| `AVAL`   | Numeric analysis value                   | Means, change from baseline, plots     |
| `AVALC`  | Character analysis value                 | Categorical results, text              |

For SBP, `AVAL = 135`, `AVALC = NA`. For a categorical pain assessment, `AVAL = NA`, `AVALC = "Moderate"`. Most BDS rows use only one of the two.

### 1.5 DTYPE — source vs derived

| `DTYPE`    | Meaning                                                       |
|------------|---------------------------------------------------------------|
| `NA`       | Source observation (raw collected data)                       |
| `LOCF`     | Last observation carried forward (imputed)                    |
| `WOCF`     | Worst observation carried forward                             |
| `MAXIMUM`  | Maximum value summary row                                     |
| `MINIMUM`  | Minimum value summary row                                     |
| `AVERAGE`  | Mean summary row                                              |

> **Critical:** when computing `lag()`, `lead()`, or worsening flags, **always** `filter(is.na(DTYPE))` first. Otherwise admiral-derived summary rows pollute time-series logic.

### 1.6 PROC TRANSPOSE vs derive_param

```
   /* SAS: wide PROC TRANSPOSE */         |   /* admiral: long stays long */
   proc transpose data=vs                 |   advs <- vs |>
     out=vs_wide                          |     derive_vars_merged_lookup(
     prefix=val_;                         |       dataset_add = param_lookup,
     by USUBJID VISIT;                    |       by_vars = exprs(VSTESTCD),
     id VSTESTCD;                         |       new_vars = exprs(PARAMCD,
     var VSSTRESN;                        |                        PARAM, PARAMN)) |>
   run;                                   |     mutate(AVAL = VSSTRESN,
                                          |            AVISIT = VISIT)
   /* now wide: val_SYSBP, val_DIABP ... */|
```

### 1.7 Demo — see the BDS shape

```{r bds-shape-demo}
vs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD, PARAM, PARAMN)
  ) |>
  mutate(AVAL = VSSTRESN, AVISIT = VISIT) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL) |>
  head(6)
```

> **Read output:** one row per (subject × parameter × visit). Same shape no matter how many parameters get added — pure long.

### 1.8 Exercise — count rows per subject × parameter

```{r bds-shape-ex, exercise=TRUE, exercise.lines=10}
vs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD)
  ) |>
  group_by(USUBJID, ___) |>
  summarise(n_rows = n(), .groups = "drop")
```

```{r bds-shape-ex-hint-1}
# Group by PARAMCD
```

```{r bds-shape-ex-solution}
vs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD)
  ) |>
  group_by(USUBJID, PARAMCD) |>
  summarise(n_rows = n(), .groups = "drop")
```

### 1.9 Knowledge check — BDS shape

```{r quiz-bds-shape, echo=FALSE}
quiz(
  question(
    "BDS keeps data long. SAS PROC TRANSPOSE puts it wide. Which wins for downstream analysis modelling?",
    answer("Wide — `PROC MIXED` reads wide"),
    answer("Long — `PROC MIXED` and `nlme::lme` both need one row per observation", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "`PARAMCD` vs `PARAM` vs `PARAMN` — why three?",
    answer("CDISC redundancy"),
    answer("Filter (PARAMCD), display label (PARAM), sort order (PARAMN) — three abstractions", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "`DTYPE = \"LOCF\"` means?",
    answer("Source raw observation"),
    answer("Last observation carried forward — admiral-derived summary row, not real measurement", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 2 — Step 1: Merge ADSL onto BDS

### 2.1 Why every BDS row needs ADSL columns

ADSL holds one row per subject. BDS holds many. Downstream derivations need treatment dates, arm, and demographics on every row.

| Bring across                | Purpose                                       |
|-----------------------------|-----------------------------------------------|
| `TRTSDT`, `TRTEDT`          | Date windowing (`ONTRTFL`, `ABLFL`)            |
| `TRT01A`, `TRT01P`          | Stratification in summaries                    |
| `AGE`, `SEX`                | Demographic subgroup analyses                  |

### 2.2 SAS-analog comparison

| admiral concept                            | SAS counterpart                                              |
|--------------------------------------------|--------------------------------------------------------------|
| `derive_vars_merged(by_vars = USUBJID)`    | `MERGE adsl(in=a) vs(in=b); BY USUBJID; if a;`               |
| `new_vars = exprs(TRTSDT, TRTEDT, TRT01A)` | `KEEP=` in MERGE statement                                   |
| One-to-many cardinality                    | Same; SAS DATA step happily duplicates ADSL columns          |

### 2.3 ASCII side-by-side

```
   data advs;                              |   advs <- vs |>
     merge vs(in=b)                        |     derive_vars_merged(
           adsl(in=a                       |       dataset_add = adsl_bds,
                keep=USUBJID TRT01A        |       by_vars  = exprs(STUDYID,
                     TRTSDT TRTEDT);       |                         USUBJID),
     by USUBJID;                           |       new_vars = exprs(TRTSDT,
     if b;                                 |                         TRTEDT,
   run;                                    |                         TRT01A))
```

### 2.4 Anatomy of the merge

```r
vs |>
  derive_vars_merged(
    dataset_add = adsl_bds,                     # source: one row per subject
    by_vars     = exprs(STUDYID, USUBJID),      # join key
    new_vars    = exprs(TRTSDT, TRTEDT, TRT01A) # columns to promote
  )
```

- `dataset_add` must be unique by `by_vars` (admiral validates). No `mode` needed.
- `new_vars` is explicit — without it ALL non-key ADSL columns come across (usually too many).

### 2.5 Demo

```{r merge-adsl-demo}
advs <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(TRTSDT, TRTEDT, TRT01A)
  )

advs |> select(USUBJID, VSTESTCD, VISIT, TRT01A, TRTSDT) |> head(6)
```

> **Read output:** TRT01A and TRTSDT appear on every VS row. Subject `01-001` has TRT01A on all rows including SCREENING and WEEK 4.

### 2.6 Exercise — promote SEX and AGE

```{r merge-adsl-ex, exercise=TRUE, exercise.lines=12}
advs_demo <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(___, ___)
  )

advs_demo |> select(USUBJID, VSTESTCD, SEX, AGE) |> head(4)
```

```{r merge-adsl-ex-hint-1}
# new_vars = exprs(SEX, AGE)
```

```{r merge-adsl-ex-solution}
advs_demo <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(SEX, AGE)
  )

advs_demo |> select(USUBJID, VSTESTCD, SEX, AGE) |> head(4)
```

### 2.7 Knowledge check

```{r quiz-merge-adsl, echo=FALSE}
quiz(
  question(
    "SAS equivalent of `derive_vars_merged(by_vars = exprs(USUBJID))`?",
    answer("`PROC TRANSPOSE`"),
    answer("`MERGE adsl vs; BY USUBJID;` with `IF a;` or PROC SQL `LEFT JOIN`", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "Omitting `new_vars`?",
    answer("Errors out"),
    answer("Brings every non-key column of `dataset_add` onto base — usually too many", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 3 — Step 2: Derive ADT and ADY

### 3.1 The DTC → DT → DY chain

SDTM stores dates as ISO 8601 character (`VSDTC = "2024-02-08"`). ADaM analysis needs numeric dates (`ADT`) and study-relative day numbers (`ADY`).

| Variable | Type    | Meaning                                |
|----------|---------|----------------------------------------|
| `VSDTC`  | char    | Source ISO 8601 string                  |
| `ADT`    | numeric | Analysis date (R numeric Date)          |
| `ADY`    | integer | Study day (Day 1 = TRTSDT; no Day 0)    |

### 3.2 SAS-analog comparison

| admiral                              | SAS counterpart                                              |
|--------------------------------------|--------------------------------------------------------------|
| `derive_vars_dt(dtc = VSDTC, new_vars_prefix = "A")` | `ADT = input(VSDTC, yymmdd10.);`             |
| `derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT))` | `IF ADT >= TRTSDT THEN ADY = ADT - TRTSDT + 1; ELSE ADY = ADT - TRTSDT;` |

> **No Day 0:** ref date is Day 1. Day before is Day -1. Day 0 does not exist in CDISC ADaM IG.

### 3.3 ASCII side-by-side

```
   data advs;                              |   advs <- advs |>
     set advs;                             |     derive_vars_dt(
     ADT = input(VSDTC, yymmdd10.);        |       dtc = VSDTC,
     if ADT >= TRTSDT                      |       new_vars_prefix = "A") |>
       then ADY = ADT - TRTSDT + 1;        |     derive_vars_dy(
       else ADY = ADT - TRTSDT;            |       reference_date = TRTSDT,
   run;                                    |       source_vars = exprs(ADT))
```

### 3.4 Demo

```{r derive-adt-demo}
advs <- advs |>
  derive_vars_dt(dtc = VSDTC, new_vars_prefix = "A") |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT))

advs |>
  mutate(ADT_human = as.Date(ADT, origin = "1970-01-01")) |>
  select(USUBJID, VISIT, ADT_human, ADY) |>
  head(6)
```

> **Read output:** SCREENING rows precede TRTSDT → ADY = -2 (two days before Day 1). WEEK 4 rows fall after TRTSDT → positive ADY. No Day 0.

### 3.5 Exercise — verify no Day 0

```{r derive-adt-ex, exercise=TRUE, exercise.lines=10}
advs |>
  summarise(
    has_day_0    = any(ADY == ___, na.rm = TRUE),
    min_negative = min(ADY[ADY < 0], na.rm = TRUE),
    max_positive = max(ADY[ADY > 0], na.rm = TRUE)
  )
```

```{r derive-adt-ex-hint-1}
# has_day_0 = any(ADY == 0)
```

```{r derive-adt-ex-solution}
advs |>
  summarise(
    has_day_0    = any(ADY == 0, na.rm = TRUE),
    min_negative = min(ADY[ADY < 0], na.rm = TRUE),
    max_positive = max(ADY[ADY > 0], na.rm = TRUE)
  )
```

### 3.6 Knowledge check

```{r quiz-derive-adt, echo=FALSE}
quiz(
  question(
    "Subject treated on 2024-01-15. AE on 2024-01-15. ADY = ?",
    answer("0"),
    answer("1 — reference date is Day 1; no Day 0", correct = TRUE),
    answer("-1"),
    random_answer_order = TRUE
  ),
  question(
    "SAS analog of `derive_vars_dy()`?",
    answer("`PROC TRANSPOSE`"),
    answer("DATA step with `IF ADT >= TRTSDT THEN ADY = ADT - TRTSDT + 1; ELSE ADY = ADT - TRTSDT;`", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 4 — Step 3: Assign PARAMCD via lookup

### 4.1 The lookup mental model — PROC FORMAT in tibble form

SAS uses `PROC FORMAT CNTLIN=` to build category lookups. admiral uses a tibble as the lookup, joined with `derive_vars_merged_lookup()`.

| SAS `PROC FORMAT`                     | admiral `derive_vars_merged_lookup`                         |
|---------------------------------------|-------------------------------------------------------------|
| `PROC FORMAT CNTLIN=lookup_ds;`        | `param_lookup <- tribble(...)`                               |
| `FORMAT VSTESTCD $param.;`             | `derive_vars_merged_lookup(dataset_add = param_lookup, ...)`|
| `START` → `LABEL`                      | `VSTESTCD` → `PARAMCD` / `PARAM` / `PARAMN`                  |

### 4.2 ASCII side-by-side

```
   data param_lookup;                      |   param_lookup <- tribble(
     length VSTESTCD $8 PARAMCD $8         |     ~VSTESTCD, ~PARAMCD, ~PARAM,
            PARAM $50;                     |     "SYSBP",   "SYSBP",  "Sys BP",
     input VSTESTCD PARAMCD                |     "DIABP",   "DIABP",  "Dia BP",
           PARAM $; datalines;             |     ...)
     SYSBP  SYSBP  Systolic BP             |
     ;                                     |   advs |>
   run;                                    |     derive_vars_merged_lookup(
                                           |       dataset_add = param_lookup,
   proc format cntlin=param_lookup;        |       by_vars  = exprs(VSTESTCD),
     value $vstparam ... ;                 |       new_vars =
   run;                                    |         exprs(PARAMCD, PARAM, PARAMN))
```

### 4.3 Anatomy

```r
advs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,                        # the lookup tibble
    by_vars     = exprs(VSTESTCD),                     # join key
    new_vars    = exprs(PARAMCD, PARAM, PARAMN)        # columns to bring across
  )
```

- The lookup must be one row per `by_vars`. Duplicates → join blows up cardinality.
- Unmatched VSTESTCD → `NA` PARAMCD silently. **Common bug.** Always audit `filter(is.na(PARAMCD))` after the join.

### 4.4 Demo

```{r paramcd-demo}
advs <- advs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD, PARAM, PARAMN)
  ) |>
  mutate(
    AVAL    = VSSTRESN,
    AVALC   = VSSTRESC,
    AVISIT  = VISIT,
    AVISITN = case_when(VISIT == "SCREENING" ~ 0,
                        VISIT == "WEEK 4"    ~ 4,
                        TRUE                 ~ NA_real_),
    ANRLO   = VSORNRLO,
    ANRHI   = VSORNRHI
  )

advs |> select(USUBJID, VSTESTCD, PARAMCD, PARAM, PARAMN, AVAL) |> head(6)
```

> **Read output:** PARAMCD now populated from lookup. PARAM holds the display label. PARAMN drives sort order.

### 4.5 Exercise — detect missing lookup entries

```{r paramcd-ex, exercise=TRUE, exercise.lines=10}
# How many rows have PARAMCD = NA after the lookup?
advs |>
  summarise(missing_paramcd = sum(is.na(___)))
```

```{r paramcd-ex-hint-1}
# Check PARAMCD column
```

```{r paramcd-ex-solution}
advs |>
  summarise(missing_paramcd = sum(is.na(PARAMCD)))
```

### 4.6 Knowledge check

```{r quiz-paramcd, echo=FALSE}
quiz(
  question(
    "Closest SAS analog of `derive_vars_merged_lookup()`?",
    answer("`PROC TRANSPOSE`"),
    answer("`PROC FORMAT CNTLIN=` and `FORMAT` statement, or `PROC SQL LEFT JOIN` to lookup", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "VSTESTCD not present in lookup → PARAMCD becomes?",
    answer("Source value `VSTESTCD`"),
    answer("`NA` silently — always audit after join", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 5 — Derived parameters: BMI and MAP

### 5.1 New rows, not new columns

In SAS you would compute BMI as a new column. In admiral BDS, BMI lives as a **new row**: `PARAMCD = "BMI"`, `AVAL = computed value`. The long shape is preserved.

```
   /* SAS: new column */                   |   /* admiral: new row */
   data advs;                              |   advs |>
     set advs;                             |     derive_param_bmi(
     by USUBJID VISIT;                     |       by_vars = exprs(STUDYID,
     if first.VISIT then BMI =             |                        USUBJID,
        weight / (height/100)**2;          |                        AVISIT, AVISITN),
   run;                                    |       weight_code = "WEIGHT",
                                           |       height_code = "HEIGHT",
                                           |       constant_by_vars = exprs(USUBJID),
                                           |       set_values_to = exprs(
                                           |         PARAMCD = "BMI",
                                           |         PARAM = "BMI (kg/m^2)",
                                           |         PARAMN = 20))
```

### 5.2 SAS-analog comparison

| admiral                          | SAS counterpart                                                          |
|----------------------------------|--------------------------------------------------------------------------|
| `derive_param_bmi()`             | DATA step computing BMI per visit + `OUTPUT;` for new row                |
| `constant_by_vars = exprs(USUBJID)` | Retain screening HEIGHT, applied to every visit                       |
| `derive_param_map()`             | `MAP = (SBP + 2*DBP) / 3;` per visit, output as new row                   |
| `set_values_to = exprs(PARAMCD = "BMI")` | Set PARAMCD on new rows; `LENGTH` + assignment in SAS              |

### 5.3 Demo — BMI

```{r bmi-demo}
advs <- advs |>
  derive_param_bmi(
    by_vars          = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    weight_code      = "WEIGHT",
    height_code      = "HEIGHT",
    constant_by_vars = exprs(USUBJID),
    set_values_to    = exprs(PARAMCD = "BMI",
                              PARAM   = "BMI (kg/m^2)",
                              PARAMN  = 20)
  )

advs |> filter(PARAMCD == "BMI") |>
  select(USUBJID, AVISIT, AVAL, PARAMCD, PARAM)
```

> **Read output:** BMI rows present even at WEEK 4 because `constant_by_vars = exprs(USUBJID)` carries screening HEIGHT forward.

### 5.4 Demo — MAP

```{r map-demo}
advs <- advs |>
  derive_param_map(
    by_vars       = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    set_values_to = exprs(PARAMCD = "MAP",
                          PARAM   = "MAP (mmHg)",
                          PARAMN  = 21)
  )

advs |> filter(PARAMCD == "MAP") |>
  select(USUBJID, AVISIT, AVAL, PARAMCD, PARAM)
```

> **Read output:** MAP = (SBP + 2 × DBP) / 3 per visit per subject. New rows appended.

### 5.5 Exercise — count derived BMI rows

```{r bmi-ex, exercise=TRUE, exercise.lines=8}
advs |>
  filter(PARAMCD == ___) |>
  group_by(USUBJID) |>
  summarise(n_bmi_rows = n(), .groups = "drop")
```

```{r bmi-ex-hint-1}
# PARAMCD == "BMI"
```

```{r bmi-ex-solution}
advs |>
  filter(PARAMCD == "BMI") |>
  group_by(USUBJID) |>
  summarise(n_bmi_rows = n(), .groups = "drop")
```

### 5.6 Knowledge check

```{r quiz-derived-params, echo=FALSE}
quiz(
  question(
    "BMI lives as a new ___ in BDS.",
    answer("Column"),
    answer("Row — `PARAMCD = \"BMI\"`, `AVAL = computed value`", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "`constant_by_vars = exprs(USUBJID)` does?",
    answer("Limits BMI to one visit"),
    answer("Carries screening height forward to every visit so BMI computes at WEEK 4 too", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "Filter to PARAMCD == SYSBP before `derive_param_map()`?",
    answer("MAP still works"),
    answer("MAP cannot compute — needs both SYSBP and DIABP rows in input", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 6 — ANRIND and ONTRTFL

### 6.1 What they encode

| Var       | Meaning                            | Values                                |
|-----------|------------------------------------|---------------------------------------|
| `ANRIND`  | Analysis normal range indicator    | `"LOW"` / `"NORMAL"` / `"HIGH"` / `NA`|
| `ONTRTFL` | On-treatment flag                  | `"Y"` / `NA` (never `"N"`)            |

### 6.2 SAS-analog comparison

| admiral                              | SAS counterpart                                                  |
|--------------------------------------|------------------------------------------------------------------|
| `derive_var_anrind()`                | `PROC FORMAT` 3-bin or DATA step `IF AVAL < ANRLO THEN ANRIND="LOW";` |
| `derive_var_ontrtfl(start, ref_start, ref_end)` | `IF TRTSDT <= ADT <= TRTEDT THEN ONTRTFL="Y";`        |
| `ref_end_window = 30`                | `IF TRTSDT <= ADT <= TRTEDT + 30 THEN ONTRTFL = "Y";`            |

### 6.3 ASCII side-by-side

```
   data advs;                              |   advs |>
     set advs;                             |     derive_var_anrind() |>
     if AVAL < ANRLO then ANRIND="LOW";    |     derive_var_ontrtfl(
     else if AVAL > ANRHI then             |       start_date     = ADT,
          ANRIND = "HIGH";                 |       ref_start_date = TRTSDT,
     else if not missing(AVAL) then        |       ref_end_date   = TRTEDT)
          ANRIND = "NORMAL";               |
     if TRTSDT <= ADT <= TRTEDT then       |
          ONTRTFL = "Y";                   |
   run;                                    |
```

### 6.4 Demo

```{r anrind-ontrtfl-demo}
advs <- advs |>
  derive_var_anrind() |>
  derive_var_ontrtfl(
    start_date     = ADT,
    ref_start_date = TRTSDT,
    ref_end_date   = TRTEDT
  )

advs |> filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVISIT, AVAL, ANRLO, ANRHI, ANRIND, ONTRTFL)
```

> **Read output:** ANRIND = LOW/NORMAL/HIGH per row. ONTRTFL = "Y" only inside treatment window; pre-treatment SCREENING rows → NA.

### 6.5 Common pitfalls

1. Forget ANRLO/ANRHI on row → ANRIND = NA.
2. `ref_end_window = TRTDURD` → wrong; window is days BEYOND TRTEDT, not duration.
3. CDISC rule: ONTRTFL never `"N"`.

### 6.6 Exercise — 30-day post-treatment safety window

```{r ontrtfl-ex, exercise=TRUE, exercise.lines=12}
advs_post <- advs |>
  derive_var_ontrtfl(
    start_date     = ADT,
    ref_start_date = TRTSDT,
    ref_end_date   = TRTEDT,
    ref_end_window = ___
  )

sum(advs_post$ONTRTFL == "Y", na.rm = TRUE)
```

```{r ontrtfl-ex-hint-1}
# ref_end_window = 30
```

```{r ontrtfl-ex-solution}
advs_post <- advs |>
  derive_var_ontrtfl(
    start_date     = ADT,
    ref_start_date = TRTSDT,
    ref_end_date   = TRTEDT,
    ref_end_window = 30
  )

sum(advs_post$ONTRTFL == "Y", na.rm = TRUE)
```

### 6.7 Knowledge check

```{r quiz-anrind, echo=FALSE}
quiz(
  question(
    "`derive_var_anrind()` ≈ which SAS?",
    answer("`PROC FREQ`"),
    answer("`PROC FORMAT` 3-bin or DATA-step `IF AVAL < ANRLO THEN \"LOW\"; ELSE IF AVAL > ANRHI THEN \"HIGH\"; ELSE \"NORMAL\";`", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "AVAL is NA — ANRIND?",
    answer("`\"NORMAL\"`"),
    answer("`NA` — cannot classify missing value", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "ONTRTFL=\"N\" allowed?",
    answer("Yes"),
    answer("No — CDISC: `\"Y\"` or `NA` only", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 7 — Baseline: ABLFL, BASE, BNRIND

### 7.1 Definition

Baseline row = **last pre-treatment observation** per (USUBJID × PARAMCD). Its `AVAL` and `ANRIND` are pulled forward to every row (`BASE`, `BNRIND`).

| Var      | Meaning                                              |
|----------|------------------------------------------------------|
| `ABLFL`  | `"Y"` on baseline row, `NA` elsewhere                 |
| `BASE`   | Baseline AVAL carried forward to all subj × param rows|
| `BNRIND` | Baseline ANRIND carried forward                       |

### 7.2 SAS-analog comparison

| admiral                                              | SAS counterpart                                                                  |
|------------------------------------------------------|----------------------------------------------------------------------------------|
| `restrict_derivation(derive_var_extreme_flag, filter = ADT <= TRTSDT)` | `PROC SORT BY USUBJID PARAMCD ADT;` + `IF LAST.PARAMCD` on filtered set |
| `mode = "last"`                                      | `IF LAST.PARAMCD THEN ABLFL = "Y";`                                              |
| `derive_var_base(source_var = AVAL, new_var = BASE)` | `RETAIN BASE; IF ABLFL = "Y" THEN BASE = AVAL;`                                  |

### 7.3 ASCII side-by-side

```
   proc sort data=advs;                    |   advs |>
     by USUBJID PARAMCD ADT;               |     restrict_derivation(
   run;                                    |       derivation =
                                           |         derive_var_extreme_flag,
   data advs;                              |       args = params(
     set advs;                             |         by_vars = exprs(STUDYID,
     by USUBJID PARAMCD;                   |                          USUBJID,
     retain BASE;                          |                          PARAMCD),
     if first.PARAMCD then BASE = .;       |         order   = exprs(ADT,
     if ADT <= TRTSDT and                  |                          AVISITN),
        not missing(AVAL) and              |         new_var = ABLFL,
        last.PARAMCD                       |         mode    = "last"),
       then do;                            |       filter =
         ABLFL = "Y"; BASE = AVAL;         |         !is.na(AVAL) & ADT <= TRTSDT)
       end;                                |
   run;                                    |   advs |>
                                           |     derive_var_base(
                                           |       by_vars = exprs(STUDYID,
                                           |                        USUBJID,
                                           |                        PARAMCD),
                                           |       source_var = AVAL,
                                           |       new_var    = BASE)
```

### 7.4 Demo

```{r baseline-demo}
advs <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  ) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD),
                  source_var = AVAL,   new_var = BASE) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD),
                  source_var = ANRIND, new_var = BNRIND)

advs |> filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVISIT, AVAL, ABLFL, BASE, ANRIND, BNRIND)
```

> **Read output:** one `ABLFL = "Y"` per subject per PARAMCD. BASE column shows same value across all rows of that group. BNRIND mirrors BASE for ANRIND.

### 7.5 Common pitfalls

1. `mode = "first"` → flags earliest screening, not last pre-treatment. Always `"last"`.
2. Forget `filter`: flags last record overall (often post-treatment).
3. Forget `PARAMCD` from `by_vars`: BASE of SYSBP overwrites BASE of DIABP.
4. Derive BASE before ABLFL → silently NA on all rows. Order: ABLFL → BASE → BNRIND.

### 7.6 Exercise — flag baseline for DIABP only

```{r ablfl-ex, exercise=TRUE, exercise.lines=14}
advs_dia <- advs |>
  filter(PARAMCD == "DIABP") |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT),
      new_var = ABLFL,
      mode    = ___
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  )

sum(advs_dia$ABLFL == "Y", na.rm = TRUE)
```

```{r ablfl-ex-hint-1}
# mode = "last"
```

```{r ablfl-ex-solution}
advs_dia <- advs |>
  filter(PARAMCD == "DIABP") |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  )

sum(advs_dia$ABLFL == "Y", na.rm = TRUE)
```

### 7.7 Knowledge check

```{r quiz-baseline, echo=FALSE}
quiz(
  question(
    "ABLFL value on non-baseline rows?",
    answer("`\"N\"`"),
    answer("`NA` — CDISC IG forbids `\"N\"`", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "BASE before ABLFL — what happens?",
    answer("Same — admiral reorders"),
    answer("BASE = NA on every row — `derive_var_base` reads ABLFL which doesn't exist yet", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "Missing `PARAMCD` from `by_vars` of `derive_var_base()`?",
    answer("Works fine"),
    answer("BASE for one param overwrites BASE for another — wrong baseline", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 8 — CHG, PCHG, and SHIFT1

### 8.1 What they compute

| Var      | Formula                                | Use                                                |
|----------|----------------------------------------|----------------------------------------------------|
| `CHG`    | `AVAL − BASE`                          | Signed change from baseline                        |
| `PCHG`   | `(AVAL − BASE) / BASE × 100`           | Percent change from baseline                       |
| `SHIFT1` | `paste0(BNRIND, " to ", ANRIND)`       | Shift cell: "NORMAL to HIGH" etc.                  |

### 8.2 SAS-analog comparison

| admiral                  | SAS counterpart                                                  |
|--------------------------|------------------------------------------------------------------|
| `derive_var_chg()`       | `CHG = AVAL - BASE;` in DATA step                                |
| `derive_var_pchg()`      | `IF BASE > 0 THEN PCHG = (AVAL - BASE) / BASE * 100;`            |
| `derive_var_shift(...)`  | `SHIFT1 = catx(" to ", BNRIND, ANRIND);`                          |

### 8.3 ASCII side-by-side

```
   data advs;                              |   advs |>
     set advs;                             |     derive_var_chg() |>
     CHG = AVAL - BASE;                    |     derive_var_pchg() |>
     if BASE > 0                           |     derive_var_shift(
       then PCHG = (AVAL - BASE)           |       new_var  = SHIFT1,
                   / BASE * 100;           |       from_var = BNRIND,
     SHIFT1 = catx(" to ", BNRIND,         |       to_var   = ANRIND)
                             ANRIND);      |
   run;                                    |
```

### 8.4 Demo

```{r chg-shift-demo}
advs <- advs |>
  derive_var_chg() |>
  derive_var_pchg() |>
  derive_var_shift(new_var = SHIFT1,
                   from_var = BNRIND, to_var = ANRIND)

advs |> filter(PARAMCD == "SYSBP" & !is.na(BASE)) |>
  select(USUBJID, AVISIT, AVAL, BASE, CHG, PCHG, BNRIND, ANRIND, SHIFT1)
```

> **Read output:** CHG signed (AVAL − BASE). PCHG percent. SHIFT1 = "BNRIND to ANRIND" string.

### 8.5 Common pitfalls

1. `BASE = 0` → PCHG NA (divide-by-zero guard), not Inf.
2. CHG before BASE derived → all NA. Order matters.
3. Sign-reversal bug in custom DATA step (`BASE - AVAL`). admiral never has this.

### 8.6 Exercise — max abs CHG per subject (DIABP)

```{r chg-ex, exercise=TRUE, exercise.lines=10}
advs |>
  filter(PARAMCD == "DIABP") |>
  group_by(USUBJID) |>
  summarise(max_abs_chg = max(abs(___), na.rm = TRUE))
```

```{r chg-ex-hint-1}
# Use CHG column
```

```{r chg-ex-solution}
advs |>
  filter(PARAMCD == "DIABP") |>
  group_by(USUBJID) |>
  summarise(max_abs_chg = max(abs(CHG), na.rm = TRUE))
```

### 8.7 Knowledge check

```{r quiz-chg, echo=FALSE}
quiz(
  question(
    "CHG formula?",
    answer("`BASE - AVAL`"),
    answer("`AVAL - BASE`", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "BASE = 0 → PCHG = ?",
    answer("`Inf`"),
    answer("`NA` — divide-by-zero guard", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "SHIFT1 separator?",
    answer("`\" -> \"`"),
    answer("`\" to \"` (hardcoded)", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 9 — `restrict_derivation` vs `slice_derivation`

### 9.1 The two patterns

| Wrapper                | Use case                                                                    |
|------------------------|-----------------------------------------------------------------------------|
| `restrict_derivation`  | One derivation, SUBSET of rows. Other rows pass through.                    |
| `slice_derivation`     | DIFFERENT derivation args per row subset. Multiple slices stacked.          |

Decision: same logic smaller scope → `restrict`. Different logic per subset → `slice`.

### 9.2 SAS-analog comparison

| admiral                                        | SAS counterpart                                              |
|------------------------------------------------|--------------------------------------------------------------|
| `restrict_derivation(filter = ONTRTFL == "Y")` | `WHERE ONTRTFL = "Y"` around derivation, MERGE back          |
| `slice_derivation + derivation_slice(filter=A) + derivation_slice(filter=B)` | DATA step `IF A THEN <d1>; ELSE IF B THEN <d2>;`|

### 9.3 Anatomy — `restrict_derivation`

```r
restrict_derivation(
  dataset    = advs,
  derivation = derive_var_extreme_flag,
  args       = params(
    by_vars  = exprs(STUDYID, USUBJID, PARAMCD),
    order    = exprs(desc(AVAL)),
    new_var  = WORSTFL,
    mode     = "first"
  ),
  filter     = ONTRTFL == "Y" & !is.na(AVAL)
)
```

- `filter` evaluated on pre-existing columns only (NOT post-derivation columns).
- Non-matching rows retain original column values (new_var = NA).

### 9.4 Anatomy — `slice_derivation`

```r
slice_derivation(
  dataset    = advs,
  derivation = derive_var_extreme_flag,
  args       = params(                              # shared across slices
    by_vars  = exprs(STUDYID, USUBJID, PARAMCD),
    new_var  = WORSTFL,
    mode     = "first"
  ),
  derivation_slice(
    filter = PARAMCD == "SYSBP" & ONTRTFL == "Y",
    args   = params(order = exprs(desc(AVAL)))
  ),
  derivation_slice(
    filter = PARAMCD == "DIABP" & ONTRTFL == "Y",
    args   = params(order = exprs(desc(AVAL)))
  )
)
```

- Top-level `args` shared. Per-slice `args` override.
- Slices evaluated top-to-bottom; first match wins.

### 9.5 Demo — `restrict_derivation`

```{r restrict-demo}
advs_worst <- advs |>
  filter(PARAMCD == "SYSBP") |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(desc(AVAL)),
      new_var = WORSTFL,
      mode    = "first"
    ),
    filter = ONTRTFL == "Y" & !is.na(AVAL)
  )

advs_worst |> select(USUBJID, AVISIT, AVAL, ONTRTFL, WORSTFL)
```

> **Read output:** WORSTFL = "Y" on highest on-treatment SBP per subject. Pre-treatment rows pass through with WORSTFL = NA.

### 9.6 Exercise — flag latest analysis observation (ANL01FL)

```{r restrict-ex, exercise=TRUE, exercise.lines=14}
advs_anl <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD, AVISIT),
      order   = exprs(desc(ADT)),
      new_var = ANL01FL,
      mode    = ___
    ),
    filter = !is.na(AVAL) & !is.na(AVISIT)
  )

sum(advs_anl$ANL01FL == "Y", na.rm = TRUE)
```

```{r restrict-ex-hint-1}
# mode = "first" (after desc(ADT), first = latest)
```

```{r restrict-ex-solution}
advs_anl <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD, AVISIT),
      order   = exprs(desc(ADT)),
      new_var = ANL01FL,
      mode    = "first"
    ),
    filter = !is.na(AVAL) & !is.na(AVISIT)
  )

sum(advs_anl$ANL01FL == "Y", na.rm = TRUE)
```

### 9.7 Knowledge check

```{r quiz-restrict-slice, echo=FALSE}
quiz(
  question(
    "Key difference between `restrict_derivation()` and `slice_derivation()`?",
    answer("`restrict` is faster"),
    answer("`slice` runs different derivation args per row subset; `restrict` applies one rule to a subset", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "Overlapping slice filters → which wins?",
    answer("Last"),
    answer("First match — slices evaluated top-to-bottom", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 10 — Full ADVS pipeline capstone

### 10.1 The 13-step canonical build

| Step | Function call                              | Adds                                                |
|------|--------------------------------------------|-----------------------------------------------------|
| 1    | `derive_vars_merged()` ← ADSL              | TRTSDT, TRTEDT, TRT01A, TRT01P                      |
| 2    | `derive_vars_dt()`                         | ADT                                                  |
| 3    | `derive_vars_dy()`                         | ADY (no Day 0)                                       |
| 4    | `derive_vars_merged_lookup()` ← param_lookup | PARAMCD, PARAM, PARAMN                             |
| 5    | `mutate()`                                 | AVAL, AVALC, AVISIT, AVISITN, ANRLO, ANRHI          |
| 6    | `derive_param_bmi()` / `derive_param_map()`| New derived parameter rows                          |
| 7    | `derive_var_anrind()`                      | ANRIND                                               |
| 8    | `derive_var_ontrtfl()`                     | ONTRTFL                                              |
| 9    | `restrict_derivation(extreme_flag)`        | ABLFL                                                |
| 10   | `derive_var_base(AVAL → BASE)`             | BASE                                                 |
| 11   | `derive_var_base(ANRIND → BNRIND)`         | BNRIND                                               |
| 12   | `derive_var_chg()`, `derive_var_pchg()`    | CHG, PCHG                                            |
| 13   | `derive_var_shift()`                       | SHIFT1                                               |

### 10.2 Order matters

- PARAMCD before `derive_param_*` (they search by source PARAMCD).
- ADT before ONTRTFL and ABLFL (date comparisons need numeric ADT).
- ABLFL before BASE (BASE reads ABLFL flag).
- BASE before CHG (CHG = AVAL − BASE).
- ANRIND before BNRIND.

### 10.3 Capstone exercise — run the full pipeline

```{r capstone-advs, exercise=TRUE, exercise.lines=60}
advs_full <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)
  ) |>
  derive_vars_dt(dtc = VSDTC, new_vars_prefix = "A") |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT)) |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD, PARAM, PARAMN)
  ) |>
  mutate(
    AVAL    = VSSTRESN,
    AVALC   = VSSTRESC,
    AVISIT  = VISIT,
    AVISITN = case_when(VISIT == "SCREENING" ~ 0,
                        VISIT == "WEEK 4"    ~ 4,
                        TRUE                 ~ NA_real_),
    ANRLO   = VSORNRLO,
    ANRHI   = VSORNRHI
  ) |>
  derive_param_bmi(
    by_vars          = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    weight_code      = "WEIGHT", height_code = "HEIGHT",
    constant_by_vars = exprs(USUBJID),
    set_values_to    = exprs(PARAMCD = "BMI",
                              PARAM   = "BMI (kg/m^2)",
                              PARAMN  = 20)
  ) |>
  derive_param_map(
    by_vars       = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    set_values_to = exprs(PARAMCD = "MAP",
                          PARAM   = "MAP (mmHg)",
                          PARAMN  = 21)
  ) |>
  derive_var_anrind() |>
  derive_var_ontrtfl(
    start_date     = ADT,
    ref_start_date = TRTSDT,
    ref_end_date   = TRTEDT
  ) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  ) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD),
                  source_var = AVAL,   new_var = BASE) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD),
                  source_var = ANRIND, new_var = BNRIND) |>
  derive_var_chg() |>
  derive_var_pchg() |>
  derive_var_shift(new_var = SHIFT1,
                   from_var = BNRIND, to_var = ANRIND)

advs_full |>
  filter(PARAMCD %in% c("SYSBP", "BMI", "MAP")) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, BASE, CHG, ANRIND, SHIFT1) |>
  arrange(USUBJID, PARAMCD, AVISIT)
```

> **Read output:** complete ADVS — every column populated. SYSBP raw + BMI/MAP derived as new rows. BASE/CHG/ANRIND/SHIFT1 all set.

### 10.4 Knowledge check — pipeline order

```{r quiz-pipeline-order, echo=FALSE}
quiz(
  question(
    "BASE derived before ABLFL — result?",
    answer("Same — admiral reorders"),
    answer("BASE = NA on every row", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "PARAMCD assigned AFTER `derive_param_bmi()` — result?",
    answer("BMI rows created normally"),
    answer("`derive_param_bmi` cannot find WEIGHT/HEIGHT source rows by PARAMCD; fails or empty rows", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 11 — ADLB and ADEG differences

### 11.1 Same BDS, different specifics

ADLB and ADEG share BDS contract (one row per subj × param × visit) but add domain specifics.

| Feature              | ADVS              | ADLB                              | ADEG                                  |
|----------------------|-------------------|-----------------------------------|---------------------------------------|
| Source SDTM domain   | VS                | LB                                | EG                                    |
| AVAL source          | `VSSTRESN`        | `LBSTRESN`                        | `EGSTRESN`                            |
| Toxicity grading     | not standard      | `ATOXGR` per NCI CTCAE            | not standard                          |
| Derived params       | BMI, MAP          | Ratios (e.g., AST/ALT), eGFR      | QTcF, QTcB                            |
| Sub-visit timepoint  | rarely            | rarely                            | `ATPT` for triplicate ECGs            |

### 11.2 ATOXGR critical quirk

In `pharmaverseadam::adlb`, `ATOXGR` is stored as **character** (`"3"`, `"-1"`). Direct `>` does alphabetic compare (`"-1" > "2"` is TRUE → WRONG). Convert with `as.integer()` BEFORE comparison.

### 11.3 QTcF correction (ADEG)

`QTcF = QT / RR^(1/3)`. SAS analog: `QTcF = QT / (RR ** (1/3));` + `OUTPUT;` for new row.

### 11.4 Triplicate ECG timepoints

| AVISIT  | ATPT       | ATPTN |
|---------|------------|-------|
| WEEK 4  | 1 MIN PRE  | -1    |
| WEEK 4  | 5 MIN POST | 5     |
| WEEK 4  | 10 MIN POST| 10    |

`AVAL` is the single value; `ATPT` orders sub-visit position.

### 11.5 Knowledge check — ADLB/ADEG

```{r quiz-adlb-adeg, echo=FALSE}
quiz(
  question(
    "ADLB AVAL maps from which SDTM variable?",
    answer("`LBORRES`"),
    answer("`LBSTRESN`", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "ATOXGR stored as character. Comparing `ATOXGR > \"2\"`?",
    answer("Works correctly"),
    answer("Alphabetic compare — `\"-1\" > \"2\"` is TRUE → wrong. Convert with `as.integer()` first.", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "Triplicate ECG sub-visit ordering uses?",
    answer("`AVISITN`"),
    answer("`ATPT` / `ATPTN`", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 12 — P21 derivation diagnostics

Three classic Pinnacle 21 bugs. Read the spec, fix the code.

### 12.1 Diagnostic 1 — ABLFL wrong mode

**Spec:** ABLFL = "Y" on LAST non-missing pre-treatment row per (USUBJID × PARAMCD).

**Broken code:**

```r
restrict_derivation(
  derivation = derive_var_extreme_flag,
  args = params(
    by_vars = exprs(STUDYID, USUBJID, PARAMCD),
    order   = exprs(ADT, AVISITN),
    new_var = ABLFL,
    mode    = "first"            # bug: should be "last"
  ),
  filter = !is.na(AVAL) & ADT <= TRTSDT
)
```

**P21 report:** *"ABLFL flagged earliest screening, not closest-to-dose pre-treatment row."*

```{r p21-bug-ablfl, exercise=TRUE, exercise.lines=14}
advs_fixed <- advs |>
  select(-ABLFL, -BASE, -BNRIND, -CHG, -PCHG, -SHIFT1) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = ___
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  )

advs_fixed |> filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVISIT, ADT, AVAL, ABLFL)
```

```{r p21-bug-ablfl-hint-1}
# mode = "last"
```

```{r p21-bug-ablfl-solution}
advs_fixed <- advs |>
  select(-ABLFL, -BASE, -BNRIND, -CHG, -PCHG, -SHIFT1) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  )

advs_fixed |> filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVISIT, ADT, AVAL, ABLFL)
```

### 12.2 Diagnostic 2 — BASE missing PARAMCD from by_vars

**Spec:** BASE per (USUBJID × PARAMCD). PARAMCD must be in `by_vars`.

**Broken code:**

```r
derive_var_base(
  by_vars    = exprs(STUDYID, USUBJID),    # bug: PARAMCD missing
  source_var = AVAL,
  new_var    = BASE
)
```

**P21 report:** *"BASE value identical across all parameters per subject — cross-parameter contamination."*

```{r p21-bug-base, exercise=TRUE, exercise.lines=12}
advs_basefix <- advs |>
  select(-BASE) |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, ___),
    source_var = AVAL,
    new_var    = BASE
  )

advs_basefix |> filter(USUBJID == "01-001") |>
  select(PARAMCD, AVISIT, AVAL, ABLFL, BASE)
```

```{r p21-bug-base-hint-1}
# Add PARAMCD to by_vars
```

```{r p21-bug-base-solution}
advs_basefix <- advs |>
  select(-BASE) |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var    = BASE
  )

advs_basefix |> filter(USUBJID == "01-001") |>
  select(PARAMCD, AVISIT, AVAL, ABLFL, BASE)
```

### 12.3 Diagnostic 3 — CHG sign reversal

**Spec:** CHG = AVAL − BASE. Positive = increase post-baseline.

**Broken code:**

```r
advs |> mutate(CHG = BASE - AVAL)   # bug: reversed
```

**P21 report:** *"CHG sign opposite of expected; increases reported as decreases."*

```{r p21-bug-chg, exercise=TRUE, exercise.lines=8}
advs_chgfix <- advs |>
  mutate(CHG = ___ - ___)

advs_chgfix |> filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVISIT, AVAL, BASE, CHG)
```

```{r p21-bug-chg-hint-1}
# CHG = AVAL - BASE
```

```{r p21-bug-chg-solution}
advs_chgfix <- advs |>
  mutate(CHG = AVAL - BASE)

advs_chgfix |> filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVISIT, AVAL, BASE, CHG)
```

---

## 13 — Final knowledge check

```{r quiz-final-16, echo=FALSE}
quiz(
  question(
    "BDS shape — one row per?",
    answer("Subject"),
    answer("(Subject × parameter × visit × timepoint)", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "PROC TRANSPOSE vs BDS long-format?",
    answer("PROC TRANSPOSE wins"),
    answer("BDS stays long; layout flips wide only at display", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "Mock SAS `IF AVAL < ANRLO THEN ANRIND=\"LOW\"; ELSE IF AVAL > ANRHI THEN ANRIND=\"HIGH\"; ELSE \"NORMAL\";` ≈ which admiral?",
    answer("`derive_param_*()`"),
    answer("`derive_var_anrind()`", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "`mode = \"last\"` in baseline flagging picks which row?",
    answer("Earliest screening"),
    answer("Closest-to-dose pre-treatment (most recent before TRTSDT)", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "BMI lives as a new ___?",
    answer("Column"),
    answer("Row", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "ATOXGR stored as character — best practice?",
    answer("Compare directly with `>`"),
    answer("Convert with `as.integer()` first", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "QTcF formula?",
    answer("`QT / RR`"),
    answer("`QT / RR^(1/3)`", correct = TRUE),
    random_answer_order = TRUE
  ),
  question(
    "ONTRTFL non-treatment row → ?",
    answer("`\"N\"`"),
    answer("`NA` — CDISC: never `\"N\"`", correct = TRUE),
    random_answer_order = TRUE
  )
)
```

---

## 14 — Lesson 16 Summary {.congratulations}

You now know how to build BDS ADaMs end-to-end:

- **BDS shape** — one row per (subj × param × visit × timepoint). Stay long. SAS PROC TRANSPOSE inversion.
- **Merge ADSL** onto every BDS row via `derive_vars_merged()`. Promote TRTSDT, TRTEDT, TRT01A.
- **ADT/ADY** via `derive_vars_dt()` and `derive_vars_dy()`. No Day 0.
- **PARAMCD via lookup** with `derive_vars_merged_lookup()`. PROC FORMAT in tibble form.
- **Derived parameter rows** with `derive_param_bmi()`, `derive_param_map()`. New rows, not columns.
- **ANRIND** via `derive_var_anrind()` (3-bin classifier). **ONTRTFL** via date-window comparison.
- **Baseline** flag + value: `restrict_derivation(extreme_flag, mode="last", filter=ADT<=TRTSDT)` then `derive_var_base()` twice (AVAL → BASE, ANRIND → BNRIND).
- **Change-from-baseline**: `derive_var_chg()`, `derive_var_pchg()`, `derive_var_shift(BNRIND → ANRIND)`.
- **Conditional derivation**: `restrict_derivation` (single rule, scoped) vs `slice_derivation` (different rules per subset).
- **Full ADVS pipeline** — 13 chained calls replace 8+ SAS DATA steps and 3+ PROC SORTs.
- **ADLB specifics** — ATOXGR character quirk, NCI CTCAE grading.
- **ADEG specifics** — ATPT for triplicate ECG sub-visit, QTcF (Fridericia correction).
- **P21 diagnostics** — wrong ABLFL mode, missing PARAMCD in by_vars, CHG sign reversal.

You also know:

- Order matters: PARAMCD → derived params → ANRIND → ABLFL → BASE → CHG.
- `restrict_derivation` filter sees only pre-existing columns.
- CDISC flag rule: `"Y"` or `NA`, **never** `"N"`.
- ATOXGR conversion to integer before any `>` comparison.

---

## What's next

Lesson 17 — **`{admiral}` Part 4: OCCDS** — covers ADAE, ADCM (occurrence datasets). New shape: one row per recorded event. Patterns: `derive_var_trtemfl` (treatment-emergent), `derive_vars_joined` (correlated last-dose-before-AE lookup), MedDRA SMQ queries.
