---
title: "Lesson 16 — admiral Part 3: BDS Datasets (ADVS, ADLB, ADEG)"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── Mock helpers from earlier lessons ────────────────────────────────────────
exprs <- function(...) {
  dots <- as.list(substitute(list(...)))[-1]
  dots
}

convert_blanks_to_na <- function(dataset) {
  dataset |>
    mutate(across(where(is.character), ~ifelse(. == "", NA_character_, .)))
}

# ── Date utilities ────────────────────────────────────────────────────────────
.impute_dtc <- function(dtc, date_imputation = "first") {
  sapply(dtc, function(d) {
    if (is.na(d)) return(NA_character_)
    if (nchar(d) == 4) {
      suffix <- if (date_imputation == "last") "-12-31" else "-01-01"
      paste0(d, suffix)
    } else if (nchar(d) == 7) {
      if (date_imputation == "last") {
        yr  <- as.integer(substr(d, 1, 4))
        mo  <- as.integer(substr(d, 6, 7))
        nxt <- if (mo == 12) paste0(yr + 1, "-01-01") else paste0(yr, "-", sprintf("%02d", mo + 1), "-01")
        as.character(as.Date(nxt) - 1)
      } else {
        paste0(d, "-01")
      }
    } else {
      d
    }
  })
}

convert_dtc_to_dt <- function(dtc, date_imputation = "first") {
  suppressWarnings(as.numeric(as.Date(.impute_dtc(dtc, date_imputation))))
}

# ── Core admiral mocks ────────────────────────────────────────────────────────
derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL,
                                order = NULL, filter_add = NULL, mode = NULL, ...) {
  by_nm   <- sapply(by_vars, function(x) deparse(x))
  join_df <- dataset_add
  if (!is.null(filter_add)) {
    fe <- substitute(filter_add)
    join_df <- tryCatch(filter(join_df, !!fe), error = function(e) join_df)
  }
  if (!is.null(new_vars)) {
    old_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d, "=")[[1]][2]) else d
    })
    new_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d, "=")[[1]][1]) else d
    })
    keep <- intersect(c(by_nm, old_nms), names(join_df))
    join_df <- join_df |> select(all_of(keep))
    for (i in seq_along(old_nms)) {
      if (old_nms[i] != new_nms[i] && old_nms[i] %in% names(join_df))
        join_df <- join_df |> rename(!!new_nms[i] := !!old_nms[i])
    }
  }
  by_present <- intersect(by_nm, names(join_df))
  left_join(dataset, join_df, by = by_present)
}

derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation = "none", highest_imputation = "n",
                            flag_imputation = "auto", ...) {
  dtc_nm  <- deparse(substitute(dtc))
  imp_rule <- if (date_imputation == "none" && highest_imputation == "n") "none"
              else if (date_imputation == "last") "last" else "first"
  raw <- if (dtc_nm %in% names(dataset)) dataset[[dtc_nm]] else rep(NA_character_, nrow(dataset))
  imputed <- if (imp_rule == "none") raw else .impute_dtc(raw, imp_rule)
  dataset[[paste0(new_vars_prefix, "DT")]] <- suppressWarnings(as.numeric(as.Date(imputed)))
  if (flag_imputation != "none") {
    orig_len <- nchar(raw); imp_len <- nchar(imputed)
    dataset[[paste0(new_vars_prefix, "DTF")]] <- case_when(
      is.na(raw)                        ~ NA_character_,
      orig_len <= 4 & !is.na(imputed)  ~ "Y",
      orig_len <= 7 & !is.na(imputed)  ~ "M",
      TRUE                              ~ NA_character_
    )
  }
  dataset
}

derive_vars_dtm <- function(dataset, dtc, new_vars_prefix,
                             date_imputation = "first", time_imputation = "first",
                             highest_imputation = "n", ...) {
  dtc_nm <- deparse(substitute(dtc))
  raw    <- if (dtc_nm %in% names(dataset)) dataset[[dtc_nm]] else rep(NA_character_, nrow(dataset))
  imputed <- .impute_dtc(raw, date_imputation)
  suffix  <- if (time_imputation == "last") "T23:59:59" else "T00:00:00"
  dataset[[paste0(new_vars_prefix, "DTM")]] <- suppressWarnings(
    as.numeric(as.POSIXct(paste0(imputed, suffix), tz = "UTC"))
  )
  dataset[[paste0(new_vars_prefix, "DTF")]] <- case_when(
    is.na(raw)                              ~ NA_character_,
    nchar(raw) <= 4 & !is.na(imputed)      ~ "Y",
    nchar(raw) <= 7 & !is.na(imputed)      ~ "M",
    TRUE                                    ~ NA_character_
  )
  dataset[[paste0(new_vars_prefix, "TMF")]] <- case_when(
    is.na(raw)                              ~ NA_character_,
    !grepl("T", raw)                        ~ "H",
    TRUE                                    ~ NA_character_
  )
  dataset
}

derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref_nm  <- deparse(substitute(reference_date))
  src_nms <- sapply(source_vars, deparse)
  for (sv in src_nms) {
    dy_nm <- sub("DT$", "DY", sv)
    if (sv %in% names(dataset) && ref_nm %in% names(dataset)) {
      v <- dataset[[sv]]; r <- dataset[[ref_nm]]
      dataset[[dy_nm]] <- ifelse(is.na(v) | is.na(r), NA_real_,
        ifelse(v >= r, as.numeric(v - r) + 1, as.numeric(v - r)))
    }
  }
  dataset
}

derive_var_extreme_flag <- function(dataset, by_vars, order, new_var, mode = "first") {
  new_nm  <- deparse(substitute(new_var))
  by_nm   <- sapply(by_vars, deparse)
  ord_nms <- sapply(order, function(x) {
    d <- deparse(x); gsub("desc\\(|\\)", "", d)
  })
  desc_flags <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
  ord_df <- dataset
  for (i in rev(seq_along(ord_nms))) {
    if (ord_nms[i] %in% names(ord_df)) {
      ord_df <- if (desc_flags[i]) arrange(ord_df, desc(.data[[ord_nms[i]]]))
                else arrange(ord_df, .data[[ord_nms[i]]])
    }
  }
  by_present <- intersect(by_nm, names(ord_df))
  if (mode == "first") {
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(1) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)], new_nm)))
  } else {
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(n()) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)], new_nm)))
  }
  if (new_nm %in% names(dataset)) dataset[[new_nm]] <- NULL
  left_join(dataset, flagged, by = intersect(names(flagged), names(dataset)))
}

restrict_derivation <- function(dataset, derivation, args = list(), filter = NULL) {
  fe <- substitute(filter)
  in_rows <- if (!is.null(fe)) {
    tryCatch(which(eval(fe, dataset, parent.frame())), error = function(e) seq_len(nrow(dataset)))
  } else seq_len(nrow(dataset))
  out_rows <- setdiff(seq_len(nrow(dataset)), in_rows)
  sub_data  <- dataset[in_rows, , drop = FALSE]
  all_args  <- c(list(dataset = sub_data), args)
  derived   <- tryCatch(do.call(derivation, all_args), error = function(e) sub_data)
  bind_rows(derived, dataset[out_rows, , drop = FALSE]) |>
    arrange(match(seq_len(nrow(dataset)),
                  c(in_rows[seq_len(nrow(derived))], out_rows)))
}

params <- function(...) list(...)

derivation_slice <- function(filter = NULL, args = list()) {
  list(filter = substitute(filter), args = args)
}

slice_derivation <- function(dataset, derivation, args = list(), ...) {
  slices  <- list(...)
  result  <- dataset
  covered <- integer(0)
  for (sl in slices) {
    fe <- sl$filter
    idx <- if (!is.null(fe)) {
      tryCatch(which(eval(fe, result, parent.frame())), error = function(e) integer(0))
    } else seq_len(nrow(result))
    idx <- setdiff(idx, covered)
    if (length(idx) == 0) next
    merged_args <- modifyList(args, sl$args)
    sub_data    <- result[idx, , drop = FALSE]
    all_args    <- c(list(dataset = sub_data), merged_args)
    derived     <- tryCatch(do.call(derivation, all_args), error = function(e) sub_data)
    result[idx, names(derived)] <- derived[seq_len(length(idx)), names(derived)]
    covered <- c(covered, idx)
  }
  result
}

# ── BDS-specific mocks ────────────────────────────────────────────────────────
derive_vars_merged_lookup <- function(dataset, dataset_add, by_vars, new_vars = NULL, ...) {
  by_nm   <- sapply(by_vars, deparse)
  if (!is.null(new_vars)) {
    nms <- sapply(new_vars, deparse)
    keep <- intersect(c(by_nm, nms), names(dataset_add))
    dataset_add <- dataset_add |> select(all_of(keep))
  }
  by_present <- intersect(by_nm, names(dataset_add))
  left_join(dataset, dataset_add, by = by_present)
}

derive_var_anrind <- function(dataset) {
  dataset |> mutate(
    ANRIND = case_when(
      is.na(AVAL)               ~ NA_character_,
      !is.na(ANRHI) & AVAL > ANRHI ~ "HIGH",
      !is.na(ANRLO) & AVAL < ANRLO ~ "LOW",
      TRUE                          ~ "NORMAL"
    )
  )
}

derive_var_ontrtfl <- function(dataset, start_date, end_date = NULL,
                                ref_start_date, ref_end_date,
                                ref_end_window = 0, ...) {
  sd_nm  <- deparse(substitute(start_date))
  rs_nm  <- deparse(substitute(ref_start_date))
  re_nm  <- deparse(substitute(ref_end_date))
  dataset |> mutate(
    ONTRTFL = case_when(
      is.na(.data[[sd_nm]]) | is.na(.data[[rs_nm]]) ~ NA_character_,
      .data[[sd_nm]] >= .data[[rs_nm]] &
        .data[[sd_nm]] <= (.data[[re_nm]] + ref_end_window) ~ "Y",
      TRUE ~ NA_character_
    )
  )
}

derive_var_base <- function(dataset, by_vars, source_var, new_var) {
  src_nm <- deparse(substitute(source_var))
  new_nm <- deparse(substitute(new_var))
  by_nm  <- sapply(by_vars, deparse)
  by_present <- intersect(by_nm, names(dataset))
  baseline <- dataset |>
    filter(!is.na(ABLFL) & ABLFL == "Y") |>
    select(all_of(c(by_present, src_nm))) |>
    rename(!!new_nm := !!src_nm)
  if (new_nm %in% names(dataset)) dataset[[new_nm]] <- NULL
  left_join(dataset, baseline, by = by_present)
}

derive_var_chg <- function(dataset) {
  dataset |> mutate(CHG = if ("BASE" %in% names(dataset)) AVAL - BASE else NA_real_)
}

derive_var_pchg <- function(dataset) {
  dataset |> mutate(
    PCHG = if ("BASE" %in% names(dataset))
      ifelse(BASE == 0, NA_real_, (AVAL - BASE) / abs(BASE) * 100)
    else NA_real_
  )
}

derive_var_shift <- function(dataset, new_var, from_var, to_var) {
  new_nm  <- deparse(substitute(new_var))
  from_nm <- deparse(substitute(from_var))
  to_nm   <- deparse(substitute(to_var))
  dataset |> mutate(
    !!new_nm := case_when(
      is.na(.data[[from_nm]]) | is.na(.data[[to_nm]]) ~ NA_character_,
      TRUE ~ paste0(.data[[from_nm]], " to ", .data[[to_nm]])
    )
  )
}

# derive_param_bmi / derive_param_map / derive_param_computed
derive_param_bmi <- function(dataset, by_vars, weight_code = "WEIGHT",
                              height_code = "HEIGHT", set_values_to = list(),
                              constant_by_vars = NULL) {
  by_nm <- sapply(by_vars, deparse)
  by_nm <- setdiff(by_nm, c("PARAMCD", "PARAM", "PARAMN", "PARCAT1"))
  w_rows <- dataset |> filter(PARAMCD == weight_code)
  h_rows <- dataset |> filter(PARAMCD == height_code)
  if (!is.null(constant_by_vars)) {
    h_key <- sapply(constant_by_vars, deparse)
    h_rows <- h_rows |>
      group_by(across(all_of(intersect(h_key, names(h_rows))))) |>
      slice(1) |> ungroup()
  }
  h_use <- h_rows |> select(any_of(c("USUBJID", "AVAL"))) |>
    rename(HEIGHT_AVAL = AVAL)
  bmi_rows <- w_rows |>
    left_join(h_use, by = "USUBJID") |>
    filter(!is.na(AVAL) & !is.na(HEIGHT_AVAL) & HEIGHT_AVAL > 0) |>
    mutate(AVAL = AVAL / (HEIGHT_AVAL / 100)^2)
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]]), error = function(e) as.character(set_values_to[[nm]]))
    bmi_rows[[nm]] <- val
  }
  bmi_rows <- bmi_rows |> select(-any_of("HEIGHT_AVAL"))
  bind_rows(dataset, bmi_rows)
}

derive_param_map <- function(dataset, by_vars, set_values_to = list(), ...) {
  by_nm <- sapply(by_vars, deparse)
  by_nm <- setdiff(by_nm, c("PARAMCD", "PARAM", "PARAMN"))
  by_key <- intersect(by_nm, names(dataset))
  sys_rows <- dataset |> filter(PARAMCD == "SYSBP") |>
    select(all_of(c(by_key, "AVAL"))) |> rename(SYS = AVAL)
  dia_rows <- dataset |> filter(PARAMCD == "DIABP") |>
    select(all_of(c(by_key, "AVAL"))) |> rename(DIA = AVAL)
  map_rows <- inner_join(sys_rows, dia_rows, by = by_key) |>
    mutate(AVAL = (SYS + 2 * DIA) / 3) |>
    select(-SYS, -DIA)
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]]), error = function(e) as.character(set_values_to[[nm]]))
    map_rows[[nm]] <- val
  }
  bind_rows(dataset, map_rows)
}

derive_param_computed <- function(dataset, by_vars, parameters, set_values_to = list(), ...) {
  by_nm  <- sapply(by_vars, deparse)
  by_key <- intersect(by_nm, names(dataset))
  wide <- dataset |>
    filter(PARAMCD %in% parameters) |>
    select(all_of(c(by_key, "PARAMCD", "AVAL"))) |>
    pivot_wider(names_from = PARAMCD, values_from = AVAL,
                names_prefix = "AVAL.")
  valid_cols <- paste0("AVAL.", parameters)
  wide <- wide |> filter(if_all(all_of(valid_cols), ~ !is.na(.)))
  if (nrow(wide) == 0) return(dataset)
  eval_env <- list2env(as.list(wide), parent = baseenv())
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]], envir = eval_env),
                    error = function(e) as.character(set_values_to[[nm]]))
    wide[[nm]] <- val
  }
  new_rows <- wide |> select(-any_of(valid_cols))
  bind_rows(dataset, new_rows)
}

# ── Sample datasets ───────────────────────────────────────────────────────────
vs <- tribble(
  ~STUDYID, ~USUBJID, ~VSTESTCD, ~VSTEST,                   ~VSSTRESN, ~VSSTRESC,
  ~VSDTC,          ~VISIT,      ~ANRLO, ~ANRHI,
  "CDISCPILOT01", "01-001", "SYSBP",  "Systolic Blood Pressure",  118, "118",
  "2024-01-05", "SCREENING",  90, 140,
  "CDISCPILOT01", "01-001", "DIABP",  "Diastolic Blood Pressure",  76, "76",
  "2024-01-05", "SCREENING",  60,  90,
  "CDISCPILOT01", "01-001", "SYSBP",  "Systolic Blood Pressure",  124, "124",
  "2024-02-05", "WEEK 4",     90, 140,
  "CDISCPILOT01", "01-001", "DIABP",  "Diastolic Blood Pressure",  80, "80",
  "2024-02-05", "WEEK 4",     60,  90,
  "CDISCPILOT01", "01-001", "WEIGHT", "Weight",                    72,  "72",
  "2024-01-05", "SCREENING",  NA,  NA,
  "CDISCPILOT01", "01-001", "HEIGHT", "Height",                   172, "172",
  "2024-01-05", "SCREENING",  NA,  NA,
  "CDISCPILOT01", "01-002", "SYSBP",  "Systolic Blood Pressure",  148, "148",
  "2024-01-08", "SCREENING",  90, 140,
  "CDISCPILOT01", "01-002", "DIABP",  "Diastolic Blood Pressure",  92, "92",
  "2024-01-08", "SCREENING",  60,  90,
  "CDISCPILOT01", "01-002", "SYSBP",  "Systolic Blood Pressure",  138, "138",
  "2024-02-08", "WEEK 4",     90, 140,
  "CDISCPILOT01", "01-002", "DIABP",  "Diastolic Blood Pressure",  86, "86",
  "2024-02-08", "WEEK 4",     60,  90,
  "CDISCPILOT01", "01-002", "WEIGHT", "Weight",                    85,  "85",
  "2024-01-08", "SCREENING",  NA,  NA,
  "CDISCPILOT01", "01-002", "HEIGHT", "Height",                   180, "180",
  "2024-01-08", "SCREENING",  NA,  NA
)

adsl_bds <- tribble(
  ~STUDYID,        ~USUBJID, ~TRTSDT, ~TRTEDT, ~TRT01A,    ~TRT01P,    ~SEX, ~AGE, ~SAFFL,
  "CDISCPILOT01", "01-001",  19367,   19548,   "Xanomeli", "Xanomeli", "F",   52,   "Y",
  "CDISCPILOT01", "01-002",  19370,   19551,   "Placebo",  "Placebo",  "M",   61,   "Y"
)

param_lookup <- tribble(
  ~VSTESTCD, ~PARAMCD, ~PARAM,                              ~PARAMN, ~PARCAT1,
  "SYSBP",   "SYSBP",  "Systolic Blood Pressure (mmHg)",    1,       "Hemodynamic",
  "DIABP",   "DIABP",  "Diastolic Blood Pressure (mmHg)",   2,       "Hemodynamic",
  "PULSE",   "PULSE",  "Pulse Rate (beats/min)",            3,       "Hemodynamic",
  "TEMP",    "TEMP",   "Temperature (C)",                   4,       "Body System",
  "WEIGHT",  "WEIGHT", "Weight (kg)",                       10,      "Anthropometric",
  "HEIGHT",  "HEIGHT", "Height (cm)",                       11,      "Anthropometric"
)
```

## Welcome

This lesson covers **BDS (Basic Data Structure) ADaMs** — the long-format analysis datasets used for vitals (ADVS), labs (ADLB), and ECGs (ADEG). You'll build a complete ADVS step by step using admiral's canonical 13-step workflow.

**Key mocks in scope:** `derive_vars_merged_lookup()`, `derive_param_bmi()`, `derive_param_map()`, `derive_param_computed()`, `derive_var_anrind()`, `derive_var_ontrtfl()`, `derive_var_base()`, `derive_var_chg()`, `derive_var_pchg()`, `derive_var_shift()`, `restrict_derivation()`, `slice_derivation()`

---

## 1 — The BDS shape

```{r bds-shape-demo}
# BDS = one row per (subject × parameter × visit)
# Compare: ADSL has one row per subject; BDS has many
vs |>
  filter(VSTESTCD %in% c("SYSBP", "DIABP")) |>
  select(USUBJID, VSTESTCD, VSSTRESN, VSDTC, VISIT) |>
  arrange(USUBJID, VSTESTCD, VSDTC)
```

```{r quiz-bds-shape, echo=FALSE}
quiz(
  question(
    "Which of the following ADaMs uses the BDS structure?",
    answer("ADSL"),
    answer("ADAE"),
    answer("ADVS", correct = TRUE),
    answer("ADTTE"),
    allow_retry = TRUE
  ),
  question(
    "In a BDS dataset, what does PARAMCD identify?",
    answer("The subject"),
    answer("The analysis visit"),
    answer("The measured parameter (e.g., SYSBP)", correct = TRUE),
    answer("The baseline flag"),
    allow_retry = TRUE
  )
)
```

---

## 2 — Step 1: Merge ADSL variables

The first step brings treatment dates and arm from ADSL so every VS row knows its treatment window.

```{r merge-adsl-demo}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

advs <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds,
    new_vars    = adsl_vars,
    by_vars     = exprs(STUDYID, USUBJID)
  )

advs |> select(USUBJID, VSTESTCD, VSDTC, TRTSDT, TRTEDT, TRT01A) |> head(4)
```

```{r merge-adsl-ex, exercise=TRUE, exercise.lines=14}
# Exercise: merge TRTSDT, TRTEDT and TRT01A from adsl_bds onto vs
# then count how many non-NA TRTSDT values result
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

advs_tmp <- vs |>
  derive_vars_merged(
    dataset_add = ___,
    new_vars    = ___,
    by_vars     = exprs(STUDYID, USUBJID)
  )

sum(!is.na(advs_tmp$TRTSDT))
```

```{r merge-adsl-ex-hint-1}
# dataset_add = adsl_bds
# new_vars    = adsl_vars
# by_vars already given
```

```{r merge-adsl-ex-solution}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

advs_tmp <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds,
    new_vars    = adsl_vars,
    by_vars     = exprs(STUDYID, USUBJID)
  )

sum(!is.na(advs_tmp$TRTSDT))
```

---

## 3 — Step 2: Derive ADT and ADY

```{r derive-adt-demo}
advs <- advs |>
  derive_vars_dt(
    new_vars_prefix = "A",   # produces ADT (and ADTF if imputation flag)
    dtc             = VSDTC
  ) |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ADT)   # produces ADY
  )

advs |> select(USUBJID, VSTESTCD, VSDTC, ADT, ADY) |> head(4)
```

```{r quiz-adt, echo=FALSE}
quiz(
  question(
    "What does `new_vars_prefix = 'A'` produce in `derive_vars_dt()`?",
    answer("AVAL and AVALC"),
    answer("ADT and ADTF", correct = TRUE),
    answer("ASTDT and AENDT"),
    answer("ABLFL and ANRIND"),
    allow_retry = TRUE
  ),
  question(
    "ADY on the treatment start date equals:",
    answer("0"),
    answer("1", correct = TRUE),
    answer("-1"),
    answer("Depends on TRTSDT"),
    allow_retry = TRUE
  )
)
```

---

## 4 — Step 3: Assign PARAMCD via lookup

`derive_vars_merged_lookup()` is like `derive_vars_merged()` but designed for static lookup tables. It maps VSTESTCD → PARAMCD/PARAM/PARAMN.

```{r param-lookup-demo}
advs <- advs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD, PARAM, PARAMN, PARCAT1)
  )

advs |> select(VSTESTCD, PARAMCD, PARAM, PARAMN) |> distinct() |> head(6)
```

```{r aval-assign}
# Assign AVAL from the standardized numeric result
advs <- advs |>
  mutate(
    AVAL   = VSSTRESN,
    AVALC  = VSSTRESC,
    AVISIT = VISIT,
    AVISITN = case_when(
      VISIT == "SCREENING" ~ 0,
      VISIT == "WEEK 4"    ~ 4,
      TRUE                 ~ NA_real_
    )
  )
```

```{r paramcd-ex, exercise=TRUE, exercise.lines=14}
# Exercise: using param_lookup, map VSTESTCD → PARAMCD and PARAM
# then verify no NA PARAMCD for SYSBP or DIABP rows

vs_tmp <- vs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(___, ___)
  )

vs_tmp |> filter(VSTESTCD %in% c("SYSBP", "DIABP")) |>
  select(VSTESTCD, PARAMCD, PARAM) |> distinct()
```

```{r paramcd-ex-hint-1}
# new_vars = exprs(PARAMCD, PARAM)
```

```{r paramcd-ex-solution}
vs_tmp <- vs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD, PARAM)
  )

vs_tmp |> filter(VSTESTCD %in% c("SYSBP", "DIABP")) |>
  select(VSTESTCD, PARAMCD, PARAM) |> distinct()
```

---

## 5 — Derived parameters: BMI and MAP

```{r derived-params-demo}
# BMI from WEIGHT and HEIGHT
advs <- advs |>
  derive_param_bmi(
    by_vars          = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    weight_code      = "WEIGHT",
    height_code      = "HEIGHT",
    constant_by_vars = exprs(USUBJID),
    set_values_to    = exprs(
      PARAMCD = "BMI",
      PARAM   = "Body Mass Index (kg/m^2)",
      PARAMN  = 20
    )
  )

# MAP from SYSBP and DIABP
advs <- advs |>
  derive_param_map(
    by_vars       = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    set_values_to = exprs(
      PARAMCD = "MAP",
      PARAM   = "Mean Arterial Pressure (mmHg)",
      PARAMN  = 21
    )
  )

advs |> filter(PARAMCD %in% c("BMI", "MAP")) |>
  select(USUBJID, PARAMCD, AVAL, AVISIT) |>
  arrange(USUBJID, PARAMCD, AVISIT)
```

```{r quiz-params, echo=FALSE}
quiz(
  question(
    "What does `constant_by_vars = exprs(USUBJID)` do in `derive_param_bmi()`?",
    answer("Filters to one height per visit"),
    answer("Uses the single height value for all visits of a subject", correct = TRUE),
    answer("Forces BMI to be recalculated once"),
    answer("Nothing — it is ignored"),
    allow_retry = TRUE
  ),
  question(
    "Which function handles arbitrary formula parameters like FLI or HOMA-IR?",
    answer("derive_param_bmi()"),
    answer("derive_param_map()"),
    answer("derive_param_computed()", correct = TRUE),
    answer("derive_var_anrind()"),
    allow_retry = TRUE
  )
)
```

```{r computed-param-ex, exercise=TRUE, exercise.lines=18}
# Exercise: use derive_param_computed() to compute MAP with the formula
# AVAL = (AVAL.SYSBP + 2 * AVAL.DIABP) / 3
# Add PARAMCD = "MAP2", PARAM = "MAP (computed)", PARAMN = 22

advs_with_map2 <- advs |>
  filter(PARAMCD %in% c("SYSBP", "DIABP")) |>
  derive_param_computed(
    by_vars       = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    parameters    = ___,
    set_values_to = exprs(
      AVAL    = ___,
      PARAMCD = "MAP2",
      PARAM   = "MAP (computed)",
      PARAMN  = 22
    )
  )

advs_with_map2 |> filter(PARAMCD == "MAP2") |>
  select(USUBJID, PARAMCD, AVAL, AVISIT)
```

```{r computed-param-ex-hint-1}
# parameters = c("SYSBP", "DIABP")
# AVAL = (AVAL.SYSBP + 2 * AVAL.DIABP) / 3
```

```{r computed-param-ex-solution}
advs_with_map2 <- advs |>
  filter(PARAMCD %in% c("SYSBP", "DIABP")) |>
  derive_param_computed(
    by_vars       = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    parameters    = c("SYSBP", "DIABP"),
    set_values_to = exprs(
      AVAL    = (AVAL.SYSBP + 2 * AVAL.DIABP) / 3,
      PARAMCD = "MAP2",
      PARAM   = "MAP (computed)",
      PARAMN  = 22
    )
  )

advs_with_map2 |> filter(PARAMCD == "MAP2") |>
  select(USUBJID, PARAMCD, AVAL, AVISIT)
```

---

## 6 — ANRIND and ONTRTFL

```{r anrind-ontrtfl-demo}
advs <- advs |>
  derive_var_anrind() |>       # reads AVAL, ANRLO, ANRHI → ANRIND
  derive_var_ontrtfl(
    start_date      = ADT,
    ref_start_date  = TRTSDT,
    ref_end_date    = TRTEDT
  )

advs |>
  filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVAL, ANRLO, ANRHI, ANRIND, ONTRTFL, AVISIT)
```

```{r anrind-ex, exercise=TRUE, exercise.lines=12}
# Exercise: add ONTRTFL with a 30-day window after TRTEDT (ref_end_window = 30)
# then count how many rows get ONTRTFL = "Y"

advs_ont <- advs |>
  derive_var_ontrtfl(
    start_date     = ADT,
    ref_start_date = TRTSDT,
    ref_end_date   = TRTEDT,
    ref_end_window = ___
  )

sum(advs_ont$ONTRTFL == "Y", na.rm = TRUE)
```

```{r anrind-ex-hint-1}
# ref_end_window = 30
```

```{r anrind-ex-solution}
advs_ont <- advs |>
  derive_var_ontrtfl(
    start_date     = ADT,
    ref_start_date = TRTSDT,
    ref_end_date   = TRTEDT,
    ref_end_window = 30
  )

sum(advs_ont$ONTRTFL == "Y", na.rm = TRUE)
```

---

## 7 — Baseline: ABLFL, BASE, BASEC, BNRIND

The baseline row is the last pre-treatment observation per subject per parameter.

```{r baseline-demo}
# Step 1: flag baseline rows with restrict_derivation
advs <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = (!is.na(AVAL) & ADT <= TRTSDT)
  )

# Step 2: copy baseline AVAL and ANRIND to every row
advs <- advs |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var    = BASE
  ) |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = ANRIND,
    new_var    = BNRIND
  )

advs |>
  filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVISIT, AVAL, ABLFL, BASE, ANRIND, BNRIND)
```

```{r quiz-baseline, echo=FALSE}
quiz(
  question(
    "What value is stored in ABLFL for the baseline row?",
    answer("1"),
    answer("TRUE"),
    answer("Y", correct = TRUE),
    answer("BASELINE"),
    allow_retry = TRUE
  ),
  question(
    "Why do you call `derive_var_base()` twice (for AVAL and ANRIND)?",
    answer("It can only copy one variable at a time", correct = TRUE),
    answer("ANRIND doesn't exist yet when BASE is derived"),
    answer("Admiral mandates two-step baseline"),
    answer("BASE needs a separate call from BASEC"),
    allow_retry = TRUE
  )
)
```

```{r ablfl-ex, exercise=TRUE, exercise.lines=20}
# Exercise: for DIABP only, flag the LAST pre-treatment row as ABLFL
# then check: how many DIABP rows have ABLFL = "Y"?

advs_bl <- advs |>
  filter(PARAMCD == "DIABP") |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT),
      new_var = ABLFL,
      mode    = "___"   # last pre-treatment = "last"
    ),
    filter = (!is.na(AVAL) & ADT <= TRTSDT)
  )

sum(advs_bl$ABLFL == "Y", na.rm = TRUE)
```

```{r ablfl-ex-hint-1}
# mode = "last"
```

```{r ablfl-ex-solution}
advs_bl <- advs |>
  filter(PARAMCD == "DIABP") |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = (!is.na(AVAL) & ADT <= TRTSDT)
  )

sum(advs_bl$ABLFL == "Y", na.rm = TRUE)
```

---

## 8 — CHG, PCHG, and SHIFT1

```{r chg-shift-demo}
advs <- advs |>
  derive_var_chg() |>
  derive_var_pchg() |>
  derive_var_shift(
    new_var  = SHIFT1,
    from_var = BNRIND,
    to_var   = ANRIND
  )

advs |>
  filter(PARAMCD == "SYSBP" & !is.na(BASE)) |>
  select(USUBJID, AVISIT, AVAL, BASE, CHG, PCHG, BNRIND, ANRIND, SHIFT1)
```

```{r chg-ex, exercise=TRUE, exercise.lines=14}
# Exercise: compute CHG and PCHG for DIABP rows, then show
# max absolute CHG per subject

advs_chg <- advs |>
  filter(PARAMCD == "DIABP") |>
  derive_var_chg() |>
  derive_var_pchg()

advs_chg |>
  group_by(USUBJID) |>
  summarise(max_abs_chg = max(abs(CHG), na.rm = TRUE))
```

```{r chg-ex-hint-1}
# derive_var_chg() reads BASE and AVAL automatically — no arguments needed
```

```{r chg-ex-solution}
advs_chg <- advs |>
  filter(PARAMCD == "DIABP") |>
  derive_var_chg() |>
  derive_var_pchg()

advs_chg |>
  group_by(USUBJID) |>
  summarise(max_abs_chg = max(abs(CHG), na.rm = TRUE))
```

---

## 9 — restrict_derivation vs slice_derivation

```{r restrict-demo}
# restrict_derivation: apply a derivation only to filtered rows
# Here: flag worst on-treatment SYSBP per subject (highest = worst)

advs_worst <- advs |>
  filter(PARAMCD == "SYSBP") |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(desc(AVAL)),
      new_var = WORSTFL,
      mode    = "first"
    ),
    filter = ONTRTFL == "Y" & !is.na(AVAL)
  )

advs_worst |> select(USUBJID, AVISIT, AVAL, ONTRTFL, WORSTFL)
```

```{r slice-demo}
# slice_derivation: different rule per slice
# Slice 1: hemodynamic params → highest AVAL is worst
# Slice 2: DIABP → same rule here for demo

advs_sliced <- advs |>
  filter(PARAMCD %in% c("SYSBP", "DIABP")) |>
  slice_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      new_var = WORSTFL,
      mode    = "first"
    ),
    derivation_slice(
      filter = PARAMCD == "SYSBP" & ONTRTFL == "Y" & !is.na(AVAL),
      args   = params(order = exprs(desc(AVAL)))
    ),
    derivation_slice(
      filter = PARAMCD == "DIABP" & ONTRTFL == "Y" & !is.na(AVAL),
      args   = params(order = exprs(desc(AVAL)))
    )
  )

advs_sliced |> filter(!is.na(WORSTFL)) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, WORSTFL)
```

```{r quiz-restrict-slice, echo=FALSE}
quiz(
  question(
    "What is the key difference between `restrict_derivation()` and `slice_derivation()`?",
    answer("restrict_derivation() is faster"),
    answer("slice_derivation() can use different derivation arguments per data slice", correct = TRUE),
    answer("restrict_derivation() creates new rows; slice_derivation() doesn't"),
    answer("They are interchangeable"),
    allow_retry = TRUE
  )
)
```

```{r restrict-ex, exercise=TRUE, exercise.lines=16}
# Exercise: using restrict_derivation, derive ANL01FL = "Y"
# for the LATEST observation per (USUBJID, PARAMCD, AVISIT)
# among rows where AVAL is not NA

advs_anl <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD, AVISIT),
      order   = exprs(desc(ADT)),
      new_var = ANL01FL,
      mode    = "___"
    ),
    filter = !is.na(AVAL) & !is.na(AVISIT)
  )

sum(advs_anl$ANL01FL == "Y", na.rm = TRUE)
```

```{r restrict-ex-hint-1}
# mode = "first" (after sorting desc(ADT), first = latest)
```

```{r restrict-ex-solution}
advs_anl <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD, AVISIT),
      order   = exprs(desc(ADT)),
      new_var = ANL01FL,
      mode    = "first"
    ),
    filter = !is.na(AVAL) & !is.na(AVISIT)
  )

sum(advs_anl$ANL01FL == "Y", na.rm = TRUE)
```

---

## 10 — Full ADVS pipeline capstone

```{r capstone-advs, exercise=TRUE, exercise.lines=70}
# Build a complete ADVS from vs + adsl_bds in one pipe:
# 1. Merge ADSL vars (TRTSDT, TRTEDT, TRT01A, TRT01P)
# 2. Derive ADT and ADY
# 3. Map PARAMCD/PARAM/PARAMN via param_lookup
# 4. Assign AVAL, AVALC, AVISIT, AVISITN
# 5. Derive BMI and MAP parameters
# 6. Derive ANRIND and ONTRTFL
# 7. Flag ABLFL (last pre-treatment row per subj × param)
# 8. Derive BASE and BNRIND
# 9. Derive CHG, PCHG, SHIFT1

adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

advs_full <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds, new_vars = adsl_vars,
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dt(new_vars_prefix = "A", dtc = VSDTC) |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT)) |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup, by_vars = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD, PARAM, PARAMN)
  ) |>
  mutate(
    AVAL    = VSSTRESN,
    AVALC   = VSSTRESC,
    AVISIT  = VISIT,
    AVISITN = case_when(
      VISIT == "SCREENING" ~ 0,
      VISIT == "WEEK 4"    ~ 4,
      TRUE                 ~ NA_real_
    )
  ) |>
  derive_param_bmi(
    by_vars = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    weight_code = "WEIGHT", height_code = "HEIGHT",
    constant_by_vars = exprs(USUBJID),
    set_values_to = exprs(PARAMCD = "BMI", PARAM = "BMI (kg/m^2)", PARAMN = 20)
  ) |>
  derive_param_map(
    by_vars = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    set_values_to = exprs(PARAMCD = "MAP", PARAM = "MAP (mmHg)", PARAMN = 21)
  ) |>
  derive_var_anrind() |>
  derive_var_ontrtfl(start_date = ADT, ref_start_date = TRTSDT, ref_end_date = TRTEDT) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL, mode = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  ) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD),
                  source_var = AVAL, new_var = BASE) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD),
                  source_var = ANRIND, new_var = BNRIND) |>
  derive_var_chg() |>
  derive_var_pchg() |>
  derive_var_shift(new_var = SHIFT1, from_var = BNRIND, to_var = ANRIND)

# Inspect the result
advs_full |>
  filter(PARAMCD %in% c("SYSBP", "BMI", "MAP")) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, BASE, CHG, ANRIND, SHIFT1) |>
  arrange(USUBJID, PARAMCD, AVISIT)
```

---

## 11 — ADLB & ADEG differences

```{r quiz-adlb-adeg, echo=FALSE}
quiz(
  question(
    "ADLB lab values: what SDTM variable maps to AVAL?",
    answer("LBORRES"),
    answer("LBSTRESN", correct = TRUE),
    answer("LBTEST"),
    answer("LBORNRHI"),
    allow_retry = TRUE
  ),
  question(
    "For ADEG with triplicate ECGs per visit, which variable captures the sub-visit timepoint?",
    answer("AVISITN"),
    answer("APERIOD"),
    answer("ATPT", correct = TRUE),
    answer("ADY"),
    allow_retry = TRUE
  ),
  question(
    "Which admiral function computes QTcF (Fridericia-corrected QT)?",
    answer("derive_param_computed()"),
    answer("derive_param_qtc()", correct = TRUE),
    answer("derive_var_anrind()"),
    answer("derive_param_map()"),
    allow_retry = TRUE
  ),
  question(
    "NCI CTCAE grading for lab toxicity uses which admiral function?",
    answer("derive_var_anrind()"),
    answer("derive_var_atoxgr()", correct = TRUE),
    answer("derive_var_base()"),
    answer("derive_var_chg()"),
    allow_retry = TRUE
  )
)
```

---

## Quiz — final

```{r quiz-final-16, echo=FALSE}
quiz(
  question(
    "What does ABLFL = 'Y' mark?",
    answer("Any post-baseline row"),
    answer("The row used as the baseline reference for that subject and parameter", correct = TRUE),
    answer("An adverse event flag"),
    answer("An on-treatment row"),
    allow_retry = TRUE
  ),
  question(
    "CHG is defined as:",
    answer("BASE - AVAL"),
    answer("AVAL / BASE"),
    answer("AVAL - BASE", correct = TRUE),
    answer("(AVAL - BASE) / BASE * 100"),
    allow_retry = TRUE
  ),
  question(
    "What does SHIFT1 = 'NORMAL to HIGH' tell you?",
    answer("AVAL exceeded the reference range"),
    answer("The subject's reference range category went from NORMAL at baseline to HIGH on-treatment", correct = TRUE),
    answer("The parameter was flagged as worst"),
    answer("The baseline was missing"),
    allow_retry = TRUE
  ),
  question(
    "Which higher-order wrapper lets you apply the same derivation with different argument sets to different data slices?",
    answer("restrict_derivation()"),
    answer("call_derivation()"),
    answer("slice_derivation()", correct = TRUE),
    answer("derive_vars_merged()"),
    allow_retry = TRUE
  )
)
```
