---
title: "Lesson 16 — admiral Part 3: BDS Datasets (ADVS, ADLB, ADEG)"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 16_admiral_part3_bds.Rmd
# PURPOSE       : admiral Part 3 — BDS datasets: ADLB (lab) and ADVS (vital signs)
#                 derivations including PARAMCD/AVAL/BASE/CHG/PCHG,
#                 toxicity grading, and visit windowing.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── Mock helpers from earlier lessons ────────────────────────────────────────
exprs <- function(...) {
  dots <- as.list(substitute(list(...)))[-1]
  dots
}

convert_blanks_to_na <- function(dataset) {
  dataset |>
    mutate(across(where(is.character), ~ifelse(. == "", NA_character_, .)))
}

# ── Date utilities ────────────────────────────────────────────────────────────
.impute_dtc <- function(dtc, date_imputation = "first") {
  sapply(dtc, function(d) {
    if (is.na(d)) return(NA_character_)
    if (nchar(d) == 4) {
      suffix <- if (date_imputation == "last") "-12-31" else "-01-01"
      paste0(d, suffix)
    } else if (nchar(d) == 7) {
      if (date_imputation == "last") {
        yr  <- as.integer(substr(d, 1, 4))
        mo  <- as.integer(substr(d, 6, 7))
        nxt <- if (mo == 12) paste0(yr + 1, "-01-01") else paste0(yr, "-", sprintf("%02d", mo + 1), "-01")
        as.character(as.Date(nxt) - 1)
      } else {
        paste0(d, "-01")
      }
    } else {
      d
    }
  })
}

convert_dtc_to_dt <- function(dtc, date_imputation = "first") {
  suppressWarnings(as.numeric(as.Date(.impute_dtc(dtc, date_imputation))))
}

# ── Core admiral mocks ────────────────────────────────────────────────────────
derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL,
                                order = NULL, filter_add = NULL, mode = NULL, ...) {
  by_nm   <- sapply(by_vars, function(x) deparse(x))
  join_df <- dataset_add
  if (!is.null(filter_add)) {
    fe <- substitute(filter_add)
    join_df <- tryCatch(filter(join_df, !!fe), error = function(e) join_df)
  }
  if (!is.null(new_vars)) {
    old_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d, "=")[[1]][2]) else d
    })
    new_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d, "=")[[1]][1]) else d
    })
    keep <- intersect(c(by_nm, old_nms), names(join_df))
    join_df <- join_df |> select(all_of(keep))
    for (i in seq_along(old_nms)) {
      if (old_nms[i] != new_nms[i] && old_nms[i] %in% names(join_df))
        join_df <- join_df |> rename(!!new_nms[i] := !!old_nms[i])
    }
  }
  by_present <- intersect(by_nm, names(join_df))
  left_join(dataset, join_df, by = by_present)
}

derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation = "none", highest_imputation = "n",
                            flag_imputation = "auto", ...) {
  dtc_nm  <- deparse(substitute(dtc))
  imp_rule <- if (date_imputation == "none" && highest_imputation == "n") "none"
              else if (date_imputation == "last") "last" else "first"
  raw <- if (dtc_nm %in% names(dataset)) dataset[[dtc_nm]] else rep(NA_character_, nrow(dataset))
  imputed <- if (imp_rule == "none") raw else .impute_dtc(raw, imp_rule)
  dataset[[paste0(new_vars_prefix, "DT")]] <- suppressWarnings(as.numeric(as.Date(imputed)))
  if (flag_imputation != "none") {
    orig_len <- nchar(raw); imp_len <- nchar(imputed)
    dataset[[paste0(new_vars_prefix, "DTF")]] <- case_when(
      is.na(raw)                        ~ NA_character_,
      orig_len <= 4 & !is.na(imputed)  ~ "Y",
      orig_len <= 7 & !is.na(imputed)  ~ "M",
      TRUE                              ~ NA_character_
    )
  }
  dataset
}

derive_vars_dtm <- function(dataset, dtc, new_vars_prefix,
                             date_imputation = "first", time_imputation = "first",
                             highest_imputation = "n", ...) {
  dtc_nm <- deparse(substitute(dtc))
  raw    <- if (dtc_nm %in% names(dataset)) dataset[[dtc_nm]] else rep(NA_character_, nrow(dataset))
  imputed <- .impute_dtc(raw, date_imputation)
  suffix  <- if (time_imputation == "last") "T23:59:59" else "T00:00:00"
  dataset[[paste0(new_vars_prefix, "DTM")]] <- suppressWarnings(
    as.numeric(as.POSIXct(paste0(imputed, suffix), tz = "UTC"))
  )
  dataset[[paste0(new_vars_prefix, "DTF")]] <- case_when(
    is.na(raw)                              ~ NA_character_,
    nchar(raw) <= 4 & !is.na(imputed)      ~ "Y",
    nchar(raw) <= 7 & !is.na(imputed)      ~ "M",
    TRUE                                    ~ NA_character_
  )
  dataset[[paste0(new_vars_prefix, "TMF")]] <- case_when(
    is.na(raw)                              ~ NA_character_,
    !grepl("T", raw)                        ~ "H",
    TRUE                                    ~ NA_character_
  )
  dataset
}

derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref_nm  <- deparse(substitute(reference_date))
  src_nms <- sapply(source_vars, deparse)
  for (sv in src_nms) {
    dy_nm <- sub("DT$", "DY", sv)
    if (sv %in% names(dataset) && ref_nm %in% names(dataset)) {
      v <- dataset[[sv]]; r <- dataset[[ref_nm]]
      dataset[[dy_nm]] <- ifelse(is.na(v) | is.na(r), NA_real_,
        ifelse(v >= r, as.numeric(v - r) + 1, as.numeric(v - r)))
    }
  }
  dataset
}

derive_var_extreme_flag <- function(dataset, by_vars, order, new_var, mode = "first") {
  # Accept either a bare name (direct call) or a string (via do.call from params())
  .e <- substitute(new_var); new_nm <- if (is.character(.e)) .e else deparse(.e)
  by_nm   <- sapply(by_vars, deparse)
  ord_nms <- sapply(order, function(x) {
    d <- deparse(x); gsub("desc\\(|\\)", "", d)
  })
  desc_flags <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
  ord_df <- dataset
  for (i in rev(seq_along(ord_nms))) {
    if (ord_nms[i] %in% names(ord_df)) {
      ord_df <- if (desc_flags[i]) arrange(ord_df, desc(.data[[ord_nms[i]]]))
                else arrange(ord_df, .data[[ord_nms[i]]])
    }
  }
  by_present <- intersect(by_nm, names(ord_df))
  if (mode == "first") {
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(1) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)], new_nm)))
  } else {
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(n()) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)], new_nm)))
  }
  if (new_nm %in% names(dataset)) dataset[[new_nm]] <- NULL
  left_join(dataset, flagged, by = intersect(names(flagged), names(dataset)))
}

restrict_derivation <- function(dataset, derivation, args = list(), filter = NULL) {
  fe <- substitute(filter)
  in_rows <- if (!is.null(fe)) {
    tryCatch(which(eval(fe, dataset, parent.frame())), error = function(e) seq_len(nrow(dataset)))
  } else seq_len(nrow(dataset))
  out_rows <- setdiff(seq_len(nrow(dataset)), in_rows)
  sub_data  <- dataset[in_rows, , drop = FALSE]
  all_args  <- c(list(dataset = sub_data), args)
  derived   <- tryCatch(do.call(derivation, all_args), error = function(e) sub_data)
  bind_rows(derived, dataset[out_rows, , drop = FALSE]) |>
    arrange(match(seq_len(nrow(dataset)),
                  c(in_rows[seq_len(nrow(derived))], out_rows)))
}

params <- function(...) {
  # Bare names (e.g. new_var = ABLFL) are captured as strings.
  # Everything else (exprs() calls, strings, numbers) is evaluated normally.
  parent_env <- parent.frame()
  dots <- as.list(substitute(list(...)))[-1]
  lapply(dots, function(x) {
    if (is.symbol(x)) deparse(x) else eval(x, envir = parent_env)
  })
}

derivation_slice <- function(filter = NULL, args = list()) {
  list(filter = substitute(filter), args = args)
}

slice_derivation <- function(dataset, derivation, args = list(), ...) {
  slices  <- list(...)
  result  <- dataset
  covered <- integer(0)
  for (sl in slices) {
    fe <- sl$filter
    idx <- if (!is.null(fe)) {
      tryCatch(which(eval(fe, result, parent.frame())), error = function(e) integer(0))
    } else seq_len(nrow(result))
    idx <- setdiff(idx, covered)
    if (length(idx) == 0) next
    merged_args <- modifyList(args, sl$args)
    sub_data    <- result[idx, , drop = FALSE]
    all_args    <- c(list(dataset = sub_data), merged_args)
    derived     <- tryCatch(do.call(derivation, all_args), error = function(e) sub_data)
    # Initialise any brand-new columns to NA before subset assignment;
    # R cannot add columns via df[idx, "new_col"] <- x on an existing df.
    new_cols <- setdiff(names(derived), names(result))
    for (nc in new_cols) result[[nc]] <- NA
    result[idx, names(derived)] <- derived[seq_len(length(idx)), names(derived)]
    covered <- c(covered, idx)
  }
  result
}

# ── BDS-specific mocks ────────────────────────────────────────────────────────
derive_vars_merged_lookup <- function(dataset, dataset_add, by_vars, new_vars = NULL, ...) {
  by_nm   <- sapply(by_vars, deparse)
  if (!is.null(new_vars)) {
    nms <- sapply(new_vars, deparse)
    keep <- intersect(c(by_nm, nms), names(dataset_add))
    dataset_add <- dataset_add |> select(all_of(keep))
  }
  by_present <- intersect(by_nm, names(dataset_add))
  left_join(dataset, dataset_add, by = by_present)
}

derive_var_anrind <- function(dataset) {
  dataset |> mutate(
    ANRIND = case_when(
      is.na(AVAL)               ~ NA_character_,
      !is.na(ANRHI) & AVAL > ANRHI ~ "HIGH",
      !is.na(ANRLO) & AVAL < ANRLO ~ "LOW",
      TRUE                          ~ "NORMAL"
    )
  )
}

derive_var_ontrtfl <- function(dataset, start_date, end_date = NULL,
                                ref_start_date, ref_end_date,
                                ref_end_window = 0, ...) {
  sd_nm  <- deparse(substitute(start_date))
  rs_nm  <- deparse(substitute(ref_start_date))
  re_nm  <- deparse(substitute(ref_end_date))
  dataset |> mutate(
    ONTRTFL = case_when(
      is.na(.data[[sd_nm]]) | is.na(.data[[rs_nm]]) ~ NA_character_,
      .data[[sd_nm]] >= .data[[rs_nm]] &
        .data[[sd_nm]] <= (.data[[re_nm]] + ref_end_window) ~ "Y",
      TRUE ~ NA_character_
    )
  )
}

derive_var_base <- function(dataset, by_vars, source_var, new_var) {
  src_nm <- deparse(substitute(source_var))
  new_nm <- deparse(substitute(new_var))
  by_nm  <- sapply(by_vars, deparse)
  by_present <- intersect(by_nm, names(dataset))
  baseline <- dataset |>
    filter(!is.na(ABLFL) & ABLFL == "Y") |>
    select(all_of(c(by_present, src_nm))) |>
    rename(!!new_nm := !!src_nm)
  if (new_nm %in% names(dataset)) dataset[[new_nm]] <- NULL
  left_join(dataset, baseline, by = by_present)
}

derive_var_chg <- function(dataset) {
  dataset |> mutate(CHG = if ("BASE" %in% names(dataset)) AVAL - BASE else NA_real_)
}

derive_var_pchg <- function(dataset) {
  dataset |> mutate(
    PCHG = if ("BASE" %in% names(dataset))
      ifelse(BASE == 0, NA_real_, (AVAL - BASE) / abs(BASE) * 100)
    else NA_real_
  )
}

derive_var_shift <- function(dataset, new_var, from_var, to_var) {
  new_nm  <- deparse(substitute(new_var))
  from_nm <- deparse(substitute(from_var))
  to_nm   <- deparse(substitute(to_var))
  dataset |> mutate(
    !!new_nm := case_when(
      is.na(.data[[from_nm]]) | is.na(.data[[to_nm]]) ~ NA_character_,
      TRUE ~ paste0(.data[[from_nm]], " to ", .data[[to_nm]])
    )
  )
}

# derive_param_bmi / derive_param_map / derive_param_computed
derive_param_bmi <- function(dataset, by_vars, weight_code = "WEIGHT",
                              height_code = "HEIGHT", set_values_to = list(),
                              constant_by_vars = NULL) {
  by_nm <- sapply(by_vars, deparse)
  by_nm <- setdiff(by_nm, c("PARAMCD", "PARAM", "PARAMN", "PARCAT1"))
  w_rows <- dataset |> filter(PARAMCD == weight_code)
  h_rows <- dataset |> filter(PARAMCD == height_code)
  if (!is.null(constant_by_vars)) {
    h_key <- sapply(constant_by_vars, deparse)
    h_rows <- h_rows |>
      group_by(across(all_of(intersect(h_key, names(h_rows))))) |>
      slice(1) |> ungroup()
  }
  h_use <- h_rows |> select(any_of(c("USUBJID", "AVAL"))) |>
    rename(HEIGHT_AVAL = AVAL)
  bmi_rows <- w_rows |>
    left_join(h_use, by = "USUBJID") |>
    filter(!is.na(AVAL) & !is.na(HEIGHT_AVAL) & HEIGHT_AVAL > 0) |>
    mutate(AVAL = AVAL / (HEIGHT_AVAL / 100)^2)
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]]), error = function(e) as.character(set_values_to[[nm]]))
    bmi_rows[[nm]] <- val
  }
  bmi_rows <- bmi_rows |> select(-any_of("HEIGHT_AVAL"))
  bind_rows(dataset, bmi_rows)
}

derive_param_map <- function(dataset, by_vars, set_values_to = list(), ...) {
  by_nm <- sapply(by_vars, deparse)
  by_nm <- setdiff(by_nm, c("PARAMCD", "PARAM", "PARAMN"))
  by_key <- intersect(by_nm, names(dataset))
  sys_rows <- dataset |> filter(PARAMCD == "SYSBP") |>
    select(all_of(c(by_key, "AVAL"))) |> rename(SYS = AVAL)
  dia_rows <- dataset |> filter(PARAMCD == "DIABP") |>
    select(all_of(c(by_key, "AVAL"))) |> rename(DIA = AVAL)
  map_rows <- inner_join(sys_rows, dia_rows, by = by_key) |>
    mutate(AVAL = (SYS + 2 * DIA) / 3) |>
    select(-SYS, -DIA)
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]]), error = function(e) as.character(set_values_to[[nm]]))
    map_rows[[nm]] <- val
  }
  bind_rows(dataset, map_rows)
}

derive_param_computed <- function(dataset, by_vars, parameters, set_values_to = list(), ...) {
  by_nm  <- sapply(by_vars, deparse)
  by_key <- intersect(by_nm, names(dataset))
  wide <- dataset |>
    filter(PARAMCD %in% parameters) |>
    select(all_of(c(by_key, "PARAMCD", "AVAL"))) |>
    pivot_wider(names_from = PARAMCD, values_from = AVAL,
                names_prefix = "AVAL.")
  valid_cols <- paste0("AVAL.", parameters)
  wide <- wide |> filter(if_all(all_of(valid_cols), ~ !is.na(.)))
  if (nrow(wide) == 0) return(dataset)
  eval_env <- list2env(as.list(wide), parent = baseenv())
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]], envir = eval_env),
                    error = function(e) as.character(set_values_to[[nm]]))
    wide[[nm]] <- val
  }
  new_rows <- wide |> select(-any_of(valid_cols))
  bind_rows(dataset, new_rows)
}

# ── Sample datasets ───────────────────────────────────────────────────────────
vs <- tribble(
  ~STUDYID, ~USUBJID, ~VSTESTCD, ~VSTEST,                   ~VSSTRESN, ~VSSTRESC,
  ~VSDTC,          ~VISIT,      ~ANRLO, ~ANRHI,
  "CDISCPILOT01", "01-001", "SYSBP",  "Systolic Blood Pressure",  118, "118",
  "2024-01-05", "SCREENING",  90, 140,
  "CDISCPILOT01", "01-001", "DIABP",  "Diastolic Blood Pressure",  76, "76",
  "2024-01-05", "SCREENING",  60,  90,
  "CDISCPILOT01", "01-001", "SYSBP",  "Systolic Blood Pressure",  124, "124",
  "2024-02-05", "WEEK 4",     90, 140,
  "CDISCPILOT01", "01-001", "DIABP",  "Diastolic Blood Pressure",  80, "80",
  "2024-02-05", "WEEK 4",     60,  90,
  "CDISCPILOT01", "01-001", "WEIGHT", "Weight",                    72,  "72",
  "2024-01-05", "SCREENING",  NA,  NA,
  "CDISCPILOT01", "01-001", "HEIGHT", "Height",                   172, "172",
  "2024-01-05", "SCREENING",  NA,  NA,
  "CDISCPILOT01", "01-002", "SYSBP",  "Systolic Blood Pressure",  148, "148",
  "2024-01-08", "SCREENING",  90, 140,
  "CDISCPILOT01", "01-002", "DIABP",  "Diastolic Blood Pressure",  92, "92",
  "2024-01-08", "SCREENING",  60,  90,
  "CDISCPILOT01", "01-002", "SYSBP",  "Systolic Blood Pressure",  138, "138",
  "2024-02-08", "WEEK 4",     90, 140,
  "CDISCPILOT01", "01-002", "DIABP",  "Diastolic Blood Pressure",  86, "86",
  "2024-02-08", "WEEK 4",     60,  90,
  "CDISCPILOT01", "01-002", "WEIGHT", "Weight",                    85,  "85",
  "2024-01-08", "SCREENING",  NA,  NA,
  "CDISCPILOT01", "01-002", "HEIGHT", "Height",                   180, "180",
  "2024-01-08", "SCREENING",  NA,  NA
)

adsl_bds <- tribble(
  ~STUDYID,        ~USUBJID, ~TRTSDT, ~TRTEDT, ~TRT01A,    ~TRT01P,    ~SEX, ~AGE, ~SAFFL,
  # TRTSDT after SCREENING (19727/19730) so SCREENING = baseline (ADT <= TRTSDT)
  # TRTEDT after WEEK 4 (19758/19761) so WEEK 4 = on-treatment (ONTRTFL = "Y")
  # 2024-01-10 = 19732; 2024-01-13 = 19735; 2024-03-01 = 19783
  "CDISCPILOT01", "01-001",  19732,   19783,   "Xanomeli", "Xanomeli", "F",   52,   "Y",
  "CDISCPILOT01", "01-002",  19735,   19783,   "Placebo",  "Placebo",  "M",   61,   "Y"
)

param_lookup <- tribble(
  ~VSTESTCD, ~PARAMCD, ~PARAM,                              ~PARAMN, ~PARCAT1,
  "SYSBP",   "SYSBP",  "Systolic Blood Pressure (mmHg)",    1,       "Hemodynamic",
  "DIABP",   "DIABP",  "Diastolic Blood Pressure (mmHg)",   2,       "Hemodynamic",
  "PULSE",   "PULSE",  "Pulse Rate (beats/min)",            3,       "Hemodynamic",
  "TEMP",    "TEMP",   "Temperature (C)",                   4,       "Body System",
  "WEIGHT",  "WEIGHT", "Weight (kg)",                       10,      "Anthropometric",
  "HEIGHT",  "HEIGHT", "Height (cm)",                       11,      "Anthropometric"
)

# ── Real pharmaverseadam datasets (for validation and exploration exercises) ───
library(pharmaverseadam)
advs_real <- pharmaverseadam::advs
adlb_real <- pharmaverseadam::adlb |>
  filter(is.na(DTYPE))            # remove admiral summary rows (AVISITN 9997/9998)
adsl_real <- pharmaverseadam::adsl
```

## Welcome

This lesson covers **BDS (Basic Data Structure) ADaMs** — the long-format analysis datasets used for vitals (ADVS), labs (ADLB), and ECGs (ADEG). You'll build a complete ADVS step by step using admiral's canonical 13-step workflow.

**Key mocks in scope:** `derive_vars_merged_lookup()`, `derive_param_bmi()`, `derive_param_map()`, `derive_param_computed()`, `derive_var_anrind()`, `derive_var_ontrtfl()`, `derive_var_base()`, `derive_var_chg()`, `derive_var_pchg()`, `derive_var_shift()`, `restrict_derivation()`, `slice_derivation()`

---

## 1 — The BDS shape

```{r bds-shape-demo}
# BDS = one row per (subject × parameter × visit)
# Compare: ADSL has one row per subject; BDS has many
vs |>
  filter(VSTESTCD %in% c("SYSBP", "DIABP")) |>
  select(USUBJID, VSTESTCD, VSSTRESN, VSDTC, VISIT) |>
  arrange(USUBJID, VSTESTCD, VSDTC)
```

```{r quiz-bds-shape, echo=FALSE}
quiz(
  question(
    "Which of the following ADaMs uses the BDS structure?",
    answer("ADSL"),
    answer("ADAE"),
    answer("ADVS", correct = TRUE),
    answer("ADTTE"),
    allow_retry = TRUE
  ),
  question(
    "In a BDS dataset, what does PARAMCD identify?",
    answer("The subject"),
    answer("The analysis visit"),
    answer("The measured parameter (e.g., SYSBP)", correct = TRUE),
    answer("The baseline flag"),
    allow_retry = TRUE
  )
)
```

---

## 2 — Step 1: Merge ADSL variables

The first step brings treatment dates and arm from ADSL so every VS row knows its treatment window.

```{r merge-adsl-demo}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

advs <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds,
    new_vars    = adsl_vars,
    by_vars     = exprs(STUDYID, USUBJID)
  )

advs |> select(USUBJID, VSTESTCD, VSDTC, TRTSDT, TRTEDT, TRT01A) |> head(4)
```

```{r merge-adsl-ex, exercise=TRUE, exercise.lines=14}
# Exercise: merge TRTSDT, TRTEDT and TRT01A from adsl_bds onto vs
# then count how many non-NA TRTSDT values result
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

advs_tmp <- vs |>
  derive_vars_merged(
    dataset_add = ___,
    new_vars    = ___,
    by_vars     = exprs(STUDYID, USUBJID)
  )

sum(!is.na(advs_tmp$TRTSDT))
```

```{r merge-adsl-ex-hint-1}
# dataset_add = adsl_bds
# new_vars    = adsl_vars
# by_vars already given
```

```{r merge-adsl-ex-solution}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

advs_tmp <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds,
    new_vars    = adsl_vars,
    by_vars     = exprs(STUDYID, USUBJID)
  )

sum(!is.na(advs_tmp$TRTSDT))
```

---

## 3 — Step 2: Derive ADT and ADY

```{r derive-adt-demo}
advs <- advs |>
  derive_vars_dt(
    new_vars_prefix = "A",   # produces ADT (and ADTF if imputation flag)
    dtc             = VSDTC
  ) |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ADT)   # produces ADY
  )

advs |> select(USUBJID, VSTESTCD, VSDTC, ADT, ADY) |> head(4)
```

```{r quiz-adt, echo=FALSE}
quiz(
  question(
    "What does `new_vars_prefix = 'A'` produce in `derive_vars_dt()`?",
    answer("AVAL and AVALC"),
    answer("ADT and ADTF", correct = TRUE),
    answer("ASTDT and AENDT"),
    answer("ABLFL and ANRIND"),
    allow_retry = TRUE
  ),
  question(
    "ADY on the treatment start date equals:",
    answer("0"),
    answer("1", correct = TRUE),
    answer("-1"),
    answer("Depends on TRTSDT"),
    allow_retry = TRUE
  )
)
```

### Exercise: Explore real ADVS parameter inventory

The toy dataset has 2 subjects and 6 test codes. The real `advs_real` from pharmaverseadam has the full CDISCPILOT01 study — 254 subjects and many more parameters and visits.

> **Orientation exercise**: before building a new BDS dataset, a programmer always audits the source SDTM to understand its shape. Answer the three questions using `advs_real`.

```{r advs-real-explore, exercise=TRUE, exercise.lines=30}
# Q1: How many unique PARAMCD values are in advs_real?
n_params <- advs_real |> distinct(PARAMCD) |> nrow()
cat("Unique parameters:", n_params, "\n")

# Q2: Which PARAMCDs are present?
advs_real |> distinct(PARAMCD, PARAM) |> arrange(PARAMCD) |> print()

# Q3: How many ONTRTFL = "Y" rows exist per PARAMCD?
advs_real |>
  filter(ONTRTFL == "Y") |>
  count(PARAMCD, name = "n_ontreatment") |>
  arrange(desc(n_ontreatment)) |>
  print()
```

```{r advs-real-explore-hint-1}
# distinct(PARAMCD) gives unique parameter codes
# count(PARAMCD) counts rows per parameter
# filter(ONTRTFL == "Y") selects on-treatment observations
```

```{r advs-real-explore-solution}
n_params <- advs_real |> distinct(PARAMCD) |> nrow()
cat("Unique parameters:", n_params, "\n\n")

advs_real |> distinct(PARAMCD, PARAM) |> arrange(PARAMCD) |> print()

advs_real |>
  filter(ONTRTFL == "Y") |>
  count(PARAMCD, name = "n_ontreatment") |>
  arrange(desc(n_ontreatment)) |>
  print()
```

---

## 4 — Step 3: Assign PARAMCD via lookup

`derive_vars_merged_lookup()` is like `derive_vars_merged()` but designed for static lookup tables. It maps VSTESTCD → PARAMCD/PARAM/PARAMN.

```{r param-lookup-demo}
advs <- advs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD, PARAM, PARAMN, PARCAT1)
  )

advs |> select(VSTESTCD, PARAMCD, PARAM, PARAMN) |> distinct() |> head(6)
```

```{r aval-assign}
# Assign AVAL from the standardized numeric result
advs <- advs |>
  mutate(
    AVAL   = VSSTRESN,
    AVALC  = VSSTRESC,
    AVISIT = VISIT,
    AVISITN = case_when(
      VISIT == "SCREENING" ~ 0,
      VISIT == "WEEK 4"    ~ 4,
      TRUE                 ~ NA_real_
    )
  )
```

```{r paramcd-ex, exercise=TRUE, exercise.lines=14}
# Exercise: using param_lookup, map VSTESTCD → PARAMCD and PARAM
# then verify no NA PARAMCD for SYSBP or DIABP rows

vs_tmp <- vs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(___, ___)
  )

vs_tmp |> filter(VSTESTCD %in% c("SYSBP", "DIABP")) |>
  select(VSTESTCD, PARAMCD, PARAM) |> distinct()
```

```{r paramcd-ex-hint-1}
# new_vars = exprs(PARAMCD, PARAM)
```

```{r paramcd-ex-solution}
vs_tmp <- vs |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    by_vars     = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD, PARAM)
  )

vs_tmp |> filter(VSTESTCD %in% c("SYSBP", "DIABP")) |>
  select(VSTESTCD, PARAMCD, PARAM) |> distinct()
```

---

## 5 — Derived parameters: BMI and MAP

```{r derived-params-demo}
# BMI from WEIGHT and HEIGHT
advs <- advs |>
  derive_param_bmi(
    by_vars          = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    weight_code      = "WEIGHT",
    height_code      = "HEIGHT",
    constant_by_vars = exprs(USUBJID),
    set_values_to    = exprs(
      PARAMCD = "BMI",
      PARAM   = "Body Mass Index (kg/m^2)",
      PARAMN  = 20
    )
  )

# MAP from SYSBP and DIABP
advs <- advs |>
  derive_param_map(
    by_vars       = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    set_values_to = exprs(
      PARAMCD = "MAP",
      PARAM   = "Mean Arterial Pressure (mmHg)",
      PARAMN  = 21
    )
  )

advs |> filter(PARAMCD %in% c("BMI", "MAP")) |>
  select(USUBJID, PARAMCD, AVAL, AVISIT) |>
  arrange(USUBJID, PARAMCD, AVISIT)
```

```{r quiz-params, echo=FALSE}
quiz(
  question(
    "What does `constant_by_vars = exprs(USUBJID)` do in `derive_param_bmi()`?",
    answer("Filters to one height per visit"),
    answer("Uses the single height value for all visits of a subject", correct = TRUE),
    answer("Forces BMI to be recalculated once"),
    answer("Nothing — it is ignored"),
    allow_retry = TRUE
  ),
  question(
    "Which function handles arbitrary formula parameters like FLI or HOMA-IR?",
    answer("derive_param_bmi()"),
    answer("derive_param_map()"),
    answer("derive_param_computed()", correct = TRUE),
    answer("derive_var_anrind()"),
    allow_retry = TRUE
  )
)
```

```{r computed-param-ex, exercise=TRUE, exercise.lines=18}
# Exercise: use derive_param_computed() to compute MAP with the formula
# AVAL = (AVAL.SYSBP + 2 * AVAL.DIABP) / 3
# Add PARAMCD = "MAP2", PARAM = "MAP (computed)", PARAMN = 22

advs_with_map2 <- advs |>
  filter(PARAMCD %in% c("SYSBP", "DIABP")) |>
  derive_param_computed(
    by_vars       = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    parameters    = ___,
    set_values_to = exprs(
      AVAL    = ___,
      PARAMCD = "MAP2",
      PARAM   = "MAP (computed)",
      PARAMN  = 22
    )
  )

advs_with_map2 |> filter(PARAMCD == "MAP2") |>
  select(USUBJID, PARAMCD, AVAL, AVISIT)
```

```{r computed-param-ex-hint-1}
# parameters = c("SYSBP", "DIABP")
# AVAL = (AVAL.SYSBP + 2 * AVAL.DIABP) / 3
```

```{r computed-param-ex-solution}
advs_with_map2 <- advs |>
  filter(PARAMCD %in% c("SYSBP", "DIABP")) |>
  derive_param_computed(
    by_vars       = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    parameters    = c("SYSBP", "DIABP"),
    set_values_to = exprs(
      AVAL    = (AVAL.SYSBP + 2 * AVAL.DIABP) / 3,
      PARAMCD = "MAP2",
      PARAM   = "MAP (computed)",
      PARAMN  = 22
    )
  )

advs_with_map2 |> filter(PARAMCD == "MAP2") |>
  select(USUBJID, PARAMCD, AVAL, AVISIT)
```

---

## 6 — ANRIND and ONTRTFL

```{r anrind-ontrtfl-demo}
advs <- advs |>
  derive_var_anrind() |>       # reads AVAL, ANRLO, ANRHI → ANRIND
  derive_var_ontrtfl(
    start_date      = ADT,
    ref_start_date  = TRTSDT,
    ref_end_date    = TRTEDT
  )

advs |>
  filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVAL, ANRLO, ANRHI, ANRIND, ONTRTFL, AVISIT)
```

```{r anrind-ex, exercise=TRUE, exercise.lines=12}
# Exercise: add ONTRTFL with a 30-day window after TRTEDT (ref_end_window = 30)
# then count how many rows get ONTRTFL = "Y"

advs_ont <- advs |>
  derive_var_ontrtfl(
    start_date     = ADT,
    ref_start_date = TRTSDT,
    ref_end_date   = TRTEDT,
    ref_end_window = ___
  )

sum(advs_ont$ONTRTFL == "Y", na.rm = TRUE)
```

```{r anrind-ex-hint-1}
# ref_end_window = 30
```

```{r anrind-ex-solution}
advs_ont <- advs |>
  derive_var_ontrtfl(
    start_date     = ADT,
    ref_start_date = TRTSDT,
    ref_end_date   = TRTEDT,
    ref_end_window = 30
  )

sum(advs_ont$ONTRTFL == "Y", na.rm = TRUE)
```

---

## 7 — Baseline: ABLFL, BASE, BASEC, BNRIND

The baseline row is the last pre-treatment observation per subject per parameter.

```{r baseline-demo}
# Step 1: flag baseline rows with restrict_derivation
advs <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = (!is.na(AVAL) & ADT <= TRTSDT)
  )

# Step 2: copy baseline AVAL and ANRIND to every row
advs <- advs |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var    = BASE
  ) |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = ANRIND,
    new_var    = BNRIND
  )

advs |>
  filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVISIT, AVAL, ABLFL, BASE, ANRIND, BNRIND)
```

```{r quiz-baseline, echo=FALSE}
quiz(
  question(
    "What value is stored in ABLFL for the baseline row?",
    answer("1"),
    answer("TRUE"),
    answer("Y", correct = TRUE),
    answer("BASELINE"),
    allow_retry = TRUE
  ),
  question(
    "Why do you call `derive_var_base()` twice (for AVAL and ANRIND)?",
    answer("It can only copy one variable at a time", correct = TRUE),
    answer("ANRIND doesn't exist yet when BASE is derived"),
    answer("Admiral mandates two-step baseline"),
    answer("BASE needs a separate call from BASEC"),
    allow_retry = TRUE
  )
)
```

```{r ablfl-ex, exercise=TRUE, exercise.lines=20}
# Exercise: for DIABP only, flag the LAST pre-treatment row as ABLFL
# then check: how many DIABP rows have ABLFL = "Y"?

advs_bl <- advs |>
  filter(PARAMCD == "DIABP") |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT),
      new_var = ABLFL,
      mode    = "___"   # last pre-treatment = "last"
    ),
    filter = (!is.na(AVAL) & ADT <= TRTSDT)
  )

sum(advs_bl$ABLFL == "Y", na.rm = TRUE)
```

```{r ablfl-ex-hint-1}
# mode = "last"
```

```{r ablfl-ex-solution}
advs_bl <- advs |>
  filter(PARAMCD == "DIABP") |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = (!is.na(AVAL) & ADT <= TRTSDT)
  )

sum(advs_bl$ABLFL == "Y", na.rm = TRUE)
```

---

## 8 — CHG, PCHG, and SHIFT1

```{r chg-shift-demo}
advs <- advs |>
  derive_var_chg() |>
  derive_var_pchg() |>
  derive_var_shift(
    new_var  = SHIFT1,
    from_var = BNRIND,
    to_var   = ANRIND
  )

advs |>
  filter(PARAMCD == "SYSBP" & !is.na(BASE)) |>
  select(USUBJID, AVISIT, AVAL, BASE, CHG, PCHG, BNRIND, ANRIND, SHIFT1)
```

```{r chg-ex, exercise=TRUE, exercise.lines=14}
# Exercise: compute CHG and PCHG for DIABP rows, then show
# max absolute CHG per subject

advs_chg <- advs |>
  filter(PARAMCD == "DIABP") |>
  derive_var_chg() |>
  derive_var_pchg()

advs_chg |>
  group_by(USUBJID) |>
  summarise(max_abs_chg = max(abs(CHG), na.rm = TRUE))
```

```{r chg-ex-hint-1}
# derive_var_chg() reads BASE and AVAL automatically — no arguments needed
```

```{r chg-ex-solution}
advs_chg <- advs |>
  filter(PARAMCD == "DIABP") |>
  derive_var_chg() |>
  derive_var_pchg()

advs_chg |>
  group_by(USUBJID) |>
  summarise(max_abs_chg = max(abs(CHG), na.rm = TRUE))
```

---

## 9 — restrict_derivation vs slice_derivation

```{r restrict-demo}
# restrict_derivation: apply a derivation only to filtered rows
# Here: flag worst on-treatment SYSBP per subject (highest = worst)

advs_worst <- advs |>
  filter(PARAMCD == "SYSBP") |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(desc(AVAL)),
      new_var = WORSTFL,
      mode    = "first"
    ),
    filter = ONTRTFL == "Y" & !is.na(AVAL)
  )

advs_worst |> select(USUBJID, AVISIT, AVAL, ONTRTFL, WORSTFL)
```

```{r slice-demo}
# slice_derivation: different rule per slice
# Slice 1: hemodynamic params → highest AVAL is worst
# Slice 2: DIABP → same rule here for demo

advs_sliced <- advs |>
  filter(PARAMCD %in% c("SYSBP", "DIABP")) |>
  slice_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      new_var = WORSTFL,
      mode    = "first"
    ),
    derivation_slice(
      filter = PARAMCD == "SYSBP" & ONTRTFL == "Y" & !is.na(AVAL),
      args   = params(order = exprs(desc(AVAL)))
    ),
    derivation_slice(
      filter = PARAMCD == "DIABP" & ONTRTFL == "Y" & !is.na(AVAL),
      args   = params(order = exprs(desc(AVAL)))
    )
  )

advs_sliced |> filter(!is.na(WORSTFL)) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, WORSTFL)
```

```{r quiz-restrict-slice, echo=FALSE}
quiz(
  question(
    "What is the key difference between `restrict_derivation()` and `slice_derivation()`?",
    answer("restrict_derivation() is faster"),
    answer("slice_derivation() can use different derivation arguments per data slice", correct = TRUE),
    answer("restrict_derivation() creates new rows; slice_derivation() doesn't"),
    answer("They are interchangeable"),
    allow_retry = TRUE
  )
)
```

```{r restrict-ex, exercise=TRUE, exercise.lines=16}
# Exercise: using restrict_derivation, derive ANL01FL = "Y"
# for the LATEST observation per (USUBJID, PARAMCD, AVISIT)
# among rows where AVAL is not NA

advs_anl <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD, AVISIT),
      order   = exprs(desc(ADT)),
      new_var = ANL01FL,
      mode    = "___"
    ),
    filter = !is.na(AVAL) & !is.na(AVISIT)
  )

sum(advs_anl$ANL01FL == "Y", na.rm = TRUE)
```

```{r restrict-ex-hint-1}
# mode = "first" (after sorting desc(ADT), first = latest)
```

```{r restrict-ex-solution}
advs_anl <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD, AVISIT),
      order   = exprs(desc(ADT)),
      new_var = ANL01FL,
      mode    = "first"
    ),
    filter = !is.na(AVAL) & !is.na(AVISIT)
  )

sum(advs_anl$ANL01FL == "Y", na.rm = TRUE)
```

---

## 10 — Full ADVS pipeline capstone

```{r capstone-advs, exercise=TRUE, exercise.lines=70}
# Build a complete ADVS from vs + adsl_bds in one pipe:
# 1. Merge ADSL vars (TRTSDT, TRTEDT, TRT01A, TRT01P)
# 2. Derive ADT and ADY
# 3. Map PARAMCD/PARAM/PARAMN via param_lookup
# 4. Assign AVAL, AVALC, AVISIT, AVISITN
# 5. Derive BMI and MAP parameters
# 6. Derive ANRIND and ONTRTFL
# 7. Flag ABLFL (last pre-treatment row per subj × param)
# 8. Derive BASE and BNRIND
# 9. Derive CHG, PCHG, SHIFT1

adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

advs_full <- vs |>
  derive_vars_merged(
    dataset_add = adsl_bds, new_vars = adsl_vars,
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  derive_vars_dt(new_vars_prefix = "A", dtc = VSDTC) |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT)) |>
  derive_vars_merged_lookup(
    dataset_add = param_lookup, by_vars = exprs(VSTESTCD),
    new_vars    = exprs(PARAMCD, PARAM, PARAMN)
  ) |>
  mutate(
    AVAL    = VSSTRESN,
    AVALC   = VSSTRESC,
    AVISIT  = VISIT,
    AVISITN = case_when(
      VISIT == "SCREENING" ~ 0,
      VISIT == "WEEK 4"    ~ 4,
      TRUE                 ~ NA_real_
    )
  ) |>
  derive_param_bmi(
    by_vars = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    weight_code = "WEIGHT", height_code = "HEIGHT",
    constant_by_vars = exprs(USUBJID),
    set_values_to = exprs(PARAMCD = "BMI", PARAM = "BMI (kg/m^2)", PARAMN = 20)
  ) |>
  derive_param_map(
    by_vars = exprs(STUDYID, USUBJID, AVISIT, AVISITN),
    set_values_to = exprs(PARAMCD = "MAP", PARAM = "MAP (mmHg)", PARAMN = 21)
  ) |>
  derive_var_anrind() |>
  derive_var_ontrtfl(start_date = ADT, ref_start_date = TRTSDT, ref_end_date = TRTEDT) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL, mode = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  ) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD),
                  source_var = AVAL, new_var = BASE) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD),
                  source_var = ANRIND, new_var = BNRIND) |>
  derive_var_chg() |>
  derive_var_pchg() |>
  derive_var_shift(new_var = SHIFT1, from_var = BNRIND, to_var = ANRIND)

# Inspect the result
advs_full |>
  filter(PARAMCD %in% c("SYSBP", "BMI", "MAP")) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, BASE, CHG, ANRIND, SHIFT1) |>
  arrange(USUBJID, PARAMCD, AVISIT)
```

---

## 11 — ADLB & ADEG differences

```{r quiz-adlb-adeg, echo=FALSE}
quiz(
  question(
    "ADLB lab values: what SDTM variable maps to AVAL?",
    answer("LBORRES"),
    answer("LBSTRESN", correct = TRUE),
    answer("LBTEST"),
    answer("LBORNRHI"),
    allow_retry = TRUE
  ),
  question(
    "For ADEG with triplicate ECGs per visit, which variable captures the sub-visit timepoint?",
    answer("AVISITN"),
    answer("APERIOD"),
    answer("ATPT", correct = TRUE),
    answer("ADY"),
    allow_retry = TRUE
  ),
  question(
    "Which admiral function computes QTcF (Fridericia-corrected QT)?",
    answer("derive_param_computed()"),
    answer("derive_param_qtc()", correct = TRUE),
    answer("derive_var_anrind()"),
    answer("derive_param_map()"),
    allow_retry = TRUE
  ),
  question(
    "NCI CTCAE grading for lab toxicity uses which admiral function?",
    answer("derive_var_anrind()"),
    answer("derive_var_atoxgr()", correct = TRUE),
    answer("derive_var_base()"),
    answer("derive_var_chg()"),
    allow_retry = TRUE
  )
)
```

### Exercise: Explore real ADLB toxicity grades

Real lab data reveals patterns that toy datasets can't: multiple parameters, sparse missing values, wide toxicity grade distributions. Use `adlb_real` to investigate.

> **Note:** `adlb_real` has admiral summary rows (`DTYPE != NA`) already removed in setup. `ATOXGR` is stored as integer (converted from character in pharmaverseadam).

```{r adlb-real-explore, exercise=TRUE, exercise.lines=35}
# Ensure ATOXGR is integer (pharmaverseadam stores it as character)
adlb_work <- adlb_real |>
  mutate(ATOXGR = suppressWarnings(as.integer(ATOXGR)),
         BTOXGR = suppressWarnings(as.integer(BTOXGR)))

# Q1: How many unique lab parameters are in this ADLB?
cat("Unique parameters:", n_distinct(adlb_work$PARAMCD), "\n\n")

# Q2: Count records by ATOXGR grade
adlb_work |>
  count(ATOXGR, name = "n_records") |>
  arrange(ATOXGR) |>
  print()

# Q3: Which parameters have the most Grade >=3 on-treatment toxicity events?
adlb_work |>
  filter(ONTRTFL == "Y" & !is.na(ATOXGR) & ATOXGR >= 3) |>
  count(PARAMCD, PARAM, name = "n_grade3plus") |>
  arrange(desc(n_grade3plus)) |>
  head(10) |>
  print()
```

```{r adlb-real-explore-hint-1}
# filter(ONTRTFL == "Y" & ATOXGR >= 3) selects on-treatment Grade 3+ events
# count(PARAMCD) then summarises per parameter
# arrange(desc(n_grade3plus)) puts worst parameters first
```

```{r adlb-real-explore-solution}
adlb_work <- adlb_real |>
  mutate(ATOXGR = suppressWarnings(as.integer(ATOXGR)),
         BTOXGR = suppressWarnings(as.integer(BTOXGR)))

cat("Unique parameters:", n_distinct(adlb_work$PARAMCD), "\n\n")

adlb_work |>
  count(ATOXGR, name = "n_records") |>
  arrange(ATOXGR) |>
  print()

adlb_work |>
  filter(ONTRTFL == "Y" & !is.na(ATOXGR) & ATOXGR >= 3) |>
  count(PARAMCD, PARAM, name = "n_grade3plus") |>
  arrange(desc(n_grade3plus)) |>
  head(10) |>
  print()
```

---

## 11.5 — P21 Derivation Diagnostics

BDS derivations are particularly prone to subtle errors that slip through code review but surface immediately in Pinnacle 21 validation. Each diagnostic below shows a derivation spec, a broken implementation, and the P21 error received.

---

### Diagnostic 1 — ABLFL: Wrong mode for pre-treatment baseline

**Derivation method (define.xml / Pinnacle 21 field):**

> *ABLFL = 'Y' for the LAST non-missing pre-treatment observation per subject per parameter (i.e., the measurement closest in time to first dose, before dosing began). This observation is used as the baseline reference. Rows where AVAL is missing must be excluded. Source: ADT, TRTSDT.*

The programmer submitted:

```r
advs_draft <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = "first"    # ← bug: should be "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  )
```

**P21 report:** *"ABLFL flagged on earliest screening record rather than the record closest to first dose. For subjects with multiple pre-treatment visits, baseline values do not match expected SCREENING 2 or pre-dose values."*

```{r p21-bug-ablfl, exercise=TRUE, exercise.lines=22}
# Fix the ABLFL derivation: baseline = LAST non-missing pre-treatment row.
# With order = exprs(ADT, AVISITN), mode = "last" picks the highest ADT.

advs_fixed_ablfl <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = "___"    # fix this
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  )

# Verify: show ABLFL-flagged rows — should be SCREENING (last pre-treatment visit)
advs_fixed_ablfl |>
  filter(!is.na(ABLFL)) |>
  select(USUBJID, PARAMCD, AVISIT, ADT, AVAL, ABLFL)
```

```{r p21-bug-ablfl-hint-1}
# mode = "last" — after ordering by ADT ascending, "last" = highest date = closest to first dose
# With "first" you get the earliest (oldest) pre-treatment record — wrong baseline
```

```{r p21-bug-ablfl-solution}
advs_fixed_ablfl <- advs |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      order   = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode    = "last"
    ),
    filter = !is.na(AVAL) & ADT <= TRTSDT
  )

advs_fixed_ablfl |>
  filter(!is.na(ABLFL)) |>
  select(USUBJID, PARAMCD, AVISIT, ADT, AVAL, ABLFL)
```

---

### Diagnostic 2 — BASE: Missing PARAMCD from by_vars

**Derivation method (define.xml / Pinnacle 21 field):**

> *BASE = AVAL at the baseline record (ABLFL = 'Y') for each subject-parameter combination. Every observation for a given subject and parameter receives the same BASE value (the one from the ABLFL = 'Y' row). Source: derived ABLFL, AVAL. By variables: STUDYID, USUBJID, PARAMCD.*

The programmer submitted:

```r
advs_draft2 <- advs |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID),   # ← PARAMCD missing!
    source_var = AVAL,
    new_var    = BASE
  )
```

**P21 report:** *"BASE values for DIABP rows match SYSBP baseline values. Multiple parameters share the same BASE within a subject. PARAMCD must be included in the by variables for baseline derivation."*

```{r p21-bug-base-byvars, exercise=TRUE, exercise.lines=18}
# Fix: PARAMCD must be in by_vars so each parameter gets its own baseline value.

advs_fixed_base <- advs |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, ___),   # add the missing variable
    source_var = AVAL,
    new_var    = BASE
  )

# Verify: SYSBP and DIABP should have different BASE values for the same subject
advs_fixed_base |>
  filter(PARAMCD %in% c("SYSBP", "DIABP") & !is.na(ABLFL)) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, BASE, ABLFL)
```

```{r p21-bug-base-byvars-hint-1}
# by_vars = exprs(STUDYID, USUBJID, PARAMCD)
# Without PARAMCD, derive_var_base() joins only on STUDYID + USUBJID,
# so the first ABLFL = "Y" row (regardless of PARAMCD) is used as BASE for ALL parameters.
```

```{r p21-bug-base-byvars-solution}
advs_fixed_base <- advs |>
  derive_var_base(
    by_vars    = exprs(STUDYID, USUBJID, PARAMCD),
    source_var = AVAL,
    new_var    = BASE
  )

advs_fixed_base |>
  filter(PARAMCD %in% c("SYSBP", "DIABP") & !is.na(ABLFL)) |>
  select(USUBJID, PARAMCD, AVISIT, AVAL, BASE, ABLFL)
```

---

### Diagnostic 3 — CHG: Sign reversal

**Derivation method (define.xml / Pinnacle 21 field):**

> *CHG = AVAL − BASE (change from baseline). Positive CHG indicates an increase relative to baseline. PCHG = (AVAL − BASE) / |BASE| × 100. BASE = 0 must result in PCHG = missing. Source: derived AVAL, BASE.*

The programmer submitted:

```r
advs_draft3 <- advs |>
  mutate(
    CHG  = BASE - AVAL,            # ← sign reversed
    PCHG = (BASE - AVAL) / abs(AVAL) * 100   # ← two bugs: sign + wrong denominator
  )
```

**P21 report:** *"CHG is negative for subjects where AVAL > BASE. Expected positive CHG for elevated post-baseline values. PCHG denominator should be |BASE|, not |AVAL|. Rule: AD0271 — CHG derivation inconsistent with spec."*

```{r p21-bug-chg, exercise=TRUE, exercise.lines=20}
# Fix both bugs:
# 1. CHG = AVAL - BASE  (not BASE - AVAL)
# 2. PCHG = (AVAL - BASE) / abs(BASE) * 100, with NA when BASE == 0

advs_fixed_chg <- advs |>
  mutate(
    CHG  = ___ - ___,
    PCHG = if_else(
      BASE == 0 | is.na(BASE),
      NA_real_,
      (___ - ___) / abs(___) * 100
    )
  )

# Verify: for WEEK 4 SYSBP, a higher AVAL than baseline should give positive CHG
advs_fixed_chg |>
  filter(PARAMCD == "SYSBP" & !is.na(BASE)) |>
  select(USUBJID, AVISIT, AVAL, BASE, CHG, PCHG)
```

```{r p21-bug-chg-hint-1}
# CHG  = AVAL - BASE
# PCHG = (AVAL - BASE) / abs(BASE) * 100
# Denominator is |BASE| — not |AVAL|. If BASE = 0, PCHG = NA (division by zero).
```

```{r p21-bug-chg-solution}
advs_fixed_chg <- advs |>
  mutate(
    CHG  = AVAL - BASE,
    PCHG = if_else(
      BASE == 0 | is.na(BASE),
      NA_real_,
      (AVAL - BASE) / abs(BASE) * 100
    )
  )

advs_fixed_chg |>
  filter(PARAMCD == "SYSBP" & !is.na(BASE)) |>
  select(USUBJID, AVISIT, AVAL, BASE, CHG, PCHG)
```

---

## Quiz — final

```{r quiz-final-16, echo=FALSE}
quiz(
  question(
    "A programmer flags ABLFL using `mode = 'first'` with `order = exprs(ADT)`. P21 reports baseline values don't match the expected pre-dose visit. What is the consequence of 'first' vs 'last'?",
    answer("'first' picks the row with the highest ADT — closest to first dose"),
    answer("'first' picks the row with the lowest ADT — earliest screening visit — not the most recent pre-treatment assessment. Fix: use mode = 'last'", correct = TRUE),
    answer("'first' and 'last' produce identical results when there is only one pre-treatment visit"),
    answer("'first' correctly picks the latest date because ADT is stored in descending order"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "BASE is derived with `by_vars = exprs(STUDYID, USUBJID)` — PARAMCD is omitted. What goes wrong?",
    answer("The function errors because PARAMCD is a required argument"),
    answer("All parameters for a subject receive the same BASE value — whichever ABLFL = 'Y' row appears first in the dataset, regardless of PARAMCD", correct = TRUE),
    answer("BASE is set to NA for all non-SYSBP parameters"),
    answer("Nothing — BASE joins on USUBJID alone and correctly finds the right row"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "P21 rule AD0271 flags CHG values. The programmer coded `CHG = BASE - AVAL`. What is the correct formula and why does the sign matter?",
    answer("CHG = AVAL - BASE. Sign matters because a positive CHG should indicate an increase post-baseline; the reversed formula reports increases as negative, misleading safety reviewers", correct = TRUE),
    answer("CHG = BASE - AVAL is correct per ADaM IG — PCHG corrects the sign"),
    answer("CHG = (AVAL + BASE) / 2 — the average of the two values"),
    answer("CHG = AVAL / BASE — the ratio from baseline"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "PCHG is coded as `(AVAL - BASE) / abs(AVAL) * 100`. P21 flags the denominator. What is the correct denominator and why?",
    answer("`abs(AVAL)` is correct — the percent change is relative to the current value"),
    answer("`abs(BASE)` — percent change from baseline is always relative to the baseline value, not the current observation. Using AVAL in the denominator gives a different (and non-standard) metric", correct = TRUE),
    answer("`(AVAL + BASE) / 2` — the average of baseline and post-baseline"),
    answer("The denominator does not matter because PCHG is informational only"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "ONTRTFL = 'Y' for a row where ADT = TRTSDT (treatment start date). Is this correct?",
    answer("No — the treatment start date is pre-treatment, so ONTRTFL should be NA"),
    answer("Yes — the standard definition is ADT >= TRTSDT AND ADT <= TRTEDT (inclusive on both ends), so the start date itself is on-treatment", correct = TRUE),
    answer("It depends on the protocol — some studies exclude the start date"),
    answer("No — ONTRTFL only applies to lab values, not vitals"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "A BDS dataset contains both SYSBP and DIABP records. The programmer filters to `PARAMCD == 'SYSBP'` before calling `derive_param_map()`. What happens?",
    answer("MAP is computed correctly from the filtered SYSBP rows"),
    answer("MAP cannot be computed — derive_param_map() needs both SYSBP and DIABP rows in the input dataset to calculate (SYS + 2×DIA) / 3. Filtering to one parameter first removes the data needed for the formula", correct = TRUE),
    answer("derive_param_map() fetches DIABP rows automatically from adsl"),
    answer("MAP is derived from SYSBP alone using a built-in reference value for DIA"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "SHIFT1 = 'NORMAL to HIGH' for a subject at Week 12. What does this mean, and which variables produce it?",
    answer("AVAL > ANRHI at Week 12 AND BNRIND = 'NORMAL' at baseline — SHIFT1 = paste0(BNRIND, ' to ', ANRIND)", correct = TRUE),
    answer("The subject's AVAL increased by more than the normal reference range"),
    answer("ANRIND = 'HIGH' at baseline and ANRIND = 'NORMAL' at Week 12 — the order is reversed"),
    answer("SHIFT1 compares two on-treatment visits, not baseline to on-treatment"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "`constant_by_vars = exprs(USUBJID)` in `derive_param_bmi()` tells admiral to:",
    answer("Compute BMI once per subject using a single height value (e.g., screening height), carried forward to all visits — rather than requiring a height measurement at every visit", correct = TRUE),
    answer("Restrict BMI computation to one visit per subject only"),
    answer("Use a constant height of 170 cm when no height record is found"),
    answer("Join the height from the ADSL dataset instead of the VS source"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "A lab ADLB has ATOXGR stored as character (e.g., '3', '-1'). The programmer compares `ATOXGR > '2'` without converting. What is the problem?",
    answer("Character comparison is alphabetic: '3' > '2' is TRUE (correct by accident), but '-1' > '2' is also TRUE because '-' sorts after digits — so below-normal grades appear as high toxicity", correct = TRUE),
    answer("R coerces character to integer automatically in > comparisons"),
    answer("The comparison errors immediately with a type mismatch"),
    answer("Only values with leading zeros are affected"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "Why does `restrict_derivation()` leave non-matching rows unchanged, while applying the derivation only to filtered rows?",
    answer("It errors on non-matching rows instead of leaving them unchanged"),
    answer("It applies the derivation to non-matching rows with a NULL result"),
    answer("It splits the dataset into matching and non-matching subsets, applies the derivation to the matching subset, then recombines — non-matching rows retain their original column values", correct = TRUE),
    answer("It is identical to filter() followed by the derivation — rows not meeting the filter are dropped permanently"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```
