---
title: "Lesson 17 — admiral Part 4: OCCDS Datasets (ADAE, ADCM)"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 17_admiral_part4_occds.Rmd
# PURPOSE       : admiral Part 4 — OCCDS datasets: ADAE derivation including serious AE
#                 flags, treatment-emergent AE (TEAE) criteria, AE
#                 severity grading, and duration calculations.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── shared mocks (from prior lessons) ────────────────────────────────────────
exprs <- function(...) {
  dots <- as.list(substitute(list(...)))[-1]
  dots
}

convert_blanks_to_na <- function(dataset) {
  dataset |> mutate(across(where(is.character), ~ifelse(. == "", NA_character_, .)))
}

.impute_dtc <- function(dtc, date_imputation = "first") {
  sapply(dtc, function(d) {
    if (is.na(d)) return(NA_character_)
    if (nchar(d) == 4) {
      paste0(d, if (date_imputation == "last") "-12-31" else "-01-01")
    } else if (nchar(d) == 7) {
      if (date_imputation == "last") {
        yr <- as.integer(substr(d, 1, 4)); mo <- as.integer(substr(d, 6, 7))
        nxt <- if (mo == 12) paste0(yr+1,"-01-01") else paste0(yr,"-",sprintf("%02d",mo+1),"-01")
        as.character(as.Date(nxt) - 1)
      } else paste0(d, "-01")
    } else d
  })
}

convert_dtc_to_dt <- function(dtc, date_imputation = "first") {
  suppressWarnings(as.numeric(as.Date(.impute_dtc(dtc, date_imputation))))
}

derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL,
                                order = NULL, filter_add = NULL, mode = NULL, ...) {
  by_nm   <- sapply(by_vars, deparse)
  join_df <- dataset_add
  if (!is.null(filter_add)) {
    fe <- substitute(filter_add)
    join_df <- tryCatch(filter(join_df, !!fe), error = function(e) join_df)
  }
  if (!is.null(new_vars)) {
    old_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d, "=")[[1]][2]) else d
    })
    new_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d, "=")[[1]][1]) else d
    })
    keep <- intersect(c(by_nm, old_nms), names(join_df))
    join_df <- join_df |> select(all_of(keep))
    for (i in seq_along(old_nms)) {
      if (old_nms[i] != new_nms[i] && old_nms[i] %in% names(join_df))
        join_df <- join_df |> rename(!!new_nms[i] := !!old_nms[i])
    }
  }
  by_present <- intersect(by_nm, names(join_df))
  left_join(dataset, join_df, by = by_present)
}

derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation = "none", highest_imputation = "n",
                            flag_imputation = "auto", ...) {
  dtc_nm  <- deparse(substitute(dtc))
  imp_rule <- if (date_imputation %in% c("none","") && highest_imputation == "n") "none"
              else if (date_imputation == "last") "last" else "first"
  raw     <- if (dtc_nm %in% names(dataset)) dataset[[dtc_nm]] else rep(NA_character_, nrow(dataset))
  imputed <- if (imp_rule == "none") raw else .impute_dtc(raw, imp_rule)
  dataset[[paste0(new_vars_prefix, "DT")]] <- suppressWarnings(as.numeric(as.Date(imputed)))
  if (flag_imputation != "none") {
    dataset[[paste0(new_vars_prefix, "DTF")]] <- case_when(
      is.na(raw)                       ~ NA_character_,
      nchar(raw) <= 4 & !is.na(imputed) ~ "Y",
      nchar(raw) <= 7 & !is.na(imputed) ~ "M",
      TRUE                              ~ NA_character_
    )
  }
  dataset
}

derive_vars_dtm <- function(dataset, dtc, new_vars_prefix,
                             date_imputation = "first", time_imputation = "first",
                             highest_imputation = "n", ...) {
  dtc_nm  <- deparse(substitute(dtc))
  raw     <- if (dtc_nm %in% names(dataset)) dataset[[dtc_nm]] else rep(NA_character_, nrow(dataset))
  imputed <- .impute_dtc(raw, date_imputation)
  suffix  <- if (time_imputation == "last") "T23:59:59" else "T00:00:00"
  dataset[[paste0(new_vars_prefix, "DTM")]] <- suppressWarnings(
    as.numeric(as.POSIXct(paste0(imputed, suffix), tz = "UTC"))
  )
  dataset
}

derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref_nm  <- deparse(substitute(reference_date))
  src_nms <- sapply(source_vars, deparse)
  for (sv in src_nms) {
    dy_nm <- sub("DT$", "DY", sv)
    if (sv %in% names(dataset) && ref_nm %in% names(dataset)) {
      v <- dataset[[sv]]; r <- dataset[[ref_nm]]
      dataset[[dy_nm]] <- ifelse(is.na(v) | is.na(r), NA_real_,
        ifelse(v >= r, as.numeric(v - r) + 1, as.numeric(v - r)))
    }
  }
  dataset
}

derive_var_extreme_flag <- function(dataset, by_vars, order, new_var, mode = "first") {
  # Accept bare name (direct call) or string (via do.call from params())
  .e <- substitute(new_var); new_nm <- if (is.character(.e)) .e else deparse(.e)
  by_nm   <- sapply(by_vars, deparse)
  ord_nms <- sapply(order, function(x) { d <- deparse(x); gsub("desc\\(|\\)","",d) })
  desc_fl <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
  ord_df  <- dataset
  for (i in rev(seq_along(ord_nms))) {
    if (ord_nms[i] %in% names(ord_df))
      ord_df <- if (desc_fl[i]) arrange(ord_df, desc(.data[[ord_nms[i]]]))
                else arrange(ord_df, .data[[ord_nms[i]]])
  }
  by_present <- intersect(by_nm, names(ord_df))
  if (mode == "first")
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(1) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)][1], new_nm)))
  else
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(n()) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)][1], new_nm)))
  if (new_nm %in% names(dataset)) dataset[[new_nm]] <- NULL
  left_join(dataset, flagged, by = intersect(names(flagged), names(dataset)))
}

restrict_derivation <- function(dataset, derivation, args = list(), filter = NULL) {
  fe <- substitute(filter)
  in_rows <- if (!is.null(fe)) {
    tryCatch(which(eval(fe, dataset, parent.frame())), error = function(e) seq_len(nrow(dataset)))
  } else seq_len(nrow(dataset))
  out_rows <- setdiff(seq_len(nrow(dataset)), in_rows)
  sub_data  <- dataset[in_rows, , drop = FALSE]
  all_args  <- c(list(dataset = sub_data), args)
  derived   <- tryCatch(do.call(derivation, all_args), error = function(e) sub_data)
  bind_rows(derived, dataset[out_rows, , drop = FALSE])
}

params <- function(...) {
  # Bare names (e.g. new_var = AOCCFL) are captured as strings.
  # Everything else (exprs() calls, strings, numbers) is evaluated normally.
  parent_env <- parent.frame()
  dots <- as.list(substitute(list(...)))[-1]
  lapply(dots, function(x) {
    if (is.symbol(x)) deparse(x) else eval(x, envir = parent_env)
  })
}

# ── OCCDS-specific mocks ─────────────────────────────────────────────────────
derive_var_trtemfl <- function(dataset, start_date, end_date = NULL,
                                trt_start_date, trt_end_date,
                                end_window = 0,
                                initial_intensity = NULL,
                                intensity = NULL, ...) {
  sd_nm  <- deparse(substitute(start_date))
  ed_nm  <- if (!is.null(substitute(end_date))) deparse(substitute(end_date)) else NULL
  ts_nm  <- deparse(substitute(trt_start_date))
  te_nm  <- deparse(substitute(trt_end_date))
  ii_nm  <- if (!is.null(substitute(initial_intensity))) deparse(substitute(initial_intensity)) else NULL
  in_nm  <- if (!is.null(substitute(intensity)))         deparse(substitute(intensity))         else NULL
  dataset |> mutate(
    TRTEMFL = case_when(
      is.na(.data[[sd_nm]]) | is.na(.data[[ts_nm]]) ~ NA_character_,
      .data[[sd_nm]] >= .data[[ts_nm]] &
        .data[[sd_nm]] <= (.data[[te_nm]] + end_window) ~ "Y",
      TRUE ~ NA_character_
    )
  )
}

derive_var_ontrtfl <- function(dataset, start_date, end_date = NULL,
                                ref_start_date, ref_end_date,
                                ref_end_window = 0, ...) {
  sd_nm <- deparse(substitute(start_date))
  rs_nm <- deparse(substitute(ref_start_date))
  re_nm <- deparse(substitute(ref_end_date))
  dataset |> mutate(
    ONTRTFL = case_when(
      is.na(.data[[sd_nm]]) | is.na(.data[[rs_nm]]) ~ NA_character_,
      .data[[sd_nm]] >= .data[[rs_nm]] &
        .data[[sd_nm]] <= (.data[[re_nm]] + ref_end_window) ~ "Y",
      TRUE ~ NA_character_
    )
  )
}

derive_var_obs_number <- function(dataset, new_var, by_vars, order,
                                   check_type = "warning") {
  new_nm  <- deparse(substitute(new_var))
  by_nm   <- sapply(by_vars, deparse)
  ord_nms <- sapply(order, function(x) { d <- deparse(x); gsub("desc\\(|\\)","",d) })
  desc_fl <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
  ord_df  <- dataset
  for (i in rev(seq_along(ord_nms))) {
    if (ord_nms[i] %in% names(ord_df))
      ord_df <- if (desc_fl[i]) arrange(ord_df, desc(.data[[ord_nms[i]]]))
                else arrange(ord_df, .data[[ord_nms[i]]])
  }
  by_present <- intersect(by_nm, names(ord_df))
  ord_df |> group_by(across(all_of(by_present))) |>
    mutate(!!new_nm := row_number()) |> ungroup()
}

derive_vars_joined <- function(dataset, dataset_add, by_vars, order,
                                new_vars = NULL, join_vars = NULL,
                                filter_join = NULL, mode = "first",
                                join_type = "all", ...) {
  by_nm  <- sapply(by_vars, deparse)
  fj_sub <- substitute(filter_join)

  # Tag each original row so we can select best dataset_add row per original row
  dataset$.ROW_ID <- seq_len(nrow(dataset))

  merged <- left_join(dataset, dataset_add,
                      by = intersect(by_nm, names(dataset_add)),
                      suffix = c("", ".add"))
  if (!is.null(fj_sub)) {
    merged <- tryCatch(filter(merged, !!fj_sub), error = function(e) merged)
  }

  # Sort by order before slicing
  ord_nms <- sapply(order, function(x) { d <- deparse(x); gsub("desc\\(|\\)","",d) })
  desc_fl <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
  for (i in rev(seq_along(ord_nms))) {
    if (ord_nms[i] %in% names(merged))
      merged <- if (desc_fl[i]) arrange(merged, desc(.data[[ord_nms[i]]]))
                else arrange(merged, .data[[ord_nms[i]]])
  }

  # Select first or last dataset_add match per original dataset row
  best <- if (mode == "first")
    merged |> group_by(.ROW_ID) |> slice(1) |> ungroup()
  else
    merged |> group_by(.ROW_ID) |> slice(n()) |> ungroup()

  # Rename new_vars — names(new_vars) holds the NEW name; deparse(value) holds the OLD name
  if (!is.null(new_vars)) {
    old_nms <- unname(sapply(new_vars, deparse))
    raw_nms <- names(new_vars)
    new_nms <- ifelse(!is.null(raw_nms) & raw_nms != "", raw_nms, old_nms)
    for (i in seq_along(old_nms)) {
      if (old_nms[i] %in% names(best))
        best <- rename(best, !!new_nms[i] := !!old_nms[i])
    }
    keep_extra <- new_nms
  } else {
    keep_extra <- setdiff(names(dataset_add), c(by_nm, names(dataset)))
  }

  keep_cols <- intersect(keep_extra, names(best))
  extra     <- best |> select(all_of(c(".ROW_ID", keep_cols)))

  if (any(keep_cols %in% names(dataset))) {
    for (nm in keep_cols[keep_cols %in% names(dataset)]) dataset[[nm]] <- NULL
  }
  result          <- left_join(dataset, extra, by = ".ROW_ID")
  result$.ROW_ID  <- NULL
  result
}

derive_vars_query <- function(dataset, dataset_queries) {
  prefixes <- unique(dataset_queries$PREFIX)
  for (px in prefixes) {
    grp_rows <- dataset_queries |> filter(PREFIX == px)
    grp_name <- grp_rows$GRPNAME[1]
    terms    <- grp_rows$TERM_NAME
    dataset <- dataset |> mutate(
      !!paste0(px, "NAM") := case_when(
        toupper(AEDECOD) %in% toupper(terms) ~ grp_name,
        TRUE ~ NA_character_
      )
    )
  }
  dataset
}

create_single_dose_dataset <- function(dataset, dose_freq, start_date,
                                        end_date, keep_source_vars = NULL) {
  sf_nm <- deparse(substitute(dose_freq))
  sd_nm <- deparse(substitute(start_date))
  ed_nm <- deparse(substitute(end_date))
  result <- vector("list", nrow(dataset))
  for (i in seq_len(nrow(dataset))) {
    row  <- dataset[i, , drop = FALSE]
    freq <- row[[sf_nm]]
    sd   <- row[[sd_nm]]; ed <- row[[ed_nm]]
    if (is.na(sd) || is.na(ed)) { result[[i]] <- row; next }
    sd_d <- as.Date(sd, origin = "1970-01-01")
    ed_d <- as.Date(ed, origin = "1970-01-01")
    if (grepl("^(QD|QD1|ONCE DAILY)$", toupper(freq %||% ""))) {
      dates <- seq(sd_d, ed_d, by = "day")
      row_rep <- row[rep(1, length(dates)), ]
      row_rep[[sd_nm]] <- as.numeric(dates)
      result[[i]] <- row_rep
    } else {
      result[[i]] <- row
    }
  }
  bind_rows(result)
}

# fallback for %||%
`%||%` <- function(a, b) if (is.null(a)) b else a

# ── Sample datasets ───────────────────────────────────────────────────────────
adsl_occ <- tribble(
  ~STUDYID,        ~USUBJID, ~TRTSDT, ~TRTEDT, ~TRT01A,    ~TRT01P,    ~AGE, ~SEX, ~RACE, ~SAFFL,
  "CDISCPILOT01", "01-001",  19367,   19548,   "Xanomeli", "Xanomeli",  52, "F", "WHITE",    "Y",
  "CDISCPILOT01", "01-002",  19370,   19551,   "Placebo",  "Placebo",   61, "M", "WHITE",    "Y",
  "CDISCPILOT01", "01-003",  19375,   19540,   "Xanomeli", "Xanomeli",  47, "F", "BLACK",    "Y"
)

ae <- tribble(
  ~STUDYID,        ~USUBJID,  ~AESEQ, ~AETERM,            ~AEDECOD,           ~AESOC,
  ~AESTDTC,     ~AEENDTC,     ~AESEV,    ~AEREL,       ~AESER, ~AETOXGR,
  "CDISCPILOT01","01-001", 1, "HEADACHE",         "Headache",          "Nervous system disorders",
  "2024-01-20", "2024-01-22", "MILD",    "UNLIKELY",   "N",    1,
  "CDISCPILOT01","01-001", 2, "NAUSEA",           "Nausea",            "Gastrointestinal disorders",
  "2024-02-01", "2024-02-05", "MODERATE","POSSIBLE",   "N",    2,
  "CDISCPILOT01","01-002", 1, "DIZZINESS",        "Dizziness",         "Nervous system disorders",
  "2024-01-25", "2024-01-28", "MILD",    "UNLIKELY",   "N",    1,
  "CDISCPILOT01","01-002", 2, "ANAPHYLACTIC SHOCK","Anaphylactic shock","Immune system disorders",
  "2024-02-10", NA,           "SEVERE",  "PROBABLE",   "Y",    3,
  "CDISCPILOT01","01-003", 1, "RASH",             "Rash",              "Skin disorders",
  "2024-01-30", "2024-02-03", "MILD",    "POSSIBLE",   "N",    1
)

ex_occ <- tribble(
  ~STUDYID,        ~USUBJID,  ~EXSEQ, ~EXTRT,      ~EXDOSE, ~EXDOSFRQ, ~EXSTDTC,      ~EXENDTC,
  "CDISCPILOT01", "01-001",   1,      "XANOMELI",   54,     "QD",      "2024-01-15",  "2024-02-15",
  "CDISCPILOT01", "01-002",   1,      "PLACEBO",     0,     "QD",      "2024-01-18",  "2024-02-18",
  "CDISCPILOT01", "01-003",   1,      "XANOMELI",   54,     "QD",      "2024-01-23",  "2024-02-20"
) |> mutate(
  EXSTDT = as.numeric(as.Date(EXSTDTC)),
  EXENDT = as.numeric(as.Date(EXENDTC))
)

queries_df <- tribble(
  ~PREFIX, ~GRPNAME,                    ~SCOPE,   ~SCOPEN, ~TERM_NAME,
  "SMQ01", "Hypersensitivity (SMQ)",   "NARROW", 2,       "ANAPHYLACTIC SHOCK",
  "SMQ01", "Hypersensitivity (SMQ)",   "NARROW", 2,       "RASH",
  "CQ02",  "GI Disorders (Customized)", NA,      NA,      "NAUSEA",
  "CQ02",  "GI Disorders (Customized)", NA,      NA,      "VOMITING"
)

# ── Real pharmaverseadam datasets (for exploration and validation exercises) ───
library(pharmaverseadam)
adae_real  <- pharmaverseadam::adae
adsl_real  <- pharmaverseadam::adsl
```

## Welcome

This lesson covers **OCCDS (Occurrence Data Structure)** ADaMs — ADAE (Adverse Events) and ADCM (Concomitant Medications). You'll build a complete ADAE step by step, learning the treatment-emergent flag, occurrence flags, MedDRA queries, and last-dose linkage.

---

## 1 — OCCDS at a glance

### 1.1 What OCCDS is

OCCDS = **Occurrence Data Structure**. It is the ADaM class designed to hold *events* — discrete things that happen to a subject during a study. The flagship OCCDS dataset is **ADAE** (Adverse Events). The other OCCDS members are:

| Dataset | What it stores            | SDTM source |
|---------|---------------------------|-------------|
| ADAE    | Adverse events            | AE          |
| ADCM    | Concomitant medications   | CM          |
| ADMH    | Medical history           | MH          |
| ADDV    | Protocol deviations       | DV          |
| ADEX    | Exposure (when modelled as events) | EX |

Each row in an OCCDS dataset is **one event for one subject** — one AE record, one medication record. There is **no PARAMCD** because the events are heterogeneous: "headache" and "nausea" are not the same parameter, they are different terms from a coded medical dictionary.

### 1.2 OCCDS vs BDS — the structural difference

This is the single most important distinction in ADaM, and the one that trips up SAS programmers first.

```
BDS  (ADLB, ADVS, ADTTE)              OCCDS (ADAE, ADCM, ADMH)
------------------------              ------------------------
one row per                           one row per
  subject x parameter x visit           subject x event
ALT visit 1                            Subject's headache
ALT visit 2                            Subject's nausea
AST visit 1                            Subject's rash
PARAMCD column                         AEDECOD / CMDECOD column
AVAL column (numeric)                  no AVAL; categorical attributes
analysis = baseline, change            analysis = incidence, counts
```

The mental model: **BDS rows are observations; OCCDS rows are happenings**. A lab value is observed; an AE happens.

### 1.3 The ADAE pipeline at a glance

```
  SDTM AE  -->  merge ADSL  -->  derive ASTDT/AENDT  -->  severity (ASEVN)
                                                              |
                                                              v
                                                        derive TRTEMFL
                                                              |
                                                              v
                                                  derive AOCCFL / AOCCSFL
                                                              |
                                                              v
                                                derive MedDRA queries (SMQ)
                                                              |
                                                              v
                                                  link last dose (LDOSEDT)
                                                              |
                                                              v
                                                       assign ASEQ
                                                              |
                                                              v
                                                       ADAE (final)
```

Each downstream step depends on the variables produced by the prior step. TRTEMFL depends on ASTDT and TRTSDT. AOCCFL depends on TRTEMFL. SMQ queries depend on AEDECOD. ASEQ is computed last because it is a sequence number over the analysis-sorted rows.

### 1.4 Incidence vs occurrence — a semantic distinction

SAS programmers often hear "incidence" and "occurrence" used interchangeably. They are **not** the same in OCCDS:

- **Occurrence**: every AE record on the dataset — "subject 01-001 had 3 AEs."
- **Incidence**: the count of subjects with **at least one** event of a given type — "12 of 86 placebo subjects had nausea."

The occurrence flags (AOCCFL, AOCCSFL, AOCCPFL) exist precisely to support **incidence** analyses: by flagging only the *first* event per subject per stratum, you can count subjects without double-counting.

### 1.5 SAS analog table for the OCCDS pipeline

| Admiral function                | What it does                              | Closest SAS pattern                              |
|---------------------------------|-------------------------------------------|--------------------------------------------------|
| `derive_vars_merged`            | Bring ADSL variables onto AE              | `PROC SQL LEFT JOIN adsl ON USUBJID`             |
| `derive_vars_dt`                | Impute partial dates to numeric ASTDT     | `INPUT(AESTDTC, YYMMDD10.)` + DATA-step IF logic |
| `derive_vars_dy`                | Compute study day (Day 1 convention)      | `ASTDY = ASTDT - TRTSDT + (ASTDT >= TRTSDT)`     |
| `mutate(ASEVN = case_when(...))`| Map MILD/MODERATE/SEVERE to 1/2/3         | `PROC FORMAT VALUE ASEVN_F`                      |
| `derive_var_trtemfl`            | Treatment-emergent flag                   | DATA step `IF ASTDT >= TRTSDT AND ASTDT <= TRTEDT THEN TRTEMFL = "Y";` |
| `derive_var_extreme_flag` (AOCCFL) | First TEAE per subject                | `PROC SORT NODUPKEY BY USUBJID;` + first.usubjid logic |
| `derive_vars_query`             | Tag SMQ / CQ membership                   | `PROC SQL LEFT JOIN smq_ref ON AEDECOD = SMQ_TERM` |
| `derive_vars_joined`            | Link last dose prior to AE                | `PROC SQL` correlated subquery (`MAX(EXSTDT) WHERE EXSTDT <= ASTDT`) |
| `create_single_dose_dataset`    | Expand QD record to one row per day       | `DATA step DO i = sd TO ed BY 1; OUTPUT;`         |
| `derive_var_obs_number` (ASEQ)  | Per-subject row counter                   | `PROC SQL ROW_NUMBER() OVER (PARTITION BY USUBJID)` |

> **SAS analog mental model:** ADAE is what you would build with about 8 SAS PROC/DATA steps glued together by `RUN; QUIT;` boundaries. Admiral collapses those into a single piped pipeline where each step's output is the next step's input — and each step is named after the *variable it produces*, not the *engine that produces it*.

### 1.6 The OCCDS-class shape

```{r occds-shape-demo}
# OCCDS = one row per event (AE, CM, MH, DV)
# No PARAMCD - events are heterogeneous
ae |> select(USUBJID, AESEQ, AEDECOD, AESOC, AESTDTC, AEENDTC, AESEV, AESER)
```

> **Read the output:** every row is one AE. Subject 01-001 has two rows (headache, nausea); subject 01-002 has two rows (dizziness, anaphylactic shock). There is no `PARAMCD` column, and there is no `AVAL`. The "value" of an event is its name (`AEDECOD`), its severity (`AESEV`), and the surrounding categorical attributes.

```{r quiz-occds-shape, echo=FALSE}
quiz(
  question(
    "Which ADaM uses the OCCDS structure?",
    answer("ADVS"),
    answer("ADLB"),
    answer("ADAE", correct = TRUE),
    answer("ADTTE"),
    allow_retry = TRUE
  ),
  question(
    "OCCDS rows represent:",
    answer("One measurement per visit"),
    answer("One time-to-event per subject"),
    answer("One event or occurrence per subject record", correct = TRUE),
    answer("One parameter per study"),
    allow_retry = TRUE
  ),
  question(
    "Why does ADAE have no PARAMCD column?",
    answer("PARAMCD is optional in OCCDS"),
    answer("Each row is a heterogeneous event identified by its coded term (AEDECOD), not a measurement of a single parameter", correct = TRUE),
    answer("PARAMCD is stored under a different name (AETERM)"),
    answer("Only BDS datasets require PARAMCD"),
    allow_retry = TRUE
  ),
  question(
    "Which best describes the distinction between 'occurrence' and 'incidence' in ADAE?",
    answer("They are synonymous"),
    answer("Occurrence = every AE record; incidence = count of unique subjects with at least one event of a given type", correct = TRUE),
    answer("Occurrence = serious AEs only; incidence = all AEs"),
    answer("Occurrence = pre-treatment AEs; incidence = treatment-emergent AEs"),
    allow_retry = TRUE
  )
)
```

### 1.7 Exercise: Identify OCCDS vs BDS shape

```{r occds-shape-ex, exercise=TRUE, exercise.lines=14}
# Two test datasets are below. For each, decide if it is OCCDS or BDS shape
# by counting rows per (USUBJID, PARAMCD-or-AEDECOD).
# Print the row count for subject "01-001" in `ae`:

n_rows_001 <- ae |>
  filter(USUBJID == "01-001") |>
  ___()

cat("Rows for 01-001 in ae:", n_rows_001, "\n")
cat("Structure: ", ifelse(n_rows_001 > 1, "OCCDS-shaped", "BDS-shaped"), "\n")
```

```{r occds-shape-ex-hint-1}
# nrow() counts rows in a data frame
```

```{r occds-shape-ex-hint-2}
# n_rows_001 <- ae |> filter(USUBJID == "01-001") |> nrow()
```

```{r occds-shape-ex-solution}
n_rows_001 <- ae |>
  filter(USUBJID == "01-001") |>
  nrow()

cat("Rows for 01-001 in ae:", n_rows_001, "\n")
cat("Structure: ", ifelse(n_rows_001 > 1, "OCCDS-shaped", "BDS-shaped"), "\n")
```

---

## 2 — Merge ADSL and derive analysis dates

### 2.1 Why this step is first

Every downstream derivation in ADAE depends on **subject-level variables** that live in ADSL: `TRTSDT`, `TRTEDT`, `TRT01A`, `SAFFL`, `AGE`, `SEX`, `RACE`. TRTEMFL needs TRTSDT/TRTEDT; subgroup analyses need AGE/SEX/RACE. The first move in any ADAE build is to attach ADSL.

### 2.2 The "SAFFL only" submission rule

Per the standard ADaM IG convention, **ADAE typically contains only Safety Population subjects** (`SAFFL == "Y"`). Subjects who were randomized but never dosed are usually removed by an upstream filter or by the join logic.

SAS analog:

```
PROC SQL;
   CREATE TABLE adae AS
     SELECT a.*, b.TRTSDT, b.TRTEDT, b.TRT01A, b.SAFFL
       FROM ae    AS a
       LEFT JOIN adsl AS b
         ON a.STUDYID = b.STUDYID AND a.USUBJID = b.USUBJID
       WHERE b.SAFFL = 'Y';
QUIT;
```

R/admiral analog: `derive_vars_merged()` + a downstream `filter(SAFFL == "Y")`.

### 2.3 Anatomy of derive_vars_merged()

```r
derive_vars_merged(
  dataset     = ae,                              # primary (events)
  dataset_add = adsl_occ,                        # source of new variables
  new_vars    = exprs(TRTSDT, TRTEDT, TRT01A),   # columns to bring across
  by_vars     = exprs(STUDYID, USUBJID)          # join keys
)
```

Read it as: *"Onto `ae`, attach the columns `new_vars` from `adsl_occ`, joining by `by_vars`."* The join is implicitly a **left join from `dataset` to `dataset_add`** — every AE row is preserved.

> **SAS analog mental model:** this is the admiral equivalent of `PROC SQL LEFT JOIN` where you list the columns you want to bring across in `SELECT`. The difference is that `new_vars` is *declarative* — you do not write the `ON` clause; you supply `by_vars`.

### 2.4 Date imputation strategy

SDTM dates are character (`AESTDTC`). Analysis dates must be numeric R Dates. SDTM dates can also be **partial** — `"2024-07"` (no day) or `"2024"` (no month or day). Admiral imputes these to a deterministic boundary:

- `date_imputation = "first"`: `"2024"` → `"2024-01-01"`, `"2024-07"` → `"2024-07-01"`. Conservative for **start dates** (push earlier → narrows the treatment-emergent window).
- `date_imputation = "last"`:  `"2024"` → `"2024-12-31"`, `"2024-07"` → `"2024-07-31"`. Conservative for **end dates** (push later → assumes the AE persisted as long as possible).

SAS analog:

```
DATA adae;
   SET ae;
   IF LENGTH(AESTDTC) = 4 THEN ASTDT = INPUT(AESTDTC || "-01-01", YYMMDD10.);
   ELSE IF LENGTH(AESTDTC) = 7 THEN ASTDT = INPUT(AESTDTC || "-01", YYMMDD10.);
   ELSE ASTDT = INPUT(AESTDTC, YYMMDD10.);
RUN;
```

### 2.5 The ASTDTF imputation flag

Whenever a date was imputed, admiral sets an **imputation flag** (`ASTDTF`) so reviewers can trace it:

| `ASTDTF` value | Meaning                                       |
|----------------|-----------------------------------------------|
| `"Y"`          | Year was the highest non-missing piece — imputed M and D |
| `"M"`          | Year-month present — imputed D only           |
| `NA` (blank)   | Full date present — no imputation             |

P21 and reviewers expect this flag for traceability. Never drop it.

### 2.6 ASTDY — the CDISC Day 1 convention

Study day in ADaM follows a specific rule: **TRTSDT is Day 1; there is no Day 0**.

```
   DAY:  -3   -2   -1    1    2    3    4    5  ...
              ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^
                         |
                         TRTSDT (first dose date)

   Formula:
     positive days   ASTDY = ASTDT - TRTSDT + 1
     negative days   ASTDY = ASTDT - TRTSDT      (no Day 0)
```

`derive_vars_dy()` implements this convention automatically — never roll your own arithmetic for study day.

### 2.7 The standard merge + date pattern

```{r adae-dates-demo}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, AGE, SEX, RACE, SAFFL)

adae <- ae |>
  derive_vars_merged(
    dataset_add = adsl_occ,
    new_vars    = adsl_vars,
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  # Start date: partial imputed to first
  derive_vars_dt(
    new_vars_prefix  = "AST",
    dtc              = AESTDTC,
    highest_imputation = "M",
    date_imputation  = "first",
    flag_imputation  = "auto"
  ) |>
  # End date: partial imputed to last (conservative)
  derive_vars_dt(
    new_vars_prefix  = "AEN",
    dtc              = AEENDTC,
    highest_imputation = "M",
    date_imputation  = "last"
  ) |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ASTDT, AENDT)
  )

adae |> select(USUBJID, AETERM, AESTDTC, ASTDT, AENDT, ASTDY) |> head(4)
```

```{r quiz-imputation, echo=FALSE}
quiz(
  question(
    "Why is `date_imputation = 'last'` appropriate for end dates?",
    answer("It makes AVAL larger"),
    answer("A partial end date like '2024-07' means the AE continued through the month — imputing to the last day is the conservative choice", correct = TRUE),
    answer("Admiral requires last for end dates"),
    answer("It matches start date imputation"),
    allow_retry = TRUE
  )
)
```

```{r dates-ex, exercise=TRUE, exercise.lines=18}
# Exercise: starting from ae, merge adsl_occ and derive ASTDT / AENDT
# then compute ASTDY = days from TRTSDT to ASTDT (with +1 convention)
# Confirm the ASTDY for subject 01-001's first AE

adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, AGE, SEX, SAFFL)

adae_tmp <- ae |>
  derive_vars_merged(dataset_add = adsl_occ, new_vars = adsl_vars,
                     by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_dt(new_vars_prefix = "AST", dtc = AESTDTC,
                 highest_imputation = "M", date_imputation = "first") |>
  derive_vars_dt(new_vars_prefix = "AEN", dtc = AEENDTC,
                 highest_imputation = "M", date_imputation = "last") |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ASTDT, AENDT))

adae_tmp |> filter(USUBJID == "01-001") |>
  select(USUBJID, AETERM, ASTDTC = AESTDTC, ASTDT, ASTDY)
```

```{r dates-ex-hint-1}
# derive_vars_dy reads TRTSDT and ASTDT automatically when you pass source_vars = exprs(ASTDT)
```

```{r dates-ex-solution}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, AGE, SEX, SAFFL)

adae_tmp <- ae |>
  derive_vars_merged(dataset_add = adsl_occ, new_vars = adsl_vars,
                     by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_dt(new_vars_prefix = "AST", dtc = AESTDTC,
                 highest_imputation = "M", date_imputation = "first") |>
  derive_vars_dt(new_vars_prefix = "AEN", dtc = AEENDTC,
                 highest_imputation = "M", date_imputation = "last") |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ASTDT, AENDT))

adae_tmp |> filter(USUBJID == "01-001") |>
  select(USUBJID, AETERM, ASTDTC = AESTDTC, ASTDT, ASTDY)
```

### 2.8 Knowledge check — merge and dates

```{r quiz-merge-dates, echo=FALSE}
quiz(
  question(
    "Why does ADAE normally include only subjects with SAFFL = 'Y'?",
    answer("ADAE is the Safety analysis dataset — randomized-but-not-dosed subjects are excluded from AE incidence", correct = TRUE),
    answer("SAFFL is an admiral requirement"),
    answer("It is required by P21"),
    answer("Non-safety subjects have no AEs by definition"),
    allow_retry = TRUE
  ),
  question(
    "A partial start date 2024-07 should be imputed to:",
    answer("2024-07-31"),
    answer("2024-07-15"),
    answer("2024-07-01 (date_imputation = 'first' — conservative push earlier)", correct = TRUE),
    answer("Cannot be imputed without month"),
    allow_retry = TRUE
  ),
  question(
    "ASTDY = 0 in ADAE is:",
    answer("Valid — represents the dosing date"),
    answer("Invalid — CDISC convention has no Day 0; first dose = Day 1 (use derive_vars_dy())", correct = TRUE),
    answer("Allowed only for screening AEs"),
    answer("Used to represent missing dates"),
    allow_retry = TRUE
  )
)
```

---

## 3 — Severity mapping

### 3.1 Why we need ASEVN (numeric severity)

`AESEV` from SDTM is character: `"MILD"`, `"MODERATE"`, `"SEVERE"`. Analysis tools (max-severity calculations, worsening flags, sort orders) need a **numeric** companion. ADaM names this `ASEVN`.

```
SDTM      AESEV   = "MILD"     "MODERATE"   "SEVERE"
                     |            |            |
                     v            v            v
ADAE      ASEVN   =  1            2            3
```

This is the exact pattern of a SAS `PROC FORMAT` `VALUE` statement applied via a `PUT()`:

```
PROC FORMAT;
   VALUE $SEVN_F
      "MILD"     = 1
      "MODERATE" = 2
      "SEVERE"   = 3;
RUN;

DATA adae;
   SET adae;
   ASEVN = INPUT(PUT(AESEV, $SEVN_F.), BEST.);
RUN;
```

R/admiral does the same with `dplyr::case_when()`.

### 3.2 ASEV vs AESEV — the copy-and-rename pattern

Many ADaM datasets create an `ASEV` column that simply copies `AESEV`. Why duplicate?

- `AESEV` is the **SDTM** source (must remain unmodified for traceability).
- `ASEV` is the **analysis** version — open to programmatic modification (e.g., re-coding if external grading dictionaries disagree).

The same pattern applies to AEREL → AREL.

### 3.3 Anatomy of the severity mutate

```r
mutate(
  ASEV  = AESEV,                       # 1:1 copy for analysis use
  AREL  = AEREL,                       # 1:1 copy for analysis use
  ASEVN = case_when(                   # numeric companion
    ASEV == "MILD"     ~ 1,
    ASEV == "MODERATE" ~ 2,
    ASEV == "SEVERE"   ~ 3,
    TRUE               ~ NA_real_      # safety net
  )
)
```

> **SAS analog mental model:** `ASEV = AESEV` is a SAS assignment statement; `ASEVN = case_when(...)` is the `PUT()` of a `$VALUE` format combined with `INPUT()` back to numeric. The `TRUE ~ NA_real_` is the SAS `OTHERWISE` arm of a `SELECT` block.

### 3.4 CTCAE grading (cross-reference)

For oncology trials, severity uses the **NCI CTCAE** 5-grade scale instead of MILD/MODERATE/SEVERE. `AETOXGR` (toxicity grade 1-5) carries this — admiralonco extensions (see lesson 20) provide specific helpers. The high-level mapping:

| CTCAE Grade | Meaning                              |
|-------------|--------------------------------------|
| 1           | Mild — asymptomatic                  |
| 2           | Moderate — minimal intervention      |
| 3           | Severe — hospitalisation possible    |
| 4           | Life-threatening                     |
| 5           | Death related to AE                  |

In ADAE.AETOXGR sits alongside AESEV; analysis tables may use either.

### 3.5 The severity derivation

```{r severity-demo}
adae <- adae |>
  mutate(
    ASEV  = AESEV,
    AREL  = AEREL,
    ASEVN = case_when(
      ASEV == "MILD"     ~ 1,
      ASEV == "MODERATE" ~ 2,
      ASEV == "SEVERE"   ~ 3,
      TRUE               ~ NA_real_
    )
  )

adae |> select(USUBJID, AETERM, ASEV, ASEVN, AREL)
```

> **Read the output:** `ASEV` is the analysis copy of `AESEV`; `ASEVN` is the numeric companion. A row with `ASEV = "SEVERE"` should have `ASEVN = 3` — verify by eye on the anaphylactic shock row.

### 3.6 Exercise: Add a relationship-numeric mapping

```{r areln-ex, exercise=TRUE, exercise.lines=18}
# Add ARELN: 1 = UNLIKELY, 2 = POSSIBLE, 3 = PROBABLE, 4 = RELATED
# Use case_when() and NA_real_ as the fallback.

adae_arel <- adae |>
  mutate(
    ARELN = case_when(
      AREL == "UNLIKELY" ~ ___,
      AREL == "POSSIBLE" ~ ___,
      AREL == "PROBABLE" ~ ___,
      AREL == "RELATED"  ~ 4,
      TRUE               ~ ___
    )
  )

adae_arel |> select(USUBJID, AETERM, AREL, ARELN)
```

```{r areln-ex-hint-1}
# 1, 2, 3 in order; fallback is NA_real_
```

```{r areln-ex-hint-2}
# AREL == "UNLIKELY" ~ 1
# AREL == "POSSIBLE" ~ 2
# AREL == "PROBABLE" ~ 3
# TRUE               ~ NA_real_
```

```{r areln-ex-solution}
adae_arel <- adae |>
  mutate(
    ARELN = case_when(
      AREL == "UNLIKELY" ~ 1,
      AREL == "POSSIBLE" ~ 2,
      AREL == "PROBABLE" ~ 3,
      AREL == "RELATED"  ~ 4,
      TRUE               ~ NA_real_
    )
  )

adae_arel |> select(USUBJID, AETERM, AREL, ARELN)
```

### 3.7 Knowledge check — severity

```{r quiz-severity, echo=FALSE}
quiz(
  question(
    "Why is ASEV typically a 1:1 copy of AESEV?",
    answer("They have to differ"),
    answer("ASEV is the analysis-modifiable copy; AESEV remains the immutable SDTM source for traceability", correct = TRUE),
    answer("ASEV uses different categories"),
    answer("AESEV is a numeric variable"),
    allow_retry = TRUE
  ),
  question(
    "The SAS analog of `case_when(ASEV == 'MILD' ~ 1, ...)` is most directly:",
    answer("PROC SQL UNION ALL"),
    answer("PROC FORMAT VALUE statement + INPUT(PUT(...))", correct = TRUE),
    answer("PROC TRANSPOSE"),
    answer("A DATA step MERGE"),
    allow_retry = TRUE
  ),
  question(
    "AETOXGR for oncology trials uses which scale?",
    answer("MILD / MODERATE / SEVERE"),
    answer("NCI CTCAE grades 1 through 5", correct = TRUE),
    answer("WHO toxicity criteria 1 through 10"),
    answer("FDA narrative grading"),
    allow_retry = TRUE
  )
)
```

---

## 4 — Treatment-emergent flag: TRTEMFL

### 4.1 What "treatment-emergent" means clinically

A **treatment-emergent adverse event (TEAE)** is an AE that **emerged during or after exposure to study treatment**. The distinction matters because regulators and reviewers want to know which AEs can plausibly be attributed to drug exposure versus which already existed at baseline.

A subject who entered the study with chronic back pain and continued to report back pain throughout treatment is **not** experiencing a *new* event; that AE is **not** treatment-emergent. A subject who developed a rash three days after first dose **is** experiencing a TEAE.

```
   AE Onset Timeline:

   Screening    TRTSDT (Day 1)   TRTEDT       TRTEDT + 30 days
       |                |                  |              |
       |  Pre-treatment |  Treatment       |  Post-trt    |
       |  (not TEAE)    |  (TEAE = Y)      |  window      |
       |                |                  |  (TEAE = Y)  |
                                                          |
                                                          v
                                                       Beyond:
                                                       not TEAE
```

### 4.2 The date-comparison logic

```
TRTEMFL = "Y"   iff   ASTDT >= TRTSDT  AND  ASTDT <= TRTEDT + end_window
TRTEMFL = NA    otherwise (or when dates are NA)
```

Two conditions, both must hold. Either ASTDT or TRTSDT being missing collapses the result to `NA`.

### 4.3 Why we use AESTDT, not AEENDT

The TEAE definition is about **onset**, not resolution. An AE that *started* before TRTSDT but resolved during treatment is **not** new — it pre-existed. An AE that *started* during treatment but resolved a year after the study ended **is** a TEAE.

Always pass `start_date = ASTDT` (the **start** date). The `end_date = AENDT` argument is used only for the optional intensity-worsening pathway (Section 4.8).

### 4.4 Pre-treatment vs treatment-emergent vs post-treatment

```
   AE              Onset           Treatment            TEAE?
   --------------- --------------- ------------------- -----
   Headache        2024-01-10      TRTSDT 2024-01-15   No  (before TRTSDT)
   Nausea          2024-01-20      TRTSDT 2024-01-15   Yes (within window)
   Rash            2024-02-20      TRTEDT 2024-02-15   Yes (within end_window=30)
   Insomnia        2024-04-01      TRTEDT 2024-02-15   No  (beyond end_window=30)
```

### 4.5 Anatomy of derive_var_trtemfl()

```r
derive_var_trtemfl(
  dataset         = adae,
  start_date      = ASTDT,     # AE onset date (analysis numeric)
  end_date        = AENDT,     # AE resolution date (analysis numeric)
  trt_start_date  = TRTSDT,    # first dose date
  trt_end_date    = TRTEDT,    # last dose date
  end_window      = 30,        # days post-last-dose still considered TEAE
  initial_intensity = NULL,    # optional: pre-trt intensity (for worsening)
  intensity         = NULL     # optional: current intensity (for worsening)
)
```

Read it as: *"Onto `adae`, add a TRTEMFL column based on the comparison `start_date >= trt_start_date AND start_date <= trt_end_date + end_window`."*

> **SAS analog mental model:** this is one DATA step `IF` statement wrapped into a callable function. The SAS form is so brief that admiral's value-add seems thin — but the magic is the consistent handling of NA, the optional intensity-worsening pathway, and the audit trail.

### 4.6 SAS equivalent: the DATA step IF/THEN

```
DATA adae;
   SET adae;
   IF NOT MISSING(ASTDT) AND NOT MISSING(TRTSDT)
      AND ASTDT >= TRTSDT
      AND ASTDT <= TRTEDT + 30
      THEN TRTEMFL = "Y";
RUN;
```

Side by side:

```
   SAS DATA step                     |  admiral
   --------------------------------- |  ----------------------------------
   DATA adae;                        |  adae |>
     SET adae;                       |
     IF NOT MISSING(ASTDT)           |    derive_var_trtemfl(
        AND NOT MISSING(TRTSDT)      |      start_date     = ASTDT,
        AND ASTDT >= TRTSDT          |      end_date       = AENDT,
        AND ASTDT <= TRTEDT + 30     |      trt_start_date = TRTSDT,
        THEN TRTEMFL = "Y";          |      trt_end_date   = TRTEDT,
   RUN;                              |      end_window     = 30
                                     |    )
```

### 4.7 Common pitfalls

1. **Wrong start date** — passing `start_date = AENDT` (resolution) instead of `ASTDT` (onset). Ongoing AEs with missing end dates silently get `TRTEMFL = NA`. Always start_date = ASTDT.
2. **Off-by-one at TRTEDT** — `end_window = 0` means "exactly up to TRTEDT". `end_window = 30` extends 30 days past last dose. Pick the protocol-specified window.
3. **Partial-date AEs** — an `AESTDTC = "2024-07"` was imputed by `derive_vars_dt(date_imputation = "first")` to `2024-07-01`. If TRTSDT was `2024-07-15`, that AE looks pre-treatment (`ASTDT < TRTSDT`). Always inspect imputation flags (ASTDTF) before drawing conclusions.
4. **Time zones / `DTM`** — never mix `DT` and `DTM` arguments. The function compares whatever you pass; if you pass an integer date for one side and a POSIX seconds-since-epoch for the other, results are nonsense.

### 4.8 Multi-period studies and intensity worsening

In some studies (cross-over, multi-period), TRTEMFL must be evaluated per treatment period. Use `derive_var_trtemfl()` once per period and combine results. The `initial_intensity` / `intensity` arguments enable the **worsening** pathway: an AE that pre-existed but worsened on-treatment (e.g., grade 1 at baseline, grade 3 mid-treatment) earns TRTEMFL = "Y" even though its onset predates TRTSDT.

### 4.9 The TRTEMFL derivation

TRTEMFL = "Y" for AEs that started **on or after** first dose and on or before last dose + end_window.

```{r trtemfl-demo}
adae <- adae |>
  derive_var_trtemfl(
    start_date      = ASTDT,
    end_date        = AENDT,
    trt_start_date  = TRTSDT,
    trt_end_date    = TRTEDT,
    end_window      = 30
  )

adae |> select(USUBJID, AETERM, ASTDT, TRTSDT, TRTEDT, TRTEMFL)
```

```{r quiz-trtemfl, echo=FALSE}
quiz(
  question(
    "An AE that started before TRTSDT but worsened during treatment: what is TRTEMFL?",
    answer("NA"),
    answer("N"),
    answer("Y — when intensity (grade) arguments are supplied and worsening is detected", correct = TRUE),
    answer("Always NA without the intensity arguments"),
    allow_retry = TRUE
  ),
  question(
    "`end_window = 30` means:",
    answer("The AE must end within 30 days"),
    answer("AEs starting up to 30 days after TRTEDT are still counted as treatment-emergent", correct = TRUE),
    answer("The window is 30 weeks"),
    answer("Only the first 30 AEs are flagged"),
    allow_retry = TRUE
  )
)
```

```{r trtemfl-ex, exercise=TRUE, exercise.lines=16}
# Exercise: apply derive_var_trtemfl with end_window = 0 (strict)
# How many AEs get TRTEMFL = "Y" vs end_window = 30?

adae_strict <- adae |>
  derive_var_trtemfl(
    start_date     = ASTDT,
    end_date       = AENDT,
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT,
    end_window     = ___
  )

# Compare counts
list(
  strict_30  = sum(adae$TRTEMFL == "Y", na.rm = TRUE),
  strict_0   = sum(adae_strict$TRTEMFL == "Y", na.rm = TRUE)
)
```

```{r trtemfl-ex-hint-1}
# end_window = 0
```

```{r trtemfl-ex-solution}
adae_strict <- adae |>
  derive_var_trtemfl(
    start_date     = ASTDT,
    end_date       = AENDT,
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT,
    end_window     = 0
  )

list(
  strict_30 = sum(adae$TRTEMFL == "Y", na.rm = TRUE),
  strict_0  = sum(adae_strict$TRTEMFL == "Y", na.rm = TRUE)
)
```

### Exercise: Explore real ADAE from pharmaverseadam

The toy `ae` dataset has 5 records and 3 subjects. The real `adae_real` from pharmaverseadam has 1200+ AE records from 254 subjects — a realistic CDISCPILOT01 submission ADAE. Use it to understand the scale and content of a real treatment-emergent AE analysis.

```{r adae-real-explore, exercise=TRUE, exercise.lines=38}
# Q1: Total treatment-emergent AE records
cat("Total TEAE (TRTEMFL = 'Y'):",
    sum(adae_real$TRTEMFL == "Y", na.rm = TRUE), "\n")
cat("Total AE records:           ", nrow(adae_real), "\n\n")

# Q2: Most frequent treatment-emergent preferred terms
adae_real |>
  filter(TRTEMFL == "Y") |>
  count(AEDECOD, name = "n_events") |>
  arrange(desc(n_events)) |>
  head(10) |>
  print()

# Q3: Breakdown of severity among TEAEs
adae_real |>
  filter(TRTEMFL == "Y") |>
  count(AESEV, name = "n") |>
  arrange(desc(n)) |>
  print()

# Q4: How many subjects had at least one TEAE? (AOCCFL = "Y")
n_subj_with_teae <- adae_real |>
  filter(!is.na(AOCCFL) & AOCCFL == "Y") |>
  n_distinct("USUBJID")
cat("\nSubjects with ≥1 TEAE (AOCCFL = 'Y'):", n_subj_with_teae, "\n")
```

```{r adae-real-explore-hint-1}
# filter(TRTEMFL == "Y") selects treatment-emergent AEs
# count(AEDECOD) tallies by preferred term
# AOCCFL = "Y" marks the first TEAE per subject — one row per subject with ≥1 TEAE
```

```{r adae-real-explore-solution}
cat("Total TEAE (TRTEMFL = 'Y'):",
    sum(adae_real$TRTEMFL == "Y", na.rm = TRUE), "\n")
cat("Total AE records:           ", nrow(adae_real), "\n\n")

adae_real |>
  filter(TRTEMFL == "Y") |>
  count(AEDECOD, name = "n_events") |>
  arrange(desc(n_events)) |>
  head(10) |>
  print()

adae_real |>
  filter(TRTEMFL == "Y") |>
  count(AESEV, name = "n") |>
  arrange(desc(n)) |>
  print()

n_subj_with_teae <- adae_real |>
  filter(!is.na(AOCCFL) & AOCCFL == "Y") |>
  n_distinct("USUBJID")
cat("\nSubjects with ≥1 TEAE (AOCCFL = 'Y'):", n_subj_with_teae, "\n")
```

### 4.10 Exercise: Pre-treatment AE flag (PRETRTFL)

Some sponsors derive a counterpart flag — PRETRTFL = "Y" for AEs that started **before** TRTSDT. Build it with a simple mutate (no admiral helper required).

```{r pretrtfl-ex, exercise=TRUE, exercise.lines=14}
# Derive PRETRTFL = "Y" if ASTDT < TRTSDT, NA otherwise (NA when ASTDT or TRTSDT missing)

adae_pre <- adae |>
  mutate(
    PRETRTFL = case_when(
      is.na(ASTDT) | is.na(TRTSDT) ~ NA_character_,
      ASTDT < ___                  ~ "Y",
      TRUE                         ~ NA_character_
    )
  )

adae_pre |> select(USUBJID, AETERM, ASTDT, TRTSDT, TRTEMFL, PRETRTFL)
```

```{r pretrtfl-ex-hint-1}
# Compare ASTDT to TRTSDT: AE that STARTED before treatment start.
```

```{r pretrtfl-ex-hint-2}
# ASTDT < TRTSDT
```

```{r pretrtfl-ex-solution}
adae_pre <- adae |>
  mutate(
    PRETRTFL = case_when(
      is.na(ASTDT) | is.na(TRTSDT) ~ NA_character_,
      ASTDT < TRTSDT               ~ "Y",
      TRUE                         ~ NA_character_
    )
  )

adae_pre |> select(USUBJID, AETERM, ASTDT, TRTSDT, TRTEMFL, PRETRTFL)
```

### 4.11 Knowledge check — TRTEMFL

```{r quiz-trtemfl-deep, echo=FALSE}
quiz(
  question(
    "TRTEMFL = NA (missing) on a row. Which is the most likely root cause?",
    answer("ASTDT or TRTSDT is missing — the date comparison returned NA", correct = TRUE),
    answer("The AE is severe"),
    answer("Subject is in placebo arm"),
    answer("end_window was set to 0"),
    allow_retry = TRUE
  ),
  question(
    "Why does TRTEMFL use ASTDT (start date) and not AENDT (end date)?",
    answer("AENDT is preferred for severe AEs"),
    answer("Treatment-emergence is defined by AE ONSET, not resolution — ASTDT determines when the event began", correct = TRUE),
    answer("AENDT is always missing"),
    answer("admiral requires ASTDT by convention only"),
    allow_retry = TRUE
  ),
  question(
    "end_window = 30 in derive_var_trtemfl()...",
    answer("Limits TRTEMFL to AEs that resolved within 30 days"),
    answer("Permits AEs starting up to 30 days AFTER TRTEDT to still be TEAE, capturing delayed-onset drug effects", correct = TRUE),
    answer("Means only the first 30 AEs per subject are flagged"),
    answer("Means severity must be moderate or worse"),
    allow_retry = TRUE
  ),
  question(
    "For a partial AESTDTC = '2024-07' with TRTSDT = 2024-07-15, what is the risk?",
    answer("Imputation of AESTDTC to 2024-07-01 makes the AE look pre-treatment (ASTDT < TRTSDT) — TRTEMFL = NA when in reality the AE may have started post-dose", correct = TRUE),
    answer("There is no risk; admiral handles it"),
    answer("Imputation is ignored when partial"),
    answer("derive_var_trtemfl errors out on partials"),
    allow_retry = TRUE
  )
)
```

---

## 5 — Occurrence flags: AOCCFL and AOCCSFL

### 5.1 The "occurrence" concept

An **occurrence flag** marks the **first event** of a given type per subject. Why "first"? Because it lets us count *subjects with at least one AE of type X* without double-counting.

If subject 01-001 had three headaches, naive counting would inflate the headache incidence to 3 — but for incidence reporting we want **"1 subject experienced headache"**. Flagging only the first headache per subject solves this.

```
Subject's AE history:                   First-occurrence flags:
   USUBJID  AETERM       ASTDT          AOCCFL  AOCCSFL  AOCCPFL
   01-001   Headache     2024-01-20      Y       Y        Y
   01-001   Nausea       2024-02-01      .       Y        Y
   01-001   Headache     2024-02-10      .       .        .       <- not the first headache
                                         ^       ^        ^
                                       first   first   first
                                       AE      per     per
                                       per     SOC     PT
                                       subj
```

### 5.2 AOCCFL — first AE per subject (any term)

`AOCCFL = "Y"` marks the chronologically **earliest** treatment-emergent AE for each subject — regardless of term or system organ class. The denominator for an incidence table built on AOCCFL is **subjects with at least one TEAE**.

### 5.3 AOCCSFL — first AE per subject per SOC

`AOCCSFL = "Y"` marks the **earliest TEAE within each SOC** (System Organ Class). One subject can have many `AOCCSFL = "Y"` rows — one per SOC they experienced.

### 5.4 AOCCPFL — first AE per subject per preferred term

`AOCCPFL = "Y"` marks the **earliest TEAE within each preferred term (AEDECOD)**. One subject can have many `AOCCPFL = "Y"` rows — one per unique AEDECOD they experienced.

The three flags are tuned for three different granularities of incidence reporting:

| Flag      | Group by                          | Used for                           |
|-----------|-----------------------------------|------------------------------------|
| AOCCFL    | USUBJID                           | "Subjects with any TEAE"           |
| AOCCSFL   | USUBJID + AESOC                   | SOC-level incidence                |
| AOCCPFL   | USUBJID + AEDECOD                 | Preferred-term-level incidence     |

### 5.5 Why occurrence flags exist (the no-double-counting reason)

A safety table titled *"AEs by Preferred Term"* lists each PT with an `n (%)` representing **subjects** with that PT. The standard query is:

```
   For each AEDECOD:
     n = count of subjects with AOCCPFL = "Y" for this AEDECOD
     % = n / (N safety population)
```

If you used raw AE rows (no occurrence flag), three headache rows in a single subject would inflate the headache subject-count by 3. AOCCPFL guarantees one row per subject per PT.

### 5.6 SAS equivalent: PROC SORT + first.usubjid

```
PROC SORT DATA=adae OUT=for_aoccfl;
   WHERE TRTEMFL = 'Y';
   BY USUBJID ASTDT AESEQ;
RUN;
DATA adae;
   MERGE adae for_aoccfl(KEEP=USUBJID ASTDT AESEQ IN=in_aocc);
   BY USUBJID ASTDT AESEQ;
   IF FIRST.USUBJID AND in_aocc THEN AOCCFL = "Y";
RUN;
```

Side-by-side:

```
   SAS                                  |  admiral
   ------------------------------------ |  ------------------------------------
   PROC SORT NODUPKEY                   |  derive_var_extreme_flag(
     BY USUBJID ASTDT AESEQ;            |    by_vars = exprs(USUBJID),
     WHERE TRTEMFL = "Y";               |    order   = exprs(ASTDT, AESEQ),
   first.USUBJID -> AOCCFL = "Y"        |    new_var = AOCCFL,
                                        |    mode    = "first"
                                        |  ) wrapped in restrict_derivation(
                                        |    filter = TRTEMFL == "Y"
                                        |  )
```

### 5.7 Anatomy of derive_var_extreme_flag()

```r
derive_var_extreme_flag(
  dataset  = adae,
  by_vars  = exprs(STUDYID, USUBJID),     # subject (one flag per subject)
  order    = exprs(ASTDT, AESEQ),         # sort key: earliest first
  new_var  = AOCCFL,                      # name of output flag
  mode     = "first"                      # "first" or "last" of the sort
)
```

Wrapped in `restrict_derivation()` to apply *only* to TEAE rows:

```r
restrict_derivation(
  derivation = derive_var_extreme_flag,
  args = params(
    by_vars = exprs(STUDYID, USUBJID),
    order   = exprs(ASTDT, AESEQ),
    new_var = AOCCFL,
    mode    = "first"
  ),
  filter = TRTEMFL == "Y"     # restrict to TEAE rows
)
```

> **SAS analog mental model:** `derive_var_extreme_flag` is `PROC SORT NODUPKEY` + a `BY` group `first.` selector packaged as a single function. The `mode` argument is what `first.` vs `last.` would be in a SAS DATA step. The `restrict_derivation()` wrapper is the SAS `WHERE` clause.

### 5.8 Order matters — how the sort key defines "first"

The `order` argument **defines** what "first" means. If `order = exprs(ASTDT, AESEQ)`:

- Sort ascending by ASTDT first, breaking ties by AESEQ → "first" = earliest onset.
- Switching to `order = exprs(desc(ASTDT), AESEQ)` would make "first" = latest onset.
- Switching to `order = exprs(ASEVN, ASTDT)` would make "first" = mildest first.

Always pick `order` that matches the **clinical question**. For AOCCFL, the clinical question is *"what was the first event chronologically"* → `order = exprs(ASTDT, AESEQ)`.

### 5.9 The occurrence-flag derivations

```{r occ-flags-demo}
# AOCCFL: first treatment-emergent AE per subject
adae <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y"
  )

# AOCCSFL: first treatment-emergent AE per subject per SOC
adae <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, AESOC),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCSFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y"
  )

adae |> arrange(USUBJID, ASTDT) |> select(USUBJID, AETERM, AESOC, TRTEMFL, AOCCFL, AOCCSFL)
```

```{r quiz-occ-flags, echo=FALSE}
quiz(
  question(
    "AOCCFL = 'Y' should equal the number of subjects with at least one treatment-emergent AE. True or false?",
    answer("True", correct = TRUE),
    answer("False — AOCCFL can be 'Y' on multiple rows per subject"),
    allow_retry = TRUE
  ),
  question(
    "AOCCSFL by-group differs from AOCCFL by-group by:",
    answer("AOCCSFL also groups by AESOC (system organ class)", correct = TRUE),
    answer("AOCCSFL uses AETOXGR"),
    answer("AOCCSFL is the last AE, not the first"),
    answer("They are identical"),
    allow_retry = TRUE
  )
)
```

```{r aocc02fl-ex, exercise=TRUE, exercise.lines=18}
# Exercise: derive AOCC02FL = first treatment-emergent AE with AETOXGR >= 2 per subject

adae_02 <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCC02FL,
      mode    = "first"
    ),
    filter = ___ & ___
  )

adae_02 |> filter(!is.na(AOCC02FL)) |>
  select(USUBJID, AETERM, AETOXGR, AOCC02FL)
```

```{r aocc02fl-ex-hint-1}
# filter = TRTEMFL == "Y" & AETOXGR >= 2
```

```{r aocc02fl-ex-solution}
adae_02 <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCC02FL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y" & AETOXGR >= 2
  )

adae_02 |> filter(!is.na(AOCC02FL)) |>
  select(USUBJID, AETERM, AETOXGR, AOCC02FL)
```

### 5.10 Exercise: Derive AOCCPFL (first AE per subject per preferred term)

```{r aoccpfl-ex, exercise=TRUE, exercise.lines=18}
# AOCCPFL = first TEAE per subject per AEDECOD (preferred term).
# Mirror AOCCSFL but group by AEDECOD instead of AESOC.

adae_pfl <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, ___),    # subject + PT
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCPFL,
      mode    = "___"
    ),
    filter = TRTEMFL == "Y"
  )

adae_pfl |> arrange(USUBJID, ASTDT) |>
  select(USUBJID, AETERM, AEDECOD, AOCCFL, AOCCSFL, AOCCPFL)
```

```{r aoccpfl-ex-hint-1}
# Group by USUBJID + AEDECOD (preferred term). mode = "first".
```

```{r aoccpfl-ex-hint-2}
# by_vars = exprs(STUDYID, USUBJID, AEDECOD)
# mode = "first"
```

```{r aoccpfl-ex-solution}
adae_pfl <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, AEDECOD),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCPFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y"
  )

adae_pfl |> arrange(USUBJID, ASTDT) |>
  select(USUBJID, AETERM, AEDECOD, AOCCFL, AOCCSFL, AOCCPFL)
```

### 5.11 Knowledge check — occurrence flags

```{r quiz-occ-deep, echo=FALSE}
quiz(
  question(
    "AOCCFL vs AOCCSFL: what is the key difference?",
    answer("AOCCSFL flags the most severe AE per subject"),
    answer("AOCCSFL groups by USUBJID + AESOC (one flag per SOC), enabling SOC-level incidence; AOCCFL groups by USUBJID only", correct = TRUE),
    answer("AOCCSFL is for serious AEs only"),
    answer("They are the same"),
    allow_retry = TRUE
  ),
  question(
    "Why does a safety-table 'AEs by Preferred Term' rely on AOCCPFL rather than raw AE rows?",
    answer("Raw rows would double-count subjects who had the same PT multiple times; AOCCPFL guarantees one Y per subject per PT", correct = TRUE),
    answer("Raw rows are too many to count"),
    answer("AOCCPFL is required by define.xml"),
    answer("Preferred terms are not in raw AE data"),
    allow_retry = TRUE
  ),
  question(
    "Changing order from exprs(ASTDT, AESEQ) to exprs(ASEVN, ASTDT) would make AOCCFL flag:",
    answer("The most severe TEAE per subject (then earliest within that severity)", correct = TRUE),
    answer("The same row (order does not matter)"),
    answer("All rows for the subject"),
    answer("Only Grade 5 AEs"),
    allow_retry = TRUE
  ),
  question(
    "The SAS analog of derive_var_extreme_flag(mode='first', by_vars=USUBJID) is:",
    answer("PROC SQL UNION"),
    answer("PROC SORT BY USUBJID + DATA step IF first.USUBJID THEN AOCCFL='Y'", correct = TRUE),
    answer("PROC MEANS NWAY"),
    answer("PROC TRANSPOSE"),
    allow_retry = TRUE
  )
)
```

---

## 6 — MedDRA queries: derive_vars_query()

```{r queries-demo}
# queries_df has PREFIX, GRPNAME, TERM_NAME for SMQ01 and CQ02
adae <- adae |>
  derive_vars_query(dataset_queries = queries_df)

adae |> select(USUBJID, AEDECOD, SMQ01NAM, CQ02NAM) |> head(6)
```

```{r quiz-queries, echo=FALSE}
quiz(
  question(
    "What variables does `derive_vars_query()` add to ADAE?",
    answer("MEDDRACD and MEDDRANM"),
    answer("SMQ01NAM, SMQ01CD, CQ02NAM, etc. — named from the PREFIX column of the queries dataset", correct = TRUE),
    answer("AETERM and AEDECOD synonyms"),
    answer("New parameter rows"),
    allow_retry = TRUE
  )
)
```

---

## 7 — Last dose linkage: derive_vars_joined()

```{r last-dose-demo}
# For each AE, attach the most recent prior dose date
adae <- adae |>
  derive_vars_joined(
    dataset_add = ex_occ |> filter(EXDOSE > 0),
    by_vars     = exprs(USUBJID),
    order       = exprs(EXSTDT),
    new_vars    = exprs(LDOSEDT = EXSTDT, LDOSEDOSE = EXDOSE),
    join_vars   = exprs(EXSTDT),
    filter_join = ASTDT >= EXSTDT,
    mode        = "last"
  )

adae |> select(USUBJID, AETERM, ASTDT, LDOSEDT, LDOSEDOSE) |> head(4)
```

```{r quiz-joined, echo=FALSE}
quiz(
  question(
    "Why use `derive_vars_joined()` instead of `derive_vars_merged()` for last-dose linkage?",
    answer("derive_vars_joined() is faster"),
    answer("derive_vars_joined() supports filter conditions that reference both datasets' variables (ASTDT >= EXSTDT)", correct = TRUE),
    answer("derive_vars_merged() cannot rename variables"),
    answer("They are equivalent for this use case"),
    allow_retry = TRUE
  )
)
```

```{r joined-ex, exercise=TRUE, exercise.lines=16}
# Exercise: for each AE, find the FIRST dose that started ON OR AFTER the AE start date
# (i.e., dose started after AE — a different temporal relationship)
# Store the date as NEXTDOSEDT

adae_nd <- adae |>
  derive_vars_joined(
    dataset_add = ex_occ |> filter(EXDOSE > 0),
    by_vars     = exprs(USUBJID),
    order       = exprs(EXSTDT),
    new_vars    = exprs(NEXTDOSEDT = EXSTDT),
    join_vars   = exprs(EXSTDT),
    filter_join = ___ <= EXSTDT,   # AE start <= dose start
    mode        = "first"
  )

adae_nd |> select(USUBJID, AETERM, ASTDT, NEXTDOSEDT) |> head(4)
```

```{r joined-ex-hint-1}
# filter_join = ASTDT <= EXSTDT
```

```{r joined-ex-solution}
adae_nd <- adae |>
  derive_vars_joined(
    dataset_add = ex_occ |> filter(EXDOSE > 0),
    by_vars     = exprs(USUBJID),
    order       = exprs(EXSTDT),
    new_vars    = exprs(NEXTDOSEDT = EXSTDT),
    join_vars   = exprs(EXSTDT),
    filter_join = ASTDT <= EXSTDT,
    mode        = "first"
  )

adae_nd |> select(USUBJID, AETERM, ASTDT, NEXTDOSEDT) |> head(4)
```

---

## 8 — ASEQ and create_single_dose_dataset

```{r aseq-demo}
adae <- adae |>
  derive_var_obs_number(
    new_var  = ASEQ,
    by_vars  = exprs(STUDYID, USUBJID),
    order    = exprs(ASTDT, AESEQ),
    check_type = "warning"
  )

adae |> select(USUBJID, AETERM, ASTDT, AESEQ, ASEQ) |> arrange(USUBJID, ASEQ)
```

```{r single-dose-demo}
# Expand QD dosing to one row per day
ex_single <- ex_occ |>
  create_single_dose_dataset(
    dose_freq        = EXDOSFRQ,
    start_date       = EXSTDT,
    end_date         = EXENDT,
    keep_source_vars = exprs(USUBJID, EXSTDT, EXENDT, EXDOSFRQ, EXDOSE)
  )

# Compare row counts
cat("Original EX rows:", nrow(ex_occ), "\n")
cat("Expanded rows:   ", nrow(ex_single), "\n")
```

```{r quiz-aseq-single, echo=FALSE}
quiz(
  question(
    "What does `create_single_dose_dataset()` do?",
    answer("Aggregates multiple doses per day into one"),
    answer("Expands summary EX records (e.g., QD from Day 1 to Day 15) to one row per dose date", correct = TRUE),
    answer("Creates a new ADaM dataset"),
    answer("Derives the number of doses"),
    allow_retry = TRUE
  )
)
```

---

## 9 — Complete ADAE pipeline

```{r capstone-adae, exercise=TRUE, exercise.lines=60}
# Build ADAE from ae + adsl_occ + ex_occ in one clean pipe
# Steps: merge ADSL → dates → severity → TRTEMFL → AOCCFL → AOCCSFL → MedDRA queries → ASEQ

adsl_v <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, AGE, SEX, RACE, SAFFL)

adae_full <- ae |>
  derive_vars_merged(dataset_add = adsl_occ, new_vars = adsl_v,
                     by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_dt(new_vars_prefix = "AST", dtc = AESTDTC,
                 highest_imputation = "M", date_imputation = "first") |>
  derive_vars_dt(new_vars_prefix = "AEN", dtc = AEENDTC,
                 highest_imputation = "M", date_imputation = "last") |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ASTDT, AENDT)) |>
  mutate(
    ASEV  = AESEV, AREL = AEREL,
    ASEVN = case_when(ASEV == "MILD" ~ 1, ASEV == "MODERATE" ~ 2, ASEV == "SEVERE" ~ 3,
                      TRUE ~ NA_real_)
  ) |>
  derive_var_trtemfl(start_date = ASTDT, end_date = AENDT,
                     trt_start_date = TRTSDT, trt_end_date = TRTEDT, end_window = 30) |>
  derive_var_ontrtfl(start_date = ASTDT, ref_start_date = TRTSDT,
                     ref_end_date = TRTEDT, ref_end_window = 30) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(by_vars = exprs(STUDYID, USUBJID),
                  order = exprs(ASTDT, AESEQ), new_var = AOCCFL, mode = "first"),
    filter = TRTEMFL == "Y"
  ) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(by_vars = exprs(STUDYID, USUBJID, AESOC),
                  order = exprs(ASTDT, AESEQ), new_var = AOCCSFL, mode = "first"),
    filter = TRTEMFL == "Y"
  ) |>
  derive_vars_query(dataset_queries = queries_df) |>
  derive_var_obs_number(new_var = ASEQ, by_vars = exprs(STUDYID, USUBJID),
                        order = exprs(ASTDT, AESEQ))

adae_full |>
  select(USUBJID, AETERM, TRTEMFL, AOCCFL, AOCCSFL, SMQ01NAM, ASEQ) |>
  arrange(USUBJID, ASEQ)
```

---

## 9.5 — P21 Derivation Diagnostics

ADAE is the most scrutinized ADaM dataset in any submission — regulators check TRTEMFL rates, AOCCFL counts, and study day derivations closely. The three diagnostics below each come with a define.xml spec and a Pinnacle 21 error message.

---

### Diagnostic 1 — AOCCFL: Wrong mode flags last event instead of first

**Derivation method (define.xml / Pinnacle 21 field):**

> *AOCCFL = 'Y' for the FIRST treatment-emergent adverse event per subject (ordered by ASTDT, then AESEQ). Used in incidence analyses where "subject with ≥1 TEAE" is the unit. Only one row per subject should have AOCCFL = 'Y'. Source: TRTEMFL, ASTDT, AESEQ.*

The programmer submitted:

```r
adae_draft <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCFL,
      mode    = "last"    # ← bug: should be "first"
    ),
    filter = TRTEMFL == "Y"
  )
```

**P21 report:** *"AOCCFL = 'Y' is assigned to the last chronological TEAE for each subject, not the earliest onset. Subject 01-001 has two TEAEs; AOCCFL = 'Y' is on the later AE (AESEQ = 2). Expected: AOCCFL marks earliest onset per subject."*

```{r p21-bug-aoccfl, exercise=TRUE, exercise.lines=22}
# Fix: AOCCFL = first (earliest onset) TEAE per subject.
# With order = exprs(ASTDT, AESEQ), mode = "first" picks the row with the smallest ADT.

adae_fixed_aoccfl <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCFL,
      mode    = "___"    # fix this
    ),
    filter = TRTEMFL == "Y"
  )

# Verify: exactly one AOCCFL = "Y" per subject, on the earliest ASTDT
adae_fixed_aoccfl |>
  filter(!is.na(AOCCFL)) |>
  select(USUBJID, AETERM, ASTDT, AESEQ, AOCCFL) |>
  arrange(USUBJID, ASTDT)
```

```{r p21-bug-aoccfl-hint-1}
# mode = "first" — after sorting by ASTDT ascending, "first" gives the earliest AE onset
# mode = "last" would give the most recent AE — wrong for AOCCFL
```

```{r p21-bug-aoccfl-solution}
adae_fixed_aoccfl <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y"
  )

adae_fixed_aoccfl |>
  filter(!is.na(AOCCFL)) |>
  select(USUBJID, AETERM, ASTDT, AESEQ, AOCCFL) |>
  arrange(USUBJID, ASTDT)
```

---

### Diagnostic 2 — TRTEMFL: Using the AE end date instead of the start date

**Derivation method (define.xml / Pinnacle 21 field):**

> *TRTEMFL = 'Y' if the AE STARTED on or after the first treatment date (TRTSDT) and on or before TRTEDT + 30 days. The onset date (ASTDT) determines treatment-emergence — not the resolution date. Source: ASTDT, TRTSDT, TRTEDT.*

The programmer submitted:

```r
adae_draft2 <- adae |>
  derive_var_trtemfl(
    start_date     = AENDT,    # ← bug: should be ASTDT (AE start, not end)
    end_date       = AENDT,
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT,
    end_window     = 30
  )
```

**P21 report:** *"TRTEMFL = NA for AE 'ANAPHYLACTIC SHOCK' (USUBJID 01-002, AEENDTC missing). An ongoing serious AE with missing end date cannot be flagged because AENDT is NA. Additionally, AEs that started on-treatment but resolved pre-treatment are incorrectly excluded. Onset date (ASTDT) must drive TRTEMFL."*

```{r p21-bug-trtemfl-startdate, exercise=TRUE, exercise.lines=18}
# Fix: use ASTDT (AE start date) as start_date — the AE onset drives treatment-emergence.

adae_fixed_trtemfl <- adae |>
  derive_var_trtemfl(
    start_date     = ___,      # fix this — should be the AE START date
    end_date       = AENDT,
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT,
    end_window     = 30
  )

# Verify: the SEVERE/SERIOUS AE for 01-002 (AEENDTC missing) should still get TRTEMFL = "Y"
adae_fixed_trtemfl |>
  select(USUBJID, AETERM, AESTDTC, AEENDTC, ASTDT, AENDT, TRTEMFL) |>
  filter(USUBJID == "01-002")
```

```{r p21-bug-trtemfl-startdate-hint-1}
# start_date = ASTDT  (the analysis AE START date, not AENDT)
# TRTEMFL checks: did the AE ONSET occur during the treatment window?
# Using AENDT as start_date means missing-end AEs (ongoing) can never be flagged.
```

```{r p21-bug-trtemfl-startdate-solution}
adae_fixed_trtemfl <- adae |>
  derive_var_trtemfl(
    start_date     = ASTDT,
    end_date       = AENDT,
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT,
    end_window     = 30
  )

adae_fixed_trtemfl |>
  select(USUBJID, AETERM, AESTDTC, AEENDTC, ASTDT, AENDT, TRTEMFL) |>
  filter(USUBJID == "01-002")
```

---

### Diagnostic 3 — ASTDY: Study day off-by-one (missing Day 1 convention)

**Derivation method (define.xml / Pinnacle 21 field):**

> *ASTDY = study day of AE onset, relative to first treatment date (TRTSDT). Follows CDISC ADaM IG §4.4.7 study day convention: TRTSDT = Day 1; the day before TRTSDT = Day −1 (no Day 0 exists for positive days). For positive study days: ASTDY = ASTDT − TRTSDT + 1. Use admiral's derive_vars_dy() which implements this convention automatically. Source: ASTDT, TRTSDT.*

The programmer submitted:

```r
adae_draft3 <- adae |>
  mutate(ASTDY = ASTDT - TRTSDT)    # ← bug: missing the +1 for Day 1 convention
```

**P21 report:** *"AD0214 — ASTDY = 0 detected for subjects where ASTDT = TRTSDT. CDISC ADaM IG prohibits study day 0 — the reference date is Day 1. All positive ASTDY values are exactly 1 less than expected."*

```{r p21-bug-astdy, exercise=TRUE, exercise.lines=20}
# Fix: use derive_vars_dy() which automatically applies the CDISC Day 1 convention.
# Positive days: ASTDY = ASTDT - TRTSDT + 1
# Negative days: ASTDY = ASTDT - TRTSDT      (no adjustment — Day -1 before treatment start)

# Broken version for comparison:
adae_broken_astdy <- adae |>
  mutate(ASTDY_wrong = ASTDT - TRTSDT)

# Fixed version:
adae_fixed_astdy <- adae |>
  derive_vars_dy(
    reference_date = ___,          # reference date for study day calculation
    source_vars    = exprs(___)    # compute ASTDY from ASTDT
  )

# Compare: ASTDY_wrong should be exactly 1 less than ASTDY for all on/post-treatment rows
left_join(
  adae_broken_astdy |> select(USUBJID, AETERM, ASTDT, TRTSDT, ASTDY_wrong),
  adae_fixed_astdy  |> select(USUBJID, AETERM, ASTDY),
  by = c("USUBJID", "AETERM")
) |>
  mutate(diff = ASTDY - ASTDY_wrong) |>
  select(USUBJID, AETERM, ASTDY_wrong, ASTDY, diff)
```

```{r p21-bug-astdy-hint-1}
# reference_date = TRTSDT
# source_vars    = exprs(ASTDT)   → produces ASTDY
# derive_vars_dy() strips the suffix to name the result: AST-DT → AST-DY → ASTDY
```

```{r p21-bug-astdy-solution}
adae_broken_astdy <- adae |>
  mutate(ASTDY_wrong = ASTDT - TRTSDT)

adae_fixed_astdy <- adae |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ASTDT)
  )

left_join(
  adae_broken_astdy |> select(USUBJID, AETERM, ASTDT, TRTSDT, ASTDY_wrong),
  adae_fixed_astdy  |> select(USUBJID, AETERM, ASTDY),
  by = c("USUBJID", "AETERM")
) |>
  mutate(diff = ASTDY - ASTDY_wrong) |>
  select(USUBJID, AETERM, ASTDY_wrong, ASTDY, diff)
```

---

## Quiz — final

```{r quiz-final-17, echo=FALSE}
quiz(
  question(
    "A programmer sets AOCCFL using `mode = 'last'` in restrict_derivation. P21 flags that AOCCFL does not mark the earliest TEAE onset per subject. What does mode = 'last' produce, and what is the fix?",
    answer("mode = 'last' marks the most recent TEAE per subject — fix: use mode = 'first' so the earliest onset (smallest ASTDT) is flagged", correct = TRUE),
    answer("mode = 'last' marks the most severe AE — fix: use mode = 'first'"),
    answer("mode = 'last' is correct for AOCCFL — only AOCCSFL uses 'first'"),
    answer("mode = 'last' and 'first' produce identical results when there is only one TEAE per subject"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "A programmer writes `start_date = AENDT` in derive_var_trtemfl(). An ongoing SAE with AEENDTC missing gets TRTEMFL = NA. What is the root cause?",
    answer("derive_var_trtemfl() requires a non-missing end date for all AEs"),
    answer("AENDT is NA when AEENDTC is missing — using it as the 'start_date' makes the on-treatment comparison NA, so no TRTEMFL is assigned. Fix: use ASTDT (AE onset) as start_date", correct = TRUE),
    answer("Missing AEENDTC always results in TRTEMFL = NA regardless of which date argument is used"),
    answer("TRTEMFL can never be 'Y' for serious AEs with missing end dates"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "P21 rule AD0214 flags ASTDY = 0 in ADAE. The programmer used `mutate(ASTDY = ASTDT - TRTSDT)`. What is the CDISC convention and the correct formula?",
    answer("CDISC study day has no Day 0 — the reference date is Day 1. For on/post-treatment days: ASTDY = ASTDT − TRTSDT + 1. Use derive_vars_dy() which implements this automatically", correct = TRUE),
    answer("ASTDY = 0 is valid — it represents the treatment start date"),
    answer("The formula is correct; ASTDY = 0 is only flagged for labs, not AEs"),
    answer("CDISC requires ASTDY to be 1-indexed but the +1 is only for BDS datasets"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "AOCCFL differs from AOCCSFL in which key respect?",
    answer("AOCCFL flags the first TEAE per subject; AOCCSFL flags the first TEAE per subject per system organ class (AESOC), enabling SOC-level incidence counting", correct = TRUE),
    answer("AOCCSFL flags the most severe AE per subject; AOCCFL flags the first"),
    answer("AOCCSFL uses AEDECOD (PT) instead of AESOC (SOC)"),
    answer("They are identical — both use the same by_vars"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "Why does `date_imputation = 'last'` apply to AEENDTC (AE end date) while `date_imputation = 'first'` applies to AESTDTC (AE start date)?",
    answer("'last' and 'first' are interchangeable — admiral chooses the appropriate value"),
    answer("'first' for start dates minimises false treatment-emergent classifications (partial start pushed earlier); 'last' for end dates is conservative — assumes the AE lasted as long as possible within the period", correct = TRUE),
    answer("'last' for end dates is required by CDISC; 'first' for start dates is optional"),
    answer("End dates must always use 'last'; start dates have no imputation in ADAE"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "`end_window = 30` in derive_var_trtemfl() means an AE starting up to 30 days after TRTEDT is still treatment-emergent. Why is a post-treatment window important clinically?",
    answer("Many drug effects (hypersensitivity, delayed toxicity) manifest after the last dose — excluding them would under-count safety events attributable to treatment", correct = TRUE),
    answer("It prevents missing AEs with partial start dates"),
    answer("It is required by ICH E3 for all AE analyses"),
    answer("It converts non-treatment-emergent AEs into treatment-emergent ones for imputation"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "derive_vars_joined() is used for last-dose linkage instead of derive_vars_merged(). What capability does derive_vars_joined() have that derive_vars_merged() lacks?",
    answer("derive_vars_joined() can rename variables; derive_vars_merged() cannot"),
    answer("derive_vars_joined() accepts a filter_join condition that references both the primary and the added dataset simultaneously (e.g., ASTDT >= EXSTDT), enabling temporal filtering across two datasets", correct = TRUE),
    answer("derive_vars_joined() is faster on large EX datasets"),
    answer("derive_vars_merged() cannot operate on exposure datasets"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "derive_vars_query() adds a column named SMQ01NAM. Where does 'SMQ01' come from?",
    answer("It is derived from the GRPNAME field in the queries dataset"),
    answer("It is derived from the PREFIX column of the queries dataset — each unique PREFIX generates a {PREFIX}NAM variable", correct = TRUE),
    answer("It is a fixed admiral naming convention for MedDRA queries"),
    answer("The programmer must specify the output column name explicitly"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "ASEQ is derived with order = exprs(ASTDT, AESEQ). A programmer uses only order = exprs(AESEQ). What is wrong?",
    answer("AESEQ alone is fine — it is always in chronological order"),
    answer("AESEQ is the source SDTM sequence number, not necessarily chronological by onset date. Using ASTDT first ensures analysis sequence matches clinical timeline; AESEQ breaks ties only", correct = TRUE),
    answer("order requires at least three variables in ADAE"),
    answer("AESEQ is not available in the ADAE dataset"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "ADCM (Concomitant Medications) uses the same OCCDS structure as ADAE but maps a different coding dictionary. What is the key difference?",
    answer("ADCM uses OCCDS like ADAE, but maps WHODrug ATC codes (anatomical therapeutic chemical hierarchy) instead of MedDRA preferred terms and SOCs", correct = TRUE),
    answer("ADCM uses BDS structure — one row per medication per subject per visit"),
    answer("ADCM has PARAMCD; ADAE does not"),
    answer("ADCM has no treatment-emergent flag equivalent"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```
