---
title: "Lesson 17 — admiral Part 4: OCCDS Datasets (ADAE, ADCM)"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── shared mocks (from prior lessons) ────────────────────────────────────────
exprs <- function(...) {
  dots <- as.list(substitute(list(...)))[-1]
  dots
}

convert_blanks_to_na <- function(dataset) {
  dataset |> mutate(across(where(is.character), ~ifelse(. == "", NA_character_, .)))
}

.impute_dtc <- function(dtc, date_imputation = "first") {
  sapply(dtc, function(d) {
    if (is.na(d)) return(NA_character_)
    if (nchar(d) == 4) {
      paste0(d, if (date_imputation == "last") "-12-31" else "-01-01")
    } else if (nchar(d) == 7) {
      if (date_imputation == "last") {
        yr <- as.integer(substr(d, 1, 4)); mo <- as.integer(substr(d, 6, 7))
        nxt <- if (mo == 12) paste0(yr+1,"-01-01") else paste0(yr,"-",sprintf("%02d",mo+1),"-01")
        as.character(as.Date(nxt) - 1)
      } else paste0(d, "-01")
    } else d
  })
}

convert_dtc_to_dt <- function(dtc, date_imputation = "first") {
  suppressWarnings(as.numeric(as.Date(.impute_dtc(dtc, date_imputation))))
}

derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL,
                                order = NULL, filter_add = NULL, mode = NULL, ...) {
  by_nm   <- sapply(by_vars, deparse)
  join_df <- dataset_add
  if (!is.null(filter_add)) {
    fe <- substitute(filter_add)
    join_df <- tryCatch(filter(join_df, !!fe), error = function(e) join_df)
  }
  if (!is.null(new_vars)) {
    old_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d, "=")[[1]][2]) else d
    })
    new_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d, "=")[[1]][1]) else d
    })
    keep <- intersect(c(by_nm, old_nms), names(join_df))
    join_df <- join_df |> select(all_of(keep))
    for (i in seq_along(old_nms)) {
      if (old_nms[i] != new_nms[i] && old_nms[i] %in% names(join_df))
        join_df <- join_df |> rename(!!new_nms[i] := !!old_nms[i])
    }
  }
  by_present <- intersect(by_nm, names(join_df))
  left_join(dataset, join_df, by = by_present)
}

derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation = "none", highest_imputation = "n",
                            flag_imputation = "auto", ...) {
  dtc_nm  <- deparse(substitute(dtc))
  imp_rule <- if (date_imputation %in% c("none","") && highest_imputation == "n") "none"
              else if (date_imputation == "last") "last" else "first"
  raw     <- if (dtc_nm %in% names(dataset)) dataset[[dtc_nm]] else rep(NA_character_, nrow(dataset))
  imputed <- if (imp_rule == "none") raw else .impute_dtc(raw, imp_rule)
  dataset[[paste0(new_vars_prefix, "DT")]] <- suppressWarnings(as.numeric(as.Date(imputed)))
  if (flag_imputation != "none") {
    dataset[[paste0(new_vars_prefix, "DTF")]] <- case_when(
      is.na(raw)                       ~ NA_character_,
      nchar(raw) <= 4 & !is.na(imputed) ~ "Y",
      nchar(raw) <= 7 & !is.na(imputed) ~ "M",
      TRUE                              ~ NA_character_
    )
  }
  dataset
}

derive_vars_dtm <- function(dataset, dtc, new_vars_prefix,
                             date_imputation = "first", time_imputation = "first",
                             highest_imputation = "n", ...) {
  dtc_nm  <- deparse(substitute(dtc))
  raw     <- if (dtc_nm %in% names(dataset)) dataset[[dtc_nm]] else rep(NA_character_, nrow(dataset))
  imputed <- .impute_dtc(raw, date_imputation)
  suffix  <- if (time_imputation == "last") "T23:59:59" else "T00:00:00"
  dataset[[paste0(new_vars_prefix, "DTM")]] <- suppressWarnings(
    as.numeric(as.POSIXct(paste0(imputed, suffix), tz = "UTC"))
  )
  dataset
}

derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref_nm  <- deparse(substitute(reference_date))
  src_nms <- sapply(source_vars, deparse)
  for (sv in src_nms) {
    dy_nm <- sub("DT$", "DY", sv)
    if (sv %in% names(dataset) && ref_nm %in% names(dataset)) {
      v <- dataset[[sv]]; r <- dataset[[ref_nm]]
      dataset[[dy_nm]] <- ifelse(is.na(v) | is.na(r), NA_real_,
        ifelse(v >= r, as.numeric(v - r) + 1, as.numeric(v - r)))
    }
  }
  dataset
}

derive_var_extreme_flag <- function(dataset, by_vars, order, new_var, mode = "first") {
  new_nm  <- deparse(substitute(new_var))
  by_nm   <- sapply(by_vars, deparse)
  ord_nms <- sapply(order, function(x) { d <- deparse(x); gsub("desc\\(|\\)","",d) })
  desc_fl <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
  ord_df  <- dataset
  for (i in rev(seq_along(ord_nms))) {
    if (ord_nms[i] %in% names(ord_df))
      ord_df <- if (desc_fl[i]) arrange(ord_df, desc(.data[[ord_nms[i]]]))
                else arrange(ord_df, .data[[ord_nms[i]]])
  }
  by_present <- intersect(by_nm, names(ord_df))
  if (mode == "first")
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(1) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)][1], new_nm)))
  else
    flagged <- ord_df |> group_by(across(all_of(by_present))) |>
      slice(n()) |> ungroup() |> mutate(!!new_nm := "Y") |>
      select(all_of(c(by_present, ord_nms[ord_nms %in% names(ord_df)][1], new_nm)))
  if (new_nm %in% names(dataset)) dataset[[new_nm]] <- NULL
  left_join(dataset, flagged, by = intersect(names(flagged), names(dataset)))
}

restrict_derivation <- function(dataset, derivation, args = list(), filter = NULL) {
  fe <- substitute(filter)
  in_rows <- if (!is.null(fe)) {
    tryCatch(which(eval(fe, dataset, parent.frame())), error = function(e) seq_len(nrow(dataset)))
  } else seq_len(nrow(dataset))
  out_rows <- setdiff(seq_len(nrow(dataset)), in_rows)
  sub_data  <- dataset[in_rows, , drop = FALSE]
  all_args  <- c(list(dataset = sub_data), args)
  derived   <- tryCatch(do.call(derivation, all_args), error = function(e) sub_data)
  bind_rows(derived, dataset[out_rows, , drop = FALSE])
}

params <- function(...) list(...)

# ── OCCDS-specific mocks ─────────────────────────────────────────────────────
derive_var_trtemfl <- function(dataset, start_date, end_date = NULL,
                                trt_start_date, trt_end_date,
                                end_window = 0,
                                initial_intensity = NULL,
                                intensity = NULL, ...) {
  sd_nm  <- deparse(substitute(start_date))
  ed_nm  <- if (!is.null(substitute(end_date))) deparse(substitute(end_date)) else NULL
  ts_nm  <- deparse(substitute(trt_start_date))
  te_nm  <- deparse(substitute(trt_end_date))
  ii_nm  <- if (!is.null(substitute(initial_intensity))) deparse(substitute(initial_intensity)) else NULL
  in_nm  <- if (!is.null(substitute(intensity)))         deparse(substitute(intensity))         else NULL
  dataset |> mutate(
    TRTEMFL = case_when(
      is.na(.data[[sd_nm]]) | is.na(.data[[ts_nm]]) ~ NA_character_,
      .data[[sd_nm]] >= .data[[ts_nm]] &
        .data[[sd_nm]] <= (.data[[te_nm]] + end_window) ~ "Y",
      TRUE ~ NA_character_
    )
  )
}

derive_var_ontrtfl <- function(dataset, start_date, end_date = NULL,
                                ref_start_date, ref_end_date,
                                ref_end_window = 0, ...) {
  sd_nm <- deparse(substitute(start_date))
  rs_nm <- deparse(substitute(ref_start_date))
  re_nm <- deparse(substitute(ref_end_date))
  dataset |> mutate(
    ONTRTFL = case_when(
      is.na(.data[[sd_nm]]) | is.na(.data[[rs_nm]]) ~ NA_character_,
      .data[[sd_nm]] >= .data[[rs_nm]] &
        .data[[sd_nm]] <= (.data[[re_nm]] + ref_end_window) ~ "Y",
      TRUE ~ NA_character_
    )
  )
}

derive_var_obs_number <- function(dataset, new_var, by_vars, order,
                                   check_type = "warning") {
  new_nm  <- deparse(substitute(new_var))
  by_nm   <- sapply(by_vars, deparse)
  ord_nms <- sapply(order, function(x) { d <- deparse(x); gsub("desc\\(|\\)","",d) })
  desc_fl <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
  ord_df  <- dataset
  for (i in rev(seq_along(ord_nms))) {
    if (ord_nms[i] %in% names(ord_df))
      ord_df <- if (desc_fl[i]) arrange(ord_df, desc(.data[[ord_nms[i]]]))
                else arrange(ord_df, .data[[ord_nms[i]]])
  }
  by_present <- intersect(by_nm, names(ord_df))
  ord_df |> group_by(across(all_of(by_present))) |>
    mutate(!!new_nm := row_number()) |> ungroup()
}

derive_vars_joined <- function(dataset, dataset_add, by_vars, order,
                                new_vars = NULL, join_vars = NULL,
                                filter_join = NULL, mode = "first",
                                join_type = "all", ...) {
  by_nm  <- sapply(by_vars, deparse)
  fj_sub <- substitute(filter_join)
  merged <- left_join(dataset, dataset_add, by = intersect(by_nm, names(dataset_add)),
                      suffix = c("", ".add"))
  if (!is.null(fj_sub)) {
    merged <- tryCatch(filter(merged, !!fj_sub), error = function(e) merged)
  }
  if (!is.null(new_vars)) {
    old_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=",d)) trimws(strsplit(d,"=")[[1]][2]) else d
    })
    new_nms <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=",d)) trimws(strsplit(d,"=")[[1]][1]) else d
    })
    for (i in seq_along(old_nms)) {
      if (old_nms[i] %in% names(merged))
        merged <- rename(merged, !!new_nms[i] := !!old_nms[i])
    }
    keep_extra <- new_nms
  } else {
    keep_extra <- setdiff(names(dataset_add), c(by_nm, names(dataset)))
  }
  ord_nms <- sapply(order, function(x) { d <- deparse(x); gsub("desc\\(|\\)","",d) })
  desc_fl <- sapply(order, function(x) grepl("^desc\\(", deparse(x)))
  for (i in rev(seq_along(ord_nms))) {
    if (ord_nms[i] %in% names(merged))
      merged <- if (desc_fl[i]) arrange(merged, desc(.data[[ord_nms[i]]]))
                else arrange(merged, .data[[ord_nms[i]]])
  }
  by_present <- intersect(by_nm, names(merged))
  if (mode == "first")
    merged <- merged |> group_by(across(all_of(by_present))) |> slice(1) |> ungroup()
  else
    merged <- merged |> group_by(across(all_of(by_present))) |> slice(n()) |> ungroup()
  keep_cols <- intersect(keep_extra, names(merged))
  extra <- merged |> select(all_of(c(by_present, keep_cols)))
  if (any(keep_cols %in% names(dataset))) {
    for (nm in keep_cols[keep_cols %in% names(dataset)]) dataset[[nm]] <- NULL
  }
  left_join(dataset, extra, by = by_present)
}

derive_vars_query <- function(dataset, dataset_queries) {
  prefixes <- unique(dataset_queries$PREFIX)
  for (px in prefixes) {
    grp_rows <- dataset_queries |> filter(PREFIX == px)
    grp_name <- grp_rows$GRPNAME[1]
    terms    <- grp_rows$TERM_NAME
    dataset <- dataset |> mutate(
      !!paste0(px, "NAM") := case_when(
        toupper(AEDECOD) %in% toupper(terms) ~ grp_name,
        TRUE ~ NA_character_
      )
    )
  }
  dataset
}

create_single_dose_dataset <- function(dataset, dose_freq, start_date,
                                        end_date, keep_source_vars = NULL) {
  sf_nm <- deparse(substitute(dose_freq))
  sd_nm <- deparse(substitute(start_date))
  ed_nm <- deparse(substitute(end_date))
  result <- vector("list", nrow(dataset))
  for (i in seq_len(nrow(dataset))) {
    row  <- dataset[i, , drop = FALSE]
    freq <- row[[sf_nm]]
    sd   <- row[[sd_nm]]; ed <- row[[ed_nm]]
    if (is.na(sd) || is.na(ed)) { result[[i]] <- row; next }
    sd_d <- as.Date(sd, origin = "1970-01-01")
    ed_d <- as.Date(ed, origin = "1970-01-01")
    if (grepl("^(QD|QD1|ONCE DAILY)$", toupper(freq %||% ""))) {
      dates <- seq(sd_d, ed_d, by = "day")
      row_rep <- row[rep(1, length(dates)), ]
      row_rep[[sd_nm]] <- as.numeric(dates)
      result[[i]] <- row_rep
    } else {
      result[[i]] <- row
    }
  }
  bind_rows(result)
}

# fallback for %||%
`%||%` <- function(a, b) if (is.null(a)) b else a

# ── Sample datasets ───────────────────────────────────────────────────────────
adsl_occ <- tribble(
  ~STUDYID,        ~USUBJID, ~TRTSDT, ~TRTEDT, ~TRT01A,    ~TRT01P,    ~AGE, ~SEX, ~RACE, ~SAFFL,
  "CDISCPILOT01", "01-001",  19367,   19548,   "Xanomeli", "Xanomeli",  52, "F", "WHITE",    "Y",
  "CDISCPILOT01", "01-002",  19370,   19551,   "Placebo",  "Placebo",   61, "M", "WHITE",    "Y",
  "CDISCPILOT01", "01-003",  19375,   19540,   "Xanomeli", "Xanomeli",  47, "F", "BLACK",    "Y"
)

ae <- tribble(
  ~STUDYID,        ~USUBJID,  ~AESEQ, ~AETERM,            ~AEDECOD,           ~AESOC,
  ~AESTDTC,     ~AEENDTC,     ~AESEV,    ~AEREL,       ~AESER, ~AETOXGR,
  "CDISCPILOT01","01-001", 1, "HEADACHE",         "Headache",          "Nervous system disorders",
  "2024-01-20", "2024-01-22", "MILD",    "UNLIKELY",   "N",    1,
  "CDISCPILOT01","01-001", 2, "NAUSEA",           "Nausea",            "Gastrointestinal disorders",
  "2024-02-01", "2024-02-05", "MODERATE","POSSIBLE",   "N",    2,
  "CDISCPILOT01","01-002", 1, "DIZZINESS",        "Dizziness",         "Nervous system disorders",
  "2024-01-25", "2024-01-28", "MILD",    "UNLIKELY",   "N",    1,
  "CDISCPILOT01","01-002", 2, "ANAPHYLACTIC SHOCK","Anaphylactic shock","Immune system disorders",
  "2024-02-10", NA,           "SEVERE",  "PROBABLE",   "Y",    3,
  "CDISCPILOT01","01-003", 1, "RASH",             "Rash",              "Skin disorders",
  "2024-01-30", "2024-02-03", "MILD",    "POSSIBLE",   "N",    1
)

ex_occ <- tribble(
  ~STUDYID,        ~USUBJID,  ~EXSEQ, ~EXTRT,      ~EXDOSE, ~EXDOSFRQ, ~EXSTDTC,      ~EXENDTC,
  "CDISCPILOT01", "01-001",   1,      "XANOMELI",   54,     "QD",      "2024-01-15",  "2024-02-15",
  "CDISCPILOT01", "01-002",   1,      "PLACEBO",     0,     "QD",      "2024-01-18",  "2024-02-18",
  "CDISCPILOT01", "01-003",   1,      "XANOMELI",   54,     "QD",      "2024-01-23",  "2024-02-20"
) |> mutate(
  EXSTDT = as.numeric(as.Date(EXSTDTC)),
  EXENDT = as.numeric(as.Date(EXENDTC))
)

queries_df <- tribble(
  ~PREFIX, ~GRPNAME,                    ~SCOPE,   ~SCOPEN, ~TERM_NAME,
  "SMQ01", "Hypersensitivity (SMQ)",   "NARROW", 2,       "ANAPHYLACTIC SHOCK",
  "SMQ01", "Hypersensitivity (SMQ)",   "NARROW", 2,       "RASH",
  "CQ02",  "GI Disorders (Customized)", NA,      NA,      "NAUSEA",
  "CQ02",  "GI Disorders (Customized)", NA,      NA,      "VOMITING"
)
```

## Welcome

This lesson covers **OCCDS (Occurrence Data Structure)** ADaMs — ADAE (Adverse Events) and ADCM (Concomitant Medications). You'll build a complete ADAE step by step, learning the treatment-emergent flag, occurrence flags, MedDRA queries, and last-dose linkage.

---

## 1 — OCCDS at a glance

```{r occds-shape-demo}
# OCCDS = one row per event (AE, CM, MH, DV)
# No PARAMCD — events are heterogeneous
ae |> select(USUBJID, AESEQ, AEDECOD, AESOC, AESTDTC, AEENDTC, AESEV, AESER)
```

```{r quiz-occds-shape, echo=FALSE}
quiz(
  question(
    "Which ADaM uses the OCCDS structure?",
    answer("ADVS"),
    answer("ADLB"),
    answer("ADAE", correct = TRUE),
    answer("ADTTE"),
    allow_retry = TRUE
  ),
  question(
    "OCCDS rows represent:",
    answer("One measurement per visit"),
    answer("One time-to-event per subject"),
    answer("One event or occurrence per subject record", correct = TRUE),
    answer("One parameter per study"),
    allow_retry = TRUE
  )
)
```

---

## 2 — Merge ADSL and derive analysis dates

```{r adae-dates-demo}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, AGE, SEX, RACE, SAFFL)

adae <- ae |>
  derive_vars_merged(
    dataset_add = adsl_occ,
    new_vars    = adsl_vars,
    by_vars     = exprs(STUDYID, USUBJID)
  ) |>
  # Start date: partial imputed to first
  derive_vars_dt(
    new_vars_prefix  = "AST",
    dtc              = AESTDTC,
    highest_imputation = "M",
    date_imputation  = "first",
    flag_imputation  = "auto"
  ) |>
  # End date: partial imputed to last (conservative)
  derive_vars_dt(
    new_vars_prefix  = "AEN",
    dtc              = AEENDTC,
    highest_imputation = "M",
    date_imputation  = "last"
  ) |>
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars    = exprs(ASTDT, AENDT)
  )

adae |> select(USUBJID, AETERM, AESTDTC, ASTDT, AENDT, ASTDY) |> head(4)
```

```{r quiz-imputation, echo=FALSE}
quiz(
  question(
    "Why is `date_imputation = 'last'` appropriate for end dates?",
    answer("It makes AVAL larger"),
    answer("A partial end date like '2024-07' means the AE continued through the month — imputing to the last day is the conservative choice", correct = TRUE),
    answer("Admiral requires last for end dates"),
    answer("It matches start date imputation"),
    allow_retry = TRUE
  )
)
```

```{r dates-ex, exercise=TRUE, exercise.lines=18}
# Exercise: starting from ae, merge adsl_occ and derive ASTDT / AENDT
# then compute ASTDY = days from TRTSDT to ASTDT (with +1 convention)
# Confirm the ASTDY for subject 01-001's first AE

adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, AGE, SEX, SAFFL)

adae_tmp <- ae |>
  derive_vars_merged(dataset_add = adsl_occ, new_vars = adsl_vars,
                     by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_dt(new_vars_prefix = "AST", dtc = AESTDTC,
                 highest_imputation = "M", date_imputation = "first") |>
  derive_vars_dt(new_vars_prefix = "AEN", dtc = AEENDTC,
                 highest_imputation = "M", date_imputation = "last") |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ASTDT, AENDT))

adae_tmp |> filter(USUBJID == "01-001") |>
  select(USUBJID, AETERM, ASTDTC = AESTDTC, ASTDT, ASTDY)
```

```{r dates-ex-hint-1}
# derive_vars_dy reads TRTSDT and ASTDT automatically when you pass source_vars = exprs(ASTDT)
```

```{r dates-ex-solution}
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, AGE, SEX, SAFFL)

adae_tmp <- ae |>
  derive_vars_merged(dataset_add = adsl_occ, new_vars = adsl_vars,
                     by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_dt(new_vars_prefix = "AST", dtc = AESTDTC,
                 highest_imputation = "M", date_imputation = "first") |>
  derive_vars_dt(new_vars_prefix = "AEN", dtc = AEENDTC,
                 highest_imputation = "M", date_imputation = "last") |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ASTDT, AENDT))

adae_tmp |> filter(USUBJID == "01-001") |>
  select(USUBJID, AETERM, ASTDTC = AESTDTC, ASTDT, ASTDY)
```

---

## 3 — Severity mapping

```{r severity-demo}
adae <- adae |>
  mutate(
    ASEV  = AESEV,
    AREL  = AEREL,
    ASEVN = case_when(
      ASEV == "MILD"     ~ 1,
      ASEV == "MODERATE" ~ 2,
      ASEV == "SEVERE"   ~ 3,
      TRUE               ~ NA_real_
    )
  )

adae |> select(USUBJID, AETERM, ASEV, ASEVN, AREL)
```

---

## 4 — Treatment-emergent flag: TRTEMFL

TRTEMFL = "Y" for AEs that started **on or after** first dose and on or before last dose + end_window.

```{r trtemfl-demo}
adae <- adae |>
  derive_var_trtemfl(
    start_date      = ASTDT,
    end_date        = AENDT,
    trt_start_date  = TRTSDT,
    trt_end_date    = TRTEDT,
    end_window      = 30
  )

adae |> select(USUBJID, AETERM, ASTDT, TRTSDT, TRTEDT, TRTEMFL)
```

```{r quiz-trtemfl, echo=FALSE}
quiz(
  question(
    "An AE that started before TRTSDT but worsened during treatment: what is TRTEMFL?",
    answer("NA"),
    answer("N"),
    answer("Y — when intensity (grade) arguments are supplied and worsening is detected", correct = TRUE),
    answer("Always NA without the intensity arguments"),
    allow_retry = TRUE
  ),
  question(
    "`end_window = 30` means:",
    answer("The AE must end within 30 days"),
    answer("AEs starting up to 30 days after TRTEDT are still counted as treatment-emergent", correct = TRUE),
    answer("The window is 30 weeks"),
    answer("Only the first 30 AEs are flagged"),
    allow_retry = TRUE
  )
)
```

```{r trtemfl-ex, exercise=TRUE, exercise.lines=16}
# Exercise: apply derive_var_trtemfl with end_window = 0 (strict)
# How many AEs get TRTEMFL = "Y" vs end_window = 30?

adae_strict <- adae |>
  derive_var_trtemfl(
    start_date     = ASTDT,
    end_date       = AENDT,
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT,
    end_window     = ___
  )

# Compare counts
list(
  strict_30  = sum(adae$TRTEMFL == "Y", na.rm = TRUE),
  strict_0   = sum(adae_strict$TRTEMFL == "Y", na.rm = TRUE)
)
```

```{r trtemfl-ex-hint-1}
# end_window = 0
```

```{r trtemfl-ex-solution}
adae_strict <- adae |>
  derive_var_trtemfl(
    start_date     = ASTDT,
    end_date       = AENDT,
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT,
    end_window     = 0
  )

list(
  strict_30 = sum(adae$TRTEMFL == "Y", na.rm = TRUE),
  strict_0  = sum(adae_strict$TRTEMFL == "Y", na.rm = TRUE)
)
```

---

## 5 — Occurrence flags: AOCCFL and AOCCSFL

```{r occ-flags-demo}
# AOCCFL: first treatment-emergent AE per subject
adae <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y"
  )

# AOCCSFL: first treatment-emergent AE per subject per SOC
adae <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, AESOC),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCSFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y"
  )

adae |> select(USUBJID, AETERM, AESOC, TRTEMFL, AOCCFL, AOCCSFL) |> arrange(USUBJID, ASTDT)
```

```{r quiz-occ-flags, echo=FALSE}
quiz(
  question(
    "AOCCFL = 'Y' should equal the number of subjects with at least one treatment-emergent AE. True or false?",
    answer("True", correct = TRUE),
    answer("False — AOCCFL can be 'Y' on multiple rows per subject"),
    allow_retry = TRUE
  ),
  question(
    "AOCCSFL by-group differs from AOCCFL by-group by:",
    answer("AOCCSFL also groups by AESOC (system organ class)", correct = TRUE),
    answer("AOCCSFL uses AETOXGR"),
    answer("AOCCSFL is the last AE, not the first"),
    answer("They are identical"),
    allow_retry = TRUE
  )
)
```

```{r aocc02fl-ex, exercise=TRUE, exercise.lines=18}
# Exercise: derive AOCC02FL = first treatment-emergent AE with AETOXGR >= 2 per subject

adae_02 <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCC02FL,
      mode    = "first"
    ),
    filter = ___ & ___
  )

adae_02 |> filter(!is.na(AOCC02FL)) |>
  select(USUBJID, AETERM, AETOXGR, AOCC02FL)
```

```{r aocc02fl-ex-hint-1}
# filter = TRTEMFL == "Y" & AETOXGR >= 2
```

```{r aocc02fl-ex-solution}
adae_02 <- adae |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCC02FL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y" & AETOXGR >= 2
  )

adae_02 |> filter(!is.na(AOCC02FL)) |>
  select(USUBJID, AETERM, AETOXGR, AOCC02FL)
```

---

## 6 — MedDRA queries: derive_vars_query()

```{r queries-demo}
# queries_df has PREFIX, GRPNAME, TERM_NAME for SMQ01 and CQ02
adae <- adae |>
  derive_vars_query(dataset_queries = queries_df)

adae |> select(USUBJID, AEDECOD, SMQ01NAM, CQ02NAM) |> head(6)
```

```{r quiz-queries, echo=FALSE}
quiz(
  question(
    "What variables does `derive_vars_query()` add to ADAE?",
    answer("MEDDRACD and MEDDRANM"),
    answer("SMQ01NAM, SMQ01CD, CQ02NAM, etc. — named from the PREFIX column of the queries dataset", correct = TRUE),
    answer("AETERM and AEDECOD synonyms"),
    answer("New parameter rows"),
    allow_retry = TRUE
  )
)
```

---

## 7 — Last dose linkage: derive_vars_joined()

```{r last-dose-demo}
# For each AE, attach the most recent prior dose date
adae <- adae |>
  derive_vars_joined(
    dataset_add = ex_occ |> filter(EXDOSE > 0),
    by_vars     = exprs(USUBJID),
    order       = exprs(EXSTDT),
    new_vars    = exprs(LDOSEDT = EXSTDT, LDOSEDOSE = EXDOSE),
    join_vars   = exprs(EXSTDT),
    filter_join = ASTDT >= EXSTDT,
    mode        = "last"
  )

adae |> select(USUBJID, AETERM, ASTDT, LDOSEDT, LDOSEDOSE) |> head(4)
```

```{r quiz-joined, echo=FALSE}
quiz(
  question(
    "Why use `derive_vars_joined()` instead of `derive_vars_merged()` for last-dose linkage?",
    answer("derive_vars_joined() is faster"),
    answer("derive_vars_joined() supports filter conditions that reference both datasets' variables (ASTDT >= EXSTDT)", correct = TRUE),
    answer("derive_vars_merged() cannot rename variables"),
    answer("They are equivalent for this use case"),
    allow_retry = TRUE
  )
)
```

```{r joined-ex, exercise=TRUE, exercise.lines=16}
# Exercise: for each AE, find the FIRST dose that started ON OR AFTER the AE start date
# (i.e., dose started after AE — a different temporal relationship)
# Store the date as NEXTDOSEDT

adae_nd <- adae |>
  derive_vars_joined(
    dataset_add = ex_occ |> filter(EXDOSE > 0),
    by_vars     = exprs(USUBJID),
    order       = exprs(EXSTDT),
    new_vars    = exprs(NEXTDOSEDT = EXSTDT),
    join_vars   = exprs(EXSTDT),
    filter_join = ___ <= EXSTDT,   # AE start <= dose start
    mode        = "first"
  )

adae_nd |> select(USUBJID, AETERM, ASTDT, NEXTDOSEDT) |> head(4)
```

```{r joined-ex-hint-1}
# filter_join = ASTDT <= EXSTDT
```

```{r joined-ex-solution}
adae_nd <- adae |>
  derive_vars_joined(
    dataset_add = ex_occ |> filter(EXDOSE > 0),
    by_vars     = exprs(USUBJID),
    order       = exprs(EXSTDT),
    new_vars    = exprs(NEXTDOSEDT = EXSTDT),
    join_vars   = exprs(EXSTDT),
    filter_join = ASTDT <= EXSTDT,
    mode        = "first"
  )

adae_nd |> select(USUBJID, AETERM, ASTDT, NEXTDOSEDT) |> head(4)
```

---

## 8 — ASEQ and create_single_dose_dataset

```{r aseq-demo}
adae <- adae |>
  derive_var_obs_number(
    new_var  = ASEQ,
    by_vars  = exprs(STUDYID, USUBJID),
    order    = exprs(ASTDT, AESEQ),
    check_type = "warning"
  )

adae |> select(USUBJID, AETERM, ASTDT, AESEQ, ASEQ) |> arrange(USUBJID, ASEQ)
```

```{r single-dose-demo}
# Expand QD dosing to one row per day
ex_single <- ex_occ |>
  create_single_dose_dataset(
    dose_freq        = EXDOSFRQ,
    start_date       = EXSTDT,
    end_date         = EXENDT,
    keep_source_vars = exprs(USUBJID, EXSTDT, EXENDT, EXDOSFRQ, EXDOSE)
  )

# Compare row counts
cat("Original EX rows:", nrow(ex_occ), "\n")
cat("Expanded rows:   ", nrow(ex_single), "\n")
```

```{r quiz-aseq-single, echo=FALSE}
quiz(
  question(
    "What does `create_single_dose_dataset()` do?",
    answer("Aggregates multiple doses per day into one"),
    answer("Expands summary EX records (e.g., QD from Day 1 to Day 15) to one row per dose date", correct = TRUE),
    answer("Creates a new ADaM dataset"),
    answer("Derives the number of doses"),
    allow_retry = TRUE
  )
)
```

---

## 9 — Complete ADAE pipeline

```{r capstone-adae, exercise=TRUE, exercise.lines=60}
# Build ADAE from ae + adsl_occ + ex_occ in one clean pipe
# Steps: merge ADSL → dates → severity → TRTEMFL → AOCCFL → AOCCSFL → MedDRA queries → ASEQ

adsl_v <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P, AGE, SEX, RACE, SAFFL)

adae_full <- ae |>
  derive_vars_merged(dataset_add = adsl_occ, new_vars = adsl_v,
                     by_vars = exprs(STUDYID, USUBJID)) |>
  derive_vars_dt(new_vars_prefix = "AST", dtc = AESTDTC,
                 highest_imputation = "M", date_imputation = "first") |>
  derive_vars_dt(new_vars_prefix = "AEN", dtc = AEENDTC,
                 highest_imputation = "M", date_imputation = "last") |>
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ASTDT, AENDT)) |>
  mutate(
    ASEV  = AESEV, AREL = AEREL,
    ASEVN = case_when(ASEV == "MILD" ~ 1, ASEV == "MODERATE" ~ 2, ASEV == "SEVERE" ~ 3,
                      TRUE ~ NA_real_)
  ) |>
  derive_var_trtemfl(start_date = ASTDT, end_date = AENDT,
                     trt_start_date = TRTSDT, trt_end_date = TRTEDT, end_window = 30) |>
  derive_var_ontrtfl(start_date = ASTDT, ref_start_date = TRTSDT,
                     ref_end_date = TRTEDT, ref_end_window = 30) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(by_vars = exprs(STUDYID, USUBJID),
                  order = exprs(ASTDT, AESEQ), new_var = AOCCFL, mode = "first"),
    filter = TRTEMFL == "Y"
  ) |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(by_vars = exprs(STUDYID, USUBJID, AESOC),
                  order = exprs(ASTDT, AESEQ), new_var = AOCCSFL, mode = "first"),
    filter = TRTEMFL == "Y"
  ) |>
  derive_vars_query(dataset_queries = queries_df) |>
  derive_var_obs_number(new_var = ASEQ, by_vars = exprs(STUDYID, USUBJID),
                        order = exprs(ASTDT, AESEQ))

adae_full |>
  select(USUBJID, AETERM, TRTEMFL, AOCCFL, AOCCSFL, SMQ01NAM, ASEQ) |>
  arrange(USUBJID, ASEQ)
```

---

## Quiz — final

```{r quiz-final-17, echo=FALSE}
quiz(
  question(
    "What does TRTEMFL stand for, and what does 'Y' mean?",
    answer("Treatment Exposure Flag; subject received treatment"),
    answer("Treatment-Emergent Flag; the AE was new or worsened during the treatment window", correct = TRUE),
    answer("Time-resolved event flag"),
    answer("Tolerability flag"),
    allow_retry = TRUE
  ),
  question(
    "ADCM differs from ADAE mainly in:",
    answer("It uses OCCDS structure like ADAE, but maps ATC codes from WHODrug instead of MedDRA terms", correct = TRUE),
    answer("ADCM has PARAMCD; ADAE does not"),
    answer("ADCM has no treatment-emergent flag"),
    answer("ADCM uses BDS structure"),
    allow_retry = TRUE
  ),
  question(
    "The function for ATC class hierarchy derivation in ADCM is:",
    answer("derive_vars_query()"),
    answer("derive_vars_transposed()"),
    answer("derive_vars_atc()", correct = TRUE),
    answer("create_single_dose_dataset()"),
    allow_retry = TRUE
  ),
  question(
    "ASEQ provides:",
    answer("The sequence number of AE records per subject", correct = TRUE),
    answer("The severity sequence"),
    answer("The chronological position of TRTEMFL"),
    answer("A link to ADTTE"),
    allow_retry = TRUE
  )
)
```
