---
title: "Lesson 18 — admiral Part 5: Time-to-Event ADaMs (ADTTE)"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tibble)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── shared mocks ─────────────────────────────────────────────────────────────
exprs <- function(...) {
  dots <- as.list(substitute(list(...)))[-1]; dots
}

params <- function(...) list(...)

convert_blanks_to_na <- function(dataset) {
  dataset |> mutate(across(where(is.character), ~ifelse(.=="", NA_character_, .)))
}

# ── ADTTE source object constructors ─────────────────────────────────────────
event_source <- function(dataset_name, filter = NULL, date, set_values_to = list()) {
  list(
    type         = "event",
    dataset_name = dataset_name,
    filter       = substitute(filter),
    date         = substitute(date),
    set_values_to = set_values_to,
    censor       = 0
  )
}

censor_source <- function(dataset_name, filter = NULL, date, set_values_to = list(),
                           censor = 1) {
  list(
    type         = "censor",
    dataset_name = dataset_name,
    filter       = substitute(filter),
    date         = substitute(date),
    set_values_to = set_values_to,
    censor       = censor
  )
}

# ── derive_param_tte ──────────────────────────────────────────────────────────
derive_param_tte <- function(dataset_adsl = NULL,
                              start_date,
                              event_conditions,
                              censor_conditions,
                              source_datasets,
                              set_values_to = list(),
                              ...) {
  start_nm <- deparse(substitute(start_date))
  adsl_use <- if (!is.null(dataset_adsl)) dataset_adsl else
              source_datasets[[names(source_datasets)[1]]]
  # helper to evaluate one source condition for all subjects
  eval_source <- function(src) {
    ds <- source_datasets[[src$dataset_name]]
    if (is.null(ds)) return(NULL)
    # apply filter
    if (!is.null(src$filter)) {
      ds <- tryCatch(filter(ds, !!src$filter), error = function(e) ds)
    }
    # evaluate date expression
    date_nm <- deparse(src$date)
    if (!date_nm %in% names(ds)) return(NULL)
    ds$..DATE.. <- ds[[date_nm]]
    ds <- ds |> filter(!is.na(..DATE..))
    if (nrow(ds) == 0) return(NULL)
    # keep just USUBJID + date + meta
    out <- ds |> select(any_of(c("USUBJID", "STUDYID", "..DATE..")))
    out$..CENSOR.. <- src$censor
    for (nm in names(src$set_values_to)) {
      val <- tryCatch(eval(src$set_values_to[[nm]]), error = function(e) as.character(src$set_values_to[[nm]]))
      out[[nm]] <- val
    }
    out
  }
  # Process events: earliest event date per subject
  event_rows <- lapply(event_conditions, eval_source)
  event_rows <- Filter(Negate(is.null), event_rows)
  event_all  <- if (length(event_rows)) bind_rows(event_rows) else NULL

  # Process censorings: latest censor date per subject
  censor_rows <- lapply(censor_conditions, eval_source)
  censor_rows <- Filter(Negate(is.null), censor_rows)
  censor_all  <- if (length(censor_rows)) bind_rows(censor_rows) else NULL

  # Determine ADT and CNSR for each subject
  subjects <- unique(adsl_use$USUBJID)
  result   <- vector("list", length(subjects))

  for (i in seq_along(subjects)) {
    subj     <- subjects[i]
    adsl_row <- adsl_use |> filter(USUBJID == subj)
    startdt  <- if (start_nm %in% names(adsl_row)) adsl_row[[start_nm]][1] else NA_real_

    # Events for this subject
    ev_sub <- if (!is.null(event_all)) filter(event_all, USUBJID == subj) else NULL
    # Censor for this subject
    ce_sub <- if (!is.null(censor_all)) filter(censor_all, USUBJID == subj) else NULL

    if (!is.null(ev_sub) && nrow(ev_sub) > 0) {
      # Take earliest event
      ev_sub  <- arrange(ev_sub, ..DATE..)
      best    <- ev_sub[1, , drop = FALSE]
      adt     <- best$..DATE..[1]
      cnsr    <- 0
      evnt    <- best$EVNTDESC[1] %||% NA_character_
      srcdom  <- best$SRCDOM[1]   %||% NA_character_
      srcvar  <- best$SRCVAR[1]   %||% NA_character_
    } else if (!is.null(ce_sub) && nrow(ce_sub) > 0) {
      # Take latest censor date
      ce_sub <- arrange(ce_sub, desc(..DATE..))
      best   <- ce_sub[1, , drop = FALSE]
      adt    <- best$..DATE..[1]
      cnsr   <- best$..CENSOR..[1]
      evnt   <- best$EVNTDESC[1] %||% NA_character_
      srcdom <- best$SRCDOM[1]   %||% NA_character_
      srcvar <- best$SRCVAR[1]   %||% NA_character_
    } else {
      adt <- NA_real_; cnsr <- NA_real_; evnt <- NA_character_
      srcdom <- NA_character_; srcvar <- NA_character_
    }

    aval <- if (!is.na(adt) && !is.na(startdt)) as.numeric(adt - startdt) + 1 else NA_real_

    row_out <- adsl_row |>
      mutate(STARTDT = startdt, ADT = adt, AVAL = aval, CNSR = cnsr,
             EVNTDESC = evnt, SRCDOM = srcdom, SRCVAR = srcvar)
    result[[i]] <- row_out
  }

  tte_rows <- bind_rows(result)
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]]), error = function(e) as.character(set_values_to[[nm]]))
    tte_rows[[nm]] <- val
  }
  tte_rows
}

# call_derivation for TTE
call_derivation <- function(dataset = NULL, derivation, variable_params, ...) {
  fixed_args <- list(...)
  result_list <- vector("list", length(variable_params))
  for (i in seq_along(variable_params)) {
    call_args <- c(fixed_args, variable_params[[i]])
    result_list[[i]] <- tryCatch(do.call(derivation, call_args), error = function(e) NULL)
  }
  bind_rows(Filter(Negate(is.null), result_list))
}

`%||%` <- function(a, b) if (is.null(a) || (length(a) == 1 && is.na(a))) b else a

# ── Sample datasets ───────────────────────────────────────────────────────────
adsl_tte <- tribble(
  ~STUDYID,        ~USUBJID,  ~TRTSDT, ~TRTEDT, ~RANDDT, ~TRT01A,
  ~DTHDT,  ~LSTALVDT, ~EOSDT,  ~DTHFL, ~SAFFL,
  "CDISCPILOT01","01-001", 19367, 19548,  19360,  "Xanomeli",
  NA_real_, 19600,     19600, "N",    "Y",
  "CDISCPILOT01","01-002", 19370, 19430,  19365,  "Placebo",
  19480,    19480,     19480, "Y",    "Y",
  "CDISCPILOT01","01-003", 19375, 19540,  19370,  "Xanomeli",
  NA_real_, 19590,     19590, "N",    "Y",
  "CDISCPILOT01","01-004", 19378, 19490,  19372,  "Placebo",
  19510,    19510,     19510, "Y",    "Y",
  "CDISCPILOT01","01-005", 19380, 19560,  19375,  "Xanomeli",
  NA_real_, 19610,     19610, "N",    "Y"
)

adrs <- tribble(
  ~STUDYID,        ~USUBJID,  ~PARAMCD, ~AVALC, ~ADT,
  "CDISCPILOT01", "01-001",  "OVR",    "SD",   19450,
  "CDISCPILOT01", "01-001",  "OVR",    "PD",   19520,
  "CDISCPILOT01", "01-002",  "OVR",    "PR",   19410,
  "CDISCPILOT01", "01-003",  "OVR",    "SD",   19440,
  "CDISCPILOT01", "01-004",  "OVR",    "PD",   19460,
  "CDISCPILOT01", "01-005",  "OVR",    "CR",   19460
)

adae_tte <- tribble(
  ~STUDYID,        ~USUBJID,  ~AETERM,              ~ASTDT, ~TRTEMFL, ~AESER, ~TRTSDT,
  "CDISCPILOT01", "01-001",  "ANAPHYLACTIC SHOCK",  19400,  "Y",     "Y",    19367,
  "CDISCPILOT01", "01-003",  "MYOCARDIAL INFARCTION",19420, "Y",     "Y",    19375
)
```

## Welcome

This lesson covers **ADTTE — Time-to-Event analysis datasets**. You'll learn admiral's declarative `event_source()` / `censor_source()` design and use `derive_param_tte()` to build Overall Survival, Progression-Free Survival, and safety TTE parameters.

---

## 1 — The ADTTE shape

```{r adtte-shape-demo}
# One row per (subject × parameter)
# Key variables: STARTDT, ADT, AVAL, CNSR, EVNTDESC, SRCDOM, SRCVAR
adsl_tte |>
  select(USUBJID, TRTSDT, DTHDT, LSTALVDT, DTHFL) |>
  arrange(USUBJID)
```

```{r quiz-adtte-shape, echo=FALSE}
quiz(
  question(
    "CNSR = 0 means:",
    answer("The subject was censored (no event observed by ADT)"),
    answer("The event occurred", correct = TRUE),
    answer("ADT is missing"),
    answer("The subject was not in the analysis"),
    allow_retry = TRUE
  ),
  question(
    "AVAL in ADTTE is typically:",
    answer("The analysis value of a measurement"),
    answer("ADT minus STARTDT plus 1 (time on study in days)", correct = TRUE),
    answer("CNSR × 100"),
    answer("A numeric version of PARAMCD"),
    allow_retry = TRUE
  )
)
```

---

## 2 — event_source() and censor_source()

```{r source-objects-demo}
# Define event: death
death_event <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(
    EVNTDESC = "DEATH",
    SRCDOM   = "ADSL",
    SRCVAR   = "DTHDT"
  )
)

# Define censoring: last known alive
last_alive_censor <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(
    EVNTDESC = "LAST DATE KNOWN ALIVE",
    SRCDOM   = "ADSL",
    SRCVAR   = "LSTALVDT"
  )
)

# Inspect structure
str(death_event)
```

```{r quiz-source-objects, echo=FALSE}
quiz(
  question(
    "What field in the source object says which actual dataset to pull data from?",
    answer("filter"),
    answer("date"),
    answer("dataset_name", correct = TRUE),
    answer("set_values_to"),
    allow_retry = TRUE
  ),
  question(
    "What differentiates an event_source() from a censor_source()?",
    answer("event_source() has a filter; censor_source() doesn't"),
    answer("event_source() produces CNSR = 0; censor_source() produces CNSR = 1+", correct = TRUE),
    answer("censor_source() uses a different date variable"),
    answer("They are identical — only the name differs"),
    allow_retry = TRUE
  )
)
```

---

## 3 — Overall Survival with derive_param_tte()

```{r os-demo}
# Build OS: event = death, censor = last alive date
adtte_os <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = TRTSDT,
  event_conditions  = list(death_event),
  censor_conditions = list(last_alive_censor),
  source_datasets   = list(adsl = adsl_tte),
  set_values_to     = exprs(
    PARAMCD = "OS",
    PARAM   = "Overall Survival"
  )
)

adtte_os |>
  select(USUBJID, PARAMCD, STARTDT, ADT, AVAL, CNSR, EVNTDESC)
```

```{r quiz-os, echo=FALSE}
quiz(
  question(
    "For a subject with DTHDT = 19480 and TRTSDT = 19370, what is AVAL?",
    answer("19480 - 19370 = 110"),
    answer("19480 - 19370 + 1 = 111", correct = TRUE),
    answer("110 / 365.25"),
    answer("1"),
    allow_retry = TRUE
  )
)
```

```{r os-ex, exercise=TRUE, exercise.lines=22}
# Exercise: build OS using adsl_tte
# event = !is.na(DTHDT), date = DTHDT
# censor = date = LSTALVDT
# STARTDT = RANDDT (not TRTSDT)

death_ev <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = ___,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST KNOWN ALIVE", SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

adtte_os2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = ___,
  event_conditions  = list(death_ev),
  censor_conditions = list(alive_cs),
  source_datasets   = list(adsl = adsl_tte),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

adtte_os2 |> select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

```{r os-ex-hint-1}
# date = DTHDT
# start_date = RANDDT
```

```{r os-ex-solution}
death_ev <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST KNOWN ALIVE", SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

adtte_os2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(death_ev),
  censor_conditions = list(alive_cs),
  source_datasets   = list(adsl = adsl_tte),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

adtte_os2 |> select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

---

## 4 — PFS: composite events

PFS uses **two event sources**: disease progression OR death, whichever comes first.

```{r pfs-demo}
# Progression event
pd_event <- event_source(
  dataset_name  = "adrs",
  filter        = PARAMCD == "OVR" & AVALC == "PD",
  date          = ADT,
  set_values_to = exprs(
    EVNTDESC = "DISEASE PROGRESSION",
    SRCDOM   = "ADRS",
    SRCVAR   = "ADT"
  )
)

# Death (reuse from above)
# last tumor assessment censor
last_assess <- censor_source(
  dataset_name  = "adrs",
  filter        = PARAMCD == "OVR" & AVALC %in% c("CR","PR","SD","NE"),
  date          = ADT,
  set_values_to = exprs(
    EVNTDESC = "LAST TUMOR ASSESSMENT",
    SRCDOM   = "ADRS",
    SRCVAR   = "ADT"
  )
)

adtte_pfs <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
)

adtte_pfs |>
  select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

```{r quiz-pfs, echo=FALSE}
quiz(
  question(
    "When both progression AND death occur for the same subject, derive_param_tte() picks:",
    answer("Death always wins"),
    answer("Progression always wins"),
    answer("The earliest of the two event dates", correct = TRUE),
    answer("The one listed first in event_conditions"),
    allow_retry = TRUE
  ),
  question(
    "A subject with no progression and no death gets CNSR =",
    answer("0"),
    answer("1 (from the censor source)", correct = TRUE),
    answer("2 — double censor"),
    answer("NA"),
    allow_retry = TRUE
  )
)
```

```{r pfs-ex, exercise=TRUE, exercise.lines=26}
# Exercise: build PFS but add a randomization-date fallback censor
# (for subjects with no assessments at all, censor = 2)
# Then count subjects by CNSR value

rand_censor <- censor_source(
  dataset_name  = "adsl",
  date          = RANDDT,
  censor        = ___,        # use censor = 2
  set_values_to = exprs(EVNTDESC = "RANDOMIZATION (NO ASSESSMENT)",
                         SRCDOM = "ADSL", SRCVAR = "RANDDT")
)

adtte_pfs2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor, rand_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

# Count by CNSR
adtte_pfs2 |> count(CNSR)
```

```{r pfs-ex-hint-1}
# censor = 2
```

```{r pfs-ex-solution}
rand_censor <- censor_source(
  dataset_name  = "adsl",
  date          = RANDDT,
  censor        = 2,
  set_values_to = exprs(EVNTDESC = "RANDOMIZATION (NO ASSESSMENT)",
                         SRCDOM = "ADSL", SRCVAR = "RANDDT")
)

adtte_pfs2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor, rand_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

adtte_pfs2 |> count(CNSR)
```

---

## 5 — Time to first serious AE

```{r ttaeser-demo}
# Event: first treatment-emergent serious AE
first_sae_event <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y" & TRTEMFL == "Y",
  date          = ASTDT,
  set_values_to = exprs(
    EVNTDESC = "FIRST SERIOUS AE",
    SRCDOM   = "ADAE",
    SRCVAR   = "ASTDT"
  )
)

eot_censor <- censor_source(
  dataset_name  = "adsl",
  date          = EOSDT,
  set_values_to = exprs(
    EVNTDESC = "END OF STUDY",
    SRCDOM   = "ADSL",
    SRCVAR   = "EOSDT"
  )
)

adtte_sae <- derive_param_tte(
  dataset_adsl      = adsl_tte |> filter(SAFFL == "Y"),
  start_date        = TRTSDT,
  event_conditions  = list(first_sae_event),
  censor_conditions = list(eot_censor, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adae = adae_tte),
  set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
)

adtte_sae |> select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

---

## 6 — call_derivation() for multiple parameters

```{r call-derivation-demo}
# Build OS and PFS together in one call_derivation call
adtte_both <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions  = list(death_event),
      censor_conditions = list(last_alive_censor),
      set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
    ),
    params(
      event_conditions  = list(pd_event, death_event),
      censor_conditions = list(last_assess, last_alive_censor),
      set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
    )
  ),
  # Shared arguments
  dataset_adsl    = adsl_tte,
  start_date      = RANDDT,
  source_datasets = list(adsl = adsl_tte, adrs = adrs)
)

adtte_both |>
  select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC) |>
  arrange(PARAMCD, USUBJID)
```

```{r quiz-call-derivation, echo=FALSE}
quiz(
  question(
    "In `call_derivation()`, the `variable_params` list contains:",
    answer("Shared arguments applied to every call"),
    answer("Per-call argument overrides (event_conditions, censor_conditions, set_values_to)", correct = TRUE),
    answer("Dataset names only"),
    answer("Filter conditions"),
    allow_retry = TRUE
  )
)
```

```{r call-derivation-ex, exercise=TRUE, exercise.lines=28}
# Exercise: use call_derivation to build OS and TTAESER in one block
# OS: death event, last alive censor, start = TRTSDT
# TTAESER: first_sae_event, eot_censor + last_alive_censor, start = TRTSDT

adtte_multi <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions  = list(death_event),
      censor_conditions = list(last_alive_censor),
      set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
    ),
    params(
      event_conditions  = list(first_sae_event),
      censor_conditions = list(eot_censor, last_alive_censor),
      set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
    )
  ),
  dataset_adsl    = adsl_tte |> filter(SAFFL == "Y"),
  start_date      = ___,
  source_datasets = list(adsl = adsl_tte, adae = adae_tte)
)

adtte_multi |> select(USUBJID, PARAMCD, AVAL, CNSR) |> arrange(PARAMCD, USUBJID)
```

```{r call-derivation-ex-hint-1}
# start_date = TRTSDT
```

```{r call-derivation-ex-solution}
adtte_multi <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions  = list(death_event),
      censor_conditions = list(last_alive_censor),
      set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
    ),
    params(
      event_conditions  = list(first_sae_event),
      censor_conditions = list(eot_censor, last_alive_censor),
      set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
    )
  ),
  dataset_adsl    = adsl_tte |> filter(SAFFL == "Y"),
  start_date      = TRTSDT,
  source_datasets = list(adsl = adsl_tte, adae = adae_tte)
)

adtte_multi |> select(USUBJID, PARAMCD, AVAL, CNSR) |> arrange(PARAMCD, USUBJID)
```

---

## 7 — AVAL in months/years and validation

```{r aval-units-demo}
# AVAL in days → months conversion
adtte_os |>
  mutate(
    AVAL_MONTHS = AVAL / (365.25 / 12),
    AVAL_YEARS  = AVAL / 365.25
  ) |>
  select(USUBJID, PARAMCD, AVAL, AVAL_MONTHS, AVAL_YEARS, CNSR)
```

```{r quiz-validation, echo=FALSE}
quiz(
  question(
    "Which CNSR distribution check is a key ADTTE validation step?",
    answer("Confirming no negative AVAL values (AVAL should always be ≥ 1)", correct = TRUE),
    answer("Confirming CNSR sums to zero"),
    answer("Confirming every subject has exactly one PARAMCD"),
    answer("Confirming ADT > STARTDT for all rows"),
    allow_retry = TRUE
  ),
  question(
    "Which check ensures completeness of the subject population?",
    answer("count(CNSR)"),
    answer("Every subject in dataset_adsl (filtered population) has exactly one row per PARAMCD", correct = TRUE),
    answer("nrow() matches ADSL"),
    answer("SRCDOM is non-missing"),
    allow_retry = TRUE
  )
)
```

---

## Quiz — final

```{r quiz-final-18, echo=FALSE}
quiz(
  question(
    "PFS is a 'composite endpoint' because:",
    answer("It requires two datasets"),
    answer("Either progression OR death qualifies as the event — two event_source objects", correct = TRUE),
    answer("It uses two censor_source objects"),
    answer("It has two CNSR values"),
    allow_retry = TRUE
  ),
  question(
    "EVNTDESC serves what purpose in ADTTE?",
    answer("It is required by CDISC as a key variable"),
    answer("It records which specific event or censoring rule fired — providing provenance and auditability", correct = TRUE),
    answer("It replaces PARAMCD"),
    answer("It maps to MedDRA terms"),
    allow_retry = TRUE
  ),
  question(
    "A `censor_source()` with `censor = 2` is useful for:",
    answer("Marking invalid censorings"),
    answer("Distinguishing different censoring reasons (e.g., last-alive vs. randomization fallback)", correct = TRUE),
    answer("Increasing AVAL"),
    answer("Flagging missing DTHDT"),
    allow_retry = TRUE
  ),
  question(
    "What does `{admiralonco}` provide for TTE derivations?",
    answer("A faster derive_param_tte() for large datasets"),
    answer("Pre-defined event_source() and censor_source() objects for oncology endpoints (OS, PFS, DOR)", correct = TRUE),
    answer("A ggplot2 extension for Kaplan-Meier plots"),
    answer("CTCAE toxicity grades"),
    allow_retry = TRUE
  )
)
```
