---
title: "Lesson 18 — admiral Part 5: Time-to-Event ADaMs (ADTTE)"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 18_admiral_part5_adtte.Rmd
# PURPOSE       : admiral Part 5 — ADTTE time-to-event datasets: derive_param_tte(), event
#                 and censoring source definitions, Kaplan-Meier
#                 readiness checks, and common oncology/survival
#                 endpoints.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tibble)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── shared mocks ─────────────────────────────────────────────────────────────
exprs <- function(...) {
  dots <- as.list(substitute(list(...)))[-1]; dots
}

params <- function(...) list(...)

convert_blanks_to_na <- function(dataset) {
  dataset |> mutate(across(where(is.character), ~ifelse(.=="", NA_character_, .)))
}

# ── ADTTE source object constructors ─────────────────────────────────────────
event_source <- function(dataset_name, filter = NULL, date, set_values_to = list()) {
  list(
    type         = "event",
    dataset_name = dataset_name,
    filter       = substitute(filter),
    date         = substitute(date),
    set_values_to = set_values_to,
    censor       = 0
  )
}

censor_source <- function(dataset_name, filter = NULL, date, set_values_to = list(),
                           censor = 1) {
  list(
    type         = "censor",
    dataset_name = dataset_name,
    filter       = substitute(filter),
    date         = substitute(date),
    set_values_to = set_values_to,
    censor       = censor
  )
}

# ── derive_param_tte ──────────────────────────────────────────────────────────
derive_param_tte <- function(dataset_adsl = NULL,
                              start_date,
                              event_conditions,
                              censor_conditions,
                              source_datasets,
                              set_values_to = list(),
                              ...) {
  .e <- substitute(start_date)
  start_nm <- if (is.character(.e)) .e else deparse(.e)
  adsl_use <- if (!is.null(dataset_adsl)) dataset_adsl else
              source_datasets[[names(source_datasets)[1]]]
  # helper to evaluate one source condition for all subjects
  eval_source <- function(src) {
    ds <- source_datasets[[src$dataset_name]]
    if (is.null(ds)) return(NULL)
    # apply filter
    if (!is.null(src$filter)) {
      ds <- tryCatch(filter(ds, !!src$filter), error = function(e) ds)
    }
    # evaluate date expression
    date_nm <- deparse(src$date)
    if (!date_nm %in% names(ds)) return(NULL)
    ds$..DATE.. <- ds[[date_nm]]
    ds <- ds |> filter(!is.na(..DATE..))
    if (nrow(ds) == 0) return(NULL)
    # keep just USUBJID + date + meta
    out <- ds |> select(any_of(c("USUBJID", "STUDYID", "..DATE..")))
    out$..CENSOR.. <- src$censor
    for (nm in names(src$set_values_to)) {
      val <- tryCatch(eval(src$set_values_to[[nm]]), error = function(e) as.character(src$set_values_to[[nm]]))
      out[[nm]] <- val
    }
    out
  }
  # Process events: earliest event date per subject
  event_rows <- lapply(event_conditions, eval_source)
  event_rows <- Filter(Negate(is.null), event_rows)
  event_all  <- if (length(event_rows)) bind_rows(event_rows) else NULL

  # Process censorings: latest censor date per subject
  censor_rows <- lapply(censor_conditions, eval_source)
  censor_rows <- Filter(Negate(is.null), censor_rows)
  censor_all  <- if (length(censor_rows)) bind_rows(censor_rows) else NULL

  # Determine ADT and CNSR for each subject
  subjects <- unique(adsl_use$USUBJID)
  result   <- vector("list", length(subjects))

  for (i in seq_along(subjects)) {
    subj     <- subjects[i]
    adsl_row <- adsl_use |> filter(USUBJID == subj)
    startdt  <- if (start_nm %in% names(adsl_row)) adsl_row[[start_nm]][1] else NA_real_

    # Events for this subject
    ev_sub <- if (!is.null(event_all)) filter(event_all, USUBJID == subj) else NULL
    # Censor for this subject
    ce_sub <- if (!is.null(censor_all)) filter(censor_all, USUBJID == subj) else NULL

    if (!is.null(ev_sub) && nrow(ev_sub) > 0) {
      # Take earliest event
      ev_sub  <- arrange(ev_sub, ..DATE..)
      best    <- ev_sub[1, , drop = FALSE]
      adt     <- best$..DATE..[1]
      cnsr    <- 0
      evnt    <- best$EVNTDESC[1] %||% NA_character_
      srcdom  <- best$SRCDOM[1]   %||% NA_character_
      srcvar  <- best$SRCVAR[1]   %||% NA_character_
    } else if (!is.null(ce_sub) && nrow(ce_sub) > 0) {
      # Take latest censor date
      ce_sub <- arrange(ce_sub, desc(..DATE..))
      best   <- ce_sub[1, , drop = FALSE]
      adt    <- best$..DATE..[1]
      cnsr   <- best$..CENSOR..[1]
      evnt   <- best$EVNTDESC[1] %||% NA_character_
      srcdom <- best$SRCDOM[1]   %||% NA_character_
      srcvar <- best$SRCVAR[1]   %||% NA_character_
    } else {
      adt <- NA_real_; cnsr <- NA_real_; evnt <- NA_character_
      srcdom <- NA_character_; srcvar <- NA_character_
    }

    aval <- if (!is.na(adt) && !is.na(startdt)) as.numeric(adt - startdt) + 1 else NA_real_

    row_out <- adsl_row |>
      mutate(STARTDT = startdt, ADT = adt, AVAL = aval, CNSR = cnsr,
             EVNTDESC = evnt, SRCDOM = srcdom, SRCVAR = srcvar)
    result[[i]] <- row_out
  }

  tte_rows <- bind_rows(result)
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]]), error = function(e) as.character(set_values_to[[nm]]))
    tte_rows[[nm]] <- val
  }
  tte_rows
}

# call_derivation for TTE
call_derivation <- function(dataset = NULL, derivation, variable_params, ...) {
  fixed_args <- list(...)
  result_list <- vector("list", length(variable_params))
  for (i in seq_along(variable_params)) {
    call_args <- c(fixed_args, variable_params[[i]])
    result_list[[i]] <- tryCatch(do.call(derivation, call_args), error = function(e) NULL)
  }
  bind_rows(Filter(Negate(is.null), result_list))
}

`%||%` <- function(a, b) if (is.null(a) || (length(a) == 1 && is.na(a))) b else a

# ── Sample datasets ───────────────────────────────────────────────────────────
adsl_tte <- tribble(
  ~STUDYID,        ~USUBJID,  ~TRTSDT, ~TRTEDT, ~RANDDT, ~TRT01A,
  ~DTHDT,  ~LSTALVDT, ~EOSDT,  ~DTHFL, ~SAFFL,
  "CDISCPILOT01","01-001", 19367, 19548,  19360,  "Xanomeli",
  NA_real_, 19600,     19600, "N",    "Y",
  "CDISCPILOT01","01-002", 19370, 19430,  19365,  "Placebo",
  19480,    19480,     19480, "Y",    "Y",
  "CDISCPILOT01","01-003", 19375, 19540,  19370,  "Xanomeli",
  NA_real_, 19590,     19590, "N",    "Y",
  "CDISCPILOT01","01-004", 19378, 19490,  19372,  "Placebo",
  19510,    19510,     19510, "Y",    "Y",
  "CDISCPILOT01","01-005", 19380, 19560,  19375,  "Xanomeli",
  NA_real_, 19610,     19610, "N",    "Y"
)

adrs <- tribble(
  ~STUDYID,        ~USUBJID,  ~PARAMCD, ~AVALC, ~ADT,
  "CDISCPILOT01", "01-001",  "OVR",    "SD",   19450,
  "CDISCPILOT01", "01-001",  "OVR",    "PD",   19520,
  "CDISCPILOT01", "01-002",  "OVR",    "PR",   19410,
  "CDISCPILOT01", "01-003",  "OVR",    "SD",   19440,
  "CDISCPILOT01", "01-004",  "OVR",    "PD",   19460,
  "CDISCPILOT01", "01-005",  "OVR",    "CR",   19460
)

adae_tte <- tribble(
  ~STUDYID,        ~USUBJID,  ~AETERM,              ~ASTDT, ~TRTEMFL, ~AESER, ~TRTSDT,
  "CDISCPILOT01", "01-001",  "ANAPHYLACTIC SHOCK",  19400,  "Y",     "Y",    19367,
  "CDISCPILOT01", "01-003",  "MYOCARDIAL INFARCTION",19420, "Y",     "Y",    19375
)

# ── Real pharmaverseadam ADSL for real-data OS exercise ───────────────────────
library(pharmaverseadam)
adsl_real <- pharmaverseadam::adsl
```

## Welcome

This lesson covers **ADTTE — Time-to-Event analysis datasets**. You'll learn admiral's declarative `event_source()` / `censor_source()` design and use `derive_param_tte()` to build Overall Survival, Progression-Free Survival, and safety TTE parameters.

---

## 1 — The ADTTE shape

```{r adtte-shape-demo}
# One row per (subject × parameter)
# Key variables: STARTDT, ADT, AVAL, CNSR, EVNTDESC, SRCDOM, SRCVAR
adsl_tte |>
  select(USUBJID, TRTSDT, DTHDT, LSTALVDT, DTHFL) |>
  arrange(USUBJID)
```

```{r quiz-adtte-shape, echo=FALSE}
quiz(
  question(
    "CNSR = 0 means:",
    answer("The subject was censored (no event observed by ADT)"),
    answer("The event occurred", correct = TRUE),
    answer("ADT is missing"),
    answer("The subject was not in the analysis"),
    allow_retry = TRUE
  ),
  question(
    "AVAL in ADTTE is typically:",
    answer("The analysis value of a measurement"),
    answer("ADT minus STARTDT plus 1 (time on study in days)", correct = TRUE),
    answer("CNSR × 100"),
    answer("A numeric version of PARAMCD"),
    allow_retry = TRUE
  )
)
```

---

## 2 — event_source() and censor_source()

```{r source-objects-demo}
# Define event: death
death_event <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(
    EVNTDESC = "DEATH",
    SRCDOM   = "ADSL",
    SRCVAR   = "DTHDT"
  )
)

# Define censoring: last known alive
last_alive_censor <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(
    EVNTDESC = "LAST DATE KNOWN ALIVE",
    SRCDOM   = "ADSL",
    SRCVAR   = "LSTALVDT"
  )
)

# Inspect structure
str(death_event)
```

```{r quiz-source-objects, echo=FALSE}
quiz(
  question(
    "What field in the source object says which actual dataset to pull data from?",
    answer("filter"),
    answer("date"),
    answer("dataset_name", correct = TRUE),
    answer("set_values_to"),
    allow_retry = TRUE
  ),
  question(
    "What differentiates an event_source() from a censor_source()?",
    answer("event_source() has a filter; censor_source() doesn't"),
    answer("event_source() produces CNSR = 0; censor_source() produces CNSR = 1+", correct = TRUE),
    answer("censor_source() uses a different date variable"),
    answer("They are identical — only the name differs"),
    allow_retry = TRUE
  )
)
```

---

## 3 — Overall Survival with derive_param_tte()

```{r os-demo}
# Build OS: event = death, censor = last alive date
adtte_os <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = TRTSDT,
  event_conditions  = list(death_event),
  censor_conditions = list(last_alive_censor),
  source_datasets   = list(adsl = adsl_tte),
  set_values_to     = exprs(
    PARAMCD = "OS",
    PARAM   = "Overall Survival"
  )
)

adtte_os |>
  select(USUBJID, PARAMCD, STARTDT, ADT, AVAL, CNSR, EVNTDESC)
```

```{r quiz-os, echo=FALSE}
quiz(
  question(
    "For a subject with DTHDT = 19480 and TRTSDT = 19370, what is AVAL?",
    answer("19480 - 19370 = 110"),
    answer("19480 - 19370 + 1 = 111", correct = TRUE),
    answer("110 / 365.25"),
    answer("1"),
    allow_retry = TRUE
  )
)
```

```{r os-ex, exercise=TRUE, exercise.lines=22}
# Exercise: build OS using adsl_tte
# event = !is.na(DTHDT), date = DTHDT
# censor = date = LSTALVDT
# STARTDT = RANDDT (not TRTSDT)

death_ev <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = ___,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST KNOWN ALIVE", SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

adtte_os2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = ___,
  event_conditions  = list(death_ev),
  censor_conditions = list(alive_cs),
  source_datasets   = list(adsl = adsl_tte),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

adtte_os2 |> select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

```{r os-ex-hint-1}
# date = DTHDT
# start_date = RANDDT
```

```{r os-ex-solution}
death_ev <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST KNOWN ALIVE", SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

adtte_os2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(death_ev),
  censor_conditions = list(alive_cs),
  source_datasets   = list(adsl = adsl_tte),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

adtte_os2 |> select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

### Exercise: Overall Survival on the real CDISCPILOT01 study

Scale up from the 5-subject toy to all 254 real subjects. `adsl_real` from pharmaverseadam contains DTHDT, LSTALVDT, SAFFL, TRTSDT, and TRT01A — everything needed for a submission-quality OS derivation.

> **Note:** In pharmaverseadam, DTHDT and LSTALVDT are Date objects. The mock `derive_param_tte()` handles Date arithmetic correctly (`as.numeric(Date − Date)` = days), so AVAL will be in days as expected.

```{r os-real, exercise=TRUE, exercise.lines=48}
# Check the death-related columns available in adsl_real
cat("Key TTE variables in adsl_real:\n")
cat(intersect(c("DTHDT","LSTALVDT","DTHFL","TRTSDT","RANDDT","SAFFL","TRT01A"),
              names(adsl_real)), "\n\n")

# Define event and censor sources using real ADSL variables
death_ev_real <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs_real <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST DATE KNOWN ALIVE",
                         SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

# Derive OS for the safety population
adtte_os_real <- derive_param_tte(
  dataset_adsl      = adsl_real |> filter(SAFFL == "Y"),
  start_date        = TRTSDT,
  event_conditions  = list(death_ev_real),
  censor_conditions = list(alive_cs_real),
  source_datasets   = list(adsl = adsl_real),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

# Summary
cat("OS subjects:", nrow(adtte_os_real), "\n")
cat("Events (CNSR = 0):", sum(adtte_os_real$CNSR == 0, na.rm = TRUE), "\n")
cat("Censored (CNSR = 1):", sum(adtte_os_real$CNSR == 1, na.rm = TRUE), "\n\n")

# Median survival in days by treatment arm
adtte_os_real |>
  group_by(TRT01A, CNSR) |>
  summarise(n = n(), median_aval = median(AVAL, na.rm = TRUE), .groups = "drop") |>
  arrange(TRT01A, CNSR) |>
  print()
```

```{r os-real-hint-1}
# adsl_real has DTHDT (Date), LSTALVDT (Date), TRTSDT (Date)
# filter(SAFFL == "Y") restricts to the safety population (~254 subjects)
# CNSR = 0 → event occurred; CNSR = 1 → censored (alive at last contact)
# median(AVAL) gives median follow-up in days per arm/event-status group
```

```{r os-real-solution}
death_ev_real <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs_real <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST DATE KNOWN ALIVE",
                         SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

adtte_os_real <- derive_param_tte(
  dataset_adsl      = adsl_real |> filter(SAFFL == "Y"),
  start_date        = TRTSDT,
  event_conditions  = list(death_ev_real),
  censor_conditions = list(alive_cs_real),
  source_datasets   = list(adsl = adsl_real),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

cat("OS subjects:", nrow(adtte_os_real), "\n")
cat("Events (CNSR = 0):", sum(adtte_os_real$CNSR == 0, na.rm = TRUE), "\n")
cat("Censored (CNSR = 1):", sum(adtte_os_real$CNSR == 1, na.rm = TRUE), "\n\n")

adtte_os_real |>
  group_by(TRT01A, CNSR) |>
  summarise(n = n(), median_aval = median(AVAL, na.rm = TRUE), .groups = "drop") |>
  arrange(TRT01A, CNSR) |>
  print()
```

---

## 4 — PFS: composite events

PFS uses **two event sources**: disease progression OR death, whichever comes first.

```{r pfs-demo}
# Progression event
pd_event <- event_source(
  dataset_name  = "adrs",
  filter        = PARAMCD == "OVR" & AVALC == "PD",
  date          = ADT,
  set_values_to = exprs(
    EVNTDESC = "DISEASE PROGRESSION",
    SRCDOM   = "ADRS",
    SRCVAR   = "ADT"
  )
)

# Death (reuse from above)
# last tumor assessment censor
last_assess <- censor_source(
  dataset_name  = "adrs",
  filter        = PARAMCD == "OVR" & AVALC %in% c("CR","PR","SD","NE"),
  date          = ADT,
  set_values_to = exprs(
    EVNTDESC = "LAST TUMOR ASSESSMENT",
    SRCDOM   = "ADRS",
    SRCVAR   = "ADT"
  )
)

adtte_pfs <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
)

adtte_pfs |>
  select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

```{r quiz-pfs, echo=FALSE}
quiz(
  question(
    "When both progression AND death occur for the same subject, derive_param_tte() picks:",
    answer("Death always wins"),
    answer("Progression always wins"),
    answer("The earliest of the two event dates", correct = TRUE),
    answer("The one listed first in event_conditions"),
    allow_retry = TRUE
  ),
  question(
    "A subject with no progression and no death gets CNSR =",
    answer("0"),
    answer("1 (from the censor source)", correct = TRUE),
    answer("2 — double censor"),
    answer("NA"),
    allow_retry = TRUE
  )
)
```

```{r pfs-ex, exercise=TRUE, exercise.lines=26}
# Exercise: build PFS but add a randomization-date fallback censor
# (for subjects with no assessments at all, censor = 2)
# Then count subjects by CNSR value

rand_censor <- censor_source(
  dataset_name  = "adsl",
  date          = RANDDT,
  censor        = ___,        # use censor = 2
  set_values_to = exprs(EVNTDESC = "RANDOMIZATION (NO ASSESSMENT)",
                         SRCDOM = "ADSL", SRCVAR = "RANDDT")
)

adtte_pfs2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor, rand_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

# Count by CNSR
adtte_pfs2 |> count(CNSR)
```

```{r pfs-ex-hint-1}
# censor = 2
```

```{r pfs-ex-solution}
rand_censor <- censor_source(
  dataset_name  = "adsl",
  date          = RANDDT,
  censor        = 2,
  set_values_to = exprs(EVNTDESC = "RANDOMIZATION (NO ASSESSMENT)",
                         SRCDOM = "ADSL", SRCVAR = "RANDDT")
)

adtte_pfs2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor, rand_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

adtte_pfs2 |> count(CNSR)
```

---

## 5 — Time to first serious AE

```{r ttaeser-demo}
# Event: first treatment-emergent serious AE
first_sae_event <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y" & TRTEMFL == "Y",
  date          = ASTDT,
  set_values_to = exprs(
    EVNTDESC = "FIRST SERIOUS AE",
    SRCDOM   = "ADAE",
    SRCVAR   = "ASTDT"
  )
)

eot_censor <- censor_source(
  dataset_name  = "adsl",
  date          = EOSDT,
  set_values_to = exprs(
    EVNTDESC = "END OF STUDY",
    SRCDOM   = "ADSL",
    SRCVAR   = "EOSDT"
  )
)

adtte_sae <- derive_param_tte(
  dataset_adsl      = adsl_tte |> filter(SAFFL == "Y"),
  start_date        = TRTSDT,
  event_conditions  = list(first_sae_event),
  censor_conditions = list(eot_censor, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adae = adae_tte),
  set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
)

adtte_sae |> select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

---

## 6 — call_derivation() for multiple parameters

```{r call-derivation-demo}
# Build OS and PFS together in one call_derivation call
adtte_both <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions  = list(death_event),
      censor_conditions = list(last_alive_censor),
      set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
    ),
    params(
      event_conditions  = list(pd_event, death_event),
      censor_conditions = list(last_assess, last_alive_censor),
      set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
    )
  ),
  # Shared arguments
  dataset_adsl    = adsl_tte,
  start_date      = "RANDDT",
  source_datasets = list(adsl = adsl_tte, adrs = adrs)
)

adtte_both |>
  select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC) |>
  arrange(PARAMCD, USUBJID)
```

```{r quiz-call-derivation, echo=FALSE}
quiz(
  question(
    "In `call_derivation()`, the `variable_params` list contains:",
    answer("Shared arguments applied to every call"),
    answer("Per-call argument overrides (event_conditions, censor_conditions, set_values_to)", correct = TRUE),
    answer("Dataset names only"),
    answer("Filter conditions"),
    allow_retry = TRUE
  )
)
```

```{r call-derivation-ex, exercise=TRUE, exercise.lines=28}
# Exercise: use call_derivation to build OS and TTAESER in one block
# OS: death event, last alive censor, start = TRTSDT
# TTAESER: first_sae_event, eot_censor + last_alive_censor, start = TRTSDT

adtte_multi <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions  = list(death_event),
      censor_conditions = list(last_alive_censor),
      set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
    ),
    params(
      event_conditions  = list(first_sae_event),
      censor_conditions = list(eot_censor, last_alive_censor),
      set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
    )
  ),
  dataset_adsl    = adsl_tte |> filter(SAFFL == "Y"),
  start_date      = ___,
  source_datasets = list(adsl = adsl_tte, adae = adae_tte)
)

adtte_multi |> select(USUBJID, PARAMCD, AVAL, CNSR) |> arrange(PARAMCD, USUBJID)
```

```{r call-derivation-ex-hint-1}
# start_date = TRTSDT
```

```{r call-derivation-ex-solution}
adtte_multi <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions  = list(death_event),
      censor_conditions = list(last_alive_censor),
      set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
    ),
    params(
      event_conditions  = list(first_sae_event),
      censor_conditions = list(eot_censor, last_alive_censor),
      set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
    )
  ),
  dataset_adsl    = adsl_tte |> filter(SAFFL == "Y"),
  start_date      = TRTSDT,
  source_datasets = list(adsl = adsl_tte, adae = adae_tte)
)

adtte_multi |> select(USUBJID, PARAMCD, AVAL, CNSR) |> arrange(PARAMCD, USUBJID)
```

---

## 7 — AVAL in months/years and validation

```{r aval-units-demo}
# AVAL in days → months conversion
adtte_os |>
  mutate(
    AVAL_MONTHS = AVAL / (365.25 / 12),
    AVAL_YEARS  = AVAL / 365.25
  ) |>
  select(USUBJID, PARAMCD, AVAL, AVAL_MONTHS, AVAL_YEARS, CNSR)
```

```{r quiz-validation, echo=FALSE}
quiz(
  question(
    "Which CNSR distribution check is a key ADTTE validation step?",
    answer("Confirming no negative AVAL values (AVAL should always be ≥ 1)", correct = TRUE),
    answer("Confirming CNSR sums to zero"),
    answer("Confirming every subject has exactly one PARAMCD"),
    answer("Confirming ADT > STARTDT for all rows"),
    allow_retry = TRUE
  ),
  question(
    "Which check ensures completeness of the subject population?",
    answer("count(CNSR)"),
    answer("Every subject in dataset_adsl (filtered population) has exactly one row per PARAMCD", correct = TRUE),
    answer("nrow() matches ADSL"),
    answer("SRCDOM is non-missing"),
    allow_retry = TRUE
  )
)
```

---

## 7.5 — P21 Derivation Diagnostics

ADTTE errors are among the most consequential in a submission: a wrong AVAL or misclassified CNSR can distort Kaplan-Meier curves and median survival estimates. Each diagnostic presents a define.xml derivation spec and the Pinnacle 21 error that revealed the bug.

---

### Diagnostic 1 — PFS: Wrong start date (TRTSDT instead of RANDDT)

**Derivation method (define.xml / Pinnacle 21 field):**

> *STARTDT for Progression-Free Survival (PFS) = RANDDT (date of randomization). The analysis clock starts at randomization, not at first dose, to capture dose-delay effects. AVAL = ADT − RANDDT + 1 days. Source: ADSL.RANDDT.*

The programmer submitted:

```r
adtte_pfs_draft <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = TRTSDT,   # ← bug: should be RANDDT
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)
```

**P21 report:** *"AVAL for PARAMCD = 'PFS' does not align with expected values. STARTDT appears equal to TRTSDT rather than RANDDT. For subjects with dose delays, AVAL is understated. Expected AVAL = ADT − RANDDT + 1."*

```{r p21-bug-pfs-startdate, exercise=TRUE, exercise.lines=22}
# Fix: PFS uses RANDDT as the start date (randomization date), not TRTSDT.

adtte_pfs_fixed <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = ___,       # fix: randomization date
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
)

# Compare AVAL from RANDDT vs TRTSDT: subjects 01-001 and 01-002 had
# RANDDT 7 days before TRTSDT, so AVAL should be 7 days longer with RANDDT.
adtte_pfs_wrong <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = TRTSDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

left_join(
  adtte_pfs_fixed |> select(USUBJID, AVAL_correct = AVAL, CNSR),
  adtte_pfs_wrong |> select(USUBJID, AVAL_wrong   = AVAL),
  by = "USUBJID"
) |> mutate(diff = AVAL_correct - AVAL_wrong)
```

```{r p21-bug-pfs-startdate-hint-1}
# start_date = RANDDT  (randomization date, not first treatment date)
# TRTSDT may differ from RANDDT if the subject had a dose delay
# RANDDT is always the correct clock start for PFS per the statistical analysis plan
```

```{r p21-bug-pfs-startdate-solution}
adtte_pfs_fixed <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
)

adtte_pfs_wrong <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = TRTSDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

left_join(
  adtte_pfs_fixed |> select(USUBJID, AVAL_correct = AVAL, CNSR),
  adtte_pfs_wrong |> select(USUBJID, AVAL_wrong   = AVAL),
  by = "USUBJID"
) |> mutate(diff = AVAL_correct - AVAL_wrong)
```

---

### Diagnostic 2 — TTAESER: Missing TRTEMFL filter includes pre-treatment SAEs

**Derivation method (define.xml / Pinnacle 21 field):**

> *TTAESER event = first treatment-emergent serious adverse event. Filter criteria: AESER = 'Y' AND TRTEMFL = 'Y'. AEs that are serious but occurred before first dose (TRTEMFL ≠ 'Y') must be excluded. Only on-treatment SAEs trigger the event. Source: ADAE.AESER, ADAE.TRTEMFL, ADAE.ASTDT.*

The programmer submitted:

```r
sae_event_draft <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y",    # ← bug: TRTEMFL == "Y" is missing
  date          = ASTDT,
  set_values_to = exprs(EVNTDESC = "FIRST SERIOUS AE",
                         SRCDOM = "ADAE", SRCVAR = "ASTDT")
)
```

**P21 report:** *"TTAESER event date precedes TRTSDT for subject 01-001. A pre-treatment SAE (TRTEMFL = 'N') is counted as the event. Event filter must include TRTEMFL = 'Y'."*

```{r p21-bug-ttaeser-filter, exercise=TRUE, exercise.lines=30}
# Self-contained: build a small AE dataset with one pre-treatment SAE
ae_with_pretrt <- tibble(
  USUBJID = c("01-001", "01-001", "01-002"),
  AETERM  = c("PRE-STUDY SAE", "ANAPHYLACTIC SHOCK", "CARDIAC ARREST"),
  ASTDT   = c(19340, 19400, 19400),   # 19340 is before TRTSDT = 19367 for 01-001
  AESER   = c("Y",   "Y",    "Y"),
  TRTEMFL = c("N",   "Y",    "Y")     # first row is pre-treatment
)

# Broken: no TRTEMFL filter → pre-treatment SAE for 01-001 triggers event
sae_ev_broken <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y",
  date          = ASTDT,
  set_values_to = exprs(EVNTDESC = "FIRST SERIOUS AE", SRCDOM = "ADAE", SRCVAR = "ASTDT")
)

# Fix: add TRTEMFL == "Y" to the filter
sae_ev_fixed <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y" & ___,     # add the missing condition
  date          = ASTDT,
  set_values_to = exprs(EVNTDESC = "FIRST SERIOUS AE", SRCDOM = "ADAE", SRCVAR = "ASTDT")
)

broken <- derive_param_tte(
  dataset_adsl = adsl_tte, start_date = TRTSDT,
  event_conditions = list(sae_ev_broken),
  censor_conditions = list(eot_censor),
  source_datasets = list(adsl = adsl_tte, adae = ae_with_pretrt),
  set_values_to = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
)

fixed <- derive_param_tte(
  dataset_adsl = adsl_tte, start_date = TRTSDT,
  event_conditions = list(sae_ev_fixed),
  censor_conditions = list(eot_censor),
  source_datasets = list(adsl = adsl_tte, adae = ae_with_pretrt),
  set_values_to = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
)

cat("--- Broken (includes pre-treatment SAE for 01-001) ---\n")
broken |> select(USUBJID, AVAL, CNSR, EVNTDESC) |> print()
cat("\n--- Fixed (only on-treatment SAEs trigger event) ---\n")
fixed  |> select(USUBJID, AVAL, CNSR, EVNTDESC) |> print()
```

```{r p21-bug-ttaeser-filter-hint-1}
# filter = AESER == "Y" & TRTEMFL == "Y"
# Without TRTEMFL == "Y", pre-treatment SAEs (TRTEMFL = "N") incorrectly trigger the event
# For subject 01-001, the event date would be ASTDT = 19340, which is before TRTSDT = 19367
```

```{r p21-bug-ttaeser-filter-solution}
ae_with_pretrt <- tibble(
  USUBJID = c("01-001", "01-001", "01-002"),
  AETERM  = c("PRE-STUDY SAE", "ANAPHYLACTIC SHOCK", "CARDIAC ARREST"),
  ASTDT   = c(19340, 19400, 19400),
  AESER   = c("Y", "Y", "Y"),
  TRTEMFL = c("N", "Y", "Y")
)

sae_ev_fixed <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y" & TRTEMFL == "Y",
  date          = ASTDT,
  set_values_to = exprs(EVNTDESC = "FIRST SERIOUS AE", SRCDOM = "ADAE", SRCVAR = "ASTDT")
)

fixed <- derive_param_tte(
  dataset_adsl = adsl_tte, start_date = TRTSDT,
  event_conditions  = list(sae_ev_fixed),
  censor_conditions = list(eot_censor),
  source_datasets   = list(adsl = adsl_tte, adae = ae_with_pretrt),
  set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
)

fixed |> select(USUBJID, AVAL, CNSR, EVNTDESC) |> print()
```

---

### Diagnostic 3 — AVAL: Missing +1 in manual duration calculation

**Derivation method (define.xml / Pinnacle 21 field):**

> *AVAL = ADT − STARTDT + 1 (days). Follows CDISC ADaM IG §4.4.7 study day convention: AVAL = 1 when ADT = STARTDT (event on Day 1 of follow-up). A subject who dies on the first day of follow-up has AVAL = 1, not 0. derive_param_tte() computes this automatically. Manual computation must include the +1.*

The programmer post-processed the ADTTE dataset:

```r
adtte_recomputed <- adtte_os |>
  mutate(AVAL = ADT - STARTDT)    # ← bug: missing +1
```

**P21 report:** *"AD0214 — AVAL = 0 detected for 2 subjects where ADT = STARTDT. CDISC ADaM IG section 4.4.7: duration of 0 is not permissible for time-to-event variables. Expected AVAL ≥ 1 for all rows."*

```{r p21-bug-aval-plusone, exercise=TRUE, exercise.lines=20}
# adtte_os was built earlier by derive_param_tte() — it has correct AVAL.
# Simulate what happens when a programmer recalculates AVAL manually.

# Introduce a subject where ADT = STARTDT to expose the bug clearly
adtte_test <- adtte_os |>
  mutate(
    ADT_orig  = ADT,
    # Force subject 01-001 to have ADT = STARTDT (event on Day 1)
    ADT       = if_else(USUBJID == "01-001", STARTDT, ADT),
    AVAL_wrong  = ADT - STARTDT,           # ← missing +1
    AVAL_correct = ADT - STARTDT + 1       # ← fix: include +1
  )

# Show the difference: AVAL_wrong = 0 for 01-001, AVAL_correct = 1
adtte_test |>
  select(USUBJID, STARTDT, ADT, AVAL_wrong, AVAL_correct, CNSR)
```

```{r p21-bug-aval-plusone-hint-1}
# AVAL = ADT - STARTDT + 1
# The +1 is mandatory: the reference date (STARTDT) is Day 1, not Day 0
# derive_param_tte() already applies this: aval = as.numeric(adt - startdt) + 1
# Only add +1 when computing AVAL manually outside of admiral functions
```

```{r p21-bug-aval-plusone-solution}
adtte_test <- adtte_os |>
  mutate(
    ADT          = if_else(USUBJID == "01-001", STARTDT, ADT),
    AVAL_wrong   = ADT - STARTDT,
    AVAL_correct = ADT - STARTDT + 1
  )

adtte_test |>
  select(USUBJID, STARTDT, ADT, AVAL_wrong, AVAL_correct, CNSR)
```

---

## Quiz — final

```{r quiz-final-18, echo=FALSE}
quiz(
  question(
    "A PFS derivation uses `start_date = TRTSDT`. P21 reports AVAL is understated for subjects with dose delays. What is the correct start date for PFS and why?",
    answer("TRTSDT is correct — PFS always starts at first dose"),
    answer("RANDDT — PFS follow-up begins at randomization; using TRTSDT shortens AVAL for subjects who started treatment after randomization", correct = TRUE),
    answer("EOSDT — the study end date anchors the survival clock"),
    answer("LSTALVDT — the last known alive date defines the PFS clock"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "A TTAESER event_source uses `filter = AESER == 'Y'`. P21 reports an event date before TRTSDT for one subject. What is the missing condition?",
    answer("AESER == 'Y' is sufficient — pre-treatment SAEs are excluded automatically by the TTE function"),
    answer("TRTEMFL == 'Y' must be added — without it, pre-treatment serious AEs (which have TRTEMFL = 'N') incorrectly trigger the event", correct = TRUE),
    answer("ASTDT > TRTSDT must be added as a date filter"),
    answer("AESER == 'Y' & AESEV == 'SEVERE' is the correct filter"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "P21 rule AD0214 flags AVAL = 0 in ADTTE. A programmer computed `AVAL = ADT - STARTDT`. What is the fix and why?",
    answer("AVAL = ADT - STARTDT + 1 — CDISC ADaM IG defines the reference date as Day 1 (AVAL = 1 when ADT = STARTDT), so no duration of 0 exists", correct = TRUE),
    answer("AVAL = ADT - STARTDT is correct; AD0214 only applies to ADSL"),
    answer("AVAL = ADT - STARTDT + 0.5, rounded up to 1 for Day 1 records"),
    answer("AVAL = (ADT - STARTDT) / 365.25 converts days to years"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "PFS uses two event_source objects (progression and death). When both occur for the same subject, derive_param_tte() selects:",
    answer("The event listed first in the event_conditions list"),
    answer("Death always takes priority over progression"),
    answer("The event with the earliest ADT — the first event to occur determines AVAL and CNSR = 0", correct = TRUE),
    answer("The last event source in the list"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "CNSR = 0 in ADTTE means the event occurred. A subject is alive at study end with no events. What values should CNSR and EVNTDESC take?",
    answer("CNSR = 0, EVNTDESC = 'NO EVENT'"),
    answer("CNSR = 1 (from the censor_source), EVNTDESC = the label defined in the applicable censor_source set_values_to (e.g., 'LAST DATE KNOWN ALIVE')", correct = TRUE),
    answer("CNSR = NA — missing indicates no event and no censoring"),
    answer("CNSR = 2 — all censor sources default to censor = 2"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "Multiple censor_source objects are listed. Which censoring date does derive_param_tte() use for a subject with no event?",
    answer("The latest date across all censor sources — the most recent evidence the subject was event-free", correct = TRUE),
    answer("The earliest date — conservative censoring"),
    answer("The censor source listed first in the list"),
    answer("All censor dates are averaged"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "EVNTDESC is not a required CDISC variable, but it is universally included. What does it provide?",
    answer("A numeric code for the event type used in regression models"),
    answer("Traceability — it records exactly which event_source or censor_source rule fired for each subject, enabling auditors to verify each row's derivation", correct = TRUE),
    answer("The MedDRA preferred term for the event"),
    answer("A link back to the SDTM source domain"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "call_derivation() builds OS and PFS in one block. What goes in variable_params vs the shared fixed arguments?",
    answer("variable_params holds shared arguments like start_date and source_datasets; fixed arguments hold per-endpoint event/censor sources"),
    answer("variable_params holds per-endpoint arguments (event_conditions, censor_conditions, set_values_to); shared arguments (start_date, source_datasets, dataset_adsl) are passed outside variable_params", correct = TRUE),
    answer("All arguments go in variable_params — the fixed argument list is ignored"),
    answer("variable_params holds only PARAMCD; all other arguments are fixed"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "A subject has DTHDT = NA and LSTALVDT = 19600 in ADSL. What CNSR and ADT should the OS row have?",
    answer("CNSR = 0, ADT = NA — missing death date means the event cannot be confirmed"),
    answer("CNSR = 1, ADT = 19600 — no death event found, so the latest censor date (LSTALVDT) is used", correct = TRUE),
    answer("CNSR = NA — the subject is excluded from the OS analysis"),
    answer("CNSR = 2, ADT = EOSDT — a second-level censor is applied"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The {admiralonco} extension package provides pre-built TTE source objects. What is the primary benefit over coding event_source() from scratch?",
    answer("Pre-built objects run faster than custom event_source() calls"),
    answer("Pre-built source objects implement validated, oncology-standard event and censoring rules (RECIST-based PFS, OS per BICR) that reduce the risk of specification deviations across studies", correct = TRUE),
    answer("admiralonco source objects work without source_datasets — they fetch data automatically"),
    answer("They eliminate the need for SRCDOM and SRCVAR metadata"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```
