---
title: "Lesson 18 — admiral Part 5: Time-to-Event ADaMs (ADTTE)"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 18_admiral_part5_adtte.Rmd
# PURPOSE       : admiral Part 5 — ADTTE time-to-event datasets: derive_param_tte(), event
#                 and censoring source definitions, Kaplan-Meier
#                 readiness checks, and common oncology/survival
#                 endpoints.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tibble)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── shared mocks ─────────────────────────────────────────────────────────────
exprs <- function(...) {
  dots <- as.list(substitute(list(...)))[-1]; dots
}

params <- function(...) list(...)

convert_blanks_to_na <- function(dataset) {
  dataset |> mutate(across(where(is.character), ~ifelse(.=="", NA_character_, .)))
}

# ── ADTTE source object constructors ─────────────────────────────────────────
event_source <- function(dataset_name, filter = NULL, date, set_values_to = list()) {
  list(
    type         = "event",
    dataset_name = dataset_name,
    filter       = substitute(filter),
    date         = substitute(date),
    set_values_to = set_values_to,
    censor       = 0
  )
}

censor_source <- function(dataset_name, filter = NULL, date, set_values_to = list(),
                           censor = 1) {
  list(
    type         = "censor",
    dataset_name = dataset_name,
    filter       = substitute(filter),
    date         = substitute(date),
    set_values_to = set_values_to,
    censor       = censor
  )
}

# ── derive_param_tte ──────────────────────────────────────────────────────────
derive_param_tte <- function(dataset_adsl = NULL,
                              start_date,
                              event_conditions,
                              censor_conditions,
                              source_datasets,
                              set_values_to = list(),
                              ...) {
  .e <- substitute(start_date)
  start_nm <- if (is.character(.e)) .e else deparse(.e)
  adsl_use <- if (!is.null(dataset_adsl)) dataset_adsl else
              source_datasets[[names(source_datasets)[1]]]
  # helper to evaluate one source condition for all subjects
  eval_source <- function(src) {
    ds <- source_datasets[[src$dataset_name]]
    if (is.null(ds)) return(NULL)
    # apply filter
    if (!is.null(src$filter)) {
      ds <- tryCatch(filter(ds, !!src$filter), error = function(e) ds)
    }
    # evaluate date expression
    date_nm <- deparse(src$date)
    if (!date_nm %in% names(ds)) return(NULL)
    ds$..DATE.. <- ds[[date_nm]]
    ds <- ds |> filter(!is.na(..DATE..))
    if (nrow(ds) == 0) return(NULL)
    # keep just USUBJID + date + meta
    out <- ds |> select(any_of(c("USUBJID", "STUDYID", "..DATE..")))
    out$..CENSOR.. <- src$censor
    for (nm in names(src$set_values_to)) {
      val <- tryCatch(eval(src$set_values_to[[nm]]), error = function(e) as.character(src$set_values_to[[nm]]))
      out[[nm]] <- val
    }
    out
  }
  # Process events: earliest event date per subject
  event_rows <- lapply(event_conditions, eval_source)
  event_rows <- Filter(Negate(is.null), event_rows)
  event_all  <- if (length(event_rows)) bind_rows(event_rows) else NULL

  # Process censorings: latest censor date per subject
  censor_rows <- lapply(censor_conditions, eval_source)
  censor_rows <- Filter(Negate(is.null), censor_rows)
  censor_all  <- if (length(censor_rows)) bind_rows(censor_rows) else NULL

  # Determine ADT and CNSR for each subject
  subjects <- unique(adsl_use$USUBJID)
  result   <- vector("list", length(subjects))

  for (i in seq_along(subjects)) {
    subj     <- subjects[i]
    adsl_row <- adsl_use |> filter(USUBJID == subj)
    startdt  <- if (start_nm %in% names(adsl_row)) adsl_row[[start_nm]][1] else NA_real_

    # Events for this subject
    ev_sub <- if (!is.null(event_all)) filter(event_all, USUBJID == subj) else NULL
    # Censor for this subject
    ce_sub <- if (!is.null(censor_all)) filter(censor_all, USUBJID == subj) else NULL

    if (!is.null(ev_sub) && nrow(ev_sub) > 0) {
      # Take earliest event
      ev_sub  <- arrange(ev_sub, ..DATE..)
      best    <- ev_sub[1, , drop = FALSE]
      adt     <- best$..DATE..[1]
      cnsr    <- 0
      evnt    <- best$EVNTDESC[1] %||% NA_character_
      srcdom  <- best$SRCDOM[1]   %||% NA_character_
      srcvar  <- best$SRCVAR[1]   %||% NA_character_
    } else if (!is.null(ce_sub) && nrow(ce_sub) > 0) {
      # Take latest censor date
      ce_sub <- arrange(ce_sub, desc(..DATE..))
      best   <- ce_sub[1, , drop = FALSE]
      adt    <- best$..DATE..[1]
      cnsr   <- best$..CENSOR..[1]
      evnt   <- best$EVNTDESC[1] %||% NA_character_
      srcdom <- best$SRCDOM[1]   %||% NA_character_
      srcvar <- best$SRCVAR[1]   %||% NA_character_
    } else {
      adt <- NA_real_; cnsr <- NA_real_; evnt <- NA_character_
      srcdom <- NA_character_; srcvar <- NA_character_
    }

    aval <- if (!is.na(adt) && !is.na(startdt)) as.numeric(adt - startdt) + 1 else NA_real_

    row_out <- adsl_row |>
      mutate(STARTDT = startdt, ADT = adt, AVAL = aval, CNSR = cnsr,
             EVNTDESC = evnt, SRCDOM = srcdom, SRCVAR = srcvar)
    result[[i]] <- row_out
  }

  tte_rows <- bind_rows(result)
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]]), error = function(e) as.character(set_values_to[[nm]]))
    tte_rows[[nm]] <- val
  }
  tte_rows
}

# call_derivation for TTE
call_derivation <- function(dataset = NULL, derivation, variable_params, ...) {
  fixed_args <- list(...)
  result_list <- vector("list", length(variable_params))
  for (i in seq_along(variable_params)) {
    call_args <- c(fixed_args, variable_params[[i]])
    result_list[[i]] <- tryCatch(do.call(derivation, call_args), error = function(e) NULL)
  }
  bind_rows(Filter(Negate(is.null), result_list))
}

`%||%` <- function(a, b) if (is.null(a) || (length(a) == 1 && is.na(a))) b else a

# ── Sample datasets ───────────────────────────────────────────────────────────
adsl_tte <- tribble(
  ~STUDYID,        ~USUBJID,  ~TRTSDT, ~TRTEDT, ~RANDDT, ~TRT01A,
  ~DTHDT,  ~LSTALVDT, ~EOSDT,  ~DTHFL, ~SAFFL,
  "CDISCPILOT01","01-001", 19367, 19548,  19360,  "Xanomeli",
  NA_real_, 19600,     19600, "N",    "Y",
  "CDISCPILOT01","01-002", 19370, 19430,  19365,  "Placebo",
  19480,    19480,     19480, "Y",    "Y",
  "CDISCPILOT01","01-003", 19375, 19540,  19370,  "Xanomeli",
  NA_real_, 19590,     19590, "N",    "Y",
  "CDISCPILOT01","01-004", 19378, 19490,  19372,  "Placebo",
  19510,    19510,     19510, "Y",    "Y",
  "CDISCPILOT01","01-005", 19380, 19560,  19375,  "Xanomeli",
  NA_real_, 19610,     19610, "N",    "Y"
)

adrs <- tribble(
  ~STUDYID,        ~USUBJID,  ~PARAMCD, ~AVALC, ~ADT,
  "CDISCPILOT01", "01-001",  "OVR",    "SD",   19450,
  "CDISCPILOT01", "01-001",  "OVR",    "PD",   19520,
  "CDISCPILOT01", "01-002",  "OVR",    "PR",   19410,
  "CDISCPILOT01", "01-003",  "OVR",    "SD",   19440,
  "CDISCPILOT01", "01-004",  "OVR",    "PD",   19460,
  "CDISCPILOT01", "01-005",  "OVR",    "CR",   19460
)

adae_tte <- tribble(
  ~STUDYID,        ~USUBJID,  ~AETERM,              ~ASTDT, ~TRTEMFL, ~AESER, ~TRTSDT,
  "CDISCPILOT01", "01-001",  "ANAPHYLACTIC SHOCK",  19400,  "Y",     "Y",    19367,
  "CDISCPILOT01", "01-003",  "MYOCARDIAL INFARCTION",19420, "Y",     "Y",    19375
)

# ── Real pharmaverseadam ADSL for real-data OS exercise ───────────────────────
library(pharmaverseadam)
adsl_real <- pharmaverseadam::adsl
```

## Welcome

This lesson covers **ADTTE — Time-to-Event analysis datasets**. You'll learn admiral's declarative `event_source()` / `censor_source()` design and use `derive_param_tte()` to build Overall Survival (OS), Progression-Free Survival (PFS), and safety TTE parameters.

**Why ADTTE deserves its own lesson.** ADTTE is not a generic BDS-style analysis dataset. It has a peculiar shape (one row per subject per endpoint), a peculiar arithmetic (`AVAL = ADT - STARTDT + 1`), and a peculiar censoring convention (`CNSR = 0` is the event, `CNSR = 1` is the censor — which is the *opposite* of how some SAS programmers code their own censor flags). The downstream analysis — `PROC LIFETEST`, `PROC PHREG`, Kaplan-Meier curves, Cox proportional hazards models — is among the most consequential in any oncology, cardiovascular, or chronic-disease submission. A misclassified censor or a wrong `STARTDT` quietly distorts the survival curve, and the bug surfaces only when the statistician asks why the median survival looks wrong.

**SAS programmer's translation guide.** In SAS, you typically built a `PROC LIFETEST`-ready dataset in a single DATA step packed with `IF/ELSE` branches: "if DTHDT is not missing then EVENT=1 and TIME = DTHDT - TRTSDT + 1; else EVENT=0 and TIME = LSTALVDT - TRTSDT + 1". admiral inverts this: you declare *each candidate event* and *each candidate censor* as a tiny named object (`event_source()`, `censor_source()`), pass them to a derivation engine, and the engine picks the right row. This separation makes the analysis specification reviewable: a statistician can read the four `event_source()` definitions and immediately verify which rules will fire.

**What you will build.** By the end of this lesson you will:

- Recognize the ADTTE row layout and the meaning of `STARTDT`, `ADT`, `AVAL`, `CNSR`, `EVNTDESC`, `SRCDOM`, `SRCVAR`
- Build an OS parameter from death events and last-known-alive censors
- Build a PFS parameter as a composite of progression OR death
- Build a TTAESER parameter for time to first serious adverse event
- Use `call_derivation()` to build multiple TTE parameters in a single DRY block
- Convert `AVAL` between days, months, and years for analysis vs display
- Run validation diagnostics that match the SAS `PROC LIFETEST` inputs your statistician expects

---

## 1 — The ADTTE shape

### 1.1 What "time-to-event" means (clinical)

A **time-to-event endpoint** measures the *duration* from some clinical starting point until the *first occurrence* of a defined event. The two most common endpoints in oncology are:

- **Overall Survival (OS)** — start = randomization or first dose; event = death from any cause
- **Progression-Free Survival (PFS)** — start = randomization; event = first of (disease progression, death)

In cardiovascular trials you measure **MACE** (Major Adverse Cardiac Events: first of MI, stroke, CV death). In safety analyses you measure **time to first SAE**. The structural pattern is always the same: a (start, end, event-or-censor) triple per subject per endpoint.

**Right-censoring** is the most important concept. A subject who is *alive and event-free* at the database lock contributes information: "we know they were event-free up to date X". That date X is the **censoring time**. The subject is *right-censored* because the event (if it ever occurs) happens to the right of X on the time axis. Survival analysis methods (`PROC LIFETEST`, `PROC PHREG`) explicitly account for censored observations — they do not just drop them.

> **SAS analog mental model.** This is the same data you used to feed into `PROC LIFETEST` with the syntax `TIME AVAL*CNSR(1);` — where `AVAL` is the duration and `CNSR(1)` declares "1 means censored". admiral builds exactly that input dataset for you.

### 1.2 The (PARAMCD, AVAL, CNSR) triple as the analytical key

Every row of ADTTE is uniquely identified by **`USUBJID + PARAMCD`**. The analytical payload of that row is the **(`AVAL`, `CNSR`)** pair:

- `AVAL` — duration in days from `STARTDT` to `ADT`, plus 1
- `CNSR` — `0` if `ADT` is an event date, `1` (or `2`, `3`, ...) if `ADT` is a censoring date

`PARAMCD` distinguishes the endpoint: `"OS"`, `"PFS"`, `"TTAESER"`, `"MACE"`, etc. A subject typically has *multiple rows* — one per endpoint they are eligible for. This is the same structure as BDS (one row per parameter) but with each row representing the *single* qualifying event or censoring (not a longitudinal series).

### 1.3 STARTDT vs ADT

Two date columns govern the duration:

- **`STARTDT`** — the **start of the risk period**. Pulled from ADSL (`RANDDT` for PFS, `TRTSDT` for OS-from-treatment, etc.)
- **`ADT`** — the **analysis date**. Either the event date (death, progression, first SAE) or the censoring date (last alive, last assessment, end of study)

The duration `ADT - STARTDT + 1` is `AVAL` (days). The `+ 1` is the CDISC ADaM IG convention that `STARTDT` is **Day 1**, not Day 0 — so an event on the same calendar day as `STARTDT` has `AVAL = 1`, never `AVAL = 0`.

> **SAS analog mental model.** `STARTDT` is what you used to write as `STARTDT = COALESCE(RANDDT, TRTSDT);` at the top of your TTE DATA step. `ADT` is the `EVENT_DT` or `CENSOR_DT` you'd assign in the IF/ELSE branches.

### 1.4 Why ADTTE is "wide-with-PARAMCD" (one row per subject per endpoint)

ADTTE behaves like a BDS dataset (`PARAMCD` is the unit of analysis), but with a critical difference: each row is a **terminal** observation. There is no longitudinal series — just one row per subject per endpoint. The `PARAMCD` discriminates which TTE the row contributes to:

```
USUBJID    PARAMCD   STARTDT     ADT         AVAL   CNSR   EVNTDESC
01-001     OS        2023-04-15  2024-03-10  331    0      DEATH
01-001     PFS       2023-04-08  2023-09-22  168    0      DISEASE PROGRESSION
01-001     TTAESER   2023-04-15  2023-05-30  46     0      FIRST SERIOUS AE
01-002     OS        2023-05-01  2025-01-10  620    1      LAST DATE KNOWN ALIVE
01-002     PFS       2023-04-24  2024-12-15  601    1      LAST TUMOR ASSESSMENT
01-002     TTAESER   2023-05-01  2025-01-10  620    1      END OF STUDY
```

Each (subject, endpoint) pair lives on exactly one row. The downstream `PROC LIFETEST` selects rows with `WHERE PARAMCD = "OS";` and runs the survival analysis on that subset.

### 1.5 Comparison with ADaM BDS

| Feature | BDS (ADLB, ADVS) | ADTTE |
|---|---|---|
| Row granularity | Subject × Parameter × Visit | Subject × Parameter (terminal) |
| Multiple rows per subject? | Yes (longitudinal) | Yes (one per endpoint) |
| `AVAL` semantics | Measurement value | Duration in days |
| Censoring | N/A | `CNSR` column required |
| Primary use | Shift tables, change from baseline | Kaplan-Meier, Cox regression |

### 1.6 The censoring convention (CNSR = 0 event, CNSR = 1 censor)

This is the single most error-prone convention in ADTTE. CDISC ADaM IG mandates:

- **`CNSR = 0`** — the row represents an **event** (death, progression, first SAE, ...)
- **`CNSR = 1`** — the row represents a **censoring** (last known alive, last assessment, end of study, ...)
- **`CNSR = 2, 3, ...`** — optional secondary censor codes for tracking *which* censoring rule fired

> **WARNING for SAS programmers.** Many homegrown SAS programs use `EVENT = 1` and `CENSOR = 0` (i.e., `1 = event` is intuitive). CDISC ADaM ADTTE is the **inverse**. Always read the `PROC LIFETEST` syntax carefully — `TIME AVAL*CNSR(1);` literally says "the values of CNSR that mean censored are listed in parentheses". For ADTTE that is `CNSR(1)`. If you flip the convention, your survival curve is computed from censoring times rather than event times — a catastrophic, often-undetected error.

### 1.7 SAS PROC LIFETEST input dataset structure

The output of admiral's `derive_param_tte()` is *exactly* the dataset shape that `PROC LIFETEST` expects as input:

```sas
/* The PROC LIFETEST call your statistician will run on ADTTE */
proc lifetest data=adtte(where=(paramcd="OS")) method=km plots=(survival);
  time aval*cnsr(1);   /* CNSR=1 declares censored */
  strata trt01a;
run;

/* PROC PHREG uses the same dataset */
proc phreg data=adtte(where=(paramcd="OS"));
  class trt01a(ref="Placebo");
  model aval*cnsr(1) = trt01a age sex;   /* Cox proportional hazards */
run;
```

**Key insight:** the entire purpose of `derive_param_tte()` is to produce a dataset that drops straight into these two PROC calls without further processing. Your validation question is always *"Does my AVAL match the survival curve the statistician will produce?"*

### 1.8 The end-to-end picture (SAS vs admiral)

| ADTTE concept | What it represents | Closest SAS analog | SAS code sketch |
|---|---|---|---|
| `event_source()` | A row that, if found, becomes the event | `IF DTHDT NE . THEN DO; EVENT=1; ADT=DTHDT; END;` | A branch of an `IF/ELSE` chain emitting `EVENT=1` |
| `censor_source()` | A row that, if found, becomes the censor | `ELSE DO; EVENT=0; ADT=LSTALVDT; END;` | The `ELSE` branch emitting `EVENT=0` |
| `start_date` | Anchor for the duration calculation | `STARTDT = COALESCE(RANDDT, TRTSDT);` | A `COALESCE()` of candidate start dates |
| `derive_param_tte()` | Picks earliest event or latest censor; computes AVAL | One DATA step that iterates candidates and emits the winner | `RETAIN` + `MIN()` logic per subject |
| `call_derivation()` | DRY: build many PARAMCDs in one call | `%MACRO` looping over a PARAMCD list | `%DO i=1 %TO &n;` macro loop |
| `AVAL = ADT - STARTDT + 1` | Days on study (Day 1 = STARTDT) | `AVAL = ADT - STARTDT + 1;` | Same arithmetic — `+1` is mandatory |
| `CNSR(1)` for `PROC LIFETEST` | Censor flag interpretation | `TIME AVAL*CNSR(1);` | Same — admiral preserves the convention |

### 1.9 ASCII picture — ADTTE row layout

```
                            ── ADTTE ROW LAYOUT ──

   STARTDT             ADT                                  AVAL = ADT - STARTDT + 1
   │                   │
   ▼                   ▼
   ●━━━━━━━━━━━━━━━━━━━●                                    CNSR = 0 (event)    PFS endpoint
   start of follow-up  ↑ disease progression date

   ●━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━━●   CNSR = 1 (censor)   OS endpoint
   start of follow-up                                   ↑ last known alive (no death observed)

   ●━━━━━━●                                                  CNSR = 0 (event)    TTAESER endpoint
   start   ↑ first serious AE date
```

### 1.10 SAS DATA step versus admiral side-by-side

```
─────────────────  THE SAS WAY  ─────────────────       ─────────────────  THE admiral WAY  ─────────────────

data adtte_os;                                           death_event <- event_source(
  set adsl;                                                dataset_name = "adsl",
  startdt = trtsdt;                                        filter       = !is.na(DTHDT),
  if dthdt ne . then do;       /* event */                  date         = DTHDT,
    adt = dthdt;                                           set_values_to = exprs(EVNTDESC = "DEATH")
    cnsr = 0;                                            )
    evntdesc = "DEATH";
  end;                                                   alive_censor <- censor_source(
  else do;                     /* censor */                dataset_name = "adsl",
    adt = lstalvdt;                                        date         = LSTALVDT,
    cnsr = 1;                                              set_values_to = exprs(EVNTDESC = "LAST KNOWN ALIVE")
    evntdesc = "LAST KNOWN ALIVE";                       )
  end;
  aval = adt - startdt + 1;                              derive_param_tte(
  paramcd = "OS";                                          dataset_adsl      = adsl,
  param   = "Overall Survival";                            start_date        = TRTSDT,
run;                                                       event_conditions  = list(death_event),
                                                           censor_conditions = list(alive_censor),
                                                           source_datasets   = list(adsl = adsl),
                                                           set_values_to     = exprs(
                                                             PARAMCD = "OS",
                                                             PARAM   = "Overall Survival"
                                                           )
                                                         )
```

> **SAS analog mental model.** The SAS DATA step *mixes* the event/censor business logic with the loop over subjects. admiral *separates* them: the event/censor rules become small, reviewable, named objects, and `derive_param_tte()` is the engine that loops. The result is more code lines but vastly less ambiguity.

### 1.11 Inspect the toy data

```{r adtte-shape-demo}
# One row per (subject × parameter)
# Key variables: STARTDT, ADT, AVAL, CNSR, EVNTDESC, SRCDOM, SRCVAR
adsl_tte |>
  select(USUBJID, TRTSDT, DTHDT, LSTALVDT, DTHFL) |>
  arrange(USUBJID)
```

> **Read the output.** Five subjects. Three are alive (`DTHFL = "N"`, `DTHDT = NA`) — they will become OS rows with `CNSR = 1` and `ADT = LSTALVDT`. Two died (`DTHFL = "Y"`, `DTHDT` populated) — they will become OS rows with `CNSR = 0` and `ADT = DTHDT`. Note that for the dead subjects, `LSTALVDT = DTHDT`: by convention the last known alive date for a deceased subject is their death date.

### 1.12 Exercise: predict the ADTTE row shape

```{r adtte-predict-ex, exercise=TRUE, exercise.lines=14}
# Without running derive_param_tte yet, predict by hand:
# For each subject, what should (PARAMCD=OS) ADT and CNSR be?
#
# Fill in the expected_cnsr and expected_adt columns.
# Hint: if DTHDT is missing, CNSR = 1 and ADT = LSTALVDT.
#       Otherwise CNSR = 0 and ADT = DTHDT.

adsl_tte |>
  mutate(
    expected_cnsr = if_else(is.na(DTHDT), ___, ___),
    expected_adt  = if_else(is.na(DTHDT), ___, ___)
  ) |>
  select(USUBJID, DTHDT, LSTALVDT, expected_cnsr, expected_adt)
```

```{r adtte-predict-ex-hint-1}
# expected_cnsr = if_else(is.na(DTHDT), 1, 0)
# CNSR = 1 when censored (no death); CNSR = 0 when event (death observed)
```

```{r adtte-predict-ex-hint-2}
# expected_adt = if_else(is.na(DTHDT), LSTALVDT, DTHDT)
# When there is no death, the last known alive date is the censoring date
```

```{r adtte-predict-ex-solution}
adsl_tte |>
  mutate(
    expected_cnsr = if_else(is.na(DTHDT), 1, 0),
    expected_adt  = if_else(is.na(DTHDT), LSTALVDT, DTHDT)
  ) |>
  select(USUBJID, DTHDT, LSTALVDT, expected_cnsr, expected_adt)
```

### 1.13 Quiz — ADTTE shape

```{r quiz-adtte-shape, echo=FALSE}
quiz(
  question(
    "CNSR = 0 means:",
    answer("The subject was censored (no event observed by ADT)"),
    answer("The event occurred", correct = TRUE),
    answer("ADT is missing"),
    answer("The subject was not in the analysis"),
    allow_retry = TRUE
  ),
  question(
    "AVAL in ADTTE is typically:",
    answer("The analysis value of a measurement"),
    answer("ADT minus STARTDT plus 1 (time on study in days)", correct = TRUE),
    answer("CNSR × 100"),
    answer("A numeric version of PARAMCD"),
    allow_retry = TRUE
  ),
  question(
    "Why does CDISC ADaM IG mandate `AVAL = ADT - STARTDT + 1` rather than just `ADT - STARTDT`?",
    answer("Because the +1 converts to months"),
    answer("Because STARTDT is defined as Day 1 (not Day 0) — an event on the same calendar day as STARTDT must have AVAL = 1, not 0", correct = TRUE),
    answer("Because SAS adds 1 by default"),
    answer("It is a historical typo that was never fixed"),
    allow_retry = TRUE
  ),
  question(
    "ADTTE has one row per:",
    answer("Subject × Visit"),
    answer("Subject × PARAMCD (endpoint)", correct = TRUE),
    answer("Subject × AETERM"),
    answer("Subject only — PARAMCD is in a separate dataset"),
    allow_retry = TRUE
  ),
  question(
    "Which SAS PROC will most commonly consume ADTTE as input?",
    answer("PROC SORT"),
    answer("PROC TABULATE"),
    answer("PROC LIFETEST (and PROC PHREG)", correct = TRUE),
    answer("PROC SUMMARY"),
    allow_retry = TRUE
  )
)
```

---

## 2 — event_source() and censor_source()

### 2.1 The source-pattern: data + filter + date + label

The core admiral idea for TTE is that you do **not** code the IF/ELSE chain directly. Instead you declare, one at a time, the *candidate rows* that could become the analysis row. Each declaration is a small named object built from four pieces:

1. **`dataset_name`** — which input dataset has the candidate row? (ADAE, ADRS, ADSL, ADTR, ...)
2. **`filter`** — which rows in that dataset are candidates? (`AESER == "Y" & TRTEMFL == "Y"`)
3. **`date`** — which column of that filtered dataset gives the candidate date? (`ASTDT`, `DTHDT`, `LSTALVDT`, ...)
4. **`set_values_to`** — what label/metadata should the resulting ADTTE row carry? (`EVNTDESC`, `SRCDOM`, `SRCVAR`)

Two flavors of this object exist: `event_source()` (the candidate is an *event*) and `censor_source()` (the candidate is a *censor*). They are nearly identical — the only difference is the implicit `censor` flag (`0` for events, `1+` for censors).

### 2.2 event_source — a row that becomes an event

```
                         ── event_source() ──

                            ┌──────────────────┐
                            │   filter rows    │   AESER == "Y" & TRTEMFL == "Y"
   ADAE  ────────────►      │   from dataset   │   ───────────────────────────────►   candidate event rows
                            └──────────────────┘
                                     │
                                     ▼
                            ┌──────────────────┐
                            │   extract date   │   date = ASTDT
                            └──────────────────┘
                                     │
                                     ▼
                            ┌──────────────────┐
                            │   tag metadata   │   EVNTDESC = "FIRST SERIOUS AE"
                            │  (set_values_to) │   SRCDOM   = "ADAE"
                            └──────────────────┘
                                     │
                                     ▼
                          { CNSR = 0, ADT = ASTDT, ... }
```

If no row in `adae` passes the filter for a subject, this `event_source` contributes nothing for that subject — the next event source (or a censor source) gets to try.

### 2.3 censor_source — a row that becomes a censoring

```
                         ── censor_source() ──

                            ┌──────────────────┐
                            │   filter rows    │   (no filter — use all ADSL rows)
   ADSL  ────────────►      │   from dataset   │   ───────────────────────────────►   candidate censor rows
                            └──────────────────┘
                                     │
                                     ▼
                            ┌──────────────────┐
                            │   extract date   │   date = LSTALVDT
                            └──────────────────┘
                                     │
                                     ▼
                            ┌──────────────────┐
                            │   tag metadata   │   EVNTDESC = "LAST KNOWN ALIVE"
                            │  (set_values_to) │   SRCDOM   = "ADSL"
                            └──────────────────┘
                                     │
                                     ▼
                          { CNSR = 1, ADT = LSTALVDT, ... }
```

### 2.4 Event-vs-censor priority

`derive_param_tte()` enforces a simple but powerful rule:

1. Look at *all* event sources. If *any* row qualifies for the subject, the **earliest** event date wins. `CNSR = 0`.
2. If no event source produces a row, look at *all* censor sources. The **latest** censor date wins. `CNSR` = the censor code of the source that produced it (usually `1`).

The asymmetry — earliest event vs latest censor — is biologically correct: the *first* event is what triggers analysis; the *most recent* evidence of being event-free is the best censor information.

### 2.5 Anatomy of `event_source()`

```r
death_event <- event_source(
  dataset_name  = "adsl",                # which dataset
  filter        = !is.na(DTHDT),         # which rows qualify
  date          = DTHDT,                 # which date column
  set_values_to = exprs(                 # row metadata
    EVNTDESC = "DEATH",
    SRCDOM   = "ADSL",
    SRCVAR   = "DTHDT"
  )
)
```

| Argument | Type | What it does | SAS analog |
|---|---|---|---|
| `dataset_name` | character | Names a dataset in `source_datasets` | `set adsl;` |
| `filter` | unevaluated expression | Subsets to candidate rows | `where DTHDT ne .;` |
| `date` | bare column name | Column whose value becomes `ADT` | `adt = dthdt;` |
| `set_values_to` | `exprs(...)` list | Columns added to the output row | `evntdesc = "DEATH";` |

> **SAS analog mental model.** Think of `event_source()` as a single branch of a giant `SELECT/WHEN` block: "*WHEN this row pattern matches, emit a candidate event row with this date and these labels.*" The `IF DTHDT NE . THEN OUTPUT EVENTS;` pattern in SAS becomes a small reusable object in R.

### 2.6 Anatomy of `censor_source()`

```r
last_alive_censor <- censor_source(
  dataset_name  = "adsl",
  filter        = NULL,                  # (optional — no filter = all rows)
  date          = LSTALVDT,
  set_values_to = exprs(
    EVNTDESC = "LAST DATE KNOWN ALIVE",
    SRCDOM   = "ADSL",
    SRCVAR   = "LSTALVDT"
  ),
  censor        = 1                      # (default 1 — secondary censors can be 2, 3, ...)
)
```

The only extra argument is `censor`, an integer that becomes `CNSR` in the output. Use `1` for the standard "censored, no event observed" — use `2` or higher when the SAP distinguishes *which* censoring rule fired (e.g., "censored at randomization due to no post-baseline assessment" might be `CNSR = 2`).

### 2.7 SAS equivalent: explicit IF/ELSE in a DATA step

The pattern below is the canonical SAS-only PFS prep — note how every business rule is interleaved with `OUTPUT` statements:

```sas
data candidates;
  set adsl;
  /* candidate 1: death event */
  if dthdt ne . then do;
    adt = dthdt; cnsr = 0; evntdesc = "DEATH"; srcdom = "ADSL"; srcvar = "DTHDT";
    output;
  end;
  /* candidate 2: last known alive censor */
  if lstalvdt ne . then do;
    adt = lstalvdt; cnsr = 1; evntdesc = "LAST KNOWN ALIVE"; srcdom = "ADSL"; srcvar = "LSTALVDT";
    output;
  end;
run;

proc sort data=candidates; by usubjid cnsr adt; run;

/* Earliest event wins (sort by cnsr asc, adt asc and take first.usubjid) */
data adtte_os;
  set candidates;
  by usubjid;
  if first.usubjid;
  startdt = trtsdt;
  aval = adt - startdt + 1;
  paramcd = "OS"; param = "Overall Survival";
run;
```

The admiral equivalent (next section) collapses all of this into one `derive_param_tte()` call.

### 2.8 Build and inspect the source objects

```{r source-objects-demo}
# Define event: death
death_event <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(
    EVNTDESC = "DEATH",
    SRCDOM   = "ADSL",
    SRCVAR   = "DTHDT"
  )
)

# Define censoring: last known alive
last_alive_censor <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(
    EVNTDESC = "LAST DATE KNOWN ALIVE",
    SRCDOM   = "ADSL",
    SRCVAR   = "LSTALVDT"
  )
)

# Inspect structure
str(death_event)
```

> **Read the output.** `str(death_event)` shows a six-element list: `type = "event"`, `dataset_name = "adsl"`, an unevaluated `filter` expression (deferred until the derivation runs), the `date` symbol, the `set_values_to` list of label expressions, and `censor = 0`. None of the source datasets have been touched yet — this object is a *recipe*, not data. The data plumbing happens later in `derive_param_tte()`.

### 2.9 Build a second event_source and inspect the contrast

```{r source-objects-contrast, exercise=TRUE, exercise.lines=18}
# Build a censor source for end-of-study and inspect the difference
eos_censor <- censor_source(
  dataset_name  = "adsl",
  date          = EOSDT,
  set_values_to = exprs(EVNTDESC = "END OF STUDY", SRCDOM = "ADSL", SRCVAR = "EOSDT")
)

cat("--- death_event ---\n")
cat("type:        ", death_event$type, "\n")
cat("censor flag: ", death_event$censor, "\n\n")

cat("--- eos_censor ---\n")
cat("type:        ", ___, "\n")
cat("censor flag: ", ___, "\n")
```

```{r source-objects-contrast-hint-1}
# eos_censor is a list; access fields the same way as death_event
# eos_censor$type   should be "censor"
# eos_censor$censor should be 1
```

```{r source-objects-contrast-hint-2}
# cat("type:        ", eos_censor$type,   "\n")
# cat("censor flag: ", eos_censor$censor, "\n")
```

```{r source-objects-contrast-solution}
eos_censor <- censor_source(
  dataset_name  = "adsl",
  date          = EOSDT,
  set_values_to = exprs(EVNTDESC = "END OF STUDY", SRCDOM = "ADSL", SRCVAR = "EOSDT")
)

cat("--- death_event ---\n")
cat("type:        ", death_event$type, "\n")
cat("censor flag: ", death_event$censor, "\n\n")

cat("--- eos_censor ---\n")
cat("type:        ", eos_censor$type, "\n")
cat("censor flag: ", eos_censor$censor, "\n")
```

### 2.10 Quiz — source objects

```{r quiz-source-objects, echo=FALSE}
quiz(
  question(
    "What field in the source object says which actual dataset to pull data from?",
    answer("filter"),
    answer("date"),
    answer("dataset_name", correct = TRUE),
    answer("set_values_to"),
    allow_retry = TRUE
  ),
  question(
    "What differentiates an event_source() from a censor_source()?",
    answer("event_source() has a filter; censor_source() doesn't"),
    answer("event_source() produces CNSR = 0; censor_source() produces CNSR = 1+", correct = TRUE),
    answer("censor_source() uses a different date variable"),
    answer("They are identical — only the name differs"),
    allow_retry = TRUE
  ),
  question(
    "If two event_source objects fire for the same subject (e.g., death AND progression both occurred), which wins?",
    answer("The one listed first in event_conditions"),
    answer("Death always wins"),
    answer("The earlier candidate date wins — the first event to occur drives AVAL", correct = TRUE),
    answer("Both are kept as two rows"),
    allow_retry = TRUE
  ),
  question(
    "Why is `filter` stored as an unevaluated expression rather than a logical vector?",
    answer("Because R requires it"),
    answer("Because the source object is a recipe — the filter is applied later inside derive_param_tte when the actual dataset is in scope (deferred evaluation)", correct = TRUE),
    answer("So that the filter can be modified after the object is built"),
    answer("For performance reasons only"),
    allow_retry = TRUE
  )
)
```

---

## 3 — Overall Survival with derive_param_tte()

### 3.1 What Overall Survival means (clinical)

**Overall Survival (OS)** is the gold-standard efficacy endpoint in oncology. It measures the time from a clinical anchor (randomization or first dose) until **death from any cause**. There is no ambiguity: any death counts, regardless of attribution. This is why OS is preferred by regulators — it cannot be confounded by adjudication bias or subjective response assessments.

A subject is **censored** for OS if, at database lock, they are *known to be alive* on some date (the last contact with the site). The censoring date is the most recent date for which we have evidence the subject was alive — typically `LSTALVDT` in ADSL.

### 3.2 OS start date: TRTSDT or RANDDT — pick wisely

The Statistical Analysis Plan (SAP) tells you which clock to start. Two common conventions:

- **Randomization-based OS** — `STARTDT = RANDDT`. Captures the entire risk period after randomization, including pre-treatment delays. Required for ITT analyses.
- **Treatment-based OS** — `STARTDT = TRTSDT`. Excludes pre-treatment time. Used for safety-population OS or for studies where treatment start is the relevant clinical anchor.

> **Common mistake.** If RANDDT is the SAP-specified clock but you use TRTSDT, you systematically *under-state* AVAL for any subject with a dose-delay. P21 will flag this (see Section 7.5 Diagnostic 1).

### 3.3 OS event: death

`event_source(dataset_name = "adsl", filter = !is.na(DTHDT), date = DTHDT, ...)`. Simple and unambiguous.

### 3.4 OS censor: last known alive date

`censor_source(dataset_name = "adsl", date = LSTALVDT, ...)`. By CDISC convention, for a deceased subject `LSTALVDT = DTHDT`, but the event source still wins because events take priority over censors.

### 3.5 The "death wins over alive" priority

For any subject with both a death event and a last-alive date populated, `derive_param_tte()` picks the **event** (because events always take priority over censors). The last-alive censor is *ignored* when an event exists.

### 3.6 Building OS step by step

The mental model:

```
   For each subject in dataset_adsl:
   ┌──────────────────────────────────────────────────┐
   │  1. Try each event_source. Collect candidates.   │
   │  2. If any candidates exist, pick earliest ADT.  │
   │     CNSR = 0. Done.                              │
   │  3. Else try each censor_source. Pick latest ADT.│
   │     CNSR = censor flag of the winning source.    │
   │  4. AVAL = ADT - STARTDT + 1.                    │
   │  5. Attach EVNTDESC, SRCDOM, SRCVAR labels.      │
   │  6. Emit one row with PARAMCD, PARAM, ...        │
   └──────────────────────────────────────────────────┘
```

### 3.7 Anatomy of `derive_param_tte()` for OS

```r
derive_param_tte(
  dataset_adsl      = adsl_tte,                 # population frame: one row per subject
  start_date        = TRTSDT,                   # the STARTDT column to pull from ADSL
  event_conditions  = list(death_event),        # list of event_source objects
  censor_conditions = list(last_alive_censor),  # list of censor_source objects
  source_datasets   = list(adsl = adsl_tte),    # name → actual data.frame
  set_values_to     = exprs(
    PARAMCD = "OS",
    PARAM   = "Overall Survival"
  )
)
```

| Argument | What it does |
|---|---|
| `dataset_adsl` | The "population frame". Defines which subjects appear in the output. Filter to the analysis population (e.g., `SAFFL == "Y"`) *here*. |
| `start_date` | A bare column name from `dataset_adsl`. Becomes `STARTDT`. |
| `event_conditions` | A `list()` of one or more `event_source()` objects. |
| `censor_conditions` | A `list()` of one or more `censor_source()` objects. |
| `source_datasets` | A named list mapping `dataset_name` strings (used in the sources) to the actual data.frames. |
| `set_values_to` | A list of expressions for row-level metadata. Always include `PARAMCD` and `PARAM`. |

> **SAS analog mental model.** `derive_param_tte()` *is* a parameterized macro. The "fixed" arguments (population frame, start date, source datasets) define the loop scope. The "variable" arguments (event_conditions, censor_conditions, set_values_to) define the business logic. In SAS you would write a macro:
>
> `%MACRO derive_tte(adsl=, startdt=, event_rules=, censor_rules=, paramcd=);`
>
> ...and call it once per endpoint. `derive_param_tte()` is the directly translated equivalent.

### 3.8 SAS equivalent: PROC SQL with COALESCE

```sas
proc sql;
  create table adtte_os as
  select  a.usubjid,
          a.trtsdt          as startdt,
          coalesce(a.dthdt, a.lstalvdt) as adt format=date9.,
          case when a.dthdt is not null then 0 else 1 end as cnsr,
          case when a.dthdt is not null then "DEATH"
               else "LAST DATE KNOWN ALIVE" end as evntdesc,
          (coalesce(a.dthdt, a.lstalvdt) - a.trtsdt + 1) as aval,
          "OS"                as paramcd,
          "Overall Survival"  as param
  from    adsl as a
  where   a.saffl = "Y";
quit;
```

Compare to the `derive_param_tte()` call: the SAS version *hardcodes* the priority (DTHDT first via `COALESCE`); admiral *declares* it (events before censors). For OS the two are equivalent — but for PFS with multiple events, the admiral declaration scales while the SAS `COALESCE` quickly becomes unreadable.

### 3.9 Common pitfalls

- **Wrong start date** — using `TRTSDT` when the SAP says `RANDDT` (or vice versa). Section 7.5 Diagnostic 1.
- **Censoring before treatment** — including last-alive dates that precede `STARTDT` produces negative `AVAL`. Always validate that `AVAL >= 1`.
- **Forgetting the +1** — Section 7.5 Diagnostic 3. The reference date is Day 1, not Day 0.
- **Mixing date types** — if `DTHDT` is a `Date` but `LSTALVDT` is a numeric, the comparison may fail silently. Standardize before passing into the sources.

### 3.10 Build OS for the toy data

```{r os-demo}
# Build OS: event = death, censor = last alive date
adtte_os <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = TRTSDT,
  event_conditions  = list(death_event),
  censor_conditions = list(last_alive_censor),
  source_datasets   = list(adsl = adsl_tte),
  set_values_to     = exprs(
    PARAMCD = "OS",
    PARAM   = "Overall Survival"
  )
)

adtte_os |>
  select(USUBJID, PARAMCD, STARTDT, ADT, AVAL, CNSR, EVNTDESC)
```

> **Read the output.** Five subjects, five rows. Subjects 01-002 and 01-004 have `CNSR = 0` (death events): `EVNTDESC = "DEATH"`, `ADT = DTHDT`. The other three have `CNSR = 1`: `EVNTDESC = "LAST DATE KNOWN ALIVE"`, `ADT = LSTALVDT`. `AVAL` is days from `TRTSDT` to `ADT` plus 1 — for subject 01-002, `19480 - 19370 + 1 = 111` days. This is precisely the dataset shape `PROC LIFETEST DATA=adtte_os; TIME AVAL*CNSR(1);` expects.

### 3.11 OS diagnostics

```{r os-diagnostics-demo}
# Three diagnostic angles your statistician will run by hand
adtte_os |>
  summarise(
    n_subjects    = n(),
    n_events      = sum(CNSR == 0, na.rm = TRUE),
    n_censored    = sum(CNSR == 1, na.rm = TRUE),
    median_aval   = median(AVAL, na.rm = TRUE),
    min_aval      = min(AVAL, na.rm = TRUE),
    max_aval      = max(AVAL, na.rm = TRUE),
    any_negative  = any(AVAL < 1, na.rm = TRUE)
  )
```

> **Read the output.** The four most common P21/QC checks: (1) every subject in the population is represented; (2) event count matches the death count in ADSL; (3) `min(AVAL) >= 1` — no negatives, no zeros; (4) `median(AVAL)` is in a clinically plausible range. If any of these look wrong, stop and recheck the source definitions before showing the dataset to your statistician.

### 3.12 Quiz — Overall Survival

```{r quiz-os, echo=FALSE}
quiz(
  question(
    "For a subject with DTHDT = 19480 and TRTSDT = 19370, what is AVAL?",
    answer("19480 - 19370 = 110"),
    answer("19480 - 19370 + 1 = 111", correct = TRUE),
    answer("110 / 365.25"),
    answer("1"),
    allow_retry = TRUE
  ),
  question(
    "Which SAS construct is the closest analog to `derive_param_tte(start_date = TRTSDT, event_conditions = list(death_event), censor_conditions = list(last_alive_censor), ...)`?",
    answer("`PROC LIFETEST DATA=adtte TIME AVAL*CNSR(1);`"),
    answer("A DATA step that builds STARTDT, then a COALESCE(DTHDT, LSTALVDT) into ADT with CNSR = 0 if DTHDT non-missing else 1, then AVAL = ADT - STARTDT + 1", correct = TRUE),
    answer("`PROC SORT BY USUBJID DESCENDING DTHDT;`"),
    answer("A `RETAIN ADT;` statement carrying forward the last visit date"),
    allow_retry = TRUE
  ),
  question(
    "A subject is alive at last contact. Their OS row should have:",
    answer("CNSR = 0, EVNTDESC = 'DEATH'"),
    answer("CNSR = 1, EVNTDESC = 'LAST DATE KNOWN ALIVE' (or the label from the censor_source)", correct = TRUE),
    answer("CNSR = NA, ADT = NA — alive subjects are not in ADTTE"),
    answer("CNSR = 2, EVNTDESC = 'ALIVE'"),
    allow_retry = TRUE
  ),
  question(
    "Your SAP specifies OS clock starts at randomization. You use `start_date = TRTSDT`. What kind of bug have you introduced?",
    answer("None — TRTSDT and RANDDT are always equal"),
    answer("AVAL is systematically understated for subjects with dose delays (time between RANDDT and TRTSDT is excluded from the risk period)", correct = TRUE),
    answer("AVAL is systematically overstated"),
    answer("CNSR will be flipped"),
    allow_retry = TRUE
  )
)
```

### 3.13 Exercise — toy OS with RANDDT as start

```{r os-ex, exercise=TRUE, exercise.lines=22}
# Exercise: build OS using adsl_tte
# event = !is.na(DTHDT), date = DTHDT
# censor = date = LSTALVDT
# STARTDT = RANDDT (not TRTSDT)

death_ev <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = ___,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST KNOWN ALIVE", SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

adtte_os2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = ___,
  event_conditions  = list(death_ev),
  censor_conditions = list(alive_cs),
  source_datasets   = list(adsl = adsl_tte),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

adtte_os2 |> select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

```{r os-ex-hint-1}
# date = DTHDT
# start_date = RANDDT
```

```{r os-ex-solution}
death_ev <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST KNOWN ALIVE", SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

adtte_os2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(death_ev),
  censor_conditions = list(alive_cs),
  source_datasets   = list(adsl = adsl_tte),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

adtte_os2 |> select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

### Exercise: Overall Survival on the real CDISCPILOT01 study

Scale up from the 5-subject toy to all 254 real subjects. `adsl_real` from pharmaverseadam contains DTHDT, LSTALVDT, SAFFL, TRTSDT, and TRT01A — everything needed for a submission-quality OS derivation.

> **Note:** In pharmaverseadam, DTHDT and LSTALVDT are Date objects. The mock `derive_param_tte()` handles Date arithmetic correctly (`as.numeric(Date − Date)` = days), so AVAL will be in days as expected.

```{r os-real, exercise=TRUE, exercise.lines=48}
# Check the death-related columns available in adsl_real
cat("Key TTE variables in adsl_real:\n")
cat(intersect(c("DTHDT","LSTALVDT","DTHFL","TRTSDT","RANDDT","SAFFL","TRT01A"),
              names(adsl_real)), "\n\n")

# Define event and censor sources using real ADSL variables
death_ev_real <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs_real <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST DATE KNOWN ALIVE",
                         SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

# Derive OS for the safety population
adtte_os_real <- derive_param_tte(
  dataset_adsl      = adsl_real |> filter(SAFFL == "Y"),
  start_date        = TRTSDT,
  event_conditions  = list(death_ev_real),
  censor_conditions = list(alive_cs_real),
  source_datasets   = list(adsl = adsl_real),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

# Summary
cat("OS subjects:", nrow(adtte_os_real), "\n")
cat("Events (CNSR = 0):", sum(adtte_os_real$CNSR == 0, na.rm = TRUE), "\n")
cat("Censored (CNSR = 1):", sum(adtte_os_real$CNSR == 1, na.rm = TRUE), "\n\n")

# Median survival in days by treatment arm
adtte_os_real |>
  group_by(TRT01A, CNSR) |>
  summarise(n = n(), median_aval = median(AVAL, na.rm = TRUE), .groups = "drop") |>
  arrange(TRT01A, CNSR) |>
  print()
```

```{r os-real-hint-1}
# adsl_real has DTHDT (Date), LSTALVDT (Date), TRTSDT (Date)
# filter(SAFFL == "Y") restricts to the safety population (~254 subjects)
# CNSR = 0 → event occurred; CNSR = 1 → censored (alive at last contact)
# median(AVAL) gives median follow-up in days per arm/event-status group
```

```{r os-real-solution}
death_ev_real <- event_source(
  dataset_name  = "adsl",
  filter        = !is.na(DTHDT),
  date          = DTHDT,
  set_values_to = exprs(EVNTDESC = "DEATH", SRCDOM = "ADSL", SRCVAR = "DTHDT")
)

alive_cs_real <- censor_source(
  dataset_name  = "adsl",
  date          = LSTALVDT,
  set_values_to = exprs(EVNTDESC = "LAST DATE KNOWN ALIVE",
                         SRCDOM = "ADSL", SRCVAR = "LSTALVDT")
)

adtte_os_real <- derive_param_tte(
  dataset_adsl      = adsl_real |> filter(SAFFL == "Y"),
  start_date        = TRTSDT,
  event_conditions  = list(death_ev_real),
  censor_conditions = list(alive_cs_real),
  source_datasets   = list(adsl = adsl_real),
  set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
)

cat("OS subjects:", nrow(adtte_os_real), "\n")
cat("Events (CNSR = 0):", sum(adtte_os_real$CNSR == 0, na.rm = TRUE), "\n")
cat("Censored (CNSR = 1):", sum(adtte_os_real$CNSR == 1, na.rm = TRUE), "\n\n")

adtte_os_real |>
  group_by(TRT01A, CNSR) |>
  summarise(n = n(), median_aval = median(AVAL, na.rm = TRUE), .groups = "drop") |>
  arrange(TRT01A, CNSR) |>
  print()
```

---

## 4 — PFS: composite events

PFS uses **two event sources**: disease progression OR death, whichever comes first.

```{r pfs-demo}
# Progression event
pd_event <- event_source(
  dataset_name  = "adrs",
  filter        = PARAMCD == "OVR" & AVALC == "PD",
  date          = ADT,
  set_values_to = exprs(
    EVNTDESC = "DISEASE PROGRESSION",
    SRCDOM   = "ADRS",
    SRCVAR   = "ADT"
  )
)

# Death (reuse from above)
# last tumor assessment censor
last_assess <- censor_source(
  dataset_name  = "adrs",
  filter        = PARAMCD == "OVR" & AVALC %in% c("CR","PR","SD","NE"),
  date          = ADT,
  set_values_to = exprs(
    EVNTDESC = "LAST TUMOR ASSESSMENT",
    SRCDOM   = "ADRS",
    SRCVAR   = "ADT"
  )
)

adtte_pfs <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
)

adtte_pfs |>
  select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

```{r quiz-pfs, echo=FALSE}
quiz(
  question(
    "When both progression AND death occur for the same subject, derive_param_tte() picks:",
    answer("Death always wins"),
    answer("Progression always wins"),
    answer("The earliest of the two event dates", correct = TRUE),
    answer("The one listed first in event_conditions"),
    allow_retry = TRUE
  ),
  question(
    "A subject with no progression and no death gets CNSR =",
    answer("0"),
    answer("1 (from the censor source)", correct = TRUE),
    answer("2 — double censor"),
    answer("NA"),
    allow_retry = TRUE
  )
)
```

```{r pfs-ex, exercise=TRUE, exercise.lines=26}
# Exercise: build PFS but add a randomization-date fallback censor
# (for subjects with no assessments at all, censor = 2)
# Then count subjects by CNSR value

rand_censor <- censor_source(
  dataset_name  = "adsl",
  date          = RANDDT,
  censor        = ___,        # use censor = 2
  set_values_to = exprs(EVNTDESC = "RANDOMIZATION (NO ASSESSMENT)",
                         SRCDOM = "ADSL", SRCVAR = "RANDDT")
)

adtte_pfs2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor, rand_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

# Count by CNSR
adtte_pfs2 |> count(CNSR)
```

```{r pfs-ex-hint-1}
# censor = 2
```

```{r pfs-ex-solution}
rand_censor <- censor_source(
  dataset_name  = "adsl",
  date          = RANDDT,
  censor        = 2,
  set_values_to = exprs(EVNTDESC = "RANDOMIZATION (NO ASSESSMENT)",
                         SRCDOM = "ADSL", SRCVAR = "RANDDT")
)

adtte_pfs2 <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor, rand_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

adtte_pfs2 |> count(CNSR)
```

---

## 5 — Time to first serious AE

```{r ttaeser-demo}
# Event: first treatment-emergent serious AE
first_sae_event <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y" & TRTEMFL == "Y",
  date          = ASTDT,
  set_values_to = exprs(
    EVNTDESC = "FIRST SERIOUS AE",
    SRCDOM   = "ADAE",
    SRCVAR   = "ASTDT"
  )
)

eot_censor <- censor_source(
  dataset_name  = "adsl",
  date          = EOSDT,
  set_values_to = exprs(
    EVNTDESC = "END OF STUDY",
    SRCDOM   = "ADSL",
    SRCVAR   = "EOSDT"
  )
)

adtte_sae <- derive_param_tte(
  dataset_adsl      = adsl_tte |> filter(SAFFL == "Y"),
  start_date        = TRTSDT,
  event_conditions  = list(first_sae_event),
  censor_conditions = list(eot_censor, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adae = adae_tte),
  set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
)

adtte_sae |> select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC)
```

---

## 6 — call_derivation() for multiple parameters

```{r call-derivation-demo}
# Build OS and PFS together in one call_derivation call
adtte_both <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions  = list(death_event),
      censor_conditions = list(last_alive_censor),
      set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
    ),
    params(
      event_conditions  = list(pd_event, death_event),
      censor_conditions = list(last_assess, last_alive_censor),
      set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
    )
  ),
  # Shared arguments
  dataset_adsl    = adsl_tte,
  start_date      = "RANDDT",
  source_datasets = list(adsl = adsl_tte, adrs = adrs)
)

adtte_both |>
  select(USUBJID, PARAMCD, AVAL, CNSR, EVNTDESC) |>
  arrange(PARAMCD, USUBJID)
```

```{r quiz-call-derivation, echo=FALSE}
quiz(
  question(
    "In `call_derivation()`, the `variable_params` list contains:",
    answer("Shared arguments applied to every call"),
    answer("Per-call argument overrides (event_conditions, censor_conditions, set_values_to)", correct = TRUE),
    answer("Dataset names only"),
    answer("Filter conditions"),
    allow_retry = TRUE
  )
)
```

```{r call-derivation-ex, exercise=TRUE, exercise.lines=28}
# Exercise: use call_derivation to build OS and TTAESER in one block
# OS: death event, last alive censor, start = TRTSDT
# TTAESER: first_sae_event, eot_censor + last_alive_censor, start = TRTSDT

adtte_multi <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions  = list(death_event),
      censor_conditions = list(last_alive_censor),
      set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
    ),
    params(
      event_conditions  = list(first_sae_event),
      censor_conditions = list(eot_censor, last_alive_censor),
      set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
    )
  ),
  dataset_adsl    = adsl_tte |> filter(SAFFL == "Y"),
  start_date      = ___,
  source_datasets = list(adsl = adsl_tte, adae = adae_tte)
)

adtte_multi |> select(USUBJID, PARAMCD, AVAL, CNSR) |> arrange(PARAMCD, USUBJID)
```

```{r call-derivation-ex-hint-1}
# start_date = TRTSDT
```

```{r call-derivation-ex-solution}
adtte_multi <- call_derivation(
  derivation = derive_param_tte,
  variable_params = list(
    params(
      event_conditions  = list(death_event),
      censor_conditions = list(last_alive_censor),
      set_values_to     = exprs(PARAMCD = "OS", PARAM = "Overall Survival")
    ),
    params(
      event_conditions  = list(first_sae_event),
      censor_conditions = list(eot_censor, last_alive_censor),
      set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
    )
  ),
  dataset_adsl    = adsl_tte |> filter(SAFFL == "Y"),
  start_date      = TRTSDT,
  source_datasets = list(adsl = adsl_tte, adae = adae_tte)
)

adtte_multi |> select(USUBJID, PARAMCD, AVAL, CNSR) |> arrange(PARAMCD, USUBJID)
```

---

## 7 — AVAL in months/years and validation

```{r aval-units-demo}
# AVAL in days → months conversion
adtte_os |>
  mutate(
    AVAL_MONTHS = AVAL / (365.25 / 12),
    AVAL_YEARS  = AVAL / 365.25
  ) |>
  select(USUBJID, PARAMCD, AVAL, AVAL_MONTHS, AVAL_YEARS, CNSR)
```

```{r quiz-validation, echo=FALSE}
quiz(
  question(
    "Which CNSR distribution check is a key ADTTE validation step?",
    answer("Confirming no negative AVAL values (AVAL should always be ≥ 1)", correct = TRUE),
    answer("Confirming CNSR sums to zero"),
    answer("Confirming every subject has exactly one PARAMCD"),
    answer("Confirming ADT > STARTDT for all rows"),
    allow_retry = TRUE
  ),
  question(
    "Which check ensures completeness of the subject population?",
    answer("count(CNSR)"),
    answer("Every subject in dataset_adsl (filtered population) has exactly one row per PARAMCD", correct = TRUE),
    answer("nrow() matches ADSL"),
    answer("SRCDOM is non-missing"),
    allow_retry = TRUE
  )
)
```

---

## 7.5 — P21 Derivation Diagnostics

ADTTE errors are among the most consequential in a submission: a wrong AVAL or misclassified CNSR can distort Kaplan-Meier curves and median survival estimates. Each diagnostic presents a define.xml derivation spec and the Pinnacle 21 error that revealed the bug.

---

### Diagnostic 1 — PFS: Wrong start date (TRTSDT instead of RANDDT)

**Derivation method (define.xml / Pinnacle 21 field):**

> *STARTDT for Progression-Free Survival (PFS) = RANDDT (date of randomization). The analysis clock starts at randomization, not at first dose, to capture dose-delay effects. AVAL = ADT − RANDDT + 1 days. Source: ADSL.RANDDT.*

The programmer submitted:

```r
adtte_pfs_draft <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = TRTSDT,   # ← bug: should be RANDDT
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)
```

**P21 report:** *"AVAL for PARAMCD = 'PFS' does not align with expected values. STARTDT appears equal to TRTSDT rather than RANDDT. For subjects with dose delays, AVAL is understated. Expected AVAL = ADT − RANDDT + 1."*

```{r p21-bug-pfs-startdate, exercise=TRUE, exercise.lines=22}
# Fix: PFS uses RANDDT as the start date (randomization date), not TRTSDT.

adtte_pfs_fixed <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = ___,       # fix: randomization date
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
)

# Compare AVAL from RANDDT vs TRTSDT: subjects 01-001 and 01-002 had
# RANDDT 7 days before TRTSDT, so AVAL should be 7 days longer with RANDDT.
adtte_pfs_wrong <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = TRTSDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

left_join(
  adtte_pfs_fixed |> select(USUBJID, AVAL_correct = AVAL, CNSR),
  adtte_pfs_wrong |> select(USUBJID, AVAL_wrong   = AVAL),
  by = "USUBJID"
) |> mutate(diff = AVAL_correct - AVAL_wrong)
```

```{r p21-bug-pfs-startdate-hint-1}
# start_date = RANDDT  (randomization date, not first treatment date)
# TRTSDT may differ from RANDDT if the subject had a dose delay
# RANDDT is always the correct clock start for PFS per the statistical analysis plan
```

```{r p21-bug-pfs-startdate-solution}
adtte_pfs_fixed <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = RANDDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "Progression-Free Survival")
)

adtte_pfs_wrong <- derive_param_tte(
  dataset_adsl      = adsl_tte,
  start_date        = TRTSDT,
  event_conditions  = list(pd_event, death_event),
  censor_conditions = list(last_assess, last_alive_censor),
  source_datasets   = list(adsl = adsl_tte, adrs = adrs),
  set_values_to     = exprs(PARAMCD = "PFS", PARAM = "PFS")
)

left_join(
  adtte_pfs_fixed |> select(USUBJID, AVAL_correct = AVAL, CNSR),
  adtte_pfs_wrong |> select(USUBJID, AVAL_wrong   = AVAL),
  by = "USUBJID"
) |> mutate(diff = AVAL_correct - AVAL_wrong)
```

---

### Diagnostic 2 — TTAESER: Missing TRTEMFL filter includes pre-treatment SAEs

**Derivation method (define.xml / Pinnacle 21 field):**

> *TTAESER event = first treatment-emergent serious adverse event. Filter criteria: AESER = 'Y' AND TRTEMFL = 'Y'. AEs that are serious but occurred before first dose (TRTEMFL ≠ 'Y') must be excluded. Only on-treatment SAEs trigger the event. Source: ADAE.AESER, ADAE.TRTEMFL, ADAE.ASTDT.*

The programmer submitted:

```r
sae_event_draft <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y",    # ← bug: TRTEMFL == "Y" is missing
  date          = ASTDT,
  set_values_to = exprs(EVNTDESC = "FIRST SERIOUS AE",
                         SRCDOM = "ADAE", SRCVAR = "ASTDT")
)
```

**P21 report:** *"TTAESER event date precedes TRTSDT for subject 01-001. A pre-treatment SAE (TRTEMFL = 'N') is counted as the event. Event filter must include TRTEMFL = 'Y'."*

```{r p21-bug-ttaeser-filter, exercise=TRUE, exercise.lines=30}
# Self-contained: build a small AE dataset with one pre-treatment SAE
ae_with_pretrt <- tibble(
  USUBJID = c("01-001", "01-001", "01-002"),
  AETERM  = c("PRE-STUDY SAE", "ANAPHYLACTIC SHOCK", "CARDIAC ARREST"),
  ASTDT   = c(19340, 19400, 19400),   # 19340 is before TRTSDT = 19367 for 01-001
  AESER   = c("Y",   "Y",    "Y"),
  TRTEMFL = c("N",   "Y",    "Y")     # first row is pre-treatment
)

# Broken: no TRTEMFL filter → pre-treatment SAE for 01-001 triggers event
sae_ev_broken <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y",
  date          = ASTDT,
  set_values_to = exprs(EVNTDESC = "FIRST SERIOUS AE", SRCDOM = "ADAE", SRCVAR = "ASTDT")
)

# Fix: add TRTEMFL == "Y" to the filter
sae_ev_fixed <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y" & ___,     # add the missing condition
  date          = ASTDT,
  set_values_to = exprs(EVNTDESC = "FIRST SERIOUS AE", SRCDOM = "ADAE", SRCVAR = "ASTDT")
)

broken <- derive_param_tte(
  dataset_adsl = adsl_tte, start_date = TRTSDT,
  event_conditions = list(sae_ev_broken),
  censor_conditions = list(eot_censor),
  source_datasets = list(adsl = adsl_tte, adae = ae_with_pretrt),
  set_values_to = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
)

fixed <- derive_param_tte(
  dataset_adsl = adsl_tte, start_date = TRTSDT,
  event_conditions = list(sae_ev_fixed),
  censor_conditions = list(eot_censor),
  source_datasets = list(adsl = adsl_tte, adae = ae_with_pretrt),
  set_values_to = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
)

cat("--- Broken (includes pre-treatment SAE for 01-001) ---\n")
broken |> select(USUBJID, AVAL, CNSR, EVNTDESC) |> print()
cat("\n--- Fixed (only on-treatment SAEs trigger event) ---\n")
fixed  |> select(USUBJID, AVAL, CNSR, EVNTDESC) |> print()
```

```{r p21-bug-ttaeser-filter-hint-1}
# filter = AESER == "Y" & TRTEMFL == "Y"
# Without TRTEMFL == "Y", pre-treatment SAEs (TRTEMFL = "N") incorrectly trigger the event
# For subject 01-001, the event date would be ASTDT = 19340, which is before TRTSDT = 19367
```

```{r p21-bug-ttaeser-filter-solution}
ae_with_pretrt <- tibble(
  USUBJID = c("01-001", "01-001", "01-002"),
  AETERM  = c("PRE-STUDY SAE", "ANAPHYLACTIC SHOCK", "CARDIAC ARREST"),
  ASTDT   = c(19340, 19400, 19400),
  AESER   = c("Y", "Y", "Y"),
  TRTEMFL = c("N", "Y", "Y")
)

sae_ev_fixed <- event_source(
  dataset_name  = "adae",
  filter        = AESER == "Y" & TRTEMFL == "Y",
  date          = ASTDT,
  set_values_to = exprs(EVNTDESC = "FIRST SERIOUS AE", SRCDOM = "ADAE", SRCVAR = "ASTDT")
)

fixed <- derive_param_tte(
  dataset_adsl = adsl_tte, start_date = TRTSDT,
  event_conditions  = list(sae_ev_fixed),
  censor_conditions = list(eot_censor),
  source_datasets   = list(adsl = adsl_tte, adae = ae_with_pretrt),
  set_values_to     = exprs(PARAMCD = "TTAESER", PARAM = "Time to First Serious AE")
)

fixed |> select(USUBJID, AVAL, CNSR, EVNTDESC) |> print()
```

---

### Diagnostic 3 — AVAL: Missing +1 in manual duration calculation

**Derivation method (define.xml / Pinnacle 21 field):**

> *AVAL = ADT − STARTDT + 1 (days). Follows CDISC ADaM IG §4.4.7 study day convention: AVAL = 1 when ADT = STARTDT (event on Day 1 of follow-up). A subject who dies on the first day of follow-up has AVAL = 1, not 0. derive_param_tte() computes this automatically. Manual computation must include the +1.*

The programmer post-processed the ADTTE dataset:

```r
adtte_recomputed <- adtte_os |>
  mutate(AVAL = ADT - STARTDT)    # ← bug: missing +1
```

**P21 report:** *"AD0214 — AVAL = 0 detected for 2 subjects where ADT = STARTDT. CDISC ADaM IG section 4.4.7: duration of 0 is not permissible for time-to-event variables. Expected AVAL ≥ 1 for all rows."*

```{r p21-bug-aval-plusone, exercise=TRUE, exercise.lines=20}
# adtte_os was built earlier by derive_param_tte() — it has correct AVAL.
# Simulate what happens when a programmer recalculates AVAL manually.

# Introduce a subject where ADT = STARTDT to expose the bug clearly
adtte_test <- adtte_os |>
  mutate(
    ADT_orig  = ADT,
    # Force subject 01-001 to have ADT = STARTDT (event on Day 1)
    ADT       = if_else(USUBJID == "01-001", STARTDT, ADT),
    AVAL_wrong  = ADT - STARTDT,           # ← missing +1
    AVAL_correct = ADT - STARTDT + 1       # ← fix: include +1
  )

# Show the difference: AVAL_wrong = 0 for 01-001, AVAL_correct = 1
adtte_test |>
  select(USUBJID, STARTDT, ADT, AVAL_wrong, AVAL_correct, CNSR)
```

```{r p21-bug-aval-plusone-hint-1}
# AVAL = ADT - STARTDT + 1
# The +1 is mandatory: the reference date (STARTDT) is Day 1, not Day 0
# derive_param_tte() already applies this: aval = as.numeric(adt - startdt) + 1
# Only add +1 when computing AVAL manually outside of admiral functions
```

```{r p21-bug-aval-plusone-solution}
adtte_test <- adtte_os |>
  mutate(
    ADT          = if_else(USUBJID == "01-001", STARTDT, ADT),
    AVAL_wrong   = ADT - STARTDT,
    AVAL_correct = ADT - STARTDT + 1
  )

adtte_test |>
  select(USUBJID, STARTDT, ADT, AVAL_wrong, AVAL_correct, CNSR)
```

---

## Quiz — final

```{r quiz-final-18, echo=FALSE}
quiz(
  question(
    "A PFS derivation uses `start_date = TRTSDT`. P21 reports AVAL is understated for subjects with dose delays. What is the correct start date for PFS and why?",
    answer("TRTSDT is correct — PFS always starts at first dose"),
    answer("RANDDT — PFS follow-up begins at randomization; using TRTSDT shortens AVAL for subjects who started treatment after randomization", correct = TRUE),
    answer("EOSDT — the study end date anchors the survival clock"),
    answer("LSTALVDT — the last known alive date defines the PFS clock"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "A TTAESER event_source uses `filter = AESER == 'Y'`. P21 reports an event date before TRTSDT for one subject. What is the missing condition?",
    answer("AESER == 'Y' is sufficient — pre-treatment SAEs are excluded automatically by the TTE function"),
    answer("TRTEMFL == 'Y' must be added — without it, pre-treatment serious AEs (which have TRTEMFL = 'N') incorrectly trigger the event", correct = TRUE),
    answer("ASTDT > TRTSDT must be added as a date filter"),
    answer("AESER == 'Y' & AESEV == 'SEVERE' is the correct filter"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "P21 rule AD0214 flags AVAL = 0 in ADTTE. A programmer computed `AVAL = ADT - STARTDT`. What is the fix and why?",
    answer("AVAL = ADT - STARTDT + 1 — CDISC ADaM IG defines the reference date as Day 1 (AVAL = 1 when ADT = STARTDT), so no duration of 0 exists", correct = TRUE),
    answer("AVAL = ADT - STARTDT is correct; AD0214 only applies to ADSL"),
    answer("AVAL = ADT - STARTDT + 0.5, rounded up to 1 for Day 1 records"),
    answer("AVAL = (ADT - STARTDT) / 365.25 converts days to years"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "PFS uses two event_source objects (progression and death). When both occur for the same subject, derive_param_tte() selects:",
    answer("The event listed first in the event_conditions list"),
    answer("Death always takes priority over progression"),
    answer("The event with the earliest ADT — the first event to occur determines AVAL and CNSR = 0", correct = TRUE),
    answer("The last event source in the list"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "CNSR = 0 in ADTTE means the event occurred. A subject is alive at study end with no events. What values should CNSR and EVNTDESC take?",
    answer("CNSR = 0, EVNTDESC = 'NO EVENT'"),
    answer("CNSR = 1 (from the censor_source), EVNTDESC = the label defined in the applicable censor_source set_values_to (e.g., 'LAST DATE KNOWN ALIVE')", correct = TRUE),
    answer("CNSR = NA — missing indicates no event and no censoring"),
    answer("CNSR = 2 — all censor sources default to censor = 2"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "Multiple censor_source objects are listed. Which censoring date does derive_param_tte() use for a subject with no event?",
    answer("The latest date across all censor sources — the most recent evidence the subject was event-free", correct = TRUE),
    answer("The earliest date — conservative censoring"),
    answer("The censor source listed first in the list"),
    answer("All censor dates are averaged"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "EVNTDESC is not a required CDISC variable, but it is universally included. What does it provide?",
    answer("A numeric code for the event type used in regression models"),
    answer("Traceability — it records exactly which event_source or censor_source rule fired for each subject, enabling auditors to verify each row's derivation", correct = TRUE),
    answer("The MedDRA preferred term for the event"),
    answer("A link back to the SDTM source domain"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "call_derivation() builds OS and PFS in one block. What goes in variable_params vs the shared fixed arguments?",
    answer("variable_params holds shared arguments like start_date and source_datasets; fixed arguments hold per-endpoint event/censor sources"),
    answer("variable_params holds per-endpoint arguments (event_conditions, censor_conditions, set_values_to); shared arguments (start_date, source_datasets, dataset_adsl) are passed outside variable_params", correct = TRUE),
    answer("All arguments go in variable_params — the fixed argument list is ignored"),
    answer("variable_params holds only PARAMCD; all other arguments are fixed"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "A subject has DTHDT = NA and LSTALVDT = 19600 in ADSL. What CNSR and ADT should the OS row have?",
    answer("CNSR = 0, ADT = NA — missing death date means the event cannot be confirmed"),
    answer("CNSR = 1, ADT = 19600 — no death event found, so the latest censor date (LSTALVDT) is used", correct = TRUE),
    answer("CNSR = NA — the subject is excluded from the OS analysis"),
    answer("CNSR = 2, ADT = EOSDT — a second-level censor is applied"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  ),
  question(
    "The {admiralonco} extension package provides pre-built TTE source objects. What is the primary benefit over coding event_source() from scratch?",
    answer("Pre-built objects run faster than custom event_source() calls"),
    answer("Pre-built source objects implement validated, oncology-standard event and censoring rules (RECIST-based PFS, OS per BICR) that reduce the risk of specification deviations across studies", correct = TRUE),
    answer("admiralonco source objects work without source_datasets — they fetch data automatically"),
    answer("They eliminate the need for SRCDOM and SRCVAR metadata"),
    random_answer_order = TRUE,
    allow_retry = TRUE
  )
)
```
