---
title: "Lesson 19 — admiral Part 6: Advanced Derivation Patterns"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# ── shared mocks ─────────────────────────────────────────────────────────────
exprs <- function(...) { dots <- as.list(substitute(list(...)))[-1]; dots }
params <- function(...) list(...)

.impute_dtc <- function(dtc, date_imputation = "first") {
  sapply(dtc, function(d) {
    if (is.na(d)) return(NA_character_)
    if (nchar(d) == 4) paste0(d, if(date_imputation=="last") "-12-31" else "-01-01")
    else if (nchar(d) == 7) {
      if (date_imputation == "last") {
        yr <- as.integer(substr(d,1,4)); mo <- as.integer(substr(d,6,7))
        nxt <- if(mo==12) paste0(yr+1,"-01-01") else paste0(yr,"-",sprintf("%02d",mo+1),"-01")
        as.character(as.Date(nxt)-1)
      } else paste0(d,"-01")
    } else d
  })
}

convert_blanks_to_na <- function(dataset) {
  dataset |> mutate(across(where(is.character), ~ifelse(.=="", NA_character_, .)))
}

derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars=NULL, ...) {
  by_nm <- sapply(by_vars, deparse)
  jdf   <- dataset_add
  if (!is.null(new_vars)) {
    old <- sapply(new_vars, function(x) { d<-deparse(x); if(grepl("=",d)) trimws(strsplit(d,"=")[[1]][2]) else d })
    new <- sapply(new_vars, function(x) { d<-deparse(x); if(grepl("=",d)) trimws(strsplit(d,"=")[[1]][1]) else d })
    keep <- intersect(c(by_nm, old), names(jdf)); jdf <- jdf |> select(all_of(keep))
    for (i in seq_along(old)) if(old[i]!=new[i] && old[i] %in% names(jdf)) jdf <- rename(jdf, !!new[i] := !!old[i])
  }
  left_join(dataset, jdf, by = intersect(by_nm, names(jdf)))
}

derive_vars_dt <- function(dataset, dtc, new_vars_prefix,
                            date_imputation="none", highest_imputation="n",
                            flag_imputation="auto", ...) {
  dtc_nm <- deparse(substitute(dtc))
  imp    <- if (date_imputation %in% c("none","") && highest_imputation=="n") "none"
            else if (date_imputation=="last") "last" else "first"
  raw    <- if (dtc_nm %in% names(dataset)) dataset[[dtc_nm]] else rep(NA_character_, nrow(dataset))
  imp_d  <- if (imp=="none") raw else .impute_dtc(raw, imp)
  dataset[[paste0(new_vars_prefix,"DT")]] <- suppressWarnings(as.numeric(as.Date(imp_d)))
  dataset
}

derive_vars_dy <- function(dataset, reference_date, source_vars) {
  ref <- deparse(substitute(reference_date))
  for (sv in sapply(source_vars, deparse)) {
    dy <- sub("DT$","DY",sv)
    if (sv %in% names(dataset) && ref %in% names(dataset)) {
      v <- dataset[[sv]]; r <- dataset[[ref]]
      dataset[[dy]] <- ifelse(is.na(v)|is.na(r), NA_real_,
        ifelse(v>=r, as.numeric(v-r)+1, as.numeric(v-r)))
    }
  }
  dataset
}

derive_var_extreme_flag <- function(dataset, by_vars, order, new_var, mode="first") {
  new_nm  <- deparse(substitute(new_var))
  by_nm   <- sapply(by_vars, deparse)
  ord_nms <- sapply(order, function(x) gsub("desc\\(|\\)","",deparse(x)))
  desc_fl <- sapply(order, function(x) grepl("^desc\\(",deparse(x)))
  odf <- dataset
  for (i in rev(seq_along(ord_nms))) if(ord_nms[i] %in% names(odf))
    odf <- if(desc_fl[i]) arrange(odf, desc(.data[[ord_nms[i]]])) else arrange(odf, .data[[ord_nms[i]]])
  byp <- intersect(by_nm, names(odf))
  if (mode=="first")
    fl <- odf |> group_by(across(all_of(byp))) |> slice(1) |> ungroup() |>
          mutate(!!new_nm := "Y") |> select(all_of(c(byp, new_nm)))
  else
    fl <- odf |> group_by(across(all_of(byp))) |> slice(n()) |> ungroup() |>
          mutate(!!new_nm := "Y") |> select(all_of(c(byp, new_nm)))
  if (new_nm %in% names(dataset)) dataset[[new_nm]] <- NULL
  left_join(dataset, fl, by = intersect(names(fl), names(dataset)))
}

restrict_derivation <- function(dataset, derivation, args=list(), filter=NULL) {
  fe <- substitute(filter)
  in_rows <- if(!is.null(fe)) tryCatch(which(eval(fe, dataset, parent.frame())), error=function(e) seq_len(nrow(dataset)))
             else seq_len(nrow(dataset))
  out_rows <- setdiff(seq_len(nrow(dataset)), in_rows)
  sub      <- dataset[in_rows,, drop=FALSE]
  derived  <- tryCatch(do.call(derivation, c(list(dataset=sub), args)), error=function(e) sub)
  bind_rows(derived, dataset[out_rows,, drop=FALSE])
}

derivation_slice <- function(filter=NULL, args=list()) list(filter=substitute(filter), args=args)

slice_derivation <- function(dataset, derivation, args=list(), ...) {
  slices  <- list(...)
  result  <- dataset
  covered <- integer(0)
  for (sl in slices) {
    fe  <- sl$filter
    idx <- if(!is.null(fe)) tryCatch(which(eval(fe, result, parent.frame())), error=function(e) integer(0))
           else seq_len(nrow(result))
    idx <- setdiff(idx, covered)
    if (!length(idx)) next
    merged_args <- modifyList(args, sl$args)
    sub <- result[idx,, drop=FALSE]
    drv <- tryCatch(do.call(derivation, c(list(dataset=sub), merged_args)), error=function(e) sub)
    result[idx, names(drv)] <- drv[seq_len(length(idx)), names(drv)]
    covered <- c(covered, idx)
  }
  result
}

derive_vars_joined <- function(dataset, dataset_add, by_vars, order,
                                new_vars=NULL, join_vars=NULL,
                                filter_join=NULL, mode="first", join_type="all", ...) {
  by_nm <- sapply(by_vars, deparse)
  fj    <- substitute(filter_join)
  mrgd  <- left_join(dataset, dataset_add, by=intersect(by_nm, names(dataset_add)), suffix=c("",".add"))
  if (!is.null(fj)) mrgd <- tryCatch(filter(mrgd, !!fj), error=function(e) mrgd)
  if (!is.null(new_vars)) {
    old <- sapply(new_vars, function(x) { d<-deparse(x); if(grepl("=",d)) trimws(strsplit(d,"=")[[1]][2]) else d })
    new <- sapply(new_vars, function(x) { d<-deparse(x); if(grepl("=",d)) trimws(strsplit(d,"=")[[1]][1]) else d })
    for (i in seq_along(old)) if(old[i] %in% names(mrgd)) mrgd <- rename(mrgd, !!new[i] := !!old[i])
  }
  ord_nms <- sapply(order, function(x) gsub("desc\\(|\\)","",deparse(x)))
  desc_fl <- sapply(order, function(x) grepl("^desc\\(",deparse(x)))
  for (i in rev(seq_along(ord_nms))) if(ord_nms[i] %in% names(mrgd))
    mrgd <- if(desc_fl[i]) arrange(mrgd, desc(.data[[ord_nms[i]]])) else arrange(mrgd, .data[[ord_nms[i]]])
  byp <- intersect(by_nm, names(mrgd))
  mrgd <- if(mode=="first") mrgd |> group_by(across(all_of(byp))) |> slice(1) |> ungroup()
          else mrgd |> group_by(across(all_of(byp))) |> slice(n()) |> ungroup()
  new_cols <- if(!is.null(new_vars)) sapply(new_vars, function(x){d<-deparse(x); if(grepl("=",d)) trimws(strsplit(d,"=")[[1]][1]) else d})
              else setdiff(names(mrgd), c(byp, names(dataset)))
  extra <- mrgd |> select(all_of(c(byp, intersect(new_cols, names(mrgd)))))
  for (nm in intersect(new_cols, names(dataset))) dataset[[nm]] <- NULL
  left_join(dataset, extra, by=byp)
}

# ── Advanced pattern mocks ────────────────────────────────────────────────────
create_period_dataset <- function(dataset, new_vars = list()) {
  # pivot APxxSDT/APxxEDT to long format
  nms <- names(dataset)
  sdt_cols <- sort(grep("^AP[0-9]+SDT$", nms, value=TRUE))
  edt_cols <- sort(grep("^AP[0-9]+EDT$", nms, value=TRUE))
  if (!length(sdt_cols)) return(dataset)
  periods <- as.integer(sub("^AP([0-9]+)SDT$","\\1", sdt_cols))
  rows <- lapply(seq_along(periods), function(i) {
    p <- periods[i]
    dataset |>
      select(any_of(c("STUDYID","USUBJID"))) |>
      mutate(
        APERIOD  = p,
        APERSDT  = dataset[[sdt_cols[i]]],
        APEREDT  = dataset[[edt_cols[i]]]
      )
  })
  bind_rows(rows) |> filter(!is.na(APERSDT))
}

derive_expected_records <- function(dataset, dataset_ref, by_vars,
                                     set_values_to = list()) {
  by_nm  <- sapply(by_vars, deparse)
  subj_vars <- intersect(c("STUDYID","USUBJID"), by_nm)
  param_vars <- setdiff(by_nm, subj_vars)
  # cartesian of subjects × parameter/visit combos
  if (length(param_vars) && all(param_vars %in% names(dataset))) {
    combos <- dataset |>
      select(all_of(param_vars)) |> distinct()
  } else {
    combos <- tibble(.rows = 1)
  }
  if (length(subj_vars)) {
    subj_df <- dataset_ref |> select(all_of(intersect(subj_vars, names(dataset_ref))))
    full_grid <- cross_join(subj_df, combos)
  } else {
    full_grid <- combos
  }
  missing <- anti_join(full_grid, dataset, by = intersect(names(full_grid), names(dataset)))
  for (nm in names(set_values_to)) {
    val <- tryCatch(eval(set_values_to[[nm]]), error=function(e) as.character(set_values_to[[nm]]))
    missing[[nm]] <- val
  }
  bind_rows(dataset, missing)
}

derive_locf_records <- function(dataset, dataset_ref = NULL, by_vars, order,
                                 keep_vars = NULL) {
  by_nm  <- sapply(by_vars, deparse)
  ord_nms <- sapply(order, function(x) gsub("desc\\(|\\)","",deparse(x)))
  kv_nms  <- if (!is.null(keep_vars)) sapply(keep_vars, deparse) else character(0)
  by_present <- intersect(by_nm, names(dataset))
  ord_present <- intersect(ord_nms, names(dataset))
  dataset |>
    arrange(across(all_of(c(by_present, ord_present)))) |>
    group_by(across(all_of(by_present))) |>
    mutate(across(all_of(intersect(kv_nms, names(dataset))),
                  ~ifelse(is.na(.) & any(!is.na(.)), zoo::na.locf(., na.rm=FALSE), .))) |>
    ungroup()
}

derive_extreme_event <- function(dataset, by_vars, events, new_vars, mode="first", order=NULL) {
  new_nm <- sapply(new_vars, deparse)
  by_nm  <- sapply(by_vars, deparse)
  by_p   <- intersect(by_nm, names(dataset))
  result <- dataset
  for (nm in new_nm) result[[nm]] <- NA_character_
  for (ev in events) {
    fe <- ev$condition
    idx <- tryCatch(which(eval(fe, result, parent.frame())), error=function(e) integer(0))
    if (!length(idx)) next
    for (nm in names(ev$set_values_to)) {
      val <- tryCatch(eval(ev$set_values_to[[nm]]),
                      error=function(e) as.character(ev$set_values_to[[nm]]))
      already_set <- if (!is.null(result[[nm]])) !is.na(result[[nm]]) else rep(FALSE, nrow(result))
      result[[nm]][idx[!already_set[idx]]] <- val
    }
  }
  result
}

event <- function(condition, set_values_to) {
  list(condition=substitute(condition), set_values_to=set_values_to)
}

# admiral options
.admiral_opts <- new.env(parent=emptyenv())
.admiral_opts$subject_keys <- exprs(STUDYID, USUBJID)

get_admiral_option <- function(option) {
  .admiral_opts[[option]]
}
set_admiral_options <- function(...) {
  opts <- list(...)
  for (nm in names(opts)) .admiral_opts[[nm]] <- opts[[nm]]
  invisible(NULL)
}

# metatools mocks (for metacore integration demo)
apply_variable_labels <- function(dataset, metacore_obj) {
  if (!is.null(metacore_obj$var_spec)) {
    lbl_map <- setNames(metacore_obj$var_spec$label, metacore_obj$var_spec$variable)
    for (v in intersect(names(lbl_map), names(dataset))) {
      attr(dataset[[v]], "label") <- lbl_map[v]
    }
  }
  dataset
}

check_variables <- function(dataset, metacore_obj) {
  if (!is.null(metacore_obj$ds_vars)) {
    expected <- metacore_obj$ds_vars$variable
    missing  <- setdiff(expected, names(dataset))
    extra    <- setdiff(names(dataset), expected)
    if (length(missing)) warning("Missing variables: ", paste(missing, collapse=", "))
    if (length(extra))   message("Extra variables (not in spec): ", paste(extra, collapse=", "))
  }
  invisible(NULL)
}

check_ct_data  <- function(dataset, metacore_obj) invisible(NULL)
check_unique_keys <- function(dataset, metacore_obj, key_vars=NULL) invisible(NULL)
build_from_derived <- function(metacore_obj, ds_list, predecessor_only=FALSE, keep=FALSE) {
  bind_rows(ds_list)
}

# ── Sample datasets ───────────────────────────────────────────────────────────
adsl_adv <- tribble(
  ~STUDYID, ~USUBJID, ~TRTSDT, ~TRTEDT, ~TRT01A,
  ~AP01SDT, ~AP01EDT, ~AP02SDT, ~AP02EDT, ~SAFFL,
  "STUDY1","01-001", 19367, 19548, "TrtA",  19367, 19456, 19457, 19548, "Y",
  "STUDY1","01-002", 19370, 19551, "TrtB",  19370, 19459, 19460, 19551, "Y",
  "STUDY1","01-003", 19375, 19540, "TrtA",  19375, 19464, 19465, 19540, "Y"
)

adae_adv <- tribble(
  ~STUDYID, ~USUBJID, ~AESEQ, ~AETERM,       ~ASTDT,  ~TRTEMFL, ~AESOC,
  ~AETOXGR, ~AESER, ~TRTSDT,
  "STUDY1","01-001", 1, "HEADACHE",     19400,  "Y", "Nervous system disorders", 1, "N", 19367,
  "STUDY1","01-001", 2, "SEVERE RASH",  19480,  "Y", "Skin disorders",           4, "Y", 19367,
  "STUDY1","01-002", 1, "NAUSEA",       19390,  "Y", "GI disorders",             2, "N", 19370,
  "STUDY1","01-003", 1, "FATIGUE",      19400,  "Y", "General disorders",        1, "N", 19375
)

advs_adv <- tribble(
  ~STUDYID, ~USUBJID, ~PARAMCD, ~AVISIT,  ~AVISITN, ~ADT,  ~AVAL, ~TRTSDT,
  "STUDY1","01-001","SYSBP","Baseline",  0, 19365, 120, 19367,
  "STUDY1","01-001","SYSBP","Week 4",    4, 19395, 125, 19367,
  "STUDY1","01-001","SYSBP","Week 8",    8, 19423,  NA, 19367,
  "STUDY1","01-002","SYSBP","Baseline",  0, 19368, 145, 19370,
  "STUDY1","01-002","SYSBP","Week 4",    4, 19398, 138, 19370,
  "STUDY1","01-002","SYSBP","Week 8",    8, 19426, 141, 19370,
  "STUDY1","01-003","SYSBP","Baseline",  0, 19373, 132, 19375,
  "STUDY1","01-003","SYSBP","Week 4",    4, 19403, 129, 19375
)

mc <- list(
  ds_vars  = tibble(dataset="ADAE", variable=c("STUDYID","USUBJID","AESEQ","AETERM","ASTDT","TRTEMFL","AESOC","AETOXGR","AESER","TRTSDT","ASEV","AREL")),
  var_spec = tibble(variable=c("STUDYID","USUBJID","AESEQ","AETERM","ASTDT","TRTEMFL"),
                    label   =c("Study ID","Unique Subject ID","Seq Num","AE Term","Analysis Start Date","TE Flag"))
)
class(mc) <- c("Metacore","list")
```

## Welcome

This lesson covers **advanced admiral patterns**: period datasets for crossover studies, expected-record insertion, LOCF imputation, `derive_extreme_event()`, `restrict_derivation()` + `slice_derivation()` combinations, metacore integration, custom function authoring, and package-level options.

---

## 1 — Period datasets

```{r period-demo}
# adsl_adv has AP01SDT, AP01EDT, AP02SDT, AP02EDT
adsl_adv |> select(USUBJID, AP01SDT, AP01EDT, AP02SDT, AP02EDT)
```

```{r create-periods-demo}
# Pivot to long format: one row per (USUBJID × APERIOD)
adperiods <- create_period_dataset(
  dataset  = adsl_adv,
  new_vars = exprs(APERSDT = APxxSDT, APEREDT = APxxEDT)
)

adperiods
```

```{r period-ae-demo}
# Assign APERIOD to each AE based on AE start date falling in period window
adae_periods <- adae_adv |>
  derive_vars_joined(
    dataset_add = adperiods,
    by_vars     = exprs(USUBJID),
    order       = exprs(APERSDT),
    filter_join = ASTDT >= APERSDT & ASTDT <= APEREDT,
    mode        = "first"
  )

adae_periods |> select(USUBJID, AETERM, ASTDT, APERIOD, APERSDT, APEREDT)
```

```{r quiz-periods, echo=FALSE}
quiz(
  question(
    "When would you use `create_period_dataset()`?",
    answer("Whenever the study has repeated measures"),
    answer("For crossover or extension studies where each subject passes through multiple distinct treatment periods", correct = TRUE),
    answer("When ADSL has more than one row per subject"),
    answer("To expand EX records per dose day"),
    allow_retry = TRUE
  ),
  question(
    "Why use `derive_vars_joined()` rather than `derive_vars_merged()` to assign APERIOD to events?",
    answer("derive_vars_joined() is faster"),
    answer("filter_join can reference ASTDT (from adae) and APERSDT/APEREDT (from adperiods) — both datasets' variables", correct = TRUE),
    answer("derive_vars_merged() can't handle periods"),
    answer("derive_vars_joined() handles multiple by_vars"),
    allow_retry = TRUE
  )
)
```

```{r period-flag-ex, exercise=TRUE, exercise.lines=18}
# Exercise: derive AOCCPRFL = first treatment-emergent AE per (USUBJID × APERIOD)
# using restrict_derivation + derive_var_extreme_flag on adae_periods

adae_pfl <- adae_periods |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, ___),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCPRFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y" & !is.na(APERIOD)
  )

adae_pfl |> filter(!is.na(AOCCPRFL)) |>
  select(USUBJID, AETERM, APERIOD, AOCCPRFL)
```

```{r period-flag-ex-hint-1}
# by_vars needs to group by both USUBJID and APERIOD → exprs(STUDYID, USUBJID, APERIOD)
```

```{r period-flag-ex-solution}
adae_pfl <- adae_periods |>
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, APERIOD),
      order   = exprs(ASTDT, AESEQ),
      new_var = AOCCPRFL,
      mode    = "first"
    ),
    filter = TRTEMFL == "Y" & !is.na(APERIOD)
  )

adae_pfl |> filter(!is.na(AOCCPRFL)) |>
  select(USUBJID, AETERM, APERIOD, AOCCPRFL)
```

---

## 2 — Expected records: derive_expected_records()

```{r expected-demo}
# advs_adv has a missing Week 8 row for 01-003
advs_adv |> select(USUBJID, PARAMCD, AVISIT, AVISITN, AVAL)
```

```{r insert-expected-demo}
# Insert expected-but-missing rows for each (USUBJID × PARAMCD × AVISIT)
ref_pop <- adsl_adv |> filter(SAFFL == "Y") |> select(STUDYID, USUBJID)

advs_exp <- advs_adv |>
  derive_expected_records(
    dataset_ref   = ref_pop,
    by_vars       = exprs(STUDYID, USUBJID, PARAMCD, AVISITN, AVISIT),
    set_values_to = exprs(DTYPE = "EXPECTED")
  )

advs_exp |>
  filter(PARAMCD == "SYSBP") |>
  select(USUBJID, AVISIT, AVISITN, AVAL, DTYPE) |>
  arrange(USUBJID, AVISITN)
```

```{r quiz-expected, echo=FALSE}
quiz(
  question(
    "What does `derive_expected_records()` do that simply filtering can't?",
    answer("It removes rows with missing AVAL"),
    answer("It inserts placeholder rows for expected (subject × visit × parameter) combinations that are absent from the data", correct = TRUE),
    answer("It merges ADSL variables"),
    answer("It computes AVAL from adjacent visits"),
    allow_retry = TRUE
  ),
  question(
    "DTYPE = 'EXPECTED' on an inserted row allows downstream analyses to:",
    answer("Exclude the row automatically"),
    answer("Include or exclude the row depending on SAP rules — it serves as a distinguishing flag", correct = TRUE),
    answer("Compute LOCF for that row"),
    answer("Flag a data quality issue"),
    allow_retry = TRUE
  )
)
```

```{r expected-ex, exercise=TRUE, exercise.lines=16}
# Exercise: count how many EXPECTED rows were inserted (DTYPE == "EXPECTED")
# compared to the total rows in advs_exp

ref_pop <- adsl_adv |> filter(SAFFL == "Y") |> select(STUDYID, USUBJID)

advs_exp2 <- advs_adv |>
  derive_expected_records(
    dataset_ref   = ref_pop,
    by_vars       = exprs(STUDYID, USUBJID, PARAMCD, AVISITN, AVISIT),
    set_values_to = exprs(DTYPE = "EXPECTED")
  )

list(
  total_rows    = nrow(advs_exp2),
  expected_rows = sum(advs_exp2$DTYPE == "EXPECTED", na.rm = TRUE),
  original_rows = nrow(advs_adv)
)
```

```{r expected-ex-hint-1}
# No changes needed — just run it to see the counts
```

```{r expected-ex-solution}
ref_pop <- adsl_adv |> filter(SAFFL == "Y") |> select(STUDYID, USUBJID)

advs_exp2 <- advs_adv |>
  derive_expected_records(
    dataset_ref   = ref_pop,
    by_vars       = exprs(STUDYID, USUBJID, PARAMCD, AVISITN, AVISIT),
    set_values_to = exprs(DTYPE = "EXPECTED")
  )

list(
  total_rows    = nrow(advs_exp2),
  expected_rows = sum(advs_exp2$DTYPE == "EXPECTED", na.rm = TRUE),
  original_rows = nrow(advs_adv)
)
```

---

## 3 — LOCF imputation

```{r locf-demo}
# advs_adv has NA AVAL at Week 8 for 01-001
advs_adv |> select(USUBJID, AVISITN, AVAL)
```

```{r locf-run-demo}
# LOCF: carry last non-NA AVAL forward for each (USUBJID × PARAMCD)
# Note: requires zoo package (installed) — mock uses simpler logic for exercises
advs_locf <- advs_adv |>
  arrange(USUBJID, PARAMCD, AVISITN) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    AVAL_LOCF = ifelse(is.na(AVAL),
                       dplyr::lag(AVAL[!is.na(AVAL)], default=NA),
                       AVAL),
    DTYPE     = ifelse(is.na(AVAL) & !is.na(lag(AVAL)), "LOCF", DTYPE %||% NA_character_)
  ) |>
  ungroup()

advs_locf |>
  filter(USUBJID == "01-001") |>
  select(USUBJID, AVISIT, AVAL, AVAL_LOCF, DTYPE)
```

```{r quiz-locf, echo=FALSE}
quiz(
  question(
    "DTYPE = 'LOCF' on an imputed row allows downstream analyses to:",
    answer("Remove the row"),
    answer("Include it in primary analysis and exclude it in sensitivity, or vice versa — the flag makes it distinguishable", correct = TRUE),
    answer("Automatically compute CHG from BASE"),
    answer("Flag a regulatory issue"),
    allow_retry = TRUE
  ),
  question(
    "LOCF is considered:",
    answer("The gold-standard imputation method"),
    answer("A simple imputation method — older protocols use it; modern guidance prefers more sophisticated approaches", correct = TRUE),
    answer("Mandatory per CDISC"),
    answer("Equivalent to multiple imputation"),
    allow_retry = TRUE
  )
)
```

---

## 4 — derive_extreme_event(): priority-based worst-event

```{r extreme-event-demo}
# Flag the worst AE per subject using priority-ordered conditions
# Grade 5 > Grade 4 > Serious

adae_worst <- adae_adv |>
  derive_extreme_event(
    by_vars  = exprs(STUDYID, USUBJID),
    events   = list(
      event(
        condition     = AETOXGR >= 4,
        set_values_to = exprs(AOWORSEN = "SEVERE")
      ),
      event(
        condition     = AESER == "Y",
        set_values_to = exprs(AOWORSEN = "SERIOUS")
      ),
      event(
        condition     = AETOXGR >= 2,
        set_values_to = exprs(AOWORSEN = "MODERATE")
      )
    ),
    new_vars = exprs(AOWORSEN),
    mode     = "first",
    order    = exprs(ASTDT)
  )

adae_worst |> select(USUBJID, AETERM, AETOXGR, AESER, AOWORSEN)
```

```{r quiz-extreme-event, echo=FALSE}
quiz(
  question(
    "How does `derive_extreme_event()` differ from `derive_var_extreme_flag()`?",
    answer("derive_extreme_event() creates new rows"),
    answer("derive_extreme_event() applies priority-ordered condition checks rather than simply sorting to find first/last", correct = TRUE),
    answer("derive_var_extreme_flag() cannot handle multiple parameters"),
    answer("They are identical"),
    allow_retry = TRUE
  )
)
```

```{r extreme-event-ex, exercise=TRUE, exercise.lines=22}
# Exercise: use derive_extreme_event to create WORSTCAT:
# priority 1: AETOXGR >= 3 → "HIGH GRADE"
# priority 2: AESER == "Y" → "SERIOUS"
# priority 3: AETOXGR >= 1 → "ANY GRADE"

adae_wc <- adae_adv |>
  derive_extreme_event(
    by_vars  = exprs(STUDYID, USUBJID),
    events   = list(
      event(condition = ___ >= 3,   set_values_to = exprs(WORSTCAT = "HIGH GRADE")),
      event(condition = ___ == "Y", set_values_to = exprs(WORSTCAT = "SERIOUS")),
      event(condition = ___ >= 1,   set_values_to = exprs(WORSTCAT = "ANY GRADE"))
    ),
    new_vars = exprs(WORSTCAT),
    mode     = "first",
    order    = exprs(ASTDT)
  )

adae_wc |> select(USUBJID, AETERM, AETOXGR, AESER, WORSTCAT)
```

```{r extreme-event-ex-hint-1}
# condition arguments: AETOXGR >= 3, AESER == "Y", AETOXGR >= 1
```

```{r extreme-event-ex-solution}
adae_wc <- adae_adv |>
  derive_extreme_event(
    by_vars  = exprs(STUDYID, USUBJID),
    events   = list(
      event(condition = AETOXGR >= 3,   set_values_to = exprs(WORSTCAT = "HIGH GRADE")),
      event(condition = AESER == "Y",   set_values_to = exprs(WORSTCAT = "SERIOUS")),
      event(condition = AETOXGR >= 1,   set_values_to = exprs(WORSTCAT = "ANY GRADE"))
    ),
    new_vars = exprs(WORSTCAT),
    mode     = "first",
    order    = exprs(ASTDT)
  )

adae_wc |> select(USUBJID, AETERM, AETOXGR, AESER, WORSTCAT)
```

---

## 5 — slice_derivation + restrict_derivation combined

```{r combined-demo}
# Different ABLFL baseline rules per parameter family
# Slice 1: SYSBP — last pre-treatment row, ordered by ADT
# (height would use a different rule, shown conceptually)
advs_bl <- advs_adv |>
  slice_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, PARAMCD),
      new_var = ABLFL,
      mode    = "last"
    ),
    derivation_slice(
      filter = PARAMCD == "SYSBP" & !is.na(AVAL) & ADT <= TRTSDT,
      args   = params(order = exprs(ADT, AVISITN))
    )
  )

advs_bl |> select(USUBJID, PARAMCD, AVISIT, AVAL, ABLFL)
```

```{r quiz-combined, echo=FALSE}
quiz(
  question(
    "Why would you use `slice_derivation()` with multiple slices over calling `restrict_derivation()` twice?",
    answer("slice_derivation() is always faster"),
    answer("slice_derivation() ensures each row is processed by at most one slice — no double derivation, and each slice can have different arguments", correct = TRUE),
    answer("slice_derivation() doesn't require a filter"),
    answer("restrict_derivation() cannot take mode arguments"),
    allow_retry = TRUE
  )
)
```

---

## 6 — Admiral options and metacore integration

```{r admiral-options-demo}
# Get the default subject keys
get_admiral_option("subject_keys")
```

```{r set-options-demo}
# Override subject keys for this project
set_admiral_options(subject_keys = exprs(STUDYID, USUBJID))

# Now use it
by_subj <- get_admiral_option("subject_keys")
by_subj
```

```{r metacore-integration-demo}
# Pattern: metacore + admiral + metatools together
# mc is a simplified Metacore object pre-built in setup

# Step 1: build base from derived datasets
adae_base <- build_from_derived(mc, ds_list = list("ae" = adae_adv),
                                 predecessor_only = FALSE)

# Step 2: admiral pipeline
adae_integrated <- adae_base |>
  mutate(ASEV = AETOXGR, AREL = AESER)

# Step 3: apply spec labels and check conformance
adae_labeled <- adae_integrated |>
  apply_variable_labels(mc)

# Step 4: check conformance (warnings for missing/extra vars)
check_variables(adae_labeled, mc)

cat("Integration pattern complete. Columns:", paste(names(adae_labeled), collapse=", "))
```

```{r quiz-metacore, echo=FALSE}
quiz(
  question(
    "What does `set_admiral_options(subject_keys = ...)` accomplish project-wide?",
    answer("Renames STUDYID everywhere"),
    answer("Sets a default so all admiral functions use the configured identifier columns — no need to pass by_vars manually every time", correct = TRUE),
    answer("Creates new variables STUDYID and USUBJID"),
    answer("Merges ADSL into every derivation"),
    allow_retry = TRUE
  ),
  question(
    "In the metacore + metatools + admiral architecture, which package is the derivation engine?",
    answer("metacore"),
    answer("metatools"),
    answer("admiral", correct = TRUE),
    answer("xportr"),
    allow_retry = TRUE
  )
)
```

---

## 7 — Custom extension functions

```{r custom-function-demo}
# Writing sponsor-specific helpers in admiral style
# Convention: derive_var_* for single new variable; derive_vars_* for multiple

derive_var_serafl <- function(dataset,
                               by_vars,
                               filter,
                               new_var,
                               mode = "first") {
  # Mock of an assert check (in real code, use admiraldev::assert_data_frame)
  stopifnot(is.data.frame(dataset))
  new_nm <- deparse(substitute(new_var))
  fe     <- substitute(filter)

  dataset |>
    restrict_derivation(
      derivation = derive_var_extreme_flag,
      args = params(
        by_vars = by_vars,
        order   = exprs(ASTDT, AESEQ),
        new_var = !!as.name(new_nm),
        mode    = mode
      ),
      filter = !!fe
    )
}

# Use the custom function
adae_adv |>
  derive_var_serafl(
    by_vars = exprs(STUDYID, USUBJID),
    filter  = AESER == "Y" & TRTEMFL == "Y",
    new_var = SERAFL
  ) |>
  filter(!is.na(SERAFL)) |>
  select(USUBJID, AETERM, AESER, SERAFL)
```

```{r custom-function-ex, exercise=TRUE, exercise.lines=24}
# Exercise: write a custom function `derive_var_rel3fl` that flags the first
# treatment-emergent AE with AETOXGR >= 3 per subject as "Y".
# It should accept: dataset, by_vars, new_var
# and internally call restrict_derivation + derive_var_extreme_flag

derive_var_rel3fl <- function(dataset, by_vars, new_var) {
  new_nm <- deparse(substitute(new_var))
  dataset |>
    restrict_derivation(
      derivation = derive_var_extreme_flag,
      args = params(
        by_vars = by_vars,
        order   = exprs(ASTDT, AESEQ),
        new_var = !!as.name(new_nm),
        mode    = "first"
      ),
      filter = TRTEMFL == "Y" & AETOXGR >= ___
    )
}

adae_adv |>
  derive_var_rel3fl(
    by_vars = exprs(STUDYID, USUBJID),
    new_var = G3FL
  ) |>
  filter(!is.na(G3FL)) |>
  select(USUBJID, AETERM, AETOXGR, G3FL)
```

```{r custom-function-ex-hint-1}
# filter uses AETOXGR >= 3
```

```{r custom-function-ex-solution}
derive_var_rel3fl <- function(dataset, by_vars, new_var) {
  new_nm <- deparse(substitute(new_var))
  dataset |>
    restrict_derivation(
      derivation = derive_var_extreme_flag,
      args = params(
        by_vars = by_vars,
        order   = exprs(ASTDT, AESEQ),
        new_var = !!as.name(new_nm),
        mode    = "first"
      ),
      filter = TRTEMFL == "Y" & AETOXGR >= 3
    )
}

adae_adv |>
  derive_var_rel3fl(
    by_vars = exprs(STUDYID, USUBJID),
    new_var = G3FL
  ) |>
  filter(!is.na(G3FL)) |>
  select(USUBJID, AETERM, AETOXGR, G3FL)
```

---

## 8 — Performance and vignettes

```{r quiz-performance, echo=FALSE}
quiz(
  question(
    "For large datasets, which is preferred when the join condition doesn't reference both datasets?",
    answer("derive_vars_joined()"),
    answer("derive_vars_merged() — it is faster when no cross-dataset filter is needed", correct = TRUE),
    answer("They have identical performance"),
    answer("Always use derive_vars_joined() for safety"),
    allow_retry = TRUE
  ),
  question(
    "Which admiral vignette covers `restrict_derivation()`, `slice_derivation()`, and `call_derivation()` in depth?",
    answer('vignette("date_time")'),
    answer('vignette("higher_order")', correct = TRUE),
    answer('vignette("imputation")'),
    answer('vignette("visit_period")'),
    allow_retry = TRUE
  )
)
```

---

## Quiz — final

```{r quiz-final-19, echo=FALSE}
quiz(
  question(
    "When would you NOT use `create_period_dataset()`?",
    answer("When the study has two treatment periods"),
    answer("When the study is a simple parallel-arm design with a single treatment window per subject", correct = TRUE),
    answer("When ADSL has APxxSDT variables"),
    answer("When the study has an extension phase"),
    allow_retry = TRUE
  ),
  question(
    "DTYPE = 'LOCF' and DTYPE = 'EXPECTED' both serve what common purpose?",
    answer("They mark rows for deletion"),
    answer("They distinguish derived/imputed rows from observed ones, enabling SAP-controlled inclusion/exclusion", correct = TRUE),
    answer("They indicate PARAMCD is missing"),
    answer("They are required by CDISC"),
    allow_retry = TRUE
  ),
  question(
    "A sponsor-internal admiral extension package should:",
    answer("Override admiral core functions"),
    answer("Depend on admiral and add sponsor-specific derive_var_*() helpers following admiral conventions", correct = TRUE),
    answer("Replace metatools"),
    answer("Skip admiraldev checks to reduce dependencies"),
    allow_retry = TRUE
  ),
  question(
    "After Module 4, the next module covers:",
    answer("TLG generation with rtables and tern"),
    answer("Therapeutic area admiral extensions: admiralonco, admiralvaccine, etc.", correct = TRUE),
    answer("SDTM validation with sdtmchecks"),
    answer("xportr and final transport"),
    allow_retry = TRUE
  )
)
```
