---
title: "Lesson 21 — admiral Vaccine Extension (admiralvaccine): Immunogenicity & Reactogenicity ADaMs"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Vaccine trial ADaM derivations using the admiralvaccine package: immunogenicity
  specimen datasets (ADIS) with HAI/MN titres, seroconversion, seroprotection,
  geometric mean titres (GMT), geometric mean fold rise (GMFR), solicited adverse
  events (ADFACE), and vaccine-specific time-to-event. Includes real-world capstone
  problems for influenza and RSV trial scenarios.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 21_admiralvaccine.Rmd
# PURPOSE       : admiralvaccine extension — vaccine immunogenicity ADaM: geometric mean
#                 titres (GMT), geometric mean fold rise (GMFR),
#                 seroconversion, seroprotection, and reactogenicity
#                 (ADFACE) endpoint derivations.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(lubridate)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# SYNTHETIC VACCINE TRIAL DATA
# Simulates a Phase III influenza + RSV bivalent vaccine trial.
# 120 subjects, 2 arms: Vaccine (VAC) vs Placebo (PBO).
# Immunogenicity measured as HAI (influenza) and ELISA (RSV) titres.
# ══════════════════════════════════════════════════════════════════════════════
set.seed(2024)
n_sub <- 120

adsl_vac <- tibble(
  STUDYID  = "VAC2024",
  USUBJID  = sprintf("VAC2024-%03d", seq_len(n_sub)),
  SUBJID   = sprintf("%03d",         seq_len(n_sub)),
  AGE      = sample(18:75, n_sub, replace = TRUE),
  AGEGR1   = cut(sample(18:75, n_sub, replace = TRUE),
                 breaks = c(17, 50, 64, Inf),
                 labels = c("18-50","51-64",">=65")),
  SEX      = sample(c("M","F"), n_sub, replace = TRUE),
  RACE     = sample(c("WHITE","ASIAN","BLACK OR AFRICAN AMERICAN"),
                    n_sub, replace = TRUE, prob = c(.60,.25,.15)),
  ARM      = rep(c("Vaccine","Placebo"), each = n_sub/2),
  ARMCD    = rep(c("VAC","PBO"),         each = n_sub/2),
  TRTSDT   = as.Date("2023-09-01") + sample(0:30, n_sub, replace = TRUE),
  SAFFL    = "Y",
  IMMFL    = "Y"   # immunogenicity evaluable flag
) |>
  mutate(TRTEDT = TRTSDT + 28L)   # single-dose trial, Day 29 = post-vaccination visit

# ── IS domain (Immunogenicity Specimens) ──────────────────────────────────────
# Three antigens: H1N1 (HAI titre), H3N2 (HAI titre), RSV-F (ELISA IU/mL)
# Two visits: Day 1 (pre-vaccination baseline) and Day 29 (post-vaccination)
antigens <- c("H1N1","H3N2","RSV-F")
visits   <- c("DAY 1","DAY 29")
days     <- c(1L, 29L)

# Helper: simulate log-normal titres. Vaccine arm has higher post-vax rise.
sim_titre <- function(arm, antigen, visit, n) {
  is_vac  <- arm == "VAC"
  is_post <- visit == "DAY 29"

  # HAI titres (H1N1, H3N2): baseline ~10-40, vaccine arm 4–8-fold rise
  if (antigen %in% c("H1N1","H3N2")) {
    base_log2 <- rnorm(n, mean = 4.0, sd = 0.8)   # ~log2(16) baseline
    rise      <- if (is_vac && is_post) rnorm(n, mean = 2.5, sd = 0.7)
                 else rnorm(n, mean = 0.2, sd = 0.4)
    titre_raw <- 2^(base_log2 + rise)
    # Round to nearest HAI dilution (10, 20, 40, 80, 160, 320, 640)
    dilutions <- c(10, 20, 40, 80, 160, 320, 640)
    sapply(titre_raw, function(t) dilutions[which.min(abs(dilutions - t))])

  # RSV-F ELISA titres (IU/mL): baseline ~50-200, vaccine arm ~10-fold rise
  } else {
    base_log  <- rnorm(n, mean = log(100), sd = 0.6)
    rise      <- if (is_vac && is_post) rnorm(n, mean = log(10), sd = 0.5)
                 else rnorm(n, mean = log(1.1), sd = 0.3)
    round(exp(base_log + rise), 1)
  }
}

is_rows <- lapply(seq_len(n_sub), function(i) {
  subj  <- adsl_vac$USUBJID[i]
  arm   <- adsl_vac$ARMCD[i]
  trtdt <- adsl_vac$TRTSDT[i]

  lapply(seq_along(antigens), function(a) {
    antigen <- antigens[a]
    lapply(seq_along(visits), function(v) {
      visit <- visits[v]; day <- days[v]
      titre <- sim_titre(arm, antigen, visit, 1)
      tibble(
        STUDYID  = "VAC2024",
        USUBJID  = subj,
        ISSEQ    = as.integer((i - 1) * 6 + (a - 1) * 2 + v),
        VISIT    = visit,
        VISITNUM = as.numeric(v),
        VISITDY  = day,
        ISTESTCD = antigen,
        ISTEST   = case_when(
          antigen == "H1N1"  ~ "H1N1 Influenza HAI Titre",
          antigen == "H3N2"  ~ "H3N2 Influenza HAI Titre",
          antigen == "RSV-F" ~ "RSV Fusion Protein IgG (ELISA)"
        ),
        ISCAT    = if_else(antigen %in% c("H1N1","H3N2"), "INFLUENZA", "RSV"),
        ISORRESU = if_else(antigen %in% c("H1N1","H3N2"), "titre (1:dilution)", "IU/mL"),
        ISSTRESN = titre,
        ISSTRESC = as.character(titre),
        ADT      = trtdt + day - 1L,
        ADY      = day,
        ARMCD    = arm,
        AVAL     = titre
      )
    }) |> bind_rows()
  }) |> bind_rows()
}) |> bind_rows()

# ── Pre-built ADIS skeleton ───────────────────────────────────────────────────
adis_base <- is_rows |>
  mutate(
    PARAMCD = ISTESTCD,
    PARAM   = ISTEST,
    DTYPE   = NA_character_,
    AVISITN = if_else(VISIT == "DAY 1", 1, 29),
    ABLFL   = if_else(VISIT == "DAY 1", "Y", NA_character_)
  )

# Baseline value per subject per antigen
adis_base <- adis_base |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    BASE  = AVAL[VISIT == "DAY 1"][1],
    BASEC = as.character(BASE)
  ) |>
  ungroup()

# ── FACE domain (Solicited Adverse Events — reactogenicity) ──────────────────
# Local: injection-site pain, redness, swelling (Days 1-7)
# Systemic: fatigue, headache, myalgia, fever (Days 1-7)
face_symptoms <- tibble(
  FACECAT  = c(rep("LOCAL", 3), rep("SYSTEMIC", 4)),
  FACETESTCD = c("PAIN","REDNESS","SWELLING","FATIGUE","HEADACHE","MYALGIA","FEVER"),
  FACETEST   = c("Injection Site Pain","Injection Site Redness",
                 "Injection Site Swelling","Fatigue","Headache","Myalgia","Fever")
)

face_rows <- lapply(seq_len(n_sub), function(i) {
  subj <- adsl_vac$USUBJID[i]
  arm  <- adsl_vac$ARMCD[i]
  trtdt <- adsl_vac$TRTSDT[i]

  lapply(seq_len(nrow(face_symptoms)), function(j) {
    sym     <- face_symptoms[j, ]
    is_vac  <- arm == "VAC"
    # Vaccine arm has higher reactogenicity
    prob_any <- if (is_vac) {
      if (sym$FACECAT == "LOCAL") 0.70 else 0.55
    } else {
      if (sym$FACECAT == "LOCAL") 0.25 else 0.20
    }
    occurred <- rbinom(1, 1, prob_any) == 1
    if (!occurred) return(NULL)

    # Grade 1–3 (1=mild, 2=moderate, 3=severe)
    grade  <- sample(1:3, 1, prob = c(.55, .35, .10))
    onset  <- sample(1:3, 1)   # onset day post-vaccination
    dur    <- sample(1:5, 1)   # duration in days

    tibble(
      STUDYID    = "VAC2024",
      USUBJID    = subj,
      FACESEQ    = as.integer(i * 10 + j),
      FACECAT    = sym$FACECAT,
      FACETESTCD = sym$FACETESTCD,
      FACETEST   = sym$FACETEST,
      FACESEV    = c("MILD","MODERATE","SEVERE")[grade],
      FACESEVN   = grade,
      FACESTDT   = trtdt + onset,
      FACEENDT   = trtdt + onset + dur - 1L,
      ADT        = trtdt + onset,
      ADURN      = dur,
      ARMCD      = arm
    )
  }) |> bind_rows()
}) |> bind_rows()

face_rows <- face_rows |> filter(!is.null(USUBJID))

# ══════════════════════════════════════════════════════════════════════════════
# ADMIRALVACCINE FUNCTION MOCKS
# ══════════════════════════════════════════════════════════════════════════════

# ── GMT/GMFR helpers ─────────────────────────────────────────────────────────

#' Geometric Mean Titre (GMT) — exp(mean(log(AVAL)))
#' For HAI titres the convention is log2; for ELISA, natural log.
gmt <- function(x, log_base = exp(1)) {
  x <- x[x > 0 & !is.na(x)]
  if (length(x) == 0) return(NA_real_)
  log_base^mean(log(x) / log(log_base))
}

#' Geometric Mean Fold Rise (GMFR) = GMT(post) / GMT(pre)
#' Equivalently: exp(mean(log(post/pre)))
gmfr <- function(post, pre) {
  ratio <- post / pre
  ratio <- ratio[ratio > 0 & !is.na(ratio)]
  if (length(ratio) == 0) return(NA_real_)
  exp(mean(log(ratio)))
}

#' Derive log2-transformed titre variable (standard for HAI)
derive_var_aval_log2 <- function(dataset, aval_var = AVAL, new_var = AVALOG2) {
  aval_nm <- deparse(substitute(aval_var))
  new_nm  <- deparse(substitute(new_var))
  dataset[[new_nm]] <- log2(pmax(dataset[[aval_nm]], 0.5))
  dataset
}

#' Derive seroconversion flag per subject per antigen
#' RECIST analogue: seroconversion = ≥4-fold rise AND post-vax titre ≥ threshold
#' HAI threshold: 40; ELISA threshold: 4× lower limit of quantitation
derive_param_seroconversion <- function(dataset,
                                         dataset_adsl,
                                         by_vars      = c("USUBJID","PARAMCD"),
                                         baseline_var = BASE,
                                         aval_var     = AVAL,
                                         fold_threshold  = 4,
                                         titre_threshold = NULL,
                                         set_values_to   = list(),
                                         ...) {
  base_nm  <- deparse(substitute(baseline_var))
  aval_nm  <- deparse(substitute(aval_var))
  by_cols  <- if (is.character(by_vars)) by_vars
              else sapply(by_vars, deparse)

  post <- dataset |> filter(VISIT != "DAY 1", !is.na(.data[[base_nm]]))

  post <- post |>
    mutate(
      FOLD_RISE = .data[[aval_nm]] / .data[[base_nm]],
      .SCFL     = FOLD_RISE >= fold_threshold &
                  (is.null(titre_threshold) | .data[[aval_nm]] >= titre_threshold)
    )

  # One row per subject per antigen
  sc_rows <- post |>
    group_by(across(all_of(by_cols))) |>
    summarise(
      AVAL  = as.integer(any(.SCFL, na.rm = TRUE)),
      AVALC = if_else(any(.SCFL, na.rm = TRUE), "Y", "N"),
      .groups = "drop"
    ) |>
    mutate(DTYPE = "SCONV")

  for (nm in names(set_values_to)) sc_rows[[nm]] <- set_values_to[[nm]]

  bind_rows(dataset, sc_rows)
}

#' Derive seroprotection flag
#' Seroprotection = post-vaccination titre ≥ threshold (e.g., HAI ≥ 40)
derive_param_seroprotection <- function(dataset,
                                         dataset_adsl,
                                         by_vars         = c("USUBJID","PARAMCD"),
                                         aval_var        = AVAL,
                                         titre_threshold = 40,
                                         set_values_to   = list(),
                                         ...) {
  aval_nm <- deparse(substitute(aval_var))
  by_cols <- if (is.character(by_vars)) by_vars
             else sapply(by_vars, deparse)

  post <- dataset |> filter(VISIT != "DAY 1")

  sp_rows <- post |>
    group_by(across(all_of(by_cols))) |>
    summarise(
      AVAL  = as.integer(any(.data[[aval_nm]] >= titre_threshold, na.rm = TRUE)),
      AVALC = if_else(any(.data[[aval_nm]] >= titre_threshold, na.rm = TRUE), "Y", "N"),
      .groups = "drop"
    ) |>
    mutate(DTYPE = "SPROT")

  for (nm in names(set_values_to)) sp_rows[[nm]] <- set_values_to[[nm]]

  bind_rows(dataset, sp_rows)
}

#' Derive GMFR parameter rows (one per subject per antigen)
derive_param_gmfr <- function(dataset,
                               by_vars       = c("USUBJID","PARAMCD"),
                               baseline_var  = BASE,
                               aval_var      = AVAL,
                               set_values_to = list(),
                               ...) {
  base_nm <- deparse(substitute(baseline_var))
  aval_nm <- deparse(substitute(aval_var))
  by_cols <- if (is.character(by_vars)) by_vars
             else sapply(by_vars, deparse)

  post <- dataset |> filter(VISIT != "DAY 1", !is.na(.data[[base_nm]]),
                             .data[[base_nm]] > 0)

  gmfr_rows <- post |>
    mutate(FOLD = .data[[aval_nm]] / .data[[base_nm]]) |>
    group_by(across(all_of(by_cols))) |>
    summarise(
      AVAL = round(exp(mean(log(FOLD[FOLD > 0 & !is.na(FOLD)]))), 3),
      .groups = "drop"
    ) |>
    mutate(DTYPE = "GMFR")

  for (nm in names(set_values_to)) gmfr_rows[[nm]] <- set_values_to[[nm]]

  bind_rows(dataset, gmfr_rows)
}

#' Derive 4-fold rise flag (simpler than full seroconversion — no titre floor)
derive_param_fourfold <- function(dataset,
                                   by_vars      = c("USUBJID","PARAMCD"),
                                   baseline_var = BASE,
                                   aval_var     = AVAL,
                                   fold         = 4,
                                   set_values_to = list(),
                                   ...) {
  base_nm <- deparse(substitute(baseline_var))
  aval_nm <- deparse(substitute(aval_var))
  by_cols <- if (is.character(by_vars)) by_vars else sapply(by_vars, deparse)

  post <- dataset |> filter(VISIT != "DAY 1", !is.na(.data[[base_nm]]),
                             .data[[base_nm]] > 0)

  ff_rows <- post |>
    mutate(.FF = (.data[[aval_nm]] / .data[[base_nm]]) >= fold) |>
    group_by(across(all_of(by_cols))) |>
    summarise(
      AVAL  = as.integer(any(.FF, na.rm = TRUE)),
      AVALC = if_else(any(.FF, na.rm = TRUE), "Y", "N"),
      .groups = "drop"
    ) |>
    mutate(DTYPE = "4FOLD")

  for (nm in names(set_values_to)) ff_rows[[nm]] <- set_values_to[[nm]]
  bind_rows(dataset, ff_rows)
}

# ── Shared admiral helpers ────────────────────────────────────────────────────
exprs <- function(...) as.list(substitute(list(...)))[-1]

derive_vars_merged <- function(dataset, dataset_add, by_vars, new_vars = NULL, ...) {
  by_nm <- if (is.character(by_vars)) by_vars else sapply(by_vars, deparse)
  jdf   <- dataset_add
  if (!is.null(new_vars)) {
    old <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d,"=")[[1]][2]) else d
    })
    new <- sapply(new_vars, function(x) {
      d <- deparse(x); if (grepl("=", d)) trimws(strsplit(d,"=")[[1]][1]) else d
    })
    keep <- intersect(c(by_nm, old), names(jdf)); jdf <- jdf |> select(all_of(keep))
    for (i in seq_along(old))
      if (old[i] != new[i] && old[i] %in% names(jdf))
        jdf <- rename(jdf, !!new[i] := !!old[i])
  }
  left_join(dataset, jdf, by = intersect(by_nm, names(jdf)))
}
```

## Overview

### The vaccine ADaM landscape

Vaccine trials differ fundamentally from therapeutic trials. Instead of measuring disease outcomes in sick patients, they measure **immune response** in healthy volunteers and track **reactogenicity** (side effects). This requires specialised ADaM datasets:

| Dataset | Full Name | Key Content |
|---------|-----------|-------------|
| **ADIS** | Immunogenicity Specimens | Antibody titres (HAI, MN, ELISA), seroconversion, seroprotection, GMT |
| **ADFACE** | Solicited Adverse Events | Reactogenicity: injection-site and systemic reactions, graded 1–3 |
| **ADCE** | Clinical Events | Unsolicited AEs, medically attended events (MAEs), disease endpoints |

> **SAS parallel:** In SAS vaccine programs, GMT and GMFR are typically computed in `PROC MEANS` with log-transformed data, then back-transformed. `derive_param_gmfr()` and `gmt()` in admiralvaccine replace this with a consistent, reproducible function interface.

### Assay types in this tutorial

| Assay | Abbrev | Measure | Scale |
|-------|--------|---------|-------|
| Hemagglutination Inhibition | HAI | Influenza neutralising antibodies | Titre (1:dilution) — rounded to series: 10, 20, 40, 80 … |
| Enzyme-Linked Immunosorbent Assay | ELISA | RSV-F protein IgG | IU/mL (continuous) |
| Microneutralisation | MN | Virus neutralisation | Titre (1:dilution) |

```{r overview-explore, exercise=FALSE}
# Pre-loaded datasets:
#   adsl_vac  — 120 subjects, arms VAC / PBO
#   adis_base — IS-derived ADIS skeleton (AVAL = raw titre)
#   face_rows — FACE-derived reactogenicity events

adis_base |>
  group_by(PARAMCD, VISIT) |>
  summarise(
    n       = n(),
    GMT     = round(gmt(AVAL), 1),
    min_val = min(AVAL),
    max_val = max(AVAL),
    .groups = "drop"
  )
```

---

## Section 1 — Immunogenicity Concepts and Titre Arithmetic

### HAI titres and the log2 scale

HAI titres are reported as reciprocal dilutions (10, 20, 40 …). The conventional analysis scale is **log₂** (so titre 40 → log₂ 40 ≈ 5.32). This linearises the multiplicative fold-rise.

```{r quiz-concepts1, echo=FALSE}
quiz(
  caption = "Quiz — Immunogenicity Concepts",

  question("A subject's Day 1 HAI titre is 20 and Day 29 titre is 160. Their fold rise is:",
    answer("140"),
    answer("8", correct = TRUE,
           message = "Fold rise = 160 / 20 = 8. On the log₂ scale this is log₂(8) = 3 doublings."),
    answer("3"),
    answer("4"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("Seroconversion for HAI is typically defined as:",
    answer("Any increase in titre from baseline"),
    answer("≥4-fold rise in titre AND post-vaccination titre ≥ 40 (1:40)", correct = TRUE,
           message = "Both conditions must be met: the ratio and the absolute titre floor. This prevents a subject with baseline 5 and post 25 (5-fold but still sub-protective) from being counted as seroconverted."),
    answer("≥2-fold rise in titre"),
    answer("Post-vaccination titre ≥ 40 only"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise 1.0 — Compute GMT from first principles

The GMT formula is `exp(mean(log(x)))`. Using `adis_base`, compute the GMT for H1N1 at each visit by arm — without using the pre-built `gmt()` helper.

```{r ex-gmt-calc, exercise=TRUE, exercise.lines=12}
# adis_base has HAI titres for H1N1 at DAY 1 and DAY 29.
# Compute the Geometric Mean Titre (GMT) for H1N1 by visit and arm.
# GMT = exp(mean(log(x))). Use filter(PARAMCD == "H1N1").

adis_base |>
  filter(PARAMCD == ___, !is.na(AVAL)) |>
  group_by(ARMCD, VISIT) |>
  summarise(
    GMT = round(exp(mean(log(___))), 2),
    n   = n(),
    .groups = "drop"
  )
```

```{r ex-gmt-calc-hint-1}
# PARAMCD == "H1N1"
# GMT = exp(mean(log(AVAL)))
```

```{r ex-gmt-calc-solution}
adis_base |>
  filter(PARAMCD == "H1N1", !is.na(AVAL)) |>
  group_by(ARMCD, VISIT) |>
  summarise(
    GMT = round(exp(mean(log(AVAL))), 2),
    n   = n(),
    .groups = "drop"
  )
```

### Exercise 1.1 — Compute fold rise per subject

Using `adis_base`, compute the fold rise (Day 29 / Day 1) for **H1N1** only. Then find the proportion of subjects in each arm who achieved ≥4-fold rise.

```{r ex-fold-rise, exercise=TRUE, exercise.lines=22}
# adis_base has USUBJID, PARAMCD, VISIT, AVAL, BASE, ARMCD
fold_rise_h1n1 <- adis_base |>
  filter(PARAMCD == ___, VISIT == "DAY 29") |>
  mutate(
    FOLD_RISE = ___ / ___,              # post / baseline
    FOURFOLD  = FOLD_RISE >= ___        # ≥4-fold flag
  ) |>
  group_by(___) |>
  summarise(
    n           = n(),
    n_fourfold  = sum(FOURFOLD, na.rm = TRUE),
    pct_fourfold = round(100 * n_fourfold / n, 1),
    .groups     = "drop"
  )

fold_rise_h1n1
```

```{r ex-fold-rise-hint-1}
# filter PARAMCD == "H1N1", VISIT == "DAY 29"
# FOLD_RISE = AVAL / BASE; FOURFOLD = FOLD_RISE >= 4
# group by ARMCD
adis_base |>
  filter(PARAMCD == "H1N1", VISIT == "DAY 29") |>
  mutate(FOLD_RISE = AVAL / BASE, FOURFOLD = FOLD_RISE >= 4) |>
  group_by(ARMCD) |>
  summarise(n = n(), n_fourfold = sum(FOURFOLD, na.rm = TRUE),
            pct_fourfold = round(100 * n_fourfold / n, 1), .groups = "drop")
```

```{r ex-fold-rise-solution}
fold_rise_h1n1 <- adis_base |>
  filter(PARAMCD == "H1N1", VISIT == "DAY 29") |>
  mutate(
    FOLD_RISE = AVAL / BASE,
    FOURFOLD  = FOLD_RISE >= 4
  ) |>
  group_by(ARMCD) |>
  summarise(
    n            = n(),
    n_fourfold   = sum(FOURFOLD, na.rm = TRUE),
    pct_fourfold = round(100 * n_fourfold / n, 1),
    .groups      = "drop"
  )
fold_rise_h1n1
```

### Exercise 1.2 — Log₂ transform and compare distributions

Derive the log₂ titre (`AVALOG2`) using `derive_var_aval_log2()`, then compare the mean log₂ titre at Day 29 by arm and antigen.

```{r ex-log2, exercise=TRUE, exercise.lines=20}
# Apply log2 transform to ALL rows in adis_base
adis_log2 <- derive_var_aval_log2(
  dataset  = ___,
  aval_var = AVAL,
  new_var  = AVALOG2
)

# Mean log2 titre at Day 29 by arm and antigen
adis_log2 |>
  filter(VISIT == ___) |>
  group_by(ARMCD, ___) |>
  summarise(
    mean_log2  = round(mean(AVALOG2, na.rm = TRUE), 2),
    sd_log2    = round(sd(AVALOG2,   na.rm = TRUE), 2),
    .groups    = "drop"
  ) |>
  arrange(PARAMCD, ARMCD)
```

```{r ex-log2-hint-1}
# dataset = adis_base; aval_var = AVAL; new_var = AVALOG2
# filter VISIT == "DAY 29"; group by ARMCD and PARAMCD
adis_log2 <- derive_var_aval_log2(dataset = adis_base, aval_var = AVAL, new_var = AVALOG2)
adis_log2 |>
  filter(VISIT == "DAY 29") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(mean_log2 = round(mean(AVALOG2, na.rm = TRUE), 2),
            sd_log2 = round(sd(AVALOG2, na.rm = TRUE), 2), .groups = "drop") |>
  arrange(PARAMCD, ARMCD)
```

```{r ex-log2-solution}
adis_log2 <- derive_var_aval_log2(
  dataset  = adis_base,
  aval_var = AVAL,
  new_var  = AVALOG2
)

adis_log2 |>
  filter(VISIT == "DAY 29") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(
    mean_log2 = round(mean(AVALOG2, na.rm = TRUE), 2),
    sd_log2   = round(sd(AVALOG2,   na.rm = TRUE), 2),
    .groups   = "drop"
  ) |>
  arrange(PARAMCD, ARMCD)
```

---

## Section 2 — GMT and GMFR

### The GMT / GMFR framework

Every immunogenicity summary table in a vaccine trial has at minimum:

| Statistic | Formula | ADIS DTYPE |
|-----------|---------|------------|
| GMT (per visit) | exp(mean(log(AVAL))) | *inline* — not a separate DTYPE |
| GMFR | exp(mean(log(Day 29 / Day 1))) | `GMFR` |
| Seroconversion rate | % with ≥4-fold rise AND titre ≥40 | `SCONV` |
| Seroprotection rate | % with titre ≥40 post-vax | `SPROT` |

```{r quiz-gmt1, echo=FALSE}
quiz(
  caption = "Quiz — GMT / GMFR",

  question("GMFR = 6.2 means:",
    answer("The average titre increased by 6.2 IU/mL"),
    answer("On average, titres increased 6.2-fold from baseline", correct = TRUE),
    answer("62% of subjects responded"),
    answer("The GMT at Day 29 is 6.2 times the lower limit of normal"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("Why is GMFR calculated on the log scale rather than directly as the ratio of GMTs?",
    answer("They are mathematically identical — it does not matter", correct = FALSE,
           message = "Actually they ARE mathematically identical: GMT(post)/GMT(pre) = exp(mean(log(post))) / exp(mean(log(pre))) = exp(mean(log(post/pre))). The log-scale approach avoids paired-ratio outliers in some edge cases."),
    answer("They are mathematically equivalent — exp(mean(log(post/pre))) = GMT(post)/GMT(pre)",
           correct = TRUE),
    answer("The ratio of GMTs is biased upward"),
    answer("Regulatory guidance mandates the log-scale formula"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise 2.0 — Compute GMFR manually

Compute GMFR for H1N1 without using `derive_param_gmfr()`. GMFR = GMT(Day 29) / GMT(Day 1). Use `pivot_wider()` to get one row per arm with Day 1 and Day 29 GMTs side by side, then divide.

```{r ex-gmfr-manual, exercise=TRUE, exercise.lines=14}
# Compute GMFR manually for H1N1: GMFR = GMT(post) / GMT(pre).
# Use DAY 29 as post, DAY 1 as pre.

gmfr_manual <- adis_base |>
  filter(PARAMCD == "H1N1", VISIT %in% c("DAY 1", "DAY 29")) |>
  group_by(ARMCD, VISIT) |>
  summarise(GMT = exp(mean(log(AVAL))), .groups = "drop") |>
  pivot_wider(names_from = VISIT, values_from = GMT,
              names_repair = "universal") |>
  mutate(GMFR = round(___ / ___, 2))  # DAY.29 / DAY.1

gmfr_manual
```

```{r ex-gmfr-manual-hint-1}
# After pivot_wider, column names become "DAY.1" and "DAY.29"
# GMFR = DAY.29 / DAY.1
```

```{r ex-gmfr-manual-solution}
gmfr_manual <- adis_base |>
  filter(PARAMCD == "H1N1", VISIT %in% c("DAY 1", "DAY 29")) |>
  group_by(ARMCD, VISIT) |>
  summarise(GMT = exp(mean(log(AVAL))), .groups = "drop") |>
  pivot_wider(names_from = VISIT, values_from = GMT,
              names_repair = "universal") |>
  mutate(GMFR = round(DAY.29 / DAY.1, 2))

gmfr_manual
```

### Exercise 2.1 — Compute GMT by arm and antigen

Using `adis_base` and the `gmt()` helper, compute GMT at **each visit** by arm and antigen.

```{r ex-gmt, exercise=TRUE, exercise.lines=18}
# gmt(x) = exp(mean(log(x))) for x > 0
# adis_base has AVAL, VISIT, PARAMCD, ARMCD

gmt_table <- adis_base |>
  group_by(___, ___, ___) |>
  summarise(
    n   = n(),
    GMT = round(gmt(___), 2),
    .groups = "drop"
  ) |>
  arrange(PARAMCD, ARMCD, VISIT)

gmt_table
```

```{r ex-gmt-hint-1}
# group by PARAMCD, ARMCD, VISIT; summarise with gmt(AVAL)
adis_base |>
  group_by(PARAMCD, ARMCD, VISIT) |>
  summarise(n = n(), GMT = round(gmt(AVAL), 2), .groups = "drop") |>
  arrange(PARAMCD, ARMCD, VISIT)
```

```{r ex-gmt-solution}
gmt_table <- adis_base |>
  group_by(PARAMCD, ARMCD, VISIT) |>
  summarise(
    n   = n(),
    GMT = round(gmt(AVAL), 2),
    .groups = "drop"
  ) |>
  arrange(PARAMCD, ARMCD, VISIT)

gmt_table
```

### Exercise 2.2 — Derive GMFR parameter rows

Use `derive_param_gmfr()` to add per-subject GMFR rows (DTYPE = `"GMFR"`) for all three antigens, then summarise GMFR by arm.

```{r ex-gmfr, exercise=TRUE, exercise.lines=22}
adis_with_gmfr <- derive_param_gmfr(
  dataset       = ___,
  by_vars       = c("USUBJID","PARAMCD"),
  baseline_var  = BASE,
  aval_var      = AVAL,
  set_values_to = list(
    PARAM   = "Geometric Mean Fold Rise",
    PARAMCD = "GMFR"
  )
)

# Summarise: geometric mean of individual GMFRs by arm and antigen
adis_with_gmfr |>
  filter(DTYPE == "GMFR") |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(___, PARAMCD) |>
  summarise(
    n      = n(),
    geo_mean_GMFR = round(gmt(___), 2),   # geometric mean of per-subject GMFRs
    .groups = "drop"
  ) |>
  arrange(PARAMCD, ARMCD)
```

```{r ex-gmfr-hint-1}
# dataset = adis_base; group by ARMCD and PARAMCD; gmt(AVAL) where DTYPE == "GMFR"
adis_with_gmfr <- derive_param_gmfr(
  dataset = adis_base,
  by_vars = c("USUBJID","PARAMCD"),
  baseline_var = BASE, aval_var = AVAL,
  set_values_to = list(PARAM = "Geometric Mean Fold Rise", PARAMCD = "GMFR")
)
adis_with_gmfr |>
  filter(DTYPE == "GMFR") |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(n = n(), geo_mean_GMFR = round(gmt(AVAL), 2), .groups = "drop") |>
  arrange(PARAMCD, ARMCD)
```

```{r ex-gmfr-solution}
adis_with_gmfr <- derive_param_gmfr(
  dataset       = adis_base,
  by_vars       = c("USUBJID","PARAMCD"),
  baseline_var  = BASE,
  aval_var      = AVAL,
  set_values_to = list(
    PARAM   = "Geometric Mean Fold Rise",
    PARAMCD = "GMFR"
  )
)

adis_with_gmfr |>
  filter(DTYPE == "GMFR") |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(
    n             = n(),
    geo_mean_GMFR = round(gmt(AVAL), 2),
    .groups       = "drop"
  ) |>
  arrange(PARAMCD, ARMCD)
```

---

## Section 3 — Seroconversion and Seroprotection

### Endpoint definitions at a glance

| Endpoint | Definition | HAI threshold | RSV-F ELISA threshold |
|----------|-----------|---------------|-----------------------|
| Seroconversion | ≥4-fold rise AND titre ≥ floor | Titre ≥ 40 | Varies by assay LLOQ |
| Seroprotection | Titre ≥ protective threshold | ≥ 40 (1:40) | Not applicable |
| 4-fold rise | ≥4-fold rise (no absolute floor) | — | Common for ELISA |

> **Gotcha:** Seroconversion ≠ 4-fold rise. A subject with baseline 5 rising to 25 (5-fold) achieves the fold criterion but NOT the titre floor (25 < 40), so they are **not** seroconverted. This is the most common mistake in vaccine immunogenicity programming.

```{r quiz-seroconv1, echo=FALSE}
quiz(
  caption = "Quiz — Seroconversion vs Seroprotection",

  question("Subject A: baseline HAI 10, Day 29 HAI 40. Subject B: baseline 20, Day 29 40. Who is seroconverted?",
    answer("Both — both reach titre ≥ 40"),
    answer("Subject A only — 4-fold rise (10→40) AND titre ≥ 40", correct = TRUE,
           message = "Subject A: 40/10 = 4× ✓ AND 40 ≥ 40 ✓. Subject B: 40/20 = 2× ✗ — does not meet the ≥4-fold criterion."),
    answer("Subject B only — higher absolute titre"),
    answer("Neither — the titre must exceed 40, not equal it"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("Subject C has baseline HAI 10, Day 29 HAI 40. Are they seroprotected?",
    answer("No — they must have been seroprotected at baseline first"),
    answer("Yes — post-vaccination titre 40 ≥ 40", correct = TRUE,
           message = "Seroprotection only checks the post-vaccination titre, not the fold rise. 40 ≥ 40 → seroprotected."),
    answer("Yes — but only if they are also seroconverted"),
    answer("No — the threshold is > 40, not ≥ 40"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise 3.0 — Manually compute seroconversion flag

Using `adis_base`, manually compute seroconversion for H1N1 without using `derive_param_seroconversion()`. Pivot to wide format so each subject has one row with DAY 1 (pre) and DAY 29 (post), then apply the two-condition rule: fold rise ≥ 4 AND post ≥ 40.

```{r ex-seroconv-flag, exercise=TRUE, exercise.lines=18}
# Seroconversion = post/pre >= 4 AND post >= 40.
# Use VISIT == "DAY 1" as pre, and VISIT == "DAY 29" as post.

seroconv_manual <- adis_base |>
  filter(PARAMCD == "H1N1", VISIT %in% c("DAY 1", "DAY 29")) |>
  select(USUBJID, VISIT, AVAL, ARMCD) |>
  pivot_wider(names_from = VISIT, values_from = AVAL,
              names_repair = "universal") |>
  rename(PRE = DAY.1, POST = DAY.29) |>
  mutate(
    FOLD_RISE   = POST / PRE,
    SEROCONV_FL = ___ >= 4 & ___ >= 40   # fold rise >= 4 AND post >= 40
  )

seroconv_manual |>
  group_by(ARMCD) |>
  summarise(n = n(), n_sc = sum(SEROCONV_FL, na.rm = TRUE),
            pct = round(100 * mean(SEROCONV_FL, na.rm = TRUE), 1),
            .groups = "drop")
```

```{r ex-seroconv-flag-hint-1}
# FOLD_RISE >= 4 AND POST >= 40
# SEROCONV_FL = FOLD_RISE >= 4 & POST >= 40
```

```{r ex-seroconv-flag-solution}
seroconv_manual <- adis_base |>
  filter(PARAMCD == "H1N1", VISIT %in% c("DAY 1", "DAY 29")) |>
  select(USUBJID, VISIT, AVAL, ARMCD) |>
  pivot_wider(names_from = VISIT, values_from = AVAL,
              names_repair = "universal") |>
  rename(PRE = DAY.1, POST = DAY.29) |>
  mutate(
    FOLD_RISE   = POST / PRE,
    SEROCONV_FL = FOLD_RISE >= 4 & POST >= 40
  )

seroconv_manual |>
  group_by(ARMCD) |>
  summarise(n = n(), n_sc = sum(SEROCONV_FL, na.rm = TRUE),
            pct = round(100 * mean(SEROCONV_FL, na.rm = TRUE), 1),
            .groups = "drop")
```

### Exercise 3.1 — Derive seroconversion (H1N1 and H3N2 only)

Apply `derive_param_seroconversion()` to HAI titres (H1N1, H3N2). Use `fold_threshold = 4` and `titre_threshold = 40`.

```{r ex-seroconv, exercise=TRUE, exercise.lines=22}
# Filter to HAI antigens only, then derive seroconversion
adis_hai <- adis_base |> filter(PARAMCD %in% c("H1N1","H3N2"))

adis_sc <- derive_param_seroconversion(
  dataset          = ___,
  dataset_adsl     = adsl_vac,
  by_vars          = c("USUBJID","PARAMCD"),
  baseline_var     = BASE,
  aval_var         = AVAL,
  fold_threshold   = ___,
  titre_threshold  = ___,
  set_values_to    = list(PARAM = "Seroconversion", PARAMCD = "SCONV_HAI")
)

# Seroconversion rate by arm and antigen
adis_sc |>
  filter(DTYPE == "SCONV") |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(n = n(), n_sc = sum(AVALC == "Y"),
            pct = round(100 * n_sc / n, 1), .groups = "drop")
```

```{r ex-seroconv-hint-1}
# dataset = adis_hai; fold_threshold = 4; titre_threshold = 40
adis_sc <- derive_param_seroconversion(
  dataset = adis_hai, dataset_adsl = adsl_vac,
  by_vars = c("USUBJID","PARAMCD"), baseline_var = BASE, aval_var = AVAL,
  fold_threshold = 4, titre_threshold = 40,
  set_values_to = list(PARAM = "Seroconversion", PARAMCD = "SCONV_HAI")
)
adis_sc |>
  filter(DTYPE == "SCONV") |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(n = n(), n_sc = sum(AVALC == "Y"), pct = round(100 * n_sc / n, 1), .groups = "drop")
```

```{r ex-seroconv-solution}
adis_hai <- adis_base |> filter(PARAMCD %in% c("H1N1","H3N2"))

adis_sc <- derive_param_seroconversion(
  dataset         = adis_hai,
  dataset_adsl    = adsl_vac,
  by_vars         = c("USUBJID","PARAMCD"),
  baseline_var    = BASE,
  aval_var        = AVAL,
  fold_threshold  = 4,
  titre_threshold = 40,
  set_values_to   = list(PARAM = "Seroconversion", PARAMCD = "SCONV_HAI")
)

adis_sc |>
  filter(DTYPE == "SCONV") |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(n    = n(), n_sc = sum(AVALC == "Y"),
            pct  = round(100 * n_sc / n, 1),
            .groups = "drop")
```

### Exercise 3.2 — Derive seroprotection

Derive seroprotection for **H1N1** using titre threshold = 40, then compare the pre-vaccination and post-vaccination seroprotection rates.

```{r ex-seroprot, exercise=TRUE, exercise.lines=28}
# For seroprotection we need to run it separately on Day 1 and Day 29 rows
sp_day29 <- derive_param_seroprotection(
  dataset         = adis_base |> filter(PARAMCD == "H1N1", VISIT == "DAY 29"),
  dataset_adsl    = adsl_vac,
  by_vars         = c("USUBJID","PARAMCD"),
  aval_var        = AVAL,
  titre_threshold = ___,
  set_values_to   = list(PARAM = "Seroprotection D29", PARAMCD = "SPROT_D29")
)

sp_day1 <- derive_param_seroprotection(
  dataset         = adis_base |> filter(PARAMCD == "H1N1", VISIT == "DAY 1"),
  dataset_adsl    = adsl_vac,
  by_vars         = c("USUBJID","PARAMCD"),
  aval_var        = AVAL,
  titre_threshold = ___,
  set_values_to   = list(PARAM = "Seroprotection D1", PARAMCD = "SPROT_D1")
)

# Compare rates by arm
bind_rows(
  sp_day29 |> filter(DTYPE == "SPROT") |> mutate(Timepoint = "Day 29"),
  sp_day1  |> filter(DTYPE == "SPROT") |> mutate(Timepoint = "Day 1")
) |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(Timepoint, ARMCD) |>
  summarise(n = n(), n_sp = sum(AVALC == "Y"),
            pct = round(100 * n_sp / n, 1), .groups = "drop") |>
  arrange(Timepoint, ARMCD)
```

```{r ex-seroprot-hint-1}
# titre_threshold = 40 for both Day 29 and Day 1
# The mock function treats ALL rows as "post-baseline" to evaluate — pass the already-filtered data
```

```{r ex-seroprot-solution}
sp_day29 <- derive_param_seroprotection(
  dataset         = adis_base |> filter(PARAMCD == "H1N1", VISIT == "DAY 29"),
  dataset_adsl    = adsl_vac,
  by_vars         = c("USUBJID","PARAMCD"),
  aval_var        = AVAL,
  titre_threshold = 40,
  set_values_to   = list(PARAM = "Seroprotection D29", PARAMCD = "SPROT_D29")
)

sp_day1 <- derive_param_seroprotection(
  dataset         = adis_base |> filter(PARAMCD == "H1N1", VISIT == "DAY 1"),
  dataset_adsl    = adsl_vac,
  by_vars         = c("USUBJID","PARAMCD"),
  aval_var        = AVAL,
  titre_threshold = 40,
  set_values_to   = list(PARAM = "Seroprotection D1", PARAMCD = "SPROT_D1")
)

bind_rows(
  sp_day29 |> filter(DTYPE == "SPROT") |> mutate(Timepoint = "Day 29"),
  sp_day1  |> filter(DTYPE == "SPROT") |> mutate(Timepoint = "Day 1")
) |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(Timepoint, ARMCD) |>
  summarise(n = n(), n_sp = sum(AVALC == "Y"),
            pct = round(100 * n_sp / n, 1),
            .groups = "drop") |>
  arrange(Timepoint, ARMCD)
```

### Exercise 3.3 — RSV: four-fold rise (no titre floor)

ELISA titres (RSV-F) use a **four-fold rise** without a titre floor (the floor varies by assay and LLOQ). Derive the 4-fold rise flag for RSV-F and compute the response rate by arm.

```{r ex-fourfold-rsv, exercise=TRUE, exercise.lines=22}
adis_rsv_ff <- derive_param_fourfold(
  dataset       = adis_base |> filter(PARAMCD == ___),
  by_vars       = c("USUBJID","PARAMCD"),
  baseline_var  = BASE,
  aval_var      = AVAL,
  fold          = ___,
  set_values_to = list(PARAMCD = "FF_RSV", PARAM = "RSV-F 4-Fold Rise")
)

adis_rsv_ff |>
  filter(DTYPE == "4FOLD") |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(ARMCD) |>
  summarise(
    n       = n(),
    n_resp  = sum(AVALC == "Y"),
    pct     = round(100 * n_resp / n, 1),
    .groups = "drop"
  )
```

```{r ex-fourfold-rsv-hint-1}
# PARAMCD == "RSV-F"; fold = 4
adis_rsv_ff <- derive_param_fourfold(
  dataset = adis_base |> filter(PARAMCD == "RSV-F"),
  by_vars = c("USUBJID","PARAMCD"), baseline_var = BASE, aval_var = AVAL, fold = 4,
  set_values_to = list(PARAMCD = "FF_RSV", PARAM = "RSV-F 4-Fold Rise")
)
adis_rsv_ff |>
  filter(DTYPE == "4FOLD") |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(ARMCD) |>
  summarise(n = n(), n_resp = sum(AVALC == "Y"),
            pct = round(100 * n_resp / n, 1), .groups = "drop")
```

```{r ex-fourfold-rsv-solution}
adis_rsv_ff <- derive_param_fourfold(
  dataset       = adis_base |> filter(PARAMCD == "RSV-F"),
  by_vars       = c("USUBJID","PARAMCD"),
  baseline_var  = BASE,
  aval_var      = AVAL,
  fold          = 4,
  set_values_to = list(PARAMCD = "FF_RSV", PARAM = "RSV-F 4-Fold Rise")
)

adis_rsv_ff |>
  filter(DTYPE == "4FOLD") |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  group_by(ARMCD) |>
  summarise(
    n       = n(),
    n_resp  = sum(AVALC == "Y"),
    pct     = round(100 * n_resp / n, 1),
    .groups = "drop"
  )
```

---

## Section 4 — ADFACE: Solicited Adverse Events (Reactogenicity)

### What is ADFACE?

ADFACE (Solicited Adverse Events) captures **pre-specified reactions** collected via electronic diary or paper diary within a defined post-vaccination surveillance window (typically 7 days). It is distinct from spontaneous AEs (captured in ADAE).

Key ADFACE variables:

| Variable | Meaning |
|----------|---------|
| `FACECAT` | LOCAL (injection site) vs SYSTEMIC |
| `FACETESTCD` | Symptom code: PAIN, REDNESS, SWELLING, FATIGUE, HEADACHE, MYALGIA, FEVER |
| `FACESEVN` | Grade 1 (mild) / 2 (moderate) / 3 (severe) |
| `ADURN` | Duration in days |

```{r quiz-adface1, echo=FALSE}
quiz(
  caption = "Quiz — ADFACE Reactogenicity",

  question("Injection site pain and redness are classified as:",
    answer("Systemic reactions — they affect the whole body"),
    answer("Local reactions — they occur at the injection site", correct = TRUE),
    answer("Unsolicited adverse events — they are not pre-specified"),
    answer("Grade 3 reactions by default"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("The 7-day post-vaccination reactogenicity surveillance period starts:",
    answer("On Day 1 (day of vaccination)", correct = TRUE,
           message = "Day 1 = day of vaccination; the diary covers Days 1–7 (inclusive). Some protocols use Day 0 = vax day and Days 1–7."),
    answer("On Day 2 — the day after vaccination"),
    answer("On Day 8 — after the acute period"),
    answer("Whenever the subject first reports a symptom"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise 4.0 — Worst symptom grade per subject

`face_rows` contains reactogenicity symptoms with `FACESEVN` (1 = mild, 2 = moderate, 3 = severe). For each subject, find their worst (maximum) symptom grade across all symptoms. Then report the proportion with any Grade 3 (severe) event by arm.

```{r ex-react-worst, exercise=TRUE, exercise.lines=14}
# face_rows has USUBJID, ARMCD, FACESEVN
# Step 1: worst grade per subject and arm
worst_grade_subj <- face_rows |>
  group_by(USUBJID, ARMCD) |>
  summarise(MAX_GRADE = max(___, na.rm = TRUE), .groups = "drop")

# Step 2: proportion with Grade 3 by arm
worst_grade_subj |>
  group_by(ARMCD) |>
  summarise(
    n        = n(),
    n_grade3 = sum(MAX_GRADE == ___),
    pct      = round(100 * n_grade3 / n, 1),
    .groups  = "drop"
  )
```

```{r ex-react-worst-hint-1}
# MAX_GRADE = max(FACESEVN, na.rm = TRUE)
# n_grade3 = sum(MAX_GRADE == 3)
```

```{r ex-react-worst-solution}
worst_grade_subj <- face_rows |>
  group_by(USUBJID, ARMCD) |>
  summarise(MAX_GRADE = max(FACESEVN, na.rm = TRUE), .groups = "drop")

worst_grade_subj |>
  group_by(ARMCD) |>
  summarise(
    n        = n(),
    n_grade3 = sum(MAX_GRADE == 3),
    pct      = round(100 * n_grade3 / n, 1),
    .groups  = "drop"
  )
```

### Exercise 4.1 — Reactogenicity incidence by grade

Using `face_rows`, compute the incidence (% of subjects with at least one occurrence) of **any Grade ≥ 2** local and systemic reaction, by arm.

```{r ex-reacto-grade, exercise=TRUE, exercise.lines=24}
# face_rows has USUBJID, FACECAT, FACETESTCD, FACESEVN, ARMCD
# Count subjects with any Grade >= 2 local OR systemic reaction

reacto_summary <- face_rows |>
  filter(FACESEVN >= ___) |>
  group_by(USUBJID, FACECAT, ARMCD) |>
  summarise(had_reaction = TRUE, .groups = "drop") |>
  # Join back to adsl to get denominator (all randomised subjects)
  right_join(
    adsl_vac |> select(USUBJID, ARMCD),
    by = c("USUBJID", "ARMCD")
  ) |>
  mutate(had_reaction = replace_na(had_reaction, FALSE)) |>
  group_by(___, ___) |>
  summarise(
    n_total    = n(),
    n_reaction = sum(had_reaction),
    pct        = round(100 * n_reaction / n_total, 1),
    .groups    = "drop"
  ) |>
  filter(!is.na(___))   # remove rows where FACECAT is NA (non-reactors)

reacto_summary
```

```{r ex-reacto-grade-hint-1}
# filter FACESEVN >= 2; group by USUBJID, FACECAT, ARMCD; right_join adsl
# then group by FACECAT and ARMCD
face_rows |>
  filter(FACESEVN >= 2) |>
  group_by(USUBJID, FACECAT, ARMCD) |>
  summarise(had_reaction = TRUE, .groups = "drop") |>
  right_join(adsl_vac |> select(USUBJID, ARMCD), by = c("USUBJID","ARMCD")) |>
  mutate(had_reaction = replace_na(had_reaction, FALSE)) |>
  group_by(FACECAT, ARMCD) |>
  summarise(n_total = n(), n_reaction = sum(had_reaction),
            pct = round(100 * n_reaction / n_total, 1), .groups = "drop") |>
  filter(!is.na(FACECAT))
```

```{r ex-reacto-grade-solution}
reacto_summary <- face_rows |>
  filter(FACESEVN >= 2) |>
  group_by(USUBJID, FACECAT, ARMCD) |>
  summarise(had_reaction = TRUE, .groups = "drop") |>
  right_join(
    adsl_vac |> select(USUBJID, ARMCD),
    by = c("USUBJID", "ARMCD")
  ) |>
  mutate(had_reaction = replace_na(had_reaction, FALSE)) |>
  group_by(FACECAT, ARMCD) |>
  summarise(
    n_total    = n(),
    n_reaction = sum(had_reaction),
    pct        = round(100 * n_reaction / n_total, 1),
    .groups    = "drop"
  ) |>
  filter(!is.na(FACECAT))

reacto_summary
```

### Exercise 4.2 — Mean duration by symptom

Compute the mean duration (in days) for each solicited symptom, limited to subjects who reported it at any grade, by arm.

```{r ex-duration, exercise=TRUE, exercise.lines=18}
face_rows |>
  group_by(___, FACETESTCD, FACETEST, FACECAT) |>
  summarise(
    n_subjects   = n_distinct(___),
    mean_dur     = round(mean(ADURN, na.rm = TRUE), 1),
    pct_grade3   = round(100 * mean(FACESEVN == 3, na.rm = TRUE), 1),
    .groups      = "drop"
  ) |>
  arrange(FACECAT, ARMCD, FACETESTCD)
```

```{r ex-duration-hint-1}
# group by ARMCD, FACETESTCD, FACETEST, FACECAT
# n_distinct(USUBJID); mean(ADURN); mean(FACESEVN == 3) for % Grade 3
face_rows |>
  group_by(ARMCD, FACETESTCD, FACETEST, FACECAT) |>
  summarise(n_subjects = n_distinct(USUBJID),
            mean_dur = round(mean(ADURN, na.rm = TRUE), 1),
            pct_grade3 = round(100 * mean(FACESEVN == 3, na.rm = TRUE), 1),
            .groups = "drop") |>
  arrange(FACECAT, ARMCD, FACETESTCD)
```

```{r ex-duration-solution}
face_rows |>
  group_by(ARMCD, FACETESTCD, FACETEST, FACECAT) |>
  summarise(
    n_subjects = n_distinct(USUBJID),
    mean_dur   = round(mean(ADURN, na.rm = TRUE), 1),
    pct_grade3 = round(100 * mean(FACESEVN == 3, na.rm = TRUE), 1),
    .groups    = "drop"
  ) |>
  arrange(FACECAT, ARMCD, FACETESTCD)
```

---

## Section 5 — Age Subgroup Analyses

### Why age subgroups matter in vaccine trials

Vaccine immunogenicity varies substantially by age: elderly subjects (≥65) often mount weaker responses due to immunosenescence. Regulators require analyses in pre-specified age strata.

```{r quiz-age1, echo=FALSE}
quiz(
  caption = "Quiz — Subgroup Analysis",

  question("In a vaccine trial, the elderly subgroup (≥65) typically shows:",
    answer("Higher seroconversion rates due to prior exposure"),
    answer("Lower GMT and seroconversion rates due to immunosenescence",
           correct = TRUE,
           message = "Immunosenescence — age-related immune decline — leads to weaker vaccine responses in older adults. This is why some vaccines require higher doses or adjuvants for this group."),
    answer("No difference compared to younger adults"),
    answer("Higher reactogenicity"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("When computing GMFR within age subgroups, which approach is statistically correct?",
    answer("Compute GMFR separately within each subgroup", correct = TRUE,
           message = "Subgroup GMFRs should be computed within-group, not post-hoc filtered from an overall model."),
    answer("Compute overall GMFR then split by subgroup"),
    answer("Average the subgroup GMFRs to get an overall GMFR"),
    answer("Use the same GMFR for all subgroups — it does not vary"),
    allow_retry = TRUE
  )
)
```

### Exercise 5.1 — GMT and seroprotection by age group

Using `adis_base` and `adsl_vac` (which has `AGEGR1`), compute GMT at Day 29 and seroprotection rate for H1N1, by arm and age group.

```{r ex-age-subgroup, exercise=TRUE, exercise.lines=28}
# Step 1: merge AGEGR1 from adsl_vac into adis_base
adis_age <- adis_base |>
  left_join(select(adsl_vac, USUBJID, AGEGR1), by = "USUBJID")

# Step 2: GMT at Day 29 by arm, age group, antigen (H1N1 only)
gmt_by_age <- adis_age |>
  filter(PARAMCD == "H1N1", VISIT == ___) |>
  group_by(ARMCD, ___) |>
  summarise(
    n   = n(),
    GMT = round(gmt(AVAL), 1),
    .groups = "drop"
  )

# Step 3: Seroprotection rate by arm and age group
sp_by_age <- adis_age |>
  filter(PARAMCD == "H1N1", VISIT == "DAY 29") |>
  group_by(ARMCD, ___) |>
  summarise(
    n        = n(),
    n_sprot  = sum(AVAL >= 40, na.rm = TRUE),
    pct_sp   = round(100 * n_sprot / n, 1),
    .groups  = "drop"
  )

# Combine
left_join(gmt_by_age, sp_by_age, by = c("ARMCD","AGEGR1")) |>
  arrange(ARMCD, AGEGR1)
```

```{r ex-age-subgroup-hint-1}
# filter VISIT == "DAY 29"; group by ARMCD and AGEGR1
# GMT: gmt(AVAL); seroprotection: sum(AVAL >= 40)
adis_age <- adis_base |> left_join(select(adsl_vac, USUBJID, AGEGR1), by = "USUBJID")
gmt_by_age <- adis_age |>
  filter(PARAMCD == "H1N1", VISIT == "DAY 29") |>
  group_by(ARMCD, AGEGR1) |>
  summarise(n = n(), GMT = round(gmt(AVAL), 1), .groups = "drop")
sp_by_age <- adis_age |>
  filter(PARAMCD == "H1N1", VISIT == "DAY 29") |>
  group_by(ARMCD, AGEGR1) |>
  summarise(n = n(), n_sprot = sum(AVAL >= 40, na.rm = TRUE),
            pct_sp = round(100 * n_sprot / n, 1), .groups = "drop")
left_join(gmt_by_age, sp_by_age, by = c("ARMCD","AGEGR1")) |> arrange(ARMCD, AGEGR1)
```

```{r ex-age-subgroup-solution}
adis_age <- adis_base |>
  left_join(select(adsl_vac, USUBJID, AGEGR1), by = "USUBJID")

gmt_by_age <- adis_age |>
  filter(PARAMCD == "H1N1", VISIT == "DAY 29") |>
  group_by(ARMCD, AGEGR1) |>
  summarise(n = n(), GMT = round(gmt(AVAL), 1), .groups = "drop")

sp_by_age <- adis_age |>
  filter(PARAMCD == "H1N1", VISIT == "DAY 29") |>
  group_by(ARMCD, AGEGR1) |>
  summarise(n = n(), n_sprot = sum(AVAL >= 40, na.rm = TRUE),
            pct_sp = round(100 * n_sprot / n, 1), .groups = "drop")

left_join(gmt_by_age, sp_by_age, by = c("ARMCD","AGEGR1")) |>
  arrange(ARMCD, AGEGR1)
```

---

## Section 6 — Capstone Problems

### Capstone 1 — Complete ADIS Assembly

**Scenario:** You are programming ADIS for a bivalent influenza + RSV vaccine submission. The statistical analysis plan requires a single ADIS containing:

| PARAMCD | Description | DTYPE |
|---------|-------------|-------|
| `H1N1`, `H3N2`, `RSV-F` | Observed titre (as-is rows) | NA |
| `GMFR` | Per-subject geometric mean fold rise for each antigen | `GMFR` |
| `SCONV` | Seroconversion (≥4-fold AND ≥40) for HAI antigens only | `SCONV` |
| `FF_RSV` | 4-fold rise for RSV-F | `4FOLD` |
| `SPROT` | Seroprotection (≥40) for HAI antigens | `SPROT` |

Additionally: `CHG` = AVAL − BASE and `PCHG` = 100 × CHG / BASE must be present on **observed titre rows** (not summary DTYPE rows).

```{r capstone1, exercise=TRUE, exercise.lines=70}
# ── Step 1: Observed titres with CHG / PCHG ───────────────────────────────────
adis_obs <- adis_base |>
  filter(is.na(DTYPE)) |>
  mutate(
    CHG  = AVAL - BASE,
    PCHG = round(100 * CHG / BASE, 2)
  )

# ── Step 2: GMFR rows for all three antigens ──────────────────────────────────
adis_gmfr <- derive_param_gmfr(
  dataset       = ___,
  by_vars       = c("USUBJID","PARAMCD"),
  baseline_var  = BASE,
  aval_var      = AVAL,
  set_values_to = list(PARAM = "Geometric Mean Fold Rise")
)

# ── Step 3: Seroconversion for HAI antigens ───────────────────────────────────
adis_sc <- derive_param_seroconversion(
  dataset         = adis_obs |> filter(PARAMCD %in% c("H1N1","H3N2")),
  dataset_adsl    = adsl_vac,
  by_vars         = c("USUBJID","PARAMCD"),
  baseline_var    = BASE,
  aval_var        = AVAL,
  fold_threshold  = ___,
  titre_threshold = ___,
  set_values_to   = list(PARAM = "Seroconversion", PARAMCD = "SCONV")
)

# ── Step 4: 4-fold rise for RSV-F ─────────────────────────────────────────────
adis_ff <- derive_param_fourfold(
  dataset       = adis_obs |> filter(PARAMCD == ___),
  by_vars       = c("USUBJID","PARAMCD"),
  baseline_var  = BASE,
  aval_var      = AVAL,
  fold          = 4,
  set_values_to = list(PARAM = "RSV-F 4-Fold Rise", PARAMCD = "FF_RSV")
)

# ── Step 5: Seroprotection for HAI ───────────────────────────────────────────
adis_sp <- derive_param_seroprotection(
  dataset         = adis_obs |> filter(PARAMCD %in% c("H1N1","H3N2"),
                                        VISIT == "DAY 29"),
  dataset_adsl    = adsl_vac,
  by_vars         = c("USUBJID","PARAMCD"),
  aval_var        = AVAL,
  titre_threshold = ___,
  set_values_to   = list(PARAM = "Seroprotection", PARAMCD = "SPROT")
)

# ── Step 6: Assemble ─────────────────────────────────────────────────────────
adis_final <- bind_rows(
  adis_obs,
  adis_gmfr |> filter(DTYPE == "GMFR"),
  adis_sc   |> filter(DTYPE == "SCONV"),
  adis_ff   |> filter(DTYPE == "4FOLD"),
  adis_sp   |> filter(DTYPE == "SPROT")
) |>
  arrange(USUBJID, PARAMCD, DTYPE, VISIT)

# ── Validation ────────────────────────────────────────────────────────────────
cat("Total rows:", nrow(adis_final), "\n")
adis_final |>
  count(PARAMCD, DTYPE) |>
  arrange(PARAMCD, DTYPE) |>
  print()
```

```{r capstone1-hint-1}
# Step 2: dataset = adis_obs (or adis_base with is.na filter)
# Step 3: fold_threshold = 4, titre_threshold = 40
# Step 4: PARAMCD == "RSV-F"
# Step 5: titre_threshold = 40
```

```{r capstone1-hint-2}
# Step 2:
adis_gmfr <- derive_param_gmfr(
  dataset = adis_obs, by_vars = c("USUBJID","PARAMCD"),
  baseline_var = BASE, aval_var = AVAL,
  set_values_to = list(PARAM = "Geometric Mean Fold Rise")
)
# Step 3: fold_threshold = 4, titre_threshold = 40
# Step 4: filter(PARAMCD == "RSV-F")
# Step 5: titre_threshold = 40
```

```{r capstone1-solution}
adis_obs <- adis_base |>
  filter(is.na(DTYPE)) |>
  mutate(CHG  = AVAL - BASE,
         PCHG = round(100 * CHG / BASE, 2))

adis_gmfr <- derive_param_gmfr(
  dataset       = adis_obs,
  by_vars       = c("USUBJID","PARAMCD"),
  baseline_var  = BASE, aval_var = AVAL,
  set_values_to = list(PARAM = "Geometric Mean Fold Rise")
)

adis_sc <- derive_param_seroconversion(
  dataset         = adis_obs |> filter(PARAMCD %in% c("H1N1","H3N2")),
  dataset_adsl    = adsl_vac,
  by_vars         = c("USUBJID","PARAMCD"),
  baseline_var    = BASE, aval_var = AVAL,
  fold_threshold  = 4, titre_threshold = 40,
  set_values_to   = list(PARAM = "Seroconversion", PARAMCD = "SCONV")
)

adis_ff <- derive_param_fourfold(
  dataset       = adis_obs |> filter(PARAMCD == "RSV-F"),
  by_vars       = c("USUBJID","PARAMCD"),
  baseline_var  = BASE, aval_var = AVAL, fold = 4,
  set_values_to = list(PARAM = "RSV-F 4-Fold Rise", PARAMCD = "FF_RSV")
)

adis_sp <- derive_param_seroprotection(
  dataset         = adis_obs |> filter(PARAMCD %in% c("H1N1","H3N2"),
                                        VISIT == "DAY 29"),
  dataset_adsl    = adsl_vac,
  by_vars         = c("USUBJID","PARAMCD"),
  aval_var        = AVAL, titre_threshold = 40,
  set_values_to   = list(PARAM = "Seroprotection", PARAMCD = "SPROT")
)

adis_final <- bind_rows(
  adis_obs,
  adis_gmfr |> filter(DTYPE == "GMFR"),
  adis_sc   |> filter(DTYPE == "SCONV"),
  adis_ff   |> filter(DTYPE == "4FOLD"),
  adis_sp   |> filter(DTYPE == "SPROT")
) |>
  arrange(USUBJID, PARAMCD, DTYPE, VISIT)

cat("Total rows:", nrow(adis_final), "\n")
adis_final |>
  count(PARAMCD, DTYPE) |>
  arrange(PARAMCD, DTYPE) |>
  print()
```

### Capstone 2 — Immunogenicity Summary Table (ISE format)

**Scenario:** Build a submission-ready immunogenicity summary table matching the format used in Integrated Summary of Efficacy (ISE) documents. The table must show for each antigen and arm:

- N (evaluable subjects)
- GMT at Day 1 and Day 29 (with 95% CI using ±1.96 SE on log scale)
- GMFR (with 95% CI)
- Seroconversion rate % (n / N)
- Seroprotection rate % at Day 29

95% CI for GMT: back-transform from `exp(mean(log) ± 1.96 × sd(log)/√N)`

```{r capstone2, exercise=TRUE, exercise.lines=60}
# ── Helper: GMT with 95% CI string ───────────────────────────────────────────
gmt_ci <- function(x) {
  x   <- x[x > 0 & !is.na(x)]
  n   <- length(x)
  if (n == 0) return(NA_character_)
  lx  <- log(x)
  mn  <- mean(lx)
  se  <- sd(lx) / sqrt(n)
  sprintf("%.1f (%.1f, %.1f)", exp(mn), exp(mn - 1.96*se), exp(mn + 1.96*se))
}

# ── Observed titre data ───────────────────────────────────────────────────────
obs <- adis_base |>
  filter(is.na(DTYPE)) |>
  left_join(select(adsl_vac, USUBJID, AGEGR1), by = "USUBJID")

# ── GMT at Day 1 and Day 29 ──────────────────────────────────────────────────
gmt_rows <- obs |>
  group_by(ARMCD, PARAMCD, VISIT) |>
  summarise(
    N         = n(),
    Stat      = paste0("GMT (", VISIT, ")"),
    Value     = gmt_ci(___),
    .groups   = "drop"
  )

# ── GMFR ────────────────────────────────────────────────────────────────────
gmfr_rows <- obs |>
  filter(VISIT == "DAY 29", BASE > 0) |>
  group_by(ARMCD, PARAMCD) |>
  summarise(
    N     = n(),
    Stat  = "GMFR (Day 1 → Day 29)",
    Value = {
      folds <- AVAL / BASE
      folds <- folds[folds > 0]
      gmt_ci(folds)
    },
    .groups = "drop"
  )

# ── Seroconversion (HAI only) ────────────────────────────────────────────────
sconv_rows <- obs |>
  filter(PARAMCD %in% c("H1N1","H3N2"), VISIT == "DAY 29") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(
    N     = n(),
    Stat  = "Seroconversion n (%)",
    Value = {
      sc <- (AVAL / BASE >= 4) & (AVAL >= 40)
      sprintf("%d (%.1f%%)", sum(sc, na.rm=TRUE),
              100*mean(sc, na.rm=TRUE))
    },
    .groups = "drop"
  )

# ── Seroprotection (HAI only, Day 29) ────────────────────────────────────────
sprot_rows <- obs |>
  filter(PARAMCD %in% c("H1N1","H3N2"), VISIT == "DAY 29") |>
  group_by(___, ___) |>
  summarise(
    N     = n(),
    Stat  = "Seroprotection n (%)",
    Value = sprintf("%d (%.1f%%)", sum(AVAL >= 40, na.rm=TRUE),
                    100*mean(AVAL >= 40, na.rm=TRUE)),
    .groups = "drop"
  )

# ── Assemble ─────────────────────────────────────────────────────────────────
ise_table <- bind_rows(gmt_rows, gmfr_rows, sconv_rows, sprot_rows) |>
  select(PARAMCD, ARMCD, N, Stat, Value) |>
  arrange(PARAMCD, ARMCD, Stat)

print(ise_table, n = Inf)
```

```{r capstone2-hint-1}
# gmt_rows: gmt_ci(AVAL)
# sprot_rows: group by ARMCD and PARAMCD (same pattern as sconv_rows)
```

```{r capstone2-solution}
gmt_ci <- function(x) {
  x  <- x[x > 0 & !is.na(x)]; n <- length(x)
  if (n == 0) return(NA_character_)
  lx <- log(x); mn <- mean(lx); se <- sd(lx) / sqrt(n)
  sprintf("%.1f (%.1f, %.1f)", exp(mn), exp(mn - 1.96*se), exp(mn + 1.96*se))
}

obs <- adis_base |>
  filter(is.na(DTYPE)) |>
  left_join(select(adsl_vac, USUBJID, AGEGR1), by = "USUBJID")

gmt_rows <- obs |>
  group_by(ARMCD, PARAMCD, VISIT) |>
  summarise(N = n(), Stat = paste0("GMT (", VISIT, ")"),
            Value = gmt_ci(AVAL), .groups = "drop")

gmfr_rows <- obs |>
  filter(VISIT == "DAY 29", BASE > 0) |>
  group_by(ARMCD, PARAMCD) |>
  summarise(N = n(), Stat = "GMFR (Day 1 → Day 29)",
            Value = { folds <- AVAL / BASE; folds <- folds[folds > 0]; gmt_ci(folds) },
            .groups = "drop")

sconv_rows <- obs |>
  filter(PARAMCD %in% c("H1N1","H3N2"), VISIT == "DAY 29") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(N = n(), Stat = "Seroconversion n (%)",
            Value = { sc <- (AVAL / BASE >= 4) & (AVAL >= 40)
                      sprintf("%d (%.1f%%)", sum(sc,na.rm=TRUE), 100*mean(sc,na.rm=TRUE)) },
            .groups = "drop")

sprot_rows <- obs |>
  filter(PARAMCD %in% c("H1N1","H3N2"), VISIT == "DAY 29") |>
  group_by(ARMCD, PARAMCD) |>
  summarise(N = n(), Stat = "Seroprotection n (%)",
            Value = sprintf("%d (%.1f%%)", sum(AVAL >= 40, na.rm=TRUE),
                            100*mean(AVAL >= 40, na.rm=TRUE)),
            .groups = "drop")

ise_table <- bind_rows(gmt_rows, gmfr_rows, sconv_rows, sprot_rows) |>
  select(PARAMCD, ARMCD, N, Stat, Value) |>
  arrange(PARAMCD, ARMCD, Stat)

print(ise_table, n = Inf)
```

---

## Section 7 — Hard Challenges

### Challenge 1 — Baseline seropositive subjects

**Background:** Subjects who are seropositive at baseline (pre-existing titre ≥ 40 for HAI) respond differently to vaccination. Regulators require efficacy analyses stratified by baseline serostatus. Flag all subjects as `SEROPRE = "Y"/"N"` at Day 1 per antigen, then re-compute seroconversion rates separately for seronegative and seropositive subjects.

```{r challenge-seropre, exercise=TRUE, exercise.lines=36}
# Step 1: Seropositivity flag at Day 1 per antigen (HAI only)
baseline_status <- adis_base |>
  filter(VISIT == "DAY 1", PARAMCD %in% c("H1N1","H3N2")) |>
  mutate(SEROPRE = if_else(AVAL >= ___, "Y", "N")) |>
  select(USUBJID, PARAMCD, SEROPRE)

# Step 2: Attach SEROPRE to post-baseline rows and compute fold rise
sc_by_seropre <- adis_base |>
  filter(VISIT == "DAY 29", PARAMCD %in% c("H1N1","H3N2")) |>
  left_join(___, by = c("USUBJID","PARAMCD")) |>
  mutate(
    FOLD_RISE = AVAL / BASE,
    SCONV     = (FOLD_RISE >= 4) & (AVAL >= 40)
  ) |>
  group_by(ARMCD, PARAMCD, ___) |>
  summarise(
    n        = n(),
    n_sc     = sum(SCONV, na.rm = TRUE),
    pct_sc   = round(100 * n_sc / n, 1),
    GMT_D29  = round(gmt(AVAL), 1),
    .groups  = "drop"
  )

sc_by_seropre |> arrange(PARAMCD, ARMCD, SEROPRE)
```

```{r challenge-seropre-hint-1}
# Threshold for seropositivity: AVAL >= 40
# left_join baseline_status by c("USUBJID","PARAMCD")
# group by ARMCD, PARAMCD, SEROPRE
```

```{r challenge-seropre-solution}
baseline_status <- adis_base |>
  filter(VISIT == "DAY 1", PARAMCD %in% c("H1N1","H3N2")) |>
  mutate(SEROPRE = if_else(AVAL >= 40, "Y", "N")) |>
  select(USUBJID, PARAMCD, SEROPRE)

sc_by_seropre <- adis_base |>
  filter(VISIT == "DAY 29", PARAMCD %in% c("H1N1","H3N2")) |>
  left_join(baseline_status, by = c("USUBJID","PARAMCD")) |>
  mutate(FOLD_RISE = AVAL / BASE,
         SCONV     = (FOLD_RISE >= 4) & (AVAL >= 40)) |>
  group_by(ARMCD, PARAMCD, SEROPRE) |>
  summarise(n = n(), n_sc = sum(SCONV, na.rm = TRUE),
            pct_sc = round(100 * n_sc / n, 1),
            GMT_D29 = round(gmt(AVAL), 1), .groups = "drop")

sc_by_seropre |> arrange(PARAMCD, ARMCD, SEROPRE)
```

### Challenge 2 — Reactogenicity severity shift analysis

**Scenario:** The DSMB requests a **worst-grade** reactogenicity profile: for each subject and each symptom, what was the worst (maximum) grade reported over the 7-day diary? Then produce a cross-tabulation of worst grade (None / Grade 1 / Grade 2 / Grade 3) by arm and symptom — similar to a CTCAE shift table.

```{r challenge-worst-grade, exercise=TRUE, exercise.lines=40}
# Step 1: Find worst grade per subject per symptom
worst_grade <- face_rows |>
  group_by(USUBJID, FACETESTCD, FACETEST, FACECAT, ARMCD) |>
  summarise(WORST_GRADE = max(FACESEVN, na.rm = TRUE), .groups = "drop")

# Step 2: Fill in subjects who did NOT report a symptom (worst grade = 0 = "None")
all_combos <- expand_grid(
  USUBJID    = adsl_vac$USUBJID,
  FACETESTCD = face_symptoms$FACETESTCD
) |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  left_join(face_symptoms, by = "FACETESTCD") |>
  left_join(___, by = c("USUBJID","FACETESTCD","ARMCD")) |>
  mutate(
    WORST_GRADE = replace_na(___, 0L),
    GRADE_LABEL = case_when(
      WORST_GRADE == 0 ~ "None",
      WORST_GRADE == 1 ~ "Grade 1",
      WORST_GRADE == 2 ~ "Grade 2",
      WORST_GRADE == 3 ~ "Grade 3"
    )
  )

# Step 3: Cross-tab
shift_tbl <- all_combos |>
  group_by(ARMCD, FACECAT, FACETESTCD, GRADE_LABEL) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = ___, values_from = n, values_fill = 0L) |>
  select(ARMCD, FACECAT, FACETESTCD, None, `Grade 1`, `Grade 2`, `Grade 3`) |>
  arrange(FACECAT, ARMCD, FACETESTCD)

print(shift_tbl, n = Inf)
```

```{r challenge-worst-grade-hint-1}
# left_join worst_grade by c("USUBJID","FACETESTCD","ARMCD")
# replace_na(WORST_GRADE, 0L)
# pivot_wider: names_from = GRADE_LABEL
```

```{r challenge-worst-grade-solution}
worst_grade <- face_rows |>
  group_by(USUBJID, FACETESTCD, FACETEST, FACECAT, ARMCD) |>
  summarise(WORST_GRADE = max(FACESEVN, na.rm = TRUE), .groups = "drop")

all_combos <- expand_grid(
  USUBJID    = adsl_vac$USUBJID,
  FACETESTCD = face_symptoms$FACETESTCD
) |>
  left_join(select(adsl_vac, USUBJID, ARMCD), by = "USUBJID") |>
  left_join(face_symptoms, by = "FACETESTCD") |>
  left_join(worst_grade, by = c("USUBJID","FACETESTCD","ARMCD")) |>
  mutate(
    WORST_GRADE = replace_na(WORST_GRADE, 0L),
    GRADE_LABEL = case_when(
      WORST_GRADE == 0 ~ "None",
      WORST_GRADE == 1 ~ "Grade 1",
      WORST_GRADE == 2 ~ "Grade 2",
      WORST_GRADE == 3 ~ "Grade 3"
    )
  )

shift_tbl <- all_combos |>
  group_by(ARMCD, FACECAT, FACETESTCD, GRADE_LABEL) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = GRADE_LABEL, values_from = n, values_fill = 0L) |>
  select(ARMCD, FACECAT, FACETESTCD,
         any_of(c("None","Grade 1","Grade 2","Grade 3"))) |>
  arrange(FACECAT, ARMCD, FACETESTCD)

print(shift_tbl, n = Inf)
```

### Challenge 3 — Vaccine efficacy against seroconversion using risk ratio

**Advanced:** Compute the **risk ratio (RR)** and its 95% CI for seroconversion (H1N1) comparing VAC to PBO. Use the log-binomial formula: RR = p_vac / p_pbo; 95% CI via `exp(log(RR) ± 1.96 × sqrt(1/x_vac − 1/n_vac + 1/x_pbo − 1/n_pbo))`.

```{r challenge-rr, exercise=TRUE, exercise.lines=36}
# Step 1: Seroconversion counts per arm (H1N1)
sc_counts <- adis_base |>
  filter(PARAMCD == "H1N1", VISIT == "DAY 29") |>
  mutate(SC = (AVAL / BASE >= 4) & (AVAL >= 40)) |>
  group_by(ARMCD) |>
  summarise(n = n(), x = sum(SC, na.rm = TRUE),
            p = x / n, .groups = "drop")

sc_counts

# Step 2: Extract components
n_vac <- sc_counts$n[sc_counts$ARMCD == "VAC"]
x_vac <- sc_counts$x[sc_counts$ARMCD == "VAC"]
n_pbo <- sc_counts$n[sc_counts$ARMCD == "PBO"]
x_pbo <- sc_counts$x[sc_counts$ARMCD == "PBO"]

p_vac <- x_vac / n_vac
p_pbo <- x_pbo / n_pbo

# Step 3: Risk Ratio and 95% CI
RR    <- ___ / ___
se_lnRR <- sqrt(1/x_vac - 1/n_vac + 1/x_pbo - 1/n_pbo)
CI_lo <- exp(log(RR) - 1.96 * se_lnRR)
CI_hi <- exp(log(RR) + 1.96 * se_lnRR)

cat(sprintf(
  "H1N1 Seroconversion\n  VAC: %d/%d (%.1f%%)\n  PBO: %d/%d (%.1f%%)\n  RR: %.2f (95%% CI: %.2f, %.2f)\n",
  x_vac, n_vac, 100*p_vac,
  x_pbo, n_pbo, 100*p_pbo,
  RR, CI_lo, CI_hi
))
```

```{r challenge-rr-hint-1}
# RR = p_vac / p_pbo
# se_lnRR = sqrt(1/x_vac - 1/n_vac + 1/x_pbo - 1/n_pbo)
# CI: exp(log(RR) ± 1.96 * se_lnRR)
```

```{r challenge-rr-solution}
sc_counts <- adis_base |>
  filter(PARAMCD == "H1N1", VISIT == "DAY 29") |>
  mutate(SC = (AVAL / BASE >= 4) & (AVAL >= 40)) |>
  group_by(ARMCD) |>
  summarise(n = n(), x = sum(SC, na.rm = TRUE), p = x / n, .groups = "drop")

sc_counts

n_vac <- sc_counts$n[sc_counts$ARMCD == "VAC"]
x_vac <- sc_counts$x[sc_counts$ARMCD == "VAC"]
n_pbo <- sc_counts$n[sc_counts$ARMCD == "PBO"]
x_pbo <- sc_counts$x[sc_counts$ARMCD == "PBO"]
p_vac <- x_vac / n_vac
p_pbo <- x_pbo / n_pbo

RR       <- p_vac / p_pbo
se_lnRR  <- sqrt(1/x_vac - 1/n_vac + 1/x_pbo - 1/n_pbo)
CI_lo    <- exp(log(RR) - 1.96 * se_lnRR)
CI_hi    <- exp(log(RR) + 1.96 * se_lnRR)

cat(sprintf(
  "H1N1 Seroconversion\n  VAC: %d/%d (%.1f%%)\n  PBO: %d/%d (%.1f%%)\n  RR: %.2f (95%% CI: %.2f, %.2f)\n",
  x_vac, n_vac, 100*p_vac, x_pbo, n_pbo, 100*p_pbo, RR, CI_lo, CI_hi
))
```

---

## Lesson Summary

```{r quiz-final, echo=FALSE}
quiz(
  caption = "Final Knowledge Check",

  question("The correct formula for Geometric Mean Titre (GMT) from individual titres x₁, x₂ … xₙ is:",
    answer("mean(x)"),
    answer("exp(mean(log(x)))", correct = TRUE),
    answer("median(x)"),
    answer("prod(x)^(1/n) — which is equivalent to exp(mean(log(x)))",
           correct = TRUE,
           message = "Both are correct and mathematically identical: the geometric mean."),
    allow_retry = TRUE
  ),

  question("Which statement correctly distinguishes seroconversion from a 4-fold rise?",
    answer("They are the same thing"),
    answer("Seroconversion requires ≥4-fold rise AND post-vaccination titre ≥ threshold; a 4-fold rise has no titre floor",
           correct = TRUE),
    answer("A 4-fold rise requires a higher fold increase (≥8-fold)"),
    answer("Seroconversion applies only to ELISA; 4-fold rise applies to HAI"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise — End-to-end immunogenicity summary

Bring it all together: for **both HAI antigens** (H1N1 and H3N2), compute a one-table summary by arm showing GMT at Day 29, GMFR, and seroconversion rate. Use the `gmt()` helper and the seroconversion two-condition rule directly (no helper functions needed).

```{r ex-final-summary, exercise=TRUE, exercise.lines=22}
# adis_base has PARAMCD, VISIT, AVAL, BASE, ARMCD
# Filter to HAI antigens and DAY 29 post-vaccination rows

adis_base |>
  filter(PARAMCD %in% c("H1N1", "H3N2"), VISIT == ___) |>
  mutate(
    FOLD_RISE = AVAL / BASE,
    SCONV     = FOLD_RISE >= ___ & AVAL >= ___   # seroconversion two conditions
  ) |>
  group_by(ARMCD, PARAMCD) |>
  summarise(
    n          = n(),
    GMT_D29    = round(gmt(___), 1),
    GMFR       = round(exp(mean(log(FOLD_RISE[FOLD_RISE > 0 & !is.na(FOLD_RISE)]))), 2),
    n_sc       = sum(SCONV, na.rm = TRUE),
    pct_sc     = round(100 * n_sc / n, 1),
    .groups    = "drop"
  ) |>
  arrange(PARAMCD, ARMCD)
```

```{r ex-final-summary-hint-1}
# VISIT == "DAY 29"
# SCONV = FOLD_RISE >= 4 & AVAL >= 40
# GMT_D29 = gmt(AVAL)
```

```{r ex-final-summary-solution}
adis_base |>
  filter(PARAMCD %in% c("H1N1", "H3N2"), VISIT == "DAY 29") |>
  mutate(
    FOLD_RISE = AVAL / BASE,
    SCONV     = FOLD_RISE >= 4 & AVAL >= 40
  ) |>
  group_by(ARMCD, PARAMCD) |>
  summarise(
    n       = n(),
    GMT_D29 = round(gmt(AVAL), 1),
    GMFR    = round(exp(mean(log(FOLD_RISE[FOLD_RISE > 0 & !is.na(FOLD_RISE)]))), 2),
    n_sc    = sum(SCONV, na.rm = TRUE),
    pct_sc  = round(100 * n_sc / n, 1),
    .groups = "drop"
  ) |>
  arrange(PARAMCD, ARMCD)
```

### Key takeaways

You have learned to:

- Understand vaccine ADaM datasets: ADIS (immunogenicity), ADFACE (reactogenicity), ADCE (clinical events)
- Apply the GMT / GMFR framework using log-normal arithmetic
- Distinguish seroconversion (fold + titre floor) from seroprotection (titre only) and 4-fold rise (fold only)
- Use `derive_param_seroconversion()`, `derive_param_seroprotection()`, `derive_param_gmfr()`, and `derive_param_fourfold()`
- Build ADFACE reactogenicity summaries with grade-based incidence and worst-grade shift tables
- Perform age-stratified immunogenicity analyses and risk ratio estimation for regulatory submissions

**Next lesson:** Lesson 22 — `admiralophtha` — ADaM derivations for ophthalmology trials (BCVA, IOP, retinal thickness, laterality handling).
