---
title: "Lesson 22 — admiral Ophthalmology Extension (admiralophtha): BCVA, IOP, OCT & Laterality"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  Ophthalmology ADaM derivations using the admiralophtha package: handling bilateral
  data (OD/OS laterality), BCVA scale conversions (ETDRS letters ↔ logMAR ↔ Snellen),
  visual acuity category endpoints (≥15-letter gain/loss), intraocular pressure (IOP),
  OCT central retinal thickness (CRT), and the unique programming challenges of
  within-subject eye-level analyses. Includes AMD and DME trial capstone problems.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 22_admiralophtha.Rmd
# PURPOSE       : admiralophtha extension — ophthalmology ADaM: BCVA (ETDRS letters and
#                 logMAR), bilateral/laterality PARAMCD encoding, study-
#                 eye designation, IOP worst-eye derivation, and retinal
#                 thickness endpoints.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(lubridate)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# SYNTHETIC OPHTHALMOLOGY TRIAL DATA
# Phase III wet AMD (Age-related Macular Degeneration) trial.
# Treatment: anti-VEGF intravitreal injection vs Sham injection.
# 100 subjects; both eyes measured at every visit.
# Primary endpoint: change from baseline in BCVA (ETDRS letters) at Week 52.
# ══════════════════════════════════════════════════════════════════════════════
set.seed(2024)
n_sub <- 100

adsl_optha <- tibble(
  STUDYID  = "AMD2024",
  USUBJID  = sprintf("AMD2024-%03d", seq_len(n_sub)),
  SUBJID   = sprintf("%03d", seq_len(n_sub)),
  AGE      = sample(55:88, n_sub, replace = TRUE),
  SEX      = sample(c("M","F"), n_sub, replace = TRUE, prob = c(.38,.62)),
  RACE     = sample(c("WHITE","ASIAN","BLACK OR AFRICAN AMERICAN"),
                    n_sub, replace = TRUE, prob = c(.72,.18,.10)),
  ARM      = rep(c("Anti-VEGF","Sham"), each = n_sub/2),
  ARMCD    = rep(c("VEGF","SHAM"), each = n_sub/2),
  # Study eye: the eye with AMD that receives treatment
  STDEYE   = sample(c("OD","OS"), n_sub, replace = TRUE),
  TRTSDT   = as.Date("2022-01-10") + sample(0:60, n_sub, replace = TRUE),
  SAFFL    = "Y",
  ITTFL    = "Y"
) |>
  mutate(
    TRTEDT   = TRTSDT + 364L,   # 52-week treatment
    LSTALVDT = TRTEDT + sample(0:90, n_sub, replace = TRUE)
  )

# ── Visit schedule ────────────────────────────────────────────────────────────
visits <- tibble(
  VISIT    = c("BASELINE","WEEK 4","WEEK 8","WEEK 12","WEEK 24","WEEK 52"),
  VISITNUM = c(1, 2, 3, 4, 5, 6),
  VISITDY  = c(1L, 29L, 57L, 85L, 169L, 365L)
)

eyes <- c("OD","OS")   # oculus dexter (right), oculus sinister (left)

# ── OE domain simulation: BCVA (ETDRS letters), IOP (mmHg), CRT (μm OCT) ────
# Study eye gets treatment benefit; fellow eye is background noise

sim_bcva <- function(arm, eye, stdeye, visit_day, baseline) {
  is_study_eye <- (eye == stdeye)
  is_vegf      <- (arm == "VEGF")

  if (visit_day == 1) return(baseline)

  # Letters gained/lost from baseline
  if (is_vegf && is_study_eye) {
    # Anti-VEGF study eye: average +8 letters at Wk52
    gain <- rnorm(1, mean = 8 * (visit_day / 365), sd = 6)
  } else if (!is_vegf && is_study_eye) {
    # Sham study eye: slow decline
    gain <- rnorm(1, mean = -3 * (visit_day / 365), sd = 5)
  } else {
    # Fellow eye: small random drift
    gain <- rnorm(1, mean = 0.5, sd = 3)
  }
  round(pmin(100, pmax(0, baseline + gain)))
}

sim_iop <- function(arm, eye, stdeye, visit_day, baseline_iop) {
  # IOP: slight reduction in treated study eye over time (anti-VEGF can mildly lower IOP)
  is_study <- (eye == stdeye) && (arm == "VEGF")
  noise <- rnorm(1, 0, 1.5)
  reduction <- if (is_study && visit_day > 1) -0.5 * (visit_day / 365) else 0
  round(pmax(6, pmin(30, baseline_iop + reduction + noise)), 1)
}

sim_crt <- function(arm, eye, stdeye, visit_day, baseline_crt) {
  # CRT (retinal thickness): elevated at baseline in study eye with AMD
  # Anti-VEGF reduces fluid → CRT decreases
  is_study <- (eye == stdeye)
  is_vegf  <- (arm == "VEGF")

  if (visit_day == 1) return(baseline_crt)

  if (is_vegf && is_study) {
    change <- rnorm(1, mean = -120 * (visit_day / 365), sd = 50)
  } else if (!is_vegf && is_study) {
    change <- rnorm(1, mean = 30 * (visit_day / 365), sd = 40)
  } else {
    change <- rnorm(1, mean = 0, sd = 20)
  }
  round(pmax(150, pmin(800, baseline_crt + change)))
}

oe_rows <- lapply(seq_len(n_sub), function(i) {
  subj   <- adsl_optha$USUBJID[i]
  arm    <- adsl_optha$ARMCD[i]
  stdeye <- adsl_optha$STDEYE[i]
  trtdt  <- adsl_optha$TRTSDT[i]

  # Baseline values per eye
  base_bcva_ode <- sample(35:75, 1)   # study eye: moderate vision loss
  base_bcva_ose <- if (stdeye == "OD") base_bcva_ode else sample(55:90, 1)
  base_bcva <- c(OD = if(stdeye=="OD") base_bcva_ode else sample(55:90,1),
                 OS = if(stdeye=="OS") base_bcva_ode else sample(55:90,1))

  base_iop  <- c(OD = round(rnorm(1, 14, 2.5), 1), OS = round(rnorm(1, 14, 2.5), 1))
  base_crt  <- c(OD = if(stdeye=="OD") sample(380:600,1) else sample(200:310,1),
                 OS = if(stdeye=="OS") sample(380:600,1) else sample(200:310,1))

  lapply(seq_len(nrow(visits)), function(v) {
    vrow <- visits[v, ]
    lapply(eyes, function(eye) {
      bcva <- sim_bcva(arm, eye, stdeye, vrow$VISITDY, base_bcva[[eye]])
      iop  <- sim_iop( arm, eye, stdeye, vrow$VISITDY, base_iop[[eye]])
      crt  <- sim_crt( arm, eye, stdeye, vrow$VISITDY, base_crt[[eye]])
      adt  <- trtdt + vrow$VISITDY - 1L

      bind_rows(
        # BCVA
        tibble(STUDYID="AMD2024", USUBJID=subj, VISIT=vrow$VISIT,
               VISITNUM=vrow$VISITNUM, VISITDY=vrow$VISITDY,
               OETESTCD="VAET", OETEST="Visual Acuity (ETDRS letters)",
               OECAT="OPHTHALMIC EXAMINATION", OEOBJ=eye,
               OESTRESN=bcva, OESTRESC=as.character(bcva),
               OESTRESU="letters", ADT=adt, AVAL=as.numeric(bcva),
               ARMCD=arm, STDEYE=stdeye),
        # IOP
        tibble(STUDYID="AMD2024", USUBJID=subj, VISIT=vrow$VISIT,
               VISITNUM=vrow$VISITNUM, VISITDY=vrow$VISITDY,
               OETESTCD="IOP", OETEST="Intraocular Pressure",
               OECAT="OPHTHALMIC EXAMINATION", OEOBJ=eye,
               OESTRESN=iop, OESTRESC=as.character(iop),
               OESTRESU="mmHg", ADT=adt, AVAL=iop,
               ARMCD=arm, STDEYE=stdeye),
        # CRT
        tibble(STUDYID="AMD2024", USUBJID=subj, VISIT=vrow$VISIT,
               VISITNUM=vrow$VISITNUM, VISITDY=vrow$VISITDY,
               OETESTCD="CRT", OETEST="Central Retinal Thickness (OCT)",
               OECAT="OPHTHALMIC EXAMINATION", OEOBJ=eye,
               OESTRESN=crt, OESTRESC=as.character(crt),
               OESTRESU="um", ADT=adt, AVAL=as.numeric(crt),
               ARMCD=arm, STDEYE=stdeye)
      )
    }) |> bind_rows()
  }) |> bind_rows()
}) |> bind_rows()

# ── ADOE skeleton (with PARAMCD encoding laterality) ─────────────────────────
adoe_base <- oe_rows |>
  mutate(
    # Laterality-aware PARAMCD: e.g. VAET_OD, IOP_OS
    PARAMCD  = paste0(OETESTCD, "_", OEOBJ),
    PARAM    = paste0(OETEST, " — ", OEOBJ),
    STUDYEYE = if_else(OEOBJ == STDEYE, "Y", "N"),
    ADY      = as.integer(ADT - (adsl_optha$TRTSDT[match(USUBJID, adsl_optha$USUBJID)])) + 1L,
    DTYPE    = NA_character_,
    ABLFL    = if_else(VISIT == "BASELINE", "Y", NA_character_)
  ) |>
  group_by(USUBJID, PARAMCD) |>
  mutate(BASE = AVAL[VISIT == "BASELINE"][1]) |>
  ungroup()

# Convenience subsets
adoe_bcva <- adoe_base |> filter(OETESTCD == "VAET")
adoe_iop  <- adoe_base |> filter(OETESTCD == "IOP")
adoe_crt  <- adoe_base |> filter(OETESTCD == "CRT")

# ══════════════════════════════════════════════════════════════════════════════
# ADMIRALOPHTHA FUNCTION MOCKS
# ══════════════════════════════════════════════════════════════════════════════

#' Convert ETDRS letters to logMAR
#' logMAR = 1.88 - 0.049 * letters  (Ferris et al. approximation)
etdrs_to_logmar <- function(letters) {
  round(1.88 - 0.049 * letters, 3)
}

#' Convert logMAR to ETDRS letters (inverse)
logmar_to_etdrs <- function(logmar) {
  round((1.88 - logmar) / 0.049)
}

#' Convert logMAR to approximate Snellen fraction (as character "20/XX")
logmar_to_snellen <- function(logmar) {
  denom <- round(20 * 10^logmar)
  paste0("20/", denom)
}

#' Derive logMAR column from ETDRS AVAL
derive_var_aval_logmar <- function(dataset,
                                    aval_var  = AVAL,
                                    new_var   = AVALOG,
                                    filter    = NULL,
                                    ...) {
  aval_nm <- deparse(substitute(aval_var))
  new_nm  <- deparse(substitute(new_var))
  filt    <- substitute(filter)

  if (!is.null(filt)) {
    idx <- tryCatch(which(eval(filt, dataset)), error = function(e) seq_len(nrow(dataset)))
    dataset[[new_nm]] <- NA_real_
    dataset[[new_nm]][idx] <- etdrs_to_logmar(dataset[[aval_nm]][idx])
  } else {
    dataset[[new_nm]] <- etdrs_to_logmar(dataset[[aval_nm]])
  }
  dataset
}

#' Derive BCVA change-from-baseline category
#' Regulatory categories: ≥15 gain, 0–14 gain, 0 change, 1–14 loss, ≥15 loss
derive_var_bcva_category <- function(dataset,
                                      chg_var    = CHG,
                                      new_var    = BCVA_CAT,
                                      thresholds = c(-15, 0, 15),
                                      ...) {
  chg_nm <- deparse(substitute(chg_var))
  new_nm <- deparse(substitute(new_var))

  dataset[[new_nm]] <- case_when(
    is.na(dataset[[chg_nm]])           ~ NA_character_,
    dataset[[chg_nm]] >= 15            ~ "Gain >=15 letters",
    dataset[[chg_nm]] >= 1             ~ "Gain 1-14 letters",
    dataset[[chg_nm]] == 0             ~ "No change",
    dataset[[chg_nm]] >= -14           ~ "Loss 1-14 letters",
    TRUE                               ~ "Loss >=15 letters"
  )
  dataset
}

#' Derive best corrected visual acuity responder flag
#' BCVA_RESPFL = "Y" if CHG >= threshold (default 15 letters = 3 lines)
derive_param_bcva_responder <- function(dataset,
                                         dataset_adsl,
                                         by_vars      = c("USUBJID","PARAMCD"),
                                         chg_var      = CHG,
                                         threshold    = 15,
                                         set_values_to = list(),
                                         ...) {
  chg_nm  <- deparse(substitute(chg_var))
  by_cols <- if (is.character(by_vars)) by_vars else sapply(by_vars, deparse)

  post <- dataset |> filter(VISIT != "BASELINE", !is.na(.data[[chg_nm]]))

  resp_rows <- post |>
    group_by(across(all_of(by_cols))) |>
    summarise(
      AVAL  = as.integer(any(.data[[chg_nm]] >= threshold, na.rm = TRUE)),
      AVALC = if_else(any(.data[[chg_nm]] >= threshold, na.rm = TRUE), "Y", "N"),
      .groups = "drop"
    ) |>
    mutate(DTYPE = "RESP")

  for (nm in names(set_values_to)) resp_rows[[nm]] <- set_values_to[[nm]]
  bind_rows(dataset, resp_rows)
}

#' Derive "study eye" subset — convenience wrapper
#' In ophthalmology analyses, analyses often focus on the study (treated) eye only
derive_study_eye <- function(dataset, study_eye_var = STUDYEYE) {
  var_nm <- deparse(substitute(study_eye_var))
  dataset |> filter(.data[[var_nm]] == "Y")
}

#' Derive worst IOP (maximum across eyes and visits)
derive_param_iop_worst <- function(dataset,
                                    by_vars       = "USUBJID",
                                    aval_var      = AVAL,
                                    set_values_to = list(),
                                    ...) {
  aval_nm <- deparse(substitute(aval_var))
  by_cols <- if (is.character(by_vars)) by_vars else sapply(by_vars, deparse)

  worst <- dataset |>
    group_by(across(all_of(by_cols))) |>
    summarise(
      AVAL  = max(.data[[aval_nm]], na.rm = TRUE),
      AVALC = as.character(max(.data[[aval_nm]], na.rm = TRUE)),
      .groups = "drop"
    ) |>
    mutate(DTYPE = "WORST")

  for (nm in names(set_values_to)) worst[[nm]] <- set_values_to[[nm]]
  bind_rows(dataset, worst)
}

# Shared helpers
exprs <- function(...) as.list(substitute(list(...)))[-1]
```

## Overview

### The ophthalmology ADaM landscape

Ophthalmology trials present a unique programming challenge: **every subject has two eyes**, and the treated (study) eye and the untreated (fellow) eye must be tracked separately but within the same subject record. This laterality dimension does not exist in other therapeutic areas.

| Dataset | Full Name | Key Content |
|---------|-----------|-------------|
| **ADOE** | Ophthalmic Examination | BCVA, IOP, OCT — one row per subject × eye × visit × parameter |
| **ADBCVA** | Best Corrected Visual Acuity | BCVA-specific analysis dataset |
| **ADOCC** | Ophthalmic Clinical Events | Injection counts, procedural events |

### Key visual acuity scales

| Scale | Range | Better = | Notes |
|-------|-------|----------|-------|
| ETDRS letters | 0–100 | Higher | Standard in clinical trials; each line = 5 letters |
| logMAR | ≥0 (typically 0–1.6) | Lower | Logarithm of Minimum Angle of Resolution |
| Snellen | 20/X (e.g. 20/20) | Lower denominator | Legacy clinical notation |

> **Conversion:** logMAR ≈ 1.88 − 0.049 × ETDRS letters (Ferris 1982). Normal vision (20/20) ≈ 84 ETDRS letters ≈ logMAR 0.

```{r overview-explore, exercise=FALSE}
# Preview ADOE baseline data
adoe_bcva |>
  filter(VISIT == "BASELINE") |>
  group_by(ARMCD, OEOBJ, STUDYEYE) |>
  summarise(
    n         = n(),
    mean_bcva = round(mean(AVAL), 1),
    mean_logmar = round(mean(etdrs_to_logmar(AVAL)), 3),
    .groups   = "drop"
  ) |>
  arrange(ARMCD, OEOBJ)
```

---

## Section 1 — Laterality: The OD/OS Challenge

### Why laterality is the hardest problem in ophthalmic ADaM

In a standard trial, `USUBJID` identifies one analysis unit (the subject). In ophthalmology, the analysis unit is often the **eye**: you need `USUBJID` + `OEOBJ` (OD/OS) to uniquely identify an observation. The PARAMCD in ADOE encodes both the measurement and the eye: e.g., `VAET_OD`, `IOP_OS`, `CRT_OD`.

> **SAS parallel:** In SAS, this was often handled with two separate datasets (one per eye) then merged. admiral handles it with laterality-aware PARAMCD naming, keeping everything in one long dataset.

```{r quiz-laterality1, echo=FALSE}
quiz(
  caption = "Quiz — Laterality Fundamentals",

  question("In CDISC SDTM, which variable records which eye was examined?",
    answer("OEOBJ — the object of the observation (OD = right eye, OS = left eye)",
           correct = TRUE,
           message = "OEOBJ (Object of Examination) stores 'OD', 'OS', or 'OU' (both eyes). In ADaM, this is often encoded into PARAMCD (e.g., 'VAET_OD')."),
    answer("OECAT — the category of the ophthalmic test"),
    answer("OELOC — the location within the eye"),
    answer("OETESTCD — the test code identifies the eye"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("The 'study eye' (STUDYEYE = 'Y') in an AMD trial is:",
    answer("Always the right eye (OD)"),
    answer("The eye with the target disease that receives the study treatment",
           correct = TRUE,
           message = "The study eye is the eye enrolled for treatment. The other eye is the 'fellow eye'. Both are measured, but the primary efficacy analyses use the study eye only."),
    answer("The eye with better baseline vision"),
    answer("Whichever eye the investigator chooses at Week 4"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise 1.0 — Count records by PARAMCD, arm, and visit

`adoe_bcva` encodes laterality in `PARAMCD` (e.g., `VAET_OD` for right eye, `VAET_OS` for left eye). Count how many records exist for each combination of PARAMCD, arm, and visit.

```{r ex-lateral-count, exercise=TRUE, exercise.lines=12}
# adoe_bcva has PARAMCD encoding the eye: e.g. "VAET_OD" (right), "VAET_OS" (left).
# Count the number of records per PARAMCD, visit, and arm.
# Which eye has more observations and why might they differ?

adoe_bcva |>
  count(___, ARMCD, VISIT) |>
  arrange(PARAMCD, ARMCD, VISIT) |>
  head(12)
```

```{r ex-lateral-count-hint-1}
# count(PARAMCD, ARMCD, VISIT)
```

```{r ex-lateral-count-solution}
adoe_bcva |>
  count(PARAMCD, ARMCD, VISIT) |>
  arrange(PARAMCD, ARMCD, VISIT) |>
  head(12)
```

### Exercise 1.1 — Identify study eye records

From `adoe_bcva`, filter to study eye records only and count subjects by arm.

```{r ex-study-eye, exercise=TRUE, exercise.lines=16}
# adoe_bcva has USUBJID, OEOBJ, STUDYEYE ("Y"/"N"), VISIT, AVAL, ARMCD

study_eye_bcva <- adoe_bcva |>
  filter(___ == "Y")   # keep only study eye rows

# Count subjects and mean baseline BCVA by arm
study_eye_bcva |>
  filter(VISIT == "BASELINE") |>
  group_by(___) |>
  summarise(
    n_subjects  = n_distinct(___),
    mean_bcva   = round(mean(AVAL), 1),
    .groups     = "drop"
  )
```

```{r ex-study-eye-hint-1}
# filter STUDYEYE == "Y"; group by ARMCD; n_distinct(USUBJID)
adoe_bcva |>
  filter(STUDYEYE == "Y") |>
  filter(VISIT == "BASELINE") |>
  group_by(ARMCD) |>
  summarise(n_subjects = n_distinct(USUBJID),
            mean_bcva = round(mean(AVAL), 1), .groups = "drop")
```

```{r ex-study-eye-solution}
study_eye_bcva <- adoe_bcva |>
  filter(STUDYEYE == "Y")

study_eye_bcva |>
  filter(VISIT == "BASELINE") |>
  group_by(ARMCD) |>
  summarise(
    n_subjects = n_distinct(USUBJID),
    mean_bcva  = round(mean(AVAL), 1),
    .groups    = "drop"
  )
```

### Exercise 1.2 — Wide format: compare OD vs OS at baseline

Sometimes a QC or analysis step requires OD and OS side by side. Pivot the baseline BCVA data to wide format with columns `BCVA_OD` and `BCVA_OS`, then compute the inter-eye difference.

```{r ex-wide-eyes, exercise=TRUE, exercise.lines=20}
# adoe_bcva is available
eye_wide <- adoe_bcva |>
  filter(VISIT == "BASELINE") |>
  select(USUBJID, OEOBJ, AVAL, STDEYE, ARMCD) |>
  pivot_wider(
    names_from  = ___,
    values_from = ___,
    names_prefix = "BCVA_"
  ) |>
  mutate(
    EYE_DIFF    = BCVA_OD - BCVA_OS,    # positive = OD better
    WORSE_EYE   = if_else(BCVA_OD <= BCVA_OS, "OD", "OS")
  )

# Summary: mean inter-eye difference by arm
eye_wide |>
  group_by(ARMCD) |>
  summarise(
    mean_diff = round(mean(EYE_DIFF, na.rm = TRUE), 1),
    .groups   = "drop"
  )
```

```{r ex-wide-eyes-hint-1}
# names_from = OEOBJ (gives OD and OS); values_from = AVAL; names_prefix = "BCVA_"
adoe_bcva |>
  filter(VISIT == "BASELINE") |>
  select(USUBJID, OEOBJ, AVAL, STDEYE, ARMCD) |>
  pivot_wider(names_from = OEOBJ, values_from = AVAL, names_prefix = "BCVA_") |>
  mutate(EYE_DIFF = BCVA_OD - BCVA_OS, WORSE_EYE = if_else(BCVA_OD <= BCVA_OS,"OD","OS"))
```

```{r ex-wide-eyes-solution}
eye_wide <- adoe_bcva |>
  filter(VISIT == "BASELINE") |>
  select(USUBJID, OEOBJ, AVAL, STDEYE, ARMCD) |>
  pivot_wider(
    names_from   = OEOBJ,
    values_from  = AVAL,
    names_prefix = "BCVA_"
  ) |>
  mutate(
    EYE_DIFF  = BCVA_OD - BCVA_OS,
    WORSE_EYE = if_else(BCVA_OD <= BCVA_OS, "OD", "OS")
  )

eye_wide |>
  group_by(ARMCD) |>
  summarise(
    mean_diff = round(mean(EYE_DIFF, na.rm = TRUE), 1),
    .groups   = "drop"
  )
```

---

## Section 2 — BCVA Scale Conversions

### ETDRS, logMAR, and Snellen

Regulatory analyses require ETDRS letters as the primary scale, but publications and clinical reports often show logMAR or Snellen. Programmers must handle all three — and know that converting back and forth introduces rounding artefacts.

```{r quiz-scales1, echo=FALSE}
quiz(
  caption = "Quiz — Visual Acuity Scales",

  question("Normal vision (20/20 Snellen) corresponds to approximately how many ETDRS letters?",
    answer("50 letters"),
    answer("70 letters"),
    answer("84–85 letters", correct = TRUE,
           message = "20/20 ≈ 84 ETDRS letters ≈ logMAR 0.0. Each line on the ETDRS chart = 5 letters; the chart goes to line 20 (20/16, ≈88 letters at the top)."),
    answer("100 letters — the maximum"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("In a trial, a subject gains 15 ETDRS letters. This is clinically equivalent to:",
    answer("1 line of vision (5 letters per line)"),
    answer("3 lines of vision", correct = TRUE,
           message = "15 letters ÷ 5 letters per ETDRS line = 3 lines. This is the standard threshold for a 'clinically meaningful' improvement in AMD/DME trials."),
    answer("Half a line of vision"),
    answer("0.15 logMAR improvement"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise 2.0 — ETDRS to logMAR conversion

The conversion formula (Ferris 1982) is: **logMAR = 1.88 − 0.049 × ETDRS letters**. Apply it to a set of letter scores.

```{r ex-logmar-convert, exercise=TRUE, exercise.lines=10}
# logMAR = 1.88 - 0.049 * ETDRS_letters
# Convert these ETDRS letter scores to logMAR:
etdrs_scores <- c(85, 70, 55, 35, 20)

tibble(
  ETDRS  = etdrs_scores,
  logMAR = round(1.88 - ___ * ___, 3)
)
```

```{r ex-logmar-convert-hint-1}
# logMAR = 1.88 - 0.049 * ETDRS
```

```{r ex-logmar-convert-solution}
etdrs_scores <- c(85, 70, 55, 35, 20)
tibble(
  ETDRS  = etdrs_scores,
  logMAR = round(1.88 - 0.049 * etdrs_scores, 3)
)
```

### Exercise 2.1 — Derive logMAR from ETDRS

Use `derive_var_aval_logmar()` on `adoe_bcva` to add a `LOGMAR` column, then show the range and mean at each visit for study eye, anti-VEGF arm.

```{r ex-logmar, exercise=TRUE, exercise.lines=20}
# derive_var_aval_logmar(dataset, aval_var, new_var)
adoe_bcva_lm <- derive_var_aval_logmar(
  dataset  = ___,
  aval_var = AVAL,
  new_var  = LOGMAR
)

# Mean logMAR at each visit, study eye, VEGF arm
adoe_bcva_lm |>
  filter(STUDYEYE == "Y", ARMCD == "VEGF") |>
  group_by(___) |>
  summarise(
    n           = n(),
    mean_logmar = round(mean(LOGMAR, na.rm = TRUE), 3),
    mean_etdrs  = round(mean(AVAL,   na.rm = TRUE), 1),
    .groups     = "drop"
  ) |>
  arrange(VISITNUM)
```

```{r ex-logmar-hint-1}
# dataset = adoe_bcva; group by VISIT (or VISITNUM for ordering)
adoe_bcva_lm <- derive_var_aval_logmar(dataset = adoe_bcva, aval_var = AVAL, new_var = LOGMAR)
adoe_bcva_lm |>
  filter(STUDYEYE == "Y", ARMCD == "VEGF") |>
  group_by(VISIT, VISITNUM) |>
  summarise(n = n(), mean_logmar = round(mean(LOGMAR,na.rm=TRUE),3),
            mean_etdrs = round(mean(AVAL,na.rm=TRUE),1), .groups = "drop") |>
  arrange(VISITNUM)
```

```{r ex-logmar-solution}
adoe_bcva_lm <- derive_var_aval_logmar(
  dataset  = adoe_bcva,
  aval_var = AVAL,
  new_var  = LOGMAR
)

adoe_bcva_lm |>
  filter(STUDYEYE == "Y", ARMCD == "VEGF") |>
  group_by(VISIT, VISITNUM) |>
  summarise(
    n           = n(),
    mean_logmar = round(mean(LOGMAR, na.rm = TRUE), 3),
    mean_etdrs  = round(mean(AVAL,   na.rm = TRUE), 1),
    .groups     = "drop"
  ) |>
  arrange(VISITNUM)
```

### Exercise 2.2 — Snellen conversion table

Build a reference table showing the Snellen fraction equivalent for each ETDRS letter score at 5-letter increments (0, 5, 10 … 100).

```{r ex-snellen-table, exercise=TRUE, exercise.lines=18}
# etdrs_to_logmar() and logmar_to_snellen() are available

snellen_ref <- tibble(
  ETDRS_letters = seq(0, 100, by = 5)
) |>
  mutate(
    logMAR  = etdrs_to_logmar(___),
    Snellen = logmar_to_snellen(___)
  )

# Show the clinically important range (40–90 letters)
snellen_ref |>
  filter(ETDRS_letters >= 40, ETDRS_letters <= 90)
```

```{r ex-snellen-hint-1}
# etdrs_to_logmar(ETDRS_letters); logmar_to_snellen(logMAR)
tibble(ETDRS_letters = seq(0, 100, by = 5)) |>
  mutate(logMAR = etdrs_to_logmar(ETDRS_letters),
         Snellen = logmar_to_snellen(logMAR)) |>
  filter(ETDRS_letters >= 40, ETDRS_letters <= 90)
```

```{r ex-snellen-solution}
snellen_ref <- tibble(
  ETDRS_letters = seq(0, 100, by = 5)
) |>
  mutate(
    logMAR  = etdrs_to_logmar(ETDRS_letters),
    Snellen = logmar_to_snellen(logMAR)
  )

snellen_ref |>
  filter(ETDRS_letters >= 40, ETDRS_letters <= 90)
```

---

## Section 3 — Change from Baseline and BCVA Categories

### Regulatory endpoint categories

For AMD and DME trials, the key BCVA endpoints are:

| Endpoint | Definition | AVAL |
|----------|-----------|------|
| Primary | Mean change from baseline in ETDRS letters at Week 52 | CHG |
| Responder (gain) | ≥15 letters gained | AVALC = "Gain >=15 letters" |
| Avoider (loss) | <15 letters lost | AVALC ≠ "Loss >=15 letters" |
| Legal blindness threshold | ≤35 letters (20/200 Snellen) | baseline categorisation |

```{r quiz-bcva-cat1, echo=FALSE}
quiz(
  caption = "Quiz — BCVA Change Categories",

  question("A subject's baseline BCVA is 62 letters; Week 52 BCVA is 74 letters. Their CHG category is:",
    answer("Gain >=15 letters"),
    answer("Gain 1-14 letters", correct = TRUE,
           message = "CHG = 74 − 62 = 12 letters gained. 12 is in the 1–14 range, not ≥15."),
    answer("No change"),
    answer("Loss 1-14 letters"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("The ≥15-letter gain threshold was chosen because:",
    answer("It equals exactly 3 lines on the ETDRS chart (5 letters × 3)",
           correct = TRUE,
           message = "3 lines of visual improvement is the consensus threshold for a clinically meaningful benefit in retinal disease trials."),
    answer("It equals 20/40 Snellen"),
    answer("FDA mandates this specific threshold for all visual acuity endpoints"),
    answer("It represents a 15% improvement in logMAR"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise 3.0 — Apply BCVA change categories manually

The five regulatory BCVA categories are defined in the table above. Use `case_when()` to apply them directly to `adoe_bcva` at Week 52.

```{r ex-bcva-cat-manual, exercise=TRUE, exercise.lines=18}
# BCVA change categories (CHG in ETDRS letters):
#   "Gain >=15 letters"  : CHG >= 15
#   "Gain 1-14 letters"  : CHG >= 1 & CHG < 15
#   "No change"          : CHG == 0
#   "Loss 1-14 letters"  : CHG > -15 & CHG < 0
#   "Loss >=15 letters"  : CHG <= -15

# Apply these manually to adoe_bcva at WEEK 52 (study eye, post-baseline).
adoe_bcva |>
  filter(VISIT == "WEEK 52", STUDYEYE == "Y", !is.na(AVAL)) |>
  mutate(
    CHG = AVAL - BASE,
    BCVACAT = case_when(
      CHG >= 15              ~ "Gain >=15 letters",
      CHG >= 1               ~ "Gain 1-14 letters",
      CHG == 0               ~ "No change",
      CHG > -15              ~ "Loss 1-14 letters",
      TRUE                   ~ "Loss >=15 letters"
    )
  ) |>
  count(___, BCVACAT) |>
  arrange(ARMCD, BCVACAT)
```

```{r ex-bcva-cat-manual-hint-1}
# count(ARMCD, BCVACAT)
```

```{r ex-bcva-cat-manual-solution}
adoe_bcva |>
  filter(VISIT == "WEEK 52", STUDYEYE == "Y", !is.na(AVAL)) |>
  mutate(
    CHG = AVAL - BASE,
    BCVACAT = case_when(
      CHG >= 15  ~ "Gain >=15 letters",
      CHG >= 1   ~ "Gain 1-14 letters",
      CHG == 0   ~ "No change",
      CHG > -15  ~ "Loss 1-14 letters",
      TRUE       ~ "Loss >=15 letters"
    )
  ) |>
  count(ARMCD, BCVACAT) |>
  arrange(ARMCD, BCVACAT)
```

### Exercise 3.1 — Derive CHG, PCHG, and BCVA category

Add `CHG`, `PCHG`, and the category variable `BCVA_CAT` to the study-eye BCVA data.

```{r ex-chg-cat, exercise=TRUE, exercise.lines=24}
# adoe_bcva has BASE already computed in the setup
bcva_study <- adoe_bcva |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(
    CHG  = ___ - ___,
    PCHG = round(100 * CHG / BASE, 1)
  )

bcva_study_cat <- derive_var_bcva_category(
  dataset = ___,
  chg_var = CHG,
  new_var = BCVA_CAT
)

# Distribution of categories at Week 52 by arm
bcva_study_cat |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD, ___) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = round(100 * n / sum(n), 1), .by = ARMCD) |>
  arrange(ARMCD, BCVA_CAT)
```

```{r ex-chg-cat-hint-1}
# CHG = AVAL - BASE; PCHG = 100 * CHG / BASE
# derive_var_bcva_category(dataset = bcva_study, chg_var = CHG, new_var = BCVA_CAT)
# group by ARMCD and BCVA_CAT
bcva_study <- adoe_bcva |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(CHG = AVAL - BASE, PCHG = round(100 * CHG / BASE, 1))
bcva_study_cat <- derive_var_bcva_category(dataset = bcva_study, chg_var = CHG, new_var = BCVA_CAT)
bcva_study_cat |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD, BCVA_CAT) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = round(100 * n / sum(n), 1), .by = ARMCD) |>
  arrange(ARMCD, BCVA_CAT)
```

```{r ex-chg-cat-solution}
bcva_study <- adoe_bcva |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(
    CHG  = AVAL - BASE,
    PCHG = round(100 * CHG / BASE, 1)
  )

bcva_study_cat <- derive_var_bcva_category(
  dataset = bcva_study,
  chg_var = CHG,
  new_var = BCVA_CAT
)

bcva_study_cat |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD, BCVA_CAT) |>
  summarise(n = n(), .groups = "drop") |>
  mutate(pct = round(100 * n / sum(n), 1), .by = ARMCD) |>
  arrange(ARMCD, BCVA_CAT)
```

### Exercise 3.2 — ≥15 letter gain responder rate over time

Compute the proportion of subjects gaining ≥15 letters at each post-baseline visit, separately by arm (study eye only).

```{r ex-responder-time, exercise=TRUE, exercise.lines=20, exercise.setup="ex-chg-cat-solution"}
# bcva_study_cat is available from the previous exercise

responder_over_time <- bcva_study_cat |>
  group_by(ARMCD, VISIT, VISITNUM) |>
  summarise(
    n             = n(),
    n_responders  = sum(___ == "Gain >=15 letters", na.rm = TRUE),
    pct_resp      = round(100 * n_responders / n, 1),
    .groups       = "drop"
  ) |>
  arrange(ARMCD, ___)

responder_over_time
```

```{r ex-responder-time-hint-1}
# sum(BCVA_CAT == "Gain >=15 letters"); arrange by VISITNUM
bcva_study_cat |>
  group_by(ARMCD, VISIT, VISITNUM) |>
  summarise(n = n(), n_responders = sum(BCVA_CAT == "Gain >=15 letters", na.rm = TRUE),
            pct_resp = round(100 * n_responders / n, 1), .groups = "drop") |>
  arrange(ARMCD, VISITNUM)
```

```{r ex-responder-time-solution}
responder_over_time <- bcva_study_cat |>
  group_by(ARMCD, VISIT, VISITNUM) |>
  summarise(
    n            = n(),
    n_responders = sum(BCVA_CAT == "Gain >=15 letters", na.rm = TRUE),
    pct_resp     = round(100 * n_responders / n, 1),
    .groups      = "drop"
  ) |>
  arrange(ARMCD, VISITNUM)

responder_over_time
```

### Exercise 3.3 — Study eye vs fellow eye mean BCVA change

`STUDYEYE = "Y"` marks the treated study eye; `"N"` marks the fellow (untreated) eye. Compare mean BCVA change at Week 52 between the two eyes, grouped by treatment arm.

```{r ex-study-eye-split, exercise=TRUE, exercise.lines=14}
# adoe_bcva has STUDYEYE = "Y" for the study eye and "N" for the fellow eye.
# Compute mean BCVA change (CHG) at WEEK 52 separately for
# study eye vs fellow eye, by treatment arm.

adoe_bcva |>
  filter(VISIT == "WEEK 52", !is.na(AVAL)) |>
  mutate(CHG = AVAL - BASE) |>
  group_by(ARMCD, ___) |>      # group by STUDYEYE
  summarise(
    n        = n(),
    mean_chg = round(mean(CHG, na.rm = TRUE), 1),
    .groups  = "drop"
  )
```

```{r ex-study-eye-split-hint-1}
# group_by(ARMCD, STUDYEYE)
```

```{r ex-study-eye-split-solution}
adoe_bcva |>
  filter(VISIT == "WEEK 52", !is.na(AVAL)) |>
  mutate(CHG = AVAL - BASE) |>
  group_by(ARMCD, STUDYEYE) |>
  summarise(
    n        = n(),
    mean_chg = round(mean(CHG, na.rm = TRUE), 1),
    .groups  = "drop"
  )
```

---

## Section 4 — Intraocular Pressure (IOP)

### IOP in ophthalmology trials

IOP (intraocular pressure, measured in mmHg) is a safety and secondary efficacy endpoint. Normal range is approximately 10–21 mmHg; sustained elevation (glaucoma) is defined as > 21 mmHg. Anti-VEGF injections can cause transient IOP spikes.

```{r quiz-iop1, echo=FALSE}
quiz(
  caption = "Quiz — IOP Analysis",

  question("IOP is typically measured in:",
    answer("μm (micrometres)"),
    answer("mmHg (millimetres of mercury)", correct = TRUE,
           message = "IOP measures the fluid pressure inside the eye using a tonometer. The clinical unit is mmHg."),
    answer("diopters"),
    answer("letters"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("In anti-VEGF trials, a clinically meaningful IOP elevation is typically defined as:",
    answer("Any increase from baseline"),
    answer("IOP > 21 mmHg at any post-injection visit, or increase ≥ 10 mmHg from baseline",
           correct = TRUE,
           message = "Both the absolute threshold (> 21 mmHg) and the change criterion (≥10 mmHg) are used. Transient spikes immediately post-injection are common and usually self-limiting."),
    answer("IOP > 30 mmHg at Week 52 only"),
    answer("Any IOP measurement outside the reference range"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise 4.1 — IOP elevation flags

Flag records where IOP > 21 mmHg OR CHG ≥ 10 mmHg from baseline, and report the proportion of eye-visits flagged by arm.

```{r ex-iop-flag, exercise=TRUE, exercise.lines=26}
# adoe_iop has AVAL (IOP mmHg), BASE, VISIT, OEOBJ, ARMCD, STUDYEYE

iop_flagged <- adoe_iop |>
  filter(VISIT != "BASELINE") |>
  mutate(
    CHG        = AVAL - BASE,
    IOP_ELEV   = (AVAL > ___) | (CHG >= ___)   # >21 mmHg OR >=10 change
  )

# Proportion of eye-visits with elevated IOP by arm
iop_flagged |>
  group_by(ARMCD, OEOBJ) |>
  summarise(
    n_eye_visits   = n(),
    n_elevated     = sum(IOP_ELEV, na.rm = TRUE),
    pct_elevated   = round(100 * n_elevated / n_eye_visits, 1),
    max_iop        = max(AVAL, na.rm = TRUE),
    .groups        = "drop"
  ) |>
  arrange(ARMCD, OEOBJ)
```

```{r ex-iop-flag-hint-1}
# IOP_ELEV = (AVAL > 21) | (CHG >= 10)
adoe_iop |>
  filter(VISIT != "BASELINE") |>
  mutate(CHG = AVAL - BASE, IOP_ELEV = (AVAL > 21) | (CHG >= 10)) |>
  group_by(ARMCD, OEOBJ) |>
  summarise(n_eye_visits = n(), n_elevated = sum(IOP_ELEV, na.rm = TRUE),
            pct_elevated = round(100 * n_elevated / n_eye_visits, 1),
            max_iop = max(AVAL, na.rm = TRUE), .groups = "drop") |>
  arrange(ARMCD, OEOBJ)
```

```{r ex-iop-flag-solution}
iop_flagged <- adoe_iop |>
  filter(VISIT != "BASELINE") |>
  mutate(
    CHG      = AVAL - BASE,
    IOP_ELEV = (AVAL > 21) | (CHG >= 10)
  )

iop_flagged |>
  group_by(ARMCD, OEOBJ) |>
  summarise(
    n_eye_visits = n(),
    n_elevated   = sum(IOP_ELEV, na.rm = TRUE),
    pct_elevated = round(100 * n_elevated / n_eye_visits, 1),
    max_iop      = max(AVAL, na.rm = TRUE),
    .groups      = "drop"
  ) |>
  arrange(ARMCD, OEOBJ)
```

---

## Section 5 — OCT: Central Retinal Thickness

### CRT and fluid resolution

**Central Retinal Thickness (CRT)** is measured by Optical Coherence Tomography (OCT). In AMD and DME, the macula accumulates fluid which thickens the retina. Anti-VEGF treatment reduces this fluid — CRT decreasing toward the normal range (< 300 μm) is a sign of treatment response.

```{r quiz-crt1, echo=FALSE}
quiz(
  caption = "Quiz — OCT / CRT",

  question("Normal central retinal thickness in a healthy adult is approximately:",
    answer("100–150 μm"),
    answer("250–300 μm", correct = TRUE,
           message = "Normal CRT ≈ 250–300 μm. In wet AMD or DME, CRT can exceed 400–600 μm due to subretinal or intraretinal fluid."),
    answer("500–600 μm"),
    answer("1000 μm — thicker is better for the retina"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("In anti-VEGF treated wet AMD, we expect CRT to:",
    answer("Increase as new vessels form"),
    answer("Decrease as VEGF-driven fluid is reabsorbed", correct = TRUE,
           message = "VEGF drives vascular leakage, causing fluid accumulation and CRT increase. Anti-VEGF blocks this, allowing fluid resolution and CRT reduction."),
    answer("Remain unchanged — CRT measures structure, not fluid"),
    answer("Increase initially then normalise over 2 years"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Exercise 5.1 — CRT change from baseline

Compute mean CRT CHG by arm and visit for the study eye. Identify which arm shows the greater reduction.

```{r ex-crt-chg, exercise=TRUE, exercise.lines=20}
# adoe_crt has AVAL (CRT in μm), BASE, VISIT, VISITNUM, OEOBJ, ARMCD, STUDYEYE

crt_summary <- adoe_crt |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(CHG = ___ - ___) |>
  group_by(___, VISIT, VISITNUM) |>
  summarise(
    n        = n(),
    mean_crt = round(mean(AVAL, na.rm = TRUE), 1),
    mean_chg = round(mean(CHG,  na.rm = TRUE), 1),
    .groups  = "drop"
  ) |>
  arrange(ARMCD, ___)

crt_summary
```

```{r ex-crt-chg-hint-1}
# CHG = AVAL - BASE; group by ARMCD and VISIT/VISITNUM; arrange by VISITNUM
adoe_crt |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(CHG = AVAL - BASE) |>
  group_by(ARMCD, VISIT, VISITNUM) |>
  summarise(n = n(), mean_crt = round(mean(AVAL,na.rm=TRUE),1),
            mean_chg = round(mean(CHG,na.rm=TRUE),1), .groups = "drop") |>
  arrange(ARMCD, VISITNUM)
```

```{r ex-crt-chg-solution}
crt_summary <- adoe_crt |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(CHG = AVAL - BASE) |>
  group_by(ARMCD, VISIT, VISITNUM) |>
  summarise(
    n        = n(),
    mean_crt = round(mean(AVAL, na.rm = TRUE), 1),
    mean_chg = round(mean(CHG,  na.rm = TRUE), 1),
    .groups  = "drop"
  ) |>
  arrange(ARMCD, VISITNUM)

crt_summary
```

### Exercise 5.2 — CRT fluid resolution flag (< 300 μm)

Derive a `DRY_RETINA` flag for post-baseline visits where CRT < 300 μm (fluid resolved), and compute the proportion of subjects with a dry retina at Week 52.

```{r ex-dry-retina, exercise=TRUE, exercise.lines=20}
dry_retina <- adoe_crt |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(DRY_RETINA = AVAL < ___) |>
  # Per-subject: was the retina ever dry at Week 52?
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD) |>
  summarise(
    n            = n(),
    n_dry        = sum(DRY_RETINA, na.rm = TRUE),
    pct_dry      = round(100 * n_dry / n, 1),
    mean_crt_w52 = round(mean(AVAL, na.rm = TRUE), 1),
    .groups      = "drop"
  )

dry_retina
```

```{r ex-dry-retina-hint-1}
# DRY_RETINA = AVAL < 300
adoe_crt |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(DRY_RETINA = AVAL < 300) |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD) |>
  summarise(n = n(), n_dry = sum(DRY_RETINA, na.rm = TRUE),
            pct_dry = round(100 * n_dry / n, 1),
            mean_crt_w52 = round(mean(AVAL, na.rm = TRUE), 1), .groups = "drop")
```

```{r ex-dry-retina-solution}
dry_retina <- adoe_crt |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(DRY_RETINA = AVAL < 300) |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD) |>
  summarise(
    n            = n(),
    n_dry        = sum(DRY_RETINA, na.rm = TRUE),
    pct_dry      = round(100 * n_dry / n, 1),
    mean_crt_w52 = round(mean(AVAL, na.rm = TRUE), 1),
    .groups      = "drop"
  )

dry_retina
```

---

## Section 6 — Capstone Problems

### Capstone 1 — Complete ADOE Assembly

**Scenario:** You are programming ADOE for an AMD NDA submission. The dataset must contain for the **study eye** (`STUDYEYE = "Y"`):

| PARAMCD | Description | DTYPE |
|---------|-------------|-------|
| `VAET_OD` / `VAET_OS` | Observed BCVA (letters) + CHG + PCHG | NA |
| `LOGMAR_OD` / `LOGMAR_OS` | logMAR transformed BCVA | NA (new derived column) |
| `RESP15` | ≥15-letter gain responder flag | `RESP` |
| `IOP_OD` / `IOP_OS` | Observed IOP + CHG | NA |
| `CRT_OD` / `CRT_OS` | Observed CRT + CHG | NA |

The final dataset must be sorted by `USUBJID`, `PARAMCD`, `VISITNUM`.

```{r capstone1, exercise=TRUE, exercise.lines=65}
# ── Step 1: BCVA with CHG, PCHG, logMAR ──────────────────────────────────────
adoe_bcva_c1 <- adoe_bcva |>
  filter(STUDYEYE == "Y") |>
  mutate(
    CHG   = AVAL - BASE,
    PCHG  = round(100 * CHG / BASE, 2)
  ) |>
  derive_var_aval_logmar(aval_var = AVAL, new_var = LOGMAR)

# ── Step 2: BCVA ≥15-letter responder (ever across all post-baseline visits) ─
adoe_resp <- derive_param_bcva_responder(
  dataset       = adoe_bcva_c1 |> filter(VISIT != "BASELINE"),
  dataset_adsl  = adsl_optha,
  by_vars       = c("USUBJID","PARAMCD"),
  chg_var       = CHG,
  threshold     = ___,
  set_values_to = list(
    PARAM   = "BCVA >=15 Letter Gain Responder",
    PARAMCD = "RESP15"
  )
)

# ── Step 3: IOP with CHG ───────────────────────────────────────────────────
adoe_iop_c1 <- adoe_iop |>
  filter(STUDYEYE == "Y") |>
  mutate(CHG = ___ - ___)

# ── Step 4: CRT with CHG ──────────────────────────────────────────────────
adoe_crt_c1 <- adoe_crt |>
  filter(STUDYEYE == "Y") |>
  mutate(CHG = ___ - ___)

# ── Step 5: Assemble ──────────────────────────────────────────────────────
adoe_final <- bind_rows(
  adoe_bcva_c1,
  adoe_resp   |> filter(DTYPE == "RESP"),
  adoe_iop_c1,
  adoe_crt_c1
) |>
  arrange(___, ___, ___)

# ── Validation ────────────────────────────────────────────────────────────
cat("Total rows:", nrow(adoe_final), "\n")
adoe_final |>
  count(OETESTCD, DTYPE) |>
  arrange(OETESTCD, DTYPE) |>
  print()
```

```{r capstone1-hint-1}
# threshold = 15
# Step 3: CHG = AVAL - BASE
# Step 4: CHG = AVAL - BASE
# arrange: USUBJID, PARAMCD, VISITNUM
```

```{r capstone1-solution}
adoe_bcva_c1 <- adoe_bcva |>
  filter(STUDYEYE == "Y") |>
  mutate(CHG  = AVAL - BASE,
         PCHG = round(100 * CHG / BASE, 2)) |>
  derive_var_aval_logmar(aval_var = AVAL, new_var = LOGMAR)

adoe_resp <- derive_param_bcva_responder(
  dataset       = adoe_bcva_c1 |> filter(VISIT != "BASELINE"),
  dataset_adsl  = adsl_optha,
  by_vars       = c("USUBJID","PARAMCD"),
  chg_var       = CHG,
  threshold     = 15,
  set_values_to = list(
    PARAM   = "BCVA >=15 Letter Gain Responder",
    PARAMCD = "RESP15"
  )
)

adoe_iop_c1 <- adoe_iop |>
  filter(STUDYEYE == "Y") |>
  mutate(CHG = AVAL - BASE)

adoe_crt_c1 <- adoe_crt |>
  filter(STUDYEYE == "Y") |>
  mutate(CHG = AVAL - BASE)

adoe_final <- bind_rows(
  adoe_bcva_c1,
  adoe_resp |> filter(DTYPE == "RESP"),
  adoe_iop_c1,
  adoe_crt_c1
) |>
  arrange(USUBJID, PARAMCD, VISITNUM)

cat("Total rows:", nrow(adoe_final), "\n")
adoe_final |>
  count(OETESTCD, DTYPE) |>
  arrange(OETESTCD, DTYPE) |>
  print()
```

### Capstone 2 — Ophthalmology Efficacy Summary Table

**Scenario:** The Data Review Committee needs an interim efficacy summary table for the study eye, containing:

1. Mean change from baseline in BCVA letters at Week 12, 24, 52 — by arm
2. % subjects gaining ≥15 letters at Week 52 — by arm
3. % subjects losing ≥15 letters at Week 52 — by arm (safety threshold)
4. Mean CRT change at Week 52 — by arm
5. Mean IOP at Week 52 and % with IOP > 21 mmHg — by arm (safety)

All in one tidy summary data frame with columns: `Endpoint`, `ARMCD`, `Value`.

```{r capstone2, exercise=TRUE, exercise.lines=70}
# ── CHG data already derived above ───────────────────────────────────────────
bcva_chg <- adoe_bcva |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(CHG = AVAL - BASE) |>
  derive_var_bcva_category(chg_var = CHG, new_var = BCVA_CAT)

# ── 1. Mean BCVA CHG at key visits ───────────────────────────────────────────
bcva_chg_tbl <- bcva_chg |>
  filter(VISIT %in% c("WEEK 12","WEEK 24","WEEK 52")) |>
  group_by(ARMCD, VISIT) |>
  summarise(
    Endpoint = paste0("Mean BCVA CHG (", VISIT, ")"),
    Value    = sprintf("%.1f letters (n=%d)", mean(CHG, na.rm=TRUE), n()),
    .groups  = "drop"
  )

# ── 2. ≥15-letter gain at Week 52 ────────────────────────────────────────────
gain15_tbl <- bcva_chg |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD) |>
  summarise(
    Endpoint = "Gained >=15 letters at Wk52",
    Value    = sprintf("%d/%d (%.1f%%)",
                       sum(BCVA_CAT == "Gain >=15 letters", na.rm=TRUE), n(),
                       100*mean(BCVA_CAT == "Gain >=15 letters", na.rm=TRUE)),
    .groups  = "drop"
  )

# ── 3. ≥15-letter loss at Week 52 (safety) ───────────────────────────────────
loss15_tbl <- bcva_chg |>
  filter(VISIT == "WEEK 52") |>
  group_by(___) |>
  summarise(
    Endpoint = "Lost >=15 letters at Wk52",
    Value    = sprintf("%d/%d (%.1f%%)",
                       sum(___ == "Loss >=15 letters", na.rm=TRUE), n(),
                       100*mean(___ == "Loss >=15 letters", na.rm=TRUE)),
    .groups  = "drop"
  )

# ── 4. CRT CHG at Week 52 ────────────────────────────────────────────────────
crt_tbl <- adoe_crt |>
  filter(STUDYEYE == "Y", VISIT == "WEEK 52") |>
  mutate(CHG = AVAL - BASE) |>
  group_by(___) |>
  summarise(
    Endpoint = "Mean CRT CHG at Wk52 (μm)",
    Value    = sprintf("%.1f (n=%d)", mean(CHG, na.rm=TRUE), n()),
    .groups  = "drop"
  )

# ── 5. IOP safety at Week 52 ─────────────────────────────────────────────────
iop_tbl <- adoe_iop |>
  filter(STUDYEYE == "Y", VISIT == "WEEK 52") |>
  group_by(___) |>
  summarise(
    Endpoint = "IOP >21 mmHg at Wk52",
    Value    = sprintf("%d/%d (%.1f%%)",
                       sum(AVAL > 21, na.rm=TRUE), n(),
                       100*mean(AVAL > 21, na.rm=TRUE)),
    .groups  = "drop"
  )

# ── Assemble ─────────────────────────────────────────────────────────────────
summary_tbl <- bind_rows(bcva_chg_tbl, gain15_tbl, loss15_tbl, crt_tbl, iop_tbl) |>
  select(Endpoint, ARMCD, Value) |>
  arrange(Endpoint, ARMCD)

print(summary_tbl, n = Inf)
```

```{r capstone2-hint-1}
# loss15_tbl: group by ARMCD; BCVA_CAT == "Loss >=15 letters"
# crt_tbl: group by ARMCD; mean(CHG)
# iop_tbl: group by ARMCD; sum(AVAL > 21)
```

```{r capstone2-solution}
bcva_chg <- adoe_bcva |>
  filter(STUDYEYE == "Y", VISIT != "BASELINE") |>
  mutate(CHG = AVAL - BASE) |>
  derive_var_bcva_category(chg_var = CHG, new_var = BCVA_CAT)

bcva_chg_tbl <- bcva_chg |>
  filter(VISIT %in% c("WEEK 12","WEEK 24","WEEK 52")) |>
  group_by(ARMCD, VISIT) |>
  summarise(Endpoint = paste0("Mean BCVA CHG (", VISIT, ")"),
            Value = sprintf("%.1f letters (n=%d)", mean(CHG,na.rm=TRUE), n()),
            .groups = "drop")

gain15_tbl <- bcva_chg |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD) |>
  summarise(Endpoint = "Gained >=15 letters at Wk52",
            Value = sprintf("%d/%d (%.1f%%)", sum(BCVA_CAT=="Gain >=15 letters",na.rm=TRUE),
                            n(), 100*mean(BCVA_CAT=="Gain >=15 letters",na.rm=TRUE)),
            .groups = "drop")

loss15_tbl <- bcva_chg |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD) |>
  summarise(Endpoint = "Lost >=15 letters at Wk52",
            Value = sprintf("%d/%d (%.1f%%)", sum(BCVA_CAT=="Loss >=15 letters",na.rm=TRUE),
                            n(), 100*mean(BCVA_CAT=="Loss >=15 letters",na.rm=TRUE)),
            .groups = "drop")

crt_tbl <- adoe_crt |>
  filter(STUDYEYE == "Y", VISIT == "WEEK 52") |>
  mutate(CHG = AVAL - BASE) |>
  group_by(ARMCD) |>
  summarise(Endpoint = "Mean CRT CHG at Wk52 (μm)",
            Value = sprintf("%.1f (n=%d)", mean(CHG,na.rm=TRUE), n()),
            .groups = "drop")

iop_tbl <- adoe_iop |>
  filter(STUDYEYE == "Y", VISIT == "WEEK 52") |>
  group_by(ARMCD) |>
  summarise(Endpoint = "IOP >21 mmHg at Wk52",
            Value = sprintf("%d/%d (%.1f%%)", sum(AVAL>21,na.rm=TRUE),
                            n(), 100*mean(AVAL>21,na.rm=TRUE)),
            .groups = "drop")

summary_tbl <- bind_rows(bcva_chg_tbl, gain15_tbl, loss15_tbl, crt_tbl, iop_tbl) |>
  select(Endpoint, ARMCD, Value) |>
  arrange(Endpoint, ARMCD)

print(summary_tbl, n = Inf)
```

---

## Section 7 — Hard Challenges

### Challenge 1 — Handling missing vision: LOCF for BCVA

**Scenario:** Per protocol, subjects who permanently lose light perception (NLP — No Light Perception) have subsequent BCVA imputed as 0. Subjects who miss a visit due to reasons other than NLP have their BCVA carried forward (LOCF). Implement both rules on `adoe_bcva` (study eye only).

```{r challenge-locf, exercise=TRUE, exercise.lines=44}
# Simulate some missing data: randomly set 8% of post-baseline BCVA to NA
set.seed(42)
bcva_with_missing <- adoe_bcva |>
  filter(STUDYEYE == "Y") |>
  mutate(
    AVAL_MISS = if_else(
      VISIT != "BASELINE" & runif(n()) < 0.08,
      NA_real_, AVAL
    )
  )

# LOCF imputation within subject (study eye)
# Rule: carry the last non-missing observation forward
bcva_locf <- bcva_with_missing |>
  arrange(USUBJID, VISITNUM) |>
  group_by(USUBJID) |>
  mutate(
    AVAL_LOCF = {
      x <- AVAL_MISS
      for (i in seq_along(x)) {
        if (is.na(x[i]) && i > 1) x[i] <- x[i - 1]
      }
      x
    },
    DTYPE = if_else(is.na(AVAL_MISS) & !is.na(AVAL_LOCF), "LOCF", NA_character_)
  ) |>
  ungroup()

# How many records were LOCF-imputed?
cat("LOCF-imputed records:", sum(bcva_locf$DTYPE == "LOCF", na.rm = TRUE), "\n")

# Compare mean BCVA at Week 52: observed vs LOCF-imputed
bcva_locf |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD) |>
  summarise(
    n_total     = n(),
    n_imputed   = sum(DTYPE == "LOCF", na.rm = TRUE),
    mean_obs    = round(mean(AVAL_MISS,  na.rm = TRUE), 1),
    mean_locf   = round(mean(___,        na.rm = TRUE), 1),
    .groups     = "drop"
  )
```

```{r challenge-locf-hint-1}
# mean_locf uses AVAL_LOCF (the imputed column)
# The LOCF loop carries x[i-1] forward when x[i] is NA
```

```{r challenge-locf-solution}
set.seed(42)
bcva_with_missing <- adoe_bcva |>
  filter(STUDYEYE == "Y") |>
  mutate(AVAL_MISS = if_else(VISIT != "BASELINE" & runif(n()) < 0.08, NA_real_, AVAL))

bcva_locf <- bcva_with_missing |>
  arrange(USUBJID, VISITNUM) |>
  group_by(USUBJID) |>
  mutate(
    AVAL_LOCF = {
      x <- AVAL_MISS
      for (i in seq_along(x)) if (is.na(x[i]) && i > 1) x[i] <- x[i - 1]
      x
    },
    DTYPE = if_else(is.na(AVAL_MISS) & !is.na(AVAL_LOCF), "LOCF", NA_character_)
  ) |>
  ungroup()

cat("LOCF-imputed records:", sum(bcva_locf$DTYPE == "LOCF", na.rm = TRUE), "\n")

bcva_locf |>
  filter(VISIT == "WEEK 52") |>
  group_by(ARMCD) |>
  summarise(n_total   = n(), n_imputed = sum(DTYPE == "LOCF", na.rm = TRUE),
            mean_obs  = round(mean(AVAL_MISS, na.rm = TRUE), 1),
            mean_locf = round(mean(AVAL_LOCF, na.rm = TRUE), 1), .groups = "drop")
```

### Challenge 2 — Fellow-eye correlation analysis

**Scenario:** A medical monitor asks whether the study eye BCVA improvement correlates with fellow-eye stability (fellow eye CHG ≈ 0 is expected). Build a paired dataset with both eyes' Week 52 CHG per subject, compute the correlation, and flag subjects where the fellow eye unexpectedly changed by ≥10 letters.

```{r challenge-fellow-eye, exercise=TRUE, exercise.lines=36}
# Build paired OD vs OS BCVA change at Week 52
paired_eyes <- adoe_bcva |>
  filter(VISIT == "WEEK 52") |>
  mutate(CHG = AVAL - BASE) |>
  select(USUBJID, OEOBJ, CHG, ARMCD, STDEYE) |>
  pivot_wider(
    names_from  = OEOBJ,
    values_from = CHG,
    names_prefix = "CHG_"
  ) |>
  mutate(
    STUDY_CHG  = if_else(STDEYE == "OD", CHG_OD, CHG_OS),
    FELLOW_CHG = if_else(STDEYE == "OD", CHG_OS, CHG_OD),
    FELLOW_ALERT = abs(___) >= ___    # fellow eye ≥10 letter change
  )

# Correlation between study and fellow eye CHG
cat("Pearson correlation (study vs fellow CHG at Wk52):",
    round(cor(paired_eyes$STUDY_CHG, paired_eyes$FELLOW_CHG, use="complete.obs"), 3), "\n")

# Subjects with unexpected fellow-eye change
cat("Subjects with |fellow CHG| >= 10 letters:",
    sum(paired_eyes$FELLOW_ALERT, na.rm = TRUE), "\n")

# Summarise by arm
paired_eyes |>
  group_by(ARMCD) |>
  summarise(
    n             = n(),
    mean_study    = round(mean(STUDY_CHG,  na.rm = TRUE), 1),
    mean_fellow   = round(mean(FELLOW_CHG, na.rm = TRUE), 1),
    n_alerts      = sum(FELLOW_ALERT, na.rm = TRUE),
    .groups       = "drop"
  )
```

```{r challenge-fellow-eye-hint-1}
# FELLOW_ALERT = abs(FELLOW_CHG) >= 10
# correlation: cor(STUDY_CHG, FELLOW_CHG, use = "complete.obs")
```

```{r challenge-fellow-eye-solution}
paired_eyes <- adoe_bcva |>
  filter(VISIT == "WEEK 52") |>
  mutate(CHG = AVAL - BASE) |>
  select(USUBJID, OEOBJ, CHG, ARMCD, STDEYE) |>
  pivot_wider(names_from = OEOBJ, values_from = CHG, names_prefix = "CHG_") |>
  mutate(
    STUDY_CHG    = if_else(STDEYE == "OD", CHG_OD, CHG_OS),
    FELLOW_CHG   = if_else(STDEYE == "OD", CHG_OS, CHG_OD),
    FELLOW_ALERT = abs(FELLOW_CHG) >= 10
  )

cat("Pearson correlation (study vs fellow CHG at Wk52):",
    round(cor(paired_eyes$STUDY_CHG, paired_eyes$FELLOW_CHG, use="complete.obs"), 3), "\n")
cat("Subjects with |fellow CHG| >= 10 letters:",
    sum(paired_eyes$FELLOW_ALERT, na.rm = TRUE), "\n")

paired_eyes |>
  group_by(ARMCD) |>
  summarise(n = n(), mean_study = round(mean(STUDY_CHG,na.rm=TRUE),1),
            mean_fellow = round(mean(FELLOW_CHG,na.rm=TRUE),1),
            n_alerts = sum(FELLOW_ALERT,na.rm=TRUE), .groups = "drop")
```

### Challenge 3 — Multi-scale BCVA shift table

**Final boss:** Produce a BCVA category shift table (baseline category × Week-52 category) for the anti-VEGF arm, study eye only. Baseline categories: ≤35 letters (legally blind), 36–69 letters (moderate vision loss), ≥70 letters (mild loss / near normal).

```{r challenge-shift, exercise=TRUE, exercise.lines=40}
# Step 1: Baseline category
bcva_baseline_cat <- adoe_bcva |>
  filter(VISIT == "BASELINE", STUDYEYE == "Y") |>
  mutate(
    BASE_CAT = case_when(
      AVAL <= 35               ~ "<=35 (legal blindness threshold)",
      AVAL <= 69               ~ "36-69 (moderate loss)",
      TRUE                     ~ ">=70 (near normal)"
    )
  ) |>
  select(USUBJID, BASE_CAT)

# Step 2: Week 52 category
bcva_wk52_cat <- adoe_bcva |>
  filter(VISIT == "WEEK 52", STUDYEYE == "Y") |>
  mutate(
    CHG = AVAL - BASE,
    WK52_CAT = case_when(
      AVAL <= 35  ~ "<=35",
      AVAL <= 69  ~ "36-69",
      TRUE        ~ ">=70"
    )
  ) |>
  select(USUBJID, ARMCD, AVAL, CHG, WK52_CAT)

# Step 3: Shift table for VEGF arm
shift_tbl <- ___ |>
  left_join(___, by = "USUBJID") |>
  filter(ARMCD == "VEGF") |>
  group_by(BASE_CAT, WK52_CAT) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = ___, values_from = n, values_fill = 0L) |>
  arrange(BASE_CAT)

print(shift_tbl)
```

```{r challenge-shift-hint-1}
# left_join bcva_wk52_cat with bcva_baseline_cat by USUBJID
# pivot_wider: names_from = WK52_CAT
```

```{r challenge-shift-solution}
bcva_baseline_cat <- adoe_bcva |>
  filter(VISIT == "BASELINE", STUDYEYE == "Y") |>
  mutate(BASE_CAT = case_when(
    AVAL <= 35 ~ "<=35 (legal blindness threshold)",
    AVAL <= 69 ~ "36-69 (moderate loss)",
    TRUE       ~ ">=70 (near normal)"
  )) |>
  select(USUBJID, BASE_CAT)

bcva_wk52_cat <- adoe_bcva |>
  filter(VISIT == "WEEK 52", STUDYEYE == "Y") |>
  mutate(CHG = AVAL - BASE,
         WK52_CAT = case_when(
           AVAL <= 35 ~ "<=35", AVAL <= 69 ~ "36-69", TRUE ~ ">=70"
         )) |>
  select(USUBJID, ARMCD, AVAL, CHG, WK52_CAT)

shift_tbl <- bcva_wk52_cat |>
  left_join(bcva_baseline_cat, by = "USUBJID") |>
  filter(ARMCD == "VEGF") |>
  group_by(BASE_CAT, WK52_CAT) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = WK52_CAT, values_from = n, values_fill = 0L) |>
  arrange(BASE_CAT)

print(shift_tbl)
```

---

## Lesson Summary

```{r quiz-final, echo=FALSE}
quiz(
  caption = "Final Knowledge Check",

  question("In ADOE, why does PARAMCD include the eye (e.g., 'VAET_OD') rather than storing laterality in a separate variable?",
    answer("Regulatory guidance requires it"),
    answer("It allows a single long dataset where each PARAMCD uniquely identifies a time series per subject, enabling standard BDS analysis functions to work without eye-specific modifications",
           correct = TRUE),
    answer("It reduces dataset size"),
    answer("The eye is always OD so a suffix is needed for the OS records only"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),

  question("In a wet AMD trial, an IOP > 21 mmHg post-injection is:",
    answer("Impossible — intravitreal injections lower IOP"),
    answer("A known complication: transient IOP spike from the injected volume, usually self-limiting",
           correct = TRUE,
           message = "Each intravitreal injection increases intraocular volume transiently, causing an IOP spike. Spikes typically resolve within 30 minutes. Sustained elevation warrants treatment."),
    answer("An automatic trial discontinuation criterion"),
    answer("Normal and expected — anti-VEGF always raises IOP"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Key takeaways

You have learned to:

- Understand the **laterality challenge**: PARAMCD encoding `OETESTCD_OEOBJ` (e.g., `VAET_OD`) keeps bilateral data in a single long dataset
- Identify and filter **study eye** vs **fellow eye** records using `STUDYEYE`
- Convert between **ETDRS letters**, **logMAR**, and **Snellen** fractions
- Derive `CHG`, `PCHG`, and the five **BCVA change categories** (≥15 gain, 1–14 gain, no change, 1–14 loss, ≥15 loss)
- Build **IOP elevation flags** and **CRT fluid-resolution flags**
- Assemble a complete **ADOE** with BCVA + logMAR + responder DTYPE + IOP + CRT
- Apply **LOCF imputation** for missing BCVA and perform **fellow-eye correlation** QC
- Build **multi-scale BCVA shift tables** for regulatory submission

**Next lesson:** Lesson 23 — `admiralpeds` — Paediatric ADaM derivations (age-adjusted parameters, growth charts, developmental milestones, weight-based dosing).
