---
title: "Lesson 24 — admiral Metabolic Disease Extension (admiralmetabolic): NASH, Glycaemia & Lipids"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Metabolic disease ADaM derivations using the admiralmetabolic package:
  liver histology endpoints for NASH/NAFLD (NAS score, fibrosis stage,
  histological response), glycaemic parameters (HbA1c, fasting glucose,
  HOMA-IR), lipid panel derivations (LDL-C via Friedewald equation, non-HDL-C,
  LDL/HDL ratio), composite metabolic syndrome flags, and cardiometabolic
  risk scoring. Includes NASH Phase IIb and T2DM capstone problems.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 24_admiralmetabolic.Rmd
# PURPOSE       : admiralmetabolic extension — metabolic disease ADaM: NASH/NAFLD liver
#                 histology (NAS score, fibrosis stage, histological
#                 response), HOMA-IR, Friedewald LDL-C, HbA1c
#                 categories, composite metabolic syndrome flags, and
#                 Hy's Law DILI signal detection.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(lubridate)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# SYNTHETIC METABOLIC TRIAL DATA — NASH Phase IIb
# Drug: semaglutide-like GLP-1 RA  (GLPx 2.4 mg s.c. weekly)  vs Placebo
# 120 adults with biopsy-confirmed NASH; 72-week treatment; biopsy at W0 & W72.
# Primary endpoint: ≥2-point NAS reduction with no worsening of fibrosis, OR
#                   fibrosis improvement ≥ 1 stage.
# ══════════════════════════════════════════════════════════════════════════════
set.seed(2025)
n_sub <- 120

adsl_meta <- tibble(
  STUDYID = "NASH2025",
  USUBJID = sprintf("NASH2025-%03d", seq_len(n_sub)),
  SUBJID  = sprintf("%03d", seq_len(n_sub)),
  SEX     = sample(c("M","F"), n_sub, replace = TRUE, prob = c(.55,.45)),
  RACE    = sample(c("WHITE","ASIAN","BLACK OR AFRICAN AMERICAN","OTHER"),
                   n_sub, replace = TRUE, prob = c(.50,.25,.15,.10)),
  ARM     = rep(c("GLPx 2.4 mg","Placebo"), each = n_sub/2),
  ARMCD   = rep(c("GLP","PBO"), each = n_sub/2),
  SAFFL   = "Y",
  ITTFL   = "Y"
) |>
  mutate(
    AGE       = sample(38:68, n_sub, replace = TRUE),
    # Baseline body weight (kg) — NASH subjects are typically overweight/obese
    WGTBL     = round(rnorm(n_sub, 98, 18), 1),
    HGTBL     = round(rnorm(n_sub, 170, 9), 1),
    BMIBL     = round(WGTBL / (HGTBL / 100)^2, 1),
    # Comorbidities
    DIABFL    = if_else(BMIBL >= 30 & sample(c(TRUE,FALSE), n_sub, replace = TRUE,
                                              prob = c(.60,.40)), "Y", "N"),
    HTNFL     = sample(c("Y","N"), n_sub, replace = TRUE, prob = c(.55,.45)),
    TRTSDT    = as.Date("2025-03-01") + sample(0:60, n_sub, replace = TRUE),
    TRTEDT    = TRTSDT + 504L  # 72 weeks
  )

# ── Visit schedule ────────────────────────────────────────────────────────────
meta_visits <- tibble(
  VISIT    = c("BASELINE","WEEK 12","WEEK 24","WEEK 48","WEEK 72"),
  VISITNUM = 1:5,
  VISITDY  = c(1L, 85L, 169L, 337L, 505L)
)

# ── Lab parameters ────────────────────────────────────────────────────────────
# Parameters: HbA1c (%), Fasting Glucose (mg/dL), Fasting Insulin (μIU/mL),
#             Total Cholesterol (mg/dL), LDL-C (mg/dL), HDL-C (mg/dL),
#             Triglycerides (mg/dL), ALT (U/L), AST (U/L), GGT (U/L)

lab_params <- tribble(
  ~PARAMCD, ~PARAM,                           ~PARAMU,  ~LBSTRESU,
  "HBA1C",  "Haemoglobin A1c",               "%",      "%",
  "GLUC",   "Fasting Glucose",               "mg/dL",  "mg/dL",
  "INS",    "Fasting Insulin",               "uIU/mL", "uIU/mL",
  "CHOL",   "Total Cholesterol",             "mg/dL",  "mg/dL",
  "LDL",    "LDL Cholesterol (calculated)",  "mg/dL",  "mg/dL",
  "HDLC",   "HDL Cholesterol",               "mg/dL",  "mg/dL",
  "TRIG",   "Triglycerides",                 "mg/dL",  "mg/dL",
  "ALT",    "Alanine Aminotransferase",      "U/L",    "U/L",
  "AST",    "Aspartate Aminotransferase",    "U/L",    "U/L",
  "GGT",    "Gamma-Glutamyl Transferase",    "U/L",    "U/L"
)

# Baseline lab values — plausible NASH/pre-diabetic population
gen_lab_base <- function(paramcd, arm, n) {
  switch(paramcd,
    HBA1C = round(rnorm(n, ifelse(arm=="GLP", 6.8, 6.7), 0.8), 1),
    GLUC  = round(rnorm(n, 112, 22)),
    INS   = round(rnorm(n, 18, 7), 1),
    CHOL  = round(rnorm(n, 205, 35)),
    LDL   = round(rnorm(n, 128, 30)),
    HDLC  = round(rnorm(n, 44, 9)),
    TRIG  = round(rnorm(n, 185, 55)),
    ALT   = round(rnorm(n, 68, 28)),
    AST   = round(rnorm(n, 55, 22)),
    GGT   = round(rnorm(n, 62, 32)),
    rep(NA_real_, n)
  )
}

# Treatment effect multipliers by visit (relative to baseline)
trt_fx <- function(paramcd, arm, visitdy) {
  base_mult <- switch(paramcd,
    HBA1C = if (arm=="GLP") -0.012 else -0.001,   # per day
    GLUC  = if (arm=="GLP") -0.08  else -0.005,
    INS   = if (arm=="GLP") -0.07  else -0.003,
    CHOL  = if (arm=="GLP") -0.04  else -0.001,
    LDL   = if (arm=="GLP") -0.035 else -0.001,
    HDLC  = if (arm=="GLP") +0.02  else +0.001,
    TRIG  = if (arm=="GLP") -0.05  else -0.003,
    ALT   = if (arm=="GLP") -0.06  else -0.002,
    AST   = if (arm=="GLP") -0.05  else -0.002,
    GGT   = if (arm=="GLP") -0.04  else -0.001,
    0
  )
  pmax(0.10, 1 + base_mult * visitdy)  # floor at 10% of baseline
}

# Build longitudinal lab dataset
lab_rows <- lapply(seq_len(n_sub), function(i) {
  subj <- adsl_meta$USUBJID[i]
  arm  <- adsl_meta$ARMCD[i]
  trtdt<- adsl_meta$TRTSDT[i]

  lapply(lab_params$PARAMCD, function(pc) {
    base_val <- gen_lab_base(pc, arm, 1)
    lapply(seq_len(nrow(meta_visits)), function(v) {
      vrow <- meta_visits[v, ]
      fx   <- trt_fx(pc, arm, vrow$VISITDY)
      val  <- round(base_val * fx + rnorm(1, 0, abs(base_val) * 0.04), 1)
      val  <- pmax(0.1, val)   # no negative lab values
      if (pc %in% c("HBA1C","INS")) val <- round(val, 1)
      else val <- round(val)
      tibble(
        STUDYID  = "NASH2025", USUBJID = subj,
        VISIT    = vrow$VISIT, VISITNUM = vrow$VISITNUM, VISITDY = vrow$VISITDY,
        PARAMCD  = pc,
        PARAM    = lab_params$PARAM[lab_params$PARAMCD == pc],
        PARAMU   = lab_params$PARAMU[lab_params$PARAMCD == pc],
        AVAL     = val,
        ADT      = trtdt + vrow$VISITDY - 1L,
        ADY      = vrow$VISITDY,
        ARMCD    = arm
      )
    }) |> bind_rows()
  }) |> bind_rows()
}) |> bind_rows()

# ── Derive BASE, ABLFL, CHG, PCHG ────────────────────────────────────────────
adlb_meta <- lab_rows |>
  group_by(USUBJID, PARAMCD) |>
  mutate(
    BASE  = AVAL[VISIT == "BASELINE"][1],
    ABLFL = if_else(VISIT == "BASELINE", "Y", NA_character_),
    CHG   = AVAL - BASE,
    PCHG  = round(100 * CHG / BASE, 2)
  ) |>
  ungroup()

# ── Body weight at each visit ─────────────────────────────────────────────────
wt_rows_meta <- lapply(seq_len(n_sub), function(i) {
  subj <- adsl_meta$USUBJID[i]
  arm  <- adsl_meta$ARMCD[i]
  bwt  <- adsl_meta$WGTBL[i]
  trtdt<- adsl_meta$TRTSDT[i]

  lapply(seq_len(nrow(meta_visits)), function(v) {
    vrow <- meta_visits[v, ]
    # GLP arm loses ~12% body weight over 72 wk; PBO ~2%
    wt_chg_pct <- if (arm=="GLP") -0.12 * vrow$VISITDY/505 else -0.02 * vrow$VISITDY/505
    wt <- round(bwt * (1 + wt_chg_pct) + rnorm(1, 0, 0.8), 1)
    tibble(STUDYID="NASH2025", USUBJID=subj, VISIT=vrow$VISIT,
           VISITNUM=vrow$VISITNUM, VISITDY=vrow$VISITDY,
           PARAMCD="WEIGHT", PARAM="Body Weight", PARAMU="kg",
           AVAL=wt, ADT=trtdt+vrow$VISITDY-1L, ADY=vrow$VISITDY, ARMCD=arm)
  }) |> bind_rows()
}) |> bind_rows()

wt_rows_meta <- wt_rows_meta |>
  group_by(USUBJID, PARAMCD) |>
  mutate(BASE=AVAL[VISIT=="BASELINE"][1], ABLFL=if_else(VISIT=="BASELINE","Y",NA_character_),
         CHG=AVAL-BASE, PCHG=round(100*CHG/BASE,2)) |>
  ungroup()

# ── Liver biopsy data (W0 and W72) ───────────────────────────────────────────
# NAS = NAFLD Activity Score: Steatosis (0-3) + Lobular Inflammation (0-3) +
#        Hepatocyte Ballooning (0-2).  Total 0-8.
# Fibrosis stage: 0 (none) to 4 (cirrhosis)
set.seed(2025)
biopsy_rows <- lapply(seq_len(n_sub), function(i) {
  subj <- adsl_meta$USUBJID[i]
  arm  <- adsl_meta$ARMCD[i]
  trtdt<- adsl_meta$TRTSDT[i]

  # Baseline scores (NASH patients have NAS ≥ 4)
  steat_b  <- sample(1:3, 1, prob=c(.20,.45,.35))
  inflam_b <- sample(1:3, 1, prob=c(.25,.45,.30))
  balloon_b<- sample(1:2, 1, prob=c(.35,.65))
  nas_b    <- steat_b + inflam_b + balloon_b
  fib_b    <- sample(1:3, 1, prob=c(.30,.40,.30))  # Stage 1-3 at baseline

  # W72 scores: GLP arm shows meaningful improvement
  delta_nas  <- if (arm=="GLP") sample(-3:0, 1, prob=c(.25,.35,.25,.15))
                else sample(-1:1, 1, prob=c(.15,.60,.25))
  delta_fib  <- if (arm=="GLP") sample(-1:0, 1, prob=c(.40,.60))
                else sample(-1:1, 1, prob=c(.10,.75,.15))

  steat_w72  <- pmax(0L, pmin(3L, as.integer(steat_b  + round(delta_nas*0.45))))
  inflam_w72 <- pmax(0L, pmin(3L, as.integer(inflam_b + round(delta_nas*0.35))))
  balloon_w72<- pmax(0L, pmin(2L, as.integer(balloon_b+ round(delta_nas*0.20))))
  nas_w72    <- steat_w72 + inflam_w72 + balloon_w72
  fib_w72    <- pmax(0L, pmin(4L, as.integer(fib_b + delta_fib)))

  bind_rows(
    tibble(STUDYID="NASH2025", USUBJID=subj, VISIT="BASELINE", VISITNUM=1L, VISITDY=1L,
           ADT=trtdt, ARMCD=arm,
           NAS_STEAT=steat_b, NAS_INFLAM=inflam_b, NAS_BALLOON=balloon_b,
           NAS_TOTAL=nas_b, FIB_STAGE=fib_b),
    tibble(STUDYID="NASH2025", USUBJID=subj, VISIT="WEEK 72", VISITNUM=5L, VISITDY=505L,
           ADT=trtdt+504L, ARMCD=arm,
           NAS_STEAT=steat_w72, NAS_INFLAM=inflam_w72, NAS_BALLOON=balloon_w72,
           NAS_TOTAL=nas_w72, FIB_STAGE=fib_w72)
  )
}) |> bind_rows()

# Wide biopsy (one row per subject with baseline and W72 columns)
biopsy_wide <- biopsy_rows |>
  pivot_wider(
    id_cols   = c(STUDYID, USUBJID, ARMCD),
    names_from = VISIT,
    values_from = c(NAS_STEAT, NAS_INFLAM, NAS_BALLOON, NAS_TOTAL, FIB_STAGE),
    names_sep  = "_"
  ) |>
  rename_with(~ gsub(" ", "_", .x))

# ══════════════════════════════════════════════════════════════════════════════
# ADMIRALMETABOLIC FUNCTION MOCKS
# ══════════════════════════════════════════════════════════════════════════════

#' Derive HOMA-IR from fasting glucose and insulin
#' Formula: HOMA-IR = (Glucose_mmol/L × Insulin_uIU/mL) / 22.5
#' Glucose conversion: mg/dL ÷ 18.016 = mmol/L
#'
#' @param dataset A data frame containing glucose and insulin values (wide format)
#' @param gluc_var Unquoted column name for fasting glucose (mg/dL)
#' @param ins_var  Unquoted column name for fasting insulin (μIU/mL)
#' @param new_var  Unquoted name for the output HOMA-IR variable
derive_var_homa_ir <- function(dataset,
                               gluc_var = GLUC,
                               ins_var  = INS,
                               new_var  = HOMAIR) {
  gluc_col <- deparse(substitute(gluc_var))
  ins_col  <- deparse(substitute(ins_var))
  out_col  <- deparse(substitute(new_var))
  dataset |>
    mutate(!!out_col := round(
      ((.data[[gluc_col]] / 18.016) * .data[[ins_col]]) / 22.5, 2
    ))
}

#' Derive LDL-C via the Friedewald equation
#' LDL = Total Cholesterol - HDL - Triglycerides/5
#' Valid only when Triglycerides < 400 mg/dL (set NA otherwise)
#'
#' @param dataset   A data frame containing CHOL, HDL, and TRIG columns
#' @param chol_var  Unquoted column name for Total Cholesterol
#' @param hdl_var   Unquoted column name for HDL-C
#' @param trig_var  Unquoted column name for Triglycerides
#' @param new_var   Unquoted output column name
#' @param trig_limit Upper TG limit for Friedewald validity (default 400)
derive_var_ldlc_friedewald <- function(dataset,
                                       chol_var   = CHOL,
                                       hdl_var    = HDLC,
                                       trig_var   = TRIG,
                                       new_var    = LDL_CALC,
                                       trig_limit = 400) {
  chol_col <- deparse(substitute(chol_var))
  hdl_col  <- deparse(substitute(hdl_var))
  trig_col <- deparse(substitute(trig_var))
  out_col  <- deparse(substitute(new_var))
  dataset |>
    mutate(!!out_col := if_else(
      .data[[trig_col]] < trig_limit,
      round(.data[[chol_col]] - .data[[hdl_col]] - .data[[trig_col]] / 5, 1),
      NA_real_
    ))
}

#' Derive Non-HDL Cholesterol
#' Non-HDL-C = Total Cholesterol - HDL-C
derive_var_non_hdlc <- function(dataset,
                                chol_var = CHOL,
                                hdl_var  = HDLC,
                                new_var  = NON_HDLC) {
  chol_col <- deparse(substitute(chol_var))
  hdl_col  <- deparse(substitute(hdl_var))
  out_col  <- deparse(substitute(new_var))
  dataset |>
    mutate(!!out_col := round(.data[[chol_col]] - .data[[hdl_col]], 1))
}

#' Derive NAS histological response flag
#' Responder: NAS reduction ≥ 2 points from baseline with no worsening of fibrosis
#' OR fibrosis improvement ≥ 1 stage (regardless of NAS change)
#'
#' @param dataset   A wide biopsy data frame (one row per subject)
#' @param nas_base  Unquoted column name: baseline NAS total
#' @param nas_post  Unquoted column name: post-treatment NAS total
#' @param fib_base  Unquoted column name: baseline fibrosis stage
#' @param fib_post  Unquoted column name: post-treatment fibrosis stage
#' @param new_var   Unquoted output flag variable ("Y"/"N")
derive_var_nash_responder <- function(dataset,
                                      nas_base = NAS_TOTAL_BASELINE,
                                      nas_post = `NAS_TOTAL_WEEK 72`,
                                      fib_base = FIB_STAGE_BASELINE,
                                      fib_post = `FIB_STAGE_WEEK 72`,
                                      new_var  = NASHRSFL) {
  nas_base_col <- deparse(substitute(nas_base))
  nas_post_col <- deparse(substitute(nas_post))
  fib_base_col <- deparse(substitute(fib_base))
  fib_post_col <- deparse(substitute(fib_post))
  out_col      <- deparse(substitute(new_var))

  dataset |>
    mutate(!!out_col := if_else(
      # Criterion 1: ≥2-point NAS drop with no fibrosis worsening
      (.data[[nas_base_col]] - .data[[nas_post_col]] >= 2 &
         .data[[fib_post_col]] >= .data[[fib_base_col]] - 1) |
        # Criterion 2: fibrosis improvement ≥1 stage
        (.data[[fib_base_col]] - .data[[fib_post_col]] >= 1),
      "Y", "N"
    ))
}

#' Derive fibrosis improvement flag (≥1 stage improvement)
derive_var_fib_improvement <- function(dataset,
                                       fib_base = FIB_STAGE_BASELINE,
                                       fib_post = `FIB_STAGE_WEEK 72`,
                                       new_var  = FIBIMPFL) {
  fib_base_col <- deparse(substitute(fib_base))
  fib_post_col <- deparse(substitute(fib_post))
  out_col      <- deparse(substitute(new_var))
  dataset |>
    mutate(!!out_col := if_else(
      .data[[fib_base_col]] - .data[[fib_post_col]] >= 1, "Y", "N"
    ))
}

#' Derive NAS resolution flag (NAS ≤ 1 at post-treatment biopsy)
derive_var_nas_resolution <- function(dataset,
                                      nas_post = `NAS_TOTAL_WEEK 72`,
                                      new_var  = NASRESFL) {
  nas_post_col <- deparse(substitute(nas_post))
  out_col      <- deparse(substitute(new_var))
  dataset |>
    mutate(!!out_col := if_else(.data[[nas_post_col]] <= 1L, "Y", "N"))
}

#' Derive metabolic syndrome flag
#' ATP-III criteria (3 of 5 required):
#'   1. Waist circumference (proxy: BMI ≥ 30 in this dataset)
#'   2. Triglycerides ≥ 150 mg/dL
#'   3. HDL-C < 40 (M) or < 50 (F)
#'   4. Blood pressure ≥ 130/85 (use HTNFL = "Y" as proxy)
#'   5. Fasting glucose ≥ 100 mg/dL
#'
#' @param dataset  A wide-format data frame with one row per subject/visit
#' @param bmi_var  BMI column
#' @param trig_var Triglycerides column
#' @param hdl_var  HDL-C column
#' @param gluc_var Fasting glucose column
#' @param htn_var  Hypertension flag column ("Y"/"N")
#' @param sex_var  Sex column ("M"/"F")
#' @param new_var  Output flag ("Y"/"N")
derive_var_metsyn <- function(dataset,
                               bmi_var  = BMIBL,
                               trig_var = TRIG_BASE,
                               hdl_var  = HDLC_BASE,
                               gluc_var = GLUC_BASE,
                               htn_var  = HTNFL,
                               sex_var  = SEX,
                               new_var  = METSYNFL) {
  bmi_col  <- deparse(substitute(bmi_var))
  trig_col <- deparse(substitute(trig_var))
  hdl_col  <- deparse(substitute(hdl_var))
  gluc_col <- deparse(substitute(gluc_var))
  htn_col  <- deparse(substitute(htn_var))
  sex_col  <- deparse(substitute(sex_var))
  out_col  <- deparse(substitute(new_var))

  dataset |>
    mutate(
      .crit_waist = .data[[bmi_col]] >= 30,
      .crit_trig  = .data[[trig_col]] >= 150,
      .crit_hdl   = if_else(.data[[sex_col]] == "M",
                            .data[[hdl_col]] < 40,
                            .data[[hdl_col]] < 50),
      .crit_bp    = .data[[htn_col]] == "Y",
      .crit_gluc  = .data[[gluc_col]] >= 100,
      .n_crit     = .crit_waist + .crit_trig + .crit_hdl + .crit_bp + .crit_gluc,
      !!out_col   := if_else(.n_crit >= 3L, "Y", "N")
    ) |>
    select(-starts_with(".crit_"), -.n_crit)
}

#' Derive percent weight loss category
#' Thresholds: < 0%, 0–<5%, 5–<10%, ≥10%
derive_var_wt_loss_cat <- function(dataset,
                                   pchg_var = PCHG,
                                   new_var  = WTLSSCAT) {
  pchg_col <- deparse(substitute(pchg_var))
  out_col  <- deparse(substitute(new_var))
  dataset |>
    mutate(!!out_col := case_when(
      .data[[pchg_col]] < 0     ~ "Weight loss",
      .data[[pchg_col]] < 5     ~ "0 to <5% loss",
      .data[[pchg_col]] < 10    ~ "5 to <10% loss",
      TRUE                      ~ ">=10% loss"
    ))
}

#' Derive HbA1c glycaemic response category (ADA 2024 targets)
#' < 5.7%: Normal; 5.7–<6.5%: Prediabetes; ≥6.5%: Diabetes threshold;
#' <7.0%: ADA target (for known T2DM); <8.0%: acceptable
derive_var_hba1c_cat <- function(dataset,
                                 hba1c_var = AVAL,
                                 new_var   = HBA1CCAT) {
  hba1c_col <- deparse(substitute(hba1c_var))
  out_col   <- deparse(substitute(new_var))
  dataset |>
    mutate(!!out_col := case_when(
      .data[[hba1c_col]] <  5.7 ~ "Normal (<5.7%)",
      .data[[hba1c_col]] <  6.5 ~ "Prediabetes (5.7-<6.5%)",
      .data[[hba1c_col]] <  7.0 ~ "Diabetes - At target (<7.0%)",
      .data[[hba1c_col]] <  8.0 ~ "Diabetes - Acceptable (7.0-<8.0%)",
      TRUE                      ~ "Diabetes - Above target (>=8.0%)"
    ))
}

#' Derive lipid panel shift: any lipid-lowering category threshold crossing
#' NCEP ATP-III LDL categories: <100 optimal, 100-129 near-optimal,
#' 130-159 borderline-high, 160-189 high, >=190 very high
derive_var_ldl_cat <- function(dataset,
                               ldl_var = AVAL,
                               new_var = LDLCAT) {
  ldl_col <- deparse(substitute(ldl_var))
  out_col <- deparse(substitute(new_var))
  dataset |>
    mutate(!!out_col := case_when(
      .data[[ldl_col]] < 100 ~ "Optimal (<100 mg/dL)",
      .data[[ldl_col]] < 130 ~ "Near-optimal (100-129 mg/dL)",
      .data[[ldl_col]] < 160 ~ "Borderline-high (130-159 mg/dL)",
      .data[[ldl_col]] < 190 ~ "High (160-189 mg/dL)",
      TRUE                   ~ "Very high (>=190 mg/dL)"
    ))
}

# ── Pre-derived helper objects for exercises ──────────────────────────────────

# Wide baseline labs (one row per subject, one column per parameter)
baseline_labs <- adlb_meta |>
  filter(VISIT == "BASELINE") |>
  select(USUBJID, PARAMCD, AVAL) |>
  pivot_wider(names_from = PARAMCD, values_from = AVAL) |>
  left_join(select(adsl_meta, USUBJID, SEX, ARMCD, BMIBL, HTNFL, DIABFL),
            by = "USUBJID") |>
  rename(CHOL_BASE=CHOL, LDL_BASE=LDL, HDLC_BASE=HDLC,
         TRIG_BASE=TRIG, GLUC_BASE=GLUC, INS_BASE=INS,
         HBA1C_BASE=HBA1C, ALT_BASE=ALT, AST_BASE=AST)

# W72 biopsy with baseline joined (for responder derivations)
biopsy_response <- biopsy_wide  # already has both BASELINE and WEEK 72 columns

# HbA1c over time (GLP arm only, for trend exercises)
hba1c_glp <- adlb_meta |>
  filter(PARAMCD == "HBA1C", ARMCD == "GLP")

# Lipid panel at all visits (wide per subject-visit)
lipid_wide <- adlb_meta |>
  filter(PARAMCD %in% c("CHOL","HDLC","TRIG","LDL")) |>
  select(USUBJID, VISIT, VISITNUM, VISITDY, PARAMCD, AVAL, ARMCD) |>
  pivot_wider(names_from=PARAMCD, values_from=AVAL)
```

## Overview

This lesson covers **admiralmetabolic** — the admiral extension package for metabolic disease trials including NASH/NAFLD, type 2 diabetes (T2DM), and cardiometabolic indications.

Metabolic trials are distinct from classic ADaM territory in several ways:

- **Liver histology** requires paired biopsy reading and multi-component composite endpoint logic (NAS score, fibrosis stage)
- **Glycaemic endpoints** (HbA1c, HOMA-IR, fasting glucose) have clinically established category thresholds from ADA/IDF guidelines
- **Lipid calculations** — particularly LDL-C via the Friedewald equation — are *derived* variables with an important validity constraint (TG < 400 mg/dL)
- **Composite metabolic syndrome** flags combine five distinct physiological criteria requiring careful coding logic
- **Weight loss** is often both a safety endpoint and a mediating variable for all other improvements

By the end of this lesson you will be able to: derive HOMA-IR, apply the Friedewald equation safely, compute NAS and fibrosis response flags, classify HbA1c and LDL-C into guideline categories, derive a composite metabolic syndrome flag, and build a complete ADLB and biopsy response dataset for a NASH trial.

### Trial Context

Our teaching dataset is a Phase IIb randomised controlled trial of **GLPx 2.4 mg** (a semaglutide-like GLP-1 RA, weekly subcutaneous injection) vs **Placebo** in 120 adults with biopsy-confirmed NASH (NAFLD Activity Score ≥ 4 and fibrosis stage 1–3).

The primary biopsy endpoint is assessed at **Week 72**; metabolic labs are collected at Weeks 0, 12, 24, 48, and 72.

---

## Section 1 — Insulin Resistance: HOMA-IR

**HOMA-IR** (Homeostatic Model Assessment of Insulin Resistance) is the standard surrogate for hepatic insulin resistance in NASH trials. It is derived from two simultaneously measured fasting values:

$$\text{HOMA-IR} = \frac{\text{Glucose (mmol/L)} \times \text{Insulin (μIU/mL)}}{22.5}$$

Because glucose is often reported in **mg/dL**, you must convert:

$$\text{Glucose (mmol/L)} = \frac{\text{Glucose (mg/dL)}}{18.016}$$

The `derive_var_homa_ir()` function handles both the unit conversion and the formula:

```r
derive_var_homa_ir(dataset, gluc_var = GLUC, ins_var = INS, new_var = HOMAIR)
```

### Practice — Apply the HOMA-IR Formula

```{r ex-homa-formula, exercise=TRUE, exercise.lines=10}
# HOMA-IR = (Glucose_mmol * Insulin_uIU/mL) / 22.5
# Glucose (mg/dL) -> mmol/L: divide by 18.016
# Compute HOMA-IR for a subject with glucose = 118 mg/dL and insulin = 22 uIU/mL.

gluc_mgdl <- 118
ins_uiu   <- 22

homa_ir <- (___ / 18.016 * ___) / 22.5
cat("HOMA-IR =", round(homa_ir, 2))
```

```{r ex-homa-formula-hint-1}
# homa_ir = (gluc_mgdl / 18.016 * ins_uiu) / 22.5
```

```{r ex-homa-formula-solution}
gluc_mgdl <- 118
ins_uiu   <- 22

homa_ir <- (gluc_mgdl / 18.016 * ins_uiu) / 22.5
cat("HOMA-IR =", round(homa_ir, 2))
```

### Quiz — HOMA-IR Formula

```{r homa-quiz-1, echo=FALSE}
quiz(caption = "HOMA-IR Fundamentals",
  question("A subject has fasting glucose = 126 mg/dL and fasting insulin = 15 μIU/mL. What is HOMA-IR (rounded to 2 dp)?",
    answer("7.87"),
    answer("4.66", correct = TRUE,
           message = "Correct! HOMA-IR = (126/18.016 × 15) / 22.5 = (6.994 × 15) / 22.5 ≈ 4.66."),
    answer("10.50"),
    answer("2.33"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question("Why does the Friedewald LDL equation become invalid when TG ≥ 400 mg/dL?",
    answer("The HDL assay is unreliable at high TG levels"),
    answer("VLDL-C is no longer well approximated by TG/5 at very high TG concentrations",
           correct = TRUE,
           message = "Friedewald assumes VLDL-C ≈ TG/5. At TG ≥ 400 mg/dL, VLDL particles change composition and this approximation breaks down, causing LDL to be underestimated."),
    answer("Total cholesterol measurement is affected by triglycerides"),
    answer("The formula was derived only for subjects with TG < 200 mg/dL"),
    allow_retry = TRUE
  )
)
```

### Practice — Derive HOMA-IR

```{r homa-derive, exercise=TRUE, exercise.lines=14}
# baseline_labs has one row per subject with columns:
# GLUC_BASE (mg/dL), INS_BASE (μIU/mL), plus SEX, ARMCD, BMIBL ...

# Step 1: Use derive_var_homa_ir() to compute HOMA-IR at baseline
# The function accepts gluc_var and ins_var arguments.
# Our columns are named GLUC_BASE and INS_BASE.

homa_baseline <- baseline_labs |>
  derive_var_homa_ir(
    gluc_var = ___,
    ins_var  = ___,
    new_var  = HOMAIR
  )

# Step 2: Report mean HOMA-IR by treatment arm
homa_baseline |>
  group_by(ARMCD) |>
  summarise(mean_HOMAIR = round(mean(HOMAIR, na.rm = TRUE), 2),
            n = n())
```

```{r homa-derive-hint-1}
# gluc_var should point to the baseline glucose column
# ins_var  should point to the baseline insulin column
```

```{r homa-derive-solution}
homa_baseline <- baseline_labs |>
  derive_var_homa_ir(
    gluc_var = GLUC_BASE,
    ins_var  = INS_BASE,
    new_var  = HOMAIR
  )

homa_baseline |>
  group_by(ARMCD) |>
  summarise(mean_HOMAIR = round(mean(HOMAIR, na.rm = TRUE), 2),
            n = n())
```

### Practice — Manual HOMA-IR and CHG

```{r homa-chg, exercise=TRUE, exercise.lines=18}
# adlb_meta has longitudinal lab values (long format).
# Create a wide-per-visit dataset for GLUC and INS,
# then compute HOMA-IR at each visit and derive CHG from baseline.

homa_wide <- adlb_meta |>
  filter(PARAMCD %in% c("GLUC","INS")) |>
  select(USUBJID, VISIT, VISITNUM, PARAMCD, AVAL, BASE, ARMCD) |>
  pivot_wider(names_from = PARAMCD, values_from = c(AVAL, BASE)) |>
  derive_var_homa_ir(
    gluc_var = ___,    # AVAL column for glucose (AVAL_GLUC)
    ins_var  = ___,    # AVAL column for insulin  (AVAL_INS)
    new_var  = HOMAIR
  ) |>
  mutate(
    HOMAIR_BASE = (___ / 18.016 * ___) / 22.5,   # baseline HOMA-IR
    HOMAIR_CHG  = round(HOMAIR - HOMAIR_BASE, 2)
  )

# Mean HOMA-IR change at Week 72 by arm
homa_wide |>
  filter(VISIT == "WEEK 72") |>
  group_by(ARMCD) |>
  summarise(mean_chg = round(mean(HOMAIR_CHG, na.rm=TRUE), 2))
```

```{r homa-chg-hint-1}
# After pivot_wider the glucose AVAL becomes AVAL_GLUC
# and insulin AVAL becomes AVAL_INS.
# Similarly the BASE columns become BASE_GLUC and BASE_INS.
```

```{r homa-chg-solution}
homa_wide <- adlb_meta |>
  filter(PARAMCD %in% c("GLUC","INS")) |>
  select(USUBJID, VISIT, VISITNUM, PARAMCD, AVAL, BASE, ARMCD) |>
  pivot_wider(names_from = PARAMCD, values_from = c(AVAL, BASE)) |>
  derive_var_homa_ir(
    gluc_var = AVAL_GLUC,
    ins_var  = AVAL_INS,
    new_var  = HOMAIR
  ) |>
  mutate(
    HOMAIR_BASE = (BASE_GLUC / 18.016 * BASE_INS) / 22.5,
    HOMAIR_CHG  = round(HOMAIR - HOMAIR_BASE, 2)
  )

homa_wide |>
  filter(VISIT == "WEEK 72") |>
  group_by(ARMCD) |>
  summarise(mean_chg = round(mean(HOMAIR_CHG, na.rm=TRUE), 2))
```

---

## Section 2 — Lipid Panel: Friedewald LDL-C

The **Friedewald equation** calculates LDL-C indirectly from three directly measured lipid values:

$$\text{LDL-C} = \text{Total Cholesterol} - \text{HDL-C} - \frac{\text{Triglycerides}}{5}$$

This approximation relies on VLDL-C ≈ TG/5. When TG ≥ 400 mg/dL, the approximation fails and LDL-C must be measured directly or flagged as missing.

> **Gotcha:** Always check `TRIG` before accepting a Friedewald LDL. The `derive_var_ldlc_friedewald()` function returns `NA` automatically when TG ≥ the specified limit.

**Non-HDL Cholesterol** = Total Cholesterol − HDL-C. This is an alternative lipid target that captures all atherogenic particles (LDL + VLDL + IDL) and does *not* require fasting.

### Practice — Manual Friedewald Calculation

```{r ex-friedewald-manual, exercise=TRUE, exercise.lines=14}
# Friedewald: LDL = CHOL - HDL - TG/5  (valid only when TG < 400)
# Apply this manually to lipid_wide at BASELINE.
# Set LDL_CALC = NA when TG >= 400.

lipid_wide |>
  filter(VISIT == "BASELINE") |>
  mutate(
    LDL_CALC = if_else(
      ___ < 400,
      round(CHOL - HDLC - ___ / 5, 1),
      NA_real_
    )
  ) |>
  select(USUBJID, CHOL, HDLC, TRIG, LDL_CALC, ARMCD) |>
  head(8)
```

```{r ex-friedewald-manual-hint-1}
# Condition: TRIG < 400
# Formula: CHOL - HDLC - TRIG / 5
```

```{r ex-friedewald-manual-solution}
lipid_wide |>
  filter(VISIT == "BASELINE") |>
  mutate(
    LDL_CALC = if_else(
      TRIG < 400,
      round(CHOL - HDLC - TRIG / 5, 1),
      NA_real_
    )
  ) |>
  select(USUBJID, CHOL, HDLC, TRIG, LDL_CALC, ARMCD) |>
  head(8)
```

### Practice — Derive LDL-C and Non-HDL-C

```{r ldl-derive, exercise=TRUE, exercise.lines=18}
# lipid_wide has columns: USUBJID, VISIT, VISITNUM, CHOL, HDLC, TRIG, LDL, ARMCD
# LDL is the "directly measured" value in our dataset.
# Derive LDL_CALC (Friedewald) and NON_HDLC, then compare LDL vs LDL_CALC.

lipid_derived <- lipid_wide |>
  derive_var_ldlc_friedewald(
    chol_var = ___,
    hdl_var  = ___,
    trig_var = ___,
    new_var  = LDL_CALC,
    trig_limit = 400
  ) |>
  derive_var_non_hdlc(
    chol_var = CHOL,
    hdl_var  = HDLC,
    new_var  = NON_HDLC
  )

# How many rows have TG >= 400 (Friedewald invalid)?
lipid_derived |>
  summarise(
    n_total       = n(),
    n_trig_hi     = sum(TRIG >= 400, na.rm = TRUE),
    n_ldl_missing = sum(is.na(LDL_CALC))
  )
```

```{r ldl-derive-hint-1}
# The column names in lipid_wide are CHOL, HDLC, TRIG (from pivot_wider).
```

```{r ldl-derive-solution}
lipid_derived <- lipid_wide |>
  derive_var_ldlc_friedewald(
    chol_var   = CHOL,
    hdl_var    = HDLC,
    trig_var   = TRIG,
    new_var    = LDL_CALC,
    trig_limit = 400
  ) |>
  derive_var_non_hdlc(
    chol_var = CHOL,
    hdl_var  = HDLC,
    new_var  = NON_HDLC
  )

lipid_derived |>
  summarise(
    n_total       = n(),
    n_trig_hi     = sum(TRIG >= 400, na.rm = TRUE),
    n_ldl_missing = sum(is.na(LDL_CALC))
  )
```

### Practice — LDL Category and Change

```{r ldl-cat, exercise=TRUE, exercise.lines=20}
# Classify LDL_CALC into NCEP ATP-III categories using derive_var_ldl_cat().
# Then compute the shift from baseline LDL category to Week 72 LDL category.

# Step 1: Classify
lipid_cat <- lipid_derived |>
  derive_var_ldl_cat(ldl_var = ___, new_var = LDLCAT)

# Step 2: Baseline and W72 categories side by side
ldl_shift <- lipid_cat |>
  filter(VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, LDLCAT, ARMCD) |>
  pivot_wider(names_from = VISIT, values_from = LDLCAT,
              names_repair = "universal") |>
  rename(LDLCAT_BASE = BASELINE, LDLCAT_W72 = WEEK.72)

# Count shift combinations by arm
ldl_shift |>
  count(ARMCD, LDLCAT_BASE, LDLCAT_W72, sort = TRUE)
```

```{r ldl-cat-hint-1}
# LDL_CALC is the Friedewald-derived variable.
# Use ldl_var = LDL_CALC in derive_var_ldl_cat().
```

```{r ldl-cat-solution}
lipid_cat <- lipid_derived |>
  derive_var_ldl_cat(ldl_var = LDL_CALC, new_var = LDLCAT)

ldl_shift <- lipid_cat |>
  filter(VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, LDLCAT, ARMCD) |>
  pivot_wider(names_from = VISIT, values_from = LDLCAT,
              names_repair = "universal") |>
  rename(LDLCAT_BASE = BASELINE, LDLCAT_W72 = WEEK.72)

ldl_shift |>
  count(ARMCD, LDLCAT_BASE, LDLCAT_W72, sort = TRUE)
```

---

## Section 3 — Glycaemic Endpoints: HbA1c

**HbA1c** (glycated haemoglobin) reflects average plasma glucose over ~90 days. It is the cornerstone endpoint for glycaemic control in T2DM and a key secondary endpoint in NASH trials (given the strong T2DM comorbidity).

ADA 2024 category thresholds:

| Range | Category |
|-------|----------|
| < 5.7% | Normal |
| 5.7–<6.5% | Prediabetes |
| ≥ 6.5% | Diabetes diagnostic threshold |
| < 7.0% | ADA glycaemic target (for known T2DM) |

A clinically meaningful HbA1c reduction is generally ≥ 0.5 percentage points.

### Practice — Manual HbA1c Categories

```{r ex-hba1c-manual, exercise=TRUE, exercise.lines=16}
# Apply ADA HbA1c categories manually using case_when.
# Thresholds: <5.7 Normal; 5.7-<6.5 Prediabetes; 6.5-<7.0 Diabetes-at-target;
#             7.0-<8.0 Diabetes-acceptable; >=8.0 Diabetes-above-target.

adlb_meta |>
  filter(PARAMCD == "HBA1C", VISIT == "WEEK 72") |>
  mutate(
    HBA1CCAT = case_when(
      AVAL <  5.7 ~ "Normal (<5.7%)",
      AVAL <  6.5 ~ ___,
      AVAL <  7.0 ~ ___,
      AVAL <  8.0 ~ "Diabetes - Acceptable (7.0-<8.0%)",
      TRUE        ~ "Diabetes - Above target (>=8.0%)"
    )
  ) |>
  count(ARMCD, HBA1CCAT) |>
  arrange(ARMCD, HBA1CCAT)
```

```{r ex-hba1c-manual-hint-1}
# 5.7-<6.5: "Prediabetes (5.7-<6.5%)"
# 6.5-<7.0: "Diabetes - At target (<7.0%)"
```

```{r ex-hba1c-manual-solution}
adlb_meta |>
  filter(PARAMCD == "HBA1C", VISIT == "WEEK 72") |>
  mutate(
    HBA1CCAT = case_when(
      AVAL < 5.7 ~ "Normal (<5.7%)",
      AVAL < 6.5 ~ "Prediabetes (5.7-<6.5%)",
      AVAL < 7.0 ~ "Diabetes - At target (<7.0%)",
      AVAL < 8.0 ~ "Diabetes - Acceptable (7.0-<8.0%)",
      TRUE       ~ "Diabetes - Above target (>=8.0%)"
    )
  ) |>
  count(ARMCD, HBA1CCAT) |>
  arrange(ARMCD, HBA1CCAT)
```

### Quiz — HbA1c

```{r hba1c-quiz, echo=FALSE}
quiz(caption = "HbA1c Endpoints",
  question("A subject with T2DM achieves HbA1c = 6.8% at Week 72 vs 8.2% at baseline. Which statement is TRUE?",
    answer("The subject has achieved the ADA target of < 7.0%", correct = TRUE),
    answer("The subject is now classified as prediabetes"),
    answer("The reduction of 1.4 pp is not considered clinically meaningful"),
    answer("HbA1c < 7.0% means T2DM has resolved"),
    allow_retry = TRUE
  ),
  question("In NASH trials, why is HbA1c collected even when patients do not have diabetes?",
    answer("Regulatory requirement for all Phase IIb trials"),
    answer("NASH and insulin resistance are tightly linked; HbA1c tracks metabolic improvement",
           correct = TRUE),
    answer("HbA1c predicts liver fibrosis stage"),
    answer("To ensure blinding is maintained"),
    allow_retry = TRUE
  )
)
```

### Practice — HbA1c Categories and Responders

```{r hba1c-cat, exercise=TRUE, exercise.lines=20}
# adlb_meta filtered to PARAMCD == "HBA1C" contains longitudinal HbA1c.
# 1. Classify each observation into ADA category using derive_var_hba1c_cat().
# 2. Flag subjects who achieve HbA1c < 7.0% at Week 72 (among DIABFL == "Y").

hba1c_data <- adlb_meta |>
  filter(PARAMCD == "HBA1C") |>
  derive_var_hba1c_cat(hba1c_var = ___, new_var = HBA1CCAT)

# Proportion in each category at baseline vs Week 72 by arm
hba1c_data |>
  filter(VISIT %in% c("BASELINE","WEEK 72")) |>
  count(ARMCD, VISIT, HBA1CCAT) |>
  group_by(ARMCD, VISIT) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

```{r hba1c-cat-hint-1}
# The AVAL column contains the HbA1c value — pass hba1c_var = AVAL.
```

```{r hba1c-cat-solution}
hba1c_data <- adlb_meta |>
  filter(PARAMCD == "HBA1C") |>
  derive_var_hba1c_cat(hba1c_var = AVAL, new_var = HBA1CCAT)

hba1c_data |>
  filter(VISIT %in% c("BASELINE","WEEK 72")) |>
  count(ARMCD, VISIT, HBA1CCAT) |>
  group_by(ARMCD, VISIT) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

---

## Section 4 — Liver Histology: NAS Score and Fibrosis Stage

The **NAFLD Activity Score (NAS)** is the primary histological endpoint in NASH trials. It is a sum of three independent components scored by a pathologist:

| Component | Range | Description |
|-----------|-------|-------------|
| Steatosis | 0–3 | < 5% / 5–33% / 33–66% / > 66% hepatocyte involvement |
| Lobular inflammation | 0–3 | Foci per 20× field |
| Hepatocyte ballooning | 0–2 | None / few / many |
| **NAS Total** | **0–8** | ≥ 4 required for NASH diagnosis |

**Fibrosis stage** (Kleiner system) is a *separate* but equally important endpoint:

| Stage | Description |
|-------|-------------|
| 0 | No fibrosis |
| 1 | Perisinusoidal or periportal fibrosis |
| 2 | Perisinusoidal + periportal fibrosis |
| 3 | Bridging fibrosis |
| 4 | Cirrhosis |

> **Key concept:** NAS and fibrosis stage are *not* the same axis. Fibrosis tends to improve more slowly than NAS. The primary composite endpoint typically requires *both* NAS improvement *and* fibrosis non-worsening (or fibrosis improvement alone).

### Practice — Verify NAS Composite Score

```{r ex-nas-composite, exercise=TRUE, exercise.lines=16}
# NAS = Steatosis (0-3) + Lobular Inflammation (0-3) + Ballooning (0-2)
# biopsy_rows (long format) has all three component scores at BASELINE and WEEK 72.
# Verify NAS_TOTAL is correctly computed as the sum of components.

biopsy_rows |>
  mutate(
    NAS_CHECK = ___ + ___ + ___   # Steat + Inflam + Balloon
  ) |>
  mutate(NAS_MATCH = NAS_TOTAL == NAS_CHECK) |>
  summarise(
    all_match = all(NAS_MATCH),
    n_rows    = n()
  )
```

```{r ex-nas-composite-hint-1}
# NAS_CHECK = NAS_STEAT + NAS_INFLAM + NAS_BALLOON
```

```{r ex-nas-composite-solution}
biopsy_rows |>
  mutate(NAS_CHECK = NAS_STEAT + NAS_INFLAM + NAS_BALLOON) |>
  mutate(NAS_MATCH = NAS_TOTAL == NAS_CHECK) |>
  summarise(
    all_match = all(NAS_MATCH),
    n_rows    = n()
  )
```

### Quiz — Histological Endpoints

```{r biopsy-quiz, echo=FALSE}
quiz(caption = "NASH Histology",
  question("A subject has NAS 6 at baseline (steatosis 2, inflammation 2, ballooning 2) and NAS 3 at Week 72 (steatosis 1, inflammation 1, ballooning 1). Fibrosis stage goes from 2 → 2. Are they a primary endpoint responder?",
    answer("Yes — NAS dropped ≥ 2 points with no fibrosis worsening",
           correct = TRUE,
           message = "NAS change = 6−3 = 3 ≥ 2, and fibrosis did not worsen (2→2 = no change). This satisfies criterion 1."),
    answer("No — fibrosis did not improve"),
    answer("No — ballooning must reach 0 for NASH resolution"),
    answer("Yes, but only if NAS drops to ≤ 2"),
    allow_retry = TRUE
  ),
  question("Which two biopsy timepoints are standard in Phase IIb NASH trials?",
    answer("Screening and Week 24"),
    answer("Baseline (Week 0) and end-of-treatment (e.g., Week 72)",
           correct = TRUE),
    answer("Week 12 and Week 52"),
    answer("Screening and Week 52"),
    allow_retry = TRUE
  )
)
```

### Practice — NAS Response Flag

```{r nash-resp, exercise=TRUE, exercise.lines=20}
# biopsy_response is a wide data frame with one row per subject.
# Columns: USUBJID, ARMCD,
#          NAS_TOTAL_BASELINE, NAS_TOTAL_WEEK 72,
#          FIB_STAGE_BASELINE, FIB_STAGE_WEEK 72

# Use derive_var_nash_responder() to create NASHRSFL.
# Then use derive_var_fib_improvement() to create FIBIMPFL.
# Report response rates by arm.

biopsy_flagged <- biopsy_response |>
  derive_var_nash_responder(
    nas_base = ___,
    nas_post = `NAS_TOTAL_WEEK 72`,
    fib_base = ___,
    fib_post = `FIB_STAGE_WEEK 72`,
    new_var  = NASHRSFL
  ) |>
  derive_var_fib_improvement(
    fib_base = ___,
    fib_post = `FIB_STAGE_WEEK 72`,
    new_var  = FIBIMPFL
  )

biopsy_flagged |>
  group_by(ARMCD) |>
  summarise(
    n_total   = n(),
    nash_resp = sum(NASHRSFL == "Y"),
    fib_impr  = sum(FIBIMPFL == "Y"),
    pct_nash  = round(100 * nash_resp / n_total, 1),
    pct_fib   = round(100 * fib_impr  / n_total, 1)
  )
```

```{r nash-resp-hint-1}
# Column names with spaces need backtick quoting: `NAS_TOTAL_WEEK 72`
# For nas_base and fib_base use the BASELINE-suffix columns.
```

```{r nash-resp-solution}
biopsy_flagged <- biopsy_response |>
  derive_var_nash_responder(
    nas_base = NAS_TOTAL_BASELINE,
    nas_post = `NAS_TOTAL_WEEK 72`,
    fib_base = FIB_STAGE_BASELINE,
    fib_post = `FIB_STAGE_WEEK 72`,
    new_var  = NASHRSFL
  ) |>
  derive_var_fib_improvement(
    fib_base = FIB_STAGE_BASELINE,
    fib_post = `FIB_STAGE_WEEK 72`,
    new_var  = FIBIMPFL
  )

biopsy_flagged |>
  group_by(ARMCD) |>
  summarise(
    n_total   = n(),
    nash_resp = sum(NASHRSFL == "Y"),
    fib_impr  = sum(FIBIMPFL == "Y"),
    pct_nash  = round(100 * nash_resp / n_total, 1),
    pct_fib   = round(100 * fib_impr  / n_total, 1)
  )
```

### Practice — NAS Resolution

```{r nas-resol, exercise=TRUE, exercise.lines=12}
# NAS resolution is defined as post-treatment NAS ≤ 1.
# Use derive_var_nas_resolution() on biopsy_response.
# Compare NAS resolution rate across arms.

biopsy_resol <- biopsy_response |>
  derive_var_nas_resolution(
    nas_post = `NAS_TOTAL_WEEK 72`,
    new_var  = NASRESFL
  )

biopsy_resol |>
  group_by(ARMCD) |>
  summarise(
    n = n(),
    n_resolved = sum(NASRESFL == "Y"),
    pct = round(100 * n_resolved / n, 1)
  )
```

```{r nas-resol-solution}
biopsy_resol <- biopsy_response |>
  derive_var_nas_resolution(
    nas_post = `NAS_TOTAL_WEEK 72`,
    new_var  = NASRESFL
  )

biopsy_resol |>
  group_by(ARMCD) |>
  summarise(
    n = n(),
    n_resolved = sum(NASRESFL == "Y"),
    pct = round(100 * n_resolved / n, 1)
  )
```

---

## Section 5 — Metabolic Syndrome and Weight Loss

### Composite Metabolic Syndrome Flag

ATP-III defines **metabolic syndrome** as ≥ 3 of these 5 criteria present simultaneously:

1. Abdominal obesity: waist circumference > 102 cm (M) / > 88 cm (F) — proxied here by BMI ≥ 30
2. Triglycerides ≥ 150 mg/dL
3. HDL-C < 40 mg/dL (M) / < 50 mg/dL (F)
4. Blood pressure ≥ 130/85 mmHg (proxied by `HTNFL = "Y"`)
5. Fasting glucose ≥ 100 mg/dL

The function `derive_var_metsyn()` evaluates all five and sets the flag.

### Practice — Manual Metabolic Syndrome Flag

```{r ex-metsyn-manual, exercise=TRUE, exercise.lines=22}
# ATP-III metabolic syndrome: 3 of 5 criteria present.
# Criteria in baseline_labs:
#   1. BMIBL >= 30 (abdominal obesity proxy)
#   2. TRIG_BASE >= 150
#   3. HDLC_BASE < 40 (M) or < 50 (F)
#   4. HTNFL == "Y"
#   5. GLUC_BASE >= 100

baseline_labs |>
  mutate(
    c_obesity = BMIBL >= 30,
    c_trig    = TRIG_BASE >= 150,
    c_hdl     = if_else(SEX == "M", HDLC_BASE < 40, HDLC_BASE < 50),
    c_bp      = HTNFL == "Y",
    c_gluc    = ___ >= 100,
    N_CRIT    = c_obesity + c_trig + c_hdl + c_bp + c_gluc,
    METSYNFL  = if_else(N_CRIT >= ___, "Y", "N")
  ) |>
  count(ARMCD, METSYNFL) |>
  group_by(ARMCD) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

```{r ex-metsyn-manual-hint-1}
# c_gluc = GLUC_BASE >= 100
# METSYNFL: N_CRIT >= 3
```

```{r ex-metsyn-manual-solution}
baseline_labs |>
  mutate(
    c_obesity = BMIBL >= 30,
    c_trig    = TRIG_BASE >= 150,
    c_hdl     = if_else(SEX == "M", HDLC_BASE < 40, HDLC_BASE < 50),
    c_bp      = HTNFL == "Y",
    c_gluc    = GLUC_BASE >= 100,
    N_CRIT    = c_obesity + c_trig + c_hdl + c_bp + c_gluc,
    METSYNFL  = if_else(N_CRIT >= 3, "Y", "N")
  ) |>
  count(ARMCD, METSYNFL) |>
  group_by(ARMCD) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

### Practice — Metabolic Syndrome Flag

```{r metsyn-derive, exercise=TRUE, exercise.lines=20}
# baseline_labs has: BMIBL, TRIG_BASE, HDLC_BASE, GLUC_BASE, HTNFL, SEX, ARMCD

metsyn_data <- baseline_labs |>
  derive_var_metsyn(
    bmi_var  = ___,
    trig_var = ___,
    hdl_var  = ___,
    gluc_var = ___,
    htn_var  = ___,
    sex_var  = SEX,
    new_var  = METSYNFL
  )

# What proportion of subjects meet metabolic syndrome criteria?
metsyn_data |>
  group_by(ARMCD) |>
  summarise(
    n_total  = n(),
    n_metsyn = sum(METSYNFL == "Y"),
    pct      = round(100 * n_metsyn / n_total, 1)
  )
```

```{r metsyn-derive-hint-1}
# Check the column names in baseline_labs:
# BMIBL, TRIG_BASE, HDLC_BASE, GLUC_BASE, HTNFL
```

```{r metsyn-derive-solution}
metsyn_data <- baseline_labs |>
  derive_var_metsyn(
    bmi_var  = BMIBL,
    trig_var = TRIG_BASE,
    hdl_var  = HDLC_BASE,
    gluc_var = GLUC_BASE,
    htn_var  = HTNFL,
    sex_var  = SEX,
    new_var  = METSYNFL
  )

metsyn_data |>
  group_by(ARMCD) |>
  summarise(
    n_total  = n(),
    n_metsyn = sum(METSYNFL == "Y"),
    pct      = round(100 * n_metsyn / n_total, 1)
  )
```

### Practice — Weight Loss Categories

Weight loss is both a **primary mediator** (GLP-1 RAs reduce liver fat via weight reduction) and a **safety signal** (excessive weight loss in cirrhotic patients is a concern).

```{r wt-loss-cat, exercise=TRUE, exercise.lines=20}
# wt_rows_meta has PARAMCD == "WEIGHT", with CHG and PCHG already derived.
# The PCHG column is negative for weight loss (convention: PCHG = (post−base)/base × 100).
# Flip sign so positive = loss.

wt_final <- wt_rows_meta |>
  filter(VISIT == "WEEK 72") |>
  mutate(PCHG_LOSS = -PCHG) |>   # positive = weight loss
  derive_var_wt_loss_cat(pchg_var = ___, new_var = WTLSSCAT)

wt_final |>
  count(ARMCD, WTLSSCAT) |>
  group_by(ARMCD) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  arrange(ARMCD, WTLSSCAT)
```

```{r wt-loss-cat-hint-1}
# PCHG_LOSS is positive for subjects who lost weight (i.e., -PCHG when PCHG < 0).
# Pass pchg_var = PCHG_LOSS.
```

```{r wt-loss-cat-solution}
wt_final <- wt_rows_meta |>
  filter(VISIT == "WEEK 72") |>
  mutate(PCHG_LOSS = -PCHG) |>
  derive_var_wt_loss_cat(pchg_var = PCHG_LOSS, new_var = WTLSSCAT)

wt_final |>
  count(ARMCD, WTLSSCAT) |>
  group_by(ARMCD) |>
  mutate(pct = round(100 * n / sum(n), 1)) |>
  arrange(ARMCD, WTLSSCAT)
```

---

## Section 6 — Liver Enzyme Trends (ALT/AST/GGT)

In NASH trials, liver enzyme normalisation is a secondary endpoint. ALT and AST elevation is a marker of hepatocyte injury; GGT reflects biliary and oxidative stress. Normalisation thresholds vary by sex:

| Analyte | Male ULN | Female ULN |
|---------|----------|------------|
| ALT | 40 U/L | 35 U/L |
| AST | 40 U/L | 35 U/L |
| GGT | 55 U/L | 38 U/L |

### Practice — ALT Normalisation Rate

```{r alt-norm, exercise=TRUE, exercise.lines=22}
# adlb_meta contains ALT values over time.
# Join SEX from adsl_meta, then flag observations where ALT is within normal limits (WNL):
#   ALTWFL = "Y" if ALT <= ULN for sex.
# Report % WNL at baseline and Week 72 by arm.

alt_uln <- tibble(SEX = c("M","F"), ALT_ULN = c(40L, 35L))

alt_norm <- adlb_meta |>
  filter(PARAMCD == "ALT") |>
  left_join(select(adsl_meta, USUBJID, SEX), by = "USUBJID") |>
  left_join(___, by = "SEX") |>     # join ULN table
  mutate(ALTWFL = if_else(___ <= ___, "Y", "N"))  # ALT <= ULN?

alt_norm |>
  filter(VISIT %in% c("BASELINE","WEEK 72")) |>
  group_by(ARMCD, VISIT) |>
  summarise(
    n_wnl = sum(ALTWFL == "Y"),
    n     = n(),
    pct   = round(100 * n_wnl / n, 1),
    .groups = "drop"
  )
```

```{r alt-norm-hint-1}
# left_join(alt_uln, by = "SEX") to attach ALT_ULN.
# Then: ALTWFL = if_else(AVAL <= ALT_ULN, "Y", "N")
```

```{r alt-norm-solution}
alt_uln <- tibble(SEX = c("M","F"), ALT_ULN = c(40L, 35L))

alt_norm <- adlb_meta |>
  filter(PARAMCD == "ALT") |>
  left_join(select(adsl_meta, USUBJID, SEX), by = "USUBJID") |>
  left_join(alt_uln, by = "SEX") |>
  mutate(ALTWFL = if_else(AVAL <= ALT_ULN, "Y", "N"))

alt_norm |>
  filter(VISIT %in% c("BASELINE","WEEK 72")) |>
  group_by(ARMCD, VISIT) |>
  summarise(
    n_wnl = sum(ALTWFL == "Y"),
    n     = n(),
    pct   = round(100 * n_wnl / n, 1),
    .groups = "drop"
  )
```

### Hard Challenge — AST/ALT Ratio (De Ritis Ratio)

The **De Ritis ratio** (AST/ALT) provides additional histological information. Values > 1 suggest more advanced fibrosis or alcoholic hepatitis; values < 1 are typical of NASH and viral hepatitis.

```{r deritis, exercise=TRUE, exercise.lines=24}
# Compute AST/ALT ratio at each visit for each subject.
# Flag subjects whose ratio crosses from <1 to >=1 between baseline and Week 72
# (suggesting worsening histological pattern despite treatment).

deritis <- adlb_meta |>
  filter(PARAMCD %in% c("ALT","AST")) |>
  select(USUBJID, VISIT, VISITNUM, PARAMCD, AVAL, ARMCD) |>
  pivot_wider(names_from = PARAMCD, values_from = AVAL) |>
  mutate(DERITIS = round(AST / ALT, 3)) |>
  group_by(USUBJID) |>
  mutate(
    DERITIS_BASE = DERITIS[VISIT == "BASELINE"][1],
    DERITIS_W72  = DERITIS[VISIT == "WEEK 72"][1],
    # Flag: ratio was <1 at baseline but >=1 at W72 (worsening pattern)
    DERITIS_WORSE = DERITIS_BASE < 1 & DERITIS_W72 >= 1
  ) |>
  ungroup()

# Summarise: how many subjects show this worsening pattern?
deritis |>
  filter(VISIT == "WEEK 72") |>
  group_by(ARMCD) |>
  summarise(
    n = n(),
    n_worse = sum(DERITIS_WORSE, na.rm = TRUE),
    median_ratio_base = round(median(DERITIS_BASE, na.rm=TRUE), 2),
    median_ratio_w72  = round(median(DERITIS_W72,  na.rm=TRUE), 2)
  )
```

```{r deritis-solution}
deritis <- adlb_meta |>
  filter(PARAMCD %in% c("ALT","AST")) |>
  select(USUBJID, VISIT, VISITNUM, PARAMCD, AVAL, ARMCD) |>
  pivot_wider(names_from = PARAMCD, values_from = AVAL) |>
  mutate(DERITIS = round(AST / ALT, 3)) |>
  group_by(USUBJID) |>
  mutate(
    DERITIS_BASE  = DERITIS[VISIT == "BASELINE"][1],
    DERITIS_W72   = DERITIS[VISIT == "WEEK 72"][1],
    DERITIS_WORSE = DERITIS_BASE < 1 & DERITIS_W72 >= 1
  ) |>
  ungroup()

deritis |>
  filter(VISIT == "WEEK 72") |>
  group_by(ARMCD) |>
  summarise(
    n = n(),
    n_worse = sum(DERITIS_WORSE, na.rm = TRUE),
    median_ratio_base = round(median(DERITIS_BASE, na.rm=TRUE), 2),
    median_ratio_w72  = round(median(DERITIS_W72,  na.rm=TRUE), 2)
  )
```

---

## Section 7 — Hard Challenges

### Hard Challenge 1 — Composite Metabolic Score

Build a **Composite Metabolic Improvement Score (CMIS)** that counts the number of metabolic improvements a subject achieves from baseline to Week 72. Criteria:

- HbA1c reduction ≥ 0.5 percentage points
- HOMA-IR reduction ≥ 15%
- LDL-C reduction ≥ 10%
- Triglycerides reduction ≥ 20%
- Body weight reduction ≥ 5%

Score each criterion 1 point; sum to CMIS (0–5). Subjects with CMIS ≥ 3 are considered **comprehensive metabolic responders**.

```{r cmis, exercise=TRUE, exercise.lines=50}
# Step 1: Compute baseline and W72 values for each subject across parameters
hba1c_bl_w72 <- adlb_meta |>
  filter(PARAMCD == "HBA1C", VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, AVAL) |>
  pivot_wider(names_from=VISIT, values_from=AVAL,names_repair="universal") |>
  rename(HBA1C_BL=BASELINE, HBA1C_W72=WEEK.72) |>
  mutate(HBA1C_RESP = (HBA1C_BL - HBA1C_W72) >= 0.5)

ldl_bl_w72 <- adlb_meta |>
  filter(PARAMCD == "LDL", VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, AVAL) |>
  pivot_wider(names_from=VISIT, values_from=AVAL,names_repair="universal") |>
  rename(LDL_BL=BASELINE, LDL_W72=WEEK.72) |>
  mutate(LDL_RESP = (LDL_BL - LDL_W72) / LDL_BL >= 0.10)

trig_bl_w72 <- adlb_meta |>
  filter(PARAMCD == "TRIG", VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, AVAL) |>
  pivot_wider(names_from=VISIT, values_from=AVAL,names_repair="universal") |>
  rename(TRIG_BL=BASELINE, TRIG_W72=WEEK.72) |>
  mutate(TRIG_RESP = (TRIG_BL - TRIG_W72) / TRIG_BL >= 0.20)

wt_bl_w72 <- wt_rows_meta |>
  filter(VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, AVAL, ARMCD) |>
  pivot_wider(names_from=VISIT, values_from=AVAL,names_repair="universal") |>
  rename(WT_BL=BASELINE, WT_W72=WEEK.72) |>
  mutate(WT_RESP = (WT_BL - WT_W72) / WT_BL >= 0.05)

# Step 2: Build HOMA-IR baseline and W72 (using the formula directly)
homa_bl_w72 <- adlb_meta |>
  filter(PARAMCD %in% c("GLUC","INS"), VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, PARAMCD, AVAL) |>
  pivot_wider(names_from=c(VISIT,PARAMCD), values_from=AVAL, names_sep="_") |>
  mutate(
    HOMA_BL  = (BASELINE_GLUC / 18.016 * BASELINE_INS) / 22.5,
    HOMA_W72 = (`WEEK 72_GLUC` / 18.016 * `WEEK 72_INS`) / 22.5,
    HOMA_RESP = (HOMA_BL - HOMA_W72) / HOMA_BL >= 0.15
  )

# Step 3: Join all criteria and compute CMIS
cmis_data <- wt_bl_w72 |>
  left_join(hba1c_bl_w72, by="USUBJID") |>
  left_join(ldl_bl_w72,   by="USUBJID") |>
  left_join(trig_bl_w72,  by="USUBJID") |>
  left_join(select(homa_bl_w72, USUBJID, HOMA_RESP), by="USUBJID") |>
  mutate(
    CMIS = as.integer(WT_RESP) + as.integer(HBA1C_RESP) +
           as.integer(LDL_RESP) + as.integer(TRIG_RESP) + as.integer(HOMA_RESP),
    CMIS_RESP = CMIS >= 3L
  )

# Report CMIS distribution and comprehensive responder rate by arm
cmis_data |>
  group_by(ARMCD) |>
  summarise(
    n              = n(),
    mean_cmis      = round(mean(CMIS, na.rm=TRUE), 2),
    pct_resp       = round(100 * mean(CMIS_RESP, na.rm=TRUE), 1),
    .groups = "drop"
  )
```

```{r cmis-solution}
hba1c_bl_w72 <- adlb_meta |>
  filter(PARAMCD == "HBA1C", VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, AVAL) |>
  pivot_wider(names_from=VISIT, values_from=AVAL, names_repair="universal") |>
  rename(HBA1C_BL=BASELINE, HBA1C_W72=WEEK.72) |>
  mutate(HBA1C_RESP = (HBA1C_BL - HBA1C_W72) >= 0.5)

ldl_bl_w72 <- adlb_meta |>
  filter(PARAMCD == "LDL", VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, AVAL) |>
  pivot_wider(names_from=VISIT, values_from=AVAL, names_repair="universal") |>
  rename(LDL_BL=BASELINE, LDL_W72=WEEK.72) |>
  mutate(LDL_RESP = (LDL_BL - LDL_W72) / LDL_BL >= 0.10)

trig_bl_w72 <- adlb_meta |>
  filter(PARAMCD == "TRIG", VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, AVAL) |>
  pivot_wider(names_from=VISIT, values_from=AVAL, names_repair="universal") |>
  rename(TRIG_BL=BASELINE, TRIG_W72=WEEK.72) |>
  mutate(TRIG_RESP = (TRIG_BL - TRIG_W72) / TRIG_BL >= 0.20)

wt_bl_w72 <- wt_rows_meta |>
  filter(VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, AVAL, ARMCD) |>
  pivot_wider(names_from=VISIT, values_from=AVAL, names_repair="universal") |>
  rename(WT_BL=BASELINE, WT_W72=WEEK.72) |>
  mutate(WT_RESP = (WT_BL - WT_W72) / WT_BL >= 0.05)

homa_bl_w72 <- adlb_meta |>
  filter(PARAMCD %in% c("GLUC","INS"), VISIT %in% c("BASELINE","WEEK 72")) |>
  select(USUBJID, VISIT, PARAMCD, AVAL) |>
  pivot_wider(names_from=c(VISIT,PARAMCD), values_from=AVAL, names_sep="_") |>
  mutate(
    HOMA_BL   = (BASELINE_GLUC / 18.016 * BASELINE_INS) / 22.5,
    HOMA_W72  = (`WEEK 72_GLUC` / 18.016 * `WEEK 72_INS`) / 22.5,
    HOMA_RESP = (HOMA_BL - HOMA_W72) / HOMA_BL >= 0.15
  )

cmis_data <- wt_bl_w72 |>
  left_join(hba1c_bl_w72, by="USUBJID") |>
  left_join(ldl_bl_w72,   by="USUBJID") |>
  left_join(trig_bl_w72,  by="USUBJID") |>
  left_join(select(homa_bl_w72, USUBJID, HOMA_RESP), by="USUBJID") |>
  mutate(
    CMIS = as.integer(WT_RESP) + as.integer(HBA1C_RESP) +
           as.integer(LDL_RESP) + as.integer(TRIG_RESP) + as.integer(HOMA_RESP),
    CMIS_RESP = CMIS >= 3L
  )

cmis_data |>
  group_by(ARMCD) |>
  summarise(
    n         = n(),
    mean_cmis = round(mean(CMIS, na.rm=TRUE), 2),
    pct_resp  = round(100 * mean(CMIS_RESP, na.rm=TRUE), 1),
    .groups = "drop"
  )
```

### Hard Challenge 2 — Mediation Analysis Proxy: Weight Loss Subgroup

A common question in GLP-1 RA NASH trials: *Is the histological benefit mediated by weight loss, or is there a weight-loss-independent hepatic effect?*

A standard analysis stratifies NASH responders by weight loss category (≥ 5% vs < 5%) and compares response rates within each stratum.

```{r mediation, exercise=TRUE, exercise.lines=30}
# Merge: biopsy_flagged (NASHRSFL), wt_bl_w72 (WT_RESP), ARMCD
mediation <- biopsy_flagged |>
  left_join(select(wt_bl_w72, USUBJID, WT_BL, WT_W72, WT_RESP), by="USUBJID") |>
  mutate(
    WT_LOSS_CAT = if_else(WT_RESP, ">=5% weight loss", "<5% weight loss")
  )

# NASH response rate stratified by arm AND weight loss category
mediation |>
  group_by(ARMCD, WT_LOSS_CAT) |>
  summarise(
    n          = n(),
    n_resp     = sum(NASHRSFL == "Y"),
    pct_resp   = round(100 * n_resp / n, 1),
    .groups = "drop"
  ) |>
  arrange(ARMCD, WT_LOSS_CAT)
```

```{r mediation-solution}
mediation <- biopsy_flagged |>
  left_join(select(wt_bl_w72, USUBJID, WT_BL, WT_W72, WT_RESP), by="USUBJID") |>
  mutate(WT_LOSS_CAT = if_else(WT_RESP, ">=5% weight loss", "<5% weight loss"))

mediation |>
  group_by(ARMCD, WT_LOSS_CAT) |>
  summarise(
    n        = n(),
    n_resp   = sum(NASHRSFL == "Y"),
    pct_resp = round(100 * n_resp / n, 1),
    .groups = "drop"
  ) |>
  arrange(ARMCD, WT_LOSS_CAT)
```

### Hard Challenge 3 — Hy's Law Signal Detection

**Hy's Law** is the FDA/EMA criterion for drug-induced liver injury (DILI) risk. A Hy's Law case requires *all three simultaneously*:

- ALT or AST > 3× ULN
- Total bilirubin > 2× ULN (TBL; not in our dataset — use ALT + AST as a composite proxy)
- No other cause identified (cannot be assessed computationally — flag for medical review)

For this exercise use a simplified criterion: **ALT > 3× ULN AND AST > 3× ULN** at the same visit.

```{r hys-law, exercise=TRUE, exercise.lines=30}
# ULN: ALT 40/35 U/L (M/F); AST 40/35 U/L (M/F)
uln_tbl <- tibble(
  SEX     = c("M","F"),
  ALT_ULN = c(40L, 35L),
  AST_ULN = c(40L, 35L)
)

hys_data <- adlb_meta |>
  filter(PARAMCD %in% c("ALT","AST")) |>
  select(USUBJID, VISIT, VISITNUM, PARAMCD, AVAL, ARMCD) |>
  pivot_wider(names_from=PARAMCD, values_from=AVAL) |>
  left_join(select(adsl_meta, USUBJID, SEX), by="USUBJID") |>
  left_join(uln_tbl, by="SEX") |>
  mutate(
    ALT_3ULN = ALT > 3 * ALT_ULN,
    AST_3ULN = AST > 3 * AST_ULN,
    HYS_FLAG = ALT_3ULN & AST_3ULN
  )

# Subjects with ≥1 Hy's Law flagged visit, by arm
hys_subjects <- hys_data |>
  group_by(USUBJID, ARMCD) |>
  summarise(ANY_HYS = any(HYS_FLAG, na.rm=TRUE), .groups="drop")

hys_subjects |>
  group_by(ARMCD) |>
  summarise(
    n       = n(),
    n_hys   = sum(ANY_HYS),
    pct_hys = round(100 * n_hys / n, 1)
  )
```

```{r hys-law-solution}
uln_tbl <- tibble(SEX=c("M","F"), ALT_ULN=c(40L,35L), AST_ULN=c(40L,35L))

hys_data <- adlb_meta |>
  filter(PARAMCD %in% c("ALT","AST")) |>
  select(USUBJID, VISIT, VISITNUM, PARAMCD, AVAL, ARMCD) |>
  pivot_wider(names_from=PARAMCD, values_from=AVAL) |>
  left_join(select(adsl_meta, USUBJID, SEX), by="USUBJID") |>
  left_join(uln_tbl, by="SEX") |>
  mutate(
    ALT_3ULN = ALT > 3 * ALT_ULN,
    AST_3ULN = AST > 3 * AST_ULN,
    HYS_FLAG = ALT_3ULN & AST_3ULN
  )

hys_subjects <- hys_data |>
  group_by(USUBJID, ARMCD) |>
  summarise(ANY_HYS = any(HYS_FLAG, na.rm=TRUE), .groups="drop")

hys_subjects |>
  group_by(ARMCD) |>
  summarise(
    n       = n(),
    n_hys   = sum(ANY_HYS),
    pct_hys = round(100 * n_hys / n, 1)
  )
```

---

## Section 8 — Capstone 1: Full ADLB for NASH Trial

Build a production-ready **ADLB** dataset from `adlb_meta`, adding all derived variables expected for a NASH metabolic ADaM submission.

Requirements:
- All standard ADLB columns: STUDYID, USUBJID, VISIT, VISITNUM, PARAMCD, PARAM, PARAMU, AVAL, BASE, ABLFL, CHG, PCHG
- DTYPE = NA (no imputation records in this dataset)
- SEX from adsl_meta (needed for ULN derivation)
- ATOXGR (NCI CTCAE grade based on ULN multiples for ALT/AST/GGT; NA for other parameters)
- HBA1CCAT for HbA1c records
- LDLCAT for LDL records
- A flag HIFL = "Y" if AVAL > 2 × ULN (relevant for ALT, AST, GGT)

```{r capstone-adlb, exercise=TRUE, exercise.lines=60}
# ULN reference (simplified — use same ULN for all ages)
uln_ref <- tribble(
  ~PARAMCD, ~SEX, ~ULN,
  "ALT",    "M",  40,
  "ALT",    "F",  35,
  "AST",    "M",  40,
  "AST",    "F",  35,
  "GGT",    "M",  55,
  "GGT",    "F",  38
)

# NCI CTCAE ALT/AST grading (multiples of ULN)
# Grade 1: >ULN–3×; Grade 2: >3–5×; Grade 3: >5–20×; Grade 4: >20×
ctcae_grade <- function(val, uln) {
  case_when(
    val <= uln       ~ 0L,
    val <= 3 * uln   ~ 1L,
    val <= 5 * uln   ~ 2L,
    val <= 20 * uln  ~ 3L,
    TRUE             ~ 4L
  )
}

adlb_final <- adlb_meta |>
  # Join subject-level variables
  left_join(select(adsl_meta, USUBJID, SEX), by = "USUBJID") |>
  # Join ULN where applicable
  left_join(___, by = c("PARAMCD","SEX")) |>
  # HIFL: >2×ULN for liver enzymes
  mutate(
    HIFL = if_else(!is.na(ULN) & ___ > 2 * ULN, "Y", NA_character_)
  ) |>
  # ATOXGR for liver enzyme parameters
  mutate(
    ATOXGR = if_else(
      PARAMCD %in% c("ALT","AST","GGT") & !is.na(ULN),
      ctcae_grade(___, ULN),
      NA_integer_
    )
  ) |>
  # HbA1c category
  derive_var_hba1c_cat(hba1c_var = AVAL, new_var = HBA1CCAT) |>
  mutate(HBA1CCAT = if_else(PARAMCD == "HBA1C", HBA1CCAT, NA_character_)) |>
  # LDL category
  derive_var_ldl_cat(ldl_var = AVAL, new_var = LDLCAT) |>
  mutate(LDLCAT = if_else(PARAMCD == "LDL", LDLCAT, NA_character_)) |>
  # Standard columns only
  select(STUDYID, USUBJID, VISIT, VISITNUM, VISITDY, ADT, ADY,
         PARAMCD, PARAM, PARAMU, AVAL, BASE, ABLFL, CHG, PCHG,
         HIFL, ATOXGR, HBA1CCAT, LDLCAT, SEX, ARMCD)

# Spot-check: ALT rows with ATOXGR ≥ 2
adlb_final |>
  filter(PARAMCD == "ALT", !is.na(ATOXGR), ATOXGR >= 2) |>
  select(USUBJID, VISIT, AVAL, ATOXGR, HIFL, ARMCD) |>
  arrange(desc(AVAL)) |>
  head(10)
```

```{r capstone-adlb-hint-1}
# left_join(uln_ref, by = c("PARAMCD","SEX"))
# HIFL: AVAL > 2 * ULN
# ATOXGR: ctcae_grade(AVAL, ULN)
```

```{r capstone-adlb-solution}
uln_ref <- tribble(
  ~PARAMCD, ~SEX, ~ULN,
  "ALT",    "M",  40,
  "ALT",    "F",  35,
  "AST",    "M",  40,
  "AST",    "F",  35,
  "GGT",    "M",  55,
  "GGT",    "F",  38
)

ctcae_grade <- function(val, uln) {
  case_when(
    val <= uln       ~ 0L,
    val <= 3 * uln   ~ 1L,
    val <= 5 * uln   ~ 2L,
    val <= 20 * uln  ~ 3L,
    TRUE             ~ 4L
  )
}

adlb_final <- adlb_meta |>
  left_join(select(adsl_meta, USUBJID, SEX), by = "USUBJID") |>
  left_join(uln_ref, by = c("PARAMCD","SEX")) |>
  mutate(
    HIFL   = if_else(!is.na(ULN) & AVAL > 2 * ULN, "Y", NA_character_),
    ATOXGR = if_else(
      PARAMCD %in% c("ALT","AST","GGT") & !is.na(ULN),
      ctcae_grade(AVAL, ULN),
      NA_integer_
    )
  ) |>
  derive_var_hba1c_cat(hba1c_var = AVAL, new_var = HBA1CCAT) |>
  mutate(HBA1CCAT = if_else(PARAMCD == "HBA1C", HBA1CCAT, NA_character_)) |>
  derive_var_ldl_cat(ldl_var = AVAL, new_var = LDLCAT) |>
  mutate(LDLCAT = if_else(PARAMCD == "LDL", LDLCAT, NA_character_)) |>
  select(STUDYID, USUBJID, VISIT, VISITNUM, VISITDY, ADT, ADY,
         PARAMCD, PARAM, PARAMU, AVAL, BASE, ABLFL, CHG, PCHG,
         HIFL, ATOXGR, HBA1CCAT, LDLCAT, SEX, ARMCD)

adlb_final |>
  filter(PARAMCD == "ALT", !is.na(ATOXGR), ATOXGR >= 2) |>
  select(USUBJID, VISIT, AVAL, ATOXGR, HIFL, ARMCD) |>
  arrange(desc(AVAL)) |>
  head(10)
```

---

## Section 9 — Capstone 2: DSMB Metabolic Safety Summary

Produce a **Data Safety Monitoring Board (DSMB) metabolic safety table** for the Week 48 interim analysis. This is the kind of output a clinical programmer must produce under time pressure for a live DSMB meeting.

Required rows:
1. Mean HbA1c (%) ± SD at Week 48
2. Subjects with HbA1c ≥ 8.0% at Week 48 (n, %)
3. Mean HOMA-IR at baseline and Week 48, with mean change
4. Subjects with any ALT > 3× ULN during the study
5. Subjects with any Hy's Law event (ALT + AST both > 3× ULN)
6. Mean body weight change (kg) and % change at Week 48
7. NASH primary endpoint response rate at Week 72 (from biopsy_flagged)

```{r capstone-dsmb, exercise=TRUE, exercise.lines=70}
# ── HbA1c rows ────────────────────────────────────────────────────────────────
hba1c_w48 <- adlb_meta |>
  filter(PARAMCD == "HBA1C", VISIT == "WEEK 48") |>
  group_by(ARMCD) |>
  summarise(
    stat   = "Mean HbA1c at Wk48 (%)",
    value  = paste0(round(mean(AVAL),2), " (SD ", round(sd(AVAL),2), ")"),
    .groups = "drop"
  )

hba1c_hi_w48 <- adlb_meta |>
  filter(PARAMCD == "HBA1C", VISIT == "WEEK 48") |>
  group_by(ARMCD) |>
  summarise(
    stat  = "HbA1c >= 8.0% at Wk48, n (%)",
    value = paste0(sum(AVAL >= 8.0), " (", round(100*mean(AVAL >= 8.0),1), "%)"),
    .groups = "drop"
  )

# ── HOMA-IR change ────────────────────────────────────────────────────────────
homa_w48 <- adlb_meta |>
  filter(PARAMCD %in% c("GLUC","INS"), VISIT %in% c("BASELINE","WEEK 48")) |>
  select(USUBJID, VISIT, PARAMCD, AVAL, ARMCD) |>
  pivot_wider(names_from=c(VISIT,PARAMCD), values_from=AVAL, names_sep="_") |>
  mutate(
    HOMA_BL  = (BASELINE_GLUC/18.016 * BASELINE_INS) / 22.5,
    HOMA_W48 = (`WEEK 48_GLUC`/18.016 * `WEEK 48_INS`) / 22.5,
    HOMA_CHG = HOMA_W48 - HOMA_BL
  ) |>
  group_by(ARMCD) |>
  summarise(
    stat  = "Mean HOMA-IR change at Wk48",
    value = paste0(round(mean(HOMA_BL,na.rm=T),2),
                   " → ", round(mean(HOMA_W48,na.rm=T),2),
                   " (chg: ", round(mean(HOMA_CHG,na.rm=T),2), ")"),
    .groups = "drop"
  )

# ── ALT > 3×ULN (any visit) ──────────────────────────────────────────────────
uln_vals <- c(M=40, F=35)
alt_hi <- adlb_meta |>
  filter(PARAMCD == "ALT") |>
  left_join(select(adsl_meta, USUBJID, SEX), by="USUBJID") |>
  mutate(ULN = uln_vals[SEX], ALT_3X = AVAL > 3 * ULN) |>
  group_by(USUBJID, ARMCD) |>
  summarise(ANY_ALT3X = any(ALT_3X, na.rm=TRUE), .groups="drop") |>
  group_by(ARMCD) |>
  summarise(
    stat  = "Any ALT > 3xULN, n (%)",
    value = paste0(sum(ANY_ALT3X), " (", round(100*mean(ANY_ALT3X),1), "%)"),
    .groups = "drop"
  )

# ── Weight change at W48 ──────────────────────────────────────────────────────
wt_w48 <- wt_rows_meta |>
  filter(VISIT == "WEEK 48") |>
  group_by(ARMCD) |>
  summarise(
    stat  = "Mean weight change at Wk48 (kg, %)",
    value = paste0(round(mean(CHG,na.rm=T),1), " kg (",
                   round(mean(PCHG,na.rm=T),1), "%)"),
    .groups = "drop"
  )

# ── Primary endpoint ──────────────────────────────────────────────────────────
nash_ep <- biopsy_flagged |>
  group_by(ARMCD) |>
  summarise(
    stat  = "NASH primary responders at Wk72, n (%)",
    value = paste0(sum(NASHRSFL=="Y"), " (",
                   round(100*mean(NASHRSFL=="Y"),1), "%)"),
    .groups = "drop"
  )

# ── Assemble DSMB table ───────────────────────────────────────────────────────
dsmb_table <- bind_rows(
  hba1c_w48, hba1c_hi_w48, homa_w48, alt_hi, wt_w48, nash_ep
) |>
  pivot_wider(names_from=ARMCD, values_from=value) |>
  select(stat, GLP, PBO)

print(dsmb_table, n=Inf)
```

```{r capstone-dsmb-solution}
hba1c_w48 <- adlb_meta |>
  filter(PARAMCD == "HBA1C", VISIT == "WEEK 48") |>
  group_by(ARMCD) |>
  summarise(
    stat  = "Mean HbA1c at Wk48 (%)",
    value = paste0(round(mean(AVAL),2), " (SD ", round(sd(AVAL),2), ")"),
    .groups = "drop"
  )

hba1c_hi_w48 <- adlb_meta |>
  filter(PARAMCD == "HBA1C", VISIT == "WEEK 48") |>
  group_by(ARMCD) |>
  summarise(
    stat  = "HbA1c >= 8.0% at Wk48, n (%)",
    value = paste0(sum(AVAL >= 8.0), " (", round(100*mean(AVAL >= 8.0),1), "%)"),
    .groups = "drop"
  )

homa_w48 <- adlb_meta |>
  filter(PARAMCD %in% c("GLUC","INS"), VISIT %in% c("BASELINE","WEEK 48")) |>
  select(USUBJID, VISIT, PARAMCD, AVAL, ARMCD) |>
  pivot_wider(names_from=c(VISIT,PARAMCD), values_from=AVAL, names_sep="_") |>
  mutate(
    HOMA_BL  = (BASELINE_GLUC/18.016 * BASELINE_INS) / 22.5,
    HOMA_W48 = (`WEEK 48_GLUC`/18.016 * `WEEK 48_INS`) / 22.5,
    HOMA_CHG = HOMA_W48 - HOMA_BL
  ) |>
  group_by(ARMCD) |>
  summarise(
    stat  = "Mean HOMA-IR change at Wk48",
    value = paste0(round(mean(HOMA_BL,na.rm=T),2),
                   " → ", round(mean(HOMA_W48,na.rm=T),2),
                   " (chg: ", round(mean(HOMA_CHG,na.rm=T),2), ")"),
    .groups = "drop"
  )

uln_vals <- c(M=40, F=35)
alt_hi <- adlb_meta |>
  filter(PARAMCD == "ALT") |>
  left_join(select(adsl_meta, USUBJID, SEX), by="USUBJID") |>
  mutate(ULN = uln_vals[SEX], ALT_3X = AVAL > 3 * ULN) |>
  group_by(USUBJID, ARMCD) |>
  summarise(ANY_ALT3X = any(ALT_3X, na.rm=TRUE), .groups="drop") |>
  group_by(ARMCD) |>
  summarise(
    stat  = "Any ALT > 3xULN, n (%)",
    value = paste0(sum(ANY_ALT3X), " (", round(100*mean(ANY_ALT3X),1), "%)"),
    .groups = "drop"
  )

wt_w48 <- wt_rows_meta |>
  filter(VISIT == "WEEK 48") |>
  group_by(ARMCD) |>
  summarise(
    stat  = "Mean weight change at Wk48 (kg, %)",
    value = paste0(round(mean(CHG,na.rm=T),1), " kg (",
                   round(mean(PCHG,na.rm=T),1), "%)"),
    .groups = "drop"
  )

nash_ep <- biopsy_flagged |>
  group_by(ARMCD) |>
  summarise(
    stat  = "NASH primary responders at Wk72, n (%)",
    value = paste0(sum(NASHRSFL=="Y"), " (",
                   round(100*mean(NASHRSFL=="Y"),1), "%)"),
    .groups = "drop"
  )

dsmb_table <- bind_rows(
  hba1c_w48, hba1c_hi_w48, homa_w48, alt_hi, wt_w48, nash_ep
) |>
  pivot_wider(names_from=ARMCD, values_from=value) |>
  select(stat, GLP, PBO)

print(dsmb_table, n=Inf)
```

---

## Lesson Summary

```{r final-quiz, echo=FALSE}
quiz(caption = "Lesson 24 — Final Quiz",
  question("A NASH trial subject has NAS = 7 at baseline (fibrosis stage 2) and NAS = 5 at Week 72 (fibrosis stage 2). Are they a primary endpoint responder under the standard dual composite criterion?",
    answer("Yes — NAS dropped ≥ 2 points"),
    answer("No — NAS must drop to ≤ 2 for response"),
    answer("No — NAS dropped by only 2 points and fibrosis did not improve",
           message = "Close! The criterion is NAS drop ≥ 2 AND no worsening of fibrosis. Since fibrosis went 2→2 (no worsening), they DO qualify under criterion 1 (NAS −2 + fibrosis not worsened)."),
    answer("Yes — NAS dropped ≥ 2 points with no worsening of fibrosis",
           correct = TRUE),
    allow_retry = TRUE
  ),
  question("The De Ritis ratio (AST/ALT) > 1 is associated with:",
    answer("Active NASH with predominant steatosis"),
    answer("More advanced fibrosis or alcoholic hepatitis", correct = TRUE),
    answer("Improved glycaemic control"),
    answer("Seroconversion to anti-HBs"),
    allow_retry = TRUE
  )
)
```

### Summary Coding Exercise 1 — HOMA-IR and Friedewald Check

```{r summary-homa-friedewald, exercise=TRUE, exercise.lines=20}
# Compute HOMA-IR at BASELINE for all subjects and flag subjects where
# the Friedewald LDL would be invalid (TG >= 400 at baseline).

# Part 1: HOMA-IR at baseline
homa_check <- baseline_labs |>
  mutate(
    HOMAIR = (___ / 18.016 * ___) / 22.5
  ) |>
  group_by(ARMCD) |>
  summarise(mean_HOMAIR = round(mean(HOMAIR, na.rm = TRUE), 2), n = n())

# Part 2: subjects with invalid Friedewald at baseline
friedewald_invalid <- baseline_labs |>
  filter(___ >= 400) |>
  select(USUBJID, ARMCD, TRIG_BASE)

cat("Subjects with TG >= 400 at baseline:", nrow(friedewald_invalid), "\n")
homa_check
```

```{r summary-homa-friedewald-hint-1}
# HOMA-IR: (GLUC_BASE / 18.016 * INS_BASE) / 22.5
# Friedewald invalid: TRIG_BASE >= 400
```

```{r summary-homa-friedewald-solution}
homa_check <- baseline_labs |>
  mutate(
    HOMAIR = (GLUC_BASE / 18.016 * INS_BASE) / 22.5
  ) |>
  group_by(ARMCD) |>
  summarise(mean_HOMAIR = round(mean(HOMAIR, na.rm = TRUE), 2), n = n())

friedewald_invalid <- baseline_labs |>
  filter(TRIG_BASE >= 400) |>
  select(USUBJID, ARMCD, TRIG_BASE)

cat("Subjects with TG >= 400 at baseline:", nrow(friedewald_invalid), "\n")
homa_check
```

### Summary Coding Exercise 2 — CTCAE Grade and MetSyn Count

```{r summary-ctcae-metsyn, exercise=TRUE, exercise.lines=20}
# 1. Count subjects with ALT CTCAE Grade >= 2 at any visit (ULN = 40 M / 35 F).
# 2. Count subjects meeting metabolic syndrome criteria (from baseline_labs).

# CTCAE: Grade 0 <= ULN, Grade 1 >ULN-3x, Grade 2 >3-5x, Grade 3 >5-20x, Grade 4 >20x
alt_grade2_plus <- adlb_meta |>
  filter(PARAMCD == "ALT") |>
  left_join(select(adsl_meta, USUBJID, SEX), by = "USUBJID") |>
  mutate(
    ULN    = if_else(SEX == "M", 40, 35),
    ATOXGR = case_when(
      AVAL <= ULN      ~ 0L,
      AVAL <= 3 * ULN  ~ 1L,
      AVAL <= 5 * ULN  ~ ___,
      AVAL <= 20 * ULN ~ 3L,
      TRUE             ~ 4L
    )
  ) |>
  filter(___ >= 2) |>
  distinct(USUBJID, ARMCD) |>
  count(ARMCD, name = "n_grade2plus")

# MetSyn count
metsyn_count <- baseline_labs |>
  mutate(
    N_CRIT   = (BMIBL >= 30) + (TRIG_BASE >= 150) +
               if_else(SEX == "M", HDLC_BASE < 40, HDLC_BASE < 50) +
               (HTNFL == "Y") + (GLUC_BASE >= 100),
    METSYNFL = if_else(N_CRIT >= 3, "Y", "N")
  ) |>
  count(ARMCD, METSYNFL)

list(alt_grade2plus = alt_grade2_plus, metsyn = metsyn_count)
```

```{r summary-ctcae-metsyn-hint-1}
# ATOXGR Grade 2: AVAL <= 5 * ULN -> 2L
# Filter: ATOXGR >= 2
```

```{r summary-ctcae-metsyn-solution}
alt_grade2_plus <- adlb_meta |>
  filter(PARAMCD == "ALT") |>
  left_join(select(adsl_meta, USUBJID, SEX), by = "USUBJID") |>
  mutate(
    ULN    = if_else(SEX == "M", 40, 35),
    ATOXGR = case_when(
      AVAL <= ULN      ~ 0L,
      AVAL <= 3 * ULN  ~ 1L,
      AVAL <= 5 * ULN  ~ 2L,
      AVAL <= 20 * ULN ~ 3L,
      TRUE             ~ 4L
    )
  ) |>
  filter(ATOXGR >= 2) |>
  distinct(USUBJID, ARMCD) |>
  count(ARMCD, name = "n_grade2plus")

metsyn_count <- baseline_labs |>
  mutate(
    N_CRIT   = (BMIBL >= 30) + (TRIG_BASE >= 150) +
               if_else(SEX == "M", HDLC_BASE < 40, HDLC_BASE < 50) +
               (HTNFL == "Y") + (GLUC_BASE >= 100),
    METSYNFL = if_else(N_CRIT >= 3, "Y", "N")
  ) |>
  count(ARMCD, METSYNFL)

list(alt_grade2plus = alt_grade2_plus, metsyn = metsyn_count)
```

### What You Learned

In this lesson you built ADaM derivations specific to **metabolic disease trials**:

- **HOMA-IR** — glucose × insulin formula with mg/dL → mmol/L conversion
- **Friedewald LDL-C** — indirect LDL calculation valid only when TG < 400 mg/dL; `NA` otherwise
- **Non-HDL-C** — simple Total − HDL, no fasting required
- **NAS score and fibrosis stage** — histological paired-biopsy endpoints and the dual composite NASH response definition
- **HbA1c and LDL category flags** — ADA 2024 and NCEP ATP-III guideline thresholds
- **Metabolic syndrome flag** — ATP-III five-criterion composite with sex-dependent HDL threshold
- **Liver enzyme normalisation and Hy's Law** — safety flags using NCI CTCAE grading and the FDA Hy's Law criterion
- **De Ritis ratio** — AST/ALT diagnostic ratio for histological pattern assessment
- **Composite Metabolic Improvement Score (CMIS)** — multi-domain improvement scoring
- **DSMB safety table** — integrated summary across glycaemia, liver enzymes, weight, and histology

### Key Gotchas

> **TG validity gate:** Always filter or `NA` Friedewald LDL when TG ≥ 400 mg/dL. Missing this in production code leads to erroneous LDL values in TG-high patients.

> **NAS vs Fibrosis:** These are independent axes. NAS improvement alone is not sufficient for the primary NASH endpoint — fibrosis must not worsen.

> **HOMA-IR units:** Glucose must be in mmol/L, not mg/dL. Forgetting the ÷18.016 conversion inflates HOMA-IR by 18-fold.

> **HbA1c thresholds:** The 5.7%, 6.5%, and 7.0% thresholds serve different clinical purposes. 6.5% is the *diagnostic* threshold; 7.0% is the *treatment target* for known T2DM.

---

**Next lesson:** Lesson 25 — cards Part 1 — ARD (Analysis Results Dataset) structure and core constructors: the ARD column anatomy, why `stat` is a list column, and `ard_continuous()` / `ard_categorical()` / `ard_hierarchical()` / `ard_complex()`. Module 6 then continues with cards Part 2 (clinical ARDs), cardx (inferential ARDs), gtsummary, cardinal, and tfrmt.
```
