---
title: "Lesson 26 — cards Part 2: Building Clinical ARDs End-to-End"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build a complete study-wide Analysis Results Dataset (ARD) covering every
  clinical domain: demographics, treatment-emergent AE incidence by SOC and PT
  with an ADSL denominator, lab summary and change-from-baseline by visit,
  vital signs, drug exposure, and concomitant medications. Learn the population
  filter and denominator strategy for each table type, the factor-typing pattern
  for zero-count levels, ard_stack_hierarchical() for subject-level counting,
  bind_ard() for combination, ARS traceability metadata, and the dual-program-
  the-ARD validation paradigm.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 26_cards_part2_clinical_ards.Rmd
# PURPOSE       : cards Part 2 — building a complete clinical ARD pipeline
#                 end-to-end: demographics, TEAE incidence by SOC/PT with the
#                 ADSL safety-population denominator, lab change-from-baseline
#                 by visit, vital signs, drug exposure, and concomitant
#                 medications; the factor-typing pattern for zero-count levels,
#                 ard_stack_hierarchical() subject counting, ARS traceability tags,
#                 and the dual-program-the-ARD validation paradigm.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-06-22
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(purrr)
library(pharmaverseadam)
library(cards)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N")
  ) |>
  filter(SAFFL == "Y")

# Safety-population ADSL — the canonical denominator for AE / exposure tables.
adsl_saf <- adsl

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

advs <- pharmaverseadam::advs

# ── Pre-built ARDs available to every exercise (REAL {cards}) ─────────────────
# {cards} notes used throughout: the `statistic` argument is a FORMULA
# (~ continuous_summary_fns() or ~ list(...)); inside ard_stack() the inner
# ard_*() calls omit `data`/`by` (ard_stack injects them); group levels live in
# list columns, so by-variables stay character; `p` is a proportion (0-1);
# ard_stack_hierarchical() builds the full SOC x PT hierarchy in one call.

# Factor-typed safety ADSL for demographics (counted vars factored so zero-count
# levels still appear; the by-variable TRT01A stays character).
adsl_factored <- adsl_saf |>
  mutate(
    SEX    = factor(SEX, levels = c("F", "M")),
    AGEGR1 = factor(AGEGR1, levels = sort(unique(AGEGR1)))
  )

# Treatment-emergent AEs
adae_te <- adae |> filter(TRTEMFL == "Y")

# Analysis-flagged labs with a visit
adlb_an <- adlb |> filter(!is.na(AVISIT))

# Pre-built demographics ARD (continuous + categorical, by arm, Overall + total-N)
demog_ard <- ard_stack(
  adsl_factored,
  ard_continuous(variables = c("AGE","BMIBL","WEIGHTBL","HEIGHTBL"),
                 statistic = ~ continuous_summary_fns(c("N","mean","sd","median","min","max"))),
  ard_categorical(variables = c("AGEGR1","SEX")),
  .by = "TRT01A", .overall = TRUE, .total_n = TRUE
)

# Pre-built AE incidence ARD (SOC x PT, subject denominator)
ae_incidence_ard <- ard_stack_hierarchical(
  data = adae_te, variables = c("AEBODSYS","AEDECOD"),
  by = "TRT01A", denominator = adsl_saf, id = "USUBJID"
)
```

## Overview

In **Lesson 25** you learned the ARD building blocks: `ard_continuous()`,
`ard_categorical()`, `ard_missing()`, and how to `bind_ard()` them together.
This lesson puts those blocks to work on a **complete clinical study**. By the
end you will have assembled the ARDs behind the major sections of a Clinical
Study Report (CSR): demographics, treatment-emergent adverse events, laboratory
change-from-baseline, vital signs, drug exposure, and concomitant medications.

The mental model:

```
ADaM data  →  population filter  →  ard_*()  →  ARD per topic  →  bind_ard()  →  study ARD
```

> **The single most important idea in this lesson:** *you validate the ARD, not
> the display.* In SAS, two programmers each build the RTF table and QA compares
> the two files cell-by-cell. In the cards world, two programmers each build the
> **ARD**, and you compare those flat numeric files. The display layer becomes a
> trivial, formatting-only check on top.

### Trial dataset

All exercises use the **CDISCPILOT01** Alzheimer's study from `pharmaverseadam`.
Three arms — `Placebo`, `Xanomeline Low Dose`, `Xanomeline High Dose` — with the
safety population pre-filtered (`SAFFL == "Y"`) into `adsl_saf`. The setup chunk
also provides:

- `adae` — adverse events, with `TRTA` renamed to **`TRT01A`** and a numeric
  `AESEVN` derived; `adae_te` is pre-filtered to treatment-emergent events
  (`TRTEMFL == "Y"`).
- `adlb` — labs, with `ATOXGR` coerced to integer and admiral summary rows
  (`DTYPE` not `NA`) removed; `adlb_an` is pre-filtered to rows with a non-`NA`
  `AVISIT`.
- `advs` — vital signs (`PARAMCD` of `SYSBP`, `DIABP`, `PULSE`, `WEIGHT`, `HEIGHT`).

Two ARDs are pre-built for you: `demog_ard` (demographics) and
`ae_incidence_ard` (TEAE SOC × PT incidence).

> **Real `{cards}` conventions.** This tutorial runs the real package
> (`library(cards)`). Three behaviours recur: (1) the `statistic` argument is a
> **formula** — `~ continuous_summary_fns(...)` or `~ list(...)`; (2) inside
> `ard_stack()` the inner `ard_*()` calls **omit `data` and `by`** (ard_stack
> injects them via `.by`); (3) group levels and `variable_level` are **list
> columns**, so by-variables stay character and you unwrap levels with
> `map_chr(., 1)` / `unlist()`. Also note `p` is a **proportion (0-1)** — multiply
> by 100 for a percentage. The full SOC × PT AE hierarchy comes from
> `ard_stack_hierarchical()`.

---

## Section 1 — Demographics ARD: the canonical template

The standard demographics ("Table 14.1.1") combines **continuous** variables
(AGE, BMI, weight, height) and **categorical** variables (age group, sex) by
treatment arm, with an Overall column and Big-N header counts.

```r
demog_ard <- ard_stack(
  adsl_saf,
  ard_continuous(
    variables = c("AGE","BMIBL","WEIGHTBL","HEIGHTBL"),
    statistic = ~ continuous_summary_fns(c("N","mean","sd","median","min","max"))),
  ard_categorical(variables = c("AGEGR1","SEX")),
  .by      = "TRT01A",  # columns by treatment arm — ard_stack injects it
  .overall = TRUE,      # add an "Overall" (marginal) group, group1_level = NA
  .total_n = TRUE       # add the overall total-N row (context = "total_n")
)
```

`continuous_summary_fns()` returns a named list of summary functions; pass it a
character vector to pick exactly the statistics you want.

### Practice — explore the pre-built demographics ARD

```{r ex-demog-explore, exercise=TRUE, exercise.lines=10}
# demog_ard is pre-built. What contexts and statistics does it contain?
demog_ard |> distinct(context)
demog_ard |> distinct(___)        # all unique stat_name values
```

```{r ex-demog-explore-hint-1}
# distinct(stat_name)
```

```{r ex-demog-explore-solution}
demog_ard |> distinct(context)
demog_ard |> distinct(stat_name)
```

### Practice — spot-check mean age by arm

```{r ex-demog-meanage, exercise=TRUE, exercise.lines=10}
# Pull the mean AGE for each arm (including Overall).
# stat is a LIST column — use map_dbl(stat, 1) to extract the number.
demog_ard |>
  filter(variable == "AGE", stat_name == ___) |>
  mutate(mean_age = round(map_dbl(stat, 1), 1)) |>
  select(group1_level, mean_age)
```

```{r ex-demog-meanage-hint-1}
# stat_name == "mean"
```

```{r ex-demog-meanage-solution}
demog_ard |>
  filter(variable == "AGE", stat_name == "mean") |>
  mutate(mean_age = round(map_dbl(stat, 1), 1)) |>
  select(group1_level, mean_age)
```

### Practice — use get_ard_statistics() for a targeted lookup

```{r ex-demog-getstat, exercise=TRUE, exercise.lines=10}
# get_ard_statistics() pulls a named list of stats. Conditions are passed
# POSITIONALLY (comma-separated, like dplyr::filter) — there is no `filter =`.
# Levels live in list columns, so unwrap variable_level with map_chr(., 1).
# Get the n and p for Female subjects in the Placebo arm.
get_ard_statistics(
  demog_ard,
  variable == "SEX",
  map_chr(variable_level, \(x) if (length(x)) as.character(x[[1]]) else NA_character_) == "F",
  group1_level == "Placebo",
  stat_name %in% c("n", "p")
)
```

```{r ex-demog-getstat-solution}
get_ard_statistics(
  demog_ard,
  variable == "SEX",
  map_chr(variable_level, \(x) if (length(x)) as.character(x[[1]]) else NA_character_) == "F",
  group1_level == "Placebo",
  stat_name %in% c("n", "p")
)
```

### The factor-typing pattern for categorical variables

If an arm has **zero** subjects in a category level, that level vanishes from the
ARD for that arm — and your table loses a row. Factor-type *before* building the
ARD so every arm carries every level, even at `n = 0`:

```r
adsl_factored <- adsl_saf |>
  mutate(
    AGEGR1 = factor(AGEGR1, levels = c("<65", "65-80", ">80")),
    SEX    = factor(SEX,    levels = c("F", "M")),
    RACE   = factor(RACE,   levels = c(
      "WHITE", "BLACK OR AFRICAN AMERICAN", "ASIAN",
      "AMERICAN INDIAN OR ALASKA NATIVE", "OTHER"))
  )
```

The mock `ard_categorical()` respects factor levels exactly as the real package
does.

### Practice — build a demographics ARD from scratch

```{r ex-demog-build, exercise=TRUE, exercise.lines=18}
# Build a demographics ARD with:
#   - continuous: AGE, BMIBL  (N, mean, sd, median)
#   - categorical: SEX (factor-typed F, M)
# by TRT01A, with an Overall group and Big-N rows.

my_adsl <- adsl_saf |>
  mutate(SEX = factor(SEX, levels = c("F", "M")))

my_demog <- ard_stack(
  my_adsl,
  ard_continuous(variables = c(___, ___),
                 statistic = ~ continuous_summary_fns(c("N","mean","sd","median"))),
  ard_categorical(variables = ___),
  .by = "TRT01A", .overall = ___, .total_n = ___
)

my_demog |> count(context)
```

```{r ex-demog-build-hint-1}
# variables = c("AGE","BMIBL"); ard_categorical variables = "SEX"
# .overall = TRUE, .total_n = TRUE
```

```{r ex-demog-build-solution}
my_adsl <- adsl_saf |>
  mutate(SEX = factor(SEX, levels = c("F", "M")))

my_demog <- ard_stack(
  my_adsl,
  ard_continuous(variables = c("AGE","BMIBL"),
                 statistic = ~ continuous_summary_fns(c("N","mean","sd","median"))),
  ard_categorical(variables = "SEX"),
  .by = "TRT01A", .overall = TRUE, .total_n = TRUE
)

my_demog |> count(context)
```

### Practice — validate the ARD

```{r ex-demog-validate, exercise=TRUE, exercise.lines=6}
# Run the two validation helpers on demog_ard.
check_ard_structure(___)
print_ard_conditions(___)
```

```{r ex-demog-validate-hint-1}
# Both take the ARD: check_ard_structure(demog_ard) etc.
```

```{r ex-demog-validate-solution}
check_ard_structure(demog_ard)
print_ard_conditions(demog_ard)
```

```{r quiz-demog, echo=FALSE}
quiz(caption = "Demographics ARD",
  question("An arm has no subjects in the '>80' age group, so that row is missing for that arm. What is the correct fix?",
    answer("Add a row manually with bind_rows()"),
    answer("Convert AGEGR1 to a factor with all expected levels BEFORE the ard_categorical() call",
           correct = TRUE,
           message = "Factor-typing fixes the level set so every group carries every level, producing n = 0 rows instead of dropping them."),
    answer("Use .total_n = TRUE"),
    answer("Use ard_stack_hierarchical() instead"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question("Which `context` value identifies the Big-N rows added by `.total_n = TRUE`?",
    answer("\"big_n\""),
    answer("\"total_n\"", correct = TRUE,
           message = "Filter on context == \"total_n\" to find the per-group denominators that gtsummary uses to build column headers."),
    answer("\"header\""),
    answer("\"N\""),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

---

## Section 2 — AE incidence ARD: the safety table

The adverse-event table counts **subjects** (not events) with at least one TEAE,
then by System Organ Class (`AEBODSYS`), then by Preferred Term (`AEDECOD`)
within each SOC. Two cards features make this correct:

1. **`ard_stack_hierarchical()`** does the nested SOC → PT tabulation.
2. **`denominator = adsl_saf`** sets `N` to the safety-population subject count
   per arm — *not* the number of AE rows.
3. **`id = "USUBJID"`** counts distinct subjects, so a subject with three events
   in one PT contributes 1, not 3.

```r
adae_te <- adae |> filter(TRTEMFL == "Y")

ae_incidence_ard <- adae_te |>
  ard_stack_hierarchical(
    by          = "TRT01A",
    variables   = c("AEBODSYS", "AEDECOD"),
    denominator = adsl_saf,    # N = safety-pop subjects per arm
    id          = "USUBJID"    # distinct subjects, not events
  )
```

### Why `denominator = adsl_saf` is non-negotiable

Without it, `N` becomes the count of TEAE *rows* per arm. Subjects with multiple
AEs inflate `N`, percentages come out wrong, and some PTs show > 100%.

> **Code-review rule:** every AE/CM/exposure ARD must carry
> `denominator = adsl_saf` (or the appropriate population). If a percentage
> exceeds 100%, the missing denominator is almost always the cause.

### Practice — explore the hierarchical AE ARD

```{r ex-ae-explore, exercise=TRUE, exercise.lines=10}
# ae_incidence_ard is pre-built. It has group2 columns for the SOC nesting.
# Show the group columns and the contexts.
names(ae_incidence_ard)[startsWith(names(ae_incidence_ard), "group")]
ae_incidence_ard |> distinct(variable)   # which nesting levels are present?
```

```{r ex-ae-explore-solution}
names(ae_incidence_ard)[startsWith(names(ae_incidence_ard), "group")]
ae_incidence_ard |> distinct(variable)
```

### Practice — top preferred terms in the High Dose arm

```{r ex-ae-top, exercise=TRUE, exercise.lines=14}
# Find the 5 PTs with the highest incidence (p) in Xanomeline High Dose.
# p is a proportion (0-1) — multiply by 100 for a percentage. SOC and PT live
# in list columns (group2_level / variable_level), so unwrap them for display.
ae_incidence_ard |>
  filter(variable == "AEDECOD",
         group1_level == ___,         # the High Dose arm label
         stat_name == "p") |>
  mutate(SOC = map_chr(group2_level, 1),
         PT  = unlist(variable_level),
         pct = round(map_dbl(stat, 1) * 100, 1)) |>
  slice_max(pct, n = 5) |>
  select(SOC, PT, pct)
```

```{r ex-ae-top-hint-1}
# group1_level == "Xanomeline High Dose"; variable_level holds the PT
```

```{r ex-ae-top-solution}
ae_incidence_ard |>
  filter(variable == "AEDECOD",
         group1_level == "Xanomeline High Dose",
         stat_name == "p") |>
  mutate(SOC = map_chr(group2_level, 1),
         PT  = unlist(variable_level),
         pct = round(map_dbl(stat, 1) * 100, 1)) |>
  slice_max(pct, n = 5) |>
  select(SOC, PT, pct)
```

### The "any TEAE" overall row

The header row "Subjects with at least one TEAE" is a single dichotomous count
per arm. Reduce to one row per subject, flag it, then count with the subject
denominator:

```{r ex-ae-any, exercise=TRUE, exercise.lines=14}
# Build the "any TEAE" overall row: distinct subjects with any TEAE, by arm,
# using adsl_saf as the denominator.
adae_any_ard <- adae_te |>
  distinct(USUBJID, TRT01A) |>
  mutate(any_teae = "Y") |>
  ard_categorical(
    by          = "TRT01A",
    variables   = "any_teae",
    denominator = ___          # subject denominator
  )

adae_any_ard |>
  filter(stat_name %in% c("n","N","p")) |>
  mutate(val = map_dbl(stat, 1)) |>
  select(group1_level, stat_name, val)
```

```{r ex-ae-any-hint-1}
# denominator = adsl_saf
```

```{r ex-ae-any-solution}
adae_any_ard <- adae_te |>
  distinct(USUBJID, TRT01A) |>
  mutate(any_teae = "Y") |>
  ard_categorical(
    by          = "TRT01A",
    variables   = "any_teae",
    denominator = adsl_saf
  )

adae_any_ard |>
  filter(stat_name %in% c("n","N","p")) |>
  mutate(val = map_dbl(stat, 1)) |>
  select(group1_level, stat_name, val)
```

### Practice — combine the overall row with the hierarchy

```{r ex-ae-combine, exercise=TRUE, exercise.lines=12}
# Use bind_ard() (NOT bind_rows()) to combine the any-TEAE row with the
# SOC x PT hierarchy into a single AE ARD.
adae_any_ard <- adae_te |>
  distinct(USUBJID, TRT01A) |>
  mutate(any_teae = "Y") |>
  ard_categorical(by = "TRT01A", variables = "any_teae", denominator = adsl_saf)

ae_full_ard <- ___(adae_any_ard, ae_incidence_ard)

ae_full_ard |> count(context)
```

```{r ex-ae-combine-hint-1}
# bind_ard(adae_any_ard, ae_incidence_ard)
```

```{r ex-ae-combine-solution}
adae_any_ard <- adae_te |>
  distinct(USUBJID, TRT01A) |>
  mutate(any_teae = "Y") |>
  ard_categorical(by = "TRT01A", variables = "any_teae", denominator = adsl_saf)

ae_full_ard <- bind_ard(adae_any_ard, ae_incidence_ard)

ae_full_ard |> count(context)
```

### Practice — serious TEAEs

```{r ex-ae-serious, exercise=TRUE, exercise.lines=12}
# Build the SOC x PT incidence ARD for SERIOUS TEAEs (AESER == "Y").
# Same denominator and id as the all-TEAE version.
ae_serious_ard <- adae |>
  filter(TRTEMFL == "Y" & AESER == ___) |>
  ard_stack_hierarchical(
    by          = "TRT01A",
    variables   = c("AEBODSYS", "AEDECOD"),
    denominator = ___,
    id          = "USUBJID"
  )

ae_serious_ard |> filter(variable == "AEDECOD", stat_name == "n") |> nrow()
```

```{r ex-ae-serious-hint-1}
# AESER == "Y"; denominator = adsl_saf
```

```{r ex-ae-serious-solution}
ae_serious_ard <- adae |>
  filter(TRTEMFL == "Y" & AESER == "Y") |>
  ard_stack_hierarchical(
    by          = "TRT01A",
    variables   = c("AEBODSYS", "AEDECOD"),
    denominator = adsl_saf,
    id          = "USUBJID"
  )

ae_serious_ard |> filter(variable == "AEDECOD", stat_name == "n") |> nrow()
```

```{r quiz-ae, echo=FALSE}
quiz(caption = "AE incidence ARD",
  question("A TEAE incidence ARD shows preferred-term percentages above 100%. What is the cause?",
    answer("The wrong AVISIT filter"),
    answer("Missing `denominator` — N defaulted to the AE row count, which exceeds the subject count",
           correct = TRUE,
           message = "Pass denominator = adsl_saf so N is the safety-population subject count, and id = \"USUBJID\" so events are collapsed to subjects."),
    answer("AESEVN was not derived"),
    answer("The arms were not factor-typed"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question("What does `id = \"USUBJID\"` accomplish in `ard_stack_hierarchical()`?",
    answer("It sorts the output by subject"),
    answer("It counts distinct subjects rather than events, so multiple AEs per subject count once",
           correct = TRUE,
           message = "AE incidence tables report subject counts, not event counts. id collapses repeated events per subject."),
    answer("It adds a USUBJID column to the ARD"),
    answer("It is the denominator"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

---

## Section 3 — Lab summary & change-from-baseline ARD

A "Laboratory Values by Visit" table reports N, mean, SD, median, min, and max
of `AVAL` *and* `CHG` (change from baseline) per `PARAMCD` × `AVISIT` × treatment
arm. This is a **three-level grouping** (`group1`/`group2` plus the parameter),
handled with `.by = c(PARAMCD, AVISIT, TRTA)` in the real package; our mock
demonstrates the two-group case so the structure is clear.

```r
adlb_analysis <- adlb_an |>
  filter(PARAMCD %in% c("ALT", "AST", "BILI", "CREAT"))

lab_ard <- ard_stack(
  adlb_analysis,
  ard_continuous(
    variables = c("AVAL", "CHG"),
    statistic = ~ continuous_summary_fns(c("N","mean","sd","median","min","max"))),
  .by = "TRT01A"
)
```

> **Population reminder:** apply the same population filter to data *and*
> denominator. For BDS labs, also keep baseline rows when the table needs a
> baseline N — filtering on `AVISITN > 0` silently drops baseline.

### Practice — lab AVAL/CHG summary for one parameter

```{r ex-lab-summary, exercise=TRUE, exercise.lines=16}
# Build an ARD of N, mean, sd, median of AVAL and CHG for ALT, by arm.
alt <- adlb_an |> filter(PARAMCD == "ALT")

lab_alt_ard <- ard_continuous(
  data      = alt,
  variables = c(___, ___),          # AVAL and CHG
  by        = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","sd","median"))
)

# Mean CHG of ALT by arm
lab_alt_ard |>
  filter(variable == "CHG", stat_name == ___) |>
  mutate(mean_chg = round(map_dbl(stat, 1), 2)) |>
  select(group1_level, mean_chg)
```

```{r ex-lab-summary-hint-1}
# variables = c("AVAL","CHG"); stat_name == "mean"
```

```{r ex-lab-summary-solution}
alt <- adlb_an |> filter(PARAMCD == "ALT")

lab_alt_ard <- ard_continuous(
  data      = alt,
  variables = c("AVAL", "CHG"),
  by        = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","sd","median"))
)

lab_alt_ard |>
  filter(variable == "CHG", stat_name == "mean") |>
  mutate(mean_chg = round(map_dbl(stat, 1), 2)) |>
  select(group1_level, mean_chg)
```

### Practice — change-from-baseline at a specific visit

```{r ex-lab-visit, exercise=TRUE, exercise.lines=14}
# Mean CHG of CREAT at "Week 24" by arm. Filter to PARAMCD and AVISIT first.
creat_w24 <- adlb_an |>
  filter(PARAMCD == ___, AVISIT == ___)

ard_continuous(
  data = creat_w24, variables = "CHG", by = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","sd"))
) |>
  filter(stat_name == "mean") |>
  mutate(mean_chg = round(map_dbl(stat, 1), 3)) |>
  select(group1_level, mean_chg)
```

```{r ex-lab-visit-hint-1}
# PARAMCD == "CREAT", AVISIT == "Week 24"
```

```{r ex-lab-visit-solution}
creat_w24 <- adlb_an |>
  filter(PARAMCD == "CREAT", AVISIT == "Week 24")

ard_continuous(
  data = creat_w24, variables = "CHG", by = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","sd"))
) |>
  filter(stat_name == "mean") |>
  mutate(mean_chg = round(map_dbl(stat, 1), 3)) |>
  select(group1_level, mean_chg)
```

### Lab toxicity-grade shift

For a shift table you need a baseline category and a post-baseline category.
`ATOXGR` is an integer toxicity grade after the setup conversion. Here we tabulate
the post-baseline grade distribution at a visit, factor-typed so all grades show:

```{r ex-lab-shift, exercise=TRUE, exercise.lines=16}
# Distribution of post-baseline ATOXGR for ALT at "Week 24", by arm.
# Factor-type the grade so every level appears even at n = 0.
alt_tox <- adlb_an |>
  filter(PARAMCD == "ALT", AVISIT == "Week 24", !is.na(ATOXGR)) |>
  mutate(ATOXGR = factor(ATOXGR, levels = c(0L, 1L, 2L, 3L, 4L)))

ard_categorical(
  data      = alt_tox,
  variables = ___,            # the toxicity grade variable
  by        = "TRT01A"
) |>
  filter(stat_name == "n") |>
  mutate(n = map_dbl(stat, 1)) |>
  select(group1_level, variable_level, n)
```

```{r ex-lab-shift-hint-1}
# variables = "ATOXGR"
```

```{r ex-lab-shift-solution}
alt_tox <- adlb_an |>
  filter(PARAMCD == "ALT", AVISIT == "Week 24", !is.na(ATOXGR)) |>
  mutate(ATOXGR = factor(ATOXGR, levels = c(0L, 1L, 2L, 3L, 4L)))

ard_categorical(
  data      = alt_tox,
  variables = "ATOXGR",
  by        = "TRT01A"
) |>
  filter(stat_name == "n") |>
  mutate(n = map_dbl(stat, 1)) |>
  select(group1_level, variable_level, n)
```

---

## Section 4 — Vital signs ARD

Vital signs follow the same continuous pattern as labs, on `advs`. Parameters are
`SYSBP`, `DIABP`, `PULSE`, `WEIGHT`, `HEIGHT`. A common variant is the **worst
post-baseline value** per subject.

```r
vs <- advs |> filter(SAFFL == "Y", !is.na(AVISIT))

vs_ard <- ard_continuous(
  vs |> filter(PARAMCD == "SYSBP"),
  variables = c("AVAL", "CHG"), by = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","sd","median","min","max")))
```

### Practice — systolic BP summary

```{r ex-vs-sbp, exercise=TRUE, exercise.lines=14}
# Summarise AVAL of SYSBP (systolic BP) by arm: N, mean, sd, min, max.
vs <- advs |> filter(SAFFL == "Y", !is.na(AVISIT))

sbp <- vs |> filter(PARAMCD == ___)

ard_continuous(
  data = sbp, variables = "AVAL", by = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","sd","min","max"))
) |>
  filter(stat_name == "mean") |>
  mutate(mean_sbp = round(map_dbl(stat, 1), 1)) |>
  select(group1_level, mean_sbp)
```

```{r ex-vs-sbp-hint-1}
# PARAMCD == "SYSBP"
```

```{r ex-vs-sbp-solution}
vs <- advs |> filter(SAFFL == "Y", !is.na(AVISIT))

sbp <- vs |> filter(PARAMCD == "SYSBP")

ard_continuous(
  data = sbp, variables = "AVAL", by = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","sd","min","max"))
) |>
  filter(stat_name == "mean") |>
  mutate(mean_sbp = round(map_dbl(stat, 1), 1)) |>
  select(group1_level, mean_sbp)
```

### Practice — worst (maximum) post-baseline value

```{r ex-vs-worst, exercise=TRUE, exercise.lines=18}
# For PULSE, take the maximum AVAL per subject post-baseline (AVISITN > 0),
# then summarise across arms.
vs <- advs |> filter(SAFFL == "Y", !is.na(AVISIT))

pulse_worst <- vs |>
  filter(PARAMCD == "PULSE", AVISITN > 0) |>
  group_by(USUBJID, TRT01A) |>
  slice_max(AVAL, n = 1, with_ties = FALSE) |>
  ungroup()

ard_continuous(
  data = pulse_worst, variables = ___, by = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","max"))
) |>
  filter(stat_name == "max") |>
  mutate(max_pulse = map_dbl(stat, 1)) |>
  select(group1_level, max_pulse)
```

```{r ex-vs-worst-hint-1}
# variables = "AVAL"
```

```{r ex-vs-worst-solution}
vs <- advs |> filter(SAFFL == "Y", !is.na(AVISIT))

pulse_worst <- vs |>
  filter(PARAMCD == "PULSE", AVISITN > 0) |>
  group_by(USUBJID, TRT01A) |>
  slice_max(AVAL, n = 1, with_ties = FALSE) |>
  ungroup()

ard_continuous(
  data = pulse_worst, variables = "AVAL", by = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","max"))
) |>
  filter(stat_name == "max") |>
  mutate(max_pulse = map_dbl(stat, 1)) |>
  select(group1_level, max_pulse)
```

---

## Section 5 — Drug exposure ARD

Treatment duration usually lives in ADSL as `TRTDURD` (treatment duration in
days). A simple continuous summary by arm gives the exposure table.

```r
expo_ard <- ard_stack(
  adsl_saf,
  ard_continuous(
    variables = "TRTDURD",
    statistic = ~ continuous_summary_fns(c("N","mean","sd","median","min","max"))),
  .by = "TRT01A", .total_n = TRUE
)
```

### Practice — exposure (treatment duration) summary

```{r ex-expo, exercise=TRUE, exercise.lines=16}
# Build an exposure ARD: summarise TRTDURD by arm with Big-N rows.
# (If TRTDURD is absent in your build, derive it as TRTEDT - TRTSDT + 1.)
adsl_exp <- adsl_saf
if (!"TRTDURD" %in% names(adsl_exp))
  adsl_exp <- adsl_exp |> mutate(TRTDURD = as.numeric(TRTEDT - TRTSDT) + 1)

expo_ard <- ard_stack(
  adsl_exp,
  ard_continuous(variables = ___,
                 statistic = ~ continuous_summary_fns(c("N","mean","sd","median","min","max"))),
  .by = "TRT01A", .total_n = ___
)

expo_ard |>
  filter(variable == "TRTDURD", stat_name == "mean") |>
  mutate(mean_days = round(map_dbl(stat, 1), 1)) |>
  select(group1_level, mean_days)
```

```{r ex-expo-hint-1}
# variables = "TRTDURD"; .total_n = TRUE
```

```{r ex-expo-solution}
adsl_exp <- adsl_saf
if (!"TRTDURD" %in% names(adsl_exp))
  adsl_exp <- adsl_exp |> mutate(TRTDURD = as.numeric(TRTEDT - TRTSDT) + 1)

expo_ard <- ard_stack(
  adsl_exp,
  ard_continuous(variables = "TRTDURD",
                 statistic = ~ continuous_summary_fns(c("N","mean","sd","median","min","max"))),
  .by = "TRT01A", .total_n = TRUE
)

expo_ard |>
  filter(variable == "TRTDURD", stat_name == "mean") |>
  mutate(mean_days = round(map_dbl(stat, 1), 1)) |>
  select(group1_level, mean_days)
```

---

## Section 6 — Concomitant medications ARD

A concomitant-medication table is hierarchical, just like AEs: medication **class**
nesting medication **term**, counting distinct subjects with the safety-population
denominator. ADCM is not always present in `pharmaverseadam`, so we demonstrate
the identical pattern on the AE hierarchy and call out the CM mapping.

The production CM call is:

```r
cm_ard <- adcm |>
  filter(SAFFL == "Y") |>
  distinct(USUBJID, TRT01A, CMCLAS, CMDECOD) |>
  ard_stack_hierarchical(
    by          = "TRT01A",
    variables   = c("CMCLAS", "CMDECOD"),   # class then preferred term
    denominator = adsl_saf,
    id          = "USUBJID"
  )
```

The structure is identical to the AE incidence ARD: `group2` carries the class,
`variable_level` carries the term, and `denominator + id` guarantee subject-level
percentages.

### Practice — a hierarchical subject-count ARD (CM pattern)

```{r ex-cm, exercise=TRUE, exercise.lines=14}
# Practise the CM/AE hierarchical pattern on adae: count distinct subjects by
# AEBODSYS (outer) and AEDECOD (inner), with the subject denominator.
cm_like_ard <- adae_te |>
  distinct(USUBJID, TRT01A, AEBODSYS, AEDECOD) |>
  ard_stack_hierarchical(
    by          = "TRT01A",
    variables   = c("AEBODSYS", "AEDECOD"),
    denominator = ___,
    id          = ___
  )

# Confirm percentages never exceed 100 (p is a proportion 0-1, so x100)
cm_like_ard |>
  filter(stat_name == "p") |>
  summarise(max_pct = max(map_dbl(stat, 1) * 100, na.rm = TRUE))
```

```{r ex-cm-hint-1}
# denominator = adsl_saf, id = "USUBJID"
```

```{r ex-cm-solution}
cm_like_ard <- adae_te |>
  distinct(USUBJID, TRT01A, AEBODSYS, AEDECOD) |>
  ard_stack_hierarchical(
    by          = "TRT01A",
    variables   = c("AEBODSYS", "AEDECOD"),
    denominator = adsl_saf,
    id          = "USUBJID"
  )

cm_like_ard |>
  filter(stat_name == "p") |>
  summarise(max_pct = max(map_dbl(stat, 1), na.rm = TRUE))
```

---

## Section 7 — ARS traceability metadata

For an ARS-aligned submission, each ARD row should carry the **Output ID**,
**Analysis ID**, and **Method ID** that produced it. Base `{cards}` does not add
these, but they are easy to attach post-hoc with `mutate()` + `case_when()`.
When you use `{siera}` or `{arsbridge}` (Lesson 27), these tags are injected
automatically from the ARS spec.

```{r ex-ars-tag, exercise=TRUE, exercise.lines=22}
# Tag demog_ard with ARS metadata: OutputId, AnalysisId, MethodId.
demog_tagged <- demog_ard |>
  mutate(
    OutputId   = "T_DEMOG",
    AnalysisId = case_when(
      variable == "AGE"    ~ "AN_DEMOG_AGE",
      variable == "BMIBL"  ~ "AN_DEMOG_BMI",
      variable == "SEX"    ~ "AN_DEMOG_SEX",
      variable == "AGEGR1" ~ "AN_DEMOG_AGEGR",
      .default = NA_character_
    ),
    MethodId = case_when(
      context == "continuous"  ~ "MTH_SUMMARY_CONTINUOUS",
      context == "categorical" ~ "MTH_COUNT_PCT",
      context == ___           ~ "MTH_SUBJECT_COUNT",   # the Big-N context
      .default = NA_character_
    )
  )

demog_tagged |> distinct(context, MethodId)
```

```{r ex-ars-tag-hint-1}
# The Big-N context value is "total_n"
```

```{r ex-ars-tag-solution}
demog_tagged <- demog_ard |>
  mutate(
    OutputId   = "T_DEMOG",
    AnalysisId = case_when(
      variable == "AGE"    ~ "AN_DEMOG_AGE",
      variable == "BMIBL"  ~ "AN_DEMOG_BMI",
      variable == "SEX"    ~ "AN_DEMOG_SEX",
      variable == "AGEGR1" ~ "AN_DEMOG_AGEGR",
      .default = NA_character_
    ),
    MethodId = case_when(
      context == "continuous"  ~ "MTH_SUMMARY_CONTINUOUS",
      context == "categorical" ~ "MTH_COUNT_PCT",
      context == "total_n"     ~ "MTH_SUBJECT_COUNT",
      .default = NA_character_
    )
  )

demog_tagged |> distinct(context, MethodId)
```

---

## Section 8 — Putting it together & validating

A complete study ARD is just every topic ARD bound together with `bind_ard()`
and validated as a group. Crucially, validation happens on the **ARD**, not on a
rendered table.

### Practice — assemble and validate a study ARD

```{r ex-study, exercise=TRUE, exercise.lines=24}
# Combine demographics + AE + a lab ARD into one study ARD, then validate.
lab_ard <- ard_continuous(
  adlb_an |> filter(PARAMCD %in% c("ALT","AST")),
  variables = c("AVAL","CHG"), by = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","sd")))

study_ard <- ___(demog_ard, ae_incidence_ard, lab_ard)

# Validate the whole package
check_ard_structure(study_ard)
print_ard_conditions(study_ard)

study_ard |> count(context)
```

```{r ex-study-hint-1}
# bind_ard(demog_ard, ae_incidence_ard, lab_ard)
```

```{r ex-study-solution}
lab_ard <- ard_continuous(
  adlb_an |> filter(PARAMCD %in% c("ALT","AST")),
  variables = c("AVAL","CHG"), by = "TRT01A",
  statistic = ~ continuous_summary_fns(c("N","mean","sd")))

study_ard <- bind_ard(demog_ard, ae_incidence_ard, lab_ard)

check_ard_structure(study_ard)
print_ard_conditions(study_ard)

study_ard |> count(context)
```

### The dual-program-the-ARD validation paradigm

Two programmers each build the demographics ARD independently. Instead of
comparing rendered RTFs, flatten each ARD to numeric values and `anti_join`:

```{r ex-dualprog, exercise=TRUE, exercise.lines=20}
# Simulate two programmers' ARDs (here both come from demog_ard, with P2
# rounded slightly differently). Flatten and find discrepancies.
flatten_ard <- function(ard) {
  ard |>
    filter(!is.na(stat_name), map_lgl(stat, ~ is.numeric(.x[1]))) |>
    mutate(stat_value = round(map_dbl(stat, 1), 4)) |>
    select(group1_level, variable, variable_level, stat_name, stat_value)
}

ard_p1 <- flatten_ard(demog_ard)
ard_p2 <- flatten_ard(demog_ard)

# Rows in P1 not matched in P2 (should be empty when ARDs agree)
___(ard_p1, ard_p2)
```

```{r ex-dualprog-hint-1}
# anti_join(ard_p1, ard_p2) returns rows in P1 absent from P2
```

```{r ex-dualprog-solution}
flatten_ard <- function(ard) {
  ard |>
    filter(!is.na(stat_name), map_lgl(stat, ~ is.numeric(.x[1]))) |>
    mutate(stat_value = round(map_dbl(stat, 1), 4)) |>
    select(group1_level, variable, variable_level, stat_name, stat_value)
}

ard_p1 <- flatten_ard(demog_ard)
ard_p2 <- flatten_ard(demog_ard)

anti_join(ard_p1, ard_p2)
```

> **Why this is the paradigm shift:** the ARD is the source of truth. Compare the
> numbers once; the display (RTF/PDF) then only needs a light formatting check
> that its cells match the ARD it was built from.

---

## Section 9 — Production checklist & migration

Use this as a code-review checklist for every clinical ARD.

| Pitfall | Detection | Fix |
|---|---|---|
| Missing `denominator` on AE ARD | percentages > 100% | `denominator = adsl_saf` in every AE/CM/exposure `ard_*()` |
| Inconsistent population filters | percentages computed on different N | apply SAFFL/ITTFL filter identically to data and denominator |
| Missing categorical levels | some arms lack a level row | `factor()` all categoricals with the full level set first |
| Post-baseline-only filter drops baseline | baseline N ≠ post-baseline N | review `AVISITN > 0`; keep baseline rows where needed |
| `bind_rows()` instead of `bind_ard()` | silent column / class loss | always combine ARDs with `bind_ard()` |
| Not checking conditions | silent per-stat errors | call `print_ard_conditions()` after building |
| Wrong `id` | event counts, not subject counts | set `id = "USUBJID"` in `ard_stack_hierarchical()` |
| Forgetting `.total_n = TRUE` | no Big-N for column headers | include `.total_n = TRUE` in `ard_stack()` |

Migration from the legacy stack:

| Old approach | New approach |
|---|---|
| `Tplyr::tplyr_table() |> build()` → wide table | `ard_stack()` → long ARD + gtsummary/tfrmt |
| `rtables::build_table()` → table object | `ard_*()` → ARD + display layer |
| recompute for each display | one ARD, many displays |
| dual program = compare RTF files | dual program = compare ARD flat files |
| study archive = pile of RTFs | study archive = one RDS per topic |

```{r quiz-pipeline, echo=FALSE}
quiz(caption = "Pipeline & validation",
  question("In the ARD-first workflow, what do two independent programmers compare during validation?",
    answer("The rendered RTF tables, cell by cell"),
    answer("Their flattened ARDs (numeric values), e.g. via anti_join / full_join",
           correct = TRUE,
           message = "You validate the ARD, not the display. Once the numbers agree, the display check is a light formatting confirmation."),
    answer("The define.xml files"),
    answer("The SAS logs"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question("Why use `bind_ard()` rather than `bind_rows()` to combine ARDs?",
    answer("bind_ard() is faster"),
    answer("bind_ard() preserves the card class and aligns the full ARD column set",
           correct = TRUE,
           message = "bind_rows() can silently drop the card class and misalign columns across ARDs from different functions."),
    answer("bind_rows() cannot handle list columns"),
    answer("There is no difference"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

---

## Capstone — a complete safety-table ARD

Assemble the full safety overview ARD: the "any TEAE" overall row, serious-TEAE
overall row, and the SOC × PT hierarchy — all subject-counted against the safety
population.

```{r ex-capstone, exercise=TRUE, exercise.lines=34}
# Step 1: any-TEAE overall row
any_ard <- adae_te |>
  distinct(USUBJID, TRT01A) |>
  mutate(any_teae = "Y") |>
  ard_categorical(by = "TRT01A", variables = "any_teae", denominator = adsl_saf)

# Step 2: any-serious-TEAE overall row
sae_ard <- adae_te |>
  filter(AESER == "Y") |>
  distinct(USUBJID, TRT01A) |>
  mutate(any_sae = "Y") |>
  ard_categorical(by = "TRT01A", variables = "any_sae", denominator = ___)

# Step 3: SOC x PT hierarchy
soc_pt_ard <- adae_te |>
  ard_stack_hierarchical(by = "TRT01A", variables = c("AEBODSYS","AEDECOD"),
                   denominator = adsl_saf, id = "USUBJID")

# Step 4: combine with bind_ard()
safety_ard <- ___(any_ard, sae_ard, soc_pt_ard)

# Step 5: validate and inspect the any-TEAE percentages
print_ard_conditions(safety_ard)
safety_ard |>
  filter(variable == "any_teae", stat_name == "p") |>
  mutate(arm = unlist(group1_level), pct = round(map_dbl(stat, 1) * 100, 1)) |>
  select(arm, pct)
```

```{r ex-capstone-hint-1}
# Step 2 denominator = adsl_saf
# Step 4 bind_ard(any_ard, sae_ard, soc_pt_ard)
```

```{r ex-capstone-solution}
any_ard <- adae_te |>
  distinct(USUBJID, TRT01A) |>
  mutate(any_teae = "Y") |>
  ard_categorical(by = "TRT01A", variables = "any_teae", denominator = adsl_saf)

sae_ard <- adae_te |>
  filter(AESER == "Y") |>
  distinct(USUBJID, TRT01A) |>
  mutate(any_sae = "Y") |>
  ard_categorical(by = "TRT01A", variables = "any_sae", denominator = adsl_saf)

soc_pt_ard <- adae_te |>
  ard_stack_hierarchical(by = "TRT01A", variables = c("AEBODSYS","AEDECOD"),
                   denominator = adsl_saf, id = "USUBJID")

safety_ard <- bind_ard(any_ard, sae_ard, soc_pt_ard)

print_ard_conditions(safety_ard)
safety_ard |>
  filter(variable == "any_teae", stat_name == "p") |>
  mutate(arm = unlist(group1_level), pct = round(map_dbl(stat, 1) * 100, 1)) |>
  select(arm, pct)
```

---

## Key takeaways

- A complete clinical ARD pipeline covers **demographics, TEAE incidence
  (hierarchical + any-AE overall), lab change-from-baseline, vital signs,
  exposure, and concomitant medications**.
- Always pass `denominator = adsl_saf` (and `id = "USUBJID"`) for AE/CM/exposure
  tables so `N` is the subject count, not the event count, and percentages stay
  ≤ 100%.
- Always `factor()` categorical variables with the full level set so zero-count
  levels still appear as `n = 0` rows.
- Use `ard_stack()` with `.by`, `.overall = TRUE`, and `.total_n = TRUE` to build
  per-arm columns, an Overall column, and Big-N (`context == "total_n"`) header
  rows in one call.
- Combine ARDs with **`bind_ard()`**, never `bind_rows()`.
- Validate with `check_ard_structure()` and `print_ard_conditions()`; pull values
  with `get_ard_statistics()` and `map_dbl(stat, 1)`.
- For ARS traceability, add `OutputId`, `AnalysisId`, `MethodId` post-hoc — or let
  `siera`/`arsbridge` inject them automatically.
- **Validate the ARD, not the display:** dual-program the ARD, flatten, and
  `anti_join` to find discrepancies.

## What's next

**Lesson 27 — `{cardx}`: inferential ARDs.** You extend the same ARD structure
to statistical tests and models: t-tests, chi-squared, Wilcoxon, regression,
and survival. The statistics come from model objects rather than simple
summaries, but the ARD columns — and everything you learned about combining and
validating ARDs — stay identical. We also introduce `{siera}` (read an ARS JSON,
auto-generate ARD programs) and `{arsbridge}` (annotated shells → formatted
tables end-to-end).

---

## Self-check questions

1. A TEAE incidence ARD shows percentages above 100% for some preferred terms.
   What caused this, and what are the two arguments that fix it?
2. Write the full `ard_stack_hierarchical()` call for *serious* TEAEs by SOC and PT
   against the safety-population denominator.
3. Your demographics ARD is missing the ">80" age-group row for the High Dose
   arm because no subjects fell in that group. What is the fix, and where in the
   pipeline does it go?
4. How would you compare two programmers' demographics ARDs to find
   discrepancies without ever rendering a table?
5. Translate to cards: "Compute N, Mean, SD, Median, Min, Max of AVAL and CHG by
   PARAMCD × AVISIT × treatment arm for the analysis-flagged safety population."
6. Which `context` value identifies the Big-N rows added by `.total_n = TRUE`,
   and what consumes them downstream?

---

## Glossary

- **`ard_stack_hierarchical()`** — nested categorical tabulation; computes
  distinct-subject counts at each nesting level (e.g. SOC then PT).
- **`ard_dichotomous()`** — single yes/no count (n, N, p) for a flag variable,
  honouring the subject denominator.
- **`continuous_summary_fns()`** — returns the named list of default summary
  functions; subset it with a character vector of statistic names.
- **`get_ard_statistics()`** — extract a named list of `stat` values matching a
  filter expression.
- **`check_ard_structure()`** — confirm required ARD columns exist and `stat` is
  a list column.
- **`print_ard_conditions()`** — report any warnings/errors carried in the ARD.
- **`denominator = adsl_saf`** — pass the ADSL safety population so `N` is the
  subject count, giving correct AE/CM/exposure percentages.
- **`id = "USUBJID"`** — count distinct subjects, not events, in hierarchical
  ARDs.
- **`.overall = TRUE`** — add an "Overall" pooled group to `ard_stack()`.
- **`.total_n = TRUE`** — add Big-N per group; rows have `context == "total_n"`.
- **`bind_ard()`** — combine ARDs while preserving the card class and columns.
- **`shuffle_ard()`** — (production) pivot `group1_level` to true wide columns,
  required by tfrmt; gtsummary consumes the long format directly.
- **`OutputId` / `AnalysisId` / `MethodId`** — ARS traceability columns linking
  ARD rows to their output, analysis, and statistical method.
- **TRT01A** — actual treatment in ADAE (renamed from `TRTA` in setup).
- **TRTEMFL** — treatment-emergent flag in ADAE (`"Y"` = treatment-emergent).
- **AESER** — serious AE flag in ADAE.
- **AEBODSYS / AEDECOD** — System Organ Class / Preferred Term in ADAE.
- **PARAMCD / AVISIT / AVAL / CHG** — parameter code, analysis visit, analysis
  value, and change-from-baseline in BDS ADaMs.
- **ATOXGR** — analysis toxicity grade (integer after the setup conversion).
- **TRTDURD** — treatment duration in days (ADSL), the exposure-table variable.
