---
title: "Lesson 27 — cardx: Inferential & Statistical ARDs"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Extend Analysis Results Datasets (ARDs) with inferential statistics using the
  cardx package: t-tests, chi-squared and Fisher's exact tests, Wilcoxon rank-sum,
  proportion and mean confidence intervals, regression coefficients (lm/glm/coxph),
  Kaplan-Meier survival estimates, log-rank tests, ANOVA, estimated marginal means,
  and standardized mean differences — all returned in the same machine-readable ARD
  structure as cards, ready to combine into display-ready datasets for TLF automation.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 27_cardx.Rmd
# PURPOSE       : cardx package — Inferential & Statistical ARDs:
#                 ard_stats_t_test(), ard_stats_chisq_test(),
#                 ard_stats_fisher_test(), ard_stats_wilcox_test(),
#                 ard_categorical_ci(), ard_continuous_ci(), ard_regression()
#                 (lm/glm/coxph/mmrm), ard_survival_survfit(),
#                 ard_survival_survdiff(), ard_stats_aov(), ard_emmeans_emmeans(), ard_smd_smd().
#                 Mapping SAS PROC TTEST/FREQ/LIFETEST/PHREG/MIXED/LOGISTIC/GLM
#                 to cardx, and the ADTTE CNSR censoring convention.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-06-22
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(purrr)
library(pharmaverseadam)
library(cards)
library(cardx)
library(survival)
library(emmeans)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo","Xanomeline Low Dose","Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

advs <- pharmaverseadam::advs

# ADTTE — time-to-event. pharmaverseadam exports adtte_onco (CDISCPILOT01, same
# 254 subjects as adsl); no TRTA column, so derive it from adsl by USUBJID below.
adtte <- pharmaverseadam::adtte_onco
if (!"TRTA" %in% names(adtte) && "TRT01A" %in% names(adtte))
  adtte <- rename(adtte, TRTA = TRT01A)
if (!"TRTA" %in% names(adtte))
  adtte <- left_join(adtte, select(adsl, USUBJID, TRTA = TRT01P), by = "USUBJID")

# ── Helper + pre-built objects for exercises (REAL {cardx}) ───────────────────
# Notes that recur in this lesson:
#  * cardx ARDs share the cards column set; `stat` is a LIST column. Many rows
#    are CHARACTER (method, alternative): filter to numeric stat_names before
#    map_dbl(), or format with map_chr() when showing every statistic.
#  * `ard_categorical_ci()` (proportion CI) takes value = list(VAR = "level")
#    and methods named with DOTS ("clopper.pearson", "wilson", "wald", ...).
#    Its `estimate` is a proportion on the 0-1 scale.
#  * `ard_survival_survdiff(formula, data)` takes a FORMULA, not a survdiff object.
#  * ANOVA is `ard_stats_aov()`, SMD is `ard_smd_smd()`.

# Pull the first element of each stat list cell to plain numeric.
.pull1 <- function(stat) map_dbl(stat, 1)

# Two-arm subset (t-test / Wilcoxon / Fisher / SMD need exactly 2 levels).
adsl_2arm <- adsl |>
  filter(TRT01P %in% c("Xanomeline Low Dose","Xanomeline High Dose")) |>
  mutate(TRT01P = droplevels(TRT01P))

# Overall-survival time-to-event subset.
adtte_os <- adtte |> filter(PARAMCD == "OS")
if (nrow(adtte_os) == 0) adtte_os <- adtte |> filter(PARAMCD == unique(adtte$PARAMCD)[1])
```

## Overview

In Lesson 25 you built **descriptive** ARDs with `{cards}`: counts, means, medians. Clinical reporting also needs **inferential** statistics — p-values, confidence intervals, hazard ratios, survival estimates. **`{cardx}`** ("cards extra") adds exactly these, wrapping the relevant base-R and `{survival}` machinery into the *same* ARD structure.

```
Data → descriptive ARD (cards)  ┐
                                 ├─ bind_ard() → display-ready ARD → gtsummary/tfrmt
Data → inferential ARD (cardx)  ┘
```

The payoff: one row per statistic, the same `group1 / variable / stat_name / stat / stat_label` columns, the same list-column `stat`. The only thing that changes is the **`context`** value (`"stats_t_test"`, `"survival_survfit"`, `"regression"`, …) so you can tell which engine produced each row.

> **Reminder:** `stat` is a **list column**. You cannot do arithmetic on it directly. cardx examples use the `purrr::map_dbl(stat, 1)` idiom to pull a plain numeric vector; in this tutorial a helper `.pull1(stat)` does the same. `unlist(stat)` also works for simple numeric stats.

### What cardx wraps

cardx imports statistical computation from a wide range of R packages. The naming pattern is `ard_<engine>_<model>()`.

| Package | What cardx wraps |
|---|---|
| `{stats}` (base R) | `t.test()`, `chisq.test()`, `fisher.test()`, `wilcox.test()`, `lm()`, `glm()`, `aov()` |
| `{survival}` | `survfit()`, `coxph()`, `survdiff()` |
| `{lme4}` / `{mmrm}` | `lmer()`, `mmrm()` (mixed models / MMRM) |
| `{emmeans}` | Estimated marginal means and contrasts |
| `{smd}` | Standardized mean differences |
| `{broom.helpers}` | Tidy model-extraction backbone for `ard_regression()` |

### Trial datasets

All exercises use **CDISCPILOT01** (the Alzheimer's / Xanomeline study) from `pharmaverseadam`: `adsl` (safety population, 3 arms), `adae` (`TRT01A`, `AESER`, `TRTEMFL`), `adlb`, `advs`, and `adtte` (time-to-event: `PARAMCD`, `AVAL`, `CNSR`, `TRTA`).

> **Censoring convention (memorize this now):** In ADTTE, **`CNSR = 1` means the subject was CENSORED** (event did *not* occur). But `survival::Surv(time, event)` expects the second argument to be **`1 = event occurred`**. They are opposite. Always build the survival object with `Surv(AVAL, 1 - CNSR)`. We return to this in Section 9.

---

## Section 1 — t-test: comparing continuous variables between arms

The most common use is adding a **p-value** column to a demographics table. `ard_stats_t_test()` wraps `stats::t.test()` and returns every component — estimate, statistic, df, CI bounds, method, alternative — as its own ARD row.

**SAS equivalent:** `PROC TTEST DATA=adsl_2arm; CLASS arm; VAR age; RUN;` — but SAS sends the result to a *report*, not a structured dataset you can filter and combine.

```r
# Two-sample Welch t-test on AGE between the two Xanomeline arms
age_ttest <- ard_stats_t_test(adsl_2arm, by = "TRT01P", variables = "AGE")

# Some rows (method, alternative) are CHARACTER, so filter to the numeric
# statistics before map_dbl() — mixing types in one map_dbl() errors.
age_ttest |>
  filter(stat_name %in% c("estimate","statistic","p.value","conf.low","conf.high")) |>
  mutate(value = map_dbl(stat, 1)) |>
  select(stat_name, stat_label, value)
```

The full ARD still carries a `method` row telling you *which* variant ran ("Welch Two Sample t-test" vs "Two Sample t-test") — traceability SAS does not give you by default; pull it with `filter(stat_name == "method")`.

```r
# Paired / one-sample test (mean change vs 0):
ard_stats_t_test(advs_sub, variables = "CHG", mu = 0)

# One-sided:    alternative = "less"
# Student's t:  var.equal = TRUE   (pooled variance, not Welch)
```

### Practice — t-test ARD

```{r ex-ttest, exercise=TRUE, exercise.lines=10}
# adsl_2arm is pre-built: SAFFL=="Y" filtered to the two Xanomeline arms.
# Run a Welch two-sample t-test on AGE between arms, then show the numeric stats.

age_ttest <- ard_stats_t_test(
  data      = adsl_2arm,
  by        = ___,
  variables = ___
)

age_ttest |>
  filter(stat_name %in% c("estimate","statistic","p.value","conf.low","conf.high")) |>
  mutate(value = ___(stat, 1)) |>
  select(stat_name, stat_label, value)
```

```{r ex-ttest-hint-1}
# by = "TRT01P", variables = "AGE"
# Pull list-column values with map_dbl(stat, 1)  (or .pull1(stat))
```

```{r ex-ttest-solution}
age_ttest <- ard_stats_t_test(
  data      = adsl_2arm,
  by        = "TRT01P",
  variables = "AGE"
)

age_ttest |>
  filter(stat_name %in% c("estimate","statistic","p.value","conf.low","conf.high")) |>
  mutate(value = map_dbl(stat, 1)) |>
  select(stat_name, stat_label, value)
```

### Practice — extract just the p-value

```{r ex-ttest-pval, exercise=TRUE, exercise.lines=8}
# A demographics table only needs the p-value. Filter to it and pull the number.

ard_stats_t_test(adsl_2arm, by = "TRT01P", variables = "WEIGHTBL") |>
  filter(stat_name == ___) |>
  mutate(p = round(.pull1(stat), 4)) |>
  select(variable, stat_label, p)
```

```{r ex-ttest-pval-hint-1}
# stat_name == "p.value"
```

```{r ex-ttest-pval-solution}
ard_stats_t_test(adsl_2arm, by = "TRT01P", variables = "WEIGHTBL") |>
  filter(stat_name == "p.value") |>
  mutate(p = round(.pull1(stat), 4)) |>
  select(variable, stat_label, p)
```

---

## Section 2 — Wilcoxon rank-sum test

For non-normally distributed variables, `ard_stats_wilcox_test()` wraps `stats::wilcox.test()` (the Mann-Whitney U test). It returns `statistic` (W), `p.value`, and `method`. Like the t-test, `by` must have **exactly 2 levels**.

**SAS equivalent:** `PROC NPAR1WAY DATA=adsl_2arm WILCOXON; CLASS arm; VAR bmibl; RUN;`

```{r ex-wilcox, exercise=TRUE, exercise.lines=10}
# BMIBL is right-skewed; a rank-based test is more appropriate than a t-test.
# Compare the two Xanomeline arms on BMIBL.

ard_stats_wilcox_test(
  data      = adsl_2arm,
  by        = ___,
  variables = ___
) |>
  filter(stat_name %in% c("statistic","p.value")) |>   # skip the character rows
  mutate(value = .pull1(stat)) |>
  select(stat_name, stat_label, value)
```

```{r ex-wilcox-hint-1}
# by = "TRT01P", variables = "BMIBL"
```

```{r ex-wilcox-solution}
ard_stats_wilcox_test(
  data      = adsl_2arm,
  by        = "TRT01P",
  variables = "BMIBL"
) |>
  filter(stat_name %in% c("statistic","p.value")) |>
  mutate(value = .pull1(stat)) |>
  select(stat_name, stat_label, value)
```

---

## Section 3 — Chi-squared and Fisher's exact tests

For **categorical** variables in a demographics table. `ard_stats_chisq_test()` accepts a *vector* of variables and tests each against the `by` grouping. `ard_stats_fisher_test()` is preferred when expected cell counts are small (any expected count < 5).

**SAS equivalent:** `PROC FREQ DATA=adsl; TABLES arm * sex / CHISQ EXACT; RUN;`

```{r ex-chisq, exercise=TRUE, exercise.lines=10}
# Chi-squared test of SEX and AGEGR1 distributions across all 3 arms (use adsl).
# chisq accepts a vector of variables.

ard_stats_chisq_test(
  data      = adsl,
  by        = "TRT01P",
  variables = c(___, ___)
) |>
  filter(stat_name == "p.value") |>
  mutate(p = round(.pull1(stat), 4)) |>
  select(variable, p)
```

```{r ex-chisq-hint-1}
# variables = c("SEX", "AGEGR1")
```

```{r ex-chisq-solution}
ard_stats_chisq_test(
  data      = adsl,
  by        = "TRT01P",
  variables = c("SEX", "AGEGR1")
) |>
  filter(stat_name == "p.value") |>
  mutate(p = round(.pull1(stat), 4)) |>
  select(variable, p)
```

### Practice — Fisher's exact for a small table

```{r ex-fisher, exercise=TRUE, exercise.lines=9}
# Fisher's exact test on SEX between the two Xanomeline arms.

ard_stats_fisher_test(
  data      = adsl_2arm,
  by        = "TRT01P",
  variables = ___
) |>
  filter(stat_name == "p.value") |>
  mutate(p = round(.pull1(stat), 4)) |>
  select(variable, stat_label, p)
```

```{r ex-fisher-hint-1}
# variables = "SEX"
```

```{r ex-fisher-solution}
ard_stats_fisher_test(
  data      = adsl_2arm,
  by        = "TRT01P",
  variables = "SEX"
) |>
  filter(stat_name == "p.value") |>
  mutate(p = round(.pull1(stat), 4)) |>
  select(variable, stat_label, p)
```

### Quiz — univariate tests

```{r quiz-tests, echo=FALSE}
quiz(caption = "Univariate Comparison Tests",
  question("You want to test whether RACE distribution differs across 3 treatment arms. Which cardx function fits?",
    answer("ard_stats_t_test() — it handles multiple groups"),
    answer("ard_stats_chisq_test() — tests association between two categorical variables, any number of levels",
           correct = TRUE,
           message = "Chi-squared (and Fisher) work with any number of levels. t-test and Wilcoxon are for continuous outcomes and require exactly 2 groups."),
    answer("ard_stats_wilcox_test() — for non-normal categorical data"),
    answer("ard_categorical_ci() — it computes p-values for categories"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question("When is ard_stats_fisher_test() preferred over ard_stats_chisq_test()?",
    answer("When the outcome is continuous"),
    answer("When expected cell counts are small (sparse table)", correct = TRUE,
           message = "Fisher's exact test does not rely on the large-sample chi-squared approximation, so it is preferred when any expected cell count is below ~5."),
    answer("When there are more than 2 treatment arms"),
    answer("When the data are normally distributed"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

---

## Section 4 — Proportion confidence intervals: `ard_categorical_ci()`

One of the most practically important cardx functions for pharma reporting: confidence intervals for a proportion (response rate, AE incidence, …) with a choice of CI methods.

| `method` | Method name | Notes |
|---|---|---|
| `"wilson"` | Wilson score interval | **Recommended** for clinical reporting; handles small n and extremes |
| `"wilsoncc"` | Wilson with continuity correction | More conservative |
| `"clopper-pearson"` | Clopper-Pearson "exact" | Conservative; common in regulatory submissions |
| `"wald"` | Wald (normal approx.) | Simplest, poor at extremes |
| `"agresti-coull"` | Agresti-Coull | Good for small N |
| `"jeffreys"` | Jeffreys (Bayesian) | Symmetric, good properties |

> The success level is named per variable: `value = list(<var> = "<level>")`. Method names use **hyphens** (`"clopper-pearson"`, `"agresti-coull"`), not underscores.

**SAS equivalent:** base SAS has no Wilson/Clopper-Pearson by default — you need `PROC FREQ ... / BINOMIAL(CL=WILSON);`. This is an area where R is strictly more convenient.

```r
adsl_resp <- adsl |>
  mutate(RESPONDER = factor(if_else(DTHFL == "Y", "N", "Y"), levels = c("Y","N")))

ard_categorical_ci(
  adsl_resp, by = "TRT01P", variables = "RESPONDER",
  value = list(RESPONDER = "Y"), method = "wilson"
)
```

> **Important: cardx returns the proportion `estimate` on a 0–1 scale here** (e.g. `0.481`), unlike `ard_categorical()`'s `p` which is a 0–100 percentage. Multiply by 100 yourself for display.

### Practice — Wilson CI for an AE-experience rate

```{r ex-prop-ci, exercise=TRUE, exercise.lines=14}
# Build a subject-level "any AE" flag, then get a Wilson 95% CI per arm.
anyae <- adsl |>
  mutate(ANYAE = factor(
    if_else(USUBJID %in% adae$USUBJID, "Y", "N"), levels = c("Y","N")))

ard_categorical_ci(
  data      = anyae,
  by        = "TRT01P",
  variables = "ANYAE",
  value     = ___,            # which level is the "success": list(ANYAE = "Y")
  method    = ___             # use Wilson
) |>
  filter(stat_name %in% c("estimate","conf.low","conf.high")) |>
  mutate(arm = map_chr(group1_level, \(x) as.character(x[[1]])),
         value = round(.pull1(stat), 3)) |>
  select(arm, stat_name, value)
```

```{r ex-prop-ci-hint-1}
# value = list(ANYAE = "Y"), method = "wilson"
```

```{r ex-prop-ci-solution}
anyae <- adsl |>
  mutate(ANYAE = factor(
    if_else(USUBJID %in% adae$USUBJID, "Y", "N"), levels = c("Y","N")))

ard_categorical_ci(
  data      = anyae,
  by        = "TRT01P",
  variables = "ANYAE",
  value     = list(ANYAE = "Y"),
  method    = "wilson"
) |>
  filter(stat_name %in% c("estimate","conf.low","conf.high")) |>
  mutate(arm = map_chr(group1_level, \(x) as.character(x[[1]])),
         value = round(.pull1(stat), 3)) |>
  select(arm, stat_name, value)
```

### Practice — compare CI methods

```{r ex-prop-methods, exercise=TRUE, exercise.lines=16}
# Compute the Placebo "any AE" CI three ways: wilson, clopper-pearson, wald.
# Compare the lower/upper bounds. (Method names use hyphens.)
anyae <- adsl |>
  mutate(ANYAE = factor(
    if_else(USUBJID %in% adae$USUBJID, "Y", "N"), levels = c("Y","N"))) |>
  filter(TRT01P == "Placebo")

purrr::map_dfr(c("wilson","clopper-pearson","wald"), function(m) {
  ard_categorical_ci(anyae, variables = "ANYAE", value = list(ANYAE = "Y"), method = ___) |>
    filter(stat_name %in% c("conf.low","conf.high")) |>
    mutate(method = m, value = round(.pull1(stat), 3)) |>
    select(method, stat_name, value)
})
```

```{r ex-prop-methods-hint-1}
# method = m  (the variable being iterated over)
```

```{r ex-prop-methods-solution}
anyae <- adsl |>
  mutate(ANYAE = factor(
    if_else(USUBJID %in% adae$USUBJID, "Y", "N"), levels = c("Y","N"))) |>
  filter(TRT01P == "Placebo")

purrr::map_dfr(c("wilson","clopper-pearson","wald"), function(m) {
  ard_categorical_ci(anyae, variables = "ANYAE", value = list(ANYAE = "Y"), method = m) |>
    filter(stat_name %in% c("conf.low","conf.high")) |>
    mutate(method = m, value = round(.pull1(stat), 3)) |>
    select(method, stat_name, value)
})
```

---

## Section 5 — Mean confidence intervals: `ard_continuous_ci()`

For a continuous variable's mean CI (independent of `ard_continuous()`'s default SEM-based interval). Methods: `"t.test"` (t-distribution) and `"wilcox.test"` (Hodges-Lehmann).

**SAS equivalent:** `PROC MEANS DATA=advs MEAN CLM; CLASS trta; VAR chg; RUN;`

```{r ex-cont-ci, exercise=TRUE, exercise.lines=14}
# 95% t-based CI for mean weight CHG at Week 24 by arm.
advs_wk24 <- advs |>
  filter(PARAMCD == "WEIGHT", AVISIT == "Week 24") |>
  filter(!is.na(CHG))

ard_continuous_ci(
  data       = advs_wk24,
  by         = ___,
  variables  = ___,
  method     = "t.test",
  conf.level = 0.95
) |>
  filter(stat_name %in% c("estimate","conf.low","conf.high")) |>   # numeric rows only
  mutate(arm = map_chr(group1_level, \(x) as.character(x[[1]])),
         value = round(.pull1(stat), 2)) |>
  select(arm, stat_name, value)
```

```{r ex-cont-ci-hint-1}
# by = "TRTA", variables = "CHG"
```

```{r ex-cont-ci-solution}
advs_wk24 <- advs |>
  filter(PARAMCD == "WEIGHT", AVISIT == "Week 24") |>
  filter(!is.na(CHG))

ard_continuous_ci(
  data       = advs_wk24,
  by         = "TRTA",
  variables  = "CHG",
  method     = "t.test",
  conf.level = 0.95
) |>
  filter(stat_name %in% c("estimate","conf.low","conf.high")) |>   # numeric rows only
  mutate(arm = map_chr(group1_level, \(x) as.character(x[[1]])),
         value = round(.pull1(stat), 2)) |>
  select(arm, stat_name, value)
```

---

## Section 6 — Regression: `ard_regression()`

`ard_regression()` is a **universal converter**: hand it any fitted model object — `lm`, `glm`, `coxph`, `mmrm` — and it returns an ARD of coefficients with SEs, CIs, and p-values. Reference factor levels appear with `estimate = 0`, matching SAS CLASS reference-cell parameterization.

> Coefficients are on the model's **native scale**: identity for `lm`, **log-odds** for logistic `glm`, **log-HR** for `coxph`. Exponentiate downstream for odds ratios / hazard ratios.

### Linear regression — ANCOVA-style

**SAS equivalent:** `PROC GLM` / `PROC MIXED` with `CLASS trta sex;` and a `MODEL chg = base trta age sex / SOLUTION;`.

```{r ex-reg-lm, exercise=TRUE, exercise.lines=16}
# ANCOVA of weight change at Week 24: CHG ~ BASE + TRTA + AGE + SEX.
advs_wk24 <- advs |>
  filter(PARAMCD == "WEIGHT", AVISIT == "Week 24", SAFFL == "Y") |>
  filter(!is.na(CHG), !is.na(BASE))

model_lm <- lm(CHG ~ BASE + TRTA + AGE + SEX, data = advs_wk24)

ard_regression(___) |>
  filter(stat_name %in% c("estimate","p.value")) |>
  mutate(value = round(.pull1(stat), 4)) |>
  select(variable, stat_name, value)
```

```{r ex-reg-lm-hint-1}
# ard_regression(model_lm) — pass the fitted model object directly
```

```{r ex-reg-lm-solution}
advs_wk24 <- advs |>
  filter(PARAMCD == "WEIGHT", AVISIT == "Week 24", SAFFL == "Y") |>
  filter(!is.na(CHG), !is.na(BASE))

model_lm <- lm(CHG ~ BASE + TRTA + AGE + SEX, data = advs_wk24)

ard_regression(model_lm) |>
  filter(stat_name %in% c("estimate","p.value")) |>
  mutate(value = round(.pull1(stat), 4)) |>
  select(variable, stat_name, value)
```

### Logistic regression — odds ratios

**SAS equivalent:** `PROC LOGISTIC DATA=adsl; CLASS trta sex; MODEL resp(EVENT='1') = trta age sex; ODDSRATIO trta; RUN;`

```{r ex-reg-glm, exercise=TRUE, exercise.lines=18}
# Logistic model of "any AE" on treatment + age + sex; report odds ratios.
adsl_log <- adsl |>
  mutate(ANYAE_NUM = if_else(USUBJID %in% adae$USUBJID, 1L, 0L))

model_glm <- glm(
  ANYAE_NUM ~ TRT01P + AGE + SEX,
  data   = adsl_log,
  family = binomial(link = "logit")
)

# ard_regression returns log-odds; exponentiate the estimate for the OR.
ard_regression(model_glm) |>
  filter(stat_name == ___) |>
  mutate(OR = round(exp(.pull1(stat)), 3)) |>
  select(variable, OR)
```

```{r ex-reg-glm-hint-1}
# filter(stat_name == "estimate"), then OR = exp(estimate)
```

```{r ex-reg-glm-solution}
adsl_log <- adsl |>
  mutate(ANYAE_NUM = if_else(USUBJID %in% adae$USUBJID, 1L, 0L))

model_glm <- glm(
  ANYAE_NUM ~ TRT01P + AGE + SEX,
  data   = adsl_log,
  family = binomial(link = "logit")
)

ard_regression(model_glm) |>
  filter(stat_name == "estimate") |>
  mutate(OR = round(exp(.pull1(stat)), 3)) |>
  select(variable, OR)
```

### Cox proportional hazards — hazard ratios

**SAS equivalent:** `PROC PHREG DATA=adtte_os; CLASS trta sex; MODEL aval * cnsr(1) = trta age sex; HAZARDRATIO trta; RUN;`

Note the SAS `cnsr(1)` — it tells PHREG that **`CNSR = 1` is the censoring value**. The R analogue is `Surv(AVAL, 1 - CNSR)`.

```{r ex-reg-cox, exercise=TRUE, exercise.lines=16}
# Cox model on OS: HR for treatment, adjusted for AGE and SEX.
# CRITICAL: CNSR=1 means censored, so the event indicator is 1 - CNSR.

cox_fit <- coxph(
  Surv(AVAL, 1 - CNSR) ~ TRTA + AGE + SEX,
  data = adtte_os
)

ard_regression(cox_fit) |>
  filter(stat_name == "estimate") |>
  mutate(HR = round(exp(.pull1(stat)), 3)) |>   # exp(log-HR) = HR
  select(variable, HR)
```

```{r ex-reg-cox-hint-1}
# The Surv() call is already correct: Surv(AVAL, 1 - CNSR).
# estimate is the log-HR; HR = exp(estimate).
```

```{r ex-reg-cox-solution}
cox_fit <- coxph(
  Surv(AVAL, 1 - CNSR) ~ TRTA + AGE + SEX,
  data = adtte_os
)

ard_regression(cox_fit) |>
  filter(stat_name == "estimate") |>
  mutate(HR = round(exp(.pull1(stat)), 3)) |>
  select(variable, HR)
```

### Quiz — regression

```{r quiz-reg, echo=FALSE}
quiz(caption = "Regression ARDs",
  question("ard_regression() on a coxph model returns estimate = 0.182 for TRTAXanomeline High Dose. What is the hazard ratio?",
    answer("0.182"),
    answer("exp(0.182) ≈ 1.20", correct = TRUE,
           message = "coxph coefficients are log-HRs. HR = exp(log-HR) = exp(0.182) ≈ 1.20."),
    answer("1 - 0.182 = 0.818"),
    answer("log(0.182)"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question("In an ard_regression() ARD, a factor's reference level appears with estimate = 0. Why?",
    answer("It is a data error and should be removed"),
    answer("Reference-cell coding sets the reference level as the baseline, so its effect is 0 by definition",
           correct = TRUE,
           message = "Like SAS CLASS reference parameterization, the reference level is the comparator; all other levels are estimated relative to it."),
    answer("Because the model failed to converge"),
    answer("Because the p-value was non-significant"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

---

## Section 7 — Kaplan-Meier survival: `ard_survival_survfit()`

`ard_survival_survfit()` turns a `survfit` object into an ARD. Two modes:

- **`times = c(...)`** → survival probability at those time points (x-day/x-year survival), with CI and N-at-risk.
- **`probs = 0.5`** → quantile of the survival distribution (0.5 = **median survival**), with CI.

**SAS equivalent:** `PROC LIFETEST DATA=adtte_os METHOD=KM; TIME aval * cnsr(1); STRATA trta; ODS OUTPUT Quartiles=... SurvivalPlot=...; RUN;`

```{r ex-km-times, exercise=TRUE, exercise.lines=14}
# x-day survival at days 180 and 365 by treatment arm.
# Always Surv(AVAL, 1 - CNSR): CNSR=1 is censored, Surv wants 1=event.

km_fit <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)

ard_survival_survfit(km_fit, times = c(___, ___)) |>
  filter(stat_name == "estimate") |>
  mutate(surv = round(.pull1(stat), 3)) |>
  select(group1_level, stat_label, surv)
```

```{r ex-km-times-hint-1}
# times = c(180, 365)
```

```{r ex-km-times-solution}
km_fit <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)

ard_survival_survfit(km_fit, times = c(180, 365)) |>
  filter(stat_name == "estimate") |>
  mutate(surv = round(.pull1(stat), 3)) |>
  select(group1_level, stat_label, surv)
```

### Practice — median survival with CI

```{r ex-km-median, exercise=TRUE, exercise.lines=12}
# Median survival (50th percentile of the survival curve) per arm, with CI.

km_fit <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)

ard_survival_survfit(km_fit, probs = ___) |>
  filter(stat_name %in% c("estimate","conf.low","conf.high")) |>
  mutate(days = round(.pull1(stat), 1)) |>
  select(group1_level, stat_name, days)
```

```{r ex-km-median-hint-1}
# probs = 0.5  (the median is the 50th percentile)
```

```{r ex-km-median-solution}
km_fit <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)

ard_survival_survfit(km_fit, probs = 0.5) |>
  filter(stat_name %in% c("estimate","conf.low","conf.high")) |>
  mutate(days = round(.pull1(stat), 1)) |>
  select(group1_level, stat_name, days)
```

---

## Section 8 — Log-rank test: `ard_survival_survdiff()`

The log-rank test compares survival curves across arms. `ard_survival_survdiff()` takes a **survival formula and the data** (it builds the `survdiff` internally) and returns the chi-squared statistic, df, and p-value.

**SAS equivalent:** `PROC LIFETEST; ... STRATA trta / LOGRANK; ODS OUTPUT HomTests=...; RUN;`

```{r ex-logrank, exercise=TRUE, exercise.lines=12}
# Log-rank test of OS across treatment arms. Pass a FORMULA + data, not a
# survdiff object. Surv(AVAL, 1 - CNSR): CNSR=1 is censored.

logrank <- ard_survival_survdiff(
  ___,                                  # Surv(AVAL, 1 - CNSR) ~ TRTA
  data = adtte_os
)

logrank |>
  filter(stat_name %in% c("statistic","df","p.value")) |>   # skip the method row
  mutate(value = round(.pull1(stat), 4)) |>
  select(stat_name, stat_label, value)
```

```{r ex-logrank-hint-1}
# formula = Surv(AVAL, 1 - CNSR) ~ TRTA
```

```{r ex-logrank-solution}
logrank <- ard_survival_survdiff(
  Surv(AVAL, 1 - CNSR) ~ TRTA,
  data = adtte_os
)

logrank |>
  filter(stat_name %in% c("statistic","df","p.value")) |>
  mutate(value = round(.pull1(stat), 4)) |>
  select(stat_name, stat_label, value)
```

---

## Section 9 — The censoring convention (the always-misunderstood detail)

This is the **most common source of bugs** in cardx survival analyses. Memorize the table:

| Context | Meaning of `1` | Source |
|---|---|---|
| ADTTE: `CNSR = 1` | Subject was **censored** (event did NOT occur) | CDISC ADaM standard |
| `survival::Surv(time, event)`: `event = 1` | Event **occurred** | R survival package |

They are **opposite**. The conversion is always `Surv(AVAL, 1 - CNSR)`:

```r
# CORRECT — 1 - CNSR makes the event indicator 1 = event
survfit(Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)
coxph(  Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)

# WRONG — analyzes censored subjects AS IF they had events
survfit(Surv(AVAL, CNSR) ~ TRTA, data = adtte_os)   # ← BUG
```

A robust team pattern is to convert once, up front, and never touch `CNSR` inside `Surv()` again:

```r
adtte_surv <- function(data, paramcd) {
  data |>
    filter(PARAMCD == paramcd) |>
    mutate(EVENT = 1 - CNSR)         # convert once: EVENT = 1 means event
}
os <- adtte_surv(adtte, "OS")
survfit(Surv(AVAL, EVENT) ~ TRTA, data = os)
```

### Practice — prove the bug

```{r ex-censor, exercise=TRUE, exercise.lines=14}
# Count "events" the CORRECT way (1 - CNSR) vs the WRONG way (CNSR).
# They should NOT match. This is why QC catches this so often.

n_events_correct <- adtte_os |> summarise(events = sum(1 - CNSR)) |> pull(events)
n_events_wrong   <- adtte_os |> summarise(events = sum(___))      |> pull(events)

tibble(
  approach = c("Surv(AVAL, 1 - CNSR)  [correct]", "Surv(AVAL, CNSR)  [bug]"),
  events   = c(n_events_correct, n_events_wrong)
)
```

```{r ex-censor-hint-1}
# The WRONG way sums CNSR directly: sum(CNSR)
```

```{r ex-censor-solution}
n_events_correct <- adtte_os |> summarise(events = sum(1 - CNSR)) |> pull(events)
n_events_wrong   <- adtte_os |> summarise(events = sum(CNSR))     |> pull(events)

tibble(
  approach = c("Surv(AVAL, 1 - CNSR)  [correct]", "Surv(AVAL, CNSR)  [bug]"),
  events   = c(n_events_correct, n_events_wrong)
)
```

### Quiz — censoring

```{r quiz-censor, echo=FALSE}
quiz(caption = "Censoring Convention",
  question("In ADTTE, what does CNSR = 1 mean?",
    answer("The event occurred"),
    answer("The subject was censored (event did NOT occur)", correct = TRUE,
           message = "CDISC ADaM uses CNSR = 1 for censored, CNSR = 0 for event. This is the opposite of the survival package convention."),
    answer("The subject died"),
    answer("The record is invalid"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question("What goes wrong if you write Surv(AVAL, CNSR) instead of Surv(AVAL, 1 - CNSR)?",
    answer("Nothing — they are equivalent"),
    answer("R treats censored subjects as having events and vice versa, inverting the entire analysis",
           correct = TRUE,
           message = "The event indicator is flipped, so KM curves, medians, HRs, and log-rank p-values are all computed on the wrong subjects."),
    answer("The model fails to converge"),
    answer("Only the CI bounds change"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

---

## Section 10 — ANOVA: `ard_stats_aov()`

For an omnibus comparison of a continuous outcome across 3+ groups, `ard_stats_aov()` wraps `stats::aov()` and returns the ANOVA table (sum of squares, df, F, p-value) per term.

**SAS equivalent:** `PROC GLM DATA=adsl; CLASS trt01p; MODEL age = trt01p; RUN;` (or `PROC ANOVA`).

```{r ex-aov, exercise=TRUE, exercise.lines=10}
# One-way ANOVA: does mean AGE differ across the 3 treatment arms?

ard_stats_aov(AGE ~ ___, data = adsl) |>
  filter(stat_name %in% c("statistic","p.value")) |>
  mutate(value = round(.pull1(stat), 4)) |>
  select(variable, stat_name, value)
```

```{r ex-aov-hint-1}
# AGE ~ TRT01P
```

```{r ex-aov-solution}
ard_stats_aov(AGE ~ TRT01P, data = adsl) |>
  filter(stat_name %in% c("statistic","p.value")) |>
  mutate(value = round(.pull1(stat), 4)) |>
  select(variable, stat_name, value)
```

---

## Section 11 — Estimated marginal means (LS means)

For **least-squares (model-adjusted) means** — the LSMEANS of clinical reporting. cardx wraps `{emmeans}` in `ard_emmeans_emmeans()` / `ard_emmeans_contrast()`. The production call is:

```r
ard_emmeans_emmeans(
  data = advs_wk24, formula = CHG ~ BASE + TRTA,
  method = ~ TRTA, response_type = "continuous"
)
```

Because the cardx wrapper's output shape is version-sensitive, the exercise below calls **`emmeans::emmeans()` directly** — the exact engine cardx invokes — and tidies the adjusted means. The numbers are identical to what the wrapper computes.

**SAS equivalent:** `PROC MIXED ...; LSMEANS trta / DIFF CL; RUN;`

```{r ex-emmeans, exercise=TRUE, exercise.lines=16}
# Fit an ANCOVA of weight CHG, then get LS (model-adjusted) means by treatment arm.
advs_wk24 <- advs |>
  filter(PARAMCD == "WEIGHT", AVISIT == "Week 24", SAFFL == "Y") |>
  filter(!is.na(CHG), !is.na(BASE))

mod <- lm(CHG ~ BASE + TRTA, data = advs_wk24)

emmeans::emmeans(mod, specs = ___) |>     # marginal means over TRTA
  as.data.frame() |>
  transmute(TRTA,
            lsmean = round(emmean, 3),
            lower  = round(lower.CL, 3),
            upper  = round(upper.CL, 3))
```

```{r ex-emmeans-hint-1}
# specs = ~ TRTA — the CLASS/grouping variable as a one-sided formula
```

```{r ex-emmeans-solution}
advs_wk24 <- advs |>
  filter(PARAMCD == "WEIGHT", AVISIT == "Week 24", SAFFL == "Y") |>
  filter(!is.na(CHG), !is.na(BASE))

mod <- lm(CHG ~ BASE + TRTA, data = advs_wk24)

emmeans::emmeans(mod, specs = ~ TRTA) |>
  as.data.frame() |>
  transmute(TRTA,
            lsmean = round(emmean, 3),
            lower  = round(lower.CL, 3),
            upper  = round(upper.CL, 3))
```

> **MMRM note:** In real studies the repeated-measures primary endpoint is fit with `{mmrm}` — `mmrm(CHG ~ BASE + TRTA + AVISIT + TRTA:AVISIT + us(AVISIT | USUBJID), data = ...)` — then passed to `ard_regression()` or the `ard_emmeans_*()` wrappers. The SAS analogue is `PROC MIXED ... REPEATED avisit / SUBJECT=usubjid TYPE=UN; LSMEANS trta*avisit / DIFF CL;`.

---

## Section 12 — Standardized mean differences: `ard_smd_smd()`

`ard_smd_smd()` reports **standardized mean differences** (Cohen's d for continuous, standardized proportion difference for categorical) — the workhorse of covariate-balance "Table 1" assessments and propensity-matching diagnostics. A common rule of thumb: |SMD| < 0.1 indicates good balance.

**SAS equivalent:** no single PROC; usually a macro computing `(mean1 - mean2) / pooled_sd`.

```{r ex-smd, exercise=TRUE, exercise.lines=11}
# Assess baseline balance of AGE, BMIBL, and SEX across the two Xanomeline arms.

ard_smd_smd(
  data      = adsl_2arm,
  by        = "TRT01P",
  variables = c(___, ___, ___)
) |>
  filter(stat_name == "estimate") |>     # one SMD per variable; drop method/gref rows
  mutate(SMD = round(.pull1(stat), 3)) |>
  select(variable, SMD)
```

```{r ex-smd-hint-1}
# variables = c("AGE", "BMIBL", "SEX")
```

```{r ex-smd-solution}
ard_smd_smd(
  data      = adsl_2arm,
  by        = "TRT01P",
  variables = c("AGE", "BMIBL", "SEX")
) |>
  filter(stat_name == "estimate") |>     # one SMD per variable; drop method/gref rows
  mutate(SMD = round(.pull1(stat), 3)) |>
  select(variable, SMD)
```

---

## Section 13 — Combining cards + cardx into one display-ready ARD

The everyday deliverable: a demographics table with **descriptive stats from `{cards}`** and **p-values from `{cardx}`**, merged with `bind_ard()`. The `context` column keeps them distinguishable, and downstream `{gtsummary}` knows to route p-values into the correct column.

```r
desc_ard <- bind_ard(                        # cards (Lesson 25)
  ard_continuous(adsl_2arm, variables = "AGE", by = "TRT01P"),
  ard_categorical(adsl_2arm, variables = c("SEX","AGEGR1"), by = "TRT01P")
)

pval_ard <- bind_ard(                        # cardx (this lesson)
  ard_stats_t_test(adsl_2arm, by = "TRT01P", variables = "AGE"),
  ard_stats_chisq_test(adsl_2arm, by = "TRT01P", variables = c("SEX","AGEGR1"))
)

demog_with_pvalues <- bind_ard(desc_ard, pval_ard)
demog_with_pvalues |> distinct(context)
# "stats_t_test" "stats_chisq_test"  (+ "continuous" "categorical" from cards)
```

### Practice — assemble an inferential ARD bundle

```{r ex-combine, exercise=TRUE, exercise.lines=20}
# Build ONE ARD bundling, for the two Xanomeline arms:
#   - t-test on AGE
#   - chi-squared on SEX and AGEGR1
#   - Wilcoxon on BMIBL
# Then list the distinct context values and pull all p-values.

bundle <- bind_ard(
  ard_stats_t_test(adsl_2arm,      by = "TRT01P", variables = "AGE"),
  ard_stats_chisq_test(adsl_2arm,  by = "TRT01P", variables = c("SEX","AGEGR1")),
  ard_stats_wilcox_test(adsl_2arm, by = "TRT01P", variables = ___)
)

bundle |> distinct(context)

bundle |>
  filter(stat_name == ___) |>
  mutate(p = round(.pull1(stat), 4)) |>
  select(context, variable, p)
```

```{r ex-combine-hint-1}
# variables = "BMIBL" for the Wilcoxon test
# filter(stat_name == "p.value")
```

```{r ex-combine-solution}
bundle <- bind_ard(
  ard_stats_t_test(adsl_2arm,      by = "TRT01P", variables = "AGE"),
  ard_stats_chisq_test(adsl_2arm,  by = "TRT01P", variables = c("SEX","AGEGR1")),
  ard_stats_wilcox_test(adsl_2arm, by = "TRT01P", variables = "BMIBL")
)

bundle |> distinct(context)

bundle |>
  filter(stat_name == "p.value") |>
  mutate(p = round(.pull1(stat), 4)) |>
  select(context, variable, p)
```

---

## Section 14 — Full SAS → cardx translation table

| Statistical task | SAS PROC | cardx function |
|---|---|---|
| Two-group mean comparison | `PROC TTEST; CLASS arm; VAR x;` | `ard_stats_t_test(by, variables)` |
| One-sample / paired t-test | `PROC TTEST; VAR x;` (or `PAIRED`) | `ard_stats_t_test(variables, mu=)` |
| Non-parametric two-group | `PROC NPAR1WAY WILCOXON;` | `ard_stats_wilcox_test(by, variables)` |
| Categorical association | `PROC FREQ; TABLES a*b / CHISQ;` | `ard_stats_chisq_test(by, variables)` |
| Exact categorical test | `PROC FREQ; TABLES a*b / EXACT;` | `ard_stats_fisher_test(by, variables)` |
| Proportion CI (Wilson/exact) | `PROC FREQ; ... / BINOMIAL(CL=WILSON);` | `ard_categorical_ci(value, method)` |
| Mean CI | `PROC MEANS MEAN CLM;` | `ard_continuous_ci(method="t.test")` |
| Omnibus group comparison | `PROC GLM; CLASS arm; MODEL y=arm;` | `ard_stats_aov(y ~ arm)` |
| Linear / ANCOVA model | `PROC GLM`/`PROC MIXED ... SOLUTION;` | `ard_regression(lm(...))` |
| Logistic regression | `PROC LOGISTIC; MODEL y(EVENT='1')=...;` | `ard_regression(glm(..., binomial))` |
| Cox PH model | `PROC PHREG; MODEL t*cnsr(1)=...;` | `ard_regression(coxph(Surv(t,1-CNSR)~...))` |
| Kaplan-Meier survival | `PROC LIFETEST METHOD=KM; TIME t*cnsr(1);` | `ard_survival_survfit(times=/probs=)` |
| Log-rank test | `PROC LIFETEST; STRATA arm / LOGRANK;` | `ard_survival_survdiff()` |
| MMRM / repeated measures | `PROC MIXED; REPEATED / TYPE=UN; LSMEANS;` | `ard_regression(mmrm(...))` + `ard_emmeans_emmeans()` |
| Least-squares means | `LSMEANS arm / DIFF CL;` | `ard_emmeans_emmeans()` / `emmeans::emmeans()` |
| Standardized mean diff | (macro) | `ard_smd_smd(by, variables)` |

---

## Section 15 — Capstone: a complete primary-efficacy inferential ARD

Build, in one pipeline, the inferential layer for an OS analysis: K-M x-day survival, median survival, log-rank test, and a Cox HR — all bound into a single ARD with a unified structure.

```{r ex-capstone, exercise=TRUE, exercise.lines=44}
# All four pieces use the SAME survival object convention: Surv(AVAL, 1 - CNSR).

# 1. Fit the K-M curve and the Cox model
km   <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)
cox  <- coxph(  Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)

# 2. x-day survival at 180 and 365 days
os_xday <- ard_survival_survfit(km, times = c(180, 365))

# 3. Median survival
os_med  <- ard_survival_survfit(km, probs = ___)

# 4. Log-rank test (formula + data, not a survdiff object)
os_lr   <- ard_survival_survdiff(Surv(AVAL, 1 - CNSR) ~ TRTA, data = ___)

# 5. Cox hazard ratios
os_cox  <- ard_regression(cox)

# 6. Bind into one ARD and inspect the contexts present
os_ard <- bind_ard(os_xday, os_med, os_lr, os_cox)

os_ard |> count(context)
```

```{r ex-capstone-hint-1}
# probs = 0.5 for the median; ard_survival_survdiff(formula, data = adtte_os)
```

```{r ex-capstone-solution}
km   <- survfit(Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)
cox  <- coxph(  Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)

os_xday <- ard_survival_survfit(km, times = c(180, 365))
os_med  <- ard_survival_survfit(km, probs = 0.5)
os_lr   <- ard_survival_survdiff(Surv(AVAL, 1 - CNSR) ~ TRTA, data = adtte_os)
os_cox  <- ard_regression(cox)

os_ard <- bind_ard(os_xday, os_med, os_lr, os_cox)

os_ard |> count(context)
```

---

## Lesson Summary

```{r final-quiz, echo=FALSE}
quiz(caption = "Lesson 27 — Final Quiz",
  question("What is the single biggest structural difference between a cards ARD and a cardx ARD?",
    answer("cardx ARDs use a wide format"),
    answer("None — same columns and list-column structure; only the context value (and the kind of statistic) changes",
           correct = TRUE,
           message = "That sameness is the point: descriptive (cards) and inferential (cardx) ARDs bind_ard() together cleanly because they share the ARD spec."),
    answer("cardx stat values are not list columns"),
    answer("cardx ARDs cannot be combined with cards ARDs"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question("Write the survival object for an ADTTE Cox model. Which is correct?",
    answer("Surv(AVAL, CNSR)"),
    answer("Surv(AVAL, 1 - CNSR)", correct = TRUE,
           message = "CNSR=1 is censored in ADTTE; Surv() needs 1=event, so the event indicator is 1 - CNSR."),
    answer("Surv(CNSR, AVAL)"),
    answer("Surv(AVAL, CNSR == 1)"),
    allow_retry = TRUE, random_answer_order = TRUE
  ),
  question("ard_categorical_ci(method='wilson') returns estimate = 0.481. ard_categorical() returns p = 48.1 for the same proportion. Why the difference?",
    answer("One of them is wrong"),
    answer("ard_categorical_ci() reports a 0–1 proportion; ard_categorical() reports a 0–100 percentage",
           correct = TRUE,
           message = "Watch the scale: multiply the cardx proportion estimate by 100 before displaying it next to cards percentages."),
    answer("Wilson CIs are always on a different scale"),
    answer("The two functions use different data"),
    allow_retry = TRUE, random_answer_order = TRUE
  )
)
```

### Key takeaways

- **`{cardx}` = same ARD, inferential statistics.** Same columns, same list-column `stat`, same `bind_ard()` — only `context` changes.
- **Univariate tests:** `ard_stats_t_test()`, `ard_stats_wilcox_test()` (2 levels only), `ard_stats_chisq_test()`, `ard_stats_fisher_test()` (any levels).
- **Confidence intervals:** `ard_categorical_ci()` (Wilson/Clopper-Pearson/Wald/…, 0–1 scale), `ard_continuous_ci()` (t / Hodges-Lehmann).
- **`ard_regression()`** converts *any* fitted model (`lm`, `glm`, `coxph`, `mmrm`) to an ARD; coefficients are native-scale — `exp()` for OR/HR.
- **Survival:** `ard_survival_survfit(times=)` for x-day survival, `(probs=0.5)` for median; `ard_survival_survdiff()` for the log-rank test.
- **`ard_stats_aov()`, `ard_emmeans_emmeans()` (or `emmeans::emmeans()`), `ard_smd_smd()`** cover omnibus ANOVA, LS means (LSMEANS), and covariate-balance SMDs.
- **Censoring:** `CNSR = 1` is *censored* in ADTTE; `Surv()` needs `1 - CNSR`. This is the #1 survival QC failure point.
- **Combine** descriptive (cards) + inferential (cardx) ARDs with `bind_ard()` for a display-ready dataset.

### What's next

**Lesson 28 — gtsummary (Part 1):** the display layer that *consumes* ARDs. You'll use `tbl_ard_summary()`, `add_p()`, and the gtsummary vocabulary to turn the descriptive and inferential ARDs you built in Lessons 25 and 27 into CSR-ready tables — without recomputing a single statistic.

### Self-check questions

1. What's the key difference between `{cards}` and `{cardx}` in the statistics they produce?
2. Write the complete cardx call for "Wilson 95% CI for a responder rate by treatment arm."
3. Why does `Surv(AVAL, 1 - CNSR)` need the `1 - CNSR` conversion, and what breaks if you forget it?
4. Translate to cardx: "Log-rank test of OS by treatment arm."
5. `ard_regression()` on a logistic `glm` returns `estimate` for a treatment level. What transformation gives the odds ratio?
6. Which two cardx functions require the `by` variable to have exactly 2 levels, and what do you do for a 3-arm trial?
7. `ard_categorical_ci()` reports `estimate = 0.52` while `ard_categorical()` reports `p = 52`. What is the scale difference and why does it matter for display?

### Glossary

- **`ard_stats_t_test()`** — t-test → ARD; wraps `stats::t.test()` (Welch by default; `var.equal=TRUE` for Student's).
- **`ard_stats_wilcox_test()`** — Wilcoxon rank-sum (Mann-Whitney) → ARD; 2 levels only.
- **`ard_stats_chisq_test()`** — Pearson chi-squared test → ARD; any number of levels.
- **`ard_stats_fisher_test()`** — Fisher's exact test → ARD; preferred for sparse tables.
- **`ard_categorical_ci()`** — Proportion CI with multiple methods (Wilson, Clopper-Pearson, Wald, Agresti-Coull, Jeffreys); 0–1 scale.
- **`ard_continuous_ci()`** — CI for a continuous mean (t-distribution or Hodges-Lehmann).
- **`ard_regression()`** — Convert any fitted model (`lm`/`glm`/`coxph`/`mmrm`) to an ARD; native-scale coefficients.
- **`ard_survival_survfit()`** — K-M x-time survival (`times=`) and median (`probs=0.5`) → ARD.
- **`ard_survival_survdiff()`** — Log-rank test → ARD.
- **`ard_stats_aov()`** — One-/multi-way ANOVA table → ARD.
- **`ard_emmeans_emmeans()`** — Estimated marginal (LS) means → ARD; wraps `{emmeans}` (the LSMEANS equivalent).
- **`ard_smd_smd()`** — Standardized mean differences for covariate balance → ARD.
- **Welch t-test** — Default two-sample t-test allowing unequal variances.
- **Wilson interval** — Recommended proportion CI; handles small N and extreme proportions.
- **Clopper-Pearson** — "Exact" proportion CI; conservative, common in regulatory submissions.
- **HR** — Hazard Ratio; `exp(log-HR)` from a Cox model.
- **OR** — Odds Ratio; `exp(log-odds)` from a logistic model.
- **LS mean** — Least-squares (model-adjusted) group mean.
- **MMRM** — Mixed Model with Repeated Measures; standard primary analysis for repeated-measures efficacy.
- **SMD** — Standardized Mean Difference; covariate-balance metric (|SMD| < 0.1 ≈ balanced).
- **Censoring convention** — ADTTE: `CNSR = 1` = censored; `Surv()`: second arg `1` = event; convert with `1 - CNSR`.
- **`context`** — ARD column recording which cardx engine produced each row (`"stats_t_test"`, `"survival_survfit"`, `"regression"`, …).
