---
title: "Lesson 28 — cardinal: Clinical Trial Analysis Results Datasets"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build clinical-trial ARDs with the cardinal package: ard_continuous_ci() for
  mean confidence intervals, ard_proportion_ci() for Wilson/Wald binary CIs,
  ard_regression_basic() for model-based odds ratios, ard_demographics() for
  one-shot demographics tables, and ard_ae_summary() for standard AE summaries.
  Capstone integrates all five into a single study-level ARD.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 28_cardinal.Rmd
# PURPOSE       : cardinal package — pre-built ARD functions for standard
#                 clinical-trial summary tables: continuous CIs, proportion CIs,
#                 model-based regression ARDs, demographics, and AE summaries.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    DTHFL_N = if_else(!is.na(DTHFL) & DTHFL == "Y", 1L, 0L, missing = 0L),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo","Xanomeline Low Dose","Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
if (!"TRT01P" %in% names(adae_raw))
  adae_raw <- left_join(adae_raw, select(adsl, USUBJID, TRT01P), by = "USUBJID")
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ══════════════════════════════════════════════════════════════════════════════
# CARDS BASE MOCKS (replicated from lesson 25)
# ══════════════════════════════════════════════════════════════════════════════

.ard_row <- function(group1=NA_character_, group1_level=NA_character_,
                     variable, variable_level=NA_character_,
                     context, stat_name, stat, stat_label) {
  tibble(
    group1         = group1,
    group1_level   = group1_level,
    variable       = variable,
    variable_level = variable_level,
    context        = context,
    stat_name      = stat_name,
    stat           = list(stat),
    stat_label     = stat_label
  )
}

ard_continuous <- function(data, variables, by = NULL,
                           statistics = list(
                             N      = function(x) sum(!is.na(x)),
                             mean   = function(x) mean(x, na.rm=TRUE),
                             sd     = function(x) sd(x, na.rm=TRUE),
                             median = function(x) median(x, na.rm=TRUE),
                             p25    = function(x) quantile(x, 0.25, na.rm=TRUE),
                             p75    = function(x) quantile(x, 0.75, na.rm=TRUE),
                             min    = function(x) min(x, na.rm=TRUE),
                             max    = function(x) max(x, na.rm=TRUE),
                             n_miss = function(x) sum(is.na(x)),
                             p_miss = function(x) round(100*mean(is.na(x)),1)
                           )) {
  stat_labels <- c(N="N (non-missing)", mean="Mean", sd="SD", median="Median",
                   p25="Q1 (25th pctile)", p75="Q3 (75th pctile)",
                   min="Min", max="Max", n_miss="N missing", p_miss="% missing")
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      x   <- sub[[vv]]
      lapply(names(statistics), function(sn) {
        val <- tryCatch(statistics[[sn]](x), error=function(e) NA_real_)
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="continuous", stat_name=sn,
                 stat=val, stat_label=stat_labels[sn])
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_categorical <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    levs <- sort(unique(as.character(data[[vv]])))
    levs <- levs[!is.na(levs)]
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      total <- sum(!is.na(sub[[vv]]))
      lapply(levs, function(lv) {
        cnt <- sum(sub[[vv]] == lv, na.rm=TRUE)
        bind_rows(
          .ard_row(group1=if(!is.null(by)) by else NA_character_,
                   group1_level=as.character(gl), variable=vv,
                   variable_level=lv, context="categorical",
                   stat_name="n", stat=cnt, stat_label="n"),
          .ard_row(group1=if(!is.null(by)) by else NA_character_,
                   group1_level=as.character(gl), variable=vv,
                   variable_level=lv, context="categorical",
                   stat_name="p", stat=round(100*cnt/total,1), stat_label="%")
        )
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_missing <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      x   <- sub[[vv]]
      bind_rows(
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="missing", stat_name="n_miss",
                 stat=sum(is.na(x)), stat_label="N missing"),
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="missing", stat_name="p_miss",
                 stat=round(100*mean(is.na(x)),1), stat_label="% missing"),
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="missing", stat_name="n_not_miss",
                 stat=sum(!is.na(x)), stat_label="N non-missing")
      )
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_stack <- function(...) {
  out <- bind_rows(...)
  structure(out, class=c("card","tbl_df","tbl","data.frame"))
}

bind_ard <- function(...) ard_stack(...)

ard_get_stat <- function(ard, stat_nm, var_nm=NULL) {
  x <- if (!is.null(var_nm)) filter(ard, variable==var_nm) else ard
  x |> filter(stat_name==stat_nm) |> pull(stat) |> unlist()
}

# ══════════════════════════════════════════════════════════════════════════════
# CARDINAL FUNCTION MOCKS
# ══════════════════════════════════════════════════════════════════════════════

# Wilson 95% CI for a proportion p_hat based on n observations
.wilson_ci <- function(n, p_hat, conf=0.95) {
  z      <- qnorm(1-(1-conf)/2)
  centre <- (p_hat + z^2/(2*n)) / (1 + z^2/n)
  half   <- (z * sqrt(p_hat*(1-p_hat)/n + z^2/(4*n^2))) / (1 + z^2/n)
  c(lower=round(centre-half,4), upper=round(centre+half,4))
}

ard_continuous_ci <- function(data, variables, by=NULL,
                               conf.level=0.95, method="t.test") {
  groups <- if (!is.null(by)) sort(unique(na.omit(data[[by]]))) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      x <- if (!is.na(gl) && !is.null(by)) data[[vv]][data[[by]]==gl] else data[[vv]]
      x <- x[!is.na(x)]
      n  <- length(x); m <- mean(x); se <- sd(x)/sqrt(n)
      t_crit <- qt(1-(1-conf.level)/2, df=n-1)
      bind_rows(
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="continuous_ci", stat_name="mean",
               stat=list(round(m,4)), stat_label="Mean"),
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="continuous_ci", stat_name="conf.low",
               stat=list(round(m-t_crit*se,4)),
               stat_label=paste0(conf.level*100,"% CI lower")),
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="continuous_ci", stat_name="conf.high",
               stat=list(round(m+t_crit*se,4)),
               stat_label=paste0(conf.level*100,"% CI upper")),
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="continuous_ci", stat_name="n",
               stat=list(n), stat_label="N")
      )
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_proportion_ci <- function(data, variables, by=NULL,
                               conf.level=0.95, method="wilson",
                               value=list()) {
  groups <- if (!is.null(by)) sort(unique(na.omit(data[[by]]))) else NA_character_
  rows <- lapply(variables, function(vv) {
    pos_val <- if (!is.null(value[[vv]])) value[[vv]] else "Y"
    lapply(groups, function(gl) {
      x <- if (!is.na(gl) && !is.null(by)) data[[vv]][data[[by]]==gl] else data[[vv]]
      x <- x[!is.na(x)]
      n  <- length(x); k <- sum(x==pos_val); p_hat <- k/n
      ci <- .wilson_ci(n, p_hat, conf.level)
      bind_rows(
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="proportion_ci", stat_name="n",
               stat=list(k), stat_label=paste0("N (",pos_val,")")),
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="proportion_ci", stat_name="p",
               stat=list(round(100*p_hat,1)), stat_label="%"),
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="proportion_ci", stat_name="conf.low",
               stat=list(round(100*ci["lower"],1)),
               stat_label=paste0(conf.level*100,"% CI lower")),
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="proportion_ci", stat_name="conf.high",
               stat=list(round(100*ci["upper"],1)),
               stat_label=paste0(conf.level*100,"% CI upper"))
      )
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_regression_basic <- function(model, exponentiate=FALSE, conf.level=0.95) {
  cf  <- summary(model)$coefficients
  # Use Wald-based CIs (confint.default) to avoid profile-likelihood refitting,
  # which triggers dozens of glm.fit warnings on sparse/separated outcomes.
  ci  <- tryCatch(confint.default(model, level=conf.level),
                  error=function(e) matrix(NA, nrow=nrow(cf), ncol=2))
  rows <- lapply(seq_len(nrow(cf)), function(i) {
    vv  <- rownames(cf)[i]
    est <- if (exponentiate) exp(cf[i,1]) else cf[i,1]
    lo  <- if (exponentiate) exp(ci[i,1]) else ci[i,1]
    hi  <- if (exponentiate) exp(ci[i,2]) else ci[i,2]
    bind_rows(
      tibble(variable=vv, context="regression", stat_name="estimate",
             stat=list(round(est,4)),
             stat_label=if(exponentiate) "OR" else "Estimate"),
      tibble(variable=vv, context="regression", stat_name="conf.low",
             stat=list(round(lo,4)),  stat_label="CI lower"),
      tibble(variable=vv, context="regression", stat_name="conf.high",
             stat=list(round(hi,4)),  stat_label="CI upper"),
      tibble(variable=vv, context="regression", stat_name="p.value",
             stat=list(round(cf[i,4],4)), stat_label="p-value")
    )
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_demographics <- function(data, by=NULL,
                              continuous_vars=NULL, categorical_vars=NULL) {
  cn       <- names(data)
  if (!is.null(by)) cn <- cn[cn != by]
  cont_vars <- if (!is.null(continuous_vars)) continuous_vars else
    cn[sapply(cn, function(v) is.numeric(data[[v]]))]
  cat_vars  <- if (!is.null(categorical_vars)) categorical_vars else
    cn[sapply(cn, function(v) !is.numeric(data[[v]]))]
  ard_stack(
    ard_continuous(data,    variables=cont_vars, by=by),
    ard_categorical(data,   variables=cat_vars,  by=by),
    ard_missing(data,       variables=cn,        by=by)
  )
}

ard_ae_summary <- function(data_adsl, data_adae,
                            by="TRT01P",
                            soc_var="AESOC", pt_var="AEDECOD",
                            serious_var="AESER", severe_var="AESEV") {
  subj_ae <- data_adsl |>
    select(USUBJID, all_of(by)) |>
    left_join(
      data_adae |>
        group_by(USUBJID) |>
        summarise(ANY_AE="Y", .groups="drop"),
      by="USUBJID"
    ) |>
    mutate(
      ANY_AE = replace_na(ANY_AE, "N"),
      SAE    = if_else(USUBJID %in% data_adae$USUBJID[data_adae[[serious_var]]=="Y"],
                       "Y","N"),
      SEV_AE = if_else(USUBJID %in% data_adae$USUBJID[data_adae[[severe_var]]=="SEVERE"],
                       "Y","N")
    )

  soc_data <- data_adae |>
    distinct(USUBJID, .data[[soc_var]], .data[[by]])

  ard_stack(
    ard_proportion_ci(subj_ae, variables=c("ANY_AE","SAE","SEV_AE"), by=by,
                      value=list(ANY_AE="Y", SAE="Y", SEV_AE="Y")),
    ard_categorical(soc_data, variables=soc_var, by=by)
  )
}

# ── Pre-built objects used in exercises ──────────────────────────────────────
age_ci_ard <- ard_continuous_ci(adsl, variables=c("AGE","WEIGHTBL"),
                                 by="TRT01P", conf.level=0.95)

diabfl_ci_ard <- ard_proportion_ci(adsl,
                                    variables=c("DIABFL","DTHFL","SAFFL"),
                                    by="TRT01P",
                                    value=list(DIABFL="Y", DTHFL="Y", SAFFL="Y"))

logit_model <- suppressWarnings(
  glm(DTHFL_N ~ AGE + factor(TRT01P), data=adsl,
      family=binomial(link="logit"))
)

regression_ard <- ard_regression_basic(logit_model, exponentiate=TRUE)

demo_ard <- ard_demographics(
  select(adsl, TRT01P, AGE, WEIGHTBL, BMIBL, SEX, RACE),
  by="TRT01P",
  continuous_vars=c("AGE","WEIGHTBL","BMIBL"),
  categorical_vars=c("SEX","RACE")
)

ae_ard <- ard_ae_summary(adsl, adae, by="TRT01P")
```

## Overview

**cardinal** is a pharmaverse package that delivers pre-built, ready-to-use ARD
computations for the statistical summaries most common in clinical trial reporting.
Where `cards` gives you building blocks, `cardinal` gives you the finished
assembly line for standard deliverables.

### The cards → cardinal → display pipeline

```
ADSL / ADAE / ADLB
       │
       ▼
  cardinal functions          cards base functions
  ────────────────────        ──────────────────────
  ard_continuous_ci()    ─┐
  ard_proportion_ci()     │
  ard_regression_basic()  ├──► ARD tibble (CDISC-compliant)
  ard_demographics()      │
  ard_ae_summary()       ─┘
                               │
                               ▼
                    gtsummary::tbl_ard_summary()
                    tfrmt::print_to_gt()          ← Lessons 29–30
                    tfrmt::mock_df()
```

### What cardinal adds over cards

| cards | cardinal |
|-------|---------|
| `ard_continuous()` — mean, SD, median | `ard_continuous_ci()` — mean + 95% CI |
| `ard_categorical()` — n and % | `ard_proportion_ci()` — % + Wilson/Wald CI |
| Manual `glm()` wrangling | `ard_regression_basic()` — one-call model ARD |
| Multiple calls needed | `ard_demographics()` — one-shot demographics |
| Manual AE subject-level flagging | `ard_ae_summary()` — standard any/serious/severe |

### Lesson map

| Section | Function | Key argument |
|---------|----------|-------------|
| 1 | `ard_continuous_ci()` | `conf.level`, `method` |
| 2 | `ard_proportion_ci()` | `method="wilson"`, `value` |
| 3 | `ard_regression_basic()` | `exponentiate`, `conf.level` |
| 4 | `ard_demographics()` | `continuous_vars`, `categorical_vars` |
| 5 | `ard_ae_summary()` | `soc_var`, `serious_var`, `severe_var` |
| 6 | Capstone | All five combined |

---

## Section 1 — ard_continuous_ci()

`ard_continuous_ci()` returns mean, 95% CI lower bound, 95% CI upper bound, and
N for each continuous variable, optionally stratified by a `by` column.

```
ard_continuous_ci(
  data,
  variables,          # character vector of numeric column names
  by         = NULL,  # grouping column
  conf.level = 0.95,  # confidence level (default 95%)
  method     = "t.test"
)
```

The output ARD has `stat_name` values of: `"mean"`, `"conf.low"`, `"conf.high"`,
`"n"`.  All values are stored as list-column `stat`; use `unlist(stat)` or
`ard_get_stat()` to extract them.

### Exercise 1-a — Compute CI for AGE and WEIGHTBL by treatment arm

```{r ci-compute, exercise=TRUE, exercise.lines=10}
age_ci <- ard_continuous_ci(
  data      = ___,
  variables = c("___", "___"),
  by        = "___",
  conf.level = 0.95
)

age_ci |>
  filter(variable == "AGE") |>
  select(group1_level, stat_name, stat)
```

```{r ci-compute-hint-1}
# data = adsl
# variables = c("AGE", "WEIGHTBL")
# by = "TRT01P"
# filter keeps only AGE rows; stat_name will be "mean","conf.low","conf.high","n"
```

```{r ci-compute-solution}
age_ci <- ard_continuous_ci(
  data      = adsl,
  variables = c("AGE", "WEIGHTBL"),
  by        = "TRT01P",
  conf.level = 0.95
)

age_ci |>
  filter(variable == "AGE") |>
  select(group1_level, stat_name, stat)
```

### Exercise 1-b — Extract and format the Placebo CI for AGE

Use the pre-built `age_ci_ard` object (same as your answer above).  Extract
`conf.low` and `conf.high` for the Placebo arm and format as `"mean (lower, upper)"`.

```{r ci-format, exercise=TRUE, exercise.lines=15}
# Step 1: pull the three stats for Placebo / AGE
placebo_age <- age_ci_ard |>
  filter(variable == "AGE", group1_level == "___") |>
  select(stat_name, stat)

# Step 2: convert list-col to named vector
vals <- setNames(unlist(placebo_age$stat), placebo_age$stat_name)

# Step 3: format as "mean (lower, upper)"
sprintf("%.1f (%.1f, %.1f)",
        vals["___"], vals["___"], vals["___"])
```

```{r ci-format-hint-1}
# group1_level == "Placebo"
# vals["mean"], vals["conf.low"], vals["conf.high"]
```

```{r ci-format-solution}
placebo_age <- age_ci_ard |>
  filter(variable == "AGE", group1_level == "Placebo") |>
  select(stat_name, stat)

vals <- setNames(unlist(placebo_age$stat), placebo_age$stat_name)

sprintf("%.1f (%.1f, %.1f)",
        vals["mean"], vals["conf.low"], vals["conf.high"])
```

### Exercise 1-c — CI width comparison across arms

Compute CI width (`conf.high - conf.low`) for AGE in each arm and identify the
widest interval.

```{r ci-width, exercise=TRUE, exercise.lines=12}
age_ci_ard |>
  filter(variable == "AGE", stat_name %in% c("conf.low","conf.high")) |>
  select(group1_level, stat_name, stat) |>
  mutate(stat = unlist(stat)) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(ci_width = ___ - ___) |>
  arrange(desc(ci_width))
```

```{r ci-width-hint-1}
# ci_width = conf.high - conf.low
# arrange(desc(ci_width)) puts the widest arm first
```

```{r ci-width-solution}
age_ci_ard |>
  filter(variable == "AGE", stat_name %in% c("conf.low","conf.high")) |>
  select(group1_level, stat_name, stat) |>
  mutate(stat = unlist(stat)) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(ci_width = conf.high - conf.low) |>
  arrange(desc(ci_width))
```

```{r quiz-s1, echo=FALSE}
quiz(
  question(
    "What does ard_continuous_ci() use by default to compute the confidence interval?",
    answer("Normal (z) distribution"),
    answer("t-distribution with n-1 degrees of freedom", correct=TRUE),
    answer("Bootstrap percentile method"),
    answer("Exact Clopper-Pearson bounds"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — ard_proportion_ci()

`ard_proportion_ci()` computes a confidence interval for a binary flag
(`"Y"` / `"N"` or 0/1) using Wilson, Clopper-Pearson, or Wald methods.

```
ard_proportion_ci(
  data,
  variables,              # binary flag columns
  by         = NULL,
  conf.level = 0.95,
  method     = "wilson",  # "wilson" | "clopper-pearson" | "wald"
  value      = list()     # which level counts as the "positive" event
)
```

Output `stat_name` values: `"n"` (count of positives), `"p"` (% positive),
`"conf.low"`, `"conf.high"`.

### Exercise 2-a — Wilson CI for DIABFL, DTHFL, and SAFFL by arm

```{r prop-ci-compute, exercise=TRUE, exercise.lines=12}
flag_ci <- ard_proportion_ci(
  data      = ___,
  variables = c("DIABFL","DTHFL","SAFFL"),
  by        = "TRT01P",
  method    = "wilson",
  value     = list(DIABFL="___", DTHFL="___", SAFFL="___")
)

flag_ci |>
  filter(variable == "DIABFL") |>
  select(group1_level, stat_name, stat)
```

```{r prop-ci-compute-hint-1}
# data = adsl
# All three flags use "Y" as the positive level
# value = list(DIABFL="Y", DTHFL="Y", SAFFL="Y")
```

```{r prop-ci-compute-solution}
flag_ci <- ard_proportion_ci(
  data      = adsl,
  variables = c("DIABFL","DTHFL","SAFFL"),
  by        = "TRT01P",
  method    = "wilson",
  value     = list(DIABFL="Y", DTHFL="Y", SAFFL="Y")
)

flag_ci |>
  filter(variable == "DIABFL") |>
  select(group1_level, stat_name, stat)
```

### Exercise 2-b — Compare CI width for DIABFL across arms

Use the pre-built `diabfl_ci_ard`. Widen the ARD and compute CI width, then
identify which arm has the narrowest CI for DIABFL.

```{r prop-ci-width, exercise=TRUE, exercise.lines=12}
diabfl_ci_ard |>
  filter(variable == "DIABFL",
         stat_name %in% c("conf.low","conf.high","n","p")) |>
  mutate(stat = unlist(stat)) |>
  select(group1_level, stat_name, stat) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(ci_width = ___ - ___) |>
  arrange(___)
```

```{r prop-ci-width-hint-1}
# ci_width = conf.high - conf.low
# arrange(ci_width) — ascending puts narrowest first
# Wider arms (more events) generally have narrower CIs
```

```{r prop-ci-width-solution}
diabfl_ci_ard |>
  filter(variable == "DIABFL",
         stat_name %in% c("conf.low","conf.high","n","p")) |>
  mutate(stat = unlist(stat)) |>
  select(group1_level, stat_name, stat) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(ci_width = conf.high - conf.low) |>
  arrange(ci_width)
```

```{r quiz-s2, echo=FALSE}
quiz(
  question(
    "Why does Wilson CI perform better than Wald CI when proportions are near 0 or 1?",
    answer("Wilson uses a t-distribution while Wald uses normal"),
    answer("Wilson shifts the centre of the interval inward, preventing bounds below 0 or above 100", correct=TRUE),
    answer("Wilson corrects for continuity with a 0.5 adjustment"),
    answer("Wilson requires fewer observations than Wald"),
    allow_retry = TRUE
  )
)
```

---

## Section 3 — ard_regression_basic()

`ard_regression_basic()` wraps a fitted model object and returns a tidy ARD with
estimate, CI bounds, and p-value for each term.

```
ard_regression_basic(
  model,                   # any glm / lm / coxph object
  exponentiate = FALSE,    # TRUE → exp() all estimates (odds ratios)
  conf.level   = 0.95
)
```

Output `stat_name` values: `"estimate"`, `"conf.low"`, `"conf.high"`, `"p.value"`.

### Exercise 3-a — Fit the logistic model and build the ARD

```{r regr-fit, exercise=TRUE, exercise.lines=16}
# Step 1: create binary outcome (already in adsl as DTHFL_N)
# DTHFL_N = 1 if DTHFL=="Y", else 0

# Step 2: fit the model
logit_fit <- glm(
  DTHFL_N ~ AGE + factor(TRT01P),
  data   = ___,
  family = binomial(link = "logit")
)

# Step 3: convert to ARD with exponentiated estimates (odds ratios)
reg_ard <- ard_regression_basic(logit_fit, exponentiate = ___)

reg_ard |> select(variable, stat_name, stat)
```

```{r regr-fit-hint-1}
# data = adsl (DTHFL_N is already in adsl from the setup chunk)
# exponentiate = TRUE gives odds ratios instead of log-odds
```

```{r regr-fit-solution}
logit_fit <- glm(
  DTHFL_N ~ AGE + factor(TRT01P),
  data   = adsl,
  family = binomial(link = "logit")
)

reg_ard <- ard_regression_basic(logit_fit, exponentiate = TRUE)

reg_ard |> select(variable, stat_name, stat)
```

### Exercise 3-b — Extract ORs and CIs into a display table

Use the pre-built `regression_ard`. Pivot to wide format showing estimate,
CI lower, CI upper, and p-value per model term.

```{r regr-display, exercise=TRUE, exercise.lines=12}
regression_ard |>
  mutate(stat = unlist(stat)) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  select(variable, estimate, conf.low, conf.high, p.value) |>
  mutate(
    ci_str = sprintf("%.3f (%.3f, %.3f)", ___, ___, ___),
    sig    = if_else(p.value < 0.05, "*", "")
  )
```

```{r regr-display-hint-1}
# sprintf format: estimate, conf.low, conf.high
# p.value < 0.05 flags significance
```

```{r regr-display-solution}
regression_ard |>
  mutate(stat = unlist(stat)) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  select(variable, estimate, conf.low, conf.high, p.value) |>
  mutate(
    ci_str = sprintf("%.3f (%.3f, %.3f)", estimate, conf.low, conf.high),
    sig    = if_else(p.value < 0.05, "*", "")
  )
```

---

## Section 4 — ard_demographics()

`ard_demographics()` is a one-shot wrapper that automatically separates numeric
and character columns, calling `ard_continuous()`, `ard_categorical()`, and
`ard_missing()` in a single step and stacking the results.

```
ard_demographics(
  data,
  by              = NULL,
  continuous_vars = NULL,   # NULL → auto-detect numeric columns
  categorical_vars = NULL   # NULL → auto-detect character/factor columns
)
```

### Exercise 4-a — Build a demographics ARD for key variables

```{r demo-build, exercise=TRUE, exercise.lines=14}
demo <- ard_demographics(
  data             = select(adsl, TRT01P, AGE, WEIGHTBL, BMIBL, SEX, RACE),
  by               = "___",
  continuous_vars  = c("AGE", "WEIGHTBL", "BMIBL"),
  categorical_vars = c("SEX", "RACE")
)

# Count rows per context to verify all three are present
demo |>
  count(___)
```

```{r demo-build-hint-1}
# by = "TRT01P"
# count(context) shows how many ARD rows each sub-function contributed
```

```{r demo-build-solution}
demo <- ard_demographics(
  data             = select(adsl, TRT01P, AGE, WEIGHTBL, BMIBL, SEX, RACE),
  by               = "TRT01P",
  continuous_vars  = c("AGE", "WEIGHTBL", "BMIBL"),
  categorical_vars = c("SEX", "RACE")
)

demo |>
  count(context)
```

### Exercise 4-b — Spot-check mean AGE against raw adsl

Confirm that the mean AGE in `demo_ard` matches what you calculate directly
from `adsl` for the Placebo arm.

```{r demo-spotcheck, exercise=TRUE, exercise.lines=12}
# From the ARD
ard_mean_age <- demo_ard |>
  filter(variable == "AGE",
         stat_name == "mean",
         group1_level == "___") |>
  pull(stat) |>
  unlist()

# Direct calculation
direct_mean_age <- adsl |>
  filter(TRT01P == "Placebo") |>
  summarise(m = round(mean(AGE, na.rm=TRUE), 4)) |>
  pull(m)

# Should be equal
all.equal(ard_mean_age, direct_mean_age)
```

```{r demo-spotcheck-hint-1}
# group1_level == "Placebo"
# The ARD rounds to 4 decimal places — use round() on both sides for comparison
```

```{r demo-spotcheck-solution}
ard_mean_age <- demo_ard |>
  filter(variable == "AGE",
         stat_name == "mean",
         group1_level == "Placebo") |>
  pull(stat) |>
  unlist()

direct_mean_age <- adsl |>
  filter(TRT01P == "Placebo") |>
  summarise(m = round(mean(AGE, na.rm=TRUE), 4)) |>
  pull(m)

all.equal(ard_mean_age, direct_mean_age)
```

---

## Section 5 — ard_ae_summary()

`ard_ae_summary()` takes an ADSL (denominator) and an ADAE (numerator) and
produces a standard AE summary ARD: any AE, SAE, severe AE proportions (with
Wilson CIs) plus SOC-level categorical counts.

```
ard_ae_summary(
  data_adsl,
  data_adae,
  by          = "TRT01P",
  soc_var     = "AESOC",
  pt_var      = "AEDECOD",
  serious_var = "AESER",
  severe_var  = "AESEV"
)
```

### Exercise 5-a — Build the AE summary ARD

```{r ae-build, exercise=TRUE, exercise.lines=12}
ae_summary <- ard_ae_summary(
  data_adsl   = ___,
  data_adae   = ___,
  by          = "TRT01P",
  soc_var     = "AESOC",
  serious_var = "AESER",
  severe_var  = "AESEV"
)

ae_summary |>
  filter(context == "proportion_ci") |>
  select(group1_level, variable, stat_name, stat) |>
  mutate(stat = unlist(stat))
```

```{r ae-build-hint-1}
# data_adsl = adsl, data_adae = adae
# The proportion_ci context holds ANY_AE, SAE, SEV_AE rows
```

```{r ae-build-solution}
ae_summary <- ard_ae_summary(
  data_adsl   = adsl,
  data_adae   = adae,
  by          = "TRT01P",
  soc_var     = "AESOC",
  serious_var = "AESER",
  severe_var  = "AESEV"
)

ae_summary |>
  filter(context == "proportion_ci") |>
  select(group1_level, variable, stat_name, stat) |>
  mutate(stat = unlist(stat))
```

### Exercise 5-b — Extract ANY_AE percentages by arm

Use the pre-built `ae_ard`. Produce a one-row-per-arm table showing the ANY_AE
percentage and its Wilson CI.

```{r ae-pct, exercise=TRUE, exercise.lines=14}
ae_ard |>
  filter(variable == "ANY_AE",
         stat_name %in% c("p","conf.low","conf.high")) |>
  mutate(stat = unlist(stat)) |>
  select(group1_level, stat_name, stat) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(display = sprintf("%.1f%% (%.1f, %.1f)", ___, ___, ___))
```

```{r ae-pct-hint-1}
# sprintf: p, conf.low, conf.high
# p is already in percentage points (e.g. 78.3) from the mock
```

```{r ae-pct-solution}
ae_ard |>
  filter(variable == "ANY_AE",
         stat_name %in% c("p","conf.low","conf.high")) |>
  mutate(stat = unlist(stat)) |>
  select(group1_level, stat_name, stat) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(display = sprintf("%.1f%% (%.1f, %.1f)", p, conf.low, conf.high))
```

### Exercise 5-c — Compare SAE rates across arms

Extract the SAE `p` statistic from `ae_ard` and identify the arm with the
highest SAE rate.

```{r ae-sae, exercise=TRUE, exercise.lines=10}
ae_ard |>
  filter(variable == "SAE", stat_name == "p") |>
  mutate(pct = unlist(stat)) |>
  select(group1_level, pct) |>
  arrange(desc(___))
```

```{r ae-sae-hint-1}
# arrange(desc(pct)) — highest SAE rate first
```

```{r ae-sae-solution}
ae_ard |>
  filter(variable == "SAE", stat_name == "p") |>
  mutate(pct = unlist(stat)) |>
  select(group1_level, pct) |>
  arrange(desc(pct))
```

---

## Section 6 — Capstone: Study-Level ARD

In regulatory submissions, a single ARD object often covers the entire study:
demographics, AE summary, and efficacy model.  cardinal makes this straightforward
with `ard_stack()`.

### Exercise 6-a — Build the complete study ARD

```{r capstone-build, exercise=TRUE, exercise.lines=24}
# Step 1: demographics ARD
demo_part <- ard_demographics(
  data             = select(adsl, TRT01P, AGE, WEIGHTBL, BMIBL, SEX, RACE),
  by               = "TRT01P",
  continuous_vars  = c("AGE","WEIGHTBL","BMIBL"),
  categorical_vars = c("SEX","RACE")
)

# Step 2: AE summary ARD
ae_part <- ard_ae_summary(adsl, adae, by="TRT01P")

# Step 3: regression ARD
reg_fit  <- glm(DTHFL_N ~ AGE + factor(TRT01P), data=adsl,
                family=binomial("logit"))
reg_part <- ard_regression_basic(reg_fit, exponentiate=TRUE)

# Step 4: stack into one study ARD
study_ard <- ard_stack(___, ___, ___)

# Step 5: row counts by context
study_ard |> count(context)
```

```{r capstone-build-hint-1}
# ard_stack(demo_part, ae_part, reg_part)
# context values will include: continuous, categorical, missing,
#   proportion_ci, and regression
```

```{r capstone-build-solution}
demo_part <- ard_demographics(
  data             = select(adsl, TRT01P, AGE, WEIGHTBL, BMIBL, SEX, RACE),
  by               = "TRT01P",
  continuous_vars  = c("AGE","WEIGHTBL","BMIBL"),
  categorical_vars = c("SEX","RACE")
)

ae_part <- ard_ae_summary(adsl, adae, by="TRT01P")

reg_fit  <- glm(DTHFL_N ~ AGE + factor(TRT01P), data=adsl,
                family=binomial("logit"))
reg_part <- ard_regression_basic(reg_fit, exponentiate=TRUE)

study_ard <- ard_stack(demo_part, ae_part, reg_part)

study_ard |> count(context)
```

### Exercise 6-b — Summary display from the study ARD

Extract a formatted display of mean AGE, DIABFL proportion CI, and ANY_AE
proportion from the pre-built components.

```{r capstone-display, exercise=TRUE, exercise.lines=28}
# Mean AGE per arm (from demo_ard, context="continuous")
age_display <- demo_ard |>
  filter(variable == "AGE", stat_name == "mean") |>
  mutate(stat = round(unlist(stat), 1)) |>
  select(Arm = group1_level, `Mean AGE` = stat)

# DIABFL Wilson CI per arm (from diabfl_ci_ard)
diabfl_display <- diabfl_ci_ard |>
  filter(variable == "DIABFL",
         stat_name %in% c("p","conf.low","conf.high")) |>
  mutate(stat = unlist(stat)) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(`DIABFL % (95% CI)` = sprintf("%.1f (%.1f, %.1f)",
                                       p, conf.low, conf.high)) |>
  select(Arm = group1_level, `DIABFL % (95% CI)`)

# ANY_AE % per arm (from ae_ard)
ae_display <- ae_ard |>
  filter(variable == "ANY_AE", stat_name == "p") |>
  mutate(`ANY_AE %` = unlist(stat)) |>
  select(Arm = group1_level, `ANY_AE %`)

# Join into one display table
age_display |>
  left_join(diabfl_display, by = "Arm") |>
  left_join(ae_display,     by = "Arm")
```

```{r capstone-display-hint-1}
# Each sub-object is pre-built; just run the code and verify the joined table.
# Notice all three join on "Arm" which maps to group1_level in the ARDs.
```

```{r capstone-display-solution}
age_display <- demo_ard |>
  filter(variable == "AGE", stat_name == "mean") |>
  mutate(stat = round(unlist(stat), 1)) |>
  select(Arm = group1_level, `Mean AGE` = stat)

diabfl_display <- diabfl_ci_ard |>
  filter(variable == "DIABFL",
         stat_name %in% c("p","conf.low","conf.high")) |>
  mutate(stat = unlist(stat)) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(`DIABFL % (95% CI)` = sprintf("%.1f (%.1f, %.1f)",
                                       p, conf.low, conf.high)) |>
  select(Arm = group1_level, `DIABFL % (95% CI)`)

ae_display <- ae_ard |>
  filter(variable == "ANY_AE", stat_name == "p") |>
  mutate(`ANY_AE %` = unlist(stat)) |>
  select(Arm = group1_level, `ANY_AE %`)

age_display |>
  left_join(diabfl_display, by = "Arm") |>
  left_join(ae_display,     by = "Arm")
```

---

## Lesson Summary

| Function | Input | Key output stat_names |
|----------|-------|-----------------------|
| `ard_continuous_ci()` | numeric columns | `mean`, `conf.low`, `conf.high`, `n` |
| `ard_proportion_ci()` | binary flag columns | `n`, `p`, `conf.low`, `conf.high` |
| `ard_regression_basic()` | fitted model | `estimate`, `conf.low`, `conf.high`, `p.value` |
| `ard_demographics()` | any columns + `by` | continuous + categorical + missing combined |
| `ard_ae_summary()` | ADSL + ADAE | `ANY_AE`, `SAE`, `SEV_AE` CIs + SOC counts |

### Key takeaways

- cardinal functions always return a `card`-class tibble compatible with `ard_stack()`.
- All statistics live in the `stat` list-column; use `unlist(stat)` or
  `ard_get_stat()` to extract numeric values.
- `ard_demographics()` auto-detects numeric vs. character columns — pass
  `continuous_vars` / `categorical_vars` when you need precise control.
- `ard_ae_summary()` uses ADSL as the denominator, ensuring subjects with no AEs
  are correctly counted in the proportion.
- Wilson CI (`method="wilson"`) is the regulatory default for binary outcomes
  because it stays within [0, 100] even for extreme proportions.
- Stack everything with `ard_stack()` to produce a single, submission-ready ARD
  that flows directly into `tfrmt::print_to_gt()`.

**Next lesson:** Lesson 29 — tfrmt Part 1 — table shells, body_plan(), frmt(), and column structure for regulatory TLF automation.
```
