---
title: "Lesson 28 — gtsummary Part 1: tbl_summary() and Clinical Demographic Tables"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build clinical demographic and baseline characteristic tables using
  gtsummary's tbl_summary(): stratify by treatment arm, customise
  statistic formats (mean/SD vs median [IQR]), add p-values with
  add_p(), add overall column with add_overall(), annotate with
  bold_labels() and italicize_levels(), extract results programmatically
  with as_tibble() and inline_text(), and assemble a complete Table
  14.1.1-style demographics table.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 28_gtsummary_part1.Rmd
# PURPOSE       : gtsummary Part 1 — tbl_summary(), add_p(), add_overall(),
#                 add_n(), modify_header(), bold_labels(), italicize_levels(),
#                 as_tibble(), inline_text(), and a full Table 14.1.1-style
#                 demographics table capstone.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo","Xanomeline Low Dose","Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ══════════════════════════════════════════════════════════════════════════════
# GTSUMMARY MOCK FUNCTIONS
# Replicate the real gtsummary API so exercises run inside learnr without the
# package installed. The mock computes real statistics from base R and returns
# an S3 "gtsummary" object whose table_body is a tibble following the same
# column layout as the real package.
# ══════════════════════════════════════════════════════════════════════════════

# Helper: format a continuous statistic string
.fmt_stat <- function(x, fmt) {
  mean_v   <- round(mean(x, na.rm=TRUE), 1)
  sd_v     <- round(sd(x, na.rm=TRUE), 1)
  median_v <- round(median(x, na.rm=TRUE), 1)
  p25_v    <- round(quantile(x, 0.25, na.rm=TRUE), 1)
  p75_v    <- round(quantile(x, 0.75, na.rm=TRUE), 1)
  fmt |>
    gsub("\\{mean\\}",   mean_v,   x = _) |>
    gsub("\\{sd\\}",     sd_v,     x = _) |>
    gsub("\\{median\\}", median_v, x = _) |>
    gsub("\\{p25\\}",    p25_v,    x = _) |>
    gsub("\\{p75\\}",    p75_v,    x = _)
}

tbl_summary <- function(data, by = NULL, include = NULL,
                        statistic = list(
                          all_continuous  = "{mean} ({sd})",
                          all_categorical = "{n} ({p}%)"
                        ),
                        digits = NULL, label = NULL, missing = "ifany") {
  all_vars <- names(data)
  if (!is.null(by)) all_vars <- all_vars[all_vars != by]
  vars <- if (!is.null(include)) include else all_vars

  cont_fmt <- if (!is.null(statistic$all_continuous))  statistic$all_continuous  else "{mean} ({sd})"
  cat_fmt  <- if (!is.null(statistic$all_categorical)) statistic$all_categorical else "{n} ({p}%)"

  arms <- if (!is.null(by)) as.character(sort(unique(na.omit(data[[by]])))) else "Overall"

  rows <- lapply(vars, function(vv) {
    x_all     <- data[[vv]]
    is_cont   <- is.numeric(x_all)
    var_label <- if (!is.null(label) && !is.null(label[[vv]])) label[[vv]] else vv

    if (is_cont) {
      r <- tibble(variable = vv, label = var_label, row_type = "label")
      for (arm in arms) {
        xi <- if (arm == "Overall") x_all else x_all[as.character(data[[by]]) == arm]
        r[[paste0("stat_", arm)]] <- .fmt_stat(xi, cont_fmt)
      }
      if (missing != "no" && sum(is.na(x_all)) > 0) {
        rm2 <- tibble(variable = vv, label = "  Unknown", row_type = "missing")
        for (arm in arms) {
          xi <- if (arm == "Overall") x_all else x_all[as.character(data[[by]]) == arm]
          rm2[[paste0("stat_", arm)]] <- as.character(sum(is.na(xi)))
        }
        r <- bind_rows(r, rm2)
      }
      r
    } else {
      header <- tibble(variable = vv, label = var_label, row_type = "label")
      for (arm in arms) header[[paste0("stat_", arm)]] <- NA_character_
      levs <- sort(unique(na.omit(as.character(x_all))))
      level_rows <- lapply(levs, function(lv) {
        lr <- tibble(variable = vv, label = paste0("  ", lv), row_type = "level")
        for (arm in arms) {
          xi    <- if (arm == "Overall") x_all else x_all[as.character(data[[by]]) == arm]
          total <- sum(!is.na(xi))
          n_lv  <- sum(xi == lv, na.rm = TRUE)
          p_lv  <- round(100 * n_lv / total, 1)
          lr[[paste0("stat_", arm)]] <- cat_fmt |>
            gsub("\\{n\\}", n_lv, x = _) |>
            gsub("\\{p\\}", p_lv, x = _)
        }
        lr
      })
      bind_rows(header, level_rows)
    }
  }) |> bind_rows()

  structure(
    list(table_body = rows, by = by, arms = arms, data = data,
         include = vars, label = label),
    class = "gtsummary"
  )
}

print.gtsummary <- function(x, ...) {
  stat_cols <- grep("^stat_", names(x$table_body), value = TRUE)
  display   <- x$table_body |> select(label, all_of(stat_cols))
  names(display)[names(display) != "label"] <-
    gsub("^stat_", "", names(display)[names(display) != "label"])
  cat("\n[gtsummary table]\n\n")
  print(display, n = Inf)
  invisible(x)
}

as_tibble.gtsummary <- function(x, ...) {
  stat_cols <- grep("^stat_", names(x$table_body), value = TRUE)
  display   <- x$table_body |>
    select(variable, label, row_type, all_of(stat_cols))
  names(display)[names(display) %in% stat_cols] <-
    gsub("^stat_", "", stat_cols)
  display
}

add_p <- function(x, test = NULL, pvalue_fun = NULL, ...) UseMethod("add_p")
add_p.gtsummary <- function(x, test = NULL, pvalue_fun = NULL, ...) {
  if (is.null(x$by)) { message("add_p() requires a 'by' variable"); return(x) }
  p_rows <- lapply(x$include, function(vv) {
    xi  <- x$data[[vv]]
    grp <- x$data[[x$by]]
    pval <- tryCatch({
      if (is.numeric(xi)) {
        levs <- unique(na.omit(grp))
        if (length(levs) == 2) {
          t.test(xi[grp == levs[1]], xi[grp == levs[2]])$p.value
        } else {
          summary(aov(xi ~ as.factor(grp)))[[1]][["Pr(>F)"]][1]
        }
      } else {
        chisq.test(table(xi, grp), simulate.p.value = TRUE)$p.value
      }
    }, error = function(e) NA_real_)
    tibble(variable  = vv,
           p.value   = round(pval, 3),
           p_display = if (!is.na(pval)) formatC(pval, digits = 3, format = "f") else "NA")
  }) |> bind_rows()
  x$table_body <- left_join(x$table_body, p_rows, by = "variable")
  x$p_added    <- TRUE
  x
}

add_overall <- function(x, col_label = "**Overall**", last = FALSE, ...) UseMethod("add_overall")
add_overall.gtsummary <- function(x, col_label = "**Overall**", last = FALSE, ...) {
  if (is.null(x$by)) return(x)
  tbl_overall  <- tbl_summary(x$data, by = NULL, include = x$include, label = x$label)
  overall_col  <- tbl_overall$table_body |>
    select(variable, label, row_type, stat_Overall)
  x$table_body <- left_join(x$table_body, overall_col,
                             by = c("variable", "label", "row_type"))
  x$overall_added <- TRUE
  x
}

add_n <- function(x, ...) UseMethod("add_n")
add_n.gtsummary <- function(x, ...) {
  n_row <- tibble(variable = "__n__", label = "N", row_type = "N")
  for (arm in x$arms) {
    n_arm <- if (arm == "Overall") nrow(x$data)
             else sum(as.character(x$data[[x$by]]) == arm, na.rm = TRUE)
    n_row[[paste0("stat_", arm)]] <- as.character(n_arm)
  }
  if (!is.null(x$table_body$stat_Overall))
    n_row$stat_Overall <- as.character(nrow(x$data))
  x$table_body <- bind_rows(n_row, x$table_body)
  x
}

modify_header <- function(x, ...) UseMethod("modify_header")
modify_header.gtsummary <- function(x, ...) {
  dots <- list(...)
  if (!is.null(dots)) x$custom_headers <- dots
  x
}

bold_labels <- function(x)           { UseMethod("bold_labels") }
bold_labels.gtsummary <- function(x) { x$bold_labels <- TRUE; x }

italicize_levels <- function(x)           { UseMethod("italicize_levels") }
italicize_levels.gtsummary <- function(x) { x$italicize_levels <- TRUE; x }

as_gt <- function(x, ...) UseMethod("as_gt")
as_gt.gtsummary <- function(x, ...) {
  cat("[as_gt(): In a live session this renders a formatted HTML/RTF table via the gt package.]\n")
  invisible(as_tibble(x))
}

inline_text <- function(x, variable, level = NULL, column = NULL, ...) UseMethod("inline_text")
inline_text.gtsummary <- function(x, variable, level = NULL, column = NULL, ...) {
  body <- x$table_body
  row  <- if (!is.null(level)) {
    body[body$variable == variable & grepl(level, body$label), ]
  } else {
    body[body$variable == variable & body$row_type %in% c("label", "N"), ]
  }
  if (nrow(row) == 0) return(NA_character_)
  col_nm <- if (!is.null(column)) paste0("stat_", column) else
    grep("^stat_", names(row), value = TRUE)[1]
  as.character(row[[col_nm]][1])
}

theme_gtsummary_compact  <- function() invisible(NULL)
set_gtsummary_theme      <- function(x) invisible(NULL)
reset_gtsummary_theme    <- function() invisible(NULL)

# Simplified selector helpers
all_continuous  <- function() "all_continuous"
all_categorical <- function() "all_categorical"
```

## Overview

**gtsummary** is the most widely used R package for producing publication-ready summary tables in clinical research. Its role in the pharmaverse pipeline sits just above the ARD layer:

```
Data / ARD  →  gtsummary (compute + format)  →  gt / flextable / Word / RTF
```

Where **cards** (Lesson 25) separates *computing* statistics from *displaying* them, gtsummary combines both steps in a concise, pipe-chainable API that can also accept a pre-built ARD via `tbl_ard_summary()`.

**Why gtsummary matters for clinical programming:**

- `tbl_summary()` produces Table 14.1.1-style demographics tables in a single call.
- Every result is extractable as a tibble (`as_tibble()`) or as inline R text (`inline_text()`), making it suitable for programmatic QC and CSR prose generation.
- The chain model (`|> add_p() |> add_overall() |> bold_labels()`) mirrors the SAS PROC REPORT + ODS pipeline but is fully reproducible and version-controllable.

**The gtsummary S3 object:**

Every gtsummary function returns a list of class `"gtsummary"`. The most important component is `$table_body` — a tibble with one row per variable (or per level for categorical variables) and one `stat_*` column per stratum.

| `$table_body` column | Contents |
|---------------------|---------|
| `variable` | Source column name |
| `label` | Display label (header for group; indented for level) |
| `row_type` | `"label"`, `"level"`, `"missing"`, or `"N"` |
| `stat_<arm>` | Formatted statistic string for that stratum |

### Trial Dataset

All exercises use the **CDISCPILOT01** Alzheimer's study from `pharmaverseadam` — 254 subjects in the safety population across three arms: `Placebo`, `Xanomeline Low Dose`, `Xanomeline High Dose`.

---

## Section 1 — tbl_summary() Basics

`tbl_summary()` is the workhorse. Its core arguments are:

| Argument | Purpose |
|----------|---------|
| `data` | A data frame (filtered safety population recommended) |
| `by` | Column name to stratify by (quoted string) |
| `include` | Character vector of variables to include |
| `statistic` | Named list of format strings |
| `label` | Named list of display labels |
| `missing` | `"ifany"` (default), `"always"`, or `"no"` |

### Exercise 1a — First demographics table

Build a table showing AGE, SEX, and RACE stratified by `TRT01P`, then print it.

```{r tbl-basic, exercise=TRUE, exercise.lines=10}
tbl1 <- tbl_summary(
  data    = ___,
  by      = ___,
  include = ___
)

tbl1
```

```{r tbl-basic-hint-1}
# Pass the adsl safety population.
# by = "TRT01P"  (quoted string, not the bare column name)
# include = c("AGE", "SEX", "RACE")
```

```{r tbl-basic-solution}
tbl1 <- tbl_summary(
  data    = adsl,
  by      = "TRT01P",
  include = c("AGE", "SEX", "RACE")
)

tbl1
```

### Exercise 1b — View table_body as a tibble

`as_tibble()` converts the gtsummary object to a plain tibble for programmatic use. Filter it to rows where `variable == "SEX"`.

```{r tbl-tibble, exercise=TRUE, exercise.lines=8}
tbl1 <- tbl_summary(adsl, by = "TRT01P", include = c("AGE", "SEX", "RACE"))

as_tibble(tbl1) |>
  filter(___ == ___)
```

```{r tbl-tibble-hint-1}
# filter(variable == "SEX")
```

```{r tbl-tibble-solution}
tbl1 <- tbl_summary(adsl, by = "TRT01P", include = c("AGE", "SEX", "RACE"))

as_tibble(tbl1) |>
  filter(variable == "SEX")
```

```{r quiz-sec1, echo=FALSE}
quiz(
  question(
    "What does `row_type = 'level'` mean in the table_body tibble?",
    answer("The row is a group header for a categorical variable"),
    answer("The row holds a formatted statistic for one level of a categorical variable",
           correct = TRUE),
    answer("The row is a missing-value count"),
    answer("The row is a continuous summary"),
    allow_retry = TRUE
  ),
  question(
    "When `by = NULL` in tbl_summary(), what column name does the mock use for the statistic?",
    answer("stat_All"),
    answer("stat_Total"),
    answer("stat_Overall", correct = TRUE),
    answer("stat_NA"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — The statistic= Argument

By default, continuous variables show `mean (SD)` and categoricals show `n (%)`. Override this with the `statistic` argument, which accepts a named list with entries `all_continuous` and `all_categorical`.

**Common format tokens:**

| Token | Meaning |
|-------|---------|
| `{mean}` | Arithmetic mean |
| `{sd}` | Standard deviation |
| `{median}` | Median |
| `{p25}` | 25th percentile |
| `{p75}` | 75th percentile |
| `{n}` | Count in level |
| `{p}` | Percentage |

### Exercise 2a — Median [IQR] format

Change the continuous statistic to `"{median} [{p25}, {p75}]"` and keep the categorical default.

```{r stat-median, exercise=TRUE, exercise.lines=10}
tbl2 <- tbl_summary(
  data      = adsl,
  by        = "TRT01P",
  include   = c("AGE", "SEX", "WEIGHTBL"),
  statistic = list(
    all_continuous  = ___,
    all_categorical = "{n} ({p}%)"
  )
)

tbl2
```

```{r stat-median-hint-1}
# all_continuous = "{median} [{p25}, {p75}]"
```

```{r stat-median-solution}
tbl2 <- tbl_summary(
  data      = adsl,
  by        = "TRT01P",
  include   = c("AGE", "SEX", "WEIGHTBL"),
  statistic = list(
    all_continuous  = "{median} [{p25}, {p75}]",
    all_categorical = "{n} ({p}%)"
  )
)

tbl2
```

### Exercise 2b — Per-variable label override

Use the `label` argument to display `AGE` as `"Age (years)"` and `WEIGHTBL` as `"Baseline Weight (kg)"`.

```{r stat-label, exercise=TRUE, exercise.lines=12}
tbl2b <- tbl_summary(
  data    = adsl,
  by      = "TRT01P",
  include = c("AGE", "SEX", "WEIGHTBL"),
  label   = list(
    AGE      = ___,
    WEIGHTBL = ___
  )
)

tbl2b
```

```{r stat-label-hint-1}
# label = list(AGE = "Age (years)", WEIGHTBL = "Baseline Weight (kg)")
```

```{r stat-label-solution}
tbl2b <- tbl_summary(
  data    = adsl,
  by      = "TRT01P",
  include = c("AGE", "SEX", "WEIGHTBL"),
  label   = list(
    AGE      = "Age (years)",
    WEIGHTBL = "Baseline Weight (kg)"
  )
)

tbl2b
```

---

## Section 3 — add_p(): P-values

`add_p()` appends a p-value column by performing an appropriate test for each variable:

- **Continuous** with 2 arms → Welch t-test; 3+ arms → one-way ANOVA.
- **Categorical** → chi-squared test (with Monte Carlo simulation for small cells).

### Exercise 3a — Add p-values

Chain `add_p()` onto a tbl_summary of AGE, SEX, and RACE.

```{r addp-basic, exercise=TRUE, exercise.lines=10}
tbl3 <- tbl_summary(
  adsl,
  by      = "TRT01P",
  include = c("AGE", "SEX", "RACE")
) |>
  ___()

tbl3
```

```{r addp-basic-hint-1}
# |> add_p()
```

```{r addp-basic-solution}
tbl3 <- tbl_summary(
  adsl,
  by      = "TRT01P",
  include = c("AGE", "SEX", "RACE")
) |>
  add_p()

tbl3
```

### Exercise 3b — Extract the AGE p-value programmatically

After calling `add_p()`, the `table_body` gains a `p.value` column. Pull the p-value for AGE.

```{r addp-extract, exercise=TRUE, exercise.lines=10}
tbl3 <- tbl_summary(adsl, by = "TRT01P", include = c("AGE", "SEX", "RACE")) |>
  add_p()

age_p <- tbl3$table_body |>
  filter(variable == ___, row_type == "label") |>
  pull(___)

age_p
```

```{r addp-extract-hint-1}
# filter(variable == "AGE", row_type == "label") |> pull(p.value)
```

```{r addp-extract-solution}
tbl3 <- tbl_summary(adsl, by = "TRT01P", include = c("AGE", "SEX", "RACE")) |>
  add_p()

age_p <- tbl3$table_body |>
  filter(variable == "AGE", row_type == "label") |>
  pull(p.value)

age_p
```

```{r quiz-sec3, echo=FALSE}
quiz(
  question(
    "Which statistical test does gtsummary use by default for a continuous variable with three treatment arms?",
    answer("Kruskal-Wallis test"),
    answer("One-way ANOVA", correct = TRUE),
    answer("Welch two-sample t-test"),
    answer("Wilcoxon rank-sum test"),
    allow_retry = TRUE
  )
)
```

---

## Section 4 — add_overall() and add_n()

`add_overall()` appends a column containing statistics computed across all strata combined.
`add_n()` prepends a row showing the subject count per stratum.

Both functions return the same `gtsummary` class, so they chain naturally.

### Exercise 4 — Overall column and N row

Chain `add_overall()` then `add_n()` onto a baseline demographics table.

```{r overall-n, exercise=TRUE, exercise.lines=12}
tbl4 <- tbl_summary(
  adsl,
  by      = "TRT01P",
  include = c("AGE", "SEX", "RACE", "WEIGHTBL")
) |>
  ___() |>
  ___()

tbl4
```

```{r overall-n-hint-1}
# |> add_overall() |> add_n()
```

```{r overall-n-solution}
tbl4 <- tbl_summary(
  adsl,
  by      = "TRT01P",
  include = c("AGE", "SEX", "RACE", "WEIGHTBL")
) |>
  add_overall() |>
  add_n()

tbl4
```

---

## Section 5 — modify_header(), bold_labels(), italicize_levels()

Three annotation functions polish the final table:

| Function | Effect |
|----------|--------|
| `modify_header()` | Override column header text |
| `bold_labels()` | Mark variable-level header rows as bold (rendered in gt/flextable) |
| `italicize_levels()` | Mark category levels as italic |

### Exercise 5 — Full annotation pipeline

Build a complete chain: `tbl_summary |> add_p() |> add_overall() |> add_n() |> bold_labels() |> italicize_levels()`.

```{r annotate, exercise=TRUE, exercise.lines=14}
tbl5 <- tbl_summary(
  adsl,
  by      = "TRT01P",
  include = c("AGE", "SEX", "RACE")
) |>
  ___() |>
  ___() |>
  ___() |>
  ___() |>
  ___()

# Check that the flags are set
cat("bold_labels flag:", isTRUE(tbl5$bold_labels), "\n")
cat("italicize_levels flag:", isTRUE(tbl5$italicize_levels), "\n")
```

```{r annotate-hint-1}
# Chain in order: add_p(), add_overall(), add_n(), bold_labels(), italicize_levels()
```

```{r annotate-solution}
tbl5 <- tbl_summary(
  adsl,
  by      = "TRT01P",
  include = c("AGE", "SEX", "RACE")
) |>
  add_p() |>
  add_overall() |>
  add_n() |>
  bold_labels() |>
  italicize_levels()

# Check that the flags are set
cat("bold_labels flag:", isTRUE(tbl5$bold_labels), "\n")
cat("italicize_levels flag:", isTRUE(tbl5$italicize_levels), "\n")
```

---

## Section 6 — as_tibble() and Programmatic Extraction

`as_tibble()` converts a gtsummary object to a plain tibble, enabling downstream filtering, joins, and QC comparisons. This is the bridge between gtsummary and the broader tidyverse.

**Use cases:**
- Export to CSV for independent verification.
- Join with ARD output from cards to cross-check values.
- Filter to specific rows for inline reporting in CSR prose.

### Exercise 6a — Extract AGE rows only

Use `as_tibble()` and then filter to keep only rows where `variable == "AGE"`.

```{r extract-age, exercise=TRUE, exercise.lines=10}
tbl6 <- tbl_summary(adsl, by = "TRT01P", include = c("AGE", "SEX", "RACE"))

age_rows <- as_tibble(___) |>
  filter(___ == "AGE")

age_rows
```

```{r extract-age-hint-1}
# as_tibble(tbl6) |> filter(variable == "AGE")
```

```{r extract-age-solution}
tbl6 <- tbl_summary(adsl, by = "TRT01P", include = c("AGE", "SEX", "RACE"))

age_rows <- as_tibble(tbl6) |>
  filter(variable == "AGE")

age_rows
```

### Exercise 6b — Coefficient of variation as a custom statistic

Compute the coefficient of variation (CV = SD / mean x 100) for AGE by treatment arm.
`tbl_summary()` format strings do not include CV directly; compute it with `summarise()` instead.

```{r cv-age, exercise=TRUE, exercise.lines=12}
adsl |>
  group_by(TRT01P) |>
  summarise(
    mean_age = mean(AGE, na.rm = TRUE),
    sd_age   = sd(AGE, na.rm = TRUE),
    cv_age   = round(___ / ___ * 100, 1)
  )
```

```{r cv-age-hint-1}
# cv_age = round(sd_age / mean_age * 100, 1)
```

```{r cv-age-solution}
adsl |>
  group_by(TRT01P) |>
  summarise(
    mean_age = mean(AGE, na.rm = TRUE),
    sd_age   = sd(AGE, na.rm = TRUE),
    cv_age   = round(sd_age / mean_age * 100, 1)
  )
```

---

## Section 7 — Capstone: Complete Table 14.1.1-Style Demographics Table

In a clinical study report, Table 14.1.1 is the baseline demographics and characteristics table. It typically includes age, sex, race, age group, BMI category, and weight, stratified by treatment arm with p-values and an overall column.

### Capstone Exercise — Full demographics table

Build a complete Table 14.1.1 for the CDISCPILOT01 safety population:

- Variables: `AGE`, `SEX`, `RACE`, `AGEGR1`, `DIABFL`, `WEIGHTBL`
- Labels: `AGE` → `"Age (years)"`, `WEIGHTBL` → `"Baseline Weight (kg)"`, `DIABFL` → `"Obese at Baseline (BMI >= 30)"`, `AGEGR1` → `"Age Group"`
- Statistic: continuous → `"{mean} ({sd})"`, categorical → `"{n} ({p}%)"`
- Chain: `add_p() |> add_overall() |> add_n() |> bold_labels() |> italicize_levels()`
- After building the table, extract the p-value for AGE and store it in `age_p_val`

```{r capstone, exercise=TRUE, exercise.lines=32}
tbl_1411 <- tbl_summary(
  data      = ___,
  by        = "TRT01P",
  include   = c("AGE", "SEX", "RACE", "AGEGR1", "DIABFL", "WEIGHTBL"),
  statistic = list(
    all_continuous  = ___,
    all_categorical = ___
  ),
  label = list(
    AGE      = ___,
    WEIGHTBL = ___,
    DIABFL   = ___,
    AGEGR1   = ___
  )
) |>
  ___() |>
  ___() |>
  ___() |>
  ___() |>
  ___()

# Extract AGE p-value
age_p_val <- tbl_1411$table_body |>
  filter(variable == "AGE", row_type == "label") |>
  pull(___)

cat("AGE p-value:", age_p_val, "\n")
tbl_1411
```

```{r capstone-hint-1}
# data = adsl
# statistic: "{mean} ({sd})" and "{n} ({p}%)"
# Chain: add_p(), add_overall(), add_n(), bold_labels(), italicize_levels()
# pull(p.value) to extract the AGE p-value
```

```{r capstone-solution}
tbl_1411 <- tbl_summary(
  data      = adsl,
  by        = "TRT01P",
  include   = c("AGE", "SEX", "RACE", "AGEGR1", "DIABFL", "WEIGHTBL"),
  statistic = list(
    all_continuous  = "{mean} ({sd})",
    all_categorical = "{n} ({p}%)"
  ),
  label = list(
    AGE      = "Age (years)",
    WEIGHTBL = "Baseline Weight (kg)",
    DIABFL   = "Obese at Baseline (BMI >= 30)",
    AGEGR1   = "Age Group"
  )
) |>
  add_p() |>
  add_overall() |>
  add_n() |>
  bold_labels() |>
  italicize_levels()

# Extract AGE p-value
age_p_val <- tbl_1411$table_body |>
  filter(variable == "AGE", row_type == "label") |>
  pull(p.value)

cat("AGE p-value:", age_p_val, "\n")
tbl_1411
```

---

## Lesson Summary

In this lesson you built clinical demographic tables using gtsummary's `tbl_summary()`:

| Concept | Key function / argument |
|---------|------------------------|
| Stratified summary | `tbl_summary(data, by="TRT01P")` |
| Variable subset | `include = c("AGE","SEX","RACE")` |
| Custom stat format | `statistic = list(all_continuous="{median} [{p25}, {p75}]")` |
| Display labels | `label = list(AGE = "Age (years)")` |
| P-values | `add_p()` |
| Overall column | `add_overall()` |
| N per arm header | `add_n()` |
| Bold / italic | `bold_labels()`, `italicize_levels()` |
| Programmatic access | `as_tibble()`, `$table_body` |
| Inline reporting | `inline_text()` |
| Export to gt | `as_gt()` |

**Key gotchas to remember:**

> `by=` takes a **quoted string** (`"TRT01P"`), not a bare column name.

> `as_tibble()` is the escape hatch for any downstream computation — always use it before trying to do arithmetic on table contents.

> `add_p()` silently returns the object unchanged (with a message) if `by=` was not set. Always stratify first.

**Next lesson:** Lesson 29 — gtsummary Part 2 — Regression tables with `tbl_regression()`, univariable screening with `tbl_uvregression()`, table merging and stacking, and building gtsummary tables directly from a cards ARD with `tbl_ard_summary()`.
