---
title: "Lesson 29 — tfrmt Part 1: Table Shells, body_plan(), and frmt()"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  Learn the tfrmt package (Table Framework): separate table formatting metadata
  from ARD data using frmt(), frmt_combine(), frmt_structure(), body_plan(),
  tfrmt() spec objects, ard_to_tfrmt() conversion, and print_to_gt() / mock_df()
  rendering. Capstone builds a full demographics table end-to-end.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 29_tfrmt_part1.Rmd
# PURPOSE       : tfrmt Part 1 — table shells, body_plan(), frmt(), column
#                 structure, ard_to_tfrmt() conversion, and print_to_gt() /
#                 mock_df() rendering for regulatory TLF automation.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo","Xanomeline Low Dose","Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ══════════════════════════════════════════════════════════════════════════════
# CARDS BASE MOCKS (replicated from lesson 25)
# ══════════════════════════════════════════════════════════════════════════════

.ard_row <- function(group1=NA_character_, group1_level=NA_character_,
                     variable, variable_level=NA_character_,
                     context, stat_name, stat, stat_label) {
  tibble(
    group1         = group1,
    group1_level   = group1_level,
    variable       = variable,
    variable_level = variable_level,
    context        = context,
    stat_name      = stat_name,
    stat           = list(stat),
    stat_label     = stat_label
  )
}

ard_continuous <- function(data, variables, by = NULL,
                           statistics = list(
                             N      = function(x) sum(!is.na(x)),
                             mean   = function(x) mean(x, na.rm=TRUE),
                             sd     = function(x) sd(x, na.rm=TRUE),
                             median = function(x) median(x, na.rm=TRUE),
                             p25    = function(x) quantile(x, 0.25, na.rm=TRUE),
                             p75    = function(x) quantile(x, 0.75, na.rm=TRUE),
                             min    = function(x) min(x, na.rm=TRUE),
                             max    = function(x) max(x, na.rm=TRUE),
                             n_miss = function(x) sum(is.na(x)),
                             p_miss = function(x) round(100*mean(is.na(x)),1)
                           )) {
  stat_labels <- c(N="N (non-missing)", mean="Mean", sd="SD", median="Median",
                   p25="Q1 (25th pctile)", p75="Q3 (75th pctile)",
                   min="Min", max="Max", n_miss="N missing", p_miss="% missing")
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      x   <- sub[[vv]]
      lapply(names(statistics), function(sn) {
        val <- tryCatch(statistics[[sn]](x), error=function(e) NA_real_)
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="continuous", stat_name=sn,
                 stat=val, stat_label=stat_labels[sn])
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_categorical <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    levs <- sort(unique(as.character(data[[vv]])))
    levs <- levs[!is.na(levs)]
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      total <- sum(!is.na(sub[[vv]]))
      lapply(levs, function(lv) {
        cnt <- sum(sub[[vv]] == lv, na.rm=TRUE)
        bind_rows(
          .ard_row(group1=if(!is.null(by)) by else NA_character_,
                   group1_level=as.character(gl), variable=vv,
                   variable_level=lv, context="categorical",
                   stat_name="n", stat=cnt, stat_label="n"),
          .ard_row(group1=if(!is.null(by)) by else NA_character_,
                   group1_level=as.character(gl), variable=vv,
                   variable_level=lv, context="categorical",
                   stat_name="p", stat=round(100*cnt/total,1), stat_label="%")
        )
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_missing <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      x   <- sub[[vv]]
      bind_rows(
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="missing", stat_name="n_miss",
                 stat=sum(is.na(x)), stat_label="N missing"),
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="missing", stat_name="p_miss",
                 stat=round(100*mean(is.na(x)),1), stat_label="% missing"),
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="missing", stat_name="n_not_miss",
                 stat=sum(!is.na(x)), stat_label="N non-missing")
      )
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_stack <- function(...) {
  out <- bind_rows(...)
  structure(out, class=c("card","tbl_df","tbl","data.frame"))
}

bind_ard <- function(...) ard_stack(...)

ard_get_stat <- function(ard, stat_nm, var_nm=NULL) {
  x <- if (!is.null(var_nm)) filter(ard, variable==var_nm) else ard
  x |> filter(stat_name==stat_nm) |> pull(stat) |> unlist()
}

# ══════════════════════════════════════════════════════════════════════════════
# TFRMT FUNCTION MOCKS
# ══════════════════════════════════════════════════════════════════════════════

# frmt: specifies a number format expression
frmt <- function(expression, missing="", scientific=NULL, ...) {
  structure(list(expression=expression, missing=missing), class="frmt")
}

# frmt_combine: combines multiple stats into one formatted cell string
frmt_combine <- function(expression, ..., missing="") {
  dots <- list(...)
  structure(list(expression=expression, frmts=dots, missing=missing),
            class="frmt_combine")
}

# frmt_structure: attaches formatting rules to specific rows
frmt_structure <- function(group_val=NULL, label_val=NULL, ...) {
  dots <- list(...)
  structure(list(group_val=group_val, label_val=label_val, frmts=dots),
            class="frmt_structure")
}

# body_plan: collects frmt_structure objects into a formatting plan
body_plan <- function(...) {
  structure(list(...), class="body_plan")
}

# col_plan: defines column ordering for the output table
col_plan <- function(...) {
  dots <- list(...)
  structure(dots, class="col_plan")
}

# span_structure: defines spanning column headers
span_structure <- function(...) {
  structure(list(...), class="span_structure")
}

# tfrmt: the main spec object
# Uses substitute() directly (not via a helper) so callers can pass bare
# unquoted column names — e.g. tfrmt(param = param, ...) — matching the
# real tfrmt NSE interface.  String literals also work: tfrmt(param = "col").
tfrmt <- function(param=NULL, value=NULL, label=NULL, column=NULL,
                  row_grp_plan=NULL, body_plan=NULL, col_plan=NULL,
                  sorting_cols=NULL, ...) {
  .s2c <- function(s) {   # symbol/expression → character
    if (is.null(s)) return(NULL)
    if (is.character(s)) return(s)
    deparse(s)
  }
  param_v  <- .s2c(substitute(param))
  value_v  <- .s2c(substitute(value))
  label_v  <- .s2c(substitute(label))
  column_v <- .s2c(substitute(column))
  structure(
    list(param=param_v, value=value_v, label=label_v, column=column_v,
         row_grp_plan=row_grp_plan, body_plan=body_plan,
         col_plan=col_plan, sorting_cols=sorting_cols),
    class="tfrmt"
  )
}

# Apply a frmt object to a single value
.apply_frmt <- function(val, frmt_obj) {
  if (inherits(frmt_obj, "frmt")) {
    if (is.na(val) || is.null(val)) return(frmt_obj$missing)
    fmt <- frmt_obj$expression
    if (grepl("\\.", fmt)) {
      dec <- nchar(sub(".*\\.", "", fmt))
      return(formatC(as.numeric(val), digits=dec, format="f"))
    }
    return(formatC(as.numeric(val), format="d"))
  }
  as.character(val)
}

# print_to_gt: apply tfrmt spec to ARD data and render a mock gt table
print_to_gt <- function(tfrmt_spec, .data, ...) {
  if (!inherits(tfrmt_spec, "tfrmt"))
    stop("tfrmt_spec must be a tfrmt object")

  param_col  <- if (!is.null(tfrmt_spec$param))  as.character(tfrmt_spec$param)  else "param"
  value_col  <- if (!is.null(tfrmt_spec$value))  as.character(tfrmt_spec$value)  else "value"
  label_col  <- if (!is.null(tfrmt_spec$label))  as.character(tfrmt_spec$label)  else "label"
  column_col <- if (!is.null(tfrmt_spec$column)) as.character(tfrmt_spec$column) else "column"

  required     <- c(param_col, value_col, label_col, column_col)
  missing_cols <- setdiff(required, names(.data))
  if (length(missing_cols) > 0) {
    warning("ARD missing columns: ", paste(missing_cols, collapse=", "))
    return(invisible(.data))
  }

  wide <- .data |>
    select(all_of(c(label_col, column_col, param_col, value_col))) |>
    pivot_wider(names_from = all_of(column_col),
                values_from = all_of(value_col))

  cat("\n[tfrmt -> gt table -- mock render]\n\n")
  print(wide, n=Inf)
  invisible(wide)
}

# mock_df: produce a blank table shell for sponsor sign-off
mock_df <- function(tfrmt_spec, .data=NULL) {
  cat("\n[mock_df() -- blank table shell for sign-off]\n")
  if (!is.null(.data)) {
    wide <- .data |>
      select(any_of(c(as.character(tfrmt_spec$label),
                      as.character(tfrmt_spec$column)))) |>
      distinct()
    cat("Rows:", nrow(wide), " Columns:", ncol(wide), "\n")
    print(wide, n=10)
  }
  invisible(tfrmt_spec)
}

# ard_to_tfrmt: convert a cards ARD to tfrmt-ready long format
ard_to_tfrmt <- function(ard, label_var="variable", group_var=NULL) {
  out <- ard |>
    mutate(value = unlist(stat)) |>
    select(
      label  = all_of(label_var),
      column = group1_level,
      param  = stat_name,
      value
    )
  if (!is.null(group_var)) out$group1 <- ard[[group_var]]
  out[!is.na(out$column), ]
}

# ══════════════════════════════════════════════════════════════════════════════
# PRE-BUILT OBJECTS USED IN EXERCISES
# ══════════════════════════════════════════════════════════════════════════════

# ARD for continuous demographics
cont_ard <- ard_continuous(
  adsl,
  variables = c("AGE","BMIBL"),
  by        = "TRT01P"
)

# ARD for categorical demographics
cat_ard <- ard_categorical(
  adsl,
  variables = c("SEX","RACE"),
  by        = "TRT01P"
)

# Combined demographics ARD
demo_ard <- ard_stack(cont_ard, cat_ard)

# tfrmt-ready version of the continuous ARD
tfrmt_cont <- ard_to_tfrmt(cont_ard, label_var="variable")

# tfrmt-ready combined demo ARD
tfrmt_demo <- ard_to_tfrmt(demo_ard, label_var="variable")

# A complete example tfrmt spec for a 3-arm demographics table
demo_spec <- tfrmt(
  param    = param,
  value    = value,
  label    = label,
  column   = column,
  body_plan = body_plan(
    frmt_structure(label_val="AGE",   frmt("xx.x")),
    frmt_structure(label_val="BMIBL", frmt("xx.x")),
    frmt_structure(label_val="SEX",
      frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x"))),
    frmt_structure(label_val="RACE",
      frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x")))
  )
)
```

## Overview

**tfrmt** (Table Framework) solves a fundamental problem in regulatory reporting:
how do you keep table *format* specifications separate from the underlying *data*?

### The core idea

In SAS, a PROC REPORT call bundles layout, column definitions, format statements,
and data together. Changing one arm label or decimal-place specification requires
touching every program. `tfrmt` splits these concerns cleanly:

```
 ARD data (from cards / cardinal)
        +
 tfrmt spec object (format metadata)
        │
        ▼
  print_to_gt()  ──►  gt table (HTML / RTF / PDF)
  mock_df()      ──►  blank shell for sponsor sign-off
```

The same `tfrmt` spec can be applied to:
- Real data for a study report
- Synthetic mock data for pre-approval sign-off
- Updated data at a database lock with zero code changes

### tfrmt vs. SAS PROC REPORT

| SAS PROC REPORT | tfrmt |
|----------------|-------|
| Column layout in `COLUMN` statement | `col_plan()` |
| `DEFINE var / DISPLAY FORMAT=` | `frmt("xx.x")` inside `frmt_structure()` |
| Combining stats in COMPUTE block | `frmt_combine("{n} ({p}%)", ...)` |
| Row grouping via `GROUP` / `ORDER` | `row_grp_plan()` |
| Hard-coded data source | Separate ARD passed to `print_to_gt()` |

### The tfrmt workflow in five steps

1. Build an ARD with `cards` / `cardinal` functions
2. Convert to tfrmt format with `ard_to_tfrmt()`
3. Define a `tfrmt()` spec with `body_plan()`, `col_plan()`, etc.
4. Preview the shell with `mock_df(spec, tfrmt_ready_ard)`
5. Render the final table with `print_to_gt(spec, tfrmt_ready_ard)`

### ARD structure required by tfrmt

tfrmt expects a long-format data frame with exactly four key columns:

| Column | Maps from | Meaning |
|--------|-----------|---------|
| `label` | `variable` in ARD | Row label (e.g. "AGE") |
| `column` | `group1_level` in ARD | Column header (e.g. treatment arm) |
| `param` | `stat_name` in ARD | Which statistic (e.g. "mean", "n") |
| `value` | `unlist(stat)` | The numeric value |

The `ard_to_tfrmt()` helper performs this reshaping automatically.

---

## Section 1 — frmt(): Format Expressions

`frmt()` defines how a single numeric value should be displayed.  The
`expression` argument is a pattern string where `x` represents a digit:

| Expression | Meaning | Example output |
|------------|---------|----------------|
| `"xx"` | Integer, 2 digits | `42` |
| `"xx.x"` | 1 decimal place | `42.3` |
| `"xx.xx"` | 2 decimal places | `42.31` |
| `"xx.xxx"` | 3 decimal places | `42.314` |
| `"x.xxxx"` | 4 decimal places | `3.1416` |

The `missing` argument controls what is displayed for `NA` values (default: `""`).

### Exercise 1-a — Create and inspect a frmt object

```{r frmt-create, exercise=TRUE, exercise.lines=10}
# Create a frmt with one decimal place
f1 <- frmt("xx.x")

# Inspect the structure
str(f1)

# What class does it have?
class(f1)
```

```{r frmt-create-hint-1}
# frmt() returns a list with elements: expression and missing
# class(f1) should be "frmt"
```

```{r frmt-create-solution}
f1 <- frmt("xx.x")
str(f1)
class(f1)
```

### Exercise 1-b — Apply .apply_frmt() and compare decimal places

```{r frmt-apply, exercise=TRUE, exercise.lines=14}
pi_val <- 3.14159265

# One decimal place
result_1 <- .apply_frmt(pi_val, frmt("___"))

# Two decimal places
result_2 <- .apply_frmt(pi_val, frmt("___"))

# Three decimal places
result_3 <- .apply_frmt(pi_val, frmt("___"))

cat(result_1, "\n", result_2, "\n", result_3, "\n")
```

```{r frmt-apply-hint-1}
# "xx.x"  -> one decimal place  -> "3.1"
# "xx.xx" -> two decimal places -> "3.14"
# "xx.xxx"-> three decimal places -> "3.142"
```

```{r frmt-apply-solution}
pi_val <- 3.14159265

result_1 <- .apply_frmt(pi_val, frmt("xx.x"))
result_2 <- .apply_frmt(pi_val, frmt("xx.xx"))
result_3 <- .apply_frmt(pi_val, frmt("xx.xxx"))

cat(result_1, "\n", result_2, "\n", result_3, "\n")
```

### Exercise 1-c — Handle missing values with the missing argument

```{r frmt-missing, exercise=TRUE, exercise.lines=10}
# Default missing: empty string
.apply_frmt(NA_real_, frmt("xx.x"))

# Custom missing string
.apply_frmt(NA_real_, frmt("xx.x", missing = "NC"))

# Verify real values still format normally
.apply_frmt(74.2, frmt("xx.x", missing = "NC"))
```

```{r frmt-missing-hint-1}
# missing = "NC" is common in regulatory tables (Not Calculable)
# Real values are unaffected by the missing argument
```

```{r frmt-missing-solution}
.apply_frmt(NA_real_, frmt("xx.x"))
.apply_frmt(NA_real_, frmt("xx.x", missing = "NC"))
.apply_frmt(74.2, frmt("xx.x", missing = "NC"))
```

```{r quiz-s1, echo=FALSE}
quiz(
  question(
    "In tfrmt, what does frmt('xx.x') control?",
    answer("The column width in the output table"),
    answer("The number of decimal places and display format for a single statistic",
           correct=TRUE),
    answer("The row grouping level of a variable"),
    answer("The column header text"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — frmt_combine(): Multi-Stat Cells

Most clinical table cells contain multiple statistics in one string: `"42 (16.7%)"`,
`"72.3 (8.1)"`, `"0.034 (0.012, 0.098)"`.  `frmt_combine()` assembles these from
individual `frmt()` objects using a glue-style `expression`.

```
frmt_combine(
  expression,     # glue string: "{n} ({p}%)"
  ...,            # named frmt objects matching placeholders
  missing = ""
)
```

### Exercise 2-a — Create a frmt_combine for n (p%)

```{r frmt-combine-create, exercise=TRUE, exercise.lines=10}
# Standard clinical table cell: "42 (16.7%)"
np_fmt <- frmt_combine(
  "___",                  # expression with {n} and {p} placeholders
  n = frmt("___"),        # integer count
  p = frmt("___")         # one decimal percentage
)

str(np_fmt)
```

```{r frmt-combine-create-hint-1}
# expression = "{n} ({p}%)"
# n = frmt("xxx")  — integer count (up to 3 digits)
# p = frmt("xx.x") — one decimal place
```

```{r frmt-combine-create-solution}
np_fmt <- frmt_combine(
  "{n} ({p}%)",
  n = frmt("xxx"),
  p = frmt("xx.x")
)

str(np_fmt)
```

### Exercise 2-b — Create a frmt_combine for mean (SD)

```{r frmt-combine-mean-sd, exercise=TRUE, exercise.lines=10}
# Cell format: "72.3 (8.1)"
mean_sd_fmt <- frmt_combine(
  "___",
  mean = frmt("___"),
  sd   = frmt("___")
)

class(mean_sd_fmt)
length(mean_sd_fmt$frmts)   # should be 2 (mean and sd)
```

```{r frmt-combine-mean-sd-hint-1}
# expression = "{mean} ({sd})"
# mean = frmt("xx.x"), sd = frmt("xx.x")
# frmts is the named list of individual frmt objects
```

```{r frmt-combine-mean-sd-solution}
mean_sd_fmt <- frmt_combine(
  "{mean} ({sd})",
  mean = frmt("xx.x"),
  sd   = frmt("xx.x")
)

class(mean_sd_fmt)
length(mean_sd_fmt$frmts)
```

```{r quiz-s2, echo=FALSE}
quiz(
  question(
    "In frmt_combine('{n} ({p}%)', n=frmt('xxx'), p=frmt('xx.x')), what do the curly braces {} around n and p signify?",
    answer("R environment variable lookup"),
    answer("Named placeholder slots that will be filled by the corresponding frmt objects",
           correct=TRUE),
    answer("Optional arguments that can be omitted"),
    answer("Regular expression capture groups"),
    allow_retry = TRUE
  )
)
```

---

## Section 3 — frmt_structure(): Row-Level Formatting Rules

`frmt_structure()` attaches a formatting rule to specific rows in the output
table, identified by `group_val` (row group) and/or `label_val` (row label).

```
frmt_structure(
  group_val = NULL,    # match on row group value
  label_val = NULL,    # match on row label value
  ...                  # one or more frmt / frmt_combine objects
)
```

### Exercise 3-a — Attach frmt to the AGE row

```{r frmt-struct-age, exercise=TRUE, exercise.lines=10}
# Format AGE mean to one decimal place
age_fmt_struct <- frmt_structure(
  label_val = "___",   # row label in the output table
  ___                  # frmt object
)

class(age_fmt_struct)
age_fmt_struct$label_val
```

```{r frmt-struct-age-hint-1}
# label_val = "AGE"
# Pass frmt("xx.x") as the second positional argument (or name it)
```

```{r frmt-struct-age-solution}
age_fmt_struct <- frmt_structure(
  label_val = "AGE",
  frmt("xx.x")
)

class(age_fmt_struct)
age_fmt_struct$label_val
```

### Exercise 3-b — Attach frmt_combine to the SEX row

```{r frmt-struct-sex, exercise=TRUE, exercise.lines=12}
# SEX row should display "n (p%)"
sex_fmt_struct <- frmt_structure(
  label_val = "___",
  frmt_combine(
    "___",
    n = frmt("___"),
    p = frmt("___")
  )
)

str(sex_fmt_struct)
```

```{r frmt-struct-sex-hint-1}
# label_val = "SEX"
# expression = "{n} ({p}%)"
# n = frmt("xxx"), p = frmt("xx.x")
```

```{r frmt-struct-sex-solution}
sex_fmt_struct <- frmt_structure(
  label_val = "SEX",
  frmt_combine(
    "{n} ({p}%)",
    n = frmt("xxx"),
    p = frmt("xx.x")
  )
)

str(sex_fmt_struct)
```

---

## Section 4 — body_plan(): Assembling the Formatting Plan

`body_plan()` collects multiple `frmt_structure()` calls into a single
formatting plan.  Rules are applied top-down; the first matching rule wins.

```
body_plan(
  frmt_structure(...),   # rule 1
  frmt_structure(...),   # rule 2
  ...
)
```

### Exercise 4-a — Build a body_plan for a demographics table

```{r body-plan-build, exercise=TRUE, exercise.lines=20}
demo_body <- body_plan(
  # AGE: mean to 1 decimal place
  frmt_structure(label_val = "AGE",   frmt("xx.x")),

  # BMIBL: mean to 1 decimal place
  frmt_structure(label_val = "BMIBL", frmt("xx.x")),

  # WEIGHTBL: mean to 1 decimal place
  frmt_structure(label_val = "WEIGHTBL", frmt("___")),

  # SEX: n (p%)
  frmt_structure(label_val = "SEX",
    frmt_combine("___", n=frmt("___"), p=frmt("___"))),

  # RACE: n (p%)
  frmt_structure(label_val = "RACE",
    frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x")))
)

class(demo_body)
length(demo_body)   # should be 5
```

```{r body-plan-build-hint-1}
# label_val = "WEIGHTBL", frmt("xx.x")
# frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x")) for SEX
# length(demo_body) counts the number of frmt_structure rules
```

```{r body-plan-build-solution}
demo_body <- body_plan(
  frmt_structure(label_val = "AGE",      frmt("xx.x")),
  frmt_structure(label_val = "BMIBL",    frmt("xx.x")),
  frmt_structure(label_val = "WEIGHTBL", frmt("xx.x")),
  frmt_structure(label_val = "SEX",
    frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x"))),
  frmt_structure(label_val = "RACE",
    frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x")))
)

class(demo_body)
length(demo_body)
```

### Exercise 4-b — Inspect a frmt_structure within the body_plan

```{r body-plan-inspect, exercise=TRUE, exercise.lines=10}
# Retrieve the second frmt_structure (BMIBL) from demo_body above
# and confirm its label_val

bp <- body_plan(
  frmt_structure(label_val="AGE",   frmt("xx.x")),
  frmt_structure(label_val="BMIBL", frmt("xx.x"))
)

# Access the second element
bp[[___]]$label_val
```

```{r body-plan-inspect-hint-1}
# body_plan is a list; bp[[2]] gets the second frmt_structure
# $label_val returns "BMIBL"
```

```{r body-plan-inspect-solution}
bp <- body_plan(
  frmt_structure(label_val="AGE",   frmt("xx.x")),
  frmt_structure(label_val="BMIBL", frmt("xx.x"))
)

bp[[2]]$label_val
```

---

## Section 5 — tfrmt(): The Spec Object

The `tfrmt()` function assembles all pieces — column mapping, body plan, column
ordering — into one spec object that can be saved, versioned, and reused.

```
tfrmt(
  param    = param,         # ARD column holding stat_name
  value    = value,         # ARD column holding numeric value
  label    = label,         # ARD column holding row label
  column   = column,        # ARD column holding treatment arm
  body_plan = body_plan(...),
  col_plan  = col_plan(...),
  row_grp_plan = NULL,
  sorting_cols = NULL
)
```

### Exercise 5-a — Build a minimal 3-arm demographics tfrmt spec

```{r tfrmt-build, exercise=TRUE, exercise.lines=24}
my_spec <- tfrmt(
  param  = ___,      # unquoted column name
  value  = ___,
  label  = ___,
  column = ___,
  body_plan = body_plan(
    frmt_structure(label_val="AGE",
                   frmt("xx.x")),
    frmt_structure(label_val="BMIBL",
                   frmt("xx.x")),
    frmt_structure(label_val="SEX",
                   frmt_combine("{n} ({p}%)",
                                n=frmt("xxx"), p=frmt("xx.x"))),
    frmt_structure(label_val="RACE",
                   frmt_combine("{n} ({p}%)",
                                n=frmt("xxx"), p=frmt("xx.x")))
  )
)

class(my_spec)
names(my_spec)
```

```{r tfrmt-build-hint-1}
# param = param, value = value, label = label, column = column
# These are the unquoted column names that ard_to_tfrmt() creates
```

```{r tfrmt-build-solution}
my_spec <- tfrmt(
  param  = param,
  value  = value,
  label  = label,
  column = column,
  body_plan = body_plan(
    frmt_structure(label_val="AGE",
                   frmt("xx.x")),
    frmt_structure(label_val="BMIBL",
                   frmt("xx.x")),
    frmt_structure(label_val="SEX",
                   frmt_combine("{n} ({p}%)",
                                n=frmt("xxx"), p=frmt("xx.x"))),
    frmt_structure(label_val="RACE",
                   frmt_combine("{n} ({p}%)",
                                n=frmt("xxx"), p=frmt("xx.x")))
  )
)

class(my_spec)
names(my_spec)
```

### Exercise 5-b — Add a col_plan to control column order

```{r tfrmt-colplan, exercise=TRUE, exercise.lines=16}
ordered_spec <- tfrmt(
  param    = param,
  value    = value,
  label    = label,
  column   = column,
  col_plan = col_plan(
    "label",
    "Placebo",
    "Xanomeline Low Dose",
    "Xanomeline High Dose"
  ),
  body_plan = body_plan(
    frmt_structure(label_val="AGE", frmt("xx.x"))
  )
)

# Inspect the col_plan slot
ordered_spec$col_plan
```

```{r tfrmt-colplan-hint-1}
# col_plan is a list of column names in the desired display order
# "label" is the row-label column; arms follow in the preferred order
```

```{r tfrmt-colplan-solution}
ordered_spec <- tfrmt(
  param    = param,
  value    = value,
  label    = label,
  column   = column,
  col_plan = col_plan(
    "label",
    "Placebo",
    "Xanomeline Low Dose",
    "Xanomeline High Dose"
  ),
  body_plan = body_plan(
    frmt_structure(label_val="AGE", frmt("xx.x"))
  )
)

ordered_spec$col_plan
```

---

## Section 6 — ard_to_tfrmt(): Converting an ARD

A cards ARD has list-column `stat` and columns named `variable`, `group1_level`,
`stat_name`.  tfrmt needs a flat data frame with `label`, `column`, `param`,
`value`.  `ard_to_tfrmt()` performs this reshaping in one step.

```
ard_to_tfrmt(
  ard,                    # a cards ARD (card-class tibble)
  label_var = "variable", # ARD column to use as row label
  group_var = NULL        # optional extra grouping column
)
```

> **Important:** `ard_to_tfrmt()` drops rows where `group1_level` is `NA`
> (i.e., unstratified rows). Always pass a `by`-stratified ARD.

### Exercise 6-a — Convert a continuous ARD to tfrmt format

```{r ard-to-tfrmt-cont, exercise=TRUE, exercise.lines=12}
# Build the ARD
age_ard <- ard_continuous(
  adsl,
  variables = c("AGE","BMIBL"),
  by        = "___"
)

# Convert to tfrmt format
age_tfrmt <- ard_to_tfrmt(age_ard, label_var = "variable")

# Inspect the result
glimpse(age_tfrmt)
```

```{r ard-to-tfrmt-cont-hint-1}
# by = "TRT01P"
# The result should have columns: label, column, param, value
# label = variable name (AGE / BMIBL), column = arm name
```

```{r ard-to-tfrmt-cont-solution}
age_ard <- ard_continuous(
  adsl,
  variables = c("AGE","BMIBL"),
  by        = "TRT01P"
)

age_tfrmt <- ard_to_tfrmt(age_ard, label_var = "variable")

glimpse(age_tfrmt)
```

### Exercise 6-b — Inspect the converted ARD

Use the pre-built `tfrmt_cont` object. Filter to `label == "AGE"` and
`param == "mean"` to verify that the three arm means appear as separate rows
with different `column` values.

```{r ard-to-tfrmt-inspect, exercise=TRUE, exercise.lines=10}
tfrmt_cont |>
  filter(label == "___", param == "___") |>
  select(label, column, param, value)
```

```{r ard-to-tfrmt-inspect-hint-1}
# label == "AGE", param == "mean"
# You should see three rows, one per treatment arm
```

```{r ard-to-tfrmt-inspect-solution}
tfrmt_cont |>
  filter(label == "AGE", param == "mean") |>
  select(label, column, param, value)
```

---

## Section 7 — print_to_gt() and mock_df()

### print_to_gt()

`print_to_gt(tfrmt_spec, .data)` applies the spec to a tfrmt-ready data frame
and renders the output as a gt table (in a real session) or prints a mock
render to the console (in this tutorial).

### mock_df()

`mock_df(tfrmt_spec, .data)` produces a **blank shell** that shows the table
structure without any data values — used for sponsor sign-off before final
data is available.  In SAS workflows this corresponds to a manually drawn
table shell in Word; tfrmt generates it programmatically from the spec.

### Exercise 7-a — Render the continuous ARD with print_to_gt()

```{r print-to-gt, exercise=TRUE, exercise.lines=14}
# Use the pre-built tfrmt_cont (tfrmt-ready continuous ARD)
# and demo_spec (tfrmt spec object)

# Filter to only mean rows for a clean render
mean_only <- tfrmt_cont |>
  filter(param == "___")

print_to_gt(demo_spec, .data = mean_only)
```

```{r print-to-gt-hint-1}
# param == "mean" — keep only the mean rows
# print_to_gt() pivots column to wide format (one column per arm)
```

```{r print-to-gt-solution}
mean_only <- tfrmt_cont |>
  filter(param == "mean")

print_to_gt(demo_spec, .data = mean_only)
```

### Exercise 7-b — Preview the table shell with mock_df()

```{r mock-df-preview, exercise=TRUE, exercise.lines=8}
# Generate a blank sign-off shell from the same spec
mock_df(demo_spec, .data = tfrmt_cont)
```

```{r mock-df-preview-hint-1}
# mock_df() shows distinct label x column combinations without values
# Use this to verify row structure before data is locked
```

```{r mock-df-preview-solution}
mock_df(demo_spec, .data = tfrmt_cont)
```

```{r quiz-s7, echo=FALSE}
quiz(
  question(
    "What is the purpose of mock_df() in a regulatory TLF workflow?",
    answer("It generates fake data to test the pipeline before real data arrives"),
    answer("It produces a blank table shell showing row and column structure for sponsor sign-off before database lock",
           correct=TRUE),
    answer("It exports the tfrmt spec to a mock PDF document"),
    answer("It checks whether the ARD conforms to CDISC standards"),
    allow_retry = TRUE
  )
)
```

---

## Section 8 — Capstone: End-to-End Demographics Table

Build a complete demographics table from scratch: ARD → tfrmt conversion → spec
definition → render.

### Exercise 8-a — Build the ARD

```{r capstone-ard, exercise=TRUE, exercise.lines=16}
# Step 1: continuous stats for AGE and BMIBL
cap_cont <- ard_continuous(
  data      = ___,
  variables = c("AGE","BMIBL"),
  by        = "TRT01P"
)

# Step 2: categorical stats for SEX and RACE
cap_cat <- ard_categorical(
  data      = ___,
  variables = c("SEX","RACE"),
  by        = "TRT01P"
)

# Step 3: stack into one ARD
cap_ard <- ard_stack(___, ___)

cap_ard |> count(context)
```

```{r capstone-ard-hint-1}
# data = adsl for both
# ard_stack(cap_cont, cap_cat)
# context should show "continuous" and "categorical"
```

```{r capstone-ard-solution}
cap_cont <- ard_continuous(
  data      = adsl,
  variables = c("AGE","BMIBL"),
  by        = "TRT01P"
)

cap_cat <- ard_categorical(
  data      = adsl,
  variables = c("SEX","RACE"),
  by        = "TRT01P"
)

cap_ard <- ard_stack(cap_cont, cap_cat)

cap_ard |> count(context)
```

### Exercise 8-b — Convert to tfrmt format

```{r capstone-convert, exercise=TRUE, exercise.lines=8}
# Convert the stacked ARD to tfrmt-ready long format
cap_tfrmt_data <- ard_to_tfrmt(___, label_var = "variable")

# Confirm the four required columns exist
names(cap_tfrmt_data)
nrow(cap_tfrmt_data)
```

```{r capstone-convert-hint-1}
# ard_to_tfrmt(cap_ard, label_var = "variable")
# Output should have columns: label, column, param, value
```

```{r capstone-convert-solution}
cap_tfrmt_data <- ard_to_tfrmt(cap_ard, label_var = "variable")

names(cap_tfrmt_data)
nrow(cap_tfrmt_data)
```

### Exercise 8-c — Define the tfrmt spec

```{r capstone-spec, exercise=TRUE, exercise.lines=28}
cap_spec <- tfrmt(
  param    = param,
  value    = value,
  label    = label,
  column   = column,
  col_plan = col_plan(
    "label",
    "Placebo",
    "Xanomeline Low Dose",
    "Xanomeline High Dose"
  ),
  body_plan = body_plan(
    # AGE and BMIBL: mean to 1 decimal place
    frmt_structure(label_val = "AGE",   frmt("___")),
    frmt_structure(label_val = "BMIBL", frmt("___")),
    # SEX and RACE: n (p%)
    frmt_structure(label_val = "SEX",
      frmt_combine("___", n=frmt("___"), p=frmt("___"))),
    frmt_structure(label_val = "RACE",
      frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x")))
  )
)

class(cap_spec)
```

```{r capstone-spec-hint-1}
# frmt("xx.x") for continuous variables
# frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x")) for categorical
```

```{r capstone-spec-solution}
cap_spec <- tfrmt(
  param    = param,
  value    = value,
  label    = label,
  column   = column,
  col_plan = col_plan(
    "label",
    "Placebo",
    "Xanomeline Low Dose",
    "Xanomeline High Dose"
  ),
  body_plan = body_plan(
    frmt_structure(label_val = "AGE",   frmt("xx.x")),
    frmt_structure(label_val = "BMIBL", frmt("xx.x")),
    frmt_structure(label_val = "SEX",
      frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x"))),
    frmt_structure(label_val = "RACE",
      frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x")))
  )
)

class(cap_spec)
```

### Exercise 8-d — Render and preview

```{r capstone-render, exercise=TRUE, exercise.lines=14}
# Use the pre-built cap objects from 8-b and 8-c equivalents
# (tfrmt_demo and demo_spec are available in the session)

# First: blank shell for review
mock_df(demo_spec, .data = tfrmt_demo)

# Then: render with mean rows only (clean demo of continuous stats)
mean_rows <- tfrmt_demo |>
  filter(param == "mean")

print_to_gt(demo_spec, .data = mean_rows)
```

```{r capstone-render-hint-1}
# mock_df() shows structure; print_to_gt() shows actual values
# Filtering to param == "mean" keeps one row per label per arm
```

```{r capstone-render-solution}
mock_df(demo_spec, .data = tfrmt_demo)

mean_rows <- tfrmt_demo |>
  filter(param == "mean")

print_to_gt(demo_spec, .data = mean_rows)
```

---

## Lesson Summary

| Object | Purpose |
|--------|---------|
| `frmt("xx.x")` | Format a single numeric stat to specified decimal places |
| `frmt_combine("{n} ({p}%)", n=frmt("xxx"), p=frmt("xx.x"))` | Combine multiple stats into one cell string |
| `frmt_structure(label_val="AGE", frmt("xx.x"))` | Attach a format rule to a specific row label |
| `body_plan(frmt_structure(...), ...)` | Collect all row format rules into one plan |
| `col_plan("label","Placebo",...)` | Specify column display order |
| `tfrmt(param, value, label, column, body_plan, col_plan)` | Assemble the complete table spec |
| `ard_to_tfrmt(ard)` | Reshape a cards ARD to tfrmt-ready format |
| `print_to_gt(spec, .data=ard)` | Apply spec to data and render a gt table |
| `mock_df(spec, .data=ard)` | Generate a blank sign-off shell |

### Key takeaways

- `tfrmt` **separates format metadata from data** — the spec object can be
  saved and version-controlled independently of the analysis results.
- The four required columns in a tfrmt-ready ARD are `label`, `column`,
  `param`, and `value`.  `ard_to_tfrmt()` maps from the standard cards ARD
  column names automatically.
- `frmt_structure()` rules match on `label_val`; the first matching rule in
  `body_plan()` wins.
- `mock_df()` replaces the manually drawn Word table shell that SAS programmers
  circulate for sponsor sign-off — the blank structure is generated directly
  from the spec, guaranteeing alignment with the final table.
- A tfrmt spec is immutable between database cut and final lock: only the ARD
  data changes; format, column layout, and row ordering remain identical.

**Next lesson:** Lesson 30 — tfrmt Part 2 — page_plan(), clinical table templates, and connecting tfrmt to RTF/PDF submission output.
```
