---
title: "Lesson 30 — cardinal Part 1: ARD-First TLG Templates Overview"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Meet cardinal (formerly falcon): the pharmaverse industry initiative that
  ships a harmonized, openly maintained catalog of Tables, Listings, and Graphs
  (TLG) templates for clinical reporting. Learn what cardinal is and is not, how
  it sits on top of the cards / cardx / gtsummary / tfrmt stack, the ARD-first
  template concept, how to browse and adapt catalog templates, how the catalog
  relates to NEST and to the emerging CDISC ARS standard, and how to build a
  basic ARD-backed summary table the cardinal way.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 30_cardinal_part1_overview.Rmd
# PURPOSE       : cardinal Part 1 — the harmonized TLG catalog. What cardinal is
#                 (an industry catalog of Quarto templates, not a function
#                 library), how it sits on cards/cardx/gtsummary/tfrmt, the
#                 ARD-first template concept, browsing and adapting templates,
#                 cardinal vs NEST, ARS alignment, and basic ARD table building.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-06-22
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    DTHFL_N = if_else(!is.na(DTHFL) & DTHFL == "Y", 1L, 0L, missing = 0L),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo","Xanomeline Low Dose","Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
if (!"TRT01P" %in% names(adae_raw))
  adae_raw <- left_join(adae_raw, select(adsl, USUBJID, TRT01P), by = "USUBJID")
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ══════════════════════════════════════════════════════════════════════════════
# CARDS BASE MOCKS (replicated from lesson 25)
#
# cardinal templates are built on the cards/cardx/gtsummary stack. Those packages
# may not be installed in this learnr container, so the functions a cardinal
# template would call are mocked below. The mocks return a `card`-class tibble
# in the same shape cards uses (group1 / variable / context / stat_name / stat /
# stat_label). This lets the exercises run standalone and lets you SEE the ARD a
# cardinal template produces before any display layer is applied.
# ══════════════════════════════════════════════════════════════════════════════

.ard_row <- function(group1=NA_character_, group1_level=NA_character_,
                     variable, variable_level=NA_character_,
                     context, stat_name, stat, stat_label) {
  tibble(
    group1         = group1,
    group1_level   = group1_level,
    variable       = variable,
    variable_level = variable_level,
    context        = context,
    stat_name      = stat_name,
    stat           = list(stat),
    stat_label     = stat_label
  )
}

ard_continuous <- function(data, variables, by = NULL,
                           statistics = list(
                             N      = function(x) sum(!is.na(x)),
                             mean   = function(x) mean(x, na.rm=TRUE),
                             sd     = function(x) sd(x, na.rm=TRUE),
                             median = function(x) median(x, na.rm=TRUE),
                             p25    = function(x) quantile(x, 0.25, na.rm=TRUE),
                             p75    = function(x) quantile(x, 0.75, na.rm=TRUE),
                             min    = function(x) min(x, na.rm=TRUE),
                             max    = function(x) max(x, na.rm=TRUE),
                             n_miss = function(x) sum(is.na(x)),
                             p_miss = function(x) round(100*mean(is.na(x)),1)
                           )) {
  stat_labels <- c(N="N (non-missing)", mean="Mean", sd="SD", median="Median",
                   p25="Q1 (25th pctile)", p75="Q3 (75th pctile)",
                   min="Min", max="Max", n_miss="N missing", p_miss="% missing")
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      x   <- sub[[vv]]
      lapply(names(statistics), function(sn) {
        val <- tryCatch(statistics[[sn]](x), error=function(e) NA_real_)
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="continuous", stat_name=sn,
                 stat=val, stat_label=stat_labels[sn])
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_categorical <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    levs <- sort(unique(as.character(data[[vv]])))
    levs <- levs[!is.na(levs)]
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      total <- sum(!is.na(sub[[vv]]))
      lapply(levs, function(lv) {
        cnt <- sum(sub[[vv]] == lv, na.rm=TRUE)
        bind_rows(
          .ard_row(group1=if(!is.null(by)) by else NA_character_,
                   group1_level=as.character(gl), variable=vv,
                   variable_level=lv, context="categorical",
                   stat_name="n", stat=cnt, stat_label="n"),
          .ard_row(group1=if(!is.null(by)) by else NA_character_,
                   group1_level=as.character(gl), variable=vv,
                   variable_level=lv, context="categorical",
                   stat_name="p", stat=round(100*cnt/total,1), stat_label="%")
        )
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_missing <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      x   <- sub[[vv]]
      bind_rows(
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="missing", stat_name="n_miss",
                 stat=sum(is.na(x)), stat_label="N missing"),
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="missing", stat_name="p_miss",
                 stat=round(100*mean(is.na(x)),1), stat_label="% missing"),
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="missing", stat_name="n_not_miss",
                 stat=sum(!is.na(x)), stat_label="N non-missing")
      )
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_stack <- function(...) {
  out <- bind_rows(...)
  structure(out, class=c("card","tbl_df","tbl","data.frame"))
}

bind_ard <- function(...) ard_stack(...)

ard_get_stat <- function(ard, stat_nm, var_nm=NULL) {
  x <- if (!is.null(var_nm)) filter(ard, variable==var_nm) else ard
  x |> filter(stat_name==stat_nm) |> pull(stat) |> unlist()
}

# ══════════════════════════════════════════════════════════════════════════════
# DISPLAY-LAYER MOCK
#
# A cardinal template ends with a gtsummary display call such as
# tbl_ard_summary(). We are not rendering a real gt/flextable here; this mock
# just turns the ARD into a tidy, human-readable preview tibble so you can see
# the "shape" of a finished table without installing gtsummary.
# ══════════════════════════════════════════════════════════════════════════════

tbl_ard_summary <- function(ard, by = NULL, label = NULL, ...) {
  out <- ard |>
    mutate(value = vapply(stat, function(s) {
      if (length(s) == 0 || is.null(s)) NA_character_ else as.character(round_if_num(s))
    }, character(1))) |>
    select(any_of(c("group1_level", "variable", "variable_level",
                    "context", "stat_name", "value")))
  structure(out, class = c("tbl_ard_preview", "tbl_df", "tbl", "data.frame"))
}

round_if_num <- function(s) if (is.numeric(s)) round(s, 2) else s

# ── Pre-built objects used in exercises ──────────────────────────────────────
demo_ard <- ard_stack(
  ard_continuous(adsl,  variables = c("AGE","WEIGHTBL","BMIBL"), by = "TRT01P"),
  ard_categorical(adsl, variables = c("SEX","RACE"),             by = "TRT01P")
)
```

## Overview

**cardinal** — formerly called **falcon** — is a pharmaverse **industry initiative**
rather than a typical R package. Its headline goal is to build and open-source a
**harmonized catalog** of Tables, Listings, and Graphs (**TLG**) for clinical
reporting.

The motivating problem is duplication. Every pharma company independently builds
essentially the same tables. The demographic table at one sponsor looks like the
demographic table at the next. Each company maintains its own templates, each
ages slightly differently, each requires its own re-validation, and reviewers see
slightly different output formats across submissions for what is conceptually the
same table.

cardinal proposes a different model: **one industry catalog** of TLG templates,
openly maintained, validated against the FDA Standard Safety Tables and Figures
guidance, that any sponsor can copy and adapt.

> **Origin.** cardinal was founded in 2023 with Roche, Boehringer Ingelheim,
> Moderna, and Sanofi. It was renamed from `{falcon}` to `{cardinal}` in 2024
> with broader pharmaverse alignment. The maintainer pool keeps growing.

### Lesson map

| Section | Topic |
|---------|-------|
| 1 | What cardinal is — and is not |
| 2 | The catalog model (Quarto, not `library(cardinal)`) |
| 3 | The Cardinal-future stack: cards / cardx / gtsummary / tfrmt / crane |
| 4 | The ARD-first template concept |
| 5 | Anatomy of a catalog template |
| 6 | Catalog categories and how to browse |
| 7 | Building a basic ARD the cardinal way |
| 8 | Customizing a template for your study |
| 9 | Contributing a template back |
| 10 | cardinal vs NEST, and CDISC ARS alignment |

---

## Section 1 — What cardinal is (and isn't)

The single most important thing to understand about cardinal is what kind of
thing it is.

cardinal is **not** a function library you install and call. You do **not** write
`library(cardinal)` and then call `cardinal::make_table()`. There is a small
`{cardinal}` R package that hosts catalog *infrastructure*, but that is not the
primary interaction model.

Instead, cardinal is a **Quarto-based catalog** of TLG templates, browsable as a
website:

```
https://pharmaverse.github.io/cardinal/
```

You **read** the catalog and **copy** templates into your study. The interaction
model is: browse → copy → adapt → render.

> **Mental model.** Think of cardinal less like a toolbox and more like a
> recipe book. The recipes (templates) are written, validated, and curated by the
> industry; you copy a recipe and cook it with your own ingredients (your study's
> ADaM data).

```{r quiz-s1, echo=FALSE}
quiz(
  question(
    "Which statement best describes how you interact with cardinal?",
    answer("Install it with install.packages() and call cardinal::make_table()"),
    answer("Browse a Quarto catalog website, copy a template, and adapt it to your study", correct=TRUE),
    answer("Add it as an admiral extension and derive variables with it"),
    answer("Submit your data to a cardinal server which returns finished RTFs"),
    allow_retry = TRUE
  ),
  question(
    "cardinal was previously known as:",
    answer("falcon", correct=TRUE),
    answer("tern"),
    answer("chevron"),
    answer("rtables"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — The catalog model

Each entry in the catalog is a Quarto document (`index.qmd`) showing:

- The **target TLG** — the specific FDA Standard Safety Tables and Figures table
  being implemented.
- The **full R code** to produce it.
- A **rendered preview** of the output.
- The **packages used** (typically cards, gtsummary, dplyr, sometimes specific
  therapeutic-area packages).
- **Notes on customization**.

Templates live under `pharmaverse/cardinal/quarto/catalog/` in the GitHub
repository, organized by FDA table number and category. Each template is a
self-contained Quarto document — you can render it on its own before integrating
it into your study.

### Exercise 2-a — Identify the catalog usage steps

The standard "use the catalog" workflow has the steps below — but two are
scrambled placeholders. Fix the `___` blanks so the ordered workflow reads
correctly.

```{r catalog-steps, exercise=TRUE, exercise.lines=10}
catalog_workflow <- c(
  "1. Navigate to the cardinal catalog website",
  "2. Browse the Catalog section and open a relevant template",
  "3. Read the methodology notes",
  "4. ___ the .qmd file into your project",
  "5. Replace test data references with your ___ data",
  "6. Customize per your SAP, then re-render"
)
cat(catalog_workflow, sep = "\n")
```

```{r catalog-steps-hint-1}
# Step 4 verb: you COPY the template file.
# Step 5: you swap pharmaverseadam test data for your STUDY data.
```

```{r catalog-steps-solution}
catalog_workflow <- c(
  "1. Navigate to the cardinal catalog website",
  "2. Browse the Catalog section and open a relevant template",
  "3. Read the methodology notes",
  "4. Copy the .qmd file into your project",
  "5. Replace test data references with your study data",
  "6. Customize per your SAP, then re-render"
)
cat(catalog_workflow, sep = "\n")
```

> **Why Quarto?** A Quarto template mixes prose, code, and rendered output in one
> document. A reviewer reading a template understands both the **methodology**
> (the *why*) and the **implementation** (the *what*) in a single place. cardinal
> templates are deliberately documentation-heavy for this reason.

---

## Section 3 — The Cardinal-future stack

cardinal templates are built on what the pharmaverse calls the **Cardinal-future
TLG stack**. Knowing which layer does what tells you, on sight, what any template
is doing.

| Layer | Packages | Job |
|-------|----------|-----|
| ARD | `{cards}`, `{cardx}` | Compute the **Analysis Results Dataset** — every statistic, tidy and machine-readable |
| Display | `{gtsummary}`, sometimes `{tfrmt}` | Turn the ARD into a formatted table |
| Pharma helpers | `{crane}` | Roche-flavoured gtsummary conventions |
| Test data | `{pharmaverseadam}`, `{pharmaversesdtm}` | CDISCPILOT01 data for examples |
| Rendering | `{quarto}`, `{rmarkdown}` | Produce the rendered document / RTF |

For graphs (the **G** in TLG), templates additionally use `{ggplot2}` (often with
`{tern}`-style themes), `{survminer}`, and `{ggsurvfit}` for Kaplan–Meier plots.

Every cardinal template uses **cards** for ARD construction; **most** use
gtsummary for display; **some** use tfrmt for complex layouts.

```{r quiz-s3, echo=FALSE}
quiz(
  question(
    "In the Cardinal-future stack, which package(s) construct the ARD?",
    answer("gtsummary"),
    answer("cards and cardx", correct=TRUE),
    answer("tfrmt"),
    answer("crane"),
    allow_retry = TRUE
  ),
  question(
    "Which package is the Roche-flavoured gtsummary helper layer?",
    answer("crane", correct=TRUE),
    answer("cardx"),
    answer("ggsurvfit"),
    answer("survminer"),
    allow_retry = TRUE
  )
)
```

---

## Section 4 — The ARD-first template concept

The defining design choice of the Cardinal-future stack is that it is
**ARD-first**. An **ARD** (Analysis Results Dataset) is a tidy, long, machine-
readable table where **one row = one statistic**. Columns describe *which* group,
*which* variable, *which* level, *what* statistic, and *its value*.

A cardinal template never jumps straight to a formatted table. It always:

1. **Builds an ARD** with cards/cardx (`ard_continuous()`, `ard_categorical()`,
   `ard_stack()`, …).
2. **Renders a display** from that ARD with gtsummary (`tbl_ard_summary()`).

This two-step separation is the whole point. The ARD is the submission artifact;
the display is a *derived rendering* of it. Save the ARD and you have the piece
CDISC ARS wants (more on that in Section 10).

Below is the structure of a cards-style ARD. The `stat` column is a **list-column**:
each cell holds the value, which you extract with `unlist(stat)`.

```{r ard-shape, exercise=TRUE, exercise.lines=8}
# demo_ard is pre-built in setup: continuous + categorical, by TRT01P
demo_ard |>
  select(group1_level, variable, variable_level, context, stat_name, stat) |>
  head(8)
```

```{r ard-shape-hint-1}
# Just run it. Note: context is "continuous" or "categorical",
# one row per statistic, and stat is a list-column.
```

```{r ard-shape-solution}
demo_ard |>
  select(group1_level, variable, variable_level, context, stat_name, stat) |>
  head(8)
```

### Exercise 4-a — Build a demographics ARD with ard_stack()

Reproduce the ARD-first first step of a cardinal demographics template: stack a
continuous ARD (AGE, WEIGHTBL, BMIBL) on top of a categorical ARD (SEX, RACE),
stratified by treatment arm.

```{r build-ard, exercise=TRUE, exercise.lines=12}
my_demo_ard <- ard_stack(
  ard_continuous(adsl,  variables = c("AGE","WEIGHTBL","BMIBL"), by = "___"),
  ard_categorical(adsl, variables = c("SEX","RACE"),             by = "___")
)

# How many rows came from each context?
my_demo_ard |>
  count(___)
```

```{r build-ard-hint-1}
# by = "TRT01P" in both calls
# count(context) shows continuous vs categorical contributions
```

```{r build-ard-solution}
my_demo_ard <- ard_stack(
  ard_continuous(adsl,  variables = c("AGE","WEIGHTBL","BMIBL"), by = "TRT01P"),
  ard_categorical(adsl, variables = c("SEX","RACE"),             by = "TRT01P")
)

my_demo_ard |>
  count(context)
```

### Exercise 4-b — Extract a single statistic from the ARD

From the pre-built `demo_ard`, pull the mean AGE for the Placebo arm. Remember
`stat` is a list-column.

```{r extract-stat, exercise=TRUE, exercise.lines=8}
demo_ard |>
  filter(variable == "AGE", stat_name == "___", group1_level == "___") |>
  pull(stat) |>
  unlist()
```

```{r extract-stat-hint-1}
# stat_name == "mean"; group1_level == "Placebo"
```

```{r extract-stat-solution}
demo_ard |>
  filter(variable == "AGE", stat_name == "mean", group1_level == "Placebo") |>
  pull(stat) |>
  unlist()
```

---

## Section 5 — Anatomy of a catalog template

A typical cardinal template `index.qmd` is documentation-heavy. Paraphrased, it
looks like this:

````
---
title: "FDA Table 14.2.1: Demographics and Baseline Characteristics"
---

## Description
This template implements the FDA Standard Safety Tables and Figures
demographics table...

## Required Packages
library(cards); library(cardx); library(gtsummary); library(dplyr)
library(pharmaverseadam)

## ADaM Inputs
- adsl: Subject-Level Analysis Dataset filtered to the safety population

## Code
adsl <- pharmaverseadam::adsl |> filter(SAFFL == "Y")
ard  <- ard_stack(adsl, ard_continuous(...), ard_categorical(...), .by = TRT01A)
tbl  <- ard |> tbl_ard_summary(...) |> bold_labels()
tbl

## Output
[rendered table preview]

## Customization Notes
- For studies with crossover, replace TRT01A with TRT01P or TRTSEQA
- Adjust footnote for sponsor terminology...
````

The five recurring sections — **Description**, **Required Packages**, **ADaM
Inputs**, **Code**, **Output**, **Customization Notes** — make every template
self-explanatory. The *why* is treated as importantly as the *what*.

```{r quiz-s5, echo=FALSE}
quiz(
  question(
    "A cardinal template is documentation-heavy primarily because:",
    answer("Quarto requires extensive YAML"),
    answer("A reader must understand both the methodology and the implementation to adapt and validate it", correct=TRUE),
    answer("FDA mandates a fixed prose length"),
    answer("gtsummary cannot run without prose"),
    allow_retry = TRUE
  )
)
```

---

## Section 6 — Catalog categories

The catalog organizes templates by TLG category:

- **fda-table_xx** — FDA Standard Safety Tables (the canonical baseline)
- **demographics** — demographics-focused tables
- **adverse-events** — AE summary tables
- **exposure** — exposure summary tables
- **laboratory** — lab tables
- **vital-signs** — VS summaries
- **survival** — survival analyses
- **subgroup** — subgroup forest plots and similar
- **listings** — subject-level listings

Each category has multiple entries, and the catalog is still growing — always
check the website for current coverage.

> The **fda-table_xx** category is the one to anchor on. If your CSR's tables are
> aligned to the FDA guide (most are, increasingly), those templates are a
> drop-in starting point. Part 2 of this lesson walks the most-used FDA safety
> table patterns in detail.

---

## Section 7 — Building a basic ARD-backed table

Let's do the full ARD-first round trip the cardinal way: build the ARD, then send
it to the display layer. In a real template the display step is `tbl_ard_summary()`
from gtsummary; here the setup chunk mocks that function to return a tidy preview
so the round trip runs standalone.

```{r basic-table, exercise=TRUE, exercise.lines=14}
# Step 1 — build the ARD (ARD-first)
ard <- ard_stack(
  ard_continuous(adsl,  variables = c("AGE","BMIBL"), by = "TRT01P"),
  ard_categorical(adsl, variables = "SEX",            by = "TRT01P")
)

# Step 2 — send the ARD to the display layer
tbl <- tbl_ard_summary(ard, by = "TRT01P")

# Preview just the continuous AGE rows
tbl |>
  filter(variable == "___", context == "continuous") |>
  select(group1_level, stat_name, value)
```

```{r basic-table-hint-1}
# variable == "AGE"
# The display layer is fed the ARD; it never recomputes the statistics.
```

```{r basic-table-solution}
ard <- ard_stack(
  ard_continuous(adsl,  variables = c("AGE","BMIBL"), by = "TRT01P"),
  ard_categorical(adsl, variables = "SEX",            by = "TRT01P")
)

tbl <- tbl_ard_summary(ard, by = "TRT01P")

tbl |>
  filter(variable == "AGE", context == "continuous") |>
  select(group1_level, stat_name, value)
```

### Exercise 7-a — Confirm the display did not recompute

A core ARD-first guarantee: the display layer reads pre-computed statistics; it
does not re-run the math. Verify that the mean SEX="F" percentage shown in the
table preview equals the value already sitting in the ARD.

```{r no-recompute, exercise=TRUE, exercise.lines=16}
ard <- ard_categorical(adsl, variables = "SEX", by = "TRT01P")
tbl <- tbl_ard_summary(ard, by = "TRT01P")

from_ard <- ard |>
  filter(variable == "SEX", variable_level == "F",
         stat_name == "p", group1_level == "Placebo") |>
  pull(stat) |>
  unlist()

from_tbl <- tbl |>
  filter(variable == "SEX", variable_level == "F",
         stat_name == "p", group1_level == "Placebo") |>
  pull(___)

c(from_ard = from_ard, from_tbl = as.numeric(from_tbl))
```

```{r no-recompute-hint-1}
# The display preview stores the rendered statistic in the `value` column.
```

```{r no-recompute-solution}
ard <- ard_categorical(adsl, variables = "SEX", by = "TRT01P")
tbl <- tbl_ard_summary(ard, by = "TRT01P")

from_ard <- ard |>
  filter(variable == "SEX", variable_level == "F",
         stat_name == "p", group1_level == "Placebo") |>
  pull(stat) |>
  unlist()

from_tbl <- tbl |>
  filter(variable == "SEX", variable_level == "F",
         stat_name == "p", group1_level == "Placebo") |>
  pull(value)

c(from_ard = from_ard, from_tbl = as.numeric(from_tbl))
```

---

## Section 8 — Customizing a template

cardinal templates are **starting points**, not finished products. They need
adaptation for your study. The four most common customizations:

**Replace test data with study data**

```r
# cardinal template (test data)
adsl <- pharmaverseadam::adsl

# your version (study data)
adsl <- haven::read_xpt("path/to/adsl.xpt")
```

**Adjust filtering for sponsor populations**

```r
# cardinal default
adsl <- adsl |> filter(SAFFL == "Y")

# sponsor with a custom analysis flag
adsl <- adsl |> filter(SAFFL == "Y" & ANL01FL == "Y")
```

**Apply sponsor styling**

```r
theme_gtsummary_compact()    # cardinal default
my_sponsor_theme()           # sponsor-internal theme function
```

**Adjust footnotes for sponsor terminology** — replace cardinal's generic
footnotes with your sponsor's SOP-required text. Purely cosmetic, but important
for SOP compliance.

### Exercise 8-a — Re-point a template's population filter

A copied template filters ADSL to the safety population. Your sponsor requires
*both* `SAFFL == "Y"` and the diabetes sub-population flag `DIABFL == "Y"`.
Adjust the filter, then count subjects per arm.

```{r custom-filter, exercise=TRUE, exercise.lines=10}
adsl_custom <- adsl |>
  filter(SAFFL == "Y" & ___ == "Y")

adsl_custom |>
  count(TRT01P)
```

```{r custom-filter-hint-1}
# The sponsor sub-population flag here is DIABFL.
```

```{r custom-filter-solution}
adsl_custom <- adsl |>
  filter(SAFFL == "Y" & DIABFL == "Y")

adsl_custom |>
  count(TRT01P)
```

> Notice the *pattern* did not change — only the filter. That is the cardinal
> value proposition: the validated structure stays; you change the inputs.

---

## Section 9 — Contributing a template back

The cardinal initiative explicitly invites contributions. To add a template:

1. Fork the `pharmaverse/cardinal` GitHub repo.
2. Create a new folder in `quarto/catalog/` with a unique name
   (e.g., `lab-creatinine-shift_01`).
3. Add `index.qmd` following the template structure (description, packages, code,
   output, notes).
4. Add a `result.png` snapshot of the rendered output.
5. Open a pull request.

The cardinal team reviews and merges. Reviewer feedback typically focuses on:

- **Methodology correctness** — does it match the FDA guide or a referenced standard?
- **Code quality** — does it use the Cardinal-future stack, is it readable, error-free?
- **Documentation completeness** — can a new programmer understand and adapt it?

> The contribution model encourages broad participation. If your sponsor has a
> useful TLG template that aligns with the FDA guide, contributing it benefits
> the whole industry — and your name appears as the contributor.

```{r quiz-s9, echo=FALSE}
quiz(
  question(
    "What does a cardinal contributor open to add a template to the catalog?",
    answer("A CRAN submission"),
    answer("A GitHub pull request against pharmaverse/cardinal", correct=TRUE),
    answer("A CDISC change-control ticket"),
    answer("An FDA pre-submission meeting request"),
    allow_retry = TRUE
  ),
  question(
    "Reviewer feedback on a contributed template usually does NOT focus on:",
    answer("Methodology correctness vs the FDA guide"),
    answer("Code quality and readability"),
    answer("Documentation completeness"),
    answer("Whether the template is faster than the equivalent SAS macro", correct=TRUE),
    allow_retry = TRUE
  )
)
```

---

## Section 10 — cardinal vs NEST, and CDISC ARS alignment

### cardinal vs NEST

A natural question: how does cardinal relate to Roche's **NEST** initiative
(rtables / tern / chevron, covered in a later module)?

| | NEST (legacy stack) | cardinal (future stack) |
|---|---|---|
| Maturity | Mature, production-proven | New, growing |
| Form | All-in-one packages | Templates, not monolithic packages |
| Built on | rtables → tern → chevron (+ r2rtf) | cards + cardx + gtsummary + tfrmt + ggplot |
| Best for | Highly customized / exotic layouts (esp. oncology) | Standard FDA-aligned tables; easy to adapt and validate |
| Direction | Existing pipelines maintained | Preferred for new starts; ARS-aligned |

Both stacks coexist. Many sponsors use rtables/tern for complex tables and
gtsummary for everything else. Strategically, cardinal is positioned to take on
NEST's role for new development, especially as CDISC ARS becomes more central.
Existing rtables pipelines won't be rewritten overnight — but new starts will
likely prefer Cardinal-future.

### CDISC ARS alignment

CDISC's **Analysis Results Standard (ARS)** is emerging as the regulatory
direction. ARS expects:

- A machine-readable **ARD** as part of every submission.
- The ARD linked to metadata describing how each statistic was computed.
- The ARD linked to the rendered table for verification.

Because cardinal templates are **ARD-first**, they produce both an ARD (via cards)
and a display (via gtsummary). If you save the intermediate ARD object, you
already have the ARS-ready artifact — the display is just a derived rendering.

When ARS becomes a formal submission requirement (anticipated 2026–2027),
cardinal-built tables are positioned to be ARS-compliant out of the box. Legacy
stack tables would need additional work to generate the corresponding ARDs.

### Exercise 10-a — Save the ARS-ready artifact

The ARD object — not the rendered table — is the ARS submission artifact. From a
demographics ARD, confirm it is a tidy long table by reporting its row count and
the distinct `context` values it contains.

```{r ars-artifact, exercise=TRUE, exercise.lines=10}
ars_artifact <- ard_stack(
  ard_continuous(adsl,  variables = c("AGE","BMIBL"), by = "TRT01P"),
  ard_categorical(adsl, variables = c("SEX","RACE"),  by = "TRT01P")
)

list(
  n_rows   = nrow(ars_artifact),
  contexts = ___(ars_artifact$context)
)
```

```{r ars-artifact-hint-1}
# Use unique() on the context column to list the distinct contexts.
```

```{r ars-artifact-solution}
ars_artifact <- ard_stack(
  ard_continuous(adsl,  variables = c("AGE","BMIBL"), by = "TRT01P"),
  ard_categorical(adsl, variables = c("SEX","RACE"),  by = "TRT01P")
)

list(
  n_rows   = nrow(ars_artifact),
  contexts = unique(ars_artifact$context)
)
```

```{r quiz-s10, echo=FALSE}
quiz(
  question(
    "Why is cardinal well-positioned for the CDISC ARS standard?",
    answer("It generates RTF faster than rtables"),
    answer("Its ARD-first templates produce a machine-readable ARD as the primary artifact, with the display as a derived rendering", correct=TRUE),
    answer("It is the only stack FDA accepts"),
    answer("It avoids gtsummary entirely"),
    allow_retry = TRUE
  ),
  question(
    "cardinal and NEST relate how?",
    answer("cardinal replaced NEST entirely in 2024"),
    answer("They coexist; NEST suits exotic layouts, cardinal suits standard FDA-aligned tables and is the likely future direction for new work", correct=TRUE),
    answer("NEST is a subset of cardinal"),
    answer("They are competing closed-source products"),
    allow_retry = TRUE
  )
)
```

---

## Lesson Summary

| Concept | Takeaway |
|---------|----------|
| What cardinal is | An industry catalog of TLG templates, not a function library |
| Interaction model | Browse → copy → adapt → render (Quarto `.qmd` files) |
| Underlying stack | cards / cardx (ARD) + gtsummary / tfrmt (display) + crane |
| ARD-first | Build the ARD, then render a display from it |
| Primary source | FDA Standard Safety Tables and Figures Integrated Guide |
| vs NEST | Coexist; cardinal is the future-leaning, ARS-aligned choice |
| ARS | The saved ARD is the submission artifact |

### Key takeaways

- **cardinal** (formerly **falcon**) is a pharmaverse industry initiative
  providing a harmonized, openly maintained catalog of TLG templates.
- It is **not** an R package you call — it is a Quarto-based template repository
  browsed on a website. You copy templates into your study.
- Templates implement the FDA Standard Safety Tables and Figures Integrated Guide,
  one table at a time, organized by FDA table number and category.
- Templates are built on the **Cardinal-future stack**: cards + cardx for the ARD,
  gtsummary (sometimes tfrmt) for display, crane for Roche helpers.
- The design is **ARD-first**: always build the ARD, then render a display from
  it. The ARD is the submission artifact.
- Use pattern: browse → copy → adapt. Customizations are usually cosmetic
  (population filter, footnotes, table numbering, styling).
- Contribution is open: fork, add a template, open a PR.
- cardinal coexists with NEST (the rtables/tern/chevron stack) and is aligned with
  the emerging CDISC ARS standard.

### What's next

**Lesson 31 — cardinal Part 2: FDA Safety Table Templates** walks through the
most heavily-used cardinal templates (Demographics, AE Overview, AE by SOC/PT,
Lab Shift, Serious-AE listings, subgroup forest plots), showing the recurring
ARD-first patterns you will see across the catalog and the customizations a
sponsor typically applies. After Part 2 we move to **Lesson 32 — tfrmt**, the
display metadata layer for complex multi-dimensional tables.

---

## Self-check questions

1. Why is cardinal described as a *catalog* rather than an R package?
2. Which industry guide is the primary source for cardinal's safety table templates?
3. Name three packages in the Cardinal-future stack and what each does.
4. What does "ARD-first" mean, and why does it matter for CDISC ARS?
5. What is the difference between cardinal and Roche's NEST stack?
6. How would you contribute a new template to the cardinal catalog?

## Glossary

- **cardinal** — Industry harmonized catalog of TLG templates (formerly `falcon`).
- **TLG** — Tables, Listings, and Graphs.
- **ARD** — Analysis Results Dataset; tidy long table where one row = one statistic.
- **ARD-first** — Build the ARD first, then derive the display from it.
- **FDA Standard Safety Tables and Figures** — FDA guide documenting canonical
  safety table layouts; cardinal's primary source.
- **Cardinal-future stack** — cards + cardx + gtsummary + tfrmt + crane + ggplot.
- **cards / cardx** — Packages that compute ARDs.
- **gtsummary** — Renders an ARD into a formatted table (`tbl_ard_summary()`).
- **tfrmt** — Display metadata layer for complex/multi-dimensional tables (Lesson 32).
- **crane** — Roche extension for pharma-specific gtsummary patterns.
- **NEST** — Roche-led legacy TLG stack (rtables, tern, chevron).
- **CDISC ARS** — Analysis Results Standard; emerging CDISC framework for ARD-based reporting.
- **Quarto** — Documentation/publishing framework; cardinal templates are Quarto documents.
- **Catalog** — The browsable collection of templates on the cardinal website.
- **Template** — A single `.qmd` file implementing one TLG; copyable into your study.
