---
title: "Lesson 30 — tfrmt Part 2: page_plan(), Clinical Templates, and RTF Output"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build on tfrmt foundations from Lesson 29: paginate long AE tables with
  page_plan() and page_structure(), organise SOC > PT hierarchies with
  row_grp_plan(), use the tfrmt_n_pct() shortcut, add column spanners with
  spanning_structure(), build clinical table templates, and produce a
  complete submission-style Table 14.3.1 AE table with RTF export via as_rtf().
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 30_tfrmt_part2.Rmd
# PURPOSE       : tfrmt Part 2 — page_plan(), row_grp_plan(), tfrmt_n_pct(),
#                 spanning_structure(), clinical table templates, mock_df()
#                 shell sign-off, and RTF output via as_rtf().
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 161, 7)), round(rnorm(n, 174, 8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 68, 12), 1), round(rnorm(n, 82, 14), 1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL = ht, WEIGHTBL = wt,
                                  BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

advs <- pharmaverseadam::advs

# ══════════════════════════════════════════════════════════════════════════════
# LESSON 29 MOCKS — frmt, frmt_combine, frmt_structure, body_plan,
#                   col_plan, tfrmt, print_to_gt, mock_df, ard_to_tfrmt
# ══════════════════════════════════════════════════════════════════════════════

# frmt: defines how a single number is printed
frmt <- function(expression = "xx", missing = "", ...) {
  structure(list(expression = expression, missing = missing, extra = list(...)),
            class = "frmt")
}

print.frmt <- function(x, ...) {
  cat(sprintf("<frmt> expression=\"%s\"  missing=\"%s\"\n",
              x$expression, x$missing))
  invisible(x)
}

# frmt_combine: combines multiple frmts into a composite cell (e.g., "n (%)")
frmt_combine <- function(expression, ..., missing = "") {
  parts <- list(...)
  structure(list(expression = expression, parts = parts, missing = missing),
            class = "frmt_combine")
}

print.frmt_combine <- function(x, ...) {
  cat(sprintf("<frmt_combine> expression=\"%s\"\n", x$expression))
  invisible(x)
}

# frmt_structure: pairs a frmt with param and group/label selectors
frmt_structure <- function(group_val = NULL, label_val = NULL,
                            frmt_combine = NULL, ...) {
  frmts <- c(list(frmt_combine), list(...))
  frmts <- frmts[!sapply(frmts, is.null)]
  structure(list(group_val = group_val, label_val = label_val, frmts = frmts),
            class = "frmt_structure")
}

# body_plan: collects frmt_structures that govern row display
body_plan <- function(...) {
  structures <- list(...)
  structure(list(structures = structures), class = "body_plan")
}

# col_plan: defines column ordering / renaming
col_plan <- function(...) {
  args <- list(...)
  structure(list(columns = args), class = "col_plan")
}

# tfrmt: the master table spec object
tfrmt <- function(label = NULL, param = NULL, value = NULL,
                  column = NULL, group = NULL,
                  body_plan = NULL, col_plan = NULL,
                  row_grp_plan = NULL, page_plan = NULL, ...) {
  # substitute() called directly in tfrmt (not via a helper) so bare unquoted
  # column names work: tfrmt(param = param, ...) or tfrmt(param = "col", ...)
  .s2c <- function(s) {
    if (is.null(s)) return(NULL)
    if (is.character(s)) return(s)
    deparse(s)
  }
  label_v  <- .s2c(substitute(label))
  param_v  <- .s2c(substitute(param))
  value_v  <- .s2c(substitute(value))
  column_v <- .s2c(substitute(column))
  group_v  <- .s2c(substitute(group))
  structure(
    list(
      label        = label_v,
      param        = param_v,
      value        = value_v,
      column       = column_v,
      group        = group_v,
      body_plan    = body_plan,
      col_plan     = col_plan,
      row_grp_plan = row_grp_plan,
      page_plan    = page_plan,
      extra        = list(...)
    ),
    class = "tfrmt"
  )
}

print.tfrmt <- function(x, ...) {
  cat("<tfrmt spec>\n")
  cat(sprintf("  label  : %s\n", x$label %||% "(not set)"))
  cat(sprintf("  param  : %s\n", x$param %||% "(not set)"))
  cat(sprintf("  value  : %s\n", x$value %||% "(not set)"))
  cat(sprintf("  column : %s\n", x$column %||% "(not set)"))
  cat(sprintf("  group  : %s\n", x$group %||% "(not set)"))
  if (!is.null(x$body_plan))
    cat(sprintf("  body_plan   : %d frmt_structure(s)\n",
                length(x$body_plan$structures)))
  if (!is.null(x$col_plan))
    cat(sprintf("  col_plan    : %d column(s)\n", length(x$col_plan$columns)))
  if (!is.null(x$row_grp_plan))
    cat(sprintf("  row_grp_plan: label_loc = %s\n", x$row_grp_plan$label_loc))
  if (!is.null(x$page_plan))
    cat(sprintf("  page_plan   : %d page_structure(s), note_loc = %s\n",
                length(x$page_plan$page_structures), x$page_plan$note_loc))
  invisible(x)
}

`%||%` <- function(a, b) if (!is.null(a)) a else b

# print_to_gt: renders tfrmt spec + data as a display table (returns tibble here)
print_to_gt <- function(tfrmt_spec, .data, ...) {
  cat("[print_to_gt()] Rendering tfrmt spec against", nrow(.data), "ARD rows\n")
  out <- .data |>
    select(any_of(c("label", "label_level", "column", "param", "value"))) |>
    head(20)
  cat("[Output preview — first 20 rows of display-ready data]\n")
  print(out)
  invisible(out)
}

# mock_df: produces a blank shell table for spec review
mock_df <- function(tfrmt_spec, .data = NULL, ...) {
  cat("[mock_df()] Generating blank shell for tfrmt spec\n")
  cat("  label     :", tfrmt_spec$label %||% "(not set)", "\n")
  cat("  param     :", tfrmt_spec$param %||% "(not set)", "\n")
  cat("  column    :", tfrmt_spec$column %||% "(not set)", "\n")
  if (!is.null(.data))
    cat("  ARD rows  :", nrow(.data), "\n")
  cat("[Shell would display here — all data cells show format expressions]\n")
  invisible(tfrmt_spec)
}

# ard_to_tfrmt: coerces an ARD tibble into the tfrmt long format
ard_to_tfrmt <- function(ard, label_col = "variable",
                          label_level_col = "variable_level",
                          column_col = "group1_level",
                          param_col  = "stat_name",
                          value_col  = "stat") {
  ard |>
    mutate(value = unlist(!!sym(value_col))) |>
    rename(
      label       = !!sym(label_col),
      label_level = !!sym(label_level_col),
      column      = !!sym(column_col),
      param       = !!sym(param_col)
    ) |>
    select(label, label_level, column, param, value) |>
    filter(!is.na(column))
}

# ══════════════════════════════════════════════════════════════════════════════
# CARDS MOCKS — ard_continuous, ard_categorical, ard_missing, ard_stack
# ══════════════════════════════════════════════════════════════════════════════

.ard_row <- function(group1 = NA_character_, group1_level = NA_character_,
                     variable, variable_level = NA_character_,
                     context, stat_name, stat, stat_label) {
  tibble(
    group1         = group1,
    group1_level   = group1_level,
    variable       = variable,
    variable_level = variable_level,
    context        = context,
    stat_name      = stat_name,
    stat           = list(stat),
    stat_label     = stat_label
  )
}

ard_continuous <- function(data, variables, by = NULL,
                           statistics = list(
                             N      = function(x) sum(!is.na(x)),
                             mean   = function(x) mean(x, na.rm = TRUE),
                             sd     = function(x) sd(x, na.rm = TRUE),
                             median = function(x) median(x, na.rm = TRUE),
                             min    = function(x) min(x, na.rm = TRUE),
                             max    = function(x) max(x, na.rm = TRUE),
                             n_miss = function(x) sum(is.na(x)),
                             p_miss = function(x) round(100 * mean(is.na(x)), 1)
                           )) {
  stat_labels <- c(N = "N (non-missing)", mean = "Mean", sd = "SD",
                   median = "Median", min = "Min", max = "Max",
                   n_miss = "N missing", p_miss = "% missing")
  groups <- if (!is.null(by)) unique(as.character(data[[by]])) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[as.character(data[[by]]) == gl, ] else data
      x   <- sub[[vv]]
      lapply(names(statistics), function(sn) {
        val <- tryCatch(statistics[[sn]](x), error = function(e) NA_real_)
        .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                 group1_level = gl, variable = vv, context = "continuous",
                 stat_name = sn, stat = val, stat_label = stat_labels[sn])
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class = c("card", "tbl_df", "tbl", "data.frame"))
}

ard_categorical <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(as.character(data[[by]])) else NA_character_
  rows <- lapply(variables, function(vv) {
    levs <- sort(unique(as.character(data[[vv]])))
    levs <- levs[!is.na(levs)]
    lapply(groups, function(gl) {
      sub   <- if (!is.na(gl) && !is.null(by)) data[as.character(data[[by]]) == gl, ] else data
      total <- sum(!is.na(sub[[vv]]))
      lapply(levs, function(lv) {
        cnt <- sum(sub[[vv]] == lv, na.rm = TRUE)
        bind_rows(
          .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                   group1_level = gl, variable = vv, variable_level = lv,
                   context = "categorical", stat_name = "n", stat = cnt, stat_label = "n"),
          .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                   group1_level = gl, variable = vv, variable_level = lv,
                   context = "categorical", stat_name = "p",
                   stat = round(100 * cnt / max(total, 1), 1), stat_label = "%")
        )
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class = c("card", "tbl_df", "tbl", "data.frame"))
}

ard_missing <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(as.character(data[[by]])) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[as.character(data[[by]]) == gl, ] else data
      x   <- sub[[vv]]
      bind_rows(
        .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                 group1_level = gl, variable = vv, context = "missing",
                 stat_name = "n_miss", stat = sum(is.na(x)), stat_label = "N missing"),
        .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                 group1_level = gl, variable = vv, context = "missing",
                 stat_name = "p_miss", stat = round(100 * mean(is.na(x)), 1),
                 stat_label = "% missing")
      )
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class = c("card", "tbl_df", "tbl", "data.frame"))
}

ard_stack <- function(...) {
  out <- bind_rows(...)
  structure(out, class = c("card", "tbl_df", "tbl", "data.frame"))
}

bind_ard <- function(...) ard_stack(...)

# ══════════════════════════════════════════════════════════════════════════════
# LESSON 30 ADDITIONAL MOCKS
# ══════════════════════════════════════════════════════════════════════════════

# page_plan: defines pagination for long tables (e.g., AE tables)
page_plan <- function(..., note_loc = c("above", "below")) {
  note_loc <- match.arg(note_loc)
  structure(list(page_structures = list(...), note_loc = note_loc),
            class = "page_plan")
}

print.page_plan <- function(x, ...) {
  cat(sprintf("<page_plan> %d page_structure(s), note_loc = \"%s\"\n",
              length(x$page_structures), x$note_loc))
  invisible(x)
}

page_structure <- function(group_val = NULL, ...) {
  structure(list(group_val = group_val, extra = list(...)),
            class = "page_structure")
}

print.page_structure <- function(x, ...) {
  cat(sprintf("<page_structure> group_val = %s\n",
              x$group_val %||% "(all rows on one page)"))
  invisible(x)
}

# row_grp_plan: grouping for row-level hierarchies (e.g., SOC > PT)
row_grp_plan <- function(..., label_loc = c("indented", "column")) {
  label_loc <- match.arg(label_loc)
  structure(list(structures = list(...), label_loc = label_loc),
            class = "row_grp_plan")
}

print.row_grp_plan <- function(x, ...) {
  cat(sprintf("<row_grp_plan> %d structure(s), label_loc = \"%s\"\n",
              length(x$structures), x$label_loc))
  invisible(x)
}

row_grp_structure <- function(group_val = NULL, label_val = NULL,
                               indent = c("indented", "column"), ...) {
  structure(list(group_val = group_val, label_val = label_val, indent = indent[1]),
            class = "row_grp_structure")
}

print.row_grp_structure <- function(x, ...) {
  cat(sprintf("<row_grp_structure> group_val = %s  label_val = %s  indent = %s\n",
              x$group_val %||% "(none)", x$label_val %||% "(none)", x$indent))
  invisible(x)
}

# spanning_structure: add a spanner header above columns
spanning_structure <- function(...) {
  structure(list(...), class = "spanning_structure")
}

print.spanning_structure <- function(x, ...) {
  lbls <- names(x)
  cat(sprintf("<spanning_structure> %d spanner(s): %s\n",
              length(x), paste(lbls, collapse = ", ")))
  invisible(x)
}

# tfrmt_n_pct: shortcut frmt_combine for n (%)
tfrmt_n_pct <- function(n_frmt   = frmt("xxx"),
                         pct_frmt = frmt("xx.x"),
                         missing  = "0 (---%)") {
  frmt_combine("{n} ({p}%)", n = n_frmt, p = pct_frmt, missing = missing)
}

# as_rtf (mock): export gt object to RTF file
as_rtf <- function(x, filename, ...) {
  cat(sprintf("[as_rtf()] Would write RTF to: %s\n", filename))
  invisible(x)
}

# make_gt: alias for print_to_gt
make_gt <- function(tfrmt_spec, .data, ...) {
  print_to_gt(tfrmt_spec, .data, ...)
}

# ══════════════════════════════════════════════════════════════════════════════
# PRE-BUILT AE ARD FOR EXERCISES
# ══════════════════════════════════════════════════════════════════════════════

adsl_trt <- adsl |> select(USUBJID, TRT01P)

# Subject-level AE SOC counts (one row per subject × SOC, avoiding double-counting)
ae_soc_subj <- adae |>
  distinct(USUBJID, AESOC) |>
  left_join(adsl_trt, by = "USUBJID")

# Subject-level AE PT counts
ae_pt_subj <- adae |>
  distinct(USUBJID, AESOC, AEDECOD) |>
  left_join(adsl_trt, by = "USUBJID")

# Combined ARD: AESOC categorical + continuous AGE/WEIGHTBL for demographics
ae_ard <- ard_stack(
  ard_categorical(ae_soc_subj, variables = "AESOC", by = "TRT01P"),
  ard_continuous(
    adsl,
    variables = c("AGE", "WEIGHTBL"),
    by = "TRT01P"
  )
)

# tfrmt-ready long format
ae_tfrmt_ready <- ae_ard |>
  mutate(value = unlist(stat)) |>
  select(
    label       = variable,
    label_level = variable_level,
    column      = group1_level,
    param       = stat_name,
    value
  ) |>
  filter(!is.na(column))

# PT-level ARD for capstone
ae_pt_ard <- ard_categorical(ae_pt_subj, variables = "AEDECOD", by = "TRT01P")

ae_pt_tfrmt_ready <- ae_pt_ard |>
  mutate(value = unlist(stat)) |>
  select(
    label       = variable,
    label_level = variable_level,
    column      = group1_level,
    param       = stat_name,
    value
  ) |>
  filter(!is.na(column))
```

## Overview

Lesson 29 introduced the core tfrmt building blocks: `frmt()`, `frmt_combine()`,
`frmt_structure()`, `body_plan()`, `col_plan()`, and `tfrmt()`. In this lesson
you go further — adding the structures clinical programmers need for real
submission tables:

| Feature | Function | Purpose |
|---------|----------|---------|
| Row hierarchy | `row_grp_plan()` + `row_grp_structure()` | SOC header rows → indented PT rows |
| Pagination | `page_plan()` + `page_structure()` | New page per SOC for long AE tables |
| n (%) shortcut | `tfrmt_n_pct()` | Pre-built `frmt_combine` for the most common cell format |
| Column spanners | `spanning_structure()` | Group visit columns under a shared label |
| Shell sign-off | `mock_df()` | Blank shell for biostatistician review |
| RTF export | `as_rtf()` | Write the rendered gt table to a Word-importable RTF file |

All exercises use the **CDISCPILOT01** Alzheimer's study from `pharmaverseadam`.

### The tfrmt pipeline

```
ARD (cards)  →  tfrmt spec  →  print_to_gt()  →  gt table  →  as_rtf()
                                  ↑
                         mock_df() for shell review
```

---

## Section 1 — row_grp_plan() and row_grp_structure()

In an AE summary table the SOC (System Organ Class) is a **group header row**
and the PT (Preferred Term) rows are **indented beneath it**. The `row_grp_plan()`
argument of `tfrmt()` controls this hierarchy.

```
NERVOUS SYSTEM DISORDERS          ← SOC header (group row)
  Headache         12  14  11     ← PT row, indented
  Dizziness         7   9   6
GASTROINTESTINAL DISORDERS
  Nausea           10  12  15
  ...
```

**Key arguments:**

- `row_grp_plan(row_grp_structure(...), ..., label_loc = "indented")` — outer
  wrapper; `label_loc = "indented"` keeps group labels in the same column,
  indented; `"column"` puts them in a separate first column.
- `row_grp_structure(group_val, label_val, indent)` — one structure per
  hierarchy level. `group_val` names the ARD column that holds the group
  identifier; `label_val` names the column that holds the display label.

```{r ex-rgp-build, exercise=TRUE, exercise.lines=14}
# Build a row_grp_plan for a SOC > PT hierarchy.
# group_val = "AESOC"  (system organ class — the group header)
# label_val = "AEDECOD"  (preferred term — the indented row label)
# Use indent = "indented" and label_loc = "indented".

rgp <- row_grp_plan(
  row_grp_structure(
    group_val = ___,
    label_val = ___,
    indent    = "indented"
  ),
  label_loc = ___
)

print(rgp)
```

```{r ex-rgp-build-hint-1}
# group_val = "AESOC", label_val = "AEDECOD", label_loc = "indented"
```

```{r ex-rgp-build-solution}
rgp <- row_grp_plan(
  row_grp_structure(
    group_val = "AESOC",
    label_val = "AEDECOD",
    indent    = "indented"
  ),
  label_loc = "indented"
)

print(rgp)
```

```{r ex-rgp-inspect, exercise=TRUE, exercise.lines=10}
# Inspect the structure of a row_grp_plan object.
# Build the same spec as above, then use str() to see the nested list.

rgp <- row_grp_plan(
  row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                    indent = "indented"),
  label_loc = "indented"
)

str(___)
```

```{r ex-rgp-inspect-hint-1}
# str(rgp)
```

```{r ex-rgp-inspect-solution}
rgp <- row_grp_plan(
  row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                    indent = "indented"),
  label_loc = "indented"
)

str(rgp)
```

```{r quiz-rgp, echo=FALSE}
quiz(
  question(
    "What does `label_loc = \"indented\"` do in `row_grp_plan()`?",
    answer("Puts group labels in a separate leftmost column"),
    answer("Indents group labels within the existing label column, keeping a single label column",
           correct = TRUE),
    answer("Applies bold formatting to group header rows"),
    answer("Hides the group label entirely"),
    allow_retry = TRUE
  ),
  question(
    "In a SOC > PT hierarchy, which argument holds the display text for indented PT rows?",
    answer("group_val", message = "group_val identifies the grouping variable (SOC level)."),
    answer("label_val", correct = TRUE,
           message = "label_val specifies the column whose values become the indented row labels (here AEDECOD)."),
    answer("indent"),
    answer("label_loc"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — page_plan() and page_structure()

AE tables can run to many pages. `page_plan()` tells tfrmt where to insert page
breaks so each SOC section starts on a new page, and `note_loc` controls whether
footnotes appear above or below each page.

**Anatomy:**
```r
page_plan(
  page_structure(group_val = "AESOC"),
  note_loc = "below"
)
```

Each `page_structure()` identifies a grouping value; tfrmt inserts a page break
whenever that group changes.

```{r ex-pp-build, exercise=TRUE, exercise.lines=14}
# Build a page_plan that breaks on AESOC and places notes below each page.

pp <- page_plan(
  page_structure(group_val = ___),
  note_loc = ___
)

print(pp)
```

```{r ex-pp-build-hint-1}
# group_val = "AESOC", note_loc = "below"
```

```{r ex-pp-build-solution}
pp <- page_plan(
  page_structure(group_val = "AESOC"),
  note_loc = "below"
)

print(pp)
```

```{r ex-pp-tfrmt, exercise=TRUE, exercise.lines=22}
# Combine page_plan with a tfrmt spec and call mock_df().
# Use ae_tfrmt_ready as the data argument.

pp <- page_plan(
  page_structure(group_val = "AESOC"),
  note_loc = "below"
)

spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default",
                   frmt_combine("{n} ({p}%)",
                                n = frmt("xxx"),
                                p = frmt("xx.x")))
  ),
  page_plan = ___
)

mock_df(spec, ae_tfrmt_ready)
```

```{r ex-pp-tfrmt-hint-1}
# page_plan = pp
```

```{r ex-pp-tfrmt-solution}
pp <- page_plan(
  page_structure(group_val = "AESOC"),
  note_loc = "below"
)

spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default",
                   frmt_combine("{n} ({p}%)",
                                n = frmt("xxx"),
                                p = frmt("xx.x")))
  ),
  page_plan = pp
)

mock_df(spec, ae_tfrmt_ready)
```

---

## Section 3 — tfrmt_n_pct(): the n (%) shortcut

Nearly every count cell in a clinical table displays as **n (x.x%)**. Writing
the `frmt_combine("{n} ({p}%)", n = frmt("xxx"), p = frmt("xx.x"))` expression
by hand is verbose. `tfrmt_n_pct()` is a convenience wrapper that produces the
same object with sensible defaults.

```{r ex-npct-compare, exercise=TRUE, exercise.lines=16}
# Compare tfrmt_n_pct() to the manually written frmt_combine.
# Both should produce identical frmt_combine objects.

manual <- frmt_combine(
  "{n} ({p}%)",
  n = frmt("xxx"),
  p = frmt("xx.x"),
  missing = "0 (---%)"
)

shortcut <- tfrmt_n_pct()

# Print both — are they the same expression?
print(manual)
print(shortcut)
identical(manual$expression, shortcut$expression)
```

```{r ex-npct-compare-hint-1}
# Both should print the same expression: "{n} ({p}%)"
# identical() returns TRUE if the expressions match.
```

```{r ex-npct-compare-solution}
manual <- frmt_combine(
  "{n} ({p}%)",
  n = frmt("xxx"),
  p = frmt("xx.x"),
  missing = "0 (---%)"
)

shortcut <- tfrmt_n_pct()

print(manual)
print(shortcut)
identical(manual$expression, shortcut$expression)
```

```{r ex-npct-customise, exercise=TRUE, exercise.lines=14}
# tfrmt_n_pct() accepts custom arguments.
# Create a version that formats n as "xxxx" (4-digit counts)
# and percentage as "xx.xx" (2 decimal places).
# Assign it to npct_wide and print it.

npct_wide <- tfrmt_n_pct(
  n_frmt   = frmt(___),
  pct_frmt = frmt(___)
)

print(npct_wide)
```

```{r ex-npct-customise-hint-1}
# n_frmt = frmt("xxxx"), pct_frmt = frmt("xx.xx")
```

```{r ex-npct-customise-solution}
npct_wide <- tfrmt_n_pct(
  n_frmt   = frmt("xxxx"),
  pct_frmt = frmt("xx.xx")
)

print(npct_wide)
```

```{r quiz-npct, echo=FALSE}
quiz(
  question(
    "What does the `missing` argument in `tfrmt_n_pct()` control?",
    answer("The label to display when a row has no subjects",
           correct = TRUE,
           message = "missing = \"0 (---%)\" is the default — displayed when n = 0 or the statistic is NA."),
    answer("Whether to omit rows entirely when n = 0"),
    answer("The placeholder text shown in mock_df() shells"),
    answer("The number of decimal places for missing-data imputations"),
    allow_retry = TRUE
  )
)
```

---

## Section 4 — spanning_structure(): Column Spanners

A **column spanner** is a merged header cell that sits above a group of columns.
In a lab table with visits Week 2, Week 4, Week 8 under each treatment arm, you
would use spanners like `"Placebo"` spanning `c("W2", "W4", "W8")`.

`spanning_structure()` is passed to the `col_plan()` of a `tfrmt()` spec. You
name the spanner label and give a vector of column names that fall beneath it.

```{r ex-span-build, exercise=TRUE, exercise.lines=20}
# Build a tfrmt spec for a hypothetical 3-visit lab table.
# Visits (columns): "W02", "W04", "W08" for each of 2 arms.
# Add a column spanner "Treatment Visits" above all three visit columns.

# First, build a spanning_structure
ss <- spanning_structure(
  "Treatment Visits" = c("W02", "W04", "W08")
)

print(ss)

# Now embed it in a tfrmt col_plan
spec_span <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  col_plan  = col_plan(___),
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xx.x"))
  )
)

print(spec_span)
```

```{r ex-span-build-hint-1}
# col_plan(spanning_structure("Treatment Visits" = c("W02","W04","W08")))
```

```{r ex-span-build-solution}
ss <- spanning_structure(
  "Treatment Visits" = c("W02", "W04", "W08")
)

print(ss)

spec_span <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  col_plan  = col_plan(spanning_structure("Treatment Visits" = c("W02", "W04", "W08"))),
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xx.x"))
  )
)

print(spec_span)
```

```{r ex-span-multi, exercise=TRUE, exercise.lines=20}
# Build a spec with TWO spanners: one for the active arm visits and one
# for the placebo arm visits.
# Active visits: "XLD_W02", "XLD_W04"
# Placebo visits: "PBO_W02", "PBO_W04"

spec_two_span <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  col_plan  = col_plan(
    spanning_structure(___ = c("XLD_W02", "XLD_W04")),
    spanning_structure(___ = c("PBO_W02", "PBO_W04"))
  ),
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xx.x"))
  )
)

print(spec_two_span)
```

```{r ex-span-multi-hint-1}
# Name each spanning_structure with the spanner label string:
# spanning_structure("Xanomeline Low Dose" = c("XLD_W02","XLD_W04"))
# spanning_structure("Placebo" = c("PBO_W02","PBO_W04"))
```

```{r ex-span-multi-solution}
spec_two_span <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  col_plan  = col_plan(
    spanning_structure("Xanomeline Low Dose" = c("XLD_W02", "XLD_W04")),
    spanning_structure("Placebo" = c("PBO_W02", "PBO_W04"))
  ),
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xx.x"))
  )
)

print(spec_two_span)
```

---

## Section 5 — Clinical Table Templates

Real submissions follow recurring patterns. Rather than building every spec from
scratch, clinical programmers maintain **template specs** that are parameterised
and reused across studies. Here you build three standard templates:

1. **Demographics table** — continuous mean (SD) and categorical n (%) rows
2. **AE summary table** — SOC-level n (%) with row grouping
3. **Lab shift table** — baseline → worst-post CTCAE grade counts

```{r ex-tmpl-demog, exercise=TRUE, exercise.lines=28}
# Build a demographics tfrmt template.
# Rows: AGE (mean/SD), SEX (n/%), RACE (n/%)
# Columns: treatment arms from TRT01P
# Use two frmt_structures:
#   label_val = "AGE" → frmt_combine("{m} ({s})", m=frmt("xx.x"), s=frmt("xx.xx"))
#   label_val = ".default" → tfrmt_n_pct()

demog_spec <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(
      label_val = ___,
      frmt_combine("{m} ({s})",
                   m = frmt("xx.x"),
                   s = frmt("xx.xx"))
    ),
    frmt_structure(
      label_val = ___,
      ___
    )
  )
)

print(demog_spec)
```

```{r ex-tmpl-demog-hint-1}
# label_val = "AGE" for the mean(SD) format
# label_val = ".default" with tfrmt_n_pct() for all other rows
```

```{r ex-tmpl-demog-solution}
demog_spec <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(
      label_val = "AGE",
      frmt_combine("{m} ({s})",
                   m = frmt("xx.x"),
                   s = frmt("xx.xx"))
    ),
    frmt_structure(
      label_val = ".default",
      tfrmt_n_pct()
    )
  )
)

print(demog_spec)
```

```{r ex-tmpl-ae, exercise=TRUE, exercise.lines=28}
# Build an AE summary tfrmt spec using ae_tfrmt_ready.
# All cells use tfrmt_n_pct().
# Add a row_grp_plan with SOC grouping and indented labels.
# Then call print_to_gt() to render it.

ae_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(
      group_val = ___,
      label_val = ___,
      indent    = "indented"
    ),
    label_loc = "indented"
  )
)

print_to_gt(ae_spec, ___)
```

```{r ex-tmpl-ae-hint-1}
# group_val = "AESOC", label_val = "AEDECOD"
# print_to_gt(ae_spec, ae_tfrmt_ready)
```

```{r ex-tmpl-ae-solution}
ae_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(
      group_val = "AESOC",
      label_val = "AEDECOD",
      indent    = "indented"
    ),
    label_loc = "indented"
  )
)

print_to_gt(ae_spec, ae_tfrmt_ready)
```

```{r ex-tmpl-lab, exercise=TRUE, exercise.lines=28}
# Build a lab-shift tfrmt spec for ATOXGR grade transitions.
# Use adlb to compute baseline (BTOXGR) × worst post-baseline (ATOXGR) counts,
# then define a spec where:
#   label = "BTOXGR_label"  (baseline grade)
#   column = "ATOXGR_label" (post-baseline grade)
#   param = "n"
#   frmt_structure: label_val = ".default" → frmt("xxx")

# Step 1: build a simple shift count ARD
shift_counts <- adlb |>
  filter(!is.na(BTOXGR), !is.na(ATOXGR), PARAMCD == "ALT") |>
  mutate(
    BTOXGR_label = paste0("Baseline Grade ", BTOXGR),
    ATOXGR_label = paste0("Post Grade ",     ATOXGR)
  ) |>
  group_by(BTOXGR_label, ATOXGR_label) |>
  summarise(value = n_distinct(USUBJID), .groups = "drop") |>
  mutate(param = "n", label = BTOXGR_label, column = ATOXGR_label) |>
  select(label, column, param, value)

# Step 2: build the tfrmt spec
shift_spec <- tfrmt(
  label     = ___,
  param     = ___,
  value     = ___,
  column    = ___,
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xxx"))
  )
)

print_to_gt(shift_spec, shift_counts)
```

```{r ex-tmpl-lab-hint-1}
# label = "label", param = "param", value = "value", column = "column"
```

```{r ex-tmpl-lab-solution}
shift_counts <- adlb |>
  filter(!is.na(BTOXGR), !is.na(ATOXGR), PARAMCD == "ALT") |>
  mutate(
    BTOXGR_label = paste0("Baseline Grade ", BTOXGR),
    ATOXGR_label = paste0("Post Grade ",     ATOXGR)
  ) |>
  group_by(BTOXGR_label, ATOXGR_label) |>
  summarise(value = n_distinct(USUBJID), .groups = "drop") |>
  mutate(param = "n", label = BTOXGR_label, column = ATOXGR_label) |>
  select(label, column, param, value)

shift_spec <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xxx"))
  )
)

print_to_gt(shift_spec, shift_counts)
```

---

## Section 6 — mock_df() for Shell Sign-Off

Before real data is available, **biostatisticians and programmers review table
shells** — blank tables that show structure, column headers, row labels, and
format expressions (e.g., `xx.x (xx.xx)`) without actual numbers.

`mock_df()` generates this shell from a `tfrmt` spec and (optionally) a sample
ARD that provides the row/column structure.

In the TLF review workflow:

1. Programmer builds the `tfrmt` spec and calls `mock_df()`.
2. Output (shell PDF or RTF) is attached to a **Table Specification Document**.
3. Biostatistician reviews row labels, column headers, and format expressions.
4. Sign-off triggers programming of the final table.

```{r ex-mock-shell, exercise=TRUE, exercise.lines=22}
# Build an AE tfrmt spec and call mock_df() to produce a blank shell.
# Pass ae_tfrmt_ready so mock_df() knows the row and column structure.

ae_tfrmt_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                      indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

mock_df(___, ___)
```

```{r ex-mock-shell-hint-1}
# mock_df(ae_tfrmt_spec, ae_tfrmt_ready)
```

```{r ex-mock-shell-solution}
ae_tfrmt_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                      indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

mock_df(ae_tfrmt_spec, ae_tfrmt_ready)
```

```{r ex-mock-review, exercise=TRUE, exercise.lines=14}
# When a biostatistician reviews a shell, they check three things:
# 1. Are all expected row labels present?
# 2. Are column headers correct (treatment labels, N= counts)?
# 3. Do the format expressions match the SAP (e.g., "xx.x (xx.xx)" for mean(SD))?
#
# Inspect ae_tfrmt_ready to answer: how many unique label_level values exist?
# (These become the row labels in the rendered table.)

ae_tfrmt_ready |>
  ___(label_level) |>
  nrow()
```

```{r ex-mock-review-hint-1}
# distinct(label_level) then nrow()
```

```{r ex-mock-review-solution}
ae_tfrmt_ready |>
  distinct(label_level) |>
  nrow()
```

---

## Section 7 — Capstone: Complete AE Table (Table 14.3.1 Style)

Bring together everything from this lesson and Lesson 29 to build a
**submission-ready AE summary table**:

- **Body:** `n (%)` cells for each treatment arm
- **Row hierarchy:** SOC header rows → indented PT sub-rows
- **Pagination:** new page per SOC
- **Rendering:** `print_to_gt()` to display the table
- **Export:** `as_rtf("Table_14_3_1.rtf")` to write the file

```{r ex-cap-ard, exercise=TRUE, exercise.lines=26}
# Step 1: Inspect ae_pt_tfrmt_ready — the PT-level ARD already built in setup.
# How many unique label_level values (PTs) are there?
# How many unique column values (treatment arms) are there?

ae_pt_tfrmt_ready |> distinct(label_level) |> nrow()
ae_pt_tfrmt_ready |> distinct(___) |> nrow()

# Also confirm that param takes values "n" and "p"
ae_pt_tfrmt_ready |> distinct(param)
```

```{r ex-cap-ard-hint-1}
# distinct(column) for the treatment arm count
```

```{r ex-cap-ard-solution}
ae_pt_tfrmt_ready |> distinct(label_level) |> nrow()
ae_pt_tfrmt_ready |> distinct(column) |> nrow()

ae_pt_tfrmt_ready |> distinct(param)
```

```{r ex-cap-spec, exercise=TRUE, exercise.lines=38}
# Step 2: Build the complete AE tfrmt spec.
# - body_plan: tfrmt_n_pct() for all rows
# - row_grp_plan: SOC (group_val) > PT (label_val), indented
# - page_plan: break on AESOC, notes below

ae_full_spec <- tfrmt(
  label     = "label_level",
  param     = ___,
  value     = ___,
  column    = ___,
  body_plan = body_plan(
    frmt_structure(label_val = ".default", ___)
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(
      group_val = ___,
      label_val = ___,
      indent    = "indented"
    ),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = ___),
    note_loc = "below"
  )
)

print(ae_full_spec)
```

```{r ex-cap-spec-hint-1}
# param = "param", value = "value", column = "column"
# tfrmt_n_pct() in the frmt_structure
# group_val = "AESOC", label_val = "AEDECOD"
# page_structure group_val = "AESOC"
```

```{r ex-cap-spec-solution}
ae_full_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(
      group_val = "AESOC",
      label_val = "AEDECOD",
      indent    = "indented"
    ),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

print(ae_full_spec)
```

```{r ex-cap-render, exercise=TRUE, exercise.lines=18}
# Step 3: Render the table with print_to_gt() and export with as_rtf().
# Use the spec built in the previous exercise (rebuild it here for completeness).

ae_full_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                      indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

# Render and assign the gt object
gt_out <- print_to_gt(___, ___)

# Export to RTF
as_rtf(gt_out, ___)
```

```{r ex-cap-render-hint-1}
# print_to_gt(ae_full_spec, ae_pt_tfrmt_ready)
# as_rtf(gt_out, "Table_14_3_1.rtf")
```

```{r ex-cap-render-solution}
ae_full_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                      indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

gt_out <- print_to_gt(ae_full_spec, ae_pt_tfrmt_ready)

as_rtf(gt_out, "Table_14_3_1.rtf")
```

---

## Lesson Summary

| Function | Purpose |
|----------|---------|
| `row_grp_plan()` | Define row group hierarchy (SOC header → indented PT rows) |
| `row_grp_structure()` | One level in the hierarchy: which column is the group, which is the label |
| `page_plan()` | Specify pagination rules for multi-page tables |
| `page_structure()` | Identify the grouping variable that triggers page breaks |
| `tfrmt_n_pct()` | Shortcut `frmt_combine` for the standard `n (x.x%)` cell format |
| `spanning_structure()` | Add spanner headers above groups of columns |
| `mock_df()` | Generate a blank shell for biostatistician sign-off |
| `print_to_gt()` | Render tfrmt spec + ARD data into a display table |
| `as_rtf()` | Export the rendered gt table to an RTF file |

**Key workflow for a standard AE table:**

```
1. Build ARD with ard_categorical() on subject-level AE data
2. Coerce to tfrmt long format (label / label_level / column / param / value)
3. Spec: body_plan(tfrmt_n_pct()) + row_grp_plan(SOC > PT) + page_plan(AESOC)
4. mock_df() → biostatistician sign-off
5. print_to_gt() → as_rtf("Table_14_3_1.rtf")
```

**Next lesson:** Lesson 31 — gt — direct table formatting with the gt package:
column labels, spanners, footnotes, conditional styling, and RTF/HTML export.
