---
title: "Lesson 31 — cardinal Part 2: FDA Safety Table Templates"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Walk the most heavily-used cardinal catalog templates implementing the FDA
  Standard Safety Tables and Figures Integrated Guide: demographics (Pattern A),
  AE overview with subject-level flags (Pattern B), AE by SOC and PT via
  ard_hierarchical() (Pattern C), lab reference-range shift tables (Pattern D),
  serious-AE listings (Pattern E), and subgroup forest plots (Pattern F). See how
  every template is the same ARD-first shape, what customizations a sponsor
  applies, and how ~10 templates assemble a complete CSR safety section.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 31_cardinal_part2_fda_safety.Rmd
# PURPOSE       : cardinal Part 2 — FDA Standard Safety Tables and Figures
#                 templates. The six recurring ARD-first patterns (demographics,
#                 AE overview, AE by SOC/PT, lab shift, listings, forest plots),
#                 sponsor customizations, and assembling a full CSR safety
#                 section from ~10 templates.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-06-22
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    DTHFL_N = if_else(!is.na(DTHFL) & DTHFL == "Y", 1L, 0L, missing = 0L),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo","Xanomeline Low Dose","Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
if (!"TRT01P" %in% names(adae_raw))
  adae_raw <- left_join(adae_raw, select(adsl, USUBJID, TRT01P), by = "USUBJID")
# Ensure TRTEMFL exists; pharmaverseadam adae carries it, but synthesise if absent.
if (!"TRTEMFL" %in% names(adae_raw))
  adae_raw <- mutate(adae_raw, TRTEMFL = "Y")
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ══════════════════════════════════════════════════════════════════════════════
# CARDS BASE MOCKS (replicated from lesson 25)
#
# cardinal templates call cards/cardx + gtsummary, which may not be installed in
# this learnr container. The functions below mock the ones the FDA-table patterns
# use, returning a `card`-class tibble in the cards shape (group1 / variable /
# context / stat_name / stat / stat_label). This lets each pattern run standalone
# and lets you inspect the ARD a template produces.
# ══════════════════════════════════════════════════════════════════════════════

.ard_row <- function(group1=NA_character_, group1_level=NA_character_,
                     variable, variable_level=NA_character_,
                     context, stat_name, stat, stat_label) {
  tibble(
    group1         = group1,
    group1_level   = group1_level,
    variable       = variable,
    variable_level = variable_level,
    context        = context,
    stat_name      = stat_name,
    stat           = list(stat),
    stat_label     = stat_label
  )
}

ard_continuous <- function(data, variables, by = NULL,
                           statistics = list(
                             N      = function(x) sum(!is.na(x)),
                             mean   = function(x) mean(x, na.rm=TRUE),
                             sd     = function(x) sd(x, na.rm=TRUE),
                             median = function(x) median(x, na.rm=TRUE),
                             p25    = function(x) quantile(x, 0.25, na.rm=TRUE),
                             p75    = function(x) quantile(x, 0.75, na.rm=TRUE),
                             min    = function(x) min(x, na.rm=TRUE),
                             max    = function(x) max(x, na.rm=TRUE)
                           )) {
  stat_labels <- c(N="N (non-missing)", mean="Mean", sd="SD", median="Median",
                   p25="Q1 (25th pctile)", p75="Q3 (75th pctile)",
                   min="Min", max="Max")
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      x   <- sub[[vv]]
      lapply(names(statistics), function(sn) {
        val <- tryCatch(statistics[[sn]](x), error=function(e) NA_real_)
        .ard_row(group1=if(!is.null(by)) by else NA_character_,
                 group1_level=as.character(gl), variable=vv,
                 context="continuous", stat_name=sn,
                 stat=val, stat_label=stat_labels[sn])
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_categorical <- function(data, variables, by = NULL, denominator = NULL) {
  groups <- if (!is.null(by)) unique(data[[by]]) else NA_character_
  rows <- lapply(variables, function(vv) {
    levs <- sort(unique(as.character(data[[vv]])))
    levs <- levs[!is.na(levs)]
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[data[[by]] == gl, ] else data
      # denominator: optional named vector / data frame providing N per group;
      # otherwise the within-group non-missing count.
      total <- if (!is.null(denominator) && !is.na(gl)) {
        d <- denominator
        if (is.data.frame(d)) sum(d[[by]] == gl, na.rm = TRUE) else as.numeric(d[[as.character(gl)]])
      } else sum(!is.na(sub[[vv]]))
      lapply(levs, function(lv) {
        cnt <- sum(sub[[vv]] == lv, na.rm=TRUE)
        bind_rows(
          .ard_row(group1=if(!is.null(by)) by else NA_character_,
                   group1_level=as.character(gl), variable=vv,
                   variable_level=lv, context="categorical",
                   stat_name="n", stat=cnt, stat_label="n"),
          .ard_row(group1=if(!is.null(by)) by else NA_character_,
                   group1_level=as.character(gl), variable=vv,
                   variable_level=lv, context="categorical",
                   stat_name="p", stat=round(100*cnt/total,1), stat_label="%")
        )
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

ard_stack <- function(...) {
  out <- bind_rows(...)
  structure(out, class=c("card","tbl_df","tbl","data.frame"))
}
bind_ard <- function(...) ard_stack(...)

# Wilson 95% CI for a proportion p_hat based on n observations
.wilson_ci <- function(n, p_hat, conf=0.95) {
  z      <- qnorm(1-(1-conf)/2)
  centre <- (p_hat + z^2/(2*n)) / (1 + z^2/n)
  half   <- (z * sqrt(p_hat*(1-p_hat)/n + z^2/(4*n^2))) / (1 + z^2/n)
  c(lower=round(centre-half,4), upper=round(centre+half,4))
}

ard_proportion_ci <- function(data, variables, by=NULL,
                               conf.level=0.95, method="wilson",
                               value=list()) {
  groups <- if (!is.null(by)) sort(unique(na.omit(data[[by]]))) else NA_character_
  rows <- lapply(variables, function(vv) {
    pos_val <- if (!is.null(value[[vv]])) value[[vv]] else "Y"
    lapply(groups, function(gl) {
      x <- if (!is.na(gl) && !is.null(by)) data[[vv]][data[[by]]==gl] else data[[vv]]
      x <- x[!is.na(x)]
      n  <- length(x); k <- sum(x==pos_val); p_hat <- k/n
      ci <- .wilson_ci(n, p_hat, conf.level)
      bind_rows(
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="proportion_ci", stat_name="n",
               stat=list(k), stat_label=paste0("N (",pos_val,")")),
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="proportion_ci", stat_name="p",
               stat=list(round(100*p_hat,1)), stat_label="%"),
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="proportion_ci", stat_name="conf.low",
               stat=list(round(100*ci["lower"],1)),
               stat_label=paste0(conf.level*100,"% CI lower")),
        tibble(group1=if(!is.null(by)) by else NA_character_,
               group1_level=as.character(gl), variable=vv,
               context="proportion_ci", stat_name="conf.high",
               stat=list(round(100*ci["upper"],1)),
               stat_label=paste0(conf.level*100,"% CI upper"))
      )
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class=c("card","tbl_df","tbl","data.frame"))
}

# Mock ard_hierarchical(): SOC x PT subject-level incidence with a denominator.
ard_hierarchical <- function(data, by, variables, denominator) {
  soc_var <- variables[1]; pt_var <- variables[2]
  groups  <- sort(unique(na.omit(data[[by]])))
  # subject-level: one row per subject per SOC/PT
  su <- distinct(data, USUBJID, !!rlang::sym(by),
                 !!rlang::sym(soc_var), !!rlang::sym(pt_var))
  rows <- lapply(groups, function(gl) {
    n_denom <- sum(denominator[[by]] == gl, na.rm = TRUE)
    sub <- su[su[[by]] == gl, ]
    pts <- distinct(sub, !!rlang::sym(soc_var), !!rlang::sym(pt_var))
    lapply(seq_len(nrow(pts)), function(i) {
      soc <- pts[[soc_var]][i]; pt <- pts[[pt_var]][i]
      k <- sum(sub[[soc_var]] == soc & sub[[pt_var]] == pt, na.rm = TRUE)
      bind_rows(
        tibble(group1 = by, group1_level = as.character(gl),
               variable = soc_var, variable_level = soc,
               context = "hierarchical", stat_name = "soc"),
        tibble(group1 = by, group1_level = as.character(gl),
               variable = pt_var, variable_level = pt,
               context = "hierarchical", stat_name = "n",
               stat = list(k), p = round(100 * k / n_denom, 1))
      )
    }) |> bind_rows() |>
      mutate(soc = soc) -> dummy
    sub_summary <- sub |>
      group_by(.data[[soc_var]], .data[[pt_var]]) |>
      summarise(n = n_distinct(USUBJID), .groups = "drop") |>
      mutate(p = round(100 * n / n_denom, 1),
             group1_level = as.character(gl),
             SOC = .data[[soc_var]], PT = .data[[pt_var]]) |>
      select(group1_level, SOC, PT, n, p)
    sub_summary
  })
  out <- bind_rows(rows)
  structure(out, class = c("card","tbl_df","tbl","data.frame"))
}

# Pre-built subject-level AE-flag dataset, used across exercises (Pattern B).
adae_te <- adae |> filter(TRTEMFL == "Y")
adsl_flags <- adsl |>
  mutate(
    ANY_TEAE     = if_else(USUBJID %in% adae_te$USUBJID, "Y", "N"),
    SERIOUS_TEAE = if_else(
      USUBJID %in% (adae_te |> filter(AESER == "Y") |> pull(USUBJID)), "Y", "N"),
    SEVERE_TEAE  = if_else(
      USUBJID %in% (adae_te |> filter(AESEV == "SEVERE") |> pull(USUBJID)), "Y", "N")
  )
```

## Overview

cardinal's primary source is the **FDA Standard Safety Tables and Figures
Integrated Guide** — a public FDA document (on FDA.gov) describing the canonical
safety tables reviewers expect to see in submissions, with mockups and methodology.

Part 1 covered *what cardinal is* (a Quarto catalog of ARD-first templates). Part
2 covers *what the templates actually do*. Rather than reproduce every template,
we focus on the **six recurring patterns**. Once you can recognize the patterns,
every table in the catalog is readable on sight.

Key principles from the FDA guide:

- Tables use **subject-level denominators** for incidence calculations.
- Severity (e.g., NCI CTCAE grades) is preferred over qualitative descriptors when available.
- **Treatment-emergent** definitions are explicit (`TRTEMFL == "Y"`).
- Footnotes document methodology.
- Multiple display formats are supported (RTF for the CSR, JSON/ARD for ARS).

### The canonical safety table set

| FDA Table # | Subject | Pattern |
|---|---|---|
| 14.1.x | Subject disposition (accountability) | — |
| 14.2.x | Demographics and baseline characteristics | **A** |
| 14.3.x | Exposure (treatment duration, total dose) | A-like |
| 14.4.x | Treatment-emergent AE overview | **B** |
| 14.5.x | AEs by SOC and PT | **C** |
| 14.6.x | Serious AEs (often a listing) | **E** |
| 14.7.x | AEs leading to discontinuation | B-like |
| 14.8.x | Deaths | B-like |
| 14.9.x | Laboratory abnormality / shift summaries | **D** |
| Figures | Subgroup forest plots | **F** |

A "complete" cardinal-driven safety section uses roughly nine templates; each is
about 50–100 lines of R, and the whole section is around 600 lines that produce
~15 RTF tables.

### Lesson map

| Section | Pattern | Template |
|---------|---------|----------|
| 1 | A | Demographics (Table 14.2.x) |
| 2 | B | AE overview (Table 14.4.x) |
| 3 | C | AE by SOC × PT (Table 14.5.x) |
| 4 | D | Lab reference-range shift (Table 14.9.x) |
| 5 | E | Serious-AE listing |
| 6 | F | Subgroup forest plot |
| 7 | — | Assembling a complete CSR safety section + customizations |

---

## Section 1 — Pattern A: Demographics (FDA Table 14.2.x)

The demographics template, paraphrased from cardinal, in real cards/cardx/gtsummary:

```r
adsl <- pharmaverseadam::adsl |> filter(SAFFL == "Y")

demog_ard <- ard_stack(
  adsl,
  ard_continuous(variables = c(AGE, BMIBL, WEIGHTBL, HEIGHTBL)),
  ard_categorical(variables = c(AGEGR1, SEX, RACE, ETHNIC)),
  .by = TRT01A, .overall = TRUE, .total_n = TRUE
)

pvalue_ard <- bind_ard(
  ard_stats_anova_oneway(adsl, by = "TRT01A",
                         variables = c(AGE, BMIBL, WEIGHTBL, HEIGHTBL)),
  ard_stats_chisq_test(adsl, by = "TRT01A",
                       variables = c(AGEGR1, SEX, RACE, ETHNIC))
)

demog_tbl <- bind_ard(demog_ard, pvalue_ard) |>
  tbl_ard_summary(by = "TRT01A",
                  type = c(AGE, BMIBL) ~ "continuous2", overall = TRUE) |>
  add_p() |> bold_labels() |>
  modify_caption("**Table 14.2.1: Demographic and Baseline Characteristics**")
```

**Pattern recognition:**

- ADSL filtered to safety pop.
- ARD built with `ard_stack()` combining continuous and categorical.
- p-values come from `cardx::ard_stats_*()` functions.
- Display via `tbl_ard_summary()` with `continuous2` type.
- Export to RTF.

Every demographics-style template follows this exact shape. The differences across
templates are only *which* variables, *which* statistical test, and *how* the
footnotes are worded.

### Exercise 1-a — Build the demographics descriptive ARD

Reproduce the descriptive half of Pattern A: a continuous ARD (AGE, BMIBL,
WEIGHTBL) stacked on a categorical ARD (SEX, RACE), stratified by treatment arm.

```{r demog-ard, exercise=TRUE, exercise.lines=12}
demog_ard <- ard_stack(
  ard_continuous(adsl,  variables = c("AGE","BMIBL","WEIGHTBL"), by = "___"),
  ard_categorical(adsl, variables = c("SEX","RACE"),            by = "___")
)

demog_ard |>
  count(context, variable)
```

```{r demog-ard-hint-1}
# by = "TRT01P" in both calls (this study uses planned treatment).
```

```{r demog-ard-solution}
demog_ard <- ard_stack(
  ard_continuous(adsl,  variables = c("AGE","BMIBL","WEIGHTBL"), by = "TRT01P"),
  ard_categorical(adsl, variables = c("SEX","RACE"),            by = "TRT01P")
)

demog_ard |>
  count(context, variable)
```

### Exercise 1-b — Format mean (SD) for AGE per arm

The `continuous2` display style shows "mean (SD)". Pull mean and SD for AGE from
the ARD and format them as `"mean (SD)"` per arm.

```{r demog-meansd, exercise=TRUE, exercise.lines=12}
ard_continuous(adsl, variables = "AGE", by = "TRT01P") |>
  filter(stat_name %in% c("mean","sd")) |>
  mutate(stat = round(unlist(stat), 1)) |>
  select(group1_level, stat_name, stat) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(display = sprintf("%.1f (%.1f)", ___, ___))
```

```{r demog-meansd-hint-1}
# sprintf order: mean then sd.
```

```{r demog-meansd-solution}
ard_continuous(adsl, variables = "AGE", by = "TRT01P") |>
  filter(stat_name %in% c("mean","sd")) |>
  mutate(stat = round(unlist(stat), 1)) |>
  select(group1_level, stat_name, stat) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(display = sprintf("%.1f (%.1f)", mean, sd))
```

```{r quiz-s1, echo=FALSE}
quiz(
  question(
    "In Pattern A, where do the p-values come from?",
    answer("Manual t.test() calls woven into mutate()"),
    answer("cardx ard_stats_*() functions (e.g., ard_stats_anova_oneway, ard_stats_chisq_test) stacked into the ARD", correct=TRUE),
    answer("gtsummary computes them during rendering with no ARD"),
    answer("They are hard-coded in the template"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — Pattern B: AE Overview (FDA Table 14.4.x)

The AE *overview* table summarizes counts and rates of AE categories per arm: any
TEAE, serious TEAE, severe TEAE, leading-to-discontinuation, leading-to-death.

```r
adae_te <- adae |> filter(TRTEMFL == "Y")

adsl_with_flags <- adsl |>
  mutate(
    ANY_TEAE     = if_else(USUBJID %in% adae_te$USUBJID, "Y", "N"),
    SERIOUS_TEAE = if_else(USUBJID %in%
                   (adae_te |> filter(AESER == "Y") |> pull(USUBJID)), "Y", "N"),
    SEVERE_TEAE  = if_else(USUBJID %in%
                   (adae_te |> filter(AESEV == "SEVERE") |> pull(USUBJID)), "Y", "N")
  )

ae_overview_ard <- ard_categorical(
  adsl_with_flags, by = "TRT01A",
  variables = c(ANY_TEAE, SERIOUS_TEAE, SEVERE_TEAE),
  denominator = adsl
)

ae_overview_tbl <- ae_overview_ard |>
  tbl_ard_summary(by = "TRT01A", type = everything() ~ "dichotomous") |>
  modify_caption("**Table 14.4.1: Adverse Event Overview**")
```

**Pattern recognition:**

- Subject-level flags computed via `mutate()` — one per AE category.
- A single `ard_categorical()` call (a `dichotomous` display shows one "Y" row per category).
- Explicit `denominator = adsl` anchors the rate to the safety-population subject count.

> **Why the explicit denominator matters.** A subject with *no* AEs has no row in
> ADAE, so they can never appear in an ADAE-driven count. The denominator must be
> the **ADSL** subject count so that subjects without AEs are correctly counted in
> the proportion. This is the single most common AE-table bug for SAS programmers
> new to the pattern.

### Exercise 2-a — Build the AE overview ARD with a subject denominator

`adsl_flags` is pre-built in setup (it already has `ANY_TEAE`, `SERIOUS_TEAE`,
`SEVERE_TEAE`). Build the overview ARD for the three flags, anchoring the
denominator to the full ADSL so AE-free subjects count.

```{r ae-overview, exercise=TRUE, exercise.lines=12}
ae_overview_ard <- ard_categorical(
  adsl_flags,
  by          = "TRT01P",
  variables   = c("ANY_TEAE","SERIOUS_TEAE","SEVERE_TEAE"),
  denominator = ___
)

ae_overview_ard |>
  filter(variable == "ANY_TEAE", variable_level == "Y") |>
  select(group1_level, stat_name, stat) |>
  mutate(stat = unlist(stat))
```

```{r ae-overview-hint-1}
# denominator = adsl  -> the safety-population subject count per arm.
```

```{r ae-overview-solution}
ae_overview_ard <- ard_categorical(
  adsl_flags,
  by          = "TRT01P",
  variables   = c("ANY_TEAE","SERIOUS_TEAE","SEVERE_TEAE"),
  denominator = adsl
)

ae_overview_ard |>
  filter(variable == "ANY_TEAE", variable_level == "Y") |>
  select(group1_level, stat_name, stat) |>
  mutate(stat = unlist(stat))
```

### Exercise 2-b — Add a Wilson CI to the AE rates

cardinal AE-overview templates often add a confidence interval to the incidence.
Use `ard_proportion_ci()` (Wilson) on the subject-level flags and pull the ANY_TEAE
rate with its CI per arm.

```{r ae-overview-ci, exercise=TRUE, exercise.lines=14}
ae_ci <- ard_proportion_ci(
  adsl_flags,
  variables = c("ANY_TEAE","SERIOUS_TEAE","SEVERE_TEAE"),
  by        = "TRT01P",
  method    = "wilson",
  value     = list(ANY_TEAE="Y", SERIOUS_TEAE="Y", SEVERE_TEAE="Y")
)

ae_ci |>
  filter(variable == "___", stat_name %in% c("p","conf.low","conf.high")) |>
  mutate(stat = unlist(stat)) |>
  select(group1_level, stat_name, stat) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(display = sprintf("%.1f%% (%.1f, %.1f)", p, conf.low, conf.high))
```

```{r ae-overview-ci-hint-1}
# variable == "ANY_TEAE"
# Wilson keeps the CI within [0, 100] even for extreme rates.
```

```{r ae-overview-ci-solution}
ae_ci <- ard_proportion_ci(
  adsl_flags,
  variables = c("ANY_TEAE","SERIOUS_TEAE","SEVERE_TEAE"),
  by        = "TRT01P",
  method    = "wilson",
  value     = list(ANY_TEAE="Y", SERIOUS_TEAE="Y", SEVERE_TEAE="Y")
)

ae_ci |>
  filter(variable == "ANY_TEAE", stat_name %in% c("p","conf.low","conf.high")) |>
  mutate(stat = unlist(stat)) |>
  select(group1_level, stat_name, stat) |>
  pivot_wider(names_from = stat_name, values_from = stat) |>
  mutate(display = sprintf("%.1f%% (%.1f, %.1f)", p, conf.low, conf.high))
```

```{r quiz-s2, echo=FALSE}
quiz(
  question(
    "Why must the AE overview pass denominator = adsl rather than letting the ADAE counts define their own base?",
    answer("To make the table render faster"),
    answer("So subjects with no AEs are still counted in the denominator, giving correct incidence", correct=TRUE),
    answer("Because ADAE has no USUBJID"),
    answer("To exclude screen failures"),
    allow_retry = TRUE
  ),
  question(
    "Wilson CI is the regulatory default for binary AE rates because:",
    answer("It is always wider than Wald"),
    answer("It stays within [0, 100] even for proportions near 0 or 1", correct=TRUE),
    answer("It requires fewer subjects"),
    answer("It uses the t-distribution"),
    allow_retry = TRUE
  )
)
```

---

## Section 3 — Pattern C: AE by SOC and PT (FDA Table 14.5.x)

The detailed AE table with the **SOC × PT** hierarchy (System Organ Class →
Preferred Term). This is the classic AE table.

```r
ae_ard <- adae |>
  filter(TRTEMFL == "Y") |>
  ard_hierarchical(
    by          = "ARM",
    variables   = c("AEBODSYS", "AEDECOD"),
    denominator = adsl
  )

ae_tbl <- ae_ard |>
  tbl_ard_summary(by = "TRT01A", overall = FALSE) |>
  modify_header(label ~ "**System Organ Class<br>&nbsp;&nbsp;Preferred Term**") |>
  modify_caption("**Table 14.5.1: Adverse Events by SOC and Preferred Term**")
```

`ard_hierarchical()` does the heavy lifting; gtsummary renders the SOC/PT
indentation automatically.

**Common customizations:**

- **Incidence threshold** — show only PTs occurring in ≥ 5% of any arm (apply at
  the ARD level or post-hoc).
- **Sort order** — alphabetical (default) vs. by total incidence.
- **Severity breakdown** — a multi-column version split by Mild/Moderate/Severe
  using grade as an additional grouping.

### Exercise 3-a — Build the SOC × PT hierarchical ARD

`ard_hierarchical()` (mocked here) takes the treatment-emergent ADAE and produces
one subject-level incidence row per SOC/PT, anchored to the ADSL denominator.

```{r soc-pt, exercise=TRUE, exercise.lines=12}
ae_socpt <- adae |>
  filter(TRTEMFL == "Y") |>
  ard_hierarchical(
    by          = "TRT01P",
    variables   = c("AEBODSYS", "AEDECOD"),
    denominator = ___
  )

# Top preferred terms by incidence in any arm
ae_socpt |>
  arrange(desc(p)) |>
  head(8)
```

```{r soc-pt-hint-1}
# denominator = adsl
# variables order is c(SOC, PT) -> AEBODSYS first, AEDECOD second.
```

```{r soc-pt-solution}
ae_socpt <- adae |>
  filter(TRTEMFL == "Y") |>
  ard_hierarchical(
    by          = "TRT01P",
    variables   = c("AEBODSYS", "AEDECOD"),
    denominator = adsl
  )

ae_socpt |>
  arrange(desc(p)) |>
  head(8)
```

### Exercise 3-b — Apply a 5% incidence threshold

A near-universal customization: keep only PTs reaching ≥ 5% incidence in *any*
arm. Filter the hierarchical ARD accordingly.

```{r soc-pt-threshold, exercise=TRUE, exercise.lines=12}
ae_socpt <- adae |>
  filter(TRTEMFL == "Y") |>
  ard_hierarchical(by = "TRT01P",
                   variables = c("AEBODSYS","AEDECOD"),
                   denominator = adsl)

keep_pts <- ae_socpt |>
  group_by(PT) |>
  summarise(max_p = max(p), .groups = "drop") |>
  filter(max_p >= ___) |>
  pull(PT)

ae_socpt |>
  filter(PT %in% keep_pts) |>
  arrange(SOC, desc(p))
```

```{r soc-pt-threshold-hint-1}
# 5% threshold -> max_p >= 5
```

```{r soc-pt-threshold-solution}
ae_socpt <- adae |>
  filter(TRTEMFL == "Y") |>
  ard_hierarchical(by = "TRT01P",
                   variables = c("AEBODSYS","AEDECOD"),
                   denominator = adsl)

keep_pts <- ae_socpt |>
  group_by(PT) |>
  summarise(max_p = max(p), .groups = "drop") |>
  filter(max_p >= 5) |>
  pull(PT)

ae_socpt |>
  filter(PT %in% keep_pts) |>
  arrange(SOC, desc(p))
```

---

## Section 4 — Pattern D: Lab Shift Table (FDA Table 14.9.x)

A lab **shift table** answers: "how many subjects shifted from NORMAL at baseline
to HIGH on treatment?" The display is a 2D cross-tab: baseline reference-range
category × worst on-treatment category, with cell counts.

```r
adlb <- pharmaverseadam::adlb |>
  filter(SAFFL == "Y" & ANL01FL == "Y" & PARAMCD == "HGB")

shift_data <- adlb |>
  filter(!is.na(BNRIND) & ONTRTFL == "Y") |>
  group_by(USUBJID, BNRIND) |>
  summarise(
    WORST_TRT_IND = case_when(
      any(ANRIND == "HIGH")   ~ "HIGH",
      any(ANRIND == "LOW")    ~ "LOW",
      any(ANRIND == "NORMAL") ~ "NORMAL",
      TRUE                    ~ NA_character_),
    .groups = "drop")

shift_ard <- ard_categorical(
  shift_data, by = c(TRT01A, BNRIND),
  variables = WORST_TRT_IND, denominator = adsl |> count(TRT01A))
```

Shift tables get visually complex — row groups (baseline categories) and column
groups (treatment × on-treatment). For the most complex layouts, switch to
**tfrmt** (Lesson 32), which handles multi-dimensional displays better than
gtsummary.

> **adlb in this tutorial.** pharmaverseadam's adlb may not carry `BNRIND` /
> `ANRIND` reference-range indicators. Below we build an equivalent shift using
> the NCI CTCAE toxicity grades that adlb *does* carry (`BTOXGR` baseline,
> `ATOXGR` analysis), collapsed to NORMAL (grade 0) vs ABNORMAL (grade != 0).

### Exercise 4-a — Build a baseline → worst-on-treatment shift

Collapse `BTOXGR`/`ATOXGR` into NORMAL vs ABNORMAL, take the worst on-treatment
status per subject, then cross-tabulate baseline vs worst-on-treatment for one
parameter.

```{r lab-shift, exercise=TRUE, exercise.lines=22}
shift_data <- adlb |>
  filter(PARAMCD == "ALT", !is.na(BTOXGR), !is.na(ATOXGR)) |>
  mutate(
    BASE_CAT = if_else(BTOXGR == 0L, "NORMAL", "ABNORMAL"),
    TRT_CAT  = if_else(ATOXGR == 0L, "NORMAL", "ABNORMAL")
  ) |>
  group_by(USUBJID, BASE_CAT) |>
  summarise(
    WORST_TRT = if_else(any(TRT_CAT == "ABNORMAL"), "ABNORMAL", "NORMAL"),
    .groups = "drop"
  )

shift_data |>
  count(BASE_CAT, ___)
```

```{r lab-shift-hint-1}
# Cross-tabulate baseline vs the worst on-treatment category:
# count(BASE_CAT, WORST_TRT)
```

```{r lab-shift-solution}
shift_data <- adlb |>
  filter(PARAMCD == "ALT", !is.na(BTOXGR), !is.na(ATOXGR)) |>
  mutate(
    BASE_CAT = if_else(BTOXGR == 0L, "NORMAL", "ABNORMAL"),
    TRT_CAT  = if_else(ATOXGR == 0L, "NORMAL", "ABNORMAL")
  ) |>
  group_by(USUBJID, BASE_CAT) |>
  summarise(
    WORST_TRT = if_else(any(TRT_CAT == "ABNORMAL"), "ABNORMAL", "NORMAL"),
    .groups = "drop"
  )

shift_data |>
  count(BASE_CAT, WORST_TRT)
```

### Exercise 4-b — The shift table as a 2D matrix

Pivot the shift counts into the familiar 2D shift matrix (baseline rows ×
on-treatment columns).

```{r lab-shift-2d, exercise=TRUE, exercise.lines=18}
shift_data <- adlb |>
  filter(PARAMCD == "ALT", !is.na(BTOXGR), !is.na(ATOXGR)) |>
  mutate(BASE_CAT = if_else(BTOXGR == 0L, "NORMAL", "ABNORMAL"),
         TRT_CAT  = if_else(ATOXGR == 0L, "NORMAL", "ABNORMAL")) |>
  group_by(USUBJID, BASE_CAT) |>
  summarise(WORST_TRT = if_else(any(TRT_CAT == "ABNORMAL"),
                                "ABNORMAL", "NORMAL"), .groups = "drop")

shift_data |>
  count(BASE_CAT, WORST_TRT) |>
  pivot_wider(names_from = ___, values_from = n, values_fill = 0)
```

```{r lab-shift-2d-hint-1}
# Columns are the worst on-treatment category: names_from = WORST_TRT
```

```{r lab-shift-2d-solution}
shift_data <- adlb |>
  filter(PARAMCD == "ALT", !is.na(BTOXGR), !is.na(ATOXGR)) |>
  mutate(BASE_CAT = if_else(BTOXGR == 0L, "NORMAL", "ABNORMAL"),
         TRT_CAT  = if_else(ATOXGR == 0L, "NORMAL", "ABNORMAL")) |>
  group_by(USUBJID, BASE_CAT) |>
  summarise(WORST_TRT = if_else(any(TRT_CAT == "ABNORMAL"),
                                "ABNORMAL", "NORMAL"), .groups = "drop")

shift_data |>
  count(BASE_CAT, WORST_TRT) |>
  pivot_wider(names_from = WORST_TRT, values_from = n, values_fill = 0)
```

```{r quiz-s4, echo=FALSE}
quiz(
  question(
    "Why do complex lab shift tables sometimes warrant tfrmt instead of gtsummary?",
    answer("tfrmt is faster"),
    answer("Shift tables have both row groups (baseline) and column groups (treatment x on-treatment), and tfrmt handles multi-dimensional layouts better", correct=TRUE),
    answer("gtsummary cannot read adlb"),
    answer("tfrmt is required by the FDA guide"),
    allow_retry = TRUE
  )
)
```

---

## Section 5 — Pattern E: Serious-AE Listing

Listings (the **L** in TLG) differ from summary tables: **one row per AE event**,
with subject identifiers and details — not aggregated.

```r
sae_listing <- adae |>
  filter(SAFFL == "Y" & TRTEMFL == "Y" & AESER == "Y") |>
  select(USUBJID, AGE, SEX, ARM, AESTDT, AEENDT, AEDECOD, AEBODSYS,
         AESEV, AESER, AEACN, AEOUT) |>
  arrange(USUBJID, AESTDT)

library(flextable)
sae_listing |>
  flextable() |>
  set_caption("Listing 14.6.1: Serious Adverse Events") |>
  save_as_rtf("outputs/l_14_6_1.rtf")
```

Listings **don't need ARDs** — they are raw event-level data with formatting.
`{flextable}` handles them directly; gtsummary is *not* the right tool. For longer
/ more complex listings (multi-page with continuation headers), `{r2rtf}` is often
preferred for its pagination control. cardinal ships both flextable- and
r2rtf-based listing templates.

### Exercise 5-a — Build a serious-AE listing

Build the event-level SAE listing: treatment-emergent serious AEs, selected
columns, sorted by subject then preferred term.

```{r sae-listing, exercise=TRUE, exercise.lines=12}
sae_listing <- adae |>
  filter(TRTEMFL == "Y" & AESER == "___") |>
  select(USUBJID, TRT01A, AEBODSYS, AEDECOD, AESEV) |>
  arrange(USUBJID, AEDECOD)

head(sae_listing, 10)
```

```{r sae-listing-hint-1}
# Serious AEs are flagged AESER == "Y".
```

```{r sae-listing-solution}
sae_listing <- adae |>
  filter(TRTEMFL == "Y" & AESER == "Y") |>
  select(USUBJID, TRT01A, AEBODSYS, AEDECOD, AESEV) |>
  arrange(USUBJID, AEDECOD)

head(sae_listing, 10)
```

```{r quiz-s5, echo=FALSE}
quiz(
  question(
    "Why doesn't gtsummary handle listings well?",
    answer("gtsummary cannot read ADAE"),
    answer("Listings are un-aggregated, event-level rows; gtsummary is built for summarized ARDs, so flextable/r2rtf are the right tools", correct=TRUE),
    answer("Listings require a Wilson CI"),
    answer("gtsummary only supports demographics"),
    allow_retry = TRUE
  )
)
```

---

## Section 6 — Pattern F: Subgroup Forest Plot (FDA Figure)

cardinal includes graph (**G**) templates. The signature graph is a **subgroup
forest plot** showing the treatment effect (e.g., hazard ratio) across subgroups,
with a vertical reference line, CIs as horizontal error bars, and a log-scaled
x-axis.

```r
subgroup_hrs <- adsl |>
  group_split(SUBGROUP) |>
  purrr::map_dfr(\(d) {
    fit <- coxph(Surv(AVAL, 1 - CNSR) ~ TRTA, data = d)
    ard_regression(fit) |> mutate(SUBGROUP = unique(d$SUBGROUP_LABEL))
  })

forest_data <- subgroup_hrs |>
  filter(stat_name %in% c("estimate","conf.low","conf.high")) |>
  pivot_wider(names_from = stat_name, values_from = stat)

ggplot(forest_data, aes(x = exp(estimate), y = SUBGROUP)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = exp(conf.low), xmax = exp(conf.high)), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(x = "Hazard Ratio (95% CI)", y = "Subgroup")
```

Forest plots are conceptually simple but visually demanding. cardinal templates
fix the standard layout; sponsor adaptations are mostly aesthetic.

> The MD lesson plan's sketch uses `EFFFL` and `TRTA`. In real CDISCPILOT01 ADSL
> there is **no** `EFFFL` — use `SAFFL` (or another genuine flag) and `TRT01A` /
> `TRT01P`. The exercise below builds the forest-plot *input data* the cardinal
> way using only real adsl variables and a descriptive effect measure
> (mean AGE difference vs Placebo) so the pattern is concrete without survival
> columns adlb/adsl don't carry here.

### Exercise 6-a — Assemble forest-plot input data across subgroups

Build a per-subgroup effect table (here: mean AGE per arm within SEX subgroups)
in the long → wide shape a forest plot consumes.

```{r forest-data, exercise=TRUE, exercise.lines=16}
forest_input <- adsl |>
  group_by(SUBGROUP = SEX, TRT01P) |>
  summarise(est = mean(AGE, na.rm = TRUE),
            lo  = est - 1.96 * sd(AGE, na.rm = TRUE) / sqrt(n()),
            hi  = est + 1.96 * sd(AGE, na.rm = TRUE) / sqrt(n()),
            .groups = "drop")

forest_input |>
  filter(TRT01P == "___") |>
  select(SUBGROUP, est, lo, hi)
```

```{r forest-data-hint-1}
# Pick a reference arm to view, e.g. TRT01P == "Placebo".
```

```{r forest-data-solution}
forest_input <- adsl |>
  group_by(SUBGROUP = SEX, TRT01P) |>
  summarise(est = mean(AGE, na.rm = TRUE),
            lo  = est - 1.96 * sd(AGE, na.rm = TRUE) / sqrt(n()),
            hi  = est + 1.96 * sd(AGE, na.rm = TRUE) / sqrt(n()),
            .groups = "drop")

forest_input |>
  filter(TRT01P == "Placebo") |>
  select(SUBGROUP, est, lo, hi)
```

```{r quiz-s6, echo=FALSE}
quiz(
  question(
    "A subgroup forest plot's standard layout includes:",
    answer("A pie chart per subgroup"),
    answer("A vertical reference line, point estimates, horizontal CI error bars, and a (often log-scaled) x-axis", correct=TRUE),
    answer("A stacked bar per arm"),
    answer("Only a single overall estimate"),
    allow_retry = TRUE
  )
)
```

---

## Section 7 — Assembling a complete CSR safety section

A complete CSR safety section is a folder of cardinal templates, each producing
one table or figure:

```
t_14_1_1_disposition.qmd     - Subject disposition
t_14_2_1_demographics.qmd    - Demographics (Pattern A)
t_14_3_1_exposure.qmd        - Exposure summary
t_14_4_1_ae_overview.qmd     - AE overview (Pattern B)
t_14_5_1_ae_soc_pt.qmd       - AE by SOC x PT (Pattern C)
t_14_6_1_sae_listing.qmd     - Serious AE listing (Pattern E)
t_14_7_1_disc_due_to_ae.qmd  - Discontinuation due to AE
t_14_8_1_deaths.qmd          - Deaths
t_14_9_1_lab_shift_hgb.qmd   - HGB shift table (Pattern D)
f_14_4_1_ae_forest.qmd       - AE incidence forest plot (Pattern F)
```

Render them all in one shot:

```r
library(quarto)
templates <- list.files("csr/safety/templates", pattern = "\\.qmd$", full.names = TRUE)
for (t in templates) quarto_render(t, quiet = TRUE)
# all RTFs / PNGs land in outputs/
```

Ten templates → ten RTFs in a few minutes. For a phase III submission this might
be 20–30 templates rendering ~50–80 RTF tables and figures.

### Common sponsor adaptations

| Adaptation | cardinal default | sponsor variant |
|---|---|---|
| Population | `filter(SAFFL == "Y")` | `filter(SAFFL == "Y" & ANLPOP == "Y")` |
| MedDRA version | test-data default | pinned study version |
| Table numbering | `Table 14.4.1: ...` | `T-14-04-01: ...` (SOP convention) |
| Footnotes | generic | SOP-mandated text |
| Column labels | `stat_1` | `**Treatment A**<br>n = {n}` |

### When to start from cardinal vs from scratch

**Start from cardinal** when the table is a standard FDA-aligned safety summary, a
template exists, and your customizations are cosmetic.

**Start from scratch** when the table is highly study-specific (exotic biomarker
or efficacy endpoints), no template exists, or your SOP demands a fundamentally
different layout. Even then, you usually borrow the cardinal *pattern* (the ARD
structure, the gtsummary configuration).

For most CSRs, ~60–70% of tables start from cardinal templates; the rest need
custom work.

### Exercise 7-a — Re-point one template to a sponsor population

Take the AE overview from Section 2 and re-point its population to the sponsor's
requirement of `SAFFL == "Y" & DIABFL == "Y"`, then recompute the ANY_TEAE rate.
Keep the denominator consistent with the new population.

```{r csr-adapt, exercise=TRUE, exercise.lines=18}
adsl_sponsor <- adsl |>
  filter(SAFFL == "Y" & ___ == "Y")

flags_sponsor <- adsl_sponsor |>
  mutate(ANY_TEAE = if_else(USUBJID %in% adae_te$USUBJID, "Y", "N"))

ard_categorical(
  flags_sponsor,
  by          = "TRT01P",
  variables   = "ANY_TEAE",
  denominator = adsl_sponsor
) |>
  filter(variable_level == "Y") |>
  select(group1_level, stat_name, stat) |>
  mutate(stat = unlist(stat))
```

```{r csr-adapt-hint-1}
# Sponsor sub-population flag is DIABFL.
# The denominator must be the SAME re-pointed population (adsl_sponsor).
```

```{r csr-adapt-solution}
adsl_sponsor <- adsl |>
  filter(SAFFL == "Y" & DIABFL == "Y")

flags_sponsor <- adsl_sponsor |>
  mutate(ANY_TEAE = if_else(USUBJID %in% adae_te$USUBJID, "Y", "N"))

ard_categorical(
  flags_sponsor,
  by          = "TRT01P",
  variables   = "ANY_TEAE",
  denominator = adsl_sponsor
) |>
  filter(variable_level == "Y") |>
  select(group1_level, stat_name, stat) |>
  mutate(stat = unlist(stat))
```

```{r quiz-s7, echo=FALSE}
quiz(
  question(
    "Roughly what fraction of a typical CSR's tables can start from a cardinal template?",
    answer("Under 10%"),
    answer("About 60-70%", correct=TRUE),
    answer("Essentially 100%"),
    answer("None; cardinal is efficacy-only"),
    allow_retry = TRUE
  ),
  question(
    "Which is typically the FIRST customization a sponsor applies to a cardinal template?",
    answer("Rewriting the statistical method"),
    answer("Re-pointing the population / data source (test data to study data and the population filter)", correct=TRUE),
    answer("Switching from gtsummary to rtables"),
    answer("Translating the code to SAS"),
    allow_retry = TRUE
  )
)
```

---

## Lesson Summary

| Pattern | Table | ARD shape |
|---------|-------|-----------|
| A — Demographics | 14.2.x | `ard_stack(continuous + categorical)` + `cardx` p-values |
| B — AE overview | 14.4.x | subject-level flags → `ard_categorical(denominator = adsl)` |
| C — AE by SOC/PT | 14.5.x | `ard_hierarchical(c(SOC, PT), denominator = adsl)` |
| D — Lab shift | 14.9.x | worst-on-treatment per subject → cross-tab |
| E — SAE listing | 14.6.x | no ARD; event-level rows via flextable / r2rtf |
| F — Forest plot | Figure | per-subgroup effect ARD → ggplot |

### Key takeaways

- The **FDA Standard Safety Tables and Figures Integrated Guide** is cardinal's
  primary reference; each template implements one FDA table number.
- **Six recurring patterns** cover the bulk of a safety section: demographics, AE
  overview, AE by SOC/PT, lab shift, listings, forest plots.
- Every summary table is **ARD-first**: build the ARD (cards/cardx), then render
  (gtsummary) — except listings, which are raw event-level data via flextable/r2rtf.
- AE incidence tables must anchor the **denominator to ADSL** so AE-free subjects
  are counted; **Wilson** CIs are the regulatory default for rates.
- `ard_hierarchical()` builds the SOC × PT incidence table; a ≥ 5% threshold is a
  near-universal customization.
- Complex multi-dimensional shift tables may warrant **tfrmt** over gtsummary.
- A complete safety section is ~10 templates rendering ~50 RTFs in minutes; ~60–70%
  of CSR tables start from cardinal templates.
- Common sponsor adaptations are mostly cosmetic: population filter, MedDRA
  version, table numbering, footnotes, column labels.

### What's next

**Lesson 32 — tfrmt** is the final lesson in this module: the display-metadata
layer for ARDs. For tables gtsummary can't render elegantly — multi-dimensional
layouts, mock-table workflows, fully sponsor-templated displays — tfrmt provides a
separate metadata-driven layer that consumes the same ARDs you built here. After
tfrmt, the module is complete and the series moves to the legacy NEST TLG stack.

---

## Self-check questions

1. Which FDA reference is cardinal primarily implementing?
2. What is the structure of an AE-overview ARD, and why is `denominator = adsl` essential?
3. Which function builds the SOC × PT hierarchical AE table?
4. Why do complex lab shift tables sometimes warrant tfrmt instead of gtsummary?
5. Why don't listings use an ARD, and what renders them instead?
6. Name three FDA Table 14.x.x subjects and the cardinal pattern each uses.
7. What is typically the first customization a sponsor applies to a cardinal template?

## Glossary

- **FDA Standard Safety Tables and Figures** — FDA guide of canonical safety
  table layouts; cardinal's primary source.
- **Subject-level denominator** — Use the safety-population subject count for AE
  incidence; subjects with no AEs still count.
- **TEAE** — Treatment-Emergent Adverse Event (`TRTEMFL == "Y"`).
- **AE Overview** — One table summarizing major AE categories (any, serious, severe, …).
- **SOC × PT** — System Organ Class (MedDRA top level) × Preferred Term (lower level).
- **`ard_hierarchical()`** — Builds the nested SOC/PT subject-level incidence ARD.
- **Shift table** — Cross-tab of baseline category × worst on-treatment category.
- **Forest plot** — Subgroup-level effect-size plot with a reference line.
- **Listing** — Subject/event-level data display, not aggregated.
- **Wilson CI** — Default binary-proportion CI; stays within [0, 100].
- **MedDRA** — Medical Dictionary for Regulatory Activities; AE coding standard.
- **CSR safety section** — The standard ~10-table safety set of a CSR.
- **tfrmt** — Display-metadata layer for complex tables (Lesson 32).
- **flextable / r2rtf** — Render listings and paginated RTF output.
