---
title: "Lesson 31 — gt: Direct Table Formatting for Clinical Output"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  Master direct gt table formatting for clinical output: create tables with gt(),
  rename columns with cols_label(), add titles and footnotes with tab_header()
  and tab_footnote(), group columns under spanners with tab_spanner(), apply
  number formatting with fmt_number() and fmt_percent(), use conditional
  styling with tab_style(), set global options with tab_options(), and export
  to HTML and RTF with gtsave(). Build a complete Table 14.1.1 demographics
  table in the capstone.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 31_gt.Rmd
# PURPOSE       : gt package — direct table formatting: cols_label(),
#                 tab_header(), tab_footnote(), tab_source_note(),
#                 tab_spanner(), fmt_number(), fmt_percent(), tab_style(),
#                 tab_options(), gtsave(). Capstone: Table 14.1.1 demographics.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 161, 7)), round(rnorm(n, 174, 8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 68, 12), 1), round(rnorm(n, 82, 14), 1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL = ht, WEIGHTBL = wt,
                                  BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ══════════════════════════════════════════════════════════════════════════════
# GT AVAILABILITY CHECK AND THIN STUBS
# ══════════════════════════════════════════════════════════════════════════════

.gt_available <- requireNamespace("gt", quietly = TRUE)

if (!.gt_available) {
  # ── Thin stubs that mimic the gt API when the package is not installed ─────

  gt <- function(data, rowname_col = NULL, groupname_col = NULL, ...) {
    structure(list(data = data, opts = list(...), rowname_col = rowname_col,
                   groupname_col = groupname_col),
              class = "gt_tbl")
  }

  print.gt_tbl <- function(x, ...) {
    cat("[gt table — install the gt package to render fully]\n\n")
    print(as.data.frame(x$data), row.names = FALSE)
    invisible(x)
  }

  cols_label <- function(.data, ...) {
    args <- list(...)
    if (inherits(.data, "gt_tbl")) {
      .data$opts$col_labels <- c(.data$opts$col_labels, args)
      cat(sprintf("[cols_label()] Relabelling %d column(s): %s\n",
                  length(args), paste(names(args), collapse = ", ")))
    }
    .data
  }

  cols_align <- function(.data, align = "left", columns = everything()) {
    if (inherits(.data, "gt_tbl"))
      cat(sprintf("[cols_align()] align = \"%s\"\n", align))
    .data
  }

  tab_header <- function(.data, title = "", subtitle = NULL) {
    if (inherits(.data, "gt_tbl")) {
      .data$opts$title    <- title
      .data$opts$subtitle <- subtitle
      cat(sprintf("[tab_header()] title = \"%s\"", title))
      if (!is.null(subtitle)) cat(sprintf("  subtitle = \"%s\"", subtitle))
      cat("\n")
    }
    .data
  }

  tab_spanner <- function(.data, label = "", columns = NULL) {
    if (inherits(.data, "gt_tbl"))
      cat(sprintf("[tab_spanner()] label = \"%s\"\n", label))
    .data
  }

  tab_footnote <- function(.data, footnote = "", locations = NULL) {
    if (inherits(.data, "gt_tbl")) {
      .data$opts$footnotes <- c(.data$opts$footnotes, footnote)
      cat(sprintf("[tab_footnote()] \"%s\"\n", footnote))
    }
    .data
  }

  tab_source_note <- function(.data, source_note = "") {
    if (inherits(.data, "gt_tbl")) {
      .data$opts$source_notes <- c(.data$opts$source_notes, source_note)
      cat(sprintf("[tab_source_note()] \"%s\"\n", source_note))
    }
    .data
  }

  fmt_number <- function(.data, columns = NULL, decimals = 2, ...) {
    if (inherits(.data, "gt_tbl"))
      cat(sprintf("[fmt_number()] decimals = %d\n", decimals))
    .data
  }

  fmt_percent <- function(.data, columns = NULL, decimals = 1, ...) {
    if (inherits(.data, "gt_tbl"))
      cat(sprintf("[fmt_percent()] decimals = %d\n", decimals))
    .data
  }

  tab_style <- function(.data, style = NULL, locations = NULL) {
    if (inherits(.data, "gt_tbl"))
      cat("[tab_style()] Style rule applied\n")
    .data
  }

  tab_options <- function(.data, ...) {
    args <- list(...)
    if (inherits(.data, "gt_tbl") && length(args) > 0)
      cat(sprintf("[tab_options()] %d option(s) set: %s\n",
                  length(args), paste(names(args), collapse = ", ")))
    .data
  }

  gtsave <- function(.data, filename, ...) {
    cat(sprintf("[gtsave()] Would save to: %s\n", filename))
    invisible(.data)
  }

  cells_body <- function(columns = NULL, rows = NULL) {
    list(type = "cells_body", columns = columns, rows = rows)
  }

  cells_column_labels <- function(columns = NULL) {
    list(type = "cells_column_labels", columns = columns)
  }

  cell_text <- function(weight = NULL, color = NULL, style = NULL, ...) {
    args <- c(list(weight = weight, color = color, style = style), list(...))
    args <- args[!sapply(args, is.null)]
    structure(args, class = "cell_text")
  }

  cell_fill <- function(color = NULL) {
    structure(list(color = color), class = "cell_fill")
  }

  everything <- function() NULL

  md <- function(text) text   # markdown passthrough

} else {
  library(gt)
}

# ══════════════════════════════════════════════════════════════════════════════
# DISPLAY TIBBLES FOR EXERCISES
# ══════════════════════════════════════════════════════════════════════════════

# Demographics summary tibble
demog_display <- adsl |>
  group_by(TRT01P) |>
  summarise(
    N          = n(),
    Age_Mean   = round(mean(AGE,      na.rm = TRUE), 1),
    Age_SD     = round(sd(AGE,        na.rm = TRUE), 1),
    Pct_Female = round(100 * mean(SEX == "F", na.rm = TRUE), 1),
    .groups    = "drop"
  )

# AE counts by SOC and arm
# Drop TRT01P first if adae already carries it (avoids .x/.y suffix collision)
ae_counts <- adae |>
  select(-any_of("TRT01P")) |>
  left_join(select(adsl, USUBJID, TRT01P), by = "USUBJID") |>
  group_by(TRT01P, AESOC) |>
  summarise(n_ae   = n(),
            n_subj = n_distinct(USUBJID),
            .groups = "drop") |>
  pivot_wider(names_from  = TRT01P,
              values_from = c(n_ae, n_subj),
              values_fill = 0)

# Simplified demographics — transposed for gt formatting exercises
# Rows = statistic, Columns = treatment arm
demog_wide <- adsl |>
  group_by(TRT01P) |>
  summarise(
    N          = n(),
    Age_Mean   = round(mean(AGE, na.rm = TRUE), 1),
    Age_SD     = round(sd(AGE, na.rm = TRUE), 1),
    Pct_Female = round(100 * mean(SEX == "F", na.rm = TRUE), 1),
    .groups    = "drop"
  ) |>
  mutate(TRT01P = as.character(TRT01P))
```

## Overview

`gt` (short for **grammar of tables**) is the table-rendering engine underlying
both `gtsummary` and `tfrmt`. When you call `print_to_gt()` in tfrmt or use a
`gtsummary` object, gt is what produces the final HTML, RTF, or PDF output.

Learning gt directly gives you **precise control** over every cell, column, and
annotation in a table. It is particularly valuable when:

- tfrmt or gtsummary cannot produce the exact layout your SAP specifies
- You need to add complex conditional formatting (bold totals, highlighted
  significant p-values)
- You are debugging why a tfrmt output does not look as expected
- You need to build a bespoke table not covered by standard templates

### gt architecture

Every gt pipeline starts with a data frame and ends with a rendered output:

```
data frame
  |> gt()                   # create gt table object
  |> cols_label()           # rename column headers
  |> tab_header()           # add title and subtitle
  |> tab_spanner()          # group columns under a shared header
  |> tab_footnote()         # add footnotes
  |> tab_source_note()      # add source reference
  |> fmt_number()           # format numeric cells
  |> fmt_percent()          # format as percentages
  |> tab_style()            # conditional cell/text formatting
  |> tab_options()          # global style options
  |> gtsave("output.rtf")   # export to file
```

Each function returns the modified gt table object, making them fully
pipe-composable.

### Trial Dataset

All exercises use **CDISCPILOT01** from `pharmaverseadam` (254 subjects, Safety
Population, three treatment arms). Two pre-built display tibbles are available:

- `demog_display` — N, mean age, SD age, % female by treatment arm (4 rows × 5 cols)
- `ae_counts` — AE counts by SOC and treatment arm

---

## Section 1 — gt() and cols_label()

`gt(data)` converts a data frame into a gt table object. The column names
from the data frame become the default headers. `cols_label()` renames them
using a `new_label = old_column_name` syntax.

```{r ex-gt-basic, exercise=TRUE, exercise.lines=12}
# Create a simple gt table from demog_display.
# Print it to see the default column names.

demog_display |> ___()
```

```{r ex-gt-basic-hint-1}
# demog_display |> gt()
```

```{r ex-gt-basic-solution}
demog_display |> gt()
```

```{r ex-gt-labels, exercise=TRUE, exercise.lines=16}
# Now rename the columns to human-readable labels using cols_label().
# TRT01P       → "Treatment"
# N            → "N"
# Age_Mean     → "Mean Age"
# Age_SD       → "SD"
# Pct_Female   → "% Female"

demog_display |>
  gt() |>
  cols_label(
    TRT01P     = ___,
    N          = ___,
    Age_Mean   = ___,
    Age_SD     = ___,
    Pct_Female = ___
  )
```

```{r ex-gt-labels-hint-1}
# cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
#            Age_SD = "SD", Pct_Female = "% Female")
```

```{r ex-gt-labels-solution}
demog_display |>
  gt() |>
  cols_label(
    TRT01P     = "Treatment",
    N          = "N",
    Age_Mean   = "Mean Age",
    Age_SD     = "SD",
    Pct_Female = "% Female"
  )
```

```{r ex-gt-align, exercise=TRUE, exercise.lines=14}
# Align the N column to the right and the TRT01P column to the left.
# cols_align(align = "right", columns = N) or cols_align(align = "left", columns = TRT01P)

demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N",
             Age_Mean = "Mean Age", Age_SD = "SD", Pct_Female = "% Female") |>
  cols_align(align = ___, columns = N) |>
  cols_align(align = ___, columns = TRT01P)
```

```{r ex-gt-align-hint-1}
# cols_align(align = "right", columns = N)
# cols_align(align = "left",  columns = TRT01P)
```

```{r ex-gt-align-solution}
demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N",
             Age_Mean = "Mean Age", Age_SD = "SD", Pct_Female = "% Female") |>
  cols_align(align = "right", columns = N) |>
  cols_align(align = "left",  columns = TRT01P)
```

```{r quiz-gt-basics, echo=FALSE}
quiz(
  question(
    "In `cols_label()`, which side of the `=` is the new display label?",
    answer("Left side — `new_label = column_name`",
           message = "In gt, cols_label() uses new_label = old_col, so the label is on the RIGHT."),
    answer("Right side — `column_name = \"New Label\"`",
           correct = TRUE,
           message = "Correct: cols_label(Age_Mean = \"Mean Age\") — the column name is on the left, the new label string on the right."),
    allow_retry = TRUE
  ),
  question(
    "Which function creates the initial gt table object from a data frame?",
    answer("gt_table()"),
    answer("as_gt()"),
    answer("gt()", correct = TRUE),
    answer("make_gt()"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — tab_header(), tab_footnote(), tab_source_note()

Clinical tables require a **title** (table number + description), **footnotes**
(explaining abbreviations, flags, or methods), and a **source note** (identifying
the input dataset). gt handles all three through dedicated functions.

```{r ex-header-add, exercise=TRUE, exercise.lines=16}
# Add a title and subtitle to the demographics table.
# title    = "Table 14.1.1"
# subtitle = "Summary of Demographics and Baseline Characteristics (Safety Population)"

demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = ___,
    subtitle = ___
  )
```

```{r ex-header-add-hint-1}
# title = "Table 14.1.1"
# subtitle = "Summary of Demographics and Baseline Characteristics (Safety Population)"
```

```{r ex-header-add-solution}
demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographics and Baseline Characteristics (Safety Population)"
  )
```

```{r ex-footnote-source, exercise=TRUE, exercise.lines=20}
# Add a footnote explaining the SD column and a source note referencing ADSL.
# Footnote: "SD = Standard Deviation"
# Source note: "Source: ADSL (data cutoff 2014-01-02)"

demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographics and Baseline Characteristics (Safety Population)"
  ) |>
  tab_footnote(footnote  = ___) |>
  tab_source_note(source_note = ___)
```

```{r ex-footnote-source-hint-1}
# tab_footnote(footnote = "SD = Standard Deviation")
# tab_source_note(source_note = "Source: ADSL (data cutoff 2014-01-02)")
```

```{r ex-footnote-source-solution}
demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographics and Baseline Characteristics (Safety Population)"
  ) |>
  tab_footnote(footnote    = "SD = Standard Deviation") |>
  tab_source_note(source_note = "Source: ADSL (data cutoff 2014-01-02)")
```

```{r ex-multi-footnote, exercise=TRUE, exercise.lines=20}
# Add TWO footnotes to a simple gt table:
# 1. "N = number of subjects in the Safety Population"
# 2. "% Female = percentage of female subjects"
# Chain two tab_footnote() calls.

demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_footnote(footnote = ___) |>
  tab_footnote(footnote = ___)
```

```{r ex-multi-footnote-hint-1}
# tab_footnote(footnote = "N = number of subjects in the Safety Population")
# tab_footnote(footnote = "% Female = percentage of female subjects")
```

```{r ex-multi-footnote-solution}
demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_footnote(footnote = "N = number of subjects in the Safety Population") |>
  tab_footnote(footnote = "% Female = percentage of female subjects")
```

---

## Section 3 — tab_spanner(): Grouping Columns Under a Shared Header

A **spanner** is a merged header cell spanning multiple columns. This is common
in clinical tables where the same statistics are shown for multiple treatment
arms or multiple visits.

```r
tab_spanner(label = "Xanomeline", columns = c(stat_Low, stat_High))
```

The `columns` argument accepts a character vector of column names or tidy
selection expressions.

```{r ex-span-ae, exercise=TRUE, exercise.lines=22}
# ae_counts has columns: AESOC, then n_ae_* and n_subj_* for each arm.
# Print ae_counts first to see the column names, then add a spanner
# "Number of AEs" above n_ae_Placebo and the first Xanomeline n_ae column.

# Step 1: inspect column names
names(ae_counts)
```

```{r ex-span-ae-hint-1}
# names(ae_counts) shows you the column names to use in the spanner.
# Look for the n_ae_* columns to use as the spanner targets.
```

```{r ex-span-ae-solution}
names(ae_counts)
```

```{r ex-span-build, exercise=TRUE, exercise.lines=22}
# Build a gt table from demog_wide with a spanner "Age Statistics"
# above the Age_Mean and Age_SD columns.
# Then add another spanner "Enrollment" above N and Pct_Female.

demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_spanner(
    label   = "Age Statistics",
    columns = c(___, ___)
  ) |>
  tab_spanner(
    label   = "Enrollment",
    columns = c(___, ___)
  )
```

```{r ex-span-build-hint-1}
# columns = c(Age_Mean, Age_SD) for the age spanner
# columns = c(N, Pct_Female) for the enrollment spanner
```

```{r ex-span-build-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_spanner(
    label   = "Age Statistics",
    columns = c(Age_Mean, Age_SD)
  ) |>
  tab_spanner(
    label   = "Enrollment",
    columns = c(N, Pct_Female)
  )
```

---

## Section 4 — fmt_number() and fmt_percent()

gt's `fmt_*` functions control how numeric values are displayed. They do **not**
change the underlying data — they only change how cells are printed.

- `fmt_number(columns, decimals)` — format as a plain number with fixed decimal places
- `fmt_percent(columns, decimals)` — multiply by 100 and append "%" (note: use only
  on columns that hold proportions 0–1, not already-computed percentages)

> **Clinical gotcha:** In pharmaverseadam summary tibbles, `Pct_Female` already
> holds values like `47.1` (i.e., it is already a percentage, not a proportion).
> Apply `fmt_number()` to it rather than `fmt_percent()`, or pre-divide by 100
> before passing to `fmt_percent()`.

```{r ex-fmt-number, exercise=TRUE, exercise.lines=16}
# Apply fmt_number() to Age_Mean and Age_SD with 1 decimal place.
# Apply fmt_number() to Pct_Female with 1 decimal place (it is already %).

demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  fmt_number(columns = c(___, ___), decimals = ___) |>
  fmt_number(columns = ___, decimals = ___)
```

```{r ex-fmt-number-hint-1}
# fmt_number(columns = c(Age_Mean, Age_SD), decimals = 1)
# fmt_number(columns = Pct_Female, decimals = 1)
```

```{r ex-fmt-number-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  fmt_number(columns = c(Age_Mean, Age_SD), decimals = 1) |>
  fmt_number(columns = Pct_Female, decimals = 1)
```

```{r ex-fmt-percent, exercise=TRUE, exercise.lines=18}
# Build a small proportion tibble and demonstrate fmt_percent().
# prop_tbl holds proportions 0–1 — fmt_percent() will multiply by 100 and add "%".

prop_tbl <- tibble(
  group    = c("Placebo", "Low Dose", "High Dose"),
  prop_sae = c(0.071, 0.095, 0.112)
)

prop_tbl |>
  gt() |>
  cols_label(group = "Treatment", prop_sae = "SAE Rate") |>
  fmt_percent(columns = ___, decimals = ___)
```

```{r ex-fmt-percent-hint-1}
# fmt_percent(columns = prop_sae, decimals = 1)
# fmt_percent() multiplies 0.071 → "7.1%"
```

```{r ex-fmt-percent-solution}
prop_tbl <- tibble(
  group    = c("Placebo", "Low Dose", "High Dose"),
  prop_sae = c(0.071, 0.095, 0.112)
)

prop_tbl |>
  gt() |>
  cols_label(group = "Treatment", prop_sae = "SAE Rate") |>
  fmt_percent(columns = prop_sae, decimals = 1)
```

```{r quiz-fmt, echo=FALSE}
quiz(
  question(
    "You have a column `pct_resp` holding values like `42.3` (already a percentage). Which function should you use?",
    answer("fmt_percent()", message = "fmt_percent() multiplies by 100 first, so 42.3 would display as 4230%. Use fmt_number() instead."),
    answer("fmt_number()", correct = TRUE,
           message = "Correct — the value is already a percentage. fmt_number() just controls decimal places without rescaling."),
    answer("fmt_integer()"),
    answer("fmt_scientific()"),
    allow_retry = TRUE
  )
)
```

---

## Section 5 — tab_style(): Conditional Formatting

`tab_style()` applies formatting rules to specific cells. It takes two arguments:

- `style` — what to apply: `cell_text(weight = "bold")`, `cell_fill(color = "#FFF0F0")`
- `locations` — where to apply it: `cells_body(rows = ...)`, `cells_column_labels()`

**Common clinical uses:**
- Bold all column headers
- Bold the "Total" row
- Highlight rows where a p-value is significant (< 0.05)
- Shade alternate rows for readability

```{r ex-style-headers, exercise=TRUE, exercise.lines=16}
# Bold all column labels (headers) in the demog_wide table.

demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_style(
    style     = cell_text(weight = ___),
    locations = ___(          )
  )
```

```{r ex-style-headers-hint-1}
# style = cell_text(weight = "bold")
# locations = cells_column_labels()
```

```{r ex-style-headers-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

```{r ex-style-rows, exercise=TRUE, exercise.lines=20}
# Highlight rows in demog_wide where Age_Mean > 75 with a light yellow fill.
# Use cells_body(rows = Age_Mean > 75) as the location.

demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_style(
    style     = cell_fill(color = ___),
    locations = cells_body(rows = ___)
  )
```

```{r ex-style-rows-hint-1}
# cell_fill(color = "#FFFACD")  -- light yellow (lemon chiffon)
# cells_body(rows = Age_Mean > 75)
```

```{r ex-style-rows-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_style(
    style     = cell_fill(color = "#FFFACD"),
    locations = cells_body(rows = Age_Mean > 75)
  )
```

```{r ex-style-combine, exercise=TRUE, exercise.lines=22}
# Combine two tab_style() calls:
# 1. Bold the column headers (cells_column_labels())
# 2. Make Age_Mean values italic in the body (cell_text(style = "italic"))

demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(style = ___),
    locations = cells_body(columns = ___)
  )
```

```{r ex-style-combine-hint-1}
# style = "italic" inside cell_text()
# cells_body(columns = Age_Mean)
```

```{r ex-style-combine-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(style = "italic"),
    locations = cells_body(columns = Age_Mean)
  )
```

---

## Section 6 — tab_options(): Global Style Options

While `tab_style()` targets specific cells, `tab_options()` sets **global**
defaults for the whole table — font size, border widths, striped rows, header
background colour, and more.

Key options for clinical output:

| Option | Typical value | Effect |
|--------|--------------|--------|
| `table.font.size` | `10` | Font size in points |
| `column_labels.font.weight` | `"bold"` | Bold all column headers |
| `row.striping.include_table_body` | `TRUE` | Alternating row shading |
| `heading.background.color` | `"#DDEEFF"` | Title block background |
| `table.border.top.width` | `2` | Top border width (px) |

```{r ex-options-basic, exercise=TRUE, exercise.lines=16}
# Set global font size to 10pt and make column labels bold.

demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_options(
    table.font.size              = ___,
    column_labels.font.weight    = ___
  )
```

```{r ex-options-basic-hint-1}
# table.font.size = 10
# column_labels.font.weight = "bold"
```

```{r ex-options-basic-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_options(
    table.font.size           = 10,
    column_labels.font.weight = "bold"
  )
```

```{r ex-options-striped, exercise=TRUE, exercise.lines=18}
# Add striped rows (alternating shading) to the demog table.
# Use row.striping.include_table_body = TRUE
# and row.striping.background_color = "#F5F5F5"

demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_options(
    row.striping.include_table_body  = ___,
    row.striping.background_color    = ___
  )
```

```{r ex-options-striped-hint-1}
# row.striping.include_table_body = TRUE
# row.striping.background_color = "#F5F5F5"
```

```{r ex-options-striped-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_options(
    row.striping.include_table_body = TRUE,
    row.striping.background_color   = "#F5F5F5"
  )
```

---

## Section 7 — gtsave(): Exporting Tables

`gtsave(.data, filename)` writes the gt table to a file. The output format is
inferred from the file extension:

| Extension | Format | Notes |
|-----------|--------|-------|
| `.html` | HTML | Self-contained; renders in any browser |
| `.rtf` | RTF | Opens in Word; required for most submission pipelines |
| `.pdf` | PDF | Requires `webshot2` + chromium |
| `.png` | PNG image | Requires `webshot2` |
| `.docx` | Word | Requires `officedown` / `officer` |

For clinical submissions, **RTF** is the standard because it integrates with
Word-based publishing workflows and preserves formatting in ISS/ISE documents.

```{r ex-save-html, exercise=TRUE, exercise.lines=18}
# Build the demog table with labels and title, then save as HTML.

demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographics (Safety Population)"
  ) |>
  gtsave(filename = ___)
```

```{r ex-save-html-hint-1}
# gtsave(filename = "Table_14_1_1.html")
```

```{r ex-save-html-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographics (Safety Population)"
  ) |>
  gtsave(filename = "Table_14_1_1.html")
```

```{r ex-save-rtf, exercise=TRUE, exercise.lines=14}
# Save the same table as RTF for submission.
# The only change is the filename extension.

demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(title = "Table 14.1.1",
             subtitle = "Summary of Demographics (Safety Population)") |>
  gtsave(filename = ___)
```

```{r ex-save-rtf-hint-1}
# gtsave(filename = "Table_14_1_1.rtf")
```

```{r ex-save-rtf-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(title = "Table 14.1.1",
             subtitle = "Summary of Demographics (Safety Population)") |>
  gtsave(filename = "Table_14_1_1.rtf")
```

```{r quiz-gtsave, echo=FALSE}
quiz(
  question(
    "Which file extension does gtsave() use for a Word-importable submission table?",
    answer(".html"),
    answer(".pdf"),
    answer(".rtf", correct = TRUE,
           message = "RTF (Rich Text Format) is the standard for clinical submission tables because it can be directly imported into Word-based CSR publishing workflows."),
    answer(".xlsx"),
    allow_retry = TRUE
  ),
  question(
    "What determines the output format when you call gtsave()?",
    answer("A format= argument you must always specify"),
    answer("The file extension in the filename argument", correct = TRUE,
           message = "gtsave() infers format from the extension: .html, .rtf, .pdf, .png, .docx."),
    answer("The theme argument in the tfrmt spec"),
    answer("The operating system locale"),
    allow_retry = TRUE
  )
)
```

---

## Section 8 — Capstone: Table 14.1.1 Demographics

Combine every element from this lesson to build a complete, publication-ready
Table 14.1.1 from `demog_display`:

- `cols_label()` — human-readable column headers
- `tab_header()` — table number and title
- `tab_spanner()` — "Age (Years)" spanner above Mean and SD
- `fmt_number()` — 1 decimal place for Age_Mean, Age_SD, Pct_Female
- `tab_style()` — bold column headers; italicise % Female values
- `tab_footnote()` — "SD = Standard Deviation"
- `tab_source_note()` — "Source: ADSL; Safety Population (SAFFL = Y)"
- `tab_options()` — font size 10, bold column labels
- `gtsave()` — export to `"Table_14_1_1.rtf"`

```{r ex-cap-step1, exercise=TRUE, exercise.lines=20}
# Step 1: Create the gt table and rename columns.

tbl_step1 <- demog_display |>
  gt() |>
  cols_label(
    TRT01P     = ___,
    N          = ___,
    Age_Mean   = ___,
    Age_SD     = ___,
    Pct_Female = ___
  )

tbl_step1
```

```{r ex-cap-step1-hint-1}
# TRT01P = "Treatment", N = "N", Age_Mean = "Mean",
# Age_SD = "SD", Pct_Female = "% Female"
```

```{r ex-cap-step1-solution}
tbl_step1 <- demog_display |>
  gt() |>
  cols_label(
    TRT01P     = "Treatment",
    N          = "N",
    Age_Mean   = "Mean",
    Age_SD     = "SD",
    Pct_Female = "% Female"
  )

tbl_step1
```

```{r ex-cap-step2, exercise=TRUE, exercise.lines=24}
# Step 2: Add the title, spanner, and number formatting.

tbl_step2 <- demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N",
             Age_Mean = "Mean", Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = ___,
    subtitle = ___
  ) |>
  tab_spanner(
    label   = ___,
    columns = c(___, ___)
  ) |>
  fmt_number(columns = c(Age_Mean, Age_SD), decimals = ___) |>
  fmt_number(columns = Pct_Female, decimals = ___)

tbl_step2
```

```{r ex-cap-step2-hint-1}
# title = "Table 14.1.1"
# subtitle = "Summary of Demographics and Baseline Characteristics (Safety Population)"
# tab_spanner label = "Age (Years)", columns = c(Age_Mean, Age_SD)
# decimals = 1 for both fmt_number calls
```

```{r ex-cap-step2-solution}
tbl_step2 <- demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N",
             Age_Mean = "Mean", Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographics and Baseline Characteristics (Safety Population)"
  ) |>
  tab_spanner(
    label   = "Age (Years)",
    columns = c(Age_Mean, Age_SD)
  ) |>
  fmt_number(columns = c(Age_Mean, Age_SD), decimals = 1) |>
  fmt_number(columns = Pct_Female, decimals = 1)

tbl_step2
```

```{r ex-cap-final, exercise=TRUE, exercise.lines=44}
# Step 3: Complete pipeline — add styling, footnotes, options, and export.

final_tbl <- demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N",
             Age_Mean = "Mean", Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographics and Baseline Characteristics (Safety Population)"
  ) |>
  tab_spanner(label = "Age (Years)", columns = c(Age_Mean, Age_SD)) |>
  fmt_number(columns = c(Age_Mean, Age_SD), decimals = 1) |>
  fmt_number(columns = Pct_Female, decimals = 1) |>
  tab_style(
    style     = cell_text(weight = ___),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(style = "italic"),
    locations = cells_body(columns = ___)
  ) |>
  tab_footnote(footnote    = ___) |>
  tab_source_note(source_note = ___) |>
  tab_options(
    table.font.size           = ___,
    column_labels.font.weight = ___
  ) |>
  gtsave(filename = ___)

final_tbl
```

```{r ex-cap-final-hint-1}
# tab_style bold: weight = "bold"
# tab_style italic: cells_body(columns = Pct_Female)
# tab_footnote: "SD = Standard Deviation"
# tab_source_note: "Source: ADSL; Safety Population (SAFFL = Y)"
# tab_options: table.font.size = 10, column_labels.font.weight = "bold"
# gtsave: "Table_14_1_1.rtf"
```

```{r ex-cap-final-solution}
final_tbl <- demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N",
             Age_Mean = "Mean", Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographics and Baseline Characteristics (Safety Population)"
  ) |>
  tab_spanner(label = "Age (Years)", columns = c(Age_Mean, Age_SD)) |>
  fmt_number(columns = c(Age_Mean, Age_SD), decimals = 1) |>
  fmt_number(columns = Pct_Female, decimals = 1) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_text(style = "italic"),
    locations = cells_body(columns = Pct_Female)
  ) |>
  tab_footnote(footnote    = "SD = Standard Deviation") |>
  tab_source_note(source_note = "Source: ADSL; Safety Population (SAFFL = Y)") |>
  tab_options(
    table.font.size           = 10,
    column_labels.font.weight = "bold"
  ) |>
  gtsave(filename = "Table_14_1_1.rtf")

final_tbl
```

---

## Lesson Summary

| Function | Purpose |
|----------|---------|
| `gt(data)` | Create a gt table object from a data frame |
| `cols_label(col = "Label")` | Rename column headers for display |
| `cols_align(align, columns)` | Align column content left / center / right |
| `tab_header(title, subtitle)` | Add table title and subtitle |
| `tab_spanner(label, columns)` | Add a merged header spanning multiple columns |
| `tab_footnote(footnote)` | Attach a footnote to the table |
| `tab_source_note(source_note)` | Attach a source reference below the table |
| `fmt_number(columns, decimals)` | Format numeric cells with fixed decimals |
| `fmt_percent(columns, decimals)` | Format proportions (0–1) as percentages |
| `tab_style(style, locations)` | Apply conditional formatting (bold, colour, italic) |
| `cell_text(weight, color, style)` | Style specification for text properties |
| `cell_fill(color)` | Style specification for cell background fill |
| `cells_body(columns, rows)` | Location selector — data body cells |
| `cells_column_labels(columns)` | Location selector — column header cells |
| `tab_options(...)` | Set global table-wide style options |
| `gtsave(filename)` | Export to HTML, RTF, PDF, or PNG |

**Key distinctions to remember:**

- `fmt_number()` for values already expressed as percentages (e.g., `47.1`)
- `fmt_percent()` for true proportions (e.g., `0.471`) — it multiplies by 100
- `tab_style()` targets specific cells; `tab_options()` applies globally
- RTF export (`gtsave("*.rtf")`) is the standard for clinical submission tables

**Next lesson:** Lesson 32 — Complete TLF Pipeline Capstone: building a full
submission-ready TLF package (Table 14.1.1, Table 14.3.1, Table 14.4.1) using
cards → tfrmt → gt → RTF.
