---
title: "Lesson 32 — tfrmt: Display Metadata & Table Formatting"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  The complete tfrmt lesson: separate table-formatting metadata from ARD data
  with frmt(), frmt_combine(), frmt_structure(), body_plan(), col_plan(),
  row_grp_plan(), page_plan(), spanning_structure(), and tfrmt() spec objects.
  Reshape ARDs with ard_to_tfrmt(), preview with mock_df(), render with
  print_to_gt(), and polish the final display with the gt package
  (cols_label, tab_header, tab_spanner, fmt_number, tab_style, gtsave).
  Capstone ties ARD -> tfrmt -> gt -> RTF for a submission-style table.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 32_tfrmt.Rmd
# PURPOSE       : tfrmt display metadata + table formatting (consolidated):
#                 frmt(), frmt_combine(), frmt_structure(), body_plan(),
#                 col_plan(), row_grp_plan(), page_plan(), spanning_structure(),
#                 tfrmt() spec objects, ard_to_tfrmt() conversion, mock_df()
#                 shell sign-off, print_to_gt() / as_rtf() rendering, and gt
#                 display formatting (cols_label, tab_header, tab_spanner,
#                 fmt_number, tab_style, gtsave) for regulatory TLF automation.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-06-22
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

`%||%` <- function(a, b) if (!is.null(a)) a else b

# ══════════════════════════════════════════════════════════════════════════════
# REAL PHARMAVERSEADAM DATA (CDISCPILOT01)
# ══════════════════════════════════════════════════════════════════════════════
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 161, 7)), round(rnorm(n, 174, 8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 68, 12), 1), round(rnorm(n, 82, 14), 1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL = ht, WEIGHTBL = wt,
                                  BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

advs <- pharmaverseadam::advs

# ══════════════════════════════════════════════════════════════════════════════
# CARDS MOCKS — ard_continuous, ard_categorical, ard_missing, ard_stack
# (replicated so exercises run without the real cards package installed)
# ══════════════════════════════════════════════════════════════════════════════

.ard_row <- function(group1 = NA_character_, group1_level = NA_character_,
                     variable, variable_level = NA_character_,
                     context, stat_name, stat, stat_label) {
  tibble(
    group1         = group1,
    group1_level   = group1_level,
    variable       = variable,
    variable_level = variable_level,
    context        = context,
    stat_name      = stat_name,
    stat           = list(stat),
    stat_label     = stat_label
  )
}

ard_continuous <- function(data, variables, by = NULL,
                           statistics = list(
                             N      = function(x) sum(!is.na(x)),
                             mean   = function(x) mean(x, na.rm = TRUE),
                             sd     = function(x) sd(x, na.rm = TRUE),
                             median = function(x) median(x, na.rm = TRUE),
                             min    = function(x) min(x, na.rm = TRUE),
                             max    = function(x) max(x, na.rm = TRUE),
                             n_miss = function(x) sum(is.na(x)),
                             p_miss = function(x) round(100 * mean(is.na(x)), 1)
                           )) {
  stat_labels <- c(N = "N (non-missing)", mean = "Mean", sd = "SD",
                   median = "Median", min = "Min", max = "Max",
                   n_miss = "N missing", p_miss = "% missing")
  groups <- if (!is.null(by)) unique(as.character(data[[by]])) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[as.character(data[[by]]) == gl, ] else data
      x   <- sub[[vv]]
      lapply(names(statistics), function(sn) {
        val <- tryCatch(statistics[[sn]](x), error = function(e) NA_real_)
        .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                 group1_level = gl, variable = vv, context = "continuous",
                 stat_name = sn, stat = val, stat_label = stat_labels[sn])
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class = c("card", "tbl_df", "tbl", "data.frame"))
}

ard_categorical <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(as.character(data[[by]])) else NA_character_
  rows <- lapply(variables, function(vv) {
    levs <- sort(unique(as.character(data[[vv]])))
    levs <- levs[!is.na(levs)]
    lapply(groups, function(gl) {
      sub   <- if (!is.na(gl) && !is.null(by)) data[as.character(data[[by]]) == gl, ] else data
      total <- sum(!is.na(sub[[vv]]))
      lapply(levs, function(lv) {
        cnt <- sum(sub[[vv]] == lv, na.rm = TRUE)
        bind_rows(
          .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                   group1_level = gl, variable = vv, variable_level = lv,
                   context = "categorical", stat_name = "n", stat = cnt, stat_label = "n"),
          .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                   group1_level = gl, variable = vv, variable_level = lv,
                   context = "categorical", stat_name = "p",
                   stat = round(100 * cnt / max(total, 1), 1), stat_label = "%")
        )
      }) |> bind_rows()
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class = c("card", "tbl_df", "tbl", "data.frame"))
}

ard_missing <- function(data, variables, by = NULL) {
  groups <- if (!is.null(by)) unique(as.character(data[[by]])) else NA_character_
  rows <- lapply(variables, function(vv) {
    lapply(groups, function(gl) {
      sub <- if (!is.na(gl) && !is.null(by)) data[as.character(data[[by]]) == gl, ] else data
      x   <- sub[[vv]]
      bind_rows(
        .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                 group1_level = gl, variable = vv, context = "missing",
                 stat_name = "n_miss", stat = sum(is.na(x)), stat_label = "N missing"),
        .ard_row(group1 = if (!is.null(by)) by else NA_character_,
                 group1_level = gl, variable = vv, context = "missing",
                 stat_name = "p_miss", stat = round(100 * mean(is.na(x)), 1),
                 stat_label = "% missing")
      )
    }) |> bind_rows()
  }) |> bind_rows()
  structure(rows, class = c("card", "tbl_df", "tbl", "data.frame"))
}

ard_stack <- function(...) {
  out <- bind_rows(...)
  structure(out, class = c("card", "tbl_df", "tbl", "data.frame"))
}

bind_ard <- function(...) ard_stack(...)

# ══════════════════════════════════════════════════════════════════════════════
# TFRMT FUNCTION MOCKS
# (the real tfrmt package is not assumed installed; these mocks reproduce its
#  API surface so all exercises run and print sensibly)
# ══════════════════════════════════════════════════════════════════════════════

# frmt: defines how a single number is printed
frmt <- function(expression = "xx", missing = "", ...) {
  structure(list(expression = expression, missing = missing, extra = list(...)),
            class = "frmt")
}

print.frmt <- function(x, ...) {
  cat(sprintf("<frmt> expression=\"%s\"  missing=\"%s\"\n",
              x$expression, x$missing))
  invisible(x)
}

# frmt_combine: combines multiple frmts into a composite cell (e.g., "n (%)")
frmt_combine <- function(expression, ..., missing = "") {
  parts <- list(...)
  structure(list(expression = expression, parts = parts, frmts = parts,
                 missing = missing),
            class = "frmt_combine")
}

print.frmt_combine <- function(x, ...) {
  cat(sprintf("<frmt_combine> expression=\"%s\"\n", x$expression))
  invisible(x)
}

# frmt_structure: pairs frmt/frmt_combine objects with group/label selectors
frmt_structure <- function(group_val = NULL, label_val = NULL,
                            frmt_combine = NULL, ...) {
  frmts <- c(list(frmt_combine), list(...))
  frmts <- frmts[!vapply(frmts, is.null, logical(1))]
  structure(list(group_val = group_val, label_val = label_val, frmts = frmts),
            class = "frmt_structure")
}

# body_plan: collects frmt_structures that govern row display
body_plan <- function(...) {
  structures <- list(...)
  structure(list(structures = structures), class = "body_plan")
}

# Length / [[ for body_plan so length() and bp[[i]] still work as in lesson 29
length.body_plan <- function(x) length(x$structures)
`[[.body_plan`    <- function(x, i) x$structures[[i]]

# col_plan: defines column ordering / renaming / spanners
col_plan <- function(...) {
  args <- list(...)
  structure(list(columns = args), class = "col_plan")
}

# tfrmt: the master table spec object. substitute() is called directly here
# (not via a helper) so bare unquoted column names work:
#   tfrmt(param = param, ...)  or  tfrmt(param = "col", ...)
tfrmt <- function(label = NULL, param = NULL, value = NULL,
                  column = NULL, group = NULL,
                  big_n = NULL, title = NULL,
                  body_plan = NULL, col_plan = NULL,
                  row_grp_plan = NULL, page_plan = NULL,
                  footnote_plan = NULL, ...) {
  .s2c <- function(s) {
    if (is.null(s)) return(NULL)
    if (is.character(s)) return(s)
    deparse(s)
  }
  label_v  <- .s2c(substitute(label))
  param_v  <- .s2c(substitute(param))
  value_v  <- .s2c(substitute(value))
  column_v <- .s2c(substitute(column))
  group_v  <- .s2c(substitute(group))
  structure(
    list(
      label         = label_v,
      param         = param_v,
      value         = value_v,
      column        = column_v,
      group         = group_v,
      big_n         = big_n,
      title         = title,
      body_plan     = body_plan,
      col_plan      = col_plan,
      row_grp_plan  = row_grp_plan,
      page_plan     = page_plan,
      footnote_plan = footnote_plan,
      extra         = list(...)
    ),
    class = "tfrmt"
  )
}

print.tfrmt <- function(x, ...) {
  cat("<tfrmt spec>\n")
  cat(sprintf("  label  : %s\n", x$label  %||% "(not set)"))
  cat(sprintf("  param  : %s\n", x$param  %||% "(not set)"))
  cat(sprintf("  value  : %s\n", x$value  %||% "(not set)"))
  cat(sprintf("  column : %s\n", x$column %||% "(not set)"))
  cat(sprintf("  group  : %s\n", x$group  %||% "(not set)"))
  if (!is.null(x$title))
    cat(sprintf("  title  : %s\n", x$title))
  if (!is.null(x$body_plan))
    cat(sprintf("  body_plan   : %d frmt_structure(s)\n",
                length(x$body_plan$structures)))
  if (!is.null(x$col_plan))
    cat(sprintf("  col_plan    : %d column(s)/spanner(s)\n",
                length(x$col_plan$columns)))
  if (!is.null(x$row_grp_plan))
    cat(sprintf("  row_grp_plan: label_loc = %s\n", x$row_grp_plan$label_loc))
  if (!is.null(x$page_plan))
    cat(sprintf("  page_plan   : %d page_structure(s), note_loc = %s\n",
                length(x$page_plan$page_structures), x$page_plan$note_loc))
  invisible(x)
}

# big_n_structure: maps Big-N rows into column-header N counts
big_n_structure <- function(param_val, n_frmt = frmt("N = xx"), ...) {
  structure(list(param_val = param_val, n_frmt = n_frmt), class = "big_n_structure")
}

# page_plan: defines pagination for long tables (e.g., AE tables)
page_plan <- function(..., note_loc = c("above", "below")) {
  note_loc <- match.arg(note_loc)
  structure(list(page_structures = list(...), note_loc = note_loc),
            class = "page_plan")
}

print.page_plan <- function(x, ...) {
  cat(sprintf("<page_plan> %d page_structure(s), note_loc = \"%s\"\n",
              length(x$page_structures), x$note_loc))
  invisible(x)
}

page_structure <- function(group_val = NULL, ...) {
  structure(list(group_val = group_val, extra = list(...)),
            class = "page_structure")
}

print.page_structure <- function(x, ...) {
  cat(sprintf("<page_structure> group_val = %s\n",
              x$group_val %||% "(all rows on one page)"))
  invisible(x)
}

# row_grp_plan: grouping for row-level hierarchies (e.g., SOC > PT)
row_grp_plan <- function(..., label_loc = c("indented", "column", "spanning")) {
  label_loc <- match.arg(label_loc)
  structure(list(structures = list(...), label_loc = label_loc),
            class = "row_grp_plan")
}

print.row_grp_plan <- function(x, ...) {
  cat(sprintf("<row_grp_plan> %d structure(s), label_loc = \"%s\"\n",
              length(x$structures), x$label_loc))
  invisible(x)
}

row_grp_structure <- function(group_val = NULL, label_val = NULL,
                               indent = c("indented", "column"), ...) {
  structure(list(group_val = group_val, label_val = label_val, indent = indent[1]),
            class = "row_grp_structure")
}

print.row_grp_structure <- function(x, ...) {
  cat(sprintf("<row_grp_structure> group_val = %s  label_val = %s  indent = %s\n",
              x$group_val %||% "(none)", x$label_val %||% "(none)", x$indent))
  invisible(x)
}

# spanning_structure: add a spanner header above columns
spanning_structure <- function(...) {
  structure(list(...), class = "spanning_structure")
}

print.spanning_structure <- function(x, ...) {
  lbls <- names(x)
  cat(sprintf("<spanning_structure> %d spanner(s): %s\n",
              length(x), paste(lbls, collapse = ", ")))
  invisible(x)
}

# footnote_plan / footnote_structure: attach footnotes to matching cells
footnote_structure <- function(footnote, column_val = NULL, label_val = NULL, ...) {
  structure(list(footnote = footnote, column_val = column_val,
                 label_val = label_val), class = "footnote_structure")
}

footnote_plan <- function(...) {
  structure(list(...), class = "footnote_plan")
}

# tfrmt_n_pct: shortcut frmt_combine for n (%)
tfrmt_n_pct <- function(n_frmt   = frmt("xxx"),
                         pct_frmt = frmt("xx.x"),
                         missing  = "0 (---%)") {
  frmt_combine("{n} ({p}%)", n = n_frmt, p = pct_frmt, missing = missing)
}

# .apply_frmt: apply a frmt object to a single value (used in Section 1)
.apply_frmt <- function(val, frmt_obj) {
  if (inherits(frmt_obj, "frmt")) {
    if (is.na(val) || is.null(val)) return(frmt_obj$missing)
    fmt <- frmt_obj$expression
    if (grepl("\\.", fmt)) {
      dec <- nchar(sub(".*\\.", "", fmt))
      return(formatC(as.numeric(val), digits = dec, format = "f"))
    }
    return(formatC(as.numeric(val), format = "d"))
  }
  as.character(val)
}

# ard_to_tfrmt: coerces a cards ARD into tfrmt long format
# (label / label_level / column / param / value)
ard_to_tfrmt <- function(ard, label_col = "variable",
                          label_level_col = "variable_level",
                          column_col = "group1_level",
                          param_col  = "stat_name",
                          value_col  = "stat") {
  ard |>
    mutate(value = unlist(.data[[value_col]])) |>
    rename(
      label       = !!sym(label_col),
      label_level = !!sym(label_level_col),
      column      = !!sym(column_col),
      param       = !!sym(param_col)
    ) |>
    select(label, label_level, column, param, value) |>
    filter(!is.na(column))
}

# print_to_gt: renders tfrmt spec + data as a display table (returns tibble here)
print_to_gt <- function(tfrmt_spec, .data, ...) {
  if (!inherits(tfrmt_spec, "tfrmt"))
    stop("tfrmt_spec must be a tfrmt object")
  cat("[print_to_gt()] Rendering tfrmt spec against", nrow(.data), "ARD rows\n")
  keep <- intersect(c("label", "label_level", "column", "param", "value"),
                    names(.data))
  out <- .data |>
    select(all_of(keep)) |>
    head(20)
  cat("[Output preview - first 20 rows of display-ready data]\n")
  print(out)
  invisible(out)
}

# print_mock_gt: render the spec with placeholder cells (no real data)
print_mock_gt <- function(tfrmt_spec, .data = NULL, ...) {
  cat("[print_mock_gt()] Blank shell for tfrmt spec (placeholder cells)\n")
  cat("  label :", tfrmt_spec$label %||% "(not set)",
      " column:", tfrmt_spec$column %||% "(not set)", "\n")
  if (!is.null(.data)) {
    keep <- intersect(c("label", "label_level", "column"), names(.data))
    shell <- .data |> select(all_of(keep)) |> distinct()
    cat("  Shell rows:", nrow(shell), "\n")
    print(head(shell, 10))
  }
  cat("[Each data cell would show its format expression, e.g. xx.x (xx.xx)]\n")
  invisible(tfrmt_spec)
}

# mock_df: alias-style blank shell generator for biostatistician sign-off
mock_df <- function(tfrmt_spec, .data = NULL, ...) {
  cat("[mock_df()] Generating blank shell for tfrmt spec\n")
  cat("  label     :", tfrmt_spec$label  %||% "(not set)", "\n")
  cat("  param     :", tfrmt_spec$param  %||% "(not set)", "\n")
  cat("  column    :", tfrmt_spec$column %||% "(not set)", "\n")
  if (!is.null(tfrmt_spec$row_grp_plan))
    cat("  row groups:", tfrmt_spec$row_grp_plan$label_loc, "\n")
  if (!is.null(tfrmt_spec$page_plan))
    cat("  pagination: breaks on each page_structure group\n")
  if (!is.null(.data))
    cat("  ARD rows  :", nrow(.data), "\n")
  cat("[Shell would display here - all data cells show format expressions]\n")
  invisible(tfrmt_spec)
}

# as_rtf (mock): export gt object to RTF file
as_rtf <- function(x, filename, ...) {
  cat(sprintf("[as_rtf()] Would write RTF to: %s\n", filename))
  invisible(x)
}

# ══════════════════════════════════════════════════════════════════════════════
# GT MOCKS / STUBS — used if the real gt package is not installed
# ══════════════════════════════════════════════════════════════════════════════

.gt_available <- requireNamespace("gt", quietly = TRUE)

if (!.gt_available) {

  gt <- function(data, rowname_col = NULL, groupname_col = NULL, ...) {
    structure(list(data = data, opts = list(...), rowname_col = rowname_col,
                   groupname_col = groupname_col),
              class = "gt_tbl")
  }

  print.gt_tbl <- function(x, ...) {
    cat("[gt table - install the gt package to render fully]\n\n")
    print(as.data.frame(x$data), row.names = FALSE)
    invisible(x)
  }

  cols_label <- function(.data, ...) {
    args <- list(...)
    if (inherits(.data, "gt_tbl")) {
      .data$opts$col_labels <- c(.data$opts$col_labels, args)
      cat(sprintf("[cols_label()] Relabelling %d column(s): %s\n",
                  length(args), paste(names(args), collapse = ", ")))
    }
    .data
  }

  cols_align <- function(.data, align = "left", columns = everything()) {
    if (inherits(.data, "gt_tbl"))
      cat(sprintf("[cols_align()] align = \"%s\"\n", align))
    .data
  }

  tab_header <- function(.data, title = "", subtitle = NULL) {
    if (inherits(.data, "gt_tbl")) {
      .data$opts$title    <- title
      .data$opts$subtitle <- subtitle
      cat(sprintf("[tab_header()] title = \"%s\"", title))
      if (!is.null(subtitle)) cat(sprintf("  subtitle = \"%s\"", subtitle))
      cat("\n")
    }
    .data
  }

  tab_spanner <- function(.data, label = "", columns = NULL) {
    if (inherits(.data, "gt_tbl"))
      cat(sprintf("[tab_spanner()] label = \"%s\"\n", label))
    .data
  }

  tab_footnote <- function(.data, footnote = "", locations = NULL) {
    if (inherits(.data, "gt_tbl")) {
      .data$opts$footnotes <- c(.data$opts$footnotes, footnote)
      cat(sprintf("[tab_footnote()] \"%s\"\n", footnote))
    }
    .data
  }

  tab_source_note <- function(.data, source_note = "") {
    if (inherits(.data, "gt_tbl")) {
      .data$opts$source_notes <- c(.data$opts$source_notes, source_note)
      cat(sprintf("[tab_source_note()] \"%s\"\n", source_note))
    }
    .data
  }

  fmt_number <- function(.data, columns = NULL, decimals = 2, ...) {
    if (inherits(.data, "gt_tbl"))
      cat(sprintf("[fmt_number()] decimals = %d\n", decimals))
    .data
  }

  fmt_percent <- function(.data, columns = NULL, decimals = 1, ...) {
    if (inherits(.data, "gt_tbl"))
      cat(sprintf("[fmt_percent()] decimals = %d\n", decimals))
    .data
  }

  tab_style <- function(.data, style = NULL, locations = NULL) {
    if (inherits(.data, "gt_tbl"))
      cat("[tab_style()] Style rule applied\n")
    .data
  }

  tab_options <- function(.data, ...) {
    args <- list(...)
    if (inherits(.data, "gt_tbl") && length(args) > 0)
      cat(sprintf("[tab_options()] %d option(s) set: %s\n",
                  length(args), paste(names(args), collapse = ", ")))
    .data
  }

  gtsave <- function(.data, filename, ...) {
    cat(sprintf("[gtsave()] Would save to: %s\n", filename))
    invisible(.data)
  }

  cells_body <- function(columns = NULL, rows = NULL) {
    list(type = "cells_body", columns = columns, rows = rows)
  }

  cells_column_labels <- function(columns = NULL) {
    list(type = "cells_column_labels", columns = columns)
  }

  cell_text <- function(weight = NULL, color = NULL, style = NULL, ...) {
    args <- c(list(weight = weight, color = color, style = style), list(...))
    args <- args[!vapply(args, is.null, logical(1))]
    structure(args, class = "cell_text")
  }

  cell_fill <- function(color = NULL) {
    structure(list(color = color), class = "cell_fill")
  }

  everything <- function() NULL

  md <- function(text) text   # markdown passthrough

} else {
  library(gt)
}

# ══════════════════════════════════════════════════════════════════════════════
# PRE-BUILT OBJECTS USED IN EXERCISES
# ══════════════════════════════════════════════════════════════════════════════

# --- Continuous + categorical demographics ARDs ---
cont_ard <- ard_continuous(adsl, variables = c("AGE", "BMIBL"), by = "TRT01P")
cat_ard  <- ard_categorical(adsl, variables = c("SEX", "RACE"), by = "TRT01P")
demo_ard <- ard_stack(cont_ard, cat_ard)

# tfrmt-ready versions (label / label_level / column / param / value)
tfrmt_cont <- ard_to_tfrmt(cont_ard)
tfrmt_demo <- ard_to_tfrmt(demo_ard)

# A complete example tfrmt spec for a 3-arm demographics table
demo_spec <- tfrmt(
  param  = "param",
  value  = "value",
  label  = "label",
  column = "column",
  body_plan = body_plan(
    frmt_structure(label_val = "AGE",   frmt("xx.x")),
    frmt_structure(label_val = "BMIBL", frmt("xx.x")),
    frmt_structure(label_val = "SEX",
      frmt_combine("{n} ({p}%)", n = frmt("xxx"), p = frmt("xx.x"))),
    frmt_structure(label_val = "RACE",
      frmt_combine("{n} ({p}%)", n = frmt("xxx"), p = frmt("xx.x")))
  )
)

# --- Subject-level AE ARDs (avoid double-counting subjects) ---
adsl_trt <- adsl |> select(USUBJID, TRT01P)

ae_soc_subj <- adae |>
  distinct(USUBJID, AESOC) |>
  left_join(adsl_trt, by = "USUBJID")

ae_pt_subj <- adae |>
  distinct(USUBJID, AESOC, AEDECOD) |>
  left_join(adsl_trt, by = "USUBJID")

# Combined SOC categorical + continuous demographics ARD
ae_ard <- ard_stack(
  ard_categorical(ae_soc_subj, variables = "AESOC", by = "TRT01P"),
  ard_continuous(adsl, variables = c("AGE", "WEIGHTBL"), by = "TRT01P")
)

ae_tfrmt_ready <- ae_ard |>
  mutate(value = unlist(stat)) |>
  select(label = variable, label_level = variable_level,
         column = group1_level, param = stat_name, value) |>
  filter(!is.na(column))

# PT-level ARD for the capstone
ae_pt_ard <- ard_categorical(ae_pt_subj, variables = "AEDECOD", by = "TRT01P")

ae_pt_tfrmt_ready <- ae_pt_ard |>
  mutate(value = unlist(stat)) |>
  select(label = variable, label_level = variable_level,
         column = group1_level, param = stat_name, value) |>
  filter(!is.na(column))

# --- Display tibbles for the gt section ---
demog_display <- adsl |>
  group_by(TRT01P) |>
  summarise(
    N          = n(),
    Age_Mean   = round(mean(AGE, na.rm = TRUE), 1),
    Age_SD     = round(sd(AGE,   na.rm = TRUE), 1),
    Pct_Female = round(100 * mean(SEX == "F", na.rm = TRUE), 1),
    .groups    = "drop"
  )

demog_wide <- demog_display |>
  mutate(TRT01P = as.character(TRT01P))

ae_counts <- adae |>
  select(-any_of("TRT01P")) |>
  left_join(select(adsl, USUBJID, TRT01P), by = "USUBJID") |>
  group_by(TRT01P, AESOC) |>
  summarise(n_ae   = n(),
            n_subj = n_distinct(USUBJID),
            .groups = "drop") |>
  pivot_wider(names_from  = TRT01P,
              values_from = c(n_ae, n_subj),
              values_fill = 0)
```

## Overview

**tfrmt** (Table Framework, developed by GSK and contributed to pharmaverse in
2022) solves a fundamental problem in regulatory reporting: how do you keep a
table's *format* specification separate from the underlying *data*?

### The core idea

In SAS, a single PROC REPORT call bundles layout, column definitions, format
statements, and data together. Changing one arm label or one decimal-place
specification means touching every program. `tfrmt` splits these concerns
cleanly: the table layout is described as **metadata**, and any ARD with a
matching shape fills it in.

```
 ARD data (from cards / cardx)
        +
 tfrmt spec object (format metadata)
        │
        ▼
  print_to_gt()    ──►  gt table (HTML / RTF / PDF)
  print_mock_gt()  ──►  blank shell for sponsor sign-off
```

The same `tfrmt` spec applies to real study data, to synthetic mock data for
pre-approval sign-off, and to updated data at database lock — all with **zero
code changes**.

### Why tfrmt exists (beyond gtsummary)

`gtsummary` (Lesson 27) covers ~70% of standard tables — demographics, AE
summaries, simple shift tables. tfrmt earns its place for the remaining ~30%:

- **Multi-dimensional column structure** — Visit → Arm → Statistic, three deep
- **Mock-then-build workflow** — design and review the shell before data exists
- **Highly bespoke layouts** — sponsor templates that fit no preset
- **Pure separation of layout from data** — metadata describes; data fills

### tfrmt vs. SAS PROC REPORT

| SAS PROC REPORT | tfrmt |
|----------------|-------|
| Column layout in `COLUMN` statement | `col_plan()` |
| `DEFINE var / DISPLAY FORMAT=` | `frmt("xx.x")` inside `frmt_structure()` |
| Combining stats in a COMPUTE block | `frmt_combine("{n} ({p}%)", ...)` |
| Row grouping via `GROUP` / `ORDER` | `row_grp_plan()` |
| Page breaks via `BREAK` | `page_plan()` |
| Hard-coded data source | Separate ARD passed to `print_to_gt()` |

### The conceptual pieces of a tfrmt

| Component | What it specifies |
|---|---|
| `label` | The row-label column (leftmost column) |
| `column` | Column variable(s) (e.g., treatment arm) |
| `param` | The stat-name column in the ARD |
| `value` | The stat-value column in the ARD |
| `group` | Row grouping variable(s) |
| `big_n` | The Big-N row for column headers |
| `body_plan` | How statistics map to cell formats |
| `row_grp_plan` | Row group ordering, indentation, spacing |
| `col_plan` | Column ordering and spanning labels |
| `page_plan` | Pagination for long tables |
| `footnote_plan` | Footnotes with positioning |

Most are optional with sensible defaults. The minimum spec sets a few core
columns plus a `body_plan` and renders.

### ARD structure required by tfrmt

tfrmt expects a long-format data frame with these key columns:

| Column | Maps from (cards ARD) | Meaning |
|--------|-----------------------|---------|
| `label` | `variable` | Row label (e.g. "AGE") |
| `label_level` | `variable_level` | Category level (e.g. "F" for SEX) |
| `column` | `group1_level` | Column header (e.g. treatment arm) |
| `param` | `stat_name` | Which statistic (e.g. "mean", "n") |
| `value` | `unlist(stat)` | The numeric value |

The `ard_to_tfrmt()` helper performs this reshaping automatically.

> All exercises use the **CDISCPILOT01** Alzheimer's / Xanomeline study from
> `pharmaverseadam` (254 subjects, Safety Population, three treatment arms).
> Pre-built objects available in the session: `cont_ard`, `cat_ard`, `demo_ard`,
> `tfrmt_cont`, `tfrmt_demo`, `demo_spec`, `ae_tfrmt_ready`, `ae_pt_tfrmt_ready`,
> `demog_display`, `demog_wide`, `ae_counts`.

---

## Section 1 — frmt(): Format Expressions

**`frmt()`** defines how a single numeric value is displayed. The `expression`
argument is a pattern string where `x` represents a digit:

| Expression | Meaning | Example output |
|------------|---------|----------------|
| `"xx"` | Integer, 2 digits | `42` |
| `"xx.x"` | 1 decimal place | `42.3` |
| `"xx.xx"` | 2 decimal places | `42.31` |
| `"xx.xxx"` | 3 decimal places | `42.314` |

The `missing` argument controls what is displayed for `NA` values (default `""`).

### Exercise 1-a — Create and inspect a frmt object

```{r s1-frmt-create, exercise=TRUE, exercise.lines=10}
# Create a frmt with one decimal place
f1 <- frmt("xx.x")

# Inspect the structure and class
str(f1)
class(f1)
```

```{r s1-frmt-create-hint-1}
# frmt() returns a list with elements: expression and missing
# class(f1) should be "frmt"
```

```{r s1-frmt-create-solution}
f1 <- frmt("xx.x")
str(f1)
class(f1)
```

### Exercise 1-b — Apply .apply_frmt() and compare decimal places

```{r s1-frmt-apply, exercise=TRUE, exercise.lines=14}
pi_val <- 3.14159265

# One decimal place
result_1 <- .apply_frmt(pi_val, frmt("___"))

# Two decimal places
result_2 <- .apply_frmt(pi_val, frmt("___"))

# Three decimal places
result_3 <- .apply_frmt(pi_val, frmt("___"))

cat(result_1, "\n", result_2, "\n", result_3, "\n")
```

```{r s1-frmt-apply-hint-1}
# "xx.x"   -> one decimal place   -> "3.1"
# "xx.xx"  -> two decimal places  -> "3.14"
# "xx.xxx" -> three decimal places -> "3.142"
```

```{r s1-frmt-apply-solution}
pi_val <- 3.14159265

result_1 <- .apply_frmt(pi_val, frmt("xx.x"))
result_2 <- .apply_frmt(pi_val, frmt("xx.xx"))
result_3 <- .apply_frmt(pi_val, frmt("xx.xxx"))

cat(result_1, "\n", result_2, "\n", result_3, "\n")
```

### Exercise 1-c — Handle missing values with the missing argument

```{r s1-frmt-missing, exercise=TRUE, exercise.lines=10}
# Default missing: empty string
.apply_frmt(NA_real_, frmt("xx.x"))

# Custom missing string
.apply_frmt(NA_real_, frmt("xx.x", missing = "NC"))

# Real values still format normally
.apply_frmt(74.2, frmt("xx.x", missing = "NC"))
```

```{r s1-frmt-missing-hint-1}
# missing = "NC" is common in regulatory tables (Not Calculable)
# Real values are unaffected by the missing argument
```

```{r s1-frmt-missing-solution}
.apply_frmt(NA_real_, frmt("xx.x"))
.apply_frmt(NA_real_, frmt("xx.x", missing = "NC"))
.apply_frmt(74.2, frmt("xx.x", missing = "NC"))
```

```{r s1-quiz, echo=FALSE}
quiz(
  question(
    "In tfrmt, what does frmt('xx.x') control?",
    answer("The column width in the output table"),
    answer("The number of decimal places and display format for a single statistic",
           correct = TRUE),
    answer("The row grouping level of a variable"),
    answer("The column header text"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — frmt_combine(): Multi-Stat Cells

Most clinical table cells contain multiple statistics in one string:
`"42 (16.7%)"`, `"72.3 (8.1)"`, `"0.034 (0.012, 0.098)"`. **`frmt_combine()`**
assembles these from individual `frmt()` objects using a glue-style
`expression` with named placeholders in curly braces.

```
frmt_combine(
  expression,     # glue string: "{n} ({p}%)"
  ...,            # named frmt objects matching placeholders
  missing = ""
)
```

The format-string mini-language is more verbose than a typical R format string
but is unambiguous: a reviewer can read the body_plan and know exactly how each
cell displays.

### Exercise 2-a — Create a frmt_combine for n (p%)

```{r s2-combine-np, exercise=TRUE, exercise.lines=10}
# Standard clinical table cell: "42 (16.7%)"
np_fmt <- frmt_combine(
  "___",                  # expression with {n} and {p} placeholders
  n = frmt("___"),        # integer count
  p = frmt("___")         # one decimal percentage
)

str(np_fmt)
```

```{r s2-combine-np-hint-1}
# expression = "{n} ({p}%)"
# n = frmt("xxx")  -- integer count (up to 3 digits)
# p = frmt("xx.x") -- one decimal place
```

```{r s2-combine-np-solution}
np_fmt <- frmt_combine(
  "{n} ({p}%)",
  n = frmt("xxx"),
  p = frmt("xx.x")
)

str(np_fmt)
```

### Exercise 2-b — Create a frmt_combine for mean (SD)

```{r s2-combine-mean-sd, exercise=TRUE, exercise.lines=10}
# Cell format: "72.3 (8.1)"
mean_sd_fmt <- frmt_combine(
  "___",
  mean = frmt("___"),
  sd   = frmt("___")
)

class(mean_sd_fmt)
length(mean_sd_fmt$frmts)   # should be 2 (mean and sd)
```

```{r s2-combine-mean-sd-hint-1}
# expression = "{mean} ({sd})"
# mean = frmt("xx.x"), sd = frmt("xx.x")
```

```{r s2-combine-mean-sd-solution}
mean_sd_fmt <- frmt_combine(
  "{mean} ({sd})",
  mean = frmt("xx.x"),
  sd   = frmt("xx.x")
)

class(mean_sd_fmt)
length(mean_sd_fmt$frmts)
```

```{r s2-quiz, echo=FALSE}
quiz(
  question(
    "In frmt_combine('{n} ({p}%)', n=frmt('xxx'), p=frmt('xx.x')), what do the curly braces around n and p signify?",
    answer("R environment variable lookup"),
    answer("Named placeholder slots filled by the corresponding frmt objects",
           correct = TRUE),
    answer("Optional arguments that can be omitted"),
    answer("Regular expression capture groups"),
    allow_retry = TRUE
  )
)
```

---

## Section 3 — frmt_structure(): Row-Level Formatting Rules

**`frmt_structure()`** attaches a formatting rule to specific rows, identified
by `group_val` (row group) and/or `label_val` (row label). The special value
`".default"` matches any group/label not matched by an earlier rule.

```
frmt_structure(
  group_val = NULL,    # match on row group value (or ".default")
  label_val = NULL,    # match on row label value (or ".default")
  ...                  # one or more frmt / frmt_combine objects
)
```

### Exercise 3-a — Attach frmt to the AGE row

```{r s3-struct-age, exercise=TRUE, exercise.lines=10}
# Format the AGE row to one decimal place
age_fmt_struct <- frmt_structure(
  label_val = "___",   # row label in the output table
  ___                  # frmt object
)

class(age_fmt_struct)
age_fmt_struct$label_val
```

```{r s3-struct-age-hint-1}
# label_val = "AGE"
# Pass frmt("xx.x") as the second argument
```

```{r s3-struct-age-solution}
age_fmt_struct <- frmt_structure(
  label_val = "AGE",
  frmt("xx.x")
)

class(age_fmt_struct)
age_fmt_struct$label_val
```

### Exercise 3-b — Attach frmt_combine to the SEX row

```{r s3-struct-sex, exercise=TRUE, exercise.lines=12}
# SEX row should display "n (p%)"
sex_fmt_struct <- frmt_structure(
  label_val = "___",
  frmt_combine(
    "___",
    n = frmt("___"),
    p = frmt("___")
  )
)

str(sex_fmt_struct)
```

```{r s3-struct-sex-hint-1}
# label_val = "SEX", expression = "{n} ({p}%)"
# n = frmt("xxx"), p = frmt("xx.x")
```

```{r s3-struct-sex-solution}
sex_fmt_struct <- frmt_structure(
  label_val = "SEX",
  frmt_combine(
    "{n} ({p}%)",
    n = frmt("xxx"),
    p = frmt("xx.x")
  )
)

str(sex_fmt_struct)
```

---

## Section 4 — body_plan(): Assembling the Formatting Plan

**`body_plan()`** collects multiple `frmt_structure()` calls into a single
formatting plan. Rules are applied top-down; the first matching rule wins, so
place specific `label_val` rules before any `".default"` catch-all.

### Exercise 4-a — Build a body_plan for a demographics table

```{r s4-body-build, exercise=TRUE, exercise.lines=20}
demo_body <- body_plan(
  # AGE: mean to 1 decimal place
  frmt_structure(label_val = "AGE",   frmt("xx.x")),

  # BMIBL: mean to 1 decimal place
  frmt_structure(label_val = "BMIBL", frmt("xx.x")),

  # WEIGHTBL: mean to 1 decimal place
  frmt_structure(label_val = "WEIGHTBL", frmt("___")),

  # SEX: n (p%)
  frmt_structure(label_val = "SEX",
    frmt_combine("___", n = frmt("___"), p = frmt("___"))),

  # RACE: n (p%)
  frmt_structure(label_val = "RACE",
    frmt_combine("{n} ({p}%)", n = frmt("xxx"), p = frmt("xx.x")))
)

class(demo_body)
length(demo_body)   # should be 5
```

```{r s4-body-build-hint-1}
# WEIGHTBL: frmt("xx.x")
# SEX: frmt_combine("{n} ({p}%)", n = frmt("xxx"), p = frmt("xx.x"))
```

```{r s4-body-build-solution}
demo_body <- body_plan(
  frmt_structure(label_val = "AGE",      frmt("xx.x")),
  frmt_structure(label_val = "BMIBL",    frmt("xx.x")),
  frmt_structure(label_val = "WEIGHTBL", frmt("xx.x")),
  frmt_structure(label_val = "SEX",
    frmt_combine("{n} ({p}%)", n = frmt("xxx"), p = frmt("xx.x"))),
  frmt_structure(label_val = "RACE",
    frmt_combine("{n} ({p}%)", n = frmt("xxx"), p = frmt("xx.x")))
)

class(demo_body)
length(demo_body)
```

### Exercise 4-b — Inspect a frmt_structure within the body_plan

```{r s4-body-inspect, exercise=TRUE, exercise.lines=10}
bp <- body_plan(
  frmt_structure(label_val = "AGE",   frmt("xx.x")),
  frmt_structure(label_val = "BMIBL", frmt("xx.x"))
)

# Access the second frmt_structure and confirm its label_val
bp[[___]]$label_val
```

```{r s4-body-inspect-hint-1}
# body_plan behaves like a list; bp[[2]] gets the second frmt_structure
```

```{r s4-body-inspect-solution}
bp <- body_plan(
  frmt_structure(label_val = "AGE",   frmt("xx.x")),
  frmt_structure(label_val = "BMIBL", frmt("xx.x"))
)

bp[[2]]$label_val
```

---

## Section 5 — tfrmt(): The Spec Object

The **`tfrmt()`** function assembles all pieces — column mapping, body plan,
column ordering — into one spec object that can be saved, version-controlled,
and reused. Pass bare unquoted ARD column names (or strings); both work.

```
tfrmt(
  param    = param,         # ARD column holding stat_name
  value    = value,         # ARD column holding numeric value
  label    = label,         # ARD column holding row label
  column   = column,        # ARD column holding treatment arm
  body_plan = body_plan(...),
  col_plan  = col_plan(...),
  row_grp_plan = NULL,
  page_plan    = NULL
)
```

### Exercise 5-a — Build a minimal 3-arm demographics tfrmt spec

```{r s5-tfrmt-build, exercise=TRUE, exercise.lines=24}
my_spec <- tfrmt(
  param  = ___,      # unquoted column name
  value  = ___,
  label  = ___,
  column = ___,
  body_plan = body_plan(
    frmt_structure(label_val = "AGE",   frmt("xx.x")),
    frmt_structure(label_val = "BMIBL", frmt("xx.x")),
    frmt_structure(label_val = "SEX",
                   frmt_combine("{n} ({p}%)",
                                n = frmt("xxx"), p = frmt("xx.x"))),
    frmt_structure(label_val = "RACE",
                   frmt_combine("{n} ({p}%)",
                                n = frmt("xxx"), p = frmt("xx.x")))
  )
)

class(my_spec)
print(my_spec)
```

```{r s5-tfrmt-build-hint-1}
# param = param, value = value, label = label, column = column
# These are the unquoted column names that ard_to_tfrmt() creates
```

```{r s5-tfrmt-build-solution}
my_spec <- tfrmt(
  param  = param,
  value  = value,
  label  = label,
  column = column,
  body_plan = body_plan(
    frmt_structure(label_val = "AGE",   frmt("xx.x")),
    frmt_structure(label_val = "BMIBL", frmt("xx.x")),
    frmt_structure(label_val = "SEX",
                   frmt_combine("{n} ({p}%)",
                                n = frmt("xxx"), p = frmt("xx.x"))),
    frmt_structure(label_val = "RACE",
                   frmt_combine("{n} ({p}%)",
                                n = frmt("xxx"), p = frmt("xx.x")))
  )
)

class(my_spec)
print(my_spec)
```

### Exercise 5-b — Add a col_plan to control column order

```{r s5-tfrmt-colplan, exercise=TRUE, exercise.lines=16}
ordered_spec <- tfrmt(
  param    = param,
  value    = value,
  label    = label,
  column   = column,
  col_plan = col_plan(
    "label",
    "Placebo",
    "Xanomeline Low Dose",
    "Xanomeline High Dose"
  ),
  body_plan = body_plan(
    frmt_structure(label_val = "AGE", frmt("xx.x"))
  )
)

# Inspect the col_plan slot
ordered_spec$col_plan
```

```{r s5-tfrmt-colplan-hint-1}
# col_plan() takes column names in display order
# "label" is the row-label column; arms follow in the preferred order
```

```{r s5-tfrmt-colplan-solution}
ordered_spec <- tfrmt(
  param    = param,
  value    = value,
  label    = label,
  column   = column,
  col_plan = col_plan(
    "label",
    "Placebo",
    "Xanomeline Low Dose",
    "Xanomeline High Dose"
  ),
  body_plan = body_plan(
    frmt_structure(label_val = "AGE", frmt("xx.x"))
  )
)

ordered_spec$col_plan
```

```{r s5-quiz, echo=FALSE}
quiz(
  question(
    "What is the main advantage of assembling layout into a saved tfrmt spec object?",
    answer("It runs faster than PROC REPORT"),
    answer("The same spec renders mock and real tables identically; only the ARD changes between database cut and lock",
           correct = TRUE),
    answer("It validates the ARD against CDISC SDTM rules"),
    answer("It removes the need for an ARD entirely"),
    allow_retry = TRUE
  )
)
```

---

## Section 6 — ard_to_tfrmt(): Converting an ARD

A cards ARD has a list-column `stat` and columns `variable`, `variable_level`,
`group1_level`, `stat_name`. tfrmt needs a flat data frame with `label`,
`label_level`, `column`, `param`, `value`. **`ard_to_tfrmt()`** performs this
reshaping in one step.

> **Important:** `ard_to_tfrmt()` drops rows where `column` (`group1_level`) is
> `NA` — i.e., unstratified rows. Always pass a `by`-stratified ARD.

### Exercise 6-a — Convert a continuous ARD to tfrmt format

```{r s6-convert-cont, exercise=TRUE, exercise.lines=12}
# Build the ARD stratified by treatment arm
age_ard <- ard_continuous(
  adsl,
  variables = c("AGE", "BMIBL"),
  by        = "___"
)

# Convert to tfrmt format
age_tfrmt <- ard_to_tfrmt(age_ard)

glimpse(age_tfrmt)
```

```{r s6-convert-cont-hint-1}
# by = "TRT01P"
# Result columns: label, label_level, column, param, value
```

```{r s6-convert-cont-solution}
age_ard <- ard_continuous(
  adsl,
  variables = c("AGE", "BMIBL"),
  by        = "TRT01P"
)

age_tfrmt <- ard_to_tfrmt(age_ard)

glimpse(age_tfrmt)
```

### Exercise 6-b — Inspect the converted ARD

Use the pre-built `tfrmt_cont`. Filter to `label == "AGE"` and `param == "mean"`
to verify the three arm means appear as separate rows with different `column`
values.

```{r s6-convert-inspect, exercise=TRUE, exercise.lines=10}
tfrmt_cont |>
  filter(label == "___", param == "___") |>
  select(label, column, param, value)
```

```{r s6-convert-inspect-hint-1}
# label == "AGE", param == "mean"
# You should see three rows, one per treatment arm
```

```{r s6-convert-inspect-solution}
tfrmt_cont |>
  filter(label == "AGE", param == "mean") |>
  select(label, column, param, value)
```

> **The cards `prep_*` family:** The real tfrmt/cards stack also ships
> `shuffle_card()` (pivots `group1`/`group1_level` into actual columns),
> `prep_big_n()` (extracts the per-arm subject N for column headers),
> `prep_combine_vars()` (consolidates several variable names into one column),
> and `prep_label()` (builds the leftmost row-label column). Our
> `ard_to_tfrmt()` mock collapses that pipeline into a single reshape so the
> exercises stay focused on the tfrmt spec itself.

---

## Section 7 — print_to_gt(), print_mock_gt() and mock_df()

**`print_to_gt(spec, .data)`** applies the spec to a tfrmt-ready data frame and
renders a gt table (in a real session) or a console preview (in this tutorial).

**`print_mock_gt(spec)`** / **`mock_df(spec, .data)`** produce a **blank shell**
showing structure without values — the artifact circulated for sponsor sign-off
before database lock. In SAS workflows this is the manually drawn Word table
shell; tfrmt generates it directly from the spec, guaranteeing the mock and the
final table share one layout.

### Exercise 7-a — Render the continuous ARD with print_to_gt()

```{r s7-print, exercise=TRUE, exercise.lines=12}
# Use pre-built tfrmt_cont and demo_spec.
# Filter to mean rows only for a clean render.
mean_only <- tfrmt_cont |>
  filter(param == "___")

print_to_gt(demo_spec, .data = mean_only)
```

```{r s7-print-hint-1}
# param == "mean" -- keep only the mean rows
```

```{r s7-print-solution}
mean_only <- tfrmt_cont |>
  filter(param == "mean")

print_to_gt(demo_spec, .data = mean_only)
```

### Exercise 7-b — Preview the table shell

```{r s7-mock, exercise=TRUE, exercise.lines=8}
# Generate a blank sign-off shell from the same spec
mock_df(demo_spec, .data = tfrmt_cont)
```

```{r s7-mock-hint-1}
# mock_df() shows distinct label x column combinations without values
```

```{r s7-mock-solution}
mock_df(demo_spec, .data = tfrmt_cont)
```

```{r s7-quiz, echo=FALSE}
quiz(
  question(
    "What is the purpose of the mock (print_mock_gt() / mock_df()) in a TLF workflow?",
    answer("It generates fake data to test the pipeline before real data arrives"),
    answer("It produces a blank table shell showing row and column structure for sponsor sign-off before database lock",
           correct = TRUE),
    answer("It exports the tfrmt spec to a PDF document"),
    answer("It checks whether the ARD conforms to CDISC standards"),
    allow_retry = TRUE
  )
)
```

---

## Section 8 — row_grp_plan(): Row Group Hierarchies

In an AE summary table the SOC (System Organ Class) is a **group header row**
and the PT (Preferred Term) rows are **indented beneath it**.
**`row_grp_plan()`** controls this hierarchy.

```
NERVOUS SYSTEM DISORDERS          ← SOC header (group row)
  Headache         12  14  11     ← PT row, indented
  Dizziness         7   9   6
GASTROINTESTINAL DISORDERS
  Nausea           10  12  15
```

Key arguments:

- `row_grp_plan(row_grp_structure(...), ..., label_loc = "indented")` — outer
  wrapper. `label_loc = "indented"` keeps group labels in the same column,
  indented; `"column"` puts them in a separate first column; `"spanning"` makes
  them a header row across the table.
- `row_grp_structure(group_val, label_val, indent)` — one structure per level.
  `group_val` names the ARD column holding the group identifier; `label_val`
  names the column holding the display label.

### Exercise 8-a — Build a SOC > PT row_grp_plan

```{r s8-rgp-build, exercise=TRUE, exercise.lines=14}
# group_val = "AESOC"  (system organ class — the group header)
# label_val = "AEDECOD"  (preferred term — the indented row label)
rgp <- row_grp_plan(
  row_grp_structure(
    group_val = ___,
    label_val = ___,
    indent    = "indented"
  ),
  label_loc = ___
)

print(rgp)
```

```{r s8-rgp-build-hint-1}
# group_val = "AESOC", label_val = "AEDECOD", label_loc = "indented"
```

```{r s8-rgp-build-solution}
rgp <- row_grp_plan(
  row_grp_structure(
    group_val = "AESOC",
    label_val = "AEDECOD",
    indent    = "indented"
  ),
  label_loc = "indented"
)

print(rgp)
```

### Exercise 8-b — Inspect the row_grp_plan structure

```{r s8-rgp-inspect, exercise=TRUE, exercise.lines=10}
rgp <- row_grp_plan(
  row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                    indent = "indented"),
  label_loc = "indented"
)

str(___)
```

```{r s8-rgp-inspect-hint-1}
# str(rgp)
```

```{r s8-rgp-inspect-solution}
rgp <- row_grp_plan(
  row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                    indent = "indented"),
  label_loc = "indented"
)

str(rgp)
```

```{r s8-quiz, echo=FALSE}
quiz(
  question(
    "What does label_loc = \"indented\" do in row_grp_plan()?",
    answer("Puts group labels in a separate leftmost column"),
    answer("Indents group labels within the existing label column, keeping a single label column",
           correct = TRUE),
    answer("Applies bold formatting to group header rows"),
    answer("Hides the group label entirely"),
    allow_retry = TRUE
  ),
  question(
    "In a SOC > PT hierarchy, which argument holds the display text for the indented PT rows?",
    answer("group_val", message = "group_val identifies the grouping variable (SOC level)."),
    answer("label_val", correct = TRUE,
           message = "label_val specifies the column whose values become the indented row labels (here AEDECOD)."),
    answer("indent"),
    answer("label_loc"),
    allow_retry = TRUE
  )
)
```

---

## Section 9 — page_plan(): Pagination for Long Tables

AE tables can run to many pages. **`page_plan()`** tells tfrmt where to insert
page breaks so each SOC section starts on a new page; `note_loc` controls
whether footnotes appear above or below each page.

```r
page_plan(
  page_structure(group_val = "AESOC"),
  note_loc = "below"
)
```

Each `page_structure()` identifies a grouping value; tfrmt inserts a page break
whenever that group changes.

### Exercise 9-a — Build a page_plan

```{r s9-pp-build, exercise=TRUE, exercise.lines=12}
# Break on AESOC, place notes below each page.
pp <- page_plan(
  page_structure(group_val = ___),
  note_loc = ___
)

print(pp)
```

```{r s9-pp-build-hint-1}
# group_val = "AESOC", note_loc = "below"
```

```{r s9-pp-build-solution}
pp <- page_plan(
  page_structure(group_val = "AESOC"),
  note_loc = "below"
)

print(pp)
```

### Exercise 9-b — Combine page_plan with a tfrmt spec

```{r s9-pp-tfrmt, exercise=TRUE, exercise.lines=22}
# Use ae_tfrmt_ready as the data argument.
pp <- page_plan(
  page_structure(group_val = "AESOC"),
  note_loc = "below"
)

spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default",
                   frmt_combine("{n} ({p}%)",
                                n = frmt("xxx"), p = frmt("xx.x")))
  ),
  page_plan = ___
)

mock_df(spec, ae_tfrmt_ready)
```

```{r s9-pp-tfrmt-hint-1}
# page_plan = pp
```

```{r s9-pp-tfrmt-solution}
pp <- page_plan(
  page_structure(group_val = "AESOC"),
  note_loc = "below"
)

spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default",
                   frmt_combine("{n} ({p}%)",
                                n = frmt("xxx"), p = frmt("xx.x")))
  ),
  page_plan = pp
)

mock_df(spec, ae_tfrmt_ready)
```

---

## Section 10 — tfrmt_n_pct(): the n (%) Shortcut

Nearly every count cell in a clinical table displays as **n (x.x%)**. Writing
`frmt_combine("{n} ({p}%)", n = frmt("xxx"), p = frmt("xx.x"))` by hand is
verbose. **`tfrmt_n_pct()`** is a convenience wrapper producing the same object
with sensible defaults (including `missing = "0 (---%)"`).

### Exercise 10-a — Compare tfrmt_n_pct() to manual frmt_combine

```{r s10-npct-compare, exercise=TRUE, exercise.lines=16}
manual <- frmt_combine(
  "{n} ({p}%)",
  n = frmt("xxx"),
  p = frmt("xx.x"),
  missing = "0 (---%)"
)

shortcut <- tfrmt_n_pct()

print(manual)
print(shortcut)
identical(manual$expression, shortcut$expression)
```

```{r s10-npct-compare-hint-1}
# Both print the same expression: "{n} ({p}%)"
# identical() returns TRUE if the expressions match
```

```{r s10-npct-compare-solution}
manual <- frmt_combine(
  "{n} ({p}%)",
  n = frmt("xxx"),
  p = frmt("xx.x"),
  missing = "0 (---%)"
)

shortcut <- tfrmt_n_pct()

print(manual)
print(shortcut)
identical(manual$expression, shortcut$expression)
```

### Exercise 10-b — Customise the shortcut

```{r s10-npct-custom, exercise=TRUE, exercise.lines=12}
# Format n as 4-digit counts ("xxxx") and percentage to 2 decimals ("xx.xx").
npct_wide <- tfrmt_n_pct(
  n_frmt   = frmt(___),
  pct_frmt = frmt(___)
)

print(npct_wide)
```

```{r s10-npct-custom-hint-1}
# n_frmt = frmt("xxxx"), pct_frmt = frmt("xx.xx")
```

```{r s10-npct-custom-solution}
npct_wide <- tfrmt_n_pct(
  n_frmt   = frmt("xxxx"),
  pct_frmt = frmt("xx.xx")
)

print(npct_wide)
```

```{r s10-quiz, echo=FALSE}
quiz(
  question(
    "What does the missing argument in tfrmt_n_pct() control?",
    answer("The label displayed when a row has no subjects (e.g. n = 0)",
           correct = TRUE,
           message = "missing = \"0 (---%)\" is the default, shown when n = 0 or the statistic is NA."),
    answer("Whether to omit rows entirely when n = 0"),
    answer("The placeholder text shown in the mock shell"),
    answer("The number of decimal places for missing-data imputations"),
    allow_retry = TRUE
  )
)
```

---

## Section 11 — spanning_structure(): Column Spanners

A **column spanner** is a merged header cell above a group of columns. In a lab
table with visits Week 2/4/8 under each treatment arm, you would span the visit
columns under each arm label. **`spanning_structure()`** is passed inside the
`col_plan()` of a `tfrmt()` spec: name the spanner label, give a vector of the
columns beneath it.

> In the real package, `col_plan()` also supports the `>` operator —
> `"Week 24" > c("Placebo", "Active")` — to nest arms under visit spanners. The
> mock here uses the equivalent `spanning_structure()` form.

### Exercise 11-a — Build a single spanner

```{r s11-span-build, exercise=TRUE, exercise.lines=20}
# Three visit columns "W02","W04","W08" under one spanner "Treatment Visits".
ss <- spanning_structure(
  "Treatment Visits" = c("W02", "W04", "W08")
)

print(ss)

spec_span <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  col_plan  = col_plan(___),
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xx.x"))
  )
)

print(spec_span)
```

```{r s11-span-build-hint-1}
# col_plan(spanning_structure("Treatment Visits" = c("W02","W04","W08")))
```

```{r s11-span-build-solution}
ss <- spanning_structure(
  "Treatment Visits" = c("W02", "W04", "W08")
)

print(ss)

spec_span <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  col_plan  = col_plan(spanning_structure("Treatment Visits" = c("W02", "W04", "W08"))),
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xx.x"))
  )
)

print(spec_span)
```

### Exercise 11-b — Two spanners (one per arm)

```{r s11-span-multi, exercise=TRUE, exercise.lines=18}
# Active visits: "XLD_W02","XLD_W04" ; Placebo visits: "PBO_W02","PBO_W04"
spec_two_span <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  col_plan  = col_plan(
    spanning_structure(___ = c("XLD_W02", "XLD_W04")),
    spanning_structure(___ = c("PBO_W02", "PBO_W04"))
  ),
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xx.x"))
  )
)

print(spec_two_span)
```

```{r s11-span-multi-hint-1}
# spanning_structure("Xanomeline Low Dose" = c("XLD_W02","XLD_W04"))
# spanning_structure("Placebo" = c("PBO_W02","PBO_W04"))
```

```{r s11-span-multi-solution}
spec_two_span <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  col_plan  = col_plan(
    spanning_structure("Xanomeline Low Dose" = c("XLD_W02", "XLD_W04")),
    spanning_structure("Placebo" = c("PBO_W02", "PBO_W04"))
  ),
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xx.x"))
  )
)

print(spec_two_span)
```

---

## Section 12 — Clinical Table Templates

Real submissions follow recurring patterns. Rather than building every spec from
scratch, clinical programmers maintain **template specs** that are parameterised
and reused across studies. Here you build three standard templates.

### Exercise 12-a — Demographics template

```{r s12-tmpl-demog, exercise=TRUE, exercise.lines=24}
# Rows: AGE (mean (SD)), all others n (%).
# label_val = "AGE"      -> frmt_combine("{m} ({s})", m=frmt("xx.x"), s=frmt("xx.xx"))
# label_val = ".default" -> tfrmt_n_pct()
demog_spec_tmpl <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(
      label_val = ___,
      frmt_combine("{m} ({s})", m = frmt("xx.x"), s = frmt("xx.xx"))
    ),
    frmt_structure(
      label_val = ___,
      ___
    )
  )
)

print(demog_spec_tmpl)
```

```{r s12-tmpl-demog-hint-1}
# label_val = "AGE" for the mean (SD) format
# label_val = ".default" with tfrmt_n_pct() for everything else
```

```{r s12-tmpl-demog-solution}
demog_spec_tmpl <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(
      label_val = "AGE",
      frmt_combine("{m} ({s})", m = frmt("xx.x"), s = frmt("xx.xx"))
    ),
    frmt_structure(
      label_val = ".default",
      tfrmt_n_pct()
    )
  )
)

print(demog_spec_tmpl)
```

### Exercise 12-b — AE summary template

```{r s12-tmpl-ae, exercise=TRUE, exercise.lines=24}
# All cells tfrmt_n_pct(); SOC > PT row grouping; then render.
ae_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(
      group_val = ___,
      label_val = ___,
      indent    = "indented"
    ),
    label_loc = "indented"
  )
)

print_to_gt(ae_spec, ___)
```

```{r s12-tmpl-ae-hint-1}
# group_val = "AESOC", label_val = "AEDECOD"
# print_to_gt(ae_spec, ae_tfrmt_ready)
```

```{r s12-tmpl-ae-solution}
ae_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(
      group_val = "AESOC",
      label_val = "AEDECOD",
      indent    = "indented"
    ),
    label_loc = "indented"
  )
)

print_to_gt(ae_spec, ae_tfrmt_ready)
```

### Exercise 12-c — Lab shift template

```{r s12-tmpl-lab, exercise=TRUE, exercise.lines=28}
# Baseline (BTOXGR) x worst post-baseline (ATOXGR) ALT grade counts.
shift_counts <- adlb |>
  filter(!is.na(BTOXGR), !is.na(ATOXGR), PARAMCD == "ALT") |>
  mutate(
    BTOXGR_label = paste0("Baseline Grade ", BTOXGR),
    ATOXGR_label = paste0("Post Grade ",     ATOXGR)
  ) |>
  group_by(BTOXGR_label, ATOXGR_label) |>
  summarise(value = n_distinct(USUBJID), .groups = "drop") |>
  mutate(param = "n", label = BTOXGR_label, column = ATOXGR_label) |>
  select(label, column, param, value)

shift_spec <- tfrmt(
  label     = ___,
  param     = ___,
  value     = ___,
  column    = ___,
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xxx"))
  )
)

print_to_gt(shift_spec, shift_counts)
```

```{r s12-tmpl-lab-hint-1}
# label = "label", param = "param", value = "value", column = "column"
```

```{r s12-tmpl-lab-solution}
shift_counts <- adlb |>
  filter(!is.na(BTOXGR), !is.na(ATOXGR), PARAMCD == "ALT") |>
  mutate(
    BTOXGR_label = paste0("Baseline Grade ", BTOXGR),
    ATOXGR_label = paste0("Post Grade ",     ATOXGR)
  ) |>
  group_by(BTOXGR_label, ATOXGR_label) |>
  summarise(value = n_distinct(USUBJID), .groups = "drop") |>
  mutate(param = "n", label = BTOXGR_label, column = ATOXGR_label) |>
  select(label, column, param, value)

shift_spec <- tfrmt(
  label     = "label",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", frmt("xxx"))
  )
)

print_to_gt(shift_spec, shift_counts)
```

---

## Section 13 — The mock workflow & shell sign-off

The "mock first" workflow is signature pharma:

```
1. Statistician designs the table structure (rows, columns, statistics)
2. Programmer builds a tfrmt object encoding the design
3. Programmer renders mock_df() / print_mock_gt() — an empty-shell version
4. Sponsor reviews and approves the layout
5. Programmer builds the ARD via cards/cardx
6. Programmer renders print_to_gt(ard) — same tfrmt, now filled with real values
```

The tfrmt object is the single source of truth for layout. Layout review happens
once; data changes don't break the layout; mock and real are guaranteed to match
in structure. In a gtsummary workflow, layout is intermingled with the data, so
producing a faithful mock without dummy data is awkward.

### Exercise 13-a — Generate a full AE shell

```{r s13-mock-shell, exercise=TRUE, exercise.lines=22}
ae_tfrmt_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                      indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

mock_df(___, ___)
```

```{r s13-mock-shell-hint-1}
# mock_df(ae_tfrmt_spec, ae_tfrmt_ready)
```

```{r s13-mock-shell-solution}
ae_tfrmt_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                      indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

mock_df(ae_tfrmt_spec, ae_tfrmt_ready)
```

### Exercise 13-b — Review the shell row labels

```{r s13-mock-review, exercise=TRUE, exercise.lines=8}
# A reviewer checks: are all expected row labels present?
# How many unique label_level values (row labels) exist in ae_tfrmt_ready?
ae_tfrmt_ready |>
  ___(label_level) |>
  nrow()
```

```{r s13-mock-review-hint-1}
# distinct(label_level) then nrow()
```

```{r s13-mock-review-solution}
ae_tfrmt_ready |>
  distinct(label_level) |>
  nrow()
```

---

## Section 14 — gt: Polishing the Final Display

tfrmt renders **through** `gt` (the *grammar of tables*). Once `print_to_gt()`
hands you a gt object — or whenever tfrmt's layout cannot express the exact
cosmetic detail your SAP requires — you reach for gt directly to control titles,
spanners, footnotes, number formatting, conditional styling, and file export.

gt is also the engine under `gtsummary`, so the same verbs apply across the
modern TLG stack. Every gt pipeline starts with a data frame and chains
pipe-composable verbs:

```
data frame
  |> gt()                   # create gt table object
  |> cols_label()           # rename column headers
  |> tab_header()           # add title and subtitle
  |> tab_spanner()          # group columns under a shared header
  |> tab_footnote()         # add footnotes
  |> tab_source_note()      # add source reference
  |> fmt_number()           # format numeric cells
  |> tab_style()            # conditional cell/text formatting
  |> tab_options()          # global style options
  |> gtsave("output.rtf")   # export to file
```

Two pre-built display tibbles are available: `demog_display` and `demog_wide`
(N, mean age, SD age, % female by arm) and `ae_counts` (AE counts by SOC).

### Exercise 14-a — gt() and cols_label()

```{r s14-gt-labels, exercise=TRUE, exercise.lines=16}
# Create a gt table and rename columns to human-readable labels.
demog_display |>
  gt() |>
  cols_label(
    TRT01P     = ___,
    N          = ___,
    Age_Mean   = ___,
    Age_SD     = ___,
    Pct_Female = ___
  )
```

```{r s14-gt-labels-hint-1}
# cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
#            Age_SD = "SD", Pct_Female = "% Female")
# Note: column name on the LEFT, new label string on the RIGHT.
```

```{r s14-gt-labels-solution}
demog_display |>
  gt() |>
  cols_label(
    TRT01P     = "Treatment",
    N          = "N",
    Age_Mean   = "Mean Age",
    Age_SD     = "SD",
    Pct_Female = "% Female"
  )
```

### Exercise 14-b — Title, footnote, and source note

```{r s14-gt-header, exercise=TRUE, exercise.lines=18}
# Add a title/subtitle, a footnote for SD, and a source note.
demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = ___,
    subtitle = "Summary of Demographics (Safety Population)"
  ) |>
  tab_footnote(footnote    = ___) |>
  tab_source_note(source_note = ___)
```

```{r s14-gt-header-hint-1}
# title = "Table 14.1.1"
# tab_footnote(footnote = "SD = Standard Deviation")
# tab_source_note(source_note = "Source: ADSL; Safety Population (SAFFL = Y)")
```

```{r s14-gt-header-solution}
demog_display |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_header(
    title    = "Table 14.1.1",
    subtitle = "Summary of Demographics (Safety Population)"
  ) |>
  tab_footnote(footnote    = "SD = Standard Deviation") |>
  tab_source_note(source_note = "Source: ADSL; Safety Population (SAFFL = Y)")
```

### Exercise 14-c — tab_spanner(): grouping columns

```{r s14-gt-spanner, exercise=TRUE, exercise.lines=18}
# Add an "Age Statistics" spanner above Age_Mean and Age_SD.
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_spanner(
    label   = ___,
    columns = c(___, ___)
  )
```

```{r s14-gt-spanner-hint-1}
# label = "Age Statistics", columns = c(Age_Mean, Age_SD)
```

```{r s14-gt-spanner-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_spanner(
    label   = "Age Statistics",
    columns = c(Age_Mean, Age_SD)
  )
```

### Exercise 14-d — fmt_number() vs fmt_percent()

> **Clinical gotcha:** `Pct_Female` already holds values like `47.1` (a
> percentage, not a proportion). Use `fmt_number()` on it. `fmt_percent()`
> multiplies by 100 and is only for true proportions (0–1).

```{r s14-gt-fmt, exercise=TRUE, exercise.lines=14}
# Format Age_Mean and Age_SD to 1 decimal; Pct_Female to 1 decimal (already a %).
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  fmt_number(columns = c(___, ___), decimals = ___) |>
  fmt_number(columns = ___, decimals = ___)
```

```{r s14-gt-fmt-hint-1}
# fmt_number(columns = c(Age_Mean, Age_SD), decimals = 1)
# fmt_number(columns = Pct_Female, decimals = 1)
```

```{r s14-gt-fmt-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  fmt_number(columns = c(Age_Mean, Age_SD), decimals = 1) |>
  fmt_number(columns = Pct_Female, decimals = 1)
```

### Exercise 14-e — tab_style(): conditional formatting

`tab_style()` takes a `style` (what: `cell_text(weight = "bold")`,
`cell_fill(color = ...)`) and `locations` (where: `cells_column_labels()`,
`cells_body(rows = ...)`).

```{r s14-gt-style, exercise=TRUE, exercise.lines=20}
# Bold all column headers, then highlight rows where Age_Mean > 75.
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_style(
    style     = cell_text(weight = ___),
    locations = ___()
  ) |>
  tab_style(
    style     = cell_fill(color = "#FFFACD"),
    locations = cells_body(rows = ___)
  )
```

```{r s14-gt-style-hint-1}
# weight = "bold"; locations = cells_column_labels()
# cells_body(rows = Age_Mean > 75)
```

```{r s14-gt-style-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#FFFACD"),
    locations = cells_body(rows = Age_Mean > 75)
  )
```

### Exercise 14-f — tab_options() and gtsave()

`tab_options()` sets table-wide defaults; `gtsave()` writes the file, inferring
the format from the extension. For submissions, **RTF** is the standard because
it imports cleanly into Word-based CSR publishing workflows.

```{r s14-gt-save, exercise=TRUE, exercise.lines=18}
# Set font size 10 and bold headers, then save as RTF.
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_options(
    table.font.size           = ___,
    column_labels.font.weight = ___
  ) |>
  gtsave(filename = ___)
```

```{r s14-gt-save-hint-1}
# table.font.size = 10, column_labels.font.weight = "bold"
# gtsave(filename = "Table_14_1_1.rtf")
```

```{r s14-gt-save-solution}
demog_wide |>
  gt() |>
  cols_label(TRT01P = "Treatment", N = "N", Age_Mean = "Mean Age",
             Age_SD = "SD", Pct_Female = "% Female") |>
  tab_options(
    table.font.size           = 10,
    column_labels.font.weight = "bold"
  ) |>
  gtsave(filename = "Table_14_1_1.rtf")
```

```{r s14-quiz, echo=FALSE}
quiz(
  question(
    "A column pct_resp holds values like 42.3 (already a percentage). Which formatter?",
    answer("fmt_percent()", message = "fmt_percent() multiplies by 100, so 42.3 would show as 4230%."),
    answer("fmt_number()", correct = TRUE,
           message = "The value is already a percentage; fmt_number() controls decimals without rescaling."),
    answer("fmt_integer()"),
    answer("fmt_scientific()"),
    allow_retry = TRUE
  ),
  question(
    "Which gtsave() extension produces a Word-importable submission table?",
    answer(".html"),
    answer(".pdf"),
    answer(".rtf", correct = TRUE,
           message = "RTF imports directly into Word-based CSR publishing workflows."),
    answer(".xlsx"),
    allow_retry = TRUE
  ),
  question(
    "How do tfrmt and gt divide labour in the modern TLG stack?",
    answer("They are competitors; pick one and never mix them"),
    answer("tfrmt describes layout metadata and renders via gt; gt provides the final display polish and file export",
           correct = TRUE),
    answer("gt builds the ARD; tfrmt formats it"),
    answer("tfrmt is for HTML only; gt is for RTF only"),
    allow_retry = TRUE
  )
)
```

---

## Section 15 — Capstone: ARD → tfrmt → gt → RTF

Bring everything together. Build a **submission-style AE table** end-to-end: an
ARD, a tfrmt spec with row grouping and pagination, a mock for sign-off, the
rendered table, and an RTF export — then show how gt would add the final
display polish.

### Exercise 15-a — Inspect the PT-level ARD

```{r s15-cap-ard, exercise=TRUE, exercise.lines=10}
# ae_pt_tfrmt_ready was built in setup. Confirm its shape.
ae_pt_tfrmt_ready |> distinct(label_level) |> nrow()   # unique PTs
ae_pt_tfrmt_ready |> distinct(___) |> nrow()           # treatment arms
ae_pt_tfrmt_ready |> distinct(param)                   # should be n and p
```

```{r s15-cap-ard-hint-1}
# distinct(column) for the treatment arm count
```

```{r s15-cap-ard-solution}
ae_pt_tfrmt_ready |> distinct(label_level) |> nrow()
ae_pt_tfrmt_ready |> distinct(column) |> nrow()
ae_pt_tfrmt_ready |> distinct(param)
```

### Exercise 15-b — Build the complete AE spec

```{r s15-cap-spec, exercise=TRUE, exercise.lines=30}
# body_plan: tfrmt_n_pct() for all rows
# row_grp_plan: SOC (group_val) > PT (label_val), indented
# page_plan: break on AESOC, notes below
ae_full_spec <- tfrmt(
  label     = "label_level",
  param     = ___,
  value     = ___,
  column    = ___,
  title     = "Table 14.3.1: Adverse Events by SOC and Preferred Term",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", ___)
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = ___, label_val = ___, indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = ___),
    note_loc = "below"
  )
)

print(ae_full_spec)
```

```{r s15-cap-spec-hint-1}
# param = "param", value = "value", column = "column"
# tfrmt_n_pct() in the frmt_structure
# group_val = "AESOC", label_val = "AEDECOD"; page_structure group_val = "AESOC"
```

```{r s15-cap-spec-solution}
ae_full_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  title     = "Table 14.3.1: Adverse Events by SOC and Preferred Term",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                      indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

print(ae_full_spec)
```

### Exercise 15-c — Mock, render, and export

```{r s15-cap-render, exercise=TRUE, exercise.lines=24}
# Rebuild the spec, then: mock for sign-off -> render -> RTF export.
ae_full_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                      indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

# 1. Mock for biostatistician sign-off
mock_df(ae_full_spec, ae_pt_tfrmt_ready)

# 2. Render and assign the gt object
gt_out <- print_to_gt(___, ___)

# 3. Export to RTF
as_rtf(gt_out, ___)
```

```{r s15-cap-render-hint-1}
# print_to_gt(ae_full_spec, ae_pt_tfrmt_ready)
# as_rtf(gt_out, "Table_14_3_1.rtf")
```

```{r s15-cap-render-solution}
ae_full_spec <- tfrmt(
  label     = "label_level",
  param     = "param",
  value     = "value",
  column    = "column",
  body_plan = body_plan(
    frmt_structure(label_val = ".default", tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = "AESOC", label_val = "AEDECOD",
                      indent = "indented"),
    label_loc = "indented"
  ),
  page_plan = page_plan(
    page_structure(group_val = "AESOC"),
    note_loc = "below"
  )
)

mock_df(ae_full_spec, ae_pt_tfrmt_ready)

gt_out <- print_to_gt(ae_full_spec, ae_pt_tfrmt_ready)

as_rtf(gt_out, "Table_14_3_1.rtf")
```

> **Where gt takes over:** In a real session `print_to_gt()` returns a live gt
> object. You can then pipe it into `tab_footnote()`, `tab_source_note()`, or
> `tab_options()` for any display polish the tfrmt spec did not encode, before
> calling `gtsave("Table_14_3_1.rtf")`. tfrmt owns the *structure*; gt owns the
> *final cosmetic layer*.

---

## Lesson Summary

| Object | Purpose |
|--------|---------|
| `frmt("xx.x")` | Format a single statistic to specified decimals |
| `frmt_combine("{n} ({p}%)", ...)` | Combine multiple stats into one cell string |
| `frmt_structure(label_val="AGE", frmt("xx.x"))` | Attach a format rule to a row |
| `body_plan(frmt_structure(...), ...)` | Collect all row format rules; first match wins |
| `tfrmt_n_pct()` | Shortcut `frmt_combine` for the standard `n (x.x%)` cell |
| `col_plan("label","Placebo",...)` | Column display order |
| `spanning_structure("Arm" = c(...))` | Spanner header above grouped columns |
| `row_grp_plan()` + `row_grp_structure()` | SOC header → indented PT rows |
| `page_plan()` + `page_structure()` | Page breaks for long tables |
| `tfrmt(label, param, value, column, ...)` | Assemble the complete table spec |
| `ard_to_tfrmt(ard)` | Reshape a cards ARD to tfrmt-ready format |
| `mock_df()` / `print_mock_gt()` | Blank shell for sponsor sign-off |
| `print_to_gt(spec, .data)` | Apply spec to data; render a gt table |
| `as_rtf(gt, "*.rtf")` | Export the rendered table to RTF |
| `gt()` + `cols_label()` / `tab_header()` / `tab_spanner()` | gt display verbs |
| `fmt_number()` / `fmt_percent()` | gt numeric formatting (decimals vs rescale) |
| `tab_style()` + `cell_text()` / `cell_fill()` | gt conditional cell formatting |
| `tab_options()` / `gtsave()` | gt global options and file export |

### Key takeaways

- `tfrmt` **separates display metadata from data** — the spec is saved and
  version-controlled independently of the analysis results, so mock and final
  tables share one layout and only the ARD changes between cut and lock.
- The tfrmt-ready ARD has columns `label`, `label_level`, `column`, `param`,
  `value`; `ard_to_tfrmt()` maps them from standard cards ARD names (the real
  stack uses `shuffle_card()` + `prep_*` helpers).
- `body_plan()` rules match top-down on `label_val` / `group_val`; place
  specific rules before any `".default"` catch-all.
- `row_grp_plan()`, `page_plan()`, and `spanning_structure()` add the row
  hierarchies, pagination, and column spanners real submission tables need.
- The **mock-then-render** workflow replaces the manually drawn Word shell SAS
  programmers circulate; the shell is generated directly from the spec.
- `gt` is the rendering engine beneath tfrmt and gtsummary — use it directly for
  final display polish (titles, spanners, footnotes, conditional styling) and
  for file export (`gtsave()`), with **RTF** as the submission standard.
- Use `fmt_number()` for values already expressed as percentages (`47.1`) and
  `fmt_percent()` only for true proportions (`0.471`).
- tfrmt objects are serializable display metadata — a natural fit for emerging
  **CDISC ARS** workflows, where machine-readable display specs travel alongside
  the ARD. (Watch CDISC ARS releases through 2026–2027.)

### When to reach for tfrmt vs gtsummary vs rtables

| Tool | Layout abstraction | Strengths |
|---|---|---|
| gtsummary | Function composition (`add_*`, `modify_*`) | Quick, fluent, standard cases |
| tfrmt | Declarative metadata object | Mock workflow, complex columns, full layout separation |
| rtables (Module 7) | Imperative layout DSL (`analyze`, `split_*`) | Maximum flexibility, NEST-stack alignment |

All three coexist. Cardinal-future templates favour gtsummary + tfrmt; the
legacy stack favours rtables + tern.

---

## What's next

Module 6 (the Cardinal-future TLG stack: cards/cardx → gtsummary → tfrmt) is
complete. **Module 7** covers the **legacy TLG stack** — starting with **Lesson
33: rtables** — followed by tern, r2rtf, Tplyr, and tidytlg. These packages
dominate production today and remain essential for sponsors invested in NEST or
similar workflows; understanding them lets you collaborate with and maintain
existing code even when new work uses Cardinal-future.

After Module 7: Module 8 (Shiny / teal), Module 9 (submission packaging —
xportr, datasetjson), Module 10 (traceability — logrx, diffdf), and a capstone
study in Module 11.

---

## Self-check questions

1. What does tfrmt provide that gtsummary does not?
2. What are the five columns a tfrmt-ready ARD must contain, and what cards ARD
   columns do they map from?
3. Explain the "mock first" workflow and why it is valuable in pharma.
4. What does `frmt_combine("{n} ({p}%)", n = frmt("xxx"), p = frmt("xx.x"))`
   produce, and how does `tfrmt_n_pct()` relate to it?
5. In `row_grp_plan()`, what is the difference between `label_loc = "indented"`
   and `label_loc = "column"`?
6. A column already holds `47.1` (a percentage). Should you apply `fmt_number()`
   or `fmt_percent()` in gt, and why?
7. Translate to tfrmt-conceptual terms: "Render a table with rows grouped by
   AESOC, indented PT labels, columns of Placebo and Active, cell format
   `n (p%)`, new page per SOC, exported to RTF."
8. How does tfrmt support CDISC ARS-aligned workflows?

## Glossary

- **tfrmt** — A metadata-driven display layer for ARDs, complementary to gtsummary.
- **`frmt()` / `frmt_combine()`** — tfrmt's format-string mini-language; `"xx.x"` = one decimal, `"{n} ({p}%)"` = combined cell.
- **`frmt_structure()`** — Pairs a frmt/frmt_combine with `group_val`/`label_val` selectors (`".default"` = catch-all).
- **`body_plan()`** — Collects frmt_structures; first matching rule per cell wins.
- **`tfrmt_n_pct()`** — Shortcut producing the standard `n (x.x%)` frmt_combine.
- **`col_plan()`** — Defines column ordering and (via `spanning_structure()` / `>`) spanner labels.
- **`spanning_structure()`** — A merged column header above a group of columns.
- **`row_grp_plan()` / `row_grp_structure()`** — Control row group display: indentation, spacing, label location (e.g. SOC > PT).
- **`page_plan()` / `page_structure()`** — Pagination rules for multi-page tables.
- **`big_n_structure()`** — Maps Big-N rows to per-column header N counts.
- **`ard_to_tfrmt()`** — Reshape a cards ARD into tfrmt long format (real stack: `shuffle_card()` + `prep_big_n()` / `prep_combine_vars()` / `prep_label()`).
- **`print_to_gt()`** — Render a tfrmt spec + ARD as a gt table.
- **`print_mock_gt()` / `mock_df()`** — Render the spec with placeholder cells (no real data) for sponsor review.
- **`as_rtf()`** — Export the rendered table to an RTF file.
- **gt** — The grammar-of-tables rendering engine under tfrmt and gtsummary.
- **`cols_label()` / `tab_header()` / `tab_spanner()`** — gt verbs to rename columns, add titles, and span columns.
- **`fmt_number()` / `fmt_percent()`** — gt numeric formatting; the former fixes decimals, the latter rescales proportions to percentages.
- **`tab_style()` / `cell_text()` / `cell_fill()`** — gt conditional formatting (bold, colour, fill) targeted via `cells_body()` / `cells_column_labels()`.
- **`tab_options()` / `gtsave()`** — gt global style options and multi-format export (`.rtf` is the submission standard).
- **Mock workflow** — Design the layout shell first, get sign-off, then plug in data; the tfrmt object guarantees mock and final match.
