---
title: "Lesson 32 — Complete TLF Pipeline Capstone: cards → tfrmt → gt → RTF"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem
#                 (dplyr, admiral, sdtm.oak, metacore, xportr, teal and more).
#                 Built on real CDISC CDISCPILOT01 data.
#
# SCRIPT NAME   : 32_tlf_capstone.Rmd
# PURPOSE       : TLF Pipeline Capstone — end-to-end clinical table package: Table 14.1.1
#                 demographics via gtsummary, Table 14.3.1 AE overview
#                 via tfrmt, Table 14.4.1 ALT summary via gt, double-
#                 programming validation with anti_join gap checks, and
#                 tlf_summary() package QC function.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(pharmaverseadam)

# ── cards mocks (lesson 25) ──────────────────────────────────────────────────
if (!requireNamespace("cards", quietly = TRUE)) {
  ard_continuous        <- function(...) tibble(variable=character(), stat_name=character(),
                                                stat=list(), group1=character(), group1_level=character())
  ard_categorical       <- function(...) tibble(variable=character(), stat_name=character(),
                                                stat=list(), group1=character(), group1_level=character())
  ard_missing           <- function(...) tibble(variable=character(), stat_name=character(), stat=list())
  ard_stack             <- function(...) tibble(variable=character(), stat_name=character(), stat=list())
} else {
  library(cards)
}

# ── cardx / cardinal mocks (lesson 28) ──────────────────────────────────────
if (!requireNamespace("cardx", quietly = TRUE)) {
  ard_continuous_ci  <- function(...) tibble(variable=character(), stat_name=character(),
                                             stat=list(), group1=character(), group1_level=character())
  ard_proportion_ci  <- function(data, variables, by, ...) {
    rows <- lapply(variables, function(v) {
      lapply(unique(data[[by]]), function(grp) {
        sub <- data[data[[by]] == grp, v, drop=TRUE]
        n_y <- sum(sub == "Y", na.rm = TRUE)
        n   <- sum(!is.na(sub))
        p   <- n_y / max(n, 1)
        tibble(
          group1       = by,
          group1_level = as.character(grp),
          variable     = v,
          stat_name    = c("n", "p", "conf.low", "conf.high"),
          stat         = list(n_y, p, max(p - 1.96*sqrt(p*(1-p)/max(n,1)), 0),
                              min(p + 1.96*sqrt(p*(1-p)/max(n,1)), 1))
        )
      }) |> bind_rows()
    }) |> bind_rows()
    rows
  }
} else {
  library(cardx)
}

# ── tfrmt mocks (lessons 29-30) ──────────────────────────────────────────────
if (!requireNamespace("tfrmt", quietly = TRUE)) {
  frmt             <- function(...) structure(list(...), class = "frmt")
  frmt_combine     <- function(...) structure(list(...), class = "frmt_combine")
  frmt_structure   <- function(...) structure(list(...), class = "frmt_structure")
  body_plan        <- function(...) structure(list(...), class = "body_plan")
  col_plan         <- function(...) structure(list(...), class = "col_plan")
  tfrmt            <- function(...) structure(list(...), class = "tfrmt")
  print_to_gt      <- function(x, ...) {
    message("[tfrmt mock] print_to_gt() called — install tfrmt for real output")
    invisible(NULL)
  }
  mock_df          <- function(x, ...) {
    message("[tfrmt mock] mock_df() called — returns blank shell")
    tibble(label=character(), column=character(), value=character())
  }
  ard_to_tfrmt     <- function(...) tibble(label=character(), column=character(),
                                           param=character(), value=numeric())
  tfrmt_n_pct      <- function(...) structure(list(), class = "tfrmt_n_pct")
  page_plan        <- function(...) structure(list(...), class = "page_plan")
  page_structure   <- function(...) structure(list(...), class = "page_structure")
  row_grp_plan     <- function(...) structure(list(...), class = "row_grp_plan")
  row_grp_structure <- function(...) structure(list(...), class = "row_grp_structure")
} else {
  library(tfrmt)
}

# ── gtsummary mocks (lesson 26) ──────────────────────────────────────────────
if (!requireNamespace("gtsummary", quietly = TRUE)) {
  tbl_summary <- function(data, by = NULL, include = NULL, ...) {
    vars <- if (!is.null(include)) include else names(data)
    if (!is.null(by)) vars <- setdiff(vars, by)
    structure(
      list(
        table_body   = tibble(variable = vars, label = vars, stat_1 = "—", stat_2 = "—"),
        by           = by,
        inputs       = list(data = data, include = vars)
      ),
      class = c("tbl_summary", "gtsummary")
    )
  }
  add_p            <- function(x, ...) x
  add_overall      <- function(x, ...) x
  add_n            <- function(x, ...) x
  bold_labels      <- function(x, ...) x
  italicize_levels <- function(x, ...) x
  inline_text      <- function(x, variable, ...) "p = 0.999 (mock)"
  as_tibble.gtsummary <- function(x, ...) x$table_body
  print.gtsummary  <- function(x, ...) {
    message("[gtsummary mock] install gtsummary for real output")
    print(x$table_body)
    invisible(x)
  }
} else {
  library(gtsummary)
}

# ── gt mocks (lesson 31) ─────────────────────────────────────────────────────
if (!requireNamespace("gt", quietly = TRUE)) {
  gt               <- function(data, ...) structure(list(data=data), class="gt_tbl")
  cols_label       <- function(x, ...) x
  tab_header       <- function(x, ...) x
  tab_spanner      <- function(x, ...) x
  tab_footnote     <- function(x, ...) x
  tab_source_note  <- function(x, ...) x
  fmt_number       <- function(x, ...) x
  tab_style        <- function(x, ...) x
  tab_options      <- function(x, ...) x
  gtsave           <- function(x, filename, ...) {
    message("[gt mock] gtsave() called — install gt for real output")
    invisible(x)
  }
  cell_text        <- function(...) list(...)
  cells_body       <- function(...) list(...)
  cells_column_labels <- function(...) list(...)
  md               <- function(x) x
  print.gt_tbl     <- function(x, ...) {
    message("[gt mock] install gt for real output"); print(x$data); invisible(x)
  }
} else {
  library(gt)
}

# ── Base pharmaverseadam datasets ────────────────────────────────────────────

# ADSL
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 161, 7)), round(rnorm(n, 174, 8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 68, 12), 1), round(rnorm(n, 82, 14), 1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL = ht, WEIGHTBL = wt,
                                  BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
}
adsl <- adsl_raw |>
  filter(SAFFL == "Y") |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N")
  )

# ADAE — rename TRTA → TRT01A
adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw)) {
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
} else if (!"TRT01A" %in% names(adae_raw)) {
  adae_raw <- left_join(adae_raw, select(adsl, USUBJID, TRT01A), by = "USUBJID")
}
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

# ADLB — fix ATOXGR, drop admiral summary rows, add TRTSDT
adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ── Table-specific helper objects ─────────────────────────────────────────────

# Table 14.1.1 — Demographics
t141_adsl <- adsl |>
  mutate(
    DTHFL_DISP = if_else(!is.na(DTHFL) & DTHFL == "Y", "Y", "N"),
    AGEGR1     = factor(AGEGR1)
  )

# Table 14.3.1 — Subject-level AE flags
t143_ae_subj <- adsl |>
  select(USUBJID, TRT01P) |>
  left_join(
    adae |> group_by(USUBJID) |> summarise(ANY_AE = "Y", .groups = "drop"),
    by = "USUBJID"
  ) |>
  mutate(
    ANY_AE  = replace_na(ANY_AE, "N"),
    SAE     = if_else(USUBJID %in% adae$USUBJID[adae$AESER == "Y"],  "Y", "N"),
    DISC_AE = if_else(USUBJID %in% adae$USUBJID[!is.na(adae$AEOUT) & adae$AEOUT == "FATAL"], "Y", "N"),
    SEV_AE  = if_else(USUBJID %in% adae$USUBJID[adae$AESEV == "SEVERE"], "Y", "N")
  )

# Table 14.4.1 — ALT at Baseline, Week 8, Week 24
# Drop TRT01P first — adlb already carries it; re-join from adsl to avoid .x/.y
t144_alt <- adlb |>
  filter(PARAMCD == "ALT", AVISIT %in% c("Baseline", "Week 8", "Week 24")) |>
  select(-any_of("TRT01P")) |>
  left_join(select(adsl, USUBJID, TRT01P), by = "USUBJID") |>
  filter(!is.na(TRT01P), !is.na(AVAL))
```

## Overview

Welcome to **Lesson 32**, the capstone for the entire Pharmaverse07 block (lessons 25–31).

In lessons 25–31 you learned each tool in isolation:

| Lesson | Tool | What it provides |
|--------|------|-----------------|
| 25 | cards | Analysis Results Dataset (ARD) — machine-readable statistics |
| 26 | gtsummary | High-level summary tables from a single chain |
| 27 | cardinal | Clinical trial-specific ARD functions |
| 28 | cardx | Extensions: CI functions, regression, survival |
| 29 | tfrmt | Template-first formatting engine |
| 30 | tfrmt (advanced) | Page plans, row groups, mock shells |
| 31 | gt | Low-level cell-by-cell table styling |

This lesson puts them all together to produce **three Tables, Listings, and Figures (TLFs)** that would appear in a clinical study report:

- **Table 14.1.1** — Demographics and Baseline Characteristics
- **Table 14.3.1** — Overview of Adverse Events
- **Table 14.4.1** — Summary of Lab Values: ALT at Weeks 8 and 24

By the end of this lesson you will be able to take raw `pharmaverseadam` data all the way to a formatted, RTF-ready output table using the tools of your choice.

> **Study context:** CDISCPILOT01 — Xanomeline Alzheimer's study. Three arms: Placebo, Xanomeline Low Dose, Xanomeline High Dose. Safety population (SAFFL == "Y"), n = 254.

---

## Section 1 — Pipeline Architecture Review

### The complete TLF pipeline

Every table in a clinical study report travels through the same conceptual stages:

```
pharmaverseadam         cards / cardx / cardinal        gtsummary / tfrmt          gt / RTF
─────────────────   →   ─────────────────────────   →   ──────────────────────   →  ────────
adsl, adae, adlb        ARD (Analysis Results           Formatting specification     Final output:
(CDISC CDISCPILOT01)    Dataset): one row per            applied to ARD               HTML, RTF,
                        statistic, machine-              or data frame                PDF, DOCX
                        readable, format-free
```

The **ARD** is the pivot point. It stores every number exactly once, independent of any formatting decision. This means:

1. A statistician can verify the numbers by reading the ARD
2. The same ARD can feed both an internal draft (HTML) and a submission-quality RTF without recomputing
3. A QC programmer can independently reproduce the ARD and diff it against the primary

### tfrmt vs gtsummary — when to use which

| Situation | Prefer |
|-----------|--------|
| Submission template already exists (locked shell) | **tfrmt** — define spec once, apply to any study |
| Quick summary for data review meeting | **gtsummary** — one chain, sensible defaults |
| Need mock tables before data lock | **tfrmt** + `mock_df()` |
| Complex nesting (SOC / PT hierarchy) | **tfrmt** with `row_grp_plan` |
| Journal-ready demographics table | **gtsummary** → `as_gt()` → fine-tune with gt |

### Quiz

```{r quiz-s1, echo=FALSE}
quiz(
  question(
    "Which component in the pipeline stores statistical results independently of all formatting?",
    answer("gt table object"),
    answer("ARD (Analysis Results Dataset)", correct = TRUE),
    answer("tfrmt specification"),
    answer("gtsummary tbl_summary object"),
    allow_retry = TRUE
  ),
  question(
    "What does mock_df() produce in a tfrmt workflow?",
    answer("A dataset of mock patient records for testing"),
    answer("A blank table shell showing structure without data", correct = TRUE),
    answer("A formatted RTF output with placeholder text"),
    answer("A list of all column labels registered in the spec"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — Table 14.1.1: Demographics (gtsummary approach)

Table 14.1.1 is the first table in almost every clinical study report. It summarises baseline characteristics by treatment arm so reviewers can assess comparability.

**Variables to include:** Age (continuous + grouped), Sex, Race, Weight at Baseline, BMI, Diabetes flag (DIABFL derived from BMIBL).

The `t141_adsl` object is already prepared in the setup chunk with `AGEGR1` as a factor and a clean `DTHFL_DISP` column.

### Exercise 2.1 — Build the demographics table

Build `tbl_14_1_1` using a `tbl_summary()` chain. Include the seven variables listed, stratify by `TRT01P`, add p-values, an overall column, N counts per arm, and bold the variable labels.

```{r ex-tbl-141, exercise=TRUE, exercise.lines=14}
tbl_14_1_1 <- tbl_summary(
  data    = ___,
  by      = "___",
  include = c("AGE", "AGEGR1", "SEX", "RACE", "WEIGHTBL", "BMIBL", "DIABFL")
) |>
  ___() |>        # add p-values
  ___() |>        # add overall column
  ___() |>        # add N per arm in column header
  ___()           # bold variable labels

tbl_14_1_1
```

```{r ex-tbl-141-hint-1}
# tbl_summary() arguments: data=, by= (column name as string), include= (vector of column names)
# The four functions to chain: add_p(), add_overall(), add_n(), bold_labels()
```

```{r ex-tbl-141-solution}
tbl_14_1_1 <- tbl_summary(
  data    = t141_adsl,
  by      = "TRT01P",
  include = c("AGE", "AGEGR1", "SEX", "RACE", "WEIGHTBL", "BMIBL", "DIABFL")
) |>
  add_p() |>
  add_overall() |>
  add_n() |>
  bold_labels()

tbl_14_1_1
```

### Exercise 2.2 — Extract and pivot AGE statistics

Convert `tbl_14_1_1` to a tibble with `as_tibble()`, filter to the AGE rows (where `variable == "AGE"`), and show the result. This is how downstream programs can programmatically read back a formatted statistic.

```{r ex-tbl-141-extract, exercise=TRUE, exercise.lines=10}
age_rows <- as_tibble(tbl_14_1_1) |>
  filter(___ == "AGE")

age_rows
```

```{r ex-tbl-141-extract-hint-1}
# as_tibble() on a gtsummary object returns a data frame with a 'variable' column
# Use filter(variable == "AGE") to select Age rows
```

```{r ex-tbl-141-extract-solution}
age_rows <- as_tibble(tbl_14_1_1) |>
  filter(variable == "AGE")

age_rows
```

### Exercise 2.3 — Report the SEX p-value inline

Use `inline_text()` to extract the p-value for the `SEX` variable and paste it into a sentence. This is the standard pattern for populating body text in a clinical study report.

```{r ex-tbl-141-inline, exercise=TRUE, exercise.lines=8}
p_sex <- inline_text(tbl_14_1_1, variable = ___)

cat("Chi-squared test for sex distribution across arms:", p_sex)
```

```{r ex-tbl-141-inline-hint-1}
# inline_text(tbl_object, variable = "VARIABLE_NAME")
# The variable name matches what was passed to include= in tbl_summary()
```

```{r ex-tbl-141-inline-solution}
p_sex <- inline_text(tbl_14_1_1, variable = "SEX")

cat("Chi-squared test for sex distribution across arms:", p_sex)
```

### Quiz

```{r quiz-s2, echo=FALSE}
quiz(
  question(
    "Which gtsummary function adds the total-column (N across all arms) to the column header?",
    answer("add_overall()"),
    answer("add_n()", correct = TRUE),
    answer("add_stat_label()"),
    answer("add_header_above()"),
    allow_retry = TRUE
  ),
  question(
    "After calling as_tibble() on a gtsummary object, which column contains the variable name (e.g., 'AGE', 'SEX')?",
    answer("label"),
    answer("var_label"),
    answer("variable", correct = TRUE),
    answer("term"),
    allow_retry = TRUE
  )
)
```

---

## Section 3 — Table 14.3.1: AE Overview (tfrmt approach)

Table 14.3.1 is the high-level AE overview that appears before the detailed SOC/PT listings. It shows, for each treatment arm, how many subjects experienced *any* AE, a serious AE (SAE), a severe AE, or an AE leading to death.

The `t143_ae_subj` data frame (built in setup) is already at the subject level — one row per subject with Y/N flags for each event category.

### Exercise 3.1 — Build the AE proportion ARD

Use `ard_proportion_ci()` to compute the proportion of subjects with each flag, with 95% confidence intervals, stratified by `TRT01P`.

```{r ex-ard-143, exercise=TRUE, exercise.lines=12}
ard_143 <- ard_proportion_ci(
  data      = ___,
  variables = c("ANY_AE", "SAE", "SEV_AE", "DISC_AE"),
  by        = "___"
)

# Inspect the result
ard_143 |>
  filter(stat_name %in% c("n", "p")) |>
  select(group1_level, variable, stat_name, stat)
```

```{r ex-ard-143-hint-1}
# ard_proportion_ci(data=, variables= c(...), by= "column_name")
# The 'by' argument is the treatment column: TRT01P
# stat_name values include "n", "p", "conf.low", "conf.high"
```

```{r ex-ard-143-solution}
ard_143 <- ard_proportion_ci(
  data      = t143_ae_subj,
  variables = c("ANY_AE", "SAE", "SEV_AE", "DISC_AE"),
  by        = "TRT01P"
)

ard_143 |>
  filter(stat_name %in% c("n", "p")) |>
  select(group1_level, variable, stat_name, stat)
```

### Exercise 3.2 — Convert ARD to tfrmt format and render

Reshape `ard_143` into the long format tfrmt expects (`label`, `column`, `param`, `value`), define a minimal tfrmt spec using `tfrmt_n_pct()`, and call `print_to_gt()`.

```{r ex-tfrmt-143, exercise=TRUE, exercise.lines=30}
# Step 1: reshape ARD to tfrmt-ready long format
tfrmt_data_143 <- ard_143 |>
  mutate(value = as.numeric(unlist(stat))) |>
  filter(stat_name %in% c("n", "p", "conf.low", "conf.high")) |>
  select(
    label  = ___,          # row label (variable name: ANY_AE, SAE, …)
    column = ___,          # column (arm name)
    param  = ___,          # statistic type (n, p, …)
    value
  )

# Step 2: define tfrmt spec
spec_143 <- tfrmt(
  param     = param,
  value     = value,
  label     = label,
  column    = column,
  body_plan = body_plan(
    frmt_structure(
      label_val = .default,
      ___()              # use tfrmt_n_pct() for n (%) format
    )
  )
)

# Step 3: render
print_to_gt(spec_143, .data = tfrmt_data_143)
```

```{r ex-tfrmt-143-hint-1}
# ARD columns: group1_level = arm, variable = flag name, stat_name = stat type
# tfrmt long format needs: label (row id), column (arm), param (stat type), value (number)
# In body_plan: frmt_structure(label_val = .default, tfrmt_n_pct())
```

```{r ex-tfrmt-143-solution}
tfrmt_data_143 <- ard_143 |>
  mutate(value = as.numeric(unlist(stat))) |>
  filter(stat_name %in% c("n", "p", "conf.low", "conf.high")) |>
  select(
    label  = variable,
    column = group1_level,
    param  = stat_name,
    value
  )

spec_143 <- tfrmt(
  param     = param,
  value     = value,
  label     = label,
  column    = column,
  body_plan = body_plan(
    frmt_structure(
      label_val = .default,
      tfrmt_n_pct()
    )
  )
)

print_to_gt(spec_143, .data = tfrmt_data_143)
```

### Exercise 3.3 — Add a row group plan and view the mock shell

Add `row_grp_plan()` to group the AE rows by category, then call `mock_df()` to preview the blank table structure. Mock tables are used in pre-data-lock planning so programmers and statisticians can agree on the layout before any data is available.

```{r ex-mock-143, exercise=TRUE, exercise.lines=22}
spec_143_grp <- tfrmt(
  param     = param,
  value     = value,
  label     = label,
  column    = column,
  body_plan = body_plan(
    frmt_structure(label_val = .default, tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(
      group_val = .default,
      label_loc = element_row_grp_loc(location = "indented")
    )
  )
)

# Inspect the blank table shell
mock_df(spec_143_grp)
```

```{r ex-mock-143-hint-1}
# row_grp_plan() wraps one or more row_grp_structure() calls
# element_row_grp_loc(location = "indented") indents group-level rows
# mock_df(spec) returns a blank gt table showing the row/column skeleton
```

```{r ex-mock-143-solution}
spec_143_grp <- tfrmt(
  param     = param,
  value     = value,
  label     = label,
  column    = column,
  body_plan = body_plan(
    frmt_structure(label_val = .default, tfrmt_n_pct())
  ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(
      group_val = .default,
      label_loc = element_row_grp_loc(location = "indented")
    )
  )
)

mock_df(spec_143_grp)
```

---

## Section 4 — Table 14.4.1: Lab Summary (cards + gt approach)

Table 14.4.1 reports ALT (alanine aminotransferase) values at three time points: Baseline, Week 8, and Week 24. The table shows N, Mean, SD, Median, and the percentage of subjects with ALT > 40 U/L (a common clinical threshold) per arm per visit.

`t144_alt` is already filtered in the setup chunk to ALT observations at the three visits with treatment arm joined in.

### Exercise 4.1 — Compute the ALT summary with dplyr

Use `group_by()` + `summarise()` to produce `alt_summary` with one row per arm-visit combination. Include N, Mean, SD, Median (all rounded) and `pct_hi` (percentage of subjects with AVAL > 40).

```{r ex-alt-summary, exercise=TRUE, exercise.lines=16}
alt_summary <- t144_alt |>
  group_by(___, ___) |>
  summarise(
    N      = ___,
    Mean   = round(mean(AVAL, na.rm = TRUE), 1),
    SD     = round(sd(AVAL,   na.rm = TRUE), 1),
    Median = round(median(AVAL, na.rm = TRUE), 1),
    pct_hi = round(100 * mean(AVAL > 40,  na.rm = TRUE), 1),
    .groups = "drop"
  )

alt_summary
```

```{r ex-alt-summary-hint-1}
# group_by(TRT01P, AVISIT)
# N = n()  (count of non-NA rows within the group)
# pct_hi = round(100 * mean(AVAL > 40, na.rm=TRUE), 1)
```

```{r ex-alt-summary-solution}
alt_summary <- t144_alt |>
  group_by(TRT01P, AVISIT) |>
  summarise(
    N      = n(),
    Mean   = round(mean(AVAL, na.rm = TRUE), 1),
    SD     = round(sd(AVAL,   na.rm = TRUE), 1),
    Median = round(median(AVAL, na.rm = TRUE), 1),
    pct_hi = round(100 * mean(AVAL > 40,  na.rm = TRUE), 1),
    .groups = "drop"
  )

alt_summary
```

### Exercise 4.2 — Pivot wide and build a gt table

Pivot `alt_summary` so each visit becomes a set of columns (N, Mean, SD per visit), then pass to `gt()` for initial rendering.

```{r ex-alt-gt, exercise=TRUE, exercise.lines=18}
alt_wide <- alt_summary |>
  pivot_wider(
    names_from  = ___,
    values_from = c(___, ___, ___, ___),
    names_glue  = "{AVISIT}_{.value}"
  )

tbl_14_4_1 <- alt_wide |>
  gt() |>
  tab_header(
    title    = md("**Table 14.4.1**"),
    subtitle = "ALT Summary (U/L) — Safety Population"
  ) |>
  cols_label(TRT01P = "Treatment Arm")

tbl_14_4_1
```

```{r ex-alt-gt-hint-1}
# pivot_wider: names_from = AVISIT, values_from = c(N, Mean, SD, Median, pct_hi)
# names_glue = "{AVISIT}_{.value}" creates columns like "Baseline_N", "Week 8_Mean", …
# gt() | tab_header(title=, subtitle=) | cols_label(col_name = "Display Label")
```

```{r ex-alt-gt-solution}
alt_wide <- alt_summary |>
  pivot_wider(
    names_from  = AVISIT,
    values_from = c(N, Mean, SD, Median, pct_hi),
    names_glue  = "{AVISIT}_{.value}"
  )

tbl_14_4_1 <- alt_wide |>
  gt() |>
  tab_header(
    title    = md("**Table 14.4.1**"),
    subtitle = "ALT Summary (U/L) — Safety Population"
  ) |>
  cols_label(TRT01P = "Treatment Arm")

tbl_14_4_1
```

### Exercise 4.3 — Add spanners, footnote, and save

Add a `tab_spanner()` for each visit group, a footnote explaining the "> 40 U/L" threshold, a source note, and save to HTML with `gtsave()`.

```{r ex-alt-save, exercise=TRUE, exercise.lines=30}
visits <- c("Baseline", "Week 8", "Week 24")

tbl_final_144 <- tbl_14_4_1

# Add a spanner for each visit
for (v in visits) {
  tbl_final_144 <- tab_spanner(
    tbl_final_144,
    label   = ___,         # use the visit name as spanner label
    columns = starts_with(paste0(v, "_"))
  )
}

tbl_final_144 <- tbl_final_144 |>
  tab_footnote(
    footnote  = "pct_hi: percentage of subjects with ALT > 40 U/L (upper limit of normal).",
    locations = cells_column_labels(columns = ends_with("_pct_hi"))
  ) |>
  tab_source_note("Source: pharmaverseadam::adlb (CDISCPILOT01). Safety Population.") |>
  tab_options(table.font.size = 11)

# Save
gtsave(tbl_final_144, "Table_14_4_1.html")
tbl_final_144
```

```{r ex-alt-save-hint-1}
# tab_spanner(tbl, label = "Visit Name", columns = starts_with("Visit_"))
# For the loop: label = v  (the visit name itself)
# tab_footnote needs a locations argument: cells_column_labels(columns = ...)
# ends_with("_pct_hi") selects all three pct_hi columns at once
```

```{r ex-alt-save-solution}
visits <- c("Baseline", "Week 8", "Week 24")

tbl_final_144 <- tbl_14_4_1

for (v in visits) {
  tbl_final_144 <- tab_spanner(
    tbl_final_144,
    label   = v,
    columns = starts_with(paste0(v, "_"))
  )
}

tbl_final_144 <- tbl_final_144 |>
  tab_footnote(
    footnote  = "pct_hi: percentage of subjects with ALT > 40 U/L (upper limit of normal).",
    locations = cells_column_labels(columns = ends_with("_pct_hi"))
  ) |>
  tab_source_note("Source: pharmaverseadam::adlb (CDISCPILOT01). Safety Population.") |>
  tab_options(table.font.size = 11)

gtsave(tbl_final_144, "Table_14_4_1.html")
tbl_final_144
```

---

## Section 5 — Hard Challenge: Integrated TLF Package

This section assembles all three tables into a single named list — the **TLF package** — and runs a programmatic quality-check function across them. This mirrors a real submission workflow where a lead programmer consolidates output objects before QC and transfer.

### Exercise 5.1 — Build and package all three tables

Complete the code below to produce all three table objects and store them in `tlf_package`. Then run the quality-check loop to verify row counts and flag any all-NA rows.

```{r ex-tlf-package, exercise=TRUE, exercise.lines=55}
# ── Table 14.1.1 ─────────────────────────────────────────────────────────────
tbl_14_1_1 <- tbl_summary(
  data    = t141_adsl,
  by      = "TRT01P",
  include = c("AGE", "AGEGR1", "SEX", "RACE", "WEIGHTBL", "DIABFL")
) |>
  add_p() |> add_overall() |> add_n() |> bold_labels()

# ── Table 14.3.1 ─────────────────────────────────────────────────────────────
ard_143 <- ard_proportion_ci(
  t143_ae_subj,
  variables = c("ANY_AE", "SAE", "SEV_AE"),
  by        = "TRT01P"
)

tbl_14_3_1_body <- ard_143 |>
  mutate(value = as.numeric(unlist(stat))) |>
  filter(stat_name %in% c("n", "p", "conf.low", "conf.high")) |>
  select(label = variable, column = group1_level, param = stat_name, value)

# ── Table 14.4.1 ─────────────────────────────────────────────────────────────
alt_summary <- t144_alt |>
  group_by(TRT01P, AVISIT) |>
  summarise(
    N      = n(),
    Mean   = round(mean(AVAL, na.rm = TRUE), 1),
    SD     = round(sd(AVAL,   na.rm = TRUE), 1),
    .groups = "drop"
  )

# ── Package ───────────────────────────────────────────────────────────────────
tlf_package <- list(
  T1411 = as_tibble(tbl_14_1_1),
  T1431 = tbl_14_3_1_body,
  T1441 = alt_summary
)

# ── Quality check ─────────────────────────────────────────────────────────────
tlf_summary <- function(pkg) {
  lapply(names(pkg), function(nm) {
    tbl        <- pkg[[nm]]
    n_na_rows  <- sum(apply(tbl, 1, function(r) all(is.na(r))))
    any_na_col <- any(sapply(tbl, function(col) anyNA(col)))
    cat(
      nm, "—",
      nrow(tbl), "rows,",
      ncol(tbl), "cols |",
      "all-NA rows:", n_na_rows, "|",
      "any cell NA:", any_na_col, "\n"
    )
  })
  invisible(pkg)
}

tlf_summary(tlf_package)
```

```{r ex-tlf-package-hint-1}
# The code is mostly complete — fill in the group_by and summarise for alt_summary.
# tlf_summary() iterates names(pkg), extracts each tibble, checks rows/cols and NAs.
# apply(tbl, 1, function(r) all(is.na(r))) checks whether every cell in a row is NA.
```

```{r ex-tlf-package-solution}
tbl_14_1_1 <- tbl_summary(
  data    = t141_adsl,
  by      = "TRT01P",
  include = c("AGE", "AGEGR1", "SEX", "RACE", "WEIGHTBL", "DIABFL")
) |>
  add_p() |> add_overall() |> add_n() |> bold_labels()

ard_143 <- ard_proportion_ci(
  t143_ae_subj,
  variables = c("ANY_AE", "SAE", "SEV_AE"),
  by        = "TRT01P"
)

tbl_14_3_1_body <- ard_143 |>
  mutate(value = as.numeric(unlist(stat))) |>
  filter(stat_name %in% c("n", "p", "conf.low", "conf.high")) |>
  select(label = variable, column = group1_level, param = stat_name, value)

alt_summary <- t144_alt |>
  group_by(TRT01P, AVISIT) |>
  summarise(
    N      = n(),
    Mean   = round(mean(AVAL, na.rm = TRUE), 1),
    SD     = round(sd(AVAL,   na.rm = TRUE), 1),
    .groups = "drop"
  )

tlf_package <- list(
  T1411 = as_tibble(tbl_14_1_1),
  T1431 = tbl_14_3_1_body,
  T1441 = alt_summary
)

tlf_summary <- function(pkg) {
  lapply(names(pkg), function(nm) {
    tbl       <- pkg[[nm]]
    n_na_rows <- sum(apply(tbl, 1, function(r) all(is.na(r))))
    any_na_col <- any(sapply(tbl, function(col) anyNA(col)))
    cat(
      nm, "—",
      nrow(tbl), "rows,",
      ncol(tbl), "cols |",
      "all-NA rows:", n_na_rows, "|",
      "any cell NA:", any_na_col, "\n"
    )
  })
  invisible(pkg)
}

tlf_summary(tlf_package)
```

### Exercise 5.2 — Extend tlf_summary with tolerance checking

Extend `tlf_summary()` to also report the **number of numeric columns** and the **range of numeric values** in each table. Numeric range checks help catch outliers that may indicate a coding error (e.g., a proportion > 1 or a count < 0).

```{r ex-tlf-extended, exercise=TRUE, exercise.lines=28}
tlf_summary_extended <- function(pkg) {
  lapply(names(pkg), function(nm) {
    tbl         <- pkg[[nm]]
    n_na_rows   <- sum(apply(tbl, 1, function(r) all(is.na(r))))
    num_cols    <- sapply(tbl, is.numeric)
    n_num       <- sum(num_cols)

    if (n_num > 0) {
      num_vals  <- unlist(tbl[, num_cols])
      val_range <- range(num_vals, na.rm = TRUE)
    } else {
      val_range <- c(NA_real_, NA_real_)
    }

    cat(
      nm, "—",
      nrow(tbl), "rows,", ncol(tbl), "cols |",
      "all-NA rows:", n_na_rows, "|",
      "numeric cols:", n_num, "|",
      "value range: [", round(val_range[1], 2), ",", round(val_range[2], 2), "]\n"
    )
  })
  invisible(pkg)
}

tlf_summary_extended(___)
```

```{r ex-tlf-extended-hint-1}
# The function is complete — just pass tlf_package as the argument
# sapply(tbl, is.numeric) returns a logical vector TRUE for numeric columns
# unlist(tbl[, num_cols]) collects all numeric values; range() gives min and max
```

```{r ex-tlf-extended-solution}
tlf_summary_extended <- function(pkg) {
  lapply(names(pkg), function(nm) {
    tbl         <- pkg[[nm]]
    n_na_rows   <- sum(apply(tbl, 1, function(r) all(is.na(r))))
    num_cols    <- sapply(tbl, is.numeric)
    n_num       <- sum(num_cols)

    if (n_num > 0) {
      num_vals  <- unlist(tbl[, num_cols])
      val_range <- range(num_vals, na.rm = TRUE)
    } else {
      val_range <- c(NA_real_, NA_real_)
    }

    cat(
      nm, "—",
      nrow(tbl), "rows,", ncol(tbl), "cols |",
      "all-NA rows:", n_na_rows, "|",
      "numeric cols:", n_num, "|",
      "value range: [", round(val_range[1], 2), ",", round(val_range[2], 2), "]\n"
    )
  })
  invisible(pkg)
}

tlf_summary_extended(tlf_package)
```

---

## Section 6 — Validation and QC

In clinical programming, **every table must be independently reproduced** by a second programmer. The two outputs are then compared cell by cell. This section practices that workflow.

### Double programming strategy

```
Primary programmer          QC programmer
────────────────            ─────────────
cards → tfrmt → gt          manual dplyr → gt
                                  ↓
                         compare_cells(tbl1, tbl2)
                                  ↓
                         Difference report: none = PASS
```

### Exercise 6.1 — Build the same ALT summary two ways

Compute the ALT mean and SD at Week 8 both with a direct `group_by()` + `summarise()` (primary) and with a different column ordering / variable naming (QC). Then compare the numeric values.

```{r ex-qc-alt, exercise=TRUE, exercise.lines=28}
# Primary approach
alt_primary <- t144_alt |>
  filter(AVISIT == "Week 8") |>
  group_by(TRT01P) |>
  summarise(
    N    = n(),
    Mean = round(mean(AVAL, na.rm = TRUE), 2),
    SD   = round(sd(AVAL,   na.rm = TRUE), 2),
    .groups = "drop"
  )

# QC approach — same calculation, different column names then renamed to match
alt_qc <- t144_alt |>
  filter(AVISIT == "Week 8") |>
  group_by(arm = TRT01P) |>
  summarise(
    count  = n(),
    avg    = round(mean(AVAL, na.rm = TRUE), 2),
    stdev  = round(sd(AVAL,   na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  rename(TRT01P = arm, N = count, Mean = avg, SD = stdev)

# Compare: any rows that differ?
comparison <- anti_join(
  alt_primary,
  alt_qc,
  by = ___          # join on all shared columns to find mismatches
)

cat("Mismatched rows:", nrow(comparison), "\n")
if (nrow(comparison) == 0) cat("QC PASSED — outputs match exactly.\n")
```

```{r ex-qc-alt-hint-1}
# anti_join(primary, qc, by = c("TRT01P","N","Mean","SD"))
# Returns rows in primary that have no match in qc — so zero rows = perfect match
# Make sure both data frames have exactly the same column names before joining
```

```{r ex-qc-alt-solution}
alt_primary <- t144_alt |>
  filter(AVISIT == "Week 8") |>
  group_by(TRT01P) |>
  summarise(
    N    = n(),
    Mean = round(mean(AVAL, na.rm = TRUE), 2),
    SD   = round(sd(AVAL,   na.rm = TRUE), 2),
    .groups = "drop"
  )

alt_qc <- t144_alt |>
  filter(AVISIT == "Week 8") |>
  group_by(arm = TRT01P) |>
  summarise(
    count  = n(),
    avg    = round(mean(AVAL, na.rm = TRUE), 2),
    stdev  = round(sd(AVAL,   na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  rename(TRT01P = arm, N = count, Mean = avg, SD = stdev)

comparison <- anti_join(
  alt_primary,
  alt_qc,
  by = c("TRT01P", "N", "Mean", "SD")
)

cat("Mismatched rows:", nrow(comparison), "\n")
if (nrow(comparison) == 0) cat("QC PASSED — outputs match exactly.\n")
```

### Exercise 6.2 — Check for data gaps with anti_join

Use `anti_join()` to find any arm-visit combinations present in the full `alt_summary` (all three visits) that are missing from a subset restricted to visits where n >= 5. Missing combinations indicate data gaps that could cause NA cells in the final table.

```{r ex-qc-gaps, exercise=TRUE, exercise.lines=20}
# Full set of arm-visit combinations
all_combos <- alt_summary |>
  distinct(TRT01P, AVISIT)

# Subset where N >= 5
adequate_combos <- alt_summary |>
  filter(N >= ___) |>
  distinct(TRT01P, AVISIT)

# Find combinations present in all_combos but missing from adequate_combos
gaps <- anti_join(
  ___,
  ___,
  by = c("TRT01P", "AVISIT")
)

cat("Combinations with N < 5 (data gaps):", nrow(gaps), "\n")
if (nrow(gaps) > 0) print(gaps)
```

```{r ex-qc-gaps-hint-1}
# filter(N >= 5) keeps only adequately-sized groups
# anti_join(all_combos, adequate_combos, by=c("TRT01P","AVISIT"))
# returns rows in all_combos with no match in adequate_combos → the gaps
```

```{r ex-qc-gaps-solution}
all_combos <- alt_summary |>
  distinct(TRT01P, AVISIT)

adequate_combos <- alt_summary |>
  filter(N >= 5) |>
  distinct(TRT01P, AVISIT)

gaps <- anti_join(
  all_combos,
  adequate_combos,
  by = c("TRT01P", "AVISIT")
)

cat("Combinations with N < 5 (data gaps):", nrow(gaps), "\n")
if (nrow(gaps) > 0) print(gaps)
```

### Exercise 6.3 — Write compare_cells()

Write a function `compare_cells(tbl1, tbl2, tol = 0.01)` that:

1. Checks that `tbl1` and `tbl2` have the same dimensions
2. Extracts all numeric columns from both
3. Returns a data frame of `(row, col, val1, val2, diff)` for any cell where `abs(val1 - val2) > tol`
4. Prints "All numeric cells match within tolerance." if no differences found

```{r ex-compare-cells, exercise=TRUE, exercise.lines=38}
compare_cells <- function(tbl1, tbl2, tol = 0.01) {
  # Dimension check
  if (!identical(dim(tbl1), dim(tbl2)))
    stop("Tables have different dimensions: ",
         paste(dim(tbl1), collapse = "x"), " vs ",
         paste(dim(tbl2), collapse = "x"))

  # Identify numeric columns (must have same names)
  num_cols <- intersect(
    names(tbl1)[sapply(tbl1, is.numeric)],
    names(tbl2)[sapply(tbl2, is.numeric)]
  )

  diffs <- lapply(num_cols, function(col) {
    v1   <- tbl1[[col]]
    v2   <- tbl2[[col]]
    diff <- abs(v1 - v2)
    idx  <- which(diff > tol)
    if (length(idx) == 0) return(NULL)
    tibble(row = idx, col = col, val1 = v1[idx], val2 = v2[idx], diff = diff[idx])
  })

  result <- bind_rows(diffs)

  if (nrow(result) == 0) {
    cat("All numeric cells match within tolerance (tol =", tol, ").\n")
  } else {
    cat(nrow(result), "cell(s) differ beyond tolerance.\n")
    print(result)
  }

  invisible(result)
}

# Test it: compare alt_primary with itself (should pass)
compare_cells(___, ___)

# Now introduce a deliberate error
alt_qc_err <- alt_primary |> mutate(Mean = Mean + 0.5)
compare_cells(alt_primary, alt_qc_err)
```

```{r ex-compare-cells-hint-1}
# compare_cells(alt_primary, alt_primary) should print "All numeric cells match"
# compare_cells(alt_primary, alt_qc_err) should report rows where Mean differs by 0.5
# The function is complete — just fill in the two compare_cells() calls
```

```{r ex-compare-cells-solution}
compare_cells <- function(tbl1, tbl2, tol = 0.01) {
  if (!identical(dim(tbl1), dim(tbl2)))
    stop("Tables have different dimensions: ",
         paste(dim(tbl1), collapse = "x"), " vs ",
         paste(dim(tbl2), collapse = "x"))

  num_cols <- intersect(
    names(tbl1)[sapply(tbl1, is.numeric)],
    names(tbl2)[sapply(tbl2, is.numeric)]
  )

  diffs <- lapply(num_cols, function(col) {
    v1   <- tbl1[[col]]
    v2   <- tbl2[[col]]
    diff <- abs(v1 - v2)
    idx  <- which(diff > tol)
    if (length(idx) == 0) return(NULL)
    tibble(row = idx, col = col, val1 = v1[idx], val2 = v2[idx], diff = diff[idx])
  })

  result <- bind_rows(diffs)

  if (nrow(result) == 0) {
    cat("All numeric cells match within tolerance (tol =", tol, ").\n")
  } else {
    cat(nrow(result), "cell(s) differ beyond tolerance.\n")
    print(result)
  }

  invisible(result)
}

compare_cells(alt_primary, alt_primary)

alt_qc_err <- alt_primary |> mutate(Mean = Mean + 0.5)
compare_cells(alt_primary, alt_qc_err)
```

### Quiz

```{r quiz-s6, echo=FALSE}
quiz(
  question(
    "In a double-programming QC workflow, anti_join(primary, qc, by=...) returns rows that...",
    answer("Exist in both primary and qc"),
    answer("Exist in primary but have no match in qc", correct = TRUE),
    answer("Exist in qc but not in primary"),
    answer("Are identical across both"),
    allow_retry = TRUE
  ),
  question(
    "Why is rounding (e.g., round(x, 2)) applied BEFORE using anti_join to compare two summary tables?",
    answer("To reduce file size"),
    answer("Because anti_join only works on integer columns"),
    answer("To avoid floating-point equality failures caused by tiny machine-precision differences",
           correct = TRUE),
    answer("To match SAS output format which rounds to 2 decimal places by default"),
    allow_retry = TRUE
  )
)
```

---

## Lesson Summary

Congratulations — you have completed the **Pharmaverse07 block capstone**.

### What each lesson in this block taught

| Lesson | Tool | Key concept |
|--------|------|-------------|
| 25 | **cards** | Build an ARD with `ard_continuous()`, `ard_categorical()`, `ard_missing()`, `ard_stack()` |
| 26 | **gtsummary** | `tbl_summary()` chain — one function from data to formatted table |
| 27 | **cardinal** | Clinical-trial ARD helpers: `ard_safety()`, SOC/PT hierarchy support |
| 28 | **cardx** | Extensions: `ard_proportion_ci()`, `ard_continuous_ci()`, regression ARDs |
| 29 | **tfrmt** | Template-first: define spec with `frmt_structure()` + `body_plan()`, apply to ARD |
| 30 | **tfrmt advanced** | `row_grp_plan()`, `page_plan()`, `mock_df()` for pre-lock table shells |
| 31 | **gt** | Cell-level control: `tab_spanner()`, `tab_style()`, `tab_options()`, `gtsave()` |
| **32** | **Capstone** | Full pipeline: data → ARD → format spec → gt output + double-programming QC |

### The three tables built in this lesson

| Table | Tool path | Key functions |
|-------|-----------|---------------|
| 14.1.1 Demographics | gtsummary → gt | `tbl_summary()`, `add_p()`, `add_overall()`, `inline_text()` |
| 14.3.1 AE Overview | cardx → tfrmt → gt | `ard_proportion_ci()`, `tfrmt()`, `print_to_gt()`, `mock_df()` |
| 14.4.1 ALT Summary | dplyr → gt | `group_by()`, `pivot_wider()`, `tab_spanner()`, `gtsave()` |

### Core principles to carry forward

1. **Separate statistics from formatting.** The ARD holds numbers; the spec holds display rules. Never mix them.
2. **Mock before you code.** `mock_df()` lets you agree on a layout before data lock — saving rework.
3. **Double-program every table.** `anti_join()` and `compare_cells()` are your allies.
4. **Use typed NAs.** `NA_real_`, `NA_character_`, `NA_integer_` prevent silent type coercion in `mutate()`.
5. **Filter DTYPE rows in adlb.** Admiral summary rows (`DTYPE = "MAXIMUM"` etc.) pollute summaries if not removed.

### Final quiz

```{r quiz-final, echo=FALSE}
quiz(
  question(
    "You need to produce 50 tables that all share the same format but come from 50 different studies. Which tool is best suited?",
    answer("gtsummary — fast defaults"),
    answer("tfrmt — define spec once, apply to any dataset", correct = TRUE),
    answer("gt — maximum cell-level control"),
    answer("cards — ARD is format-free"),
    allow_retry = TRUE
  ),
  question(
    "The ARD pattern says statistics should be stored how?",
    answer("One row per table cell (wide format)"),
    answer("One column per statistic type"),
    answer("One row per statistic, independent of formatting", correct = TRUE),
    answer("In a nested list, one list element per variable"),
    allow_retry = TRUE
  )
)
```

---

**Next lesson:** Lesson 33 — **rtables** — high-performance clinical tables with the rtables package used in Roche and pharma submissions.
