---
title: "Lesson 33 — rtables: Clinical Table Layout Engine"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build complex multi-column clinical tables with rtables: the layout-then-build
  paradigm, column/row splits, analyze functions, custom afuns, pruning, sorting,
  rendering, and a full Demographics Table 14.1.1 capstone.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 33_rtables.Rmd
# PURPOSE       : rtables — clinical table layout engine: basic_table(),
#                 split_cols_by(), split_rows_by(), analyze(), build_table(),
#                 pruning, sorting, rendering, and a Demographics 14.1.1 capstone.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.blanks = TRUE)

# ── ADSL ──────────────────────────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl

if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014)
  n  <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 161, 7)), round(rnorm(n, 174, 8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 68, 12), 1), round(rnorm(n, 82, 14), 1))
  adsl_raw <- adsl_raw |>
    mutate(HEIGHTBL = ht, WEIGHTBL = wt,
           BMIBL    = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N")
  )

# ── ADAE ──────────────────────────────────────────────────────────────────────
adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw)) {
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
} else if (!"TRT01A" %in% names(adae_raw)) {
  adae_raw <- left_join(adae_raw, select(adsl, USUBJID, TRT01A), by = "USUBJID")
}
adae <- adae_raw |>
  filter(SAFFL == "Y") |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

# ── ADLB ──────────────────────────────────────────────────────────────────────
adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ── MOCK rtables ──────────────────────────────────────────────────────────────
# rtables may not be installed in the learnr sandbox.
# These mocks faithfully replicate the API surface used in this lesson.

basic_table <- function(title = "", subtitles = character(),
                        main_footer = character(), prov_footer = character()) {
  structure(
    list(title = title, col_var = NULL, overall_col = NULL,
         row_splits = character(), analyze_vars = character(),
         afun = NULL),
    class = "rtable_layout"
  )
}

split_cols_by <- function(lyt, var, labels_var = NULL, split_fun = NULL) {
  lyt$col_var <- var
  lyt
}

add_overall_col <- function(lyt, label = "All Patients") {
  lyt$overall_col <- label
  lyt
}

split_rows_by <- function(lyt, var, labels_var = NULL, split_fun = NULL,
                          label_pos = "topleft", indent_mod = 0L) {
  lyt$row_splits <- c(lyt$row_splits, var)
  lyt
}

analyze <- function(lyt, vars, afun = NULL, ...) {
  lyt$analyze_vars <- c(lyt$analyze_vars, vars)
  if (!is.null(afun)) lyt$afun <- afun
  lyt
}

in_rows <- function(..., .list = NULL, .labels = NULL) {
  vals <- if (!is.null(.list)) .list else list(...)
  structure(vals, labels = .labels, class = "in_rows")
}

rcell <- function(x, format = NULL, ...) {
  structure(list(value = x, format = format), class = "rcell")
}

mean_sd <- function(x, ...) {
  in_rows(
    .list   = list(rcell(c(mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)),
                         format = "xx.x (xx.x)")),
    .labels = "Mean (SD)"
  )
}

median_range <- function(x, ...) {
  in_rows(
    .list   = list(rcell(c(median(x, na.rm = TRUE),
                           min(x, na.rm = TRUE),
                           max(x, na.rm = TRUE)),
                         format = "xx.x [xx.x, xx.x]")),
    .labels = "Median [Min, Max]"
  )
}

count_fraction <- function(x, ...) {
  n <- length(x)
  in_rows(
    .list   = list(rcell(c(n, n / n * 100), format = "xx (xx.x%)")),
    .labels = "n (%)"
  )
}

# build_table: constructs an rtable object from a layout + data frame
build_table <- function(lyt, df, ...) {
  col_var   <- lyt$col_var
  col_lvls  <- if (!is.null(col_var)) sort(unique(df[[col_var]])) else "All"
  if (!is.null(lyt$overall_col)) col_lvls <- c(col_lvls, lyt$overall_col)

  # Compute row groups
  if (length(lyt$row_splits) > 0) {
    row_grps <- unique(df[, lyt$row_splits, drop = FALSE])
    n_rows   <- nrow(row_grps) * length(lyt$analyze_vars)
  } else {
    n_rows <- length(lyt$analyze_vars)
  }

  structure(
    list(
      layout     = lyt,
      data       = df,
      col_levels = col_lvls,
      n_rows     = n_rows,
      n_cols     = length(col_lvls)
    ),
    class = "rtable"
  )
}

print.rtable <- function(x, ...) {
  cat("rtable:", x$n_rows, "analysis row(s) x", x$n_cols, "col(s)\n")
  cat("Columns :", paste(x$col_levels, collapse = " | "), "\n")
  if (length(x$layout$row_splits) > 0)
    cat("Row splits on:", paste(x$layout$row_splits, collapse = " > "), "\n")
  if (length(x$layout$analyze_vars) > 0)
    cat("Analyzed vars:", paste(x$layout$analyze_vars, collapse = ", "), "\n")
  invisible(x)
}

nrow.rtable   <- function(x, ...) x$n_rows
ncol.rtable   <- function(x, ...) x$n_cols

toString.rtable <- function(x, ...) {
  paste0("rtable with ", x$n_rows, " row(s) and ", x$n_cols, " col(s); ",
         "cols: ", paste(x$col_levels, collapse = " | "))
}

as_html <- function(tbl, ...) {
  cat("[HTML table rendered — ", tbl$n_rows, "rows x", tbl$n_cols, "cols]\n")
  invisible(tbl)
}

prune_table <- function(tbl, ...) {
  # In the real package this removes all-NA/zero rows.
  # Mock: return the table unchanged (no actual zero rows in the mocked engine).
  tbl
}

all_obs_pruning_fun <- function() NULL

sort_at_path <- function(tbl, path, scorefun, ...) {
  # Mock: real package sorts a subtree; here we return unchanged.
  tbl
}

row_paths_summary <- function(tbl) {
  cat("Row path summary — use path strings to address subtrees.\n")
  invisible(NULL)
}

cell_values <- function(tbl, row_path = NULL, col_path = NULL) {
  list()
}
```

## Overview

This lesson introduces **rtables**, Roche's open-source R package for building
complex multi-column clinical tables that match the layout conventions of ICH E3,
FDA guidance, and typical CSR Table 14.1.x specifications.

By the end of this lesson you will be able to:

- Describe the **layout → build** paradigm and explain why it matters
- Create column and row splits to mirror SAS `PROC TABULATE` / `PROC REPORT`
  designs
- Apply built-in and custom analyze functions to compute cell statistics
- Prune empty rows, sort frequency tables, and render output as text or HTML
- Build a complete Demographics Table 14.1.1 from real CDISCPILOT01 `adsl` data

**Packages used:** `rtables` (mocked so exercises run without installation),
`dplyr`, `pharmaverseadam`

---

## Section 1 — The rtables Mental Model

### Layout → Build

The defining feature of rtables is the separation between *what* a table looks
like and *what data* it summarises.

```
layout ──► build_table(lyt, data) ──► table object
```

You first declare a **layout** — a tree of column splits, row splits, and
analysis instructions. Then you **build** the table by passing a data frame
through the layout. The same layout can be re-used on different data frames
(e.g., different study populations).

This is a fundamental shift from SAS `PROC REPORT`:

| Concept | SAS PROC REPORT | rtables |
|---------|----------------|---------|
| Column definition | `DEFINE col / ACROSS` | `split_cols_by()` |
| Row grouping | `DEFINE grp / GROUP` | `split_rows_by()` |
| Computed cell | `COMPUTE block` | `analyze()` + afun |
| Output | `ODS` destination | `as_html()` / `toString()` |

The three layout verbs you will use most are:

```r
basic_table()       # creates the root / title block
split_cols_by()     # one column per level of a variable
split_rows_by()     # one row group per level of a variable
analyze()           # computes statistics and adds analysis rows
```

And the execution verb:

```r
build_table(lyt, df)
```

### Minimal working example

```{r demo-basic, exercise=FALSE}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl)
tbl
```

The table has **one analysis row** (mean/SD of AGE) and **three columns**
(one per treatment arm in `TRT01P`).

### Exercise 1.1 — Add a row split

Add `split_rows_by("SEX")` between `split_cols_by("TRT01P")` and
`analyze("AGE", mean_sd)` so that AGE statistics are computed separately
for each sex.

After building the table, print it and note the value of `nrow(tbl)`.

```{r ex-1-1, exercise=TRUE, exercise.lines=10}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  ___ |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl)
tbl
nrow(tbl)
```

```{r ex-1-1-hint-1}
# Insert split_rows_by("SEX") between split_cols_by and analyze.
# split_rows_by() takes the variable name as a quoted string.
```

```{r ex-1-1-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  split_rows_by("SEX") |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl)
tbl
nrow(tbl)
# nrow is 2 — one analysis row per SEX level (F, M)
```

### Exercise 1.2 — Verify column count

Using the same layout from Exercise 1.1, confirm that `ncol(tbl)` equals 3
(one column per treatment arm).

```{r ex-1-2, exercise=TRUE, exercise.lines=8}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  split_rows_by("SEX") |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl)
ncol(tbl)
```

```{r ex-1-2-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  split_rows_by("SEX") |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl)
ncol(tbl)
# 3 — Placebo | Xanomeline High Dose | Xanomeline Low Dose
```

### Quiz

```{r quiz-1, echo=FALSE}
quiz(
  question(
    "In rtables, which function EXECUTES the layout against a data frame?",
    answer("basic_table()"),
    answer("analyze()"),
    answer("build_table()", correct = TRUE),
    answer("split_cols_by()"),
    allow_retry = TRUE
  ),
  question(
    "What does split_rows_by() do in a layout?",
    answer("Creates one column per level of a variable"),
    answer("Groups rows into one sub-table per level of a variable", correct = TRUE),
    answer("Applies a summary function to each cell"),
    answer("Removes empty rows from the table"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — Column Splits

### Splitting columns by a treatment variable

`split_cols_by("TRT01P")` creates **one column per unique level** of `TRT01P`
in the data you pass to `build_table()`. The CDISCPILOT01 `adsl` dataset has
three arms:

- Placebo
- Xanomeline High Dose
- Xanomeline Low Dose

### Adding an "All Patients" column

`add_overall_col()` appends a grand-total column. Place it after
`split_cols_by()`:

```r
basic_table() |>
  split_cols_by("TRT01P") |>
  add_overall_col("All Patients") |>
  analyze("AGE", mean_sd)
```

This pattern is standard in Demographics Table 14.1.1.

### Exercise 2.1 — Mean AGE by treatment arm

Build a table of AGE mean/SD split by `TRT01P`. Use `adsl |> filter(SAFFL == "Y")`
as the data. Print the result and confirm `ncol(tbl)` is 3.

```{r ex-2-1, exercise=TRUE, exercise.lines=12}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by(___) |>
  analyze("AGE", ___)

tbl <- build_table(lyt, adsl_saf)
tbl
ncol(tbl)
```

```{r ex-2-1-hint-1}
# split_cols_by("TRT01P") creates one column per arm.
# mean_sd is the built-in analyze function for Mean (SD).
```

```{r ex-2-1-solution}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl_saf)
tbl
ncol(tbl)
# 3 — one column per arm
```

### Exercise 2.2 — Add an overall column

Extend the layout from Exercise 2.1 by adding an "All Patients" overall column.
Confirm `ncol(tbl)` is now 4.

```{r ex-2-2, exercise=TRUE, exercise.lines=12}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  ___ |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl_saf)
tbl
ncol(tbl)
```

```{r ex-2-2-hint-1}
# add_overall_col("All Patients") goes between split_cols_by() and analyze().
```

```{r ex-2-2-solution}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  add_overall_col("All Patients") |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl_saf)
tbl
ncol(tbl)
# 4 — Placebo | Xanomeline High Dose | Xanomeline Low Dose | All Patients
```

---

## Section 3 — Row Splits and Nesting

### Grouping rows with split_rows_by()

`split_rows_by()` divides the table into labelled row groups — one group per
level of the splitting variable. In AE tables you routinely nest two splits:
body system (`AEBODSYS`) → preferred term (`AEDECOD`).

```r
basic_table() |>
  split_cols_by("TRT01A") |>
  split_rows_by("AEBODSYS") |>
  split_rows_by("AEDECOD") |>
  analyze("USUBJID", count_fraction)
```

The layout tree is:

```
root
 └─ columns: TRT01A levels
     └─ rows: AEBODSYS levels
         └─ rows: AEDECOD levels
             └─ analysis: count_fraction(USUBJID)
```

### Flat vs nested row count

A flat layout (no row split) over `adae` with one analyze variable produces
**1 analysis row** per column. A nested layout with body system → preferred term
produces **one row per unique AEDECOD** — potentially hundreds of rows.

### Exercise 3.1 — Nested AE layout

Build the nested AE layout below and compare `nrow()` of the flat version
(no row splits) vs the nested version.

```{r ex-3-1, exercise=TRUE, exercise.lines=20}
# Flat version
lyt_flat <- basic_table() |>
  split_cols_by("TRT01A") |>
  analyze("USUBJID", count_fraction)

tbl_flat <- build_table(lyt_flat, adae)

# Nested version
lyt_nested <- basic_table() |>
  split_cols_by("TRT01A") |>
  split_rows_by(___) |>
  split_rows_by(___) |>
  analyze("USUBJID", count_fraction)

tbl_nested <- build_table(lyt_nested, adae)

cat("Flat nrow:", nrow(tbl_flat), "\n")
cat("Nested nrow:", nrow(tbl_nested), "\n")
```

```{r ex-3-1-hint-1}
# First split on "AEBODSYS" (body system), then on "AEDECOD" (preferred term).
```

```{r ex-3-1-solution}
lyt_flat <- basic_table() |>
  split_cols_by("TRT01A") |>
  analyze("USUBJID", count_fraction)

tbl_flat <- build_table(lyt_flat, adae)

lyt_nested <- basic_table() |>
  split_cols_by("TRT01A") |>
  split_rows_by("AEBODSYS") |>
  split_rows_by("AEDECOD") |>
  analyze("USUBJID", count_fraction)

tbl_nested <- build_table(lyt_nested, adae)

cat("Flat nrow:", nrow(tbl_flat), "\n")
cat("Nested nrow:", nrow(tbl_nested), "\n")
# Nested nrow is much larger — one row per unique AEDECOD
```

### Exercise 3.2 — Row split on SEX then AGEGR1

Build a layout that splits rows on `SEX` then `AGEGR1`, analyzing `AGE` with
`mean_sd`, columns by `TRT01P`. How many analysis rows does the result have?

```{r ex-3-2, exercise=TRUE, exercise.lines=12}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  split_rows_by(___) |>
  split_rows_by(___) |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl)
nrow(tbl)
```

```{r ex-3-2-hint-1}
# split_rows_by("SEX") then split_rows_by("AGEGR1").
# nrow counts the total number of analysis rows (one per combination).
```

```{r ex-3-2-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  split_rows_by("SEX") |>
  split_rows_by("AGEGR1") |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl)
nrow(tbl)
# Number of unique (SEX x AGEGR1) combinations in adsl
```

---

## Section 4 — Analyze Functions

### Built-in afuns

rtables ships with three commonly used analysis functions:

| Function | Output | Typical use |
|----------|--------|-------------|
| `mean_sd` | Mean (SD) | Continuous endpoints |
| `median_range` | Median [Min, Max] | Skewed variables |
| `count_fraction` | n (%) | Categorical flags |

These are passed as the `afun` argument to `analyze()`.

### Writing a custom afun

An `afun` receives the **column-filtered vector** `x` and must return an
`in_rows()` object. Use `rcell()` to wrap each value with an optional format
string, and `.labels` to name the rows:

```r
my_q1 <- function(x, ...) {
  in_rows(
    .list   = list(rcell(quantile(x, 0.25, na.rm = TRUE), format = "xx.xx")),
    .labels = "Q1"
  )
}
```

### Exercise 4.1 — Custom Q1/Q3 afun

Write a custom `afun` called `q1_q3` that returns **two rows**: Q1 and Q3 for
`AVAL`. Apply it to `adlb` filtered to the `ALT` parameter, with columns split
by `TRT01A`.

```{r ex-4-1, exercise=TRUE, exercise.lines=24}
q1_q3 <- function(x, ...) {
  in_rows(
    .list = list(
      rcell(quantile(x, ___, na.rm = TRUE), format = "xx.xx"),
      rcell(quantile(x, ___, na.rm = TRUE), format = "xx.xx")
    ),
    .labels = c(___, ___)
  )
}

adlb_alt <- adlb |> filter(PARAMCD == "ALT")

lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  analyze("AVAL", ___)

tbl <- build_table(lyt, adlb_alt)
tbl
```

```{r ex-4-1-hint-1}
# quantile(x, 0.25) = Q1, quantile(x, 0.75) = Q3.
# .labels should be c("Q1", "Q3").
# Pass q1_q3 (without parentheses) as the afun argument to analyze().
```

```{r ex-4-1-solution}
q1_q3 <- function(x, ...) {
  in_rows(
    .list = list(
      rcell(quantile(x, 0.25, na.rm = TRUE), format = "xx.xx"),
      rcell(quantile(x, 0.75, na.rm = TRUE), format = "xx.xx")
    ),
    .labels = c("Q1", "Q3")
  )
}

adlb_alt <- adlb |> filter(PARAMCD == "ALT")

lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  analyze("AVAL", q1_q3)

tbl <- build_table(lyt, adlb_alt)
tbl
```

### Exercise 4.2 — Median and range for WEIGHTBL

Write a layout using the built-in `median_range` afun on `WEIGHTBL` in `adsl`,
columns by `TRT01P`. Print the table.

```{r ex-4-2, exercise=TRUE, exercise.lines=10}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze(___, ___)

tbl <- build_table(lyt, adsl)
tbl
```

```{r ex-4-2-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("WEIGHTBL", median_range)

tbl <- build_table(lyt, adsl)
tbl
```

---

## Section 5 — Multiple Analyze Variables

### Calling analyze() more than once

You can chain multiple `analyze()` calls on the **same** layout to produce
multiple analysis row blocks. Each call may use a different afun:

```r
basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("AGE",      mean_sd)      |>
  analyze("WEIGHTBL", median_range)
```

This mirrors a SAS `PROC TABULATE` table with multiple statistics rows stacked
vertically.

### Exercise 5.1 — Demographics layout with AGE and SEX

Build a layout that analyzes:
- `AGE` with `mean_sd`
- `SEX` with `count_fraction`

Columns split by `TRT01P`, using `adsl |> filter(SAFFL == "Y")`.

```{r ex-5-1, exercise=TRUE, exercise.lines=14}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze(___, ___) |>
  analyze(___, ___)

tbl <- build_table(lyt, adsl_saf)
tbl
nrow(tbl)
```

```{r ex-5-1-hint-1}
# First analyze("AGE", mean_sd), then analyze("SEX", count_fraction).
# nrow will equal the number of analysis rows summed across both analyze() calls.
```

```{r ex-5-1-solution}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("AGE",  mean_sd) |>
  analyze("SEX",  count_fraction)

tbl <- build_table(lyt, adsl_saf)
tbl
nrow(tbl)
# 2 analyze rows (1 per analyze() call)
```

### Exercise 5.2 — Three continuous variables

Extend the layout to also analyze `WEIGHTBL` with `mean_sd`. How many analysis
rows does the table now have?

```{r ex-5-2, exercise=TRUE, exercise.lines=14}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("AGE",      mean_sd) |>
  analyze("SEX",      count_fraction) |>
  analyze(___, ___)

tbl <- build_table(lyt, adsl_saf)
nrow(tbl)
```

```{r ex-5-2-solution}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("AGE",      mean_sd) |>
  analyze("SEX",      count_fraction) |>
  analyze("WEIGHTBL", mean_sd)

tbl <- build_table(lyt, adsl_saf)
nrow(tbl)
# 3 analysis rows
```

---

## Section 6 — Pruning and Sorting

### Why prune?

When you split rows by `AEBODSYS` and `AEDECOD`, some combinations may have
**zero observations** in a treatment arm. These empty rows clutter the output.
`prune_table()` removes them.

```r
tbl_pruned <- prune_table(tbl, all_obs_pruning_fun())
```

`all_obs_pruning_fun()` returns a predicate that prunes rows where **all cells**
across all columns are empty or zero.

### Sorting

`sort_at_path()` reorders a subtree of the table based on a scoring function.
For AE frequency tables you typically sort body systems by **descending total
subject count**:

```r
tbl_sorted <- sort_at_path(
  tbl,
  path     = c("AEBODSYS"),
  scorefun = cont_n_allcols
)
```

The `path` argument identifies the row split node to sort.

> **Note:** Pruning should always come *before* sorting. If you sort first,
> pruning may reorder rows unexpectedly.

### Exercise 6.1 — Build, prune, and inspect

Build the two-level AE table (body system → preferred term), prune it, and
compare `nrow()` before and after pruning.

```{r ex-6-1, exercise=TRUE, exercise.lines=20}
lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  split_rows_by("AEBODSYS") |>
  split_rows_by("AEDECOD") |>
  analyze("USUBJID", count_fraction)

tbl        <- build_table(lyt, adae)
tbl_pruned <- prune_table(tbl, ___)

cat("Before pruning:", nrow(tbl), "\n")
cat("After  pruning:", nrow(tbl_pruned), "\n")
```

```{r ex-6-1-hint-1}
# pass all_obs_pruning_fun() as the second argument to prune_table().
# With the mock, nrow will be the same (no truly empty rows in adae).
```

```{r ex-6-1-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  split_rows_by("AEBODSYS") |>
  split_rows_by("AEDECOD") |>
  analyze("USUBJID", count_fraction)

tbl        <- build_table(lyt, adae)
tbl_pruned <- prune_table(tbl, all_obs_pruning_fun())

cat("Before pruning:", nrow(tbl), "\n")
cat("After  pruning:", nrow(tbl_pruned), "\n")
```

### Exercise 6.2 — Sort then inspect row paths

Apply `sort_at_path()` to the pruned table on the `AEBODSYS` node, then call
`row_paths_summary()`.

```{r ex-6-2, exercise=TRUE, exercise.lines=14}
lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  split_rows_by("AEBODSYS") |>
  analyze("USUBJID", count_fraction)

tbl        <- build_table(lyt, adae)
tbl_pruned <- prune_table(tbl, all_obs_pruning_fun())
tbl_sorted <- sort_at_path(tbl_pruned, path = c(___), scorefun = NULL)

row_paths_summary(tbl_sorted)
```

```{r ex-6-2-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  split_rows_by("AEBODSYS") |>
  analyze("USUBJID", count_fraction)

tbl        <- build_table(lyt, adae)
tbl_pruned <- prune_table(tbl, all_obs_pruning_fun())
tbl_sorted <- sort_at_path(tbl_pruned, path = c("AEBODSYS"), scorefun = NULL)

row_paths_summary(tbl_sorted)
```

---

## Section 7 — Rendering Output

### Three ways to render an rtable

| Function | Output | Use case |
|----------|--------|----------|
| `toString(tbl)` | Character string | Logging, unit tests |
| `as_html(tbl)` | HTML output | R Markdown, Shiny |
| `print(tbl)` | Console summary | Interactive sessions |

`nrow()` and `ncol()` let you assert table dimensions in unit tests:

```r
stopifnot(ncol(tbl) == 3L)
```

### Exercise 7.1 — Render all three ways

Build a simple AGE mean/SD table (columns by `TRT01P`), then:
1. Call `toString(tbl)` and assign to `s`
2. Call `as_html(tbl)`
3. Assert `ncol(tbl) == 3L` with `stopifnot()`

```{r ex-7-1, exercise=TRUE, exercise.lines=18}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl)

# 1) toString
s <- ___(tbl)
cat(s, "\n")

# 2) as_html
___(tbl)

# 3) assert ncol == 3
stopifnot(___ == 3L)
```

```{r ex-7-1-hint-1}
# toString(tbl), as_html(tbl), ncol(tbl) == 3L
```

```{r ex-7-1-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("AGE", mean_sd)

tbl <- build_table(lyt, adsl)

s <- toString(tbl)
cat(s, "\n")

as_html(tbl)

stopifnot(ncol(tbl) == 3L)
```

### Exercise 7.2 — Test dimensions programmatically

Build a table with columns by `TRT01P` and an overall column, analyze `AGE` and
`WEIGHTBL`. Use `stopifnot()` to assert `ncol(tbl) == 4L`.

```{r ex-7-2, exercise=TRUE, exercise.lines=14}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  add_overall_col("All Patients") |>
  analyze("AGE",      mean_sd) |>
  analyze("WEIGHTBL", mean_sd)

tbl <- build_table(lyt, adsl)
stopifnot(ncol(tbl) == ___)
cat("ncol:", ncol(tbl), "— assertion passed\n")
```

```{r ex-7-2-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  add_overall_col("All Patients") |>
  analyze("AGE",      mean_sd) |>
  analyze("WEIGHTBL", mean_sd)

tbl <- build_table(lyt, adsl)
stopifnot(ncol(tbl) == 4L)
cat("ncol:", ncol(tbl), "— assertion passed\n")
```

---

## Section 8 — Capstone: Demographics Table 14.1.1

### Objective

Build a standard Demographics Table 14.1.1 from `adsl` using the Safety
population (`SAFFL == "Y"`):

| Row | Statistic | Function |
|-----|-----------|----------|
| N | Subject count per arm | Custom afun |
| Age | Mean (SD) | `mean_sd` |
| Age | Median [Min, Max] | `median_range` |
| Sex | n (%) | `count_fraction` |
| Race | n (%) | `count_fraction` |

Columns: Placebo | Xanomeline High Dose | Xanomeline Low Dose | All Patients

### Expected N per arm (Safety population)

```{r capstone-n, exercise=FALSE}
adsl |>
  filter(SAFFL == "Y") |>
  count(TRT01P)
```

### Custom N afun

```{r capstone-n-afun, exercise=FALSE}
n_afun <- function(x, ...) {
  in_rows(
    .list   = list(rcell(length(x), format = "xx")),
    .labels = "n"
  )
}
```

### Exercise 8.1 — Assemble the full layout

Complete the layout by filling in the four `___` blanks (two variable names, two
afun names) and build the table.

```{r ex-8-1, exercise=TRUE, exercise.lines=28}
n_afun <- function(x, ...) {
  in_rows(
    .list   = list(rcell(length(x), format = "xx")),
    .labels = "n"
  )
}

adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table(title = "Table 14.1.1  Summary of Demographics") |>
  split_cols_by("TRT01P") |>
  add_overall_col("All Patients") |>
  analyze("USUBJID",  n_afun)       |>
  analyze(___,        mean_sd)      |>   # Age mean (SD)
  analyze("AGE",      ___) |>            # Age median [range]
  analyze(___,        count_fraction) |> # Sex n (%)
  analyze(___,        count_fraction)    # Race n (%)

tbl <- build_table(lyt, adsl_saf)
tbl
```

```{r ex-8-1-hint-1}
# Variables: "AGE", "AGE", "SEX", "RACE"
# Afuns: mean_sd, median_range, count_fraction, count_fraction
```

```{r ex-8-1-solution}
n_afun <- function(x, ...) {
  in_rows(
    .list   = list(rcell(length(x), format = "xx")),
    .labels = "n"
  )
}

adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table(title = "Table 14.1.1  Summary of Demographics") |>
  split_cols_by("TRT01P") |>
  add_overall_col("All Patients") |>
  analyze("USUBJID",  n_afun)        |>
  analyze("AGE",      mean_sd)       |>
  analyze("AGE",      median_range)  |>
  analyze("SEX",      count_fraction)|>
  analyze("RACE",     count_fraction)

tbl <- build_table(lyt, adsl_saf)
tbl
```

### Exercise 8.2 — Verify expected N

Use `dplyr` to count subjects per arm in `adsl_saf`, then use `stopifnot()` to
assert that `ncol(tbl)` equals 4 (three arms + overall).

```{r ex-8-2, exercise=TRUE, exercise.lines=22}
adsl_saf <- adsl |> filter(SAFFL == "Y")

# Step 1: count per arm
arm_counts <- adsl_saf |>
  count(TRT01P, name = "n_subjects")
print(arm_counts)

# Step 2: build the table (reuse layout from 8.1)
n_afun <- function(x, ...) {
  in_rows(.list = list(rcell(length(x), format = "xx")), .labels = "n")
}

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  add_overall_col("All Patients") |>
  analyze("USUBJID",  n_afun) |>
  analyze("AGE",      mean_sd)

tbl <- build_table(lyt, adsl_saf)

# Step 3: assert ncol == 4
stopifnot(ncol(tbl) == ___)
cat("Table has", ncol(tbl), "columns — assertion passed\n")
```

```{r ex-8-2-solution}
adsl_saf <- adsl |> filter(SAFFL == "Y")

arm_counts <- adsl_saf |>
  count(TRT01P, name = "n_subjects")
print(arm_counts)

n_afun <- function(x, ...) {
  in_rows(.list = list(rcell(length(x), format = "xx")), .labels = "n")
}

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  add_overall_col("All Patients") |>
  analyze("USUBJID",  n_afun) |>
  analyze("AGE",      mean_sd)

tbl <- build_table(lyt, adsl_saf)

stopifnot(ncol(tbl) == 4L)
cat("Table has", ncol(tbl), "columns — assertion passed\n")
```

---

## Lesson Summary

You have covered the complete rtables workflow:

| Step | Code |
|------|------|
| Create layout root | `basic_table(title = "...")` |
| Split columns | `split_cols_by("TRT01P")` |
| Add overall column | `add_overall_col("All Patients")` |
| Group rows | `split_rows_by("AEBODSYS")` |
| Compute cells | `analyze("AGE", mean_sd)` |
| Run layout | `build_table(lyt, df)` |
| Remove empties | `prune_table(tbl, all_obs_pruning_fun())` |
| Re-order | `sort_at_path(tbl, path, scorefun)` |
| Render text | `toString(tbl)` |
| Render HTML | `as_html(tbl)` |

**Next lesson (34):** `tern` — pre-built analyze functions that drop directly into
rtables layouts, covering `analyze_vars()`, `count_occurrences()`,
`count_patients_with_flags()`, and confidence-interval functions.
