---
title: "Lesson 34 — tern: Analysis Functions for rtables"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  tern provides pre-built afuns (analyze_vars, count_occurrences,
  count_patients_with_flags, s_summary) that plug into rtables layouts.
  Covers continuous and categorical summarisation, AE counting, serious AE
  flags, custom afuns via s_summary(), and a full AE Summary Table capstone.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 34_tern.Rmd
# PURPOSE       : tern — pre-built rtables analysis functions: analyze_vars(),
#                 count_occurrences(), count_patients_with_flags(), s_summary(),
#                 control_analyze_vars(), estimate_proportion(), and a full
#                 AE Summary Table 14.3.1 capstone.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER:
# This program and its contents represent independent work product. They contain
# absolutely no proprietary items, confidential materials, or intellectual
# property belonging to any past, current, or specific organization or corporate
# entity. All data mapping, standards, and logic used herein are derived
# strictly from public documentation, open-source ecosystems, or simulated data.
# ==============================================================================
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.blanks = TRUE)

# ── ADSL ──────────────────────────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl

if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014)
  n  <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 161, 7)), round(rnorm(n, 174, 8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 68, 12), 1), round(rnorm(n, 82, 14), 1))
  adsl_raw <- adsl_raw |>
    mutate(HEIGHTBL = ht, WEIGHTBL = wt,
           BMIBL    = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
}

adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N")
  )

# ── ADAE ──────────────────────────────────────────────────────────────────────
adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw)) {
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
} else if (!"TRT01A" %in% names(adae_raw)) {
  adae_raw <- left_join(adae_raw, select(adsl, USUBJID, TRT01A), by = "USUBJID")
}
adae <- adae_raw |>
  filter(SAFFL == "Y") |>
  mutate(
    AESEVN  = case_when(
      AESEV == "MILD"     ~ 1L,
      AESEV == "MODERATE" ~ 2L,
      AESEV == "SEVERE"   ~ 3L,
      TRUE                ~ NA_integer_
    ),
    AESERFL = if_else(AESER == "Y", TRUE, FALSE, missing = FALSE)
  )

# ── ADLB ──────────────────────────────────────────────────────────────────────
adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ── MOCK rtables ──────────────────────────────────────────────────────────────

basic_table <- function(title = "", subtitles = character(),
                        main_footer = character(), prov_footer = character()) {
  structure(
    list(title = title, col_var = NULL, overall_col = NULL,
         row_splits = character(), analyze_vars = character(),
         afun = NULL, flag_vars = character(), count_var = NULL,
         row_summary = FALSE, analyze_stats = NULL),
    class = "rtable_layout"
  )
}

split_cols_by <- function(lyt, var, labels_var = NULL, split_fun = NULL) {
  lyt$col_var <- var; lyt
}

add_overall_col <- function(lyt, label = "All Patients") {
  lyt$overall_col <- label; lyt
}

split_rows_by <- function(lyt, var, labels_var = NULL, split_fun = NULL,
                          label_pos = "topleft", indent_mod = 0L) {
  lyt$row_splits <- c(lyt$row_splits, var); lyt
}

analyze <- function(lyt, vars, afun = NULL, ...) {
  lyt$analyze_vars <- c(lyt$analyze_vars, vars)
  if (!is.null(afun)) lyt$afun <- afun
  lyt
}

in_rows <- function(..., .list = NULL, .labels = NULL) {
  vals <- if (!is.null(.list)) .list else list(...)
  structure(vals, labels = .labels, class = "in_rows")
}

rcell <- function(x, format = NULL, ...) {
  structure(list(value = x, format = format), class = "rcell")
}

mean_sd <- function(x, ...) {
  in_rows(
    .list   = list(rcell(c(mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)),
                         format = "xx.x (xx.x)")),
    .labels = "Mean (SD)"
  )
}

median_range <- function(x, ...) {
  in_rows(
    .list   = list(rcell(c(median(x, na.rm = TRUE),
                           min(x, na.rm = TRUE),
                           max(x, na.rm = TRUE)),
                         format = "xx.x [xx.x, xx.x]")),
    .labels = "Median [Min, Max]"
  )
}

count_fraction <- function(x, ...) {
  n <- length(x)
  in_rows(
    .list   = list(rcell(c(n, 100), format = "xx (xx.x%)")),
    .labels = "n (%)"
  )
}

build_table <- function(lyt, df, ...) {
  col_var   <- lyt$col_var
  col_lvls  <- if (!is.null(col_var)) sort(unique(df[[col_var]])) else "All"
  if (!is.null(lyt$overall_col)) col_lvls <- c(col_lvls, lyt$overall_col)

  if (length(lyt$row_splits) > 0) {
    row_grps <- unique(df[, lyt$row_splits, drop = FALSE])
    n_rows   <- max(1L, nrow(row_grps)) * max(1L, length(lyt$analyze_vars))
  } else {
    n_rows <- max(1L, length(lyt$analyze_vars))
  }
  if (isTRUE(lyt$row_summary)) n_rows <- n_rows + 1L

  structure(
    list(layout = lyt, data = df, col_levels = col_lvls,
         n_rows = n_rows, n_cols = length(col_lvls)),
    class = "rtable"
  )
}

print.rtable <- function(x, ...) {
  cat("rtable:", x$n_rows, "analysis row(s) x", x$n_cols, "col(s)\n")
  cat("Columns :", paste(x$col_levels, collapse = " | "), "\n")
  if (length(x$layout$row_splits) > 0)
    cat("Row splits on:", paste(x$layout$row_splits, collapse = " > "), "\n")
  if (length(x$layout$analyze_vars) > 0)
    cat("Analyzed vars:", paste(x$layout$analyze_vars, collapse = ", "), "\n")
  if (length(x$layout$flag_vars) > 0)
    cat("Flag vars    :", paste(x$layout$flag_vars, collapse = ", "), "\n")
  invisible(x)
}

nrow.rtable    <- function(x, ...) x$n_rows
ncol.rtable    <- function(x, ...) x$n_cols
toString.rtable <- function(x, ...) {
  paste0("rtable with ", x$n_rows, " row(s) and ", x$n_cols, " col(s); ",
         "cols: ", paste(x$col_levels, collapse = " | "))
}
as_html <- function(tbl, ...) {
  cat("[HTML table rendered —", tbl$n_rows, "rows x", tbl$n_cols, "cols]\n")
  invisible(tbl)
}
prune_table           <- function(tbl, ...) tbl
all_obs_pruning_fun   <- function() NULL
sort_at_path          <- function(tbl, path, scorefun, ...) tbl
row_paths_summary     <- function(tbl) {
  cat("Row path summary — use path strings to address subtrees.\n")
  invisible(NULL)
}

# ── MOCK tern ─────────────────────────────────────────────────────────────────
# tern functions are curried: they return a function(lyt) that modifies the
# layout.  This mirrors how tern integrates with the rtables pipe in practice.

analyze_vars <- function(vars, var_labels = NULL, show_labels = "default",
                         .stats = c("n", "mean", "sd", "median", "range"),
                         na.rm = TRUE, .indent_mods = NULL, ...) {
  function(lyt) {
    lyt$analyze_vars   <- c(lyt$analyze_vars, vars)
    lyt$analyze_stats  <- .stats
    lyt$show_labels    <- show_labels
    lyt
  }
}

count_occurrences <- function(vars, drop = TRUE, .indent_mods = NULL, ...) {
  function(lyt) {
    lyt$count_var  <- vars
    lyt$count_drop <- drop
    lyt$analyze_vars <- c(lyt$analyze_vars, vars)
    lyt
  }
}

summarize_row_groups <- function(cfun = NULL, ...) {
  function(lyt) {
    lyt$row_summary <- TRUE
    lyt
  }
}

count_patients_with_flags <- function(var, flag_variables,
                                      .indent_mods = NULL, ...) {
  function(lyt) {
    lyt$flag_var   <- flag_variables
    lyt$analyze_vars <- c(lyt$analyze_vars, var)
    lyt
  }
}

# s_summary: the core tern statistics engine for a numeric vector
s_summary <- function(x, na.rm = TRUE, ...) {
  x_clean <- if (na.rm) x[!is.na(x)] else x
  n       <- length(x_clean)
  list(
    n         = n,
    mean      = if (n > 0) mean(x_clean) else NA_real_,
    sd        = if (n > 1) sd(x_clean)   else NA_real_,
    median    = if (n > 0) median(x_clean) else NA_real_,
    min       = if (n > 0) min(x_clean)  else NA_real_,
    max       = if (n > 0) max(x_clean)  else NA_real_,
    range     = if (n > 0) range(x_clean) else c(NA_real_, NA_real_),
    q1_q3     = if (n > 0) quantile(x_clean, c(0.25, 0.75)) else c(NA_real_, NA_real_),
    n_missing = sum(is.na(x))
  )
}

control_analyze_vars <- function(stats = c("n", "mean", "sd", "median", "range"),
                                  ...) {
  list(stats = stats)
}

estimate_proportion <- function(x, conf_level = 0.95, ...) {
  x    <- as.logical(x)
  n    <- sum(x, na.rm = TRUE)
  N    <- sum(!is.na(x))
  prop <- if (N > 0) n / N else NA_real_
  z    <- qnorm(1 - (1 - conf_level) / 2)
  if (!is.na(prop) && N > 0) {
    denom  <- 1 + z^2 / N
    centre <- (prop + z^2 / (2 * N)) / denom
    margin <- z * sqrt(prop * (1 - prop) / N + z^2 / (4 * N^2)) / denom
    ci     <- c(max(0, centre - margin), min(1, centre + margin))
  } else {
    ci <- c(NA_real_, NA_real_)
  }
  list(n = n, N = N, prop = prop, ci = ci)
}

estimate_proportion_diff <- function(x, y, conf_level = 0.95, ...) {
  x  <- as.logical(x); y <- as.logical(y)
  p1 <- mean(x, na.rm = TRUE); p2 <- mean(y, na.rm = TRUE)
  n1 <- sum(!is.na(x));        n2 <- sum(!is.na(y))
  d  <- p1 - p2
  se <- sqrt(p1 * (1 - p1) / n1 + p2 * (1 - p2) / n2)
  z  <- qnorm(1 - (1 - conf_level) / 2)
  list(diff = d, ci = c(d - z * se, d + z * se))
}

format_fraction <- function(num, denom) {
  sprintf("%d/%d (%.1f%%)", as.integer(num), as.integer(denom),
          100 * num / denom)
}

format_count_fraction <- function(x, N) {
  sprintf("%d (%.1f%%)", as.integer(x), 100 * x / N)
}

# build_tern_table: extended build that actually runs s_summary per group
build_tern_table <- function(lyt, df, alt_counts_df = NULL) {
  col_var <- lyt$col_var
  groups  <- if (!is.null(col_var)) split(df, df[[col_var]]) else list(All = df)
  if (!is.null(lyt$overall_col)) groups[[lyt$overall_col]] <- df

  results <- lapply(groups, function(grp) {
    if (length(lyt$analyze_vars) > 0) {
      lapply(lyt$analyze_vars, function(v) {
        if (is.numeric(grp[[v]])) s_summary(grp[[v]])
        else list(n = nrow(grp), levels = unique(grp[[v]]))
      })
    } else {
      list(n = nrow(grp))
    }
  })

  col_lvls <- names(groups)
  n_rows   <- max(1L, length(lyt$analyze_vars))
  if (length(lyt$row_splits) > 0) {
    rg     <- unique(df[, lyt$row_splits, drop = FALSE])
    n_rows <- nrow(rg) * max(1L, length(lyt$analyze_vars))
  }
  if (!is.null(lyt$flag_var) && length(lyt$flag_var) > 0)
    n_rows <- n_rows + length(lyt$flag_var)

  structure(
    list(layout = lyt, data = df, results = results,
         col_levels = col_lvls, n_rows = n_rows, n_cols = length(col_lvls)),
    class = "rtable"
  )
}
```

## Overview

This lesson covers **tern**, the pharmaverse package that sits *on top of*
rtables and provides production-ready analysis functions (afuns) for
ICH-compliant clinical study report tables.

In Lesson 33 you wrote `in_rows() / rcell()` from scratch. tern replaces that
boilerplate with a single function call — `analyze_vars()`, `count_occurrences()`,
or `count_patients_with_flags()` — while keeping full compatibility with the
rtables layout pipeline.

By the end of this lesson you will be able to:

- Explain how tern afuns relate to rtables layouts
- Use `analyze_vars()` for both continuous and categorical variables
- Use `count_occurrences()` for subject-level AE counting
- Use `count_patients_with_flags()` for serious-AE and death-flag rows
- Build a custom afun using `s_summary()`
- Assemble a complete AE Summary Table 14.3.1

**Packages used:** `tern` (mocked), `rtables` (mocked), `dplyr`,
`pharmaverseadam`

---

## Section 1 — tern in the rtables Ecosystem

### The relationship

```
rtables                  tern
───────────────          ───────────────────────────────────────
basic_table()        +   analyze_vars()
split_cols_by()          count_occurrences()
split_rows_by()          count_patients_with_flags()
analyze()                summarize_row_groups()
build_table()            s_summary()
```

tern functions are **not** passed to `analyze()` as a plain afun. Instead they
are **curried** — they return a function that accepts a layout and returns a
modified layout. You use them as pipe steps directly:

```r
basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars("AGE", .stats = c("n", "mean", "sd"))()
```

> The `()` at the end calls the curried function with the layout built so far.
> In production code with real tern installed, tern uses S4 dispatch so this
> syntax is handled transparently. In these exercises the mock works the same way.

### Why tern instead of raw rtables afuns?

With raw rtables you would write:

```r
my_age_afun <- function(x, ...) {
  in_rows(
    .list = list(
      rcell(length(x[!is.na(x)]),                           format = "xx"),
      rcell(c(mean(x, na.rm=TRUE), sd(x, na.rm=TRUE)),      format = "xx.x (xx.x)"),
      rcell(c(min(x,na.rm=TRUE), max(x,na.rm=TRUE)),        format = "xx.x, xx.x")
    ),
    .labels = c("n", "Mean (SD)", "Min, Max")
  )
}
analyze("AGE", my_age_afun)
```

With tern you write:

```r
analyze_vars("AGE", .stats = c("n", "mean", "sd", "range"))()
```

### Exercise 1.1 — Match function to use-case

For each use-case, write the tern function name you would use. Answer by
running code that prints your answers.

```{r ex-1-1, exercise=TRUE, exercise.lines=10}
# Complete the character vector with tern function names:
# Options: analyze_vars, count_occurrences, count_patients_with_flags

answers <- c(
  "Continuous lab values (mean, SD, median, range)"  = "___",
  "AE counts by body system (subject-level)"         = "___",
  "Subjects with serious AEs (AESER == 'Y')"         = "___"
)
print(answers)
```

```{r ex-1-1-hint-1}
# Continuous variables -> analyze_vars
# Subject-level occurrence counts -> count_occurrences
# Flag-based counts -> count_patients_with_flags
```

```{r ex-1-1-solution}
answers <- c(
  "Continuous lab values (mean, SD, median, range)"  = "analyze_vars",
  "AE counts by body system (subject-level)"         = "count_occurrences",
  "Subjects with serious AEs (AESER == 'Y')"         = "count_patients_with_flags"
)
print(answers)
```

### Exercise 1.2 — Side-by-side comparison

Below is a raw-rtables layout for AGE mean/SD. Rewrite the `analyze()` line
using tern's `analyze_vars()` instead.

```{r ex-1-2, exercise=TRUE, exercise.lines=14}
# Raw rtables version (given — do not change):
lyt_raw <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze("AGE", mean_sd)

# tern version — replace analyze("AGE", mean_sd) with analyze_vars():
lyt_tern <- basic_table() |>
  split_cols_by("TRT01P") |>
  ___("AGE", .stats = c("n", "mean", "sd"))(___)

tbl_tern <- build_tern_table(lyt_tern, adsl)
cat("tern ncol:", ncol(tbl_tern), "\n")
```

```{r ex-1-2-hint-1}
# analyze_vars("AGE", .stats = c("n","mean","sd")) returns a function.
# Call it with (lyt_tern_so_far) by appending () to the pipe result,
# or pass the layout: analyze_vars("AGE", .stats = ...)(lyt_so_far).
# In the mock, the pipe step looks like: analyze_vars("AGE", ...)()
# where the () applies the returned function to the layout built so far.
```

```{r ex-1-2-solution}
lyt_tern <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars("AGE", .stats = c("n", "mean", "sd"))()

tbl_tern <- build_tern_table(lyt_tern, adsl)
cat("tern ncol:", ncol(tbl_tern), "\n")
```

### Quiz

```{r quiz-1, echo=FALSE}
quiz(
  question(
    "What does tern's analyze_vars() return when called?",
    answer("An rtable object ready to print"),
    answer("A function that accepts a layout and returns a modified layout",
           correct = TRUE),
    answer("A data frame of summary statistics"),
    answer("A character vector of statistic labels"),
    allow_retry = TRUE
  ),
  question(
    "Which tern function is specifically designed to count SUBJECTS (not rows) with AEs?",
    answer("analyze_vars()"),
    answer("s_summary()"),
    answer("count_occurrences()", correct = TRUE),
    answer("summarize_row_groups()"),
    allow_retry = TRUE
  )
)
```

---

## Section 2 — analyze_vars() for Continuous Variables

### Default statistics

`analyze_vars()` detects numeric variables automatically and produces:
`n`, `mean`, `sd`, `median`, `min`, `max`, and `range`.

You can customise with `.stats`:

```r
analyze_vars("AGE",
             .stats = c("n", "mean", "sd", "median", "range"))()
```

The `control_analyze_vars()` helper builds the stats list for you, which is
useful when applying the same configuration to many variables:

```r
ctrl <- control_analyze_vars(stats = c("n", "mean", "sd"))
analyze_vars("AGE", .stats = ctrl$stats)()
```

### Exercise 2.1 — AGE and WEIGHTBL with custom stats

Build a layout with `analyze_vars()` for both `AGE` and `WEIGHTBL`, requesting
only `n`, `mean`, `sd`, and `range`. Columns split by `TRT01P`. Use
`build_tern_table()`.

```{r ex-2-1, exercise=TRUE, exercise.lines=16}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars(___, .stats = c("n", "mean", "sd", "range"))() |>
  analyze_vars(___, .stats = c("n", "mean", "sd", "range"))()

tbl <- build_tern_table(lyt, adsl)
tbl
```

```{r ex-2-1-hint-1}
# First analyze_vars("AGE", ...) then analyze_vars("WEIGHTBL", ...).
# Each call is a separate pipe step, each followed by ().
```

```{r ex-2-1-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars("AGE",      .stats = c("n", "mean", "sd", "range"))() |>
  analyze_vars("WEIGHTBL", .stats = c("n", "mean", "sd", "range"))()

tbl <- build_tern_table(lyt, adsl)
tbl
```

### Exercise 2.2 — s_summary() exploration

Call `s_summary()` directly on the `AGE` column of the safety population.
Inspect the returned list and print the `n`, `mean`, and `q1_q3` elements.

```{r ex-2-2, exercise=TRUE, exercise.lines=12}
adsl_saf <- adsl |> filter(SAFFL == "Y")

stats <- s_summary(adsl_saf$AGE)

cat("n       :", stats$n,    "\n")
cat("mean    :", round(stats$___, 2), "\n")
cat("Q1, Q3  :", round(stats$___[1], 2), ",", round(stats$___[2], 2), "\n")
```

```{r ex-2-2-hint-1}
# Access list elements with $mean and $q1_q3.
# q1_q3 is a length-2 vector: q1_q3[1] = Q1, q1_q3[2] = Q3.
```

```{r ex-2-2-solution}
adsl_saf <- adsl |> filter(SAFFL == "Y")

stats <- s_summary(adsl_saf$AGE)

cat("n       :", stats$n,               "\n")
cat("mean    :", round(stats$mean, 2),  "\n")
cat("Q1, Q3  :", round(stats$q1_q3[1], 2), ",",
                 round(stats$q1_q3[2], 2), "\n")
```

---

## Section 3 — analyze_vars() for Categorical Variables

### Auto-detection

When the variable passed to `analyze_vars()` is a **character or factor**,
tern automatically switches to count/percentage output (equivalent to
`count_fraction` from Lesson 33).

The `show_labels` argument controls row labelling:

| Value | Effect |
|-------|--------|
| `"default"` | Labels shown based on context |
| `"visible"` | Variable name always appears as a header row |
| `"hidden"` | Variable name suppressed; counts appear without header |

### Exercise 3.1 — Add SEX as categorical variable

Extend the layout from Section 2 Exercise 2.1 to also include `SEX` with
`show_labels = "visible"`. Use `adsl |> filter(SAFFL == "Y")`.

```{r ex-3-1, exercise=TRUE, exercise.lines=16}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars("AGE",      .stats = c("n", "mean", "sd"))() |>
  analyze_vars("WEIGHTBL", .stats = c("n", "mean", "sd"))() |>
  analyze_vars(___, show_labels = ___)()

tbl <- build_tern_table(lyt, adsl_saf)
tbl
```

```{r ex-3-1-hint-1}
# analyze_vars("SEX", show_labels = "visible")()
```

```{r ex-3-1-solution}
adsl_saf <- adsl |> filter(SAFFL == "Y")

lyt <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars("AGE",      .stats = c("n", "mean", "sd"))() |>
  analyze_vars("WEIGHTBL", .stats = c("n", "mean", "sd"))() |>
  analyze_vars("SEX",      show_labels = "visible")()

tbl <- build_tern_table(lyt, adsl_saf)
tbl
```

### Exercise 3.2 — Compare show_labels settings

Build the same single-variable layout twice — once with `show_labels = "visible"`
and once with `show_labels = "hidden"` — and compare the `n_rows` values.

```{r ex-3-2, exercise=TRUE, exercise.lines=20}
lyt_visible <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars("SEX", show_labels = ___)()

lyt_hidden <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars("SEX", show_labels = ___)()

tbl_v <- build_tern_table(lyt_visible, adsl)
tbl_h <- build_tern_table(lyt_hidden,  adsl)

cat("show_labels = 'visible' nrow:", nrow(tbl_v), "\n")
cat("show_labels = 'hidden'  nrow:", nrow(tbl_h), "\n")
```

```{r ex-3-2-solution}
lyt_visible <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars("SEX", show_labels = "visible")()

lyt_hidden <- basic_table() |>
  split_cols_by("TRT01P") |>
  analyze_vars("SEX", show_labels = "hidden")()

tbl_v <- build_tern_table(lyt_visible, adsl)
tbl_h <- build_tern_table(lyt_hidden,  adsl)

cat("show_labels = 'visible' nrow:", nrow(tbl_v), "\n")
cat("show_labels = 'hidden'  nrow:", nrow(tbl_h), "\n")
```

---

## Section 4 — count_occurrences() for AE Tables

### Subject-level counting

AE datasets contain **one row per event**, not per subject. A single subject can
contribute many rows to `adae`. If you naively count rows you overcount subjects.

`count_occurrences()` counts **unique USUBJID values** within each column-group
for each row-split category. This is the subject-level incidence count required
by FDA guidance for AE summary tables.

Key argument: `drop = TRUE` removes row categories that have zero subjects in
**every** column (equivalent to `prune_table()` after the fact).

### Typical AE table layout

```r
basic_table() |>
  split_cols_by("TRT01A") |>
  split_rows_by("AEBODSYS", label_pos = "topleft") |>
  summarize_row_groups()() |>
  count_occurrences("AEDECOD")()
```

`summarize_row_groups()` adds a body-system subtotal row above the preferred
terms — this is the "Any AE in <body system>" row.

### Exercise 4.1 — Count subjects per body system

Build a layout with columns by `TRT01A`, rows split on `AEBODSYS`, and
`count_occurrences("AEDECOD")`. Use `build_tern_table(lyt, adae)`.
Print the result.

```{r ex-4-1, exercise=TRUE, exercise.lines=14}
lyt <- basic_table() |>
  split_cols_by(___) |>
  split_rows_by("AEBODSYS", label_pos = "topleft") |>
  count_occurrences(___)()

tbl <- build_tern_table(lyt, adae)
tbl
```

```{r ex-4-1-hint-1}
# split_cols_by("TRT01A") — TRTA was renamed to TRT01A in the setup chunk.
# count_occurrences("AEDECOD")() counts subjects per preferred term.
```

```{r ex-4-1-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  split_rows_by("AEBODSYS", label_pos = "topleft") |>
  count_occurrences("AEDECOD")()

tbl <- build_tern_table(lyt, adae)
tbl
```

### Exercise 4.2 — NERVOUS SYSTEM DISORDERS subject count

Using `dplyr`, count the number of **unique subjects** with at least one AE in
the "NERVOUS SYSTEM DISORDERS" body system for each treatment arm. This gives
you the expected subject count that `count_occurrences()` would place in each
cell.

```{r ex-4-2, exercise=TRUE, exercise.lines=12}
adae |>
  filter(AEBODSYS == ___) |>
  group_by(___) |>
  summarise(n_subjects = n_distinct(___), .groups = "drop")
```

```{r ex-4-2-hint-1}
# filter(AEBODSYS == "NERVOUS SYSTEM DISORDERS")
# group_by(TRT01A)
# n_distinct(USUBJID) counts unique subjects per arm
```

```{r ex-4-2-solution}
adae |>
  filter(AEBODSYS == "NERVOUS SYSTEM DISORDERS") |>
  group_by(TRT01A) |>
  summarise(n_subjects = n_distinct(USUBJID), .groups = "drop")
```

---

## Section 5 — count_patients_with_flags()

### Flagging specific AE categories

For regulatory AE tables you need rows like:

- Subjects with any serious AE (AESER == "Y")
- Subjects with an AE leading to death (AESDTH == "Y")
- Subjects with an AE leading to drug withdrawal (AEACN == "DRUG WITHDRAWN")

`count_patients_with_flags()` takes:

- `var`: the subject identifier variable (usually `"USUBJID"`)
- `flag_variables`: a named character vector mapping display label → column name

```r
count_patients_with_flags(
  "USUBJID",
  flag_variables = c("Serious AEs" = "AESERFL",
                     "Fatal AEs"   = "AEDTHFL")
)()
```

The flag columns must be **logical** (`TRUE` / `FALSE`) in the data frame.

### Exercise 5.1 — Derive AESERFL and count serious AEs

`adae` in this lesson already has `AESERFL = (AESER == "Y")` (derived in
the setup chunk). Count the number of subjects with a serious AE per arm
using `dplyr`. This is the expected value that `count_patients_with_flags()`
would produce.

```{r ex-5-1, exercise=TRUE, exercise.lines=12}
# Verify AESERFL exists and is logical
cat("Class of AESERFL:", class(adae$AESERFL), "\n")

adae |>
  filter(___) |>
  group_by(TRT01A) |>
  summarise(n_serious = n_distinct(USUBJID), .groups = "drop")
```

```{r ex-5-1-hint-1}
# filter(AESERFL == TRUE) or filter(AESERFL) to keep only serious AE rows.
# Then count unique USUBJIDs per arm.
```

```{r ex-5-1-solution}
cat("Class of AESERFL:", class(adae$AESERFL), "\n")

adae |>
  filter(AESERFL) |>
  group_by(TRT01A) |>
  summarise(n_serious = n_distinct(USUBJID), .groups = "drop")
```

### Exercise 5.2 — Add flags row to a layout

Build a layout with columns by `TRT01A` and two analyze steps:
1. `count_occurrences("AEBODSYS")` for all-AE body-system counts
2. `count_patients_with_flags("USUBJID", flag_variables = c("Serious AEs" = "AESERFL"))`

```{r ex-5-2, exercise=TRUE, exercise.lines=16}
lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  count_occurrences("AEBODSYS")() |>
  count_patients_with_flags(
    ___,
    flag_variables = c("Serious AEs" = ___)
  )()

tbl <- build_tern_table(lyt, adae)
tbl
```

```{r ex-5-2-hint-1}
# var = "USUBJID", flag_variables = c("Serious AEs" = "AESERFL")
```

```{r ex-5-2-solution}
lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  count_occurrences("AEBODSYS")() |>
  count_patients_with_flags(
    "USUBJID",
    flag_variables = c("Serious AEs" = "AESERFL")
  )()

tbl <- build_tern_table(lyt, adae)
tbl
```

---

## Section 6 — s_summary() and Custom Afuns

### When to use s_summary() directly

`s_summary()` is the underlying statistics engine called by `analyze_vars()`.
You use it directly when you need a **custom row set** not covered by
`analyze_vars()`'s built-in stats options, for example:

- Mean (SD) on the first row, Median [Q1, Q3] on the second, N Missing on a
  third row

### Pattern

```r
my_afun <- function(x, ...) {
  s <- s_summary(x)
  in_rows(
    .list = list(
      rcell(c(s$mean, s$sd),        format = "xx.x (xx.x)"),
      rcell(c(s$median,
              s$q1_q3[1], s$q1_q3[2]),  format = "xx.x [xx.x, xx.x]"),
      rcell(s$n_missing,            format = "xx")
    ),
    .labels = c("Mean (SD)", "Median [Q1, Q3]", "N Missing")
  )
}
```

> **Tip:** Always type-check `x` at the start of a custom afun with
> `stopifnot(is.numeric(x))` to catch accidental non-numeric variables early.

### Exercise 6.1 — Write a three-row custom afun

Write an afun called `custom_stats` using `s_summary()` that returns three rows:
1. Mean (SD)
2. Median [Q1, Q3]
3. N Missing

Apply it to `AVAL` in `adlb` filtered to `PARAMCD == "ALT"`, columns by `TRT01A`.

```{r ex-6-1, exercise=TRUE, exercise.lines=28}
custom_stats <- function(x, ...) {
  s <- s_summary(x)
  in_rows(
    .list = list(
      rcell(c(s$___, s$___),
            format = "xx.x (xx.x)"),
      rcell(c(s$median, s$q1_q3[___], s$q1_q3[___]),
            format = "xx.x [xx.x, xx.x]"),
      rcell(s$___,
            format = "xx")
    ),
    .labels = c("Mean (SD)", "Median [Q1, Q3]", "N Missing")
  )
}

adlb_alt <- adlb |> filter(PARAMCD == "ALT")

lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  analyze("AVAL", ___)

tbl <- build_table(lyt, adlb_alt)
tbl
```

```{r ex-6-1-hint-1}
# s$mean, s$sd, s$median, s$q1_q3[1] (Q1), s$q1_q3[2] (Q3), s$n_missing
# Pass custom_stats (without parentheses) as afun to analyze().
```

```{r ex-6-1-solution}
custom_stats <- function(x, ...) {
  s <- s_summary(x)
  in_rows(
    .list = list(
      rcell(c(s$mean, s$sd),
            format = "xx.x (xx.x)"),
      rcell(c(s$median, s$q1_q3[1], s$q1_q3[2]),
            format = "xx.x [xx.x, xx.x]"),
      rcell(s$n_missing,
            format = "xx")
    ),
    .labels = c("Mean (SD)", "Median [Q1, Q3]", "N Missing")
  )
}

adlb_alt <- adlb |> filter(PARAMCD == "ALT")

lyt <- basic_table() |>
  split_cols_by("TRT01A") |>
  analyze("AVAL", custom_stats)

tbl <- build_table(lyt, adlb_alt)
tbl
```

### Exercise 6.2 — estimate_proportion() for a binary endpoint

Use `estimate_proportion()` to estimate the proportion and 95% Wilson CI of
subjects with any serious AE in the Placebo arm.

```{r ex-6-2, exercise=TRUE, exercise.lines=16}
placebo_ae <- adae |>
  filter(TRT01A == "Placebo") |>
  # one row per subject (take max flag per subject)
  group_by(USUBJID) |>
  summarise(any_serious = any(AESERFL), .groups = "drop")

res <- estimate_proportion(placebo_ae$___)

cat(sprintf("n serious  : %d\n",   res$n))
cat(sprintf("N total    : %d\n",   res$N))
cat(sprintf("Proportion : %.3f\n", res$prop))
cat(sprintf("95%% CI     : [%.3f, %.3f]\n", res$ci[1], res$ci[2]))
```

```{r ex-6-2-hint-1}
# estimate_proportion() expects a logical vector (TRUE/FALSE).
# Use placebo_ae$any_serious as the argument.
```

```{r ex-6-2-solution}
placebo_ae <- adae |>
  filter(TRT01A == "Placebo") |>
  group_by(USUBJID) |>
  summarise(any_serious = any(AESERFL), .groups = "drop")

res <- estimate_proportion(placebo_ae$any_serious)

cat(sprintf("n serious  : %d\n",   res$n))
cat(sprintf("N total    : %d\n",   res$N))
cat(sprintf("Proportion : %.3f\n", res$prop))
cat(sprintf("95%% CI     : [%.3f, %.3f]\n", res$ci[1], res$ci[2]))
```

---

## Section 7 — Capstone: AE Summary Table 14.3.1

### Objective

Build a standard AE Summary Table (Table 14.3.1 style) using `adae`:

```
                          Placebo   Xan. High   Xan. Low
                          (n=XX)    (n=XX)      (n=XX)
Any Adverse Event          XX (XX%)  XX (XX%)    XX (XX%)
  Serious AEs              XX        XX          XX

NERVOUS SYSTEM DISORDERS
  Headache                 XX        XX          XX
  Dizziness                XX        XX          XX
  ...

GASTROINTESTINAL DISORDERS
  ...
```

Columns: TRT01A arms (no overall column for this table)

Rows:
1. "Subjects with any AE" flag row — `count_patients_with_flags` on `ANYAEFL`
   (derived below)
2. "Serious AEs" flag row — `count_patients_with_flags` on `AESERFL`
3. Body system → preferred term — `split_rows_by + count_occurrences`

### Derive ANYAEFL

```{r capstone-anyaefl, exercise=FALSE}
# Each row of adae is an AE — every USUBJID in adae has at least one AE.
adae <- adae |> mutate(ANYAEFL = TRUE)
```

### Expected subject counts (dplyr validation)

```{r capstone-dplyr-check, exercise=FALSE}
# Subjects with any AE per arm
adae |>
  group_by(TRT01A) |>
  summarise(n_any_ae = n_distinct(USUBJID), .groups = "drop")
```

### Exercise 7.1 — Complete the capstone layout

Fill in the four `___` blanks to build the full AE Summary Table.

```{r ex-7-1, exercise=TRUE, exercise.lines=30}
adae <- adae |> mutate(ANYAEFL = TRUE)

lyt <- basic_table(title = "Table 14.3.1  Summary of Adverse Events") |>
  split_cols_by(___) |>
  # --- Flag rows at the top ---
  count_patients_with_flags(
    "USUBJID",
    flag_variables = c("Subjects with any AE" = "ANYAEFL",
                       "Subjects with serious AE" = ___)
  )() |>
  # --- Body system -> preferred term ---
  split_rows_by("AEBODSYS", label_pos = "topleft") |>
  summarize_row_groups()() |>
  count_occurrences(___)()

tbl <- build_tern_table(lyt, ___)
tbl
```

```{r ex-7-1-hint-1}
# split_cols_by("TRT01A")
# flag = "AESERFL"
# count_occurrences("AEDECOD")
# build_tern_table(lyt, adae)
```

```{r ex-7-1-solution}
adae <- adae |> mutate(ANYAEFL = TRUE)

lyt <- basic_table(title = "Table 14.3.1  Summary of Adverse Events") |>
  split_cols_by("TRT01A") |>
  count_patients_with_flags(
    "USUBJID",
    flag_variables = c("Subjects with any AE"      = "ANYAEFL",
                       "Subjects with serious AE"  = "AESERFL")
  )() |>
  split_rows_by("AEBODSYS", label_pos = "topleft") |>
  summarize_row_groups()() |>
  count_occurrences("AEDECOD")()

tbl <- build_tern_table(lyt, adae)
tbl
```

### Exercise 7.2 — Count subjects with any AE in Placebo arm

Use `dplyr` to compute the number of unique subjects with any AE in the Placebo
arm. This is the expected value for the "Subjects with any AE" cell in the
Placebo column of Table 14.3.1.

```{r ex-7-2, exercise=TRUE, exercise.lines=10}
adae |>
  filter(TRT01A == ___) |>
  summarise(n_subjects_any_ae = n_distinct(___))
```

```{r ex-7-2-hint-1}
# filter(TRT01A == "Placebo")
# n_distinct(USUBJID) gives the subject count
```

```{r ex-7-2-solution}
adae |>
  filter(TRT01A == "Placebo") |>
  summarise(n_subjects_any_ae = n_distinct(USUBJID))
```

---

## Lesson Summary

| tern function | Purpose | Equivalent in raw rtables |
|---------------|---------|--------------------------|
| `analyze_vars()` | Continuous and categorical summaries | `analyze(var, custom_afun)` |
| `count_occurrences()` | Subject-level AE counts | Manual `n_distinct()` afun |
| `count_patients_with_flags()` | Flag-based subject counts | Manual logical filter afun |
| `summarize_row_groups()` | Subtotal row per split group | Manual `in_rows()` with group total |
| `s_summary()` | Named statistics list (building block) | Direct `mean()`, `sd()`, etc. |
| `estimate_proportion()` | Proportion + Wilson CI | Manual `binom.test()` |
| `control_analyze_vars()` | Reusable stats config | Hard-coded `.stats` vector |

**Key rule:** `count_occurrences()` counts *subjects*, not rows. Always use it
(not `count_fraction`) when the same subject can contribute multiple AE rows.

**Next lesson (35):** `r2rtf` — R-to-RTF table rendering for submission-ready
Word/RTF output.
