---
title: "Lesson 35 — r2rtf: RTF Output for Clinical Submission Tables"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  Build submission-grade RTF clinical tables using r2rtf: the data-first
  philosophy, page setup, titles, column headers, body formatting, footnotes,
  and a full capstone AE summary table.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 35_r2rtf.Rmd
# PURPOSE       : Teach r2rtf for producing submission-grade RTF clinical tables:
#                 page setup, titles, column headers, body formatting,
#                 footnotes, number formatting, and a capstone AE table.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.blanks = TRUE)

# ── Null-coalesce operator ────────────────────────────────────────────────────
`%||%` <- function(a, b) if (!is.null(a)) a else b

# ── r2rtf mock functions ──────────────────────────────────────────────────────

rtf_page <- function(df, orientation = "landscape", margin = rep(1, 4),
                     nrow = 40, font_size = 9, ...) {
  attr(df, "rtf_page") <- list(
    orientation = orientation,
    margin      = margin,
    nrow        = nrow,
    font_size   = font_size
  )
  df
}

rtf_title <- function(df, title = "", ...) {
  prev <- attr(df, "rtf_titles") %||% character()
  attr(df, "rtf_titles") <- c(prev, title)
  df
}

rtf_colheader <- function(df, colheader, col_rel_width = NULL,
                           border_top = NULL, border_bottom = NULL,
                           text_justification = NULL, ...) {
  prev <- attr(df, "rtf_colheaders") %||% character()
  attr(df, "rtf_colheaders") <- c(prev, colheader)
  df
}

rtf_body <- function(df, col_rel_width = NULL, text_justification = NULL,
                     border_top = NULL, border_bottom = NULL,
                     cell_height = 0.15, stripes = NULL, ...) {
  attr(df, "rtf_body") <- list(
    col_rel_width      = col_rel_width,
    text_justification = text_justification
  )
  df
}

rtf_footnote <- function(df, footnote = "", ...) {
  prev <- attr(df, "rtf_footnotes") %||% character()
  attr(df, "rtf_footnotes") <- c(prev, footnote)
  df
}

rtf_source <- function(df, source = "", ...) {
  attr(df, "rtf_source") <- source
  df
}

rtf_encode <- function(df, ...) {
  structure(
    list(
      data       = df,
      page       = attr(df, "rtf_page"),
      titles     = attr(df, "rtf_titles"),
      colheaders = attr(df, "rtf_colheaders"),
      body       = attr(df, "rtf_body"),
      footnotes  = attr(df, "rtf_footnotes"),
      source     = attr(df, "rtf_source"),
      nrow_body  = nrow(df),
      ncol_body  = ncol(df)
    ),
    class = "rtf_doc"
  )
}

rtf_write <- function(rtf, path = "output.rtf") {
  cat(sprintf("RTF document written to: %s\n", path))
  cat(sprintf("  Page        : %s, font_size = %s pt\n",
      rtf$page$orientation %||% "landscape",
      rtf$page$font_size   %||% 9))
  cat(sprintf("  Titles      : %d line(s)\n",   length(rtf$titles)))
  cat(sprintf("  Col headers : %d row(s)\n",    length(rtf$colheaders)))
  cat(sprintf("  Body rows   : %d\n",            rtf$nrow_body))
  cat(sprintf("  Footnotes   : %d line(s)\n",   length(rtf$footnotes)))
  invisible(rtf)
}

print.rtf_doc <- function(x, ...) {
  cat("rtf_doc:\n")
  cat("  Titles :", paste(x$titles, collapse = " | "), "\n")
  cat("  Rows   :", x$nrow_body, " Cols:", x$ncol_body, "\n")
  invisible(x)
}

# ── Real pharmaverseadam data ─────────────────────────────────────────────────

adsl <- pharmaverseadam::adsl

# Age summary by treatment arm
age_summary <- adsl |>
  group_by(TRT01P) |>
  summarise(
    N      = n(),
    Mean   = round(mean(AGE, na.rm = TRUE), 1),
    SD     = round(sd(AGE,   na.rm = TRUE), 1),
    Median = round(median(AGE, na.rm = TRUE), 1),
    Min    = min(AGE, na.rm = TRUE),
    Max    = max(AGE, na.rm = TRUE),
    .groups = "drop"
  )

# Wide-format demographics table (r2rtf-ready character columns)
demog_rtf <- tribble(
  ~Parameter,     ~Statistic,  ~Placebo,      ~XanoLow,      ~XanoHigh,
  "Age (years)",  "Mean (SD)", "75.2 (8.6)",  "75.7 (8.3)",  "74.4 (7.9)",
  "",             "Median",    "76.0",         "77.5",         "76.0",
  "",             "Range",     "[52, 89]",     "[56, 88]",     "[56, 88]",
  "Sex, n (%)",   "Male",      "37 (43.0%)",   "35 (41.7%)",   "35 (41.7%)",
  "",             "Female",    "49 (57.0%)",   "49 (58.3%)",   "49 (58.3%)",
  "Race, n (%)",  "White",     "68 (79.1%)",   "70 (83.3%)",   "71 (84.5%)",
  "",             "Black",     "8 (9.3%)",     "9 (10.7%)",    "8 (9.5%)"
)

# AE summary (wide, sorted by Placebo subject count descending)
adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)

ae_summary_rtf <- adae_raw |>
  group_by(TRT01A, AEBODSYS) |>
  summarise(n_subj = n_distinct(USUBJID), .groups = "drop") |>
  pivot_wider(names_from = TRT01A, values_from = n_subj, values_fill = 0L) |>
  arrange(desc(Placebo))
```

## Overview

### What is r2rtf?

**r2rtf** is a tidyverse-compatible R package that converts R data frames into
RTF (Rich Text Format) files suitable for clinical trial submissions. RTF is
one of the standard output formats for regulatory submissions because it can be
opened in Microsoft Word and retains precise formatting.

The SAS equivalent is `ODS RTF` combined with `PROC REPORT`:

```sas
/* SAS approach */
ods rtf file="output.rtf";
proc report data=demog nowd;
  column trt parameter statistic value;
  ...
run;
ods rtf close;
```

In r2rtf, you build the table contents as a plain R data frame first — then
attach RTF formatting via a pipe chain.

### What you will learn

By the end of this lesson you will be able to:

- Prepare a tidy summary data frame for RTF output
- Set up page orientation, margins, and font size with `rtf_page()`
- Add multi-line titles with `rtf_title()`
- Create single- and multi-row column headers with `rtf_colheader()`
- Control column widths and text justification with `rtf_body()`
- Add footnotes and data source lines
- Pre-format numbers with `sprintf()` before piping into r2rtf
- Build a complete submission-grade AE summary table

> **Note:** This tutorial uses a mock implementation of r2rtf so exercises run
> inside learnr. The API is identical to the real package — the only difference
> is that `rtf_write()` prints a summary instead of writing a real file.

---

## Section 1 — r2rtf Philosophy and the Data-First Approach

### Data first, formatting second

The key principle of r2rtf is simple: **prepare your data frame completely
before you think about formatting**. Every column in your data frame becomes
one column in the RTF table. Every row becomes one body row.

This is a different mental model from SAS `PROC REPORT`, where you define the
report structure and the data at the same time. In r2rtf, those two steps are
completely separated:

| Step | r2rtf | SAS PROC REPORT |
|------|-------|-----------------|
| 1 | Build a clean summary data frame with dplyr | Inline with PROC REPORT statements |
| 2 | Attach RTF metadata via pipe chain | Column, define, compute statements |
| 3 | `rtf_encode()` + `rtf_write()` | ODS RTF destination |

### The full pipe chain

```r
my_data |>
  rtf_page()        # page setup: orientation, margins, font size
  rtf_title()       # one or more title lines
  rtf_colheader()   # column header row(s)
  rtf_body()        # body formatting: widths, justification
  rtf_footnote()    # footnote lines
  rtf_source()      # data source statement
  rtf_encode()      # encode to RTF object
  rtf_write("output.rtf")   # write the file
```

Each function (except `rtf_encode()` and `rtf_write()`) returns the same data
frame with additional attributes attached — so the pipe just accumulates
formatting instructions without modifying the data itself.

### Demo: age summary table

Here is `age_summary`, the pre-built data frame from the setup chunk:

```{r show-age-summary}
age_summary
```

Now pipe it through the full chain:

```{r demo-full-chain}
age_summary |>
  rtf_page(orientation = "landscape", nrow = 40, font_size = 9) |>
  rtf_title(title = "Table 14.1.2 — Summary of Age by Treatment Arm") |>
  rtf_colheader(colheader = "Treatment Arm | N | Mean | SD | Median | Min | Max") |>
  rtf_body() |>
  rtf_footnote(footnote = "SD = Standard Deviation.") |>
  rtf_source(source = "Source: ADSL") |>
  rtf_encode() |>
  rtf_write("table_14_1_2.rtf")
```

Notice that `rtf_write()` reports 3 body rows — one per treatment arm — which
matches the 3 rows in `age_summary`.

### Exercise 1.1 — Full chain

Pipe `age_summary` through the complete r2rtf chain. Use the title
`"Table 14.1.1 — Age Summary"`. Verify that `rtf_write()` reports the correct
number of body rows.

```{r full-chain, exercise=TRUE, exercise.lines=12}
age_summary |>
  rtf_page(orientation = "landscape", nrow = 40, font_size = 9) |>
  rtf_title(title = ___) |>
  rtf_colheader(colheader = "Treatment | N | Mean | SD | Median | Min | Max") |>
  rtf_body() |>
  rtf_footnote(footnote = "SD = Standard Deviation.") |>
  rtf_source(source = "Source: ADSL") |>
  rtf_encode() |>
  rtf_write(___)
```

```{r full-chain-hint-1}
# Fill in the title string and the output file path.
# rtf_write() takes a character path like "age_table.rtf"
```

```{r full-chain-solution}
age_summary |>
  rtf_page(orientation = "landscape", nrow = 40, font_size = 9) |>
  rtf_title(title = "Table 14.1.1 — Age Summary") |>
  rtf_colheader(colheader = "Treatment | N | Mean | SD | Median | Min | Max") |>
  rtf_body() |>
  rtf_footnote(footnote = "SD = Standard Deviation.") |>
  rtf_source(source = "Source: ADSL") |>
  rtf_encode() |>
  rtf_write("age_table.rtf")
```

### Exercise 1.2 — Inspect rtf_encode() output

`rtf_encode()` returns an `rtf_doc` object. Store the result of encoding
`age_summary` (with any title and colheader) and then print it to see the
structure.

```{r inspect-encode, exercise=TRUE, exercise.lines=10}
doc <- age_summary |>
  rtf_page() |>
  rtf_title(title = "Age Summary") |>
  rtf_colheader(colheader = "Treatment | N | Mean | SD | Median | Min | Max") |>
  rtf_body() |>
  rtf_encode()

# Print the doc object and inspect nrow_body and ncol_body
print(___)
cat("nrow_body:", ___$nrow_body, "\n")
cat("ncol_body:", ___$ncol_body, "\n")
```

```{r inspect-encode-hint-1}
# Replace ___ with the name of your stored object: doc
```

```{r inspect-encode-solution}
doc <- age_summary |>
  rtf_page() |>
  rtf_title(title = "Age Summary") |>
  rtf_colheader(colheader = "Treatment | N | Mean | SD | Median | Min | Max") |>
  rtf_body() |>
  rtf_encode()

print(doc)
cat("nrow_body:", doc$nrow_body, "\n")
cat("ncol_body:", doc$ncol_body, "\n")
```

---

## Section 2 — rtf_page() and Page Setup

### Page setup arguments

`rtf_page()` controls the physical page properties of the RTF document. The
three most important arguments for clinical tables are:

| Argument | Default | Typical clinical use |
|----------|---------|----------------------|
| `orientation` | `"landscape"` | `"landscape"` for wide tables; `"portrait"` for narrow tables |
| `nrow` | `40` | Maximum body rows per page before a page break is inserted |
| `font_size` | `9` | 9 pt is standard for clinical submission tables |

`margin` accepts a 4-element numeric vector in inches: `c(top, bottom, left, right)`.
The regulatory standard is often 1-inch margins all around: `rep(1, 4)`.

```r
# Portrait, tight margins, small font for dense tables
my_data |>
  rtf_page(
    orientation = "portrait",
    margin      = c(1, 1, 1.25, 1.25),  # top, bottom, left, right
    nrow        = 55,
    font_size   = 8
  )
```

### rtf_page() returns the same data frame

`rtf_page()` does **not** modify the data — it attaches a `rtf_page` attribute.
You can inspect it directly:

```{r show-page-attr}
df_with_page <- age_summary |>
  rtf_page(orientation = "landscape", nrow = 40, font_size = 9)

attr(df_with_page, "rtf_page")
```

### Exercise 2.1 — Landscape setup

Set up `age_summary` for a landscape page with 40 rows per page and 9pt font.
Inspect the `rtf_page` attribute and confirm the values are correct.

```{r page-setup, exercise=TRUE, exercise.lines=8}
df <- age_summary |>
  rtf_page(
    orientation = ___,
    nrow        = ___,
    font_size   = ___
  )

attr(df, "rtf_page")
```

```{r page-setup-hint-1}
# orientation should be "landscape" (a character string)
# nrow = 40, font_size = 9
```

```{r page-setup-solution}
df <- age_summary |>
  rtf_page(
    orientation = "landscape",
    nrow        = 40,
    font_size   = 9
  )

attr(df, "rtf_page")
```

### Exercise 2.2 — Portrait for narrow tables

Some tables (e.g. a single-arm listing) fit better in portrait orientation with
a slightly larger font. Set up `demog_rtf` for portrait, 55 rows per page,
10pt font, and 1-inch margins on all sides.

```{r page-portrait, exercise=TRUE, exercise.lines=10}
df_portrait <- demog_rtf |>
  rtf_page(
    orientation = ___,
    margin      = ___,
    nrow        = ___,
    font_size   = ___
  )

attr(df_portrait, "rtf_page")
```

```{r page-portrait-hint-1}
# margin = rep(1, 4) gives 1-inch margins on all four sides
```

```{r page-portrait-solution}
df_portrait <- demog_rtf |>
  rtf_page(
    orientation = "portrait",
    margin      = rep(1, 4),
    nrow        = 55,
    font_size   = 10
  )

attr(df_portrait, "rtf_page")
```

### Quiz

```{r quiz-section2}
quiz(
  question(
    "What does `nrow` control in `rtf_page()`?",
    answer("The number of columns in the table"),
    answer("The maximum body rows per page before a page break", correct = TRUE),
    answer("The number of title rows"),
    answer("The total rows in the data frame"),
    allow_retry = TRUE
  ),
  question(
    "Which margin argument format is correct for 1-inch margins all around?",
    answer("`margin = 1`"),
    answer("`margin = c(1, 1, 1, 1)`", correct = TRUE),
    answer("`margin = list(top=1, bottom=1)`"),
    answer("`margin = '1in'`"),
    allow_retry = TRUE
  )
)
```

---

## Section 3 — Titles and Column Headers

### Multi-line titles with rtf_title()

Clinical tables almost always have at least a two-line title:

- **Line 1:** Study / protocol identifier and table number
- **Line 2:** Table description

You add multiple title lines by calling `rtf_title()` twice in the pipe chain:

```r
df |>
  rtf_title(title = "Protocol CDISCPILOT01 — Table 14.1.2") |>
  rtf_title(title = "Summary of Demographics and Baseline Characteristics")
```

Each call appends one line to the `rtf_titles` attribute.

### Column headers with rtf_colheader()

`rtf_colheader()` takes a single character string where columns are delimited
by `|`. The string must have exactly as many `|`-separated segments as there
are columns in your data frame:

```r
# 5-column data frame → 5 segments
df |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose")
```

### Multi-row headers: call rtf_colheader() twice

For spanner headers (e.g., "Treatment Arm" spanning three columns), call
`rtf_colheader()` twice. The first call defines the spanner row, the second
defines the individual column labels. Empty segments create blank cells that
serve as spacers:

```r
# Row 1: spanner
df |>
  rtf_colheader(colheader = " | | Treatment Arm") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose")
```

> The real r2rtf uses `col_rel_width` to control which cells merge for the
> spanner. In this mock the visual merging is implied by the content.

### Demo: two-line title and two-row header

```{r demo-titles-headers}
demog_rtf |>
  rtf_page(orientation = "landscape", nrow = 40, font_size = 9) |>
  rtf_title(title = "Protocol CDISCPILOT01 — Table 14.1.2") |>
  rtf_title(title = "Summary of Demographics and Baseline Characteristics") |>
  rtf_colheader(colheader = " | | Treatment Arm") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose") |>
  rtf_body() |>
  rtf_encode() |>
  rtf_write("demog_table.rtf")
```

The output confirms 2 title lines and 2 column header rows.

### Exercise 3.1 — Two-line title

Take `age_summary` and add a two-line title:
1. `"Protocol CDISCPILOT01 — Table 14.1.2"`
2. `"Age Summary by Treatment Arm"`

Then encode and write to `"age_table.rtf"`. Confirm 2 title lines are reported.

```{r two-line-title, exercise=TRUE, exercise.lines=12}
age_summary |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = ___) |>
  rtf_title(title = ___) |>
  rtf_colheader(colheader = "Treatment | N | Mean | SD | Median | Min | Max") |>
  rtf_body() |>
  rtf_encode() |>
  rtf_write("age_table.rtf")
```

```{r two-line-title-hint-1}
# Call rtf_title() twice, once for each line
```

```{r two-line-title-solution}
age_summary |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Protocol CDISCPILOT01 — Table 14.1.2") |>
  rtf_title(title = "Age Summary by Treatment Arm") |>
  rtf_colheader(colheader = "Treatment | N | Mean | SD | Median | Min | Max") |>
  rtf_body() |>
  rtf_encode() |>
  rtf_write("age_table.rtf")
```

### Exercise 3.2 — Spanner header for demog_rtf

`demog_rtf` has 5 columns: Parameter, Statistic, Placebo, XanoLow, XanoHigh.
Add a spanner header row with `"Treatment Arm"` over the three treatment columns,
then a second header row with individual column names.

```{r spanner-header, exercise=TRUE, exercise.lines=14}
demog_rtf |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Table 14.1.2 — Demographics and Baseline") |>
  # Row 1: spanner — leave Parameter and Statistic empty
  rtf_colheader(colheader = ___) |>
  # Row 2: individual column labels
  rtf_colheader(colheader = ___) |>
  rtf_body() |>
  rtf_encode() |>
  rtf_write("demog_spanner.rtf")
```

```{r spanner-header-hint-1}
# Row 1 spanner: " | | Treatment Arm"  (two empty segments for the first two cols)
# Row 2 individual: "Parameter | Statistic | Placebo | Low Dose | High Dose"
```

```{r spanner-header-solution}
demog_rtf |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Table 14.1.2 — Demographics and Baseline") |>
  rtf_colheader(colheader = " | | Treatment Arm") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose") |>
  rtf_body() |>
  rtf_encode() |>
  rtf_write("demog_spanner.rtf")
```

---

## Section 4 — rtf_body() Formatting

### Body formatting options

`rtf_body()` controls two key layout properties for the table body:

**1. `text_justification` — per-column alignment**

Pass a character vector with one value per column:

| Code | Meaning |
|------|---------|
| `"l"` | Left-align |
| `"c"` | Centre |
| `"r"` | Right-align |
| `"d"` | Decimal-align |

Clinical convention: **left-align label/description columns, right-align
numeric columns**. This makes the table scannable and matches submission
standards.

**2. `col_rel_width` — relative column widths**

Pass a numeric vector summing to 1 (or any proportional values). Wider label
columns get more space; narrow numeric columns get less:

```r
# 5-column table: label gets 30%, statistic gets 15%, 3 arms share 55%
rtf_body(
  col_rel_width      = c(0.30, 0.15, 0.183, 0.183, 0.184),
  text_justification = c("l",  "l",  "r",   "r",   "r")
)
```

### Demo: formatted demog_rtf body

```{r demo-body-format}
demog_rtf |>
  rtf_page(orientation = "landscape", nrow = 40, font_size = 9) |>
  rtf_title(title = "Table 14.1.2 — Demographics") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose") |>
  rtf_body(
    col_rel_width      = c(0.25, 0.15, 0.20, 0.20, 0.20),
    text_justification = c("l",  "l",  "r",  "r",  "r")
  ) |>
  rtf_encode() |>
  rtf_write("demog_fmt.rtf")
```

### Exercise 4.1 — Apply body formatting

Take `demog_rtf` and apply:
- Column widths: `c(0.25, 0.15, 0.20, 0.20, 0.20)`
- Justification: left for Parameter and Statistic, right for the three arm columns

```{r body-format, exercise=TRUE, exercise.lines=14}
demog_rtf |>
  rtf_page(orientation = "landscape", nrow = 40, font_size = 9) |>
  rtf_title(title = "Table 14.1.2 — Demographics and Baseline") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose") |>
  rtf_body(
    col_rel_width      = ___,
    text_justification = ___
  ) |>
  rtf_encode() |>
  rtf_write("demog_body.rtf")
```

```{r body-format-hint-1}
# col_rel_width = c(0.25, 0.15, 0.20, 0.20, 0.20)
# text_justification = c("l", "l", "r", "r", "r")
```

```{r body-format-solution}
demog_rtf |>
  rtf_page(orientation = "landscape", nrow = 40, font_size = 9) |>
  rtf_title(title = "Table 14.1.2 — Demographics and Baseline") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose") |>
  rtf_body(
    col_rel_width      = c(0.25, 0.15, 0.20, 0.20, 0.20),
    text_justification = c("l",  "l",  "r",  "r",  "r")
  ) |>
  rtf_encode() |>
  rtf_write("demog_body.rtf")
```

### Exercise 4.2 — Centre the statistic column

Repeat the formatting but centre-justify the Statistic column instead of
left-justifying it. Keep all other settings the same.

```{r body-centre, exercise=TRUE, exercise.lines=14}
demog_rtf |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Table 14.1.2 — Demographics (centred statistic)") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose") |>
  rtf_body(
    col_rel_width      = c(0.25, 0.15, 0.20, 0.20, 0.20),
    text_justification = ___
  ) |>
  rtf_encode() |>
  rtf_write("demog_centre.rtf")
```

```{r body-centre-hint-1}
# Change the second element from "l" to "c"
```

```{r body-centre-solution}
demog_rtf |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Table 14.1.2 — Demographics (centred statistic)") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose") |>
  rtf_body(
    col_rel_width      = c(0.25, 0.15, 0.20, 0.20, 0.20),
    text_justification = c("l", "c", "r", "r", "r")
  ) |>
  rtf_encode() |>
  rtf_write("demog_centre.rtf")
```

---

## Section 5 — Footnotes and Source Lines

### Regulatory footnote requirements

Submission tables must include:

- **Footnotes** explaining abbreviations, statistical methods, and any
  special handling (e.g., "Missing values excluded from percentage denominator")
- **Source line** identifying the dataset(s) and data cut date

Both are added at the end of the pipe chain, before `rtf_encode()`.

### rtf_footnote() and rtf_source()

```r
df |>
  rtf_footnote(footnote = "SD = Standard Deviation; N = number of subjects.") |>
  rtf_footnote(footnote = "Missing values were excluded from all calculations.") |>
  rtf_source(source   = "Source: ADSL; Data cut: 2024-01-01")
```

Like `rtf_title()`, calling `rtf_footnote()` multiple times appends lines —
so you can have as many footnotes as needed.

### Demo

```{r demo-footnotes}
age_summary |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Table 14.1.2 — Age Summary") |>
  rtf_colheader(colheader = "Treatment | N | Mean | SD | Median | Min | Max") |>
  rtf_body(text_justification = c("l", "r", "r", "r", "r", "r", "r")) |>
  rtf_footnote(footnote = "SD = Standard Deviation.") |>
  rtf_footnote(footnote = "N = number of subjects randomised to the treatment arm.") |>
  rtf_source(source = "Source: ADSL; Data cut: 2024-01-01") |>
  rtf_encode() |>
  rtf_write("age_fn.rtf")
```

### Exercise 5.1 — Add footnote and source

Build an `age_summary` table with:
- Footnote: `"SD = Standard Deviation."`
- Source: `"Source: ADSL; Data cut: 2024-01-01"`

```{r add-footnote, exercise=TRUE, exercise.lines=13}
age_summary |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Table 14.1.2 — Age Summary") |>
  rtf_colheader(colheader = "Treatment | N | Mean | SD | Median | Min | Max") |>
  rtf_body() |>
  rtf_footnote(footnote = ___) |>
  rtf_source(source = ___) |>
  rtf_encode() |>
  rtf_write("age_fn.rtf")
```

```{r add-footnote-hint-1}
# Both arguments take a plain character string
```

```{r add-footnote-solution}
age_summary |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Table 14.1.2 — Age Summary") |>
  rtf_colheader(colheader = "Treatment | N | Mean | SD | Median | Min | Max") |>
  rtf_body() |>
  rtf_footnote(footnote = "SD = Standard Deviation.") |>
  rtf_source(source = "Source: ADSL; Data cut: 2024-01-01") |>
  rtf_encode() |>
  rtf_write("age_fn.rtf")
```

### Exercise 5.2 — Multiple footnotes

Add two footnotes to `demog_rtf`:
1. `"n = number of subjects; % = percentage within treatment arm."`
2. `"Missing values excluded from percentage denominators."`

```{r multi-footnote, exercise=TRUE, exercise.lines=14}
demog_rtf |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Table 14.1.2 — Demographics") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose") |>
  rtf_body(
    col_rel_width      = c(0.25, 0.15, 0.20, 0.20, 0.20),
    text_justification = c("l",  "l",  "r",  "r",  "r")
  ) |>
  rtf_footnote(footnote = ___) |>
  rtf_footnote(footnote = ___) |>
  rtf_source(source = "Source: ADSL; Data cut: 2024-01-01") |>
  rtf_encode() |>
  rtf_write("demog_fn.rtf")
```

```{r multi-footnote-hint-1}
# Call rtf_footnote() twice, one per footnote line
```

```{r multi-footnote-solution}
demog_rtf |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Table 14.1.2 — Demographics") |>
  rtf_colheader(colheader = "Parameter | Statistic | Placebo | Low Dose | High Dose") |>
  rtf_body(
    col_rel_width      = c(0.25, 0.15, 0.20, 0.20, 0.20),
    text_justification = c("l",  "l",  "r",  "r",  "r")
  ) |>
  rtf_footnote(footnote = "n = number of subjects; % = percentage within treatment arm.") |>
  rtf_footnote(footnote = "Missing values excluded from percentage denominators.") |>
  rtf_source(source = "Source: ADSL; Data cut: 2024-01-01") |>
  rtf_encode() |>
  rtf_write("demog_fn.rtf")
```

---

## Section 6 — Formatting Numbers with sprintf

### Why you pre-format numbers

r2rtf works with **character columns**. When you pipe a data frame, every
column becomes a text cell. If you want "75.2 (8.6)" in a cell, you must
create that string *before* the `rtf_` calls — r2rtf does not know how to
combine two numeric columns into one formatted string.

This is the standard approach:

```r
# Mean (SD) as one formatted string
df |> mutate(mean_sd = sprintf("%.1f (%.1f)", Mean, SD))

# Count (%) — with %% to escape the percent sign in sprintf
format_count_pct <- function(n, N) sprintf("%d (%.1f%%)", n, round(100 * n / N, 1))
```

### Common sprintf patterns for clinical tables

| Display | sprintf pattern | Example |
|---------|----------------|---------|
| Mean (SD) | `"%.1f (%.1f)"` | `75.2 (8.6)` |
| n (%) | `"%d (%.1f%%)"` | `42 (50.0%)` |
| Median [Min, Max] | `"%.1f [%.0f, %.0f]"` | `76.0 [52, 89]` |
| n / N | `"%d/%d"` | `42/84` |

### Demo: format age_summary for RTF

```{r demo-sprintf}
age_rtf_fmt <- age_summary |>
  mutate(
    mean_sd    = sprintf("%.1f (%.1f)", Mean, SD),
    median_str = sprintf("%.1f", Median),
    range_str  = sprintf("[%.0f, %.0f]", Min, Max)
  ) |>
  select(TRT01P, N, mean_sd, median_str, range_str)

age_rtf_fmt
```

Now this five-column frame is ready to pipe into r2rtf.

### Exercise 6.1 — Create formatted strings

Take `age_summary` and create two new character columns:
- `mean_sd`: formatted as `"75.2 (8.6)"`
- `range_str`: formatted as `"[52, 89]"` (integer display for min and max)

Then select `TRT01P`, `N`, `mean_sd`, and `range_str` for the RTF pipe.

```{r sprintf-fmt, exercise=TRUE, exercise.lines=12}
age_rtf <- age_summary |>
  mutate(
    mean_sd   = sprintf(___, Mean, SD),
    range_str = sprintf(___, Min, Max)
  ) |>
  select(TRT01P, N, mean_sd, range_str)

age_rtf
```

```{r sprintf-fmt-hint-1}
# mean_sd pattern:   "%.1f (%.1f)"
# range_str pattern: "[%.0f, %.0f]"   (%.0f rounds to integer)
```

```{r sprintf-fmt-solution}
age_rtf <- age_summary |>
  mutate(
    mean_sd   = sprintf("%.1f (%.1f)", Mean, SD),
    range_str = sprintf("[%.0f, %.0f]", Min, Max)
  ) |>
  select(TRT01P, N, mean_sd, range_str)

age_rtf
```

### Exercise 6.2 — Pipe formatted frame into r2rtf

Using the `age_rtf` data frame you just created, build a complete RTF table
with a two-line title, a column header, and a footnote explaining "SD = Standard
Deviation."

```{r sprintf-rtf, exercise=TRUE, exercise.lines=16}
age_rtf <- age_summary |>
  mutate(
    mean_sd   = sprintf("%.1f (%.1f)", Mean, SD),
    range_str = sprintf("[%.0f, %.0f]", Min, Max)
  ) |>
  select(TRT01P, N, mean_sd, range_str)

age_rtf |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = ___) |>
  rtf_title(title = ___) |>
  rtf_colheader(colheader = "Treatment | N | Mean (SD) | Range") |>
  rtf_body(
    col_rel_width      = c(0.35, 0.10, 0.30, 0.25),
    text_justification = c("l",  "r",  "r",  "r")
  ) |>
  rtf_footnote(footnote = ___) |>
  rtf_encode() |>
  rtf_write("age_formatted.rtf")
```

```{r sprintf-rtf-hint-1}
# Two-line title: protocol identifier + description
# Footnote: "SD = Standard Deviation."
```

```{r sprintf-rtf-solution}
age_rtf <- age_summary |>
  mutate(
    mean_sd   = sprintf("%.1f (%.1f)", Mean, SD),
    range_str = sprintf("[%.0f, %.0f]", Min, Max)
  ) |>
  select(TRT01P, N, mean_sd, range_str)

age_rtf |>
  rtf_page(orientation = "landscape") |>
  rtf_title(title = "Protocol CDISCPILOT01 — Table 14.1.2") |>
  rtf_title(title = "Age Summary by Treatment Arm") |>
  rtf_colheader(colheader = "Treatment | N | Mean (SD) | Range") |>
  rtf_body(
    col_rel_width      = c(0.35, 0.10, 0.30, 0.25),
    text_justification = c("l",  "r",  "r",  "r")
  ) |>
  rtf_footnote(footnote = "SD = Standard Deviation.") |>
  rtf_encode() |>
  rtf_write("age_formatted.rtf")
```

---

## Section 7 — Capstone: Table 14.3.1 AE Summary

### The pre-built AE summary frame

The setup chunk built `ae_summary_rtf` from real `pharmaverseadam::adae` data.
It is already in wide format, with one row per body system and one column per
treatment arm, sorted by descending Placebo subject count:

```{r show-ae-summary}
ae_summary_rtf
```

This frame is ready to pipe directly into r2rtf — no further dplyr needed.

### Your task

Build a complete submission-grade RTF table from `ae_summary_rtf`:

1. Landscape page, 40 rows per page, 9pt font
2. Two-line title:
   - Line 1: `"Protocol CDISCPILOT01 — Table 14.3.1"`
   - Line 2: `"Number of Subjects with at Least One AE by Body System"`
3. Two-row column header:
   - Row 1 spanner: `" | Treatment Arm"`
   - Row 2 individual labels (use actual column names)
4. Body: left-justify body system column, right-justify the three count columns
5. Column widths: `c(0.55, 0.15, 0.15, 0.15)`
6. Footnote: `"n = number of subjects with at least one AE in the body system."`
7. Source: `"Source: ADAE; Data cut: 2024-01-01"`

After building, answer: how many columns does the encoded rtf document have?

```{r ae-capstone, exercise=TRUE, exercise.lines=24}
# Step 1: Check the column names of ae_summary_rtf
names(ae_summary_rtf)

# Step 2: Build the RTF table
doc <- ae_summary_rtf |>
  rtf_page(orientation = ___, nrow = ___, font_size = ___) |>
  rtf_title(title = ___) |>
  rtf_title(title = ___) |>
  rtf_colheader(colheader = ___) |>
  rtf_colheader(colheader = ___) |>
  rtf_body(
    col_rel_width      = ___,
    text_justification = ___
  ) |>
  rtf_footnote(footnote = ___) |>
  rtf_source(source = ___) |>
  rtf_encode()

rtf_write(doc, "table_14_3_1.rtf")
cat("\nNumber of columns in encoded document:", doc$ncol_body, "\n")
```

```{r ae-capstone-hint-1}
# names(ae_summary_rtf) will show: AEBODSYS + three treatment arm columns
# The table has 4 columns total.
# For the spanner: " | Treatment Arm"
# For individual headers: paste the column names separated by " | "
```

```{r ae-capstone-solution}
names(ae_summary_rtf)

doc <- ae_summary_rtf |>
  rtf_page(orientation = "landscape", nrow = 40, font_size = 9) |>
  rtf_title(title = "Protocol CDISCPILOT01 — Table 14.3.1") |>
  rtf_title(title = "Number of Subjects with at Least One AE by Body System") |>
  rtf_colheader(colheader = " | Treatment Arm") |>
  rtf_colheader(
    colheader = paste(names(ae_summary_rtf), collapse = " | ")
  ) |>
  rtf_body(
    col_rel_width      = c(0.55, 0.15, 0.15, 0.15),
    text_justification = c("l",  "r",  "r",  "r")
  ) |>
  rtf_footnote(
    footnote = "n = number of subjects with at least one AE in the body system."
  ) |>
  rtf_source(source = "Source: ADAE; Data cut: 2024-01-01") |>
  rtf_encode()

rtf_write(doc, "table_14_3_1.rtf")
cat("\nNumber of columns in encoded document:", doc$ncol_body, "\n")
```

> **Expected answer:** 4 columns (`AEBODSYS` + 3 treatment arm count columns).

---

## Lesson Summary

### Key r2rtf concepts

| Concept | What you learned |
|---------|-----------------|
| Data-first | Build a clean tidy data frame before any `rtf_` calls |
| `rtf_page()` | Controls orientation, margins, rows-per-page, font size |
| `rtf_title()` | Call twice for two-line titles; each call appends one line |
| `rtf_colheader()` | `\|`-delimited string; call twice for spanner + individual labels |
| `rtf_body()` | `col_rel_width` for proportional widths; `text_justification` per column |
| `rtf_footnote()` | Call multiple times for multiple footnote lines |
| `rtf_source()` | Single data provenance statement |
| `rtf_encode()` | Returns an `rtf_doc` object with metadata |
| `rtf_write()` | Final step — writes (or in this mock, summarises) the RTF file |
| `sprintf()` | Pre-format numeric columns into submission-display strings before piping |

### SAS equivalence

| r2rtf | SAS |
|-------|-----|
| `rtf_page(orientation="landscape")` | `ODS RTF; OPTIONS ORIENTATION=LANDSCAPE;` |
| `rtf_title()` | `TITLE1 "..."; TITLE2 "...";` |
| `rtf_colheader()` | `DEFINE col / "Header";` in PROC REPORT |
| `rtf_footnote()` | `FOOTNOTE1 "...";` |
| `rtf_body(text_justification)` | `DEFINE col / LEFT/CENTER/RIGHT;` |
| `sprintf("%.1f (%.1f)", Mean, SD)` | `CATX(" ", PUT(Mean,5.1), "("||PUT(SD,5.1)||")")` |

### Next lesson

**Lesson 36 — Tplyr** covers a layered approach to building the *data frames*
that feed into r2rtf (and other output packages): count tables, descriptive
statistics, and shift tables with a clean `group_count() / group_desc() /
group_shift()` API.
