---
title: "Lesson 36 — Tplyr: Layered Clinical Summary Table Pipeline"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build clinical summary tables with Tplyr's layered API: group_count() for
  categorical variables, group_desc() for continuous variables, group_shift()
  for shift tables, f_str() format strings, ordering, and a demographic + AE
  capstone table.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 36_Tplyr.Rmd
# PURPOSE       : Teach Tplyr's layered table-building API: tplyr_table(),
#                 group_count(), group_desc(), group_shift(), f_str(),
#                 set_format_strings(), ordering, set_pop_data(), and build().
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.blanks = TRUE)

# ── Tplyr mock — complete implementation ─────────────────────────────────────

# Internal layer constructors
new_count_layer <- function(target, by = NULL) {
  structure(list(type = "count", target = target, by = by,
                 formats = NULL, order = "byvar", distinct_by = "USUBJID",
                 missing_count = FALSE, pop_data = NULL),
            class = "tplyr_layer")
}

new_desc_layer <- function(target, by = NULL) {
  structure(list(type = "desc", target = target, by = by,
                 formats = NULL),
            class = "tplyr_layer")
}

new_shift_layer <- function(target, by = NULL) {
  structure(list(type = "shift", target = target, by = by),
            class = "tplyr_layer")
}

# f_str: format string helper
f_str <- function(fmt, ...) {
  vars <- sapply(substitute(list(...))[-1], deparse)
  structure(list(fmt = fmt, vars = vars), class = "f_str")
}

# tplyr_table constructor
tplyr_table <- function(dataset, treat_var, where = TRUE, cols = NULL) {
  tv <- deparse(substitute(treat_var))
  structure(
    list(
      dataset   = dataset,
      treat_var = tv,
      layers    = list(),
      cols      = cols,
      pop_data  = NULL,
      pop_treat_var = tv
    ),
    class = "tplyr_table"
  )
}

# Fluent layer adder
add_layer <- function(table, layer) {
  table$layers <- c(table$layers, list(layer))
  table
}

# Layer factory functions
group_count <- function(target, by = NULL, where = TRUE) {
  tgt <- deparse(substitute(target))
  by_str <- if (!missing(by)) deparse(substitute(by)) else NULL
  new_count_layer(target = tgt, by = by_str)
}

group_desc <- function(target, by = NULL, where = TRUE) {
  tgt <- deparse(substitute(target))
  by_str <- if (!missing(by)) deparse(substitute(by)) else NULL
  new_desc_layer(target = tgt, by = by_str)
}

group_shift <- function(target, by = NULL, where = TRUE) {
  tgt <- deparse(substitute(target))
  by_str <- if (!missing(by)) deparse(substitute(by)) else NULL
  new_shift_layer(target = tgt, by = by_str)
}

# Layer customisation helpers
set_format_strings <- function(layer, ...) {
  layer$formats <- list(...)
  layer
}

set_order_count_method <- function(layer, method = "byvarn") {
  layer$order <- method
  layer
}

set_result_order_var <- function(layer, var) {
  layer$result_order_var <- deparse(substitute(var))
  layer
}

set_distinct_by <- function(layer, ...) {
  vars <- sapply(substitute(list(...))[-1], deparse)
  layer$distinct_by <- vars
  layer
}

set_missing_count <- function(layer, ...) {
  layer$missing_count <- TRUE
  layer
}

set_pop_data <- function(table, pop_data, treat_var = NULL) {
  table$pop_data <- pop_data
  if (!is.null(treat_var))
    table$pop_treat_var <- deparse(substitute(treat_var))
  table
}

set_pop_treat_var <- function(table, treat_var) {
  table$pop_treat_var <- deparse(substitute(treat_var))
  table
}

# build(): execute all layers and bind results
build <- function(table, ...) {
  dataset   <- table$dataset
  treat_var <- table$treat_var

  # Denominator population for % calculations
  pop_data <- if (!is.null(table$pop_data)) table$pop_data else dataset
  pop_tv   <- table$pop_treat_var %||% treat_var

  treat_lvls <- sort(unique(dataset[[treat_var]]))
  # Subject counts per arm for denominator
  arm_N <- pop_data |>
    group_by(.data[[pop_tv]]) |>
    summarise(N = n_distinct(USUBJID), .groups = "drop") |>
    rename(treat_arm = !!pop_tv)

  layer_idx <- 0L

  all_rows <- lapply(table$layers, function(layer) {
    layer_idx <<- layer_idx + 1L
    var <- gsub('^"|"$|^\'|\'$', "", layer$target)
    if (var == "NULL" || !(var %in% names(dataset))) return(NULL)

    if (layer$type == "count") {
      # Subject-level counts per treatment × category
      raw <- dataset |>
        group_by(.data[[treat_var]], .data[[var]]) |>
        summarise(n = n_distinct(USUBJID), .groups = "drop")

      # Attach denominator
      raw <- raw |>
        rename(treat_arm = !!treat_var) |>
        left_join(arm_N, by = "treat_arm") |>
        mutate(pct = round(100 * n / N, 1))

      # Ordering
      if (!is.null(layer$order) && layer$order == "bycount") {
        cat_order <- raw |>
          group_by(.data[[var]]) |>
          summarise(total_n = sum(n), .groups = "drop") |>
          arrange(desc(total_n)) |>
          mutate(ord = seq_len(n()))
        raw <- left_join(raw, cat_order, by = var)
      } else {
        cat_order <- raw |>
          distinct(.data[[var]]) |>
          mutate(ord = seq_len(n()))
        raw <- left_join(raw, cat_order, by = var)
      }

      out <- raw |>
        pivot_wider(
          id_cols     = c(all_of(var), "ord"),
          names_from  = treat_arm,
          values_from = c(n, pct),
          values_fill = 0
        ) |>
        rename(row_label_1 = !!var) |>
        arrange(ord) |>
        mutate(
          ord_layer_index = layer_idx,
          ord_layer_1     = row_label_1
        ) |>
        select(-ord)

    } else if (layer$type == "desc") {
      stat_names <- c("n", "mean", "sd", "median", "min", "max")

      out <- dataset |>
        group_by(.data[[treat_var]]) |>
        summarise(
          n      = n(),
          mean   = round(mean(.data[[var]], na.rm = TRUE), 1),
          sd     = round(sd(.data[[var]],   na.rm = TRUE), 1),
          median = round(median(.data[[var]], na.rm = TRUE), 1),
          min    = min(.data[[var]], na.rm = TRUE),
          max    = max(.data[[var]], na.rm = TRUE),
          .groups = "drop"
        ) |>
        pivot_longer(cols = -!!treat_var,
                     names_to  = "stat_name",
                     values_to = "value") |>
        pivot_wider(names_from  = !!treat_var,
                    values_from = value) |>
        rename(row_label_1 = stat_name) |>
        mutate(
          row_label_1     = factor(row_label_1, levels = stat_names),
          ord_layer_index = layer_idx,
          ord_layer_1     = as.character(row_label_1)
        ) |>
        arrange(row_label_1) |>
        mutate(row_label_1 = as.character(row_label_1))

    } else {
      # shift — return placeholder
      out <- tibble(
        row_label_1     = character(),
        ord_layer_index = integer(),
        ord_layer_1     = character()
      )
    }

    out
  })

  bind_rows(all_rows)
}

# Null-coalesce
`%||%` <- function(a, b) if (!is.null(a)) a else b

# ── Real pharmaverseadam data ─────────────────────────────────────────────────

adsl <- pharmaverseadam::adsl

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw

# Synthetic shift dataset for Section 7
set.seed(2014)
shift_demo <- tibble(
  USUBJID  = rep(paste0("SUBJ-", sprintf("%03d", 1:60)), each = 2),
  TRT01P   = rep(c(rep("Placebo", 40), rep("Xanomeline High Dose", 80)), length.out = 120),
  VISIT    = rep(c("BASELINE", "WEEK 12"), 60),
  ATOXGR_C = sample(c("0","1","2","3"), 120, replace = TRUE, prob = c(0.5,0.3,0.15,0.05))
) |>
  mutate(ATOXGR = as.integer(ATOXGR_C)) |>
  select(-ATOXGR_C)

shift_wide <- shift_demo |>
  pivot_wider(names_from = VISIT, values_from = ATOXGR,
              names_prefix = "GR_") |>
  rename(BASELINE_GR = `GR_BASELINE`, WEEK12_GR = `GR_WEEK 12`)
```

## Overview

### What is Tplyr?

**Tplyr** (from Atorus Research) is a tidyverse-based package for building
clinical summary tables using a **layered approach**. It handles the most
common clinical table patterns — count/%, mean/SD, and shift tables — with
minimal code, while keeping the output as a plain R tibble that you can pass
to any downstream tool (r2rtf, gt, flextable, etc.).

The SAS equivalents are `PROC FREQ`, `PROC MEANS`, and custom shift-table
macros. Tplyr replaces all three with a consistent pipe-friendly API.

### Core Tplyr workflow

```r
tplyr_table(dataset, treat_var) |>   # 1. Define the base table
  add_layer(group_count(SEX)) |>     # 2. Add a count layer (Sex)
  add_layer(group_desc(AGE))  |>     # 3. Add a descriptive stats layer (Age)
  build()                            # 4. Execute and return a tibble
```

### What you will learn

- The layered mental model: each `add_layer()` = one block of rows
- `group_count()` for categorical variables (demographics, AEs)
- `group_desc()` for continuous variables (age, lab values)
- `f_str()` format strings to control how statistics are displayed
- `set_pop_data()` for correct AE percentage denominators
- `set_order_count_method()` to sort rows
- `group_shift()` for baseline vs. post-baseline category shift tables
- Post-build: the result is a plain tibble you can `filter()` and `arrange()`

> **Note:** This tutorial uses a mock implementation of Tplyr so exercises run
> inside learnr. The API mirrors the real package.

---

## Section 1 — Tplyr Philosophy

### The layered mental model

Tplyr's core idea is that a clinical summary table is made up of **blocks of
rows**. Each block covers one variable:

```
┌────────────────────────────────────────────────────────────┐
│  Age (years)   n = 86      n = 84      n = 84              │  ← group_desc(AGE)
│                Mean 75.2   Mean 75.7   Mean 74.4            │
│                SD   8.6    SD   8.3    SD   7.9             │
├────────────────────────────────────────────────────────────┤
│  Sex     Male  37 (43%)   35 (42%)   35 (42%)              │  ← group_count(SEX)
│         Female 49 (57%)   49 (58%)   49 (58%)              │
└────────────────────────────────────────────────────────────┘
```

You add layers in the order you want them to appear in the final table. Tplyr
stacks them vertically when you call `build()`.

### Tplyr vs gtsummary

| Feature | Tplyr | gtsummary |
|---------|-------|-----------|
| Mental model | Explicit layers | Infers from variable type |
| Output | Plain tibble | Formatted gt object |
| AE tables | Native with `set_pop_data()` | Requires extra steps |
| Customisation | Fine-grained via `set_*()` | Via `modify_*()` helpers |
| Downstream use | Any package (r2rtf, gt, flextable) | Primarily gt / Word |

### Demo: minimal two-layer table

```{r demo-two-layer}
tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  add_layer(group_count(SEX)) |>
  build()
```

### Exercise 1.1 — Count rows returned by build()

Run the demo table above and capture the result. How many total rows does
`build()` return? (6 rows from `group_desc` — n, mean, sd, median, min, max —
plus 2 rows from `group_count` for SEX categories.)

```{r layer-row-count, exercise=TRUE, exercise.lines=10}
result <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  add_layer(group_count(SEX)) |>
  build()

# How many rows?
nrow(___)
```

```{r layer-row-count-hint-1}
# Replace ___ with the name of your result object
```

```{r layer-row-count-solution}
result <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  add_layer(group_count(SEX)) |>
  build()

nrow(result)
```

### Exercise 1.2 — Inspect column names

Build a simple `group_count(SEX)` table and check its column names. Identify
the treatment-arm count columns and the ordering columns.

```{r inspect-cols, exercise=TRUE, exercise.lines=8}
tbl <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_count(SEX)) |>
  build()

names(tbl)
```

```{r inspect-cols-solution}
tbl <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_count(SEX)) |>
  build()

names(tbl)
```

> Tplyr output always includes `row_label_1` (the category label),
> `ord_layer_index` (which layer the row came from), and `ord_layer_1`
> (sort key within the layer), plus n\_ and pct\_ columns for each arm.

---

## Section 2 — group_count() for Categorical Variables

### Basic count layer

`group_count(variable)` produces subject counts and percentages for each level
of a categorical variable:

```r
tplyr_table(adsl, TRT01P) |>
  add_layer(group_count(SEX)) |>
  build()
```

Each category of `SEX` becomes one row. The output has `n_<arm>` and
`pct_<arm>` columns for each treatment arm.

### set_distinct_by() — subject-level counts

By default the mock counts distinct subjects. In the real Tplyr package, you
must explicitly request subject-level counting with `set_distinct_by(USUBJID)`.
This is **critical for AE tables** — without it you count AE records, not subjects:

```r
# Wrong for AEs: counts records
group_count(AEBODSYS)

# Correct for AEs: counts distinct subjects with that AE
group_count(AEBODSYS) |>
  set_distinct_by(USUBJID)
```

### set_format_strings() with f_str()

`f_str()` controls how multiple statistics are combined into one display cell.
For count layers:

```r
group_count(SEX) |>
  set_format_strings(f_str("xx (xx.x%)", n, pct))
# Result: "42 (50.0%)"
```

> In this mock, `set_format_strings()` stores the format for reference — the
> `build()` output still returns separate `n_` and `pct_` columns. The real
> Tplyr applies the format string and merges them.

### Exercise 2.1 — Count RACE

Build a count table for `RACE` using `adsl` with `TRT01P` as the treatment
variable. Print the result.

```{r count-race, exercise=TRUE, exercise.lines=8}
tplyr_table(adsl, TRT01P) |>
  add_layer(group_count(___)) |>
  build()
```

```{r count-race-hint-1}
# Replace ___ with the variable name: RACE
```

```{r count-race-solution}
tplyr_table(adsl, TRT01P) |>
  add_layer(group_count(RACE)) |>
  build()
```

### Exercise 2.2 — Subject count for WHITE in Placebo

Using the RACE count table, filter to the "WHITE" row and inspect the Placebo
subject count.

```{r count-white, exercise=TRUE, exercise.lines=10}
race_tbl <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_count(RACE)) |>
  build()

# Filter to WHITE and check n_Placebo
race_tbl |>
  filter(row_label_1 == ___) |>
  select(row_label_1, starts_with("n_"))
```

```{r count-white-hint-1}
# The category value is "WHITE" (all caps in pharmaverseadam)
```

```{r count-white-solution}
race_tbl <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_count(RACE)) |>
  build()

race_tbl |>
  filter(row_label_1 == "WHITE") |>
  select(row_label_1, starts_with("n_"))
```

### Exercise 2.3 — Apply set_distinct_by and set_format_strings

Build a SEX count table with `set_distinct_by(USUBJID)` and a format string
`f_str("xx (xx.x%)", n, pct)`. Print the result.

```{r count-sex-fmt, exercise=TRUE, exercise.lines=10}
tplyr_table(adsl, TRT01P) |>
  add_layer(
    group_count(SEX) |>
      set_distinct_by(___) |>
      set_format_strings(f_str(___, n, pct))
  ) |>
  build()
```

```{r count-sex-fmt-hint-1}
# set_distinct_by(USUBJID)
# f_str format for count/percent: "xx (xx.x%)"
```

```{r count-sex-fmt-solution}
tplyr_table(adsl, TRT01P) |>
  add_layer(
    group_count(SEX) |>
      set_distinct_by(USUBJID) |>
      set_format_strings(f_str("xx (xx.x%)", n, pct))
  ) |>
  build()
```

---

## Section 3 — group_desc() for Continuous Variables

### What group_desc() produces

`group_desc(variable)` produces one row per summary statistic per treatment
arm — specifically: n, mean, sd, median, min, max. Each statistic becomes a
row with `row_label_1` equal to the statistic name:

```
row_label_1  Placebo  Xanomeline Low Dose  Xanomeline High Dose
n             86        84                   84
mean          75.2      75.7                 74.4
sd             8.6       8.3                  7.9
median        76.0      77.5                 76.0
min           52        56                   56
max           89        88                   88
```

### set_format_strings() for desc layers

```r
group_desc(AGE) |>
  set_format_strings(
    f_str("xx.x (xx.x)", mean, sd)   # "75.2 (8.6)"
  )
```

You can provide multiple format strings to control different rows.

### Demo

```{r demo-desc}
tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  build()
```

### Exercise 3.1 — Descriptive stats for AGE

Build a `group_desc(AGE)` table and identify the mean AGE for the
`"Xanomeline High Dose"` arm.

```{r desc-age, exercise=TRUE, exercise.lines=10}
age_desc <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(___)) |>
  build()

# Filter to the 'mean' row
age_desc |>
  filter(row_label_1 == ___) |>
  select(row_label_1, `Xanomeline High Dose`)
```

```{r desc-age-hint-1}
# group_desc takes the variable name: AGE
# Filter row_label_1 == "mean"
```

```{r desc-age-solution}
age_desc <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  build()

age_desc |>
  filter(row_label_1 == "mean") |>
  select(row_label_1, `Xanomeline High Dose`)
```

### Exercise 3.2 — Add a format string

Rebuild the AGE descriptive table with a format string that shows mean and SD
as `"xx.x (xx.x)"`. Print the resulting layer (note the formats are stored in
the layer object before `build()` is called).

```{r desc-fmt, exercise=TRUE, exercise.lines=10}
tplyr_table(adsl, TRT01P) |>
  add_layer(
    group_desc(AGE) |>
      set_format_strings(f_str(___, mean, sd))
  ) |>
  build()
```

```{r desc-fmt-hint-1}
# The format string for "75.2 (8.6)" is "xx.x (xx.x)"
```

```{r desc-fmt-solution}
tplyr_table(adsl, TRT01P) |>
  add_layer(
    group_desc(AGE) |>
      set_format_strings(f_str("xx.x (xx.x)", mean, sd))
  ) |>
  build()
```

---

## Section 4 — Format Strings with f_str()

### What f_str() does

`f_str()` is Tplyr's compact way to specify how multiple statistics combine
into a single display cell. The first argument is a display format template;
subsequent arguments name which statistics to slot in.

```r
f_str("xx (xx.x%)", n, pct)     # "42 (50.0%)"
f_str("xx.x (xx.x)", mean, sd)  # "75.2 (8.6)"
f_str("xx.x, xx.x", min, max)   # "52.0, 89.0"
```

The `xx` placeholder means integer; `xx.x` means one decimal place; `xx.xx`
means two decimal places.

### Common clinical f_str patterns

| Display | f_str call | Example output |
|---------|------------|----------------|
| Count (%) | `f_str("xx (xx.x%)", n, pct)` | `42 (50.0%)` |
| Mean (SD) | `f_str("xx.x (xx.x)", mean, sd)` | `75.2 (8.6)` |
| Median [Min, Max] | `f_str("xx.x [xx.x, xx.x]", median, min, max)` | `76.0 [52.0, 89.0]` |
| Range only | `f_str("xx.x, xx.x", min, max)` | `52.0, 89.0` |
| n only | `f_str("xx", n)` | `42` |

### Exercise 4.1 — f_str for Mean (SD)

Write an `f_str()` that produces `"75.2 (8.6)"` — i.e., mean to one decimal
place, followed by SD in parentheses.

```{r fstr-mean-sd, exercise=TRUE, exercise.lines=6}
# Write the f_str and store it — inspect its structure
fs <- f_str(___, mean, sd)
str(fs)
```

```{r fstr-mean-sd-hint-1}
# Template for "75.2 (8.6)": "xx.x (xx.x)"
```

```{r fstr-mean-sd-solution}
fs <- f_str("xx.x (xx.x)", mean, sd)
str(fs)
```

### Exercise 4.2 — f_str for Median [Q1, Q3]

Write an `f_str()` that would display median with [Q1, Q3] in the format
`"75.0 [70.0, 81.0]"`. Use `median`, `min`, and `max` as placeholders for Q1
and Q3 (we use min/max here as stand-ins).

```{r fstr-median-iqr, exercise=TRUE, exercise.lines=6}
fs_median <- f_str(___, median, min, max)
str(fs_median)
```

```{r fstr-median-iqr-hint-1}
# Pattern: "xx.x [xx.x, xx.x]"  — three placeholders in order: median, min, max
```

```{r fstr-median-iqr-solution}
fs_median <- f_str("xx.x [xx.x, xx.x]", median, min, max)
str(fs_median)
```

### Exercise 4.3 — f_str for range only

Write an `f_str()` that produces `"52, 89"` (integer display for min and max,
comma-separated, no brackets).

```{r fstr-range, exercise=TRUE, exercise.lines=6}
fs_range <- f_str(___, min, max)
str(fs_range)
```

```{r fstr-range-hint-1}
# Integer display: xx  (no decimal); pattern: "xx, xx"
```

```{r fstr-range-solution}
fs_range <- f_str("xx, xx", min, max)
str(fs_range)
```

### Quiz

```{r quiz-fstr}
quiz(
  question(
    "Which f_str template produces '42 (50.0%)'?",
    answer("`f_str(\"xx (xx.x)\", n, pct)`"),
    answer("`f_str(\"xx (xx.x%)\", n, pct)`", correct = TRUE),
    answer("`f_str(\"xx.x (xx.x)\", n, pct)`"),
    answer("`f_str(\"xx%\", pct)`"),
    allow_retry = TRUE
  )
)
```

---

## Section 5 — AE Tables with group_count() and set_pop_data()

### The AE denominator problem

For an AE table, the percentage denominator should be the **number of subjects
randomised to each arm** (from ADSL), not the number of AE records in ADAE.

Without `set_pop_data()`, Tplyr would use the ADAE dataset to compute
denominators — which undercounts subjects who had no AEs.

```r
# Wrong: denominator = subjects who had at least one AE
tplyr_table(adae, TRT01A) |>
  add_layer(group_count(AEBODSYS)) |>
  build()

# Correct: denominator = all randomised subjects (from ADSL)
tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID)
  ) |>
  build()
```

### set_pop_data() arguments

- **First argument:** the population dataset (e.g., `adsl`)
- **Second argument:** the treatment variable in *that* population dataset
  (e.g., `TRT01P` — note: ADSL uses `TRT01P`, ADAE uses `TRT01A`)

### set_distinct_by() for AE subject counts

For AE tables, you always want to count distinct subjects, not AE records. A
subject who had 3 SKIN events should count as 1, not 3:

```r
group_count(AEBODSYS) |>
  set_distinct_by(USUBJID)
```

### Exercise 5.1 — AE body system count table

Build an AE body-system count table using `adae` as the primary dataset and
`adsl` as the population denominator. Use `set_distinct_by(USUBJID)`.

```{r ae-count, exercise=TRUE, exercise.lines=12}
ae_tbl <- tplyr_table(adae, TRT01A) |>
  set_pop_data(___, ___) |>
  add_layer(
    group_count(___) |>
      set_distinct_by(___)
  ) |>
  build()

ae_tbl
```

```{r ae-count-hint-1}
# set_pop_data(adsl, TRT01P)  — population dataset + treatment var in that dataset
# group_count(AEBODSYS)
# set_distinct_by(USUBJID)
```

```{r ae-count-solution}
ae_tbl <- tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID)
  ) |>
  build()

ae_tbl
```

### Exercise 5.2 — Highest body system in Placebo

From the AE table you just built, find the body system with the highest
subject count in the Placebo arm.

```{r ae-top, exercise=TRUE, exercise.lines=12}
ae_tbl <- tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID)
  ) |>
  build()

ae_tbl |>
  select(row_label_1, n_Placebo) |>
  arrange(desc(___))
```

```{r ae-top-hint-1}
# arrange(desc(n_Placebo)) to sort highest first
```

```{r ae-top-solution}
ae_tbl <- tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID)
  ) |>
  build()

ae_tbl |>
  select(row_label_1, n_Placebo) |>
  arrange(desc(n_Placebo))
```

---

## Section 6 — Ordering

### set_order_count_method()

By default, Tplyr preserves the order in which categories appear in the data
(`"byvar"`). You can change this with `set_order_count_method()`:

| Method | Behaviour |
|--------|-----------|
| `"byvar"` | Preserve data order (default) |
| `"byvarn"` | Sort by the underlying numeric coding of the variable |
| `"bycount"` | Sort by descending total count across all arms |

For AE tables, `"bycount"` is the most common choice — it puts the most
frequent body systems at the top, matching regulatory expectations.

### set_result_order_var()

For more control, `set_result_order_var()` sorts by a specific column in the
built result — typically the count in the reference arm:

```r
group_count(AEBODSYS) |>
  set_order_count_method("bycount") |>
  set_result_order_var(n_Placebo)
```

### Exercise 6.1 — Sort RACE by descending Placebo count

Build a RACE count table sorted by descending Placebo count. Check that the
first row is `"WHITE"`.

```{r sort-race, exercise=TRUE, exercise.lines=10}
race_sorted <- tplyr_table(adsl, TRT01P) |>
  add_layer(
    group_count(RACE) |>
      set_order_count_method(___)
  ) |>
  build()

# Show row_label_1 and n_Placebo in order
race_sorted |>
  select(row_label_1, n_Placebo) |>
  arrange(desc(n_Placebo))
```

```{r sort-race-hint-1}
# set_order_count_method("bycount") sorts by total count descending
```

```{r sort-race-solution}
race_sorted <- tplyr_table(adsl, TRT01P) |>
  add_layer(
    group_count(RACE) |>
      set_order_count_method("bycount")
  ) |>
  build()

race_sorted |>
  select(row_label_1, n_Placebo) |>
  arrange(desc(n_Placebo))
```

### Exercise 6.2 — Sort AE body systems by bycount

Apply `"bycount"` ordering to the AE body system table and print the top 5
body systems by total count.

```{r sort-ae, exercise=TRUE, exercise.lines=14}
ae_sorted <- tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID) |>
      set_order_count_method(___)
  ) |>
  build()

ae_sorted |>
  select(row_label_1, starts_with("n_")) |>
  slice_head(n = 5)
```

```{r sort-ae-hint-1}
# set_order_count_method("bycount")
```

```{r sort-ae-solution}
ae_sorted <- tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID) |>
      set_order_count_method("bycount")
  ) |>
  build()

ae_sorted |>
  select(row_label_1, starts_with("n_")) |>
  slice_head(n = 5)
```

---

## Section 7 — group_shift() for Shift Tables

### What is a shift table?

A **shift table** cross-tabulates where subjects started (baseline category)
against where they ended up (post-baseline category). It is most commonly used
for CTCAE toxicity grades:

```
                    Post-baseline toxicity grade
Baseline grade   Grade 0  Grade 1  Grade 2  Grade 3+
Grade 0           n        n        n        n
Grade 1           n        n        n        n
Grade 2           n        n        n        n
```

This is much more informative than a simple "worst grade" summary because it
captures worsening, improvement, and stability at the same time.

### group_shift() in Tplyr

```r
tplyr_table(dataset, treat_var) |>
  add_layer(
    group_shift(vars(row = BASELINE_GR, column = WEEK12_GR), by = PARAM)
  ) |>
  build()
```

The `vars()` argument specifies which variable drives the *rows* (baseline)
and which drives the *columns* (post-baseline) within each treatment arm.

### When would you use this?

- Lab toxicity grades: `BTOXGR` (baseline) vs `ATOXGR` (worst on-treatment)
- Vital signs: normal/abnormal at baseline vs post-baseline
- Cardiac: QTc category at baseline vs worst post-baseline

### The `by` argument

Use `by` to produce a separate shift table for each value of a grouping
variable. For lab data, `by = PARAM` gives one block per lab parameter:

```r
group_shift(vars(row = BTOXGR, column = ATOXGR), by = PARAM)
```

### Demo: inspect shift_wide structure

The setup chunk created `shift_wide` — a synthetic dataset with one row per
subject containing `BASELINE_GR` and `WEEK12_GR`:

```{r show-shift-wide}
head(shift_wide)
```

### Exercise 7.1 — Define a shift layer

Using `shift_demo` (the long-format version), describe (in words) how you
would structure the data before passing it to `group_shift()`. What format
does the dataset need to be in?

```{r shift-structure, exercise=TRUE, exercise.lines=6}
# Inspect shift_demo structure — what variables does it have?
glimpse(shift_demo)
```

```{r shift-structure-solution}
glimpse(shift_demo)
# shift_demo is in long format with VISIT column.
# For group_shift(), you need wide format: one column for baseline grade,
# one column for post-baseline grade, per subject.
# shift_wide already has BASELINE_GR and WEEK12_GR as separate columns.
```

### Exercise 7.2 — Build a shift layer (group_shift)

Use `shift_wide` to build a shift table with `group_shift()`. The row variable
is `BASELINE_GR` and the result should show how subjects moved between grades.

```{r shift-build, exercise=TRUE, exercise.lines=10}
# group_shift stores layer metadata; build() returns a placeholder in this mock
shift_result <- tplyr_table(shift_wide, TRT01P) |>
  add_layer(
    group_shift(___)
  ) |>
  build()

# Inspect what the layer metadata contains
shift_result
```

```{r shift-build-hint-1}
# group_shift(BASELINE_GR) — pass the baseline grade variable
```

```{r shift-build-solution}
shift_result <- tplyr_table(shift_wide, TRT01P) |>
  add_layer(
    group_shift(BASELINE_GR)
  ) |>
  build()

shift_result
```

> In the real Tplyr package, `group_shift()` returns a full cross-tabulation.
> The mock returns a placeholder — the concept and code structure are identical.

### Quiz

```{r quiz-shift}
quiz(
  question(
    "For a lab toxicity shift table, which variable should drive the ROWS?",
    answer("Worst on-treatment grade (ATOXGR)"),
    answer("Baseline grade (BTOXGR)", correct = TRUE),
    answer("The parameter name (PARAM)"),
    answer("The treatment arm (TRT01A)"),
    allow_retry = TRUE
  )
)
```

---

## Section 8 — Capstone: Demographics + AE Table

### The task

Build a single three-layer table combining:

1. **Layer 1:** Age (`group_desc(AGE)`) — continuous demographics
2. **Layer 2:** Sex (`group_count(SEX)`) — categorical demographics
3. **Layer 3:** AE body system (`group_count(AEBODSYS)`) — AE summary with
   correct denominator from ADSL

After building, arrange by `ord_layer_index` so all layer-1 rows come first,
then layer-2, then layer-3.

### Approach

Because Layer 3 uses a *different* dataset (`adae`) from Layers 1 and 2
(`adsl`), you build it as a separate table and then `bind_rows()` the results.

This is the standard Tplyr pattern for multi-dataset tables:

```r
# Demographics layers
demog_tbl <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  add_layer(group_count(SEX)) |>
  build()

# AE layer (separate table with its own dataset)
ae_tbl <- tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID) |>
      set_order_count_method("bycount")
  ) |>
  build()

# Combine
combined <- bind_rows(demog_tbl, ae_tbl)
```

### Exercise 8.1 — Build the three-layer combined table

Complete the code below to build and combine the demographics and AE layers.

```{r capstone-build, exercise=TRUE, exercise.lines=26}
# Layer 1 + 2: Demographics
demog_tbl <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(___)) |>
  add_layer(group_count(___)) |>
  build()

# Layer 3: AE body system
ae_tbl <- tplyr_table(adae, TRT01A) |>
  set_pop_data(___, ___) |>
  add_layer(
    group_count(___) |>
      set_distinct_by(___) |>
      set_order_count_method(___)
  ) |>
  build()

# Combine both tables
combined <- bind_rows(demog_tbl, ae_tbl)

# Arrange by ord_layer_index and show row counts
combined |>
  arrange(ord_layer_index) |>
  select(ord_layer_index, row_label_1) |>
  print(n = 30)

cat("\nTotal rows in combined table:", nrow(combined), "\n")
```

```{r capstone-build-hint-1}
# Layers 1+2: group_desc(AGE), group_count(SEX)
# AE layer: set_pop_data(adsl, TRT01P), group_count(AEBODSYS),
#           set_distinct_by(USUBJID), set_order_count_method("bycount")
```

```{r capstone-build-solution}
demog_tbl <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  add_layer(group_count(SEX)) |>
  build()

ae_tbl <- tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID) |>
      set_order_count_method("bycount")
  ) |>
  build()

combined <- bind_rows(demog_tbl, ae_tbl)

combined |>
  arrange(ord_layer_index) |>
  select(ord_layer_index, row_label_1) |>
  print(n = 30)

cat("\nTotal rows in combined table:", nrow(combined), "\n")
```

### Exercise 8.2 — Filter to just the AE rows

From `combined`, filter to only the AE body system rows (those with
`ord_layer_index` equal to the value set by the AE layer — which will be 1
since it's the only layer in `ae_tbl`). Show body system and Placebo count.

```{r capstone-filter, exercise=TRUE, exercise.lines=20}
demog_tbl <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  add_layer(group_count(SEX)) |>
  build()

ae_tbl <- tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID) |>
      set_order_count_method("bycount")
  ) |>
  build()

# The AE table has ord_layer_index = 1 (it's its own table)
# Filter and display
ae_tbl |>
  select(row_label_1, n_Placebo) |>
  arrange(___)
```

```{r capstone-filter-hint-1}
# arrange(desc(n_Placebo)) to show highest AE body systems first
```

```{r capstone-filter-solution}
demog_tbl <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  add_layer(group_count(SEX)) |>
  build()

ae_tbl <- tplyr_table(adae, TRT01A) |>
  set_pop_data(adsl, TRT01P) |>
  add_layer(
    group_count(AEBODSYS) |>
      set_distinct_by(USUBJID) |>
      set_order_count_method("bycount")
  ) |>
  build()

ae_tbl |>
  select(row_label_1, n_Placebo) |>
  arrange(desc(n_Placebo))
```

---

## Lesson Summary

### Key Tplyr concepts

| Concept | What you learned |
|---------|-----------------|
| `tplyr_table()` | Defines the base dataset and treatment variable |
| `add_layer()` | Stacks a layer; each layer = one row block in the final table |
| `group_count()` | Subject count + % for categorical variables |
| `group_desc()` | Descriptive statistics (n, mean, sd, median, min, max) |
| `group_shift()` | Baseline vs. post-baseline cross-tabulation |
| `f_str()` | Compact format-string spec: `f_str("xx (xx.x%)", n, pct)` |
| `set_format_strings()` | Attaches an `f_str` to a layer |
| `set_distinct_by()` | Count distinct subjects (critical for AE tables) |
| `set_pop_data()` | Set denominator population for % calculations |
| `set_order_count_method()` | `"byvar"`, `"byvarn"`, `"bycount"` |
| `build()` | Executes all layers; returns a plain tibble |
| `ord_layer_index` | Integer column identifying which layer each row came from |

### Tplyr vs SAS equivalents

| Tplyr | SAS |
|-------|-----|
| `group_count(SEX)` | `PROC FREQ DATA=adsl; TABLE SEX * TRT01P;` |
| `group_desc(AGE)` | `PROC MEANS DATA=adsl N MEAN STD MEDIAN MIN MAX;` |
| `group_shift(...)` | Custom shift-table macro |
| `set_pop_data(adsl)` | `PROC FREQ; WEIGHT from denominator dataset;` |
| `set_order_count_method("bycount")` | `ORDER=FREQ` in PROC FREQ |
| `build()` | `RUN;` — executes and returns output |

### Typical production workflow

```r
# 1. Build with Tplyr
demog <- tplyr_table(adsl, TRT01P) |>
  add_layer(group_desc(AGE)) |>
  add_layer(group_count(SEX)) |>
  add_layer(group_count(RACE)) |>
  build()

# 2. Post-process with dplyr
demog_clean <- demog |>
  arrange(ord_layer_index, ord_layer_1) |>
  select(row_label_1, starts_with("n_"))

# 3. Format and output with r2rtf
demog_clean |>
  rtf_page(orientation = "landscape") |>
  rtf_title("Table 14.1.2 — Demographics") |>
  ...
  rtf_encode() |>
  rtf_write("t_14_1_2.rtf")
```

### Next lesson

**Lesson 37 — tidytlg** covers an alternative tidyverse-native table-generation
framework with a declarative column-definition API, designed for outputs that
go directly to HTML or PDF.
