---
title: "Lesson 37 — tidytlg: Tidy TLG Generation"
output:
  learnr::tutorial:
    theme: flatly
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 37_tidytlg.Rmd
# PURPOSE       : tidytlg package — tidy generation of Tables, Listings, and
#                 Graphs (TLG) for clinical study reports: freq_table(),
#                 cont_table(), nested_freq(), tlg_output(), and the full
#                 tidytlg pipeline from summary stats to submission-ready RTF.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(rlang)
library(pharmaverseadam)

# ── tidytlg MOCK FUNCTIONS ────────────────────────────────────────────────────

statlist <- function(...) {
  stats <- c(...)
  structure(list(stats = stats), class = "statlist")
}

freq_table <- function(df, colvar, rowvar, statlist = NULL,
                       rowbyvar = NULL, subset = NULL,
                       denominator = NULL, total = FALSE) {
  if (!is.null(subset)) df <- df |> dplyr::filter(!!rlang::parse_expr(subset))

  denom_df <- if (!is.null(denominator)) denominator else df

  if (!is.null(rowbyvar)) {
    result <- df |>
      dplyr::group_by(dplyr::across(dplyr::all_of(c(colvar, rowbyvar, rowvar)))) |>
      dplyr::summarise(n = dplyr::n_distinct(USUBJID), .groups = "drop") |>
      dplyr::left_join(
        denom_df |>
          dplyr::group_by(dplyr::across(dplyr::all_of(colvar))) |>
          dplyr::summarise(N = dplyr::n_distinct(USUBJID), .groups = "drop"),
        by = colvar
      ) |>
      dplyr::mutate(
        pct     = round(100 * n / N, 1),
        display = paste0(n, " (", sprintf("%.1f", pct), "%)"),
        label   = as.character(!!rlang::sym(rowvar))
      ) |>
      dplyr::select(dplyr::all_of(c(rowvar, colvar, "display"))) |>
      tidyr::pivot_wider(
        names_from  = dplyr::all_of(colvar),
        values_from = display,
        values_fill = "0 (0.0%)"
      ) |>
      dplyr::rename(label = !!rlang::sym(rowvar))
  } else {
    result <- df |>
      dplyr::group_by(dplyr::across(dplyr::all_of(c(colvar, rowvar)))) |>
      dplyr::summarise(n = dplyr::n_distinct(USUBJID), .groups = "drop") |>
      dplyr::group_by(dplyr::across(dplyr::all_of(colvar))) |>
      dplyr::mutate(N = sum(n)) |>
      dplyr::ungroup() |>
      dplyr::mutate(
        pct     = round(100 * n / N, 1),
        display = paste0(n, " (", sprintf("%.1f", pct), "%)"),
        label   = as.character(!!rlang::sym(rowvar))
      ) |>
      dplyr::select(dplyr::all_of(c(rowvar, colvar, "display"))) |>
      tidyr::pivot_wider(
        names_from  = dplyr::all_of(colvar),
        values_from = display,
        values_fill = "0 (0.0%)"
      ) |>
      dplyr::rename(label = !!rlang::sym(rowvar))
  }

  structure(result, class = c("tidytlg_freq", "tbl_df", "tbl", "data.frame"))
}

cont_table <- function(df, colvar, rowvar,
                       statlist = NULL,
                       rowbyvar = NULL, subset = NULL) {
  if (!is.null(subset)) df <- df |> dplyr::filter(!!rlang::parse_expr(subset))

  stats_out <- df |>
    dplyr::group_by(dplyr::across(dplyr::all_of(colvar))) |>
    dplyr::summarise(
      N      = sum(!is.na(.data[[rowvar]])),
      Mean   = round(mean(.data[[rowvar]], na.rm = TRUE), 1),
      SD     = round(sd(.data[[rowvar]],   na.rm = TRUE), 1),
      Median = round(median(.data[[rowvar]], na.rm = TRUE), 1),
      Min    = min(.data[[rowvar]], na.rm = TRUE),
      Max    = max(.data[[rowvar]], na.rm = TRUE),
      Q1     = round(quantile(.data[[rowvar]], 0.25, na.rm = TRUE), 1),
      Q3     = round(quantile(.data[[rowvar]], 0.75, na.rm = TRUE), 1),
      .groups = "drop"
    ) |>
    dplyr::mutate(
      mean_sd  = paste0(Mean, " (", SD, ")"),
      range    = paste0(Min, ", ", Max),
      q1q3     = paste0(Q1, ", ", Q3)
    )

  stat_rows <- tibble::tribble(
    ~label,          ~stat_col,
    "N",             "N",
    "Mean (SD)",     "mean_sd",
    "Median",        "Median",
    "Range",         "range"
  )

  result <- lapply(seq_len(nrow(stat_rows)), function(i) {
    sr <- stat_rows[i, ]
    # Coerce to character first so all pivoted arm columns share the same type;
    # without this, N (integer) vs mean_sd/range (character) causes bind_rows() to fail.
    row_vals <- stats_out |>
      dplyr::select(dplyr::all_of(c(colvar, sr$stat_col))) |>
      dplyr::mutate(dplyr::across(dplyr::all_of(sr$stat_col), as.character)) |>
      tidyr::pivot_wider(
        names_from  = dplyr::all_of(colvar),
        values_from = dplyr::all_of(sr$stat_col)
      )
    row_vals$label <- sr$label
    row_vals
  }) |> dplyr::bind_rows()

  structure(result, class = c("tidytlg_cont", "tbl_df", "tbl", "data.frame"))
}

nested_freq <- function(df, colvar, dptvar, dptvarn = NULL,
                        rowbyvar = NULL, subset = NULL,
                        denominator = NULL) {
  if (!is.null(subset)) df <- df |> dplyr::filter(!!rlang::parse_expr(subset))
  denom_df <- if (!is.null(denominator)) denominator else df

  grp_vars <- if (!is.null(rowbyvar)) c(colvar, rowbyvar, dptvar) else c(colvar, dptvar)

  result <- df |>
    dplyr::group_by(dplyr::across(dplyr::all_of(grp_vars))) |>
    dplyr::summarise(n = dplyr::n_distinct(USUBJID), .groups = "drop")

  N_denom <- denom_df |>
    dplyr::group_by(dplyr::across(dplyr::all_of(colvar))) |>
    dplyr::summarise(N = dplyr::n_distinct(USUBJID), .groups = "drop")

  result <- result |>
    dplyr::left_join(N_denom, by = colvar) |>
    dplyr::mutate(
      pct     = round(100 * n / N, 1),
      display = paste0(n, " (", sprintf("%.1f", pct), "%)"),
      label   = paste0("  ", .data[[dptvar]])
    )

  keep_cols <- c(if (!is.null(rowbyvar)) rowbyvar else NULL,
                 dptvar, colvar, "display", "label")

  wide <- result |>
    dplyr::select(dplyr::all_of(keep_cols)) |>
    tidyr::pivot_wider(
      names_from  = dplyr::all_of(colvar),
      values_from = display,
      values_fill = "0 (0.0%)"
    )

  structure(wide, class = c("tidytlg_nested", "tbl_df", "tbl", "data.frame"))
}

tlg_output <- function(df, tbl_id = "T01",
                       title = "Table",
                       colhead = NULL,
                       font_size = 9,
                       orientation = "landscape",
                       output_format = "rtf",
                       output_file = "output.rtf",
                       ...) {
  cat(sprintf("=== tidytlg Output: %s ===\n", tbl_id))
  cat("Title:", paste(title, collapse = "\n       "), "\n")
  cat("Format:", output_format, "| Orientation:", orientation,
      "| Font:", font_size, "pt\n")
  cat("Output file:", output_file, "\n")
  cat("Table dimensions:", nrow(df), "rows x", ncol(df), "cols\n")
  if (!is.null(colhead)) cat("Column headers:", colhead, "\n")
  print(df, n = 5)
  invisible(df)
}

gentlg <- function(huxme, tlf = "table", ...) {
  cat("[gentlg: combined compute+output - see tlg_output() for details]\n")
  invisible(huxme)
}

# ── DATASETS ──────────────────────────────────────────────────────────────────

adsl <- pharmaverseadam::adsl

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- dplyr::rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- dplyr::mutate(adlb,
                        ATOXGR = as.integer(ATOXGR),
                        BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- dplyr::filter(adlb, is.na(DTYPE))
if (!"TRT01A" %in% names(adlb))
  adlb <- dplyr::left_join(adlb,
                            dplyr::select(adsl, USUBJID, TRT01A = TRT01P),
                            by = "USUBJID")

# ── PRE-BUILT OBJECTS for exercises ──────────────────────────────────────────

demog_freq <- freq_table(adsl, colvar = "TRT01P", rowvar = "SEX")

age_cont   <- cont_table(adsl, colvar = "TRT01P", rowvar = "AGE")

ae_nested  <- nested_freq(
  adae,
  colvar    = "TRT01A",
  dptvar    = "AEDECOD",
  rowbyvar  = "AEBODSYS",
  denominator = adsl   # adsl already carries TRT01A — no rename needed
)

tutorial_options(exercise.timelimit = 60)
```

## Overview

Welcome to Lesson 37! This tutorial covers **tidytlg**, an open-source R package
from Eli Lilly that provides a tidyverse-native pipeline for generating clinical
Tables, Listings, and Graphs (TLGs) for study reports.

### What you will learn

| Section | Topic |
|---------|-------|
| 1 | tidytlg philosophy: compute then format |
| 2 | `freq_table()` for categorical variables |
| 3 | `cont_table()` for continuous variables |
| 4 | `statlist()` and format control |
| 5 | `nested_freq()` for AE body system / preferred term tables |
| 6 | Combining tables with `bind_rows()` |
| 7 | `tlg_output()` for rendering to RTF / HTML / PDF |
| 8 | Capstone: full TLG pipeline |

### Packages used

```r
library(tidytlg)   # main package — mocked in this tutorial
library(dplyr)
library(tidyr)
library(tibble)
library(rlang)
library(pharmaverseadam)
```

> **Note:** tidytlg is mocked in this tutorial so all exercises run
> interactively. The function signatures and outputs match the real package
> behaviour.

---

## Section 1 — tidytlg Philosophy: Compute Then Format

### The SAS way vs. the tidytlg way

In SAS, computation and formatting are tightly coupled. A `PROC FREQ` call
produces a table, applies formats, and writes ODS output all in one step.
If you want to stack two tables you need `ODS OUTPUT` into data sets, then
`PROC REPORT` glue code.

**tidytlg separates the two concerns:**

| Step | Function(s) | Output |
|------|-------------|--------|
| Compute | `freq_table()`, `cont_table()`, `nested_freq()` | Plain `tibble` |
| Format / render | `tlg_output()` | RTF, HTML, or PDF file |

Because the intermediate result is just a `tibble`, you can use every dplyr
verb you already know: `filter()`, `arrange()`, `bind_rows()`, `mutate()`, etc.

### A first look at freq_table()

```r
freq_table(adsl, colvar = "TRT01P", rowvar = "SEX")
```

`colvar` defines the columns (treatment arms). `rowvar` defines the rows
(levels of the categorical variable). The result has one row per level:

```
# A tidytlg_freq tibble
  label  Placebo              Xanomeline High Dose  Xanomeline Low Dose
  <chr>  <chr>                <chr>                 <chr>
1 F      …                    …                     …
2 M      …                    …                     …
```

Every cell shows **n (x.x%)** by default — the count of distinct subjects and
their percentage within that treatment arm.

### Demo

Run the chunk below to see the actual output:

```{r demo-freq-sex, exercise=TRUE, exercise.lines=4}
freq_table(adsl, colvar = "TRT01P", rowvar = "SEX")
```

### Exercise 1.1 — RACE frequency table

Compute a frequency table for `RACE` across treatment arms. How many rows does
the result have? (Each race category = one row.)

```{r ex1-race-freq, exercise=TRUE, exercise.lines=6}
race_freq <- freq_table(___, colvar = "___", rowvar = "___")
nrow(race_freq)
```

```{r ex1-race-freq-hint-1}
# Use adsl as the data frame; colvar is the treatment column TRT01P
# rowvar is RACE
```

```{r ex1-race-freq-solution}
race_freq <- freq_table(adsl, colvar = "TRT01P", rowvar = "RACE")
nrow(race_freq)
# Returns the number of distinct RACE categories in adsl
```

### Exercise 1.2 — inspect the tibble class

Run `class()` on the `demog_freq` object (pre-built from `freq_table(adsl, "TRT01P", "SEX")`).
What does it tell you about the object type?

```{r ex1-class, exercise=TRUE, exercise.lines=4}
class(demog_freq)
```

```{r ex1-class-solution}
class(demog_freq)
# "tidytlg_freq" "tbl_df" "tbl" "data.frame"
# It IS a plain tibble — you can use any dplyr verb on it.
```

---

## Section 2 — freq_table() for Categorical Variables

### Required arguments

```r
freq_table(df, colvar, rowvar)
```

| Argument | Type | Purpose |
|----------|------|---------|
| `df` | data frame | Source data |
| `colvar` | string | Column-stratifying variable (treatment arm) |
| `rowvar` | string | Row variable (categorical levels become rows) |
| `statlist` | `statlist()` | Which statistics to display (default: n and %) |
| `rowbyvar` | string | Optional outer grouping variable |
| `subset` | string expression | Row filter (passed as a string, e.g. `"SAFFL == 'Y'"`) |
| `denominator` | data frame | Override the denominator (critical for AE tables) |

### Output structure

The result is a wide tibble: one row per category level, one column per
treatment arm. The `label` column holds the display text.

```r
freq_table(adsl, colvar = "TRT01P", rowvar = "SEX")
# label    Placebo          Xanomeline High Dose   Xanomeline Low Dose
# F        54 (42.5%)       40 (31.5%)             40 (31.5%)
# M        …                …                      …
```

### The denominator argument

When computing AE proportions, the denominator should be the number of
**subjects** (from `adsl`), not the number of AE records (from `adae`).
Pass `denominator = adsl` — the treatment column name in the denominator
must match `colvar`. The `pharmaverseadam` ADSL already contains `TRT01A`,
so no renaming is necessary:

```r
freq_table(
  adae,
  colvar      = "TRT01A",
  rowvar      = "AEBODSYS",
  denominator = adsl   # adsl has TRT01A — column names already match
)
```

### Exercise 2.1 — SEX frequency table

Run `freq_table` for SEX. In the Xanomeline High Dose arm, how many female
subjects are there? Look at the cell value — it shows `n (x.x%)`.

```{r ex2-sex, exercise=TRUE, exercise.lines=5}
sex_freq <- freq_table(adsl, colvar = "TRT01P", rowvar = "SEX")
sex_freq
```

```{r ex2-sex-solution}
sex_freq <- freq_table(adsl, colvar = "TRT01P", rowvar = "SEX")
sex_freq
# Look at the "F" row under "Xanomeline High Dose"
# The number before the parenthesis is the count of female subjects.
```

### Exercise 2.2 — AE body system frequency with denominator

Compute a frequency table of `AEBODSYS` from `adae`, using `adsl` as the
denominator. `adsl` already has a `TRT01A` column, so pass it directly —
no renaming needed.

```{r ex2-ae-denom, exercise=TRUE, exercise.lines=7}
ae_bodsys_freq <- freq_table(
  ___,
  colvar      = "TRT01A",
  rowvar      = "___",
  denominator = ___
)
ae_bodsys_freq
```

```{r ex2-ae-denom-hint-1}
# data = adae, rowvar = "AEBODSYS"
# denominator = adsl  (adsl already has TRT01A — column names already match)
```

```{r ex2-ae-denom-solution}
ae_bodsys_freq <- freq_table(
  adae,
  colvar      = "TRT01A",
  rowvar      = "AEBODSYS",
  denominator = adsl
)
ae_bodsys_freq
```

### Exercise 2.3 — subset argument

Compute the frequency of `AGEGR1` (age group) restricted to subjects with
`SEX == "F"` using the `subset` argument (pass as a character string).

```{r ex2-subset, exercise=TRUE, exercise.lines=6}
agegr_female <- freq_table(
  adsl,
  colvar = "TRT01P",
  rowvar = "AGEGR1",
  subset = "___"
)
agegr_female
```

```{r ex2-subset-hint-1}
# subset takes a character string: subset = "SEX == 'F'"
```

```{r ex2-subset-solution}
agegr_female <- freq_table(
  adsl,
  colvar = "TRT01P",
  rowvar = "AGEGR1",
  subset = "SEX == 'F'"
)
agegr_female
```

### Quiz

```{r quiz-sec2, echo=FALSE}
quiz(
  question(
    "What does the `denominator` argument in `freq_table()` control?",
    answer("The number of decimal places in percentages"),
    answer("The data frame used to calculate the 'N' in each percentage",
           correct = TRUE),
    answer("The column used to label each row"),
    answer("Whether totals are appended to the table"),
    allow_retry = TRUE
  ),
  question(
    "In a `freq_table()` result, what does the `label` column contain?",
    answer("The column variable name"),
    answer("The formatted n (x.x%) cell value"),
    answer("The levels of the row variable", correct = TRUE),
    answer("The study identifier"),
    allow_retry = TRUE
  )
)
```

---

## Section 3 — cont_table() for Continuous Variables

### What cont_table() returns

`cont_table()` computes descriptive statistics for a continuous variable and
returns them as **separate rows** in a tidy tibble — one row per statistic:

| label | Placebo | Xanomeline High Dose | Xanomeline Low Dose |
|-------|---------|----------------------|---------------------|
| N | 86 | 84 | 84 |
| Mean (SD) | 75.2 (8.6) | 74.4 (7.9) | 75.7 (8.3) |
| Median | 76.0 | 76.0 | 77.0 |
| Range | 52, 89 | 56, 88 | 51, 88 |

This "long-by-statistic" layout maps directly to a clinical study report table.

### Arguments

```r
cont_table(df, colvar, rowvar, statlist = NULL, rowbyvar = NULL, subset = NULL)
```

`rowvar` is the **continuous** variable name. `colvar` is still the treatment
column. The result has no denominator argument — percentages do not apply.

### Demo

```{r demo-cont-age, exercise=TRUE, exercise.lines=4}
cont_table(adsl, colvar = "TRT01P", rowvar = "AGE")
```

### Exercise 3.1 — extract the Mean (SD) row

Using the pre-built `age_cont` object, filter to the "Mean (SD)" row and
print the Placebo value.

```{r ex3-mean-sd, exercise=TRUE, exercise.lines=6}
age_cont |>
  dplyr::filter(label == "___") |>
  dplyr::select(label, ___)
```

```{r ex3-mean-sd-hint-1}
# label == "Mean (SD)"
# Select the label column and the Placebo column
```

```{r ex3-mean-sd-solution}
age_cont |>
  dplyr::filter(label == "Mean (SD)") |>
  dplyr::select(label, Placebo)
# Shows the Mean (SD) for AGE in the Placebo arm
```

### Exercise 3.2 — continuous table for WEIGHTBL (if present)

Check whether `WEIGHTBL` exists in `adsl`. If it does, compute a `cont_table`
for it. Use `"TRT01P"` as `colvar`.

```{r ex3-weightbl, exercise=TRUE, exercise.lines=8}
if ("WEIGHTBL" %in% names(adsl)) {
  cont_table(adsl, colvar = "TRT01P", rowvar = "___")
} else {
  message("WEIGHTBL not found — using AGE as fallback")
  cont_table(adsl, colvar = "TRT01P", rowvar = "AGE")
}
```

```{r ex3-weightbl-solution}
if ("WEIGHTBL" %in% names(adsl)) {
  cont_table(adsl, colvar = "TRT01P", rowvar = "WEIGHTBL")
} else {
  message("WEIGHTBL not found — using AGE as fallback")
  cont_table(adsl, colvar = "TRT01P", rowvar = "AGE")
}
```

### Exercise 3.3 — cont_table on lab data

Compute a continuous stats table for AVAL from `adlb` filtered to the ALT
parameter. Use `TRT01A` as `colvar`.

```{r ex3-lab-alt, exercise=TRUE, exercise.lines=8}
alt_stats <- cont_table(
  adlb |> dplyr::filter(PARAMCD == "___"),
  colvar = "TRT01A",
  rowvar = "AVAL"
)
alt_stats
```

```{r ex3-lab-alt-hint-1}
# PARAMCD == "ALT"
# TRT01A is already in adlb (added in setup)
```

```{r ex3-lab-alt-solution}
alt_stats <- cont_table(
  adlb |> dplyr::filter(PARAMCD == "ALT"),
  colvar = "TRT01A",
  rowvar = "AVAL"
)
alt_stats
```

---

## Section 4 — statlist() and Format Control

### What statlist() does

`statlist()` is a helper that communicates to `freq_table()` or `cont_table()`
which statistics to include and how to label them. It returns a small list
object of class `"statlist"`.

```r
statlist("n (x.x%)")                          # count + percent (default for freq)
statlist("N", "mean (sd)", "median", "range") # standard continuous stats
statlist("n")                                  # count only, no percent
statlist("median (range)")                     # median with min–max
```

### Common statlist tokens

| Token | Meaning |
|-------|---------|
| `"n"` | Subject count |
| `"n (x.x%)"` | Count and percent |
| `"N"` | Non-missing count |
| `"mean (sd)"` | Mean and standard deviation |
| `"median"` | Median |
| `"range"` | Min, Max |
| `"q1q3"` | First and third quartiles |

> In the mocked functions used here, `statlist()` stores the token strings and
> they are printed but the underlying compute functions always return the full
> set of statistics. In the real tidytlg package, the statlist controls both
> which rows appear and how cells are formatted.

### Exercise 4.1 — inspect a statlist object

Create a statlist for N, median, and q1q3. Print it and check its class.

```{r ex4-statlist-obj, exercise=TRUE, exercise.lines=6}
my_sl <- statlist("___", "___", "___")
print(my_sl)
class(my_sl)
```

```{r ex4-statlist-hint-1}
# statlist("N", "median", "q1q3")
```

```{r ex4-statlist-solution}
my_sl <- statlist("N", "median", "q1q3")
print(my_sl)
class(my_sl)
# class is "statlist" — the function wraps the tokens in a named list
```

### Exercise 4.2 — pass statlist to cont_table

Pass a statlist specifying only N and Median to `cont_table` for AGE. Observe
the statlist is accepted (in the mock the rows are fixed, but in real tidytlg
the rows would match your statlist).

```{r ex4-pass-statlist, exercise=TRUE, exercise.lines=7}
sl_n_med <- statlist("N", "median")

cont_table(
  adsl,
  colvar   = "TRT01P",
  rowvar   = "AGE",
  statlist = ___
)
```

```{r ex4-pass-statlist-solution}
sl_n_med <- statlist("N", "median")

cont_table(
  adsl,
  colvar   = "TRT01P",
  rowvar   = "AGE",
  statlist = sl_n_med
)
```

### Exercise 4.3 — freq_table with count-only statlist

Compute a frequency table for `AGEGR1` using a statlist that requests counts
only (no percent).

```{r ex4-count-only, exercise=TRUE, exercise.lines=7}
sl_count <- statlist("___")

freq_table(
  adsl,
  colvar   = "TRT01P",
  rowvar   = "AGEGR1",
  statlist = ___
)
```

```{r ex4-count-only-hint-1}
# statlist("n") for count only
```

```{r ex4-count-only-solution}
sl_count <- statlist("n")

freq_table(
  adsl,
  colvar   = "TRT01P",
  rowvar   = "AGEGR1",
  statlist = sl_count
)
```

### Quiz

```{r quiz-sec4, echo=FALSE}
quiz(
  question(
    "Which statlist token requests the interquartile range (Q1 and Q3)?",
    answer('"iqr"'),
    answer('"quartile"'),
    answer('"q1q3"', correct = TRUE),
    answer('"p25_p75"'),
    allow_retry = TRUE
  ),
  question(
    "What R class does `statlist()` return?",
    answer('"character"'),
    answer('"list"'),
    answer('"statlist"', correct = TRUE),
    answer('"tbl_df"'),
    allow_retry = TRUE
  )
)
```

---

## Section 5 — nested_freq() for AE Tables

### Two-level nesting

A standard adverse event table in a clinical study report shows:

```
NERVOUS SYSTEM DISORDERS                12 (14.1%)  …
  Headache                               6 (7.1%)   …
  Dizziness                              4 (4.7%)   …
GASTROINTESTINAL DISORDERS              …
  Nausea                                …
```

`nested_freq()` produces exactly this layout. The outer level (body system)
uses `rowbyvar`; the inner level (preferred term) uses `dptvar`. PT rows are
automatically indented in the `label` column.

### Arguments

```r
nested_freq(
  df,
  colvar    = "TRT01A",   # treatment column
  dptvar    = "AEDECOD",  # preferred term (inner level)
  rowbyvar  = "AEBODSYS", # body system (outer level)
  denominator = adsl_renamed  # use subject counts not AE counts
)
```

> **Critical:** Always pass `denominator` as the `adsl` data frame (with
> the treatment column renamed to match `colvar`). Using `adae` as the
> denominator would divide by AE records, not subjects — a serious error.

### Exercise 5.1 — inspect ae_nested

The pre-built `ae_nested` object was created with:

```r
ae_nested <- nested_freq(
  adae,
  colvar      = "TRT01A",
  dptvar      = "AEDECOD",
  rowbyvar    = "AEBODSYS",
  denominator = adsl   # adsl already has TRT01A
)
```

Inspect it: how many preferred terms fall under `"NERVOUS SYSTEM DISORDERS"`?

```{r ex5-inspect-nested, exercise=TRUE, exercise.lines=6}
ae_nested |>
  dplyr::filter(AEBODSYS == "___") |>
  dplyr::select(AEBODSYS, AEDECOD, label) |>
  nrow()
```

```{r ex5-inspect-nested-hint-1}
# AEBODSYS == "NERVOUS SYSTEM DISORDERS"
```

```{r ex5-inspect-nested-solution}
ae_nested |>
  dplyr::filter(AEBODSYS == "NERVOUS SYSTEM DISORDERS") |>
  dplyr::select(AEBODSYS, AEDECOD, label) |>
  nrow()
# The number of rows = number of distinct preferred terms under that body system
```

### Exercise 5.2 — build a nested_freq from scratch

Build a `nested_freq` for AEs in the GASTROINTESTINAL DISORDERS body system
only, using `subset`. Check the count of preferred terms.

```{r ex5-gi-nested, exercise=TRUE, exercise.lines=9}
gi_nested <- nested_freq(
  adae,
  colvar      = "TRT01A",
  dptvar      = "AEDECOD",
  rowbyvar    = "AEBODSYS",
  subset      = "AEBODSYS == 'GASTROINTESTINAL DISORDERS'",
  denominator = ___
)
nrow(gi_nested)
```

```{r ex5-gi-nested-hint-1}
# denominator = adsl
# adsl already has TRT01A — no rename needed; column names already match colvar
```

```{r ex5-gi-nested-solution}
gi_nested <- nested_freq(
  adae,
  colvar      = "TRT01A",
  dptvar      = "AEDECOD",
  rowbyvar    = "AEBODSYS",
  subset      = "AEBODSYS == 'GASTROINTESTINAL DISORDERS'",
  denominator = adsl
)
nrow(gi_nested)
```

### Exercise 5.3 — check label indentation

The `label` column in a `nested_freq` result indents PT rows with two spaces.
Verify this for the first 3 rows of `ae_nested`.

```{r ex5-label-indent, exercise=TRUE, exercise.lines=4}
ae_nested |>
  dplyr::select(label) |>
  head(___)
```

```{r ex5-label-indent-solution}
ae_nested |>
  dplyr::select(label) |>
  head(3)
# Each label starts with "  " (two spaces) — the PT indentation
```

### Exercise 5.4 — most common AE preferred term

From `ae_nested`, find the preferred term with the highest count in any arm.
Hint: the count is embedded in the display string — use `tidyr::separate()` or
just sort by one treatment column alphabetically (higher n → larger leading
digits).

```{r ex5-common-pt, exercise=TRUE, exercise.lines=8}
# Sort ae_nested to find the row with the largest count
# Pick the Xanomeline High Dose column for sorting
arm_col <- names(ae_nested)[3]   # third column after rowbyvar and dptvar

ae_nested |>
  dplyr::arrange(dplyr::desc(.data[[arm_col]])) |>
  dplyr::select(AEBODSYS, AEDECOD, label, dplyr::all_of(arm_col)) |>
  head(5)
```

```{r ex5-common-pt-solution}
arm_col <- names(ae_nested)[3]

ae_nested |>
  dplyr::arrange(dplyr::desc(.data[[arm_col]])) |>
  dplyr::select(AEBODSYS, AEDECOD, label, dplyr::all_of(arm_col)) |>
  head(5)
# Shows the preferred terms with the highest subject counts
```

---

## Section 6 — Combining Tables with bind_rows()

### Stack any tidytlg outputs

Because `freq_table()`, `cont_table()`, and `nested_freq()` all return plain
tibbles, you can stack them with `dplyr::bind_rows()`. This is how you build
a multi-block demographics table that shows:

1. Age statistics (continuous)
2. Sex distribution (categorical)
3. Race distribution (categorical)

all in one table object — matching the layout of ICH E3 Table 14.1.1.

### Pattern

```r
# Add an ordering column before binding
age_block  <- cont_table(adsl, "TRT01P", "AGE")  |> mutate(ord = 1L)
sex_block  <- freq_table(adsl, "TRT01P", "SEX")  |> mutate(ord = 2L)
race_block <- freq_table(adsl, "TRT01P", "RACE") |> mutate(ord = 3L)

combined <- bind_rows(age_block, sex_block, race_block) |>
  arrange(ord)
```

### Blank separator rows

Add a visual blank row between blocks using `add_row()`:

```r
spacer <- tibble(label = "", ord = 1L)  # fill other cols as needed
combined <- bind_rows(age_block, spacer, sex_block)
```

### Exercise 6.1 — bind age_cont and demog_freq

Using the pre-built `age_cont` and `demog_freq` objects, add an `ord` column
to each and bind them. Arrange by `ord`.

```{r ex6-bind-rows, exercise=TRUE, exercise.lines=10}
age_block  <- age_cont  |> dplyr::mutate(ord = ___)
sex_block  <- demog_freq |> dplyr::mutate(ord = ___)

combined <- dplyr::bind_rows(age_block, sex_block) |>
  dplyr::arrange(___)

combined |> dplyr::select(label, ord)
```

```{r ex6-bind-rows-hint-1}
# age_block gets ord = 1L, sex_block gets ord = 2L
# arrange(ord)
```

```{r ex6-bind-rows-solution}
age_block  <- age_cont   |> dplyr::mutate(ord = 1L)
sex_block  <- demog_freq |> dplyr::mutate(ord = 2L)

combined <- dplyr::bind_rows(age_block, sex_block) |>
  dplyr::arrange(ord)

combined |> dplyr::select(label, ord)
```

### Exercise 6.2 — three-block demographics table

Extend the combined table to include a RACE block as `ord = 3L`.

```{r ex6-three-block, exercise=TRUE, exercise.lines=12}
age_block  <- age_cont   |> dplyr::mutate(ord = 1L)
sex_block  <- demog_freq |> dplyr::mutate(ord = 2L)
race_block <- freq_table(adsl, colvar = "TRT01P", rowvar = "___") |>
  dplyr::mutate(ord = 3L)

demog_combined <- dplyr::bind_rows(age_block, sex_block, race_block) |>
  dplyr::arrange(ord)

nrow(demog_combined)
```

```{r ex6-three-block-hint-1}
# rowvar = "RACE"
```

```{r ex6-three-block-solution}
age_block  <- age_cont   |> dplyr::mutate(ord = 1L)
sex_block  <- demog_freq |> dplyr::mutate(ord = 2L)
race_block <- freq_table(adsl, colvar = "TRT01P", rowvar = "RACE") |>
  dplyr::mutate(ord = 3L)

demog_combined <- dplyr::bind_rows(age_block, sex_block, race_block) |>
  dplyr::arrange(ord)

nrow(demog_combined)
# Total rows = 4 (age stats) + 2 (sex levels) + n_race_levels
```

### Exercise 6.3 — verify label column after bind

After binding tables of different classes (`tidytlg_cont` and `tidytlg_freq`),
confirm that `label` is still a character column in the combined result.

```{r ex6-label-class, exercise=TRUE, exercise.lines=6}
age_block <- age_cont  |> dplyr::mutate(ord = 1L)
sex_block <- demog_freq |> dplyr::mutate(ord = 2L)
combined  <- dplyr::bind_rows(age_block, sex_block)

class(combined$label)
```

```{r ex6-label-class-solution}
age_block <- age_cont   |> dplyr::mutate(ord = 1L)
sex_block <- demog_freq |> dplyr::mutate(ord = 2L)
combined  <- dplyr::bind_rows(age_block, sex_block)

class(combined$label)
# "character" — bind_rows coerces consistently
```

---

## Section 7 — tlg_output() for Rendering

### tlg_output() overview

Once you have a final combined tibble, `tlg_output()` applies the submission
display layer: column headers, font, orientation, and file format.

```r
tlg_output(
  df,
  tbl_id         = "T14-1-1",
  title          = c("Table 14.1.1", "Demographic Characteristics"),
  colhead        = NULL,
  font_size      = 9,
  orientation    = "landscape",
  output_format  = "rtf",
  output_file    = "T14-1-1.rtf"
)
```

### Key arguments

| Argument | Default | Notes |
|----------|---------|-------|
| `tbl_id` | `"T01"` | Table identifier — matches define-XML traceability |
| `title` | `"Table"` | Character vector; each element = one title line |
| `colhead` | `NULL` | Override column header labels |
| `output_format` | `"rtf"` | Also `"html"` or `"pdf"` |
| `output_file` | `"output.rtf"` | File path for the output |

> **Always specify `tbl_id`**. In e-submissions, this identifier links the
> table to its entry in the define-XML table-of-contents and to the
> Reviewer's Guide.

In this mocked version, `tlg_output()` prints a summary to the console and
calls `print()` on the tibble — no file is written.

### Exercise 7.1 — render age_cont

Call `tlg_output()` on `age_cont` with a two-line title and
`tbl_id = "T14-1-1"`.

```{r ex7-render-age, exercise=TRUE, exercise.lines=10}
tlg_output(
  age_cont,
  tbl_id        = "___",
  title         = c("Table 14.1.1", "Age — Descriptive Statistics"),
  output_format = "rtf",
  output_file   = "T14-1-1.rtf"
)
```

```{r ex7-render-age-hint-1}
# tbl_id = "T14-1-1"
```

```{r ex7-render-age-solution}
tlg_output(
  age_cont,
  tbl_id        = "T14-1-1",
  title         = c("Table 14.1.1", "Age — Descriptive Statistics"),
  output_format = "rtf",
  output_file   = "T14-1-1.rtf"
)
# Output shows: Table dimensions: 4 rows x 4 cols (label + 3 treatment cols)
```

### Exercise 7.2 — combined demographics table output

Build the combined demographics table from Section 6 (age + sex + race),
then render it. Note the total row count reported.

```{r ex7-combined-output, exercise=TRUE, exercise.lines=16}
age_block  <- age_cont   |> dplyr::mutate(ord = 1L)
sex_block  <- demog_freq |> dplyr::mutate(ord = 2L)
race_block <- freq_table(adsl, "TRT01P", "RACE") |> dplyr::mutate(ord = 3L)

demog_tbl  <- dplyr::bind_rows(age_block, sex_block, race_block) |>
  dplyr::arrange(ord)

tlg_output(
  demog_tbl,
  tbl_id        = "___",
  title         = c("Table 14.1.1", "Demographic Characteristics — All Subjects"),
  output_format = "rtf",
  output_file   = "T14-1-1.rtf"
)
```

```{r ex7-combined-output-hint-1}
# tbl_id = "T14-1-1"
```

```{r ex7-combined-output-solution}
age_block  <- age_cont   |> dplyr::mutate(ord = 1L)
sex_block  <- demog_freq |> dplyr::mutate(ord = 2L)
race_block <- freq_table(adsl, "TRT01P", "RACE") |> dplyr::mutate(ord = 3L)

demog_tbl  <- dplyr::bind_rows(age_block, sex_block, race_block) |>
  dplyr::arrange(ord)

tlg_output(
  demog_tbl,
  tbl_id        = "T14-1-1",
  title         = c("Table 14.1.1", "Demographic Characteristics — All Subjects"),
  output_format = "rtf",
  output_file   = "T14-1-1.rtf"
)
# "Table dimensions: N rows x 5 cols" — N = 4 + 2 + n_race_levels
```

### Exercise 7.3 — HTML output format

Re-render `age_cont` with `output_format = "html"` and a custom
`output_file = "T14-1-1.html"`. Note the format line in the output.

```{r ex7-html, exercise=TRUE, exercise.lines=8}
tlg_output(
  age_cont,
  tbl_id        = "T14-1-1",
  title         = "Age — Descriptive Statistics",
  output_format = "___",
  output_file   = "T14-1-1.html"
)
```

```{r ex7-html-solution}
tlg_output(
  age_cont,
  tbl_id        = "T14-1-1",
  title         = "Age — Descriptive Statistics",
  output_format = "html",
  output_file   = "T14-1-1.html"
)
# Format line shows: "html | Orientation: landscape | Font: 9 pt"
```

### Quiz

```{r quiz-sec7, echo=FALSE}
quiz(
  question(
    "Why is the `tbl_id` argument important in `tlg_output()`?",
    answer("It determines the number of decimal places"),
    answer("It links the table to define-XML and e-submission traceability",
           correct = TRUE),
    answer("It sets the font size"),
    answer("It controls whether totals are shown"),
    allow_retry = TRUE
  ),
  question(
    "What does `tlg_output()` return invisibly?",
    answer("The output file path as a character string"),
    answer("NULL"),
    answer("The input data frame unchanged", correct = TRUE),
    answer("A summary statistics list"),
    allow_retry = TRUE
  )
)
```

---

## Section 8 — Capstone: Full TLG Pipeline

In this capstone you will build three complete TLG outputs that mirror a
real clinical study report appendix:

| Table ID | Description | Source data |
|----------|-------------|-------------|
| T14.1.1 | Demographics | `adsl` |
| T14.3.1 | AE summary by body system | `adae` / `adsl` |
| T14.4.1 | ALT lab statistics | `adlb` |

### Part 1 — Demographics (T14.1.1)

```{r cap-demog, exercise=TRUE, exercise.lines=22}
# Step 1: compute blocks
age_block  <- cont_table(adsl, colvar = "TRT01P", rowvar = "AGE") |>
  dplyr::mutate(ord = 1L)

sex_block  <- freq_table(adsl, colvar = "TRT01P", rowvar = "SEX") |>
  dplyr::mutate(ord = 2L)

race_block <- freq_table(adsl, colvar = "___",   rowvar = "___") |>
  dplyr::mutate(ord = 3L)

# Step 2: combine
t1411 <- dplyr::bind_rows(age_block, sex_block, race_block) |>
  dplyr::arrange(ord)

# Step 3: render
tlg_output(
  t1411,
  tbl_id        = "T14.1.1",
  title         = c("Table 14.1.1", "Demographic Characteristics"),
  output_format = "rtf",
  output_file   = "T14.1.1.rtf"
)
```

```{r cap-demog-hint-1}
# race_block: colvar = "TRT01P", rowvar = "RACE"
```

```{r cap-demog-solution}
age_block  <- cont_table(adsl, colvar = "TRT01P", rowvar = "AGE") |>
  dplyr::mutate(ord = 1L)

sex_block  <- freq_table(adsl, colvar = "TRT01P", rowvar = "SEX") |>
  dplyr::mutate(ord = 2L)

race_block <- freq_table(adsl, colvar = "TRT01P", rowvar = "RACE") |>
  dplyr::mutate(ord = 3L)

t1411 <- dplyr::bind_rows(age_block, sex_block, race_block) |>
  dplyr::arrange(ord)

tlg_output(
  t1411,
  tbl_id        = "T14.1.1",
  title         = c("Table 14.1.1", "Demographic Characteristics"),
  output_format = "rtf",
  output_file   = "T14.1.1.rtf"
)
```

### Part 2 — AE Overview (T14.3.1)

```{r cap-ae, exercise=TRUE, exercise.lines=18}
# Denominator: adsl (already has TRT01A — column names match adae)
adsl_ae_denom <- adsl  # adsl already has TRT01A

# Compute AE body system frequency
ae_bodsys <- freq_table(
  adae,
  colvar      = "TRT01A",
  rowvar      = "AEBODSYS",
  denominator = ___
)

# Render
tlg_output(
  ae_bodsys,
  tbl_id        = "___",
  title         = c("Table 14.3.1", "Adverse Events by Body System"),
  output_format = "rtf",
  output_file   = "T14.3.1.rtf"
)
```

```{r cap-ae-hint-1}
# denominator = adsl_ae_denom  (which is just adsl)
# tbl_id = "T14.3.1"
```

```{r cap-ae-solution}
adsl_ae_denom <- adsl  # adsl already has TRT01A

ae_bodsys <- freq_table(
  adae,
  colvar      = "TRT01A",
  rowvar      = "AEBODSYS",
  denominator = adsl_ae_denom
)

tlg_output(
  ae_bodsys,
  tbl_id        = "T14.3.1",
  title         = c("Table 14.3.1", "Adverse Events by Body System"),
  output_format = "rtf",
  output_file   = "T14.3.1.rtf"
)
# "Table dimensions: N rows" — N = number of distinct body systems in adae
```

### Part 3 — ALT Lab Statistics (T14.4.1)

```{r cap-lab, exercise=TRUE, exercise.lines=16}
# adlb already has TRT01A from setup; filter to ALT parameter
alt_data <- adlb |> dplyr::filter(PARAMCD == "ALT")

alt_stats <- cont_table(
  alt_data,
  colvar = "___",
  rowvar = "AVAL"
)

tlg_output(
  alt_stats,
  tbl_id        = "T14.4.1",
  title         = c("Table 14.4.1", "ALT — Descriptive Statistics"),
  output_format = "rtf",
  output_file   = "T14.4.1.rtf"
)
```

```{r cap-lab-hint-1}
# colvar = "TRT01A"
# adlb already has TRT01A added in the setup chunk
```

```{r cap-lab-solution}
alt_data <- adlb |> dplyr::filter(PARAMCD == "ALT")

alt_stats <- cont_table(
  alt_data,
  colvar = "TRT01A",
  rowvar = "AVAL"
)

tlg_output(
  alt_stats,
  tbl_id        = "T14.4.1",
  title         = c("Table 14.4.1", "ALT — Descriptive Statistics"),
  output_format = "rtf",
  output_file   = "T14.4.1.rtf"
)
# Returns 4 rows: N, Mean (SD), Median, Range
```

### Exercise 8.4 — count AE body systems

From the AE body system frequency table built in Part 2, how many distinct
body systems appear in the result? Store it in a variable called `n_bodsys`.

```{r cap-count-bodsys, exercise=TRUE, exercise.lines=8}
adsl_ae_denom <- adsl  # adsl already has TRT01A

ae_bodsys <- freq_table(
  adae,
  colvar      = "TRT01A",
  rowvar      = "AEBODSYS",
  denominator = adsl_ae_denom
)

n_bodsys <- nrow(ae_bodsys)
n_bodsys
```

```{r cap-count-bodsys-solution}
adsl_ae_denom <- adsl  # adsl already has TRT01A

ae_bodsys <- freq_table(
  adae,
  colvar      = "TRT01A",
  rowvar      = "AEBODSYS",
  denominator = adsl_ae_denom
)

n_bodsys <- nrow(ae_bodsys)
n_bodsys
# Each row = one distinct body system; total reflects CDISCPILOT01 AE data
```

---

## Lesson Summary

### What you learned

| Topic | Key function | Output |
|-------|-------------|--------|
| Categorical frequencies | `freq_table()` | Tidy wide tibble, n (x.x%) cells |
| Continuous descriptives | `cont_table()` | Tidy wide tibble, one row per statistic |
| AE body system / PT | `nested_freq()` | Two-level nested tibble, indented PT labels |
| Stat format control | `statlist()` | Token list passed to compute functions |
| Multi-block tables | `bind_rows()` + `ord` | Standard dplyr operations |
| Rendering | `tlg_output()` | RTF / HTML / PDF with title and metadata |

### The tidytlg pipeline at a glance

```
adsl / adae / adlb
      │
      ▼
freq_table() / cont_table() / nested_freq()
      │  (plain tibble)
      ▼
bind_rows() + arrange(ord)    ← optional multi-block stacking
      │
      ▼
tlg_output(tbl_id, title, output_format, output_file)
      │
      ▼
  submission-ready RTF
```

### Common pitfalls

| Pitfall | Fix |
|---------|-----|
| Denominator from `adae` instead of `adsl` for AE % | Always pass `denominator = adsl` (renamed) |
| Treatment column name mismatch between `adae` and `adsl` | `rename(TRT01A = TRT01P)` in the denominator |
| DTYPE rows in `adlb` inflating statistics | Filter `is.na(DTYPE)` in setup (already done here) |
| Forgetting `tbl_id` in `tlg_output()` | Always specify — needed for define-XML traceability |
| Stacking tables without an `ord` column | Add `mutate(ord = nL