---
title: "Lesson 38 — Shiny Foundations for Clinical Programmers"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 38_shiny_foundations.Rmd
# PURPOSE       : Shiny foundations for clinical programmers: reactive
#                 programming model, UI layout, input/output widgets,
#                 reactive expressions, validation, dynamic UI, modules,
#                 download handlers, clinical ADaM data explorers, and
#                 bslib theming — with real pharmaverseadam examples.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================

library(learnr)
library(shiny)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(pharmaverseadam)
tryCatch(library(bslib), error = function(e) NULL)

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.blanks = TRUE)

# ── ADSL ──────────────────────────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 161, 7)), round(rnorm(n, 174, 8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 68, 12), 1), round(rnorm(n, 82, 14), 1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL = ht, WEIGHTBL = wt,
                                  BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
}
adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    TRT01P = factor(TRT01P, levels = c("Placebo", "Xanomeline Low Dose",
                                        "Xanomeline High Dose"))
  ) |>
  filter(SAFFL == "Y")

# ── ADAE ──────────────────────────────────────────────────────────────────────
adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE ~ NA_integer_
  ))

# ── ADLB ──────────────────────────────────────────────────────────────────────
adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
adlb <- adlb |> select(-any_of("TRT01P"))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb,
                    select(adsl, USUBJID, TRTSDT, TRT01P = TRT01P),
                    by = "USUBJID")
```

## Overview

Welcome to **Lesson 38 — Shiny Foundations for Clinical Programmers**.

This lesson bridges your SAS programming experience to building interactive clinical data applications in R using Shiny. By the end you will be able to:

- Explain the **reactive programming model** and how it differs from SAS batch execution
- Build **UI layouts** using `fluidPage()`, `navbarPage()`, and Bootstrap grid columns
- Wire **input widgets** (dropdowns, sliders, date ranges, action buttons) to server logic
- Write **server functions** with `reactive()`, `observe()`, `observeEvent()`, and render functions
- Apply **validation** patterns with `req()`, `validate()`, and `need()`
- Build **dynamic UI** with `renderUI()` and `updateSelectInput()`
- Handle **file downloads** with `downloadHandler()` and **uploads** with `fileInput()`
- Structure large apps with **Shiny modules** and namespace isolation
- Polish apps with **notifications**, **modal dialogs**, **progress bars**, and `bslib` theming
- Read and extend a complete **clinical data explorer** for CDISCPILOT01

All examples use real `pharmaverseadam` data: `adsl` (254 subjects, 3 treatment arms), `adae`, and `adlb`.

> **SAS analogy**: Shiny is to R what SAS/AF or Stored Process Web Applications were to SAS — but far more accessible, and the standard for clinical exploratory tools in modern pharma.

---

## Section 1 — The Reactive Programming Model (SAS vs Shiny)

### Paradigm shift: batch vs reactive

In SAS, a program runs **top-to-bottom, once**, producing static output. Every change means re-submitting the entire job.

In Shiny, code lives in a **reactive graph**. Think of it as a live spreadsheet:

- **Inputs** are the source cells (user controls: dropdowns, sliders, buttons)
- **Reactive expressions** are intermediate calculated cells — they cache their value and only recalculate when their dependencies change
- **Outputs** are the result cells (tables, plots, text rendered in the browser)

When a user changes an input, Shiny automatically re-runs **only the downstream portions of the graph** that depend on it.

```
[User selects treatment arm]
         |
         v  input$trt_arm
[reactive() filters adsl]     <- cached reactive expression
         |
         +---> [renderTable()  -> output$demog_table]
         +---> [renderPlot()   -> output$age_plot]
```

### The three server objects

Every Shiny server function receives three arguments:

```r
server <- function(input, output, session) { ... }
```

| Object | Role | Access |
|--------|------|--------|
| `input` | Named list of current widget values | Read-only; creates reactive dependency |
| `output` | Named list of render function assignments | Write-only from server |
| `session` | Connection to the user's browser | Used for `update*Input()`, modules, URLs |

### `reactive()` vs `observe()`

| Function | Returns | Runs when | Use for |
|----------|---------|-----------|---------|
| `reactive({...})` | A value (call with `()`) | Called from reactive context, dependencies changed | Shared computed data |
| `observe({...})` | Nothing (side effect) | Dependencies change, eagerly | Logging, updating inputs |
| `observeEvent(ev, {...})` | Nothing | `ev` fires (e.g., button click) | Gating on button press |
| `eventReactive(ev, {...})` | A value | `ev` fires | Value that only updates on click |

### Reactive expressions are lazy and cached

A `reactive()` expression does **not** run when it is defined. It runs the first time it is called from a reactive context, caches the result, and only invalidates (and recalculates) when one of its dependencies changes. This is why putting expensive operations in a shared `reactive()` — rather than repeating them inside each `renderPlot()` and `renderTable()` — is the correct pattern.

### Exercises

**Exercise 1.1** — UI functions return R tag objects. Inspect the classes returned by `fluidPage()` and `selectInput()`:

```{r ex-tag-class, exercise=TRUE, exercise.lines=8}
# What class does fluidPage() return?
cls_page <- class(fluidPage())

# What class does selectInput() return?
cls_input <- class(selectInput("arm", "Treatment Arm",
                               choices = c("Placebo", "Xanomeline Low Dose")))

cat("fluidPage class: ", cls_page,  "\n")
cat("selectInput class:", cls_input, "\n")
```

```{r ex-tag-class-solution}
cls_page  <- class(fluidPage())
cls_input <- class(selectInput("arm", "Treatment Arm",
                               choices = c("Placebo", "Xanomeline Low Dose")))
cat("fluidPage class: ", cls_page,  "\n")
cat("selectInput class:", cls_input, "\n")
```

---

**Exercise 1.2** — `shiny::isolate()` reads a reactive value without creating a dependency. Create a `reactive()` and call it safely outside an app context:

```{r ex-isolate, exercise=TRUE, exercise.lines=10}
# Create a reactive expression
mean_age_r <- reactive({
  mean(adsl$AGE, na.rm = TRUE)
})

# Call it safely outside a reactive context using isolate()
result <- isolate(mean_age_r())
cat("Mean age:", round(result, 1), "\n")
```

```{r ex-isolate-solution}
mean_age_r <- reactive({
  mean(adsl$AGE, na.rm = TRUE)
})
result <- isolate(mean_age_r())
cat("Mean age:", round(result, 1), "\n")
```

---

**Exercise 1.3** — `observe()` runs for side effects. Write an `observe()` that records the subject count, then trigger it once with `isolate()`:

```{r ex-observe, exercise=TRUE, exercise.lines=12}
n_subjects <- 0L

obs <- observe({
  n_subjects <<- nrow(adsl)
  cat("Side effect triggered: n =", n_subjects, "\n")
})

# Trigger the observer once
isolate(obs)
cat("n_subjects after observe:", n_subjects, "\n")
```

```{r ex-observe-solution}
n_subjects <- 0L
obs <- observe({
  n_subjects <<- nrow(adsl)
  cat("Side effect triggered: n =", n_subjects, "\n")
})
isolate(obs)
cat("n_subjects after observe:", n_subjects, "\n")
```

---

**Exercise 1.4** — Reactive chains: create a `reactive()` that filters `adsl` to subjects aged 70 or older, and a second `reactive()` that computes the mean BMI of that filtered subset. Call both with `isolate()`:

```{r ex-reactive-chain-basic, exercise=TRUE, exercise.lines=14}
older_subjects <- reactive({
  adsl |> filter(AGE >= ___)
})

mean_bmi_older <- reactive({
  mean(older_subjects()$BMIBL, na.rm = TRUE)
})

cat("N subjects age >= 70:", isolate(nrow(older_subjects())), "\n")
cat("Mean BMI (age >= 70):", isolate(round(mean_bmi_older(), 1)), "\n")
```

```{r ex-reactive-chain-basic-solution}
older_subjects <- reactive({
  adsl |> filter(AGE >= 70)
})
mean_bmi_older <- reactive({
  mean(older_subjects()$BMIBL, na.rm = TRUE)
})
cat("N subjects age >= 70:", isolate(nrow(older_subjects())), "\n")
cat("Mean BMI (age >= 70):", isolate(round(mean_bmi_older(), 1)), "\n")
```

---

## Section 2 — UI Layout Functions

### Building UI from nested function calls

Shiny UI is constructed by nesting R function calls that generate HTML. The result is a tag tree that Shiny serialises and sends to the browser.

**Core layout hierarchy:**

```r
fluidPage(                         # responsive Bootstrap page
  titlePanel("My App"),            # page title bar
  sidebarLayout(                   # two-column layout
    sidebarPanel(...),             # left panel: filters & controls
    mainPanel(                     # right panel: outputs
      tabsetPanel(                 # in-page tabs
        tabPanel("Tab A", ...),
        tabPanel("Tab B", ...)
      )
    )
  )
)
```

**Multi-tab navigation** (`navbarPage`) is the standard structure for clinical study dashboards:

```r
navbarPage(
  title = "Clinical Study Report",
  tabPanel("Demographics", ...),
  tabPanel("Safety",        ...),
  tabPanel("Lab Data",      ...)
)
```

**Bootstrap 12-column grid** — use `fluidRow()` + `column(width, ...)` for custom layouts. Column widths must sum to 12:

```r
fluidRow(
  column(4, wellPanel(...)),   # 1/3 width
  column(8, plotOutput(...))   # 2/3 width
)
```

**Useful HTML helpers:** `tags$h3()`, `tags$p()`, `tags$br()`, `hr()`, `HTML()`, `wellPanel()`.

### Exercises

**Exercise 2.1** — Complete the sidebarLayout skeleton for a clinical data explorer. Fill in the two blanks:

```{r ex-sidebar-layout, exercise=TRUE, exercise.eval=FALSE, exercise.lines=20}
# Note: exercise.eval=FALSE — copy to RStudio to run the app.
# Here we inspect the tag structure only.

ui <- fluidPage(
  titlePanel("CDISCPILOT01 — Clinical Data Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput("trt_arm", "Treatment Arm",
                  choices = c("All", "Placebo",
                              "Xanomeline Low Dose",
                              "Xanomeline High Dose")),
      # Use sliderInput for the age range (min 50, max 90)
      ___(  "age_range", "Age Range",
            min = 50, max = 90, value = c(50, 90)),
      actionButton("update_btn", "Update", class = "btn-primary")
    ),
    mainPanel(
      # Use tabsetPanel with two tabPanels: "Table" and "Plot"
      ___(
        tabPanel("Table", tableOutput("demog_table")),
        tabPanel("Plot",  plotOutput("age_hist"))
      )
    )
  )
)
class(ui)
```

```{r ex-sidebar-layout-solution}
ui <- fluidPage(
  titlePanel("CDISCPILOT01 — Clinical Data Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput("trt_arm", "Treatment Arm",
                  choices = c("All", "Placebo",
                              "Xanomeline Low Dose",
                              "Xanomeline High Dose")),
      sliderInput("age_range", "Age Range",
                  min = 50, max = 90, value = c(50, 90)),
      actionButton("update_btn", "Update", class = "btn-primary")
    ),
    mainPanel(
      tabsetPanel(
        tabPanel("Table", tableOutput("demog_table")),
        tabPanel("Plot",  plotOutput("age_hist"))
      )
    )
  )
)
class(ui)
```

---

**Exercise 2.2** — Build a `navbarPage` with three clinical tabs. Fill in the tab names:

```{r ex-navbar, exercise=TRUE, exercise.eval=FALSE, exercise.lines=22}
ui <- navbarPage(
  title = "Clinical Study Report",

  tabPanel("Demographics",
    fluidRow(
      column(4, wellPanel(
        selectInput("demog_trt", "Treatment", choices = levels(adsl$TRT01P))
      )),
      column(8, tableOutput("demog_out"))
    )
  ),

  tabPanel("___",         # Safety / AE tab
    plotOutput("ae_plot")
  ),

  tabPanel("___",         # Lab Data tab
    selectInput("param", "Parameter", choices = c("ALT", "AST", "WBC")),
    plotOutput("lab_plot")
  )
)
cat("navbarPage tabs defined\n")
```

```{r ex-navbar-solution}
ui <- navbarPage(
  title = "Clinical Study Report",
  tabPanel("Demographics",
    fluidRow(
      column(4, wellPanel(
        selectInput("demog_trt", "Treatment", choices = levels(adsl$TRT01P))
      )),
      column(8, tableOutput("demog_out"))
    )
  ),
  tabPanel("Safety",
    plotOutput("ae_plot")
  ),
  tabPanel("Lab Data",
    selectInput("param", "Parameter", choices = c("ALT", "AST", "WBC")),
    plotOutput("lab_plot")
  )
)
cat("navbarPage tabs defined\n")
```

---

**Exercise 2.3** — `column()` grid — place two equal-width panels side by side in the 12-column Bootstrap grid:

```{r ex-column-grid, exercise=TRUE, exercise.eval=FALSE, exercise.lines=12}
# Two equal-width columns: each takes half the 12-column grid
row <- fluidRow(
  column(___, plotOutput("plot_left")),
  column(___, plotOutput("plot_right"))
)

# How many Bootstrap columns wide is each panel?
cat("Each panel width:", ___, "of 12\n")
```

```{r ex-column-grid-solution}
row <- fluidRow(
  column(6, plotOutput("plot_left")),
  column(6, plotOutput("plot_right"))
)
cat("Each panel width:", 6, "of 12\n")
```

---

**Exercise 2.4** — `wellPanel()` and `tags$` helpers — build a styled information box:

```{r ex-well-panel, exercise=TRUE, exercise.lines=14}
info_box <- wellPanel(
  tags$h4("Study: CDISCPILOT01"),
  tags$p("Indication: Alzheimer's Disease"),
  tags$p(paste("Total subjects (Safety population):", nrow(adsl))),
  hr(),
  tags$small("Data source: pharmaverseadam")
)

# wellPanel returns a shiny.tag — check its class and name
cat("class:", class(info_box), "\n")
cat("tag name:", info_box$name, "\n")
```

```{r ex-well-panel-solution}
info_box <- wellPanel(
  tags$h4("Study: CDISCPILOT01"),
  tags$p("Indication: Alzheimer's Disease"),
  tags$p(paste("Total subjects (Safety population):", nrow(adsl))),
  hr(),
  tags$small("Data source: pharmaverseadam")
)
cat("class:", class(info_box), "\n")
cat("tag name:", info_box$name, "\n")
```

---

## Section 3 — Input Widgets

### Widget anatomy

Every Shiny input widget follows the same pattern:

```r
widgetFunction(inputId, label, ...)
```

The `inputId` string becomes the key in the `input` list on the server side. If you write `selectInput("trt_arm", ...)` in the UI, the server reads `input$trt_arm`.

> **Rule**: all `inputId` values must be unique within an app.

### Clinically relevant widgets

| Widget | Function | Clinical Use |
|--------|----------|-------------|
| Dropdown | `selectInput()` | Treatment arm, analysis parameter |
| Multi-select | `selectInput(..., multiple=TRUE)` | Multiple SOC / PT filter |
| Checkbox group | `checkboxGroupInput()` | Flag filters (SAFFL, ITTFL) |
| Slider | `sliderInput()` | Age range, visit window |
| Date range | `dateRangeInput()` | Study date window filter |
| Text | `textInput()` | Subject ID search |
| Numeric | `numericInput()` | Threshold or grade value |
| Radio buttons | `radioButtons()` | Plot type selection |
| Action button | `actionButton()` | "Run analysis" / event gate |
| File upload | `fileInput()` | Upload external dataset |
| Download | `downloadButton()` | Export filtered data |

### Exercises

**Exercise 3.1** — Build a complete filter panel using four widgets. Fill in the blanks:

```{r ex-input-panel, exercise=TRUE, exercise.eval=FALSE, exercise.lines=26}
filter_panel <- sidebarPanel(

  # Treatment arm selector
  selectInput(
    inputId  = "___",
    label    = "Treatment Arm",
    choices  = c("All Patients", levels(adsl$TRT01P)),
    selected = "All Patients"
  ),

  # Age range slider: 50 to 90, default full range
  sliderInput("age_range", "Age Range (years)",
              min = ___, max = ___, value = c(50, 90), step = 1),

  # Sex checkbox group — select both by default
  checkboxGroupInput("sex_filter", "Sex",
                     choices  = c("Male" = "M", "Female" = "F"),
                     selected = ___),

  # Action button
  actionButton("run_btn", "Apply Filters",
               icon = icon("filter"), class = "btn-success")
)

cat("sidebarPanel children:", length(filter_panel$children[[1]]), "\n")
```

```{r ex-input-panel-solution}
filter_panel <- sidebarPanel(
  selectInput(
    inputId  = "trt_arm",
    label    = "Treatment Arm",
    choices  = c("All Patients", levels(adsl$TRT01P)),
    selected = "All Patients"
  ),
  sliderInput("age_range", "Age Range (years)",
              min = 50, max = 90, value = c(50, 90), step = 1),
  checkboxGroupInput("sex_filter", "Sex",
                     choices  = c("Male" = "M", "Female" = "F"),
                     selected = c("M", "F")),
  actionButton("run_btn", "Apply Filters",
               icon = icon("filter"), class = "btn-success")
)
cat("sidebarPanel children:", length(filter_panel$children[[1]]), "\n")
```

---

**Exercise 3.2** — `dateRangeInput()` for the CDISCPILOT01 study window:

```{r ex-date-range, exercise=TRUE, exercise.lines=14}
dr <- dateRangeInput(
  inputId   = "study_window",
  label     = "Study Date Window",
  start     = as.Date("___"),    # CDISCPILOT01 start ~ 2012-01-01
  end       = as.Date("___"),    # End ~ 2014-12-31
  format    = "yyyy-mm-dd",
  separator = " to "
)

# The inputId is stored in the attribs list
cat("inputId:", dr$attribs$id, "\n")
```

```{r ex-date-range-solution}
dr <- dateRangeInput(
  inputId   = "study_window",
  label     = "Study Date Window",
  start     = as.Date("2012-01-01"),
  end       = as.Date("2014-12-31"),
  format    = "yyyy-mm-dd",
  separator = " to "
)
cat("inputId:", dr$attribs$id, "\n")
```

---

**Exercise 3.3** — `selectInput` with `multiple = TRUE` for multi-parameter lab selection:

```{r ex-multi-select, exercise=TRUE, exercise.lines=14}
lab_params <- sort(unique(adlb$PARAMCD))

param_input <- selectInput(
  inputId  = "lab_params",
  label    = "Lab Parameters",
  choices  = ___,                    # use the lab_params vector
  selected = c("ALT", "AST"),
  multiple = ___                     # allow multiple selections
)

cat("Total available parameters:", length(lab_params), "\n")
cat("Default selected: ALT, AST\n")
```

```{r ex-multi-select-solution}
lab_params <- sort(unique(adlb$PARAMCD))
param_input <- selectInput(
  inputId  = "lab_params",
  label    = "Lab Parameters",
  choices  = lab_params,
  selected = c("ALT", "AST"),
  multiple = TRUE
)
cat("Total available parameters:", length(lab_params), "\n")
cat("Default selected: ALT, AST\n")
```

---

**Exercise 3.4** — `radioButtons()` for plot type selection. Fill in the default selection:

```{r ex-radio, exercise=TRUE, exercise.lines=14}
plot_type_input <- radioButtons(
  inputId  = "plot_type",
  label    = "Display Type",
  choices  = c(
    "Box Plot"   = "box",
    "Violin Plot" = "violin",
    "Mean +/- SD" = "mean_sd"
  ),
  selected = "___",   # default to box plot
  inline   = TRUE
)

cat("inputId:", plot_type_input$attribs$id, "\n")
```

```{r ex-radio-solution}
plot_type_input <- radioButtons(
  inputId  = "plot_type",
  label    = "Display Type",
  choices  = c(
    "Box Plot"   = "box",
    "Violin Plot" = "violin",
    "Mean +/- SD" = "mean_sd"
  ),
  selected = "box",
  inline   = TRUE
)
cat("inputId:", plot_type_input$attribs$id, "\n")
```

---

## Section 4 — Server Basics and Render Functions

### How the server populates outputs

The server assigns render functions to output slots:

```r
output$my_table <- renderTable({ ... })
output$my_plot  <- renderPlot({ ... })
```

The body of each render function is a **reactive expression**: it automatically tracks any `input$...` values it reads and re-runs when those values change.

### Render function pairing

| Render function | Output function | Use |
|-----------------|----------------|-----|
| `renderTable()` | `tableOutput()` | Simple HTML table |
| `renderDataTable()` | `dataTableOutput()` | Searchable/pageable DT |
| `renderPlot()` | `plotOutput()` | ggplot2 or base R |
| `renderText()` | `textOutput()` | Inline text value |
| `renderPrint()` | `verbatimTextOutput()` | Console/`glimpse()` output |
| `renderUI()` | `uiOutput()` | Dynamically generated UI |
| `downloadHandler()` | `downloadButton()` | File download |

### `shiny::testServer()`

`testServer()` lets you test server logic without launching a browser. It creates a mock session, lets you set input values with `session$setInputs()`, and then read output values. It **does not** start a full Shiny server — it runs fine inside learnr exercises.

```r
testServer(my_server, {
  session$setInputs(param = "ALT")
  cat("Output:", output$my_text, "\n")
})
```

### Exercises

**Exercise 4.1** — Write a server function that filters `adsl` by treatment arm and renders a summary table. Fill in the reactive reference:

```{r ex-server-table, exercise=TRUE, exercise.lines=30}
server <- function(input, output, session) {

  # Reactive: filter adsl by selected treatment arm
  filtered_adsl <- reactive({
    if (input$trt_arm == "All Patients") return(adsl)
    adsl |> filter(as.character(TRT01P) == ___)
  })

  # Render summary table
  output$demog_table <- renderTable({
    filtered_adsl() |>
      summarise(
        N        = n(),
        Age_Mean = round(mean(AGE, na.rm = TRUE), 1),
        Age_SD   = round(sd(AGE,  na.rm = TRUE), 1),
        Pct_F    = round(100 * mean(SEX == "F"), 1),
        .groups  = "drop"
      )
  })
}

# Test with testServer
testServer(server, args = list(), {
  session$setInputs(trt_arm = "Placebo")
  result <- filtered_adsl()
  cat("Placebo N:", nrow(result), "\n")   # should be ~86
})
```

```{r ex-server-table-solution}
server <- function(input, output, session) {
  filtered_adsl <- reactive({
    if (input$trt_arm == "All Patients") return(adsl)
    adsl |> filter(as.character(TRT01P) == input$trt_arm)
  })
  output$demog_table <- renderTable({
    filtered_adsl() |>
      summarise(
        N        = n(),
        Age_Mean = round(mean(AGE, na.rm = TRUE), 1),
        Age_SD   = round(sd(AGE,  na.rm = TRUE), 1),
        Pct_F    = round(100 * mean(SEX == "F"), 1),
        .groups  = "drop"
      )
  })
}
testServer(server, args = list(), {
  session$setInputs(trt_arm = "Placebo")
  result <- filtered_adsl()
  cat("Placebo N:", nrow(result), "\n")
})
```

---

**Exercise 4.2** — Write a `renderPlot()` for an AGE histogram. Fill in the bin width:

```{r ex-render-plot, exercise=TRUE, exercise.lines=22}
server_plot <- function(input, output, session) {

  output$age_hist <- renderPlot({
    df <- adsl |>
      filter(input$trt_arm == "All Patients" |
             as.character(TRT01P) == input$trt_arm)

    ggplot(df, aes(x = AGE, fill = as.character(TRT01P))) +
      geom_histogram(binwidth = ___, colour = "white", alpha = 0.8) +
      labs(title = paste("Age Distribution —", input$trt_arm),
           x = "Age (years)", y = "Count", fill = "Treatment") +
      theme_minimal()
  })
}

testServer(server_plot, {
  session$setInputs(trt_arm = "All Patients")
  output$age_hist   # triggers the render; no error = pass
  cat("renderPlot: OK\n")
})
```

```{r ex-render-plot-solution}
server_plot <- function(input, output, session) {
  output$age_hist <- renderPlot({
    df <- adsl |>
      filter(input$trt_arm == "All Patients" |
             as.character(TRT01P) == input$trt_arm)
    ggplot(df, aes(x = AGE, fill = as.character(TRT01P))) +
      geom_histogram(binwidth = 5, colour = "white", alpha = 0.8) +
      labs(title = paste("Age Distribution —", input$trt_arm),
           x = "Age (years)", y = "Count", fill = "Treatment") +
      theme_minimal()
  })
}
testServer(server_plot, {
  session$setInputs(trt_arm = "All Patients")
  output$age_hist
  cat("renderPlot: OK\n")
})
```

---

**Exercise 4.3** — `renderPrint()` for a `glimpse()` output. Fill in the render function name:

```{r ex-render-print, exercise=TRUE, exercise.lines=14}
server_print <- function(input, output, session) {
  output$data_glimpse <- ___(
    dplyr::glimpse(adsl |> select(USUBJID, TRT01P, AGE, SEX, RACE))
  )
}

testServer(server_print, {
  out <- output$data_glimpse
  cat("renderPrint output class:", class(out), "\n")
})
```

```{r ex-render-print-solution}
server_print <- function(input, output, session) {
  output$data_glimpse <- renderPrint(
    dplyr::glimpse(adsl |> select(USUBJID, TRT01P, AGE, SEX, RACE))
  )
}
testServer(server_print, {
  out <- output$data_glimpse
  cat("renderPrint output class:", class(out), "\n")
})
```

---

**Exercise 4.4** — `renderText()` for a dynamic subject count display:

```{r ex-render-text, exercise=TRUE, exercise.lines=16}
server_text <- function(input, output, session) {

  output$subject_count <- renderText({
    n_total <- nrow(adsl)
    n_arm   <- adsl |>
      filter(as.character(TRT01P) == input$trt_arm) |>
      nrow()
    paste0("Arm: ", input$trt_arm,
           " | N = ", n_arm,
           " of ", n_total, " total")
  })
}

testServer(server_text, {
  session$setInputs(trt_arm = "Placebo")
  cat(output$subject_count, "\n")
})
```

```{r ex-render-text-solution}
server_text <- function(input, output, session) {
  output$subject_count <- renderText({
    n_total <- nrow(adsl)
    n_arm   <- adsl |>
      filter(as.character(TRT01P) == input$trt_arm) |>
      nrow()
    paste0("Arm: ", input$trt_arm,
           " | N = ", n_arm,
           " of ", n_total, " total")
  })
}
testServer(server_text, {
  session$setInputs(trt_arm = "Placebo")
  cat(output$subject_count, "\n")
})
```

---

## Section 5 — Reactive Expressions, observe(), and observeEvent()

### Three reactive consumer types

**`reactive()`** — memoised, returns a value, called like a function:

```r
filtered_data <- reactive({
  adsl |> filter(as.character(TRT01P) == input$arm)
})
# Consumed by outputs:
output$table <- renderTable(filtered_data())
output$plot  <- renderPlot(ggplot(filtered_data(), ...))
```

**`observe()`** — runs for side effects (logging, file writes, `update*Input()` calls); returns nothing:

```r
observe({
  cat("Input changed to:", input$arm, "\n")
})
```

**`observeEvent(eventExpr, handlerExpr)`** — runs `handlerExpr` only when `eventExpr` fires. Essential for gating expensive computations behind a button:

```r
observeEvent(input$run_btn, {
  showNotification("Analysis running...")
})
```

**`eventReactive(eventExpr, valueExpr)`** — like `reactive()` but only recalculates when the event fires:

```r
results <- eventReactive(input$run_btn, {
  adlb |> group_by(TRT01P, PARAMCD) |> summarise(Mean = mean(AVAL, na.rm=TRUE))
})
```

> **Performance rule**: put expensive operations (large merges, model fits) inside `eventReactive()` gated by an `actionButton`. Use plain `reactive()` for lightweight, sub-second filtering.

**`reactiveValues()`** — a mutable list of reactive values, analogous to a named environment that Shiny watches:

```r
rv <- reactiveValues(log = character(0), count = 0L)
observeEvent(input$btn, { rv$count <- rv$count + 1L })
output$count_text <- renderText(rv$count)
```

### Exercises

**Exercise 5.1** — Create a reactive filter chain where a single `reactive()` feeds two outputs. Fill in the reactive reference (both blanks are the same):

```{r ex-reactive-chain, exercise=TRUE, exercise.lines=24}
server <- function(input, output, session) {

  # Step 1: one reactive filter used by both outputs
  filtered <- reactive({
    adsl |> filter(
      AGE >= input$age_min,
      AGE <= input$age_max
    )
  })

  output$n_subjects <- renderText({
    paste("N =", nrow(___()))      # call the reactive
  })

  output$mean_age <- renderText({
    paste("Mean Age =", round(mean(___()$AGE), 1))
  })
}

testServer(server, {
  session$setInputs(age_min = 70, age_max = 90)
  cat("n_subjects:", output$n_subjects, "\n")
  cat("mean_age:  ", output$mean_age,   "\n")
})
```

```{r ex-reactive-chain-solution}
server <- function(input, output, session) {
  filtered <- reactive({
    adsl |> filter(
      AGE >= input$age_min,
      AGE <= input$age_max
    )
  })
  output$n_subjects <- renderText({
    paste("N =", nrow(filtered()))
  })
  output$mean_age <- renderText({
    paste("Mean Age =", round(mean(filtered()$AGE), 1))
  })
}
testServer(server, {
  session$setInputs(age_min = 70, age_max = 90)
  cat("n_subjects:", output$n_subjects, "\n")
  cat("mean_age:  ", output$mean_age,   "\n")
})
```

---

**Exercise 5.2** — `observeEvent()` with `actionButton` — only recompute when the button is clicked. Fill in the event expression:

```{r ex-observe-event, exercise=TRUE, exercise.lines=24}
server <- function(input, output, session) {

  # eventReactive: only fires when run_btn is clicked
  analysis_result <- eventReactive(___, {
    adsl |>
      filter(input$trt_arm == "All" |
             as.character(TRT01P) == input$trt_arm) |>
      summarise(N = n(), Mean_Age = round(mean(AGE, na.rm = TRUE), 1))
  })

  output$result_table <- renderTable(analysis_result())

  # Log each button click as a side effect
  observeEvent(input$run_btn, {
    cat("Button clicked at:", format(Sys.time()), "\n")
  })
}

testServer(server, {
  session$setInputs(trt_arm = "Placebo", run_btn = 0)
  session$setInputs(run_btn = 1)      # simulate button click
  cat("After click, result rows:", nrow(analysis_result()), "\n")
})
```

```{r ex-observe-event-solution}
server <- function(input, output, session) {
  analysis_result <- eventReactive(input$run_btn, {
    adsl |>
      filter(input$trt_arm == "All" |
             as.character(TRT01P) == input$trt_arm) |>
      summarise(N = n(), Mean_Age = round(mean(AGE, na.rm = TRUE), 1))
  })
  output$result_table <- renderTable(analysis_result())
  observeEvent(input$run_btn, {
    cat("Button clicked at:", format(Sys.time()), "\n")
  })
}
testServer(server, {
  session$setInputs(trt_arm = "Placebo", run_btn = 0)
  session$setInputs(run_btn = 1)
  cat("After click, result rows:", nrow(analysis_result()), "\n")
})
```

---

**Exercise 5.3** — `reactiveValues()` — accumulate a log of subject IDs viewed. Fill in the increment value:

```{r ex-reactive-values, exercise=TRUE, exercise.lines=24}
server <- function(input, output, session) {

  rv <- reactiveValues(
    viewed_ids  = character(0),
    click_count = 0L
  )

  observeEvent(input$select_id, {
    rv$viewed_ids  <- c(rv$viewed_ids, input$select_id)
    rv$click_count <- rv$click_count + ___    # increment by 1
  })

  output$log_text <- renderText({
    paste("Clicks:", rv$click_count,
          "| Last ID:", tail(rv$viewed_ids, 1))
  })
}

testServer(server, {
  session$setInputs(select_id = "CDISCPILOT01-01-701-1015")
  session$setInputs(select_id = "CDISCPILOT01-01-701-1023")
  cat("click_count:", rv$click_count, "\n")   # should be 2
  cat("viewed_ids: ", paste(rv$viewed_ids, collapse = ", "), "\n")
})
```

```{r ex-reactive-values-solution}
server <- function(input, output, session) {
  rv <- reactiveValues(
    viewed_ids  = character(0),
    click_count = 0L
  )
  observeEvent(input$select_id, {
    rv$viewed_ids  <- c(rv$viewed_ids, input$select_id)
    rv$click_count <- rv$click_count + 1L
  })
  output$log_text <- renderText({
    paste("Clicks:", rv$click_count,
          "| Last ID:", tail(rv$viewed_ids, 1))
  })
}
testServer(server, {
  session$setInputs(select_id = "CDISCPILOT01-01-701-1015")
  session$setInputs(select_id = "CDISCPILOT01-01-701-1023")
  cat("click_count:", rv$click_count, "\n")
  cat("viewed_ids: ", paste(rv$viewed_ids, collapse = ", "), "\n")
})
```

---

## Section 6 — Validation with req(), validate(), and need()

### Three layers of graceful validation

**`req(x)`** — "require" that `x` is truthy. If not, silently stops the reactive chain and shows a blank output (no error message, no red traceback). This is the first line of every render function:

```r
output$plot <- renderPlot({
  req(input$param)    # nothing selected yet? show blank, not an error
  ...
})
```

**`validate(need(condition, message))`** — shows a friendly user-facing message in the output area when `condition` is `FALSE`. Not an R error — renders as grey text in the output panel:

```r
validate(
  need(nrow(df) > 0,
       paste("No records found for parameter:", input$param))
)
```

**`req()` with multiple arguments** — all must be truthy:

```r
req(input$trt_arm, input$param)   # both must be non-NULL and non-empty
```

**File upload guard:**

```r
req(input$upload_file)            # blocks until file has been uploaded
df <- read.csv(input$upload_file$datapath)
```

### Clinical validation pattern

```r
output$lab_plot <- renderPlot({
  req(input$param)                         # 1. guard: widget populated?
  df <- adlb |> filter(PARAMCD == input$param)
  validate(                                # 2. guard: data exists?
    need(nrow(df) > 0,
         paste("No data for parameter:", input$param))
  )
  ggplot(df, aes(x = AVISIT, y = AVAL, fill = TRT01P)) +
    geom_boxplot() + theme_minimal()
})
```

### Exercises

**Exercise 6.1** — Add `req()` and `validate()` to a render function. Fill in the two blanks:

```{r ex-validation, exercise=TRUE, exercise.lines=22}
server <- function(input, output, session) {

  output$lab_plot <- renderPlot({

    # Require that a parameter has been selected (1st blank)
    ___(input$param)

    df <- adlb |> filter(PARAMCD == input$param)

    # Friendly message if no data (2nd blank = the bad input)
    validate(
      need(nrow(df) > 0,
           paste("No records found for parameter:", ___))
    )

    ggplot(df, aes(x = AVISIT, y = AVAL, fill = TRT01P)) +
      geom_boxplot() + theme_minimal() +
      labs(title = paste("Lab Values —", input$param))
  })
}

testServer(server, {
  session$setInputs(param = "ALT")
  cat("ALT render: OK\n")
})
```

```{r ex-validation-solution}
server <- function(input, output, session) {
  output$lab_plot <- renderPlot({
    req(input$param)
    df <- adlb |> filter(PARAMCD == input$param)
    validate(
      need(nrow(df) > 0,
           paste("No records found for parameter:", input$param))
    )
    ggplot(df, aes(x = AVISIT, y = AVAL, fill = TRT01P)) +
      geom_boxplot() + theme_minimal() +
      labs(title = paste("Lab Values —", input$param))
  })
}
testServer(server, {
  session$setInputs(param = "ALT")
  cat("ALT render: OK\n")
})
```

---

**Exercise 6.2** — `req()` with multiple conditions — both treatment arm and parameter must be selected before rendering:

```{r ex-req-multi, exercise=TRUE, exercise.lines=16}
server <- function(input, output, session) {

  output$filtered_n <- renderText({
    req(input$trt_arm, input$param)   # BOTH must be truthy
    n <- adlb |>
      filter(PARAMCD == input$param,
             as.character(TRT01P) == input$trt_arm) |>
      nrow()
    paste("N records:", ___)
  })
}

testServer(server, {
  session$setInputs(trt_arm = "Placebo", param = "ALT")
  cat(output$filtered_n, "\n")
})
```

```{r ex-req-multi-solution}
server <- function(input, output, session) {
  output$filtered_n <- renderText({
    req(input$trt_arm, input$param)
    n <- adlb |>
      filter(PARAMCD == input$param,
             as.character(TRT01P) == input$trt_arm) |>
      nrow()
    paste("N records:", n)
  })
}
testServer(server, {
  session$setInputs(trt_arm = "Placebo", param = "ALT")
  cat(output$filtered_n, "\n")
})
```

---

**Exercise 6.3** — `validate()` with multiple `need()` calls — check two conditions and report both failures:

```{r ex-validate-multi, exercise=TRUE, exercise.lines=18}
server <- function(input, output, session) {

  output$ae_table <- renderTable({
    df <- adae |>
      filter(as.character(TRT01A) == input$ae_trt | input$ae_trt == "All")

    validate(
      need(nrow(df) > 0,
           "No adverse events found for the selected treatment arm."),
      need("AEBODSYS" %in% names(df),
           "___")   # message if body system column is missing
    )

    df |> count(AEBODSYS, sort = TRUE) |> head(10)
  })
}

testServer(server, {
  session$setInputs(ae_trt = "All")
  cat("AE table rows:", nrow(adae |> count(AEBODSYS) |> head(10)), "\n")
})
```

```{r ex-validate-multi-solution}
server <- function(input, output, session) {
  output$ae_table <- renderTable({
    df <- adae |>
      filter(as.character(TRT01A) == input$ae_trt | input$ae_trt == "All")
    validate(
      need(nrow(df) > 0,
           "No adverse events found for the selected treatment arm."),
      need("AEBODSYS" %in% names(df),
           "Column AEBODSYS not found in the AE dataset.")
    )
    df |> count(AEBODSYS, sort = TRUE) |> head(10)
  })
}
testServer(server, {
  session$setInputs(ae_trt = "All")
  cat("AE table rows:", nrow(adae |> count(AEBODSYS) |> head(10)), "\n")
})
```

---

## Section 7 — Dynamic UI with renderUI() and updateInput()

### Two ways to make UI dynamic

**1. `renderUI()` / `uiOutput()`** — generate entire UI widgets from the server. Use when widget choices depend on filtered data (e.g., the PT list changes as the body system changes):

```r
# UI:
uiOutput("pt_selector")

# Server:
output$pt_selector <- renderUI({
  pt_choices <- adae |> filter(AEBODSYS == input$soc) |>
    pull(AEDECOD) |> unique() |> sort()
  selectInput("pt", "Preferred Term", choices = pt_choices)
})
```

**2. `update*Input()`** — update an existing widget's choices or value without recreating it. More efficient for simple changes:

```r
observeEvent(input$soc, {
  pt_choices <- adae |> filter(AEBODSYS == input$soc) |>
    pull(AEDECOD) |> unique() |> sort()
  updateSelectInput(session, "pt", choices = pt_choices)
})
```

**Cascading filter pattern** (common in clinical apps): SOC dropdown drives PT dropdown which drives the output table.

> **When to use which**: use `renderUI()` when you need to change the *type* of widget or its *structure*. Use `update*Input()` when you only need to change the *choices* or *selected value* of an existing widget.

### Exercises

**Exercise 7.1** — Build a `renderUI()` that generates a PT `selectInput` whose choices derive from filtering `adae` by body system. Fill in the two blanks:

```{r ex-render-ui, exercise=TRUE, exercise.lines=22}
server <- function(input, output, session) {

  output$pt_selector <- ___(   # which render function for dynamic UI?
    {
      req(input$soc)

      pt_choices <- adae |>
        filter(AEBODSYS == input$soc) |>
        pull(AEDECOD) |>
        unique() |> sort()

      selectInput("pt", "Preferred Term",
                  choices  = ___,     # use pt_choices
                  multiple = TRUE)
    }
  )
}

testServer(server, {
  soc_val <- adae$AEBODSYS[1]
  session$setInputs(soc = soc_val)
  ui_out <- output$pt_selector
  cat("renderUI class:", class(ui_out), "\n")
})
```

```{r ex-render-ui-solution}
server <- function(input, output, session) {
  output$pt_selector <- renderUI({
    req(input$soc)
    pt_choices <- adae |>
      filter(AEBODSYS == input$soc) |>
      pull(AEDECOD) |>
      unique() |> sort()
    selectInput("pt", "Preferred Term",
                choices  = pt_choices,
                multiple = TRUE)
  })
}
testServer(server, {
  soc_val <- adae$AEBODSYS[1]
  session$setInputs(soc = soc_val)
  ui_out <- output$pt_selector
  cat("renderUI class:", class(ui_out), "\n")
})
```

---

**Exercise 7.2** — `updateSelectInput()` — update treatment arm choices to only include arms with at least one subject older than a minimum age. Fill in the blank:

```{r ex-update-input, exercise=TRUE, exercise.lines=20}
server <- function(input, output, session) {

  observeEvent(input$min_age, {
    # Arms with at least one subject at or above min_age
    eligible_arms <- adsl |>
      filter(AGE >= input$min_age) |>
      pull(TRT01P) |>
      as.character() |>
      unique() |> sort()

    updateSelectInput(
      session  = ___,       # session object
      inputId  = "trt_arm",
      choices  = eligible_arms,
      selected = eligible_arms[1]
    )
  })
}

testServer(server, {
  session$setInputs(min_age = 75, trt_arm = "Placebo")
  cat("updateSelectInput test: OK\n")
})
```

```{r ex-update-input-solution}
server <- function(input, output, session) {
  observeEvent(input$min_age, {
    eligible_arms <- adsl |>
      filter(AGE >= input$min_age) |>
      pull(TRT01P) |>
      as.character() |>
      unique() |> sort()
    updateSelectInput(
      session  = session,
      inputId  = "trt_arm",
      choices  = eligible_arms,
      selected = eligible_arms[1]
    )
  })
}
testServer(server, {
  session$setInputs(min_age = 75, trt_arm = "Placebo")
  cat("updateSelectInput test: OK\n")
})
```

---

**Exercise 7.3** — Cascading SOC → PT filter using `observeEvent()` and `updateSelectInput()`:

```{r ex-cascading-filter, exercise=TRUE, exercise.lines=22}
server <- function(input, output, session) {

  # When SOC changes, update the PT choices
  observeEvent(input$soc, {
    pt_choices <- if (input$soc == "All") {
      sort(unique(adae$AEDECOD))
    } else {
      adae |>
        filter(AEBODSYS == ___) |>    # filter to selected SOC
        pull(AEDECOD) |>
        unique() |> sort()
    }
    updateSelectInput(session, "pt",
                      choices  = c("All", pt_choices),
                      selected = "All")
  })

  output$ae_n <- renderText({
    req(input$soc, input$pt)
    paste("Filtered AE records: (select SOC and PT)")
  })
}

testServer(server, {
  first_soc <- adae$AEBODSYS[1]
  session$setInputs(soc = first_soc, pt = "All")
  cat("Cascading filter test: OK\n")
})
```

```{r ex-cascading-filter-solution}
server <- function(input, output, session) {
  observeEvent(input$soc, {
    pt_choices <- if (input$soc == "All") {
      sort(unique(adae$AEDECOD))
    } else {
      adae |>
        filter(AEBODSYS == input$soc) |>
        pull(AEDECOD) |>
        unique() |> sort()
    }
    updateSelectInput(session, "pt",
                      choices  = c("All", pt_choices),
                      selected = "All")
  })
  output$ae_n <- renderText({
    req(input$soc, input$pt)
    paste("Filtered AE records: (select SOC and PT)")
  })
}
testServer(server, {
  first_soc <- adae$AEBODSYS[1]
  session$setInputs(soc = first_soc, pt = "All")
  cat("Cascading filter test: OK\n")
})
```

---

## Section 8 — Download Handlers and File Upload

### Exporting data: `downloadHandler()`

```r
# UI
downloadButton("dl_csv", "Export CSV", class = "btn-success")

# Server
output$dl_csv <- downloadHandler(
  filename = function() paste0("ADSL_filtered_", Sys.Date(), ".csv"),
  content  = function(file) {
    write.csv(filtered_adsl(), file, row.names = FALSE)
  }
)
```

Key points:
- `filename` is a **function**, not a string — so it can reference `input$...` to make the name dynamic
- `content` receives a `file` path and **must write to it** (write.csv, write_xlsx, ggsave, etc.)
- Works for any file type: CSV, XLSX, PDF, PNG

### Receiving data: `fileInput()`

```r
# UI
fileInput("upload", "Upload ADSL CSV", accept = ".csv")

# Server
uploaded <- reactive({
  req(input$upload)
  read.csv(input$upload$datapath)
})
```

`input$upload` is a data frame with columns: `name` (original filename), `size` (bytes), `type` (MIME), `datapath` (temp file path on server).

### Exercises

**Exercise 8.1** — Write a complete `downloadHandler()` for filtered adsl as CSV. Fill in the two blanks:

```{r ex-download, exercise=TRUE, exercise.lines=22}
server <- function(input, output, session) {

  filtered <- reactive({
    if (input$trt_arm == "All") return(adsl)
    adsl |> filter(as.character(TRT01P) == input$trt_arm)
  })

  output$download_adsl <- downloadHandler(
    filename = function() {
      paste0("ADSL_", gsub(" ", "_", input$trt_arm),
             "_", ___, ".csv")       # append today's date
    },
    content = function(file) {
      write.csv(___, file,            # write the filtered data
                row.names = FALSE)
    }
  )
}

testServer(server, {
  session$setInputs(trt_arm = "Placebo")
  cat("downloadHandler defined:", !is.null(output$download_adsl), "\n")
})
```

```{r ex-download-solution}
server <- function(input, output, session) {
  filtered <- reactive({
    if (input$trt_arm == "All") return(adsl)
    adsl |> filter(as.character(TRT01P) == input$trt_arm)
  })
  output$download_adsl <- downloadHandler(
    filename = function() {
      paste0("ADSL_", gsub(" ", "_", input$trt_arm),
             "_", Sys.Date(), ".csv")
    },
    content = function(file) {
      write.csv(filtered(), file, row.names = FALSE)
    }
  )
}
testServer(server, {
  session$setInputs(trt_arm = "Placebo")
  cat("downloadHandler defined:", !is.null(output$download_adsl), "\n")
})
```

---

**Exercise 8.2** — Handle an uploaded file with column validation. Fill in the blank:

```{r ex-file-upload, exercise=TRUE, exercise.lines=22}
server <- function(input, output, session) {

  uploaded <- reactive({
    req(input$upload_file)
    df <- read.csv(input$upload_file$datapath, stringsAsFactors = FALSE)

    required_cols <- c("USUBJID", "TRT01P", "AGE", "SEX")
    missing_cols  <- setdiff(required_cols, names(df))

    validate(
      need(length(missing_cols) == 0,
           paste("Missing required columns:",
                 paste(___, collapse = ", ")))   # list the missing columns
    )
    df
  })

  output$upload_summary <- renderPrint({
    dplyr::glimpse(uploaded())
  })
}

cat("File upload server: defined OK\n")
```

```{r ex-file-upload-solution}
server <- function(input, output, session) {
  uploaded <- reactive({
    req(input$upload_file)
    df <- read.csv(input$upload_file$datapath, stringsAsFactors = FALSE)
    required_cols <- c("USUBJID", "TRT01P", "AGE", "SEX")
    missing_cols  <- setdiff(required_cols, names(df))
    validate(
      need(length(missing_cols) == 0,
           paste("Missing required columns:",
                 paste(missing_cols, collapse = ", ")))
    )
    df
  })
  output$upload_summary <- renderPrint({
    dplyr::glimpse(uploaded())
  })
}
cat("File upload server: defined OK\n")
```

---

**Exercise 8.3** — Dynamic filename in `downloadHandler()` — generate a filename that includes the selected parameter and the current date:

```{r ex-download-dynamic, exercise=TRUE, exercise.lines=18}
server <- function(input, output, session) {

  output$dl_lab <- downloadHandler(
    filename = function() {
      # Format: "ADLB_ALT_2026-05-24.csv"
      paste0("ADLB_", ___, "_", Sys.Date(), ".csv")
    },
    content = function(file) {
      df <- adlb |> filter(PARAMCD == input$param)
      write.csv(df, file, row.names = FALSE)
    }
  )
}

testServer(server, {
  session$setInputs(param = "ALT")
  cat("Dynamic filename handler: OK\n")
})
```

```{r ex-download-dynamic-solution}
server <- function(input, output, session) {
  output$dl_lab <- downloadHandler(
    filename = function() {
      paste0("ADLB_", input$param, "_", Sys.Date(), ".csv")
    },
    content = function(file) {
      df <- adlb |> filter(PARAMCD == input$param)
      write.csv(df, file, row.names = FALSE)
    }
  )
}
testServer(server, {
  session$setInputs(param = "ALT")
  cat("Dynamic filename handler: OK\n")
})
```

---

## Section 9 — Shiny Modules

### Why modules?

Large clinical Shiny apps — a demographics tab, a safety tab, a lab tab, a data export tab — grow unwieldy without structure. **Modules** let you package a UI + server pair into a reusable, namespaced unit.

### Module anatomy

A module has two R functions:

```r
# 1. UI function — wrap all inputIds/outputIds in NS(id)
myModuleUI <- function(id) {
  ns <- NS(id)
  tagList(
    selectInput(ns("arm"), "Treatment Arm", choices = levels(adsl$TRT01P)),
    textOutput(ns("n_display"))
  )
}

# 2. Server function — use moduleServer()
myModuleServer <- function(id, data) {
  moduleServer(id, function(input, output, session) {
    # inside here, input$arm refers to ns("arm") automatically
    filtered <- reactive({
      data() |> filter(as.character(TRT01P) == input$arm)
    })
    output$n_display <- renderText(paste("N =", nrow(filtered())))
    return(filtered)   # return reactive to parent app
  })
}
```

**Using the module in the main app:**

```r
# UI:    myModuleUI("mod1")
# Server: filtered_data <- myModuleServer("mod1", reactive(adsl))
```

### Namespace isolation

`NS("demog")` returns a function that prepends `"demog-"` to any string:

```r
ns <- NS("demog")
ns("age_range")   # → "demog-age_range"
```

This means two modules with internal input `"arm"` become `"mod1-arm"` and `"mod2-arm"` — completely isolated.

### Exercises

**Exercise 9.1** — What does `NS()` produce? Inspect the namespaced ID:

```{r ex-ns, exercise=TRUE, exercise.lines=10}
# NS() creates a namespace function for a given module id
ns <- NS("demog")

# Apply the namespace to an inputId
namespaced_id <- ns("age_range")
cat("Namespaced ID:", namespaced_id, "\n")
# Expected: "demog-age_range"

cat("NS function class:", class(ns), "\n")
```

```{r ex-ns-solution}
ns <- NS("demog")
namespaced_id <- ns("age_range")
cat("Namespaced ID:", namespaced_id, "\n")
cat("NS function class:", class(ns), "\n")
```

---

**Exercise 9.2** — Complete a module UI function for a lab parameter selector panel. Fill in the two blanks:

```{r ex-module-ui, exercise=TRUE, exercise.lines=22}
labParamUI <- function(id) {
  ns <- ___(id)       # create the namespace function

  tagList(
    selectInput(
      inputId  = ns(___),     # namespace the "param" inputId
      label    = "Lab Parameter",
      choices  = sort(unique(adlb$PARAMCD)),
      selected = "ALT"
    ),
    sliderInput(
      inputId = ns("visit_range"),
      label   = "Visit number range",
      min = 1, max = 20, value = c(1, 20)
    ),
    tableOutput(ns("param_summary"))
  )
}

# Test: confirm the UI can be created without error
ui_lab <- labParamUI("lab1")
cat("Module UI created: OK\n")
```

```{r ex-module-ui-solution}
labParamUI <- function(id) {
  ns <- NS(id)
  tagList(
    selectInput(
      inputId  = ns("param"),
      label    = "Lab Parameter",
      choices  = sort(unique(adlb$PARAMCD)),
      selected = "ALT"
    ),
    sliderInput(
      inputId = ns("visit_range"),
      label   = "Visit number range",
      min = 1, max = 20, value = c(1, 20)
    ),
    tableOutput(ns("param_summary"))
  )
}
ui_lab <- labParamUI("lab1")
cat("Module UI created: OK\n")
```

---

**Exercise 9.3** — Write a complete module server that returns a filtered reactive for the parent app. Fill in the blank:

```{r ex-module-server, exercise=TRUE, exercise.lines=26}
labParamServer <- function(id, adlb_data) {
  moduleServer(id, function(input, output, session) {

    # Filter lab data to the selected parameter
    param_data <- reactive({
      req(input$param)
      adlb_data() |> filter(PARAMCD == ___)   # use input$param
    })

    output$param_summary <- renderTable({
      param_data() |>
        group_by(TRT01P) |>
        summarise(N    = n(),
                  Mean = round(mean(AVAL, na.rm = TRUE), 1),
                  .groups = "drop")
    })

    return(param_data)   # expose to parent app
  })
}

# Smoke test
testServer(labParamServer, args = list(adlb_data = reactive(adlb)), {
  session$setInputs(param = "ALT")
  df <- param_data()
  cat("ALT rows:", nrow(df), "\n")
})
```

```{r ex-module-server-solution}
labParamServer <- function(id, adlb_data) {
  moduleServer(id, function(input, output, session) {
    param_data <- reactive({
      req(input$param)
      adlb_data() |> filter(PARAMCD == input$param)
    })
    output$param_summary <- renderTable({
      param_data() |>
        group_by(TRT01P) |>
        summarise(N    = n(),
                  Mean = round(mean(AVAL, na.rm = TRUE), 1),
                  .groups = "drop")
    })
    return(param_data)
  })
}
testServer(labParamServer, args = list(adlb_data = reactive(adlb)), {
  session$setInputs(param = "ALT")
  df <- param_data()
  cat("ALT rows:", nrow(df), "\n")
})
```

---

## Section 10 — Clinical App Polish: Theming, Notifications, Modals, Progress

### Production-quality clinical apps need polish

**1. bslib theming** — Bootstrap 5 support via `bslib`:

```r
library(bslib)
ui <- fluidPage(
  theme = bs_theme(version = 5, bootswatch = "flatly",
                   primary = "#0066CC", font_scale = 0.9),
  ...
)
```

Replace `shinythemes::themeSelector()` with `bslib::bs_theme()` for all new apps.

**2. Notifications** — brief banner messages:

```r
showNotification("Data exported successfully", type = "message", duration = 4)
showNotification("No data for selection",      type = "warning")
showNotification("Server error: contact admin", type = "error")
```

**3. Modal dialogs** — used for confirmations, subject drill-down, help text:

```r
showModal(modalDialog(
  title  = "Confirm: Delete Filter",
  "Are you sure you want to reset all filters?",
  footer = tagList(
    modalButton("Cancel"),
    actionButton("confirm_delete", "Yes, reset", class = "btn-danger")
  )
))
removeModal()   # close it programmatically
```

**4. Progress bars** — for long-running computations:

```r
withProgress(message = "Computing summary...", value = 0, {
  incProgress(0.3, detail = "Filtering")
  ...
  incProgress(0.7, detail = "Summarising")
  ...
  incProgress(1.0, detail = "Done")
})
```

**5. `conditionalPanel()`** — show/hide UI server-free, based on JS condition:

```r
conditionalPanel(
  condition = "input.show_advanced == true",
  numericInput("grade_threshold", "Min toxicity grade", value = 2)
)
```

### Exercises

**Exercise 10.1** — Write a `showNotification()` inside `observeEvent()` when the export button is clicked. Fill in the notification type:

```{r ex-notification, exercise=TRUE, exercise.lines=16}
server <- function(input, output, session) {

  observeEvent(input$export_btn, {
    showNotification(
      ui       = paste("Exported", nrow(adsl), "subjects to CSV"),
      type     = "___",    # "message", "warning", or "error"
      duration = 4
    )
  })
}

testServer(server, {
  session$setInputs(export_btn = 1)
  cat("showNotification test: OK\n")
})
```

```{r ex-notification-solution}
server <- function(input, output, session) {
  observeEvent(input$export_btn, {
    showNotification(
      ui       = paste("Exported", nrow(adsl), "subjects to CSV"),
      type     = "message",
      duration = 4
    )
  })
}
testServer(server, {
  session$setInputs(export_btn = 1)
  cat("showNotification test: OK\n")
})
```

---

**Exercise 10.2** — Write a confirmation `modalDialog()` before a destructive operation. Fill in the modal title:

```{r ex-modal, exercise=TRUE, exercise.lines=26}
server <- function(input, output, session) {

  observeEvent(input$clear_filters_btn, {
    showModal(
      modalDialog(
        title     = "___",    # "Confirm: Clear All Filters"
        "This will reset all treatment arm and date filters to defaults.",
        footer    = tagList(
          modalButton("Cancel"),
          actionButton("confirm_clear", "Yes, clear filters",
                       class = "btn-danger")
        ),
        easyClose = TRUE
      )
    )
  })

  observeEvent(input$confirm_clear, {
    updateSelectInput(session, "trt_arm", selected = "All Patients")
    removeModal()
    showNotification("Filters cleared", type = "message")
  })
}

cat("Modal server: defined OK\n")
```

```{r ex-modal-solution}
server <- function(input, output, session) {
  observeEvent(input$clear_filters_btn, {
    showModal(
      modalDialog(
        title     = "Confirm: Clear All Filters",
        "This will reset all treatment arm and date filters to defaults.",
        footer    = tagList(
          modalButton("Cancel"),
          actionButton("confirm_clear", "Yes, clear filters",
                       class = "btn-danger")
        ),
        easyClose = TRUE
      )
    )
  })
  observeEvent(input$confirm_clear, {
    updateSelectInput(session, "trt_arm", selected = "All Patients")
    removeModal()
    showNotification("Filters cleared", type = "message")
  })
}
cat("Modal server: defined OK\n")
```

---

**Exercise 10.3** — `withProgress()` wrapping a multi-step summarise. Fill in the progress message:

```{r ex-progress, exercise=TRUE, exercise.lines=20}
server <- function(input, output, session) {

  output$heavy_table <- renderTable({
    withProgress(
      message = "___",   # "Computing summary..."
      value   = 0,
      {
        incProgress(0.3, detail = "Filtering data")
        df <- adlb |> filter(PARAMCD %in% c("ALT", "AST", "WBC"))

        incProgress(0.6, detail = "Summarising by treatment and parameter")
        result <- df |>
          group_by(PARAMCD, TRT01P) |>
          summarise(N    = n(),
                    Mean = round(mean(AVAL, na.rm = TRUE), 1),
                    .groups = "drop")

        incProgress(1.0, detail = "Done")
        result
      }
    )
  })
}

cat("withProgress server: defined OK\n")
```

```{r ex-progress-solution}
server <- function(input, output, session) {
  output$heavy_table <- renderTable({
    withProgress(
      message = "Computing summary...",
      value   = 0,
      {
        incProgress(0.3, detail = "Filtering data")
        df <- adlb |> filter(PARAMCD %in% c("ALT", "AST", "WBC"))
        incProgress(0.6, detail = "Summarising by treatment and parameter")
        result <- df |>
          group_by(PARAMCD, TRT01P) |>
          summarise(N    = n(),
                    Mean = round(mean(AVAL, na.rm = TRUE), 1),
                    .groups = "drop")
        incProgress(1.0, detail = "Done")
        result
      }
    )
  })
}
cat("withProgress server: defined OK\n")
```

---

## Section 11 — Capstone: Complete Clinical Data Explorer App

The block below is a complete, production-style clinical data explorer for CDISCPILOT01. It demonstrates all patterns from this lesson working together: `navbarPage`, cascading filters, `reactive()`, `renderUI()`, `downloadHandler()`, `validate()`, and `bslib` theming.

> **Note**: This code uses `exercise.eval=FALSE` — it will not execute inside the learnr tutorial. Copy the entire block into a file called `app.R` in RStudio and run it with `shiny::runApp()`.

```{r capstone-full-app, exercise=TRUE, exercise.eval=FALSE, exercise.lines=175}
# ==================================================================
# CDISCPILOT01 Clinical Data Explorer
# Copy this entire block to app.R and run in RStudio:
#   shiny::runApp("app.R")
# ==================================================================
library(shiny)
library(dplyr)
library(ggplot2)
library(pharmaverseadam)
tryCatch(library(bslib), error = function(e) NULL)

# ── Data preparation ───────────────────────────────────────────────
adsl_app <- pharmaverseadam::adsl |>
  mutate(TRT01P = factor(TRT01P, levels = c(
    "Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"
  ))) |>
  filter(SAFFL == "Y")

adae_app <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_app) && !"TRT01A" %in% names(adae_app))
  adae_app <- rename(adae_app, TRT01A = TRTA)

adlb_app <- pharmaverseadam::adlb |>
  filter(is.na(DTYPE)) |>
  select(-any_of("TRT01P")) |>
  left_join(select(adsl_app, USUBJID, TRT01P), by = "USUBJID")
if (is.character(adlb_app$ATOXGR))
  adlb_app <- mutate(adlb_app,
                     ATOXGR = as.integer(ATOXGR),
                     BTOXGR = as.integer(BTOXGR))

# ── UI ─────────────────────────────────────────────────────────────
app_theme <- if (requireNamespace("bslib", quietly = TRUE)) {
  bslib::bs_theme(version = 5, bootswatch = "flatly")
} else {
  NULL
}

ui <- navbarPage(
  title = "CDISCPILOT01 Clinical Explorer",
  theme = app_theme,

  # ── Tab 1: Demographics ─────────────────────────────────────────
  tabPanel("Demographics",
    sidebarLayout(
      sidebarPanel(
        selectInput("demog_trt", "Treatment Arm",
                    choices  = c("All", levels(adsl_app$TRT01P)),
                    selected = "All"),
        sliderInput("demog_age", "Age Range",
                    min = 50, max = 90, value = c(50, 90)),
        checkboxGroupInput("demog_sex", "Sex",
                           choices  = c("Male" = "M", "Female" = "F"),
                           selected = c("M", "F")),
        hr(),
        downloadButton("dl_adsl", "Export ADSL CSV",
                       class = "btn-sm btn-success")
      ),
      mainPanel(
        tabsetPanel(
          tabPanel("Summary Table", tableOutput("demog_table")),
          tabPanel("Age Distribution", plotOutput("demog_age_plot",
                                                   height = "350px"))
        )
      )
    )
  ),

  # ── Tab 2: Safety (AEs) ─────────────────────────────────────────
  tabPanel("Safety",
    sidebarLayout(
      sidebarPanel(
        selectInput("ae_trt", "Treatment Arm",
                    choices  = c("All", sort(unique(adae_app$TRT01A))),
                    selected = "All"),
        selectInput("ae_soc", "Body System (SOC)",
                    choices  = c("All", sort(unique(adae_app$AEBODSYS))),
                    selected = "All"),
        uiOutput("ae_pt_selector"),
        radioButtons("ae_severity", "Minimum Severity",
                     choices  = c("Any"        = "any",
                                  "Moderate+"  = "moderate",
                                  "Severe only" = "severe"),
                     selected = "any")
      ),
      mainPanel(
        tabsetPanel(
          tabPanel("AE Summary",  tableOutput("ae_summary_table")),
          tabPanel("AE Bar Chart", plotOutput("ae_bar_plot", height = "400px"))
        )
      )
    )
  ),

  # ── Tab 3: Lab Data ─────────────────────────────────────────────
  tabPanel("Lab Data",
    sidebarLayout(
      sidebarPanel(
        selectInput("lab_param", "Parameter",
                    choices  = sort(unique(adlb_app$PARAMCD)),
                    selected = "ALT"),
        selectInput("lab_trt", "Treatment Arm",
                    choices  = c("All", levels(adsl_app$TRT01P)),
                    selected = "All"),
        radioButtons("lab_plot_type", "Plot Type",
                     choices  = c("Boxplot"    = "box",
                                  "Violin"     = "violin",
                                  "Mean +/- SE" = "mean_se"),
                     selected = "box")
      ),
      mainPanel(
        plotOutput("lab_plot", height = "400px"),
        verbatimTextOutput("lab_stats")
      )
    )
  )
)

# ── Server ─────────────────────────────────────────────────────────
server <- function(input, output, session) {

  # ── Demographics ──────────────────────────────────────────────
  filtered_adsl <- reactive({
    df <- adsl_app
    if (input$demog_trt != "All")
      df <- df |> filter(as.character(TRT01P) == input$demog_trt)
    df <- df |>
      filter(AGE >= input$demog_age[1],
             AGE <= input$demog_age[2],
             SEX %in% input$demog_sex)
    df
  })

  output$demog_table <- renderTable({
    req(nrow(filtered_adsl()) > 0)
    filtered_adsl() |>
      group_by(TRT01P) |>
      summarise(
        N       = n(),
        Age_Mean = round(mean(AGE, na.rm = TRUE), 1),
        Age_SD   = round(sd(AGE,  na.rm = TRUE), 1),
        Pct_F    = round(100 * mean(SEX == "F"), 1),
        .groups = "drop"
      )
  })

  output$demog_age_plot <- renderPlot({
    req(nrow(filtered_adsl()) > 0)
    ggplot(filtered_adsl(), aes(x = TRT01P, y = AGE, fill = TRT01P)) +
      geom_boxplot(alpha = 0.7) + coord_flip() +
      labs(title = "Age by Treatment Arm", x = "", y = "Age (years)") +
      theme_minimal() + theme(legend.position = "none")
  })

  output$dl_adsl <- downloadHandler(
    filename = function() paste0("ADSL_", Sys.Date(), ".csv"),
    content  = function(f) write.csv(filtered_adsl(), f, row.names = FALSE)
  )

  # ── Safety ────────────────────────────────────────────────────
  output$ae_pt_selector <- renderUI({
    choices <- if (input$ae_soc == "All") {
      sort(unique(adae_app$AEDECOD))
    } else {
      adae_app |> filter(AEBODSYS == input$ae_soc) |>
        pull(AEDECOD) |> unique() |> sort()
    }
    selectInput("ae_pt", "Preferred Term",
                choices = c("All", choices), selected = "All")
  })

  filtered_ae <- reactive({
    df <- adae_app
    if (input$ae_trt != "All")
      df <- df |> filter(TRT01A == input$ae_trt)
    if (input$ae_soc != "All")
      df <- df |> filter(AEBODSYS == input$ae_soc)
    if (input$ae_severity == "moderate")
      df <- df |> filter(AESEV %in% c("MODERATE", "SEVERE"))
    if (input$ae_severity == "severe")
      df <- df |> filter(AESEV == "SEVERE")
    df
  })

  output$ae_summary_table <- renderTable({
    req(nrow(filtered_ae()) > 0)
    filtered_ae() |>
      group_by(AEBODSYS) |>
      summarise(
        N_Events   = n(),
        N_Subjects = n_distinct(USUBJID),
        .groups    = "drop"
      ) |>
      arrange(desc(N_Subjects)) |>
      head(15)
  })

  output$ae_bar_plot <- renderPlot({
    req(nrow(filtered_ae()) > 0)
    top_soc <- filtered_ae() |>
      count(AEBODSYS, sort = TRUE) |>
      head(10) |>
      pull(AEBODSYS)
    filtered_ae() |>
      filter(AEBODSYS %in% top_soc) |>
      count(AEBODSYS, TRT01A) |>
      ggplot(aes(x = reorder(AEBODSYS, n), y = n, fill = TRT01A)) +
      geom_col(position = "dodge") + coord_flip() +
      labs(title = "AE Count by Body System (Top 10)",
           x = "", y = "N Events", fill = "Treatment") +
      theme_minimal()
  })

  # ── Lab Data ──────────────────────────────────────────────────
  filtered_lab <- reactive({
    req(input$lab_param)
    df <- adlb_app |> filter(PARAMCD == input$lab_param)
    if (input$lab_trt != "All")
      df <- df |> filter(as.character(TRT01P) == input$lab_trt)
    validate(need(nrow(df) > 0,
                  paste("No records for parameter:", input$lab_param)))
    df
  })

  output$lab_plot <- renderPlot({
    df <- filtered_lab()
    if (input$lab_plot_type == "box") {
      ggplot(df, aes(x = AVISIT, y = AVAL, fill = TRT01P)) +
        geom_boxplot(alpha = 0.7) + theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = paste(input$lab_param, "— Boxplot"),
             x = "Visit", y = "Value", fill = "Treatment")
    } else if (input$lab_plot_type == "violin") {
      ggplot(df, aes(x = AVISIT, y = AVAL, fill = TRT01P)) +
        geom_violin(alpha = 0.6) + theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = paste(input$lab_param, "— Violin Plot"),
             x = "Visit", y = "Value", fill = "Treatment")
    } else {
      ggplot(df, aes(x = AVISIT, y = AVAL,
                     colour = TRT01P, group = TRT01P)) +
        stat_summary(fun = mean, geom = "line") +
        stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        labs(title = paste(input$lab_param, "— Mean +/- SE"),
             x = "Visit", y = "Value", colour = "Treatment")
    }
  })

  output$lab_stats <- renderPrint({
    filtered_lab() |>
      group_by(TRT01P) |>
      summarise(
        N    = n(),
        Mean = round(mean(AVAL, na.rm = TRUE), 2),
        SD   = round(sd(AVAL,   na.rm = TRUE), 2),
        Min  = round(min(AVAL,  na.rm = TRUE), 2),
        Max  = round(max(AVAL,  na.rm = TRUE), 2),
        .groups = "drop"
      )
  })
}

shinyApp(ui, server)
```

### Capstone exercises

**Exercise 11.1** — From the capstone app server, `filtered_adsl()` is consumed by three output IDs. List all three:

```{r ex-capstone-deps, exercise=TRUE, exercise.lines=8}
# From the capstone app, filtered_adsl() feeds which three output IDs?
dependent_outputs <- c("___", "___", "___")
cat(paste(dependent_outputs, collapse = ", "), "\n")
```

```{r ex-capstone-deps-solution}
dependent_outputs <- c("demog_table", "demog_age_plot", "dl_adsl")
cat(paste(dependent_outputs, collapse = ", "), "\n")
```

---

**Exercise 11.2** — The AE dynamic PT selector uses `renderUI()`. What should the `req()` call guard against?

```{r ex-capstone-req, exercise=TRUE, exercise.lines=12}
# The dynamic PT selector renderUI needs input$ae_soc to be populated.
# Write the req() call that guards it:

output_pt_selector_stub <- function(input, output, session) {
  output$ae_pt_selector <- renderUI({
    req(___)   # what input must be available?
    choices <- sort(unique(adae$AEDECOD))
    selectInput("ae_pt", "Preferred Term", choices = c("All", choices))
  })
}

cat("req() should guard: input$ae_soc\n")
```

```{r ex-capstone-req-solution}
output_pt_selector_stub <- function(input, output, session) {
  output$ae_pt_selector <- renderUI({
    req(input$ae_soc)
    choices <- sort(unique(adae$AEDECOD))
    selectInput("ae_pt", "Preferred Term", choices = c("All", choices))
  })
}
cat("req() should guard: input$ae_soc\n")
```

---

**Exercise 11.3** — The capstone lab plot uses three different plot types controlled by `input$lab_plot_type`. What are the three string values that trigger each branch?

```{r ex-capstone-plot-types, exercise=TRUE, exercise.lines=8}
# The radioButtons in the capstone lab tab uses three choice values.
# What are the three internal values (not the labels)?
plot_type_values <- c("___", "___", "___")
cat(paste(plot_type_values, collapse = ", "), "\n")
```

```{r ex-capstone-plot-types-solution}
plot_type_values <- c("box", "violin", "mean_se")
cat(paste(plot_type_values, collapse = ", "), "\n")
```

---

## Lesson Summary

### Shiny function quick-reference

| Function | Category | Clinical use case |
|----------|----------|-------------------|
| `fluidPage()`, `navbarPage()` | Layout | App shell; multi-tab clinical dashboard |
| `sidebarLayout()` | Layout | Filter panel + output panel |
| `fluidRow()` + `column()` | Layout | Bootstrap grid for custom panel arrangements |
| `selectInput()` | Input | Treatment arm, parameter selection |
| `sliderInput()` | Input | Age range, visit window |
| `checkboxGroupInput()` | Input | Population flag filters (SAFFL, ITTFL) |
| `dateRangeInput()` | Input | Study date window |
| `actionButton()` | Input | Gate expensive computations |
| `downloadButton()` | Input/Output | Export filtered dataset |
| `fileInput()` | Input | Upload external data |
| `reactive()` | Reactivity | Shared filtered dataset used by multiple outputs |
| `observe()` | Reactivity | Logging, `update*Input()` side effects |
| `observeEvent()` | Reactivity | Respond to button clicks |
| `eventReactive()` | Reactivity | Value that only updates on button click |
| `reactiveValues()` | Reactivity | Mutable state (session log, counter) |
| `req()` | Validation | Guard: widget not yet populated |
| `validate()` + `need()` | Validation | Friendly user-facing error message |
| `renderTable()` | Output | Demographics or AE summary table |
| `renderPlot()` | Output | ggplot2 boxplot, histogram |
| `renderText()` | Output | Inline subject count, status message |
| `renderPrint()` | Output | `glimpse()`, `summary()` console output |
| `renderUI()` | Output | Dynamic widget (cascading SOC → PT filter) |
| `downloadHandler()` | Output | CSV/XLSX export |
| `NS()` + `moduleServer()` | Modules | Reusable namespaced UI + server pairs |
| `updateSelectInput()` | Dynamic UI | Update dropdown choices from server |
| `showNotification()` | Polish | Export confirmation, warning banner |
| `showModal()` / `removeModal()` | Polish | Confirmation dialog, subject drill-down |
| `withProgress()` | Polish | Progress bar for long computations |
| `bslib::bs_theme()` | Theming | Bootstrap 5 styling, corporate colours |

### Key rules to remember

1. **`req()` first** — put `req(input$x)` as the first line of every render function
2. **`reactive()` for shared data** — if two outputs need the same filtered data, compute it once in a `reactive()`
3. **`eventReactive()` for expensive ops** — gate model fits and large merges behind an `actionButton`
4. **Module namespacing** — every inputId and outputId inside a module must go through `ns()` in the UI function
5. **`exercise.eval=FALSE` for `shinyApp()`** — full apps cannot run inside learnr; inspect structure in exercises, run in RStudio
6. **`testServer()` works in learnr** — use it to test server logic and reactive expressions without a browser

### Next steps

- Official Shiny documentation and function reference: <https://shiny.posit.co/r/reference/shiny/>
- Mastering Shiny (free online book): <https://mastering-shiny.org/>
- Shiny modules deep-dive: <https://shiny.posit.co/r/articles/improve/modules/>
- bslib theming gallery: <https://rstudio.github.io/bslib/>
- Lesson 39 will cover **teal architecture** — the pharmaverse framework for building standardised clinical study explorers on top of Shiny modules.
