---
title: "Lesson 39 — teal: Architecture & App Structure"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Master the teal framework for building clinical data review applications.
  Understand the app structure, data connectors, filtering infrastructure,
  and how teal modules plug into a unified interactive platform.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 39_teal_architecture.Rmd
# PURPOSE       : teal framework architecture — cdisc_dataset/cdisc_data connectors,
#                 teal_slice filter infrastructure, modules/tab_group
#                 nesting, and deploying study apps on Posit Connect.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================
library(learnr)
library(dplyr)
library(tibble)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.eval = FALSE)

# ── Pharmaverseadam datasets ────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}
adsl <- adsl_raw |> mutate(
  ITTFL  = SAFFL,
  DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"))

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw)) {
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
}
adae <- adae_raw |> mutate(AESEVN = case_when(
  AESEV == "MILD"     ~ 1L, AESEV == "MODERATE" ~ 2L,
  AESEV == "SEVERE"   ~ 3L, TRUE ~ NA_integer_))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

advs <- pharmaverseadam::advs

# ── Mock teal package infrastructure ────────────────────────────────────────
# teal is a large Shiny-based framework; we mock its core API so exercises
# run inside learnr without requiring the full installation.

# teal.data mocks ----
cdisc_dataset <- function(dataname, x, keys = character(0), label = NULL) {
  structure(
    list(dataname = dataname, data = x, keys = keys,
         label = label %||% dataname),
    class = c("cdisc_dataset", "teal_dataset")
  )
}

cdisc_data <- function(..., check = FALSE) {
  datasets <- list(...)
  names(datasets) <- vapply(datasets, function(d) d$dataname, character(1))
  structure(list(datasets = datasets, check = check),
            class = c("cdisc_data", "teal_data"))
}

relational_data <- function(...) {
  datasets <- list(...)
  structure(list(datasets = datasets), class = c("relational_data", "teal_data"))
}

`%||%` <- function(a, b) if (is.null(a)) b else a

# teal module mocks ----
tm_data_table <- function(label = "Data Table", ...) {
  structure(list(label = label, type = "data_table", args = list(...)),
            class = c("tm_data_table", "teal_module"))
}

tm_variable_browser <- function(label = "Variable Browser", ...) {
  structure(list(label = label, type = "variable_browser", args = list(...)),
            class = c("tm_variable_browser", "teal_module"))
}

tm_t_summary <- function(label = "Summary Table",
                         dataname = "ADSL",
                         arm_var = "TRT01P",
                         summarize_vars = NULL, ...) {
  structure(list(label = label, type = "t_summary",
                 dataname = dataname, arm_var = arm_var,
                 summarize_vars = summarize_vars, args = list(...)),
            class = c("tm_t_summary", "teal_module"))
}

tm_g_scatterplot <- function(label = "Scatter Plot",
                              dataname = "ADSL",
                              x = "AGE", y = "HEIGHTBL", ...) {
  structure(list(label = label, type = "scatterplot",
                 dataname = dataname, x = x, y = y, args = list(...)),
            class = c("tm_g_scatterplot", "teal_module"))
}

tm_missing_data <- function(label = "Missing Data", ...) {
  structure(list(label = label, type = "missing_data", args = list(...)),
            class = c("tm_missing_data", "teal_module"))
}

# teal filter mock ----
teal_slice <- function(dataname, varname, selected = NULL, keep_na = FALSE, ...) {
  structure(list(dataname = dataname, varname = varname,
                 selected = selected, keep_na = keep_na, args = list(...)),
            class = "teal_slice")
}

teal_slices <- function(..., include_varnames = NULL, exclude_varnames = NULL) {
  slices <- list(...)
  structure(list(slices = slices,
                 include_varnames = include_varnames,
                 exclude_varnames = exclude_varnames),
            class = "teal_slices")
}

# main teal() function mock ----
teal <- function(data, modules, title = "teal app",
                 filter = NULL, header = NULL, footer = NULL, ...) {
  structure(
    list(data = data, modules = modules, title = title,
         filter = filter, header = header, footer = footer),
    class = c("teal_app", "shiny.appobj")
  )
}

# print method for teal app
print.teal_app <- function(x, ...) {
  cat("── teal Application ──────────────────────────────────\n")
  cat("  Title:   ", x$title, "\n")
  n_ds <- length(x$data$datasets)
  cat("  Datasets:", n_ds, "dataset(s):", paste(names(x$data$datasets), collapse=", "), "\n")
  cat("  Modules: ", length(x$modules), "module(s)\n")
  for (m in x$modules) cat("    •", m$label, "\n")
  if (!is.null(x$filter)) {
    cat("  Filters: ", length(x$filter$slices), "pre-set filter(s)\n")
  }
  invisible(x)
}

# modules() container ----
modules <- function(..., label = "root") {
  mods <- list(...)
  # flatten named list if passed as list(...)
  if (length(mods) == 1 && is.list(mods[[1]]) && !inherits(mods[[1]], "teal_module"))
    mods <- mods[[1]]
  structure(list(modules = mods, label = label), class = "teal_modules")
}

# length method for teal_modules
length.teal_modules <- function(x) length(x$modules)

# tab_group mock (nested tabs)
tab_group <- function(label, ...) {
  children <- list(...)
  structure(list(label = label, children = children), class = c("tab_group", "teal_module"))
}

# describe helpers ----
describe_teal_data <- function(td) {
  cat("teal_data object\n")
  cat("  Datasets:\n")
  for (nm in names(td$datasets)) {
    ds <- td$datasets[[nm]]
    cat(sprintf("    %s: %d rows × %d cols (keys: %s)\n",
                nm,
                nrow(ds$data),
                ncol(ds$data),
                if (length(ds$keys)) paste(ds$keys, collapse=", ") else "none"))
  }
}
```

## Overview

**teal** is the pharmaverse framework for building interactive, filter-driven clinical data review applications. Where Shiny gives you raw building blocks, teal gives you an entire clinical-data application scaffold — datasets wired together, a filter panel, module tabs, and reproducibility — in a fraction of the code.

> **Where does teal fit in the pharmaverse?**
>
> ```
> Raw data (SDTM/ADaM)
>      │
>      ▼
> teal.data ──── teal.code ──── teal.widgets
>      │                              │
>      └─────────── teal ─────────────┘
>                     │
>              teal.modules.general
>              teal.modules.clinical
>              teal.modules.pkpd
>                     │
>                     ▼
>           Interactive Shiny App
> ```

### What you will learn

- The four-layer teal package ecosystem
- `cdisc_data()` and `cdisc_dataset()` — the data connectors
- The `teal()` call and its arguments
- The filtering infrastructure: `teal_slice()` / `teal_slices()`
- How modules register themselves as app tabs
- How to run and share a teal app

---

## 1 — The teal Ecosystem

### 1.1 Four packages, one platform

teal is not a single package — it is a coordinated family:

| Package | Role |
|---------|------|
| **teal.data** | Data connectors: `cdisc_data()`, `cdisc_dataset()`, `relational_data()` |
| **teal.code** | Reproducible code tracking — every filter and module action is logged |
| **teal.widgets** | Reusable UI components (filter panel widget, variable selector, etc.) |
| **teal** | The orchestrator: assembles data + modules + filters into a Shiny app |
| **teal.modules.general** | Ready-made exploratory modules (data tables, scatter plots, missing data) |
| **teal.modules.clinical** | Clinical-specific modules: summary tables, forest plots, Kaplan-Meier |

In practice you load only the packages you need:

```{r load-teal-packages, eval=FALSE}
library(teal)
library(teal.data)
library(teal.modules.general)
library(teal.modules.clinical)
```

### 1.2 The SAS clinical programmer analogy

| SAS world | teal world |
|-----------|-----------|
| `libname ADAM "/path/to/adam";` | `cdisc_data(cdisc_dataset("ADSL", adsl, ...))` |
| Proc Report / proc tabulate program | A teal module |
| Selection macro parameter | A filter slice |
| SAS Enterprise Guide project | A teal app |
| Submitted output file | Downloaded reproducible code from teal |

### 1.3 Quiz

```{r quiz-ecosystem, echo=FALSE}
quiz(
  question(
    "Which teal package is responsible for wiring SDTM/ADaM datasets together with their key relationships?",
    answer("teal.widgets"),
    answer("teal.data", correct = TRUE),
    answer("teal.code"),
    answer("teal.modules.clinical"),
    allow_retry = TRUE
  ),
  question(
    "What does `teal.code` provide that plain Shiny does not?",
    answer("Bootstrap theming"),
    answer("Automatic reproducible code tracking of every user action", correct = TRUE),
    answer("Pre-built clinical summary tables"),
    answer("Dataset key validation"),
    allow_retry = TRUE
  )
)
```

---

## 2 — teal.data: Connecting Datasets

### 2.1 `cdisc_dataset()`

Every dataset in a teal CDISC app is wrapped with `cdisc_dataset()`:

```{r cdisc-dataset-demo, eval=FALSE}
adsl_ds <- cdisc_dataset(
  dataname = "ADSL",        # name referenced by modules
  x        = adsl,          # the actual data frame
  keys     = c("STUDYID", "USUBJID"),   # primary key columns
  label    = "Subject-Level Analysis Dataset"
)
```

The `keys` argument matters enormously — teal uses them to:

1. Join datasets correctly in the filter panel
2. Display the correct subject count after filtering
3. Validate that merge operations preserve subject integrity

**Key rules for common ADaM datasets:**

| Dataset | Keys |
|---------|------|
| ADSL | `c("STUDYID", "USUBJID")` |
| ADAE | `c("STUDYID", "USUBJID", "ASTDTM", "AETERM", "AESEQ")` |
| ADLB | `c("STUDYID", "USUBJID", "PARAMCD", "AVISIT", "ADT")` |
| ADVS | `c("STUDYID", "USUBJID", "PARAMCD", "AVISIT", "ADT")` |
| ADTTE | `c("STUDYID", "USUBJID", "PARAMCD")` |

### 2.2 Exercise — create a dataset connector

Create a `cdisc_dataset` for `adae` with appropriate keys.

```{r cdisc-dataset-ex, exercise=TRUE, exercise.lines=8}
adae_ds <- cdisc_dataset(
  dataname = ___,
  x        = ___,
  keys     = c("STUDYID", "USUBJID", "AESEQ"),
  label    = "Adverse Events Analysis Dataset"
)

# Inspect the result
str(adae_ds)
```

```{r cdisc-dataset-ex-solution}
adae_ds <- cdisc_dataset(
  dataname = "ADAE",
  x        = adae,
  keys     = c("STUDYID", "USUBJID", "AESEQ"),
  label    = "Adverse Events Analysis Dataset"
)

str(adae_ds)
```

### 2.3 `cdisc_data()` — the data container

`cdisc_data()` collects all datasets and declares how they are joined:

```{r cdisc-data-demo, eval=FALSE}
app_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID")),
  cdisc_dataset("ADAE", adae, keys = c("STUDYID", "USUBJID", "AESEQ")),
  cdisc_dataset("ADLB", adlb, keys = c("STUDYID", "USUBJID", "PARAMCD", "ADT")),
  check = TRUE   # validate key uniqueness at startup
)
```

When `check = TRUE`, teal validates that:

- All declared keys uniquely identify rows in each dataset
- Merge paths exist between related datasets
- Variable types are consistent

### 2.4 Exercise — build a multi-dataset container

Build a `cdisc_data` container with ADSL, ADAE, and ADLB. Print it to see the summary.

```{r cdisc-data-ex, exercise=TRUE, exercise.lines=16}
app_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl,
                keys  = c("STUDYID", "USUBJID")),
  cdisc_dataset("ADAE", ___,
                keys  = c("STUDYID", "USUBJID", "AESEQ")),
  cdisc_dataset("ADLB", ___,
                keys  = c("STUDYID", "USUBJID", ___, "ADT")),
  check = FALSE
)

describe_teal_data(app_data)
```

```{r cdisc-data-ex-solution}
app_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl,
                keys  = c("STUDYID", "USUBJID")),
  cdisc_dataset("ADAE", adae,
                keys  = c("STUDYID", "USUBJID", "AESEQ")),
  cdisc_dataset("ADLB", adlb,
                keys  = c("STUDYID", "USUBJID", "PARAMCD", "ADT")),
  check = FALSE
)

describe_teal_data(app_data)
```

---

## 3 — The `teal()` Call

### 3.1 Anatomy of `teal()`

```{r teal-anatomy, eval=FALSE}
app <- teal(
  data    = app_data,          # cdisc_data() object
  modules = modules(           # list of teal modules
    tm_data_table("Data Tables"),
    tm_variable_browser("Variable Browser"),
    tm_t_summary("Demographics Summary"),
    tm_g_scatterplot("Scatter Plots")
  ),
  title  = "CDISCPILOT01 Clinical Review",
  filter = teal_slices(        # pre-set filters (optional)
    teal_slice("ADSL", "SAFFL", selected = "Y")
  ),
  header = tags$h3("CDISCPILOT01 — Xanomeline Study"),
  footer = tags$p("Data cut: 2026-01-01 | Internal use only")
)
```

### 3.2 Argument reference

| Argument | Type | Purpose |
|----------|------|---------|
| `data` | `cdisc_data` | Datasets and their relationships |
| `modules` | `modules(...)` | List of tab modules |
| `title` | character | Browser tab / window title |
| `filter` | `teal_slices` | Pre-populated filter panel entries |
| `header` | `shiny.tag` | Custom header HTML shown above tabs |
| `footer` | `shiny.tag` | Custom footer HTML |

### 3.3 Exercise — build and inspect a teal app

Build a minimal teal app with ADSL data and two modules. Use `print()` to inspect it.

```{r teal-build-ex, exercise=TRUE, exercise.lines=20}
my_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID"))
)

my_app <- teal(
  data    = ___,
  modules = modules(
    tm_data_table("Data Table"),
    ___("Variable Browser")
  ),
  title   = "My First teal App"
)

print(my_app)
```

```{r teal-build-ex-solution}
my_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID"))
)

my_app <- teal(
  data    = my_data,
  modules = modules(
    tm_data_table("Data Table"),
    tm_variable_browser("Variable Browser")
  ),
  title   = "My First teal App"
)

print(my_app)
```

---

## 4 — The Filtering Infrastructure

### 4.1 Why teal filters are special

In plain Shiny you write `filter()` logic yourself. In teal, the filter panel is:

1. **Automatically generated** from all dataset variables
2. **Cross-dataset aware** — filtering ADSL by SAFFL="Y" automatically propagates to ADAE and ADLB via USUBJID
3. **Reproducible** — every filter state generates R code that can be downloaded
4. **Restorable** — bookmarks save filter state

### 4.2 `teal_slice()` — a single filter condition

```{r teal-slice-demo, eval=FALSE}
# Simple categorical filter
saffl_slice <- teal_slice(
  dataname = "ADSL",
  varname  = "SAFFL",
  selected = "Y"          # pre-select this level
)

# Numeric range filter
age_slice <- teal_slice(
  dataname = "ADSL",
  varname  = "AGE",
  selected = c(60, 85)    # pre-set range [60, 85]
)

# With NA handling
tox_slice <- teal_slice(
  dataname = "ADLB",
  varname  = "ATOXGR",
  selected = c(1L, 2L, 3L, 4L),
  keep_na  = FALSE          # exclude NA grades
)
```

### 4.3 `teal_slices()` — collection of filters

```{r teal-slices-demo, eval=FALSE}
pre_filters <- teal_slices(
  teal_slice("ADSL", "SAFFL",  selected = "Y"),
  teal_slice("ADSL", "AGE",    selected = c(50, 85)),
  teal_slice("ADLB", "PARAMCD", selected = c("ALT", "CREAT", "WBC")),
  # Lock certain variables out of the filter panel entirely:
  exclude_varnames = list(
    ADSL = c("STUDYID", "USUBJID", "SUBJID"),
    ADLB = c("STUDYID", "USUBJID")
  )
)
```

### 4.4 Filter propagation diagram

```
ADSL filter (SAFFL == "Y")
        │
        │  propagates via USUBJID
        ├──────────────┬──────────────┐
        ▼              ▼              ▼
  ADAE filtered  ADLB filtered  ADVS filtered
  (only SAF pts) (only SAF pts) (only SAF pts)
```

All modules automatically receive the filtered data — the programmer writes no `filter()` calls inside modules.

### 4.5 Exercise — define pre-set filters

Define a `teal_slices` object that:
- Pre-selects `SAFFL == "Y"` from ADSL
- Pre-selects PARAMCD values `"ALT"` and `"CREAT"` from ADLB
- Excludes `STUDYID` and `USUBJID` from the filter panel for both datasets

```{r teal-slices-ex, exercise=TRUE, exercise.lines=18}
clinical_filters <- teal_slices(
  teal_slice(___, "SAFFL",   selected = ___),
  teal_slice("ADLB", ___,    selected = c("ALT", ___)),
  exclude_varnames = list(
    ADSL = c(___, "USUBJID"),
    ADLB = c("STUDYID", ___)
  )
)

# inspect
str(clinical_filters)
```

```{r teal-slices-ex-solution}
clinical_filters <- teal_slices(
  teal_slice("ADSL", "SAFFL",   selected = "Y"),
  teal_slice("ADLB", "PARAMCD", selected = c("ALT", "CREAT")),
  exclude_varnames = list(
    ADSL = c("STUDYID", "USUBJID"),
    ADLB = c("STUDYID", "USUBJID")
  )
)

str(clinical_filters)
```

### 4.6 Quiz

```{r quiz-filter, echo=FALSE}
quiz(
  question(
    "When you filter ADSL to SAFFL='Y' in the teal filter panel, what happens to ADAE?",
    answer("Nothing — ADAE is filtered separately"),
    answer("ADAE is automatically filtered to the same subjects via the USUBJID key", correct = TRUE),
    answer("ADAE is removed from the app"),
    answer("You must write a join manually to propagate the filter"),
    allow_retry = TRUE
  ),
  question(
    "What does `teal.code` do automatically when a user applies a filter in teal?",
    answer("Sends the filter to a remote database"),
    answer("Records reproducible R code for every filter and module action", correct = TRUE),
    answer("Emails the filtered dataset to the statistician"),
    answer("Validates the filter against ICH E9 guidelines"),
    allow_retry = TRUE
  )
)
```

---

## 5 — Modules and the `modules()` Container

### 5.1 What is a teal module?

A **teal module** is a self-contained Shiny module packaged as a tab in the teal app. It receives:

- The filtered data (automatically, via the filter panel)
- A namespace (from the module's label/id)
- Access to teal's reproducibility machinery

From the module's perspective, data always arrives pre-filtered — no `req()` or filter logic needed.

### 5.2 Module families

**teal.modules.general** (exploratory, any dataset):

| Function | Purpose |
|----------|---------|
| `tm_data_table()` | Browse raw rows of any dataset |
| `tm_variable_browser()` | Distribution plot for every variable |
| `tm_missing_data()` | Missingness heatmap and summary |
| `tm_g_scatterplot()` | X-Y scatter with optional colour/size |
| `tm_g_lineplot()` | Longitudinal line plots |
| `tm_g_pp_*()` | Patient profile modules |

**teal.modules.clinical** (clinical reporting):

| Function | Purpose |
|----------|---------|
| `tm_t_summary()` | Demographic summary tables (like Table 14.1.1) |
| `tm_t_events()` | AE frequency tables by SOC/PT |
| `tm_t_shift_by_grade()` | Lab toxicity shift tables |
| `tm_g_km()` | Kaplan-Meier survival curves |
| `tm_t_exposure()` | Exposure (treatment duration) tables |
| `tm_t_tte()` | Time-to-event analysis tables |

### 5.3 Nested modules with `tab_group()`

For large apps, group modules under labelled tab parents:

```{r nested-modules-demo, eval=FALSE}
app <- teal(
  data    = app_data,
  modules = modules(
    tab_group("Data Explorer",
      tm_data_table("Raw Data"),
      tm_variable_browser("Variable Distributions"),
      tm_missing_data("Missingness Report")
    ),
    tab_group("Safety",
      tm_t_summary("Demographics"),
      tm_t_events("Adverse Events"),
      tm_t_shift_by_grade("Lab Toxicity Shifts")
    ),
    tab_group("Efficacy",
      tm_g_km("Kaplan-Meier"),
      tm_t_tte("Time-to-Event Table")
    )
  ),
  title = "Full Clinical Review Application"
)
```

### 5.4 Exercise — assemble a three-group app

Build a teal app with three tab groups (Explorer, Safety, Labs) using the module stubs defined in the setup chunk. Print it.

```{r assemble-app-ex, exercise=TRUE, exercise.lines=30}
full_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID")),
  cdisc_dataset("ADAE", adae, keys = c("STUDYID", "USUBJID", "AESEQ")),
  cdisc_dataset("ADLB", adlb, keys = c("STUDYID", "USUBJID", "PARAMCD", "ADT"))
)

full_app <- teal(
  data  = full_data,
  modules = modules(
    tab_group("Explorer",
      tm_data_table("Raw Data"),
      ___("Variable Distributions")
    ),
    tab_group("Safety",
      tm_t_summary("Demographics", dataname = "ADSL"),
      ___("Adverse Events")
    ),
    tab_group("Labs",
      tm_t_summary("Lab Summary", dataname = "ADLB"),
      tm_g_scatterplot("Lab Scatter", dataname = "ADLB",
                       x = "ADY", y = "AVAL")
    )
  ),
  title  = "CDISCPILOT01 Clinical Review",
  filter = teal_slices(
    teal_slice("ADSL", "SAFFL", selected = "Y")
  )
)

print(full_app)
```

```{r assemble-app-ex-solution}
full_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID")),
  cdisc_dataset("ADAE", adae, keys = c("STUDYID", "USUBJID", "AESEQ")),
  cdisc_dataset("ADLB", adlb, keys = c("STUDYID", "USUBJID", "PARAMCD", "ADT"))
)

full_app <- teal(
  data  = full_data,
  modules = modules(
    tab_group("Explorer",
      tm_data_table("Raw Data"),
      tm_variable_browser("Variable Distributions")
    ),
    tab_group("Safety",
      tm_t_summary("Demographics", dataname = "ADSL"),
      tm_missing_data("Adverse Events")
    ),
    tab_group("Labs",
      tm_t_summary("Lab Summary", dataname = "ADLB"),
      tm_g_scatterplot("Lab Scatter", dataname = "ADLB",
                       x = "ADY", y = "AVAL")
    )
  ),
  title  = "CDISCPILOT01 Clinical Review",
  filter = teal_slices(
    teal_slice("ADSL", "SAFFL", selected = "Y")
  )
)

print(full_app)
```

---

## 6 — Running, Deploying, and Sharing

### 6.1 Running locally

```{r run-app, eval=FALSE}
# Launch in browser
shinyApp(ui = app$ui, server = app$server)

# OR (preferred shorthand with recent teal):
runApp(app)
```

> **Important:** `teal()` returns a Shiny app object. You can pass it directly to `shinyApp()`, wrap it in an `app.R` file, or use `runApp()`.

### 6.2 A complete `app.R` pattern

```{r app-r-pattern, eval=FALSE}
# app.R — drop this in its own folder and deploy to Posit Connect
library(shiny)
library(teal)
library(teal.data)
library(teal.modules.general)
library(teal.modules.clinical)
library(pharmaverseadam)

# ── Data ──────────────────────────────────────────────────────
adsl <- pharmaverseadam::adsl
adae <- pharmaverseadam::adae
adlb <- pharmaverseadam::adlb |> filter(is.na(DTYPE))

app_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID")),
  cdisc_dataset("ADAE", adae, keys = c("STUDYID", "USUBJID", "AESEQ")),
  cdisc_dataset("ADLB", adlb, keys = c("STUDYID", "USUBJID", "PARAMCD", "ADT"))
)

# ── Modules ───────────────────────────────────────────────────
app_modules <- modules(
  tm_data_table("Data Tables"),
  tm_variable_browser("Variable Browser"),
  tab_group("Safety",
    tm_t_summary("Demographics", dataname = "ADSL"),
    tm_t_events("Adverse Events", dataname = "ADAE",
                arm_var = "TRT01A", llt_var = "AEDECOD",
                hlt_var = "AESOC")
  )
)

# ── Filters ───────────────────────────────────────────────────
app_filters <- teal_slices(
  teal_slice("ADSL", "SAFFL", selected = "Y"),
  exclude_varnames = list(ADSL = c("STUDYID","USUBJID","SUBJID"))
)

# ── Launch ────────────────────────────────────────────────────
app <- teal(
  data    = app_data,
  modules = app_modules,
  filter  = app_filters,
  title   = "CDISCPILOT01 Clinical Review"
)

shinyApp(ui = app$ui, server = app$server)
```

### 6.3 Deploying to Posit Connect

```{r deploy-connect, eval=FALSE}
# One-time publish from RStudio
rsconnect::deployApp(
  appDir      = "path/to/app/",
  appName     = "cdiscpilot01-review",
  server      = "connect.company.com",
  account     = "htavakoli",
  forceUpdate = TRUE
)

# Or via the Publish button in RStudio → Posit Connect
```

**renv considerations for deployment:**

```{r renv-deploy, eval=FALSE}
# Ensure renv.lock captures all teal dependencies
renv::snapshot()

# teal family packages to include in renv.lock:
# teal, teal.data, teal.code, teal.widgets, teal.slice
# teal.modules.general, teal.modules.clinical
# teal.reporter (for PDF report generation within app)
```

### 6.4 The reproducibility download

One of teal's most powerful features for regulated environments: **every user session generates downloadable R code**. A reviewer can click "Show R Code" and get a complete, standalone script that:

1. Loads the same datasets
2. Applies every filter the user set
3. Runs the analysis shown in the active module

This satisfies GxP audit trail requirements without extra effort.

### 6.5 Exercise — complete the app.R structure

Fill in the missing pieces of a production `app.R` pattern:

```{r app-r-ex, exercise=TRUE, exercise.lines=35}
# Imagine we have adsl, adae, advs loaded in the workspace

clinical_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", ___)),
  cdisc_dataset("ADAE", adae, keys = c("STUDYID", "USUBJID", ___)),
  cdisc_dataset("ADVS", advs, keys = c("STUDYID", "USUBJID", ___, "ADT"))
)

app_modules <- modules(
  ___("Data Tables"),
  tab_group("Safety",
    tm_t_summary("Demographics", dataname = ___),
    tm_missing_data("Missingness")
  ),
  tab_group("Vitals",
    tm_g_scatterplot("VS Scatter", dataname = "ADVS",
                     x = "ADY", y = "AVAL")
  )
)

app_filters <- teal_slices(
  teal_slice("ADSL", "SAFFL", selected = ___)
)

my_app <- teal(
  data    = clinical_data,
  modules = ___,
  filter  = app_filters,
  title   = "Clinical Data Review"
)

print(my_app)
```

```{r app-r-ex-solution}
clinical_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID")),
  cdisc_dataset("ADAE", adae, keys = c("STUDYID", "USUBJID", "AESEQ")),
  cdisc_dataset("ADVS", advs, keys = c("STUDYID", "USUBJID", "PARAMCD", "ADT"))
)

app_modules <- modules(
  tm_data_table("Data Tables"),
  tab_group("Safety",
    tm_t_summary("Demographics", dataname = "ADSL"),
    tm_missing_data("Missingness")
  ),
  tab_group("Vitals",
    tm_g_scatterplot("VS Scatter", dataname = "ADVS",
                     x = "ADY", y = "AVAL")
  )
)

app_filters <- teal_slices(
  teal_slice("ADSL", "SAFFL", selected = "Y")
)

my_app <- teal(
  data    = clinical_data,
  modules = app_modules,
  filter  = app_filters,
  title   = "Clinical Data Review"
)

print(my_app)
```

---

## 7 — Architecture Deep Dive

### 7.1 What happens when teal starts up

```
1. teal() is called
        │
        ▼
2. cdisc_data validated
   └── key uniqueness checked (if check=TRUE)
   └── join graph inferred from shared keys
        │
        ▼
3. Shiny app created
   ├── Filter panel UI auto-generated from all dataset variables
   ├── Tabs created from modules()
   └── Bookmark/restore infrastructure wired up
        │
        ▼
4. User opens app
   ├── Pre-set teal_slices applied to filter panel
   ├── Each module tab initialises with filtered data
   └── teal.code begins recording reproducible session log
```

### 7.2 Data flow inside the app

```
User adjusts filter panel
        │
        ▼
teal.slice reactive triggers
        │
        ▼
Filtered reactive data re-computed for each dataset
        │
        ▼
All module servers receive updated data via reactive()
        │
        ▼
Outputs re-render automatically (Shiny reactive graph)
        │
        ▼
Reproducible code log updated
```

This is pure Shiny reactive programming — teal simply sets up the reactive plumbing automatically so module authors don't have to.

### 7.3 Quiz

```{r quiz-arch, echo=FALSE}
quiz(
  question(
    "In a teal app, when does a module author need to write filter() calls to subset data?",
    answer("In every module — they must filter explicitly"),
    answer("Only for ADLB — other datasets are filtered automatically"),
    answer("Never — teal's filter panel delivers pre-filtered data to each module", correct = TRUE),
    answer("Only when using teal.modules.general"),
    allow_retry = TRUE
  ),
  question(
    "What is the primary benefit of teal's reproducibility feature in a regulated (GxP) environment?",
    answer("It speeds up rendering by caching results"),
    answer("It generates a downloadable R script of every filter and analysis action, supporting audit trails", correct = TRUE),
    answer("It automatically validates analysis results against CDISC standards"),
    answer("It enables offline use without R installed"),
    allow_retry = TRUE
  ),
  question(
    "What does teal use the 'keys' argument in cdisc_dataset() for?",
    answer("Only for display purposes in the UI"),
    answer("To join datasets and propagate filter selections cross-dataset, and to validate row uniqueness", correct = TRUE),
    answer("To encrypt the dataset before transmission"),
    answer("To determine the column sort order in data tables"),
    allow_retry = TRUE
  )
)
```

---

## 8 — Comparing teal to Plain Shiny

### 8.1 Side-by-side: demographics tab

**Plain Shiny approach (you write everything):**

```{r plain-shiny-demo, eval=FALSE}
# Plain Shiny — ~80 lines for one demographics table
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      selectInput("saffl", "Safety Population", choices = c("Y","N","All"),
                  selected = "Y"),
      selectInput("arm", "Treatment Arm",
                  choices = c("All", unique(adsl$TRT01P)), selected = "All")
    ),
    mainPanel(tableOutput("demo_table"))
  )
)

server <- function(input, output, session) {
  filtered_data <- reactive({
    d <- adsl
    if (input$saffl != "All") d <- filter(d, SAFFL == input$saffl)
    if (input$arm   != "All") d <- filter(d, TRT01P == input$arm)
    d
  })
  output$demo_table <- renderTable({
    filtered_data() |>
      group_by(TRT01P) |>
      summarise(N = n(),
                `Mean Age` = round(mean(AGE, na.rm=TRUE), 1),
                `Female %` = round(mean(SEX=="F")*100, 1))
  })
}

shinyApp(ui, server)
```

**teal approach (framework handles the plumbing):**

```{r teal-demo-approach, eval=FALSE}
# teal — ~15 lines for a full demographics tab with filter panel
app <- teal(
  data    = cdisc_data(
    cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID"))
  ),
  modules = modules(
    tm_t_summary("Demographics",
                 dataname      = "ADSL",
                 arm_var       = "TRT01P",
                 summarize_vars = c("AGE", "SEX", "RACE", "BMIBL"))
  ),
  filter  = teal_slices(teal_slice("ADSL", "SAFFL", selected = "Y"))
)
```

**What teal gives you for free:**
- Filter panel with all ADSL variables
- Subject count updates as you filter
- Downloadable reproducible R code
- Bookmarkable URL with filter state
- Consistent UX across all tabs

### 8.2 Exercise — evaluate the tradeoff

```{r quiz-tradeoff, echo=FALSE}
quiz(
  question(
    "A clinical programmer needs to build a one-off exploratory scatter plot for a single analysis. Which approach is more appropriate?",
    answer("teal — always use teal for clinical data"),
    answer("Plain Shiny — teal's overhead is justified only for multi-module apps with shared filters", correct = TRUE),
    answer("Neither — use SAS ODS Graphics instead"),
    answer("Python Dash — it is faster than both"),
    allow_retry = TRUE
  ),
  question(
    "A clinical data review platform needs to serve 50 analysts reviewing AE, lab, and efficacy data with consistent subject-level filtering. Which approach is more appropriate?",
    answer("Plain Shiny — more flexible"),
    answer("teal — the cross-dataset filter propagation, reproducibility, and pre-built clinical modules justify the investment", correct = TRUE),
    answer("A static PDF report"),
    answer("Excel pivot tables"),
    allow_retry = TRUE
  )
)
```

---

## 9 — Capstone: Wire a Complete Study App

Assemble a full CDISCPILOT01 teal app with:

1. Four datasets: ADSL, ADAE, ADLB, ADVS
2. Three tab groups: Explorer, Safety, Lab & Vitals
3. Pre-set filters: SAFFL="Y" on ADSL; ALT and CREAT on ADLB
4. A custom title and exclude STUDYID/USUBJID from filter panel

```{r capstone-app, exercise=TRUE, exercise.eval=FALSE, exercise.lines=50}
# Step 1: Build the data container
pilot_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl,
                keys  = c("STUDYID", "USUBJID"),
                label = "Subject-Level Analysis Dataset"),
  cdisc_dataset("ADAE", ___,
                keys  = c("STUDYID", "USUBJID", "AESEQ"),
                label = "Adverse Events"),
  cdisc_dataset("ADLB", ___,
                keys  = c("STUDYID", "USUBJID", "PARAMCD", "ADT"),
                label = "Lab Data"),
  cdisc_dataset("ADVS", ___,
                keys  = c("STUDYID", "USUBJID", "PARAMCD", "ADT"),
                label = "Vital Signs")
)

# Step 2: Define filters
pilot_filters <- teal_slices(
  teal_slice("ADSL", "SAFFL",   selected = ___),
  teal_slice("ADLB", "PARAMCD", selected = c("ALT", ___)),
  exclude_varnames = list(
    ADSL = c("STUDYID", ___),
    ADLB = c(___, "USUBJID")
  )
)

# Step 3: Assemble modules
pilot_modules <- modules(
  tab_group("Explorer",
    tm_data_table("Raw Data"),
    ___("Variable Distributions"),
    tm_missing_data("Missingness")
  ),
  tab_group("Safety",
    tm_t_summary("Demographics",    dataname = "ADSL"),
    tm_t_summary("AE Overview",     dataname = "ADAE")
  ),
  tab_group("Lab & Vitals",
    tm_g_scatterplot("Lab Values",  dataname = "ADLB",
                     x = "ADY", y = "AVAL"),
    tm_g_scatterplot("Vital Signs", dataname = "ADVS",
                     x = "ADY", y = "AVAL")
  )
)

# Step 4: Launch
pilot_app <- teal(
  data    = ___,
  modules = ___,
  filter  = ___,
  title   = "CDISCPILOT01 — Xanomeline Study Review"
)

print(pilot_app)
```

```{r capstone-app-solution}
pilot_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl,
                keys  = c("STUDYID", "USUBJID"),
                label = "Subject-Level Analysis Dataset"),
  cdisc_dataset("ADAE", adae,
                keys  = c("STUDYID", "USUBJID", "AESEQ"),
                label = "Adverse Events"),
  cdisc_dataset("ADLB", adlb,
                keys  = c("STUDYID", "USUBJID", "PARAMCD", "ADT"),
                label = "Lab Data"),
  cdisc_dataset("ADVS", advs,
                keys  = c("STUDYID", "USUBJID", "PARAMCD", "ADT"),
                label = "Vital Signs")
)

pilot_filters <- teal_slices(
  teal_slice("ADSL", "SAFFL",   selected = "Y"),
  teal_slice("ADLB", "PARAMCD", selected = c("ALT", "CREAT")),
  exclude_varnames = list(
    ADSL = c("STUDYID", "USUBJID"),
    ADLB = c("STUDYID", "USUBJID")
  )
)

pilot_modules <- modules(
  tab_group("Explorer",
    tm_data_table("Raw Data"),
    tm_variable_browser("Variable Distributions"),
    tm_missing_data("Missingness")
  ),
  tab_group("Safety",
    tm_t_summary("Demographics",    dataname = "ADSL"),
    tm_t_summary("AE Overview",     dataname = "ADAE")
  ),
  tab_group("Lab & Vitals",
    tm_g_scatterplot("Lab Values",  dataname = "ADLB",
                     x = "ADY", y = "AVAL"),
    tm_g_scatterplot("Vital Signs", dataname = "ADVS",
                     x = "ADY", y = "AVAL")
  )
)

pilot_app <- teal(
  data    = pilot_data,
  modules = pilot_modules,
  filter  = pilot_filters,
  title   = "CDISCPILOT01 — Xanomeline Study Review"
)

print(pilot_app)
```

---

## Summary

You now understand the full teal architecture:

- **teal.data** wraps datasets with `cdisc_dataset()` / `cdisc_data()`, declaring keys and relationships
- **`teal()`** takes data + modules + filters and produces a Shiny app
- **`teal_slices()`** / **`teal_slice()`** pre-populate and configure the filter panel
- The filter panel **propagates cross-dataset** via shared keys — no manual filter logic
- **Modules** are self-contained tabs that receive pre-filtered data
- **teal.code** automatically logs reproducible R code for every user action
- Apps deploy to **Posit Connect** as standard Shiny apps

Next: **Lesson 40** covers writing your own custom teal modules — the `ui_args` / `server_args` pattern, accessing filtered data inside a module, and integrating with `teal.code`.
