---
title: "Lesson 40 — teal: Custom Modules"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  Learn to write your own teal modules. Understand the module API (ui_args /
  server_args), accessing filtered datasets, interacting with teal.code for
  reproducibility, and using teal.widgets helpers inside custom modules.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 40_teal_modules.Rmd
# PURPOSE       : Custom teal module authoring — teal_module() constructor,
#                 ui_args/server_args pattern, reactive data access
#                 (data()[[...]]), qenv reproducibility, testServer().
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================
library(learnr)
library(dplyr)
library(ggplot2)
library(tibble)
library(shiny)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.eval = FALSE)

# ── Pharmaverseadam datasets ────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}
adsl <- adsl_raw |> mutate(
  ITTFL  = SAFFL,
  DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"))

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw)) {
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
}
adae <- adae_raw |> mutate(AESEVN = case_when(
  AESEV == "MILD"     ~ 1L, AESEV == "MODERATE" ~ 2L,
  AESEV == "SEVERE"   ~ 3L, TRUE ~ NA_integer_))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

advs <- pharmaverseadam::advs

# ── Mock teal infrastructure ────────────────────────────────────────────────
`%||%` <- function(a, b) if (is.null(a)) b else a

# teal.data mocks
cdisc_dataset <- function(dataname, x, keys = character(0), label = NULL) {
  structure(list(dataname=dataname, data=x, keys=keys,
                 label=label %||% dataname),
            class=c("cdisc_dataset","teal_dataset"))
}

cdisc_data <- function(..., check = FALSE) {
  datasets <- list(...)
  names(datasets) <- vapply(datasets, function(d) d$dataname, character(1))
  structure(list(datasets=datasets, check=check),
            class=c("cdisc_data","teal_data"))
}

teal_slice <- function(dataname, varname, selected=NULL, keep_na=FALSE, ...) {
  structure(list(dataname=dataname, varname=varname,
                 selected=selected, keep_na=keep_na),
            class="teal_slice")
}

teal_slices <- function(..., include_varnames=NULL, exclude_varnames=NULL) {
  slices <- list(...)
  structure(list(slices=slices,
                 include_varnames=include_varnames,
                 exclude_varnames=exclude_varnames),
            class="teal_slices")
}

modules <- function(..., label="root") {
  mods <- list(...)
  if (length(mods)==1 && is.list(mods[[1]]) && !inherits(mods[[1]],"teal_module"))
    mods <- mods[[1]]
  structure(list(modules=mods, label=label), class="teal_modules")
}

tab_group <- function(label, ...) {
  children <- list(...)
  structure(list(label=label, children=children),
            class=c("tab_group","teal_module"))
}

# ── teal_module() — the custom module constructor ───────────────────────────
# In real teal, you call teal_module() to register your module.
# Signature:  teal_module(label, ui, server, datanames="all", ui_args=list(), server_args=list(), ...)
teal_module <- function(label,
                        ui,
                        server,
                        datanames  = "all",
                        ui_args    = list(),
                        server_args = list(),
                        ...) {
  structure(
    list(label       = label,
         ui          = ui,
         server      = server,
         datanames   = datanames,
         ui_args     = ui_args,
         server_args = server_args,
         extra       = list(...)),
    class = c("teal_module_custom", "teal_module")
  )
}

# print method for custom modules
print.teal_module_custom <- function(x, ...) {
  cat("── Custom teal Module ──────────────────────────────────\n")
  cat("  Label:       ", x$label, "\n")
  cat("  Datanames:   ", paste(x$datanames, collapse=", "), "\n")
  cat("  UI args:     ", paste(names(x$ui_args), collapse=", "),
      if (length(x$ui_args)==0) "(none)" else "", "\n")
  cat("  Server args: ", paste(names(x$server_args), collapse=", "),
      if (length(x$server_args)==0) "(none)" else "", "\n")
  invisible(x)
}

# teal main app mock
teal <- function(data, modules, title="teal app",
                 filter=NULL, header=NULL, footer=NULL, ...) {
  structure(
    list(data=data, modules=modules, title=title,
         filter=filter, header=header, footer=footer),
    class=c("teal_app","shiny.appobj"))
}

print.teal_app <- function(x, ...) {
  cat("── teal Application ──────────────────────────────────\n")
  cat("  Title:", x$title, "\n")
  n_ds <- length(x$data$datasets)
  cat("  Datasets:", n_ds, "—", paste(names(x$data$datasets), collapse=", "), "\n")
  mods <- if (inherits(x$modules, "teal_modules")) x$modules$modules else list(x$modules)
  cat("  Modules:\n")
  for (m in mods) {
    if (inherits(m, "tab_group")) {
      cat("    [group]", m$label, "\n")
      for (c in m$children) cat("      •", c$label, "\n")
    } else {
      cat("    •", m$label, "\n")
    }
  }
  invisible(x)
}

# Mock filtered_data accessor (simulates what teal passes to server)
# In real teal, the server receives `data` as a reactive filtered_data object.
make_filtered_data <- function(cdisc_data_obj) {
  # Returns a reactive-like environment; in exercises we use a simple function
  env <- new.env(parent = emptyenv())
  env$datasets <- cdisc_data_obj$datasets
  env$get_data <- function(dataname) {
    if (dataname %in% names(env$datasets))
      env$datasets[[dataname]]$data
    else
      stop(sprintf("Dataset '%s' not found in filtered_data", dataname))
  }
  env
}

# teal.code mock — qenv (quoted environment)
qenv <- function() {
  env <- new.env(parent = baseenv())
  env$.code_log <- character(0)
  env$eval <- function(expr_str) {
    env$.code_log <- c(env$.code_log, expr_str)
    eval(parse(text = expr_str), envir = env)
    invisible(env)
  }
  env$get_code <- function() paste(env$.code_log, collapse = "\n")
  class(env) <- c("qenv", "environment")
  env
}

print.qenv <- function(x, ...) {
  cat("── qenv (reproducible code log) ──────────────────────\n")
  if (length(x$.code_log) == 0) {
    cat("  (empty)\n")
  } else {
    for (i in seq_along(x$.code_log)) cat(sprintf("  [%d] %s\n", i, x$.code_log[i]))
  }
  invisible(x)
}
```

## Overview

Lesson 39 showed how to wire together ready-made modules. This lesson teaches you to **write your own**. Custom teal modules are essential when:

- You need analysis logic not covered by `teal.modules.clinical`
- Your statistical team has house-style outputs that differ from the defaults
- You are embedding a validated R package output directly into the app

By the end of this lesson you will be able to author a teal module from scratch, pass configuration through `ui_args` / `server_args`, access filtered data, and log reproducible code with `teal.code`.

---

## 1 — Module Anatomy

### 1.1 The three-part structure

Every custom teal module has three parts:

```{r module-anatomy, eval=FALSE}
my_module <- teal_module(

  # 1. Label (appears as the tab title)
  label = "My Analysis",

  # 2. UI function — receives id and any ui_args
  ui = function(id, ...) {
    ns <- NS(id)
    tagList(
      selectInput(ns("arm"), "Treatment Arm",
                  choices = c("Placebo", "Xanomeline Low Dose",
                              "Xanomeline High Dose")),
      plotOutput(ns("plot"))
    )
  },

  # 3. Server function — receives id, data (reactive), reporter, ...
  server = function(id, data, reporter, ...) {
    moduleServer(id, function(input, output, session) {

      # data() returns the filtered cdisc_data object
      adsl <- reactive(data()[["ADSL"]])

      output$plot <- renderPlot({
        adsl() |>
          filter(TRT01P == input$arm) |>
          ggplot(aes(AGE)) +
          geom_histogram(bins = 20)
      })
    })
  },

  # 4. Which datasets this module needs (enables optimised lazy loading)
  datanames = "ADSL"
)
```

### 1.2 The `data` argument

Inside the server function, `data` is a **reactive** that returns the filtered data. Access individual datasets by name:

```{r data-access, eval=FALSE}
server = function(id, data, reporter, ...) {
  moduleServer(id, function(input, output, session) {

    # Correct: data is reactive — call it with ()
    adsl_filtered <- reactive(data()[["ADSL"]])
    adae_filtered <- reactive(data()[["ADAE"]])

    # Wrong: data[["ADSL"]] without () — not reactive
    # adsl_wrong <- data[["ADSL"]]   # <- static, won't update when filters change
  })
}
```

> **Key insight:** `data` is a reactive expression, not a list. Always call `data()` with parentheses inside reactive contexts (`reactive()`, `observe()`, `render*()`).

### 1.3 Quiz

```{r quiz-anatomy, echo=FALSE}
quiz(
  question(
    "Inside a teal module server, how do you get the filtered ADSL data?",
    answer("`data[[\"ADSL\"]]`"),
    answer("`data()[\"ADSL\"]`"),
    answer("`data()[[\"ADSL\"]]`", correct = TRUE),
    answer("`filtered_data(\"ADSL\")`"),
    allow_retry = TRUE
  ),
  question(
    "What does the `datanames` argument in `teal_module()` control?",
    answer("The column names that the module can use"),
    answer("Which datasets teal pre-loads for this module, enabling optimised lazy loading", correct = TRUE),
    answer("The datasets that the filter panel shows"),
    answer("The order of tabs in the app"),
    allow_retry = TRUE
  )
)
```

---

## 2 — Building Your First Custom Module

### 2.1 A demographics bar chart module

Let's build a module that shows a bar chart of a chosen demographic variable by treatment arm.

```{r first-module-build, eval=FALSE}
# ── Module definition ──────────────────────────────────────────────────────
tm_demographics_bar <- function(label    = "Demographics Bar Chart",
                                dataname = "ADSL",
                                arm_var  = "TRT01P",
                                vars     = c("SEX", "RACE", "AGEGR1")) {
  teal_module(
    label      = label,
    datanames  = dataname,

    # Pass configuration as ui_args / server_args
    ui_args    = list(vars = vars),
    server_args = list(dataname = dataname,
                       arm_var  = arm_var,
                       vars     = vars),

    ui = function(id, vars, ...) {
      ns <- NS(id)
      tagList(
        fluidRow(
          column(4,
            selectInput(ns("var"), "Variable",
                        choices  = vars,
                        selected = vars[1])
          )
        ),
        plotOutput(ns("bar_chart"), height = "400px")
      )
    },

    server = function(id, data, reporter, dataname, arm_var, vars, ...) {
      moduleServer(id, function(input, output, session) {

        adsl_r <- reactive(data()[[dataname]])

        output$bar_chart <- renderPlot({
          req(adsl_r(), input$var)
          adsl_r() |>
            count(.data[[arm_var]], .data[[input$var]]) |>
            ggplot(aes(x = .data[[arm_var]],
                       y = n,
                       fill = .data[[input$var]])) +
            geom_col(position = "dodge") +
            labs(x = arm_var, y = "Count",
                 fill = input$var,
                 title = paste("Distribution of", input$var)) +
            theme_minimal(base_size = 14) +
            theme(axis.text.x = element_text(angle = 30, hjust = 1))
        })
      })
    }
  )
}
```

### 2.2 `ui_args` and `server_args`

Configuration flows from the module constructor to `ui` and `server` via named argument lists:

```{r args-flow, eval=FALSE}
# How args flow:
#
#  tm_my_module(arm_var = "TRT01P", ...)
#          │
#          ├─ ui_args    = list(arm_var = "TRT01P")
#          │       └──────────► ui(id, arm_var, ...)
#          │
#          └─ server_args = list(arm_var = "TRT01P")
#                  └──────────► server(id, data, reporter, arm_var, ...)
```

This pattern means:
- Module constructors become simple **wrappers** with documented arguments
- The same `teal_module()` core can be re-used with different configurations
- Argument names in `ui` and `server` **must match** the names in `ui_args` / `server_args`

### 2.3 Exercise — complete the module definition

Complete the `tm_ae_severity_bar` module. It should show a bar chart of AE severity counts by treatment arm.

```{r module-complete-ex, exercise=TRUE, exercise.lines=45}
tm_ae_severity_bar <- function(label    = "AE Severity",
                               dataname = "ADAE",
                               arm_var  = "TRT01A") {
  teal_module(
    label       = label,
    datanames   = ___,
    ui_args     = list(),
    server_args = list(dataname = ___, arm_var = ___),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        plotOutput(ns("sev_plot"), height = "400px")
      )
    },

    server = function(id, data, reporter, dataname, arm_var, ...) {
      moduleServer(id, function(input, output, session) {

        adae_r <- reactive(___[[___]])

        output$sev_plot <- renderPlot({
          req(adae_r())
          adae_r() |>
            count(.data[[arm_var]], AESEV) |>
            ggplot(aes(x = .data[[arm_var]],
                       y = n,
                       fill = AESEV)) +
            geom_col(position = "stack") +
            labs(x = arm_var, y = "Number of AEs",
                 fill = "Severity",
                 title = "AE Severity by Treatment Arm") +
            theme_minimal(base_size = 14)
        })
      })
    }
  )
}

# Test: print the module definition
mod <- tm_ae_severity_bar()
print(mod)
```

```{r module-complete-ex-solution}
tm_ae_severity_bar <- function(label    = "AE Severity",
                               dataname = "ADAE",
                               arm_var  = "TRT01A") {
  teal_module(
    label       = label,
    datanames   = dataname,
    ui_args     = list(),
    server_args = list(dataname = dataname, arm_var = arm_var),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        plotOutput(ns("sev_plot"), height = "400px")
      )
    },

    server = function(id, data, reporter, dataname, arm_var, ...) {
      moduleServer(id, function(input, output, session) {

        adae_r <- reactive(data()[[dataname]])

        output$sev_plot <- renderPlot({
          req(adae_r())
          adae_r() |>
            count(.data[[arm_var]], AESEV) |>
            ggplot(aes(x = .data[[arm_var]],
                       y = n,
                       fill = AESEV)) +
            geom_col(position = "stack") +
            labs(x = arm_var, y = "Number of AEs",
                 fill = "Severity",
                 title = "AE Severity by Treatment Arm") +
            theme_minimal(base_size = 14)
        })
      })
    }
  )
}

mod <- tm_ae_severity_bar()
print(mod)
```

---

## 3 — Accessing Multiple Datasets

### 3.1 Cross-dataset modules

Some modules need data from more than one dataset — for example, merging ADLB with ADSL to display treatment labels.

```{r multi-dataset-module, eval=FALSE}
tm_lab_by_arm <- function(label    = "Lab by Arm",
                          param    = "ALT") {
  teal_module(
    label       = label,
    datanames   = c("ADSL", "ADLB"),   # declare both
    server_args = list(param = param),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        plotOutput(ns("lab_plot"), height = "400px")
      )
    },

    server = function(id, data, reporter, param, ...) {
      moduleServer(id, function(input, output, session) {

        adsl_r <- reactive(data()[["ADSL"]])
        adlb_r <- reactive(data()[["ADLB"]])

        # Merged dataset — both adsl_r() and adlb_r() are reactive
        lab_data <- reactive({
          req(adsl_r(), adlb_r())
          adlb_r() |>
            filter(PARAMCD == param) |>
            select(-any_of("TRT01P")) |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by = "USUBJID")
        })

        output$lab_plot <- renderPlot({
          req(lab_data())
          lab_data() |>
            filter(!is.na(AVAL), !is.na(AVISIT)) |>
            ggplot(aes(x = AVISIT, y = AVAL,
                       colour = TRT01P, group = TRT01P)) +
            stat_summary(fun = mean, geom = "line", linewidth = 1) +
            stat_summary(fun = mean, geom = "point", size = 3) +
            labs(title = paste("Mean", param, "over Time"),
                 x = "Visit", y = "Observed Value",
                 colour = "Treatment") +
            theme_minimal(base_size = 14) +
            theme(axis.text.x = element_text(angle = 45, hjust = 1))
        })
      })
    }
  )
}
```

### 3.2 The `select(-any_of(...))` guard

Always remove any column that already exists in the receiving dataset before a join:

```{r join-guard, eval=FALSE}
# ADLB already has TRT01P (from admiral derivation)
# Joining adsl's TRT01P without removing it first creates TRT01P.x / TRT01P.y

# WRONG:
adlb_r() |>
  left_join(select(adsl_r(), USUBJID, TRT01P), by = "USUBJID")
  # → TRT01P.x (from adlb) and TRT01P.y (from adsl) — downstream breaks

# CORRECT:
adlb_r() |>
  select(-any_of("TRT01P")) |>        # remove existing copy first
  left_join(select(adsl_r(), USUBJID, TRT01P), by = "USUBJID")
  # → single TRT01P from adsl
```

### 3.3 Exercise — multi-dataset module

Complete the server of a module that merges ADLB with ADSL and produces a boxplot of AVAL by treatment arm for a specified parameter.

```{r multi-ds-ex, exercise=TRUE, exercise.lines=40}
tm_lab_boxplot <- function(label = "Lab Boxplot",
                           param = "ALT") {
  teal_module(
    label       = label,
    datanames   = c(___, "ADLB"),
    server_args = list(param = param),

    ui = function(id, ...) {
      ns <- NS(id)
      plotOutput(ns("box_plot"), height = "400px")
    },

    server = function(id, data, reporter, param, ...) {
      moduleServer(id, function(input, output, session) {

        adsl_r <- reactive(data()[[___]])
        adlb_r <- reactive(data()[[___]])

        lab_merged <- reactive({
          req(adsl_r(), adlb_r())
          adlb_r() |>
            filter(PARAMCD == ___) |>
            select(-any_of(___)) |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by = "USUBJID")
        })

        output$box_plot <- renderPlot({
          req(lab_merged())
          lab_merged() |>
            filter(!is.na(AVAL)) |>
            ggplot(aes(x = TRT01P, y = AVAL, fill = TRT01P)) +
            geom_boxplot(outlier.shape = 21, outlier.size = 2) +
            labs(title = paste(param, "Distribution by Treatment"),
                 x = NULL, y = "Observed Value") +
            theme_minimal(base_size = 14) +
            theme(legend.position = "none",
                  axis.text.x = element_text(angle = 30, hjust = 1))
        })
      })
    }
  )
}

mod <- tm_lab_boxplot(param = "ALT")
print(mod)
```

```{r multi-ds-ex-solution}
tm_lab_boxplot <- function(label = "Lab Boxplot",
                           param = "ALT") {
  teal_module(
    label       = label,
    datanames   = c("ADSL", "ADLB"),
    server_args = list(param = param),

    ui = function(id, ...) {
      ns <- NS(id)
      plotOutput(ns("box_plot"), height = "400px")
    },

    server = function(id, data, reporter, param, ...) {
      moduleServer(id, function(input, output, session) {

        adsl_r <- reactive(data()[["ADSL"]])
        adlb_r <- reactive(data()[["ADLB"]])

        lab_merged <- reactive({
          req(adsl_r(), adlb_r())
          adlb_r() |>
            filter(PARAMCD == param) |>
            select(-any_of("TRT01P")) |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by = "USUBJID")
        })

        output$box_plot <- renderPlot({
          req(lab_merged())
          lab_merged() |>
            filter(!is.na(AVAL)) |>
            ggplot(aes(x = TRT01P, y = AVAL, fill = TRT01P)) +
            geom_boxplot(outlier.shape = 21, outlier.size = 2) +
            labs(title = paste(param, "Distribution by Treatment"),
                 x = NULL, y = "Observed Value") +
            theme_minimal(base_size = 14) +
            theme(legend.position = "none",
                  axis.text.x = element_text(angle = 30, hjust = 1))
        })
      })
    }
  )
}

mod <- tm_lab_boxplot(param = "ALT")
print(mod)
```

---

## 4 — teal.code: Reproducible Code Logging

### 4.1 What is teal.code?

`teal.code` provides the `qenv` (quoted environment) class. Every data manipulation and analysis step that a module performs can be logged to a `qenv`, so the app can present users with a downloadable R script that exactly reproduces what they see.

In regulated environments this is critical: a clinical data reviewer can see the output **and** download the code that produced it.

### 4.2 Basic `qenv` usage

```{r qenv-demo, eval=FALSE}
library(teal.code)

# Create a fresh quoted environment
q <- qenv()

# Execute code inside the qenv — it runs AND is recorded
q <- q |>
  eval_code("adsl_sub <- dplyr::filter(adsl, SAFFL == 'Y')") |>
  eval_code("n_subjects <- nrow(adsl_sub)") |>
  eval_code("cat('N =', n_subjects, '\\n')")

# Retrieve the logged code
get_code(q)
# [1]
# adsl_sub <- dplyr::filter(adsl, SAFFL == 'Y')
# n_subjects <- nrow(adsl_sub)
# cat('N =', n_subjects, '\n')
```

### 4.3 `qenv` inside a module server

In real teal, the `data` reactive returns a `teal_data` object that includes a `qenv`. Modules append their analysis code to it:

```{r qenv-module-pattern, eval=FALSE}
server = function(id, data, reporter, param, ...) {
  moduleServer(id, function(input, output, session) {

    # data() is a reactive teal_data object
    # data()$env is its qenv

    result <- reactive({
      req(data())
      # Get the pre-filtered data + its code history
      tdata <- data()

      # Extend the qenv with module-specific code
      q <- tdata$env |>
        eval_code(
          sprintf(
            'adlb_param <- dplyr::filter(ADLB, PARAMCD == "%s")',
            param
          )
        ) |>
        eval_code(
          'p <- ggplot2::ggplot(adlb_param, ggplot2::aes(x=AVISIT, y=AVAL)) +
             ggplot2::geom_boxplot()'
        )

      list(
        data  = q$adlb_param,
        plot  = q$p,
        code  = get_code(q)
      )
    })

    output$lab_plot <- renderPlot(result()$plot)

    # Register with teal's reporter so "Show R Code" works
    reporter$add_card({
      teal.reporter::rcode_card(
        code = result()$code,
        title = paste("Lab Module —", param)
      )
    })
  })
}
```

### 4.4 Simplified `qenv` exercise

Use the mock `qenv` defined in the setup chunk to build a reproducible analysis pipeline.

```{r qenv-exercise, exercise=TRUE, exercise.lines=25}
# Build a qenv that:
# 1. Filters adsl to SAFFL == "Y"
# 2. Computes mean age by TRT01P
# 3. Prints the result

q <- qenv()

q$eval('adsl_saf <- dplyr::filter(adsl, ___ == "Y")')
q$eval('age_summary <- adsl_saf |>
          dplyr::group_by(___) |>
          dplyr::summarise(mean_age = round(mean(AGE, na.rm=TRUE), 1),
                           n = dplyr::n())')
q$eval('print(___)')

# Show the code log
cat("\n── Recorded code ────────────────────────────────\n")
cat(q$get_code(), "\n")
```

```{r qenv-exercise-solution}
q <- qenv()

q$eval('adsl_saf <- dplyr::filter(adsl, SAFFL == "Y")')
q$eval('age_summary <- adsl_saf |>
          dplyr::group_by(TRT01P) |>
          dplyr::summarise(mean_age = round(mean(AGE, na.rm=TRUE), 1),
                           n = dplyr::n())')
q$eval('print(age_summary)')

cat("\n── Recorded code ────────────────────────────────\n")
cat(q$get_code(), "\n")
```

---

## 5 — Dynamic UI Inside Modules

### 5.1 Variable selectors driven by data

A common pattern: populate a `selectInput` with the actual column names of the filtered dataset.

```{r dynamic-ui-module, eval=FALSE}
tm_dynamic_scatter <- function(label    = "Dynamic Scatter",
                               dataname = "ADSL") {
  teal_module(
    label       = label,
    datanames   = dataname,
    server_args = list(dataname = dataname),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        fluidRow(
          column(3, selectInput(ns("x_var"), "X Variable", choices = NULL)),
          column(3, selectInput(ns("y_var"), "Y Variable", choices = NULL)),
          column(3, selectInput(ns("col_var"), "Colour by",
                                choices = c("(none)" = ""), selected = ""))
        ),
        plotOutput(ns("scatter"), height = "450px")
      )
    },

    server = function(id, data, reporter, dataname, ...) {
      moduleServer(id, function(input, output, session) {

        dataset_r <- reactive(data()[[dataname]])

        # Populate variable selectors from actual data columns
        observe({
          req(dataset_r())
          num_vars <- names(select(dataset_r(), where(is.numeric)))
          cat_vars <- c("(none)" = "",
                        names(select(dataset_r(), where(is.character))))

          updateSelectInput(session, "x_var",
                            choices = num_vars, selected = num_vars[1])
          updateSelectInput(session, "y_var",
                            choices = num_vars,
                            selected = num_vars[min(2, length(num_vars))])
          updateSelectInput(session, "col_var", choices = cat_vars)
        })

        output$scatter <- renderPlot({
          req(dataset_r(), input$x_var, input$y_var)
          p <- dataset_r() |>
            ggplot(aes(x = .data[[input$x_var]],
                       y = .data[[input$y_var]]))
          if (nchar(input$col_var) > 0)
            p <- p + aes(colour = .data[[input$col_var]])
          p + geom_point(alpha = 0.6, size = 2) +
            theme_minimal(base_size = 14)
        })
      })
    }
  )
}
```

### 5.2 Exercise — add a visit selector

Complete the server of a module that shows ADLB data for a user-selected PARAMCD and AVISIT.

```{r dynamic-visit-ex, exercise=TRUE, exercise.lines=40}
tm_lab_visit <- function(label = "Lab by Visit") {
  teal_module(
    label      = label,
    datanames  = c("ADSL", "ADLB"),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        fluidRow(
          column(3, selectInput(ns("param"),  "Parameter", choices = NULL)),
          column(3, selectInput(ns("visit"),  "Visit",     choices = NULL))
        ),
        plotOutput(ns("lab_plot"), height = "400px")
      )
    },

    server = function(id, data, reporter, ...) {
      moduleServer(id, function(input, output, session) {

        adsl_r <- reactive(data()[["ADSL"]])
        adlb_r <- reactive(data()[["ADLB"]])

        # Populate param choices
        observe({
          req(adlb_r())
          params <- sort(unique(adlb_r()$___))
          updateSelectInput(session, "param", choices = params,
                            selected = params[1])
        })

        # Populate visit choices based on selected param
        observe({
          req(adlb_r(), input$param)
          visits <- adlb_r() |>
            filter(PARAMCD == ___) |>
            pull(AVISIT) |>
            unique() |> sort()
          updateSelectInput(session, "visit", choices = visits,
                            selected = visits[1])
        })

        # Merged reactive
        plot_data <- reactive({
          req(adlb_r(), adsl_r(), input$param, input$visit)
          adlb_r() |>
            filter(PARAMCD == ___, AVISIT == ___) |>
            select(-any_of("TRT01P")) |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by = "USUBJID")
        })

        output$lab_plot <- renderPlot({
          req(plot_data())
          plot_data() |>
            filter(!is.na(AVAL)) |>
            ggplot(aes(x = TRT01P, y = AVAL, fill = TRT01P)) +
            geom_boxplot() +
            labs(title = paste(input$param, "at", input$visit),
                 x = NULL, y = "Value") +
            theme_minimal(base_size = 14) +
            theme(legend.position = "none",
                  axis.text.x = element_text(angle = 30, hjust = 1))
        })
      })
    }
  )
}

mod <- tm_lab_visit()
print(mod)
```

```{r dynamic-visit-ex-solution}
tm_lab_visit <- function(label = "Lab by Visit") {
  teal_module(
    label      = label,
    datanames  = c("ADSL", "ADLB"),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        fluidRow(
          column(3, selectInput(ns("param"),  "Parameter", choices = NULL)),
          column(3, selectInput(ns("visit"),  "Visit",     choices = NULL))
        ),
        plotOutput(ns("lab_plot"), height = "400px")
      )
    },

    server = function(id, data, reporter, ...) {
      moduleServer(id, function(input, output, session) {

        adsl_r <- reactive(data()[["ADSL"]])
        adlb_r <- reactive(data()[["ADLB"]])

        observe({
          req(adlb_r())
          params <- sort(unique(adlb_r()$PARAMCD))
          updateSelectInput(session, "param", choices = params,
                            selected = params[1])
        })

        observe({
          req(adlb_r(), input$param)
          visits <- adlb_r() |>
            filter(PARAMCD == input$param) |>
            pull(AVISIT) |>
            unique() |> sort()
          updateSelectInput(session, "visit", choices = visits,
                            selected = visits[1])
        })

        plot_data <- reactive({
          req(adlb_r(), adsl_r(), input$param, input$visit)
          adlb_r() |>
            filter(PARAMCD == input$param, AVISIT == input$visit) |>
            select(-any_of("TRT01P")) |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by = "USUBJID")
        })

        output$lab_plot <- renderPlot({
          req(plot_data())
          plot_data() |>
            filter(!is.na(AVAL)) |>
            ggplot(aes(x = TRT01P, y = AVAL, fill = TRT01P)) +
            geom_boxplot() +
            labs(title = paste(input$param, "at", input$visit),
                 x = NULL, y = "Value") +
            theme_minimal(base_size = 14) +
            theme(legend.position = "none",
                  axis.text.x = element_text(angle = 30, hjust = 1))
        })
      })
    }
  )
}

mod <- tm_lab_visit()
print(mod)
```

---

## 6 — Testing Custom Modules

### 6.1 Testing with `testServer()`

teal modules are Shiny modules at heart, so Shiny's `testServer()` works directly:

```{r test-server-pattern, eval=FALSE}
library(shiny)
library(testthat)

test_that("tm_ae_severity_bar server runs without error", {

  # Simulate filtered data
  mock_data <- list(
    ADAE = adae |> filter(TRT01A %in% c("Placebo", "Xanomeline High Dose"))
  )
  mock_reactive <- function() mock_data

  mod_def <- tm_ae_severity_bar(arm_var = "TRT01A")

  testServer(
    mod_def$server,
    args = list(
      data     = mock_reactive,
      reporter = list(),   # mock reporter
      dataname = "ADAE",
      arm_var  = "TRT01A"
    ),
    {
      # Check output renders
      expect_s3_class(output$sev_plot, "recordedplot")
    }
  )
})
```

### 6.2 Unit-testing reactive data flow

```{r reactive-test, eval=FALSE}
test_that("lab_merged filters and joins correctly", {
  mock_data <- list(
    ADSL = adsl,
    ADLB = adlb
  )
  mock_reactive <- function() mock_data

  mod_def <- tm_lab_boxplot(param = "ALT")

  testServer(
    mod_def$server,
    args = list(data = mock_reactive, reporter = list(), param = "ALT"),
    {
      # Access internal reactive
      session$setInputs()   # no inputs to set
      result <- lab_merged()

      expect_true(all(result$PARAMCD == "ALT"))
      expect_true("TRT01P" %in% names(result))
      expect_false(any(duplicated(names(result))))   # no .x/.y suffixes
    }
  )
})
```

### 6.3 Quiz

```{r quiz-testing, echo=FALSE}
quiz(
  question(
    "Which Shiny testing function lets you test the server logic of a teal module?",
    answer("`runTests()`"),
    answer("`testServer()`", correct = TRUE),
    answer("`shinytest2::AppDriver()`"),
    answer("`evaluate_promise()`"),
    allow_retry = TRUE
  ),
  question(
    "When testing a module that joins ADLB with ADSL, what should you check to catch the .x/.y column suffix bug?",
    answer("`expect_equal(ncol(result), 10)`"),
    answer("`expect_false(any(duplicated(names(result))))`", correct = TRUE),
    answer("`expect_true(\"PARAMCD\" %in% names(result))`"),
    answer("`expect_s3_class(result, \"data.frame\")`"),
    allow_retry = TRUE
  )
)
```

---

## 7 — Module Best Practices

### 7.1 Design checklist

Before shipping a custom teal module:

**Data safety**
- [ ] Declare all needed datasets in `datanames`
- [ ] Use `select(-any_of(...))` before any join that adds a column that may already exist
- [ ] Wrap all reactives with `req()` to prevent errors on empty filter results
- [ ] Use `.data[[var]]` (tidy eval) instead of `df$var` for programmatic column access

**Reproducibility**
- [ ] Log all data manipulation steps to `qenv` / `eval_code`
- [ ] Register output with `teal.reporter` if the module produces analysable results

**UX**
- [ ] Populate variable selectors from the actual filtered data (not hardcoded values)
- [ ] Show an informative message when `nrow(data()) == 0` after filtering
- [ ] Use `withProgress()` for computations > ~1 second

**Testing**
- [ ] Write at least one `testServer()` test per module
- [ ] Test with an empty dataset (0 rows after filter) — should not crash

### 7.2 Exercise — add an empty-data guard

Add a guard to this module's `renderPlot` that shows a message when the filtered data has 0 rows.

```{r empty-guard-ex, exercise=TRUE, exercise.lines=30}
# Module server with empty-data guard
safe_plot_server <- function(input, output, session, data_r) {
  output$plot <- renderPlot({
    req(data_r())

    d <- data_r()

    # Add guard here: if d has 0 rows, show a message instead of plotting
    if (___ == 0) {
      plot.new()
      text(0.5, 0.5,
           "No data after applying filters.\nAdjust the filter panel.",
           cex = 1.4, col = "grey50")
      return(invisible(NULL))
    }

    d |>
      count(TRT01P, SEX) |>
      ggplot(aes(x = TRT01P, y = n, fill = SEX)) +
      geom_col(position = "dodge") +
      theme_minimal(base_size = 14)
  })
}

# Test it with real data
safe_data <- reactive(adsl)

testServer(
  function(input, output, session) {
    safe_plot_server(input, output, session, safe_data)
  },
  {
    expect_silent(output$plot)
  }
)

cat("Module with guard compiled successfully\n")
```

```{r empty-guard-ex-solution}
safe_plot_server <- function(input, output, session, data_r) {
  output$plot <- renderPlot({
    req(data_r())

    d <- data_r()

    if (nrow(d) == 0) {
      plot.new()
      text(0.5, 0.5,
           "No data after applying filters.\nAdjust the filter panel.",
           cex = 1.4, col = "grey50")
      return(invisible(NULL))
    }

    d |>
      count(TRT01P, SEX) |>
      ggplot(aes(x = TRT01P, y = n, fill = SEX)) +
      geom_col(position = "dodge") +
      theme_minimal(base_size = 14)
  })
}

safe_data <- reactive(adsl)

testServer(
  function(input, output, session) {
    safe_plot_server(input, output, session, safe_data)
  },
  {
    expect_silent(output$plot)
  }
)

cat("Module with guard compiled successfully\n")
```

---

## 8 — Capstone: Build a Complete Custom Module Suite

Build a three-module custom suite and wire it into a teal app.

### Requirements:

1. **`tm_subj_count`** — shows N by treatment arm as a simple table (ADSL)
2. **`tm_ae_rate`** — shows % subjects with ≥1 AE by treatment arm (ADSL + ADAE)
3. **`tm_lab_shift`** — shows a simple shift table (BTOXGR → ATOXGR) for a chosen parameter (ADSL + ADLB)

```{r capstone-modules, exercise=TRUE, exercise.eval=FALSE, exercise.lines=100}
# ── Module 1: Subject Counts ──────────────────────────────────────────────
tm_subj_count <- function(label = "Subject Counts") {
  teal_module(
    label      = label,
    datanames  = "ADSL",
    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        h4("Subject Counts by Treatment Arm"),
        tableOutput(ns("count_table"))
      )
    },
    server = function(id, data, reporter, ...) {
      moduleServer(id, function(input, output, session) {
        adsl_r <- reactive(data()[[___]])
        output$count_table <- renderTable({
          req(adsl_r())
          adsl_r() |>
            count(___, name = "N") |>
            rename(`Treatment Arm` = TRT01P)
        })
      })
    }
  )
}

# ── Module 2: AE Rate ──────────────────────────────────────────────────────
tm_ae_rate <- function(label = "AE Rate") {
  teal_module(
    label      = label,
    datanames  = c("ADSL", "ADAE"),
    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        h4("% Subjects with ≥1 AE"),
        tableOutput(ns("ae_table"))
      )
    },
    server = function(id, data, reporter, ...) {
      moduleServer(id, function(input, output, session) {
        adsl_r <- reactive(data()[[___]])
        adae_r <- reactive(data()[[___]])

        output$ae_table <- renderTable({
          req(adsl_r(), adae_r())
          denom <- adsl_r() |>
            count(TRT01P, name = "N_total")

          subj_with_ae <- adae_r() |>
            distinct(USUBJID) |>
            select(-any_of("TRT01P")) |>
            left_join(select(adsl_r(), USUBJID, ___), by = "USUBJID") |>
            count(___, name = "N_ae")

          denom |>
            left_join(___, by = "TRT01P") |>
            mutate(N_ae = coalesce(N_ae, 0L),
                   Pct  = round(100 * N_ae / N_total, 1)) |>
            rename(`Treatment Arm` = TRT01P,
                   `N (total)`     = N_total,
                   `N (≥1 AE)`     = N_ae,
                   `%`             = Pct)
        })
      })
    }
  )
}

# ── Module 3: Lab Shift ────────────────────────────────────────────────────
tm_lab_shift <- function(label = "Lab Shift", default_param = "ALT") {
  teal_module(
    label       = label,
    datanames   = c("ADSL", "ADLB"),
    server_args = list(default_param = default_param),
    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        selectInput(ns("param"), "Parameter", choices = NULL),
        tableOutput(ns("shift_table"))
      )
    },
    server = function(id, data, reporter, default_param, ...) {
      moduleServer(id, function(input, output, session) {
        adsl_r <- reactive(data()[["ADSL"]])
        adlb_r <- reactive(data()[["ADLB"]])

        observe({
          req(adlb_r())
          params <- sort(unique(adlb_r()$PARAMCD))
          updateSelectInput(session, "param", choices = params,
                            selected = default_param)
        })

        output$shift_table <- renderTable({
          req(adlb_r(), input$param)
          adlb_r() |>
            filter(PARAMCD == input$param,
                   !is.na(BTOXGR), !is.na(ATOXGR)) |>
            count(BTOXGR, ATOXGR) |>
            tidyr::pivot_wider(names_from  = ATOXGR,
                               values_from = n,
                               values_fill = 0L,
                               names_prefix = "Post: Grade ") |>
            rename(`Baseline Grade` = BTOXGR)
        }, striped = TRUE, hover = TRUE, bordered = TRUE)
      })
    }
  )
}

# ── Assemble the app ──────────────────────────────────────────────────────
suite_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID")),
  cdisc_dataset("ADAE", adae, keys = c("STUDYID", "USUBJID", "AESEQ")),
  cdisc_dataset("ADLB", adlb, keys = c("STUDYID", "USUBJID", "PARAMCD", "ADT"))
)

suite_app <- teal(
  data    = suite_data,
  modules = modules(
    tm_subj_count("Subject Counts"),
    tm_ae_rate("AE Rate"),
    tm_lab_shift("Lab Shift", default_param = "ALT")
  ),
  filter  = teal_slices(
    teal_slice("ADSL", "SAFFL", selected = "Y")
  ),
  title   = "Custom Module Suite — CDISCPILOT01"
)

print(suite_app)
```

```{r capstone-modules-solution}
tm_subj_count <- function(label = "Subject Counts") {
  teal_module(
    label      = label,
    datanames  = "ADSL",
    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        h4("Subject Counts by Treatment Arm"),
        tableOutput(ns("count_table"))
      )
    },
    server = function(id, data, reporter, ...) {
      moduleServer(id, function(input, output, session) {
        adsl_r <- reactive(data()[["ADSL"]])
        output$count_table <- renderTable({
          req(adsl_r())
          adsl_r() |>
            count(TRT01P, name = "N") |>
            rename(`Treatment Arm` = TRT01P)
        })
      })
    }
  )
}

tm_ae_rate <- function(label = "AE Rate") {
  teal_module(
    label      = label,
    datanames  = c("ADSL", "ADAE"),
    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        h4("% Subjects with ≥1 AE"),
        tableOutput(ns("ae_table"))
      )
    },
    server = function(id, data, reporter, ...) {
      moduleServer(id, function(input, output, session) {
        adsl_r <- reactive(data()[["ADSL"]])
        adae_r <- reactive(data()[["ADAE"]])

        output$ae_table <- renderTable({
          req(adsl_r(), adae_r())
          denom <- adsl_r() |>
            count(TRT01P, name = "N_total")

          subj_with_ae <- adae_r() |>
            distinct(USUBJID) |>
            select(-any_of("TRT01P")) |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by = "USUBJID") |>
            count(TRT01P, name = "N_ae")

          denom |>
            left_join(subj_with_ae, by = "TRT01P") |>
            mutate(N_ae = coalesce(N_ae, 0L),
                   Pct  = round(100 * N_ae / N_total, 1)) |>
            rename(`Treatment Arm` = TRT01P,
                   `N (total)`     = N_total,
                   `N (≥1 AE)`     = N_ae,
                   `%`             = Pct)
        })
      })
    }
  )
}

tm_lab_shift <- function(label = "Lab Shift", default_param = "ALT") {
  teal_module(
    label       = label,
    datanames   = c("ADSL", "ADLB"),
    server_args = list(default_param = default_param),
    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        selectInput(ns("param"), "Parameter", choices = NULL),
        tableOutput(ns("shift_table"))
      )
    },
    server = function(id, data, reporter, default_param, ...) {
      moduleServer(id, function(input, output, session) {
        adsl_r <- reactive(data()[["ADSL"]])
        adlb_r <- reactive(data()[["ADLB"]])

        observe({
          req(adlb_r())
          params <- sort(unique(adlb_r()$PARAMCD))
          updateSelectInput(session, "param", choices = params,
                            selected = default_param)
        })

        output$shift_table <- renderTable({
          req(adlb_r(), input$param)
          adlb_r() |>
            filter(PARAMCD == input$param,
                   !is.na(BTOXGR), !is.na(ATOXGR)) |>
            count(BTOXGR, ATOXGR) |>
            tidyr::pivot_wider(names_from  = ATOXGR,
                               values_from = n,
                               values_fill = 0L,
                               names_prefix = "Post: Grade ") |>
            rename(`Baseline Grade` = BTOXGR)
        }, striped = TRUE, hover = TRUE, bordered = TRUE)
      })
    }
  )
}

suite_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl, keys = c("STUDYID", "USUBJID")),
  cdisc_dataset("ADAE", adae, keys = c("STUDYID", "USUBJID", "AESEQ")),
  cdisc_dataset("ADLB", adlb, keys = c("STUDYID", "USUBJID", "PARAMCD", "ADT"))
)

suite_app <- teal(
  data    = suite_data,
  modules = modules(
    tm_subj_count("Subject Counts"),
    tm_ae_rate("AE Rate"),
    tm_lab_shift("Lab Shift", default_param = "ALT")
  ),
  filter  = teal_slices(
    teal_slice("ADSL", "SAFFL", selected = "Y")
  ),
  title   = "Custom Module Suite — CDISCPILOT01"
)

print(suite_app)
```

---

## Summary

You can now author production-quality custom teal modules:

- **`teal_module()`** takes a `label`, `ui`, `server`, `datanames`, `ui_args`, and `server_args`
- Configuration flows from the constructor to `ui` and `server` through `ui_args` / `server_args`
- Inside the server, `data()[[\"DATANAME\"]]` retrieves the current filtered dataset as a reactive
- **Always** use `select(-any_of(...))` before joining a column that may already exist
- `teal.code`'s `qenv` / `eval_code` logs every analysis step for reproducible R code download
- Dynamic UI (variable selectors driven by real data) uses `observe` + `updateSelectInput`
- Empty-data guards (`if (nrow(d) == 0)`) prevent cryptic errors after aggressive filtering
- `testServer()` validates reactive data flow without launching the app

Next: **Lesson 41** covers `teal.data` in depth — `relational_data`, pull-from-server data connectors, lazy data loading, and deploying teal apps to Posit Connect with renv and environment variables.
