---
title: "Lesson 41 — teal: Data Connectors & Deployment"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Deep dive into teal.data — cdisc_data, relational_data, pull-from-server
  data connectors, lazy loading, dataset relationships, and full deployment
  workflows for Posit Connect, Docker, and internal clinical environments.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 41_teal_data_deployment.Rmd
# PURPOSE       : teal data layer and deployment — teal_data() + join_keys() API,
#                 dataset_connector() lazy loading, code reproducibility,
#                 and production deployment on Posit Connect / Docker.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================
library(learnr)
library(dplyr)
library(tibble)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.eval = FALSE)

# ── Pharmaverseadam datasets ────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}
adsl <- adsl_raw |> mutate(
  ITTFL  = SAFFL,
  DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"))

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw)) {
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
}
adae <- adae_raw |> mutate(AESEVN = case_when(
  AESEV == "MILD"     ~ 1L, AESEV == "MODERATE" ~ 2L,
  AESEV == "SEVERE"   ~ 3L, TRUE ~ NA_integer_))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

advs <- pharmaverseadam::advs

# ── Mock teal.data infrastructure ────────────────────────────────────────────
`%||%` <- function(a, b) if (is.null(a)) b else a

# ── cdisc_dataset / cdisc_data (review from lesson 39) ──────────────────────
cdisc_dataset <- function(dataname, x, keys=character(0), label=NULL,
                          code=NULL, parent=NULL) {
  structure(
    list(dataname=dataname, data=x, keys=keys,
         label=label %||% dataname, code=code, parent=parent),
    class=c("cdisc_dataset","teal_dataset"))
}

cdisc_data <- function(..., check=FALSE, code=NULL) {
  datasets <- list(...)
  names(datasets) <- vapply(datasets, function(d) d$dataname, character(1))
  structure(list(datasets=datasets, check=check, code=code),
            class=c("cdisc_data","teal_data"))
}

# ── relational_data ─────────────────────────────────────────────────────────
dataset <- function(dataname, x, keys=character(0), label=NULL, code=NULL) {
  structure(
    list(dataname=dataname, data=x, keys=keys,
         label=label %||% dataname, code=code),
    class=c("dataset","teal_dataset"))
}

relational_data <- function(..., check=FALSE) {
  datasets <- list(...)
  names(datasets) <- vapply(datasets, function(d) d$dataname, character(1))
  structure(list(datasets=datasets, check=check),
            class=c("relational_data","teal_data"))
}

# ── dataset_connector (simulates deferred/lazy loading) ─────────────────────
dataset_connector <- function(dataname, pull_callable, keys=character(0), label=NULL) {
  structure(
    list(dataname=dataname,
         pull_callable=pull_callable,
         keys=keys,
         label=label %||% dataname,
         .pulled=FALSE,
         .data=NULL),
    class=c("dataset_connector","teal_dataset"))
}

# Simulate pulling data from a connector
pull_dataset_connector <- function(connector, ...) {
  connector$.data   <- connector$pull_callable(...)
  connector$.pulled <- TRUE
  connector
}

cdisc_dataset_connector <- function(dataname, pull_callable, keys=character(0),
                                    label=NULL) {
  structure(
    list(dataname=dataname,
         pull_callable=pull_callable,
         keys=keys,
         label=label %||% dataname,
         .pulled=FALSE,
         .data=NULL),
    class=c("cdisc_dataset_connector","dataset_connector","teal_dataset"))
}

# ── join_keys ────────────────────────────────────────────────────────────────
join_key <- function(dataset_1, dataset_2, keys) {
  structure(list(dataset_1=dataset_1, dataset_2=dataset_2, keys=keys),
            class="join_key")
}

join_keys <- function(...) {
  jk_list <- list(...)
  structure(list(keys=jk_list), class="join_keys")
}

print.join_keys <- function(x, ...) {
  cat("── join_keys ─────────────────────────────────────────\n")
  for (jk in x$keys) {
    cat(sprintf("  %s ↔ %s  [%s]\n",
                jk$dataset_1, jk$dataset_2,
                paste(names(jk$keys), collapse=", ")))
  }
  invisible(x)
}

# ── teal_data() — new-style API (teal >= 0.15) ───────────────────────────────
# In teal >= 0.15, teal_data() replaces cdisc_data() + relational_data()
teal_data <- function(..., code=NULL, join_keys=NULL) {
  datasets <- list(...)
  nms <- sapply(datasets, function(d) {
    if (inherits(d, "teal_dataset")) d$dataname
    else deparse(substitute(d))
  })
  # For datasets passed as named arguments, use the argument names
  call_names <- names(match.call(expand.dots=FALSE)$`...`)
  if (!is.null(call_names) && length(call_names) == length(datasets))
    nms <- call_names
  names(datasets) <- nms
  structure(list(datasets=datasets, code=code, join_keys=join_keys),
            class=c("teal_data","list"))
}

# Convenience: wrap a data frame without explicit cdisc_dataset()
# (new teal_data() API accepts data frames directly)
verify <- function(tdata) {
  cat("── teal_data verification ──────────────────────────\n")
  cat("  Datasets:", paste(names(tdata$datasets), collapse=", "), "\n")
  if (!is.null(tdata$join_keys)) {
    cat("  Join keys defined:", length(tdata$join_keys$keys), "\n")
  }
  cat("  Status: OK\n")
  invisible(tdata)
}

# ── teal mock ────────────────────────────────────────────────────────────────
modules <- function(..., label="root") {
  mods <- list(...)
  structure(list(modules=mods, label=label), class="teal_modules")
}
tab_group <- function(label, ...) {
  structure(list(label=label, children=list(...)),
            class=c("tab_group","teal_module"))
}
tm_data_table        <- function(label="Data Table", ...)    structure(list(label=label,type="data_table"),   class=c("tm_data_table","teal_module"))
tm_variable_browser  <- function(label="Variable Browser",...) structure(list(label=label,type="var_browser"), class=c("tm_variable_browser","teal_module"))
tm_t_summary         <- function(label="Summary",...) structure(list(label=label,type="t_summary"),  class=c("tm_t_summary","teal_module"))
tm_g_scatterplot     <- function(label="Scatter",...)  structure(list(label=label,type="scatter"),    class=c("tm_g_scatterplot","teal_module"))
teal_slice  <- function(dataname,varname,selected=NULL,...) structure(list(dataname=dataname,varname=varname,selected=selected),class="teal_slice")
teal_slices <- function(...,exclude_varnames=NULL) structure(list(slices=list(...),exclude_varnames=exclude_varnames),class="teal_slices")

teal <- function(data, modules, title="teal app", filter=NULL, ...) {
  structure(list(data=data,modules=modules,title=title,filter=filter),
            class=c("teal_app","shiny.appobj"))
}
print.teal_app <- function(x,...) {
  cat("── teal Application ──────────────────────────────────\n")
  cat("  Title:", x$title,"\n")
  ds_names <- if (inherits(x$data,"teal_data")) names(x$data$datasets)
              else if (!is.null(x$data$datasets)) names(x$data$datasets)
              else "?"
  cat("  Datasets:", paste(ds_names, collapse=", "),"\n")
  invisible(x)
}
```

## Overview

This lesson covers the `teal.data` layer in depth — the connectors that load and relate datasets — and the complete deployment workflow for clinical applications. You will learn the difference between static and dynamic data connectors, how to declare join relationships, and how to ship a teal app securely to Posit Connect.

---

## 1 — `teal.data` in Depth

### 1.1 Evolution of the API

teal.data has gone through two API generations:

| Version | Style | Key functions |
|---------|-------|---------------|
| teal < 0.15 | "Old API" | `cdisc_data()`, `cdisc_dataset()`, `relational_data()`, `dataset()` |
| teal ≥ 0.15 | "New API" | `teal_data()` (unified), `join_keys()`, `join_key()` |

Both APIs are in active use in the field. This lesson covers both.

### 1.2 `cdisc_data()` — old API (still common)

```{r old-api-demo, eval=FALSE}
# Old API — explicit cdisc_dataset() wrappers
app_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl,
                keys  = c("STUDYID", "USUBJID"),
                label = "Subject-Level Analysis Dataset"),
  cdisc_dataset("ADAE", adae,
                keys  = c("STUDYID", "USUBJID", "AESEQ"),
                label = "Adverse Events"),
  check = TRUE
)
```

### 1.3 `teal_data()` — new API (teal ≥ 0.15)

```{r new-api-demo, eval=FALSE}
# New API — pass data frames directly, declare keys separately
app_data <- teal_data(
  ADSL = adsl,
  ADAE = adae,
  ADLB = adlb,
  join_keys = join_keys(
    join_key("ADSL", "ADAE", c("STUDYID" = "STUDYID",
                                "USUBJID" = "USUBJID")),
    join_key("ADSL", "ADLB", c("STUDYID" = "STUDYID",
                                "USUBJID" = "USUBJID")),
    join_key("ADAE", "ADLB", c("STUDYID" = "STUDYID",
                                "USUBJID" = "USUBJID"))
  )
)

# Verify integrity
verify(app_data)
```

> **Which API to use?** Use the new `teal_data()` API for new projects — it is cleaner and actively maintained. The old `cdisc_data()` API will remain supported but receives fewer enhancements.

### 1.4 Exercise — new API

Build a `teal_data` object for ADSL, ADAE, and ADVS with the new API.

```{r new-api-ex, exercise=TRUE, exercise.lines=18}
pilot_data <- teal_data(
  ADSL = ___,
  ADAE = ___,
  ADVS = advs,
  join_keys = join_keys(
    join_key("ADSL", "ADAE",
             c("STUDYID" = "STUDYID", ___ = "USUBJID")),
    join_key("ADSL", "ADVS",
             c("STUDYID" = "STUDYID", "USUBJID" = ___))
  )
)

verify(pilot_data)
```

```{r new-api-ex-solution}
pilot_data <- teal_data(
  ADSL = adsl,
  ADAE = adae,
  ADVS = advs,
  join_keys = join_keys(
    join_key("ADSL", "ADAE",
             c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID")),
    join_key("ADSL", "ADVS",
             c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID"))
  )
)

verify(pilot_data)
```

---

## 2 — `relational_data()` — Non-CDISC Datasets

### 2.1 When to use `relational_data()`

`cdisc_data()` is designed for ADaM datasets with CDISC-standard join paths. For other data (site metadata, patient registries, operational data), use `relational_data()`:

```{r relational-demo, eval=FALSE}
# Non-CDISC supplemental datasets
site_info <- tibble::tribble(
  ~SITEID, ~COUNTRY,  ~PI_NAME,
  "001",   "USA",     "Dr. Smith",
  "002",   "Germany", "Dr. Müller",
  "003",   "Japan",   "Dr. Tanaka"
)

protocol_defs <- tibble::tribble(
  ~PARAMCD, ~PARAM_LONG,            ~UNIT,  ~LLOQ, ~ULOQ,
  "ALT",    "Alanine Aminotransferase", "U/L", 7,    56,
  "CREAT",  "Creatinine",            "mg/dL", 0.6,  1.2,
  "WBC",    "White Blood Cells",      "10^9/L", 3.5, 10.5
)

# relational_data for non-ADAM supplemental data
supp_data <- relational_data(
  dataset("SITE",  site_info,      keys = "SITEID"),
  dataset("PARAM", protocol_defs,  keys = "PARAMCD")
)
```

### 2.2 Mixing CDISC and relational data

In practice, a clinical app often needs both types. The new `teal_data()` API handles them uniformly:

```{r mixed-data-demo, eval=FALSE}
mixed_data <- teal_data(
  ADSL  = adsl,
  ADLB  = adlb,
  SITE  = site_info,
  PARAM = protocol_defs,
  join_keys = join_keys(
    join_key("ADSL", "ADLB", c("STUDYID","USUBJID")),
    join_key("ADSL", "SITE", c("SITEID" = "SITEID")),
    join_key("ADLB", "PARAM", c("PARAMCD" = "PARAMCD"))
  )
)
```

### 2.3 Exercise — build a relational dataset

Create a `relational_data` object with two supplemental tables.

```{r relational-ex, exercise=TRUE, exercise.lines=20}
# Supplemental datasets
arm_labels <- tibble(
  ARMCD = c("Pbo", "Xan_Lo", "Xan_Hi"),
  ARM   = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"),
  DOSE_MG = c(0, 54, 81)
)

visit_schedule <- tibble(
  AVISIT = c("Baseline", "Week 2", "Week 4", "Week 8",
             "Week 16", "Week 24", "End of Treatment"),
  AVISITN = c(0, 2, 4, 8, 16, 24, 99)
)

supp <- relational_data(
  dataset("ARM_LABELS",     ___, keys = "ARMCD"),
  dataset("VISIT_SCHEDULE", ___, keys = c("AVISIT", ___))
)

cat("Supplemental data created\n")
cat("Datasets:", paste(names(supp$datasets), collapse=", "), "\n")
```

```{r relational-ex-solution}
arm_labels <- tibble(
  ARMCD = c("Pbo", "Xan_Lo", "Xan_Hi"),
  ARM   = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"),
  DOSE_MG = c(0, 54, 81)
)

visit_schedule <- tibble(
  AVISIT = c("Baseline", "Week 2", "Week 4", "Week 8",
             "Week 16", "Week 24", "End of Treatment"),
  AVISITN = c(0, 2, 4, 8, 16, 24, 99)
)

supp <- relational_data(
  dataset("ARM_LABELS",     arm_labels,     keys = "ARMCD"),
  dataset("VISIT_SCHEDULE", visit_schedule, keys = c("AVISIT", "AVISITN"))
)

cat("Supplemental data created\n")
cat("Datasets:", paste(names(supp$datasets), collapse=", "), "\n")
```

---

## 3 — Data Connectors: Deferred & Dynamic Loading

### 3.1 The problem with static data

Loading all datasets at app startup creates problems in production:

- Large datasets (ADLB can be millions of rows) slow the startup
- Datasets may need to be refreshed from an external source during the session
- Credentials for databases should not be embedded in the app code

**Data connectors** solve this by deferring the data pull until it is actually needed.

### 3.2 `dataset_connector()` / `cdisc_dataset_connector()`

```{r connector-demo, eval=FALSE}
# Define a connector — data is NOT loaded yet
adlb_connector <- cdisc_dataset_connector(
  dataname = "ADLB",
  keys     = c("STUDYID", "USUBJID", "PARAMCD", "ADT"),

  # pull_callable: a function or JoinKeys-aware callable that loads data
  pull_callable = function(...) {
    # Could be a DBI query, haven::read_xpt, or anything
    message("Pulling ADLB from server...")
    pharmaverseadam::adlb |>
      dplyr::filter(is.na(DTYPE))
  }
)

# The data is pulled when teal first needs it:
adlb_connector <- pull_dataset_connector(adlb_connector)
cat("Pulled:", nrow(adlb_connector$.data), "rows\n")
```

### 3.3 Real-world connector patterns

**SAS XPT files from a study folder:**

```{r xpt-connector, eval=FALSE}
adam_path <- Sys.getenv("ADAM_PATH", "/mnt/clinical/study001/adam")

adlb_from_xpt <- cdisc_dataset_connector(
  dataname = "ADLB",
  keys     = c("STUDYID","USUBJID","PARAMCD","ADT"),
  pull_callable = function(...) {
    haven::read_xpt(file.path(adam_path, "adlb.xpt")) |>
      dplyr::filter(is.na(DTYPE))
  }
)
```

**Database via DBI:**

```{r db-connector, eval=FALSE}
adlb_from_db <- cdisc_dataset_connector(
  dataname = "ADLB",
  keys     = c("STUDYID","USUBJID","PARAMCD","ADT"),
  pull_callable = function(...) {
    con <- DBI::dbConnect(
      odbc::odbc(),
      dsn      = Sys.getenv("DB_DSN"),
      uid      = Sys.getenv("DB_USER"),
      pwd      = Sys.getenv("DB_PASS")
    )
    on.exit(DBI::dbDisconnect(con))
    DBI::dbReadTable(con, "ADLB_FILTERED")
  }
)
```

> **Security rule:** Never hard-code credentials in app code. Always use `Sys.getenv()` with environment variables set in the deployment environment.

### 3.4 Exercise — build a file-based connector

Build a connector that simulates loading ADSL from a file path stored in an environment variable. Use `Sys.getenv()` and fall back to `pharmaverseadam::adsl` if the variable is not set.

```{r connector-ex, exercise=TRUE, exercise.lines=20}
adsl_connector <- cdisc_dataset_connector(
  dataname = "ADSL",
  keys     = c("STUDYID", "USUBJID"),
  label    = "Subject-Level Analysis Dataset (from file)",

  pull_callable = function(...) {
    path <- Sys.getenv("ADSL_XPT_PATH", unset = "")

    if (nchar(path) > 0 && file.exists(path)) {
      message("Loading ADSL from: ", path)
      haven::read_xpt(___)
    } else {
      message("ADSL_XPT_PATH not set — using pharmaverseadam fallback")
      ___
    }
  }
)

# Simulate pulling (no env var set → fallback)
adsl_connector <- pull_dataset_connector(adsl_connector)

cat("Pulled ADSL:", nrow(adsl_connector$.data), "rows\n")
cat("Columns:", ncol(adsl_connector$.data), "\n")
```

```{r connector-ex-solution}
adsl_connector <- cdisc_dataset_connector(
  dataname = "ADSL",
  keys     = c("STUDYID", "USUBJID"),
  label    = "Subject-Level Analysis Dataset (from file)",

  pull_callable = function(...) {
    path <- Sys.getenv("ADSL_XPT_PATH", unset = "")

    if (nchar(path) > 0 && file.exists(path)) {
      message("Loading ADSL from: ", path)
      haven::read_xpt(path)
    } else {
      message("ADSL_XPT_PATH not set — using pharmaverseadam fallback")
      pharmaverseadam::adsl
    }
  }
)

adsl_connector <- pull_dataset_connector(adsl_connector)

cat("Pulled ADSL:", nrow(adsl_connector$.data), "rows\n")
cat("Columns:", ncol(adsl_connector$.data), "\n")
```

---

## 4 — `join_keys()` and Dataset Relationships

### 4.1 Why explicit join keys matter

Without declared join keys:
- teal cannot infer how to propagate filters across datasets
- Cross-dataset modules must do their own joins manually
- Subject count in the filter panel is wrong or missing

### 4.2 Declaring CDISC join paths

Standard CDISC join topology for a typical clinical study:

```{r join-keys-demo, eval=FALSE}
adam_join_keys <- join_keys(
  # ADSL is the subject backbone — joins to everything via USUBJID
  join_key("ADSL", "ADAE",  c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID")),
  join_key("ADSL", "ADLB",  c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID")),
  join_key("ADSL", "ADVS",  c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID")),
  join_key("ADSL", "ADTTE", c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID")),

  # ADAE self-key (for within-ADAE filtering)
  join_key("ADAE", "ADAE",  c("STUDYID","USUBJID","AESEQ")),

  # Lab and VS share time-point structure but are not directly joined
  join_key("ADLB", "ADVS",  c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID"))
)

print(adam_join_keys)
```

### 4.3 Self-keys (within-dataset key declarations)

teal uses self-keys (where both `dataset_1` and `dataset_2` are the same name) to understand the primary key of a dataset:

```{r self-keys, eval=FALSE}
# Tell teal that ADLB rows are uniquely identified by these 4 columns
join_key("ADLB", "ADLB",  c("STUDYID", "USUBJID", "PARAMCD", "ADT"))
```

Always declare a self-key for every dataset so the filter panel shows correct subject counts.

### 4.4 Exercise — declare a full join_keys object

Declare join keys for ADSL, ADAE, ADLB, and ADVS including self-keys for each.

```{r join-keys-ex, exercise=TRUE, exercise.lines=22}
study_keys <- join_keys(
  # Self-keys (primary keys)
  join_key("ADSL", "ADSL", c("STUDYID", "USUBJID")),
  join_key("ADAE", "ADAE", c("STUDYID", "USUBJID", ___)),
  join_key("ADLB", "ADLB", c("STUDYID", "USUBJID", "PARAMCD", ___)),
  join_key("ADVS", "ADVS", c("STUDYID", "USUBJID", ___, "ADT")),

  # Cross-dataset links
  join_key("ADSL", "ADAE", c("STUDYID" = "STUDYID", ___ = "USUBJID")),
  join_key("ADSL", "ADLB", c("STUDYID" = "STUDYID", "USUBJID" = ___)),
  join_key("ADSL", "ADVS", c(___ = "STUDYID", "USUBJID" = "USUBJID"))
)

print(study_keys)
```

```{r join-keys-ex-solution}
study_keys <- join_keys(
  join_key("ADSL", "ADSL", c("STUDYID", "USUBJID")),
  join_key("ADAE", "ADAE", c("STUDYID", "USUBJID", "AESEQ")),
  join_key("ADLB", "ADLB", c("STUDYID", "USUBJID", "PARAMCD", "ADT")),
  join_key("ADVS", "ADVS", c("STUDYID", "USUBJID", "PARAMCD", "ADT")),

  join_key("ADSL", "ADAE", c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID")),
  join_key("ADSL", "ADLB", c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID")),
  join_key("ADSL", "ADVS", c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID"))
)

print(study_keys)
```

---

## 5 — Reproducible `code` Argument

### 5.1 Embedded code in `teal_data()`

The new API accepts a `code` argument — R code (as a string or expression) that reproduces the data preparation steps. This is logged by teal.code and available in the downloaded reproducible script.

```{r code-arg-demo, eval=FALSE}
app_data <- teal_data(
  ADSL = adsl,
  ADAE = adae,
  ADLB = adlb,

  # Reproducible code string: what a user would need to run to recreate ADSL
  code = "
    library(pharmaverseadam)
    library(dplyr)

    adsl <- pharmaverseadam::adsl |>
      mutate(
        ITTFL  = SAFFL,
        DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, 'Y', 'N', missing = 'N')
      )

    adae <- pharmaverseadam::adae |>
      rename(TRT01A = TRTA) |>
      mutate(AESEVN = case_when(
        AESEV == 'MILD'     ~ 1L,
        AESEV == 'MODERATE' ~ 2L,
        AESEV == 'SEVERE'   ~ 3L,
        TRUE ~ NA_integer_))

    adlb <- pharmaverseadam::adlb |>
      filter(is.na(DTYPE)) |>
      mutate(ATOXGR = as.integer(ATOXGR),
             BTOXGR = as.integer(BTOXGR))
  "
)
```

> When users click "Show R Code" in the teal app, this preparation code appears at the top of the downloaded script, followed by all filter and analysis steps recorded during the session.

### 5.2 Per-dataset code with `cdisc_dataset()`

In the old API, reproducible code is attached per-dataset:

```{r per-dataset-code, eval=FALSE}
adlb_ds <- cdisc_dataset(
  dataname = "ADLB",
  x        = adlb,
  keys     = c("STUDYID", "USUBJID", "PARAMCD", "ADT"),
  code     = "
    adlb <- pharmaverseadam::adlb |>
      dplyr::filter(is.na(DTYPE)) |>
      dplyr::mutate(ATOXGR = as.integer(ATOXGR),
                    BTOXGR = as.integer(BTOXGR))
  "
)
```

---

## 6 — Deploying to Posit Connect

### 6.1 Deployment checklist

Before deploying a teal app to Posit Connect:

**Code preparation**
- [ ] All data loading uses `Sys.getenv()` for paths and credentials
- [ ] `renv::snapshot()` run to capture all dependencies in `renv.lock`
- [ ] `app.R` as the single entry point in a clean directory
- [ ] No absolute local paths in the code

**Posit Connect configuration**
- [ ] Environment variables set in the Connect Content Settings
- [ ] Correct R version specified in `renv.lock`
- [ ] `rsconnect::deployApp()` or Git-backed deployment configured

**Access control**
- [ ] Viewer permissions set (individuals, groups, or public)
- [ ] Data access audited — does the Posit Connect server have access to the ADaM file path?

### 6.2 A production `app.R`

```{r production-app-r, eval=FALSE}
# ── app.R ── deploy this directory to Posit Connect ────────────────────────
library(shiny)
library(teal)
library(teal.data)
library(teal.modules.general)
library(teal.modules.clinical)
library(dplyr)
library(haven)

# ── Data loading: use env vars for paths ───────────────────────────────────
adam_dir <- Sys.getenv("ADAM_DIR", unset = "")

load_adam <- function(dataset_name) {
  if (nchar(adam_dir) > 0) {
    path <- file.path(adam_dir, paste0(tolower(dataset_name), ".xpt"))
    if (file.exists(path)) return(read_xpt(path))
  }
  # Fallback for development: pharmaverseadam
  message(sprintf("%s: using pharmaverseadam fallback", dataset_name))
  get(tolower(dataset_name), envir = asNamespace("pharmaverseadam"))
}

adsl <- load_adam("ADSL") |>
  mutate(ITTFL  = SAFFL,
         DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"))

adae <- load_adam("ADAE") |>
  { d <- .; if ("TRTA" %in% names(d) && !"TRT01A" %in% names(d)) rename(d, TRT01A=TRTA) else d }() |>
  mutate(AESEVN = case_when(AESEV=="MILD"~1L, AESEV=="MODERATE"~2L,
                             AESEV=="SEVERE"~3L, TRUE~NA_integer_))

adlb <- load_adam("ADLB") |>
  { d <- .; if ("DTYPE" %in% names(d)) filter(d, is.na(DTYPE)) else d }() |>
  mutate(ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))

# ── teal.data ──────────────────────────────────────────────────────────────
app_data <- teal_data(
  ADSL = adsl,
  ADAE = adae,
  ADLB = adlb,
  join_keys = join_keys(
    join_key("ADSL","ADSL", c("STUDYID","USUBJID")),
    join_key("ADAE","ADAE", c("STUDYID","USUBJID","AESEQ")),
    join_key("ADLB","ADLB", c("STUDYID","USUBJID","PARAMCD","ADT")),
    join_key("ADSL","ADAE", c("STUDYID"="STUDYID","USUBJID"="USUBJID")),
    join_key("ADSL","ADLB", c("STUDYID"="STUDYID","USUBJID"="USUBJID"))
  )
)

# ── Modules ────────────────────────────────────────────────────────────────
app_modules <- modules(
  tab_group("Data Explorer",
    tm_data_table("Raw Data"),
    tm_variable_browser("Variables")
  ),
  tab_group("Safety",
    tm_t_summary("Demographics", dataname="ADSL"),
    tm_t_summary("AE Summary",   dataname="ADAE")
  ),
  tab_group("Labs",
    tm_g_scatterplot("Lab Scatter", dataname="ADLB", x="ADY", y="AVAL")
  )
)

# ── App ────────────────────────────────────────────────────────────────────
app <- teal(
  data    = app_data,
  modules = app_modules,
  filter  = teal_slices(
    teal_slice("ADSL", "SAFFL", selected = "Y"),
    exclude_varnames = list(ADSL = c("STUDYID","USUBJID","SUBJID"))
  ),
  title   = "CDISCPILOT01 Clinical Review",
  header  = tags$h4(
    "Xanomeline Study — Internal Review Platform",
    style = "color:#2c3e50; margin:10px 0;"
  ),
  footer  = tags$p(
    sprintf("Data cut: %s | For Internal Use Only",
            Sys.getenv("DATA_CUT_DATE", "2026-01-01")),
    style = "font-size:11px; color:#7f8c8d;"
  )
)

shinyApp(ui = app$ui, server = app$server)
```

### 6.3 `rsconnect` deployment commands

```{r deploy-commands, eval=FALSE}
# Step 1: Snapshot dependencies
renv::snapshot()

# Step 2: Add Posit Connect server (one-time)
rsconnect::addServer(
  url  = "https://connect.company.com",
  name = "company-connect"
)

# Step 3: Deploy
rsconnect::deployApp(
  appDir      = "path/to/app/",
  appName     = "cdiscpilot01-clinical-review",
  server      = "company-connect",
  account     = "htavakoli",
  appFiles    = c("app.R", "renv.lock", "renv/activate.R"),
  forceUpdate = TRUE
)
```

### 6.4 Setting environment variables on Posit Connect

After deployment, go to the content item's settings on Posit Connect:

```
Content Settings → Environment Variables → Add Variable

ADAM_DIR        /mnt/study001/adam
DATA_CUT_DATE   2026-03-15
DB_DSN          clinical_db
DB_USER         rstudio_ro
DB_PASS         <from secrets manager>
```

> **GxP note:** Storing credentials as environment variables (not in code) is required under 21 CFR Part 11 and ICH E6 good programming practice.

### 6.5 Exercise — write a deployment-ready data loader

Write a `load_adam_dataset()` function that:
- Reads an XPT from `Sys.getenv("ADAM_DIR")/{dataset}.xpt` if available
- Falls back to `pharmaverseadam::{dataset}` otherwise
- Converts the dataset name to lowercase for the file path

```{r deploy-loader-ex, exercise=TRUE, exercise.lines=25}
load_adam_dataset <- function(dataname) {
  # dataname: e.g. "ADSL", "ADLB"
  adam_dir <- Sys.getenv(___, unset = "")
  fname    <- paste0(tolower(___), ".xpt")

  if (nchar(adam_dir) > 0) {
    full_path <- file.path(___, fname)
    if (file.exists(full_path)) {
      message(sprintf("Loading %s from %s", dataname, full_path))
      return(haven::read_xpt(full_path))
    }
  }

  # Fallback
  message(sprintf("%s: ADAM_DIR not set or file not found — using pharmaverseadam",
                  dataname))
  ns <- asNamespace("pharmaverseadam")
  if (exists(tolower(dataname), envir = ns, inherits = FALSE)) {
    get(tolower(___), envir = ns)
  } else {
    stop(sprintf("Dataset '%s' not found in pharmaverseadam", dataname))
  }
}

# Test the fallback
adsl_loaded <- load_adam_dataset("adsl")
cat("Rows:", nrow(adsl_loaded), "| Cols:", ncol(adsl_loaded), "\n")
```

```{r deploy-loader-ex-solution}
load_adam_dataset <- function(dataname) {
  adam_dir <- Sys.getenv("ADAM_DIR", unset = "")
  fname    <- paste0(tolower(dataname), ".xpt")

  if (nchar(adam_dir) > 0) {
    full_path <- file.path(adam_dir, fname)
    if (file.exists(full_path)) {
      message(sprintf("Loading %s from %s", dataname, full_path))
      return(haven::read_xpt(full_path))
    }
  }

  message(sprintf("%s: ADAM_DIR not set or file not found — using pharmaverseadam",
                  dataname))
  ns <- asNamespace("pharmaverseadam")
  if (exists(tolower(dataname), envir = ns, inherits = FALSE)) {
    get(tolower(dataname), envir = ns)
  } else {
    stop(sprintf("Dataset '%s' not found in pharmaverseadam", dataname))
  }
}

adsl_loaded <- load_adam_dataset("adsl")
cat("Rows:", nrow(adsl_loaded), "| Cols:", ncol(adsl_loaded), "\n")
```

---

## 7 — renv for teal Applications

### 7.1 Why renv is mandatory for regulated apps

In clinical development, every software component must be version-controlled and reproducible. `renv` does this for R packages:

```{r renv-workflow, eval=FALSE}
# ── One-time project setup ──────────────────────────────────────────────────
renv::init()        # creates renv.lock, renv/ folder, .Rprofile

# Install all teal packages
install.packages(c(
  "teal",
  "teal.data",
  "teal.code",
  "teal.widgets",
  "teal.modules.general",
  "teal.modules.clinical",
  "teal.reporter",
  "pharmaverseadam"
))

# Snapshot after any install
renv::snapshot()

# ── Reproducible restore on new machine / CI ────────────────────────────────
renv::restore()     # installs exact versions from renv.lock
```

### 7.2 `renv.lock` for a teal app — what to include

```{r renv-check, eval=FALSE}
# Check which packages are recorded
renv::status()

# Key teal-family packages that MUST be in renv.lock:
required_pkgs <- c(
  "teal", "teal.data", "teal.code", "teal.widgets", "teal.slice",
  "teal.modules.general", "teal.modules.clinical",
  "shiny", "bslib", "ggplot2", "dplyr",
  "pharmaverseadam", "admiral", "haven"
)

# Verify all are snapshotted
renv_lock <- renv::lockfile_read()
missing <- setdiff(required_pkgs, names(renv_lock$Packages))
if (length(missing) > 0) {
  warning("Not in renv.lock: ", paste(missing, collapse=", "))
} else {
  message("All required packages in renv.lock")
}
```

### 7.3 Docker deployment pattern

```{r docker-pattern, eval=FALSE}
# Dockerfile for a teal app
# FROM rocker/r-ver:4.4.0
#
# RUN apt-get update && apt-get install -y \
#     libssl-dev libcurl4-openssl-dev libxml2-dev
#
# WORKDIR /app
# COPY renv.lock renv.lock
# COPY renv/activate.R renv/activate.R
# RUN Rscript -e "renv::restore()"
#
# COPY . .
# EXPOSE 3838
# CMD ["Rscript", "-e", "shiny::runApp('.', host='0.0.0.0', port=3838)"]
```

### 7.4 Quiz

```{r quiz-deploy, echo=FALSE}
quiz(
  question(
    "Under 21 CFR Part 11 and GxP good practice, where should database credentials be stored?",
    answer("Directly in app.R as R character strings"),
    answer("In a config.R file committed to git"),
    answer("As environment variables set in the deployment environment, never in code", correct = TRUE),
    answer("In a hardcoded list inside the server function"),
    allow_retry = TRUE
  ),
  question(
    "What does `renv::restore()` do when run on a new server?",
    answer("Upgrades all packages to the latest version"),
    answer("Installs the exact package versions recorded in renv.lock, ensuring reproducibility", correct = TRUE),
    answer("Removes all packages and re-installs from CRAN"),
    answer("Validates the R code syntax"),
    allow_retry = TRUE
  ),
  question(
    "In the new teal_data() API, what does the `join_keys()` argument tell teal?",
    answer("The column order for display in data tables"),
    answer("How to join datasets for cross-dataset filter propagation and subject counting", correct = TRUE),
    answer("Which datasets to load first for performance"),
    answer("The dataset labels shown in the filter panel"),
    allow_retry = TRUE
  )
)
```

---

## 8 — Capstone: Production-Ready Data Layer

Build a complete, deployment-ready data layer for a three-dataset clinical app using the new API, with proper join keys, a fallback data loader, and reproducible code.

```{r capstone-data, exercise=TRUE, exercise.eval=FALSE, exercise.lines=60}
# ── 1. Flexible data loader ───────────────────────────────────────────────
load_study_data <- function(dataname) {
  adam_dir <- Sys.getenv("ADAM_DIR", unset = "")
  fname    <- paste0(tolower(___), ".xpt")

  if (nchar(adam_dir) > 0 && file.exists(file.path(adam_dir, fname))) {
    haven::read_xpt(file.path(adam_dir, fname))
  } else {
    get(tolower(dataname), envir = asNamespace("pharmaverseadam"))
  }
}

# ── 2. Load and prepare datasets ─────────────────────────────────────────
adsl_prod <- load_study_data(___) |>
  mutate(ITTFL  = SAFFL,
         DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"))

adae_prod <- load_study_data("adae") |>
  { d <- .; if ("TRTA" %in% names(d) && !"TRT01A" %in% names(d))
      rename(d, TRT01A = TRTA) else d }() |>
  mutate(AESEVN = case_when(AESEV=="MILD"~1L, AESEV=="MODERATE"~2L,
                             AESEV=="SEVERE"~3L, TRUE~NA_integer_))

adlb_prod <- load_study_data(___) |>
  { d <- .; if ("DTYPE" %in% names(d)) filter(d, is.na(DTYPE)) else d }() |>
  mutate(ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))

# ── 3. Build teal_data with full join_keys ────────────────────────────────
study_keys <- join_keys(
  join_key("ADSL", "ADSL", c("STUDYID", "USUBJID")),
  join_key("ADAE", "ADAE", c("STUDYID", "USUBJID", ___)),
  join_key("ADLB", "ADLB", c("STUDYID", "USUBJID", "PARAMCD", ___)),
  join_key("ADSL", "ADAE", c("STUDYID" = "STUDYID", ___ = "USUBJID")),
  join_key("ADSL", "ADLB", c("STUDYID" = "STUDYID", "USUBJID" = ___))
)

prod_data <- teal_data(
  ADSL = ___,
  ADAE = ___,
  ADLB = ___,
  join_keys = ___,
  code = '
    adsl <- pharmaverseadam::adsl |>
      dplyr::mutate(ITTFL=SAFFL,
                    DIABFL=dplyr::if_else(!is.na(BMIBL)&BMIBL>=30,"Y","N",missing="N"))
    adae <- pharmaverseadam::adae |>
      dplyr::rename(TRT01A=TRTA)
    adlb <- pharmaverseadam::adlb |>
      dplyr::filter(is.na(DTYPE)) |>
      dplyr::mutate(ATOXGR=as.integer(ATOXGR), BTOXGR=as.integer(BTOXGR))
  '
)

verify(prod_data)
```

```{r capstone-data-solution}
load_study_data <- function(dataname) {
  adam_dir <- Sys.getenv("ADAM_DIR", unset = "")
  fname    <- paste0(tolower(dataname), ".xpt")

  if (nchar(adam_dir) > 0 && file.exists(file.path(adam_dir, fname))) {
    haven::read_xpt(file.path(adam_dir, fname))
  } else {
    get(tolower(dataname), envir = asNamespace("pharmaverseadam"))
  }
}

adsl_prod <- load_study_data("adsl") |>
  mutate(ITTFL  = SAFFL,
         DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"))

adae_prod <- load_study_data("adae") |>
  { d <- .; if ("TRTA" %in% names(d) && !"TRT01A" %in% names(d))
      rename(d, TRT01A = TRTA) else d }() |>
  mutate(AESEVN = case_when(AESEV=="MILD"~1L, AESEV=="MODERATE"~2L,
                             AESEV=="SEVERE"~3L, TRUE~NA_integer_))

adlb_prod <- load_study_data("adlb") |>
  { d <- .; if ("DTYPE" %in% names(d)) filter(d, is.na(DTYPE)) else d }() |>
  mutate(ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))

study_keys <- join_keys(
  join_key("ADSL", "ADSL", c("STUDYID", "USUBJID")),
  join_key("ADAE", "ADAE", c("STUDYID", "USUBJID", "AESEQ")),
  join_key("ADLB", "ADLB", c("STUDYID", "USUBJID", "PARAMCD", "ADT")),
  join_key("ADSL", "ADAE", c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID")),
  join_key("ADSL", "ADLB", c("STUDYID" = "STUDYID", "USUBJID" = "USUBJID"))
)

prod_data <- teal_data(
  ADSL = adsl_prod,
  ADAE = adae_prod,
  ADLB = adlb_prod,
  join_keys = study_keys,
  code = '
    adsl <- pharmaverseadam::adsl |>
      dplyr::mutate(ITTFL=SAFFL,
                    DIABFL=dplyr::if_else(!is.na(BMIBL)&BMIBL>=30,"Y","N",missing="N"))
    adae <- pharmaverseadam::adae |>
      dplyr::rename(TRT01A=TRTA)
    adlb <- pharmaverseadam::adlb |>
      dplyr::filter(is.na(DTYPE)) |>
      dplyr::mutate(ATOXGR=as.integer(ATOXGR), BTOXGR=as.integer(BTOXGR))
  '
)

verify(prod_data)
```

---

## Summary

You now have a complete picture of `teal.data` and deployment:

- **Old API** (`cdisc_data()` / `cdisc_dataset()`) and **new API** (`teal_data()` / `join_keys()`) both work; prefer the new API for new projects
- **`relational_data()`** handles non-CDISC supplemental datasets
- **Data connectors** (`cdisc_dataset_connector()`) defer loading to runtime and support any pull source — files, databases, APIs
- **`join_keys()`** declares the full join topology so teal can propagate filters correctly
- The `code` argument embeds reproducible preparation steps into the downloaded R script
- **Deployment** requires `Sys.getenv()` for all credentials, `renv::snapshot()` for dependency locking, and environment variables set on Posit Connect
- **Docker** provides fully reproducible, OS-level isolation for regulated deployment

Next: **Lesson 42** is the teal capstone — building a complete, multi-module CDISCPILOT01 clinical data review application with custom modules, full deployment configuration, and end-to-end testing.
