---
title: "Lesson 42 — teal Capstone: Full Clinical Data Review App"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Build a complete, production-quality CDISCPILOT01 clinical data review
  application from scratch using teal. Covers custom modules, teal.data
  connectors, filter panel configuration, teal.reporter integration,
  deployment packaging, and end-to-end testing patterns.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 42_teal_capstone.Rmd
# PURPOSE       : teal Capstone — full CDISCPILOT01 clinical data review app with
#                 4 custom modules (BMI scatter, AE severity, lab over
#                 time, shift table), unit tests, and debug challenges.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================
library(learnr)
library(dplyr)
library(ggplot2)
library(tibble)
library(tidyr)
library(shiny)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 120, exercise.eval = FALSE)

# ── Pharmaverseadam datasets ────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}
adsl <- adsl_raw |> mutate(
  ITTFL  = SAFFL,
  DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"))

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw)) {
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
}
adae <- adae_raw |> mutate(AESEVN = case_when(
  AESEV == "MILD"     ~ 1L, AESEV == "MODERATE" ~ 2L,
  AESEV == "SEVERE"   ~ 3L, TRUE ~ NA_integer_))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

advs <- pharmaverseadam::advs

# ── Full teal mock infrastructure ────────────────────────────────────────────
`%||%` <- function(a, b) if (is.null(a)) b else a

# teal.data
cdisc_dataset <- function(dataname, x, keys=character(0), label=NULL, code=NULL) {
  structure(list(dataname=dataname,data=x,keys=keys,
                 label=label%||%dataname,code=code),
            class=c("cdisc_dataset","teal_dataset"))
}
cdisc_data <- function(...,check=FALSE,code=NULL) {
  datasets <- list(...)
  names(datasets) <- vapply(datasets,function(d)d$dataname,character(1))
  structure(list(datasets=datasets,check=check,code=code),
            class=c("cdisc_data","teal_data"))
}
join_key <- function(d1,d2,keys)
  structure(list(dataset_1=d1,dataset_2=d2,keys=keys),class="join_key")
join_keys <- function(...) structure(list(keys=list(...)),class="join_keys")
teal_data <- function(...,code=NULL,join_keys=NULL) {
  datasets <- list(...)
  call_names <- tryCatch(names(match.call(expand.dots=FALSE)$`...`),error=function(e)NULL)
  if (!is.null(call_names) && length(call_names)==length(datasets))
    names(datasets) <- call_names
  structure(list(datasets=datasets,code=code,join_keys=join_keys),
            class=c("teal_data","list"))
}
verify <- function(tdata) {
  cat("── teal_data verified ─────────────────────────────────\n")
  cat("  Datasets:", paste(names(tdata$datasets),collapse=", "),"\n")
  invisible(tdata)
}

# teal modules
teal_slice  <- function(dataname,varname,selected=NULL,keep_na=FALSE,...)
  structure(list(dataname=dataname,varname=varname,selected=selected,keep_na=keep_na),class="teal_slice")
teal_slices <- function(...,exclude_varnames=NULL)
  structure(list(slices=list(...),exclude_varnames=exclude_varnames),class="teal_slices")
modules <- function(...,label="root") {
  mods <- list(...)
  structure(list(modules=mods,label=label),class="teal_modules")
}
tab_group <- function(label,...) structure(list(label=label,children=list(...)),class=c("tab_group","teal_module"))
tm_data_table       <- function(label="Data Table",...)      structure(list(label=label,type="data_table"),class=c("tm_data_table","teal_module"))
tm_variable_browser <- function(label="Variable Browser",...) structure(list(label=label,type="var_browser"),class=c("tm_variable_browser","teal_module"))
tm_t_summary        <- function(label="Summary",...)          structure(list(label=label,type="t_summary"),class=c("tm_t_summary","teal_module"))
tm_t_events         <- function(label="AE Events",...)        structure(list(label=label,type="t_events"),class=c("tm_t_events","teal_module"))
tm_t_shift_by_grade <- function(label="Shift Table",...)      structure(list(label=label,type="t_shift"),class=c("tm_t_shift","teal_module"))
tm_g_km             <- function(label="Kaplan-Meier",...)     structure(list(label=label,type="km"),class=c("tm_g_km","teal_module"))
tm_g_scatterplot    <- function(label="Scatter",...)          structure(list(label=label,type="scatter"),class=c("tm_g_scatter","teal_module"))
tm_missing_data     <- function(label="Missing Data",...)     structure(list(label=label,type="missing"),class=c("tm_missing_data","teal_module"))

# Custom module constructor
teal_module <- function(label,ui,server,datanames="all",
                        ui_args=list(),server_args=list(),...) {
  structure(list(label=label,ui=ui,server=server,datanames=datanames,
                 ui_args=ui_args,server_args=server_args),
            class=c("teal_module_custom","teal_module"))
}

# teal app
teal <- function(data,modules,title="teal app",filter=NULL,header=NULL,footer=NULL,...) {
  structure(list(data=data,modules=modules,title=title,
                 filter=filter,header=header,footer=footer),
            class=c("teal_app","shiny.appobj"))
}
print.teal_app <- function(x,...) {
  cat("╔══════════════════════════════════════════════════════╗\n")
  cat("║  teal Application:", x$title, "\n")
  cat("╠══════════════════════════════════════════════════════╣\n")
  ds <- if (!is.null(x$data$datasets)) names(x$data$datasets) else "?"
  cat("║  Datasets:", paste(ds,collapse=", "),"\n")
  mods <- if (inherits(x$modules,"teal_modules")) x$modules$modules else list()
  cat("║  Modules:\n")
  for (m in mods) {
    if (inherits(m,"tab_group")) {
      cat("║    [group]",m$label,"\n")
      for (c in m$children) cat("║      •",c$label,"\n")
    } else {
      cat("║    •",m$label,"\n")
    }
  }
  if (!is.null(x$filter)) cat("║  Pre-filters:",length(x$filter$slices),"slice(s)\n")
  cat("╚══════════════════════════════════════════════════════╝\n")
  invisible(x)
}

# qenv mock
qenv <- function() {
  env <- new.env(parent=baseenv())
  env$.log <- character(0)
  env$eval <- function(expr_str) {
    env$.log <- c(env$.log, expr_str)
    eval(parse(text=expr_str), envir=env)
    invisible(env)
  }
  env$get_code <- function() paste(env$.log, collapse="\n")
  class(env) <- c("qenv","environment")
  env
}
```

## Overview

This capstone brings together every concept from Lessons 39–41 into a single, production-quality CDISCPILOT01 clinical data review application. You will build all layers from scratch: the data connector, join keys, three custom modules, the filter panel, and the deployment package.

The finished app will have four tab groups covering demographics, safety, lab data, and a patient profile, using both ready-made modules from `teal.modules.clinical` and custom modules written to the `teal_module()` API.

---

## 1 — App Design

### 1.1 Scope

| Tab Group | Modules | Datasets |
|-----------|---------|---------|
| Explorer | Data Table, Variable Browser | ADSL, ADAE, ADLB |
| Demographics | Summary Table (ADSL), BMI scatter | ADSL |
| Safety | AE events table, AE severity bar chart | ADSL, ADAE |
| Lab Data | Lab by visit (line), Lab toxicity shift table | ADSL, ADLB |

### 1.2 App file structure

```
cdiscpilot_review/
├── app.R                  ← main entry point
├── R/
│   ├── data_loader.R      ← load_study_data()
│   ├── tm_ae_severity.R   ← custom AE severity module
│   ├── tm_lab_line.R      ← custom lab mean-over-time module
│   └── tm_bmi_scatter.R   ← custom BMI scatter module
├── renv.lock
└── renv/activate.R
```

In this lesson we write everything in-line for clarity; in production, split into the `R/` files shown.

---

## 2 — Data Layer

### 2.1 Build the production data connector

```{r data-layer-ex, exercise=TRUE, exercise.lines=50}
# ── Load & prepare ──────────────────────────────────────────────────────────
load_study_data <- function(dataname) {
  adam_dir <- Sys.getenv("ADAM_DIR", unset = "")
  fname    <- paste0(tolower(dataname), ".xpt")
  full     <- file.path(adam_dir, fname)

  if (nchar(adam_dir) > 0 && file.exists(full)) {
    haven::read_xpt(full)
  } else {
    get(tolower(dataname), envir = asNamespace("pharmaverseadam"))
  }
}

adsl_app <- load_study_data("adsl") |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N")
  )

adae_app <- load_study_data("adae") |>
  { d <- .
    if ("TRTA" %in% names(d) && !"TRT01A" %in% names(d)) rename(d, TRT01A=TRTA)
    else d
  }() |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE                ~ NA_integer_
  ))

adlb_app <- load_study_data("adlb") |>
  { d <- .; if ("DTYPE" %in% names(d)) filter(d, is.na(DTYPE)) else d }() |>
  mutate(ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))

# ── teal_data with join keys ─────────────────────────────────────────────────
app_data <- teal_data(
  ADSL = adsl_app,
  ADAE = adae_app,
  ADLB = adlb_app,
  join_keys = join_keys(
    join_key("ADSL","ADSL", c("STUDYID","USUBJID")),
    join_key("ADAE","ADAE", c("STUDYID","USUBJID","AESEQ")),
    join_key("ADLB","ADLB", c("STUDYID","USUBJID","PARAMCD","ADT")),
    join_key("ADSL","ADAE", c("STUDYID"="STUDYID","USUBJID"="USUBJID")),
    join_key("ADSL","ADLB", c("STUDYID"="STUDYID","USUBJID"="USUBJID"))
  )
)

verify(app_data)
```

---

## 3 — Custom Module 1: BMI Scatter

This module shows a BMI vs. Age scatter plot coloured by treatment arm.

```{r bmi-module, exercise=TRUE, exercise.lines=40}
tm_bmi_scatter <- function(label    = "BMI vs Age",
                            dataname = "ADSL") {
  teal_module(
    label      = label,
    datanames  = dataname,
    server_args = list(dataname = dataname),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        fluidRow(
          column(4, checkboxGroupInput(ns("arms"), "Treatment Arms",
                                       choices = c("Placebo",
                                                   "Xanomeline Low Dose",
                                                   "Xanomeline High Dose"),
                                       selected = c("Placebo",
                                                    "Xanomeline Low Dose",
                                                    "Xanomeline High Dose")))
        ),
        plotOutput(ns("bmi_plot"), height = "450px")
      )
    },

    server = function(id, data, reporter, dataname, ...) {
      moduleServer(id, function(input, output, session) {

        adsl_r <- reactive(data()[[dataname]])

        output$bmi_plot <- renderPlot({
          req(adsl_r(), input$arms)
          d <- adsl_r() |>
            filter(TRT01P %in% input$arms, !is.na(BMIBL), !is.na(AGE))

          if (nrow(d) == 0) {
            plot.new()
            text(0.5, 0.5, "No data for selected arms", cex=1.4, col="grey50")
            return(invisible(NULL))
          }

          ggplot(d, aes(x=AGE, y=BMIBL, colour=TRT01P)) +
            geom_point(alpha=0.7, size=2.5) +
            geom_smooth(method="lm", se=TRUE, linewidth=1, alpha=0.2) +
            labs(title="BMI vs Age by Treatment Arm",
                 x="Age (years)", y="BMI at Baseline (kg/m²)",
                 colour="Treatment") +
            theme_minimal(base_size=14) +
            theme(legend.position="bottom")
        })
      })
    }
  )
}

mod <- tm_bmi_scatter()
cat("Module:", mod$label, "| Datanames:", mod$datanames, "\n")
```

---

## 4 — Custom Module 2: AE Severity Bar

```{r ae-module, exercise=TRUE, exercise.lines=50}
tm_ae_severity <- function(label   = "AE Severity Distribution",
                            arm_var = "TRT01A") {
  teal_module(
    label       = label,
    datanames   = c("ADSL","ADAE"),
    server_args = list(arm_var = arm_var),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        fluidRow(
          column(4,
            selectInput(ns("soc"), "System Organ Class",
                        choices = c("All SOCs" = "ALL"), selected = "ALL")
          ),
          column(4,
            radioButtons(ns("count_type"), "Count by",
                         choices = c("Number of AEs"     = "ae",
                                     "Number of Subjects" = "subj"),
                         selected = "ae")
          )
        ),
        plotOutput(ns("sev_plot"), height = "450px")
      )
    },

    server = function(id, data, reporter, arm_var, ...) {
      moduleServer(id, function(input, output, session) {

        adsl_r <- reactive(data()[["ADSL"]])
        adae_r <- reactive(data()[["ADAE"]])

        # Populate SOC choices
        observe({
          req(adae_r())
          socs <- c("All SOCs" = "ALL", sort(unique(adae_r()$AESOC)))
          updateSelectInput(session, "soc", choices = socs)
        })

        plot_data <- reactive({
          req(adae_r(), adsl_r())
          d <- adae_r() |>
            select(-any_of("TRT01P")) |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by="USUBJID")
          if (input$soc != "ALL")
            d <- filter(d, AESOC == input$soc)
          if (input$count_type == "subj")
            d <- distinct(d, USUBJID, TRT01P, AESEV)
          d
        })

        output$sev_plot <- renderPlot({
          req(plot_data())
          d <- plot_data()

          if (nrow(d) == 0) {
            plot.new()
            text(0.5, 0.5, "No AEs for selected filters", cex=1.4, col="grey50")
            return(invisible(NULL))
          }

          y_label <- if (input$count_type == "ae") "Number of AEs"
                     else "Number of Subjects with AE"

          d |>
            count(TRT01P, AESEV) |>
            mutate(AESEV = factor(AESEV,
                                  levels = c("MILD","MODERATE","SEVERE"))) |>
            ggplot(aes(x=TRT01P, y=n, fill=AESEV)) +
            geom_col(position="dodge") +
            scale_fill_manual(values=c(MILD="#3498db", MODERATE="#e67e22",
                                       SEVERE="#e74c3c")) +
            labs(title = if (input$soc == "ALL") "AE Severity — All SOCs"
                         else paste("AE Severity —", input$soc),
                 x = NULL, y = y_label, fill = "Severity") +
            theme_minimal(base_size=14) +
            theme(axis.text.x = element_text(angle=30, hjust=1))
        })
      })
    }
  )
}

mod <- tm_ae_severity()
cat("Module:", mod$label, "\n")
```

---

## 5 — Custom Module 3: Lab Mean-Over-Time

```{r lab-module, exercise=TRUE, exercise.lines=55}
tm_lab_over_time <- function(label          = "Lab Values Over Time",
                              default_param  = "ALT") {
  teal_module(
    label       = label,
    datanames   = c("ADSL","ADLB"),
    server_args = list(default_param = default_param),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        fluidRow(
          column(3, selectInput(ns("param"), "Parameter", choices=NULL)),
          column(3, checkboxInput(ns("show_ci"), "Show 95% CI", value=TRUE)),
          column(3, checkboxInput(ns("log_y"),   "Log Y-axis",  value=FALSE))
        ),
        plotOutput(ns("lab_plot"), height="450px"),
        hr(),
        h5("Summary by Visit"),
        tableOutput(ns("visit_table"))
      )
    },

    server = function(id, data, reporter, default_param, ...) {
      moduleServer(id, function(input, output, session) {

        adsl_r <- reactive(data()[["ADSL"]])
        adlb_r <- reactive(data()[["ADLB"]])

        observe({
          req(adlb_r())
          params <- sort(unique(adlb_r()$PARAMCD))
          updateSelectInput(session, "param", choices=params,
                            selected=default_param)
        })

        merged <- reactive({
          req(adsl_r(), adlb_r(), input$param)
          adlb_r() |>
            filter(PARAMCD == input$param, !is.na(AVAL), !is.na(AVISITN)) |>
            filter(AVISITN < 900) |>        # exclude sentinel visits
            select(-any_of("TRT01P")) |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by="USUBJID") |>
            filter(!is.na(TRT01P))
        })

        summary_data <- reactive({
          req(merged())
          merged() |>
            group_by(TRT01P, AVISIT, AVISITN) |>
            summarise(
              N    = n(),
              Mean = mean(AVAL, na.rm=TRUE),
              SD   = sd(AVAL,   na.rm=TRUE),
              SE   = SD / sqrt(N),
              LCI  = Mean - 1.96 * SE,
              UCI  = Mean + 1.96 * SE,
              .groups = "drop"
            ) |>
            arrange(TRT01P, AVISITN)
        })

        output$lab_plot <- renderPlot({
          req(summary_data())
          d <- summary_data()

          if (nrow(d) == 0) {
            plot.new()
            text(0.5, 0.5, "No data after filtering", cex=1.4, col="grey50")
            return(invisible(NULL))
          }

          # Order visits by AVISITN
          visit_order <- d |> distinct(AVISIT, AVISITN) |> arrange(AVISITN) |> pull(AVISIT)
          d <- mutate(d, AVISIT = factor(AVISIT, levels=visit_order))

          p <- ggplot(d, aes(x=AVISIT, y=Mean, colour=TRT01P, group=TRT01P)) +
            geom_line(linewidth=1) +
            geom_point(size=3) +
            labs(title = paste("Mean", input$param, "over Time"),
                 subtitle = "Vertical bars = 95% CI",
                 x="Visit", y=paste(input$param, "[mean]"),
                 colour="Treatment") +
            theme_minimal(base_size=14) +
            theme(axis.text.x = element_text(angle=45, hjust=1),
                  legend.position="bottom")

          if (input$show_ci)
            p <- p + geom_errorbar(aes(ymin=LCI, ymax=UCI),
                                   width=0.3, alpha=0.6)
          if (input$log_y)
            p <- p + scale_y_log10()

          p
        })

        output$visit_table <- renderTable({
          req(summary_data())
          summary_data() |>
            mutate(across(c(Mean,SD,SE,LCI,UCI), ~round(.x,2))) |>
            rename(`Treatment` = TRT01P, `Visit` = AVISIT,
                   `Mean`=Mean, `SD`=SD, `95% LCI`=LCI, `95% UCI`=UCI) |>
            select(Treatment, Visit, N, Mean, SD, `95% LCI`, `95% UCI`)
        }, striped=TRUE, hover=TRUE, bordered=TRUE)
      })
    }
  )
}

mod <- tm_lab_over_time()
cat("Module:", mod$label, "\n")
```

---

## 6 — Custom Module 4: Lab Toxicity Shift Table

```{r shift-module, exercise=TRUE, exercise.lines=40}
tm_lab_shift_table <- function(label         = "Lab Toxicity Shift",
                                default_param = "ALT") {
  teal_module(
    label       = label,
    datanames   = "ADLB",
    server_args = list(default_param = default_param),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        fluidRow(
          column(3, selectInput(ns("param"), "Parameter", choices=NULL)),
          column(3, selectInput(ns("arm"),   "Treatment Arm",
                                choices=NULL))
        ),
        h5("Baseline Grade → Post-Baseline Grade Shift"),
        tableOutput(ns("shift_tbl"))
      )
    },

    server = function(id, data, reporter, default_param, ...) {
      moduleServer(id, function(input, output, session) {

        adlb_r <- reactive(data()[["ADLB"]])

        observe({
          req(adlb_r())
          params <- sort(unique(adlb_r()$PARAMCD))
          arms   <- sort(unique(adlb_r()$TRT01P))
          updateSelectInput(session, "param", choices=params,
                            selected=default_param)
          updateSelectInput(session, "arm",   choices=arms,
                            selected=arms[1])
        })

        output$shift_tbl <- renderTable({
          req(adlb_r(), input$param, input$arm)
          adlb_r() |>
            filter(PARAMCD == input$param,
                   TRT01P  == input$arm,
                   !is.na(BTOXGR), !is.na(ATOXGR)) |>
            count(BTOXGR, ATOXGR) |>
            mutate(
              BTOXGR = paste("Baseline Grade", BTOXGR),
              ATOXGR = paste("Post Grade", ATOXGR)
            ) |>
            pivot_wider(names_from=ATOXGR, values_from=n, values_fill=0L) |>
            rename(`Baseline / Post-Baseline` = BTOXGR)
        }, striped=TRUE, hover=TRUE, bordered=TRUE)
      })
    }
  )
}

mod <- tm_lab_shift_table()
cat("Module:", mod$label, "\n")
```

---

## 7 — Assemble the Full App

Now wire everything together into the complete application.

```{r full-app, exercise=TRUE, exercise.eval=FALSE, exercise.lines=80}
# ── Data ──────────────────────────────────────────────────────────────────────
full_app_data <- teal_data(
  ADSL = adsl,
  ADAE = adae,
  ADLB = adlb,
  join_keys = join_keys(
    join_key("ADSL","ADSL", c("STUDYID","USUBJID")),
    join_key("ADAE","ADAE", c("STUDYID","USUBJID","AESEQ")),
    join_key("ADLB","ADLB", c("STUDYID","USUBJID","PARAMCD","ADT")),
    join_key("ADSL","ADAE", c("STUDYID"="STUDYID","USUBJID"="USUBJID")),
    join_key("ADSL","ADLB", c("STUDYID"="STUDYID","USUBJID"="USUBJID"))
  )
)

# ── Modules ───────────────────────────────────────────────────────────────────
full_app_modules <- modules(

  # Data Explorer
  tab_group("Data Explorer",
    tm_data_table("Raw Data Tables"),
    tm_variable_browser("Variable Distributions"),
    tm_missing_data("Missingness Report")
  ),

  # Demographics
  tab_group("Demographics",
    tm_t_summary("Summary Table",
                 dataname="ADSL"),
    tm_bmi_scatter("BMI vs Age")
  ),

  # Safety
  tab_group("Safety",
    tm_t_events("AE Frequency Table",
                dataname="ADAE"),
    tm_ae_severity("AE Severity Distribution")
  ),

  # Lab Data
  tab_group("Lab Data",
    tm_lab_over_time("Lab Values Over Time",    default_param="ALT"),
    tm_lab_over_time("Creatinine Over Time",    default_param="CREAT"),
    tm_lab_shift_table("Toxicity Shift — ALT",  default_param="ALT"),
    tm_lab_shift_table("Toxicity Shift — CREAT",default_param="CREAT")
  )
)

# ── Filters ───────────────────────────────────────────────────────────────────
full_app_filters <- teal_slices(
  teal_slice("ADSL", "SAFFL",   selected = "Y"),
  teal_slice("ADLB", "PARAMCD", selected = c("ALT","CREAT","WBC","CA")),
  exclude_varnames = list(
    ADSL = c("STUDYID","USUBJID","SUBJID","RFSTDTC","RFENDTC"),
    ADAE = c("STUDYID","USUBJID"),
    ADLB = c("STUDYID","USUBJID")
  )
)

# ── Build app ─────────────────────────────────────────────────────────────────
full_app <- teal(
  data    = full_app_data,
  modules = full_app_modules,
  filter  = full_app_filters,
  title   = "CDISCPILOT01 — Xanomeline Clinical Data Review",
  header  = tags$div(
    style = "background:#2c3e50; padding:8px 16px; color:white;",
    tags$strong("CDISCPILOT01 — Xanomeline Study "),
    tags$span("| Internal Use Only",
              style="font-size:12px; color:#bdc3c7;")
  ),
  footer  = tags$p(
    sprintf("Pharmaverse tutorial | Data: pharmaverseadam | %s",
            format(Sys.Date(), "%Y-%m-%d")),
    style="font-size:11px; color:#7f8c8d; text-align:center; padding:4px;"
  )
)

print(full_app)
```

---

## 8 — Deployment Package

### 8.1 Complete `app.R` for production

```{r prod-app-r, eval=FALSE}
# ── app.R — CDISCPILOT01 Clinical Data Review ─────────────────────────────
# Deploy to Posit Connect via rsconnect::deployApp()

library(shiny)
library(teal)
library(teal.data)
library(teal.modules.general)
library(teal.modules.clinical)
library(dplyr)
library(ggplot2)
library(tidyr)

# Source custom modules
source("R/data_loader.R")
source("R/tm_bmi_scatter.R")
source("R/tm_ae_severity.R")
source("R/tm_lab_over_time.R")
source("R/tm_lab_shift_table.R")

# ── Load data ─────────────────────────────────────────────────────────────
adsl <- load_study_data("adsl") |>
  mutate(ITTFL=SAFFL,
         DIABFL=if_else(!is.na(BMIBL)&BMIBL>=30,"Y","N",missing="N"))
adae <- load_study_data("adae") |>
  (\(d) if("TRTA"%in%names(d)&&!"TRT01A"%in%names(d)) rename(d,TRT01A=TRTA) else d)() |>
  mutate(AESEVN=case_when(AESEV=="MILD"~1L,AESEV=="MODERATE"~2L,
                           AESEV=="SEVERE"~3L,TRUE~NA_integer_))
adlb <- load_study_data("adlb") |>
  (\(d) if("DTYPE"%in%names(d)) filter(d,is.na(DTYPE)) else d)() |>
  mutate(ATOXGR=as.integer(ATOXGR),BTOXGR=as.integer(BTOXGR))

# ── teal_data ─────────────────────────────────────────────────────────────
app_data <- teal_data(
  ADSL=adsl, ADAE=adae, ADLB=adlb,
  join_keys=join_keys(
    join_key("ADSL","ADSL",c("STUDYID","USUBJID")),
    join_key("ADAE","ADAE",c("STUDYID","USUBJID","AESEQ")),
    join_key("ADLB","ADLB",c("STUDYID","USUBJID","PARAMCD","ADT")),
    join_key("ADSL","ADAE",c("STUDYID"="STUDYID","USUBJID"="USUBJID")),
    join_key("ADSL","ADLB",c("STUDYID"="STUDYID","USUBJID"="USUBJID"))
  )
)

# ── Modules ───────────────────────────────────────────────────────────────
app <- teal(
  data    = app_data,
  modules = modules(
    tab_group("Data Explorer",
      tm_data_table("Raw Data"),
      tm_variable_browser("Variables"),
      tm_missing_data("Missingness")
    ),
    tab_group("Demographics",
      tm_t_summary("Summary Table", dataname="ADSL"),
      tm_bmi_scatter("BMI vs Age")
    ),
    tab_group("Safety",
      tm_t_events("AE Frequency", dataname="ADAE"),
      tm_ae_severity("AE Severity")
    ),
    tab_group("Lab Data",
      tm_lab_over_time("Lab Values Over Time"),
      tm_lab_shift_table("Toxicity Shift")
    )
  ),
  filter=teal_slices(
    teal_slice("ADSL","SAFFL",selected="Y"),
    exclude_varnames=list(
      ADSL=c("STUDYID","USUBJID"),
      ADAE=c("STUDYID","USUBJID"),
      ADLB=c("STUDYID","USUBJID")
    )
  ),
  title="CDISCPILOT01 Clinical Review"
)

shinyApp(ui=app$ui, server=app$server)
```

### 8.2 Deploy commands

```{r deploy-ex, eval=FALSE}
# 1. Lock dependencies
renv::snapshot()

# 2. Deploy to Posit Connect
rsconnect::deployApp(
  appDir      = "cdiscpilot_review/",
  appName     = "cdiscpilot01-review",
  server      = "connect.company.com",
  account     = "htavakoli",
  appFiles    = c("app.R", "R/", "renv.lock", "renv/activate.R"),
  forceUpdate = TRUE
)
```

---

## 9 — End-to-End Testing

### 9.1 Module unit tests with `testServer()`

```{r e2e-tests, exercise=TRUE, exercise.lines=45}
# Test custom BMI scatter module
bmi_mod <- tm_bmi_scatter()

# Simulate teal's filtered data delivery
mock_data_fn <- function() list(ADSL = adsl)

testServer(
  bmi_mod$server,
  args = list(
    data     = mock_data_fn,
    reporter = list(),
    dataname = "ADSL"
  ),
  {
    # Check the output renders without error
    session$setInputs(arms = c("Placebo", "Xanomeline High Dose"))
    expect_silent(output$bmi_plot)
    cat("✓ BMI scatter renders with selected arms\n")

    # Test empty-data guard
    session$setInputs(arms = character(0))
    expect_silent(output$bmi_plot)
    cat("✓ BMI scatter handles empty arm selection\n")
  }
)
```

### 9.2 Lab module tests

```{r lab-module-test, exercise=TRUE, exercise.lines=35}
# Test lab over time module
lab_mod <- tm_lab_over_time(default_param = "ALT")

mock_data_fn <- function() list(ADSL = adsl, ADLB = adlb)

testServer(
  lab_mod$server,
  args = list(
    data          = mock_data_fn,
    reporter      = list(),
    default_param = "ALT"
  ),
  {
    session$setInputs(param="ALT", show_ci=TRUE, log_y=FALSE)

    # Plot renders
    expect_silent(output$lab_plot)
    cat("✓ Lab plot renders for ALT\n")

    # Summary table renders
    expect_silent(output$visit_table)
    cat("✓ Visit summary table renders\n")

    # Switch to log scale
    session$setInputs(log_y = TRUE)
    expect_silent(output$lab_plot)
    cat("✓ Lab plot renders on log scale\n")
  }
)
```

### 9.3 Data integrity test

```{r data-integrity-test, exercise=TRUE, exercise.lines=30}
# Verify no .x/.y column collisions in merged data
test_join_integrity <- function(adsl_df, adlb_df, param="ALT") {

  merged <- adlb_df |>
    filter(PARAMCD == param) |>
    select(-any_of("TRT01P")) |>
    left_join(select(adsl_df, USUBJID, TRT01P), by="USUBJID")

  # No duplicate column names
  dups <- names(merged)[duplicated(names(merged))]
  if (length(dups) > 0) {
    stop(sprintf("Duplicate columns after join: %s", paste(dups, collapse=", ")))
  }

  # TRT01P exists and is not all NA
  if (!"TRT01P" %in% names(merged)) stop("TRT01P missing after join")
  na_trt <- mean(is.na(merged$TRT01P))
  if (na_trt > 0.5) stop(sprintf("%.0f%% NA in TRT01P — join may have failed", na_trt*100))

  cat(sprintf("✓ ALT data: %d rows, %d cols, %.0f%% matched to TRT01P\n",
              nrow(merged), ncol(merged), 100*(1-na_trt)))
  invisible(merged)
}

test_join_integrity(adsl, adlb)
```

---

## 10 — Final Challenge: Fix the Module

The module below has three bugs. Identify and fix them all.

```{r debug-challenge, exercise=TRUE, exercise.lines=45}
# BUGGY MODULE — find and fix 3 bugs
tm_ae_count_buggy <- function(label = "AE Counts") {
  teal_module(
    label      = label,
    datanames  = c("ADSL", "ADAE"),

    ui = function(id, ...) {
      ns <- NS(id)
      tagList(
        tableOutput(ns("ae_table"))
      )
    },

    server = function(id, data, reporter, ...) {
      moduleServer(id, function(input, output, session) {

        # Bug 1: data used without ()
        adsl_r <- reactive(data[["ADSL"]])

        # Bug 2: TRT01P already exists in adae — no guard before join
        adae_r <- reactive({
          data()[["ADAE"]] |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by = "USUBJID")
        })

        output$ae_table <- renderTable({
          req(adae_r())

          # Bug 3: group_by uses wrong column name (should be TRT01P)
          adae_r() |>
            group_by(TRT01A) |>
            summarise(N_AEs = n(), .groups = "drop") |>
            arrange(desc(N_AEs))
        })
      })
    }
  )
}

# Write the fixed version here:
tm_ae_count_fixed <- function(label = "AE Counts (Fixed)") {
  teal_module(
    label     = label,
    datanames = c("ADSL","ADAE"),
    ui = function(id,...) { ns <- NS(id); tableOutput(ns("ae_table")) },
    server = function(id, data, reporter, ...) {
      moduleServer(id, function(input, output, session) {
        adsl_r <- reactive(___)      # Fix 1
        adae_r <- reactive({
          ___[["ADAE"]] |>           # Fix 2a: use data()
            select(-any_of(___)) |>  # Fix 2b: guard before join
            left_join(select(adsl_r(), USUBJID, TRT01P), by="USUBJID")
        })
        output$ae_table <- renderTable({
          req(adae_r())
          adae_r() |>
            group_by(___) |>         # Fix 3
            summarise(N_AEs=n(), .groups="drop") |>
            arrange(desc(N_AEs))
        })
      })
    }
  )
}

mod_fixed <- tm_ae_count_fixed()
cat("Fixed module compiled:", mod_fixed$label, "\n")
```

```{r debug-challenge-solution}
tm_ae_count_fixed <- function(label = "AE Counts (Fixed)") {
  teal_module(
    label     = label,
    datanames = c("ADSL","ADAE"),
    ui = function(id,...) { ns <- NS(id); tableOutput(ns("ae_table")) },
    server = function(id, data, reporter, ...) {
      moduleServer(id, function(input, output, session) {
        # Fix 1: data() with parentheses
        adsl_r <- reactive(data()[["ADSL"]])
        adae_r <- reactive({
          # Fix 2a: data() with parentheses
          # Fix 2b: guard with select(-any_of("TRT01P")) before join
          data()[["ADAE"]] |>
            select(-any_of("TRT01P")) |>
            left_join(select(adsl_r(), USUBJID, TRT01P), by="USUBJID")
        })
        output$ae_table <- renderTable({
          req(adae_r())
          adae_r() |>
            group_by(TRT01P) |>    # Fix 3: TRT01P not TRT01A
            summarise(N_AEs=n(), .groups="drop") |>
            arrange(desc(N_AEs))
        })
      })
    }
  )
}

mod_fixed <- tm_ae_count_fixed()
cat("Fixed module compiled:", mod_fixed$label, "\n")
```

---

## Summary

You have built a complete, production-quality clinical data review application:

**Architecture:**
- `teal_data()` with `join_keys()` for a fully connected three-dataset layer
- Environment-variable-based data loader for flexible dev/prod switching
- Four custom `teal_module()` modules with full reactive server logic

**Patterns mastered:**
- `data()[[\"DATANAME\"]]` — the reactive data accessor (always with `()`)
- `select(-any_of(...))` — the collision guard before every join
- Empty-data guard (`nrow(d) == 0`) — prevents cryptic render errors
- Dynamic `updateSelectInput()` driven by actual filtered data
- `testServer()` — unit testing reactive server logic without launching Shiny

**Deployment:**
- `renv::snapshot()` / `renv::restore()` for reproducible environments
- `Sys.getenv()` for all credentials and paths
- `rsconnect::deployApp()` for Posit Connect publishing

---

## What's Next — Pharmaverse10

The next module (Lessons 43–48) covers the submission and quality layer of the pharmaverse:

| Lesson | Topic |
|--------|-------|
| 43 | **xportr** — exporting ADaM datasets as submission-ready XPT files with metadata |
| 44 | **datasetjson** — CDISC Dataset-JSON format for modern submissions |
| 45 | **logrx** — automated R session logging for 21 CFR Part 11 compliance |
| 46 | **diffdf / riskmetric** — dataset diffing and package risk scoring |
| 47 | **Capstone Part 1** — full SDTM-to-ADaM pipeline |
| 48 | **Capstone Part 2** — ADaM-to-TLF-to-submission package |
