---
title: "Lesson 43 — xportr: Submission-Ready XPT Export"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  Master xportr for producing FDA/EMA submission-ready SAS XPT transport
  files from R. Covers variable type coercion, label application, length
  assignment, format codes, variable ordering, and the full xportr() pipeline
  integrated with metacore metadata specifications.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 43_xportr.Rmd
# PURPOSE       : xportr package — preparing ADaM datasets for eCTD submission:
#                 xportr_type/label/length/format/order pipeline,
#                 unified xportr() wrapper, date handling, validation.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================
library(learnr)
library(dplyr)
library(tibble)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.eval = FALSE)

# ── Pharmaverseadam datasets ────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}
adsl <- adsl_raw |>
  mutate(ITTFL  = SAFFL,
         DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N")) |>
  select(STUDYID, USUBJID, SUBJID, SITEID, ARM, ARMCD,
         TRT01P, TRT01A, AGE, AGEGR1, SEX, RACE, ETHNIC,
         SAFFL, ITTFL, DIABFL, HEIGHTBL, WEIGHTBL, BMIBL,
         TRTSDT, TRTEDT, TRTDURD, RFSTDTC, RFENDTC,
         DTHFL, COUNTRY)

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(AESEV=="MILD"~1L, AESEV=="MODERATE"~2L,
                             AESEV=="SEVERE"~3L, TRUE~NA_integer_)) |>
  select(STUDYID, USUBJID, AESEQ, AEDECOD, AESOC, AESEV, AESEVN,
         AESTDTC, AEENDTC, AESER, AEREL, AEOUT,
         ASTDT, AENDT, ASTDY, AENDY, TRT01A,
         any_of(c("TRTA","TRTEMFL")))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR=as.integer(ATOXGR), BTOXGR=as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))

# ── Synthetic metadata spec (mirrors a real ADaM spec workbook) ────────────
# adsl_spec: one row per variable in ADSL
adsl_spec <- tribble(
  ~dataset,  ~variable,   ~label,                              ~type,     ~length, ~format,
  "ADSL",   "STUDYID",   "Study Identifier",                  "character", 12,    NA_character_,
  "ADSL",   "USUBJID",   "Unique Subject Identifier",         "character", 50,    NA_character_,
  "ADSL",   "SUBJID",    "Subject Identifier for the Study",  "character", 10,    NA_character_,
  "ADSL",   "SITEID",    "Study Site Identifier",             "character",  3,    NA_character_,
  "ADSL",   "ARM",       "Description of Planned Arm",        "character", 200,   NA_character_,
  "ADSL",   "ARMCD",     "Planned Arm Code",                  "character",  20,   NA_character_,
  "ADSL",   "TRT01P",    "Planned Treatment for Period 01",   "character", 200,   NA_character_,
  "ADSL",   "TRT01A",    "Actual Treatment for Period 01",    "character", 200,   NA_character_,
  "ADSL",   "AGE",       "Age",                               "numeric",     8,   NA_character_,
  "ADSL",   "AGEGR1",    "Pooled Age Group 1",                "character",  10,   NA_character_,
  "ADSL",   "SEX",       "Sex",                               "character",   2,   NA_character_,
  "ADSL",   "RACE",      "Race",                              "character",  64,   NA_character_,
  "ADSL",   "ETHNIC",    "Ethnicity",                         "character",  64,   NA_character_,
  "ADSL",   "SAFFL",     "Safety Population Flag",            "character",   2,   NA_character_,
  "ADSL",   "ITTFL",     "Intent-To-Treat Population Flag",   "character",   2,   NA_character_,
  "ADSL",   "DIABFL",    "Diabetic Flag",                     "character",   2,   NA_character_,
  "ADSL",   "HEIGHTBL",  "Baseline Height (cm)",              "numeric",     8,   NA_character_,
  "ADSL",   "WEIGHTBL",  "Baseline Weight (kg)",              "numeric",     8,   NA_character_,
  "ADSL",   "BMIBL",     "Baseline BMI (kg/m^2)",             "numeric",     8,   NA_character_,
  "ADSL",   "TRTSDT",    "Date of First Exposure to Treatment","numeric",   8,   "DATE9.",
  "ADSL",   "TRTEDT",    "Date of Last Exposure to Treatment","numeric",    8,   "DATE9.",
  "ADSL",   "TRTDURD",   "Total Treatment Duration (Days)",   "numeric",     8,   NA_character_,
  "ADSL",   "RFSTDTC",   "Subject Reference Start Date/Time","character",   20,   NA_character_,
  "ADSL",   "RFENDTC",   "Subject Reference End Date/Time",  "character",   20,   NA_character_,
  "ADSL",   "DTHFL",     "Death Flag",                       "character",    2,   NA_character_,
  "ADSL",   "COUNTRY",   "Country",                          "character",    3,   NA_character_
)

# adae_spec
adae_spec <- tribble(
  ~dataset, ~variable,  ~label,                                  ~type,      ~length, ~format,
  "ADAE",  "STUDYID",  "Study Identifier",                      "character",  12,   NA_character_,
  "ADAE",  "USUBJID",  "Unique Subject Identifier",             "character",  50,   NA_character_,
  "ADAE",  "AESEQ",    "Sequence Number",                       "numeric",     8,   NA_character_,
  "ADAE",  "AEDECOD",  "Dictionary-Derived Term",               "character", 200,   NA_character_,
  "ADAE",  "AESOC",    "Primary System Organ Class",            "character", 200,   NA_character_,
  "ADAE",  "AESEV",    "Severity/Intensity",                    "character",  10,   NA_character_,
  "ADAE",  "AESEVN",   "Severity/Intensity (N)",                "numeric",     8,   NA_character_,
  "ADAE",  "AESTDTC",  "Start Date/Time of Adverse Event",     "character",  20,   NA_character_,
  "ADAE",  "AEENDTC",  "End Date/Time of Adverse Event",       "character",  20,   NA_character_,
  "ADAE",  "AESER",    "Serious Adverse Event",                 "character",   2,   NA_character_,
  "ADAE",  "AEREL",    "Causality",                             "character",  16,   NA_character_,
  "ADAE",  "AEOUT",    "Outcome of Adverse Event",             "character",  60,   NA_character_,
  "ADAE",  "ASTDT",    "Analysis Start Date",                   "numeric",    8,   "DATE9.",
  "ADAE",  "AENDT",    "Analysis End Date",                     "numeric",    8,   "DATE9.",
  "ADAE",  "ASTDY",    "Analysis Start Relative Day",           "numeric",    8,   NA_character_,
  "ADAE",  "AENDY",    "Analysis End Relative Day",             "numeric",    8,   NA_character_,
  "ADAE",  "TRT01A",   "Actual Treatment for Period 01",        "character", 200,   NA_character_
)

# ── Mock xportr functions ─────────────────────────────────────────────────
# xportr is a package for writing SAS XPT files with metadata.
# We mock its core functions so exercises run inside learnr.
# Real xportr writes actual XPT binary files via haven::write_xpt().

# Helper to look up spec for a variable
.xp_lookup <- function(spec, dataset_name, variable) {
  spec[spec$dataset == toupper(dataset_name) & spec$variable == variable, ]
}

# xportr_metadata() — attach spec to dataset as attribute
xportr_metadata <- function(.df, metadata, domain = NULL) {
  domain <- domain %||% deparse(substitute(.df))
  attr(.df, "xportr.df_metadata") <- metadata
  attr(.df, "xportr.domain")      <- toupper(domain)
  .df
}

# xportr_type() — coerce columns to types declared in spec
xportr_type <- function(.df, metadata = NULL, domain = NULL,
                         verbose = "none") {
  spec <- metadata %||% attr(.df, "xportr.df_metadata")
  dom  <- domain   %||% attr(.df, "xportr.domain") %||% "UNKNOWN"

  if (is.null(spec)) {
    message("xportr_type: no metadata — skipping type coercion")
    return(.df)
  }

  spec_dom <- spec[toupper(spec$dataset) == toupper(dom), ]
  changes  <- character(0)

  for (v in intersect(names(.df), spec_dom$variable)) {
    target_type <- spec_dom$type[spec_dom$variable == v]
    if (length(target_type) == 0 || is.na(target_type)) next

    current_class <- class(.df[[v]])[1]

    if (target_type == "numeric" && !is.numeric(.df[[v]])) {
      suppressWarnings({ .df[[v]] <- as.numeric(.df[[v]]) })
      changes <- c(changes, sprintf("%s: %s → numeric", v, current_class))
    } else if (target_type == "character" && !is.character(.df[[v]])) {
      .df[[v]] <- as.character(.df[[v]])
      changes <- c(changes, sprintf("%s: %s → character", v, current_class))
    }
  }

  if (verbose == "message" && length(changes) > 0) {
    message(sprintf("xportr_type [%s]: %d column(s) coerced:\n  %s",
                    dom, length(changes), paste(changes, collapse="\n  ")))
  }
  .df
}

# xportr_label() — apply variable labels from spec
xportr_label <- function(.df, metadata = NULL, domain = NULL,
                          verbose = "none") {
  spec <- metadata %||% attr(.df, "xportr.df_metadata")
  dom  <- domain   %||% attr(.df, "xportr.domain") %||% "UNKNOWN"

  if (is.null(spec)) return(.df)

  spec_dom <- spec[toupper(spec$dataset) == toupper(dom), ]
  labelled <- character(0)

  for (v in intersect(names(.df), spec_dom$variable)) {
    lbl <- spec_dom$label[spec_dom$variable == v]
    if (length(lbl) == 0 || is.na(lbl)) next
    attr(.df[[v]], "label") <- lbl
    labelled <- c(labelled, v)
  }

  if (verbose == "message")
    message(sprintf("xportr_label [%s]: labels applied to %d variable(s)",
                    dom, length(labelled)))
  .df
}

# xportr_length() — apply display lengths from spec
xportr_length <- function(.df, metadata = NULL, domain = NULL,
                           length_source = "spec", verbose = "none") {
  spec <- metadata %||% attr(.df, "xportr.df_metadata")
  dom  <- domain   %||% attr(.df, "xportr.domain") %||% "UNKNOWN"

  if (is.null(spec)) return(.df)

  spec_dom <- spec[toupper(spec$dataset) == toupper(dom), ]

  for (v in intersect(names(.df), spec_dom$variable)) {
    len <- spec_dom$length[spec_dom$variable == v]
    if (length(len) == 0 || is.na(len)) next
    attr(.df[[v]], "width") <- as.integer(len)
  }

  if (verbose == "message")
    message(sprintf("xportr_length [%s]: lengths applied", dom))
  .df
}

# xportr_format() — apply SAS format codes from spec
xportr_format <- function(.df, metadata = NULL, domain = NULL,
                           verbose = "none") {
  spec <- metadata %||% attr(.df, "xportr.df_metadata")
  dom  <- domain   %||% attr(.df, "xportr.domain") %||% "UNKNOWN"

  if (is.null(spec)) return(.df)

  spec_dom <- spec[toupper(spec$dataset) == toupper(dom), ]
  formatted <- character(0)

  for (v in intersect(names(.df), spec_dom$variable)) {
    fmt <- spec_dom$format[spec_dom$variable == v]
    if (length(fmt) == 0 || is.na(fmt)) next
    attr(.df[[v]], "format.sas") <- fmt
    formatted <- c(formatted, sprintf("%s=%s", v, fmt))
  }

  if (verbose == "message" && length(formatted) > 0)
    message(sprintf("xportr_format [%s]: SAS formats applied: %s",
                    dom, paste(formatted, collapse=", ")))
  .df
}

# xportr_order() — reorder columns to match spec variable order
xportr_order <- function(.df, metadata = NULL, domain = NULL,
                          verbose = "none") {
  spec <- metadata %||% attr(.df, "xportr.df_metadata")
  dom  <- domain   %||% attr(.df, "xportr.domain") %||% "UNKNOWN"

  if (is.null(spec)) return(.df)

  spec_vars <- spec$variable[toupper(spec$dataset) == toupper(dom)]
  ordered   <- intersect(spec_vars, names(.df))
  extra     <- setdiff(names(.df), spec_vars)

  if (length(extra) > 0 && verbose == "message")
    message(sprintf("xportr_order [%s]: %d extra column(s) not in spec moved to end: %s",
                    dom, length(extra), paste(extra, collapse=", ")))

  .df[, c(ordered, extra), drop=FALSE]
}

# xportr_write() — write XPT file (mocked: prints what would be written)
xportr_write <- function(.df, path, label = NULL, metadata = NULL,
                          domain = NULL, verbose = "none") {
  dom   <- domain %||% attr(.df, "xportr.domain") %||%
           toupper(tools::file_path_sans_ext(basename(path)))
  label <- label %||% sprintf("%s dataset", dom)

  # Collect variable metadata from attributes
  var_info <- lapply(names(.df), function(v) {
    list(
      variable = v,
      type     = if (is.numeric(.df[[v]])) "numeric" else "character",
      label    = attr(.df[[v]], "label")   %||% "",
      length   = attr(.df[[v]], "width")   %||% NA_integer_,
      format   = attr(.df[[v]], "format.sas") %||% ""
    )
  })

  cat(sprintf(
    "\n── xportr_write ──────────────────────────────────────\n",
  ))
  cat(sprintf("  Output path : %s\n", path))
  cat(sprintf("  Dataset     : %s\n", dom))
  cat(sprintf("  Label       : %s\n", label))
  cat(sprintf("  Rows        : %d\n", nrow(.df)))
  cat(sprintf("  Variables   : %d\n", ncol(.df)))

  labelled_n <- sum(vapply(var_info, function(v) nchar(v$label)>0, logical(1)))
  formatted_n <- sum(vapply(var_info, function(v) nchar(v$format)>0, logical(1)))
  cat(sprintf("  Labelled    : %d / %d variables\n", labelled_n, ncol(.df)))
  cat(sprintf("  SAS formats : %d variable(s)\n", formatted_n))
  cat(sprintf("  [mock] XPT write to '%s' (haven::write_xpt)\n", path))
  cat("──────────────────────────────────────────────────────\n")

  invisible(.df)
}

# xportr() — the pipe-friendly unified wrapper (teal >= 1.0.0)
# Applies type, label, length, format, order, and write in one call
xportr <- function(.df, metadata, domain = NULL, path = NULL,
                   label = NULL, verbose = "none") {
  dom <- domain %||% deparse(substitute(.df))

  result <- .df |>
    xportr_metadata(metadata, domain = dom) |>
    xportr_type(    verbose = verbose) |>
    xportr_label(   verbose = verbose) |>
    xportr_length(  verbose = verbose) |>
    xportr_format(  verbose = verbose) |>
    xportr_order(   verbose = verbose)

  if (!is.null(path))
    xportr_write(result, path=path, label=label, verbose=verbose)

  invisible(result)
}

# xportr_df_label() — set dataset-level label (SAS PROC format LABEL= equivalent)
xportr_df_label <- function(.df, label) {
  attr(.df, "label") <- label
  .df
}

# Utility: show metadata summary for a dataset
xportr_var_names <- function(.df) {
  cat(sprintf("Variables in dataset: %d\n", ncol(.df)))
  cat(sprintf("  %-20s %-12s %-8s %-12s %s\n",
              "Variable", "Type", "Length", "Format", "Label"))
  cat(strrep("-", 80), "\n")
  for (v in names(.df)) {
    cat(sprintf("  %-20s %-12s %-8s %-12s %s\n",
                v,
                if (is.numeric(.df[[v]])) "numeric" else "character",
                attr(.df[[v]], "width") %||% NA,
                attr(.df[[v]], "format.sas") %||% "",
                attr(.df[[v]], "label") %||% ""))
  }
  invisible(.df)
}

# Validation summary (mock Pinnacle 21 / P21 checks)
xportr_validate <- function(.df, metadata, domain = NULL) {
  dom <- domain %||% attr(.df, "xportr.domain") %||% "UNKNOWN"
  spec_dom <- metadata[toupper(metadata$dataset) == toupper(dom), ]

  issues <- character(0)

  # Check 1: all spec variables present
  missing_from_data <- setdiff(spec_dom$variable, names(.df))
  if (length(missing_from_data) > 0)
    issues <- c(issues, sprintf("[WARN] Variables in spec but not in data: %s",
                                paste(missing_from_data, collapse=", ")))

  # Check 2: extra variables not in spec
  extra_in_data <- setdiff(names(.df), spec_dom$variable)
  if (length(extra_in_data) > 0)
    issues <- c(issues, sprintf("[INFO] Variables in data but not in spec: %s",
                                paste(extra_in_data, collapse=", ")))

  # Check 3: character variables exceeding declared length
  for (v in intersect(names(.df), spec_dom$variable)) {
    if (is.character(.df[[v]])) {
      max_nchar <- max(nchar(.df[[v]]), na.rm=TRUE)
      spec_len  <- spec_dom$length[spec_dom$variable == v]
      if (length(spec_len) > 0 && !is.na(spec_len) && max_nchar > spec_len)
        issues <- c(issues, sprintf(
          "[ERROR] %s: max length %d exceeds spec length %d",
          v, max_nchar, spec_len))
    }
  }

  # Check 4: labels missing
  unlabelled <- intersect(spec_dom$variable[is.na(spec_dom$label)], names(.df))
  if (length(unlabelled) > 0)
    issues <- c(issues, sprintf("[WARN] No label in spec for: %s",
                                paste(unlabelled, collapse=", ")))

  cat(sprintf("── xportr validation [%s] ──────────────────────────\n", dom))
  if (length(issues) == 0) {
    cat("  ✓ No issues found\n")
  } else {
    for (iss in issues) cat(" ", iss, "\n")
  }
  cat(sprintf("  Total issues: %d\n", length(issues)))
  cat("──────────────────────────────────────────────────────\n")
  invisible(issues)
}

`%||%` <- function(a, b) if (is.null(a)) b else a
```

## Overview

**xportr** is the pharmaverse package that turns an R data frame into a submission-ready SAS XPT transport file. It bridges the gap between the open-source R workflow and the FDA/EMA requirement that electronic submissions include data in SAS Version 5 transport format.

> **Why not just `haven::write_xpt()`?**
>
> `haven::write_xpt()` writes the XPT binary, but it cannot apply variable labels, SAS format codes, or declaration lengths — all required by the CDISC Submission Data Standards and validated by Pinnacle 21. xportr adds that metadata layer using your ADaM specification workbook.

### What you will learn

- The five xportr transformation steps: type → label → length → format → order
- `xportr_metadata()` for attaching a spec to a pipeline
- `xportr_write()` for producing the XPT file
- The `xportr()` unified wrapper
- Validation with `xportr_validate()`
- Integration with metacore spec objects
- Common submission gotchas: date numeric encoding, flag lengths, label truncation

---

## 1 — The XPT Submission Standard

### 1.1 What FDA requires in an XPT file

SAS Version 5 transport format (XPT) has strict rules enforced by Pinnacle 21 Enterprise during submission validation:

| Requirement | Rule |
|-------------|------|
| Variable names | ≤ 8 characters, uppercase, no special chars |
| Variable labels | ≤ 200 characters |
| Character length | Declared width ≥ actual max character length |
| Date variables | Stored as **numeric** (SAS date integer = days since 1960-01-01) |
| SAS formats | Date variables must carry `DATE9.` or `DATETIME20.` format |
| Variable order | Must match define.xml / ADaM spec variable order |
| Dataset label | ≤ 200 characters |

### 1.2 The SAS date encoding problem

This is the most common source of XPT validation failures when migrating from SAS to R:

```{r sas-date-demo, eval=FALSE}
# SAS stores dates as integers: days since 1960-01-01
# R stores dates as Date objects (days since 1970-01-01)

# A SAS date of 20000 means: 1960-01-01 + 20000 days = 2014-10-13
as.Date("1960-01-01") + 20000   # → 2014-10-13

# haven::write_xpt() converts R Date → SAS integer automatically
# BUT: in your ADaM data frame the date must be NUMERIC (integer),
# not a Date class object, to get the right format code applied.

# Wrong type (Date class — haven converts OK but P21 misses format code):
class(adsl$TRTSDT)   # "Date"

# Correct for xportr pipeline (numeric with DATE9. format):
# xportr_type() + xportr_format() handle this conversion
```

### 1.3 Quiz

```{r quiz-xpt, echo=FALSE}
quiz(
  question(
    "What is the maximum allowed length for a variable label in a SAS XPT submission file?",
    answer("40 characters"),
    answer("200 characters", correct = TRUE),
    answer("256 characters"),
    answer("There is no limit"),
    allow_retry = TRUE
  ),
  question(
    "How does SAS (and XPT format) store date values internally?",
    answer("As ISO 8601 character strings (e.g., '2024-03-15')"),
    answer("As integers representing days since 1960-01-01", correct = TRUE),
    answer("As integers representing days since 1970-01-01"),
    answer("As double-precision floating point timestamps"),
    allow_retry = TRUE
  ),
  question(
    "What Pinnacle 21 message would you expect if a character variable's actual max length exceeds its declared spec length?",
    answer("A WARNING about variable ordering"),
    answer("An ERROR about dataset label"),
    answer("An ERROR about character length exceeding declared width", correct = TRUE),
    answer("An INFO about missing format code"),
    allow_retry = TRUE
  )
)
```

---

## 2 — Metadata Specifications

### 2.1 The spec data frame

xportr works from a metadata data frame — typically exported from your ADaM specification workbook. The minimum required columns are:

| Column | Contents |
|--------|---------|
| `dataset` | Dataset name in UPPERCASE (e.g., `"ADSL"`) |
| `variable` | Variable name in UPPERCASE (e.g., `"AGE"`) |
| `label` | Variable label ≤ 200 characters |
| `type` | `"character"` or `"numeric"` |
| `length` | Integer — declared width for XPT |
| `format` | SAS format code (e.g., `"DATE9."`) — `NA` for most variables |

> The `adsl_spec` and `adae_spec` objects shown in these exercises are already loaded in the setup chunk. They mirror what a real `ADSL` and `ADAE` specification workbook would produce.

### 2.2 Examine the ADSL spec

```{r spec-examine, exercise=TRUE, exercise.lines=12}
# Look at the first several rows of adsl_spec
head(adsl_spec, 10)
```

```{r spec-examine-solution}
head(adsl_spec, 10)
```

### 2.3 Exercise — inspect date and flag variables in spec

Pull only the date-type variables (those with a `format` value) and the flag variables (ending in `FL`).

```{r spec-dates-flags, exercise=TRUE, exercise.lines=12}
# Date variables (have a SAS format code)
date_vars <- adsl_spec |>
  filter(!is.na(___)) |>
  select(variable, label, type, format)

cat("Date variables:\n")
print(date_vars)

# Flag variables
flag_vars <- adsl_spec |>
  filter(grepl(___,  variable)) |>
  select(variable, label, length)

cat("\nFlag variables:\n")
print(flag_vars)
```

```{r spec-dates-flags-solution}
date_vars <- adsl_spec |>
  filter(!is.na(format)) |>
  select(variable, label, type, format)

cat("Date variables:\n")
print(date_vars)

flag_vars <- adsl_spec |>
  filter(grepl("FL$", variable)) |>
  select(variable, label, length)

cat("\nFlag variables:\n")
print(flag_vars)
```

---

## 3 — Step-by-Step: The Five xportr Functions

### 3.1 `xportr_metadata()` — attach spec to dataset

Rather than passing `metadata =` to every function call, attach the spec once:

```{r metadata-attach-demo, eval=FALSE}
adsl_ready <- adsl |>
  xportr_metadata(metadata = adsl_spec, domain = "ADSL")

# Now all subsequent xportr_*() calls infer metadata automatically
```

### 3.2 `xportr_type()` — coerce column types

```{r type-demo, eval=FALSE}
# Before: check current types
sapply(select(adsl, AGE, TRT01P, TRTSDT, SAFFL), class)
# AGE    TRT01P  TRTSDT  SAFFL
# integer character Date    character

# After xportr_type() with spec declaring TRTSDT as numeric:
adsl_typed <- adsl |>
  xportr_metadata(adsl_spec, "ADSL") |>
  xportr_type(verbose = "message")
# xportr_type [ADSL]: TRTSDT: Date → numeric, TRTEDT: Date → numeric
```

### 3.3 `xportr_label()` — attach variable labels

```{r label-demo, eval=FALSE}
adsl_labelled <- adsl_typed |>
  xportr_label(verbose = "message")

# Verify a label
attr(adsl_labelled$AGE, "label")
# [1] "Age"

attr(adsl_labelled$USUBJID, "label")
# [1] "Unique Subject Identifier"
```

### 3.4 `xportr_length()` — set declared widths

```{r length-demo, eval=FALSE}
adsl_lengthed <- adsl_labelled |>
  xportr_length(verbose = "message")

attr(adsl_lengthed$STUDYID, "width")   # 12
attr(adsl_lengthed$USUBJID, "width")   # 50
attr(adsl_lengthed$AGE,     "width")   # 8
```

### 3.5 `xportr_format()` — apply SAS format codes

```{r format-demo, eval=FALSE}
adsl_formatted <- adsl_lengthed |>
  xportr_format(verbose = "message")
# xportr_format [ADSL]: SAS formats applied: TRTSDT=DATE9., TRTEDT=DATE9.

attr(adsl_formatted$TRTSDT, "format.sas")   # "DATE9."
attr(adsl_formatted$AGE,    "format.sas")   # ""  (NA → no format)
```

### 3.6 `xportr_order()` — reorder to match spec

```{r order-demo, eval=FALSE}
# Confirm variable order matches spec
adsl_ordered <- adsl_formatted |>
  xportr_order(verbose = "message")

# Check first 5 variables in order
names(adsl_ordered)[1:5]
# [1] "STUDYID" "USUBJID" "SUBJID"  "SITEID"  "ARM"
```

### 3.7 `xportr_write()` — write the XPT

```{r write-demo, eval=FALSE}
xportr_write(adsl_ordered,
             path  = "adam/adsl.xpt",
             label = "Subject-Level Analysis Dataset",
             verbose = "message")
```

### 3.8 Exercise — apply all five steps manually

Apply all five xportr transformations to ADSL in sequence and print the result summary.

```{r five-steps-ex, exercise=TRUE, exercise.lines=30}
adsl_xpt <- adsl |>
  xportr_metadata(___, domain = "ADSL") |>
  xportr_type(   verbose = "message") |>
  ___( verbose = "message") |>   # label
  ___( verbose = "message") |>   # length
  ___( verbose = "message") |>   # format
  xportr_order(  verbose = "message")

# Inspect the processed dataset
cat("\n── Variable metadata after processing ──\n")
xportr_var_names(adsl_xpt)
```

```{r five-steps-ex-solution}
adsl_xpt <- adsl |>
  xportr_metadata(adsl_spec, domain = "ADSL") |>
  xportr_type(   verbose = "message") |>
  xportr_label(  verbose = "message") |>
  xportr_length( verbose = "message") |>
  xportr_format( verbose = "message") |>
  xportr_order(  verbose = "message")

cat("\n── Variable metadata after processing ──\n")
xportr_var_names(adsl_xpt)
```

---

## 4 — The Unified `xportr()` Wrapper

### 4.1 One function, all steps

From xportr ≥ 1.0.0, the unified `xportr()` call replaces five separate calls:

```{r unified-wrapper, eval=FALSE}
# Equivalent to all five steps + write:
adsl |>
  xportr(
    metadata = adsl_spec,
    domain   = "ADSL",
    path     = "adam/adsl.xpt",
    label    = "Subject-Level Analysis Dataset",
    verbose  = "message"
  )
```

### 4.2 Exercise — use the unified wrapper for ADAE

```{r unified-adae-ex, exercise=TRUE, exercise.lines=15}
adae_xpt <- adae |>
  xportr(
    metadata = ___,
    domain   = ___,
    path     = "adam/adae.xpt",
    label    = "Adverse Events Analysis Dataset",
    verbose  = "message"
  )

cat("\nDimensions:", nrow(adae_xpt), "×", ncol(adae_xpt), "\n")
```

```{r unified-adae-ex-solution}
adae_xpt <- adae |>
  xportr(
    metadata = adae_spec,
    domain   = "ADAE",
    path     = "adam/adae.xpt",
    label    = "Adverse Events Analysis Dataset",
    verbose  = "message"
  )

cat("\nDimensions:", nrow(adae_xpt), "×", ncol(adae_xpt), "\n")
```

### 4.3 Dataset-level label with `xportr_df_label()`

XPT files carry a dataset-level label in addition to variable-level labels. Apply it with `xportr_df_label()`:

```{r df-label-ex, exercise=TRUE, exercise.lines=10}
adsl_final <- adsl |>
  xportr_metadata(adsl_spec, "ADSL") |>
  xportr_type() |>
  xportr_label() |>
  xportr_length() |>
  xportr_format() |>
  xportr_order() |>
  xportr_df_label(label = ___)  # "Subject-Level Analysis Dataset"

cat("Dataset label:", attr(adsl_final, "label"), "\n")
```

```{r df-label-ex-solution}
adsl_final <- adsl |>
  xportr_metadata(adsl_spec, "ADSL") |>
  xportr_type() |>
  xportr_label() |>
  xportr_length() |>
  xportr_format() |>
  xportr_order() |>
  xportr_df_label(label = "Subject-Level Analysis Dataset")

cat("Dataset label:", attr(adsl_final, "label"), "\n")
```

---

## 5 — Validation

### 5.1 `xportr_validate()` — pre-submission checks

Before writing the XPT, validate the dataset against the spec:

```{r validate-demo, eval=FALSE}
issues <- xportr_validate(adsl_xpt, metadata = adsl_spec, domain = "ADSL")
# ── xportr validation [ADSL] ──────────────────────────────────────────
#   ✓ No issues found
#   Total issues: 0
```

Common issues caught by `xportr_validate()`:

| Issue level | Cause | Impact |
|------------|-------|--------|
| ERROR | Character variable exceeds declared spec length | XPT will fail P21 |
| ERROR | Numeric variable stored as character | Type mismatch in XPT |
| WARN | Variable in spec but missing from data | Missing variable warning in P21 |
| WARN | Variable label missing in spec | P21 message about unlabelled variable |
| INFO | Variable in data but not in spec | Extra variable — may be intentional |

### 5.2 Exercise — validate a dataset with deliberate issues

We'll create a modified ADSL with an intentional length violation and observe the validation output.

```{r validate-ex, exercise=TRUE, exercise.lines=25}
# Create a dataset with a deliberate issue:
# Add a new variable not in spec (INFO)
# Make COUNTRY have a value longer than its spec length (3)
adsl_bad <- adsl |>
  mutate(
    COUNTRY    = if_else(COUNTRY == "USA", "UNITED STATES OF AMERICA", COUNTRY),
    NEW_VAR    = "not in spec"
  )

# Run validation
issues <- xportr_validate(
  adsl_bad,
  metadata = ___,
  domain   = "ADSL"
)

cat("\nIssues found:", length(issues), "\n")
```

```{r validate-ex-solution}
adsl_bad <- adsl |>
  mutate(
    COUNTRY    = if_else(COUNTRY == "USA", "UNITED STATES OF AMERICA", COUNTRY),
    NEW_VAR    = "not in spec"
  )

issues <- xportr_validate(
  adsl_bad,
  metadata = adsl_spec,
  domain   = "ADSL"
)

cat("\nIssues found:", length(issues), "\n")
```

### 5.3 Fix and re-validate

```{r fix-validate-ex, exercise=TRUE, exercise.lines=20}
# Fix the issues:
# 1. Truncate COUNTRY back to ISO 3-char code
# 2. Remove NEW_VAR
adsl_fixed <- adsl |>
  mutate(
    COUNTRY = substr(COUNTRY, 1, ___)   # truncate to spec length 3
  ) |>
  select(-any_of("NEW_VAR"))

# Re-validate
xportr_validate(adsl_fixed, metadata = adsl_spec, domain = "ADSL")
```

```{r fix-validate-ex-solution}
adsl_fixed <- adsl |>
  mutate(
    COUNTRY = substr(COUNTRY, 1, 3)
  ) |>
  select(-any_of("NEW_VAR"))

xportr_validate(adsl_fixed, metadata = adsl_spec, domain = "ADSL")
```

---

## 6 — Working with Date Variables

### 6.1 R Date → SAS numeric

This is the most common type of submission error. In pharmaverseadam, date variables like `TRTSDT` are `Date` class. For XPT they must be **numeric** with `DATE9.` format.

```{r date-conversion-demo, exercise=TRUE, exercise.lines=18}
# Check current type of TRTSDT in adsl
cat("Class of TRTSDT:", class(adsl$TRTSDT), "\n")
cat("First 3 values:", format(head(adsl$TRTSDT, 3)), "\n\n")

# How xportr converts: as.numeric(Date) = days since 1970-01-01
# SAS needs days since 1960-01-01, so offset = 3653 days
# haven::write_xpt handles the offset automatically during write

# Show the numeric representation
adsl_dates <- adsl |>
  select(USUBJID, TRTSDT, TRTEDT) |>
  mutate(
    TRTSDT_NUM = as.numeric(TRTSDT),  # days since 1970-01-01
    TRTEDT_NUM = as.numeric(TRTEDT)
  )

head(adsl_dates, 5)
```

### 6.2 Type coercion in the pipeline

`xportr_type()` converts `Date` → `numeric` automatically when the spec declares the variable as `"numeric"`. Verify this:

```{r date-type-ex, exercise=TRUE, exercise.lines=15}
# Before: TRTSDT is Date
cat("Before - TRTSDT class:", class(adsl$TRTSDT), "\n")

adsl_typed <- adsl |>
  xportr_metadata(adsl_spec, "ADSL") |>
  xportr_type(verbose = "message")

# After: TRTSDT should be numeric
cat("\nAfter  - TRTSDT class:", class(adsl_typed$TRTSDT), "\n")
cat("First value:", adsl_typed$TRTSDT[1], "\n")
cat("Decoded back:", format(as.Date(adsl_typed$TRTSDT[1], origin="1970-01-01")), "\n")
```

### 6.3 Quiz

```{r quiz-dates, echo=FALSE}
quiz(
  question(
    "After `xportr_type()` converts an R Date variable to numeric for XPT, what additional step ensures Pinnacle 21 recognises it as a date?",
    answer("`xportr_label()` — the label must say 'Date'"),
    answer("`xportr_format()` — applies the DATE9. SAS format code", correct = TRUE),
    answer("`xportr_length()` — date variables need length=10"),
    answer("`xportr_order()` — date variables must come last"),
    allow_retry = TRUE
  ),
  question(
    "A date variable in your spec has `type = 'numeric'` but `format = NA`. What will happen after `xportr_type()` + `xportr_format()`?",
    answer("The variable will be left as Date class"),
    answer("The variable will be converted to numeric with no SAS format code — P21 may warn about missing format", correct = TRUE),
    answer("An error is thrown because format is required for numeric dates"),
    answer("The variable is dropped from the dataset"),
    allow_retry = TRUE
  )
)
```

---

## 7 — Integration with metacore

### 7.1 `metacore` as a metadata source

In Lesson 12 you learned `metacore` — a package that stores structured ADaM specification objects. xportr can consume a `metacore` object directly instead of a raw data frame.

```{r metacore-integration, eval=FALSE}
library(metacore)
library(metatools)

# Load a metacore spec from an Excel workbook
mc <- spec_to_metacore(
  path     = "specs/adam_specs.xlsx",
  where_sep_sheet = FALSE
)

# xportr accepts metacore directly in any metadata argument:
adsl |>
  xportr_metadata(mc, domain = "ADSL") |>
  xportr_type() |>
  xportr_label() |>
  xportr_length() |>
  xportr_format() |>
  xportr_order() |>
  xportr_write("adam/adsl.xpt",
               label = "Subject-Level Analysis Dataset")
```

### 7.2 Extracting a flat spec from metacore

If you have a metacore object but need a flat data frame (for inspection or logging):

```{r metacore-flat, eval=FALSE}
# Extract variable-level metadata as a flat tibble
flat_spec <- mc |>
  select_dataset("ADSL") |>
  get_variables()

# flat_spec has columns: dataset, variable, label, type, length, format, ...
head(flat_spec)
```

---

## 8 — Building a Submission Package: Multiple Datasets

### 8.1 The `purrr::walk()` pattern

In a real submission you export 10+ ADaM datasets. Use `purrr::walk()` for clean iteration:

```{r multi-dataset-export, eval=FALSE}
library(purrr)

# Named list of datasets
adam_datasets <- list(
  ADSL = adsl,
  ADAE = adae,
  ADLB = adlb,
  ADVS = advs
)

# Combined spec (bind rows from each dataset spec)
full_spec <- bind_rows(adsl_spec, adae_spec, adlb_spec, advs_spec)

# Dataset labels
ds_labels <- c(
  ADSL = "Subject-Level Analysis Dataset",
  ADAE = "Adverse Events Analysis Dataset",
  ADLB = "Laboratory Test Results",
  ADVS = "Vital Signs Analysis Dataset"
)

# Export all datasets
output_dir <- "submission/adam/"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

walk(names(adam_datasets), function(dsname) {
  message(sprintf("Exporting %s...", dsname))
  adam_datasets[[dsname]] |>
    xportr(
      metadata = full_spec,
      domain   = dsname,
      path     = file.path(output_dir, paste0(tolower(dsname), ".xpt")),
      label    = ds_labels[[dsname]],
      verbose  = "message"
    )
})

message("All datasets exported to ", output_dir)
```

### 8.2 Exercise — export two datasets in a loop

Export ADSL and ADAE using the loop pattern. Use the mock specs and `xportr()`.

```{r multi-export-ex, exercise=TRUE, exercise.lines=28}
library(purrr)

adam_list <- list(
  ADSL = adsl,
  ADAE = ___
)

combined_spec <- bind_rows(___, adae_spec)

ds_labels <- c(
  ADSL = "Subject-Level Analysis Dataset",
  ADAE = "Adverse Events Analysis Dataset"
)

walk(names(adam_list), function(dsname) {
  cat(sprintf("\n── Exporting %s ──────────────────\n", dsname))
  adam_list[[dsname]] |>
    xportr(
      metadata = ___,
      domain   = ___,
      path     = paste0("adam/", tolower(dsname), ".xpt"),
      label    = ds_labels[[___]],
      verbose  = "none"
    )
  cat(sprintf("  ✓ %s: %d rows × %d cols exported\n",
              dsname,
              nrow(adam_list[[dsname]]),
              ncol(adam_list[[dsname]])))
})
```

```{r multi-export-ex-solution}
library(purrr)

adam_list <- list(
  ADSL = adsl,
  ADAE = adae
)

combined_spec <- bind_rows(adsl_spec, adae_spec)

ds_labels <- c(
  ADSL = "Subject-Level Analysis Dataset",
  ADAE = "Adverse Events Analysis Dataset"
)

walk(names(adam_list), function(dsname) {
  cat(sprintf("\n── Exporting %s ──────────────────\n", dsname))
  adam_list[[dsname]] |>
    xportr(
      metadata = combined_spec,
      domain   = dsname,
      path     = paste0("adam/", tolower(dsname), ".xpt"),
      label    = ds_labels[[dsname]],
      verbose  = "none"
    )
  cat(sprintf("  ✓ %s: %d rows × %d cols exported\n",
              dsname,
              nrow(adam_list[[dsname]]),
              ncol(adam_list[[dsname]])))
})
```

---

## 9 — Common Submission Issues

### 9.1 The eight most common P21 xportr failures

```{r p21-issues, eval=FALSE}
# 1. Variable name > 8 characters
# Solution: rename before xportr pipeline
adsl <- rename(adsl, USUBJID = UNIQUE_SUBJECT_ID)  # if it was too long

# 2. Character variable value exceeds declared width
# Solution: check max nchar() and increase spec width or truncate
adsl |>
  summarise(across(where(is.character), ~max(nchar(.), na.rm=TRUE))) |>
  tidyr::pivot_longer(everything(),
                      names_to="variable", values_to="max_len") |>
  left_join(select(adsl_spec, variable, length), by="variable") |>
  filter(max_len > length)

# 3. Date variable left as Date class (not numeric)
# Symptom: P21 "Variable format does not match type"
# Solution: xportr_type() with spec declaring type="numeric"

# 4. Flag variable (SAFFL, ITTFL) declared with length=1 but has "Y" / "N" / ""
# "" has nchar 0, but "Y" has nchar 1 — spec length=2 covers both safely
# Solution: always declare flag variables with length >= 2

# 5. Label > 200 characters
# Solution: truncate in spec
adsl_spec |>
  filter(nchar(label) > 200)   # should be empty; fix if not

# 6. Missing format code for date variable
# Solution: spec must have format="DATE9." for any numeric date

# 7. Variables in wrong order
# Solution: xportr_order() sorts to spec order; ensure spec is complete

# 8. Dataset label missing or > 200 chars
# Solution: xportr_df_label("Subject-Level Analysis Dataset")
```

### 9.2 Exercise — detect length violations before export

Write code to find any character variables in ADSL whose actual maximum length exceeds the spec-declared length.

```{r length-violations-ex, exercise=TRUE, exercise.lines=20}
# Step 1: compute actual max character length per variable
actual_lengths <- adsl |>
  select(where(is.character)) |>
  summarise(across(everything(), ~max(nchar(.), na.rm=TRUE))) |>
  tidyr::pivot_longer(everything(),
                      names_to  = "variable",
                      values_to = "actual_max")

# Step 2: join to spec
violations <- actual_lengths |>
  left_join(select(adsl_spec, variable, ___), by = "variable") |>
  filter(!is.na(length), ___ > length) |>
  arrange(desc(actual_max - length))

if (nrow(violations) == 0) {
  cat("✓ No length violations found\n")
} else {
  cat("Length violations:\n")
  print(violations)
}
```

```{r length-violations-ex-solution}
actual_lengths <- adsl |>
  select(where(is.character)) |>
  summarise(across(everything(), ~max(nchar(.), na.rm=TRUE))) |>
  tidyr::pivot_longer(everything(),
                      names_to  = "variable",
                      values_to = "actual_max")

violations <- actual_lengths |>
  left_join(select(adsl_spec, variable, length), by = "variable") |>
  filter(!is.na(length), actual_max > length) |>
  arrange(desc(actual_max - length))

if (nrow(violations) == 0) {
  cat("✓ No length violations found\n")
} else {
  cat("Length violations:\n")
  print(violations)
}
```

---

## 10 — The Full Submission Export Pipeline

### 10.1 End-to-end pipeline

```{r full-pipeline, eval=FALSE}
# ── Full ADaM XPT export pipeline ─────────────────────────────────────────
library(xportr)
library(dplyr)
library(purrr)

# ── 1. Load and prepare data ─────────────────────────────────────────────
source("R/data_prep.R")   # loads adsl, adae, adlb, advs

# ── 2. Load metadata spec ────────────────────────────────────────────────
library(readxl)
spec <- bind_rows(
  read_excel("specs/adam_spec.xlsx", sheet="ADSL"),
  read_excel("specs/adam_spec.xlsx", sheet="ADAE"),
  read_excel("specs/adam_spec.xlsx", sheet="ADLB"),
  read_excel("specs/adam_spec.xlsx", sheet="ADVS")
) |>
  mutate(
    type   = tolower(type),        # normalise "Char"/"Text" → "character"
    format = na_if(format, ""),    # blank → NA
    length = as.integer(length)
  )

# ── 3. Validate all datasets ─────────────────────────────────────────────
datasets <- list(ADSL=adsl, ADAE=adae, ADLB=adlb, ADVS=advs)

all_issues <- imap(datasets, function(ds, nm) {
  xportr_validate(ds, metadata=spec, domain=nm)
}) |>
  compact()

if (length(all_issues) > 0) {
  warning("Validation issues found — review before submitting")
}

# ── 4. Export ────────────────────────────────────────────────────────────
ds_labels <- c(
  ADSL = "Subject-Level Analysis Dataset",
  ADAE = "Adverse Events Analysis Dataset",
  ADLB = "Laboratory Test Results",
  ADVS = "Vital Signs Analysis Dataset"
)

output_dir <- "submission/adam"
dir.create(output_dir, recursive=TRUE, showWarnings=FALSE)

walk(names(datasets), function(nm) {
  datasets[[nm]] |>
    xportr(
      metadata = spec,
      domain   = nm,
      path     = file.path(output_dir, paste0(tolower(nm), ".xpt")),
      label    = ds_labels[[nm]],
      verbose  = "message"
    )
})

# ── 5. Verify written files ──────────────────────────────────────────────
xpt_files <- list.files(output_dir, pattern="\\.xpt$", full.names=TRUE)
file_info <- tibble(
  file    = basename(xpt_files),
  size_kb = round(file.size(xpt_files)/1024, 1)
)
print(file_info)
```

### 10.2 Capstone: build and validate the ADSL export

Build the complete ADSL XPT export pipeline: validate → fix issues → apply all xportr steps → write → verify.

```{r capstone-export, exercise=TRUE, exercise.lines=45}
# Step 1: Pre-export validation
cat("=== Step 1: Pre-export validation ===\n")
issues_before <- xportr_validate(adsl, metadata = adsl_spec, domain = "ADSL")

# Step 2: Prepare clean dataset
cat("\n=== Step 2: Prepare dataset ===\n")
adsl_clean <- adsl |>
  # Keep only variables that are in the spec
  select(any_of(adsl_spec$variable)) |>
  # Ensure COUNTRY is at most 3 chars (spec length)
  mutate(COUNTRY = substr(COUNTRY, 1, 3))

cat("Variables retained:", ncol(adsl_clean), "\n")
cat("Variables dropped:", ncol(adsl) - ncol(adsl_clean), "\n")

# Step 3: Apply xportr pipeline
cat("\n=== Step 3: Apply xportr ===\n")
adsl_submission <- adsl_clean |>
  xportr_metadata(___, domain = "ADSL") |>
  xportr_type() |>
  xportr_label() |>
  ___() |>    # length
  ___() |>    # format
  xportr_order() |>
  xportr_df_label("Subject-Level Analysis Dataset")

# Step 4: Post-pipeline validation
cat("\n=== Step 4: Post-pipeline validation ===\n")
xportr_validate(adsl_submission, metadata = ___, domain = "ADSL")

# Step 5: Write (mock)
cat("\n=== Step 5: Write XPT ===\n")
xportr_write(adsl_submission,
             path    = "adam/adsl.xpt",
             label   = "Subject-Level Analysis Dataset",
             verbose = "none")

cat("\n✓ Pipeline complete\n")
cat("  Rows:      ", nrow(adsl_submission), "\n")
cat("  Variables: ", ncol(adsl_submission), "\n")
cat("  Label:     ", attr(adsl_submission, "label"), "\n")
```

```{r capstone-export-solution}
cat("=== Step 1: Pre-export validation ===\n")
issues_before <- xportr_validate(adsl, metadata = adsl_spec, domain = "ADSL")

cat("\n=== Step 2: Prepare dataset ===\n")
adsl_clean <- adsl |>
  select(any_of(adsl_spec$variable)) |>
  mutate(COUNTRY = substr(COUNTRY, 1, 3))

cat("Variables retained:", ncol(adsl_clean), "\n")
cat("Variables dropped:", ncol(adsl) - ncol(adsl_clean), "\n")

cat("\n=== Step 3: Apply xportr ===\n")
adsl_submission <- adsl_clean |>
  xportr_metadata(adsl_spec, domain = "ADSL") |>
  xportr_type() |>
  xportr_label() |>
  xportr_length() |>
  xportr_format() |>
  xportr_order() |>
  xportr_df_label("Subject-Level Analysis Dataset")

cat("\n=== Step 4: Post-pipeline validation ===\n")
xportr_validate(adsl_submission, metadata = adsl_spec, domain = "ADSL")

cat("\n=== Step 5: Write XPT ===\n")
xportr_write(adsl_submission,
             path    = "adam/adsl.xpt",
             label   = "Subject-Level Analysis Dataset",
             verbose = "none")

cat("\n✓ Pipeline complete\n")
cat("  Rows:      ", nrow(adsl_submission), "\n")
cat("  Variables: ", ncol(adsl_submission), "\n")
cat("  Label:     ", attr(adsl_submission, "label"), "\n")
```

---

## Summary

You can now produce FDA/EMA submission-ready XPT files from R:

**The five xportr transformations (applied in order):**
1. `xportr_type()` — coerce column types (especially Date → numeric for date vars)
2. `xportr_label()` — attach variable labels from spec
3. `xportr_length()` — declare character widths
4. `xportr_format()` — apply SAS format codes (critical: `DATE9.` for dates)
5. `xportr_order()` — reorder variables to match spec

**Key patterns:**
- `xportr_metadata(.df, spec, domain)` — attach spec once, skip `metadata=` in every call
- `xportr(.df, metadata, domain, path, label)` — unified wrapper runs all steps + write
- `xportr_validate(.df, metadata, domain)` — pre/post validation against spec
- `xportr_df_label(.df, label)` — set the dataset-level XPT label
- `purrr::walk()` loop — export multiple datasets cleanly

**Most common submission errors:**
- Date variable stays as `Date` class (needs `type="numeric"` in spec + `format="DATE9."`)
- Flag variable declared with `length=1` but `"Y"`/`"N"` need at least 2
- Character value exceeds declared spec length — detect with `max(nchar(.))`
- Variable order wrong — always run `xportr_order()` last

Next: **Lesson 44** covers `datasetjson` — the modern CDISC Dataset-JSON format for electronic submissions, and how xportr and datasetjson complement each other.
