---
title: "Lesson 44 — datasetjson: CDISC Dataset-JSON"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Master the CDISC Dataset-JSON format and the datasetjson R package.
  Covers the Dataset-JSON v1.1 specification, writing and reading JSON
  datasets, metadata attachment, round-trip fidelity, comparison with XPT,
  and integration with xportr for dual-format submission pipelines.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 44_datasetjson.Rmd
# PURPOSE       : CDISC Dataset-JSON v1.1 — constructing, writing, and reading
#                 JSON submission datasets: dataset_json(), write/read,
#                 NA handling, round-trip fidelity, define.xml linkage.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================
library(learnr)
library(dplyr)
library(tibble)
library(jsonlite)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.eval = FALSE)

# ── Pharmaverseadam datasets ────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}
adsl <- adsl_raw |>
  mutate(ITTFL  = SAFFL,
         DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N")) |>
  select(STUDYID, USUBJID, SUBJID, SITEID, ARM, TRT01P, TRT01A,
         AGE, AGEGR1, SEX, RACE, ETHNIC,
         SAFFL, ITTFL, DIABFL, HEIGHTBL, WEIGHTBL, BMIBL,
         TRTSDT, TRTEDT, TRTDURD, RFSTDTC, RFENDTC,
         DTHFL, COUNTRY)

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  select(STUDYID, USUBJID, AESEQ, AEDECOD, AESOC, AESEV,
         AESTDTC, AEENDTC, AESER, AEREL, AEOUT,
         ASTDT, AENDT, ASTDY, AENDY, TRT01A)

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))

`%||%` <- function(a, b) if (is.null(a)) b else a

# ── Metadata spec (reused from lesson 43) ────────────────────────────────────
adsl_spec <- tribble(
  ~dataset,  ~variable,  ~label,                              ~type,       ~length, ~format,       ~data_type,
  "ADSL","STUDYID",  "Study Identifier",                  "character",  12, NA_character_, "string",
  "ADSL","USUBJID",  "Unique Subject Identifier",         "character",  50, NA_character_, "string",
  "ADSL","SUBJID",   "Subject Identifier for the Study",  "character",  10, NA_character_, "string",
  "ADSL","SITEID",   "Study Site Identifier",             "character",   3, NA_character_, "string",
  "ADSL","ARM",      "Description of Planned Arm",        "character", 200, NA_character_, "string",
  "ADSL","TRT01P",   "Planned Treatment for Period 01",   "character", 200, NA_character_, "string",
  "ADSL","TRT01A",   "Actual Treatment for Period 01",    "character", 200, NA_character_, "string",
  "ADSL","AGE",      "Age",                               "numeric",     8, NA_character_, "integer",
  "ADSL","AGEGR1",   "Pooled Age Group 1",                "character",  10, NA_character_, "string",
  "ADSL","SEX",      "Sex",                               "character",   2, NA_character_, "string",
  "ADSL","RACE",     "Race",                              "character",  64, NA_character_, "string",
  "ADSL","ETHNIC",   "Ethnicity",                         "character",  64, NA_character_, "string",
  "ADSL","SAFFL",    "Safety Population Flag",            "character",   2, NA_character_, "string",
  "ADSL","ITTFL",    "Intent-To-Treat Population Flag",   "character",   2, NA_character_, "string",
  "ADSL","DIABFL",   "Diabetic Flag",                     "character",   2, NA_character_, "string",
  "ADSL","HEIGHTBL", "Baseline Height (cm)",              "numeric",     8, NA_character_, "decimal",
  "ADSL","WEIGHTBL", "Baseline Weight (kg)",              "numeric",     8, NA_character_, "decimal",
  "ADSL","BMIBL",    "Baseline BMI (kg/m^2)",             "numeric",     8, NA_character_, "decimal",
  "ADSL","TRTSDT",   "Date of First Exposure to Treatment","numeric",   8, "DATE9.",       "date",
  "ADSL","TRTEDT",   "Date of Last Exposure to Treatment","numeric",    8, "DATE9.",       "date",
  "ADSL","TRTDURD",  "Total Treatment Duration (Days)",   "numeric",     8, NA_character_, "integer",
  "ADSL","RFSTDTC",  "Subject Reference Start Date/Time","character",  20, NA_character_, "string",
  "ADSL","RFENDTC",  "Subject Reference End Date/Time",  "character",  20, NA_character_, "string",
  "ADSL","DTHFL",    "Death Flag",                       "character",   2, NA_character_, "string",
  "ADSL","COUNTRY",  "Country",                          "character",   3, NA_character_, "string"
)

# ── Mock datasetjson infrastructure ──────────────────────────────────────────
# datasetjson writes/reads the CDISC Dataset-JSON v1.1 format.
# We mock the core API so exercises run in learnr without the package.

# CDISC Dataset-JSON v1.1 structure:
#
# {
#   "datasetJSONVersion": "1.1.0",
#   "fileOID": "...",
#   "creationDateTime": "...",
#   "originator": "...",
#   "sourceSystem": {"name": "R", "version": "..."},
#   "studyOID": "...",
#   "metaDataRef": "define.xml",
#   "clinicalData": {
#     "studyOID": "...",
#     "metaDataVersionOID": "...",
#     "itemGroupData": {
#       "IG.ADSL": {
#         "records": 254,
#         "label": "Subject-Level Analysis Dataset",
#         "items": [
#           {"OID": "IT.ADSL.STUDYID", "name": "STUDYID", "label": "Study Identifier",
#            "type": "string", "length": 12},
#           ...
#         ],
#         "itemData": [
#           ["CDISCPILOT01", "CDISCPILOT01-01-701", "701", ...],
#           ...
#         ]
#       }
#     }
#   }
# }

# dataset_json() — build a Dataset-JSON list object
dataset_json <- function(data,
                         file_oid         = NULL,
                         label            = NULL,
                         domain           = NULL,
                         originator       = "Clinical Programmer",
                         source_system    = "R",
                         source_system_version = R.version.string,
                         study_oid        = NULL,
                         metadata_version = "MDV.1",
                         metadata_ref     = "define.xml",
                         item_metadata    = NULL) {
  domain <- domain %||% deparse(substitute(data))
  domain <- toupper(domain)

  # Build item metadata from data types if not supplied
  if (is.null(item_metadata)) {
    item_metadata <- tibble(
      name  = names(data),
      label = vapply(data, function(v) attr(v,"label") %||% "", character(1)),
      type  = vapply(data, function(v) {
        if (inherits(v, c("Date","POSIXct","POSIXlt"))) "date"
        else if (is.numeric(v)) "decimal"
        else "string"
      }, character(1)),
      length = vapply(data, function(v) {
        w <- attr(v,"width")
        if (!is.null(w)) as.integer(w)
        else if (is.character(v)) max(nchar(v), na.rm=TRUE)
        else 8L
      }, integer(1))
    )
  }

  structure(
    list(
      data             = data,
      file_oid         = file_oid %||% paste0("OID.", domain, ".", format(Sys.time(),"%Y%m%d%H%M%S")),
      label            = label %||% domain,
      domain           = domain,
      originator       = originator,
      source_system    = source_system,
      source_system_version = source_system_version,
      study_oid        = study_oid %||% paste0("STUDY.", unique(data$STUDYID)[1]),
      metadata_version = metadata_version,
      metadata_ref     = metadata_ref,
      item_metadata    = item_metadata,
      created          = format(Sys.time(), "%Y-%m-%dT%H:%M:%S")
    ),
    class = "dataset_json"
  )
}

# print method
print.dataset_json <- function(x, ...) {
  cat("── Dataset-JSON Object ────────────────────────────────\n")
  cat("  Domain         :", x$domain, "\n")
  cat("  Label          :", x$label, "\n")
  cat("  Records        :", nrow(x$data), "\n")
  cat("  Variables      :", ncol(x$data), "\n")
  cat("  fileOID        :", x$file_oid, "\n")
  cat("  studyOID       :", x$study_oid, "\n")
  cat("  metaDataRef    :", x$metadata_ref, "\n")
  cat("  Created        :", x$created, "\n")
  cat("  Source system  :", x$source_system, "\n")
  cat("  Item types     :",
      paste(sort(unique(x$item_metadata$type)), collapse=", "), "\n")
  invisible(x)
}

# write_dataset_json() — serialize to JSON file (mocked)
write_dataset_json <- function(x, path, pretty = FALSE, na = "null") {
  if (!inherits(x, "dataset_json"))
    stop("x must be a dataset_json object; use dataset_json() first")

  # Build the JSON structure
  items <- lapply(seq_len(nrow(x$item_metadata)), function(i) {
    row <- x$item_metadata[i, ]
    item <- list(
      OID    = paste0("IT.", x$domain, ".", row$name),
      name   = row$name,
      label  = row$label,
      type   = row$type,
      length = row$length
    )
    if (row$type == "date") item$displayFormat <- "yyyy-mm-dd"
    item
  })

  # Row data as list of lists
  row_data <- apply(x$data, 1, function(r) as.list(r), simplify = FALSE)

  json_obj <- list(
    datasetJSONVersion = "1.1.0",
    fileOID            = x$file_oid,
    creationDateTime   = x$created,
    originator         = x$originator,
    sourceSystem       = list(
      name    = x$source_system,
      version = x$source_system_version
    ),
    studyOID      = x$study_oid,
    metaDataRef   = x$metadata_ref,
    clinicalData  = list(
      studyOID           = x$study_oid,
      metaDataVersionOID = x$metadata_version,
      itemGroupData      = setNames(
        list(list(
          records  = nrow(x$data),
          label    = x$label,
          items    = items,
          itemData = row_data
        )),
        paste0("IG.", x$domain)
      )
    )
  )

  cat(sprintf("── write_dataset_json ──────────────────────────────────\n"))
  cat(sprintf("  Path       : %s\n", path))
  cat(sprintf("  Domain     : %s\n", x$domain))
  cat(sprintf("  Records    : %d\n", nrow(x$data)))
  cat(sprintf("  Variables  : %d\n", ncol(x$data)))
  cat(sprintf("  JSON ver.  : 1.1.0\n"))
  cat(sprintf("  pretty     : %s\n", pretty))
  cat(sprintf("  File size  : ~%.0f KB (estimated)\n",
              object.size(json_obj) / 1024))

  # Actually serialize (writes to the mock object, not disk in learnr)
  json_str <- if (pretty)
    jsonlite::toJSON(json_obj, auto_unbox=TRUE, pretty=TRUE, na=na)
  else
    jsonlite::toJSON(json_obj, auto_unbox=TRUE, pretty=FALSE, na=na)

  attr(x, "json_string") <- json_str
  cat(sprintf("  [mock] JSON serialized (%d chars)\n", nchar(json_str)))
  cat("──────────────────────────────────────────────────────\n")
  invisible(x)
}

# read_dataset_json() — deserialize JSON file back to data frame (mocked)
read_dataset_json <- function(path) {
  # In real use: reads JSON from disk. Here we simulate with a small inline example.
  cat(sprintf("[mock] Reading Dataset-JSON from: %s\n", path))
  cat("  Returns: dataset_json object with $data data frame\n")
  invisible(NULL)
}

# as_tibble.dataset_json — extract data frame
as_tibble.dataset_json <- function(x, ...) x$data

# dataset_json_metadata() — extract item metadata
dataset_json_metadata <- function(x) x$item_metadata

# set_dataset_metadata() — attach metadata from spec data frame
set_dataset_metadata <- function(dj_obj, metadata, domain = NULL) {
  dom <- domain %||% dj_obj$domain
  spec_dom <- metadata[toupper(metadata$dataset) == toupper(dom), ]

  item_meta <- tibble(
    name   = spec_dom$variable,
    label  = spec_dom$label,
    type   = spec_dom$data_type %||% if_else(spec_dom$type=="numeric","decimal","string"),
    length = as.integer(spec_dom$length)
  )

  dj_obj$item_metadata <- item_meta
  dj_obj
}

# dataset_json_from_spec() — convenience constructor with metadata
dataset_json_from_spec <- function(data, metadata, domain, label = NULL,
                                   study_oid = NULL, ...) {
  dom      <- toupper(domain)
  spec_dom <- metadata[toupper(metadata$dataset) == dom, ]

  # infer data_type if column absent
  if (!"data_type" %in% names(spec_dom)) {
    spec_dom <- spec_dom |>
      mutate(data_type = if_else(type == "numeric", "decimal", "string"))
  }

  item_meta <- tibble(
    name   = spec_dom$variable,
    label  = spec_dom$label,
    type   = spec_dom$data_type,
    length = as.integer(spec_dom$length)
  )

  # Reorder data to spec variable order
  ordered_vars <- intersect(spec_dom$variable, names(data))
  data_ordered <- data[, ordered_vars, drop = FALSE]

  dataset_json(
    data          = data_ordered,
    label         = label %||% dom,
    domain        = dom,
    study_oid     = study_oid,
    item_metadata = item_meta,
    ...
  )
}

# Summarise JSON object — pretty diagnostics
summarise_dataset_json <- function(dj_obj) {
  cat(sprintf("Dataset-JSON Summary — %s\n", dj_obj$domain))
  cat(strrep("─", 60), "\n")
  cat(sprintf("  Records    : %d\n", nrow(dj_obj$data)))
  cat(sprintf("  Variables  : %d\n", nrow(dj_obj$item_metadata)))
  type_tbl <- table(dj_obj$item_metadata$type)
  cat("  Data types :\n")
  for (nm in names(type_tbl))
    cat(sprintf("    %-12s %d variable(s)\n", nm, type_tbl[[nm]]))
  unlabelled <- sum(is.na(dj_obj$item_metadata$label) |
                    dj_obj$item_metadata$label == "")
  cat(sprintf("  Unlabelled : %d variable(s)\n", unlabelled))
  cat(strrep("─", 60), "\n")
  invisible(dj_obj)
}

# compare_xpt_json() — compare two exports for round-trip fidelity
compare_formats <- function(df_original, df_restored, domain = "DATASET") {
  cat(sprintf("── Round-trip comparison: %s ──────────────────────────\n", domain))

  # Row count
  if (nrow(df_original) == nrow(df_restored)) {
    cat(sprintf("  ✓ Row count: %d\n", nrow(df_original)))
  } else {
    cat(sprintf("  ✗ Row count mismatch: original=%d, restored=%d\n",
                nrow(df_original), nrow(df_restored)))
  }

  # Column count
  common_cols <- intersect(names(df_original), names(df_restored))
  if (length(common_cols) == ncol(df_original)) {
    cat(sprintf("  ✓ All %d columns present\n", ncol(df_original)))
  } else {
    missing <- setdiff(names(df_original), names(df_restored))
    cat(sprintf("  ✗ Missing columns: %s\n", paste(missing, collapse=", ")))
  }

  # Value fidelity for character columns
  char_cols  <- common_cols[sapply(df_original[common_cols], is.character)]
  char_match <- 0L
  for (v in char_cols) {
    if (isTRUE(all.equal(df_original[[v]], df_restored[[v]]))) char_match <- char_match + 1L
  }
  cat(sprintf("  ✓ Character columns matching: %d / %d\n",
              char_match, length(char_cols)))
  cat("──────────────────────────────────────────────────────\n")
  invisible(NULL)
}

# peek_json() — show first N characters of a serialised JSON string
peek_json <- function(dj_obj, n = 800) {
  js <- attr(dj_obj, "json_string")
  if (is.null(js)) {
    cat("No JSON string attached — run write_dataset_json() first\n")
    return(invisible(NULL))
  }
  cat(substr(js, 1, n))
  if (nchar(js) > n) cat(sprintf("\n... [%d more chars]\n", nchar(js)-n))
  invisible(js)
}
```

## Overview

**CDISC Dataset-JSON** is a modern, open-standard format for submitting clinical trial data to regulatory authorities. It is designed to complement — and eventually succeed — the SAS XPT transport format. The **datasetjson** R package provides the tools to read and write Dataset-JSON files within the pharmaverse workflow.

> **Why a new format?**
>
> The SAS Version 5 XPT format was designed in the 1980s. Its limitations — 8-character variable names, 200-character labels, no native support for Unicode, maximum file size restrictions — have become pain points as datasets grow and submission standards evolve. Dataset-JSON is JSON-based, human-readable, natively Unicode, and supports modern data types.

### The two formats side by side

| Feature | XPT (SAS V5) | Dataset-JSON v1.1 |
|---------|-------------|------------------|
| Encoding | Binary (SAS transport) | UTF-8 JSON text |
| Variable name max | 8 characters | 32 characters |
| Label max | 200 characters | Unlimited |
| Date type | Numeric (days since 1960-01-01) | ISO 8601 string |
| Human-readable | No | Yes |
| Software dependency | SAS / haven | Any JSON parser |
| FDA acceptance | Required (current) | Pilot submissions accepted |
| CDISC standard | CDISC Submission Data Standards | CDISC Dataset-JSON v1.1 |

---

## 1 — The Dataset-JSON v1.1 Specification

### 1.1 Document structure

A Dataset-JSON file is a single JSON document with this top-level structure:

```{r json-structure-demo, eval=FALSE}
# Conceptual structure of a Dataset-JSON v1.1 file:
{
  "datasetJSONVersion": "1.1.0",
  "fileOID":            "OID.ADSL.20260524",
  "creationDateTime":   "2026-05-24T09:00:00",
  "originator":         "Clinical Programming Team",
  "sourceSystem":       {"name": "R", "version": "4.4.0"},
  "studyOID":           "CDISCPILOT01",
  "metaDataRef":        "define.xml",
  "clinicalData": {
    "studyOID":           "CDISCPILOT01",
    "metaDataVersionOID": "MDV.1",
    "itemGroupData": {
      "IG.ADSL": {
        "records":  254,
        "label":    "Subject-Level Analysis Dataset",
        "items": [
          {
            "OID":    "IT.ADSL.STUDYID",
            "name":   "STUDYID",
            "label":  "Study Identifier",
            "type":   "string",
            "length": 12
          },
          {
            "OID":    "IT.ADSL.AGE",
            "name":   "AGE",
            "label":  "Age",
            "type":   "integer"
          },
          {
            "OID":             "IT.ADSL.TRTSDT",
            "name":            "TRTSDT",
            "label":           "Date of First Exposure to Treatment",
            "type":            "date",
            "displayFormat":   "yyyy-mm-dd"
          }
        ],
        "itemData": [
          ["CDISCPILOT01", "CDISCPILOT01-01-701", 67, "2014-01-02"],
          ["CDISCPILOT01", "CDISCPILOT01-01-702", 72, "2014-01-15"],
          ...
        ]
      }
    }
  }
}
```

### 1.2 Key structural elements

| Element | Purpose |
|---------|---------|
| `datasetJSONVersion` | Spec version — must be `"1.1.0"` for current standard |
| `fileOID` | Unique identifier for this file |
| `studyOID` | Links to the study's define.xml |
| `metaDataRef` | Path/name of the accompanying define.xml |
| `clinicalData.itemGroupData` | Map of dataset OIDs to dataset contents |
| `items` | Variable-level metadata (name, label, type, length) |
| `itemData` | Row data — each row is a JSON array of values |

### 1.3 Dataset-JSON data types

| JSON type | Maps to | SAS equivalent |
|-----------|---------|---------------|
| `"string"` | character | $CHAR |
| `"integer"` | integer numeric | numeric (format `8.`) |
| `"decimal"` | double numeric | numeric (format `8.2`) |
| `"date"` | Date (ISO 8601 `"yyyy-mm-dd"`) | numeric + `DATE9.` format |
| `"datetime"` | POSIXct (ISO 8601) | numeric + `DATETIME20.` |
| `"time"` | character HH:MM:SS | $TIME |

### 1.4 Quiz

```{r quiz-structure, echo=FALSE}
quiz(
  question(
    "In Dataset-JSON v1.1, how are date values represented in the `itemData` arrays?",
    answer("As SAS date integers (days since 1960-01-01)"),
    answer("As ISO 8601 date strings (e.g., '2024-03-15')", correct = TRUE),
    answer("As Unix timestamps (seconds since 1970-01-01)"),
    answer("As character strings in the format 'DDMMMYYYY'"),
    allow_retry = TRUE
  ),
  question(
    "What is the key structural difference between Dataset-JSON and XPT regarding variable names?",
    answer("XPT allows 32-character names; Dataset-JSON allows only 8"),
    answer("Both allow only 8-character names"),
    answer("Dataset-JSON allows 32-character names while XPT is limited to 8", correct = TRUE),
    answer("Dataset-JSON has no limit on variable name length"),
    allow_retry = TRUE
  ),
  question(
    "Which field in a Dataset-JSON file links the data to its corresponding define.xml?",
    answer("`datasetJSONVersion`"),
    answer("`studyOID`"),
    answer("`metaDataRef`", correct = TRUE),
    answer("`fileOID`"),
    allow_retry = TRUE
  )
)
```

---

## 2 — Creating a `dataset_json` Object

### 2.1 `dataset_json()` — the basic constructor

```{r basic-constructor-demo, eval=FALSE}
# Simplest form: data frame + domain
adsl_dj <- dataset_json(
  data      = adsl,
  domain    = "ADSL",
  label     = "Subject-Level Analysis Dataset",
  study_oid = "CDISCPILOT01",
  originator = "Clinical Programming — Pharmasave"
)

print(adsl_dj)
```

### 2.2 Exercise — create a dataset_json object

Create a `dataset_json` object for ADSL with appropriate metadata fields.

```{r basic-constructor-ex, exercise=TRUE, exercise.lines=16}
adsl_dj <- dataset_json(
  data       = ___,
  domain     = "ADSL",
  label      = ___,
  study_oid  = "CDISCPILOT01",
  originator = "Clinical Programming"
)

print(___)
```

```{r basic-constructor-ex-solution}
adsl_dj <- dataset_json(
  data       = adsl,
  domain     = "ADSL",
  label      = "Subject-Level Analysis Dataset",
  study_oid  = "CDISCPILOT01",
  originator = "Clinical Programming"
)

print(adsl_dj)
```

### 2.3 `dataset_json_from_spec()` — metadata-driven constructor

The spec-driven constructor is preferred in production because it attaches variable labels and types in one call:

```{r spec-constructor-demo, eval=FALSE}
adsl_dj <- dataset_json_from_spec(
  data      = adsl,
  metadata  = adsl_spec,
  domain    = "ADSL",
  label     = "Subject-Level Analysis Dataset",
  study_oid = "CDISCPILOT01"
)

# Inspect item metadata
dataset_json_metadata(adsl_dj)
```

### 2.4 Exercise — spec-driven constructor

```{r spec-constructor-ex, exercise=TRUE, exercise.lines=14}
adsl_dj_spec <- dataset_json_from_spec(
  data      = ___,
  metadata  = adsl_spec,
  domain    = ___,
  label     = "Subject-Level Analysis Dataset",
  study_oid = "CDISCPILOT01"
)

cat("── Item metadata (first 8 rows) ──\n")
print(head(dataset_json_metadata(___), 8))
```

```{r spec-constructor-ex-solution}
adsl_dj_spec <- dataset_json_from_spec(
  data      = adsl,
  metadata  = adsl_spec,
  domain    = "ADSL",
  label     = "Subject-Level Analysis Dataset",
  study_oid = "CDISCPILOT01"
)

cat("── Item metadata (first 8 rows) ──\n")
print(head(dataset_json_metadata(adsl_dj_spec), 8))
```

---

## 3 — Writing Dataset-JSON Files

### 3.1 `write_dataset_json()`

```{r write-demo, eval=FALSE}
adsl_dj <- dataset_json_from_spec(
  data      = adsl,
  metadata  = adsl_spec,
  domain    = "ADSL",
  label     = "Subject-Level Analysis Dataset",
  study_oid = "CDISCPILOT01"
)

write_dataset_json(
  x      = adsl_dj,
  path   = "submission/adam/adsl.json",
  pretty = FALSE    # compact for submission; TRUE for debugging
)
```

### 3.2 Exercise — write ADSL to JSON

```{r write-ex, exercise=TRUE, exercise.lines=20}
adsl_dj <- dataset_json_from_spec(
  data     = adsl,
  metadata = adsl_spec,
  domain   = "ADSL",
  label    = "Subject-Level Analysis Dataset"
)

result <- write_dataset_json(
  x      = ___,
  path   = "adam/adsl.json",
  pretty = ___     # use FALSE for compact output
)

cat("\nObject class:", class(result), "\n")
```

```{r write-ex-solution}
adsl_dj <- dataset_json_from_spec(
  data     = adsl,
  metadata = adsl_spec,
  domain   = "ADSL",
  label    = "Subject-Level Analysis Dataset"
)

result <- write_dataset_json(
  x      = adsl_dj,
  path   = "adam/adsl.json",
  pretty = FALSE
)

cat("\nObject class:", class(result), "\n")
```

### 3.3 Inspect the JSON output with `peek_json()`

```{r peek-demo, exercise=TRUE, exercise.lines=14}
adsl_dj <- dataset_json_from_spec(
  data     = adsl,
  metadata = adsl_spec,
  domain   = "ADSL",
  label    = "Subject-Level Analysis Dataset"
)

adsl_written <- write_dataset_json(adsl_dj, path="adam/adsl.json", pretty=TRUE)

cat("\n── First 1000 chars of JSON ──\n")
peek_json(adsl_written, n = 1000)
```

### 3.4 Pretty vs compact output

```{r pretty-vs-compact, eval=FALSE}
# For submission: compact (smaller file, faster to parse by agency systems)
write_dataset_json(adsl_dj, path="submission/adsl.json", pretty=FALSE)
# ~50% smaller file size than pretty-printed

# For review / debugging: pretty-printed (human-readable)
write_dataset_json(adsl_dj, path="review/adsl_debug.json", pretty=TRUE)

# File size comparison:
# adsl.json (compact): ~145 KB for 254 subjects × 25 vars
# adsl_debug.json (pretty): ~290 KB
```

---

## 4 — Reading Dataset-JSON Files

### 4.1 `read_dataset_json()`

```{r read-demo, eval=FALSE}
# Read a Dataset-JSON file back into R
adsl_restored <- read_dataset_json("submission/adam/adsl.json")

# Inspect the restored object
print(adsl_restored)

# Extract the data frame
adsl_df <- as_tibble(adsl_restored)
head(adsl_df)
```

### 4.2 Round-trip fidelity

A critical property: data must survive a write → read cycle intact. Use `compare_formats()` to verify:

```{r round-trip-ex, exercise=TRUE, exercise.lines=22}
# Simulate a round-trip: build → write → restore
adsl_dj <- dataset_json_from_spec(
  data     = adsl,
  metadata = adsl_spec,
  domain   = "ADSL",
  label    = "Subject-Level Analysis Dataset"
)

# Write (mock — stores JSON string in attribute)
adsl_written <- write_dataset_json(adsl_dj, path="adam/adsl.json")

# In real usage, read_dataset_json() restores from disk.
# Here we simulate by extracting data directly from the object:
adsl_restored_df <- as_tibble(adsl_dj)   # same data (round-trip fidelity check)

# Compare original to restored
compare_formats(
  df_original = ___,
  df_restored = ___,
  domain      = "ADSL"
)
```

```{r round-trip-ex-solution}
adsl_dj <- dataset_json_from_spec(
  data     = adsl,
  metadata = adsl_spec,
  domain   = "ADSL",
  label    = "Subject-Level Analysis Dataset"
)

adsl_written <- write_dataset_json(adsl_dj, path="adam/adsl.json")
adsl_restored_df <- as_tibble(adsl_dj)

compare_formats(
  df_original = adsl,
  df_restored = adsl_restored_df,
  domain      = "ADSL"
)
```

---

## 5 — NA Handling in Dataset-JSON

### 5.1 The `na` argument

JSON has no native concept of a missing value for all types. Dataset-JSON handles NAs via the `na` argument to `write_dataset_json()`:

| `na` value | JSON output for NA | When to use |
|-----------|-------------------|-------------|
| `"null"` (default) | `null` | FDA/EMA submission — most common |
| `""` | `""` | Legacy compatibility |
| `"NA"` | `"NA"` | Debugging only — creates string values |

```{r na-handling-demo, eval=FALSE}
# Submission-ready: NA → null
write_dataset_json(adsl_dj, path="adsl.json", na="null")

# Results in JSON like:
# {"DTHFL": null, "TRTDURD": null}

# NOT this (never use na="NA" for submissions):
# {"DTHFL": "NA", "TRTDURD": "NA"}  ← creates string "NA", wrong type
```

### 5.2 Quiz

```{r quiz-na, echo=FALSE}
quiz(
  question(
    "When writing a Dataset-JSON file for FDA submission, which `na` argument value should you use?",
    answer("`na = \"NA\"` — explicit string"),
    answer("`na = \"\"` — empty string"),
    answer("`na = \"null\"` — JSON null", correct = TRUE),
    answer("`na = 0` — zero for numeric NAs"),
    allow_retry = TRUE
  ),
  question(
    "A Dataset-JSON file has `\"type\": \"date\"` for variable TRTSDT. How should date values appear in `itemData`?",
    answer("As SAS date integers: `20454`"),
    answer("As ISO 8601 strings: `\"2026-01-15\"`", correct = TRUE),
    answer("As R Date objects: `Date(20454)`"),
    answer("As character strings with SAS format: `\"15JAN2026\"`"),
    allow_retry = TRUE
  )
)
```

---

## 6 — Dataset-JSON vs XPT: When to Use Each

### 6.1 Decision guide

```
Is this a current FDA eCTD submission?
├── Yes → XPT is REQUIRED (FDA Study Data Technical Conformance Guide)
│          Dataset-JSON may ALSO be included as supplementary
└── No  → use Dataset-JSON for:
           • EMA submissions (increasingly accepting JSON)
           • Internal data warehouses and review platforms
           • Data exchange with clinical data management systems
           • Datasets > 5 GB (XPT has de facto file size limits)
           • Non-ASCII characters in labels / values
```

### 6.2 Dual-format export pipeline

Many sponsors produce **both** XPT and Dataset-JSON from the same source:

```{r dual-format-demo, eval=FALSE}
library(xportr)
library(datasetjson)
library(purrr)

export_dual_format <- function(data, domain, spec, output_dir,
                               label, study_oid) {

  # ── XPT (required for FDA eCTD) ───────────────────────────────────────
  xpt_path <- file.path(output_dir, "xpt", paste0(tolower(domain), ".xpt"))

  data |>
    xportr_metadata(spec, domain = domain) |>
    xportr_type()   |>
    xportr_label()  |>
    xportr_length() |>
    xportr_format() |>
    xportr_order()  |>
    xportr_df_label(label) |>
    xportr_write(xpt_path, label = label)

  # ── Dataset-JSON (supplementary / modern pipeline) ────────────────────
  json_path <- file.path(output_dir, "json", paste0(tolower(domain), ".json"))

  dataset_json_from_spec(
    data      = data,
    metadata  = spec,
    domain    = domain,
    label     = label,
    study_oid = study_oid
  ) |>
    write_dataset_json(path = json_path, na = "null")

  message(sprintf("[%s] XPT: %s | JSON: %s", domain, xpt_path, json_path))
}
```

### 6.3 Exercise — dual format export

Build a dual-format export for ADSL using the mock functions.

```{r dual-format-ex, exercise=TRUE, exercise.lines=40}
# Helper: run xportr pipeline (from lesson 43 mocks in setup)
# We re-define a minimal version here for clarity
mock_xportr_pipeline <- function(data, spec, domain, label) {
  cat(sprintf("[xportr] %s: %d rows × %d cols → adam/%s.xpt\n",
              domain, nrow(data), ncol(data), tolower(domain)))
  invisible(data)
}

# Step 1: XPT export
cat("=== XPT Export ===\n")
mock_xportr_pipeline(
  data   = ___,
  spec   = adsl_spec,
  domain = "ADSL",
  label  = "Subject-Level Analysis Dataset"
)

# Step 2: Dataset-JSON export
cat("\n=== Dataset-JSON Export ===\n")
adsl_dj <- dataset_json_from_spec(
  data      = ___,
  metadata  = ___,
  domain    = "ADSL",
  label     = "Subject-Level Analysis Dataset",
  study_oid = "CDISCPILOT01"
)

write_dataset_json(adsl_dj, path = "adam/adsl.json", na = ___)

# Step 3: Summarise both outputs
cat("\n=== Summary ===\n")
summarise_dataset_json(___)
```

```{r dual-format-ex-solution}
mock_xportr_pipeline <- function(data, spec, domain, label) {
  cat(sprintf("[xportr] %s: %d rows × %d cols → adam/%s.xpt\n",
              domain, nrow(data), ncol(data), tolower(domain)))
  invisible(data)
}

cat("=== XPT Export ===\n")
mock_xportr_pipeline(
  data   = adsl,
  spec   = adsl_spec,
  domain = "ADSL",
  label  = "Subject-Level Analysis Dataset"
)

cat("\n=== Dataset-JSON Export ===\n")
adsl_dj <- dataset_json_from_spec(
  data      = adsl,
  metadata  = adsl_spec,
  domain    = "ADSL",
  label     = "Subject-Level Analysis Dataset",
  study_oid = "CDISCPILOT01"
)

write_dataset_json(adsl_dj, path = "adam/adsl.json", na = "null")

cat("\n=== Summary ===\n")
summarise_dataset_json(adsl_dj)
```

---

## 7 — Working with Multiple Datasets

### 7.1 The `purrr::walk()` pattern for JSON export

```{r multi-json-demo, eval=FALSE}
library(purrr)
library(datasetjson)

datasets <- list(ADSL=adsl, ADAE=adae, ADLB=adlb)

full_spec <- bind_rows(adsl_spec, adae_spec, adlb_spec)

ds_labels <- c(
  ADSL = "Subject-Level Analysis Dataset",
  ADAE = "Adverse Events Analysis Dataset",
  ADLB = "Laboratory Test Results"
)

output_dir <- "submission/json"
dir.create(output_dir, recursive=TRUE, showWarnings=FALSE)

walk(names(datasets), function(nm) {
  dataset_json_from_spec(
    data      = datasets[[nm]],
    metadata  = full_spec,
    domain    = nm,
    label     = ds_labels[[nm]],
    study_oid = "CDISCPILOT01"
  ) |>
    write_dataset_json(
      path = file.path(output_dir, paste0(tolower(nm), ".json")),
      na   = "null"
    )

  message(sprintf("✓ %s exported", nm))
})
```

### 7.2 Exercise — export ADSL and ADAE

```{r multi-json-ex, exercise=TRUE, exercise.lines=30}
library(purrr)

datasets <- list(
  ADSL = adsl,
  ADAE = ___
)

# Build a combined spec (ADSL only for this exercise — ADAE spec not loaded)
# Use adsl_spec for ADSL and a minimal inline spec for ADAE
adae_mini_spec <- adae |>
  names() |>
  tibble::tibble(variable=___, dataset="ADAE") |>
  mutate(
    label     = variable,
    type      = if_else(sapply(adae[variable], is.numeric), "numeric", "character"),
    data_type = if_else(type=="numeric", "decimal", "string"),
    length    = 200L,
    format    = NA_character_
  )

combined_spec <- bind_rows(adsl_spec, adae_mini_spec)

walk(names(datasets), function(nm) {
  dj <- dataset_json_from_spec(
    data      = datasets[[nm]],
    metadata  = ___,
    domain    = ___,
    study_oid = "CDISCPILOT01"
  )
  write_dataset_json(dj, path=paste0("adam/",tolower(nm),".json"), na="null")
})

cat("Export complete\n")
```

```{r multi-json-ex-solution}
library(purrr)

datasets <- list(
  ADSL = adsl,
  ADAE = adae
)

adae_mini_spec <- adae |>
  names() |>
  (\(nms) tibble::tibble(variable=nms, dataset="ADAE"))() |>
  mutate(
    label     = variable,
    type      = if_else(sapply(adae[variable], is.numeric), "numeric", "character"),
    data_type = if_else(type=="numeric", "decimal", "string"),
    length    = 200L,
    format    = NA_character_
  )

combined_spec <- bind_rows(adsl_spec, adae_mini_spec)

walk(names(datasets), function(nm) {
  dj <- dataset_json_from_spec(
    data      = datasets[[nm]],
    metadata  = combined_spec,
    domain    = nm,
    study_oid = "CDISCPILOT01"
  )
  write_dataset_json(dj, path=paste0("adam/",tolower(nm),".json"), na="null")
})

cat("Export complete\n")
```

---

## 8 — Dataset-JSON and define.xml

### 8.1 The relationship

Every Dataset-JSON submission package must include a `define.xml` that:

1. Declares all datasets and their variables
2. Provides the `OID` values referenced in each JSON file's `metaDataRef` field
3. Includes codelist, value level metadata, and computational algorithms

```{r define-relationship, eval=FALSE}
# The flow:
#
#   define.xml ←──────────────────────── metaDataRef
#       │                                     │
#       │  declares                           │  references
#       ▼                                     │
#   IG.ADSL (ItemGroupDef OID)          adsl.json
#   IT.ADSL.USUBJID (ItemDef OID)     "IT.ADSL.USUBJID" in items[]
#   IT.ADSL.AGE (ItemDef OID)         "IT.ADSL.AGE" in items[]
#
# The OIDs in the JSON items[] MUST match OIDs in define.xml

# Correct OID pattern:
# fileOID:     "OID.STUDY001.ADSL.20260524"
# itemGroup:   "IG.ADSL"
# item OIDs:   "IT.ADSL.STUDYID", "IT.ADSL.USUBJID", ...
```

### 8.2 Generating define.xml with `metacore` + `definexml`

In a full pharmaverse submission pipeline:

```{r define-gen, eval=FALSE}
library(metacore)
library(definexml)   # pharmaverse package for define.xml generation

# Load specs from Excel workbook
mc <- spec_to_metacore("specs/adam_specs.xlsx")

# Generate define.xml
define <- mc |>
  metacore_to_define(
    study_oid     = "CDISCPILOT01",
    metadata_version_oid = "MDV.1",
    std_oid       = "CDISC/ADAM/3.2"
  )

write_define(define, path = "submission/define.xml")

# The Dataset-JSON files then reference this define.xml:
adsl_dj <- dataset_json_from_spec(
  data         = adsl,
  metadata     = mc,
  domain       = "ADSL",
  metadata_ref = "define.xml"   # must match the written file name
)
```

---

## 9 — Validation and Quality Checks

### 9.1 What to validate before submission

```{r validation-checklist, eval=FALSE}
# 1. All OIDs match define.xml
#    → Check item OIDs in JSON match ItemDef OIDs in define.xml

# 2. Variable count matches define.xml
#    → itemGroupData.IG.ADSL records count == nrow(adsl)

# 3. No unlabelled variables
adsl_dj$item_metadata |>
  filter(is.na(label) | label == "")
# Should be empty

# 4. All date variables have type="date" (not "decimal")
adsl_dj$item_metadata |>
  filter(name %in% c("TRTSDT","TRTEDT","ASTDT","AENDT")) |>
  select(name, type)
# type should be "date" not "decimal"

# 5. NA encoding is "null" (not "NA" or "")
write_dataset_json(adsl_dj, path="check.json", na="null")

# 6. fileOID is unique across all files in the package
# (enforce in your export pipeline by including a timestamp)
```

### 9.2 Exercise — validate item metadata completeness

Check whether all variables in `adsl_dj` have labels and correct types.

```{r validate-meta-ex, exercise=TRUE, exercise.lines=20}
adsl_dj <- dataset_json_from_spec(
  data     = adsl,
  metadata = adsl_spec,
  domain   = "ADSL",
  label    = "Subject-Level Analysis Dataset"
)

meta <- dataset_json_metadata(adsl_dj)

# Check 1: any unlabelled variables?
unlabelled <- meta |> filter(is.na(label) | label == "")
cat("Unlabelled variables:", nrow(___), "\n")

# Check 2: date variables with wrong type
date_vars  <- c("TRTSDT", "TRTEDT")
date_types <- meta |>
  filter(name %in% ___) |>
  select(name, ___)

cat("\nDate variable types:\n")
print(date_types)

# Check 3: string variables with length > 200
long_strings <- meta |>
  filter(type == "string", ___ > 200) |>
  arrange(desc(length))

cat("\nString variables > 200 chars:", nrow(long_strings), "\n")
```

```{r validate-meta-ex-solution}
adsl_dj <- dataset_json_from_spec(
  data     = adsl,
  metadata = adsl_spec,
  domain   = "ADSL",
  label    = "Subject-Level Analysis Dataset"
)

meta <- dataset_json_metadata(adsl_dj)

unlabelled <- meta |> filter(is.na(label) | label == "")
cat("Unlabelled variables:", nrow(unlabelled), "\n")

date_vars  <- c("TRTSDT", "TRTEDT")
date_types <- meta |>
  filter(name %in% date_vars) |>
  select(name, type)

cat("\nDate variable types:\n")
print(date_types)

long_strings <- meta |>
  filter(type == "string", length > 200) |>
  arrange(desc(length))

cat("\nString variables > 200 chars:", nrow(long_strings), "\n")
```

---

## 10 — Capstone: Full Submission JSON Package

Build a complete, validated Dataset-JSON submission package for ADSL and ADAE.

```{r capstone-json, exercise=TRUE, exercise.eval=FALSE, exercise.lines=70}
library(purrr)

# ── Step 1: Prepare datasets ──────────────────────────────────────────────
cat("=== Step 1: Prepare datasets ===\n")

adsl_clean <- adsl |>
  select(any_of(adsl_spec$variable)) |>
  mutate(COUNTRY = substr(COUNTRY, 1, 3))

adae_clean <- adae |>
  select(STUDYID, USUBJID, AESEQ, AEDECOD, AESOC, AESEV,
         AESTDTC, AEENDTC, AESER, AEREL, AEOUT,
         ASTDT, AENDT, ASTDY, AENDY, TRT01A)

cat(sprintf("  ADSL: %d rows × %d cols\n", nrow(adsl_clean), ncol(adsl_clean)))
cat(sprintf("  ADAE: %d rows × %d cols\n", nrow(adae_clean), ncol(adae_clean)))

# ── Step 2: Build minimal ADAE spec ──────────────────────────────────────
adae_mini <- tibble(
  dataset   = "ADAE",
  variable  = names(adae_clean),
  label     = c("Study Identifier", "Unique Subject Identifier",
                "Sequence Number", "Dictionary-Derived Term",
                "Primary System Organ Class", "Severity/Intensity",
                "Start Date/Time of AE", "End Date/Time of AE",
                "Serious AE", "Causality", "Outcome of AE",
                "Analysis Start Date", "Analysis End Date",
                "Analysis Start Relative Day", "Analysis End Relative Day",
                "Actual Treatment for Period 01"),
  type      = c(rep("character",2), "numeric", rep("character",3),
                rep("character",4), rep("character",2),
                rep("numeric",4), "character"),
  data_type = c(rep("string",2), "integer", rep("string",3),
                rep("string",4), rep("string",2),
                rep("date",2), rep("integer",2), "string"),
  length    = c(12L,50L,8L,200L,200L,10L,20L,20L,2L,16L,60L,8L,8L,8L,8L,200L),
  format    = c(rep(NA_character_,11), "DATE9.","DATE9.",
                rep(NA_character_,3))
)

combined_spec <- bind_rows(adsl_spec, adae_mini)

# ── Step 3: Build dataset_json objects ────────────────────────────────────
cat("\n=== Step 3: Build JSON objects ===\n")

datasets <- list(
  ADSL = list(data=___, label="Subject-Level Analysis Dataset"),
  ADAE = list(data=___, label="Adverse Events Analysis Dataset")
)

dj_objects <- imap(datasets, function(ds, nm) {
  cat(sprintf("  Building %s...\n", nm))
  dataset_json_from_spec(
    data      = ds$data,
    metadata  = ___,
    domain    = nm,
    label     = ds$label,
    study_oid = "CDISCPILOT01"
  )
})

# ── Step 4: Validate ──────────────────────────────────────────────────────
cat("\n=== Step 4: Validate ===\n")
walk(names(dj_objects), function(nm) {
  meta <- dataset_json_metadata(dj_objects[[nm]])
  unlabelled <- sum(is.na(meta$label) | meta$label == "")
  cat(sprintf("  %s: %d variables, %d unlabelled\n",
              nm, nrow(meta), unlabelled))
})

# ── Step 5: Write ─────────────────────────────────────────────────────────
cat("\n=== Step 5: Write JSON files ===\n")
walk(names(dj_objects), function(nm) {
  write_dataset_json(
    dj_objects[[nm]],
    path = paste0("submission/", tolower(nm), ".json"),
    na   = ___
  )
})

# ── Step 6: Summary ──────────────────────────────────────────────────────
cat("\n=== Final Summary ===\n")
walk(names(dj_objects), function(nm) summarise_dataset_json(dj_objects[[nm]]))
```

```{r capstone-json-solution}
library(purrr)

cat("=== Step 1: Prepare datasets ===\n")
adsl_clean <- adsl |>
  select(any_of(adsl_spec$variable)) |>
  mutate(COUNTRY = substr(COUNTRY, 1, 3))

adae_clean <- adae |>
  select(STUDYID, USUBJID, AESEQ, AEDECOD, AESOC, AESEV,
         AESTDTC, AEENDTC, AESER, AEREL, AEOUT,
         ASTDT, AENDT, ASTDY, AENDY, TRT01A)

cat(sprintf("  ADSL: %d rows × %d cols\n", nrow(adsl_clean), ncol(adsl_clean)))
cat(sprintf("  ADAE: %d rows × %d cols\n", nrow(adae_clean), ncol(adae_clean)))

adae_mini <- tibble(
  dataset   = "ADAE",
  variable  = names(adae_clean),
  label     = c("Study Identifier", "Unique Subject Identifier",
                "Sequence Number", "Dictionary-Derived Term",
                "Primary System Organ Class", "Severity/Intensity",
                "Start Date/Time of AE", "End Date/Time of AE",
                "Serious AE", "Causality", "Outcome of AE",
                "Analysis Start Date", "Analysis End Date",
                "Analysis Start Relative Day", "Analysis End Relative Day",
                "Actual Treatment for Period 01"),
  type      = c(rep("character",2), "numeric", rep("character",3),
                rep("character",4), rep("character",2),
                rep("numeric",4), "character"),
  data_type = c(rep("string",2), "integer", rep("string",3),
                rep("string",4), rep("string",2),
                rep("date",2), rep("integer",2), "string"),
  length    = c(12L,50L,8L,200L,200L,10L,20L,20L,2L,16L,60L,8L,8L,8L,8L,200L),
  format    = c(rep(NA_character_,11), "DATE9.","DATE9.",
                rep(NA_character_,3))
)

combined_spec <- bind_rows(adsl_spec, adae_mini)

cat("\n=== Step 3: Build JSON objects ===\n")
datasets <- list(
  ADSL = list(data=adsl_clean, label="Subject-Level Analysis Dataset"),
  ADAE = list(data=adae_clean, label="Adverse Events Analysis Dataset")
)

dj_objects <- imap(datasets, function(ds, nm) {
  cat(sprintf("  Building %s...\n", nm))
  dataset_json_from_spec(
    data      = ds$data,
    metadata  = combined_spec,
    domain    = nm,
    label     = ds$label,
    study_oid = "CDISCPILOT01"
  )
})

cat("\n=== Step 4: Validate ===\n")
walk(names(dj_objects), function(nm) {
  meta <- dataset_json_metadata(dj_objects[[nm]])
  unlabelled <- sum(is.na(meta$label) | meta$label == "")
  cat(sprintf("  %s: %d variables, %d unlabelled\n",
              nm, nrow(meta), unlabelled))
})

cat("\n=== Step 5: Write JSON files ===\n")
walk(names(dj_objects), function(nm) {
  write_dataset_json(
    dj_objects[[nm]],
    path = paste0("submission/", tolower(nm), ".json"),
    na   = "null"
  )
})

cat("\n=== Final Summary ===\n")
walk(names(dj_objects), function(nm) summarise_dataset_json(dj_objects[[nm]]))
```

---

## Summary

You can now produce CDISC Dataset-JSON v1.1 files from R for modern electronic submissions:

**Format fundamentals:**
- Dataset-JSON stores data as UTF-8 JSON with `items[]` (variable metadata) and `itemData[]` (row arrays)
- Dates are ISO 8601 strings (`"yyyy-mm-dd"`), not SAS numeric integers
- JSON data types: `"string"`, `"integer"`, `"decimal"`, `"date"`, `"datetime"`
- `na = "null"` is the correct NA encoding for submission files

**Key functions:**
- `dataset_json(data, domain, label, ...)` — basic constructor
- `dataset_json_from_spec(data, metadata, domain, ...)` — metadata-driven constructor (preferred)
- `write_dataset_json(x, path, pretty, na)` — serialize to JSON file
- `read_dataset_json(path)` — deserialize from JSON file
- `dataset_json_metadata(dj)` — extract item metadata table
- `summarise_dataset_json(dj)` — diagnostics summary

**Submission strategy:**
- **XPT is still required** for FDA eCTD submissions — Dataset-JSON is supplementary or for EMA/pilot
- Dual-format pipelines produce both XPT and JSON from the same source data
- Every Dataset-JSON package must include a matching `define.xml` with consistent OIDs
- `metaDataRef = "define.xml"` in the JSON links it to the define file

Next: **Lesson 45** covers `logrx` — automated R session logging for 21 CFR Part 11 compliance, capturing package versions, environment details, and execution metadata alongside every analysis run.
