---
title: "Lesson 45 — logrx: Automated R Session Logging"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Master logrx for producing audit-ready logs from R analysis scripts.
  Covers axecute(), log structure, session metadata capture, error/warning
  recording, GxP compliance patterns, log review, and integration with
  CI/CD pipelines for clinical submissions.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 45_logrx.Rmd
# PURPOSE       : logrx package — GxP-compliant program logging with axecute():
#                 log structure (HEADER/SESSION/MESSAGES/WARNINGS/
#                 ERRORS/STATUS), batch execution, CI/CD integration.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================
library(learnr)
library(dplyr)
library(tibble)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.eval = FALSE)

`%||%` <- function(a, b) if (is.null(a)) b else a

# ── Pharmaverseadam datasets ────────────────────────────────────────────────
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}
adsl <- adsl_raw |> mutate(
  ITTFL  = SAFFL,
  DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"))

# ── Mock logrx infrastructure ────────────────────────────────────────────────
# logrx wraps R script execution and produces structured log files.
# Core function: axecute(file) — runs a script and writes a log.
# We mock the full log generation pipeline for learnr exercises.

# Log section builder
.log_section <- function(title, ...) {
  lines <- c(strrep("─", 60),
             paste0("  ", title),
             strrep("─", 60),
             ...)
  paste(lines, collapse="\n")
}

# Capture session info for log header
.log_session_info <- function() {
  si <- sessionInfo()
  pkgs <- si$otherPkgs
  pkg_str <- if (length(pkgs) > 0)
    paste(vapply(pkgs, function(p) sprintf("%s %s", p$Package, p$Version),
                 character(1)), collapse=", ")
  else "(none)"
  list(
    r_version  = paste(si$R.version$major, si$R.version$minor, sep="."),
    platform   = si$platform,
    locale     = Sys.getlocale("LC_TIME"),
    packages   = pkg_str,
    user       = Sys.info()[["user"]] %||% "unknown",
    node       = Sys.info()[["nodename"]] %||% "unknown",
    datetime   = format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
  )
}

# Main axecute() mock
axecute <- function(file,
                    log_name       = NULL,
                    log_path       = getwd(),
                    working_dir    = NULL,
                    quiet          = FALSE,
                    show_log       = TRUE,
                    log_breaks     = TRUE,
                    remove_log_na  = FALSE) {

  if (!file.exists(file))
    stop(sprintf("Script not found: %s", file))

  log_name <- log_name %||%
    paste0(tools::file_path_sans_ext(basename(file)), ".log")
  log_file  <- file.path(log_path, log_name)
  script_nm <- basename(file)

  si        <- .log_session_info()
  start_tm  <- proc.time()
  start_dt  <- Sys.time()

  # Capture output + errors from script execution
  msgs <- warnings_caught <- errors_caught <- character(0)

  script_result <- tryCatch({
    capture.output({
      withCallingHandlers(
        source(file, echo = TRUE, local = new.env(parent = globalenv())),
        message = function(m) { msgs <<- c(msgs, conditionMessage(m)); invokeRestart("muffleMessage") },
        warning = function(w) { warnings_caught <<- c(warnings_caught, conditionMessage(w)); invokeRestart("muffleWarning") }
      )
    })
  }, error = function(e) conditionMessage(e))

  elapsed   <- (proc.time() - start_tm)[["elapsed"]]
  end_dt    <- Sys.time()
  succeeded <- !is.character(script_result) || length(grep("Error", script_result)) == 0

  # Build log lines
  log_lines <- c(
    strrep("═", 62),
    sprintf("  LOG: %s", log_name),
    sprintf("  Script: %s", script_nm),
    strrep("═", 62),
    "",
    .log_section("HEADER",
      sprintf("  Script        : %s", file),
      sprintf("  Log           : %s", log_file),
      sprintf("  User          : %s", si$user),
      sprintf("  Node          : %s", si$node),
      sprintf("  Start time    : %s", format(start_dt, "%Y-%m-%d %H:%M:%S")),
      sprintf("  End time      : %s", format(end_dt,   "%Y-%m-%d %H:%M:%S")),
      sprintf("  Elapsed (s)   : %.2f", elapsed)
    ),
    "",
    .log_section("SESSION INFO",
      sprintf("  R version     : %s", si$r_version),
      sprintf("  Platform      : %s", si$platform),
      sprintf("  Attached pkgs : %s", si$packages)
    ),
    "",
    .log_section("MESSAGES",
      if (length(msgs) > 0) paste0("  ", msgs) else "  (none)"
    ),
    "",
    .log_section("WARNINGS",
      if (length(warnings_caught) > 0)
        paste0("  WARNING: ", warnings_caught)
      else "  (none)"
    ),
    "",
    .log_section("ERRORS",
      if (!succeeded)
        paste0("  ERROR: ", tail(script_result, 1))
      else "  (none)"
    ),
    "",
    .log_section("STATUS",
      sprintf("  Status : %s", if(succeeded) "SUCCESS" else "FAILURE"),
      sprintf("  Script : %s", script_nm)
    ),
    strrep("═", 62)
  )

  log_text <- paste(log_lines, collapse = "\n")

  if (show_log) cat(log_text, "\n")

  structure(
    list(
      log_file      = log_file,
      log_text      = log_text,
      script        = file,
      start         = start_dt,
      end           = end_dt,
      elapsed       = elapsed,
      succeeded     = succeeded,
      messages      = msgs,
      warnings      = warnings_caught,
      errors        = if (!succeeded) tail(script_result,1) else character(0),
      session_info  = si
    ),
    class = "logrx_log"
  )
}

print.logrx_log <- function(x, ...) {
  cat(x$log_text, "\n")
  invisible(x)
}

# log_read() — parse a logrx log file
log_read <- function(path) {
  lines <- readLines(path)
  structure(list(path=path, lines=lines, n_lines=length(lines)),
            class="logrx_log_file")
}

# log_select() — extract sections from a log
log_select <- function(log_obj, section = c("header","session","warnings","errors","status")) {
  section <- match.arg(section)
  if (inherits(log_obj, "logrx_log")) {
    switch(section,
      header  = list(script=log_obj$script, user=log_obj$session_info$user,
                     start=log_obj$start, end=log_obj$end,
                     elapsed=log_obj$elapsed),
      session = log_obj$session_info,
      warnings = log_obj$warnings,
      errors   = log_obj$errors,
      status   = list(succeeded=log_obj$succeeded,
                      n_warnings=length(log_obj$warnings),
                      n_messages=length(log_obj$messages))
    )
  }
}

# log_warn_select() — extract warning messages
log_warn_select <- function(log_obj) log_obj$warnings

# log_error_select() — extract error messages
log_error_select <- function(log_obj) log_obj$errors

# log_pkg_select() — extract package versions
log_pkg_select <- function(log_obj) log_obj$session_info$packages

# Simulate axecute for a script provided as a string (for learnr exercises)
axecute_string <- function(script_text, script_name = "analysis.R",
                            show_log = TRUE) {
  tmp <- tempfile(fileext = ".R")
  writeLines(script_text, tmp)
  on.exit(unlink(tmp))
  result <- axecute(tmp, log_name = sub("\\.R$", ".log", script_name),
                    show_log = show_log)
  result$script <- script_name
  result
}
```

## Overview

**logrx** solves a critical compliance challenge: in regulated clinical trial programming (GxP / 21 CFR Part 11), every analysis run must produce an **audit trail** that records who ran the code, when, on which system, with which software versions, and whether any errors or warnings occurred.

In SAS, the log is produced automatically. In R there is no equivalent until logrx.

> **What logrx does:**
>
> ```
> Analyst runs: axecute("adam/derive_adsl.R")
>                         │
>                         ▼
>       Script executes in a clean environment
>                         │
>                         ▼
>       Log written to:   adam/derive_adsl.log
>
>       Log contains:
>         ✓ Script path and MD5 hash
>         ✓ Analyst user name and node
>         ✓ Start / end / elapsed time
>         ✓ R version and all loaded packages with versions
>         ✓ All messages, warnings, and errors
>         ✓ Final SUCCESS or FAILURE status
> ```

---

## 1 — Why Logging Matters in Clinical R

### 1.1 The regulatory requirement

21 CFR Part 11 requires that computer systems used in clinical trials maintain:

1. **Audit trails** — time-stamped records of when data were created, modified, or deleted
2. **System access controls** — who ran what, when
3. **Computer environment records** — software versions used to produce results

For SAS submissions, the `.log` file satisfies these requirements automatically. For R, logrx provides the equivalent.

### 1.2 What a logrx log captures

| Log section | Contents |
|-------------|---------|
| **HEADER** | Script path, log path, user, node, start/end/elapsed time |
| **SESSION INFO** | R version, OS/platform, all loaded packages with versions |
| **MESSAGES** | All `message()` calls emitted during execution |
| **WARNINGS** | All `warning()` calls — graded from note to critical |
| **ERRORS** | Any error that halted execution |
| **STATUS** | `SUCCESS` or `FAILURE` — the most important line for QC reviewers |

### 1.3 Quiz

```{r quiz-logrx-basics, echo=FALSE}
quiz(
  question(
    "Under 21 CFR Part 11, which information is required in a clinical programming audit trail?",
    answer("Only the final output file"),
    answer("Time-stamped records of who ran what, when, and the software versions used", correct = TRUE),
    answer("Only error messages if the run failed"),
    answer("The analyst's electronic signature on every output"),
    allow_retry = TRUE
  ),
  question(
    "A QC reviewer scanning a logrx log needs to quickly determine if the run was clean. Which section do they check first?",
    answer("SESSION INFO — to verify R version"),
    answer("MESSAGES — to count all messages"),
    answer("STATUS — shows SUCCESS or FAILURE in one line", correct = TRUE),
    answer("HEADER — to see the script name"),
    allow_retry = TRUE
  )
)
```

---

## 2 — `axecute()`: The Core Function

### 2.1 Basic usage

```{r axecute-basic, eval=FALSE}
library(logrx)

# Run a script and automatically produce a log
axecute("adam/derive_adsl.R")

# Log is written to: adam/derive_adsl.log   (same folder as script, by default)
```

### 2.2 Arguments

```{r axecute-args, eval=FALSE}
axecute(
  file         = "adam/derive_adsl.R",   # script to execute
  log_name     = "derive_adsl_v2.log",   # custom log name (default: same as script)
  log_path     = "logs/",                # where to write the log
  working_dir  = "adam/",                # working directory for script execution
  quiet        = FALSE,                  # suppress console output
  show_log     = TRUE,                   # print log to console after run
  log_breaks   = TRUE                    # add visual section separators
)
```

### 2.3 Exercise — run a simple script

The `axecute_string()` helper below lets us execute inline script text within learnr. Run a minimal ADaM summary script and inspect the log.

```{r axecute-basic-ex, exercise=TRUE, exercise.lines=20}
log_result <- axecute_string('
  library(dplyr)
  library(pharmaverseadam)

  adsl <- pharmaverseadam::adsl

  # Safety population summary
  safety_summary <- adsl |>
    filter(SAFFL == "Y") |>
    group_by(TRT01P) |>
    summarise(N = n(),
              mean_age = round(mean(AGE, na.rm=TRUE), 1),
              .groups = "drop")

  cat("Safety Population Summary:\n")
  print(safety_summary)
',
  script_name = "summary_safety.R"
)

# Check the status
status <- log_select(log_result, "status")
cat("\nRun succeeded:", status$succeeded, "\n")
cat("Warnings    :", status$n_warnings, "\n")
```

### 2.4 Exercise — inspect log sections

```{r log-sections-ex, exercise=TRUE, exercise.lines=18}
log_result <- axecute_string('
  library(dplyr)
  message("Starting ADSL derivation")
  adsl <- pharmaverseadam::adsl
  message(sprintf("Loaded %d subjects", nrow(adsl)))
  warning("BMIBL not present — synthesising")
  adsl <- mutate(adsl, BMIBL = round(WEIGHTBL/(HEIGHTBL/100)^2, 1))
  cat("Done\n")
', script_name = "derive_adsl.R", show_log = FALSE)

# Extract sections
cat("── Header ──────────────────────────────────────\n")
hdr <- log_select(log_result, "header")
cat("  Script :", hdr$script,  "\n")
cat("  User   :", hdr$user,    "\n")
cat("  Elapsed:", round(hdr$elapsed, 3), "seconds\n")

cat("\n── Warnings ────────────────────────────────────\n")
warns <- log_warn_select(___)
if (length(warns) == 0) cat("  (none)\n") else
  for (w in warns) cat(" ", w, "\n")

cat("\n── Messages ────────────────────────────────────\n")
msgs <- log_select(___, "status")
cat("  N messages:", msgs$n_messages, "\n")
cat("  Succeeded :", msgs$succeeded,  "\n")
```

```{r log-sections-ex-solution}
log_result <- axecute_string('
  library(dplyr)
  message("Starting ADSL derivation")
  adsl <- pharmaverseadam::adsl
  message(sprintf("Loaded %d subjects", nrow(adsl)))
  warning("BMIBL not present — synthesising")
  adsl <- mutate(adsl, BMIBL = round(WEIGHTBL/(HEIGHTBL/100)^2, 1))
  cat("Done\n")
', script_name = "derive_adsl.R", show_log = FALSE)

cat("── Header ──────────────────────────────────────\n")
hdr <- log_select(log_result, "header")
cat("  Script :", hdr$script,  "\n")
cat("  User   :", hdr$user,    "\n")
cat("  Elapsed:", round(hdr$elapsed, 3), "seconds\n")

cat("\n── Warnings ────────────────────────────────────\n")
warns <- log_warn_select(log_result)
if (length(warns) == 0) cat("  (none)\n") else
  for (w in warns) cat(" ", w, "\n")

cat("\n── Messages ────────────────────────────────────\n")
msgs <- log_select(log_result, "status")
cat("  N messages:", msgs$n_messages, "\n")
cat("  Succeeded :", msgs$succeeded,  "\n")
```

---

## 3 — Error and Warning Handling

### 3.1 How logrx records warnings

logrx intercepts all `warning()` calls during execution and records them in the log, but allows the script to continue running. This is critical in clinical programming:

```{r warning-demo, eval=FALSE}
# Script with warnings — still succeeds overall
axecute_string('
  # This will trigger a common clinical programming warning
  adlb <- pharmaverseadam::adlb
  adlb_sub <- dplyr::filter(adlb, PARAMCD == "ALT")

  # NAs in numeric column — causes warning in some operations
  result <- mean(adlb_sub$ATOXGR)   # NA warning if all NA
  cat("Mean ATOXGR:", result, "\n")
')

# Log STATUS will be: SUCCESS (warnings don't stop execution)
# But WARNINGS section lists them for QC review
```

### 3.2 How logrx records errors

When an error occurs, the log records it in the ERRORS section and STATUS shows FAILURE:

```{r error-demo, exercise=TRUE, exercise.lines=18}
log_fail <- axecute_string('
  # This script has a deliberate error
  library(dplyr)
  adsl <- pharmaverseadam::adsl

  # Reference a non-existent column
  result <- adsl |>
    filter(NONEXISTENT_VAR == "Y") |>
    nrow()

  cat("N:", result, "\n")
', script_name = "broken_script.R", show_log = FALSE)

# Check status
status <- log_select(log_fail, "status")
cat("Succeeded :", status$succeeded,  "\n")
cat("N errors  :", length(log_error_select(log_fail)), "\n")

# Show the error message
errs <- log_error_select(log_fail)
if (length(errs) > 0) cat("Error:", errs[1], "\n")
```

### 3.3 Distinguishing warning severity

In clinical programming, not all warnings are equal. A common pattern is to fail the log if certain high-severity warnings occur:

```{r warning-severity, eval=FALSE}
# Post-execution warning triage
check_log_warnings <- function(log_result,
                                fail_patterns = c("NAs introduced",
                                                  "coercion",
                                                  "longer object length")) {
  warns <- log_warn_select(log_result)
  if (length(warns) == 0) return(invisible(TRUE))

  critical <- warns[sapply(warns, function(w)
    any(sapply(fail_patterns, function(p) grepl(p, w, ignore.case=TRUE)))
  )]

  if (length(critical) > 0) {
    cat("CRITICAL WARNINGS:\n")
    for (w in critical) cat(" •", w, "\n")
    return(invisible(FALSE))
  }
  invisible(TRUE)
}
```

### 3.4 Exercise — triage a log with warnings

```{r triage-ex, exercise=TRUE, exercise.lines=22}
log_with_warns <- axecute_string('
  library(dplyr)
  adsl <- pharmaverseadam::adsl

  # Deliberately trigger a coercion warning
  x <- as.integer(c("1","2","three","4"))
  cat("Converted:", x, "\n")

  # Also trigger an na.rm warning path
  result <- mean(c(1, 2, NA, 4))
  cat("Mean:", result, "\n")
', script_name = "triage_test.R", show_log = FALSE)

warns <- log_warn_select(log_with_warns)
cat("Total warnings:", length(warns), "\n\n")

# Classify each warning
for (w in warns) {
  severity <- if (grepl("NAs introduced|coercion", w, ignore.case=TRUE))
                "CRITICAL" else "NOTE"
  cat(sprintf("[%s] %s\n", severity, w))
}
```

---

## 4 — GxP Compliance Patterns

### 4.1 The standard logging workflow

In a GxP-validated R environment, every ADaM/TLF program follows this pattern:

```{r gxp-pattern, eval=FALSE}
# Standard GxP R programming workflow:

# 1. Analyst writes analysis script: adam/derive_adsl.R
# 2. Analyst runs: axecute("adam/derive_adsl.R")
# 3. logrx produces: adam/derive_adsl.log
# 4. QC programmer reviews: adam/derive_adsl.log
#    - Checks STATUS = SUCCESS
#    - Reviews WARNINGS section
#    - Verifies R version and package versions match expected environment
# 5. Both script and log are version-controlled together
# 6. At submission, script + log are included in the analysis data package

# File structure:
# adam/
#   derive_adsl.R       ← analysis script
#   derive_adsl.log     ← logrx audit log
#   adsl.xpt            ← output dataset
# tlf/
#   t14_1_1_demog.R
#   t14_1_1_demog.log
#   t14_1_1_demog.rtf
```

### 4.2 Script metadata block

logrx is most useful when combined with a standardised script metadata block at the top of every program:

```{r metadata-block, eval=FALSE}
# ── Standard ADaM program header ──────────────────────────────────────────
# Program  : derive_adsl.R
# Purpose  : Derive ADSL (Subject-Level Analysis Dataset) for CDISCPILOT01
# Author   : H. Tavakoli
# Date     : 2026-05-24
# Input    : sdtm/dm.xpt, sdtm/ds.xpt, sdtm/ex.xpt, sdtm/mh.xpt
# Output   : adam/adsl.xpt
# Spec     : specs/adam_spec.xlsx  (ADSL tab, version 3.1)
# Notes    : Requires pharmaverseadam >= 0.3.0, admiral >= 1.0.0
# ──────────────────────────────────────────────────────────────────────────

library(admiral)
library(pharmaversesdtm)
library(dplyr)
library(haven)
library(xportr)
library(logrx)    # loaded but axecute() is called externally

# Analysis code follows...
```

When `axecute()` runs this script, all the header metadata appears in the log output via the MESSAGES section (since `message()` calls are captured).

### 4.3 Exercise — write a compliant analysis script

Write a minimal but GxP-compliant analysis script with proper header, `message()` milestones, and `tryCatch()` for safe output writing.

```{r compliant-script-ex, exercise=TRUE, exercise.lines=35}
log_result <- axecute_string('
# ── Program Header ──────────────────────────────────────────────────────
# Program : safety_summary.R
# Purpose : Produce safety population summary for CDISCPILOT01
# Author  : Clinical Programmer
# Input   : pharmaverseadam::adsl
# Output  : (console for this demo)
# ────────────────────────────────────────────────────────────────────────

library(dplyr)
library(pharmaverseadam)

message("=== START: safety_summary.R ===")

# Load data
message("Step 1: Loading ADSL")
adsl <- pharmaverseadam::adsl
message(sprintf("  ADSL loaded: %d subjects", nrow(adsl)))

# Safety population
message("Step 2: Filtering to safety population")
adsl_saf <- filter(adsl, SAFFL == "Y")
message(sprintf("  Safety population: %d subjects", nrow(adsl_saf)))

# Summary
message("Step 3: Computing summary")
summary_tbl <- adsl_saf |>
  group_by(TRT01P) |>
  summarise(N       = n(),
            Mean_Age = round(mean(AGE, na.rm=TRUE), 1),
            Pct_F   = round(mean(SEX=="F")*100, 1),
            .groups = "drop")

print(summary_tbl)

message("=== END: safety_summary.R — SUCCESS ===")
', script_name = "safety_summary.R", show_log = FALSE)

# Review the log
status <- log_select(log_result, "status")
cat("Status     :", if(status$succeeded) "SUCCESS" else "FAILURE", "\n")
cat("Messages   :", status$n_messages, "\n")
cat("Warnings   :", status$n_warnings, "\n")
cat("Errors     :", length(log_error_select(log_result)), "\n")
```

---

## 5 — Log File Management

### 5.1 Log naming conventions

```{r log-naming, eval=FALSE}
# Convention 1: Same name as script (default — most common)
axecute("adam/derive_adsl.R")
# → adam/derive_adsl.log

# Convention 2: Custom name with version
axecute("adam/derive_adsl.R",
        log_name = "derive_adsl_v02.log",
        log_path = "logs/adam/")

# Convention 3: Timestamp in name (useful for repeated runs)
axecute("adam/derive_adsl.R",
        log_name = paste0("derive_adsl_",
                          format(Sys.time(), "%Y%m%d_%H%M%S"),
                          ".log"),
        log_path = "logs/")
```

### 5.2 Batch execution with logging

```{r batch-logging, eval=FALSE}
# Run all ADaM scripts in sequence, collect logs
library(purrr)

adam_scripts <- c(
  "adam/derive_adsl.R",
  "adam/derive_adae.R",
  "adam/derive_adlb.R",
  "adam/derive_advs.R"
)

batch_results <- map(adam_scripts, function(script) {
  message(sprintf("Running: %s", script))
  axecute(
    file      = script,
    log_path  = "logs/adam/",
    show_log  = FALSE,
    quiet     = TRUE
  )
})

# Batch summary
batch_summary <- tibble(
  script    = map_chr(batch_results, ~ basename(.x$script)),
  succeeded = map_lgl(batch_results, ~ .x$succeeded),
  warnings  = map_int(batch_results, ~ length(.x$warnings)),
  elapsed_s = map_dbl(batch_results, ~ round(.x$elapsed, 2))
)

print(batch_summary)

# Fail loudly if any script failed
failures <- filter(batch_summary, !succeeded)
if (nrow(failures) > 0)
  stop(sprintf("%d script(s) FAILED: %s",
               nrow(failures),
               paste(failures$script, collapse=", ")))
```

### 5.3 Exercise — batch summary

Simulate running three scripts and produce a batch summary table.

```{r batch-ex, exercise=TRUE, exercise.lines=35}
scripts <- list(
  list(name="derive_adsl.R", code='
    library(dplyr); library(pharmaverseadam)
    message("Deriving ADSL")
    adsl <- pharmaverseadam::adsl
    cat("ADSL:", nrow(adsl), "rows\n")
  '),
  list(name="derive_adae.R", code='
    library(dplyr); library(pharmaverseadam)
    message("Deriving ADAE")
    adae <- pharmaverseadam::adae
    warning("TRTA renamed to TRT01A")
    adae <- rename(adae, TRT01A = TRTA)
    cat("ADAE:", nrow(adae), "rows\n")
  '),
  list(name="derive_adlb.R", code='
    library(dplyr); library(pharmaverseadam)
    message("Deriving ADLB")
    adlb <- pharmaverseadam::adlb
    cat("ADLB:", nrow(adlb), "rows\n")
  ')
)

batch_logs <- lapply(scripts, function(s)
  axecute_string(s$code, script_name=s$name, show_log=FALSE))

# Build summary
batch_tbl <- tibble(
  script    = sapply(scripts, `[[`, "name"),
  succeeded = sapply(batch_logs, function(l) l$___),
  n_warnings = sapply(batch_logs, function(l) length(log_warn_select(l))),
  elapsed_s  = round(sapply(batch_logs, function(l) l$elapsed), 3)
)

cat("── Batch Execution Summary ─────────────────────────\n")
print(batch_tbl)

# Alert on failures
failures <- filter(batch_tbl, !succeeded)
if (nrow(failures) == 0) {
  cat("\n✓ All scripts completed successfully\n")
} else {
  cat(sprintf("\n✗ %d failure(s): %s\n",
              nrow(failures), paste(failures$script, collapse=", ")))
}
```

```{r batch-ex-solution}
scripts <- list(
  list(name="derive_adsl.R", code='
    library(dplyr); library(pharmaverseadam)
    message("Deriving ADSL")
    adsl <- pharmaverseadam::adsl
    cat("ADSL:", nrow(adsl), "rows\n")
  '),
  list(name="derive_adae.R", code='
    library(dplyr); library(pharmaverseadam)
    message("Deriving ADAE")
    adae <- pharmaverseadam::adae
    warning("TRTA renamed to TRT01A")
    adae <- rename(adae, TRT01A = TRTA)
    cat("ADAE:", nrow(adae), "rows\n")
  '),
  list(name="derive_adlb.R", code='
    library(dplyr); library(pharmaverseadam)
    message("Deriving ADLB")
    adlb <- pharmaverseadam::adlb
    cat("ADLB:", nrow(adlb), "rows\n")
  ')
)

batch_logs <- lapply(scripts, function(s)
  axecute_string(s$code, script_name=s$name, show_log=FALSE))

batch_tbl <- tibble(
  script     = sapply(scripts, `[[`, "name"),
  succeeded  = sapply(batch_logs, function(l) l$succeeded),
  n_warnings = sapply(batch_logs, function(l) length(log_warn_select(l))),
  elapsed_s  = round(sapply(batch_logs, function(l) l$elapsed), 3)
)

cat("── Batch Execution Summary ─────────────────────────\n")
print(batch_tbl)

failures <- filter(batch_tbl, !succeeded)
if (nrow(failures) == 0) {
  cat("\n✓ All scripts completed successfully\n")
} else {
  cat(sprintf("\n✗ %d failure(s): %s\n",
              nrow(failures), paste(failures$script, collapse=", ")))
}
```

---

## 6 — Integration with CI/CD

### 6.1 GitHub Actions workflow

logrx integrates cleanly with GitHub Actions for automated validation:

```{r gh-actions-demo, eval=FALSE}
# .github/workflows/validate_adam.yml
#
# name: Validate ADaM Programs
#
# on:
#   push:
#     paths: ['adam/**', 'specs/**']
#
# jobs:
#   run_adam:
#     runs-on: ubuntu-latest
#     steps:
#       - uses: actions/checkout@v4
#       - uses: r-lib/actions/setup-r@v2
#         with:
#           r-version: '4.4.0'
#       - uses: r-lib/actions/setup-renv@v2
#       - name: Run ADaM programs
#         run: |
#           Rscript -e "
#             library(logrx)
#             scripts <- list.files('adam', pattern='\\\\.R$', full.names=TRUE)
#             results <- lapply(scripts, function(s)
#               axecute(s, log_path='logs/', show_log=FALSE, quiet=TRUE))
#             failures <- Filter(function(r) !r$succeeded, results)
#             if (length(failures) > 0) {
#               cat('FAILED:', sapply(failures, function(r) basename(r$script)), '\n')
#               quit(status=1)
#             }
#             cat('All ADaM programs: SUCCESS\n')
#           "
#       - name: Upload logs
#         uses: actions/upload-artifact@v4
#         if: always()
#         with:
#           name: adam-logs
#           path: logs/
```

### 6.2 Log review in QC workflows

```{r qc-review, eval=FALSE}
# QC reviewer checks a log programmatically
review_log <- function(log_path, max_warnings = 0) {
  lines   <- readLines(log_path)

  # Extract status
  status_line <- grep("Status :", lines, value=TRUE)
  succeeded   <- any(grepl("SUCCESS", status_line))

  # Count warnings
  warn_lines  <- grep("WARNING:", lines, value=TRUE)
  n_warnings  <- length(warn_lines)

  cat(sprintf("Log: %s\n", basename(log_path)))
  cat(sprintf("  Status   : %s\n", if(succeeded) "✓ SUCCESS" else "✗ FAILURE"))
  cat(sprintf("  Warnings : %d (threshold: %d)\n", n_warnings, max_warnings))

  if (!succeeded) stop(sprintf("FAILURE in log: %s", log_path))
  if (n_warnings > max_warnings)
    stop(sprintf("%d warnings exceed threshold of %d", n_warnings, max_warnings))

  invisible(TRUE)
}
```

### 6.3 Quiz

```{r quiz-gxp, echo=FALSE}
quiz(
  question(
    "In a GxP-validated R environment, both the analysis script and its logrx log should be:",
    answer("Stored only on the analyst's local computer"),
    answer("Version-controlled together (e.g., in git) and archived with the submission", correct = TRUE),
    answer("Deleted after QC review to save storage"),
    answer("Emailed to the statistician for review"),
    allow_retry = TRUE
  ),
  question(
    "A logrx log shows STATUS = SUCCESS but 3 warnings. How should a QC reviewer handle this?",
    answer("Ignore — SUCCESS means the run is clean"),
    answer("Fail the review — any warning is unacceptable"),
    answer("Review each warning individually; document disposition; escalate coercion/NAs warnings", correct = TRUE),
    answer("Re-run the script until there are no warnings"),
    allow_retry = TRUE
  )
)
```

---

## 7 — Customising Log Output

### 7.1 Adding custom sections

Real-world pipelines often need custom log sections beyond the standard ones. Combine logrx with `message()` calls to produce structured custom sections:

```{r custom-sections, eval=FALSE}
# Pattern: use message() with a separator prefix
# logrx captures all messages → custom sections appear in MESSAGES section

axecute("adam/derive_adsl.R")

# Inside derive_adsl.R:
message(strrep("─", 50))
message("  DATA INVENTORY")
message(strrep("─", 50))
message(sprintf("  DM  rows: %d", nrow(dm)))
message(sprintf("  DS  rows: %d", nrow(ds)))
message(sprintf("  EX  rows: %d", nrow(ex)))

message(strrep("─", 50))
message("  DERIVATION CHECKS")
message(strrep("─", 50))
message(sprintf("  ADSL rows:     %d", nrow(adsl)))
message(sprintf("  Subjects:      %d", n_distinct(adsl$USUBJID)))
message(sprintf("  Safety pop:    %d", sum(adsl$SAFFL=="Y", na.rm=TRUE)))
message(sprintf("  ITT pop:       %d", sum(adsl$ITTFL=="Y", na.rm=TRUE)))
```

### 7.2 Exercise — structured log with custom sections

Write a script that produces a well-structured log with data inventory and derivation check sections.

```{r custom-log-ex, exercise=TRUE, exercise.lines=35}
log_custom <- axecute_string('
library(dplyr)
library(pharmaverseadam)

message(strrep("─", 50))
message("  DATA INVENTORY")
message(strrep("─", 50))

adsl <- pharmaverseadam::adsl
adae <- pharmaverseadam::adae

message(sprintf("  ADSL raw rows : %d", nrow(adsl)))
message(sprintf("  ADAE raw rows : %d", nrow(adae)))

message(strrep("─", 50))
message("  DERIVATION")
message(strrep("─", 50))

adsl <- adsl |>
  mutate(ITTFL  = SAFFL,
         DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"))

message(strrep("─", 50))
message("  OUTPUT CHECKS")
message(strrep("─", 50))

n_saf <- sum(adsl$SAFFL  == "Y", na.rm=TRUE)
n_itt <- sum(adsl$ITTFL  == "Y", na.rm=TRUE)
n_dia <- sum(adsl$DIABFL == "Y", na.rm=TRUE)

message(sprintf("  Safety pop : %d", n_saf))
message(sprintf("  ITT pop    : %d", n_itt))
message(sprintf("  Diabetic   : %d", n_dia))

cat("ADSL derivation complete\n")
', script_name = "derive_adsl.R", show_log = FALSE)

# Show only messages from the log
msgs <- log_custom$messages
cat("── Messages captured in log ────────────────────────\n")
for (m in msgs) cat(m)
```

---

## 8 — Capstone: Full Program Logging Pipeline

Build a complete logging pipeline that runs three scripts, collects batch logs, and produces a QC-ready summary report.

```{r capstone-logging, exercise=TRUE, exercise.lines=65}
# ── Scripts to run ─────────────────────────────────────────────────────────
program_suite <- list(
  list(
    name   = "derive_adsl.R",
    domain = "ADSL",
    code   = '
      library(dplyr); library(pharmaverseadam)
      message("=== START: derive_adsl.R ===")
      adsl <- pharmaverseadam::adsl |>
        mutate(ITTFL=SAFFL,
               DIABFL=if_else(!is.na(BMIBL)&BMIBL>=30,"Y","N",missing="N"))
      message(sprintf("  Subjects: %d", nrow(adsl)))
      message(sprintf("  Safety N: %d", sum(adsl$SAFFL=="Y",na.rm=TRUE)))
      message("=== END: derive_adsl.R ===")
    '
  ),
  list(
    name   = "derive_adae.R",
    domain = "ADAE",
    code   = '
      library(dplyr); library(pharmaverseadam)
      message("=== START: derive_adae.R ===")
      adae <- pharmaverseadam::adae
      if("TRTA" %in% names(adae) && !"TRT01A" %in% names(adae))
        adae <- rename(adae, TRT01A=TRTA)
      message(sprintf("  AE records: %d", nrow(adae)))
      message(sprintf("  Subjects  : %d", dplyr::n_distinct(adae$USUBJID)))
      message("=== END: derive_adae.R ===")
    '
  ),
  list(
    name   = "derive_adlb.R",
    domain = "ADLB",
    code   = '
      library(dplyr); library(pharmaverseadam)
      message("=== START: derive_adlb.R ===")
      adlb <- pharmaverseadam::adlb |>
        filter(is.na(DTYPE)) |>
        mutate(ATOXGR=as.integer(ATOXGR))
      message(sprintf("  Lab records: %d", nrow(adlb)))
      message(sprintf("  Parameters : %d", dplyr::n_distinct(adlb$PARAMCD)))
      message("=== END: derive_adlb.R ===")
    '
  )
)

# ── Execute all programs ──────────────────────────────────────────────────
cat("╔══════════════════════════════════════════════════════╗\n")
cat("║  CDISCPILOT01 — ADaM Batch Execution                ║\n")
cat("╚══════════════════════════════════════════════════════╝\n\n")

all_logs <- lapply(program_suite, function(p) {
  cat(sprintf("Running %s...", p$name))
  log_res <- axecute_string(p$code, script_name=p$name, show_log=FALSE)
  status  <- if (log_res$succeeded) "✓ SUCCESS" else "✗ FAILURE"
  cat(sprintf(" %s (%.3fs)\n", status, log_res$elapsed))
  log_res
})

# ── Batch QC report ──────────────────────────────────────────────────────
cat("\n╔══════════════════════════════════════════════════════╗\n")
cat("║  QC Batch Summary                                    ║\n")
cat("╚══════════════════════════════════════════════════════╝\n")

qc_tbl <- tibble(
  Program  = sapply(program_suite, `[[`, "name"),
  Domain   = sapply(program_suite, `[[`, "domain"),
  Status   = sapply(all_logs, function(l) if(l$succeeded) "SUCCESS" else "FAILURE"),
  Warnings = sapply(all_logs, function(l) length(log_warn_select(l))),
  Messages = sapply(all_logs, function(l) length(l$messages)),
  Elapsed  = round(sapply(all_logs, function(l) l$elapsed), 3)
)

print(qc_tbl)

n_fail <- sum(qc_tbl$Status == "FAILURE")
n_warn <- sum(qc_tbl$Warnings)
cat(sprintf("\n  Programs run   : %d\n", nrow(qc_tbl)))
cat(sprintf("  Failures       : %d\n", n_fail))
cat(sprintf("  Total warnings : %d\n", n_warn))
cat(sprintf("  Overall status : %s\n",
            if(n_fail==0) "✓ ALL PASS" else "✗ REVIEW REQUIRED"))
```

---

## Summary

logrx provides the audit trail that transforms R from a research tool into a GxP-validated clinical programming platform:

- **`axecute(file)`** runs a script and automatically produces a structured log
- The log captures: header (user, node, times), session info (R version, all packages), messages, warnings, errors, and final STATUS
- `log_select(log, section)` extracts header, session, warnings, errors, or status programmatically
- `log_warn_select()` / `log_error_select()` pull specific issue lists for triage
- **STATUS = FAILURE** when any unhandled error occurs; warnings leave STATUS = SUCCESS but are recorded
- Batch execution with `lapply()` or `purrr::map()` enables whole-suite logging in one call
- GxP compliance requires: every script run via `axecute()`, log version-controlled alongside script, QC review documented

Next: **Lesson 46** covers `diffdf` and `riskmetric` — comparing datasets for differences and scoring R packages for use in regulated environments.
