---
title: "Lesson 46 — diffdf & riskmetric: Dataset QC and Package Risk"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: flatly
runtime: shiny_prerendered
description: >
  Master diffdf for dataset comparison and QC, and riskmetric for scoring
  R packages for use in regulated clinical environments. Covers row/column
  diffs, tolerance-based numeric comparison, QC workflows, package risk
  scores, cohort assessments, and validation documentation.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 46_diffdf_riskmetric.Rmd
# PURPOSE       : diffdf + riskmetric — double-programming QC with diffdf() and
#                 package risk scoring with riskmetric: pkg_ref() →
#                 assess() → score() → risk tier classification.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================
library(learnr)
library(dplyr)
library(tibble)
library(pharmaverseadam)

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
tutorial_options(exercise.timelimit = 90, exercise.eval = FALSE)

`%||%` <- function(a, b) if (is.null(a)) b else a

# ── Reference and modified ADSL for diffing ─────────────────────────────────
adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl_raw$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl_raw <- adsl_raw |> mutate(HEIGHTBL=ht, WEIGHTBL=wt,
                                  BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}

adsl <- adsl_raw |>
  mutate(ITTFL  = SAFFL,
         DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing="N")) |>
  select(STUDYID, USUBJID, SUBJID, SITEID,
         TRT01P, TRT01A, ARM, ARMCD,
         AGE, AGEGR1, SEX, RACE, ETHNIC,
         SAFFL, ITTFL, DIABFL,
         HEIGHTBL, WEIGHTBL, BMIBL,
         TRTSDT, TRTEDT, TRTDURD,
         COUNTRY, DTHFL)

# Make a "QC programmer's version" with deliberate differences for exercises
set.seed(42)
adsl_qc <- adsl |>
  mutate(
    # Difference 1: rounding differences in BMIBL for 5 subjects
    BMIBL   = if_else(row_number() %in% c(10,20,30,40,50),
                      round(BMIBL + 0.05, 1), BMIBL),
    # Difference 2: DIABFL recoded for 2 subjects
    DIABFL  = if_else(row_number() %in% c(15,25), "Y", DIABFL),
    # Difference 3: TRTDURD off by 1 for 3 subjects
    TRTDURD = if_else(row_number() %in% c(100,101,102),
                      TRTDURD + 1L, TRTDURD)
  )

# ADAE for diffing
adae <- pharmaverseadam::adae
if ("TRTA" %in% names(adae) && !"TRT01A" %in% names(adae))
  adae <- rename(adae, TRT01A = TRTA)
adae <- select(adae, STUDYID, USUBJID, AESEQ, AEDECOD, AESOC,
               AESEV, AESER, AEREL, TRT01A, ASTDT, AENDT)

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR=as.integer(ATOXGR), BTOXGR=as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))

# ── Mock diffdf infrastructure ────────────────────────────────────────────────
# diffdf compares two data frames and returns structured difference information.

diffdf <- function(base, compare,
                   keys         = NULL,
                   suppress_warnings = FALSE,
                   tolerance    = NULL,
                   check_column_order = FALSE) {

  result <- list(
    base_name    = deparse(substitute(base)),
    compare_name = deparse(substitute(compare)),
    keys         = keys,
    tolerance    = tolerance
  )

  # ── Check 1: column differences ─────────────────────────────────────────
  cols_base    <- names(base)
  cols_compare <- names(compare)
  result$cols_in_base_not_compare <- setdiff(cols_base, cols_compare)
  result$cols_in_compare_not_base <- setdiff(cols_compare, cols_base)

  # ── Check 2: row count ───────────────────────────────────────────────────
  result$nrow_base    <- nrow(base)
  result$nrow_compare <- nrow(compare)

  # ── Check 3: type differences ────────────────────────────────────────────
  common_cols <- intersect(cols_base, cols_compare)
  type_diffs <- tibble(
    variable   = common_cols,
    class_base = sapply(base[common_cols],    function(v) class(v)[1]),
    class_comp = sapply(compare[common_cols], function(v) class(v)[1])
  ) |>
    filter(class_base != class_comp)
  result$type_diffs <- type_diffs

  # ── Check 4: value differences ───────────────────────────────────────────
  value_diffs <- list()

  # Key-based matching
  if (!is.null(keys) && all(keys %in% names(base)) && all(keys %in% names(compare))) {
    base_keyed    <- base[, common_cols, drop=FALSE]
    compare_keyed <- compare[, common_cols, drop=FALSE]

    # Join on keys
    joined <- inner_join(
      mutate(base_keyed,    .row_base = row_number()),
      mutate(compare_keyed, .row_comp = row_number()),
      by = keys,
      suffix = c("_base", "_comp")
    )

    non_key_cols <- setdiff(common_cols, keys)

    for (col in non_key_cols) {
      base_col <- paste0(col, "_base")
      comp_col <- paste0(col, "_comp")
      if (!base_col %in% names(joined) || !comp_col %in% names(joined)) next

      b_vals <- joined[[base_col]]
      c_vals <- joined[[comp_col]]

      # Determine differences
      if (is.numeric(b_vals) && is.numeric(c_vals)) {
        tol <- tolerance %||% 0
        diffs_idx <- which(!is.na(b_vals) & !is.na(c_vals) & abs(b_vals - c_vals) > tol)
        both_na   <- which(is.na(b_vals) != is.na(c_vals))
        diff_rows <- unique(c(diffs_idx, both_na))
      } else {
        diff_rows <- which(!mapply(identical, b_vals, c_vals))
      }

      if (length(diff_rows) > 0) {
        diff_df <- joined[diff_rows, c(keys, base_col, comp_col), drop=FALSE]
        names(diff_df)[names(diff_df)==base_col] <- "base_value"
        names(diff_df)[names(diff_df)==comp_col] <- "compare_value"
        diff_df$variable <- col
        value_diffs[[col]] <- diff_df
      }
    }
  } else {
    # Positional matching (no keys)
    min_rows <- min(nrow(base), nrow(compare))
    for (col in common_cols) {
      b_vals <- base[[col]][seq_len(min_rows)]
      c_vals <- compare[[col]][seq_len(min_rows)]
      if (is.numeric(b_vals) && is.numeric(c_vals)) {
        tol <- tolerance %||% 0
        diff_rows <- which(!is.na(b_vals) & !is.na(c_vals) & abs(b_vals-c_vals)>tol)
      } else {
        diff_rows <- which(!mapply(identical, b_vals, c_vals))
      }
      if (length(diff_rows) > 0) {
        value_diffs[[col]] <- tibble(
          row          = diff_rows,
          variable     = col,
          base_value   = as.character(b_vals[diff_rows]),
          compare_value = as.character(c_vals[diff_rows])
        )
      }
    }
  }

  result$value_diffs  <- value_diffs
  result$n_var_diffs  <- length(value_diffs)
  result$is_identical <- (
    length(result$cols_in_base_not_compare) == 0 &&
    length(result$cols_in_compare_not_base) == 0 &&
    result$nrow_base == result$nrow_compare &&
    nrow(result$type_diffs) == 0 &&
    length(value_diffs) == 0
  )

  class(result) <- "diffdf_result"
  result
}

print.diffdf_result <- function(x, ...) {
  cat("── diffdf Result ────────────────────────────────────\n")
  if (x$is_identical) {
    cat("  ✓ Datasets are IDENTICAL\n")
    return(invisible(x))
  }

  if (length(x$cols_in_base_not_compare) > 0)
    cat(sprintf("  Columns in base not in compare: %s\n",
                paste(x$cols_in_base_not_compare, collapse=", ")))
  if (length(x$cols_in_compare_not_base) > 0)
    cat(sprintf("  Columns in compare not in base: %s\n",
                paste(x$cols_in_compare_not_base, collapse=", ")))
  if (x$nrow_base != x$nrow_compare)
    cat(sprintf("  Row count: base=%d, compare=%d\n", x$nrow_base, x$nrow_compare))
  if (nrow(x$type_diffs) > 0) {
    cat("  Type differences:\n")
    for (i in seq_len(nrow(x$type_diffs)))
      cat(sprintf("    %s: base=%s, compare=%s\n",
                  x$type_diffs$variable[i],
                  x$type_diffs$class_base[i],
                  x$type_diffs$class_comp[i]))
  }
  if (x$n_var_diffs > 0) {
    cat(sprintf("  Variable differences: %d variable(s) have differing values\n",
                x$n_var_diffs))
    for (nm in names(x$value_diffs)) {
      nd <- nrow(x$value_diffs[[nm]])
      cat(sprintf("    %-20s %d row(s) differ\n", nm, nd))
    }
  }
  cat("────────────────────────────────────────────────────\n")
  invisible(x)
}

# Accessor helpers
diffdf_has_issues  <- function(x) !x$is_identical
diffdf_n_diffs     <- function(x) x$n_var_diffs
diffdf_issue_cols  <- function(x) names(x$value_diffs)
diffdf_var_diffs   <- function(x, variable) {
  if (variable %in% names(x$value_diffs)) x$value_diffs[[variable]]
  else tibble()
}
diffdf_is_identical <- function(x) x$is_identical

# ── Mock riskmetric infrastructure ────────────────────────────────────────────
# riskmetric scores R packages for use in regulated environments.

pkg_ref <- function(pkg_name, source = "cran", ...) {
  structure(
    list(name = pkg_name, source = source, extra = list(...)),
    class = "pkg_ref"
  )
}

# Metrics calculators (mocked)
.score_has_news     <- function(pkg) if (pkg %in% c("admiral","dplyr","ggplot2","tidyr","xportr","logrx","diffdf","pharmaverseadam")) 1 else 0.5
.score_has_vignette <- function(pkg) if (pkg %in% c("admiral","dplyr","ggplot2","tidyr","xportr","logrx")) 1 else 0.6
.score_has_tests    <- function(pkg) if (pkg %in% c("admiral","dplyr","ggplot2","xportr","logrx","diffdf")) 1 else 0.7
.score_covr         <- function(pkg) if (pkg %in% c("admiral","dplyr","ggplot2")) 0.95 else if (pkg %in% c("xportr","logrx","diffdf")) 0.85 else 0.70
.score_downloads    <- function(pkg) {
  d <- c(admiral=25000, dplyr=5000000, ggplot2=6000000, tidyr=4000000,
         xportr=8000, logrx=5000, diffdf=12000, pharmaverseadam=3500,
         metacore=6000, metatools=5500, teal=9000)
  d[[pkg]] %||% 1000
}
.score_cran_checks  <- function(pkg) if (pkg %in% c("dplyr","ggplot2","tidyr")) 1 else 0.9

# assess() — compute all metrics
assess <- function(pkg_ref_list) {
  if (inherits(pkg_ref_list, "pkg_ref")) pkg_ref_list <- list(pkg_ref_list)
  lapply(pkg_ref_list, function(p) {
    nm <- p$name
    list(
      pkg_name       = nm,
      has_news       = .score_has_news(nm),
      has_vignette   = .score_has_vignette(nm),
      has_tests      = .score_has_tests(nm),
      covr_coverage  = .score_covr(nm),
      downloads_1yr  = .score_downloads(nm),
      cran_checks    = .score_cran_checks(nm)
    )
  })
}

# score() — aggregate into a 0–1 risk score
score <- function(assessed_list) {
  lapply(assessed_list, function(a) {
    metric_scores <- c(
      a$has_news,
      a$has_vignette,
      a$has_tests,
      a$covr_coverage,
      min(1, log10(a$downloads_1yr + 1) / log10(1e7)),
      a$cran_checks
    )
    risk_score <- mean(metric_scores)
    a$risk_score <- round(risk_score, 3)
    a$risk_label <- if (risk_score >= 0.75) "Low Risk"
                    else if (risk_score >= 0.50) "Medium Risk"
                    else "High Risk"
    class(a) <- "pkg_score"
    a
  })
}

print.pkg_score <- function(x, ...) {
  cat(sprintf("Package : %s\n", x$pkg_name))
  cat(sprintf("  Risk score  : %.3f  [%s]\n", x$risk_score, x$risk_label))
  cat(sprintf("  Has NEWS    : %.2f\n", x$has_news))
  cat(sprintf("  Vignettes   : %.2f\n", x$has_vignette))
  cat(sprintf("  Unit tests  : %.2f\n", x$has_tests))
  cat(sprintf("  Test cov    : %.2f\n", x$covr_coverage))
  cat(sprintf("  Downloads   : %d /yr\n", x$downloads_1yr))
  cat(sprintf("  CRAN checks : %.2f\n", x$cran_checks))
  invisible(x)
}

# pkg_score_tbl() — combine scores into a tibble
pkg_score_tbl <- function(scored_list) {
  bind_rows(lapply(scored_list, function(s) {
    tibble(
      Package        = s$pkg_name,
      Risk_Score     = s$risk_score,
      Risk_Label     = s$risk_label,
      Has_News       = s$has_news,
      Has_Tests      = s$has_tests,
      Cov_Pct        = round(s$covr_coverage * 100, 0),
      Downloads_1yr  = s$downloads_1yr
    )
  }))
}

# cohort workflow helpers
pkg_cohort <- function(...) {
  pkgs <- list(...)
  lapply(pkgs, function(p) {
    if (is.character(p)) pkg_ref(p) else p
  })
}
```

## Overview

This lesson covers two packages used in QC and governance for clinical R programming:

**diffdf** — compare two data frames to find rows, columns, or values that differ. Essential for double-programming QC, where a second programmer independently derives the same dataset and the two versions must be compared.

**riskmetric** — score R packages across dimensions like documentation, test coverage, download volume, and CRAN checks to produce a risk-based assessment for regulated use. Required by many organisations' R validation frameworks (e.g., the R Validation Hub's white paper).

---

## 1 — diffdf: Dataset Comparison

### 1.1 Why double-programming uses diffdf

In clinical trials, critical ADaM datasets (especially ADSL, ADAE, primary endpoint datasets) are derived **independently by two programmers** and the results compared. If the datasets are identical, the derivation is validated. Any differences must be investigated and resolved.

```
Programmer 1: derive_adsl.R  →  adsl_p1.xpt
Programmer 2: derive_adsl_qc.R → adsl_p2.xpt
                                        │
                                        ▼
                        diffdf(adsl_p1, adsl_p2,
                               keys = c("STUDYID","USUBJID"))
                                        │
                                ┌───────┴──────────┐
                           ✓ Identical         ✗ Differences found
                         → QC passed            → Investigate & resolve
```

### 1.2 Basic `diffdf()` usage

```{r diffdf-basic, eval=FALSE}
library(diffdf)

result <- diffdf(
  base    = adsl_p1,     # reference (Programmer 1)
  compare = adsl_p2,     # new version (Programmer 2 / QC)
  keys    = c("STUDYID", "USUBJID")
)

# Check if identical
diffdf_is_identical(result)   # TRUE → pass; FALSE → investigate
print(result)
```

### 1.3 Exercise — basic comparison

Compare `adsl` (reference) with `adsl_qc` (has deliberate differences). The QC dataset has BMIBL rounding differences, DIABFL changes, and TRTDURD off-by-one errors.

```{r diffdf-basic-ex, exercise=TRUE, exercise.lines=14}
result <- diffdf(
  base    = ___,
  compare = ___,
  keys    = c("STUDYID", "USUBJID")
)

print(result)

cat("\nIdentical?    ", diffdf_is_identical(result), "\n")
cat("N vars differ:", diffdf_n_diffs(result), "\n")
cat("Diff columns: ", paste(diffdf_issue_cols(result), collapse=", "), "\n")
```

```{r diffdf-basic-ex-solution}
result <- diffdf(
  base    = adsl,
  compare = adsl_qc,
  keys    = c("STUDYID", "USUBJID")
)

print(result)

cat("\nIdentical?    ", diffdf_is_identical(result), "\n")
cat("N vars differ:", diffdf_n_diffs(result), "\n")
cat("Diff columns: ", paste(diffdf_issue_cols(result), collapse=", "), "\n")
```

---

## 2 — Exploring Differences

### 2.1 Drilling into a specific variable

```{r drill-diff-ex, exercise=TRUE, exercise.lines=16}
result <- diffdf(adsl, adsl_qc, keys=c("STUDYID","USUBJID"))

# Inspect BMIBL differences
bmibl_diffs <- diffdf_var_diffs(result, "BMIBL")

cat("BMIBL differences:\n")
print(bmibl_diffs)

cat("\nDIABFL differences:\n")
diabfl_diffs <- diffdf_var_diffs(result, ___)
print(diabfl_diffs)
```

```{r drill-diff-ex-solution}
result <- diffdf(adsl, adsl_qc, keys=c("STUDYID","USUBJID"))

cat("BMIBL differences:\n")
bmibl_diffs <- diffdf_var_diffs(result, "BMIBL")
print(bmibl_diffs)

cat("\nDIABFL differences:\n")
diabfl_diffs <- diffdf_var_diffs(result, "DIABFL")
print(diabfl_diffs)
```

### 2.2 Numeric tolerance

When comparing numeric variables, rounding differences between SAS and R outputs are expected. Use `tolerance` to allow small differences:

```{r tolerance-ex, exercise=TRUE, exercise.lines=16}
# Without tolerance: BMIBL differences flagged
result_strict <- diffdf(adsl, adsl_qc,
                         keys      = c("STUDYID","USUBJID"),
                         tolerance = 0)
cat("Strict (tol=0) — vars with diffs:", diffdf_n_diffs(result_strict), "\n")

# With tolerance of 0.1: minor rounding forgiven
result_lenient <- diffdf(adsl, adsl_qc,
                          keys      = c("STUDYID","USUBJID"),
                          tolerance = ___)
cat("Lenient (tol=0.1) — vars with diffs:", diffdf_n_diffs(result_lenient), "\n")
cat("Remaining diff columns:", paste(diffdf_issue_cols(result_lenient), collapse=", "), "\n")
```

```{r tolerance-ex-solution}
result_strict <- diffdf(adsl, adsl_qc,
                         keys      = c("STUDYID","USUBJID"),
                         tolerance = 0)
cat("Strict (tol=0) — vars with diffs:", diffdf_n_diffs(result_strict), "\n")

result_lenient <- diffdf(adsl, adsl_qc,
                          keys      = c("STUDYID","USUBJID"),
                          tolerance = 0.1)
cat("Lenient (tol=0.1) — vars with diffs:", diffdf_n_diffs(result_lenient), "\n")
cat("Remaining diff columns:", paste(diffdf_issue_cols(result_lenient), collapse=", "), "\n")
```

### 2.3 Column differences

```{r col-diff-ex, exercise=TRUE, exercise.lines=16}
# Create a version with an extra column and a missing column
adsl_modified <- adsl |>
  select(-DIABFL) |>              # remove one column
  mutate(NEW_SCORE = AGE * 2)     # add a new column

result_cols <- diffdf(
  base    = ___,
  compare = ___,
  keys    = c("STUDYID","USUBJID")
)

cat("Cols in base not compare:", result_cols$cols_in_base_not_compare, "\n")
cat("Cols in compare not base:", result_cols$cols_in_compare_not_base, "\n")
```

```{r col-diff-ex-solution}
adsl_modified <- adsl |>
  select(-DIABFL) |>
  mutate(NEW_SCORE = AGE * 2)

result_cols <- diffdf(
  base    = adsl,
  compare = adsl_modified,
  keys    = c("STUDYID","USUBJID")
)

cat("Cols in base not compare:", result_cols$cols_in_base_not_compare, "\n")
cat("Cols in compare not base:", result_cols$cols_in_compare_not_base, "\n")
```

### 2.4 Quiz

```{r quiz-diffdf, echo=FALSE}
quiz(
  question(
    "In a double-programming QC workflow, what does it mean when `diffdf_is_identical()` returns TRUE?",
    answer("The datasets have the same column names only"),
    answer("The two independently-derived datasets are identical — QC passed", correct = TRUE),
    answer("The row counts are equal but values may differ"),
    answer("The column types match but values are not checked"),
    allow_retry = TRUE
  ),
  question(
    "When would you set `tolerance = 0.01` in a diffdf() call?",
    answer("To ignore all numeric differences regardless of size"),
    answer("To allow small floating-point differences due to SAS vs R rounding", correct = TRUE),
    answer("To skip character variable comparisons"),
    answer("To compare only the first 0.01% of rows"),
    allow_retry = TRUE
  )
)
```

---

## 3 — diffdf in a QC Pipeline

### 3.1 Automated QC comparison

```{r qc-pipeline, eval=FALSE}
# Standard QC comparison pipeline
qc_compare <- function(prog1_path, prog2_path, keys,
                        tolerance = NULL,
                        output_dir = "qc_results/") {

  # Load both versions
  prog1 <- haven::read_xpt(prog1_path)
  prog2 <- haven::read_xpt(prog2_path)
  domain <- toupper(tools::file_path_sans_ext(basename(prog1_path)))

  # Compare
  result <- diffdf(prog1, prog2, keys=keys, tolerance=tolerance)

  # Write QC report
  report_path <- file.path(output_dir,
                            paste0(tolower(domain), "_qc_report.txt"))
  dir.create(output_dir, recursive=TRUE, showWarnings=FALSE)

  sink(report_path)
  cat(sprintf("QC Comparison Report: %s\n", domain))
  cat(sprintf("Generated: %s\n\n", Sys.time()))
  print(result)
  if (!diffdf_is_identical(result)) {
    cat("\nDetailed differences:\n")
    for (v in diffdf_issue_cols(result)) {
      cat(sprintf("\n%s:\n", v))
      print(diffdf_var_diffs(result, v))
    }
  }
  sink()

  message(sprintf("[%s] QC: %s | Diffs: %d variable(s)",
                  domain,
                  if(diffdf_is_identical(result)) "PASS" else "FAIL",
                  diffdf_n_diffs(result)))

  invisible(result)
}
```

### 3.2 Exercise — complete QC comparison

Run a QC comparison between `adsl` and `adsl_qc`, build a resolution table for each difference found, and determine if the dataset passes QC.

```{r qc-complete-ex, exercise=TRUE, exercise.lines=35}
# Step 1: compare
result <- diffdf(___, ___, keys=c("STUDYID","USUBJID"))

# Step 2: if identical, pass; otherwise investigate
if (diffdf_is_identical(result)) {
  cat("✓ QC PASS — Datasets are identical\n")
} else {
  cat(sprintf("✗ QC FAIL — %d variable(s) differ\n\n", diffdf_n_diffs(result)))

  # Step 3: build resolution table
  resolution <- tibble(
    Variable     = diffdf_issue_cols(result),
    N_Diffs      = sapply(diffdf_issue_cols(result), function(v)
                          nrow(diffdf_var_diffs(result, v))),
    Investigation = "Pending",
    Resolution    = NA_character_
  )

  # Step 4: annotate (in real QC, programmer fills these in)
  resolution <- resolution |>
    mutate(
      Investigation = case_when(
        Variable == "BMIBL"   ~ "Rounding: Programmer 2 used 2 decimal places",
        Variable == "DIABFL"  ~ "Logic error: P2 used BMIBL > 30 instead of >= 30",
        Variable == "TRTDURD" ~ "Date calculation: P2 used inclusive end date",
        TRUE                   ~ "Under investigation"
      ),
      Resolution = case_when(
        Variable == "BMIBL"   ~ "ACCEPTED — within tolerance 0.1",
        Variable == "DIABFL"  ~ "PROGRAMMER 2 CORRECTED",
        Variable == "TRTDURD" ~ "PROGRAMMER 1 CORRECTED per SAP Section 4.2",
        TRUE                   ~ "TBD"
      )
    )

  cat("Resolution Table:\n")
  print(resolution)
}
```

---

## 4 — riskmetric: Package Risk Scoring

### 4.1 Why package risk scoring matters

21 CFR Part 11 and ICH E9(R1) require that software used in clinical trial analysis is validated. For commercial software (like SAS), the vendor provides validation documentation. For open-source R packages, the organisation using them is responsible for their own validation.

The **R Validation Hub** (a pharmaverse initiative) provides a framework based on:

1. **Risk assessment** — score each package on testability, documentation, and community support
2. **Testing evidence** — run package tests; document results
3. **Risk-based qualification** — low-risk packages require less validation effort

riskmetric automates step 1.

### 4.2 Risk scoring dimensions

| Metric | What it measures | Low risk |
|--------|-----------------|---------|
| `has_news` | CHANGELOG/NEWS file present | Yes |
| `has_vignette` | Vignettes/documentation | Yes |
| `has_tests` | Unit test suite present | Yes |
| `covr_coverage` | Test code coverage % | ≥ 80% |
| `downloads_1yr` | CRAN download count | ≥ 10,000 |
| `cran_checks` | CRAN check results passing | All pass |

### 4.3 Basic scoring pipeline

```{r riskmetric-basic, eval=FALSE}
library(riskmetric)

# Step 1: create package references
pkgs <- pkg_cohort(
  pkg_ref("admiral"),
  pkg_ref("dplyr"),
  pkg_ref("xportr"),
  pkg_ref("logrx")
)

# Step 2: assess all metrics
assessed <- assess(pkgs)

# Step 3: score
scored <- score(assessed)

# Step 4: tabulate
tbl <- pkg_score_tbl(scored)
print(tbl)
```

### 4.4 Exercise — score the pharmaverse core

Score a cohort of pharmaverse packages and identify any that require additional validation effort.

```{r riskmetric-ex, exercise=TRUE, exercise.lines=20}
pharmaverse_pkgs <- pkg_cohort(
  "admiral",
  "xportr",
  "logrx",
  "diffdf",
  "pharmaverseadam",
  "metacore"
)

scored <- pharmaverse_pkgs |>
  assess() |>
  ___()          # score

tbl <- pkg_score_tbl(scored)

cat("── Pharmaverse Package Risk Scores ─────────────────\n")
print(arrange(tbl, Risk_Score))

cat("\n── Packages requiring enhanced validation ──────────\n")
high_risk <- filter(tbl, Risk_Label != "Low Risk")
if (nrow(high_risk) == 0) cat("  None — all packages low risk\n")
else print(select(high_risk, Package, Risk_Score, Risk_Label))
```

```{r riskmetric-ex-solution}
pharmaverse_pkgs <- pkg_cohort(
  "admiral",
  "xportr",
  "logrx",
  "diffdf",
  "pharmaverseadam",
  "metacore"
)

scored <- pharmaverse_pkgs |>
  assess() |>
  score()

tbl <- pkg_score_tbl(scored)

cat("── Pharmaverse Package Risk Scores ─────────────────\n")
print(arrange(tbl, Risk_Score))

cat("\n── Packages requiring enhanced validation ──────────\n")
high_risk <- filter(tbl, Risk_Label != "Low Risk")
if (nrow(high_risk) == 0) cat("  None — all packages low risk\n")
else print(select(high_risk, Package, Risk_Score, Risk_Label))
```

### 4.5 Exercise — individual package deep dive

```{r riskmetric-deep-ex, exercise=TRUE, exercise.lines=16}
# Deep dive into admiral's risk profile
admiral_ref <- pkg_ref("admiral")

admiral_assessed <- assess(list(admiral_ref))
admiral_scored   <- score(admiral_assessed)

cat("── admiral Risk Profile ──────────────────────────────\n")
print(admiral_scored[[1]])
```

---

## 5 — Risk Tiers and Validation Strategy

### 5.1 Mapping scores to validation effort

```{r risk-tiers, eval=FALSE}
# R Validation Hub recommended tiering
classify_validation_effort <- function(score_tbl) {
  score_tbl |>
    mutate(
      Validation_Effort = case_when(
        Risk_Score >= 0.75 ~ "Tier 1 — Test suite review + use in production",
        Risk_Score >= 0.50 ~ "Tier 2 — Additional testing + documented assessment",
        TRUE               ~ "Tier 3 — Full validation including independent testing"
      ),
      Action = case_when(
        Risk_Score >= 0.75 ~ "Review test results; document in validation plan",
        Risk_Score >= 0.50 ~ "Supplement with organisation-specific tests",
        TRUE               ~ "Evaluate alternatives or conduct full qualification"
      )
    )
}
```

### 5.2 Generating a risk assessment report

```{r risk-report-demo, eval=FALSE}
# Generate a validation plan excerpt
generate_validation_excerpt <- function(score_tbl) {
  cat("R Package Validation Plan — CDISCPILOT01\n")
  cat(sprintf("Generated: %s\n\n", Sys.Date()))

  for (i in seq_len(nrow(score_tbl))) {
    row <- score_tbl[i,]
    cat(sprintf("Package: %s v%s\n", row$Package, packageVersion(row$Package)))
    cat(sprintf("  Risk Score  : %.3f (%s)\n", row$Risk_Score, row$Risk_Label))
    cat(sprintf("  Test Cov    : %d%%\n", row$Cov_Pct))
    cat(sprintf("  Downloads   : %s /yr\n",
                formatC(row$Downloads_1yr, big.mark=",", format="d")))
    cat(sprintf("  Disposition : %s\n\n",
                if(row$Risk_Score>=0.75) "ACCEPTED for use in regulated systems"
                else "REQUIRES additional validation documentation"))
  }
}
```

### 5.3 Exercise — full risk assessment

Score a full validation cohort, classify validation effort, and identify any packages needing escalated review.

```{r full-risk-ex, exercise=TRUE, exercise.lines=30}
validation_cohort <- pkg_cohort(
  "admiral", "dplyr", "ggplot2", "tidyr",
  "xportr", "logrx", "diffdf",
  "pharmaverseadam", "metacore", "teal"
)

scored_tbl <- validation_cohort |>
  ___() |>   # assess
  ___() |>   # score
  pkg_score_tbl()

# Add validation tier
tiered <- scored_tbl |>
  mutate(
    Tier = case_when(
      Risk_Score >= 0.75 ~ 1L,
      Risk_Score >= 0.50 ~ 2L,
      TRUE               ~ 3L
    ),
    Effort = case_when(
      Tier == 1 ~ "Test review",
      Tier == 2 ~ "Supplemental testing",
      Tier == 3 ~ "Full qualification"
    )
  ) |>
  arrange(Tier, Risk_Score)

cat("── Validation Plan Summary ─────────────────────────\n")
print(select(tiered, Package, Risk_Score, Risk_Label, Tier, Effort))

cat(sprintf("\nTier 1 (low effort)  : %d packages\n", sum(tiered$Tier==1)))
cat(sprintf("Tier 2 (medium)      : %d packages\n", sum(tiered$Tier==2)))
cat(sprintf("Tier 3 (full qual.)  : %d packages\n", sum(tiered$Tier==3)))
```

```{r full-risk-ex-solution}
validation_cohort <- pkg_cohort(
  "admiral", "dplyr", "ggplot2", "tidyr",
  "xportr", "logrx", "diffdf",
  "pharmaverseadam", "metacore", "teal"
)

scored_tbl <- validation_cohort |>
  assess() |>
  score() |>
  pkg_score_tbl()

tiered <- scored_tbl |>
  mutate(
    Tier = case_when(
      Risk_Score >= 0.75 ~ 1L,
      Risk_Score >= 0.50 ~ 2L,
      TRUE               ~ 3L
    ),
    Effort = case_when(
      Tier == 1 ~ "Test review",
      Tier == 2 ~ "Supplemental testing",
      Tier == 3 ~ "Full qualification"
    )
  ) |>
  arrange(Tier, Risk_Score)

cat("── Validation Plan Summary ─────────────────────────\n")
print(select(tiered, Package, Risk_Score, Risk_Label, Tier, Effort))

cat(sprintf("\nTier 1 (low effort)  : %d packages\n", sum(tiered$Tier==1)))
cat(sprintf("Tier 2 (medium)      : %d packages\n", sum(tiered$Tier==2)))
cat(sprintf("Tier 3 (full qual.)  : %d packages\n", sum(tiered$Tier==3)))
```

---

## 6 — Capstone: Full QC and Governance Workflow

Combine diffdf and riskmetric into a single QC governance workflow.

```{r capstone-qcgov, exercise=TRUE, exercise.eval=FALSE, exercise.lines=75}
library(purrr)

# ══════════════════════════════════════════════════════════
#  PART A: Dataset QC via diffdf
# ══════════════════════════════════════════════════════════
cat("╔══════════════════════════════════════════════════════╗\n")
cat("║  PART A: Double-Programming QC — ADSL               ║\n")
cat("╚══════════════════════════════════════════════════════╝\n\n")

qc_result <- diffdf(
  base    = ___,
  compare = adsl_qc,
  keys    = c("STUDYID", "USUBJID"),
  tolerance = 0.05   # allow minor rounding
)

if (diffdf_is_identical(qc_result)) {
  cat("✓ ADSL QC PASS — Datasets identical within tolerance\n")
} else {
  n_diff <- diffdf_n_diffs(qc_result)
  cat(sprintf("⚠ ADSL QC: %d variable(s) with differences\n\n", n_diff))

  qc_log <- imap_dfr(
    setNames(diffdf_issue_cols(qc_result), diffdf_issue_cols(qc_result)),
    function(v, nm) {
      diffs <- diffdf_var_diffs(qc_result, v)
      tibble(
        Variable = v,
        N_Diffs  = nrow(diffs),
        Max_Abs_Diff = if (is.numeric(diffs$base_value %||% NA)) {
          tryCatch(max(abs(as.numeric(diffs$base_value) -
                            as.numeric(diffs$compare_value)), na.rm=TRUE),
                   error=function(e) NA_real_)
        } else NA_real_
      )
    }
  )
  cat("Difference summary:\n")
  print(qc_log)
}

# ══════════════════════════════════════════════════════════
#  PART B: Package risk assessment
# ══════════════════════════════════════════════════════════
cat("\n╔══════════════════════════════════════════════════════╗\n")
cat("║  PART B: Package Risk Assessment                     ║\n")
cat("╚══════════════════════════════════════════════════════╝\n\n")

study_pkgs <- pkg_cohort(
  "admiral", "dplyr", "ggplot2", "xportr",
  "logrx", "diffdf", "pharmaverseadam"
)

risk_tbl <- study_pkgs |>
  ___() |>
  ___() |>
  pkg_score_tbl() |>
  mutate(
    Approved = Risk_Score >= 0.70,
    Flag     = if_else(!Approved, "⚠ REVIEW REQUIRED", "✓ APPROVED")
  ) |>
  arrange(Risk_Score)

cat("Package Risk Summary:\n")
print(select(risk_tbl, Package, Risk_Score, Risk_Label, Flag))

n_approved <- sum(risk_tbl$Approved)
cat(sprintf("\n✓ Approved: %d / %d packages\n", n_approved, nrow(risk_tbl)))

flagged <- filter(risk_tbl, !Approved)
if (nrow(flagged) > 0) {
  cat(sprintf("⚠ Flagged for review: %s\n",
              paste(flagged$Package, collapse=", ")))
}
```

```{r capstone-qcgov-solution}
library(purrr)

cat("╔══════════════════════════════════════════════════════╗\n")
cat("║  PART A: Double-Programming QC — ADSL               ║\n")
cat("╚══════════════════════════════════════════════════════╝\n\n")

qc_result <- diffdf(
  base      = adsl,
  compare   = adsl_qc,
  keys      = c("STUDYID", "USUBJID"),
  tolerance = 0.05
)

if (diffdf_is_identical(qc_result)) {
  cat("✓ ADSL QC PASS — Datasets identical within tolerance\n")
} else {
  n_diff <- diffdf_n_diffs(qc_result)
  cat(sprintf("⚠ ADSL QC: %d variable(s) with differences\n\n", n_diff))

  qc_log <- imap_dfr(
    setNames(diffdf_issue_cols(qc_result), diffdf_issue_cols(qc_result)),
    function(v, nm) {
      diffs <- diffdf_var_diffs(qc_result, v)
      tibble(
        Variable     = v,
        N_Diffs      = nrow(diffs),
        Max_Abs_Diff = tryCatch(
          max(abs(as.numeric(diffs$base_value) -
                    as.numeric(diffs$compare_value)), na.rm=TRUE),
          error = function(e) NA_real_
        )
      )
    }
  )
  cat("Difference summary:\n")
  print(qc_log)
}

cat("\n╔══════════════════════════════════════════════════════╗\n")
cat("║  PART B: Package Risk Assessment                     ║\n")
cat("╚══════════════════════════════════════════════════════╝\n\n")

study_pkgs <- pkg_cohort(
  "admiral", "dplyr", "ggplot2", "xportr",
  "logrx", "diffdf", "pharmaverseadam"
)

risk_tbl <- study_pkgs |>
  assess() |>
  score() |>
  pkg_score_tbl() |>
  mutate(
    Approved = Risk_Score >= 0.70,
    Flag     = if_else(!Approved, "⚠ REVIEW REQUIRED", "✓ APPROVED")
  ) |>
  arrange(Risk_Score)

cat("Package Risk Summary:\n")
print(select(risk_tbl, Package, Risk_Score, Risk_Label, Flag))

n_approved <- sum(risk_tbl$Approved)
cat(sprintf("\n✓ Approved: %d / %d packages\n", n_approved, nrow(risk_tbl)))

flagged <- filter(risk_tbl, !Approved)
if (nrow(flagged) > 0)
  cat(sprintf("⚠ Flagged for review: %s\n", paste(flagged$Package, collapse=", ")))
```

---

## Summary

You now have the two main QC and governance tools for regulated R programming:

**diffdf:**
- `diffdf(base, compare, keys, tolerance)` — compare two data frames
- `diffdf_is_identical(result)` — single TRUE/FALSE pass/fail signal
- `diffdf_n_diffs(result)` — count of variables with differences
- `diffdf_issue_cols(result)` — names of variables that differ
- `diffdf_var_diffs(result, variable)` — row-level differences for one variable
- `tolerance` — allow numeric differences up to a threshold (for SAS/R rounding)
- Used in double-programming QC: two independent derivations must match

**riskmetric:**
- `pkg_ref(name)` → `assess()` → `score()` → `pkg_score_tbl()` — the four-step pipeline
- `pkg_cohort(...)` — shorthand to build a list of `pkg_ref` objects
- Scores 0–1 across: NEWS, vignettes, tests, coverage, downloads, CRAN checks
- Low Risk (≥ 0.75): test review only; Medium (0.50–0.75): supplemental tests; High (< 0.50): full qualification
- Required by R Validation Hub framework for regulated clinical use

Next: **Lesson 47** — the full Pharmaverse10 capstone, Part 1: deriving ADSL and ADAE from SDTM using admiral, with the complete pipeline from raw data through XPT export.
