---
title: "Lesson 48 — Capstone Part 2: ADaM → TLF → Submission Package"
author: "Hamid Tavakoli, MD, MSc"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: www/custom.css
runtime: shiny_prerendered
description: >
  Tie together the entire Pharmaverse10 toolchain: derive ADLB from pharmaversesdtm,
  produce submission-ready Tables/Listings/Figures with gtsummary + tfrmt, export
  with r2rtf, log every program with logrx, QC outputs with diffdf, export ADaMs as
  XPT and Dataset-JSON, and assemble a complete eCTD submission package.
---

```{r setup, include=FALSE}
# ==============================================================================
# PROJECT       : pharmaverse learnr Tutorials
# DESCRIPTION   : Interactive learnr tutorials for clinical R programmers
#                 transitioning from SAS to the pharmaverse ecosystem.
#
# SCRIPT NAME   : 48_capstone_part2.Rmd
# PURPOSE       : Capstone Part 2 — ADaM → TLF → Submission Package: demographics
#                 and TEAE tables (gtsummary + tfrmt), RTF export
#                 (r2rtf), logrx logging, diffdf QC, XPT + Dataset-JSON.
#
# AUTHOR        : Hamid Tavakoli
# DATE CREATED  : 2026-05-24
# ==============================================================================
# DISCLAIMER: Independent work. No proprietary IP. Public docs only.
# LICENSE       : MIT License
# ==============================================================================
library(learnr)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(purrr)
library(pharmaverseadam)

# ── learnr options ──────────────────────────────────────────────────────────
tutorial_options(exercise.timelimit = 120, exercise.lines = 30)
knitr::opts_chunk$set(echo = FALSE)

# ══════════════════════════════════════════════════════════════════════════════
# SECTION A — pharmaverseadam datasets (real data, pre-cleaned)
# ══════════════════════════════════════════════════════════════════════════════

adsl_raw <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl_raw))) {
  set.seed(2014); n <- nrow(adsl_raw)
  ht <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 161, 7)), round(rnorm(n, 174, 8)))
  wt <- ifelse(adsl_raw$SEX == "F", round(rnorm(n, 68, 12), 1), round(rnorm(n, 82, 14), 1))
  adsl_raw <- adsl_raw |>
    mutate(HEIGHTBL = ht, WEIGHTBL = wt,
           BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
} else if (!"BMIBL" %in% names(adsl_raw)) {
  adsl_raw <- mutate(adsl_raw, BMIBL = round(WEIGHTBL / (HEIGHTBL / 100)^2, 1))
}
adsl <- adsl_raw |>
  mutate(
    ITTFL  = SAFFL,
    DIABFL = if_else(!is.na(BMIBL) & BMIBL >= 30, "Y", "N", missing = "N"),
    TRT01P = factor(TRT01P,
                    levels = c("Placebo", "Xanomeline Low Dose", "Xanomeline High Dose"))
  )

adae_raw <- pharmaverseadam::adae
if ("TRTA" %in% names(adae_raw) && !"TRT01A" %in% names(adae_raw))
  adae_raw <- rename(adae_raw, TRT01A = TRTA)
adae <- adae_raw |>
  mutate(AESEVN = case_when(
    AESEV == "MILD"     ~ 1L,
    AESEV == "MODERATE" ~ 2L,
    AESEV == "SEVERE"   ~ 3L,
    TRUE ~ NA_integer_
  ))

adlb <- pharmaverseadam::adlb
if (is.character(adlb$ATOXGR))
  adlb <- mutate(adlb, ATOXGR = as.integer(ATOXGR), BTOXGR = as.integer(BTOXGR))
if ("DTYPE" %in% names(adlb))
  adlb <- filter(adlb, is.na(DTYPE))
if (!"TRTSDT" %in% names(adlb))
  adlb <- left_join(adlb, select(adsl, USUBJID, TRTSDT), by = "USUBJID")

# ══════════════════════════════════════════════════════════════════════════════
# SECTION B — gtsummary mock (API-compatible)
# ══════════════════════════════════════════════════════════════════════════════

# Internal helpers
.fmt_mean_sd <- function(x) {
  x <- x[!is.na(x)]
  sprintf("%.1f (%.2f)", mean(x), sd(x))
}
.fmt_n_pct <- function(flag, denom) {
  n <- sum(flag == "Y", na.rm = TRUE)
  sprintf("%d (%.1f%%)", n, 100 * n / denom)
}

# tbl_summary — summarise a dataset by a grouping column
tbl_summary <- function(data, by = NULL, include = NULL,
                        label = list(), statistic = list(),
                        digits = NULL, missing = "no", ...) {
  # Capture 'by' via substitute() BEFORE any is.null() check to avoid
  # forcing the promise (which would error if 'by' is a bare column name)
  by_expr <- substitute(by)
  by_col  <- if (!identical(by_expr, quote(NULL)) && length(by_expr) > 0) {
    as.character(by_expr)
  } else NULL

  # resolve 'include' to column names
  inc_expr <- substitute(include)
  inc_is_null <- identical(inc_expr, quote(NULL)) || length(inc_expr) == 0
  if (!inc_is_null) {
    if (is.call(inc_expr)) {
      inc_vars <- eval(inc_expr, envir = list(c = c), enclos = baseenv())
      if (is.character(inc_vars)) {
        data <- data[, c(if (!is.null(by_col)) by_col, inc_vars), drop = FALSE]
      }
    } else if (is.character(eval(inc_expr))) {
      data <- data[, c(if (!is.null(by_col)) by_col, eval(inc_expr)), drop = FALSE]
    }
  }

  vars <- setdiff(names(data), by_col)
  if (!is.null(by_col)) {
    groups <- sort(unique(data[[by_col]]))
  } else {
    groups <- "Overall"
  }

  rows <- list()
  for (v in vars) {
    col <- data[[v]]
    lbl <- if (!is.null(label[[v]])) label[[v]] else v
    if (is.numeric(col)) {
      if (!is.null(by_col)) {
        vals <- sapply(groups, function(g) {
          sub <- col[data[[by_col]] == g]
          .fmt_mean_sd(sub)
        })
        rows[[length(rows) + 1]] <- c(Variable = lbl, setNames(vals, as.character(groups)))
      } else {
        rows[[length(rows) + 1]] <- c(Variable = lbl, Overall = .fmt_mean_sd(col))
      }
    } else {
      levs <- if (is.factor(col)) levels(col) else sort(unique(col[!is.na(col)]))
      for (lv in levs) {
        flag_n <- sum(col == lv, na.rm = TRUE)
        lbl_lv <- sprintf("  %s", lv)
        if (!is.null(by_col)) {
          vals <- sapply(groups, function(g) {
            n <- sum(col[data[[by_col]] == g] == lv, na.rm = TRUE)
            tot <- sum(data[[by_col]] == g, na.rm = TRUE)
            sprintf("%d (%.1f%%)", n, 100 * n / max(tot, 1))
          })
          rows[[length(rows) + 1]] <- c(Variable = lbl_lv, setNames(vals, as.character(groups)))
        } else {
          tot <- length(col[!is.na(col)])
          rows[[length(rows) + 1]] <- c(Variable = lbl_lv,
                                         Overall = sprintf("%d (%.1f%%)", flag_n,
                                                           100 * flag_n / max(tot, 1)))
        }
      }
    }
  }

  tbl <- as_tibble(bind_rows(rows))
  structure(list(table_body = tbl, by = by_col, groups = as.character(groups),
                 n = nrow(data)),
            class = "tbl_summary")
}

add_overall <- function(x, ...) {
  # append an Overall column
  tbl  <- x$table_body
  data <- environment(tbl_summary)  # not accessible; just fake it
  x$has_overall <- TRUE
  x
}

add_p <- function(x, ...) { x$has_p <- TRUE; x }
add_n <- function(x, ...) { x$has_n <- TRUE; x }

bold_labels   <- function(x, ...) { x$bold_labels   <- TRUE; x }
italicize_labels <- function(x, ...) { x$italicize  <- TRUE; x }

modify_header <- function(x, ...) { x$custom_header <- list(...); x }
modify_caption <- function(x, caption = NULL, ...) { x$caption <- caption; x }
modify_spanning_header <- function(x, ...) { x }
modify_footnote <- function(x, ...) { x }

as_gt <- function(x, ...) {
  structure(list(tbl_summary_obj = x), class = c("gt_tbl", "tbl_summary_gt"))
}
as_flex_table <- function(x, ...) {
  structure(list(tbl_summary_obj = x), class = "flextable")
}

print.tbl_summary <- function(x, ...) {
  if (!is.null(x$caption)) cat(sprintf("\nTable: %s\n", x$caption))
  cat(sprintf("N = %d\n\n", x$n))
  print(x$table_body, n = 40)
  invisible(x)
}

# tbl_merge — merge two tbl_summary objects side-by-side
tbl_merge <- function(tbls, tab_spanner = NULL) {
  combined <- reduce(map(tbls, "table_body"), function(a, b) {
    bind_cols(a, select(b, -any_of("Variable")))
  })
  structure(list(table_body = combined, merged = TRUE), class = "tbl_summary")
}

tbl_stack <- function(tbls, group_header = NULL) {
  combined <- bind_rows(map(tbls, "table_body"))
  structure(list(table_body = combined), class = "tbl_summary")
}

# ══════════════════════════════════════════════════════════════════════════════
# SECTION C — tfrmt mock (submission formatting layer)
# ══════════════════════════════════════════════════════════════════════════════

# Core tfrmt objects
tfrmt <- function(label = NULL, param = NULL, column = NULL, value = NULL,
                  sorting_cols = NULL, body_plan = NULL, col_plan = NULL,
                  col_style_plan = NULL, row_grp_plan = NULL,
                  title = NULL, subtitle = NULL, footnote_plan = NULL, ...) {
  # Capture column-selector args via substitute() before any forced evaluation
  # so callers can pass bare names (label, param, TRT01P, value) as in real tfrmt
  .sym <- function(e) {
    ex <- substitute(e)
    if (identical(ex, quote(NULL)) || length(ex) == 0) NULL else as.character(ex)
  }
  label_str  <- tryCatch(as.character(substitute(label)),  error = function(e) NULL)
  param_str  <- tryCatch(as.character(substitute(param)),  error = function(e) NULL)
  column_str <- tryCatch(as.character(substitute(column)), error = function(e) NULL)
  value_str  <- tryCatch(as.character(substitute(value)),  error = function(e) NULL)
  # If the argument was the literal NULL default, set to NULL
  if (identical(label_str,  "NULL")) label_str  <- NULL
  if (identical(param_str,  "NULL")) param_str  <- NULL
  if (identical(column_str, "NULL")) column_str <- NULL
  if (identical(value_str,  "NULL")) value_str  <- NULL
  structure(
    list(label = label_str, param = param_str, column = column_str, value = value_str,
         body_plan = body_plan, col_plan = col_plan, title = title,
         subtitle = subtitle, footnote_plan = footnote_plan),
    class = "tfrmt"
  )
}

body_plan <- function(...) {
  frmt_structs <- list(...)
  structure(list(frmt_structs = frmt_structs), class = "body_plan")
}

frmt_structure <- function(group_val = NULL, label_val = NULL, ...) {
  structure(list(group_val = group_val, label_val = label_val, frmts = list(...)),
            class = "frmt_structure")
}

frmt <- function(expression, missing = "") {
  structure(list(expression = expression, missing = missing), class = "frmt")
}

frmt_combine <- function(expression, ..., missing = "") {
  structure(list(expression = expression, params = list(...), missing = missing),
            class = "frmt_combine")
}

frmt_when <- function(...) {
  conditions <- list(...)
  structure(list(conditions = conditions), class = "frmt_when")
}

col_plan <- function(...) {
  structure(list(cols = list(...)), class = "col_plan")
}

row_grp_plan <- function(..., label_loc = element_row_grp_loc("indented")) {
  structure(list(args = list(...), label_loc = label_loc), class = "row_grp_plan")
}

element_row_grp_loc <- function(location = "indented") {
  structure(list(location = location), class = "element_row_grp_loc")
}

footnote_plan <- function(..., marks = "numbers") {
  structure(list(footnotes = list(...), marks = marks), class = "footnote_plan")
}

footnote_structure <- function(footnote_text, column_val = NULL, label_val = NULL) {
  structure(list(text = footnote_text, column_val = column_val,
                 label_val = label_val), class = "footnote_structure")
}

print_to_gt <- function(data, .tfrmt, ...) {
  # Simulate applying tfrmt formatting
  title_str <- if (!is.null(.tfrmt$title)) .tfrmt$title else "Clinical Table"
  cat(sprintf("\n[gt table: \"%s\"]\n", title_str))
  cat(sprintf("  Rows: %d  |  Columns: %d\n", nrow(data), ncol(data)))
  if (!is.null(.tfrmt$subtitle)) cat(sprintf("  Subtitle: %s\n", .tfrmt$subtitle))
  invisible(structure(list(tfrmt = .tfrmt, data = data), class = "gt_tbl"))
}

# data_bindings helper used in ARD-based tfrmt workflow
make_big_n_df <- function(data, column, ...) {
  data |>
    count(!!sym(column)) |>
    mutate(param = "big_n", label = as.character(!!sym(column)))
}

# ══════════════════════════════════════════════════════════════════════════════
# SECTION D — r2rtf mock (RTF output)
# ══════════════════════════════════════════════════════════════════════════════

rtf_body <- function(x, ...) { attr(x, "rtf_body") <- list(...); x }
rtf_title <- function(x, title = "", subtitle = "") {
  attr(x, "rtf_title")    <- title
  attr(x, "rtf_subtitle") <- subtitle
  x
}
rtf_footnote <- function(x, footnote = "") { attr(x, "rtf_footnote") <- footnote; x }
rtf_colheader <- function(x, colheader = NULL, col_rel_width = NULL, ...) {
  attr(x, "rtf_colheader") <- colheader; x
}
rtf_page <- function(x, ...) { attr(x, "rtf_page") <- list(...); x }

rtf_encode <- function(x, ...) {
  attr(x, "rtf_encoded") <- TRUE
  x
}

write_rtf <- function(x, file, ...) {
  content <- sprintf(
    "{\\rtf1\\ansi\\deff0\n{\\b Clinical Table}\n%s rows x %s cols\n}",
    nrow(x), ncol(x)
  )
  writeLines(content, file)
  invisible(file)
}

# ══════════════════════════════════════════════════════════════════════════════
# SECTION E — logrx mock (program logging)
# ══════════════════════════════════════════════════════════════════════════════

axecute <- function(file, log_file = NULL, ...) {
  if (!file.exists(file)) stop(sprintf("Script not found: %s", file))
  if (is.null(log_file)) log_file <- sub("\\.R$", ".log", file)

  messages <- character(0)
  warnings <- character(0)
  errors   <- character(0)
  status   <- "SUCCESS"

  output <- tryCatch(
    withCallingHandlers(
      capture.output(source(file, local = new.env(parent = globalenv()))),
      message = function(m) {
        messages <<- c(messages, conditionMessage(m))
        invokeRestart("muffleMessage")
      },
      warning = function(w) {
        warnings <<- c(warnings, conditionMessage(w))
        invokeRestart("muffleWarning")
      }
    ),
    error = function(e) {
      errors  <<- c(errors, conditionMessage(e))
      status  <<- "ERROR"
      character(0)
    }
  )

  log_obj <- structure(
    list(
      file      = file,
      log_file  = log_file,
      status    = status,
      messages  = messages,
      warnings  = warnings,
      errors    = errors,
      output    = output,
      session   = list(R_version = R.version.string,
                       date      = format(Sys.time(), "%Y-%m-%d %H:%M:%S"))
    ),
    class = "logrx_log"
  )
  invisible(log_obj)
}

axecute_string <- function(code, log_file = NULL) {
  tmp <- tempfile(fileext = ".R")
  on.exit(unlink(tmp))
  writeLines(code, tmp)
  axecute(tmp, log_file = log_file)
}

log_select   <- function(x) x$output
log_warn_select  <- function(x) x$warnings
log_error_select <- function(x) x$errors

print.logrx_log <- function(x, ...) {
  cat(sprintf("── logrx log: %s ──\n", basename(x$file)))
  cat(sprintf("Status  : %s\n", x$status))
  cat(sprintf("Messages: %d\n", length(x$messages)))
  cat(sprintf("Warnings: %d\n", length(x$warnings)))
  cat(sprintf("Errors  : %d\n", length(x$errors)))
  cat(sprintf("Run at  : %s\n", x$session$date))
  invisible(x)
}

# ══════════════════════════════════════════════════════════════════════════════
# SECTION F — diffdf mock (output QC)
# ══════════════════════════════════════════════════════════════════════════════

diffdf <- function(base, compare, keys = NULL, tolerance = NULL,
                   suppress_warnings = FALSE) {
  # Align rows on keys (if provided)
  if (!is.null(keys)) {
    base    <- arrange(base,    across(all_of(keys)))
    compare <- arrange(compare, across(all_of(keys)))
  }

  shared_cols <- intersect(names(base), names(compare))
  only_base   <- setdiff(names(base),    names(compare))
  only_cmp    <- setdiff(names(compare), names(base))

  type_diffs <- tibble(
    VARIABLE    = character(),
    CLASS.BASE  = character(),
    CLASS.COMP  = character()
  )
  val_diffs <- list()

  for (col in shared_cols) {
    bc <- base[[col]]; cc <- compare[[col]]
    if (!identical(class(bc), class(cc))) {
      type_diffs <- add_row(type_diffs,
                            VARIABLE   = col,
                            CLASS.BASE = class(bc)[1],
                            CLASS.COMP = class(cc)[1])
      next
    }
    n_rows <- min(nrow(base), nrow(compare))
    bc <- bc[seq_len(n_rows)]; cc <- cc[seq_len(n_rows)]
    if (is.numeric(bc) && !is.null(tolerance)) {
      differ <- abs(bc - cc) > tolerance & !(is.na(bc) & is.na(cc))
    } else {
      differ <- !mapply(identical, bc, cc) & !(is.na(bc) & is.na(cc))
    }
    if (any(differ, na.rm = TRUE)) {
      idx <- which(differ)
      val_diffs[[col]] <- tibble(
        VARIABLE    = col,
        ROW         = idx,
        BASE_VALUE  = as.character(bc[idx]),
        COMP_VALUE  = as.character(cc[idx])
      )
    }
  }

  structure(
    list(
      is_identical    = length(type_diffs) == 0 && length(val_diffs) == 0 &&
                        length(only_base) == 0   && length(only_cmp) == 0,
      vars_only_base  = only_base,
      vars_only_comp  = only_cmp,
      type_diffs      = type_diffs,
      val_diffs       = val_diffs,
      n_diffs         = sum(sapply(val_diffs, nrow)),
      issue_cols      = names(val_diffs)
    ),
    class = "diffdf"
  )
}

diffdf_is_identical <- function(x) x$is_identical
diffdf_n_diffs      <- function(x) x$n_diffs
diffdf_issue_cols   <- function(x) x$issue_cols
diffdf_var_diffs    <- function(x, var) x$val_diffs[[var]]

print.diffdf <- function(x, ...) {
  cat(sprintf("diffdf: %s\n", if (x$is_identical) "IDENTICAL" else "DIFFERENCES FOUND"))
  if (!x$is_identical) {
    if (length(x$vars_only_base) > 0) cat("  Vars only in base:   ", paste(x$vars_only_base, collapse=", "), "\n")
    if (length(x$vars_only_comp) > 0) cat("  Vars only in compare:", paste(x$vars_only_comp, collapse=", "), "\n")
    if (nrow(x$type_diffs) > 0) { cat("  Type differences:\n"); print(x$type_diffs) }
    cat(sprintf("  Value differences: %d cells across %d variable(s)\n",
                x$n_diffs, length(x$issue_cols)))
    for (vc in x$issue_cols) {
      cat(sprintf("    %s:\n", vc)); print(x$val_diffs[[vc]])
    }
  }
  invisible(x)
}

# ══════════════════════════════════════════════════════════════════════════════
# SECTION G — xportr mock (submission export)
# ══════════════════════════════════════════════════════════════════════════════

xportr_metadata <- function(.df, metadata, domain = NULL, verbose = "warn") {
  attr(.df, "xportr.df_metadata") <- metadata
  attr(.df, "xportr.domain")      <- domain %||% deparse(substitute(.df))
  .df
}

xportr_type <- function(.df, metadata = NULL, domain = NULL, verbose = "warn") {
  md <- metadata %||% attr(.df, "xportr.df_metadata")
  if (is.null(md)) return(.df)
  for (i in seq_len(nrow(md))) {
    v <- md$variable[i]; tp <- md$type[i]
    if (!v %in% names(.df)) next
    .df[[v]] <- switch(tp,
      "text"    = as.character(.df[[v]]),
      "integer" = as.integer(.df[[v]]),
      "float"   = as.numeric(.df[[v]]),
      "numeric" = as.numeric(.df[[v]]),
      {
        if (inherits(.df[[v]], "Date")) as.numeric(.df[[v]])
        else .df[[v]]
      }
    )
  }
  .df
}

xportr_label <- function(.df, metadata = NULL, domain = NULL, verbose = "warn") {
  md <- metadata %||% attr(.df, "xportr.df_metadata")
  if (is.null(md) || !"label" %in% names(md)) return(.df)
  for (i in seq_len(nrow(md))) {
    v <- md$variable[i]
    if (v %in% names(.df)) attr(.df[[v]], "label") <- md$label[i]
  }
  .df
}

xportr_length <- function(.df, metadata = NULL, domain = NULL, verbose = "warn") {
  md <- metadata %||% attr(.df, "xportr.df_metadata")
  if (!is.null(md) && "length" %in% names(md)) {
    for (i in seq_len(nrow(md))) {
      v <- md$variable[i]
      if (v %in% names(.df)) attr(.df[[v]], "width") <- md$length[i]
    }
  }
  .df
}

xportr_format <- function(.df, metadata = NULL, domain = NULL, verbose = "warn") {
  md <- metadata %||% attr(.df, "xportr.df_metadata")
  if (!is.null(md) && "format" %in% names(md)) {
    for (i in seq_len(nrow(md))) {
      v <- md$variable[i]
      if (v %in% names(.df) && !is.na(md$format[i]) && nchar(md$format[i]) > 0)
        attr(.df[[v]], "format.sas") <- md$format[i]
    }
  }
  .df
}

xportr_order <- function(.df, metadata = NULL, domain = NULL, verbose = "warn") {
  md <- metadata %||% attr(.df, "xportr.df_metadata")
  if (is.null(md)) return(.df)
  ordered_vars <- intersect(md$variable, names(.df))
  extra_vars   <- setdiff(names(.df), md$variable)
  .df[, c(ordered_vars, extra_vars), drop = FALSE]
}

xportr_write <- function(.df, path, metadata = NULL, domain = NULL,
                         label = NULL, ...) {
  domain_name <- domain %||% attr(.df, "xportr.domain") %||%
                 toupper(deparse(substitute(.df)))
  message(sprintf("xportr: writing %s.xpt  [%d obs x %d vars]",
                  domain_name, nrow(.df), ncol(.df)))
  saveRDS(.df, gsub("\\.xpt$", ".rds", path))
  invisible(.df)
}

xportr <- function(.df, metadata, domain = NULL, label = NULL, path, ...) {
  .df |>
    xportr_metadata(metadata, domain) |>
    xportr_type()     |>
    xportr_label()    |>
    xportr_length()   |>
    xportr_format()   |>
    xportr_order()    |>
    xportr_write(path, domain = domain, label = label)
}

# ══════════════════════════════════════════════════════════════════════════════
# SECTION H — dataset.json mock (submission JSON)
# ══════════════════════════════════════════════════════════════════════════════

# type mapping helper
.r_type_to_json_type <- function(col) {
  if (inherits(col, "Date"))    return("date")
  if (is.integer(col))          return("integer")
  if (is.numeric(col))          return("decimal")
  return("string")
}

dataset_json <- function(data, file_oid = NULL, dataset_oid = NULL,
                          description = NULL, ...) {
  domain <- toupper(deparse(substitute(data)))
  structure(
    list(
      fileOID           = file_oid %||% paste0("www.example.com/", domain),
      datasetOID        = dataset_oid %||% domain,
      description       = description %||% paste(domain, "Dataset"),
      domain            = domain,
      data              = data,
      items             = tibble(
        variable    = names(data),
        data_type   = sapply(data, .r_type_to_json_type),
        length      = sapply(data, function(x) if (is.character(x)) max(nchar(x[!is.na(x)]), 0) else 8L),
        label       = names(data)
      )
    ),
    class = "dataset_json"
  )
}

write_dataset_json <- function(x, path, na = "null", ...) {
  require(jsonlite, quietly = TRUE)
  payload <- list(
    datasetJSONVersion = "1.1.0",
    fileOID            = x$fileOID,
    datasetOID         = x$datasetOID,
    description        = x$description,
    records            = nrow(x$data),
    columns            = ncol(x$data),
    items              = as.list(x$items),
    itemData           = lapply(seq_len(nrow(x$data)), function(i)
                            unname(as.list(x$data[i, ])))
  )
  jsonlite::write_json(payload, path, auto_unbox = TRUE, na = na, ...)
  message(sprintf("dataset.json: wrote %s  [%d records]", basename(path), nrow(x$data)))
  invisible(path)
}

print.dataset_json <- function(x, ...) {
  cat(sprintf("<dataset_json>  domain=%s  records=%d  vars=%d\n",
              x$datasetOID, nrow(x$data), ncol(x$data)))
  cat(sprintf("  fileOID: %s\n", x$fileOID))
  invisible(x)
}

# ══════════════════════════════════════════════════════════════════════════════
# SECTION I — Submission metadata specs
# ══════════════════════════════════════════════════════════════════════════════

adsl_spec <- tribble(
  ~dataset, ~variable, ~label,                          ~type,    ~length, ~format,
  "ADSL",   "STUDYID", "Study Identifier",              "text",   12,      "",
  "ADSL",   "USUBJID", "Unique Subject Identifier",     "text",   50,      "",
  "ADSL",   "SUBJID",  "Subject Identifier",            "text",   10,      "",
  "ADSL",   "SITEID",  "Study Site Identifier",         "text",   10,      "",
  "ADSL",   "AGE",     "Age",                           "float",  8,       "",
  "ADSL",   "AGEU",    "Age Units",                     "text",   5,       "",
  "ADSL",   "SEX",     "Sex",                           "text",   1,       "",
  "ADSL",   "RACE",    "Race",                          "text",   32,      "",
  "ADSL",   "SAFFL",   "Safety Population Flag",        "text",   1,       "",
  "ADSL",   "ITTFL",   "ITT Population Flag",           "text",   1,       "",
  "ADSL",   "TRT01P",  "Planned Treatment for Period 1","text",   40,      "",
  "ADSL",   "TRT01A",  "Actual Treatment for Period 1", "text",   40,      "",
  "ADSL",   "TRTSDT",  "Date of First Exposure to Treatment","float",8,    "DATE9.",
  "ADSL",   "TRTEDT",  "Date of Last Exposure to Treatment", "float",8,    "DATE9.",
  "ADSL",   "TRTDURD", "Total Treatment Duration (Days)","float", 8,       "",
  "ADSL",   "HEIGHTBL","Baseline Height (cm)",          "float",  8,       "",
  "ADSL",   "WEIGHTBL","Baseline Weight (kg)",          "float",  8,       "",
  "ADSL",   "BMIBL",   "Baseline BMI (kg/m2)",          "float",  8,       "",
  "ADSL",   "DTHFL",   "Subject Death Flag",            "text",   1,       "",
  "ADSL",   "DIABFL",  "Diabetes Flag (BMI >= 30)",     "text",   1,       ""
)

adae_spec <- tribble(
  ~dataset, ~variable, ~label,                          ~type,    ~length, ~format,
  "ADAE",   "STUDYID", "Study Identifier",              "text",   12,      "",
  "ADAE",   "USUBJID", "Unique Subject Identifier",     "text",   50,      "",
  "ADAE",   "AESEQ",   "Sequence Number",               "float",  8,       "",
  "ADAE",   "AEDECOD", "Dictionary-Derived Term",       "text",   200,     "",
  "ADAE",   "AEBODSYS","Body System or Organ Class",    "text",   200,     "",
  "ADAE",   "AESEV",   "Severity/Intensity",            "text",   8,       "",
  "ADAE",   "AESEVN",  "Severity (Numeric)",            "float",  8,       "",
  "ADAE",   "AESER",   "Serious Event",                 "text",   1,       "",
  "ADAE",   "AEREL",   "Causality",                     "text",   16,      "",
  "ADAE",   "TRTEMFL", "Treatment Emergent Analysis Flag","text", 1,       "",
  "ADAE",   "ASTDT",   "Analysis Start Date",           "float",  8,       "DATE9.",
  "ADAE",   "AENDT",   "Analysis End Date",             "float",  8,       "DATE9."
)

# ══════════════════════════════════════════════════════════════════════════════
# SECTION J — Reference outputs for QC exercises
# ══════════════════════════════════════════════════════════════════════════════

# Reference demographics table (used in diffdf QC exercises)
ref_demog_counts <- adsl |>
  filter(SAFFL == "Y") |>
  count(TRT01P, SEX, name = "n")

# Reference TEAE summary
ref_teae <- adae |>
  filter(!is.na(TRTEMFL) & TRTEMFL == "Y") |>
  count(AEBODSYS, name = "n_ae") |>
  arrange(desc(n_ae)) |>
  slice_head(n = 10)

# Helper for file paths
out_dir <- function(fname) {
  file.path(tempdir(), fname)
}
```

## Welcome {-}

This is the **final lesson** in the Pharmaverse10 series — and in the entire curriculum.

You've built the ADaM datasets. Now you'll take them all the way through to a **complete eCTD submission package**: clinical tables formatted to FDA standards, RTF outputs, logrx-logged programs, diffdf-verified QC, and a folder structure ready to submit.

By the end of this lesson you will have:

- Produced **Table 14.1.1** (Demographics) and **Table 14.3.1** (TEAE by SOC) using `gtsummary`
- Applied `tfrmt` formatting templates that match FDA/ICH submission style
- Exported tables to RTF with `r2rtf`
- Logged every TLF program with `logrx::axecute()`
- QC'd your output against a reference using `diffdf`
- Exported ADSL and ADAE as XPT and Dataset-JSON
- Assembled the full submission folder tree

---

## 1 — The Submission Package Blueprint

Before writing a single line of code, let's understand what we're building.

An eCTD Study Data Package for a clinical trial typically contains:

```
submission/
├── adam/
│   ├── adsl.xpt
│   ├── adae.xpt
│   └── adlb.xpt
├── json/
│   ├── adsl.json
│   └── adae.json
├── define/
│   └── define.xml          ← (stub — out of scope here)
├── tlf/
│   ├── t_14_1_1_demog.rtf
│   └── t_14_3_1_teae.rtf
├── logs/
│   ├── t_14_1_1_demog.log
│   └── t_14_3_1_teae.log
└── qc/
    ├── qc_demog.rds
    └── qc_teae.rds
```

> **Each layer has its tool:** `xportr` owns `adam/`, `dataset.json` owns `json/`, `gtsummary`+`tfrmt`+`r2rtf` own `tlf/`, `logrx` owns `logs/`, and `diffdf` owns `qc/`.

### Quiz: Package anatomy

```{r quiz-package-anatomy, echo=FALSE}
quiz(
  question(
    "Which tool is responsible for writing the XPT transport files in the adam/ folder?",
    answer("gtsummary"),
    answer("xportr", correct = TRUE),
    answer("r2rtf"),
    answer("tfrmt"),
    random_answer_order = TRUE
  ),
  question(
    "What does logrx record when you call axecute()?",
    answer("Only errors"),
    answer("Messages, warnings, errors, session info, and execution status", correct = TRUE),
    answer("R console output only"),
    answer("Package versions only"),
    random_answer_order = TRUE
  )
)
```

---

## 2 — Table 14.1.1: Demographics with gtsummary

The ICH E3 guideline requires a demographics and baseline characteristics table as the first table in Section 14.1. Let's build it.

### 2.1 Basic structure

A demographics table cross-tabulates continuous variables (mean ± SD) and categorical variables (n, %) by treatment arm, with an Overall column.

```{r demog-demo, exercise=FALSE, echo=TRUE}
# Safety population only
adsl_saf <- adsl |> filter(SAFFL == "Y")

demog_tbl <- tbl_summary(
  adsl_saf,
  by    = TRT01P,
  include = c("AGE", "SEX", "RACE", "BMIBL", "DTHFL"),
  label = list(
    AGE    ~ "Age (years)",
    SEX    ~ "Sex",
    RACE   ~ "Race",
    BMIBL  ~ "BMI at Baseline (kg/m²)",
    DTHFL  ~ "Death Flag"
  )
)

demog_tbl
```

### 2.2 Adding N and p-values

```{r demog-add-n, exercise=FALSE, echo=TRUE}
demog_tbl_enhanced <- tbl_summary(
  adsl |> filter(SAFFL == "Y"),
  by    = TRT01P,
  include = c("AGE", "SEX", "RACE", "BMIBL")
) |>
  add_n() |>
  add_p() |>
  modify_caption("**Table 14.1.1** — Demographics and Baseline Characteristics (Safety Population)")

demog_tbl_enhanced
```

### Exercise 2.1: Build the ITT demographics table

Build a demographics table for the **ITT population** (`ITTFL == "Y"`). Include `AGE`, `SEX`, `RACE`, `HEIGHTBL`, `WEIGHTBL`, and `BMIBL`. Add a caption "Table 14.1.1 (ITT Population)".

```{r ex-demog-itt, exercise=TRUE, exercise.lines=20}
# Filter to ITT population
adsl_itt <- adsl |> filter(___ == "Y")

# Build the table
itt_tbl <- tbl_summary(
  ___,
  by      = TRT01P,
  include = c("AGE", "SEX", "RACE", "HEIGHTBL", "WEIGHTBL", "BMIBL"),
  label   = list(
    AGE      ~ "Age (years)",
    HEIGHTBL ~ "Height at Baseline (cm)",
    WEIGHTBL ~ "Weight at Baseline (kg)",
    BMIBL    ~ "BMI at Baseline (kg/m²)"
  )
) |>
  ___(caption = "Table 14.1.1 (ITT Population)")

itt_tbl
```

```{r ex-demog-itt-hint-1}
# Filter: ITTFL == "Y"
# modify_caption() sets the table title
```

```{r ex-demog-itt-solution}
adsl_itt <- adsl |> filter(ITTFL == "Y")

itt_tbl <- tbl_summary(
  adsl_itt,
  by      = TRT01P,
  include = c("AGE", "SEX", "RACE", "HEIGHTBL", "WEIGHTBL", "BMIBL"),
  label   = list(
    AGE      ~ "Age (years)",
    HEIGHTBL ~ "Height at Baseline (cm)",
    WEIGHTBL ~ "Weight at Baseline (kg)",
    BMIBL    ~ "BMI at Baseline (kg/m²)"
  )
) |>
  modify_caption("Table 14.1.1 (ITT Population)")

itt_tbl
```

---

## 3 — Table 14.3.1: TEAE Summary

Treatment-emergent adverse events (TEAEs) are the first safety table reviewers look at. We need incidence by body system and preferred term.

### 3.1 Build the TEAE dataset

```{r teae-build, exercise=FALSE, echo=TRUE}
# Treatment-emergent AEs (TRTEMFL = "Y")
teae <- adae |>
  filter(!is.na(TRTEMFL), TRTEMFL == "Y") |>
  select(USUBJID, AEBODSYS, AEDECOD, AESEV, AESER, TRT01A)

# Subject-level incidence: deduplicate by subject x SOC x PT
teae_subj <- teae |>
  distinct(USUBJID, AEBODSYS, AEDECOD, TRT01A)

cat("Unique TEAE subject-PT pairs:", nrow(teae_subj), "\n")
cat("Body systems affected:", n_distinct(teae$AEBODSYS), "\n")
```

### 3.2 Count subjects with at least one TEAE by SOC

```{r teae-count, exercise=FALSE, echo=TRUE}
teae_soc_counts <- teae_subj |>
  count(AEBODSYS, TRT01A, name = "n_subj") |>
  pivot_wider(names_from = TRT01A, values_from = n_subj, values_fill = 0) |>
  arrange(desc(rowSums(across(where(is.integer)), na.rm = TRUE)))

teae_soc_counts
```

### Exercise 3.1: Top-10 AEs by preferred term

Produce a count of **unique subjects** experiencing each preferred term (`AEDECOD`) among **treatment-emergent, serious AEs** (`AESER == "Y"`). Keep only the top 10 PTs by total subjects affected across arms.

```{r ex-teae-pt, exercise=TRUE, exercise.lines=20}
serious_teae_pt <- adae |>
  filter(!is.na(TRTEMFL), TRTEMFL == "Y",
         ___ == "Y") |>           # serious AEs only
  distinct(USUBJID, AEDECOD, TRT01A) |>
  count(___, TRT01A, name = "n_subj") |>
  pivot_wider(names_from = TRT01A, values_from = n_subj, values_fill = 0L)

# Total across arms
serious_teae_pt <- serious_teae_pt |>
  mutate(TOTAL = rowSums(across(where(is.integer)))) |>
  arrange(___) |>
  slice_head(n = ___)

serious_teae_pt
```

```{r ex-teae-pt-hint-1}
# AESER == "Y" is the serious AE filter
# group by AEDECOD, then TRT01A
# arrange(desc(TOTAL)) to put highest first
# slice_head(n = 10)
```

```{r ex-teae-pt-solution}
serious_teae_pt <- adae |>
  filter(!is.na(TRTEMFL), TRTEMFL == "Y",
         AESER == "Y") |>
  distinct(USUBJID, AEDECOD, TRT01A) |>
  count(AEDECOD, TRT01A, name = "n_subj") |>
  pivot_wider(names_from = TRT01A, values_from = n_subj, values_fill = 0L)

serious_teae_pt <- serious_teae_pt |>
  mutate(TOTAL = rowSums(across(where(is.integer)))) |>
  arrange(desc(TOTAL)) |>
  slice_head(n = 10)

serious_teae_pt
```

---

## 4 — Formatting with tfrmt

`gtsummary` gives you correct numbers. `tfrmt` gives you the **submission-style formatting**: controlled decimal places, parentheses around percentages, combining mean ± SD into a single cell, spanning headers, and footnotes — all defined in a reusable template, separate from the data.

### 4.1 The ARD pattern

`tfrmt` works with **Analysis Results Datasets (ARDs)**: long-format tables with columns for group, label, parameter (mean, sd, n, pct…), and value. You format the ARD; `tfrmt` assembles the display.

```{r tfrmt-ard-build, exercise=FALSE, echo=TRUE}
# Build a minimal ARD for age by treatment
age_ard <- adsl |>
  filter(SAFFL == "Y") |>
  group_by(TRT01P) |>
  summarise(
    mean_val = mean(AGE, na.rm = TRUE),
    sd_val   = sd(AGE,   na.rm = TRUE),
    n_val    = n(),
    .groups  = "drop"
  ) |>
  pivot_longer(cols = c(mean_val, sd_val, n_val),
               names_to  = "param",
               values_to = "value") |>
  mutate(
    param = recode(param, mean_val = "mean", sd_val = "sd", n_val = "n"),
    label = "Age (years)"
  )

age_ard
```

### 4.2 Define a tfrmt template

```{r tfrmt-template, exercise=FALSE, echo=TRUE}
# Template: display "mean (sd)" in one cell, n separately
demog_fmt <- tfrmt(
  label   = label,
  param   = param,
  column  = TRT01P,
  value   = value,
  title   = "Table 14.1.1  Demographics and Baseline Characteristics",
  subtitle = "Safety Population",
  body_plan = body_plan(
    frmt_structure(
      label_val = ".default",
      frmt_combine(
        "{mean} ({sd})",
        mean = frmt("xx.x"),
        sd   = frmt("xx.xx")
      )
    ),
    frmt_structure(
      label_val = ".default",
      n = frmt("xx")
    )
  )
)

demog_fmt
```

### 4.3 Apply tfrmt to produce a gt table

```{r tfrmt-apply, exercise=FALSE, echo=TRUE}
age_display <- print_to_gt(age_ard, .tfrmt = demog_fmt)
```

### Exercise 4.1: Build an ARD for BMI and format it

1. Build a `bmi_ard` tibble with `mean`, `sd`, `n`, and `median` parameters for `BMIBL` by `TRT01P` (Safety population).
2. Create a `bmi_fmt` tfrmt that combines mean and sd as `"{mean} ({sd})"`.
3. Apply `print_to_gt()`.

```{r ex-tfrmt-bmi, exercise=TRUE, exercise.lines=35}
# Step 1: ARD
bmi_ard <- adsl |>
  filter(SAFFL == "Y") |>
  group_by(___) |>
  summarise(
    mean_val   = mean(BMIBL,   na.rm = TRUE),
    sd_val     = sd(BMIBL,     na.rm = TRUE),
    n_val      = n(),
    median_val = median(BMIBL, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  pivot_longer(cols = c(mean_val, sd_val, n_val, median_val),
               names_to  = "param",
               values_to = "value") |>
  mutate(
    param = recode(param,
                   mean_val   = "mean",
                   sd_val     = "sd",
                   n_val      = "n",
                   median_val = "median"),
    label = "BMI at Baseline (kg/m²)"
  )

# Step 2: tfrmt template
bmi_fmt <- tfrmt(
  label    = label,
  param    = ___,
  column   = TRT01P,
  value    = value,
  title    = "BMI Summary",
  body_plan = body_plan(
    frmt_structure(
      label_val = ".default",
      frmt_combine(
        "___",
        mean = frmt("xx.x"),
        sd   = frmt("xx.xx")
      )
    )
  )
)

# Step 3: render
print_to_gt(bmi_ard, .tfrmt = ___)
```

```{r ex-tfrmt-bmi-hint-1}
# group_by(TRT01P)
# frmt_combine expression: "{mean} ({sd})"
# print_to_gt(bmi_ard, .tfrmt = bmi_fmt)
```

```{r ex-tfrmt-bmi-solution}
bmi_ard <- adsl |>
  filter(SAFFL == "Y") |>
  group_by(TRT01P) |>
  summarise(
    mean_val   = mean(BMIBL,   na.rm = TRUE),
    sd_val     = sd(BMIBL,     na.rm = TRUE),
    n_val      = n(),
    median_val = median(BMIBL, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  pivot_longer(cols = c(mean_val, sd_val, n_val, median_val),
               names_to  = "param",
               values_to = "value") |>
  mutate(
    param = recode(param,
                   mean_val   = "mean",
                   sd_val     = "sd",
                   n_val      = "n",
                   median_val = "median"),
    label = "BMI at Baseline (kg/m²)"
  )

bmi_fmt <- tfrmt(
  label    = label,
  param    = param,
  column   = TRT01P,
  value    = value,
  title    = "BMI Summary",
  body_plan = body_plan(
    frmt_structure(
      label_val = ".default",
      frmt_combine(
        "{mean} ({sd})",
        mean = frmt("xx.x"),
        sd   = frmt("xx.xx")
      )
    )
  )
)

print_to_gt(bmi_ard, .tfrmt = bmi_fmt)
```

---

## 5 — RTF Output with r2rtf

FDA accepts RTF (Rich Text Format) for clinical study report tables. `r2rtf` converts R data frames to RTF with full control over fonts, borders, column widths, and pagination.

### 5.1 The r2rtf pipeline

```{r r2rtf-pipeline-demo, exercise=FALSE, echo=TRUE}
# Build the display data frame
demog_display <- adsl |>
  filter(SAFFL == "Y") |>
  group_by(TRT01P) |>
  summarise(
    N            = n(),
    `Age Mean(SD)` = sprintf("%.1f (%.2f)", mean(AGE, na.rm = TRUE), sd(AGE, na.rm = TRUE)),
    `Female n(%)`  = sprintf("%d (%.1f%%)",
                             sum(SEX == "F", na.rm = TRUE),
                             100 * mean(SEX == "F", na.rm = TRUE)),
    `BMI Mean(SD)` = sprintf("%.1f (%.2f)", mean(BMIBL, na.rm = TRUE), sd(BMIBL, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  mutate(TRT01P = as.character(TRT01P))

# r2rtf pipeline
rtf_path <- out_dir("t_14_1_1_demog.rtf")

demog_display |>
  rtf_title("Table 14.1.1  Demographics and Baseline Characteristics",
            "Safety Population") |>
  rtf_colheader("Treatment Group | N | Age | Sex | BMI") |>
  rtf_body() |>
  rtf_footnote("Source: ADSL; Safety population (SAFFL=Y)") |>
  rtf_encode() |>
  write_rtf(rtf_path)

cat("RTF written to:", rtf_path, "\n")
cat("File size (bytes):", file.info(rtf_path)$size, "\n")
```

### Exercise 5.1: Export the TEAE table to RTF

Using the `teae_soc_counts` data frame (Body System × Treatment Arm counts) built earlier, run the full `r2rtf` pipeline to write an RTF file called `t_14_3_1_teae.rtf`.

```{r ex-r2rtf-teae, exercise=TRUE, exercise.lines=20}
# Recreate counts
teae_display <- adae |>
  filter(!is.na(TRTEMFL), TRTEMFL == "Y") |>
  distinct(USUBJID, AEBODSYS, TRT01A) |>
  count(AEBODSYS, TRT01A, name = "n") |>
  pivot_wider(names_from = TRT01A, values_from = n, values_fill = 0L) |>
  arrange(desc(rowSums(across(where(is.integer)))))

rtf_path <- out_dir("t_14_3_1_teae.rtf")

teae_display |>
  rtf_title("Table 14.3.1  Treatment-Emergent Adverse Events by System Organ Class",
            "Safety Population") |>
  ___(colheader = "Body System | Placebo | Xanomeline Low | Xanomeline High") |>
  ___() |>
  rtf_footnote("TEAE = adverse event with onset on or after first study treatment dose.") |>
  ___() |>
  ___(rtf_path)

cat("Written:", rtf_path, "\n")
```

```{r ex-r2rtf-teae-hint-1}
# rtf_colheader(), rtf_body(), rtf_encode(), write_rtf() in that order
```

```{r ex-r2rtf-teae-solution}
teae_display <- adae |>
  filter(!is.na(TRTEMFL), TRTEMFL == "Y") |>
  distinct(USUBJID, AEBODSYS, TRT01A) |>
  count(AEBODSYS, TRT01A, name = "n") |>
  pivot_wider(names_from = TRT01A, values_from = n, values_fill = 0L) |>
  arrange(desc(rowSums(across(where(is.integer)))))

rtf_path <- out_dir("t_14_3_1_teae.rtf")

teae_display |>
  rtf_title("Table 14.3.1  Treatment-Emergent Adverse Events by System Organ Class",
            "Safety Population") |>
  rtf_colheader(colheader = "Body System | Placebo | Xanomeline Low | Xanomeline High") |>
  rtf_body() |>
  rtf_footnote("TEAE = adverse event with onset on or after first study treatment dose.") |>
  rtf_encode() |>
  write_rtf(rtf_path)

cat("Written:", rtf_path, "\n")
```

---

## 6 — Logging TLF Programs with logrx

Every submission program must be logged. `logrx::axecute()` runs a script and captures messages, warnings, errors, and session metadata into a structured log object.

### 6.1 Writing and logging a TLF script

```{r logrx-tlf, exercise=FALSE, echo=TRUE}
# Write a self-contained demographics script to a temp file
demog_script <- '
library(dplyr)
library(tidyr)
library(pharmaverseadam)

message("Program: t_14_1_1_demog.R")
message("Population: Safety (SAFFL=Y)")

adsl <- pharmaverseadam::adsl

# Synthesise BMIBL if absent
if (!all(c("HEIGHTBL", "WEIGHTBL") %in% names(adsl))) {
  set.seed(2014); n <- nrow(adsl)
  ht <- ifelse(adsl$SEX == "F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl$SEX == "F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl <- dplyr::mutate(adsl, HEIGHTBL=ht, WEIGHTBL=wt,
                         BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}

adsl_saf <- dplyr::filter(adsl, SAFFL == "Y")

result <- adsl_saf |>
  dplyr::group_by(TRT01P) |>
  dplyr::summarise(N = dplyr::n(),
                   mean_age = mean(AGE, na.rm = TRUE),
                   .groups = "drop")

message(sprintf("Subjects in safety population: %d", nrow(adsl_saf)))
print(result)
'

script_path <- out_dir("t_14_1_1_demog.R")
writeLines(demog_script, script_path)

# Execute and log
log <- axecute(script_path)
print(log)
```

### 6.2 Inspecting the log

```{r logrx-inspect, exercise=FALSE, echo=TRUE}
# Check messages captured
cat("--- Messages ---\n")
cat(paste(log_select(log), collapse = "\n"), "\n")

# Any warnings?
warns <- log_warn_select(log)
cat(sprintf("\n--- Warnings (%d) ---\n", length(warns)))
if (length(warns)) cat(paste(warns, collapse = "\n"), "\n") else cat("None\n")

# Status
cat(sprintf("\nStatus: %s\n", log$status))
```

### Exercise 6.1: Log a failing script and catch its error

Write a script that intentionally tries to join on a non-existent column (`MISSINGCOL`). Use `axecute_string()` to run it, then check `log$status` and `log_error_select()`.

```{r ex-logrx-error, exercise=TRUE, exercise.lines=18}
bad_script <- '
library(dplyr)
library(pharmaverseadam)
adsl <- pharmaverseadam::adsl
# This will error — MISSINGCOL does not exist
result <- adsl |> select(USUBJID, MISSINGCOL)
print(result)
'

log_bad <- axecute_string(___)

cat("Status:", log_bad$___, "\n")
cat("Errors:\n")
cat(paste(___(___ ), collapse = "\n"), "\n")
```

```{r ex-logrx-error-hint-1}
# axecute_string(bad_script)
# log_bad$status
# log_error_select(log_bad)
```

```{r ex-logrx-error-solution}
bad_script <- '
library(dplyr)
library(pharmaverseadam)
adsl <- pharmaverseadam::adsl
result <- adsl |> select(USUBJID, MISSINGCOL)
print(result)
'

log_bad <- axecute_string(bad_script)

cat("Status:", log_bad$status, "\n")
cat("Errors:\n")
cat(paste(log_error_select(log_bad), collapse = "\n"), "\n")
```

---

## 7 — QC with diffdf

Double-programming is mandatory in regulated submissions. A QC programmer reproduces each output independently, then `diffdf` compares the two versions cell by cell.

### 7.1 Set up a comparison scenario

```{r diffdf-qc-setup, exercise=FALSE, echo=TRUE}
# "Primary" demographics count
primary_demog <- adsl |>
  filter(SAFFL == "Y") |>
  count(TRT01P, SEX, name = "n") |>
  arrange(TRT01P, SEX)

# "QC programmer's" version — introduce a deliberate error in one cell
qc_demog <- primary_demog |>
  mutate(n = if_else(row_number() == 2L, n + 1L, n))  # off by 1 in row 2

diff_result <- diffdf(primary_demog, qc_demog, keys = c("TRT01P", "SEX"))
diff_result
```

### 7.2 Checking and resolving diffs

```{r diffdf-resolve, exercise=FALSE, echo=TRUE}
cat("Identical?", diffdf_is_identical(diff_result), "\n")
cat("Total cell differences:", diffdf_n_diffs(diff_result), "\n")
cat("Affected columns:", paste(diffdf_issue_cols(diff_result), collapse = ", "), "\n")

# Inspect the affected column
diffdf_var_diffs(diff_result, "n")
```

### Exercise 7.1: QC the TEAE SOC counts

A QC programmer has produced `qc_teae_soc` by recounting from `adae` but accidentally included non-emergent AEs (`TRTEMFL != "Y"`). Build the primary count, build the QC count (without the TRTEMFL filter), run `diffdf`, and report how many cells differ.

```{r ex-diffdf-teae, exercise=TRUE, exercise.lines=25}
# Primary count (TEAE only)
primary_teae <- adae |>
  filter(!is.na(TRTEMFL), ___ == "Y") |>
  distinct(USUBJID, AEBODSYS) |>
  count(AEBODSYS, name = "n_subj") |>
  arrange(AEBODSYS)

# QC count — missing the TRTEMFL filter (intentional bug)
qc_teae <- adae |>
  distinct(USUBJID, AEBODSYS) |>     # no TRTEMFL filter
  count(AEBODSYS, name = "n_subj") |>
  arrange(AEBODSYS)

# diffdf comparison
diff_teae <- diffdf(___, ___, keys = "AEBODSYS")

cat("Identical?", diffdf_is_identical(___), "\n")
cat("Cell diffs:", diffdf_n_diffs(___), "\n")
```

```{r ex-diffdf-teae-hint-1}
# diffdf(primary_teae, qc_teae, keys = "AEBODSYS")
# diffdf_is_identical(diff_teae)
# diffdf_n_diffs(diff_teae)
```

```{r ex-diffdf-teae-solution}
primary_teae <- adae |>
  filter(!is.na(TRTEMFL), TRTEMFL == "Y") |>
  distinct(USUBJID, AEBODSYS) |>
  count(AEBODSYS, name = "n_subj") |>
  arrange(AEBODSYS)

qc_teae <- adae |>
  distinct(USUBJID, AEBODSYS) |>
  count(AEBODSYS, name = "n_subj") |>
  arrange(AEBODSYS)

diff_teae <- diffdf(primary_teae, qc_teae, keys = "AEBODSYS")

cat("Identical?", diffdf_is_identical(diff_teae), "\n")
cat("Cell diffs:", diffdf_n_diffs(diff_teae), "\n")
```

### Quiz: QC best practice

```{r quiz-diffdf, echo=FALSE}
quiz(
  question(
    "Why must you keep your QC program completely independent of the primary program?",
    answer("To save disk space"),
    answer("So that a coding bug in one program doesn't propagate to the other, ensuring the comparison is meaningful", correct = TRUE),
    answer("Because regulators require two different programming languages"),
    answer("To double the number of log files generated"),
    random_answer_order = TRUE
  ),
  question(
    "diffdf_is_identical() returns TRUE. What does this tell you?",
    answer("The datasets have the same column names only"),
    answer("The datasets have no row-count differences, no type differences, and no value differences", correct = TRUE),
    answer("The datasets were created by the same programmer"),
    answer("The datasets have the same number of rows only"),
    random_answer_order = TRUE
  )
)
```

---

## 8 — Export: XPT and Dataset-JSON

Submissions require both XPT (legacy, for SAS-based reviewers) and Dataset-JSON (modern, v1.1) formats.

### 8.1 ADSL XPT export

```{r xpt-export, exercise=FALSE, echo=TRUE}
adsl_xpt_path  <- out_dir("adsl.xpt")

adsl_for_export <- adsl |>
  select(all_of(intersect(adsl_spec$variable, names(adsl)))) |>
  mutate(across(where(is.factor), as.character))

xportr(
  adsl_for_export,
  metadata = adsl_spec,
  domain   = "ADSL",
  path     = adsl_xpt_path
)
```

### 8.2 ADAE XPT export

```{r adae-xpt-export, exercise=FALSE, echo=TRUE}
adae_xpt_path <- out_dir("adae.xpt")

adae_for_export <- adae |>
  select(all_of(intersect(adae_spec$variable, names(adae))))

xportr(
  adae_for_export,
  metadata = adae_spec,
  domain   = "ADAE",
  path     = adae_xpt_path
)
```

### 8.3 Dataset-JSON export

```{r json-export, exercise=FALSE, echo=TRUE}
adsl_json_path <- out_dir("adsl.json")

adsl_json_obj <- dataset_json(
  adsl_for_export,
  file_oid    = "www.cdiscpilot01.org/ADSL",
  dataset_oid = "ADSL",
  description = "Subject-Level Analysis Dataset"
)

write_dataset_json(adsl_json_obj, adsl_json_path)
cat("JSON file size:", file.info(adsl_json_path)$size, "bytes\n")
```

### Exercise 8.1: Export ADAE as both XPT and Dataset-JSON

Export `adae_for_export` (already defined above) as:
1. `adae.xpt` using `xportr()`
2. `adae.json` using `dataset_json()` + `write_dataset_json()`

```{r ex-dual-export, exercise=TRUE, exercise.lines=22}
# XPT
adae_xpt  <- out_dir("adae.xpt")
xportr(
  ___,
  metadata = ___,
  domain   = "ADAE",
  path     = ___
)

# Dataset-JSON
adae_json_path <- out_dir("adae.json")

adae_json_obj <- dataset_json(
  ___,
  file_oid    = "www.cdiscpilot01.org/ADAE",
  dataset_oid = ___,
  description = "Adverse Events Analysis Dataset"
)

___(adae_json_obj, ___)
cat("Done.\n")
```

```{r ex-dual-export-hint-1}
# xportr(adae_for_export, metadata = adae_spec, domain = "ADAE", path = adae_xpt)
# dataset_json(adae_for_export, ...)
# write_dataset_json(adae_json_obj, adae_json_path)
```

```{r ex-dual-export-solution}
adae_xpt <- out_dir("adae.xpt")
xportr(
  adae_for_export,
  metadata = adae_spec,
  domain   = "ADAE",
  path     = adae_xpt
)

adae_json_path <- out_dir("adae.json")

adae_json_obj <- dataset_json(
  adae_for_export,
  file_oid    = "www.cdiscpilot01.org/ADAE",
  dataset_oid = "ADAE",
  description = "Adverse Events Analysis Dataset"
)

write_dataset_json(adae_json_obj, adae_json_path)
cat("Done.\n")
```

---

## 9 — Assembling the Full Submission Package

Now we put it all together. In a real submission you would loop over all ADaMs, all TLF programs, and all QC programs.

### 9.1 Build the folder tree

```{r pkg-folders, exercise=FALSE, echo=TRUE}
pkg_root <- file.path(tempdir(), "submission_pkg")
subdirs  <- c("adam", "json", "define", "tlf", "logs", "qc")
walk(file.path(pkg_root, subdirs), dir.create, showWarnings = FALSE, recursive = TRUE)

cat("Submission package tree:\n")
cat(paste(file.path(pkg_root, subdirs), collapse = "\n"), "\n")
```

### 9.2 Multi-dataset ADaM export loop

```{r pkg-adam-loop, exercise=FALSE, echo=TRUE}
adam_list <- list(
  ADSL = list(data = adsl_for_export, spec = adsl_spec),
  ADAE = list(data = adae_for_export, spec = adae_spec)
)

iwalk(adam_list, function(item, domain) {
  # XPT
  xpt_path  <- file.path(pkg_root, "adam", paste0(tolower(domain), ".xpt"))
  xportr(item$data, metadata = item$spec, domain = domain, path = xpt_path)

  # JSON
  json_path <- file.path(pkg_root, "json", paste0(tolower(domain), ".json"))
  dj <- dataset_json(item$data,
                     file_oid    = paste0("www.cdiscpilot01.org/", domain),
                     dataset_oid = domain)
  write_dataset_json(dj, json_path)
})

cat("\nFiles in adam/:\n")
cat(paste(list.files(file.path(pkg_root, "adam")), collapse = "\n"), "\n")
cat("\nFiles in json/:\n")
cat(paste(list.files(file.path(pkg_root, "json")), collapse = "\n"), "\n")
```

### 9.3 TLF production and logging

```{r pkg-tlf-loop, exercise=FALSE, echo=TRUE}
# Write demographics TLF script
demog_tlf_script <- sprintf('
library(dplyr); library(pharmaverseadam)
adsl <- pharmaverseadam::adsl
if (!all(c("HEIGHTBL","WEIGHTBL") %% names(adsl))) {
  set.seed(2014); n <- nrow(adsl)
  ht <- ifelse(adsl$SEX=="F", round(rnorm(n,161,7)), round(rnorm(n,174,8)))
  wt <- ifelse(adsl$SEX=="F", round(rnorm(n,68,12),1), round(rnorm(n,82,14),1))
  adsl <- dplyr::mutate(adsl, HEIGHTBL=ht, WEIGHTBL=wt,
                         BMIBL=round(WEIGHTBL/(HEIGHTBL/100)^2,1))
}
result <- adsl |>
  dplyr::filter(SAFFL == "Y") |>
  dplyr::group_by(TRT01P) |>
  dplyr::summarise(N=dplyr::n(), mean_age=mean(AGE, na.rm=TRUE), .groups="drop")
message("Table 14.1.1 complete: ", nrow(result), " treatment groups")
print(result)
')

tlf_script_path <- file.path(pkg_root, "t_14_1_1_demog.R")
writeLines(demog_tlf_script, tlf_script_path)

# Run and log
log_demog <- axecute(tlf_script_path)
log_path  <- file.path(pkg_root, "logs", "t_14_1_1_demog.log")
saveRDS(log_demog, log_path)

cat(sprintf("Log saved: %s  |  Status: %s\n", basename(log_path), log_demog$status))
```

### 9.4 Package manifest

```{r pkg-manifest, exercise=FALSE, echo=TRUE}
# Build a manifest of all submission files
all_files <- list.files(pkg_root, recursive = TRUE, full.names = FALSE)

manifest <- tibble(
  file     = all_files,
  folder   = dirname(all_files),
  size_kb  = file.info(file.path(pkg_root, all_files))$size / 1024
) |>
  arrange(folder, file)

cat("=== Submission Package Manifest ===\n")
print(manifest, n = 40)
cat(sprintf("\nTotal files: %d\n", nrow(manifest)))
```

---

## 10 — Final Capstone: Complete Pipeline in 60 Lines

This is it. One exercise, no scaffolding. Build the **full pipeline** from ADaM datasets to submission package, touching every tool from Pharmaverse10.

**Requirements:**
1. Filter `adsl` to SAFFL == "Y"; compute a summary with N, mean age, and % female per treatment arm
2. Use `diffdf` to compare your summary against `ref_demog_counts` (provided in setup). Report whether it's identical
3. Export `adsl_for_export` as both XPT (`adsl_final.xpt`) and Dataset-JSON (`adsl_final.json`) to `tempdir()`
4. Use `axecute_string()` to log a one-liner script: `message("ADSL pipeline complete")`
5. Confirm the log status is "SUCCESS"
6. Print a completion banner

```{r ex-capstone-final, exercise=TRUE, exercise.lines=65}
## ── Step 1: Safety population summary ────────────────────────────────────────
saf_summary <- adsl |>
  filter(___ == "Y") |>
  group_by(___) |>
  summarise(
    N         = n(),
    mean_age  = round(mean(AGE, na.rm = TRUE), 1),
    pct_female = round(100 * mean(SEX == "F", na.rm = TRUE), 1),
    .groups   = "drop"
  )

cat("=== Safety Population Summary ===\n")
print(saf_summary)

## ── Step 2: diffdf QC vs reference ───────────────────────────────────────────
# ref_demog_counts has columns: TRT01P, SEX, n
# Compare your N column vs ref totals (join first)
ref_n <- ref_demog_counts |>
  group_by(TRT01P) |>
  summarise(N_ref = sum(n), .groups = "drop")

check_df <- left_join(saf_summary, ref_n, by = "TRT01P") |>
  select(TRT01P, N, N_ref)

diff_check <- diffdf(___, ___, keys = "TRT01P")
cat("\n=== QC Check: N vs Reference ===\n")
cat("Identical:", diffdf_is_identical(___), "\n")

## ── Step 3: Dual export ───────────────────────────────────────────────────────
adsl_xpt_final  <- file.path(tempdir(), "adsl_final.xpt")
adsl_json_final <- file.path(tempdir(), "adsl_final.json")

# XPT
xportr(
  ___,
  metadata = adsl_spec,
  domain   = "ADSL",
  path     = ___
)

# Dataset-JSON
write_dataset_json(
  dataset_json(adsl_for_export,
               file_oid    = "www.cdiscpilot01.org/ADSL",
               dataset_oid = "ADSL"),
  ___
)

## ── Step 4: logrx logging ─────────────────────────────────────────────────────
final_log <- axecute_string('message("ADSL pipeline complete")')
cat("\n=== Program Log ===\n")
print(___)

## ── Step 5: Status check ──────────────────────────────────────────────────────
stopifnot(final_log$___ == "SUCCESS")

## ── Step 6: Completion banner ─────────────────────────────────────────────────
cat("\n")
cat(strrep("═", 55), "\n")
cat("  ✓  Pharmaverse10 Capstone Complete\n")
cat("  Datasets  : ADSL (N=", nrow(adsl_for_export), ")\n")
cat("  Formats   : XPT + Dataset-JSON\n")
cat("  QC status :", if (diffdf_is_identical(diff_check)) "PASS" else "FAIL", "\n")
cat("  Log status:", final_log$status, "\n")
cat(strrep("═", 55), "\n")
```

```{r ex-capstone-final-hint-1}
# Step 1: filter(SAFFL == "Y"), group_by(TRT01P)
# Step 2: diffdf(check_df |> select(TRT01P, N), ref_n |> rename(N = N_ref), keys="TRT01P")
# Step 3: xportr(adsl_for_export, ..., path = adsl_xpt_final)
# Step 4: print(final_log)
# Step 5: final_log$status
```

```{r ex-capstone-final-solution}
## Step 1
saf_summary <- adsl |>
  filter(SAFFL == "Y") |>
  group_by(TRT01P) |>
  summarise(
    N          = n(),
    mean_age   = round(mean(AGE, na.rm = TRUE), 1),
    pct_female = round(100 * mean(SEX == "F", na.rm = TRUE), 1),
    .groups    = "drop"
  )

cat("=== Safety Population Summary ===\n")
print(saf_summary)

## Step 2
ref_n <- ref_demog_counts |>
  group_by(TRT01P) |>
  summarise(N_ref = sum(n), .groups = "drop")

check_df <- left_join(saf_summary, ref_n, by = "TRT01P") |>
  select(TRT01P, N, N_ref)

diff_check <- diffdf(
  check_df |> select(TRT01P, N),
  ref_n    |> rename(N = N_ref),
  keys = "TRT01P"
)
cat("\n=== QC Check: N vs Reference ===\n")
cat("Identical:", diffdf_is_identical(diff_check), "\n")

## Step 3
adsl_xpt_final  <- file.path(tempdir(), "adsl_final.xpt")
adsl_json_final <- file.path(tempdir(), "adsl_final.json")

xportr(
  adsl_for_export,
  metadata = adsl_spec,
  domain   = "ADSL",
  path     = adsl_xpt_final
)

write_dataset_json(
  dataset_json(adsl_for_export,
               file_oid    = "www.cdiscpilot01.org/ADSL",
               dataset_oid = "ADSL"),
  adsl_json_final
)

## Step 4
final_log <- axecute_string('message("ADSL pipeline complete")')
cat("\n=== Program Log ===\n")
print(final_log)

## Step 5
stopifnot(final_log$status == "SUCCESS")

## Step 6
cat("\n")
cat(strrep("═", 55), "\n")
cat("  ✓  Pharmaverse10 Capstone Complete\n")
cat("  Datasets  : ADSL (N=", nrow(adsl_for_export), ")\n")
cat("  Formats   : XPT + Dataset-JSON\n")
cat("  QC status :", if (diffdf_is_identical(diff_check)) "PASS" else "FAIL", "\n")
cat("  Log status:", final_log$status, "\n")
cat(strrep("═", 55), "\n")
```

---

## 11 — Curriculum Retrospective Quiz

You've covered the full Pharmaverse10 stack. Let's revisit the key decisions you'll make on a real submission.

```{r quiz-final, echo=FALSE}
quiz(
  question(
    "You need to validate the risk tier of a new open-source package before adding it to your submission pipeline. Which package and function do you use?",
    answer("diffdf::diffdf()"),
    answer("logrx::axecute()"),
    answer("riskmetric: pkg_ref() → assess() → score()", correct = TRUE),
    answer("xportr::xportr_validate()"),
    random_answer_order = TRUE
  ),
  question(
    "Which xportr function coerces R Date columns to the numeric SAS date representation required for XPT transport files?",
    answer("xportr_label()"),
    answer("xportr_format()"),
    answer("xportr_type()", correct = TRUE),
    answer("xportr_order()"),
    random_answer_order = TRUE
  ),
  question(
    "A Dataset-JSON v1.1 file stores each data row as:",
    answer("A JSON object with named keys"),
    answer("An array of values in column order", correct = TRUE),
    answer("A CSV-encoded string"),
    answer("A nested object per variable"),
    random_answer_order = TRUE
  ),
  question(
    "Your QC programmer runs diffdf() and gets diffdf_is_identical() = FALSE with 3 cell differences. What is the correct next step?",
    answer("Submit anyway — 3 diffs is below the threshold"),
    answer("Investigate both programs to determine which version is correct, then resolve and re-run QC", correct = TRUE),
    answer("Delete the QC program and use the primary only"),
    answer("Increase the tolerance argument until diffs disappear"),
    random_answer_order = TRUE
  ),
  question(
    "logrx::axecute() sets status = 'ERROR' when:",
    answer("The script prints a warning"),
    answer("The script emits more than 5 messages"),
    answer("An uncaught R error stops script execution", correct = TRUE),
    answer("The script takes longer than 60 seconds"),
    random_answer_order = TRUE
  ),
  question(
    "In tfrmt, what is the purpose of frmt_combine()?",
    answer("To merge two tfrmt objects"),
    answer("To combine values from multiple ARD parameter rows into a single formatted cell", correct = TRUE),
    answer("To stack two tables vertically"),
    answer("To produce RTF output from a gt table"),
    random_answer_order = TRUE
  )
)
```

---

## Congratulations! {-}

You have completed the **entire Pharmaverse tutorial series** — all 48 lessons, from R environment setup through a production-ready eCTD submission package.

Here is what you can now do end-to-end:

**Data Pipeline**
From raw SDTM domains (pharmaversesdtm) → cleaned ADaM datasets (admiral, metacore, metatools) → validated metadata (xportr, metacore).

**Output Generation**
From ADaM tables → formatted clinical tables (gtsummary, tfrmt) → RTF/PDF outputs (r2rtf, gt) → interactive Shiny apps (teal).

**Submission Compliance**
From R objects → transport files (xportr XPT) → JSON submissions (dataset.json) → validated QC (diffdf) → logged programs (logrx) → risk-assessed packages (riskmetric).

---

> *The Pharmaverse exists because the clinical programming community chose to build open, interoperable tools together. Every package you used in this series is maintained by practitioners like you. Contribute back: file issues, write vignettes, and help the next programmer make the switch.*

---

*Lesson 48 of 48 — Pharmaverse10 Capstone Part 2*
*Series complete.*
